### 2017-11-06 HPC_Paper/paper.tex
Working on paper, ready to collect results for MLP Regressor and Support Vector Regressor.

%**************************************************************************
%* SpringSim 2017 Author Kit
%*
%* Word Processing System: TeXnicCenter and MiKTeX
%*
%**************************************************************************

\documentclass{scspaperproc}

\usepackage{latexsym}
\usepackage{graphicx}
\usepackage{mathptmx}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{amsbsy}
\usepackage{amsthm}
%\usepackage[pdftex,colorlinks=true,urlcolor=blue,citecolor=black,anchorcolor=black,linkcolor=black]{hyperref}
%% \usepackage[dvips,colorlinks=true,urlcolor=blue,citecolor=black,%
%% anchorcolor=black,linkcolor=black]{hyperref}

% custom hyphenation rules
\usepackage{hyphenat}
\hyphenation{op-tical net-works semi-conduc-tor}

% theorem style
\newtheoremstyle{scsthe}% hnamei
{8pt}% hSpace abovei
{8pt}% hSpace belowi
{\it}% hBody fonti
{}% hIndent amounti1
{\bf}% hTheorem head fontbf
{.}% hPunctuation after theorem headi
{.5em}% hSpace after theorem headi2
{}% hTheorem head spec (can be left empty, meaning `normal')i
\theoremstyle{scsthe}
\newtheorem{theorem}{Theorem}
\renewcommand{\thetheorem}{\arabic{theorem}}
\newtheorem{corollary}[theorem]{Corollary}
\renewcommand{\thecorollary}{\arabic{corollary}}
\newtheorem{definition}{Definition}
\renewcommand{\thedefinition}{\arabic{definition}}

% avoid overrunning the right margin
\sloppy

%% ***** NOTE *****

%% The use of the long citation format (e.g. "Brown and Edwards
%% (1993)" rather than "[5]") and at the same time using the hyperref
%% package can lead to hard to trace bugs in case the citation is
%% broken accross the line (usually this will mark the entire
%% paragraph as a hyperlink (clickable) which is easily noticeable and
%% fixed if using colorlinks, but not if the color is black -- as it
%% is now). Worse yet, if a citation spans page boundary, LaTeX
%% compilation can fail, with an obscure error message. Since this
%% depends a lot on the flow of the text and wording, these bugs come
%% and go and can be extremely hard for a beginner to trace. The error
%% message can look like this:
%%
%%    ! pdfTeX error (ext4): \pdfendlink ended up in different nesting
%%    level than \pdfstartlink.  \AtBegShi@Output ...ipout \box
%%    \AtBeginShipoutBox \fi \fi
%%    l.174 
%%    ! ==> Fatal error occurred, no output PDF file produced!
%%
%% and can be universally fixed by putting an \mbox{} around the
%% citation in question (in this case, at line 174) and maybe adapting
%% the wording a little bit to improve the paragraph typesetting,
%% which is perhaps not immediately obvious.
%****************************************************************************

% begin document
\begin{document}

% Page header (author list)
\SCSpagesetup{Lux, Watson, Chang, Bernard, Li, Xu, Back, Butt, Cameron, and Hong}

% Conference info
\def\SCSconferenceacro{SpringSim}
\def\SCSpublicationyear{2018}
\def\SCSconferencedates{April 15-18}
\def\SCSconferencevenue{Baltimore, MD, USA}
\def\SCSsymposiumacro{HPC} % High Performance Computing Symposium

% title
\title{Comparison of Models of I/O Characteristics \\ in High
  Performance Computing Systems}

% AUTHOR LIST
% *** NOTE: May need to adjust titlevboxsize in the preamble
\author{Thomas C. H. Lux \\ [12pt]
Dept. of Computer Science \\
Virginia Polytechnic Institute\\
\& State University \\
Blacksburg, VA 24061 \\
tchlux@vt.edu \\
\and
Layne T. Watson \\[12pt]
Dept. of Computer Science\\
Dept. of Mathematics\\
Dept. of Aerospace \& Ocean Eng.\\ 
Virginia Polytechnic Institute\\
\& State University \\
\and
Tyler H. Chang\\
Jon Bernard\\
Bo Li\\[12pt]
Dept. of Computer Science\\ 
Virginia Polytechnic Institute\\
\& State University \\
\and
Li Xu\\[12pt]
Dept. of Statistics\\ 
Virginia Polytechnic Institute\\
\& State University \\
\and
Godmar Back\\
Ali R. Butt\\
Kirk W. Cameron\\[12pt]
Dept. of Computer Science\\ 
Virginia Polytechnic Institute\\
\& State University \\
\and
Yili Hong\\[12pt]
Dept. of Statistics\\ 
Virginia Polytechnic Institute\\
\& State University \\
}

\maketitle

\section*{Abstract}

Each of high performance computing, cloud computing, and computer
security have their own interests in modeling and predicting the
performance of computers with respect to how they are configured. An
effective model might infer internal mechanics, minimize power
consumption, or maximize computational throughput of a given
system. This paper analyzes a seven-dimensional dataset measuring the
input/output (I/O) characteristics of a cluster of identical computers
using the benchmark IOzone. The I/O performance characteristics are
modelled with respect to system configuration using multivariate
interpolation and approximation techniques. The analysis reveals that
accurate models of I/O characteristics for a computer system may be
created from a small fraction of possible configurations, and that
some modeling techniques will continue to perform well as the number
of system parameters being modeled increases. These results have
strong implications for future predictive analyses based on more
comprehensive sets of system parameters.

\textbf{Keywords:} Regression, approximation, other stuff

%     Introduction     
%======================
\section{Introduction}

Performance tuning is often an experimentally complex and time-intense
chore necessary for configuring HPC systems. The procedures for this
tuning vary largely from system to system and are often subjectively
guided by the system engineer(s). Once a desired level of performance
is achieved, HPC systems often remain in that base configuration with
all subsequent modifications being incremental. The changes made are
often associated with updates and job-specific customizations. In the
case that a system has changing workloads or non-stationary
performance objectives that range from maximizing computational
throughput to minimizing power consumption and system variability, it
becomes obvious that a more effective and automated tool is needed for
configuring systems. This scenario presents a challenging and
important application of multivariate approximation and interpolation
techniques.

\begin{enumerate}
\item The value of multivariate modlling
\item The data context
\item The proposed method for using multivariate models
\item The impact of effective models
\end{enumerate}

\subsection{Approximation}

This paper compares five multivariate approximation techniques that
operate on inputs in $\mathbb{R}^d$ (vectors of $d$ real numbers) and
produce predicted responses in $\mathbb{R}^1$. Three of the techniques
are regression-based and produce models that do not perfectly
reproduce the inputs, but are hopefully more generalizable. The
ramaining two techniques are interpolation techniques that reproduce
input data exactly at the expense of higher model complexity. The
sections below outline the mathematical formulations of each technique
and provide computational complexity bounds.

\subsubsection{Multivariate Adaptive Regression Splines}
\subsubsection{Multi-Layer Perceptron Regressor}
\subsubsection{Support Vector Regressor}

\subsection{Interpolation}
\subsubsection{Linear Shepard}
\subsubsection{Delaunay}

%     Related Work     
%======================
\section{Related Work}
\begin{enumerate}
\item Not sure how much to include here? Shooting for thoroughness or
  simply necessary coverage?
\end{enumerate}

%     Methodology     
%=====================
\section{Methodology}
\subsection{Data}
\subsection{Dimensional Analysis}
\begin{enumerate}
\item Cycling the categorical settings
\item Selecting subsets of 1,2,3 up to 4 dimensions
\item Cycling different training : testing ratios (5:95 $\rightarrow$ 95:5)
\item Generating 200 random training : testing splits that ensure the
  testing points are not outside the convex hull of the training.
\item Selecting training points to be well-spaced using QNSTOPP algorithm.
\end{enumerate}

\subsection{Prediction}
\begin{enumerate}
\item For each file generated from the dimensional analysis, train on
  the training data, evaluate at the testing data points
\end{enumerate}

%     Results     
%=================
\section{Results}
\subsection{I/O Throughput Mean}
\subsection{I/O Throughput Variance}
\subsection{Increasing Dimension}

%     Discussion     
%====================
\section{Discussion}
\subsection{Modelling the System}
\subsection{Quantifying Variability}
\subsection{Extending the Analysis}

%     Future Work     
%=====================
\section{Future Work}

\bibliographystyle{scsproc}
\bibliography{paper}

\end{document}

### 2017-11-06 TestingAlgs/py_performance_test.py

@@ -7,7 +7,10 @@ from py_settings import *
 fit_times = {}
 eval_times = {}
 max_iters = MAX_ITERS
-surfs = ALGORITHMS.copy()
+surfs = [NearestNeighbor(), LinearModel(), FitBoxMesh(),
+         qHullDelaunay(), Delaunay(), LSHEP(), MARS(),
+         BayesTree(), NearestPlane(), MaxBoxMesh(), tgp(),
+         dynaTree(), BBS(), MLPRegressor()]
 num_points = NUM_POINTS
 dim = DIM
 surfs = surfs[:4]

### 2017-11-06 TestingAlgs/py_test_plot.py

@@ -8,7 +8,7 @@ from util.plotly import *
 from py_settings import *
 
 # Preparing the data
-normalize_points = False
+normalize_points = True
 random_points = False
 random_perturbation = False
 tight_plot = False
@@ -34,7 +34,9 @@ multiplier = 10
 # algorithms = [NearestNeighbor()]
 # algorithms = [LSHEP()]
 # algorithms = [LinearModel()]
-algorithms = [Delaunay()]
+# algorithms = [Delaunay()]
+# algorithms = [ MLPRegressor() ] 
+algorithms = [ SVR() ] 
 
 # algorithms = [NearestPlane()]
 
@@ -51,6 +53,13 @@ else:
 
 if random_perturbation:
     points += np.random.random(size=points.shape)*perturbation_multiplier
+
+FUN = lambda x: 100 * np.sum(x)**2
+# FUN = lambda x: (x[0]-2)*(x[1]-1) if ((x[0] > 2) and (x[1] > 1)) else 0
+# FUN = lambda x: 1*x[0]
+# FUN = lambda x: (x[0]-num_points/3)*(x[1]-num_points/2)**2
+# FUN = lambda x: np.cos(x[0]) * (np.sin(x[1]) if len(x) > 1 else 1.0) + 1
+# FUN = lambda x: np.product(np.sin(np.array(x))/10)
 
 points *= 0.05
 # Calculate the associated response values

### 2017-11-10 HPC_Paper/paper.tex
Renamed TestingAlgs to be Algorithm_Comparison. Modified the comparison code to store raw data files in a parallel directory structure to the MDA tests.

@@
 % title
-\title{Comparison of Models of I/O Characteristics \\ in High
+\title{Predictive Modeling of I/O Characteristics \\ in High
   Performance Computing Systems}
@@
 using the benchmark IOzone. The I/O performance characteristics are
-modelled with respect to system configuration using multivariate
+modeled with respect to system configuration using multivariate
 interpolation and approximation techniques. The analysis reveals that
@@
 
-\textbf{Keywords:} Regression, approximation, other stuff
+\textbf{Keywords:} Regression, approximation, interpolation,
+performance modeling
 
@@
 is achieved, HPC systems often remain in that base configuration with
-all subsequent modifications being incremental. The changes made are
-often associated with updates and job-specific customizations. In the
-case that a system has changing workloads or non-stationary
+all subsequent modifications being incremental. Future changes made
+are often associated with updates and job-specific customizations. In
+the case that a system has changing workloads or non-stationary
 performance objectives that range from maximizing computational
@@
 \begin{enumerate}
-\item The value of multivariate modlling
+\item The value of multivariate Modeling
 \item The data context
@@
 
-\subsection{Approximation}
-
 This paper compares five multivariate approximation techniques that
 operate on inputs in $\mathbb{R}^d$ (vectors of $d$ real numbers) and
-produce predicted responses in $\mathbb{R}^1$. Three of the techniques
-are regression-based and produce models that do not perfectly
-reproduce the inputs, but are hopefully more generalizable. The
-ramaining two techniques are interpolation techniques that reproduce
-input data exactly at the expense of higher model complexity. The
-sections below outline the mathematical formulations of each technique
-and provide computational complexity bounds.
+produce predicted responses in $\mathbb{R}^1$. In order to provide
+coverage of the varied mathematical strategies that can be employed to
+solve the continuous modeling problem, three of the techniques are
+regression-based and the remaining two techniques produce
+interpolants. The sections below outline the mathematical formulations
+of each technique and provide computational complexity bounds.
 
+\subsection{Approximation}
 \subsubsection{Multivariate Adaptive Regression Splines}
@@
 \subsection{Interpolation}
+\subsubsection{Delaunay}
 \subsubsection{Linear Shepard}
-\subsubsection{Delaunay}
+
+The remainder of the paper is broken up into \_ parts. \textit{(will
+  fill in the parts and descriptions once they are finalized)}
 
@@
 \item Not sure how much to include here? Shooting for thoroughness or
-  simply necessary coverage?
+  simply necessary coverage? How much background should I expect the
+  readers of this paper to have in the ``multivariate modeling of
+  systems'' area?
 \end{enumerate}
@@
 \subsection{Data}
+
+\begin{table}
+  \centering
+  \begin{tabular}{|c|c|}
+
+  \end{tabular}
+  \caption{}
+  \label{tab:data_type}
+\end{table}
+
 \subsection{Dimensional Analysis}
+This work utilizes an extension to standard k-fold cross validation
+that allows for a more thorough investigation of the expected model
+performance in a variety of real-world situations. Alongside
+randomized splits, two extra components are considered: the amount of
+training data provided, and the dimension of the input data. It is
+important to consider that algorithms which perform well with less
+training input also require less experimentation. Although, the amount
+of training data required may change as a function of the number of
+input dimensions and this needs to be studied as well.
+
 \begin{enumerate}
@@
 \item Cycling different training : testing ratios (5:95 $\rightarrow$ 95:5)
-\item Generating 200 random training : testing splits that ensure the
-  testing points are not outside the convex hull of the training.
-\item Selecting training points to be well-spaced using QNSTOPP algorithm.
+\item Generating 200 random training : testing splits
+\item Ensuring the testing points are on/inside the convex hull of the training.
+\item Ensuring the training points are well-spaced.
 \end{enumerate}
@@
 \subsection{I/O Throughput Mean}
+
+%% \begin{center}
+%%   \includegraphics[width=\textwidth]{Prediction_Performance_Dim.pdf}
+%%   \caption{In this figure we see how the distribution of errors }
+%% \end{center}
+
+\begin{figure}
+  \centering
+  \includegraphics[width=\textwidth]{Prediction_Performance_TT_Ratio.pdf}
+  \caption{In this figure, it can be seen how the distribution of
+    prediction errors for each algorithm is affected by the quantity
+    of training data provided. Notice that for all amounts of training
+    data, Delaunay has the largest likelihoods around 0 error.}
+  \label{fig:tt_ratio}
+\end{figure}
+
+\begin{table}
+  \centering
+  \begin{tabular}{|c||c|c|c|}
+    %% \hline
+    %% \multicolumn{4}{|c|}{Table Title}\\
+    \hline
+    \textbf{Algorithm} & \textbf{Input Dimension} & \textbf{Mean Absolute Error} & \textbf{Mean Relative Error}\\
+    \hline
+    Delaunay     & 1 & 2060267   & 0.03 \\
+                 & 2 & 2916224   & 0.07 \\
+                 & 3 & 3001794   & 0.09 \\
+                 & 4 & 3134779   & 0.10 \\
+    \hline
+    LSHEP        & 1 & 6353148   & 0.06 \\
+                 & 2 & 13038882  & 1.52 \\
+                 & 3 & 5446207   & 1.13 \\
+                 & 4 & 5133648   & 1.07 \\
+    \hline
+    MARS         & 1 & 2176368   & 0.03 \\
+                 & 2 & 6213369   & 0.61 \\
+                 & 3 & 4024964   & 0.65 \\
+                 & 4 & 4497414   & 1.01 \\
+    \hline
+    MLPRegressor & 1 & 2461097   & 0.06 \\
+                 & 2 & 8883580   & 0.81 \\
+                 & 3 & 11190977  & 2.03 \\
+                 & 4 & 11831721  & 2.20 \\
+    \hline
+    SVR          & 1 & 19306731  & 0.96 \\
+                 & 2 & 78213963  & 18.49\\
+                 & 3 & 97761266  & 31.27\\
+                 & 4 & 113551300 & 37.2 \\
+    \hline
+  \end{tabular}
+  \caption{Most models experience a decay in predictive performance as
+    the dimension of the data increases. This is expected because
+    higher dimension input has exponentially more possible patterns to
+    identify. The MLP Regressor and SVR experience the worst decay in
+    performance with increasing dimension.}
+  \label{tab:perf_by_dim}
+\end{table}
+
+\begin{table}
+  \centering
+  \begin{tabular}{|c||c|c|c|}
+    %% \hline
+    %% \multicolumn{4}{|c|}{Table Title}\\
+    \hline
+    \textbf{Algorithm} & \textbf{Training Data} & \textbf{Mean Absolute Error} & \textbf{Mean Relative Error}\\
+    \hline
+    Delaunay     & [5-21]\%  & 5274933   & 0.11 \\
+                 & [25-46]\% & 2898330   & 0.08 \\
+                 & [50-74]\% & 1518477   & 0.07 \\
+                 & [75-96]\% & 886346    & 0.08 \\
+    \hline
+    LSHEP        & [5-21]\%  & 13218469  & 2.53 \\
+                 & [25-46]\% & 4936381   & 0.85 \\
+                 & [50-74]\% & 2251409   & 0.26 \\
+                 & [75-96]\% & 1415034   & 0.15 \\
+    \hline
+    MARS         & [5-21]\%  & 6543153   & 0.61 \\
+                 & [25-46]\% & 4324916   & 0.80 \\
+                 & [50-74]\% & 3115943   & 0.95 \\
+                 & [75-96]\% & 2591451   & 0.74 \\
+    \hline
+    MLPRegressor & [5-21]\%  & 17348982  & 2.58 \\
+                 & [25-46]\% & 12504957  & 2.29 \\
+                 & [50-74]\% & 5292177   & 1.17 \\
+                 & [75-96]\% & 2996253   & 1.02 \\
+    \hline
+    SVR          & [5-21]\%  & 170906944 & 49.45\\
+                 & [25-46]\% & 99851846  & 31.28\\
+                 & [50-74]\% & 51565405  & 19.63\\
+                 & [75-96]\% & 27125130  & 12.6 \\
+    \hline
+  \end{tabular}
+  \caption{All models experience a reduction in error with increasing
+    amounts of training data. This suggests that the data being
+    modeled is pattern-dense, over-fitting is \textit{not} occurring,
+    and that oversimplifications will tend towards worse predictions.}
+  \label{tab:perf_by_tt}
+\end{table}
+
 \subsection{I/O Throughput Variance}
@@
 \section{Discussion}
-\subsection{Modelling the System}
+\subsection{Modeling the System}
 \subsection{Quantifying Variability}
@@
 \end{document}
+
+%%  ALGORITHM     MEAN ABSOLUTE ERROR    MEAN RELATIVE ERROR
+
+%% 1 Dimension                                        
+%%  Delaunay         2060267             0.03 
+%%  MARS             2176368             0.03 
+%%  MLPRegressor     2461097             0.06 
+%%  LSHEP            6353148             0.06 
+%%  SVR             19306731             0.96
+
+%% 2 Dimensions                                       
+%%  Delaunay         2916224             0.07 
+%%  MARS             6213369             0.61 
+%%  MLPRegressor     8883580             0.81 
+%%  LSHEP           13038882             1.52 
+%%  SVR             78213963             18.49
+
+%% 3 Dimensions                                       
+%%  Delaunay         3001794             0.09 
+%%  MARS             4024964             0.65 
+%%  LSHEP            5446207             1.13 
+%%  MLPRegressor    11190977             2.03 
+%%  SVR             97761266             31.27
+
+%% 4 Dimensions                                       
+%%  Delaunay         3134779             0.10 
+%%  MARS             4497414             1.01 
+%%  LSHEP            5133648             1.07 
+%%  MLPRegressor    11831721             2.20 
+%%  SVR            113551300             37.26
+
+%% [5 - 21%] Training Data               
+%%  Delaunay         5274933             0.11 
+%%  MARS             6543153             0.61 
+%%  LSHEP           13218469             2.53 
+%%  MLPRegressor    17348982             2.58 
+%%  SVR            170906944             49.45
+
+%% [25 - 46%] Training Data              
+%%  Delaunay         2898330             0.08 
+%%  MARS             4324916             0.80 
+%%  LSHEP            4936381             0.85 
+%%  MLPRegressor    12504957             2.29 
+%%  SVR             99851846             31.28
+
+%% [50 - 74%] Training Data              
+%%  Delaunay         1518477             0.07 
+%%  LSHEP            2251409             0.26 
+%%  MARS             3115943             0.95 
+%%  MLPRegressor     5292177             1.17 
+%%  SVR             51565405             19.63
+
+%% [75 - 96%] Training Data              
+%%  Delaunay          886346             0.08 
+%%  LSHEP            1415034             0.15 
+%%  MARS             2591451             0.74 
+%%  MLPRegressor     2996253             1.02 
+%%  SVR             27125130             12.64
+
+%% ALGORITHM   AVG ABSOLUTE MEAN ERROR   AVG RELATIVE ERROR
+%% Delaunay         2711394              0.08  
+%% MARS             4185947              0.72  
+%% LSHEP            6474147              0.94  
+%% MLPRegressor     9063718              1.48  
+%% SVR             82285323              25.12

### 2017-11-10 Algorithm_Comparison/py_test_on_data.py

@@ -1,0 +1,218 @@
+from util.plotly import Plot, multiplot
+from py_settings import *
+import numpy as np
+import time
+
+# MANUALLY TESTING FAILURE CASE
+# train = "/Users/thomaslux/Desktop/TestingAlgs/VarSys-2016-Mean_MDA_Data_0/3-F_Size--R_Size--Freq/10.42--89.58/Train-0.csv"
+# test = "/Users/thomaslux/Desktop/TestingAlgs/VarSys-2016-Mean_MDA_Data_0/3-F_Size--R_Size--Freq/10.42--89.58/Test-0.csv"
+# # ^^ FAILURE CASE FOR MARS
+
+# i = 150
+# train = "/Users/thomaslux/Desktop/TestingAlgs/VarSys-2016-Mean_MDA_Data_0/2-Threads--Freq/50--50/Train-%s.csv"%i
+# test = "/Users/thomaslux/Desktop/TestingAlgs/VarSys-2016-Mean_MDA_Data_0/2-Threads--Freq/50--50/Test-%s.csv"%i
+
+# train = "/Users/thomaslux/Desktop/TestingAlgs/VarSys-2016-Mean_MDA_Data_0/4-F_Size--R_Size--Threads--Freq/25--75/Train-169.csv"
+# test = "/Users/thomaslux/Desktop/TestingAlgs/VarSys-2016-Mean_MDA_Data_0/4-F_Size--R_Size--Threads--Freq/25--75/Test-169.csv"
+
+# train = "/Users/thomaslux/Desktop/TestingAlgs/VarSys-2016-Mean_MDA_Data_0/4-F_Size--R_Size--Threads--Freq/20.02--79.98/Train-7.csv"
+# test = "/Users/thomaslux/Desktop/TestingAlgs/VarSys-2016-Mean_MDA_Data_0/4-F_Size--R_Size--Threads--Freq/20.02--79.98/Test-7.csv"
+
+# train = "/Users/thomaslux/Desktop/TestingAlgs/VarSys-2016-Mean_MDA_Data_0/4-F_Size--R_Size--Threads--Freq/9.95--90.05/Train-134.csv"
+# test = "/Users/thomaslux/Desktop/TestingAlgs/VarSys-2016-Mean_MDA_Data_0/4-F_Size--R_Size--Threads--Freq/9.95--90.05/Test-134.csv"
+
+# i = 3
[…]

### 2017-11-10 Prediction_Performance/make_plots.py

@@ -39,10 +39,13 @@ print(len(data))
 
 algorithms = sorted(np.unique(data["Algorithm"]))
 
-lower = -5000000
-upper =  5000000
+# lower = -5000000
+# upper =  5000000
+lower = -2
+upper =  2
 y_bins = dict(start=lower, end=upper, size=(upper - lower) / 100)
-
+lower_x = 0
+upper_x = .21
 
 # ============================================
 #      Making histogram plot by dimension     
@@ -57,26 +60,27 @@ brief_title = "Distributions of Prediction Error with Increasing Dimension Data"
 full_text_title = "<b>"+brief_title+"</b><br>"+\
                   "Where the individual x-axes are the probability ranges"+\
                   " for the bars in each histogram."
+error_col = "Relative_Mean_Error"
 
-for dim in range(1,5):
+for dim in range(1,np.max(data["Dimension"])+1):
     dim_data = data[(data["Dimension"] == dim)]
     title = (brief_title if dim == 4 else "")
     x_axis = "%i Dimension"%dim + ("s" if dim > 1 else "")
-    y_axis = ("Error in Predicted System Throughput" if dim == 1 else "")
+    y_axis = ("Relative Error in Predicted System Throughput" if dim == 1 else "")
     p = Plot(title, x_axis, y_axis)
     for alg in algorithms:
         alg_data = dim_data[(dim_data["Algorithm"] == alg)]
-        values = alg_data["Mean_Error"]
+        values = alg_data[error_col]
         y_bins['start'] = min(values)
         y_bins['end'] = max(values)
         results[((dim,),alg)] = abs(values).mean()
         p.add_histogram(alg, values, bar_spacing="y", group=alg,
                         histnorm='probability',
                         ybins=y_bins, show_in_legend=(dim==1))
-    plots.append(p.plot(y_range=(lower,upper), html=False,
-                        layout={'title':full_text_title}))
+    plots.append(p.plot(y_range=(lower,upper), x_range=(lower_x,upper_x),
+                         html=False, layout={'title':full_text_title}))
 
-multiplot(plots, shared_y=True, file_name=out_name,show=False)
+# multiplot(plots, shared_y=True, file_name=out_name,show=False)
 
 # ======================================================
 #      Making histogram plot by training percentage     
@@ -103,22 +107,22 @@ for min_train,max_train in training_ranges:
     min_val = min(set_data["Train_Percentage"])
[…]

### 2017-11-10 TestingAlgs/py_alg_evaluation.py

@@ -1,372 +1,0 @@
-import numpy as np
-import pickle, sys, pathlib
-from util.multi_dim_analysis import make_test_data, MDA_Iterator
-from notifications import send_email
-from util.decorators import timeout_after
-import util.algorithms
-from py_settings import *
-
-# 1) using all 4 continuous dimensions (file size, record size,
-#    threads, frequency), normalized to be inside the unit
-#    (hyper)cube, predict variance and then mean for a single
-#    selection of categorical settings (CFQ hypervisor scheduler, NOOP
-#    virtual machine Scheduler, File-read test). Potentially modify
-#    these categorical settings later.
-
-# 1a) perform leave-1-out testing for each datum *inside the convex
-#     hull* of the 4D data 
-
-# 1b) perform a sliding range of (95-5, 90-10, 85-15, ...) 
-#     training-testing splits of the data ensuring *all* testing data
-#     is inside the convex hull of the training data. Using N trials
-#     for each split, where N is determined by the smallest number of
-#     trials that gives us 95% confidence in the predicted results for
-#     *all* splits (assuming normally distributed results) 
-
-# 1c) train and test each algorithm with each of the splits from above
-#     (each algorithm is given the same input and asked to predict the
-#     same values)
-
-# 2) repeat step 1 using all 4 pairs of 3 continuous dimensions,
-#    averaging the variance / mean for overlapping configurations 
-
-# 3) repeat step 1 using all 6 pairs of 2 continuous dimensions,
-#    averaging the variance / mean for overlapping configurations 
-
-# 4) repeat step 1 using all 4 continuous dimensions, averaging the
-#    variance / mean for overlapping configurations 
-
-DATA = "CFQ_NOOP_Fread.pkl"
-TARGET = "Throughput"
-# Two data sets, one for mean throughput data, the other for variance
-# of throughput data.
-MEAN_MDA_DIR = "VarSys-2016-Mean_MDA_Data_0"
-VAR_MDA_DIR = "VarSys-2016-Variance_MDA_Data_0"
-
-# DATA = "Stage_1_Fread.pkl"
-# TARGET = "Throughput"
-# # Two data sets, one for mean throughput data, the other for variance
-# # of throughput data.
-# MEAN_MDA_DIR = "VarSys-2017-Mean_MDA_Data_0"
-# VAR_MDA_DIR = "VarSys-2017-Variance_MDA_Data_0"
-
-
[…]

### 2017-11-10 TestingAlgs/py_performance_test.py

@@ -1,66 +1,0 @@
-import numpy as np
-from util.plotly import *
-import pickle
-
-from py_settings import *
-
-fit_times = {}
-eval_times = {}
-max_iters = MAX_ITERS
-surfs = [NearestNeighbor(), LinearModel(), FitBoxMesh(),
-         qHullDelaunay(), Delaunay(), LSHEP(), MARS(),
-         BayesTree(), NearestPlane(), MaxBoxMesh(), tgp(),
-         dynaTree(), BBS(), MLPRegressor()]
-num_points = NUM_POINTS
-dim = DIM
-surfs = surfs[:4]
-max_iters = 100
-
-if __name__ == "__main__":
-    while (max_iters > 0) and (len(surfs) > 1):
-        max_iters -= 1
-        if MEASURE_PTS:
-            num_points += 100
-        elif MEASURE_DIM:
-            dim += 1
-
-        # Generate testing points randomly spaced
-        points = np.random.random(size=(num_points,dim)) * (num_points-1)
-        # Calculate the associated response values
-        values = np.array([FUN(pt) for pt in points])
-
-        print()
-        print("Points:   %s"%(str(points.shape)))
-        print("Response: %s"%(str(values.shape)))
-        print("Beginning to fit the data...")
-
-        surfs = [Delaunay(), MaxBoxMesh()]
-
-        to_remove = []
-        for s in surfs:
-            name = GET_CLASS_NAME(s)
-            # Test time-to-fit
-            start = time.time()
-            s.fit(points, values, *EXTRA_ARGS.get(name,[]))
-            fit_time = time.time()-start
-            # Test time-to-evaluate
-            start = time.time()
-            s( points[:NUM_EVALS] )
-            eval_time = time.time()-start
-            # Depending on what we're measuring, record appropriately
-            idx = 0 if MEASURE_PTS else 1
-            fit_times[name] = fit_times.get(name,[]) + [[points.shape[idx],fit_time]]
-            eval_times[name] = eval_times.get(name,[]) + [[points.shape[idx],eval_time]]
-            # Calculate the total time to give to the user
-            total = fit_time + eval_time
-            print("%s: %s seconds"%(name,total))
-            # Mark algorithms that take more than "max_compute_time_sec"
-            if total > MAX_COMPUTE_TIME_SEC:
-                to_remove.append(s)
-
-        # Remove algorithms that are taking too long to compute
-        for s in to_remove:
[…]

### 2017-11-10 TestingAlgs/py_prepare_stage_1.py

@@ -1,17 +1,0 @@
-import numpy as np
-from util.data import read_struct
-import pickle
-
-data = read_struct("stage1.csv")
-print(data.dtype.names)
-print(data.shape)
-# Reduce to interesting columns
-cols_to_keep = ["Frequency", "Record_Size", "Threads", "Throughput", "Test_Type"]
-data = data[cols_to_keep]
-# Only look at "Fread"
-data = data[data["Test_Type"] == "freaders"]
-# Remove the "Test_Type" column entirely
-data = data[cols_to_keep[:-1]]
-# Save the data pickle to file
-with open("Stage_1_Fread.pkl", "wb") as f:
-    pickle.dump(data, f)

### 2017-11-10 TestingAlgs/py_show_times.py

@@ -1,70 +1,0 @@
-from util.plotly import *
-import pickle
-from py_settings import *
-
-# Get the names of the series, order them for aesthetics
-with open(FILE_NAME,"rb") as f:
-    fit_times, eval_times = pickle.load(f)
-names = list(fit_times.keys())
-
-# # Fixing bad data
-# index = 0
-# while (fit_times[names[0]][index][0] <= 9310): index += 1
-# for n in names:
-#     fit_times[n] = fit_times[n][:index]
-#     eval_times[n] = eval_times[n][:index]
-# with open(FILE_NAME,"wb") as f:
-#     pickle.dump((fit_times,eval_times),f)
-# exit()
-
-# Sort by (longest length first, longest execution time first)
-names.sort( key=lambda n: (-len(fit_times[n]),-(eval_times[n][-1][1]+fit_times[n][-1][1])) )
-
-# Setup title based on measurements
-if MEASURE_PTS:
-    first = "increasing numbers of points"
-    second = "%i dimensions"%DIM
-    x_title = "Number of points"
-elif MEASURE_DIM:
-    first = "%i points"%NUM_POINTS
-    second = "increasing dimension"
-    x_title = "Number of dimensions"
-title = "Fitting %s in %s, evaluating at %i points"%(
-    first,second,NUM_EVALS)
-
-
-
-p = Plot(title, x_title, "Execution Time (seconds)")
-
-for n in names:
-    p.color_num += 1
-    alpha = 0.5
-    eval_brightness = 1.0
-    fit_brightness = 0.7
-
-    color = p.color(p.color_num, alpha=alpha, brightness=eval_brightness)
-    p.add(n+" Eval", *list(zip(*eval_times[n])), mode='lines',
-          fill='tonexty', color=color, group=n)
-    color = p.color(p.color_num, alpha=alpha, brightness=fit_brightness)
-    p.add(n+" Fit", *list(zip(*fit_times[n])), mode='lines',
-          fill='tozeroy', color=color, group=n)
-
-    if n == "Delaunay":
-        from scipy.optimize import minimize
-        print("FITTING")
-        # Collect the dimension (x) and total time (y) into arrays
-        dims = np.array(fit_times[n])
-        dims, times = dims[:,0], dims[:,1]
-        times += np.array(eval_times[n])[:,1]
-        print(dims)
-        print(times)
-        coefs = np.zeros(4)
[…]

### 2017-11-10 TestingAlgs/py_test_on_data.py

@@ -1,218 +1,0 @@
-from util.plotly import Plot, multiplot
-from py_settings import *
-import numpy as np
-import time
-
-# MANUALLY TESTING FAILURE CASE
-# train = "/Users/thomaslux/Desktop/TestingAlgs/VarSys-2016-Mean_MDA_Data_0/3-F_Size--R_Size--Freq/10.42--89.58/Train-0.csv"
-# test = "/Users/thomaslux/Desktop/TestingAlgs/VarSys-2016-Mean_MDA_Data_0/3-F_Size--R_Size--Freq/10.42--89.58/Test-0.csv"
-# # ^^ FAILURE CASE FOR MARS
-
-# i = 150
-# train = "/Users/thomaslux/Desktop/TestingAlgs/VarSys-2016-Mean_MDA_Data_0/2-Threads--Freq/50--50/Train-%s.csv"%i
-# test = "/Users/thomaslux/Desktop/TestingAlgs/VarSys-2016-Mean_MDA_Data_0/2-Threads--Freq/50--50/Test-%s.csv"%i
-
-# train = "/Users/thomaslux/Desktop/TestingAlgs/VarSys-2016-Mean_MDA_Data_0/4-F_Size--R_Size--Threads--Freq/25--75/Train-169.csv"
-# test = "/Users/thomaslux/Desktop/TestingAlgs/VarSys-2016-Mean_MDA_Data_0/4-F_Size--R_Size--Threads--Freq/25--75/Test-169.csv"
-
-# train = "/Users/thomaslux/Desktop/TestingAlgs/VarSys-2016-Mean_MDA_Data_0/4-F_Size--R_Size--Threads--Freq/20.02--79.98/Train-7.csv"
-# test = "/Users/thomaslux/Desktop/TestingAlgs/VarSys-2016-Mean_MDA_Data_0/4-F_Size--R_Size--Threads--Freq/20.02--79.98/Test-7.csv"
-
-# train = "/Users/thomaslux/Desktop/TestingAlgs/VarSys-2016-Mean_MDA_Data_0/4-F_Size--R_Size--Threads--Freq/9.95--90.05/Train-134.csv"
-# test = "/Users/thomaslux/Desktop/TestingAlgs/VarSys-2016-Mean_MDA_Data_0/4-F_Size--R_Size--Threads--Freq/9.95--90.05/Test-134.csv"
-
-# i = 3
[…]

### 2017-11-11 HPC_Paper/paper.tex
Updated the algorithm comparison code to create the variance MDA data and also require an option saying which you want to predict. Will test local, then run on pawnee to collect variance prediction results.

@@
 guided by the system engineer(s). Once a desired level of performance
-is achieved, HPC systems often remain in that base configuration with
-all subsequent modifications being incremental. Future changes made
-are often associated with updates and job-specific customizations. In
-the case that a system has changing workloads or non-stationary
-performance objectives that range from maximizing computational
-throughput to minimizing power consumption and system variability, it
-becomes obvious that a more effective and automated tool is needed for
-configuring systems. This scenario presents a challenging and
-important application of multivariate approximation and interpolation
-techniques.
+is achieved, an HPC system may only be incremenetally reconfigured as
+required by updates or specific jobs. In the case that a system has
+changing workloads or non-stationary performance objectives that range
+from maximizing computational throughput to minimizing power
+consumption and system variability, it becomes clear that a more
+effective and automated tool is needed for configuring systems. This
+scenario presents a challenging and important application of
+multivariate approximation and interpolation techniques.
 
-\begin{enumerate}
-\item The value of multivariate Modeling
-\item The data context
-\item The proposed method for using multivariate models
-\item The impact of effective models
-\end{enumerate}
+Predicting the performance of an HPC system is a challenging problem
+that is primarily attempted in one of two ways: (1) build a
+statistical model of the performance by running experiments on the
+system at select settings or (2) use a performance simulator to run
+artificial experiments and search for optimal configurations. In this
+paper the proposed multivariate modelling techniques rest in the first
+category, however they represent a notable increase in the ability to
+model more complex interactions between system parameters.
 
-This paper compares five multivariate approximation techniques that
-operate on inputs in $\mathbb{R}^d$ (vectors of $d$ real numbers) and
-produce predicted responses in $\mathbb{R}^1$. In order to provide
-coverage of the varied mathematical strategies that can be employed to
-solve the continuous modeling problem, three of the techniques are
-regression-based and the remaining two techniques produce
-interpolants. The sections below outline the mathematical formulations
-of each technique and provide computational complexity bounds.
+Many previous works attempting to model system performance have used
+simulated environments to estimate the performance of a
+system. \cite{grobelny2007fase} \cite{wang2009simulation}
+\cite{wang2013towards} Some of these works cite statistical models as
+being over-simplified and not capabel of capturing the true complexity
+of the underlying system. This claim is partially correct, noticing
+that a large portion of predictive statistical models rely on
+simplifying the system to one or two
+parameters. \cite{snavely2002framework} \cite{bailey2005performance}
+\cite{barker2009using} \cite{ye2010analyzing} These limited
+statistical models have provided satisfactory performance in very
+narrow application settings. Many of the aforementioned statistical
+modeling techniques claim to generalize, while simultaneously
+requiring additional code annotations, hardware abstractions, or
+additional application level understandings in order to generate
+models. The approach presented here requires no modifications of the
+application, no architecture abstractions, nor any structural
+descriptions of the input data being modelled. The techniques used are
+purely mathematical and only need formatted data as input.
+
+Among the statistcal models presented in prior works,
+\cite{bailey2005performance} specifically mentioned that it is
+difficult for the models to capture variability introduced by
+I/O. System variability in general has become a problem of increasing
+interest to the systems and HPC communities, however most of the work
+has focused on Operating System induced variability.
+\cite{beckman2008benchmarking} \cite{de2007identifying} The work that
+has focused on managing I/O variability, \cite{lofstead2010managing},
+does not use any sophisticated modeling techniques. Hence, this paper
+presents a case study applying advanced mathematical and statistical
+modeling techniques to the domain of HPC system I/O
+characteristics. The models are used to predict the mean throughput of
+a system and the variance in throughput of a system. The discussion
+section outlines how the exact techniques presented can be applied to
+any performance metric and any system.
+
+%% Multivariate models for an HPC system would be a function of the
+%% tunable parameters built to accurately model some desired performance
+%% metric.
+
+%% \begin{enumerate}
+%% \item The value of multivariate Modeling
+%% \item The data context
+%% \item The proposed method for using multivariate models
+%% \item The impact of effective models
+%% \end{enumerate}
+
+In general, this paper compares five multivariate approximation
+techniques that operate on inputs in $\mathbb{R}^d$ (vectors of $d$
+real numbers) and produce predicted responses in $\mathbb{R}^1$. In
+order to provide coverage of the varied mathematical strategies that
+can be employed to solve the continuous modeling problem, three of the
+techniques are regression-based and the remaining two techniques
+produce interpolants. The sections below outline the mathematical
+formulations of each technique and provide computational complexity
+bounds with respect to the size (number of points and dimension) of
+input data.
 
 \subsection{Approximation}
+Multivariate approximations are capable of accurately modelling
+complex relationships between the dimensions of any set of points in
+$\mathbb{R}^n$ with respect to some response in $\mathbb{R}$. The
+approximations produce a function $f: \mathbb{R}^n \rightarrow
+\mathbb{R}$ which minimizes some metric related to the provided data.
+
 \subsubsection{Multivariate Adaptive Regression Splines}
@@
 
-%     Related Work     
-%======================
-\section{Related Work}
-\begin{enumerate}
-\item Not sure how much to include here? Shooting for thoroughness or
-  simply necessary coverage? How much background should I expect the
-  readers of this paper to have in the ``multivariate modeling of
-  systems'' area?
-\end{enumerate}
+%% %     Related Work     
+%% %======================
+%% \section{Related Work}
+%% \begin{enumerate}
+%% \item Not sure how much to include here? Shooting for thoroughness or
+%%   simply necessary coverage? How much background should I expect the
+%%   readers of this paper to have in the ``multivariate modeling of
+%%   systems'' area?
+%% \end{enumerate}
 
@@
 \subsection{Data}
+The summary of the data used in the experiments for this paper can be
+seen in Table \ref{tab:data_type}.
 
@@
   \centering
-  \begin{tabular}{|c|c|}
-
+  \begin{tabular}{|c|c|c|}
+    \hline
+    \textbf{System Parameter} & \textbf{Type} & \textbf{Values}\\
+    \hline
+    Hypervisor Scheduler & Categorical & CFQ, NOOP, DEAD\\
+    Operating System Scheduler & Categorical & CFQ, NOOP, DEAD\\
+    IOZone Test Type & Categorical & Fread, Fwrite, RandomRead\\
+    File Size & Continuous & 64, 128, 256, 512, 1024\\
+    Record Size & Continuous & 32 64, 128, 256, 512\\
+    Thread Count & Continuous & 1,2,4,8,16,32,64\\
+    Frequency & Continuous & 2.4, 2.5, 2.6, 2.7, 2.8, 2.9, 3.0, 3.1\\
+    \hline
   \end{tabular}
-  \caption{}
+  \caption{A description of the system parameters being considered in
+    the IOZone tests. There are three categorical settings and four
+    continuous settings. Notice that the ranges of values for
+    continuous settings are different.}
   \label{tab:data_type}
@@
 input dimensions and this needs to be studied as well.
+
+The framework used in this paper will be referred to as a
+multi-dimensional analysis (MDA) of the IOZone data presented in this
+study. However, the MDA framework can be applied to other datasets.
 
@@
 \section{Future Work}
+The most severe limitation to the presented work is the present
+inability to model the relationship between categorical system
+parameters. This could be seen as a limitation of the models selected,
+because only MARS is capable of handling non-numeric variables, or as
+a limitation of the MDA framework.
 

### 2017-11-11 Algorithm_Comparison/py_alg_evaluation.py

@@ -1,5 +1,5 @@
 import numpy as np
-import pickle, sys, pathlib
+import pickle, sys, pathlib, os
 from util.multi_dim_analysis import make_test_data, MDA_Iterator
 from notifications import send_email
 from util.decorators import timeout_after
@@ -36,20 +36,21 @@ from py_settings import *
 # 4) repeat step 1 using all 4 continuous dimensions, averaging the
 #    variance / mean for overlapping configurations 
 
-DATA = "CFQ_NOOP_Fread.pkl"
+SOURCE_DIR = "Data_Source"
+
+DATA = os.path.join(SOURCE_DIR, "CFQ_NOOP_Fread.pkl")
 TARGET = "Throughput"
 # Two data sets, one for mean throughput data, the other for variance
 # of throughput data.
 MEAN_MDA_DIR = "VarSys-2016-Mean_MDA_Data_0"
 VAR_MDA_DIR = "VarSys-2016-Variance_MDA_Data_0"
 
-# DATA = "Stage_1_Fread.pkl"
+# DATA = os.path.join(SOURCE_DIR, "Stage_1_Fread.pkl")
 # TARGET = "Throughput"
 # # Two data sets, one for mean throughput data, the other for variance
 # # of throughput data.
 # MEAN_MDA_DIR = "VarSys-2017-Mean_MDA_Data_0"
 # VAR_MDA_DIR = "VarSys-2017-Variance_MDA_Data_0"
-
 
 # Load the original data file (and get the column names)
 with open(DATA,"rb") as f:
@@ -58,7 +59,7 @@ with open(DATA,"rb") as f:
     settings.remove(TARGET)
 
 # Generate the MDA data set if necessary
-if ((not os.path.exists(MEAN_MDA_DIR)) and
+if ((not os.path.exists(MEAN_MDA_DIR)) or
     (not os.path.exists(VAR_MDA_DIR))):
     # Filter down the data into the unique x-coordinates
     unique_pts = {}
@@ -99,9 +100,9 @@ if ((not os.path.exists(MEAN_MDA_DIR)) and
     if not(os.path.exists(MEAN_MDA_DIR)):
         output = stat_data["Mean"]
         make_test_data(inputs, output, "VarSys-2016"+"-Mean", settings)
-    # if not(os.path.exists(VAR_MDA_DIR)):
-    #     output = stat_data["Variance"]
[…]

### 2017-11-12 HPC_Paper/paper.tex
Working on paper, updated CSV result files to have decimal points in all predicted columns, checked function of data collection for SVR polynomial.

@@
 consumption, or maximize computational throughput of a given
-system. This paper analyzes a seven-dimensional dataset measuring the
+system. This paper analyzes a four-dimensional dataset measuring the
 input/output (I/O) characteristics of a cluster of identical computers
@@
 simulated environments to estimate the performance of a
-system. \cite{grobelny2007fase} \cite{wang2009simulation}
-\cite{wang2013towards} Some of these works cite statistical models as
-being over-simplified and not capabel of capturing the true complexity
-of the underlying system. This claim is partially correct, noticing
-that a large portion of predictive statistical models rely on
-simplifying the system to one or two
-parameters. \cite{snavely2002framework} \cite{bailey2005performance}
-\cite{barker2009using} \cite{ye2010analyzing} These limited
-statistical models have provided satisfactory performance in very
-narrow application settings. Many of the aforementioned statistical
-modeling techniques claim to generalize, while simultaneously
-requiring additional code annotations, hardware abstractions, or
-additional application level understandings in order to generate
-models. The approach presented here requires no modifications of the
-application, no architecture abstractions, nor any structural
-descriptions of the input data being modelled. The techniques used are
-purely mathematical and only need formatted data as input.
+system. \cite{grobelny2007fase,wang2009simulation,wang2013towards}
+Some of these works cite statistical models as being over-simplified
+and not capabel of capturing the true complexity of the underlying
+system. This claim is partially correct, noting that a large portion
+of predictive statistical models rely on simplifying the system to one
+or two parameters.
+\cite{snavely2002framework,bailey2005performance,barker2009using,ye2010analyzing}
+These limited statistical models have provided satisfactory
+performance in very narrow application settings. Many of the
+aforementioned statistical modeling techniques claim to generalize,
+while simultaneously requiring additional code annotations, hardware
+abstractions, or additional application level understandings in order
+to generate models. The approach presented here requires no
+modifications of the application, no architecture abstractions, nor
+any structural descriptions of the input data being modelled. The
+techniques used are purely mathematical and only need formatted data
+as input.
 
 Among the statistcal models presented in prior works,
-\cite{bailey2005performance} specifically mentioned that it is
-difficult for the models to capture variability introduced by
-I/O. System variability in general has become a problem of increasing
-interest to the systems and HPC communities, however most of the work
-has focused on Operating System induced variability.
-\cite{beckman2008benchmarking} \cite{de2007identifying} The work that
+\cite{bailey2005performance} specifically mentions that it is
+difficult for the simplified models to capture variability introduced
+by I/O. System variability in general has become a problem of
+increasing interest to the HPC and systems communities, however most
+of the work has focused on Operating System (OS) induced variability.
+\cite{beckman2008benchmarking,de2007identifying} The work that
 has focused on managing I/O variability, \cite{lofstead2010managing},
@@
 presents a case study applying advanced mathematical and statistical
-modeling techniques to the domain of HPC system I/O
-characteristics. The models are used to predict the mean throughput of
-a system and the variance in throughput of a system. The discussion
-section outlines how the exact techniques presented can be applied to
-any performance metric and any system.
+modeling techniques to the domain of HPC I/O characteristics. The
+models are used to predict the mean throughput of a system and the
+variance in throughput of a system. The discussion section outlines
+how the exact techniques presented can be applied to any performance
+metric and any system.
 
@@
 techniques that operate on inputs in $\mathbb{R}^d$ (vectors of $d$
-real numbers) and produce predicted responses in $\mathbb{R}^1$. In
+real numbers) and produce predicted responses in $\mathbb{R}$. In
 order to provide coverage of the varied mathematical strategies that
@@
 bounds with respect to the size (number of points and dimension) of
-input data.
+input data. Throughout the sections, $d$ will refer to the dimension
+of the input data, $n$ is the number of points in the input data,
+$x_i \in \mathbb{R}^d$ is the $i^{th}$ input data point, and $y_i$ is
+the response value of the $i^{th}$ input data point.
 
@@
 complex relationships between the dimensions of any set of points in
-$\mathbb{R}^n$ with respect to some response in $\mathbb{R}$. The
-approximations produce a function $f: \mathbb{R}^n \rightarrow
-\mathbb{R}$ which minimizes some metric related to the provided data.
+$\mathbb{R}^d$ with respect to some response in $\mathbb{R}$. The
+approximations produce a function $f: \mathbb{R}^d \rightarrow
+\mathbb{R}$ which minimizes some selected error metric related to the
+provided data. Often, as in least-squares regression, the distance
+metric chosen is the sum of squared differences between modeled
+response values and true response values.
 
 \subsubsection{Multivariate Adaptive Regression Splines}
+This algorithm was introduced by statistician J. Friedman in
+\cite{friedman1991multivariate} and subsequently improved to it's
+current version in \cite{stanford1993fast}. This work uses the Fast
+Multivariate Adaptive Regression Splines (Fast MARS) formulation. In
+Fast MARS, a least-squares fit model is iteratively built by beginning
+with a single constant valued function and adding two new basis
+functions at each iteration of the form
+
+$$ B_{2s-1}(x) = B_l(x) [c(x_i-v)]_+ $$
+$$ B_{2s}(x) = B_k(x) [c(x_i-v)]_- $$
+
+Where $s$ is the iteration number, $B_l(x)$ and $B_k(x)$ are some
+previously generated basis functions, $c, v \in \mathbb{R}$, and $x_i$
+with $0 < i \leq d$ is the $i^{th}$ component (1 indexed) of input
+vector $x$. After iteratively constructing a model, MARS then
+iteratively removes basis functions that do not contribute to goodness
+of fit. In effect, MARS creates a locally component-wise linear
+approximation of the data that is $C^0$. The overall computational
+complexity of Fast MARS is $\mathcal{O}(n d m^3)$ where $m$ is a user
+selected value that is the maximum complexity of the model in terms of
+the number of underlying basis functions. This paper uses a default
+max number of basis functions of $n$, however with larger datasets ($n
+> 1000$)this value would need to be reduced.
+
 \subsubsection{Multi-Layer Perceptron Regressor}
+The neural network is a well studied and widely used mathematical tool
+for both regression and classification tasks.
+\cite{hornik1989multilayer} When using the rectified linear unit
+(ReLU) activation functions \cite{dahl2013improving} and trained using
+the BFGS technique \cite{moller1993scaled}, the model built by a
+multi-layer perceptron uses layers of the form
+
+$$l(x) : \mathbb{R}^{i} \rightarrow \mathbb{R}^{j} = \text{max}\big( x
+W_l, 0 \big)$$
+
+Where $W_l$ is the $i$ by $j$ weight matrix of layer $l$ and the max
+function is applied element-wise to the resulting vector. In this
+form, the multi-layer perceptron (MLP) produces a locally linear model
+of the input data that is $C^0$. The computational complexity of
+training a multi-layer perceptron is approximately $\mathcal{O}(n d
+m)$ where $m$ is a constant determined by a linear function of the
+sizes of the layers of the network and the convergence stopping
+criteria of the BFGS minimization used for finding weights. This paper
+uses a single hidden layer with 100 nodes, ReLU activation, and BFGS
+for training.
+
 \subsubsection{Support Vector Regressor}
+Support vector machines are a common tool used in machine learning
+classification tasks that can be adapted for the purpose of regression
+\cite{basak2007support}. The mathematical formalisms for exactly how
+to build a support vector regressor (SVR) are beyond the scope of this
+summary, but the core of the technique solves an optimization problem
+of the form
+
+$$ \text{Minimize } \bigg\{ \frac{1}{2}\sum_{i=1}^{n}\sum_{j=1}^{n}
+a_i a_j K(x_i, x_j) + \epsilon \sum_{i=1}^{n} a_i - \sum_{i=1}^{n} y_i
+a_i \bigg\} $$
+
+$$ \text{Subject to } \sum_{i=1}^{n}a_i = 0 \text{ } \text{ and }
+\text{ } a_i \in [0,C] $$
+
+Where $K$ is the selected kernel function, $a \in \mathbb{R}^n$ is the
+solved coefficient vector, and $\epsilon$ is the error tolerance.
 

### 2017-11-12 Algorithm_Comparison/py_alg_evaluation.py

@@ -296,47 +296,47 @@ for alg_name in to_test:
         data_scale = np.max(train_points, axis=0)
         train_points /= data_scale
         test_points /= data_scale
-        try:
-            # Fit the training points
-            # Get any extra arguments to pass to individual alrogithm fits
-            extra_args = EXTRA_ARGS.get(name,[])
-            # Wrap the fit function to monitor timeouts
-            fit_func = timeout_after(FIT_TIMEOUT_SECONDS,FAIL_VALUE)(alg.fit)
-            # Time the fit operation
-            fit_time = time.time()
-            fit_output = fit_func(train_points, train_values, *extra_args)
-            fit_time = time.time() - fit_time
-            if ((type(fit_output) == type(FAIL_VALUE)) and
-                (fit_output == FAIL_VALUE)):
-                print()
-                print('ERROR: %s failed to fit within %i seconds.'%(
-                    name,FIT_TIMEOUT_SECONDS))
-                print("Training file: '%s'"%(train))
-                print("Testing file:  '%s'"%(test))
-                print()
-                break # Cancel the rest of data collection for this algorithm
-            # Wrap the approximation function to monitor timeouts
-            approx_func = timeout_after(APPROX_TIMEOUT_SECONDS,FAIL_VALUE)(alg)
-            # Evaluate at the testing points
-            eval_time = time.time()
-            test_approx = approx_func(test_points)
-            eval_time = time.time() - eval_time
-            if ((type(test_approx) == type(FAIL_VALUE)) and
-                (test_approx == FAIL_VALUE)):
-                print()
-                print('ERROR: %s failed to create approximation within %i seconds.'%(
-                    name, APPROX_TIMEOUT_SECONDS))
-                print("Training file: '%s'"%(train))
-                print("Testing file:  '%s'"%(test))
-                print()
-                break # Cancel the rest of data collection for this algorithm
-        except:
+        # try:
+        # Fit the training points
+        # Get any extra arguments to pass to individual alrogithm fits
+        extra_args = EXTRA_ARGS.get(name,[])
+        # Wrap the fit function to monitor timeouts
+        fit_func = timeout_after(FIT_TIMEOUT_SECONDS,FAIL_VALUE)(alg.fit)
+        # Time the fit operation
+        fit_time = time.time()
+        fit_output = fit_func(train_points, train_values, *extra_args)
+        fit_time = time.time() - fit_time
+        if ((type(fit_output) == type(FAIL_VALUE)) and
+            (fit_output == FAIL_VALUE)):
[…]

### 2017-11-12 Algorithm_Comparison/py_test_plot.py

@@ -32,11 +32,11 @@ multiplier = 10
 # algorithms = [MaxBoxMesh()]
 # algorithms = [SplineMesh()]
 # algorithms = [NearestNeighbor()]
-algorithms = [LSHEP()]
+# algorithms = [LSHEP()]
 # algorithms = [LinearModel()]
 # algorithms = [Delaunay()]
 # algorithms = [ MLPRegressor() ] 
-# algorithms = [ SVR() ] 
+algorithms = [ SVR() ] 
 
 # algorithms = [NearestPlane()]
 

### 2017-11-12 HPC_Paper/HPC_Paper_Results/preprocess_data.py

@@ -1,0 +1,20 @@
+import os
+from util.data import read_struct
+from util.paper import latex_table
+
+
+predictors = ["Mean", "Variance"]
+algorithms = ["LSHEP", "MARS", "MLPRegressor", "qHullDelaunay", "SVR"]
+
+# Process all raw files into pkl files that hold the contents (for
+# easier loading later on)
+for p in predictors:
+    summary_output = p+"_all_summary.pkl"
+    all_data = None
+    for a in algorithms:
+        source = "Output-%s_%s-MDA_results.csv"%(a,p)
+        a_data = read_struct(source)
+        print(len(a_data), source)
+        
+# Read in the data files, merge them into one pkl
+# Process pkl as I do normally 

### 2017-11-12 HPC_Paper/paper.tex
Working on paper, deleted all test data folders for recreation on pawnee.

@@
 selected value that is the maximum complexity of the model in terms of
-the number of underlying basis functions. This paper uses a default
-max number of basis functions of $n$, however with larger datasets ($n
-> 1000$)this value would need to be reduced.
+the number of underlying basis functions. This paper uses an Earth
+implementation of MARS \cite{rudy2017pyearth} with a max number of
+basis functions of $n$, however with larger datasets ($n > 1000$) this
+value would need to be reduced.
 
@@
 criteria of the BFGS minimization used for finding weights. This paper
-uses a single hidden layer with 100 nodes, ReLU activation, and BFGS
-for training.
+uses the scikit-learn MLP regressor \cite{scikit-learn} with a single
+hidden layer with 100 nodes, ReLU activation, and BFGS for training.
 
@@
 to build a support vector regressor (SVR) are beyond the scope of this
-summary, but the core of the technique solves an optimization problem
-of the form
+summary, but the resulting functional fit is of the form
 
-$$ \text{Minimize } \bigg\{ \frac{1}{2}\sum_{i=1}^{n}\sum_{j=1}^{n}
-a_i a_j K(x_i, x_j) + \epsilon \sum_{i=1}^{n} a_i - \sum_{i=1}^{n} y_i
-a_i \bigg\} $$
+$$ f(x) : \mathbb{R}^d \rightarrow \mathbb{R} = \sum_{i=1}^{n}a_i
+K(x,x_i) $$
 
-$$ \text{Subject to } \sum_{i=1}^{n}a_i = 0 \text{ } \text{ and }
-\text{ } a_i \in [0,C] $$
+%% $$ \text{Minimize } \bigg\{ \frac{1}{2}\sum_{i=1}^{n}\sum_{j=1}^{n}
+%% a_i a_j K(x_i, x_j) + \epsilon \sum_{i=1}^{n} a_i - \sum_{i=1}^{n} y_i
+%% a_i \bigg\} $$
+
+%% $$ \text{Subject to } \sum_{i=1}^{n}a_i = 0 \text{ } \text{ and }
+%% \text{ } a_i \in [0,C] $$
 
 Where $K$ is the selected kernel function, $a \in \mathbb{R}^n$ is the
-solved coefficient vector, and $\epsilon$ is the error tolerance.
+solved coefficient vector, and $\epsilon$ is the error tolerance. The
+SVR is always fitting a linear function, however the kernel mapping of
+the data can be non-linear to create non-linear fits. The
+computational complexity of the SVR minimization (for the
+implementation used) is $\mathcal{O}(n^2dm)$ with $m$ being a constant
+determined by the minimization convergence criteria. This paper uses
+the scikit-learn SVR \cite{scikit-learn} with a polynomial kernel
+function.
 
 \subsection{Interpolation}
+
 \subsubsection{Delaunay}

### 2017-11-12 Algorithm_Comparison/py_alg_evaluation.py

@@ -82,7 +82,7 @@ if ((not os.path.exists(MEAN_MDA_DIR)) or
     stat_data = []
     for key in data_keys:
         vals = np.array(unique_pts[key])
-        stat_data.append( key + (vals.mean(), vals.var()) )
+        stat_data.append( key + (vals.mean(), vals.var(ddof=1)) )
     # List of the column names and their data types, simply adding two
     # columns for mean and variance for conversion back into numpy
     # structured arrays.

### 2017-11-27 HPC_Paper/paper.tex
Working on the HPC paper. Processing results and generating visuals. Added code for a new type of interpolation that might be promising (because of its scalability to large dimension even with few points).

--- 2017-11-11 HPC_Paper/paper.tex
@@
 \usepackage{amsthm}
+\usepackage{algorithm}     % needed for the algorithm block
+\usepackage{algpseudocode} % algorithmicx psuedo-code styling
 %\usepackage[pdftex,colorlinks=true,urlcolor=blue,citecolor=black,anchorcolor=black,linkcolor=black]{hyperref}
@@
 \renewcommand{\thedefinition}{\arabic{definition}}
+\newcommand{\rpm}{\raisebox{.2ex}{$\scriptstyle\pm$}}
 
@@
 %======================
-\section{Introduction}
+\section{Introduction and related work}
+\label{sec:introduction}
 
-Performance tuning is often an experimentally complex and time-intense
-chore necessary for configuring HPC systems. The procedures for this
-tuning vary largely from system to system and are often subjectively
-guided by the system engineer(s). Once a desired level of performance
-is achieved, an HPC system may only be incremenetally reconfigured as
-required by updates or specific jobs. In the case that a system has
-changing workloads or non-stationary performance objectives that range
-from maximizing computational throughput to minimizing power
-consumption and system variability, it becomes clear that a more
-effective and automated tool is needed for configuring systems. This
-scenario presents a challenging and important application of
-multivariate approximation and interpolation techniques.
+Performance tuning is often an experimentally complex and time
+intensive chore necessary for configuring HPC systems. The procedures
+for this tuning vary largely from system to system and are often
+subjectively guided by the system engineer(s). Once a desired level of
+performance is achieved, an HPC system may only be incrementally
+reconfigured as required by updates or specific jobs. In the case that
+a system has changing workloads or non stationary performance
+objectives that range from maximizing computational throughput to
+minimizing power consumption and system variability, it becomes clear
+that a more effective and automated tool is needed for configuring
+systems. This scenario presents a challenging and important
+application of multivariate approximation and interpolation
+techniques.
 
@@
 statistical model of the performance by running experiments on the
-system at select settings or (2) use a performance simulator to run
-artificial experiments and search for optimal configurations. In this
-paper the proposed multivariate modelling techniques rest in the first
-category, however they represent a notable increase in the ability to
-model more complex interactions between system parameters.
+system at select settings or (2) run artificial experiments using a
+simplified simulation of the target system to estimate architecture
+and application bottlenecks. In this paper the proposed multivariate
+modeling techniques rest in the first category, and they represent a
+notable increase in the ability to model complex interactions between
+system parameters.
 
 Many previous works attempting to model system performance have used
-simulated environments to estimate the performance of a
-system. \cite{grobelny2007fase} \cite{wang2009simulation}
-\cite{wang2013towards} Some of these works cite statistical models as
-being over-simplified and not capabel of capturing the true complexity
-of the underlying system. This claim is partially correct, noticing
-that a large portion of predictive statistical models rely on
-simplifying the system to one or two
-parameters. \cite{snavely2002framework} \cite{bailey2005performance}
-\cite{barker2009using} \cite{ye2010analyzing} These limited
-statistical models have provided satisfactory performance in very
-narrow application settings. Many of the aforementioned statistical
-modeling techniques claim to generalize, while simultaneously
-requiring additional code annotations, hardware abstractions, or
-additional application level understandings in order to generate
-models. The approach presented here requires no modifications of the
-application, no architecture abstractions, nor any structural
-descriptions of the input data being modelled. The techniques used are
-purely mathematical and only need formatted data as input.
+simulated environments to estimate the performance of a system
+\shortcite{grobelny2007fase,wang2009simulation,wang2013towards}. Some
+of these works refer to statistical models as being oversimplified and
+not capable of capturing the true complexity of the underlying
+system. This claim is partially correct, noting that a large portion
+of predictive statistical models rely on simplifying the model to one
+or two parameters
+\shortcite{snavely2002framework,bailey2005performance,barker2009using,ye2010analyzing}.
+These limited statistical models have provided satisfactory
+performance in very narrow application settings. Many of the
+aforementioned statistical modeling techniques claim to generalize,
+while simultaneously requiring additional code annotations, hardware
+abstractions, or additional application level understandings in order
+to generate models. The approach presented here requires no
+modifications of the application, no architectural abstractions, nor
+any structural descriptions of the input data being modeled. The
+techniques used are purely mathematical and only need performance data
+as input.
 
-Among the statistcal models presented in prior works,
-\cite{bailey2005performance} specifically mentioned that it is
-difficult for the models to capture variability introduced by
-I/O. System variability in general has become a problem of increasing
-interest to the systems and HPC communities, however most of the work
-has focused on Operating System induced variability.
-\cite{beckman2008benchmarking} \cite{de2007identifying} The work that
-has focused on managing I/O variability, \cite{lofstead2010managing},
-does not use any sophisticated modeling techniques. Hence, this paper
-presents a case study applying advanced mathematical and statistical
-modeling techniques to the domain of HPC system I/O
+Among the statistical models presented in prior works
+\shortciteN{bailey2005performance} specifically mention that it is
+difficult for the simplified models to capture variability introduced
+by I/O. System variability in general has become a problem of
+increasing interest to the HPC and systems communities, however most
+of the work has focused on operating system (OS) induced variability
+\shortcite{beckman2008benchmarking,de2007identifying}. The work that
+has focused on managing I/O variability does not use any sophisticated
+modeling techniques \shortcite{lofstead2010managing}. Hence, this
+paper presents a case study applying advanced mathematical and
+statistical modeling techniques to the domain of HPC I/O
 characteristics. The models are used to predict the mean throughput of
 a system and the variance in throughput of a system. The discussion
-section outlines how the exact techniques presented can be applied to
-any performance metric and any system.
+section outlines how the techniques presented can be applied to any
+performance metric and any system.
 
@@
 In general, this paper compares five multivariate approximation
-techniques that operate on inputs in $\mathbb{R}^d$ (vectors of $d$
-real numbers) and produce predicted responses in $\mathbb{R}^1$. In
+techniques that operate on inputs in $\mathbb{R}^d$ ($d$-tuples of
+real numbers) and produce predicted responses in $\mathbb{R}$. In
 order to provide coverage of the varied mathematical strategies that
 can be employed to solve the continuous modeling problem, three of the
-techniques are regression-based and the remaining two techniques
-produce interpolants. The sections below outline the mathematical
-formulations of each technique and provide computational complexity
-bounds with respect to the size (number of points and dimension) of
-input data.
+techniques are regression based and the remaining two are
+interpolants. The sections below outline the mathematical formulation
+of each technique and provide computational complexity bounds with
+respect to the size (number of points and dimension) of input
+data. Throughout the sections, $d$ will refer to the dimension of the
+input data, $n$ is the number of points in the input data, $x^{(i)}
+\in \mathbb{R}^d$ is the $i$-th input data point, $x^{(i)}_j$ is the
+$j$-th component of $x^{(i)}$, and $f(x^{(i)})$ is the response value
+of the $i$-th input data point.
 
-\subsection{Approximation}
-Multivariate approximations are capable of accurately modelling
-complex relationships between the dimensions of any set of points in
-$\mathbb{R}^n$ with respect to some response in $\mathbb{R}$. The
-approximations produce a function $f: \mathbb{R}^n \rightarrow
-\mathbb{R}$ which minimizes some metric related to the provided data.
+The remainder of the paper is broken up into four major parts. Section
+\ref{sec:multivariate} provides an overview of the multivariate
+modeling techniques, Section \ref{sec:methodology} outlines the
+methodology for comparing and evaluating the performance of the
+models, Section \ref{sec:results} presents the IOzone predictions,
+Section \ref{sec:discussion} discusses the obvious and subtle
+implications of the models' performance, and Section
+\ref{sec:conclusion} concludes and offers directions for future work.
+
+\section{Multivariate Models}
+\label{sec:multivariate}
+
+\subsection{Regression}
+Multivariate approximations are capable of accurately modeling a
+complex dependence of a response (in $\mathbb{R}$) on multiple
+variables (represented as a points in $\mathbb{R}^{d}$). The
+approximations to some (unknown) underlying function $f: \mathbb{R}^d
+\rightarrow \mathbb{R}$ are chosen to minimize some error measure
+related to data samples $f(x^{(i)})$. For example, least squares
+regression uses the sum of squared differences between modeled
+response values and true response values as an error measure.
 
 \subsubsection{Multivariate Adaptive Regression Splines}
-\subsubsection{Multi-Layer Perceptron Regressor}
+This approximation was introduced in
+\shortciteN{friedman1991multivariate} and subsequently improved to its
+current version in \shortciteN{stanford1993fast}, called fast
+multivariate adaptive regression splines (Fast MARS). In Fast MARS, a
+least squares fit model is iteratively built by beginning with a
+single constant valued function and adding two new basis functions at
+each iteration of the form
+
+$$ B_{2s-1}(x) = B_l(x) [c(x_i-v)]_+ ,$$
+$$ B_{2s}(x) = B_k(x) [c(x_i-v)]_- ,$$
+
+where $s$ is the iteration number, $B_l(x)$ and $B_k(x)$ are basis
+functions from the previous iteration, $c, v \in \mathbb{R}$, $w_+ =
+\{w, w \geq 0, 0, w < 0\},$ and $w_+ = (-w)_+$. After iteratively
+constructing a model, MARS then iteratively removes basis functions
+that do not contribute to goodness of fit. In effect, MARS creates a
+locally component-wise tensor product approximation of the data. The
+overall computational complexity of Fast MARS is $\mathcal{O}(n d
+m^3)$ where $m$ is the maximum number of underlying basis
+functions. This paper uses an implementation of MARS
+\shortcite{rudy2017pyearth} with $m = 200$.
+
+\subsubsection{Multilayer Perceptron Regressor}
+The neural network is a well studied and widely used method for both
+regression and classification tasks
+\shortcite{hornik1989multilayer}. When using the rectified linear unit
+(ReLU) activation function \shortcite{dahl2013improving} and training
+with the BFGS minimization technique \shortcite{moller1993scaled}, the
+model built by a multilayer perceptron uses layers $l : \mathbb{R}^{i}
+\rightarrow \mathbb{R}^{j}$ defined by
+
+$$ l(u) = \big( u^t W_l \big)_+ $$
+
+where $W_l$ is the $i$ by $j$ weight matrix for layer $l$. In this
+form, the multi layer perceptron (MLP) produces a piecewise linear
+model of the input data. The computational complexity of training a
+multi layer perceptron is $\mathcal{O}(n d m)$, where $m$ is
+determined by the sizes of the layers of the network and the stopping
+criterion of the BFGS minimization used for finding weights. This
+paper uses the scikit-learn MLP regressor \shortcite{scikit-learn}, a
+single hidden layer with 100 nodes, ReLU activation, and BFGS for
+training.
+
 \subsubsection{Support Vector Regressor}
+Support vector machines are a common method used in machine learning
+classification tasks that can be adapted for the purpose of regression
+\shortcite{basak2007support}. How to build a support vector regressor
+(SVR) is beyond the scope of this summary, but the resulting
+functional fit $p : \mathbb{R}^d \rightarrow \mathbb{R}$ has the form
+
+$$ p(x)  = \sum_{i=1}^{n}a_i K(x,x^{(i)}) + b ,$$
+
+%% $$ \text{Minimize } \bigg\{ \frac{1}{2}\sum_{i=1}^{n}\sum_{j=1}^{n}
+%% a_i a_j K(x_i, x_j) + \epsilon \sum_{i=1}^{n} a_i - \sum_{i=1}^{n} y_i
+%% a_i \bigg\} $$
+
+%% $$ \text{Subject to } \sum_{i=1}^{n}a_i = 0 \text{ } \text{ and }
+%% \text{ } a_i \in [0,C] $$
+
+where $K$ is the selected kernel function, $a \in \mathbb{R}^n$, $b
+\in \mathbb{R}$ are coefficients to be solved for simultaneously with
+$b \in \mathbb{R}$, and $\epsilon$ is the error tolerance. The
+computational complexity of the SVR is $\mathcal{O}(n^2dm)$, with $m$
+being determined by the minimization convergence criterion. This paper
+uses the scikit-learn SVR \shortcite{scikit-learn} with a polynomial
+kernel function.
 
 \subsection{Interpolation}
+
+In some cases it is desirable to have a model that can recreate the
+input data exactly. This is especially the case when the confidence in
+the response values for known data is high. Both interpolation models
+analyzed in this paper are based on linear functions.
+
 \subsubsection{Delaunay}
+
+The Delaunay method of interpolation is a well studied geometric
+technique for producing an interpolant \shortcite{lee1980two}. The
+Delaunay triangulation of a set of data points into simplices is such
+that the sphere defined by the vertices of each simplex contains no
+data points in the sphere's interior. For a $d$-simplex S with
+vertices $v^{(0)}, v^{(1)}, \ldots, v^{(d)}$, $x \in S$, and data
+values $f(v^{(i)})$, $i=0,\ldots,d$, $x$ is a unique convex
+combination of the vertices,
+
+$$ x = \sum_{i=0}^{d} w_i v^{(i)}; \quad \sum_{i=0}^{d} w_i = 1; \quad
+w_i \geq 0; \quad i=0,\ldots,d $$
+
+and the Delaunay interpolant to $f$ at $x$ is given by
+
+$$ p(x) = \sum_{i=0}^{d} w_i f(v^{(i)}). $$
+
+The computational complexity of the Delaunay triangulation (for the
+implementation used here) is $\mathcal{O}(n^{\lceil d/2 \rceil})$,
+which is not scalable to $d > 10$ \shortcite{sartipizadeh2016computing}.
+The scipy interface \shortcite{scipy} to the QuickHull implementation
+\shortcite{barber1996qhull} of the Delaunay triangulation is used here.
+
 \subsubsection{Linear Shepard}
 
-The remainder of the paper is broken up into \_ parts. \textit{(will
-  fill in the parts and descriptions once they are finalized)}
+The linear Shepard method (LSHEP) is a blending function using local
+linear interpolants, a special case of the general Shepard algorithm
+\shortcite{thacker2010algorithm}. The interpolant has the form
+
+$$ p(x) = \frac{\sum_{k=1}^{n}W_k(x)P_k(x)}{\sum_{k=1}^{n}W_k(x)} ,$$
+
+where $W_k(x)$ is a locally supported weighting function and $P_k(x)$
+is a local linear approximation to the data satisfying $P_k(x^{(k)} =
+f(x^{(x)})$. The computational complexity LSHEP is
+$\mathcal{O}(n^2d^3)$. This paper uses the FORTRAN95 implementation of
+LSHEP in SHEPPACK \shortcite{thacker2010algorithm}.
 
@@
 \section{Methodology}
+\label{sec:methodology}
 \subsection{Data}
-The summary of the data used in the experiments for this paper can be
-seen in Table \ref{tab:data_type}.
+In order to evaluate the viability of multivariate models for
+predicting system performance, this paper presents a case study of a
+four-dimensional dataset produced by executing the IOzone benchmark
+from \shortciteN{iozone} on a homogeneous cluster of computers. The
+system performance data was collected by executing IOzone 40 times for
+each of a select set of system configurations. A single IOzone
+execution reports the max I/O throughput seen for the selected
+test. The 40 executions for each system configuration are converted
+into the mean and sample variance, both values in $\mathbb{R}$ capable
+of being modeled individually by the multivariate approximation
+techniques presented in Section \ref{sec:multivariate}. The summary of
+data used in the experiments for this paper can be seen in Table
+\ref{tab:data_type}.  Distributions of raw throughput values being
+modeled can be seen in Figure \ref{fig:raw_throughput}.
 
@@
   \centering
-  \begin{tabular}{|c|c|c|}
+  \begin{tabular}{c|c}
     \hline
-    \textbf{System Parameter} & \textbf{Type} & \textbf{Values}\\
+    \textbf{System Parameter} & \textbf{Values}\\
     \hline
-    Hypervisor Scheduler & Categorical & CFQ, NOOP, DEAD\\
-    Operating System Scheduler & Categorical & CFQ, NOOP, DEAD\\
-    IOZone Test Type & Categorical & Fread, Fwrite, RandomRead\\
-    File Size & Continuous & 64, 128, 256, 512, 1024\\
-    Record Size & Continuous & 32 64, 128, 256, 512\\
-    Thread Count & Continuous & 1,2,4,8,16,32,64\\
-    Frequency & Continuous & 2.4, 2.5, 2.6, 2.7, 2.8, 2.9, 3.0, 3.1\\
+    File Size & 64, 256, 1024\\
+    Record Size & 32, 128, 512\\
+    Thread Count & 1, 2, 4, 8, 16, 32, 64, 128, 256\\
+    Frequency & \{12, 14, 15, 16, 18, 19, 20, 21, 23, 24, 25, 27, 28, 29, 30, 30.01\} $\times 10^5$\\
+    \hline
+    \textbf{Response Values} & \\
+    \hline
+    Throughput Mean & [$2.6 \times 10^5$, $5.9 \times 10^8$]\\
+    Throughput Sample Variance & [$5.9\times 10^{10} $, $4.7 \times 10^{16}$]\\
     \hline
@@
   \caption{A description of the system parameters being considered in
-    the IOZone tests. There are three categorical settings and four
-    continuous settings. Notice that the ranges of values for
-    continuous settings are different.}
+    the IOzone tests. Record size must not be greater than file size
+    and hence there are only six valid combinations of the two. In
+    total there are $6 \times 9 \times 16 = 864$ unique system
+    configurations.}
   \label{tab:data_type}
@@
 
+\begin{figure}
+  \centering
+  \includegraphics[width=\textwidth,trim={0 .5in 0 1in}]{Raw_Throughput.pdf}
+  \caption{Histograms of 100-bin reductions of the PMF of I/O
+    throughput mean (top) and I/O throughput sample variance
+    (bottom). In the mean plot, the first 1\% bin (truncated in plot)
+    has a probability mass of .45. In the variance plot, the first 1\%
+    bin has a probability mass of .58. It can be seen that the
+    distributions of throughputs are primarily of lower magnitude with
+    occasional extreme outliers.}
+  \label{fig:raw_throughput}
+\end{figure}
+
 \subsection{Dimensional Analysis}
-This work utilizes an extension to standard k-fold cross validation
+This work utilizes an extension to standard $k$-fold cross validation
 that allows for a more thorough investigation of the expected model
@@
 training data provided, and the dimension of the input data. It is
-important to consider that algorithms which perform well with less
-training input also require less experimentation. Although, the amount
+important to consider that algorithms that perform well with less
+training input also require less experimentation, although the amount
 of training data required may change as a function of the number of
-input dimensions and this needs to be studied as well.
+input dimensions, and this needs to be studied as well. The framework
+used here will be referred to as a multidimensional analysis (MDA) of
+the IOzone data.
 
-The framework used in this paper will be referred to as a
-multi-dimensional analysis (MDA) of the IOZone data presented in this
-study. However, the MDA framework can be applied to other datasets.
-
+\subsubsection{Multidimensional Analysis}
+This procedure combines random selection of training and testing
+splits with changes in the input dimension and the ratio of training
+size to testing size. Given an input data matrix with $n$ rows
+(points) and $d$ columns (features), the multidimensional analysis
+(MDA) is performed as follows:
 \begin{enumerate}
-\item Cycling the categorical settings
-\item Selecting subsets of 1,2,3 up to 4 dimensions
-\item Cycling different training : testing ratios (5:95 $\rightarrow$ 95:5)
-\item Generating 200 random training : testing splits
-\item Ensuring the testing points are on/inside the convex hull of the training.
-\item Ensuring the training points are well-spaced.
+\item for all $k = 1, \ldots, d$ and for all subsets of input features
+  $F = \{ j_1, j_2, \ldots j_k \}$; reduce the input data to unique
+  points $z \in \mathbb{R}^k$ with $f(z) = E[ \{ f(x^{(i)}) \quad
+    \forall i \quad s.t. \quad (x^{(i)}_F = z) \} ]$.
+\item for all $r$ in $\{5, 10, \ldots, 95\}$; generate $N$ random
+  splits $(train, test)$ of the reduced data with $r$ percentage for
+  training and $100 - r$ percentage for testing.
+\item when generate each random $(train, test)$ split, ensure that all
+  points from $test$ are on or inside the convex hull of points in
+  $train$; also ensure that the points in $train$ are well spaced.
 \end{enumerate}
 
-\subsection{Prediction}
+In order to ensure that the testing points are inside the convex hull
+of training points, this paper identifies the set of (reduced
+dimension) points that compose the convex hull and forcibly places
+those points in the training set. In order to ensure that training
+points are well spaced, this paper relies on the statistical method
+for picking points from \shortciteN{amos2014algorithm} that procedes
+as follows:
 \begin{enumerate}
-\item For each file generated from the dimensional analysis, train on
-  the training data, evaluate at the testing data points
+\item Generate a sequence of all pairs of points sorted by ascending
+  pairwise $L_2$ distance between points, ${(x^{(i_1)},x^{(j_1)}),
+    (x^{(i_2)},x^{(j_2)}), \ldots}$ such that
+  $||x^{(i_k)}-x^{(j_k)}||_2 \leq ||x^{(i_{k+1})}-x^{(j_{k+1})}||_2$
+\item Iteratively remove points from candidacy until only $N$ remain
+  by sequentially selecting one point from the pair $(x^{(i_k)},
+  x^{(j_k)})$ for $k = 1,\ldots$ if both $x^{(i_k)}$ and $x^{(j_k)}$
+  are still candidates.
 \end{enumerate}
+
+Given the large number of constraints, level of reduction, and use of
+randomness in the MDA procedure, occasionally $N$ unique training
+testing splits may not be created or may not exist. In these cases, if
+there are fewer than $N$ possible splits, those deterministically
+generated splits are used. Otherwise after $3*N$ attempts, only the
+unique splits are kept for analysis. The MDA procedure has been
+implemented in python3 while most regression and interpolation methods
+are FORTRAN wrapped with python. All randomness has been seeded for
+reproducability.
+
+For any input feature subset $F$ (of size $k$) and selected value of
+$r$, MDA will generate up to $N$ predictions that each multivariate
+model has made for each point $z^{(i)} \in \mathbb{R}^k$. There may be
+fewer than $N$ predictions made for any given point. Points in the
+convex hull for the selected subest of features will always be used
+for training, never for testing. Points that do not have any close
+neighbors will often be used for training in order to ensure well
+spacedness. Finally, as mentioned before, some subsets of features do
+not readily generate $N$ unique training and testing splits.
+
+MDA results in a distribution of possible estimates that a given
+multivariate model could make for a point in $\mathbb{R}^k$, $k \leq
+d$ given feature set $F$. The summary results presented in this work
+use the expected value of the distributions at each point as the model
+estimate for error analysis.
+
+%% is open source, and is freely available at the url
+%% \textit{https://github.com/tchlux/VarSys/HPC\_Paper/Code/multi\_dim\_analysis.py}
+
+%% \begin{enumerate}
+%% \item Cycling the categorical settings
+%% \item Selecting subsets of 1,2,3 up to 4 dimensions
+%% \item Cycling different training : testing ratios (5:95 $\rightarrow$ 95:5)
+%% \item Generating 200 random training : testing splits
+%% \item Ensuring the testing points are on/inside the convex hull of the training.
+%% \item Ensuring the training points are well-spaced.
+%% \end{enumerate}
+
+%% \subsection{Prediction}
+%% \begin{enumerate}
+%% \item For each file generated from the dimensional analysis, train on
+%%   the training data, evaluate at the testing data points
+%% \end{enumerate}
 
@@
 \section{Results}
+\label{sec:results}
+
+A naive multivariate prediction technique such as nearest neighbor
+could experience relative errors in the range $[0,10^m]$ where $m$ is
+$log_{10}(\text{max}_f - \text{min}_f)$ when modeling a function $f$
+from data. The IOzone mean data response values span 3 orders of
+magnitude (as can be seen in Table \ref{tab:data_type}) while variance
+data response values span 6 orders of magnitude. It is expected
+therefore, that all studied multivariate models perform better than a
+naive approach, achieving relative errors strictly less than $10^3$
+for mean throughput and $10^6$ when for throughput sample
+variance. Ideally, models will yield relative errors significantly
+smaller than 1.
+
 \subsection{I/O Throughput Mean}
 
-%% \begin{center}
-%%   \includegraphics[width=\textwidth]{Prediction_Performance_Dim.pdf}
-%%   \caption{In this figure we see how the distribution of errors }
-%% \end{center}
+Almost all multivariate models analyzed make predictions with a
+relative error less than 1 for most system configurations when
+predicting the mean I/O throughput of a system given varying amounts
+of training data. The overall best of the multivariate models,
+Delaunay, consistently makes predictions with relative error less than
+$.05$ (5\% error). In figure \ref{fig:mean_tt_ratio} it can also be
+seen that the Delaunay model consistently makes good predictions even
+with as low as 5\% training data (43 system configurations) regardless
+of the dimension of the data.
 
@@
   \centering
-  \includegraphics[width=\textwidth]{Prediction_Performance_TT_Ratio.pdf}
-  \caption{In this figure, it can be seen how the distribution of
-    prediction errors for each algorithm is affected by the quantity
-    of training data provided. Notice that for all amounts of training
-    data, Delaunay has the largest likelihoods around 0 error.}
-  \label{fig:tt_ratio}
+  \includegraphics[width=\textwidth,trim={0 1in 0 1in}]{Mean_TT_Ratio.pdf}
+  \caption{These box plots show the prediction error of means with
+    increasing amounts of training data provided to the models. Notice
+    that the response values being predicted span 3 orders of
+    magnitude and hence relative error in that range would not be
+    unexpected. For SVR the top box whisker goes from around 100 to 50
+    from left to right and is chopped in order to maintain focus on
+    the more performant models.}
+  \label{fig:mean_tt_ratio}
 \end{figure}
 
-\begin{table}
+\subsection{I/O Throughput Variance}
+
+The prediction results for variance resemble that of predicting I/O
+throughput mean, with each of the errors being scaled proportional to
+the change in scale of the response values. Delaunay remains the best
+overall predictor with LSHEP closely competing, but yielding much more
+outliers.
+
+\subsection{Increasing Dimension}
+
+As can be seen in figure \ref{fig:mean_dim}, all of the models suffer
+increasing error rates in higher dimension. This is expected, because
+the number of possible interactions to model grows
+exponentially. However, LSHEP and Delaunay maintain the slowest
+increase in relative error. The particularly slow increase in error
+seen for Delaunay appears to suggest that it is capable of making
+predictions with a range of relative errors approximately spanning
+\rpm $.2*\text{\textit{dim}}$.
+
+\begin{figure}
   \centering
-  \begin{tabular}{|c||c|c|c|}
-    %% \hline
-    %% \multicolumn{4}{|c|}{Table Title}\\
-    \hline
-    \textbf{Algorithm} & \textbf{Input Dimension} & \textbf{Mean Absolute Error} & \textbf{Mean Relative Error}\\
-    \hline
-    Delaunay     & 1 & 2060267   & 0.03 \\
-                 & 2 & 2916224   & 0.07 \\
-                 & 3 & 3001794   & 0.09 \\
-                 & 4 & 3134779   & 0.10 \\
-    \hline
-    LSHEP        & 1 & 6353148   & 0.06 \\
-                 & 2 & 13038882  & 1.52 \\
-                 & 3 & 5446207   & 1.13 \\
-                 & 4 & 5133648   & 1.07 \\
-    \hline
-    MARS         & 1 & 2176368   & 0.03 \\
-                 & 2 & 6213369   & 0.61 \\
-                 & 3 & 4024964   & 0.65 \\
-                 & 4 & 4497414   & 1.01 \\
-    \hline
-    MLPRegressor & 1 & 2461097   & 0.06 \\
-                 & 2 & 8883580   & 0.81 \\
-                 & 3 & 11190977  & 2.03 \\
-                 & 4 & 11831721  & 2.20 \\
-    \hline
-    SVR          & 1 & 19306731  & 0.96 \\
-                 & 2 & 78213963  & 18.49\\
-                 & 3 & 97761266  & 31.27\\
-                 & 4 & 113551300 & 37.2 \\
-    \hline
-  \end{tabular}
-  \caption{Most models experience a decay in predictive performance as
-    the dimension of the data increases. This is expected because
-    higher dimension input has exponentially more possible patterns to
-    identify. The MLP Regressor and SVR experience the worst decay in
-    performance with increasing dimension.}
-  \label{tab:perf_by_dim}
-\end{table}
-
-\begin{table}
-  \centering
-  \begin{tabular}{|c||c|c|c|}
-    %% \hline
-    %% \multicolumn{4}{|c|}{Table Title}\\
-    \hline
-    \textbf{Algorithm} & \textbf{Training Data} & \textbf{Mean Absolute Error} & \textbf{Mean Relative Error}\\
-    \hline
-    Delaunay     & [5-21]\%  & 5274933   & 0.11 \\
-                 & [25-46]\% & 2898330   & 0.08 \\
-                 & [50-74]\% & 1518477   & 0.07 \\
-                 & [75-96]\% & 886346    & 0.08 \\
-    \hline
-    LSHEP        & [5-21]\%  & 13218469  & 2.53 \\
-                 & [25-46]\% & 4936381   & 0.85 \\
-                 & [50-74]\% & 2251409   & 0.26 \\
-                 & [75-96]\% & 1415034   & 0.15 \\
-    \hline
-    MARS         & [5-21]\%  & 6543153   & 0.61 \\
-                 & [25-46]\% & 4324916   & 0.80 \\
-                 & [50-74]\% & 3115943   & 0.95 \\
-                 & [75-96]\% & 2591451   & 0.74 \\
-    \hline
-    MLPRegressor & [5-21]\%  & 17348982  & 2.58 \\
-                 & [25-46]\% & 12504957  & 2.29 \\
-                 & [50-74]\% & 5292177   & 1.17 \\
-                 & [75-96]\% & 2996253   & 1.02 \\
-    \hline
-    SVR          & [5-21]\%  & 170906944 & 49.45\\
-                 & [25-46]\% & 99851846  & 31.28\\
-                 & [50-74]\% & 51565405  & 19.63\\
-                 & [75-96]\% & 27125130  & 12.6 \\
-    \hline
-  \end{tabular}
-  \caption{All models experience a reduction in error with increasing
-    amounts of training data. This suggests that the data being
-    modeled is pattern-dense, over-fitting is \textit{not} occurring,
-    and that oversimplifications will tend towards worse predictions.}
-  \label{tab:perf_by_tt}
-\end{table}
-
-\subsection{I/O Throughput Variance}
-\subsection{Increasing Dimension}
+  \includegraphics[width=\textwidth,trim={0 1in 0 2in}]{Mean_Dim.pdf}
+  \caption{These box plots show the prediction error of means with
+    increasing dimension. }
+  \label{fig:mean_dim}
+\end{figure}
 
@@
 \section{Discussion}
+\label{sec:discussion}
+
 \subsection{Modeling the System}
-\subsection{Quantifying Variability}
+\begin{enumerate}
+\item Promise of the models, especially Delaunay
+\item Possible reasons for the performance differences between models
+\item The need for as few as 50 experiments to get less than 5\% error
+\item The unfortunate inability of Delaunay to scale to higher dimension
+\end{enumerate}
+
 \subsection{Extending the Analysis}
+\begin{enumerate}
+\item Measuring different system characteristics other than I/O
+\item Recording system performance constantly to maintain model accuracy
+\end{enumerate}
+
+\section{Conclusion}
+\label{sec:conclusion}
+\begin{enumerate}
+\item Delaunay is the best model, capable of predicting the
+  performance of 800 different system configurations accurately after
+  only being trained on a well-spaced set of 50 configurations.
+\item More comparisons of models similar to Delaunay should be done to
+  evaluate what mathematical characteristics of multivariate models are
+  most effective.
+\end{enumerate}
 
@@
 %=====================
-\section{Future Work}
+\subsection{Future Work}
 The most severe limitation to the presented work is the present
@@
 because only MARS is capable of handling non-numeric variables, or as
-a limitation of the MDA framework.
+a limitation of the MDA framework. Future works could attempt to
+identify the viability of reducing categorical variables to numerical
+ones to expand set of predictable system configurations.
+
+\begin{enumerate}
+\item Using models to predict more complex responses, such as functions
+\end{enumerate}
+
+%% \newpage
+%% \_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\\
+
+%% \vspace{3in}
+
+%% \texttt{\textbf{L}oc\textbf{A}ll\textbf{Y} li\textbf{NE}ar
+%%   \textbf{W}eighted \textbf{A}pproxima\textbf{T}ion\textbf{S}
+%%   \textbf{O}f fu\textbf{N}tions}
+%% %% L  A  Y   NE   W        A        T   S O    N
+%% \newpage
 

### 2017-11-27 Algorithm_Comparison/py_alg_evaluation.py

@@ -274,6 +274,7 @@ for alg_name in to_test:
         # - load the training and testing data files
         # - train the algorithm on the training data
         # - test the algorithm on the testing data
+        # - record the prediction performance for each point predicted
         perc_complete = 100.0*file_iterator.current/len(file_iterator)
 
         print("%.2f%% of %i tests"%(perc_complete, len(file_iterator)),
@@ -296,47 +297,47 @@ for alg_name in to_test:
         data_scale = np.max(train_points, axis=0)
         train_points /= data_scale
         test_points /= data_scale
-        # try:
-        # Fit the training points
-        # Get any extra arguments to pass to individual alrogithm fits
-        extra_args = EXTRA_ARGS.get(name,[])
-        # Wrap the fit function to monitor timeouts
-        fit_func = timeout_after(FIT_TIMEOUT_SECONDS,FAIL_VALUE)(alg.fit)
-        # Time the fit operation
-        fit_time = time.time()
-        fit_output = fit_func(train_points, train_values, *extra_args)
-        fit_time = time.time() - fit_time
-        if ((type(fit_output) == type(FAIL_VALUE)) and
-            (fit_output == FAIL_VALUE)):
+        try:
+            # Fit the training points
+            # Get any extra arguments to pass to individual alrogithm fits
+            extra_args = EXTRA_ARGS.get(name,[])
+            # Wrap the fit function to monitor timeouts
+            fit_func = timeout_after(FIT_TIMEOUT_SECONDS,FAIL_VALUE)(alg.fit)
+            # Time the fit operation
+            fit_time = time.time()
+            fit_output = fit_func(train_points, train_values, *extra_args)
+            fit_time = time.time() - fit_time
+            if ((type(fit_output) == type(FAIL_VALUE)) and
+                (fit_output == FAIL_VALUE)):
+                print()
+                print('ERROR: %s failed to fit within %i seconds.'%(
+                    name,FIT_TIMEOUT_SECONDS))
+                print("Training file: '%s'"%(train))
+                print("Testing file:  '%s'"%(test))
+                print()
+                break # Cancel the rest of data collection for this algorithm
+            # Wrap the approximation function to monitor timeouts
+            approx_func = timeout_after(APPROX_TIMEOUT_SECONDS,FAIL_VALUE)(alg)
+            # Evaluate at the testing points
+            eval_time = time.time()
+            test_approx = approx_func(test_points)
+            eval_time = time.time() - eval_time
+            if ((type(test_approx) == type(FAIL_VALUE)) and
[…]

### 2017-11-27 Algorithm_Comparison/py_test_plot.py

@@ -7,17 +7,19 @@ import numpy as np
 from util.plotly import *
 from py_settings import *
 
+from nearest import Linn
+
 # Preparing the data
 normalize_points = True
-random_points = False
-random_perturbation = False
+random_points = True
+random_perturbation = True
 tight_plot = False
 plot_results = True
 perform_test = False
 
 perturbation_multiplier = 1.0
 plot_points = 4000
-num_points = 10
+num_points = 4
 dim = 2
 multiplier = 10
 
@@ -36,11 +38,12 @@ multiplier = 10
 # algorithms = [LinearModel()]
 # algorithms = [Delaunay()]
 # algorithms = [ MLPRegressor() ] 
-algorithms = [ SVR() ] 
+# algorithms = [ SVR() ] 
+algorithms = [ Linn() ]
 
 # algorithms = [NearestPlane()]
 
-np.random.seed(0)
+np.random.seed(2)
 
 # Generate testing points in a grid pattern
 plot_range = [[-0.3,1.3]]*dim # [[0,1]]*dim # 
@@ -114,7 +117,10 @@ if plot_results:
     print("Plotting..")
     print()
     p = Plot()
+    p2 = Plot()
+    p2.add("Input Data", *(points[:,:-1].T))
     p.add("Raw data",*(points.T))
+
     use_gradient = len(algorithms) <= 1
     surf_x = np.meshgrid(*[
         np.linspace(lower,upper,PLOT_POINTS**(1/dim))
@@ -131,12 +137,29 @@ if plot_results:
         print(surf_y.shape)
         total = time.time() - start
         print("  %.2f seconds."%total)
-        p.add(name, *(surf_x.T), surf_y, use_gradient=use_gradient,
-              plot_type='surface',
-              # mode="markers", marker_size=3, marker_line_width=1,
-              **marker_args,
-        )
-    p.plot()
+        # p.add(name, *(surf_x.T), surf_y, use_gradient=use_gradient,
+        #       plot_type='surface', opacity=0.9,
+        #       # mode="markers", marker_size=3, marker_line_width=1,
+        #       **marker_args,
+        # )
+        surf_y = s.weights(surf_x.copy())
+        for i in range(len(points)):
+            print("Adding region for %i.."%i)
+            color = p2.color(i+1)
+            p2.add("p%i"%i, [points[i,0]], [points[i,1]], color=color,
+                   show_in_legend=False, shade=False, group=str(i))
+            p2.add_region("Support for point %i"%i, 
[…]

### 2017-11-29 HPC_Paper/paper.tex
Finished the HPC paper, ready for review by Dr.Watson. Updated all the figures to have uniform font sizes that should remain legible in print.

@@
 \usepackage{amsthm}
-\usepackage{algorithm}     % needed for the algorithm block
-\usepackage{algpseudocode} % algorithmicx psuedo-code styling
+%% \usepackage{color}   % Needed to color text for drafts
+
+%% \usepackage{algorithm}     % needed for the algorithm block
+%% \usepackage{algpseudocode} % algorithmicx psuedo-code styling
 %\usepackage[pdftex,colorlinks=true,urlcolor=blue,citecolor=black,anchorcolor=black,linkcolor=black]{hyperref}
@@
 functions from the previous iteration, $c, v \in \mathbb{R}$, $w_+ =
-\{w, w \geq 0, 0, w < 0\},$ and $w_+ = (-w)_+$. After iteratively
+\{w, w \geq 0;\quad 0, w < 0\},$ and $w_+ = (-w)_+$. After iteratively
 constructing a model, MARS then iteratively removes basis functions
@@
   \centering
-  \includegraphics[width=\textwidth,trim={0 .5in 0 1in}]{Raw_Throughput.pdf}
+  \includegraphics[width=\textwidth,trim={0 .5in 0 .4in}]{Raw_Throughput.pdf}
   \caption{Histograms of 100-bin reductions of the PMF of I/O
@@
 important to consider that algorithms that perform well with less
-training input also require less experimentation, although the amount
+training input also require less experimentation. Although, the amount
 of training data required may change as a function of the number of
-input dimensions, and this needs to be studied as well. The framework
+input dimensions and this needs to be studied as well. The framework
 used here will be referred to as a multidimensional analysis (MDA) of
 the IOzone data.
+
+%% \color{red}
 
@@
 size to testing size. Given an input data matrix with $n$ rows
-(points) and $d$ columns (features), the multidimensional analysis
-(MDA) is performed as follows:
+(points) and $d$ columns (components), MDA proceeds as follows:
 \begin{enumerate}
-\item for all $k = 1, \ldots, d$ and for all subsets of input features
-  $F = \{ j_1, j_2, \ldots j_k \}$; reduce the input data to unique
-  points $z \in \mathbb{R}^k$ with $f(z) = E[ \{ f(x^{(i)}) \quad
-    \forall i \quad s.t. \quad (x^{(i)}_F = z) \} ]$.
+\item for all $k = 1, \ldots, d$ and for all subsets of input
+  components $F = \{ j_1, j_2, \ldots j_k \}$; reduce the input data
+  to unique points $z \in \mathbb{R}^k$ with $f(z) = E[ \{ f(x^{(i)})
+    \quad \forall i \quad s.t. \quad (x^{(i)}_F = z) \} ]$ where
+  $x^{(i)}_F$ is the $k$-tuple containing only those components of
+  $x^{(i)}$ in $F$.
 \item for all $r$ in $\{5, 10, \ldots, 95\}$; generate $N$ random
@@
   training and $100 - r$ percentage for testing.
-\item when generate each random $(train, test)$ split, ensure that all
-  points from $test$ are on or inside the convex hull of points in
+\item when generating each random $(train, test)$ split, ensure that
+  all points from $test$ are on or inside the convex hull of points in
   $train$; also ensure that the points in $train$ are well spaced.
@@
 points are well spaced, this paper relies on the statistical method
-for picking points from \shortciteN{amos2014algorithm} that procedes
+for picking points from \shortciteN{amos2014algorithm} that proceeds
 as follows:
@@
   $||x^{(i_k)}-x^{(j_k)}||_2 \leq ||x^{(i_{k+1})}-x^{(j_{k+1})}||_2$
-\item Iteratively remove points from candidacy until only $N$ remain
-  by sequentially selecting one point from the pair $(x^{(i_k)},
+\item Sequentially remove points from candidacy until only $N$ remain
+  by randomly selecting one point from the pair $(x^{(i_k)},
   x^{(j_k)})$ for $k = 1,\ldots$ if both $x^{(i_k)}$ and $x^{(j_k)}$
@@
 testing splits may not be created or may not exist. In these cases, if
-there are fewer than $N$ possible splits, those deterministically
+there are fewer than $N$ possible splits then deterministically
 generated splits are used. Otherwise after $3*N$ attempts, only the
@@
 are FORTRAN wrapped with python. All randomness has been seeded for
-reproducability.
+repeatability.
 
-For any input feature subset $F$ (of size $k$) and selected value of
+For any input component subset $F$ (of size $k$) and selected value of
 $r$, MDA will generate up to $N$ predictions that each multivariate
@@
 fewer than $N$ predictions made for any given point. Points in the
-convex hull for the selected subest of features will always be used
+convex hull for the selected subset of components will always be used
 for training, never for testing. Points that do not have any close
 neighbors will often be used for training in order to ensure well
-spacedness. Finally, as mentioned before, some subsets of features do
-not readily generate $N$ unique training and testing splits.
+spacedness. Finally, as mentioned before, some subsets of components
+do not readily generate $N$ unique training and testing splits.
 
@@
 multivariate model could make for a point in $\mathbb{R}^k$, $k \leq
-d$ given feature set $F$. The summary results presented in this work
+d$ given component set $F$. The summary results presented in this work
 use the expected value of the distributions at each point as the model
@@
 
-A naive multivariate prediction technique such as nearest neighbor
-could experience relative errors in the range $[0,10^m]$ where $m$ is
-$log_{10}(\text{max}_f - \text{min}_f)$ when modeling a function $f$
-from data. The IOzone mean data response values span 3 orders of
-magnitude (as can be seen in Table \ref{tab:data_type}) while variance
-data response values span 6 orders of magnitude. It is expected
-therefore, that all studied multivariate models perform better than a
-naive approach, achieving relative errors strictly less than $10^3$
-for mean throughput and $10^6$ when for throughput sample
-variance. Ideally, models will yield relative errors significantly
-smaller than 1.
+A na\"{\i}ve multivariate prediction technique such as nearest
+neighbor could experience relative errors in the range $[0,
+  (\text{max}_f - \text{min}_f) / \text{min}_f ]$ when modeling a non
+negative function $f$ from data. The IOzone mean data response values
+span three orders of magnitude (as can be seen in Table
+\ref{tab:data_type}) while variance data response values span six
+orders of magnitude. It is expected therefore, that all studied
+multivariate models perform better than a na\"{\i}ve approach,
+achieving relative errors strictly less than $10^3$ for throughput
+mean and $10^6$ for throughput sample variance. Ideally, models will
+yield relative errors significantly smaller than 1.
 
@@
 Delaunay, consistently makes predictions with relative error less than
-$.05$ (5\% error). In figure \ref{fig:mean_tt_ratio} it can also be
+$.05$ (5\% error). In Figure \ref{fig:mean_tt_ratio} it can also be
 seen that the Delaunay model consistently makes good predictions even
-with as low as 5\% training data (43 system configurations) regardless
-of the dimension of the data.
+with as low as 5\% training data (43 of the 864 system configurations)
+regardless of the dimension of the data.
 
@@
   \centering
-  \includegraphics[width=\textwidth,trim={0 1in 0 1in}]{Mean_TT_Ratio.pdf}
-  \caption{These box plots show the prediction error of means with
+  \includegraphics[width=\textwidth,trim={0 .5in 0 .5in}]{Mean_TT_Ratio.pdf}
+  \caption{These box plots show the prediction error of mean with
     increasing amounts of training data provided to the models. Notice
-    that the response values being predicted span 3 orders of
-    magnitude and hence relative error in that range would not be
-    unexpected. For SVR the top box whisker goes from around 100 to 50
-    from left to right and is chopped in order to maintain focus on
-    the more performant models.}
+    that MARS is the only model whose average spread of performance
+    decreases with more training data. Recall that the response values
+    being predicted span three orders of magnitude and hence relative
+    errors should certainly remain within that range. For SVR the top
+    box whisker goes from around 100 to 50 from left to right and is
+    truncated in order to maintain focus on more performant models.}
   \label{fig:mean_tt_ratio}
+\end{figure}
+
+\begin{figure}
+  \centering
+  \includegraphics[width=\textwidth,trim={0 .5in 0 .5in}]{Mean_Dim.pdf}
+  \caption{These box plots show the prediction error of mean with
+    increasing dimension. The top box whisker for SVR transitions
+    \{40, 80, 90\} for dimensions 2, 3, and 4 respectively. Notice
+    that each model consistently experiences greater magnitude error
+    with increasing dimension.}
+  \label{fig:mean_dim}
 \end{figure}
@@
 
-The prediction results for variance resemble that of predicting I/O
-throughput mean, with each of the errors being scaled proportional to
-the change in scale of the response values. Delaunay remains the best
-overall predictor with LSHEP closely competing, but yielding much more
-outliers.
+The prediction results for variance resemble those for predicting
+mean. Delaunay remains the best overall predictor with relative error
+of .6 on average and LSHEP closely competes with Delaunay having an
+average relative error of -2. Outliers in prediction error are much
+larger for all models. Delaunay produces relative errors as large as
+78 and other models achieve relative errors around $10^3$. The
+relative errors for many models scaled proportional to the increased
+orders of magnitude spanned by the variance response compared with
+mean response. All models are more sensitive to the amount of training
+data provided than their counterparts for predicting mean.
 
-\subsection{Increasing Dimension}
+\subsection{Increasing Dimension and Decreasing Training Data}
 
-As can be seen in figure \ref{fig:mean_dim}, all of the models suffer
+As can be seen in Figure \ref{fig:mean_dim}, all of the models suffer
 increasing error rates in higher dimension. This is expected, because
@@
 exponentially. However, LSHEP and Delaunay maintain the slowest
-increase in relative error. The particularly slow increase in error
-seen for Delaunay appears to suggest that it is capable of making
-predictions with a range of relative errors approximately spanning
-\rpm $.2*\text{\textit{dim}}$.
-
-\begin{figure}
-  \centering
-  \includegraphics[width=\textwidth,trim={0 1in 0 2in}]{Mean_Dim.pdf}
-  \caption{These box plots show the prediction error of means with
-    increasing dimension. }
-  \label{fig:mean_dim}
-\end{figure}
+increase in relative error. The increase in error seen for Delaunay
+suggests that it is capable of making predictions with a range of
+relative errors that grows approximately linearly with increasing
+dimension input. This trend suggests that Delaunay would remain a
+viable technique for accurately modeling systems with 10's of
+parameters given only small amounts of training data. All models, with
+the exception of MARS, produce smaller errors given more training
+data. Increasing the amount of training data most notably reduces the
+number of prediction error outliers.
 
@@
 
+The present results demonstrate that a straightforward application of
+multivariate modeling techniques can be used to effectively predict
+HPC system performance. Little effort on the part of a systems
+engineer combined with a minimal amount experimentation can yield a
+model capable of accurately tuning an HPC system to the desired
+performance specification.
+
 \subsection{Modeling the System}
-\begin{enumerate}
-\item Promise of the models, especially Delaunay
-\item Possible reasons for the performance differences between models
-\item The need for as few as 50 experiments to get less than 5\% error
-\item The unfortunate inability of Delaunay to scale to higher dimension
-\end{enumerate}
+
+The modeling techniques generated estimates of drastically different
+quality when predicting I/O throughput mean and sample variance. A few
+notable characteristics include: the SVR has the largest range of
+errors for all selections of dimension and amount of training data;
+MARS and LSHEP produce similar magnitude errors while the former
+consistently underestimates and the latter consistently overestimates
+in predictions; and Delaunay has considerably fewer outliers than all
+other methods. SVR likely experiences the greatest quantity of errors
+because the underlying parametric representation is global and
+oversimplified (a single polynomial), making it unable to capture the
+complex local behaviors of system I/O. It is still unclear however
+what causes the behaviors of LSHEP, MARS, and Delaunay. An exploration
+of this topic is left to future work.
+
+While the Delaunay method appears to be the best predictor in the
+presented IOzone case study, it is important to note its significantly
+higher computational complexity with respect to the dimension of the
+input. The implementation of Delaunay used would experience
+unacceptably large training times beyond seven dimensions of
+input. This leaves much room for other techniques to perform best in
+higher dimension unless a more performant implementation of Delaunay
+can be used.
+
+Finally, the ability of the models to predict variance was
+significantly worse for I/O sample variance. The larger scale in
+variance responses alone do not account for the increase in relative
+errors witnessed. This suggests that system variability has an greater
+underlying functional complexity when compared with the system mean
+and that latent factors are reducing prediction performance
+considerably.
 
 \subsection{Extending the Analysis}
-\begin{enumerate}
-\item Measuring different system characteristics other than I/O
-\item Recording system performance constantly to maintain model accuracy
-\end{enumerate}
+
+System I/O throughput mean and throughput sample variance are simple
+and useful system characteristics to model. The process presented in
+this work is equally applicable to predicting other useful performance
+characteristics of HPC systems such as: computational throughput,
+power consumption, processor idle time, context switches, RAM usage,
+or any other ordinal performance metric. For each of these there is
+the potential to model system variability as well. This work has
+chosen variance as a measure of variability, but the techniques used
+in this paper could be used to predict more precise measures of
+variability such as the percentiles of the performance distribution. A
+thorough exploration of HPC systems applications of multivariate
+modeling is left to future work.
+
+%% \begin{enumerate}
+%% \item Measuring different system characteristics other than I/O
+%% \item Recording system performance constantly to maintain model accuracy
+%% \end{enumerate}
 
@@
 \label{sec:conclusion}
-\begin{enumerate}
-\item Delaunay is the best model, capable of predicting the
-  performance of 800 different system configurations accurately after
-  only being trained on a well-spaced set of 50 configurations.
-\item More comparisons of models similar to Delaunay should be done to
-  evaluate what mathematical characteristics of multivariate models are
-  most effective.
-\end{enumerate}
+Multivariate models of HPC systems produce promising results when
+applied to predicting I/O throughput mean and variability. These
+multivariate techniques significantly expand the comprehensiveness and
+portability of statistical models for predicting computer system
+performance over previous works. In the IOzone case study presented,
+the Delaunay method produces the best overall results making
+predictions for 821 system configurations with less than 5\% error
+when trained on only 43 configurations. Analysis also suggests that
+the error in the Delaunay method will remain manageable as the number
+of system parameters being modeled increases. These multivariate
+techniques can and should be applied to HPC systems with more than
+four tunable parameters in order to identify optimal system
+configurations that may not be discoverable via previous methods nor
+by manual performance tuning.
 
@@
 \subsection{Future Work}
-The most severe limitation to the presented work is the present
-inability to model the relationship between categorical system
-parameters. This could be seen as a limitation of the models selected,
-because only MARS is capable of handling non-numeric variables, or as
-a limitation of the MDA framework. Future works could attempt to
-identify the viability of reducing categorical variables to numerical
-ones to expand set of predictable system configurations.
+The most severe limitation to the present work is the restriction to
+modeling strictly ordinal (not categorical) system parameters. This
+could be seen as a limitation of the multivariate models selected,
+because only MARS is capable of handling non numeric components, or as
+a limitation of the MDA framework. Future work could attempt to
+identify the viability of different techniques for making predictions
+across varying categorical system paramters.
 
-\begin{enumerate}
-\item Using models to predict more complex responses, such as functions
-\end{enumerate}
+There remain many other multivariate modeling techniques not included
+in this work that should be evaluated and applied to HPC performance
+tuning. For I/O alone, there are far more than the four tunable
+parameters studied in this work. Alongside experimentation with more
+models, there is room for a theoretical characterisation of the
+combined model and data properties that allow for the greatest
+predictive power.
+
+%% \color{black}
 
 %% \newpage
-%% \_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\\
+%% \ \\
 
@@
 %%   \textbf{W}eighted \textbf{A}pproxima\textbf{T}ion\textbf{S}
-%%   \textbf{O}f fu\textbf{N}tions}
-%% %% L  A  Y   NE   W        A        T   S O    N
+%%   \textbf{O}f fu\textbf{N}tions}\\
+%% %% \texttt{L\ \ A\ \ Y \ \ NE\ \ \  W\ \ \ \ \ \ \ \ A\ \ \ \ \ \ \ \ T\ \ \ S O\ \ \ \ N\ \ \ \ \ \ }
 %% \newpage
@@
 \end{document}
-
-%%  ALGORITHM     MEAN ABSOLUTE ERROR    MEAN RELATIVE ERROR
-
-%% 1 Dimension                                        
-%%  Delaunay         2060267             0.03 
-%%  MARS             2176368             0.03 
-%%  MLPRegressor     2461097             0.06 
-%%  LSHEP            6353148             0.06 
-%%  SVR             19306731             0.96
-
-%% 2 Dimensions                                       
-%%  Delaunay         2916224             0.07 
-%%  MARS             6213369             0.61 
-%%  MLPRegressor     8883580             0.81 
-%%  LSHEP           13038882             1.52 
-%%  SVR             78213963             18.49
-
-%% 3 Dimensions                                       
-%%  Delaunay         3001794             0.09 
-%%  MARS             4024964             0.65 
-%%  LSHEP            5446207             1.13 
-%%  MLPRegressor    11190977             2.03 
-%%  SVR             97761266             31.27
-
-%% 4 Dimensions                                       
-%%  Delaunay         3134779             0.10 
-%%  MARS             4497414             1.01 
-%%  LSHEP            5133648             1.07 
-%%  MLPRegressor    11831721             2.20 
-%%  SVR            113551300             37.26
-
-%% [5 - 21%] Training Data               
-%%  Delaunay         5274933             0.11 
-%%  MARS             6543153             0.61 
-%%  LSHEP           13218469             2.53 
-%%  MLPRegressor    17348982             2.58 
-%%  SVR            170906944             49.45
-
-%% [25 - 46%] Training Data              
-%%  Delaunay         2898330             0.08 
-%%  MARS             4324916             0.80 
-%%  LSHEP            4936381             0.85 
-%%  MLPRegressor    12504957             2.29 
-%%  SVR             99851846             31.28
-
-%% [50 - 74%] Training Data              
-%%  Delaunay         1518477             0.07 
-%%  LSHEP            2251409             0.26 
-%%  MARS             3115943             0.95 
-%%  MLPRegressor     5292177             1.17 
-%%  SVR             51565405             19.63
-
-%% [75 - 96%] Training Data              
-%%  Delaunay          886346             0.08 
-%%  LSHEP            1415034             0.15 
-%%  MARS             2591451             0.74 
-%%  MLPRegressor     2996253             1.02 
-%%  SVR             27125130             12.64
-
-%% ALGORITHM   AVG ABSOLUTE MEAN ERROR   AVG RELATIVE ERROR
-%% Delaunay         2711394              0.08  
-%% MARS             4185947              0.72  
-%% LSHEP            6474147              0.94  
-%% MLPRegressor     9063718              1.48  
-%% SVR             82285323              25.12

### 2017-12-02 Paper_[2017-11]_HPC/paper.tex
I have now properly added all my relevant VarSys work to this repo under a common format. This will make my work as transferrable to another person as it can possibly be. I chunked large files so that they all fit into the repo.

--- 2017-11-29 HPC_Paper/paper.tex
@@
     (bottom). In the mean plot, the first 1\% bin (truncated in plot)
-    has a probability mass of .45. In the variance plot, the first 1\%
-    bin has a probability mass of .58. It can be seen that the
+    has a probability mass of .45. In the variance plot, the second
+    1\% bin has a probability mass of .58. It can be seen that the
     distributions of throughputs are primarily of lower magnitude with
@@
 training input also require less experimentation. Although, the amount
-of training data required may change as a function of the number of
-input dimensions and this needs to be studied as well. The framework
-used here will be referred to as a multidimensional analysis (MDA) of
-the IOzone data.
+of training data required may change as a function of the dimension of
+the input and this needs to be studied as well. The framework used
+here will be referred to as a multidimensional analysis (MDA) of the
+IOzone data.
 
@@
   training and $100 - r$ percentage for testing.
-\item when generating each random $(train, test)$ split, ensure that
-  all points from $test$ are on or inside the convex hull of points in
-  $train$; also ensure that the points in $train$ are well spaced.
+\item when generating each of $N$ random $(train, test)$ splits,
+  ensure that all points from $test$ are on or inside the convex hull
+  of points in $train$; also ensure that the points in $train$ are
+  well spaced.
 \end{enumerate}
@@
   $||x^{(i_k)}-x^{(j_k)}||_2 \leq ||x^{(i_{k+1})}-x^{(j_{k+1})}||_2$
-\item Sequentially remove points from candidacy until only $N$ remain
-  by randomly selecting one point from the pair $(x^{(i_k)},
+\item Sequentially remove points from candidacy until only $|train|$
+  remain by randomly selecting one point from the pair $(x^{(i_k)},
   x^{(j_k)})$ for $k = 1,\ldots$ if both $x^{(i_k)}$ and $x^{(j_k)}$
@@
 in predictions; and Delaunay has considerably fewer outliers than all
-other methods. SVR likely experiences the greatest quantity of errors
+other methods. SVR likely produces the poorest quality predictions
 because the underlying parametric representation is global and
@@
 variance responses alone do not account for the increase in relative
-errors witnessed. This suggests that system variability has an greater
+errors witnessed. This suggests that system variability has a greater
 underlying functional complexity when compared with the system mean
-and that latent factors are reducing prediction performance
-considerably.
+and that latent factors are reducing prediction performance.
 
@@
 in this work that should be evaluated and applied to HPC performance
-tuning. For I/O alone, there are far more than the four tunable
+prediction. For I/O alone, there are far more than the four tunable
 parameters studied in this work. Alongside experimentation with more

### 2017-12-02 Report_[2017-03]_Convergence/First_100_Trial_Report/Report.tex
I have now properly added all my relevant VarSys work to this repo under a common format. This will make my work as transferrable to another person as it can possibly be. I chunked large files so that they all fit into the repo.

\documentclass[a4paper,landscape]{article}

%% Required to make the text fill the page better
\usepackage[textwidth=20cm]{geometry}

% Required for specifying captions to tables and figures
\usepackage[font=small,labelfont=bf]{caption}

%% Required to display some mathematical formula
\usepackage{amsmath,amssymb}

%% So that I can draw the graphs for the confidence in difference
\usepackage{tkz-graph}
\renewcommand*{\EdgeLineWidth}{0.15pt}

%% So that I can display the plots
\usepackage{graphicx}

%% Sets the font family for the document to a sans-serif font
\renewcommand{\familydefault}{\sfdefault}

%% pdflatex hw2_tchlux.tex && evince hw2_tchlux.pdf

\begin{document}

\subsection*{\hfil Comparing Throughput CDF's of 3 System Configurations \hfil}
\begin{center}
  March $22^{nd}$, 2017\\
  Thomas Lux
\end{center}

\subsection*{Summary}

In this report, we compare the Cumulative Distribution Functions
(CDF's) of the Throughput of three different system
configurations. All three configurations have the same settings for:
\begin{align*}
  Frequency  &-  2700000\\
  Hyp\ Sched &-  CFQ\\
  VM\ Sched  &-  NOOP\\
  Threads    &-  64\\
  Mode       &-  Fread
\end{align*}

While the file and record sizes for the three system settings were:
\begin{align}
  File\ Size - 64  \ \  &\ \ Record\ Size - 32\\
  File\ Size - 1024\ \  &\ \ Record\ Size - 32\\
  File\ Size - 1024\ \  &\ \ Record\ Size - 512
\end{align}

Most notably, from this in-depth comparison, it appears that our
independently drawn samples from the systems are not always conforming
to the same underlying distribution. If this trend continues, then it
will prohibit the effective modelling goals of the VarSys project.

\subsection*{Source Data}

For each of these system configurations we have 3 independently
gathered sets of data:
\begin{itemize}
\item The \textit{First} \textbf{40} runs collected for the VarSys data mid-last
  year.
\item The \textit{Second} set of \textbf{40} runs collected December of last year.
\item The newest, \textit{True} set of \textbf{420} runs collected two weeks ago.
\end{itemize}

\pagebreak
\subsection*{Figures and Plots}

For each of the three system configurations, there is 1 figure
followed by 3 plots:
\begin{enumerate}
\item The \textbf{Paired KS Solved for Alpha Graph} shows the trials
  \textit{first}, \textit{second}, and \textit{true} as nodes with the
  edges between them being the result of solving the KS Two Sample
  test for alpha (see explanation in item 4 below). This will be
  followed by the 'Confidence' interpretation below.

\item The \textbf{CDF Comparison Plot} shows the Throughput CDF's of
  each of the \textit{first}, \textit{second}, and \textit{true} data
  sets. Each CDF plot also has a red-tinted cloud~\footnote{The clouds
    were generated by taking 10,000 random 40 sample subsets of the
    (\textit{true}) 420, calculating the CDF's of each subset, then
    calculating the percentiles of possible \textit{true} CDF values
    based on the sapmle CDF's.} of the possible positions that the
  \textit{true} CDF could take on when collecting 40 sample subsets.

\item The \textbf{Kolmogorov-Smirnov (KS) Difference Plot} shows the
  decrease in KS difference~\footnote{The two-sample KS Difference is
    simply the maximum difference between the two CDF's of the
    samples. From this difference, a test is performed to determine if
    the two distributions are different.}  between a subsapmle and the
  \textit{true} CDF with increasing subsample size (x-axis).

\item The \textbf{Confidence in KS Difference Plot} shows what happens
  when we take the typical KS test for difference:
  \begin{align*}
    \text{KS Diff} &> c(\alpha) \sqrt{\frac{n_1 + n_2}{n_1 \cdot n_2}}\\
    \text{where}\ c(\alpha) &= \sqrt{-\frac{1}{2}ln\bigg(\frac{\alpha}{2}\bigg)}
  \end{align*}
  and use our knowledge of $n_1$, $n_2$ and the KS Difference to solve
  for the minimum alpha such that this inequality holds. Giving us:
  \begin{align*}
    \alpha\ \  >\ \  2 \cdot e^{-2 \cdot \bigg(\frac{\text{KS Diff}}{\sqrt{\frac{n_1 + n_2}{n_1 \cdot n_2}}}\bigg)^2}
  \end{align*}
  This final plot shows $100 \cdot (1 - \alpha)$, or the approximate
  confidence, in percentage, that the two distributions being compared
  are different. This data was collected by comparing the CDF's of 100
  random subsamples of size $\textless$x-axis$\textgreater$ to the
  \textit{true} CDF. The corresponding y-values are displayed in shaded
  regions by percentile.Ideally we want this quantity to go to zero.

\end{enumerate}

%% \textbf{Commentary:}\\
%% Notice that something.

\pagebreak
\vspace*{0.5cm}
\subsection*{\hfil System Setting \#1: File Size 64 and Record Size 32 \hfil}
\vspace*{1.0cm}

\begin{center}
  Paired KS Solved for Alpha: \\
  \vspace{0.5cm}

  \begin{tikzpicture}
    \GraphInit[vstyle=Empty]
    %% Draw the nodes
    \node (A) at (5,3) {$First$};
    \node (B) at (3,0) {$Second$};
    \node (C) at (7,0) {$True$};
    %% Draw the alpha values
    \node (AB) at (3,1.5) {0.19366};
    \node (AC) at (7,1.5) {0.00002};
    \node (BC) at (5,-.5) {0.01904};
    %% Draw the edges
    \begin{scope}[every path/.style={->}]
      \draw (A) -- (B);
      \draw (B) -- (A);
      \draw (A) -- (C);
      \draw (C) -- (A);
      \draw (B) -- (C);
      \draw (C) -- (B);
    \end{scope}
  \end{tikzpicture}

  \vspace{1cm}
  Translated to confidence in paired difference: \\
  \vspace{0.5cm}

  \begin{tikzpicture}
    \GraphInit[vstyle=Empty]
    %% Draw the nodes
    \node (A) at (5,3) {$First$};
    \node (B) at (3,0) {$Second$};
    \node (C) at (7,0) {$True$};
    %% Draw the alpha values
    \node (AB) at (3,1.5) {80.633\%};
    \node (AC) at (7,1.5) {99.998\%};
    \node (BC) at (5,-.5) {98.096\%};
    %% Draw the edges
    \begin{scope}[every path/.style={->}]
      \draw (A) -- (B);
      \draw (B) -- (A);
      \draw (A) -- (C);
      \draw (C) -- (A);
      \draw (B) -- (C);
      \draw (C) -- (B);
    \end{scope}
  \end{tikzpicture}
\end{center}

\vspace*{-3.5cm}
\hspace*{-3.5cm}
\includegraphics[page=1,width=27cm]{Raw-and-Fit-Report-Convergence-Over-100-Trials}
\captionof{figure}{In this CDF comparison, the \textit{first} CDF
  exceeds the reasonable bounds of \textit{true} the most, followed
  closely by \textit{second}.}

\vspace*{-3.5cm}
\hspace*{-3.5cm}
\includegraphics[page=2,width=27cm]{Raw-and-Fit-Report-Convergence-Over-100-Trials}
\captionof{figure}{In this KS Difference Plot, notice that most of the
  convergence happens with approximately 30 samples from the true
  distribution. \\This trend is common across all three system settings.}

\vspace*{-3.5cm}
\hspace*{-3.5cm}
\includegraphics[page=3,width=27cm]{Raw-and-Fit-Report-Convergence-Over-100-Trials}
\captionof{figure}{In this Confidence in KS Difference plot notice
  that we do not guarantee large improvements in confidence even with
  hundreds of samples from the distribution. This may seem to
  contradict the decreasing KS difference as seen in \textit{figure 2},
  but the slower decrease in confidence is due to the fact that the
  KS-difference test becomes more strict with increasingnly large
  sample sizes.}

%================================
%     Begin System Setting 2     
%================================

\pagebreak
\vspace*{0.5cm}
\subsection*{\hfil System Setting \#2: File Size 1024 and Record Size 32 \hfil}
\vspace*{1cm}

\begin{center}
  Paired KS Solved for Alpha:\\
  \vspace{0.5cm}

  \begin{tikzpicture}
    \GraphInit[vstyle=Empty]
    %% Draw the nodes
    \node (A) at (5,3) {$First$};
    \node (B) at (3,0) {$Second$};
    \node (C) at (7,0) {$True$};
    %% Draw the alpha values
    \node (AB) at (3,1.5) {0.00791};
    \node (AC) at (7,1.5) {0.09033};
    \node (BC) at (5,-.5) {0.00009};
    %% Draw the edges
    \begin{scope}[every path/.style={->}]
      \draw (A) -- (B);
      \draw (B) -- (A);
      \draw (A) -- (C);
      \draw (C) -- (A);
      \draw (B) -- (C);
      \draw (C) -- (B);
    \end{scope}
  \end{tikzpicture}

  \vspace{1cm}
  Translated to confidence in paired difference: \\
  \vspace{0.5cm}

  \begin{tikzpicture}
    \GraphInit[vstyle=Empty]
    %% Draw the nodes
    \node (A) at (5,3) {$First$};
    \node (B) at (3,0) {$Second$};
    \node (C) at (7,0) {$True$};
    %% Draw the alpha values
    \node (AB) at (3,1.5) {99.209\%};
    \node (AC) at (7,1.5) {90.967\%};
    \node (BC) at (5,-.5) {99.991\%};
    %% Draw the edges
    \begin{scope}[every path/.style={->}]
      \draw (A) -- (B);
      \draw (B) -- (A);
      \draw (A) -- (C);
      \draw (C) -- (A);
      \draw (B) -- (C);
      \draw (C) -- (B);
    \end{scope}
  \end{tikzpicture}
\end{center}

\vspace*{-3.5cm}
\hspace*{-3.5cm}
\includegraphics[page=4,width=27cm]{Raw-and-Fit-Report-Convergence-Over-100-Trials}
\captionof{figure}{In this CDF comparison, the \textit{second} CDF is
  most notably an outlier. This is reflected in the solved alpha graph
  above.}

\vspace*{-3.5cm}
\hspace*{-3.5cm}
\includegraphics[page=5,width=27cm]{Raw-and-Fit-Report-Convergence-Over-100-Trials}
\captionof{figure}{This KS Difference Plot follows similar trends to
  the previous.}

\vspace*{-3.5cm}
\hspace*{-3.5cm}
\includegraphics[page=6,width=27cm]{Raw-and-Fit-Report-Convergence-Over-100-Trials}
\captionof{figure}{This Confidence in KS Difference plot follows
  similar trends to the previous.}

%================================
%     Begin System Setting 3     
%================================

\pagebreak
\vspace*{0.5cm}
\subsection*{\hfil System Setting \#3: File Size 1024 and Record Size 512 \hfil}
\vspace*{1cm}

\begin{center}
  Paired KS Solved for Alpha:\\
  \vspace{0.5cm}

  \begin{tikzpicture}
    \GraphInit[vstyle=Empty]
    %% Draw the nodes
    \node (A) at (5,3) {$First$};
    \node (B) at (3,0) {$Second$};
    \node (C) at (7,0) {$True$};
    %% Draw the alpha values
    \node (AB) at (3,1.5) {0.01341};
    \node (AC) at (7,1.5) {1.32547};
    \node (BC) at (5,-.5) {0.00190};
    %% Draw the edges
    \begin{scope}[every path/.style={->}]
      \draw (A) -- (B);
      \draw (B) -- (A);
      \draw (A) -- (C);
      \draw (C) -- (A);
      \draw (B) -- (C);
      \draw (C) -- (B);
    \end{scope}
  \end{tikzpicture}

  \vspace{1cm}
  Translated to confidence in paired difference: \\
  \vspace{0.5cm}

  \begin{tikzpicture}
    \GraphInit[vstyle=Empty]
    %% Draw the nodes
    \node (A) at (5,3) {$First$};
    \node (B) at (3,0) {$Second$};
    \node (C) at (7,0) {$True$};
    %% Draw the alpha values
    \node (AB) at (3,1.5) {98.659\%};
    \node (AC) at (7,1.5) {\textless\ 1.0\%};
    \node (BC) at (5,-.5) {99.810\%};
    %% Draw the edges
    \begin{scope}[every path/.style={->}]
      \draw (A) -- (B);
      \draw (B) -- (A);
      \draw (A) -- (C);
      \draw (C) -- (A);
      \draw (B) -- (C);
      \draw (C) -- (B);
    \end{scope}
  \end{tikzpicture}
\end{center}

\vspace*{-3.5cm}
\hspace*{-3.5cm}
\includegraphics[page=7,width=27cm]{Raw-and-Fit-Report-Convergence-Over-100-Trials}
\captionof{figure}{In this CDF comparison, the \textit{second} CDF is
  once again the extreme outlier.}

\vspace*{-3.5cm}
\hspace*{-3.5cm}
\includegraphics[page=8,width=27cm]{Raw-and-Fit-Report-Convergence-Over-100-Trials}
\captionof{figure}{This KS Difference Plot follows similar trends to
  the previous 2.}

\vspace*{-3.5cm}
\hspace*{-3.5cm}
\includegraphics[page=9,width=27cm]{Raw-and-Fit-Report-Convergence-Over-100-Trials}
\captionof{figure}{This Confidence in KS Difference plot follows
  similar trends to the previous 2.}

\end{document}

%% pdflatex Report.tex && evince Report.pdf

### 2017-12-02 Report_[2017-05-29]_Box_Spline_Mesh/BS_Report.tex
I have now properly added all my relevant VarSys work to this repo under a common format. This will make my work as transferrable to another person as it can possibly be. I chunked large files so that they all fit into the repo.

%% pdflatex BS_Report.tex && open BS_Report.pdf

\documentclass[11pt,margin=1in]{article}
\usepackage[margin=1in]{geometry}         % Document setup
\usepackage{graphicx}                     % Including pictures
\usepackage{amsmath}                      % Math mode
\usepackage{tikz}                         % For drawing example boxes
\usepackage{algorithm2e}                  % For including box mesh algorithm
\usetikzlibrary{shapes}                   % For the 'circle' shape
\renewcommand{\familydefault}{\sfdefault} % Setting the font

\begin{document}
\begin{center}
  \textbf{Report on Box Spline Interpolation and Approximation}\\
  May $\text{29}^\text{th}$, 2017\\
  Thomas Lux
\end{center}

\begin{table}[!htbp]
\centering
\begin{tabular}{cc}

  Linear Box Mesh Interpolation & Quadratic Box Mesh Interpolation \\
  \vspace{1mm} & \\

  \includegraphics[width=7cm]{Linear_Box_Mesh} &
  \includegraphics[width=7cm]{Quadratic_Box_Mesh}

\end{tabular}
\end{table}

\vspace{-10mm}

\section{Introduction}

This documents contains a brief explanation and analysis of the
\textit{Box Mesh} algorithm for generating an interpolating and/or
approximating surface to data given a desired error tolerance and
smoothness. The box mesh algorithm attempts to generate the smallest
set of overlapping boxes with the largest minimum side length (least
skinny) such that each box contains only one data point and the error
at all data points is below a given tolerance. After constructing this
set of boxes, we can use a variety of different functions to generate
the weight of a control point within each individual box. Taking a
weighted sum of control point values using these box functions
generates the surfaces seen above.

First we discuss the specification of each individual box and analyze
the use of linear and quadratic Box Splines as the box functions.
Then, we outline the box mesh construction algorithm, providing a
visual example. Following the example we discuss some mechanisms for
smoothing the surface. Finally, we conclude with a runtime analysis of
the box mesh algorithm.

\section{Box Specification}

A box mesh is composed of individual boxes. Each n-dimensional box has
a \textit{center} (c), \textit{lower width} (l), and \textit{upper
  width} (u). Given we are working in $d$ dimensions, a point $p$ is
inside of a box $b$ with center $c$, lower width $l$, and upper width
$u$ if $(\forall$ $i$ $|$ $0 \leq i < d)(c_i-l_i < p_i < c_i+u_i)$.
See the following visualized example of a box in 2 dimensions.

\vspace{3mm}

\begin{center}
\begin{tikzpicture}[scale=1]

  \draw[thick] (3,1.1) circle (0.7mm);

  %% Draw arrows for each width
  \draw[thick,->] (3,1.2) -- (3,2.4);
  \draw[thick,->] (3,1.0) -- (3,0.1);
  \draw[thick,->] (3.1,1.1) -- (9.9,1.1);
  \draw[thick,->] (2.9,1.1) -- (0.1,1.1);

  %% Draw boundary of box
  \draw[thick,-] (0,0) -- (10,0);
  \draw[thick,-] (10,0) -- (10,2.5);
  \draw[thick,-] (10,2.5) -- (0,2.5);
  \draw[thick,-] (0,2.5) -- (0,0);

  %% Add text to the picture
  \node at (3.25,1.35) {$c$};
  \node at (1.3,1.35) {$l_0$};
  \node at (6.3,1.35) {$u_0$};
  \node at (2.75,0.55) {$l_1$};
  \node at (2.75,1.8) {$u_1$};

\end{tikzpicture}
\end{center}

\subsubsection*{Box Functions}

Allowing for the conditional scaling of points, we can assume that
every box is a unit cube with center $(\frac{1}{2},\frac{1}{2})$. All
of our box functions then only need to be defined on the unit cube in
$d$ dimensions. The two functions that we utilize are: the linear box
spline defined by the direction vector set $[ I I ]$ where $I$ is the
d-dimensional identity matrix; and the quadratic box spline defined by
the direction vector set $[ I I I ]$. The convenience of restricting
ourselves to box splines of this form is that the box spline value at
any $x$ is simply the multaplicative sum $\Pi_{i=0}^{d-1} f(x_i)$
where $f$ is the piecewise polynomial that defines the associated
linear or quadratic B-Spline in one dimension.

We chose these two box functions because the linear box spline is
$c_0$ and the quadratic box spline is $c_1$. When normalizing any
arbitrary box into the unit cube with center
$(\frac{1}{2},\frac{1}{2})$ it is necessary that values be scaled
differently depending on which side of the center they are on. It is
clear that scaling the $x$ has no effect on the function value (we
maintain $c_0$), and since the first derivative of any box spline at
the center coordinate is 0 we do not break $c_1$ continuity with this
non-uniform scaling. But, the continuity of any derivative that is
non-zero at the center would lost with higher order box splines.

The output of the box functions will be used as the weight for the
control point value at the center of each box. Thus, the response
value associated with a given $x$ is computed by taking:

$$ \frac{\sum_{i=0}^{\mathbf{b}-1} \mathbf{b}_i(x) \cdot \text{boxval}_i}{\sum_{i=0}^{\mathbf{b}} \mathbf{b}_i(x) } $$

where $\mathbf{b}$ is the set of boxes that contain $x$,
$\mathbf{b}_i(x)$ is the value of the $i^{\text{th}}$ box function
evaluated at $x$, and $\text{boxval}_i$ is the response value
associated with the $i^{\text{th}}$ box.

\begin{table}[!htbp]
\centering
\begin{tabular}{cc}
  Linear Box Splines in 1 and 2 Dimensions & Quadratic Box Splines in 1 and 2 Dimensions \\

  \vspace{1mm} & \\

  \includegraphics[width=7cm]{Linear_1D} & \includegraphics[width=7cm]{Quadratic_1D} \\
  \includegraphics[width=7cm]{Linear_2D} & \includegraphics[width=7cm]{Quadratic_2D}

\end{tabular}
\end{table}

\section{Box Mesh Construction}

Given a data set with $N$ points in $D$ dimensions, we want to
construct a set of $B$ boxes centered about $B$ points from the data,
referred to as control points. In order to create good response
approximations, we want to ensure that the boxes are both reasonably
shaped (not extremely skinny) and maximally overlapping. The box mesh
construction algorithm creates a set of boxes that each meet these two
criteria as well as contain strictly \textit{one} control point. This
final extra condition allows for strict interpolation.

The algorithm for the box mesh construction proceeds as follows:

\vspace{3mm}

\begin{algorithm}[H]
 \KwData{N by D+1 matrix of data where the last dimension is the response.}
 \KwResult{B-box mesh}
 Initialize one box with upper and lower widths of infinity at median response point\;
 \While{max error $>$ tolerance}{
  Identify point $p$ with max error\;
  Identify boxes $\mathbf{b}$ that contain $p$\;
  Initialize box \textit{new} with center $p$ that contains all boxes in $\mathbf{b}$\;
  \For{$b$ in $\mathbf{b}$}{
    split dimension := $\text{max}_i$ $|\text{b-center}_{i}-\text{\textit{new}-center}_{i}|$\;
    Adjust the widths of $b$ and \textit{new} in \textit{split dimension}\\
    such that $b$ and \textit{new} do not contain each other's center\;
  }
 }
 \caption{Box Mesh Construction}
\end{algorithm}

\vspace{10mm}

\subsubsection*{Visual Example of Box Mesh Construction}

The following 3 examples are all for x with dimension 2. The initial
boxes seen on the left are the $\mathbf{b}$. These boxes contain the
new point that will be added. The premise is that the new point has
already been identified as the one with the most error in
response. The resulting boxes seen on the right demonstrate the
reshaping of the new and old boxes.

\begin{table}[!htbp]
  \centering
  \begin{tabular}{cc}
    \includegraphics[width=7.5cm]{1-Before} & \includegraphics[width=7.5cm]{1-After} \\
  \end{tabular}
  \caption{Adding the blue point, reshaping the x-dimension of the orange box.}
\end{table}

\begin{table}[!htbp]
  \centering
  \begin{tabular}{cc}
    \includegraphics[width=7.5cm]{2-Before} & \includegraphics[width=7.5cm]{2-After} \\
  \end{tabular}
  \caption{Adding the orange point, reshaping the y-dimension of the
    blue and green boxes.}
\end{table}

\begin{table}[!htbp]
  \centering
  \begin{tabular}{cc}
    \includegraphics[width=7.5cm]{3-Before} & \includegraphics[width=7.5cm]{3-After} \\
  \end{tabular}
  \caption{Adding the brown point, reshaping the x dimension of the
    blue box and the y dimension of the orange box.}
\end{table}

\section{Smoothing the Box Mesh Surface}

Given our box mesh as it is computed, we have created a strictly
interpolating surface. It might be of interest to create a surface
that approximates instead of interpolating in order to create greater
smoothness and hence robustness to noise in the data.

Initially, we propose the use of a scaling factor $s$ over the
existing widths of each box as a smoothing parameter. An $s$ equal to
$1.0$ produces the interpolatory surfaces mentioned before. As the
smoothing paramater grows, neighboring boxes begin to influence each
other more and thus the response estimates tend towards average values
of larger neighborhoods of points.

Take note of what happens as we increase the smoothing paramater of
our initial 3D surface plots.

\begin{table}[!htbp]
  \centering
  \begin{tabular}{ccc}
    \includegraphics[width=5cm]{Linear_Box_Mesh} & \includegraphics[width=5cm]{Lin-1-1} & \includegraphics[width=5cm]{Lin-1-2} \\
  \end{tabular}
\end{table}

\begin{table}[!htbp]
  \centering
  \begin{tabular}{ccc}
    \includegraphics[width=5cm]{Lin-1-3} & \includegraphics[width=5cm]{Lin-1-4} & \includegraphics[width=5cm]{Lin-1-5} \\
  \end{tabular}
  \caption{Linear box mesh increasing smoothness from 1.0 to 1.5 by
    increments of 0.1}
\end{table}

\begin{table}[!htbp]
  \centering
  \begin{tabular}{ccc}
    \includegraphics[width=5cm]{Quadratic_Box_Mesh} & \includegraphics[width=5cm]{Quad-1-1} & \includegraphics[width=5cm]{Quad-1-2} \\
  \end{tabular}
\end{table}

\begin{table}[!htbp]
  \centering
  \begin{tabular}{ccc}
    \includegraphics[width=5cm]{Quad-1-3} & \includegraphics[width=5cm]{Quad-1-4} & \includegraphics[width=5cm]{Quad-1-5} \\
  \end{tabular}
  \caption{Quadratic box mesh increasing smoothness from 1.0 to 1.5 by
    increments of 0.1}
\end{table}

\section{Runtime Analysis}

The complexity of generating a box mesh is $O(N^2 + N D^2)$ with
respect to the number of points N and the dimension of the data
D. This is a worst case bound and further analysis needs to be done
in order to determine (a) tighter and more accurate / insightful bound(s).

In the following analysis, B is the current number of boxes. Note that
at each iteration, the number of boxes increases as more points from
the data are designated control points.

\begin{table}[!htbp]
  \centering
  \caption{Runtime Analysis of Box Mesh construction}
  \label{table-runtime}
  \begin{tabular}{|l|c|}
    \hline
    Initialize BM with single point                       & --   \\
    \begin{tabular}{|l|c|}
      \hline      
      Check model error (at modified boxes)               & (N-B) x 2D x D \\
      Add new point $p$ to the Box Mesh                   &  \\
      1. Identify boxes $b$ that contain $p$ (at most 2D) & B x D \\
      2. Get single bounding box for all boxes in $b$     & 2D x D \\
      3. Split largest dimension between box and boxes    & 2D x D\\
      \hline
    \end{tabular}                                         & N    \\
    \hline
  \end{tabular}
\end{table}

\section{Summary}

Overall, the benefits of using the box mesh algorithm are its low
computational complexity and flexibility in terms of interpolating and
smoothing. Further testing needs to be done in order to better
quantify the relative performance of the box mesh versus other
interpolation and approximation techniques.

\vspace{10mm}

\textbf{Potential Alternative Names:}
\begin{itemize}
\item[--] Incremental Box Spline Interpolation and Approximation
\item[--] Max Box Interpolation and Approximation with Box Splines
\item[--] Adaptive Box Partitioning for Interpolation and Approximation
\item[--] Adaptive Box Spline Regression
\item[--] Box Mesh Regression Interpolation and Approximation
\item[--] Adaptive Box Mesh
\item[--] Minimal Box Mesh
\item[--] Full Box Mesh
\item[--] Dynamic Box Mesh
\item[--] Constructive Box Mesh
\item[--] Incremental Box Mesh
\end{itemize}

\end{document}

%% \begin{center}
%%   $A = \begin{bmatrix} 1 & 0 \\ 0 & 1 \end{bmatrix}$\\
%%   \includegraphics[width=7cm]{Box-2}
%%   \includegraphics[width=12cm]{T-Splines}
%% \end{center}

### 2017-12-02 Report_[2017-06-05]_Box_Mesh_Summary/BS_Report.tex
I have now properly added all my relevant VarSys work to this repo under a common format. This will make my work as transferrable to another person as it can possibly be. I chunked large files so that they all fit into the repo.

--- 2017-12-02 Report_[2017-05-29]_Box_Spline_Mesh/BS_Report.tex
@@
 \usepackage{amsmath}                      % Math mode
+\usepackage{amssymb}                      % Math mode symbols
 \usepackage{tikz}                         % For drawing example boxes
@@
 \begin{document}
+
 \begin{center}
-  \textbf{Report on Box Spline Interpolation and Approximation}\\
-  May $\text{29}^\text{th}$, 2017\\
+  \textbf{Box Spline Mesh Algorithm Summary}\\
+  June $\text{5}^\text{th}$, 2017\\
   Thomas Lux
 \end{center}
+
+Given data $N = \{x_1, x_2, \dots, x_n\}\ |\  x_i \in \mathbb{R}^d$, and
+associated response values $\{y_1, y_2, \dots, y_n \}$.
+
+\section{Construct Box Mesh}
+Each box in the box mesh will be specified by a center $c \in \mathbb{R}^d$,
+lower width $l \in \mathbb{R}^d$, and upper width $u \in \mathbb{R}^d$. 
+A point $x_i$ is contained by a box if $(1 \leq j \leq d) (c_j - l_j < x_{i,j} < c_j + u_j)$.
+A two dimensional visual example of the containment region of a box, as we refer to it:
+
+\vspace{1mm}
+
+\begin{center}
+\begin{tikzpicture}[scale=1]
+
+  \draw[thick] (3,1.1) circle (0.7mm);
+
+  %% Draw arrows for each width
+  \draw[thick,->] (3,1.2) -- (3,2.4);
+  \draw[thick,->] (3,1.0) -- (3,0.1);
+  \draw[thick,->] (3.1,1.1) -- (9.9,1.1);
+  \draw[thick,->] (2.9,1.1) -- (0.1,1.1);
+
+  %% Draw boundary of box
+  \draw[thick,-] (0,0) -- (10,0);
+  \draw[thick,-] (10,0) -- (10,2.5);
+  \draw[thick,-] (10,2.5) -- (0,2.5);
+  \draw[thick,-] (0,2.5) -- (0,0);
+
+  %% Add text to the picture
+  \node at (3.25,1.35) {$c$};
+  \node at (1.3,1.35) {$l_0$};
+  \node at (6.3,1.35) {$u_0$};
+  \node at (2.75,0.55) {$l_1$};
+  \node at (2.75,1.8) {$u_1$};
+
+\end{tikzpicture}
+\end{center}
+
+The only restriction is that all widths must be greater than
+zero. This means that $\text{min}_i\ l_i > 0 $ and $\text{min}_i\ u_i
+> 0$. First, order the data in $N$ appropriately (discussed later),
+initialize a single box with $c = x_1$ and lower and upper widths of
+infinity.  Now, proceed to add the rest of the points in $N$ by:
+
+\vspace{3mm}
+
+\begin{algorithm}[H]
+ \For{$i = 2, \cdots, n$}{
+  Identify boxes $\mathbf{B}$ that contain $x_i$\;
+  Initialize box \textit{m} with center $x_i$ that contains all boxes in $\mathbf{B}$\;
+  \For{$b$ in $\mathbf{B}$}{
+    split dimension := $\text{max}_j$ $|\text{b-center}_{j}-\text{m-center}_{j}|$\;
+    Adjust the widths of $b$ and \textit{m} in \textit{split dimension}\\
+    to be the difference between the centers in that dimension.
+  }
+ }
+\end{algorithm}
+
+\section{Evaluating A Box}
+Interpolating and approximating requires using a set of boxes as
+regions of influence for response values. Given a new data point $x$
+is contained by a box $b$ (with associated $c$,$l$,and $u$), we
+evaluate $b(x)$ as:
+
+\begin{enumerate}
+\item Scale $x$ according to the lower and upper widths of $b$ such
+  that all $x_j < c_j$ map to $(0,0.5)$ and all $x_j > c_j$ map to
+  $(0.5,1)$.
+\item Scale this box-normalized $x$ to be in the appropriate range for
+  the order of B-Spline chosen ($\times 2$ for order 2,$\times 3$ for order 3, etc.)
+\item Take the product of the B-Splines along each dimension for the
+  box spline normalized $x$ value. $\prod_{j=1}^{d} f(x_j)$ where $f$
+  is the piecewise polynomial defining the b-spline of the desired order.
+
+\end{enumerate}
+
+\section{Interpolate $y$ for new $x$}
+
+Identify the set of boxes $\mathbf{B} = \{b_i\ |\ b_i\ \text{contains}\ x\}$.
+Calculate the estimated response value by
+$$\frac{\sum_{b\in\mathbf{B}}b\text{-response} \times b(x)}{\sum_{b\in\mathbf{B}}b(x)}$$
+where $b\text{-response}$ is the response value associated with each
+box. Notice that this is using the box functions as a weighted sum.
+
+\section*{Notes}
+
+The ordering that has been used when adding points to the box mesh thus far is:
+
+\begin{itemize}
+\item Add the $x_i$ with the median valued $y_i$.
+\item Compute the approximation surface error at all remaining $x_i$,
+  add the $x_i$ to the surface with the most (relative) error.
+\item Repeat previous step until small enough maximum error is obtained.
+\end{itemize}
+
+It is important to note that this constant recalculation of error
+causes an increase in the runtime by a factor of $n$.
+
+\end{document}
+
+%===============================
+%     Old Report (too long)     
+%===============================
 
@@
 See the following visualized example of a box in 2 dimensions.
-
-\vspace{3mm}
-
-\begin{center}
-\begin{tikzpicture}[scale=1]
-
-  \draw[thick] (3,1.1) circle (0.7mm);
-
-  %% Draw arrows for each width
-  \draw[thick,->] (3,1.2) -- (3,2.4);
-  \draw[thick,->] (3,1.0) -- (3,0.1);
-  \draw[thick,->] (3.1,1.1) -- (9.9,1.1);
-  \draw[thick,->] (2.9,1.1) -- (0.1,1.1);
-
-  %% Draw boundary of box
-  \draw[thick,-] (0,0) -- (10,0);
-  \draw[thick,-] (10,0) -- (10,2.5);
-  \draw[thick,-] (10,2.5) -- (0,2.5);
-  \draw[thick,-] (0,2.5) -- (0,0);
-
-  %% Add text to the picture
-  \node at (3.25,1.35) {$c$};
-  \node at (1.3,1.35) {$l_0$};
-  \node at (6.3,1.35) {$u_0$};
-  \node at (2.75,0.55) {$l_1$};
-  \node at (2.75,1.8) {$u_1$};
-
-\end{tikzpicture}
-\end{center}
 

### 2017-12-02 Code_[2017-07-11]_Data_Comparison_Check/compare.py

@@ -1,0 +1,403 @@
+import pickle, sys, os, time, webbrowser
+import numpy as np
+from Resources.plotly_interface import *
+
+CURR_DIR = os.path.abspath(os.path.curdir)
+FILE_DELIM = ","
+RESOURCE_DIR = os.path.join(CURR_DIR, "Resources")
+OLD_DATA = os.path.join(RESOURCE_DIR,"VarSys-2016.txt")
+OLD_DATA_NP = os.path.join(RESOURCE_DIR, "VarSys-2016.pkl")
+LOAD_RAW_OLD_DATA = False # False -> use saved pkl file if it exists
+DATA_TYPE_KEYS = ["Machine", "Store", "Journal", "Hyp", "Hyp_Sched",
+                  "VM_Sched", "RCU", "F_Size", "R_Size", "Threads",
+                  "Mode", "Freq", "Throughput"]
+MEAN_OPACITY = 0.3
+DATA_TYPES = {
+    "Machine":'S3',
+    "Store":'S3',
+    "Journal":'S3',
+    "Hyp":'S3',
+    "Hyp_Sched":'S4',
+    "VM_Sched":'S4',
+    "RCU":float,
+    "F_Size":float,
+    "R_Size":float,
+    "Threads":float,
+    "Mode":'S13',
+    "Freq":float,
+    "Throughput":float
+}
+
+# Read the help string for the program
+with open("README.txt") as f: HELP_STRING = f.read()
+
+HTML_INFO_ALERT = '''
+<div class="alert alert-info alert-dismissable">
+  <a href="#" class="close" data-dismiss="alert" aria-label="close">&times;</a>
+  <strong>%s</strong>%s
+</div>
+'''
+
+# Takes: .format(title, mean_diff, stdev_diff, var_diff, fig_str)
+HTML_BEGIN = '''
+<!DOCTYPE html>
+<html lang="en">
+  <head>
+    <title>Report on F Size and R Size Variance</title>
+    <!-- Default setup for a webpage with basic bootstrap -->
+    <meta charset="utf-8">
+    <meta name="viewport" content="width=device-width, initial-scale=1">
+    <link rel="stylesheet" href="{0}">
+    <link rel="stylesheet" href="{1}">
+
+    <script src="{2}"></script>
+    <script src="{3}"></script>
+
+    <style type="text/css">
+        #Brand {{ cursor: default; }}
+        #Project {{ float: right; }}
+        #Project-Name:hover {{ color: #555; }}
[…]

### 2017-12-02 Documents_[2016]/Bspline.m

@@ -1,0 +1,8 @@
+bspl[j_Integer,1,t_List,x_Real] := If[t[[j]] <= x
+< t[[j+1]], 1.0,0.0]
+
+bspl[i_Integer,k_Integer,t_List,x_Real] := If[
+t[[i+k-1]] == t[[i]], 0.0, (x-t[[i]])/(t[[i+k-1]] - t[[i]])*
+bspl[i,k-1,t,x]] + If[t[[i+k]] == t[[i+1]], 0.0, (t[[i+k]]-x)
+/(t[[i+k]]-t[[i+1]])*bspl[i+1,k-1,t,x]]
+

### 2017-12-02 Documents_[2016]/time_Delaunay.py

@@ -1,0 +1,72 @@
+import time
+from scipy.spatial import Delaunay
+import numpy as np
+import pygal
+
+def test(pts, dim):
+    a = np.random.rand(pts, dim)*10
+    start = time.time()
+    hull = Delaunay(a)
+    return time.time() - start
+
+if __name__ == "__main__":
+    # Should take about 5 minutes to run
+
+    t_v_p = pygal.Line()
+    time_vs_pts_2 = []
+    time_vs_pts_3 = []
+    time_vs_pts_4 = []
+    time_vs_pts_5 = []
+    now = time.time()
+    for pts in range(0, 10001, 500):
+        if pts == 0: pts = 100
+        print("%.1f%% (%i points) at %.2f"%(
+            100.0*pts/10001, pts,
+            time.time() - now))
+        now = time.time()
+        time_vs_pts_2.append( test(pts, 2) )
+        time_vs_pts_3.append( test(pts, 3) )
+        time_vs_pts_4.append( test(pts, 4) )
+        time_vs_pts_5.append( test(pts, 5) )
+    t_v_p.add("Dim 5", time_vs_pts_5)
+    t_v_p.add("Dim 4", time_vs_pts_4)
+    t_v_p.add("Dim 3", time_vs_pts_3)
+    t_v_p.add("Dim 2", time_vs_pts_2)
+    t_v_p.title = "Delaunay Runtime"
+    t_v_p.x_labels =  map(str,[100]+list(range(500, 10001, 500)))
+    t_v_p.x_title = "Number of Points"
+    t_v_p.y_title = "Computation Time (sec)"
+    t_v_p.render_in_browser()
+    t_v_p.render_to_file("Delaunay_time-vs-pts.svg")
+
+
+    t_v_d = pygal.Line()
+    time_vs_dim_20 = []
+    time_vs_dim_50 = []
+    time_vs_dim_80 = []
+    time_vs_dim_100 = []
+    now = time.time()
+    dim = 2
+    for i in range(9):
+        print("(%i dimension) at %.2f"%(
+            dim-1, time.time() - now))
+        now = time.time()
+        time_vs_dim_20.append(test(20, dim))
+        time_vs_dim_50.append(test(50, dim))
+        time_vs_dim_80.append(test(80, dim))
+        time_vs_dim_100.append(test(100, dim))
+        dim += 1
+    print("(%i dimension) at %.2f"%(
[…]

### 2017-12-02 Documents_[2017]/cheney_light.py

@@ -1,0 +1,26 @@
+import numpy as np
+from plotly_interface import Plot
+
+data = np.array([
+    [1.2, 1.2],
+    [3.6, 3.1],
+    [4.3,-1.3],
+    [6.1, 2.7],
+    [7.8, 1.4]
+])
+
+def u1(x):
+    return abs(x)
+
+def u2(x):
+    return np.exp(-x**2)
+
+def F(x,u=u1,d=data):
+    return sum(cj*u(x - xj) for (xj,cj) in d)
+
+p = Plot()
+
+p.add("Data", data[:,0], data[:,1])
+p.add_func("Abs", lambda x: F(x,u1), [min(data[:,0])-1, max(data[:,0])+1])
+p.add_func("e", lambda x: F(x,u2), [min(data[:,0])-1, max(data[:,0])+1])
+p.plot(fixed=False)

### 2017-12-02 Documents_[2017]/plot_basis.py

@@ -1,0 +1,33 @@
+import pygal
+
+CUSTOM_CSS_FILE = 'file:///home/thomas/Git_Analytics/Analytics-Vis/Code/ViewData/Value_Distribution/no_dot.css'
+num_points = 500
+min_val = 0
+width = 10
+
+on_side = lambda x,knot,side: (x - knot) * side > 0
+basis_1 = lambda x,knot,side=1: on_side(x,knot,side) * (x - knot)
+basis_2 = lambda x,knot,side=1: on_side(x,knot,side) * (x - knot)**2
+basis_3 = lambda x,knot,side=1: on_side(x,knot,side) * (x - knot)**3
+
+funcs = [
+    ("B-1, k-(4)",   lambda x: basis_1(x,4)),
+    ("B-2, k-(4)",   lambda x: basis_2(x,4)),
+    ("B-2, k-(4,2)", lambda x: basis_2(x,4) + basis_1(x,4,-1)),
+    ("B-3, k-(2)",   lambda x: 0.01 * basis_3(x,2)),
+]
+
+
+# Create the plot
+config = pygal.Config(fill=False, show_legend=True)
+config.css.append(CUSTOM_CSS_FILE)
+plot = pygal.XY(config)
+
+for f_name, func in funcs:
+    pts = []
+    for i in range(num_points+1):
+        x = (i / num_points) * width + min_val
+        pts.append( (x, func(x)) )
+    plot.add(f_name,pts)
+
+plot.render_in_browser()

### 2017-12-02 Paper_[2017-10-22]_MOANA/bootstrapped_predictions.py

@@ -1,0 +1,489 @@
+import pickle, os
+import numpy as np
+from scipy.interpolate import splrep, splev
+from util.algorithms import Delaunay
+from util.data import read_struct
+from util.plotly import Plot, multiplot
+
+# Path to raw text data
+HOME_DIR = os.path.expanduser("~")
+DATA_PATH = os.path.join(HOME_DIR, *["Git_Analytics", "Analytics-Vis",
+                                     "Data", "VirginiaTech",
+                                     "Many_Trials", "Data",
+                                     "Clean_Data.txt"])
+
+# DATA_PATH = os.path.join(HOME_DIR, *["Git_Analytics", "Analytics-Vis",
+#                                      "Data", "VirginiaTech",
+#                                      "Raw_Sampled", "Full_Data",
+#                                      "Clean_Data.txt"])
+
+# Paths to local cache files (for speed of re-execution)
+LOCAL_FILE = "IOzone_data_2016.pkl"
+ORGANIZED_DATA_FILE = "data_dict.pkl"
+# LOCAL_FILE = "Many_Trials_IOzone_data_2016.pkl"
+# ORGANIZED_DATA_FILE = "many_trials_data_dict.pkl"
+PREDICTED_RUNS_FILE = "predicted_runs.pkl"
+NUM_RUNS_DICT = "%i_runs_%i_convergence_(%s,%s,%s).pkl"
+
+# Known hard-coded data settings
+OS_SCHEDULER = "Hyp_Sched"
+VM_SCHEDULER = "VM_Sched"
+FILE_SIZE = "F_size"
+RECORD_SIZE = "R_Size"
+THREAD_COUNT = "Threads"
+TEST_TYPE = "Mode"
+CPU_FREQUENCY = "Freq"
+IO_THROUGHPUT = "Throughput"
+# Data columns broken up by type of interest
+CATEGORICAL_COLS = [OS_SCHEDULER, VM_SCHEDULER, TEST_TYPE]
+CONTINUOUS_COLS = [FILE_SIZE, RECORD_SIZE, THREAD_COUNT, CPU_FREQUENCY]
+TARGET_COLUMN = [IO_THROUGHPUT]
+# Print settings (for showing more or less specific information)
+SHOW_VALUES = False
+PLOT_BOOTSTRAPPED = False
+PLOT_CONVERGENCE = True
+PLOT_CONVERGENCE_EXAMPLE = False
+COLLECT_RUNS_DATA = False
[…]

### 2017-12-02 Report_[2017-02-05]_Variance_Itemsets/variance_itemset_report.py

@@ -1,0 +1,153 @@
+# Global variables
+import sys
+sys.path = ['/home/thomas/Git_Analytics/Analytics-Vis/Code'] + sys.path
+sys.path = ['/home/thomas/Dropbox/Research/TL ML/Apriori_Tree'] + sys.path
+
+import numpy as np
+
+from settings import read_data_settings, data_for_spec, spec_str_to_dict, dict_to_spec_str
+from tl_apriori_tree import AprioriTree
+
+
+# Get the three primary command line arguments
+group_id = "VirginiaTech"
+project_name = ["VarSys", "VarSys_Per_Thread"]
+
+# Bottom 20% of Variance
+bot_20_var = "2.5e+08_&le;_V_&le;_7e+11"
+# Mid 60% of Variance (20th - 80th percentile)
+mid_60_var = "7e+11_&le;_V_&le;_7.7e+14"
+# Top 20% of Variance
+top_20_var = "7.7e+14_&le;_V_&le;_2.4e+17"
+
+# # Top 10% of Variance
+# top_10_var = "3.5e+15_&le;_V_&le;_2.4e+17"
+
+var_settings = [(0.0, 20.0), (20.0, 80.0), (80.0, 100.0)]
+var_setting_names = ("Bottom 20%", "Mid 60%", "Top 20%")
+
+default_data_settings = read_data_settings(group_id, project_name[1])
+data_settings_dict = dict(default_data_settings)
+data_settings_order = [s[0] for s in default_data_settings]
+# Update the settings that we want to lock down
+data_settings_dict.update({
+    "Mode":["Fread"],
+    "Stdev":[],
+    "Mean":[],
+})
+
+data_req = dict_to_spec_str(data_settings_dict, data_settings_order)
+
+# Read the data from file to pass to plotting functions,
+# change stdout so that the print statements from
+# "data_for_spec" do not make their way to the terminal
+normal_out = sys.stdout
+sys.stdout = open("/dev/null",'w')
+raw_data, header, types,_ = data_for_spec(
+    group_id, project_name[1], spec_str_to_dict(data_req))
+sys.stdout = normal_out
+
+print("Raw data shape: (%s, %s)"%(raw_data.shape[0], len(header)))
+print("Variance = [%s to %s]"%(min(raw_data["Variance"]), max(raw_data["Variance"])))
+print()
+print()
+# Remove "Variance" and "Mode" attributes from header
+header = [h for h in header if h not in ["Variance", "Mode"]]
+
[…]

### 2017-12-02 Report_[2017-02-08]_F_R_Size_Differences/3D-Scatter.handler.js

@@ -1,0 +1,106 @@
+var graph3d_array = graph3d_array || [];
+
+function redraw_plots() {
+    for (var i = 0; i < graph3d_array.length; i++) {
+	graph3d_array[i].redraw();
+    }
+}
+
+// Redraw the plots if the window changes size
+window.addEventListener('onresize', redraw_plots);
+
+function add_3D_Scatter(group, project, view, name, selected_data) {
+// Function for creating 3D plot based on JSON returned by API call.
+    function create_options (x_axis, y_axis, z_axis) {
+    // Generate an "options" object for visualizing the 3D scatter plot
+	var options = {	
+	    align: 'center',
+	    width:  '100%',
+	    height: '60vmin',
+	    style: "dot",
+	    backgroundColor: {fill:'transparent', stroke:'transparent', strokWidth:0},
+	    cameraPosition: {horizontal: 1.0, vertical: 0.5, distance: 2.3},
+            dotSizeRatio: 0.015,
+            keepAspectRatio: false,
+            // verticalAspectRatio: 0.8,
+	    // showShadow: false,
+            xValueLabel: function (value) {
+		return (value.toExponential(1));
+            },
+            yValueLabel: function (value) {
+		return (value.toExponential(1));
+            },
+            zValueLabel: function (value) {
+		return (value.toExponential(1));
+            },
+            xLabel: x_axis,
+            yLabel: y_axis,
+            zLabel: z_axis,
+	    // Option tooltip can be true, false, or a function returning a string with HTML contents
+            tooltip: function (point) {
+		// parameter point contains properties x, y, z
+		return ('('+(point.x).toFixed(0)+','+
+			(point.y).toFixed(0)+'):  '+
+			(point.z).toExponential(1)) ;
+            }
+	};
+	return options;
+    };
+
+    // Beginning of 'add_3D_Scatter' function
+    var plot_id = name + "--" + "3D-Scatter"
+    var plot_title = document.createElement("div");
+    var plot = document.createElement("div");
+    var holder = document.createElement("div");
+
+    plot_title.id = plot_id + "--title";
+    plot.id = plot_id + "--plot";
+    plot_title.style.paddingTop = "10px";
+    holder.style.backgroundColor = "#fafafa";
+    // holder.style.borderWidth =  "1px";
+    // holder.style.borderColor = "#e5e5e5";
+    // holder.style.borderStyle = "solid";
+    // holder.style.borderRadius = "10px";
[…]

### 2017-12-02 Report_[2017-03]_Convergence/code.py

@@ -1,0 +1,279 @@
+# Global variables
+import sys, os, time, pickle, webbrowser
+import numpy as np
+# Get relative path to the data and this directory
+home = os.path.expanduser("~")
+print("Using home:", home)
+sys.path += [home+"/Dropbox/Sync/bin"]
+print("Using path:", sys.path)
+from plotly_interface import *
+
+
+sys.path = [home + "/Git_Analytics/Analytics-Vis/Code"] + sys.path
+curr_dir = os.path.abspath(os.path.dirname(__file__))
+from settings import data_for_spec, spec_str_to_dict
+
+NUM_PLOT_POINTS = 1000
+NUM_SUBSAMPLES = 1000
+NUM_TRIALS = 2
+
+sys.path += [home + "/Git_Analytics/Analytics-Vis/Data/VirginiaTech/Many_Trials/Data/"]
+from plot_throughput import read_raw_file
+
+# Code for generating the view file for a requested view
+if __name__ == "__main__":
+    # If there is an existing report, store it as 'old'
+    os.system("mv report.html old-report.html")
+
+    # Get the three primary command line arguments
+    group_id = "VirginiaTech"
+    project_names = ["Raw_CFQ_NOOP_64_Fread", "F_R_Size_Raw", "Many_Trials"]
+    data_req_base = "[F_size={0}][R_Size={1}][Hyp_Sched=CFQ][VM_Sched=NOOP][Freq=2700000][Mode=Fread][Throughput=-inf_&le;_inf]"
+
+    # Ordered in terms of largest to smallest discrepancy variance
+    f_r_sizes = [(64,32),(1024,32),
+                         (1024,512)]
+
+    study = "Throughput"
+    yaxis = "Percent Less"
+    names = ["First",
+             "Second",
+             "True",
+             # "All Together"
+    ]
+
+    data_file_names = [home+"/Git_Analytics/Analytics-Vis/Data/VirginiaTech/Raw_CFQ_NOOP_64_Fread/Data/Clean_Data.txt",
+                       home+"/Git_Analytics/Analytics-Vis/Data/VirginiaTech/F_R_Size_Raw/Data/Clean_Data.txt",
+                       home+"/Git_Analytics/Analytics-Vis/Data/VirginiaTech/Many_Trials/Data/Clean_Data.txt"
+    ]
+
+    # Cycle through the file and record sizes, generating a plot for each
+    for f_size, r_size in f_r_sizes:
+        # Get all three data sets from file
[…]

### 2017-12-02 Report_[2017-04]_Modes/code.py

@@ -1,0 +1,170 @@
+# Global variables
+import sys, os, time, pickle, webbrowser
+import numpy as np
+# Get relative path to the data and this directory
+home = os.path.expanduser("~")
+print("Using home:", home)
+sys.path += [home+"/Dropbox/Sync/bin"]
+print("Using path:", sys.path)
+from plotly_interface import *
+
+
+sys.path.append(home + "/Git_Analytics/Analytics-Vis/Code")
+curr_dir = os.path.abspath(os.path.dirname(__file__))
+from settings import data_for_spec, spec_str_to_dict
+
+NUM_PLOT_POINTS = 1000
+NUM_SUBSAMPLES = 1000
+NUM_TRIALS = 2
+NUM_MODES = 3
+
+sys.path += [home + "/Git_Analytics/Analytics-Vis/Data/VirginiaTech/Many_Trials/Data/"]
+from plot_throughput import read_raw_file
+
+# Code for generating the view file for a requested view
+if __name__ == "__main__":
+    # If there is an existing report, store it as 'old'
+    os.system("mv report.html old-report.html")
+
+    # Get the three primary command line arguments
+    group_id = "VirginiaTech"
+    project_names = ["Raw_CFQ_NOOP_64_Fread", "F_R_Size_Raw", "Many_Trials"]
+    data_req_base = "[F_size={0}][R_Size={1}][Hyp_Sched=CFQ][VM_Sched=NOOP][Freq=2700000][Mode=Fread][Throughput=-inf_&le;_inf]"
+
+    # Ordered in terms of largest to smallest discrepancy variance
+    f_r_sizes = [(64,32),(1024,32),
+                         (1024,512)]
+
+    study = "Throughput"
+    yaxis = "Percent Less"
+    names = ["First",
+             "Second",
+             "True",
+             # "All Together"
+    ]
+
+    data_file_names = [home+"/Git_Analytics/Analytics-Vis/Data/VirginiaTech/Raw_CFQ_NOOP_64_Fread/Data/Clean_Data.txt",
+                       home+"/Git_Analytics/Analytics-Vis/Data/VirginiaTech/F_R_Size_Raw/Data/Clean_Data.txt",
+                       home+"/Git_Analytics/Analytics-Vis/Data/VirginiaTech/Many_Trials/Data/Clean_Data.txt"
+    ]
+
+    # Cycle through the file and record sizes, generating a plot for each
+    for f_size, r_size in f_r_sizes:
+        # Get all three data sets from file
[…]

### 2017-12-02 Report_[2017-09-20]_Dangers_of_Exponential_Scaling/log_scale.py

@@ -1,0 +1,49 @@
+from pyasis.algorithms import Delaunay
+from pyasis.plotly import Plot
+import numpy as np
+
+x = np.array([1,3,3.5,5,6,7])
+y = np.array([1,0,1.8,2.5,5.5,6])
+min_max = [min(x), max(x)]
+shift = min_max[0]
+scale = min_max[1] - min_max[0]
+
+# Define a truth function
+fun = lambda new_x: np.interp(new_x, x, y)
+
+# Pick a number of samples
+num_samples = 8
+
+# Generate the evenly and logarithmically spaced sample points
+even_x = np.linspace(min(x),max(x),num_samples)
+even_y = fun(even_x)
+even_x = even_x.reshape((num_samples, 1))
+even_d = Delaunay()
+even_d.fit(even_x, even_y)
+even_d_out = lambda x: even_d(x)
+
+# Fit the exponentially spaced points
+log_x = np.linspace(np.log(min(x)), np.log(max(x)), num_samples)
+log_y = np.log( fun(np.exp(log_x)) )
+log_x = log_x.reshape((num_samples,1))
+# Generate a non-trasnformed fit
+log_d = Delaunay()
+log_d.fit(np.exp(log_x), np.exp(log_y))
+log_d_out = lambda x: log_d(x)
+# Generate a transformed fit
+log_d_trans = Delaunay()
+log_d_trans.fit(log_x, log_y)
+log_d_trans_out = lambda x: np.exp( log_d_trans( np.log(x) ) )
+
+even_color = "rgb(30,180,80)"
+log_color = "rgb(100,50,200)"
+
+p = Plot("The Dangers Associated with Log-Scaling")
+p.add_func("Truth", fun, min_max, group="Truth")
+p.add("Evenly Spaced Points", even_x[:,0], even_y, color=even_color)
+p.add_func("Delaunay Even", even_d_out, min_max, color=even_color)
+p.add("Exponentially Spaced Points", np.exp(log_x[:,0]), np.exp(log_y), color=log_color)
+p.add_func("Delaunay Log Normal", log_d_out, min_max, color=log_color)
+p.add_func("Delaunay Log Transformed", log_d_trans_out, min_max, color=log_color)
+
+p.plot()

### 2017-12-02 Report_[2017-09-20]_Dangers_of_Exponential_Scaling/run_test.py

@@ -1,0 +1,222 @@
+import pickle, os
+from util.data import read_struct
+import numpy as np
+# Add a field to the unique columns that will hold the throughput std
+from numpy.lib.recfunctions import append_fields
+
+data_file = "Li_Results.txt"
+train_pkl = "Train.pkl"
+test_pkl = "Test.pkl"
+train = "Train.txt"
+test = "Test.txt"
+target = "Throughput"
+numerical_cols = ['F_size', 'R_Size']
+categorical_cols = ['Hyp_Sched', 'VM_Sched', 'Mode']
+testing_f_r_sizes = [(512,32),(512,128),(512,256),(768,32),(768,128)]
+output_file = "New_Predictions.csv"
+
+
+
+# SUMMARY:
+#    Return a structured array with only the unique combinations of
+#    desired columns and a summary statistic describing the set of
+#    values that all share similar settings.
+# 
+# INPUTS:
+#    data: 
+#       A sorted numpy structured array (lexigraphic by column)
+#    non_target_cols:
+#       A list of the column headers that should be used to identify 
+#       unique rows
+#    target:
+#       A target column header for a numerical column in "data".
+#    statistic: (Default = np.std)
+#       A function that converts a numpy array of numbers into a
+#       single number. 
+#    verbose: (Default = True)
+#       A boolean indicating whether or not print statements about
+#       progress should be produced during the conversion.
+# 
+# RETURNS:
+#    A sorted numpy structured array (same order as "data")  with the
+#    "target" column representing a summary statistic of the repeated
+#    settings and an additional column, 'Count' that gives the number
+#    of points encountered at each individual unique setting. 
+def reduce_to_unique_vals(data, non_target_cols, target,
+                          statistic=np.std, verbose=True):
+    unique_data = np.unique(data[non_target_cols])
+    unique_data = append_fields(unique_data, target,
+                                np.zeros(unique_data.shape[0]))
+    unique_data = append_fields(unique_data, "Count",
+                                np.zeros(unique_data.shape[0], dtype=int))
+    # Initialize indices for collecting sets of identical rows
+    group_start_idx = 0
+    group_end_idx = 0
+    unique_current_idx = 0
+    unique_setting = unique_data[non_target_cols][unique_current_idx]
+    # Coalesce all data into the statistics
+    while (group_end_idx < len(data)):
+        if verbose: print("\r%i / %i"%(group_end_idx, len(data)), end="")
[…]

### 2017-12-11 Paper_[2017-11]_HPC/paper.tex
Added some articles from this semester, deleted a duplicate report, and put finishing touches on HPC paper.

@@
 % *** NOTE: May need to adjust titlevboxsize in the preamble
-\author{Thomas C. H. Lux \\ [12pt]
+\author{Thomas C. H. Lux \\ [8pt]
 Dept. of Computer Science \\
@@
 \and
-Layne T. Watson \\[12pt]
+Layne T. Watson \\[8pt]
 Dept. of Computer Science\\
@@
 \and
+\vspace{-30pt}\\
 Tyler H. Chang\\
 Jon Bernard\\
-Bo Li\\[12pt]
+Bo Li\\[8pt]
 Dept. of Computer Science\\ 
@@
 \and
-Li Xu\\[12pt]
+\vspace{-30pt}\\
+Li Xu\\[8pt]
 Dept. of Statistics\\ 
@@
 \and
+\vspace{-30pt}\\
 Godmar Back\\
 Ali R. Butt\\
-Kirk W. Cameron\\[12pt]
+Kirk W. Cameron\\[8pt]
 Dept. of Computer Science\\ 
@@
 \and
-Yili Hong\\[12pt]
+\vspace{-30pt}\\
+Yili Hong\\[8pt]
 Dept. of Statistics\\ 
@@
 \maketitle
+
+\vspace{-40pt}
 
@@
 reconfigured as required by updates or specific jobs. In the case that
-a system has changing workloads or non stationary performance
+a system has changing workloads or nonstationary performance
 objectives that range from maximizing computational throughput to
@@
 statistical model of the performance by running experiments on the
-system at select settings or (2) run artificial experiments using a
+system at select settings, or (2) run artificial experiments using a
 simplified simulation of the target system to estimate architecture
@@
 
-Among the statistical models presented in prior works
+Among the statistical models presented in prior works,
 \shortciteN{bailey2005performance} specifically mention that it is
@@
 respect to the size (number of points and dimension) of input
-data. Throughout the sections, $d$ will refer to the dimension of the
-input data, $n$ is the number of points in the input data, $x^{(i)}
-\in \mathbb{R}^d$ is the $i$-th input data point, $x^{(i)}_j$ is the
+data. Throughout, $d$ will refer to the dimension of the input data,
+$n$ is the number of points in the input data, $x^{(i)} \in
+\mathbb{R}^d$ is the $i$-th input data point, $x^{(i)}_j$ is the
 $j$-th component of $x^{(i)}$, and $f(x^{(i)})$ is the response value
@@
 
-The remainder of the paper is broken up into four major parts. Section
+The remainder of the paper is broken up into five major parts. Section
 \ref{sec:multivariate} provides an overview of the multivariate
@@
 \subsection{Regression}
+\vspace{-10pt}
 Multivariate approximations are capable of accurately modeling a
@@
 
+\vspace{-10pt}
 \subsubsection{Multivariate Adaptive Regression Splines}
+\vspace{-10pt}
 This approximation was introduced in
@@
 where $s$ is the iteration number, $B_l(x)$ and $B_k(x)$ are basis
-functions from the previous iteration, $c, v \in \mathbb{R}$, $w_+ =
-\{w, w \geq 0;\quad 0, w < 0\},$ and $w_+ = (-w)_+$. After iteratively
-constructing a model, MARS then iteratively removes basis functions
-that do not contribute to goodness of fit. In effect, MARS creates a
-locally component-wise tensor product approximation of the data. The
-overall computational complexity of Fast MARS is $\mathcal{O}(n d
-m^3)$ where $m$ is the maximum number of underlying basis
-functions. This paper uses an implementation of MARS
-\shortcite{rudy2017pyearth} with $m = 200$.
+functions from the previous iteration, $c, v \in \mathbb{R}$,
 
+$$w_+ = \begin{cases} w, & w \geq 0 \\ 0, & w < 0 \end{cases},$$
+
+and $w_- = (-w)_+$. After iteratively constructing a model, MARS then
+iteratively removes basis functions that do not contribute to goodness
+of fit. In effect, MARS creates a locally component-wise tensor
+product approximation of the data. The overall computational
+complexity of Fast MARS is $\mathcal{O}(n d m^3)$ where $m$ is the
+maximum number of underlying basis functions. This paper uses an
+implementation of Fast MARS \shortcite{rudy2017pyearth} with $m =
+200$.
+
+\vspace{-10pt}
 \subsubsection{Multilayer Perceptron Regressor}
+\vspace{-10pt}
 The neural network is a well studied and widely used method for both
@@
 
-$$ l(u) = \big( u^t W_l \big)_+ $$
+$$ l(u) = \big( u^t W_l \big)_+ ,$$
 
 where $W_l$ is the $i$ by $j$ weight matrix for layer $l$. In this
-form, the multi layer perceptron (MLP) produces a piecewise linear
+form, the multilayer perceptron (MLP) produces a piecewise linear
 model of the input data. The computational complexity of training a
-multi layer perceptron is $\mathcal{O}(n d m)$, where $m$ is
+multilayer perceptron is $\mathcal{O}(n d m)$, where $m$ is
 determined by the sizes of the layers of the network and the stopping
@@
 \subsubsection{Support Vector Regressor}
+\vspace{-10pt}
 Support vector machines are a common method used in machine learning
@@
 where $K$ is the selected kernel function, $a \in \mathbb{R}^n$, $b
-\in \mathbb{R}$ are coefficients to be solved for simultaneously with
-$b \in \mathbb{R}$, and $\epsilon$ is the error tolerance. The
+\in \mathbb{R}$ are coefficients to be solved for simultaneously. The
 computational complexity of the SVR is $\mathcal{O}(n^2dm)$, with $m$
@@
 
+\vspace{-10pt}
 \subsection{Interpolation}
-
+\vspace{-10pt}
 In some cases it is desirable to have a model that can recreate the
@@
 
+\vspace{-10pt}
 \subsubsection{Delaunay}
-
+\vspace{-10pt}
 The Delaunay method of interpolation is a well studied geometric
@@
 vertices $v^{(0)}, v^{(1)}, \ldots, v^{(d)}$, $x \in S$, and data
-values $f(v^{(i)})$, $i=0,\ldots,d$, $x$ is a unique convex
+values $f(v^{(i)})$, $i=0$,$\ldots$,$d$, $x$ is a unique convex
 combination of the vertices,
 
-$$ x = \sum_{i=0}^{d} w_i v^{(i)}; \quad \sum_{i=0}^{d} w_i = 1; \quad
-w_i \geq 0; \quad i=0,\ldots,d $$
+$$ x = \sum_{i=0}^{d} w_i v^{(i)}, \quad \sum_{i=0}^{d} w_i = 1, \quad
+w_i \geq 0, \quad i=0,\ldots,d ,$$
 
@@
 \subsubsection{Linear Shepard}
-
+\vspace{-10pt}
 The linear Shepard method (LSHEP) is a blending function using local
@@
 where $W_k(x)$ is a locally supported weighting function and $P_k(x)$
-is a local linear approximation to the data satisfying $P_k(x^{(k)} =
-f(x^{(x)})$. The computational complexity LSHEP is
-$\mathcal{O}(n^2d^3)$. This paper uses the FORTRAN95 implementation of
-LSHEP in SHEPPACK \shortcite{thacker2010algorithm}.
+is a local linear approximation to the data satisfying
+$P_k\big(x^{(k)}\big) = f\big(x^{(x)}\big)$. The computational
+complexity of LSHEP is $\mathcal{O}(n^2d^3)$. This paper uses the
+Fortran\#95 implementation of LSHEP in SHEPPACK
+\shortcite{thacker2010algorithm}.
 
@@
 \subsection{Data}
+\vspace{-10pt}
 In order to evaluate the viability of multivariate models for
@@
 test. The 40 executions for each system configuration are converted
-into the mean and sample variance, both values in $\mathbb{R}$ capable
-of being modeled individually by the multivariate approximation
+into the mean and variance, both values in $\mathbb{R}$ capable of
+being modeled individually by the multivariate approximation
 techniques presented in Section \ref{sec:multivariate}. The summary of
@@
     Throughput Mean & [$2.6 \times 10^5$, $5.9 \times 10^8$]\\
-    Throughput Sample Variance & [$5.9\times 10^{10} $, $4.7 \times 10^{16}$]\\
+    Throughput Variance & [$5.9\times 10^{10} $, $4.7 \times 10^{16}$]\\
     \hline
@@
   \caption{Histograms of 100-bin reductions of the PMF of I/O
-    throughput mean (top) and I/O throughput sample variance
-    (bottom). In the mean plot, the first 1\% bin (truncated in plot)
-    has a probability mass of .45. In the variance plot, the second
-    1\% bin has a probability mass of .58. It can be seen that the
-    distributions of throughputs are primarily of lower magnitude with
-    occasional extreme outliers.}
+    throughput mean (top) and I/O throughput variance (bottom). In the
+    mean plot, the first 1\% bin (truncated in plot) has a probability
+    mass of .45. In the variance plot, the second 1\% bin has a
+    probability mass of .58. It can be seen that the distributions of
+    throughputs are primarily of lower magnitude with occasional
+    extreme outliers.}
   \label{fig:raw_throughput}
@@
 
+\vspace{-10pt}
 \subsection{Dimensional Analysis}
+\vspace{-10pt}
 This work utilizes an extension to standard $k$-fold cross validation
@@
 \subsubsection{Multidimensional Analysis}
+\vspace{-10pt}
 This procedure combines random selection of training and testing
@@
 \begin{enumerate}
-\item for all $k = 1, \ldots, d$ and for all subsets of input
-  components $F = \{ j_1, j_2, \ldots j_k \}$; reduce the input data
-  to unique points $z \in \mathbb{R}^k$ with $f(z) = E[ \{ f(x^{(i)})
-    \quad \forall i \quad s.t. \quad (x^{(i)}_F = z) \} ]$ where
-  $x^{(i)}_F$ is the $k$-tuple containing only those components of
-  $x^{(i)}$ in $F$.
-\item for all $r$ in $\{5, 10, \ldots, 95\}$; generate $N$ random
+\item For all $k = 1$, $\ldots$, $d$ and for all nonempty subsets $F
+  \subset \{1, 2, \ldots, d\}$, reduce the input data to points $(z,
+  f_F(z))$ with $z \in \mathbb{R}^k$ and $f_F(z) = E[ \{ f(x^{(i)})
+    \mid (x^{(i)}_F = z) \} ]$, where $E[\cdot]$ denotes the mean and
+  $x^{(i)}_F$ is the subvector of $x^{(i)}$ indexed by $F$.
+\item For all $r$ in $\{5, 10, \ldots, 95\}$, generate $N$ random
   splits $(train, test)$ of the reduced data with $r$ percentage for
   training and $100 - r$ percentage for testing.
-\item when generating each of $N$ random $(train, test)$ splits,
-  ensure that all points from $test$ are on or inside the convex hull
-  of points in $train$; also ensure that the points in $train$ are
-  well spaced.
+\item When generating each of $N$ random $(train, test)$ splits,
+  ensure that all points from $test$ are in the convex hull of points
+  in $train$ (to prevent extrapolation); also ensure that the points
+  in $train$ are well spaced.
 \end{enumerate}
 
-In order to ensure that the testing points are inside the convex hull
-of training points, this paper identifies the set of (reduced
-dimension) points that compose the convex hull and forcibly places
-those points in the training set. In order to ensure that training
-points are well spaced, this paper relies on the statistical method
-for picking points from \shortciteN{amos2014algorithm} that proceeds
-as follows:
+In order to ensure that the testing points are in the convex hull of
+the training points, the convex hull vertices of each set of (reduced
+dimension) points are forcibly placed in the training set. In order to
+ensure that training points are well spaced, a statistical method for
+picking points from \shortciteN{amos2014algorithm} is used:
 \begin{enumerate}
-\item Generate a sequence of all pairs of points sorted by ascending
-  pairwise $L_2$ distance between points, ${(x^{(i_1)},x^{(j_1)}),
-    (x^{(i_2)},x^{(j_2)}), \ldots}$ such that
-  $||x^{(i_k)}-x^{(j_k)}||_2 \leq ||x^{(i_{k+1})}-x^{(j_{k+1})}||_2$
+\item Generate a sequence of all pairs of points
+  ${(z^{(i_1)},z^{(j_1)}), (z^{(i_2)},z^{(j_2)}), \ldots}$ sorted by
+  ascending pairwise Euclidean distance between points, so that
+  $||z^{(i_k)}-z^{(j_k)}||_2 \leq ||z^{(i_{k+1})}-z^{(j_{k+1})}||_2$.
 \item Sequentially remove points from candidacy until only $|train|$
-  remain by randomly selecting one point from the pair $(x^{(i_k)},
-  x^{(j_k)})$ for $k = 1,\ldots$ if both $x^{(i_k)}$ and $x^{(j_k)}$
-  are still candidates.
+  remain by randomly selecting one point from the pair $(z^{(i_m)},
+  z^{(j_m)})$ for $m = 1,\ldots$ if both $z^{(i_m)}$ and $z^{(j_m)}$
+  are still candidates for removal.
 \end{enumerate}
@@
 Given the large number of constraints, level of reduction, and use of
-randomness in the MDA procedure, occasionally $N$ unique training
+randomness in the MDA procedure, occasionally $N$ unique training /
 testing splits may not be created or may not exist. In these cases, if
-there are fewer than $N$ possible splits then deterministically
-generated splits are used. Otherwise after $3*N$ attempts, only the
+there are fewer than $N$ possible splits, then deterministically
+generated splits are used. Otherwise after $3N$ attempts, only the
 unique splits are kept for analysis. The MDA procedure has been
-implemented in python3 while most regression and interpolation methods
-are FORTRAN wrapped with python. All randomness has been seeded for
-repeatability.
+implemented in Python\#3 while most regression and interpolation
+methods are Fortran wrapped with Python. All randomness has been
+seeded for repeatability.
 
-For any input component subset $F$ (of size $k$) and selected value of
-$r$, MDA will generate up to $N$ predictions that each multivariate
-model has made for each point $z^{(i)} \in \mathbb{R}^k$. There may be
-fewer than $N$ predictions made for any given point. Points in the
-convex hull for the selected subset of components will always be used
-for training, never for testing. Points that do not have any close
-neighbors will often be used for training in order to ensure well
-spacedness. Finally, as mentioned before, some subsets of components
-do not readily generate $N$ unique training and testing splits.
-
-MDA results in a distribution of possible estimates that a given
-multivariate model could make for a point in $\mathbb{R}^k$, $k \leq
-d$ given component set $F$. The summary results presented in this work
-use the expected value of the distributions at each point as the model
+For any index subset $F$ (of size $k$) and selected value of $r$, MDA
+will generate up to $N$ multivariate models $f_F(z)$ and predictions
+$\hat{f}_F\big(z^{(i)}\big)$ for a point $z^{(i)} \in \mathbb{R}^k$.
+There may be fewer than $N$ predictions made for any given
+point. Extreme points of the convex hull for the selected index subset
+will always be used for training, never for testing. Points that do
+not have any close neighbors will often be used for training in order
+to ensure well-spacing. Finally, as mentioned before, some index
+subsets do not readily generate $N$ unique training and testing
+splits. The summary results presented in this work use the median of
+the ($N$ or fewer) values $\hat{f}_F(z)$ at each point as the model
 estimate for error analysis.
@@
 A na\"{\i}ve multivariate prediction technique such as nearest
-neighbor could experience relative errors in the range $[0,
-  (\text{max}_f - \text{min}_f) / \text{min}_f ]$ when modeling a non
-negative function $f$ from data. The IOzone mean data response values
-span three orders of magnitude (as can be seen in Table
+neighbor could experience relative errors in the range $\displaystyle
+[0, \big(\max_x f(x) - \min_x f(x)\big) / \min_x f(x) ]$ when modeling
+a nonnegative function $f(x)$ from data. The IOzone mean data response
+values span three orders of magnitude (as can be seen in Table
 \ref{tab:data_type}) while variance data response values span six
@@
 
+\vspace{-10pt}
 \subsection{I/O Throughput Mean}
-
+\vspace{-10pt}
 Almost all multivariate models analyzed make predictions with a
@@
   \centering
-  \includegraphics[width=\textwidth,trim={0 .5in 0 .5in}]{Mean_TT_Ratio.pdf}
+  \includegraphics[width=\textwidth,trim={0 .5in 0 .3in}]{Mean_Dim.pdf}
+  \caption{These box plots show the prediction error of mean with
+    increasing dimension. The top box whisker for SVR is 40, 80, 90
+    for dimensions 2, 3, and 4, respectively. Notice that each model
+    consistently experiences greater magnitude error with increasing
+    dimension. Results for all training percentages are aggregated.}
+  \label{fig:mean_dim}
+\end{figure}
+
+\begin{figure}
+  \centering
+  \includegraphics[width=\textwidth,trim={0 .5in 0 .3in}]{Mean_TT_Ratio.pdf}
   \caption{These box plots show the prediction error of mean with
     increasing amounts of training data provided to the models. Notice
-    that MARS is the only model whose average spread of performance
-    decreases with more training data. Recall that the response values
+    that MARS is the only model whose primary spread of performance
+    increases with more training data. Recall that the response values
     being predicted span three orders of magnitude and hence relative
@@
     box whisker goes from around 100 to 50 from left to right and is
-    truncated in order to maintain focus on more performant models.}
+    truncated in order to maintain focus on better models. Results for
+    all dimensions are aggregated. Max training percentage is 96\% due
+    to rounding training set size.}
   \label{fig:mean_tt_ratio}
@@
   \centering
-  \includegraphics[width=\textwidth,trim={0 .5in 0 .5in}]{Mean_Dim.pdf}
-  \caption{These box plots show the prediction error of mean with
-    increasing dimension. The top box whisker for SVR transitions
-    \{40, 80, 90\} for dimensions 2, 3, and 4 respectively. Notice
-    that each model consistently experiences greater magnitude error
-    with increasing dimension.}
-  \label{fig:mean_dim}
+  \includegraphics[width=\textwidth,trim={0 .5in 0 .3in}]{Var_TT_Ratio.pdf}
+  \caption{These box plots show the prediction error of variance with
+    increasing amounts of training data provided to the models. The
+    response values being predicted span six orders of magnitude. For
+    SVR the top box whisker goes from around 6000 to 400 (decreasing
+    by factors of 2) from left to right and is truncated in order to
+    maintain focus on better models. Results for all dimensions are
+    aggregated. Max training percentage is 96\% due to rounding
+    training set size.}
+  \label{fig:var_tt_ratio}
 \end{figure}
 
+\vspace{-10pt}
 \subsection{I/O Throughput Variance}
+\vspace{-10pt} The prediction results for variance resemble those for
+predicting mean. Delaunay remains the best overall predictor
+(aggregated across training percentages and dimensions) with median
+relative error of .47 and LSHEP closely competes with Delaunay having
+a median signed relative error of -.92. Outliers in prediction error
+are much larger for all models. Delaunay produces relative errors as
+large as 78 and other models achieve relative errors around
+$10^3$. The relative errors for many models scaled proportional to the
+increased orders of magnitude spanned by the variance response
+compared with mean response. As can be seen in Figure
+\ref{fig:var_tt_ratio}, all models are more sensitive to the amount of
+training data provided than their counterparts for predicting mean.
 
-The prediction results for variance resemble those for predicting
-mean. Delaunay remains the best overall predictor with relative error
-of .6 on average and LSHEP closely competes with Delaunay having an
-average relative error of -2. Outliers in prediction error are much
-larger for all models. Delaunay produces relative errors as large as
-78 and other models achieve relative errors around $10^3$. The
-relative errors for many models scaled proportional to the increased
-orders of magnitude spanned by the variance response compared with
-mean response. All models are more sensitive to the amount of training
-data provided than their counterparts for predicting mean.
-
+\vspace{-10pt}
 \subsection{Increasing Dimension and Decreasing Training Data}
-
+\vspace{-10pt}
 As can be seen in Figure \ref{fig:mean_dim}, all of the models suffer
@@
 multivariate modeling techniques can be used to effectively predict
-HPC system performance. Little effort on the part of a systems
-engineer combined with a minimal amount experimentation can yield a
-model capable of accurately tuning an HPC system to the desired
-performance specification.
+HPC system performance. Some modeling effort on the part of a systems
+engineer combined with a significant amount of experimentation (weeks
+of CPU time for the IOzone data used here) can yield a model capable
+of accurately tuning an HPC system to the desired performance
+specification, although qualitatively correct predictions can be
+achieved with mush less (10\%, say) effort.
 
+\vspace{-10pt}
 \subsection{Modeling the System}
-
+\vspace{-10pt}
 The modeling techniques generated estimates of drastically different
-quality when predicting I/O throughput mean and sample variance. A few
-notable characteristics include: the SVR has the largest range of
-errors for all selections of dimension and amount of training data;
-MARS and LSHEP produce similar magnitude errors while the former
-consistently underestimates and the latter consistently overestimates
-in predictions; and Delaunay has considerably fewer outliers than all
-other methods. SVR likely produces the poorest quality predictions
-because the underlying parametric representation is global and
-oversimplified (a single polynomial), making it unable to capture the
-complex local behaviors of system I/O. It is still unclear however
-what causes the behaviors of LSHEP, MARS, and Delaunay. An exploration
-of this topic is left to future work.
+quality when predicting I/O throughput mean and variance. A few
+observations: SVR has the largest range of errors for all selections
+of dimension and amounts of training data; MARS and LSHEP produce
+similar magnitude errors while the former consistently underestimates
+and the latter consistently overestimates; Delaunay has considerably
+fewer outliers than all other methods. SVR likely produces the poorest
+quality predictions because the underlying parametric representation
+is global and oversimplified (a single polynomial), making it unable
+to capture the complex local behaviors of system I/O. It is still
+unclear, however, what causes the behaviors of LSHEP, MARS, and
+Delaunay. An exploration of this topic is left to future work.
 
 While the Delaunay method appears to be the best predictor in the
-presented IOzone case study, it is important to note its significantly
-higher computational complexity with respect to the dimension of the
-input. The implementation of Delaunay used would experience
-unacceptably large training times beyond seven dimensions of
-input. This leaves much room for other techniques to perform best in
-higher dimension unless a more performant implementation of Delaunay
-can be used.
+present IOzone case study, it is important to note that the Delaunay
+computational complexity increases with the dimension of the input
+more rapidly than other techniques. The implementation of Delaunay
+(QuickHull) used would experience unacceptably large training times
+beyond ten-dimensional input. This leaves much room for other
+techniques to perform best in higher dimension unless a more efficient
+implementation of Delaunay can be used.
 
 Finally, the ability of the models to predict variance was
-significantly worse for I/O sample variance. The larger scale in
+significantly worse than for the I/O mean. The larger scale in
 variance responses alone do not account for the increase in relative
 errors witnessed. This suggests that system variability has a greater
-underlying functional complexity when compared with the system mean
-and that latent factors are reducing prediction performance.
+underlying functional complexity than the system mean and that latent
+factors are reducing prediction performance.
 
+\vspace{-10pt}
 \subsection{Extending the Analysis}
-
-System I/O throughput mean and throughput sample variance are simple
-and useful system characteristics to model. The process presented in
-this work is equally applicable to predicting other useful performance
+\vspace{-10pt}
+System I/O throughput mean and variance are simple and useful system
+characteristics to model. The process presented in this work is
+equally applicable to predicting other useful performance
 characteristics of HPC systems such as: computational throughput,
@@
 chosen variance as a measure of variability, but the techniques used
-in this paper could be used to predict more precise measures of
-variability such as the percentiles of the performance distribution. A
-thorough exploration of HPC systems applications of multivariate
-modeling is left to future work.
+in this paper could be applied to more precise measures of variability
+such as the percentiles of the performance distribution or the entire
+distribution itself. A thorough exploration of HPC systems
+applications of multivariate modeling constitutes future work.
 
@@
 \label{sec:conclusion}
-Multivariate models of HPC systems produce promising results when
-applied to predicting I/O throughput mean and variability. These
-multivariate techniques significantly expand the comprehensiveness and
-portability of statistical models for predicting computer system
-performance over previous works. In the IOzone case study presented,
-the Delaunay method produces the best overall results making
-predictions for 821 system configurations with less than 5\% error
-when trained on only 43 configurations. Analysis also suggests that
-the error in the Delaunay method will remain manageable as the number
-of system parameters being modeled increases. These multivariate
-techniques can and should be applied to HPC systems with more than
-four tunable parameters in order to identify optimal system
-configurations that may not be discoverable via previous methods nor
-by manual performance tuning.
+Multivariate models of HPC system performance can effectively predict
+I/O throughput mean and variance. These multivariate techniques
+significantly expand the scope and portability of statistical models
+for predicting computer system performance over previous work. In the
+IOzone case study presented, the Delaunay method produces the best
+overall results making predictions for 821 system configurations with
+less than 5\% error when trained on only 43 configurations. Analysis
+also suggests that the error in the Delaunay method will remain
+acceptable as the number of system parameters being modeled
+increases. These multivariate techniques can and should be applied to
+HPC systems with more than four tunable parameters in order to
+identify optimal system configurations that may not be discoverable
+via previous methods nor by manual performance tuning.
 
@@
 %=====================
+\vspace{-10pt}
 \subsection{Future Work}
+\vspace{-10pt}
 The most severe limitation to the present work is the restriction to
-modeling strictly ordinal (not categorical) system parameters. This
-could be seen as a limitation of the multivariate models selected,
-because only MARS is capable of handling non numeric components, or as
-a limitation of the MDA framework. Future work could attempt to
-identify the viability of different techniques for making predictions
-across varying categorical system paramters.
+modeling strictly ordinal (not categorical) system
+parameters. Existing statistical approaches for including categorical
+variables are inadequate for nonlinear interactions in high
+dimensions. Future work could attempt to identify the viability of
+different techniques for making predictions including categorical
+system parameters.
 
@@
 parameters studied in this work. Alongside experimentation with more
-models, there is room for a theoretical characterisation of the
+models, there is room for a theoretical characterization of the
 combined model and data properties that allow for the greatest

### 2017-12-11 Code_[2017]_Algorithm_Comparison/py_nearest.py

@@ -1,0 +1,130 @@
+import random
+import numpy as np
+from scipy.optimize import minimize
+
+SMALL_NUM = .00001
+BIG_NUM = 1000000
+
+L1 = lambda xs,x: np.sum(abs(xs-x),axis=1)
+L2 = lambda xs,x: np.sqrt(np.sum((xs-x)**2,axis=1))
+L3 = lambda xs,x: (np.sum(abs(xs-x)**3,axis=1))**(1/3)
+L4 = lambda xs,x: (np.sum((xs-x)**4,axis=1))**(1/4)
+L5 = lambda xs,x: (np.sum(abs(xs-x)**5,axis=1))**(1/5)
+L6 = lambda xs,x: (np.sum((xs-x)**6,axis=1))**(1/6)
+L2SQ = lambda xs,x: np.sum((xs-x)**2, axis=1)
+
+LINEAR = lambda r: max(0,(1+SMALL_NUM) - r)
+QUADRATIC = lambda r: max(0,(1+SMALL_NUM) - r)**2
+
+BASIS_SIZE = 2.0
+
+class Linn:
+    # Initialize the Linn regressor.
+    def __init__(self, radius=LINEAR, distance=L2, basis_size=BASIS_SIZE):
+        self.radius = radius
+        self.distance = distance
+        self.basis_size = basis_size
+
+    # Fit a set of points
+    def fit(self, points, values):
+        self.points = points.copy()
+        self.values = values.copy()
+
+    # Wrapper for 'predict' that returns a single value for a single
+    # prediction, or an array of values for an array of predictions
+    def __call__(self, x:np.ndarray, *args, **kwargs):
+        single_response = len(x.shape) == 1
+        if single_response:
+            x = np.array([x])
+        if len(x.shape) != 2:
+            raise(Exception("ERROR: Bad input shape."))
+        response = np.asarray(self.predict(x, *args, **kwargs), dtype=float)
+        # Return the response values
+        return response[0] if single_response else response
+
+
+    # Generate a prediction for a new point
+    def predict(self, xs):
+        predictions = []
+        for x in xs:
+            # nearest_ind = 0 # np.argmin(self.distance(self.points, x))
+            weights = []
+            for nearest_ind in range(len(self.points)):
+                nearest = self.points[nearest_ind]
+                dist_to_point = self.distance([nearest], x)[0]
+                # Get the unit-length vector from (nearest -> x)
+                vector = (x - nearest)
+                vector /= np.sum(vector**2)**.5
+
[…]

### 2017-12-11 Code_[2017]_Algorithm_Comparison/py_test_plot.py

@@ -7,49 +7,48 @@ import numpy as np
 from util.plotly import *
 from py_settings import *
 
-from nearest import Linn
-
 # Preparing the data
 normalize_points = True
 random_points = True
-random_perturbation = True
+random_perturbation = False
 tight_plot = False
 plot_results = True
+autoshow_plot = True
 perform_test = False
 
 perturbation_multiplier = 1.0
 plot_points = 4000
-num_points = 4
+num_points = 10
 dim = 2
 multiplier = 10
+rand_seed = 6 # 2
 
 # from sklearn.linear_model import RANSACRegressor as sklearn_alg
 # class Alg(sklearn_alg):
 #     def __call__(self,x):
 #         return self.predict(x)
 
-# algorithms = [Alg()]
-# algorithms = [BSB()]
-# algorithms = [FitBoxMesh()]
-# algorithms = [MaxBoxMesh()]
-# algorithms = [SplineMesh()]
-# algorithms = [NearestNeighbor()]
-# algorithms = [LSHEP()]
-# algorithms = [LinearModel()]
-# algorithms = [Delaunay()]
-# algorithms = [ MLPRegressor() ] 
-# algorithms = [ SVR() ] 
-algorithms = [ Linn() ]
+algorithms = []
+# algorithms += [ Alg() ]
+algorithms += [ SVR() ] 
+algorithms += [ MARS() ]
+algorithms += [ LSHEP() ]
+algorithms += [ Delaunay() ]
+# algorithms += [ FitBoxMesh() ]
+algorithms += [ MaxBoxMesh() ]
+# algorithms += [ LinearModel() ]
+algorithms += [ VoronoiMesh() ]
+algorithms += [ MLPRegressor() ] 
+# algorithms += [ NearestPlane() ]
+algorithms += [ NearestNeighbor() ]
 
-# algorithms = [NearestPlane()]
-
-np.random.seed(2)
+np.random.seed(rand_seed)
 
 # Generate testing points in a grid pattern
 plot_range = [[-0.3,1.3]]*dim # [[0,1]]*dim # 
 if random_points:
     # Generate testing points randomly spaced
-    points = np.random.random(size=(num_points**dim,dim)) * (num_points-1)*num_points
+    points = np.random.random(size=(num_points**dim,dim)) * num_points
 else:
     points = np.meshgrid(*[np.linspace(0,num_points-1,num_points) for i in range(dim)])
[…]

### 2018-01-08 Paper_[2017-11]_HPC/paper.tex
Finished submitted draft of ACM SE paper.

@@
 data points in the sphere's interior. For a $d$-simplex S with
-vertices $v^{(0)}, v^{(1)}, \ldots, v^{(d)}$, $x \in S$, and data
-values $f(v^{(i)})$, $i=0$,$\ldots$,$d$, $x$ is a unique convex
+vertices $v^{(0)}$, $v^{(1)}$, $\ldots$, $v^{(d)}$, $x \in S$, and
+data values $f(v^{(i)})$, $i=0$, $\ldots$, $d$, $x$ is a unique convex
 combination of the vertices,
@@
   \subset \{1, 2, \ldots, d\}$, reduce the input data to points $(z,
-  f_F(z))$ with $z \in \mathbb{R}^k$ and $f_F(z) = E[ \{ f(x^{(i)})
-    \mid (x^{(i)}_F = z) \} ]$, where $E[\cdot]$ denotes the mean and
-  $x^{(i)}_F$ is the subvector of $x^{(i)}$ indexed by $F$.
+  f_F(z))$ with $z \in \mathbb{R}^k$ and $f_F(z) = E\bigl[ \bigl\{
+    f\bigl(x^{(i)}\bigr) \bigm| \bigl(x^{(i)}_F = z\bigr) \bigr\}
+    \bigr]$, where $E[\cdot]$ denotes the mean and $x^{(i)}_F$ is the
+  subvector of $x^{(i)}$ indexed by $F$.
 \item For all $r$ in $\{5, 10, \ldots, 95\}$, generate $N$ random
@@
 \item Generate a sequence of all pairs of points
-  ${(z^{(i_1)},z^{(j_1)}), (z^{(i_2)},z^{(j_2)}), \ldots}$ sorted by
-  ascending pairwise Euclidean distance between points, so that
-  $||z^{(i_k)}-z^{(j_k)}||_2 \leq ||z^{(i_{k+1})}-z^{(j_{k+1})}||_2$.
+  $\bigl(z^{(i_1)},z^{(j_1)}\bigr), \bigl(z^{(i_2)},z^{(j_2)}\bigr),
+  \ldots$ sorted by ascending pairwise Euclidean distance between
+  points, so that $\bigl|\bigl|z^{(i_k)}-z^{(j_k)}\bigr|\bigr|_2 \leq
+  \bigl|\bigl|z^{(i_{k+1})}-z^{(j_{k+1})}\bigr|\bigr|_2$.
 \item Sequentially remove points from candidacy until only $|train|$
-  remain by randomly selecting one point from the pair $(z^{(i_m)},
-  z^{(j_m)})$ for $m = 1,\ldots$ if both $z^{(i_m)}$ and $z^{(j_m)}$
-  are still candidates for removal.
+  remain by randomly selecting one point from the pair
+  $\bigl(z^{(i_m)}, z^{(j_m)}\bigr)$ for $m = 1,\ldots$ if both
+  $z^{(i_m)}$ and $z^{(j_m)}$ are still candidates for removal.
 \end{enumerate}
@@
 Given the large number of constraints, level of reduction, and use of
-randomness in the MDA procedure, occasionally $N$ unique training /
-testing splits may not be created or may not exist. In these cases, if
-there are fewer than $N$ possible splits, then deterministically
-generated splits are used. Otherwise after $3N$ attempts, only the
-unique splits are kept for analysis. The MDA procedure has been
-implemented in Python\#3 while most regression and interpolation
-methods are Fortran wrapped with Python. All randomness has been
-seeded for repeatability.
+randomness in the MDA procedure, occasionally $N$ unique
+training/testing splits may not be created or may not exist. In these
+cases, if there are fewer than $N$ possible splits, then
+deterministically generated splits are used. Otherwise after $3N$
+attempts, only the unique splits are kept for analysis. The MDA
+procedure has been implemented in Python\#3 while most regression and
+interpolation methods are Fortran wrapped with Python. All randomness
+has been seeded for repeatability.
 
@@
 achieving relative errors strictly less than $10^3$ for throughput
-mean and $10^6$ for throughput sample variance. Ideally, models will
+mean and $10^6$ for throughput variance. Ideally, models will
 yield relative errors significantly smaller than 1.
@@
 HPC system performance. Some modeling effort on the part of a systems
-engineer combined with a significant amount of experimentation (weeks
+engineer combined with a significant amount of experimentation (days
 of CPU time for the IOzone data used here) can yield a model capable

### 2018-01-08 Paper_[2018-01]_ACMSE/meshes_body.tex
Finished submitted draft of ACM SE paper.

\section{Introduction}
\label{sec_introduction}

Regression and interpolation are problems of considerable importance that find applications across many fields of science. Pollution and air quality analysis \cite{de2008field}, energy consumption management \cite{lazos2014optimisation}, and student performance prediction \cite{cortez2008using} are a few examples of interdisciplinary applications of multivariate regression for predictive analysis. As discussed later, these techniques can also be applied to prediction problems related to high performance computing (HPC) file input/output (I/O), Parkinson's patient clinical evaluations \cite{tsanas2010accurate}, and forest fire risk assessment \cite{cortez2007data}.

Multivariate interpolation is formally defined when there exists some function $f:\mathbb{R}^d \rightarrow \mathbb{R}$ and a set $X$ of $n$ points in $\mathbb{R}^d$ along with associated response values $f(x)$ for all $x \in X$. The problem is to construct an approximation $\hat f: \mathbb{R}^d \rightarrow \mathbb{R}$ such that $\hat f(x) = f(x)$ for all $x \in X$. It is often the case that the form of the true underlying function $f$ is unknown, however it is still desirable to construct an approximation $\hat f$ with minimum approximation error at $y \notin X$.

Multivariate regression is often used when the underlying function is presumed to be stochastic, or stochastic error is introduced in the evaluation of $f$. Hence, multivariate regression relaxes the conditions of interpolation by minimizing the error in $\hat f$ at $x \in X$ while maintaining some parametric form with parameters $P$. This can be written as $\min_{P} ||\hat f(X) - f(X)||$, where $f(X)$ is a vector of $f(x)$ for all $x \in X$ and $||\cdot||$ is an appropriate measure. The difficult question in the case of regression is often what parametric form to adopt for any given application. This paper proposes basis functions with overlapping regions of support as the general parametric form.

As the dimension of data increases, the number of possible interactions between dimensions grows exponentially. Quantifying all possible interactions becomes intractable and hence beyond three-dimensional data, mostly linear models are used. That is not to say nonlinear models are absent, but nonlinearities are often either preconceived or model pairwise interactions between dimensions at most. Similarly, higher order models of single variables can be generated from the techniques proposed in this paper, however only first-order interactions between variables are considered.

Regression and interpolation have a considerable theoretical base in one dimension \cite{cheney2009course}. Splines in particular are well understood as an interpolation technique in one dimension \cite{de1978practical}, particularly B-splines. Fewer techniques have been discussed and evaluated for two dimensional problems. Though in two and three dimensions tensor product splines remain computable \cite{unther1996interpolating}, tensor products have an unfortunate exponential scaling in parameterization with increasing dimension. Exponential scaling prohibits tensor products from being reasonably applied beyond three-dimensional data. In order to address this dimensional scaling challenge, C. de Boor and others have recently proposed box splines \cite{de2013box}. Two of the three meshes introduced in this work involve box-shaped regions of support and use box splines as the underlying basis functions.

% Art is a subjective form of personal expression and computers are just robots that are going to take over the world. \textit{*deep breath*} In order to communicate my frustration with the technological and information revolution, I'm going to write the rest of this paper in poetry.

% Star light, star bright \ldots

\subsection{Box Splines}
\label{sec_box_splines}

\begin{figure}
  \includegraphics[width=0.23\textwidth,height=2.5cm]{1D-linear.pdf}
  \includegraphics[width=0.23\textwidth,height=2.5cm]{1D-quadratic.pdf}
  \caption{1D linear (order 2) and quadratic (order 3) box splines with direction vector sets $\bigl( 1 \ 1 \bigr)$ and $\bigl( 1 \ 1 \ 1 \bigr)$ respectively. Notice that these direction vector sets form the B-Spline analogues, order 2 composed of two linear components and order 3 composed of 3 quadratic components (colored in plot).
  \vspace{-.5cm}}
  \label{fig_1D_boxes}
\end{figure}

A box spline in $\mathbb{R}^d$ is defined by its \textit{direction vector set} $A$, composed of $s$ $d$-vectors where $s \geq d$. Further, $A$ will be written as a $d \times s$ matrix. The first $m$ column vectors of $A$ are denoted by $A_m$, $m \leq s$. $A_d$ is required to be nonsingular. Consider the unit cube in $s$ dimensions $Q_s = [0,1)^s$. $A_s \bigl( Q_s \bigr)$ is now the image (in $d$ dimensions) of $Q_s$ under the linear map $A$. This image is the region of support for the box spline defined by $A_s$ in $d$ dimensions. The box spline function in $d$ dimensions for $A_d$ is defined as
\begin{equation}
B(x \mid A_d) = \begin{cases} 
(\det(A_d))^{-1}, & x \in A_d(Q_d), \\
0,                & \text{otherwise.}
\end{cases}
\label{eq_box_base}
\end{equation}
For $A_s$ when $s > d$ the box spline is computed as
\begin{equation}
B(x \mid A_s) = \int_0^1 B(x - t v_s \mid A_{s-1}) dt,
\label{eq_box_recursive}
\end{equation}
where $v_s$ is the $s$th direction vector of $A$.

The application of box splines presented in this paper always utilizes the $d$-dimensional identity matrix as $A_d$. This simplifies the computation in Equation \ref{eq_box_base} to be the indicator function over the unit cube. Composing $A$ strictly out of $k$ repetitions of the identity matrix forms the $k$th order B-spline with knots located at $0$, $1$, $\ldots$, $k-1$, $k$ along each axis (see Figure \ref{fig_1D_boxes}). Furthermore, while the number of subregions for the $k$th order $d$-dimensional box spline grows as $k^d$ (see Figure \ref{fig_2D_boxes}), the symmetry provided by direction vector sets composed of repeated identity matrices allows the computation of box splines to be simplified. The value of a box spline at any location is then the product of all axis-aligned 1-dimensional $k$th order box splines.

The box splines as presented are viable basis functions. Each box spline can be shifted and scaled without modifying the underlying computation (similar to wavelets), yet the underlying computation is simple and scales linearly with dimension. For a more thorough introduction and exploration of box splines in their more general form, readers are referred to \cite{de2013box}.

\begin{figure}
  \vspace{-.5cm}
  \includegraphics[width=0.23\textwidth,height=3cm]{2D-linear.pdf}
  \includegraphics[width=0.23\textwidth,height=3cm]{2D-quadratic.pdf}
  \caption{2D linear (order 2) and quadratic (order 3) box splines with direction vector sets $\bigl( I \ I \bigr)$ and $\bigl( I \ I \ I \bigr)$ respectively, where $I$ is the identity matrix in two dimensions. Notice that these direction vector sets also produce boxes with $\text{order}^2$ subregions (colored in plot).
  \vspace{-0.1cm}}
  \label{fig_2D_boxes}
\end{figure}

\section{Interpolation and Regression}
\label{sec_mesh_construction}

Throughout this section, the same notation will be used as introduced in Section \ref{sec_introduction}. $X \subset \mathbb{R}^d$ is a finite set of points with known response values $f(x)$ for all $x \in X$.

Define a box $b^c = (c,l^c,u^c)$ in $d$ dimensions with center $c \in \mathbb{R}^d$, lower width vector $l^c \in \mathbb{R}^d_+$, and upper width vector $u^c \in \mathbb{R}^d_+$ (where $u^c_i$ refers to the $i$th component of $u^c$). A visual example of a box in two dimensions can be seen in Figure \ref{fig_example_box}. Now, define a componentwise rescaling function $g: \mathbb{R}^d \rightarrow \mathbb{R}^d$ at point $x \in \mathbb{R}^d$ to be
\begin{equation}
  \bigl(g^c(x)\bigr)_r = {k \over 2} \left( 1 + {(x_r - c_r)_- \over l^c_r} - {(x_r - c_r)_+ \over u^c_r} \right)
\end{equation}
where $y_+=\max\{y,0\}$, $y_-=(-y)_+$, $k$ is the order of the box spline as described in Section \ref{sec_box_splines}. Finally, each box spline in a box mesh can be evaluated as $B(g^c(x) \mid A)$ presuming the order of approximation implies $A$. Both box meshes described in the following subsections use box spline basis functions of this form.

A notable property of boxes defined with the linear rescaling function $g^c$, is that $C^0$ and $C^1$ continuity of the underlying box spline are maintained. $C^0$ continuity is maintained through scaling. $C^1$ continuity is maintained for all box splines with $C^1$ continuity (order $\geq 3$) because the scaling discontinuity is located at $c$, where all box splines (of the presented form) have first derivative zero. All continuity beyond the first derivative is lost through the rescaling function $g^c$.

\begin{figure}
  \centering
  \begin{tikzpicture}[scale=1]
    \draw[thick] (3,1.1) circle (0.7mm);

    %% Draw arrows for each width
    \draw[thick,->] (3,1.2) -- (3,2.4);
    \draw[thick,->] (3,1.0) -- (3,0.1);
    \draw[thick,->] (3.1,1.1) -- (6.9,1.1);
    \draw[thick,->] (2.9,1.1) -- (0.1,1.1);

    %% Draw boundary of box
    \draw[thick,-] (0,0) -- (7,0);
    \draw[thick,-] (7,0) -- (7,2.5);
    \draw[thick,-] (7,2.5) -- (0,2.5);
    \draw[thick,-] (0,2.5) -- (0,0);

    %% Add text to the picture
    \node at (3.25,1.35) {$c$};
    \node at (1.3,1.35) {$l^c_1$};
    \node at (5.1,1.35) {$u^c_1$};
    \node at (2.75,0.55) {$l^c_2$};
    \node at (2.75,1.8) {$u^c_2$};
  \end{tikzpicture}
  \caption{An example box in two dimensions with center $c$, upper widths $u^c_1$, $u^c_2$, and lower widths $l^c_1$, $l^c_2$. Notice that $c$ is not required to be equidistant from opposing sides of the box, that is $u^c_i \not= l^c_i$ is allowed.}
  \label{fig_example_box}
\end{figure}

\subsection{Max Box Mesh}

The first of the three meshes produces a set of boxes around chosen control points and each box has maximal distance between the control point and the nearest side of the box. This centrality property is one mechanism of creating the largest reasonable regions of support for the underlying basis functions. The individual boxes are constructed via the following procedure given a set of control points $C \subseteq X$, $c^{(i)} \in C$,
\begin{enumerate}
  \item Initialize a box $b^{c^{(1)}} \supset X$ with center $c^{(1)}$. \label{step_init}
  \item Identify $c^{(i)}$ over $\bigl\{ j \bigm| j \ne 1$, $B^{c^{(1)}}\bigl( c^{(j)} \bigr) \ne 0 \bigr\}$ that minimizes $\bigl\|c^{(j)}-c^{(1)}\bigr\|_\infty$.  \label{step_closest}
  \item Change the the box $b^{c^{(1)}}$ along the first dimension $r$ such that $\bigl|\bigl|c^{(1)} - c^{(i)}\bigr|\bigr|_\infty = \bigl\vert c^{(1)} - c^{(i)}\bigr\vert_r $, to exclude $c^{(i)}$ from the support of $B^{c^{(1)}}$. \label{step_shrink}
  \item Repeat steps \ref{step_closest} and \ref{step_shrink} until no point in $C$ is in the support of $B^{c^{(1)}}$ (at most $2d$ times, once for each boundary of a box).
\end{enumerate}

The same process is used to construct boxes around all control points in $C$. In order to improve the generality of the approximation, a set of control points is initially chosen to be well-spaced using a statistical method from \cite{amos2014algorithm}:

\begin{enumerate}
\item Generate a sequence of all pairs of points sorted by ascending pairwise Euclidean distance between points $\bigl(x^{(i_1)},x^{(j_1)}\bigr)$, $\bigl(x^{(i_2)},x^{(j_2)}\bigr)$, $\ldots$ , so that $\bigl\|x^{(i_k)}-x^{(j_k)}\bigr\|_2 \leq \bigl\|x^{(i_{k+1})}-x^{(j_{k+1})}\bigr\|_2$.
\item Sequentially remove points from candidacy until only $|C|$ remain by randomly selecting a single point from each pair $\bigl(x^{(i_m)}, x^{(j_m)}\bigr)$ for $m = 1,\ldots$ if both $x^{(i_m)}$ and $x^{(j_m)}$ are still candidates for removal.
\end{enumerate}

Once the boxes for a max box mesh have been constructed, the parameters can be identified via a least squares fit. The max box mesh (denoted $MBM$) is used to generate a $|X| \times |C|$ matrix $M$ of box spline basis function evaluations at all points in $X$. The solution to the least squares problem $\min_P \bigl\| M \ P - f(X) \bigr\|_2$ is the parameterization of $MBM$. When $C = X$, $M$ is the $|X| \times |X|$ identity matrix, making the max box mesh approximation $\hat f$ an interpolant.

While setting the number of boxes equal to the number of points causes the max box mesh to be an interpolant, the generality of the max box mesh approximation can often be improved by bootstrapping the selection of control points. Given a user-selected batch size $s \leq |X|$, start with $s$ well-spaced control points. Next, measure the approximation error at $x \notin C$ and if the error is too large (determined by user), pick $s$ points at which the magnitude of approximation error is largest, add those points to $C$, and recompute the max box mesh. The user is left to decide the batch size $s$ and the error tolerance based on validation performance and computability. Next, the max box mesh is shown to be nearly space filling.

\begin{definition}
A box $b^y$ is the upper \textit{boundary} of box $b^x$ along dimension $r$ if and only if $u^x_r = (y_r - x_r)$. A box $b^y$ is the lower \textit{boundary} of box $b^x$ along dimension $r$ if and only if $l^x_r = (x_r - y_r)$.
\end{definition}

\begin{definition}
A \textit{proper} upper boundary for box $b^x$ at point $y$ occurs when $u^x_r = ||x - y||_\infty$ and $r$ is the first dimension such that $u^x_r = ||x - y||_\infty$, assuming dimensions are deterministically ordered. The same definition applies to lower boundaries.
\end{definition}

\begin{theorem}
A max box mesh composed of $n$ unique control points in $d$ dimensions, $C \subset \mathbb{R}^d$ where all boundaries are proper, is nearly space filling.
\label{theorem_mutual_space_filling}
\end{theorem}
\begin{proof}
Pick an arbitrary point $x \in \mathbb{R}^d$. Now let $c^{(1)}$ be the $c \in C$ that minimizes $|| c - x ||_\infty$, making the nearest box (by max norm) $b^{c^{(1)}}$. Either $x$ is in the support of $b^{c^{(1)}}$ or on the boundary of $b^{c^{(1)}}$ and this is done, or $b^{c^{(1)}}$ has a proper boundary along dimension $r$ with box $b^{c^{(2)}}$ such that $|c^{(2)}_r - x_r| < |c^{(1)}_r - x_r|$. The sequence of bounds $c^{(2)}$, $\ldots$, must be finite for the strict inequality to hold and the last box in this sequence supports $x$. Therefore, the $MBM$ is nearly space filling.
\end{proof}

The max box mesh as proposed is still capable of producing coefficients of 0 for $x \in \mathbb{R}^d$ placed along the boundary of boxes, however this problem is solved by increasing the upper and lower widths of all boxes by the unit roundoff during computation. For this reason, the max box mesh remains a viable strategy for computed approximations. Given a maximum of $c$ control points in $d$ dimensions with $n$ points, the computational complexities are: $\mathcal{O}(c^2 d)$ for computing boxes, $\mathcal{O}(c d^2 + d^3)$ for a least squares fit, and $\mathcal{O}(n / s)$ for bootstrapping (which is multiplicative over the fitting complexities). Evaluating the max box mesh requires $\mathcal{O}(c d)$ computations.

\subsection{Iterative Box Mesh}

The iterative box mesh ($IBM$) composes box-shaped regions of support over control points that each support strictly one control point just as in the $MBM$, however the mesh is completely space filling by construction and places boxes in a way that reduces apparent error. The boxes are constructed via the following procedure given a finite set of points $X \subset \mathbb{R}^d$,
\begin{enumerate}
\item Add a box with total support around the most central point $x^{(1)} \in X$ to $IBM$ and fit the mesh to all $X$.
\item Add a box centered at $x^{(i)} \notin IBM$ that has the most approximation error, reshaping all boxes in $b^{x^{(j)}} \in IBM$ that contain $x^{(i)}$ by bounding the first dimension $r$ such that $\bigl| x^{(j)}_r - x^{(i)}_r \bigr| = \bigl| \bigl| x^{(j)} - x^{(i)} \bigr| \bigr|_\infty$ and then fit the mesh to all $X$. \label{step_add_box}
\item Repeat Step \ref{step_add_box} until approximation error is below tolerance $t$.
\end{enumerate}

Just as the $MBM$, the parameters can be identified via a least squares solve. The iterative box mesh is used to generate a $|X| \times |C|$ matrix $M$ of box spline function evaluations at all points in $X$. Now the set of coefficients for each box is the solution to the least squares problem $\min_P \bigl|\bigl| M\ P - f(X) \bigr|\bigr|_2$. Also as for the $MBM$, $C = X$ causes $M$ to equal the $|X| \times |X|$ identity, making the iterative box mesh approximation $\hat f$ an interpolant.

As opposed to the max box mesh, the bootstrapping procedure is built into the iterative box mesh. The user is left to decide the most appropriate error tolerance, however a decision mechanism and analysis is presented in Section \ref{sec_performance_analysis}. Next, it is shown that the $IBM$ is space filling.

\begin{definition}
A \textit{mutual} boundary between boxes $b^x$ and $b^y$ is when there exists an $r$ such that $u^x_r = l^y_r$ or $l^x_r = u^y_r$.
\end{definition}

\begin{theorem} An iterative box mesh composed of $n$ unique control points in $d$ dimensions, $C \subset \mathbb{R}^d$, is space filling. \end{theorem}
\begin{proof}
The first box added to an $IBM$ is space filling by construction. Assume that an $IBM$ composed of $n$ boxes is space filling and a new box centered at $x \in \mathbb{R}^d$ is added. Consider a point $y \in \mathbb{R}^d$. Every box $b^{x^{(i)}}$ that supported $y$ before the addition of $b^x$ and was reshaped along dimension $r \in [1,d]$ has a mutual boundary with $b^x$ by construction. Now either $\bigl| x_r - x^{(i)}_r \bigr| \geq \bigl| y_r - x^{(i)}_r \bigr|$ and $b^{x^{(i)}}$ still supports $y$, or $\bigl| x_r - x^{(i)}_r \bigr| < \bigl| y_r - x^{(i)}_r \bigr|$ and $b^x$ supports $y$. Therefore, the $IBM$ composed of $n+1$ boxes is still space filling.
\end{proof}

After fitting for parameters, an $IBM$ can generate approximations for a set of new points $Z \subset \mathbb{R}^d$ by evaluating all box basis functions multiplied by the coefficients, $P$. The computational complexity for generating the mesh is $\mathcal{O}(c^2 n d)$ where $c$ is the number of control points determined by the minimum error threshold and $n = |X|$. The computational complexity of evaluating the mesh at a singple point is $\mathcal{O}(c d)$.

\subsection{Voronoi Mesh}

The final of the three meshes utilizes $\ell2$ norm distances to define boundaries rather than max norm distances. A well-studied technique for classification and approximation is the nearest neighbor algorithm \cite{cover1967nearest}. The convex region surrounding a central control point in which all points are closer to that center than any other control point is also called the Voronoi cell \cite{dirichlet1850reduction}. The Voronoi mesh ($VM$) utilizes Voronoi cells to define support via a generic basis function $v: \mathbb{R}^d \rightarrow \mathbb{R}$ with

\begin{equation}
v^c(z) = \left(1 - \frac{||z - c||_2}{2 \ dist(z - c \mid c)} \right)_+
\label{eq_voronoi_basis}
\end{equation}

where $c \in \mathbb{R}^d$ is the center of the Voronoi cell, $z \in \mathbb{R}^d$ is an interpolation point, and $dist(\ \cdot \mid c)$ is the distance between $c$ and the nearest boundary of the cell along the vector $\cdot\ $. While all basis functions $v^c(c) = 1$, the $2$ in the denominator causes all basis functions to go linearly to $0$ at the boundary of the twice-expanded Voronoi cell. Note that this basis function is $C^0$ because the boundaries of the Voronoi cell are $C^0$. In the case that there is no boundary along the vector $x-c$, the basis function value is always $1$.

While the cost of computing all Voronoi cells for any given set of points grows exponentially \cite{dutour2009complexity}, the calculation of $dist$ is linear with respect to the number of control points and dimensions. Given any center $c^{(1)} \in \mathbb{R}^d$, set of control points $C \subset \mathbb{R}^d$, and interpolation point $z \in \mathbb{R}^d$, $dist\bigl((z-c^{(1)}) \mid c^{(1)}\bigr)$ is the solution to
\begin{equation}
  \max_{\{c \in C\}\backslash c^{(1)}} \frac{|z|}{2} \ \frac{z \cdot \bigl(c - c^{(1)}\bigr) - c^{(1)} \cdot \bigl(c - c^{(1)}\bigr)}{c \cdot \bigl(c - c^{(1)}\bigr) - c^{(1)} \cdot \bigl(c - c^{(1)}\bigr)}.
\end{equation}

The parameters of the $VM$ can now be computed exactly as for the $MBM$ and $IBM$. The vornonoi mesh is used to generate a $|X| \times |C|$ matrix $M$ of basis function evaluations at all points in $X$. Now the set of coefficients for each basis function is the solution to the least squares problem $\min_P \bigl\| M \ P - f(X) \bigr\|_2$. When $X = C$, $M$ is the identity making the mesh an interpolant. Bootstrapping can be performed with an identical procedure to that of the $IBM$.
\begin{enumerate}
\item Pick the most central point $x^{(1)} \in X$ to be the first control point in $VM$ and fit the mesh to all $X$.
\item Add a control point $x^{(i)} \notin VM$ that has the most approximation error and then fit the mesh to all $X$. \label{step_add_control}
\item Repeat Step \ref{step_add_control} until approximation error is below tolerance $t$.
\end{enumerate}

Any $VM$ is na\"{\i}vely space filling, since any possible interpolation point will have a nearest neighbor control point when the set of control points is nonempty.

\begin{theorem}
  A Voronoi mesh composed of $n$ unique control points in $d$ dimensions, $C \subset \mathbb{R}^d$, is space filling.
\end{theorem}
\begin{proof}
  Consider interpolating at point $z$. Let $c^{(1)}$ be the $c \in C$ that minimizes $||z-c||_2$. Now, $z$ must be in the support of the basis function $v^{c^{(1)}}$. Therefore the Voronoi mesh is space filling.
\end{proof}

The computational complexity of evaluating a parameterized Voronoi mesh with $c$ control points is $\mathcal{O}(c^2 d)$. Bootrsapping the generation of a Voronoi mesh requres $\mathcal{O}(c^2 n d)$ computations for a maximum number of basis functions $c$ determined by the error threshold.

% \subsection{Boxed Natural Neighbor}
% \begin{itemize}
% \item How the mesh is generated
% \item How interpolation is done
% \end{itemize}

\section{Data and Analysis}
In order to evaluate the proposed interpolation and approximation techniques, this paper utilizes three data sets of varying dimension and application. In the following subsections the sources and targets of each data set are described, as well as challenges and limitations related to doing interpolation and approximation on these data sets. The distributions of response values being modeled can be seen in Figure \ref{fig_response_hists}. The preprocessing and approximation processes are described in Section \ref{sec_performance_analysis}.

\begin{figure}
  \includegraphics[width=.5\textwidth,height=2cm]{p-hist.pdf}
  \includegraphics[width=.5\textwidth,height=2cm]{f-hist.pdf}
  \includegraphics[width=.5\textwidth,height=2cm]{h-hist.pdf}
  \caption{Histograms of Parkinsons (total UPDRS), forest fire (area), and HPC I/O (mean throughput) response values respectively. Notice that both the forest fire and HPC I/O data sets are heavily skewed.
  \vspace{-.5cm}}
  \label{fig_response_hists}
\end{figure}

\subsection{Data Summary}

\subsubsection{High Performance Computing I/O}
$$N = 532, D = 4$$
The first of three data sets is a four-dimensional dataset produced by executing the IOzone benchmark from \cite{iozone} on a homogeneous cluster of computers. The system performance data was collected by executing IOzone 40 times for each of a select set of system configurations. A single IOzone execution reports the max I/O file-read throughput seen. The 40 executions for each system configuration are converted to their mean, which is capable of being modeled by each of the multivariate approximation techniques presented in Section \ref{sec_mesh_construction}. The four dimensions being modelled to predict throughput mean are file size, record size, thread count, frequency.

\subsubsection{Forest Fire}
$$N = 517, D = 12$$

The forest fire data set provided by \cite{cortez2007data}, describes the area of Montesinho park burned on specific days and months of the year in terms of the environmental conditions. The twelve dimensions being used to model burn area are the $x$ and $y$ spatial coordinates of burns in the park, month and day of year, the FFMC, DMC, DC, and ISI indices (see source for details), the temperature in celcius, relative humidity, wind speed, and outdoor rain. The original analysis of this data set demonstrated it to be difficult to model, likely due to the skew in response values.

\subsubsection{Parkinson's Telemonitoring}
$$N = 468, D = 16$$

The final data set for evaluation is derived from a speech monitoring study with the intent to automatically estimate Parkinson's disease symptom development in parkinsons patients from \cite{tsanas2010accurate}. Particularly, a time-consuming clincal evaluation measure referred to as the UPDRS score. The total UPDRS score given by a clinical evaluation is estimated through 16 real numbers generated from biomedical voice measures of in-home sound recordings.

\subsection{Performance Analysis}
\label{sec_performance_analysis}

\begin{figure*}[htb]
  \begin{tikzpicture}
    \node (img)  {\includegraphics[width=0.95\textwidth,height=9.5cm]{all-performance.pdf}};
    \node[below=of img, node distance=0cm, yshift=1cm] {Relative Error Tolerance while Bootstrapping};
    \node[left=of img, node distance=0cm, rotate=90, anchor=center,yshift=-0.7cm] {Signed Relative Error};
  \end{tikzpicture}
  \caption{The performance of all three techniques with varied relative error tolerance for the bootstrapping parameter. The columns are for Max Box Mesh, Iterative Box Mesh, and Voronoi Mesh, respectively. The rows are for HPC I/O, Forest Fire, and Parkinsons respectively. Notice that the techniques exibit counterintuitive behavior on the Parkinsons and Forest Fire data sets, performance increases with increasing error tolerance.}
  \label{fig_all_performance}
\end{figure*}

The performance of the approximation techniques varies considerably across the three evaluation data sets. Relative errors for the most na\"{\i}ve approximators such as nearest neighbor can range $\displaystyle [0, \big(\max_x f(x) - \min_x f(x)\big) / \min_x f(x) ]$ when modeling a nonnegative function $f(x)$ from data. Each of the approximation techniques presented remain within these bounds and all errors are presented in signed relative form $(\hat f(x) - f(x)) / f(x)$. Before the models are constructed all data $X$ values are shifted and scaled to be in the unit cube $[0,1]$, while the response values are taken in their original form. All models are evaluated with a 10-fold cross validation over 80:20 splits of the data.

\begin{table}
  \centering
  \begin{tabular}{c|c|c|c}
    \hline
    \textbf{Data Set} & \textbf{Technique} & \textbf{Tolerance} & \textbf{Average Errror}\\
    \hline
    HPC I/O & MBM & 1.2 & 0.597\\
    Forest Fire & MBM & 1.8 & 3.517\\
    Parkinsons & MBM & 0.6 & 0.114\\
    \hline
    HPC I/O & IBM & 0.4 & 0.419\\
    Forest Fire & IBM & 1.8 & 3.615\\
    Parkinsons & IBM & 1.8 & 0.121\\
    \hline
    HPC I/O & VM & 0.2 & 0.382\\
    Forest Fire & VM & 1.0 & 4.783\\
    Parkinsons & VM & 2.0 & 1.824\\
    \hline
  \end{tabular}
  \caption{The optimial error tolerance bootstrapping parameters for each technique and each data set as well as the average absolute relative errors achieved by that tolerance. Notice that large relative error tolerances occasionally yield even lower evaluation errors, demonstrating the benefits of approximation over interpolation for noisy data sets.
  \vspace{-1cm}}
  \label{tab_optimal_tolerance}
\end{table}

Each of the approximation techniques presented incorporates a bootstrapping based on an allowable error tolerance $t$. An analysis of the effects of bootstrapping error tolerances on validation accuracy can be seen in Figure \ref{fig_all_performance}. The approximation meshes perform best on the forest fire and Parkinson's data sets when the error tolerance used for fitting is large (regressing rather than interpolating), while near-interpolation generally produces the most accurate models for HPC I/O. Another performance result of note is that the $MBM$ and $IBM$ have very similar basis functions with largely different outputs.

\begin{figure*}[htb]
  \begin{tikzpicture}
    \node (img)  {\includegraphics[width=0.95\textwidth,height=9.5cm]{perf-sample.pdf}};
    \node[below=of img, node distance=0cm, yshift=1cm] {Signed Relative Error in Prediction};
    \node[left=of img, node distance=0cm, rotate=90, anchor=center,yshift=-0.7cm] {Count};
  \end{tikzpicture}
  \caption{A sample of relative errors for all three techniques with optimal selections of error tolerance. The columns are for Max Box Mesh, Iterative Box Mesh, and Voronoi Mesh, respectively. The rows are for HPC I/O, Forest Fire, and Parkinsons respectively. Notice that the Max Box Mesh exibits the least skewed distribution of errors across all three data sets.
    \vspace{-.3cm}}
  \label{fig_perf_sample}
\end{figure*}

The selection of bootstrapping error tolerance also effects the computation time required to fit each of the models to data. Figure \ref{fig_eval_times} presents the time required to construct approximations for each model and each data set with varying $t$. The rapid reducting in computation time required for the forest fire and HPC I/O data sets suggests that large reductions in error can be achieved with relatively few basis functions. The Parkinson's data set however presents a more noisy response, with increasing number of basis functions reducing error much less quickly.

The distributions of errors experienced by each approximation technique when the optimal bootstrapping relative error tolerance is selected can be seen in Figure \ref{fig_perf_sample}. HPC I/O exibits the most normal approximation errors, which suggests that the models are converging on the latent noise of the response for the data set. The worst relative approximation errors are produced by the Voronoi mesh on the forest fire data set. The small magnitude true response values contribute to the larger relative errors, however the $VM$ performances are still unacceptably large.

\begin{figure*}
  \begin{tikzpicture}
    \node (img)  {\includegraphics[width=0.95\textwidth,height=4cm]{eval-times.pdf}};
    \node[below=of img, node distance=0cm, yshift=1cm] {Relative Error Tolerance while Bootstrapping};
    \node[left=of img, node distance=0cm, rotate=90, anchor=center,yshift=-0.7cm] {Fit Time};
  \end{tikzpicture}
  \caption{Time required to generate model fits for each technique with varying relative error tolerance during bootstrapping.
    \vspace{-.3cm}}
  \label{fig_eval_times}
\end{figure*}

\section{Discussion}

The bootstrapping procdure presented for each approximation technique still has much room for improvement. Initial analysis suggests that the appropriate relative error tolerance needs to be discovered emperically for each application of a modelling technique. Further analytic studies could arrive at methods for determining optimal error tolerances at runtime, however increases in runtime complexity may not be afforded in many applications.

The box-shaped basis functions and the construction algorithms used for the $MBM$ and $IBM$ could become a source of error when $d$ (the dimension of the data $X$) is much larger than $n$ (the number of known points). The blending regions in which multiple basis functions overlap are always axis aligned and in applications such as image analysis, any single dimension may be unsuitable for approximating the true underlying function. The Voronoi mesh attempts to address this problem by utilizing boundaries between points in multiple dimensions simultaneously. However, it is empirically unclear whether true benefits are seen in applications where $d < n$.

Each of the case studies presented have fewer than $1000$ points. The complexities of the presented approximation techniques are suitable for large dimension, but the increased complexity associated with brute-force bootstrapping currently prohibits their use on larger data sets. The Voronoi mesh in particular has a large complexity with respect to $n$ which could be significantly improved via more greedy bootstrapping. While each technique requires less than ten seconds on average to produce a fit in the presented case studies, the fit time required quickly grows into minutes around $1000$ points. While these initial results appear somewhat limiting, they demonstrate the viability of each mesh and leave room for further theoretical exploration of techniques to reduce the runtime complexity while maintaining the approximation power and flexibility.

\section{Conclusion}

The Max Box Mesh, Iterative Box Mesh, and Voronoi Mesh each provide novel strategies for effectively approximating multivariate phenomonon. The underlying constructions are theoretically straightforward, yet powerful and flexible. The computational complexities of each make them particularly suitable for applications in many dimensions, while the bootstrapping error tolerance parameter allows a balance between approximation and interplation to be explored empirically with each application.

\subsection{Future Work}

A thorough comparison with constituent multivariate approximation techniques including but not limited to, Linear Shepard interpolation, Multivariate Adaptive Regression Splines, Multi Layer Perceptron regression, and Delaunay Triangulation constitutes future work. A more detailed study of alternative bootstrapping techniques may also provide valuable insight.

### 2018-01-08 Report_[2017-05-13]_Box_Spline_Initial_Analysis/box_mesh[6-8-2017 max-side-len verbose].py

@@ -1,0 +1,550 @@
+# WARNING: This 'max-side-len' box mesh construction code cannot
+#          handle data with shared x or y coordinates (allowing for
+#          more than 2*dimension neighbors)
+# WARNING: This 'max-side-len' box mesh construction code does not
+#          properly expand into higher dimension than x \in
+#          R^2. During the 'growing' phase, all dimensions need to be
+#          checked for expansion, not just the current study.
+
+import numpy as np
+from plotly_interface import *
+import os
+
+# Pre:  "x" is a 1 x N dimensional numpy array or list in the range [0,1]
+# Post: The value of the quadratic box spline in N dimensions
+#       is computed for the given point.
+def linear(x):
+    x *= 2
+    # Compute the box spline function value for the x-coordinate based
+    # on which region of the box that it is in
+    func_val = 1.0
+    for d in range(len(x)):
+        if (0 <= x[d] < 1):
+            func_val *= x[d]
+        elif (1 <= x[d] < 2):
+            func_val *= (2 - x[d])
+    return func_val
+
+# Pre:  "x" is a 1 x N dimensional numpy array or list in the range [0,1]
+# Post: The value of the quadratic box spline in N dimensions
+#       is computed for the given point.
+def quadratic(x):
+    x *= 3
+    # Compute the box spline function value for the x-coordinate based
+    # on which region of the box that it is in
+    func_val = 1 / 2**(x.shape[0])
+    for d in range(len(x)):
+        if (0 <= x[d] < 1):
+            func_val *= x[d]**2
+        elif (1 <= x[d] < 2):
+            func_val *= -(2*x[d]**2 - 6*x[d] + 3)
+        elif (2 <= x[d] <= 3):
+            func_val *= (x[d] - 3)**2
+    return func_val
+
+# Pre:  "x" is a 1 x N dimensional numpy array or list in the range [0,1]
+# Post: The value of the classic cubic box spline in N dimensions
+#       is computed for the given point.
+def cubic(x):
+    x *= 4
+    # Compute the box spline function value for the x-coordinate based
+    # on which region of the box that it is in
+    func_val = 1 / (2*3**(x.shape[0]))
+    for d in range(len(x)):
+        if (0 <= x[d] < 1):
+            func_val *= (x[d]**3)
+        elif (1 <= x[d] < 2):
+            func_val *= (4 - 12*x[d] + 12*x[d]**2 - 3*x[d]**3)
[…]

### 2018-01-21 Paper_[2018-01-09]_ACMSE/paper.tex
Added IEEE paper.

--- 2018-01-08 Paper_[2018-01]_ACMSE/meshes_body.tex
@@
   \caption{1D linear (order 2) and quadratic (order 3) box splines with direction vector sets $\bigl( 1 \ 1 \bigr)$ and $\bigl( 1 \ 1 \ 1 \bigr)$ respectively.
-Notice that these direction vector sets form the B-Spline analogues, order 2 composed of two linear components and order 3 composed of 3 quadratic components (colored in plot).
+Notice that these direction vector sets form the B-Spline analogues, order 2 composed of two linear components and order 3 composed of 3 quadratic components (colored and styled in plot).
   \vspace{-.5cm}}

### 2018-01-21 Paper_[2018-01-20]_IEEESE/paper.tex
Added IEEE paper.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%2345678901234567890123456789012345678901234567890123456789012345678901234567890
%        1         2         3         4         5         6         7         8

\documentclass[letterpaper, 10 pt, conference]{ieeeconf}  % Comment this line out
                                                          % if you need a4paper
%\documentclass[a4paper, 10pt, conference]{ieeeconf}      % Use this line for a4 paper
\IEEEoverridecommandlockouts                              % This command is only needed if you want to use the \thanks command
\overrideIEEEmargins
% See the \addtolength command later in the file to balance the column lengths on the last page of the document

% The following packages can be found on http:\\www.ctan.org
\usepackage{tikz}     % For adding axes to plots
\usepackage{graphicx} % For including PDFs
\usetikzlibrary{positioning}

%% \usepackage{graphics} % for pdf, bitmapped graphics files
%\usepackage{epsfig} % for postscript graphics files
%\usepackage{mathptmx} % assumes new font selection scheme installed
%\usepackage{times} % assumes new font selection scheme installed
\usepackage{amsmath} % assumes amsmath package installed
\usepackage{amssymb}  % assumes amsmath package installed

%% Command for creating multiplie-line cells inside a table environment
\newcommand{\multilinecell}[2][c]{%
  \begin{tabular}[#1]{@{}c@{}}#2\end{tabular}}

\title{\LARGE \bf Nonparametric Distribution Models for Predicting and
  \\ Managing Computational Performance Variability* }

\author{Thomas C. H. Lux$^{1}$, Layne Watson$^{2}$, Jon Bernard$^{1}$, Tyler H. Chang$^{1}$, Bo Li$^{1}$, \\ Li Xu$^{3}$, Godmar Back$^{2}$, Ali R. Butt$^{2}$, Kirk W. Cameron$^{2}$, Yili Hong$^{4}$% <-this % stops a space
\thanks{*This work was supported by the National Science Foundation}%
\thanks{$^{1}$T. Lux, J. Bernard, T. Chang, B. Li are doctoral students of Computer Science at Virginia Polytechnic Institute and State University, Blacksburg, Virginia 24060, USA {\tt\small (tchlux at vt.edu)}}%
\thanks{$^{2}$L. Watson, G. Back, A. Butt, and K. Cameron are Faculty of Computer Science at Virginia Polytechnic Institute and State University, Blacksburg, Virginia 24060, USA}%
\thanks{$^{3}$L. Xu is a doctoral student of Statistics at Virginia Polytechnic Institute and State University, Blacksburg, Virginia 24060, USA}%
\thanks{$^{4}$Y. Hong is with the Faculty of Statistics at Virginia Polytechnic Institute and State University, Blacksburg, Virginia 24060, USA}%
}

\begin{document}

\maketitle
\thispagestyle{empty}
\pagestyle{empty}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{abstract}
Performance variability can have a significant impact on many applications of computing. Cloud Computing, High Performance Computing, and Computer Security communities each exert considerable effort managing and analyzing variability throughout the system stack. This work presents and evaluates a methodology for predicting precise characteristics of the computational performance variability of an input/output (I/O) application over varying system configurations. Results demonstrate that the presented methodology is capable of precisely modeling performance variability, which could allow applications that tighten service level agreements, maximize computational throughput, and obfuscate system configurations against malicious users.
\end{abstract}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section{INTRODUCTION}
\label{sec:introduction}

Computational variability presents itself in a variety of forms. Processes that are apparently identical in a cloud computing or high performance computing (HPC) environment may take different amounts of time to complete the same job. This variability can cause unintentional violations of service level agreements in cloud computing applications or indicate suboptimal performance in HPC applications. The sources of variability, however, are distributed throughout the system stack and often difficult to identify. The methodology presented in this work is applicable to modeling the expected variability of useful computer system performance metrics without any prior knowledge of system architecture. Some examples of interesting performance metrics that could be modeled with the techniques in this work include computational throughput, power consumption, processor idle time, number of context switches, and RAM usage, as well as any other ordinal performance metric.

Predicting performance variability in a computer system is a challenging problem that has primarily been attempted in one of two ways: (1) build a statistical model of the performance data collected by running experiments on the system at select settings, or (2) run artificial experiments using a simplified simulation of the target system to estimate architecture and application bottlenecks. In this paper, the proposed modeling techniques rest in the first category and they represent a notable increase in the ability to model precise characteristics of variability.

Many previous works attempting to model system performance have used simulated environments to estimate the performance of a system \cite{grobelny2007fase,wang2009simulation,wang2013towards}. Some refer to statistical models as being oversimplified and not capable of capturing the true complexity of the underlying system. This claim is partially correct, noting that a large portion of predictive statistical models rely on simplifying the machine to one or two parameters \cite{snavely2002framework,bailey2005performance,barker2009using,ye2010analyzing}. These limited statistical models have provided satisfactory performance in very narrow application settings. Many of the aforementioned statistical modeling techniques claim to generalize, while simultaneously requiring additional code annotations, hardware abstractions, or additional application level understandings in order to generate models. The approach presented here requires no modifications of the application, no architectural abstractions, nor any structural descriptions of the input data being modeled. The techniques used are purely mathematical and only need performance data as input.

Among the statistical models presented in prior works, \cite{bailey2005performance} specifically mentions that it is difficult for the simplified models to capture variability introduced by I/O. System variability in general has become a problem of increasing interest to the cloud and HPC systems communities, however most existing work has focused on operating system (OS) induced variability \cite{beckman2008benchmarking,de2007identifying}. The work that has focused on managing I/O variability does not use any sophisticated modeling techniques \cite{lofstead2010managing}.

This paper presents an application of advanced statistical modeling to the domain of computer system I/O throughput. The models are used to predict the cumulative distribution function (CDF) of the expected I/O throughput for a system at previously unseen configurations. The techniques in this paper can tractably model tens of interacting system parameters with tens of thousands of unique configurations.

Section \ref{sec:variability_modeling} details the mathematical techniques used to approximate variability, the chosen measurement of error, and an optimization strategy for improving model performance. Section \ref{sec:data} summarizes the data collection process as well as provides summary statistics of the data used for the case study. Section \ref{sec:results} presents the results of the I/O case study including analyses of approximation accuracy, model convergence, and system parameter importance. Section \ref{sec:discussion} discusses possible explanations for witnessed results, potential applications of the predictive methodology, and future improvements of the models.

%% ===================================================================
\section{VARIABILITY MODELING}
\label{sec:variability_modeling}

Performance and its variability can be summarized by a variety of statistics. Mean, range, standard deviation, variance, and interquartile range are a few summary statistics that describe performance and variability. However, the most precise characterization of any ordinal performance metric is the cumulative distribution function (CDF), or its derivative the probability density function (PDF). Previous techniques for predicting system performance have strictly modeled real-valued summary statistics because there exists a large base of mathematical techniques capable of approximating functions of the form $f: \mathbb{R}^d \rightarrow \mathbb{R}$. There is a smaller theoretical foundation for approximating functions of the form $f: \mathbb{R}^d \rightarrow g$ with $g: \mathbb{R} \rightarrow \mathbb{R}$. The latter is the approximation of functions.

A major hurdle when modeling functions such as the CDF or PDF, is that certain properties must be maintained in order for the functions to remain valid. It is necessary that a PDF $f: \mathbb{R} \rightarrow \mathbb{R}$ have the properties $f(x) > 0$ and $\int_{-\infty}^{\infty}f(x)dx = 1$. However, for a CDF $F: \mathbb{R} \rightarrow \mathbb{R}$ the properties are $F(x) \in [0,1]$ and $F'(x) \geq 0$. For the purpose of function approximation, the properties of the CDF are more straightforward to maintain. This work utilizes the fact that a convex combination of CDFs results in a valid CDF. Given $G(x) = \sum_{i}w_i F_i(x)$, $\sum_{i} w_i = 1$, $w_i \geq 0$, and each $F_i$ is a valid CDF, $G$ must also be a valid CDF. A demonstration of how this is applied can be seen in Figure \ref{fig:prediction_example}. Next, the three techniques used to model CDFs are summarized, a mechanism for identifying important features is proposed, and the error measure used throughout this work is presented. For each of the following, $n \in \mathcal{N}$ refers to the number of training samples (known system configurations), $d \in \mathcal{N}$ refers to the dimension of the data (ordinal system parameters), $x^{(i)} \in \mathbb{R}^d$ is a known system configuration with known CDF $F_{x^{(i)}}$, and $y \in \mathbb{R}^d$ refers to a previously unseen system configuration with CDF $F_y$.

%% ===================================================================
\subsection{Delaunay}
\label{sec:delaunay}

The Delaunay method of interpolation is a well studied geometric technique for producing an interpolant \cite{lee1980two}. The Delaunay triangulation of a set of data points into simplices is such that the sphere defined by the vertices of each simplex contains no data points in the sphere's interior. For a $d$-simplex S with vertices $x^{(0)}$, $x^{(1)}$, $\ldots$, $x^{(d)}$, and functions $F_{x^{(i)}}$, $i=0$, $\ldots$, $d$, $y$ is a unique convex combination of the vertices,

$$ y = \sum_{i=0}^{d} w_i x^{(i)}, \quad \sum_{i=0}^{d} w_i = 1, \quad w_i \geq 0, \quad i=0,\ldots,d, $$

and the Delaunay interpolant $F_y$ at $x$ is given by

$$ F_y = \sum_{i=0}^{d} w_i F_{x^{(i)}}. $$

\begin{figure}[htb]
  \begin{tikzpicture}
    \node (img)  {\includegraphics[width=0.4\textwidth,]{plot-delaunay-example-prediction.pdf}};
    \node[below=of img, node distance=1cm, yshift=1cm] {I/O Throughput};
    \node[left=of img, node distance=0cm, rotate=90, anchor=center,yshift=-0.7cm] {CDF Value};
  \end{tikzpicture}
  \vspace{-0.3cm}
  \caption{In this example, the general methodology for predicting a CDF and evaluating error can be seen. The Delaunay method chose three source system configurations (dotted lines) and assigned weights \{.3, .4, .3\} (top to bottom). The weighted sum of the three known CDFs produces the predicted CDF (dashed line). The KS Statistic (red arrow) computed between the true CDF (solid line) and predicted CDF (dashed line) is 0.2 for this example. This prediction is deemed a failure by p-values 0.05 and 0.01, however it is not a failure according to p-values 0.001 and 1.0e-6.
  \vspace{-.1cm}}
  \label{fig:prediction_example}
\end{figure}

The computational complexity of interpolating via Delaunay (for the implementation used) is approximately $\mathcal{O}(n d^4 log d)$, which is capable of scaling reasonably to $d \leq 50$. In the present application, Delaunay is used to select up to $d+1$ system configurations and assign weights to each in order to generate a prediction for a new system configuration.

%% ===================================================================
\subsection{Max Box Mesh}
\label{sec:max_box_mesh}

The Max Box mesh is an interpolation technique that utilizes overlapping box splines \cite{de2013box} as basis functions that are shifted and scaled to support box shaped regions. The boxes are constructed such that each box has strictly one point on its interior while maintaining the largest minimum distance between the interior point and the closest face of the box (a measure of centrality). Given a set of box splines $b^{x^{(i)}}$ with the max box properties anchored at each $x^{(i)}$,

$$ F_y = \frac{\sum_i b^{x^{(i)}}(y) F_{x^{(i)}}}{\sum_i b^{x^{(i)}}(y)}, $$

where $b^{x^{(i)}}(y) \geq 0$ and the denominator causes the sum of weights to be one. The computational complexity of interpolation via the Max Box mesh is $\mathcal{O}(n^2 d)$. The Max Box mesh interpolates a function $F_y$ by producing a convex combination of known functions $F_{x^{(i)}}$, where the weights are determined by the box splines.

%% ===================================================================
\subsection{Voronoi Mesh}
\label{sec:voronoi_mesh}

A well-studied technique for classification and approximation is the nearest neighbor algorithm \cite{cover1967nearest}. Nearest neighbor inherently utilizes the convex region surrounding a central control point in whose interior all points are closer to that center than any other control point, which is also called the Voronoi cell \cite{dirichlet1850reduction}. The Voronoi mesh smooths the nearest neighbor approximation by utilizing the Voronoi cells to define support via a generic basis function $v: \mathbb{R}^d \rightarrow \mathbb{R}_+$ with

$$ v^{x^{(i)}}(y) = \left(1 - \frac{||y - x^{(i)}||_2}{2 \ d(y - x^{(i)} \mid x^{(i)})} \right), $$

where $d(\ \cdot \mid x^{(i)})$ is the distance between $x^{(i)}$ and the nearest boundary of the Voronoi cell along the vector $\cdot\ $. While all basis functions $v^{x^{(i)}}(x^{(i)}) = 1$, the $2$ in the denominator causes all basis functions to go linearly to $0$ at the boundary of the twice-expanded Voronoi cell. Now we have

$$ F_y = \frac{\sum_i v^{x^{(i)}}(y) F_{x^{(i)}}}{\sum_i v^{x^{(i)}}(y)}, $$

where $v^{x^{(i)}}(y) \geq 0$ by definition and the denominator causes the sum of weights to be one. The computational complexity of interpolation via Voronoi mesh is $\mathcal{O}(n^2 d)$. The Voronoi mesh interpolates a function $F_y$ just as the Max Box mesh and Delaunay triangulation, by a convex combination of known functions $F_{x^{(i)}}$.
%% ===================================================================
\subsection{Feature Weighting}
\label{sec:feature_weighting}

Machine learning has demonstrated that an important procedure in any application of predictive methodologies is identifying those features of training data that are most relevant to making accurate predictions \cite{guyon2003introduction}. Selection strategies such as the floating searches studied in \cite{pudil1994floating} or others compared in \cite{ferri1994comparative} can be too expensive for large approximation problems. Rather, this work poses feature selection as a continuous optimization problem. Let $X$ be an $n \times d$ matrix of $n$ known system configurations with $d$ parameters each normalized to be in the unit cube. Define an error function $err: \mathbb{R}^d \rightarrow \mathbb{R}_+$ that estimates the error of a predictive model trained on $X diag(w)$, $w \in \mathbb{R}^d$,  by performing 10-fold cross validation using 80\% of the rows of $X diag(w)$ as training and 20\% as testing. A minimal solution to this error function could be considered an optimal weighting of the features of $X$. Minimization is performed using the modified simulated annealing technique presented in \cite{lux2016convergence} because of its empirically fast convergence in the absence of a readily computable gradient.

%% ===================================================================
\subsection{Measuring Error}

The performance of approximation techniques that predict functions can be analyzed through a variety of summary statistics. Mean error, mean absolute error, mean squared error, and the max absolute error are all popular measures. This work uses the max absolute error, also known as the Kolmogorov-Smirnov (KS) statistic \cite{lilliefors1967kolmogorov} for its compatibility with the KS test.

The two-sample KS test is a useful nonparametric test for declaring when two CDFs do not originate from the same underlying distribution while only assuming stationarity, finite mean, and finite variance. The null hypothesis (that two CDFs come from the same underlying distribution) is rejected at level $p \in [0,1]$ when

$$ KS > \sqrt{-\frac{1}{2}ln\big(\frac{p}{2}\big)} \sqrt{\frac{1}{n_1} + \frac{1}{n_2}}, $$

with distribution sample sizes $n_1,n_2 \in \mathcal{N}$. For all applications of the KS test presented in this work $n_1 = n_2 = 150$. An example of the round-trip prediction methodology from known and predicted distributions to the calculation of error can be seen in Figure \ref{fig:prediction_example}.

%% The majority of results presented in this work are absent any asymetric preprocessing operations on the training data, some findings related to automatic identification of important system parameters are presented in Table \ref{tab:optimized_p_value_failure_rate} and analyzed in \ref{sec:results}.

%% ===================================================================
\section{DATA}
\label{sec:data}

\begin{table}
  \centering
  \begin{tabular}{c|c}
    \hline
    \textbf{System Parameter} & \textbf{Values}\\
    \hline
    File Size (KB) & 4, 16, 64, 256, 1024, 4096, 8192, 16384\\
    \hline
    Record Size (KB) & \multilinecell{4, 8, 16, 32, 64, 128, 256, 512,\\ 1024, 2048, 4096, 8192, 16384}\\
    \hline
    Thread Count & 1, 8, 16, 24, 32, 40, 48, 56, 64\\
    \hline
    Frequency (GHz) & 1.2, 1.6, 2, 2.3, 2.8, 3.2, 3.5\\
    \hline
    Test Type & \multilinecell{Readers, Rereaders, Random Readers, \\ Initial Writers, Rewriters, Random Writers}\\
    \hline
  \end{tabular}
  \caption{A description of system parameters considered for IOzone. Record size must be less than file size during execution.
    \vspace{-.5cm}}
  \label{tab:data_description}
\end{table}

\begin{figure}
  \includegraphics[width=.5\textwidth]{plot-histogram-throughput.pdf}
  \caption{Histogram of the raw throughput values recorded during all IOzone tests across all system configurations. The distribution is skewed right, with few tests having significantly higher throughput than most others.
  \vspace{-.5cm}}
  \label{fig:throughput_histogram}
\end{figure}

This paper presents a variability modeling case study with a five-dimensional dataset produced by executing the IOzone benchmark \cite{iozone} on a homogeneous cluster of computers. Each node contains two Intel Xeon E5-2637 CPUs offering a total of 16 CPU cores with 16GB of DRAM. While the CPU frequency varies depending on the test configuration, the I/O from IOzone is performed by an ext4 filesystem sitting above an Intel SSDSC2BA20 SSD drive. At the time of data collection, a Linux kernel version of 4.13.0 was used. The system performance data was collected over two weeks by executing IOzone 150 times for each of a select set of approximately 18K system configurations, for a total of approximately 2.7M executions of IOzone. A single IOzone execution reports the max I/O throughput in kilobytes per second seen for the selected test. The summary of the data generated in the experiments for this paper can be seen in Table \ref{tab:data_description}. Distributions of raw throughput values being modeled can be seen in Figure \ref{fig:throughput_histogram}.

Some mild preprocessing was necessary to prepare the data for modeling and analysis. All features were shifted by their minimum value and scaled by their range, mapping each feature independently into $[0,1]$. This normalization ensures each feature is treated equally by the interpolation techniques and should be performed on all data before building models and making predictions regardless of application. All 150 repeated trials for a system configuration were grouped with that configuration. The only non-ordinal feature in this data is the test. All tests were treated as different applications and were separated for modeling and analysis. (i.e. predictions for the ``readers'' test were made using known configurations for the ``readers'' test only.) The division of non-ordinal features into separate models is necessary for the presented methodology, however future improvements are discussed in Section \ref{sec:discussion}.

%% ===================================================================

\begin{figure}
  \begin{tikzpicture}
    \node (img)  {\includegraphics[width=0.5\textwidth,trim={1.5cm 0 0 0}]{plot-histogram-80_20-KS.pdf}};
    \node[below=of img, node distance=1cm, yshift=1cm] {KS Statistic for Predicted vs. Actual};
    \node[left=of img, node distance=0cm, rotate=90, anchor=center,yshift=-0.7cm] {Count};
  \end{tikzpicture}
  \caption{Histograms of the prediction error averaged over 10 folds for each modeling algorithm when trained with 80\% of the data. The distributions show the KS statistics for the predicted throughput distribution versus the actual throughput distribution. The four vertical red lines represent commonly used p-values \{0.05, 0.01, 0.001, 1.0e-6\} respectively. All predictions to the right of a red line represent failed predictions that are significantly different (by respective p-value) from the actual distribution according to the KS test.
  \vspace{-.1cm}}
  \label{fig:ks_histogram_80_20}
\end{figure}

\section{RESULTS}
\label{sec:results}

All three interpolation techniques are used to predict the distribution of I/O throughput values at previously unseen system configurations. In order to improve robustness of the error analysis, 10-fold cross validation is used. 10 random selections of 80\% of the IOzone data are used to train each model and the remaining 20\% provide estimations of approximation error for each model. The recorded errors are averaged over the 10 random folds and grouped by unique system configuration. The folds are identically seeded for each interpolation technique, ensuring consistency in the training and testing sets.

The aggregation of errors across all IOzone tests given 80\% of the data as training can be seen in Figure \ref{fig:ks_histogram_80_20}. Agglomerate errors for each technique present similarly, each resembles a Gamma distribution. The percentages of failed predictions with varying p-values are on display in Table \ref{tab:p_value_failure_rate}. The primary p-value used for analyses in this work is 0.001. This p-value is chosen because close to 2K predictions are made for each test. Also, applications executed in cloud and HPC systems that could benefit from statistical modeling will be executed at least on the order of 1000's of times. In line with this knowledge, it is important to ensure that only a small fraction of interpretable results could occur solely under the influence of random chance. When considering the p=0.001 results for each technique, a little under half of the generated predictions fail. In a traditional application of machine learning techniques a failure rate of 45\% would be a poor result, however in this situation the complexity of the problem warrants a slightly different interpretation. These predictions are of a very \textit{precise} characterization of performance variability, in fact the strongest possible characterization of variability. Globally, only a little under half of the predictions fail to capture \textit{all} of the characteristics of performance variability at new system configurations. It is also demonstrated later in this Section that this result can likely be improved.

\begin{table}
  \centering
  \begin{tabular}{c|c|c}
    \hline
    \textbf{Algorithm} & \textbf{P-Value} & \textbf{\% Failed Predictions} \\
    \hline
    \multilinecell{Delaunay\\Max Box Mesh\\Voronoi Mesh} & .05 & \multilinecell{58.4\\69.3\\61.9}\\
    \hline
    \multilinecell{Delaunay\\Max Box Mesh\\Voronoi Mesh} & .01 & \multilinecell{51.1\\58.4\\53.4}\\
    \hline
    \multilinecell{Delaunay\\Max Box Mesh\\Voronoi Mesh} & .001 & \multilinecell{44.1\\46.9\\45.1}\\
    \hline
    \multilinecell{Delaunay\\Max Box Mesh\\Voronoi Mesh} & 1.0e-6 & \multilinecell{31.4\\26.6\\28.7}\\
    \hline
  \end{tabular}
  \caption{The prediction failure rate by the KS-test when provided different selections of p-values. These accompany the results from Figure \ref{fig:ks_histogram_80_20}. 
    \vspace{-.5cm}}
  \label{tab:p_value_failure_rate}
\end{table}

While interpreting failure rates of these interpolation techniques, it is important to consider at what rate the error reduces with increasing amounts of training data. Figure \ref{fig:ks_failure_by_training} displays the change in p=0.001 failure rate with increasing density of training data up to the maximum density allowed by this training set. Delaunay provides the best results with the least training data by about 5\%, but these low-density prediction error rates are unacceptably high (90\% failure). Increased density demonstrates a roughly $(1 - \sqrt{n})$ rate of reduction in error that will level off around 20-30\% with a training set twice the size of the one collected for this study. However, constructing a training set of such a size would likely require months of compute time and be infeasible for many applications. The rate of convergence seen also suggests that these models are successfully converging on the lowest possible error given the (presumed) Lipschitz constant of the data and any latent variables that may perfectly describe system I/O throughput variability.

\begin{figure}
  \begin{tikzpicture}
    \node (img)  {\includegraphics[width=0.4\textwidth,]{plot-KS-Failure-by-Training.pdf}};
    \node[below=of img, node distance=1cm, yshift=1cm] {Percentage of Training Data};
    \node[left=of img, node distance=0cm, rotate=90, anchor=center,yshift=-0.7cm] {Percentage of Failed Predictions};
  \end{tikzpicture}
  \caption{The performance of each algorithm with increasing amounts of training data estimated using 10-fold cross validation and the KS test with p=0.001. The KS statistic for any given training percentage is achieved by averaging over the KS statistic for all tests and system configurations. The training percentages range from 5\% to 95\% by increments of 5\%. Delaunay is the best performer until 95\% of data is used for training, at which Max Box mesh becomes the best performer by a fraction of a percent.
  \vspace{-.1cm}}
  \label{fig:ks_failure_by_training}
\end{figure}

It may be misleading to consider the global performance of each prediction technique across all tests, as some tests are more difficult than others to predict and have more apparent latent activity. In Figure \ref{fig:ks_failure_by_training_and_test}, the relative difficulty of each IOzone test can be compared. The I/O tests analyzing reads are typically approximated with lower error than those tests analyzing writes. Regardless of test, in the aggregate results the modal KS statistics hover consistently around 0.15, demonstrating an impressively low KS statistic for modal predictions. In order to address the opacity of aggregate analysis, another case study and an application of the methodology from Section \ref{sec:feature_weighting} is presented in Table \ref{tab:optimized_p_value_failure_rate}.

The results presented in Table \ref{tab:optimized_p_value_failure_rate} are achieved by permitting each approximation technique 300 iterations of modified simulated annealing. In each iteration, the impact of potential weights on average validation KS statistic were considered. All weights were kept in the range [0,2], and were applied to the normalized features for frequency, file size, record size, and number of threads. All three techniques converged on similar optimal weights for the validation sets of approximately \{.001, 2, 1.7, 1.5\} for each respective feature. Recall that each interpolation technique uses smaller distances to denote larger influence on predictions, meaning that frequency was the most important feature when predicting variability for the ``readers'' test, followed not-so-closely by number of threads, then record size.

The ``readers'' test results demonstrate that the underlying prediction techniques work and are capable of seeing failure rates below 5\% when optimized for a given application. It is important to emphasize that the roughly 95\% of predictions that did not fail are predicting the \textit{precise} distribution of I/O throughput that will be witnessed at a previously unseen system configuration. To present knowledge, there is no existing methodology that is generically applicable to any system performance metric, agnostic of system architecture, and capable of making such powerful predictions.

\begin{table}
  \centering
  \begin{tabular}{c|c|c|c}
    \hline
    \textbf{Algorithm} & \textbf{P-Value} & \multilinecell{\textbf{Unweighted}\\\textbf{\% Failure}} & \multilinecell{\textbf{Weighted}\\\textbf{\% Failure}}\\
    \hline
    \multilinecell{Delaunay\\Max Box Mesh\\Voronoi Mesh} & .05 & \multilinecell{24.9\\21.3\\18.7} & \multilinecell{30.2\\21.2\\11.3}\\
    \hline
    \multilinecell{Delaunay\\Max Box Mesh\\Voronoi Mesh} & .01 & \multilinecell{21.6\\16.4\\14.9} & \multilinecell{27.4\\16.4\\7.0}\\
    \hline
    \multilinecell{Delaunay\\Max Box Mesh\\Voronoi Mesh} & .001 & \multilinecell{19.7\\13.1\\12.3} & \multilinecell{25.4\\13.1\\4.6}\\
    \hline
    \multilinecell{Delaunay\\Max Box Mesh\\Voronoi Mesh} & 1.0e-6 & \multilinecell{17.9\\11.3\\8.5} & \multilinecell{23.4\\11.3\\2.3}\\
    \hline
  \end{tabular}
  \caption{The prediction failure rates by various p-values for the KS-test. These results are strictly for the ``readers'' IOzone test and show unweighted results as well as the results with weights optimized for minimum error (KS statistic) by 300 iterations of modified simulated annealing. Notice that the weights identified for the Delaunay model cause overfitting and reduction in apparent performance. MaxBoxMesh performance is improved by a negligible amount. VoronoiMesh predictions are notably improved.
    \vspace{-.5cm}}
  \label{tab:optimized_p_value_failure_rate}
\end{table}

%% \begin{itemize}
%% \item Case studies of good and bad predictions of distributions made, present 3 (best, median, worst)
%% \item Table / Figure showing error for each of the tests (hardest test to predict)
%% \item Presentation of the top most difficult configurations to predict for each test (and the bad performance at those points)
%% %% \item Comparison to predictions made using the "test" numerical column
%% \end{itemize}

\section{DISCUSSION}
\label{sec:discussion}

The results of the IOzone case study indicate that predicting the CDF of I/O throughput at previously unseen system configurations is a challenging problem. The KS statistic captures the worst part of any prediction and hence provides a conservatively large estimate of approximation error. The average absolute error in the predicted CDFs are always lower than the KS statistics. However, the KS statistic was chosen because of the important statistical theory surrounding it as an error measure. Considering this circumstance, a nonnegligible volume of predictions provide impressively low levels of error. Powerful predictive tools such as those presented in this work allow for more in-depth analysis of system performance variability. For example, system configurations that are most difficult to predict in these tests are likely ``outlier'' configurations that do not resemble those configurations that share many similar parameters. Analysis of these configurations may provide valuable insight into effective application specific operation of computer systems.

As mentioned in Section \ref{sec:introduction}, no prior work has attempted to model an arbitrary performance metric for a system to such a high degree of precision. All previous statistical modeling attempts capture few (\textless 3) ordinal measures of performance. Generating models that have such high degrees of accuracy allows system engineers to identify previously unused configurations that present desired characteristics. Service level agreements (SLAs) in cloud computing environments are cause for capital competition that is impacted heavily by system performance \cite{patel2009service}. Users prefer SLAs that allow the most computing power per capital expense, incentivizing service providers to guarantee the greatest possible performance. Overscheduling and irregular usage patterns force cloud service providers to occasionally overload machines, in which case precise models of system performance can be used to statistically minimize the probability of SLA violation. Similar targeted performance tuning techniques can be applied to HPC system configuration to maximize application throughput or minimize system power consumption.

A final application domain that may also be affected by this methodology is computer security. Collocated users on cloud systems are a source of recent attention \cite{ali2015security}. If a malicious collocated user is capable of achieving specific insight into the configuration of the system, or the activity of other collocated users by executing performance evaluation programs (i.e. IOzone), a new attack vector may present itself. Malicious users could be capable of identifying common performance distributions of vulnerable system configurations and vulnerable active user jobs. This knowledge may allow targeted exploits to be executed. Light inspection of raw IOzone I/O throughputs provides substantial evidence that distinct performance distributions coincide closely with specific system configuration parameters. Conversely, a service provider may defend against such attacks by deliberately obfuscating the performance of the machine. Models such as those presented in this paper could identify optimal staggering and time-delay whose introduction into the system would prevent malicious users from identifying system configurations and active jobs.

\begin{figure}
  \begin{tikzpicture}
    \node (img)  {\includegraphics[width=0.4\textwidth,]{plot-KS-failure-by-training-and-test.pdf}};
    \node[below=of img, node distance=1cm, yshift=1cm] {Percentage of Training Data};
    \node[left=of img, node distance=0cm, rotate=90, anchor=center,yshift=-0.7cm] {Percentage of Failed Predictions};
  \end{tikzpicture}
  \caption{The performance of each algorithm on different IOzone tests with increasing amounts of training data measured using 10-fold cross validation and the KS test with p=0.001. The KS statistic for any given training percentage is achieved by averaging over the KS statistic for all tests and system configurations. The training percentages range from 5\% to 95\% by increments of 5\%. The ``readers'' tests tend to allow lower failure rates than the ``writers'' tests.
  \vspace{-.1cm}}
  \label{fig:ks_failure_by_training_and_test}
\end{figure}

It remains unclear what mathematical properties of each approximation technique contribute the most to overall accuracy. Results presented in Table \ref{tab:optimized_p_value_failure_rate} are particularly interesting, demonstrating that Delaunay appears most vulnerable to overfitting, Max Box mesh is largely invariant to feature scaling, and Voronoi mesh generalizes feature weights well to unseen configurations.

There are many avenues for extending this modeling methodology. One of the most severe limitations of the present framework is the restriction to modeling ordinal system parameters. Many applications have nonordinal settings and individually modeling all combinations of those settings is not a tractable solution. Automatic processing of categorical settings into an appropriate set of ordinal features constitutes future work. Presently the failure rate of distribution predictions can only be reduced with large volumes of performance data, however future work may be able to effectively reduce the amount of training data required by introducing appropriate expectations of stochastic noise in the measured performance. Finally, more case studies need to be explored in order to test the robustness of the present modeling techniques to changes in domain and performance metric.

\section{CONCLUSION}
\label{sec:conclusion}

The presented methodology and application is capable of providing new insights, extending existing analyses, and improving the management of computational performance variability. Each of Delaunay, Max Box, and Voronoi meshes are viable techniques for constructing approximations of performance cumulative distribution functions. A case study on I/O throughput demonstrated that models are capable of effectively predicting most unseen system configurations for any of the available I/O tests. The present methodology represents a notable increase in the ability to statistically model arbitrary system performance metrics involving the interaction of many ordinal system parameters.
%% \begin{itemize}
%% \item Modeling the distributions of system performance is valuable.
%% \item This work presents a steep increase in the power of predictions made by statistical modeling techniques.
%% \end{itemize}

\bibliographystyle{IEEEtran}
\bibliography{paper}

\end{document}

### 2018-01-21 Report_[2018-01-15]_System_Throughput_Sample/code.py

@@ -1,0 +1,173 @@
+import os, pickle
+import numpy as np
+from util.data import read_struct
+from util.plotly import Plot
+from util.stats import cdf_fit_func
+
+CLEAN_DATA_CSV = "clean_data.csv"
+CLEAN_DATA_CSV_PKL = "clean_data.csv.pkl"
+
+if not os.path.exists(CLEAN_DATA_CSV_PKL):
+    data = read_struct(CLEAN_DATA_CSV)
+    print("Saving '%s'..."%(CLEAN_DATA_CSV_PKL))
+    with open(CLEAN_DATA_CSV_PKL, "wb") as f:
+        pickle.dump(data,f)
+else:
+    print("Loading '%s'..."%(CLEAN_DATA_CSV_PKL))
+    with open(CLEAN_DATA_CSV_PKL, "rb") as f: 
+        data = pickle.load(f)
+
+header = list(data.dtype.names)
+h_config = header[:4]
+h_test = header[4]
+h_values = header[5:]
+tests = sorted(np.unique(data[h_test]))
+test_len = max(map(len, tests))
+# Extract a subset of data, only the "readers" test
+# data = data[data["Test"] == "readers"]
+
+# ============================================================
+
+count = 0
+failures = 0
+slow_configs = []
+throughput_threshold = 40000
+total_threshold = 80
+
+p = Plot("Low throughput configurations", "Sample number", "Throughput")
+for row in data:
+    total = sum(1 for h in h_values if row[h] < throughput_threshold)
+    if total > 0:
+        slow_configs.append((tuple(row[h] for h in h_config), total))
+        if total >= total_threshold:
+            raw_values = np.array([row[h] for h in h_values])
+            name = str(slow_configs[-1][0] + (row["Test"],))
+            p.add("",np.arange(len(raw_values)), raw_values, 
+                  frame=name, group="")
+    if (row["File Size"] == 4 and row["Record Size"] == 4): 
+        count += 1
+        if total > 0:
+            failures += total
+slow_configs = sorted(slow_configs,key=lambda r: r[1])[-16:]
+for (c,t) in slow_configs:
+    print(c,t)
+
+p.plot(show_slider_labels=False, frame_label="System Configuration: ",
+       data_easing=True, show_legend=False, bounce=True)
+
+print()
+print("Number of slow runs: ", failures)
+print("Number of (4,4) runs:", count*150)
+print("Total number of runs:", len(data)*150)
+print()
[…]

### 2018-02-13 Paper_[2018-01-09]_ACMSE/paper.tex
Finalized some more modifications on the ACM paper. Ready to deliver to Dr. Watson for further review.

@@
+\def\mn#1{\left\Vert #1\right\Vert_\infty} 
 \section{Introduction}
@@
 
-Throughout this section, the same notation will be used as introduced in Section \ref{sec_introduction}.
+Throughout this section, the notation will be reused from Section \ref{sec_introduction}.
 $X \subset \mathbb{R}^d$ is a finite set of points with known response values $f(x)$ for all $x \in X$.
+Also let $L,U \in \mathbb{R}^d$ define a bounding box for $X$ such that $L < x < U$ for all $x \in X$ (and comparison is componentwise).
 
-Define a box $b^c = (c,l^c,u^c)$ in $d$ dimensions with center $c \in \mathbb{R}^d$, lower width vector $l^c \in \mathbb{R}^d_+$, and upper width vector $u^c \in \mathbb{R}^d_+$ (where $u^c_i$ refers to the $i$th component of $u^c$).
+Define a box $b^c = (c,l^c,u^c)$ in $d$ dimensions with anchor $c \in \mathbb{R}^d$, lower width vector $l^c \in \mathbb{R}^d_+$, and upper width vector $u^c \in \mathbb{R}^d_+$ (where $u^c_i$ refers to the $i$th component of $u^c$).
 A visual example of a box in two dimensions can be seen in Figure \ref{fig_example_box}.
@@
 \begin{equation}
-  \bigl(g^c(x)\bigr)_r = {k \over 2} \left( 1 + {(x_r - c_r)_- \over l^c_r} - {(x_r - c_r)_+ \over u^c_r} \right)
+  \bigl(g^c(x)\bigr)_r = {k \over 2} \left( 1 - {(x_r - c_r)_- \over l^c_r} + {(x_r - c_r)_+ \over u^c_r} \right),
 \end{equation}
 where $y_+=\max\{y,0\}$, $y_-=(-y)_+$, $k$ is the order of the box spline as described in Section \ref{sec_box_splines}.
-Finally, each box spline in a box mesh can be evaluated as $B(g^c(x) \mid A)$ presuming the order of approximation implies $A$.
+Finally, each box spline in a box mesh can be evaluated as $B^c(x)$, where $B^c(x) = B(g^c(x) \mid A)$ presuming the order of approximation implies $A$.
 Both box meshes described in the following subsections use box spline basis functions of this form.
@@
   \end{tikzpicture}
-  \caption{An example box in two dimensions with center $c$, upper widths $u^c_1$, $u^c_2$, and lower widths $l^c_1$, $l^c_2$.
+  \caption{An example box in two dimensions with anchor $c$, upper widths $u^c_1$, $u^c_2$, and lower widths $l^c_1$, $l^c_2$.
 Notice that $c$ is not required to be equidistant from opposing sides of the box, that is $u^c_i \not= l^c_i$ is allowed.}
@@
 The first of the three meshes produces a set of boxes around chosen control points and each box has maximal distance between the control point and the nearest side of the box.
-This centrality property is one mechanism of creating the largest reasonable regions of support for the underlying basis functions.
+This centrality property is one mechanism for creating the largest reasonable regions of support for the underlying basis functions.
 The individual boxes are constructed via the following procedure given a set of control points $C \subseteq X$, $c^{(i)} \in C$,
 \begin{enumerate}
-  \item Initialize a box $b^{c^{(1)}} \supset X$ with center $c^{(1)}$. \label{step_init}
+  \item Initialize a box $b^{c^{(1)}} = \bigl(c^{(1)}, (c^{(1)} - L), (U - c^{(1)})\bigr)$. \label{step_init}
   \item Identify $c^{(i)}$ over $\bigl\{ j \bigm| j \ne 1$, $B^{c^{(1)}}\bigl( c^{(j)} \bigr) \ne 0 \bigr\}$ that minimizes $\bigl\|c^{(j)}-c^{(1)}\bigr\|_\infty$.  \label{step_closest}
@@
 The user is left to decide the batch size $s$ and the error tolerance based on validation performance and computability.
-Next, the max box mesh is shown to be nearly space filling.
+This work uses a batch size of one.
 
@@
 
-\begin{definition}
-A \textit{proper} upper boundary for box $b^x$ at point $y$ occurs when $u^x_r = ||x - y||_\infty$ and $r$ is the first dimension such that $u^x_r = ||x - y||_\infty$, assuming dimensions are deterministically ordered.
-The same definition applies to lower boundaries.
-\end{definition}
+Throughout all experiments and all repeated trials conducted for this study, all tested interpolation points were covered by at least one box in the $MBM$.
+However, it is possible for the $MBM$ to not form a covering of $(L,U)$ when there are cyclic boundaries.
+Consider the following example in three dimensions:
+\begin{align*}
+  C         &= \bigl\{(0,0,0), (1,0,2/3), (1,1,4/3)\bigr\} \\
+  b^{c^{(1)}} &= \bigl( (0,0,0),  \ (*,*,*),  \ (1,*,4/3) \bigr) \\
+  b^{c^{(2)}} &= \bigl( (1,0,2/3),\ (1,*,*),  \ (*,1,*)   \bigr) \\
+  b^{c^{(3)}} &= \bigl( (1,1,4/3),\ (*,1,4/3),\ (*,*,*)   \bigr)
+\end{align*}
+Asterisks are used to represent boxes that are not bounded by other boxes along some dimensions.
+The point $(2,2,-3)$ is not in any of the max boxes defined above.
+In this case, there is a cycle in box boundaries that looks like $\bigl( b^{c^{(1)}} \bigm\vert_1 b^{c^{(2)}} \bigm\vert_2 b^{c^{(3)}} \bigm\vert_3 b^{c^{(1)}} \bigr)$, where $\bigl( b^x \bigm\vert_r b^y \bigr)$ is used to denote ``$b^x$ is bounded on dimension $r$ by $b^y$''.
+This example demonstrates that it is geometrically possible for the max box mesh to fail to cover a space, however experiments demonstrate that it is empirically unlikely.
 
-\begin{theorem}
-A max box mesh composed of $n$ unique control points in $d$ dimensions, $C \subset \mathbb{R}^d$ where all boundaries are proper, is nearly space filling.
-\label{theorem_mutual_space_filling}
-\end{theorem}
-\begin{proof}
-Pick an arbitrary point $x \in \mathbb{R}^d$.
-Now let $c^{(1)}$ be the $c \in C$ that minimizes $|| c - x ||_\infty$, making the nearest box (by max norm) $b^{c^{(1)}}$.
-Either $x$ is in the support of $b^{c^{(1)}}$ or on the boundary of $b^{c^{(1)}}$ and this is done, or $b^{c^{(1)}}$ has a proper boundary along dimension $r$ with box $b^{c^{(2)}}$ such that $|c^{(2)}_r - x_r| < |c^{(1)}_r - x_r|$.
-The sequence of bounds $c^{(2)}$, $\ldots$, must be finite for the strict inequality to hold and the last box in this sequence supports $x$.
-Therefore, the $MBM$ is nearly space filling.
-\end{proof}
-
-The max box mesh as proposed is still capable of producing coefficients of 0 for $x \in \mathbb{R}^d$ placed along the boundary of boxes, however this problem is solved by increasing the upper and lower widths of all boxes by the unit roundoff during computation.
-For this reason, the max box mesh remains a viable strategy for computed approximations.
+The max box mesh remains a viable strategy for computed approximations.
 Given a maximum of $c$ control points in $d$ dimensions with $n$ points, the computational complexities are: $\mathcal{O}(c^2 d)$ for computing boxes, $\mathcal{O}(c d^2 + d^3)$ for a least squares fit, and $\mathcal{O}(n / s)$ for bootstrapping (which is multiplicative over the fitting complexities).
@@
 
-The iterative box mesh ($IBM$) composes box-shaped regions of support over control points that each support strictly one control point just as in the $MBM$, however the mesh is completely space filling by construction and places boxes in a way that reduces apparent error.
-The boxes are constructed via the following procedure given a finite set of points $X \subset \mathbb{R}^d$,
+The iterative box mesh ($IBM$) composes box-shaped regions that each contain strictly one control point just as in the $MBM$.
+However, the mesh is a covering for $(L,U)$ by construction and places boxes in a way that reduces apparent error.
+The boxes are constructed via the following procedure given a finite set of points $X \subset \mathbb{R}^d$, where $C \subseteq X$ is the (initially empty) set of control points
 \begin{enumerate}
-\item Add a box with total support around the most central point $x^{(1)} \in X$ to $IBM$ and fit the mesh to all $X$.
-\item Add a box centered at $x^{(i)} \notin IBM$ that has the most approximation error, reshaping all boxes in $b^{x^{(j)}} \in IBM$ that contain $x^{(i)}$ by bounding the first dimension $r$ such that $\bigl| x^{(j)}_r - x^{(i)}_r \bigr| = \bigl| \bigl| x^{(j)} - x^{(i)} \bigr| \bigr|_\infty$ and then fit the mesh to all $X$.
+\item Add a box that covers $(L,U)$ anchored at the most central point $x^{(1)} \in X$ to $C$ and fit the $IBM$ to all $X$.
+\item Add a new box that covers $(L,U)$ anchored at $x^{(i)} \notin C$ such that $x^{(i)} = \max_{x \in X \backslash C} \bigl| \hat f(x) - f(x) \bigr|$, reshaping all boxes $b^{x^{(j)}} \in C$ that contain $x^{(i)}$ by bounding the first dimension $r$ such that $\bigl| x^{(j)}_r - x^{(i)}_r \bigr| = \bigl\| x^{(j)} - x^{(i)} \bigr\|_\infty$ (also reshaping the box $b^{x^{(i)}}$ symmetrically) and then fit the mesh to all $X$.
 \label{step_add_box}
@@
 The user is left to decide the most appropriate error tolerance, however a decision mechanism and analysis is presented in Section \ref{sec_performance_analysis}.
-Next, it is shown that the $IBM$ is space filling.
-
-\begin{definition}
-A \textit{mutual} boundary between boxes $b^x$ and $b^y$ is when there exists an $r$ such that $u^x_r = l^y_r$ or $l^x_r = u^y_r$.
-\end{definition}
-
-\begin{theorem} An iterative box mesh composed of $n$ unique control points in $d$ dimensions, $C \subset \mathbb{R}^d$, is space filling. \end{theorem}
-\begin{proof}
-The first box added to an $IBM$ is space filling by construction.
-Assume that an $IBM$ composed of $n$ boxes is space filling and a new box centered at $x \in \mathbb{R}^d$ is added.
-Consider a point $y \in \mathbb{R}^d$.
-Every box $b^{x^{(i)}}$ that supported $y$ before the addition of $b^x$ and was reshaped along dimension $r \in [1,d]$ has a mutual boundary with $b^x$ by construction.
-Now either $\bigl| x_r - x^{(i)}_r \bigr| \geq \bigl| y_r - x^{(i)}_r \bigr|$ and $b^{x^{(i)}}$ still supports $y$, or $\bigl| x_r - x^{(i)}_r \bigr| < \bigl| y_r - x^{(i)}_r \bigr|$ and $b^x$ supports $y$.
-Therefore, the $IBM$ composed of $n+1$ boxes is still space filling.
-\end{proof}
+As mentioned earlier, the iterative box mesh is a covering for $(L,U)$ by construction and this can be proved by an inductive argument.
 
@@
 A well-studied technique for classification and approximation is the nearest neighbor algorithm \cite{cover1967nearest}.
-The convex region surrounding a central control point in which all points are closer to that center than any other control point is also called the Voronoi cell \cite{dirichlet1850reduction}.
-The Voronoi mesh ($VM$) utilizes Voronoi cells to define support via a generic basis function $v: \mathbb{R}^d \rightarrow \mathbb{R}$ with
+Nearest neighbor inherently utilizes the convex region $v^{x^{(i)}}$ (Voronoi cell \cite{dirichlet1850reduction}) consisting of all points closer to $x^{(i)}$ than any other point $x^{(j)}$.
+The Voronoi mesh smooths the nearest neighbor approximation by utilizing the Voronoi cells to define support via a generic basis function $V: \mathbb{R}^d \rightarrow \mathbb{R}_+$ given by
 
+$$ V^{x^{(i)}}(y) = \left(1 - \frac{||y - x^{(i)}||_2}{2 \ d(y - x^{(i)} \mid x^{(i)})} \right)_+, $$
+where $x^{(i)}$ is the center of the Voronoi cell, $y \in \mathbb{R}^d$ is an interpolation point, and $d(w \mid x^{(i)})$ is the distance between $x^{(i)}$ and the boundary of the Voronoi cell $v^{x^{(i)}}$ in the direction $w$.
+$V^{x^{(i)}}\bigl(x^{(j)}\bigr) = 0$ and $V^{x^{(i)}}$ has local support.
+While all basis functions $V^{x^{(i)}}(x^{(i)}) = 1$, the $2$ in the denominator causes all basis functions to go linearly to $0$ at the boundary of the twice-expanded Voronoi cell.
+Note that this basis function is $C^0$ because the boundaries of the Voronoi cell are $C^0$.
+In the case that there is no boundary along the vector $w$, the basis function value is always $1$.
+
+While the cost of computing the exact Voronoi cells for any given set of points grows exponentially \cite{dutour2009complexity}, the calculation of $d$ is linear with respect to the number of control points and dimensions.
+Given any center $x^{(i)} \in \mathbb{R}^d$, set of control points $C \subseteq X$, and interpolation point $y \in \mathbb{R}^d$, $d\bigl((y-x^{(i)}) \mid x^{(i)}\bigr)$ is the solution to
 \begin{equation}
-v^c(z) = \left(1 - \frac{||z - c||_2}{2 \ dist(z - c \mid c)} \right)_+
-\label{eq_voronoi_basis}
-\end{equation}
-
-where $c \in \mathbb{R}^d$ is the center of the Voronoi cell, $z \in \mathbb{R}^d$ is an interpolation point, and $dist(\ \cdot \mid c)$ is the distance between $c$ and the nearest boundary of the cell along the vector $\cdot\ $.
-While all basis functions $v^c(c) = 1$, the $2$ in the denominator causes all basis functions to go linearly to $0$ at the boundary of the twice-expanded Voronoi cell.
-Note that this basis function is $C^0$ because the boundaries of the Voronoi cell are $C^0$.
-In the case that there is no boundary along the vector $x-c$, the basis function value is always $1$.
-
-While the cost of computing all Voronoi cells for any given set of points grows exponentially \cite{dutour2009complexity}, the calculation of $dist$ is linear with respect to the number of control points and dimensions.
-Given any center $c^{(1)} \in \mathbb{R}^d$, set of control points $C \subset \mathbb{R}^d$, and interpolation point $z \in \mathbb{R}^d$, $dist\bigl((z-c^{(1)}) \mid c^{(1)}\bigr)$ is the solution to
-\begin{equation}
-  \max_{\{c \in C\}\backslash c^{(1)}} \frac{|z|}{2} \ \frac{z \cdot \bigl(c - c^{(1)}\bigr) - c^{(1)} \cdot \bigl(c - c^{(1)}\bigr)}{c \cdot \bigl(c - c^{(1)}\bigr) - c^{(1)} \cdot \bigl(c - c^{(1)}\bigr)}.
+  \max_{\{c \in C\}\backslash x^{(i)}} \frac{\|y\|_2}{2} \ \frac{y \cdot \bigl(c - x^{(i)}\bigr) - x^{(i)} \cdot \bigl(c - x^{(i)}\bigr)}{c \cdot \bigl(c - x^{(i)}\bigr) - x^{(i)} \cdot \bigl(c - x^{(i)}\bigr)}.
 \end{equation}
@@
 The parameters of the $VM$ can now be computed exactly as for the $MBM$ and $IBM$.
-The vornonoi mesh is used to generate a $|X| \times |C|$ matrix $M$ of basis function evaluations at all points in $X$.
+The Voronoi mesh is used to generate a $|X| \times |C|$ matrix $M$ of basis function evaluations at all points in $X$.
 Now the set of coefficients for each basis function is the solution to the least squares problem $\min_P \bigl\| M \ P - f(X) \bigr\|_2$.
@@
 \begin{enumerate}
-\item Pick the most central point $x^{(1)} \in X$ to be the first control point in $VM$ and fit the mesh to all $X$.
-\item Add a control point $x^{(i)} \notin VM$ that has the most approximation error and then fit the mesh to all $X$. \label{step_add_control}
+\item Pick the most central point $x^{(1)} \in X$ to be the first control point in $C$ and fit the $VM$ to all $X$.
+\item Add a control point $x^{(i)} \notin C$ such that $x^{(i)} = \max_{x \in X \backslash C} \bigl| \hat f(x) - f(x) \bigr|$ and then fit the $VM$ to all $X$. \label{step_add_control}
 \item Repeat Step \ref{step_add_control} until approximation error is below tolerance $t$.
@@
 
-Any $VM$ is na\"{\i}vely space filling, since any possible interpolation point will have a nearest neighbor control point when the set of control points is nonempty.
-
-\begin{theorem}
-  A Voronoi mesh composed of $n$ unique control points in $d$ dimensions, $C \subset \mathbb{R}^d$, is space filling.
-\end{theorem}
-\begin{proof}
-  Consider interpolating at point $z$.
-Let $c^{(1)}$ be the $c \in C$ that minimizes $||z-c||_2$.
-Now, $z$ must be in the support of the basis function $v^{c^{(1)}}$.
-Therefore the Voronoi mesh is space filling.
-\end{proof}
-
+Any $VM$ is na\"{\i}vely a covering for $(L,U)$, since any possible interpolation point will have a nearest neighbor control point.
 The computational complexity of evaluating a parameterized Voronoi mesh with $c$ control points is $\mathcal{O}(c^2 d)$.
-Bootrsapping the generation of a Voronoi mesh requres $\mathcal{O}(c^2 n d)$ computations for a maximum number of basis functions $c$ determined by the error threshold.
+Bootstrapping the generation of a Voronoi mesh requires $\mathcal{O}(c^2 n d)$ computations for a maximum number of basis functions $c$ determined by the error threshold.
 
@@
 In order to evaluate the proposed interpolation and approximation techniques, this paper utilizes three data sets of varying dimension and application.
-In the following subsections the sources and targets of each data set are described, as well as challenges and limitations related to doing interpolation and approximation on these data sets.
+In the following subsections the sources and targets of each data set are described, as well as challenges and limitations related to interpolating and approximating these data sets.
 The distributions of response values being modeled can be seen in Figure \ref{fig_response_hists}.
@@
 $$N = 532, D = 4$$
-The first of three data sets is a four-dimensional dataset produced by executing the IOzone benchmark from \cite{iozone} on a homogeneous cluster of computers.
+The first of three data sets is a four-dimensional data set produced by executing the IOzone benchmark from \cite{iozone} on a homogeneous cluster of computers.
 The system performance data was collected by executing IOzone 40 times for each of a select set of system configurations.
@@
 The 40 executions for each system configuration are converted to their mean, which is capable of being modeled by each of the multivariate approximation techniques presented in Section \ref{sec_mesh_construction}.
-The four dimensions being modelled to predict throughput mean are file size, record size, thread count, frequency.
+The four dimensions being modeled to predict throughput mean are file size, record size, thread count, and frequency.
 
@@
 The forest fire data set provided by \cite{cortez2007data}, describes the area of Montesinho park burned on specific days and months of the year in terms of the environmental conditions.
-The twelve dimensions being used to model burn area are the $x$ and $y$ spatial coordinates of burns in the park, month and day of year, the FFMC, DMC, DC, and ISI indices (see source for details), the temperature in celcius, relative humidity, wind speed, and outdoor rain.
+The twelve dimensions being used to model burn area are the $x$ and $y$ spatial coordinates of burns in the park, month and day of year, the FFMC, DMC, DC, and ISI indices (see source for details), the temperature in Celsius, relative humidity, wind speed, and outdoor rain.
 The original analysis of this data set demonstrated it to be difficult to model, likely due to the skew in response values.
@@
 
-The final data set for evaluation is derived from a speech monitoring study with the intent to automatically estimate Parkinson's disease symptom development in parkinsons patients from \cite{tsanas2010accurate}.
-Particularly, a time-consuming clincal evaluation measure referred to as the UPDRS score.
+The final data set for evaluation is derived from a speech monitoring study with the intent to automatically estimate Parkinson's disease symptom development in Parkinson's patients from \cite{tsanas2010accurate}.
+Particularly, a time-consuming clinical evaluation measure referred to as the UPDRS score.
 The total UPDRS score given by a clinical evaluation is estimated through 16 real numbers generated from biomedical voice measures of in-home sound recordings.
@@
 The columns are for Max Box Mesh, Iterative Box Mesh, and Voronoi Mesh, respectively.
-The rows are for HPC I/O, Forest Fire, and Parkinsons respectively.
-Notice that the techniques exibit counterintuitive behavior on the Parkinsons and Forest Fire data sets, performance increases with increasing error tolerance.}
+The rows are for HPC I/O, Forest Fire, and Parkinson's respectively.
+Notice that the techniques exhibit counterintuitive behavior on the Parkinson's and Forest Fire data sets, performance increases with larger error tolerance.}
   \label{fig_all_performance}
@@
 Before the models are constructed all data $X$ values are shifted and scaled to be in the unit cube $[0,1]$, while the response values are taken in their original form.
-All models are evaluated with a 10-fold cross validation over 80:20 splits of the data.
+All models are evaluated with $10$ random $80/20$ splits of the data.
 
@@
     \hline
-    \textbf{Data Set} & \textbf{Technique} & \textbf{Tolerance} & \textbf{Average Errror}\\
+    \textbf{Data Set} & \textbf{Technique} & \textbf{Tolerance} & \textbf{Average Error}\\
     \hline
@@
     Forest Fire & MBM & 1.8 & 3.517\\
-    Parkinsons & MBM & 0.6 & 0.114\\
+    Parkinson's & MBM & 0.6 & 0.114\\
     \hline
@@
     Forest Fire & IBM & 1.8 & 3.615\\
-    Parkinsons & IBM & 1.8 & 0.121\\
+    Parkinson's & IBM & 1.8 & 0.121\\
     \hline
@@
     Forest Fire & VM & 1.0 & 4.783\\
-    Parkinsons & VM & 2.0 & 1.824\\
+    Parkinson's & VM & 2.0 & 1.824\\
     \hline
   \end{tabular}
-  \caption{The optimial error tolerance bootstrapping parameters for each technique and each data set as well as the average absolute relative errors achieved by that tolerance.
+  \caption{The optimal error tolerance bootstrapping parameters for each technique and each data set as well as the average absolute relative errors achieved by that tolerance.
 Notice that large relative error tolerances occasionally yield even lower evaluation errors, demonstrating the benefits of approximation over interpolation for noisy data sets.
@@
 The columns are for Max Box Mesh, Iterative Box Mesh, and Voronoi Mesh, respectively.
-The rows are for HPC I/O, Forest Fire, and Parkinsons respectively.
+The rows are for HPC I/O, Forest Fire, and Parkinson's respectively.
 Notice that the Max Box Mesh exibits the least skewed distribution of errors across all three data sets.
@@
 Figure \ref{fig_eval_times} presents the time required to construct approximations for each model and each data set with varying $t$.
-The rapid reducting in computation time required for the forest fire and HPC I/O data sets suggests that large reductions in error can be achieved with relatively few basis functions.
+The rapid reduction in computation time required for the forest fire and HPC I/O data sets suggests that large reductions in error can be achieved with relatively few basis functions.
 The Parkinson's data set however presents a more noisy response, with increasing number of basis functions reducing error much less quickly.
@@
 The distributions of errors experienced by each approximation technique when the optimal bootstrapping relative error tolerance is selected can be seen in Figure \ref{fig_perf_sample}.
-HPC I/O exibits the most normal approximation errors, which suggests that the models are converging on the latent noise of the response for the data set.
+HPC I/O exhibits the most normal approximation errors, which suggests that the models are converging on the random noise of the response for the data set.
 The worst relative approximation errors are produced by the Voronoi mesh on the forest fire data set.
-The small magnitude true response values contribute to the larger relative errors, however the $VM$ performances are still unacceptably large.
+The small magnitude true response values contribute to the larger relative errors.
+Regardless, the $VM$ errors are unacceptably large.
 
@@
 
-The bootstrapping procdure presented for each approximation technique still has much room for improvement.
-Initial analysis suggests that the appropriate relative error tolerance needs to be discovered emperically for each application of a modelling technique.
+The bootstrapping procedure presented for each approximation technique still has much room for improvement.
+Initial analysis suggests that the appropriate relative error tolerance needs to be discovered empirically for each application of a modeling technique.
 Further analytic studies could arrive at methods for determining optimal error tolerances at runtime, however increases in runtime complexity may not be afforded in many applications.
@@
 The Voronoi mesh attempts to address this problem by utilizing boundaries between points in multiple dimensions simultaneously.
-However, it is empirically unclear whether true benefits are seen in applications where $d < n$.
+However, it is empirically unclear whether the true benefits of the $VM$ are seen in applications where $d < n$.
 

### 2018-02-13 Paper_[2018-01-20]_IEEESE/paper.tex
Finalized some more modifications on the ACM paper. Ready to deliver to Dr. Watson for further review.

@@
 H.
-Lux$^{1}$, Layne Watson$^{2}$, Jon Bernard$^{1}$, Tyler H.
-Chang$^{1}$, Bo Li$^{1}$, \\ Li Xu$^{3}$, Godmar Back$^{2}$, Ali R.
+Lux$^{1}$, Layne T.
+Watson$^{2}$, Jon Bernard$^{1}$, Tyler H.
+Chang$^{1}$, Bo Li$^{1}$, Xiaodong Yu$^{1}$,\\ Li Xu$^{3}$, Godmar Back$^{2}$, Ali R.
 Butt$^{2}$, Kirk W.
-Cameron$^{2}$, Yili Hong$^{4}$% <-this % stops a space
-\thanks{*This work was supported by the National Science Foundation}%
-\thanks{$^{1}$T.
-Lux, J.
-Bernard, T.
-Chang, B.
-Li are doctoral students of Computer Science at Virginia Polytechnic Institute and State University, Blacksburg, Virginia 24060, USA {\tt\small (tchlux at vt.edu)}}%
-\thanks{$^{2}$L. Watson, G. Back, A. Butt, and K. Cameron are Faculty of Computer Science at Virginia Polytechnic Institute and State University, Blacksburg, Virginia 24060, USA}%
-\thanks{$^{3}$L. Xu is a doctoral student of Statistics at Virginia Polytechnic Institute and State University, Blacksburg, Virginia 24060, USA}%
-\thanks{$^{4}$Y. Hong is with the Faculty of Statistics at Virginia Polytechnic Institute and State University, Blacksburg, Virginia 24060, USA}%
+Cameron$^{2}$, Yili Hong$^{4}$, Danfeng Yao$^{2}$% <-this % stops a space
+\thanks{*This work was supported by the National Science Foundation Grant CNS-1565314.}%
+\thanks{$^{1}$Doctoral student, Department of Computer Science, Virginia Polytechnic Institute and State University, Blacksburg, Virginia 24060, USA {\tt\small (tchlux at vt.edu)}}%
+\thanks{$^{2}$Department of Computer Science, Virginia Polytechnic Institute and State University, Blacksburg, Virginia 24060, USA}%
+\thanks{$^{3}$Doctoral student, Department of Statistics, Virginia Polytechnic Institute and State University, Blacksburg, Virginia 24060, USA}%
+\thanks{$^{4}$Department of Statistics, Virginia Polytechnic Institute and State University, Blacksburg, Virginia 24060, USA}%
 }
@@
 Performance variability can have a significant impact on many applications of computing.
-Cloud Computing, High Performance Computing, and Computer Security communities each exert considerable effort managing and analyzing variability throughout the system stack.
+Cloud computing, high performance computing, and computer security communities each exert considerable effort managing and analyzing variability throughout the system stack.
 This work presents and evaluates a methodology for predicting precise characteristics of the computational performance variability of an input/output (I/O) application over varying system configurations.
@@
 Predicting performance variability in a computer system is a challenging problem that has primarily been attempted in one of two ways: (1) build a statistical model of the performance data collected by running experiments on the system at select settings, or (2) run artificial experiments using a simplified simulation of the target system to estimate architecture and application bottlenecks.
-In this paper, the proposed modeling techniques rest in the first category and they represent a notable increase in the ability to model precise characteristics of variability.
+In this paper, the proposed modeling techniques rest in the first category and represent a notable increase in the ability to model precise characteristics of variability.
 
@@
 Previous techniques for predicting system performance have strictly modeled real-valued summary statistics because there exists a large base of mathematical techniques capable of approximating functions of the form $f: \mathbb{R}^d \rightarrow \mathbb{R}$.
-There is a smaller theoretical foundation for approximating functions of the form $f: \mathbb{R}^d \rightarrow g$ with $g: \mathbb{R} \rightarrow \mathbb{R}$.
-The latter is the approximation of functions.
+There is little systems work in approximating functions of the form $f: \mathbb{R}^d \rightarrow \{g \mid g: \mathbb{R} \rightarrow \mathbb{R}\}$.
 
-A major hurdle when modeling functions such as the CDF or PDF, is that certain properties must be maintained in order for the functions to remain valid.
-It is necessary that a PDF $f: \mathbb{R} \rightarrow \mathbb{R}$ have the properties $f(x) > 0$ and $\int_{-\infty}^{\infty}f(x)dx = 1$.
-However, for a CDF $F: \mathbb{R} \rightarrow \mathbb{R}$ the properties are $F(x) \in [0,1]$ and $F'(x) \geq 0$.
+A major hurdle when modeling functions such as the CDF or PDF is that certain properties must be maintained.
+It is necessary that a PDF $f: \mathbb{R} \rightarrow \mathbb{R}$ have the properties $f(x) \geq 0$ and $\int_{-\infty}^{\infty}f(x)dx = 1$.
+However, for a CDF $F: \mathbb{R} \rightarrow \mathbb{R}$ the properties are $F(x) \in [0,1]$ and $F(x)$ is monotone increasing.
 For the purpose of function approximation, the properties of the CDF are more straightforward to maintain.
@@
 Next, the three techniques used to model CDFs are summarized, a mechanism for identifying important features is proposed, and the error measure used throughout this work is presented.
-For each of the following, $n \in \mathcal{N}$ refers to the number of training samples (known system configurations), $d \in \mathcal{N}$ refers to the dimension of the data (ordinal system parameters), $x^{(i)} \in \mathbb{R}^d$ is a known system configuration with known CDF $F_{x^{(i)}}$, and $y \in \mathbb{R}^d$ refers to a previously unseen system configuration with CDF $F_y$.
+For each of the following, $n \in \mathcal{N}$ refers to the number of training samples (known system configurations), $d \in \mathcal{N}$ refers to the dimension of the data (ordinal system parameters), $x^{(i)} \in \mathbb{R}^d$ is a known system configuration with known performance CDF $F_{x^{(i)}}$, and $y \in \mathbb{R}^d$ refers to a previously unseen system configuration with performance CDF $F_y$.
 
@@
 
-The Delaunay method of interpolation is a well studied geometric technique for producing an interpolant \cite{lee1980two}.
-The Delaunay triangulation of a set of data points into simplices is such that the sphere defined by the vertices of each simplex contains no data points in the sphere's interior.
-For a $d$-simplex S with vertices $x^{(0)}$, $x^{(1)}$, $\ldots$, $x^{(d)}$, and functions $F_{x^{(i)}}$, $i=0$, $\ldots$, $d$, $y$ is a unique convex combination of the vertices,
+The Delaunay triangulation is a well-studied geometric technique for producing an interpolant \cite{lee1980two}.
+The Delaunay triangulation of a set of data points into simplices is such that there are no data points inside the sphere defined by the vertices of each simplex.
+For a $d$-simplex S with vertices $x^{(0)}$, $x^{(1)}$, $\ldots$, $x^{(d)}$, and functions $F_{x^{(i)}}$, $i=0$, $\ldots$, $d$, $y \in S$ is a unique convex combination of the vertices,
 
@@
 
-and the Delaunay interpolant $F_y$ at $x$ is given by
+and the Delaunay interpolant $F_y$ at $y$ is given by
 
@@
 The KS Statistic (red arrow) computed between the true CDF (solid line) and predicted CDF (dashed line) is 0.2 for this example.
-This prediction is deemed a failure by p-values 0.05 and 0.01, however it is not a failure according to p-values 0.001 and 1.0e-6.
+The KS test null hypothesis is rejected by $p$-value 0.01, however it is not rejected by $p$-value 0.001.
   \vspace{-.1cm}}
@@
 
-The computational complexity of interpolating via Delaunay (for the implementation used) is approximately $\mathcal{O}(n d^4 log d)$, which is capable of scaling reasonably to $d \leq 50$.
-In the present application, Delaunay is used to select up to $d+1$ system configurations and assign weights to each in order to generate a prediction for a new system configuration.
+The computational complexity of Delaunay interpolation (for the implementation used) is $\mathcal{O}(n d^4 \log d)$, which is capable of scaling reasonably to $d \leq 50$.
+In the present application, a Delaunay simplex $S$ containing $y$ is found, then the $d+1$ vertices (system configurations) of $S$ are used to assign weights to each vertex and produce the predicted CDF for system configuration $y$.
 
@@
 
-The Max Box mesh is an interpolation technique that utilizes overlapping box splines \cite{de2013box} as basis functions that are shifted and scaled to support box shaped regions.
-The boxes are constructed such that each box has strictly one point on its interior while maintaining the largest minimum distance between the interior point and the closest face of the box (a measure of centrality).
-Given a set of box splines $b^{x^{(i)}}$ with the max box properties anchored at each $x^{(i)}$,
+The Max Box mesh is an interpolation technique that utilizes overlapping box splines \cite{de2013box} as basis functions that are shifted and scaled to have support over box shaped regions.
+The boxes are constructed such that each box has exactly one point in its interior while maintaining the largest minimum distance between the interior point and the closest face of the box (a measure of centrality).
+Given a set of box splines $\bigl\{b^{x^{(i)}}\bigr\}$ with the max box properties anchored at interior points $\bigl\{x^{(i)}\bigr\}$,
 
-$$ F_y = \frac{\sum_i b^{x^{(i)}}(y) F_{x^{(i)}}}{\sum_i b^{x^{(i)}}(y)}, $$
-
-where $b^{x^{(i)}}(y) \geq 0$ and the denominator causes the sum of weights to be one.
-The computational complexity of interpolation via the Max Box mesh is $\mathcal{O}(n^2 d)$.
-The Max Box mesh interpolates a function $F_y$ by producing a convex combination of known functions $F_{x^{(i)}}$, where the weights are determined by the box splines.
+$$ F_y = \frac{\sum_i b^{x^{(i)}}(y) F_{x^{(i)}}}{\sum_i b^{x^{(i)}}(y)}. $$
+Note that the box splines always satisfy $b^{x^{(i)}}(y) \geq 0$. The computational complexity of interpolation via the Max Box mesh is $\mathcal{O}(n^2 d)$.
 
@@
 A well-studied technique for classification and approximation is the nearest neighbor algorithm \cite{cover1967nearest}.
-Nearest neighbor inherently utilizes the convex region surrounding a central control point in whose interior all points are closer to that center than any other control point, which is also called the Voronoi cell \cite{dirichlet1850reduction}.
-The Voronoi mesh smooths the nearest neighbor approximation by utilizing the Voronoi cells to define support via a generic basis function $v: \mathbb{R}^d \rightarrow \mathbb{R}_+$ with
+Nearest neighbor inherently utilizes the convex region $C^{x^{(i)}}$ (Voronoi cell \cite{dirichlet1850reduction}) consisting of all points closer to $x^{(i)}$ than any other point $x^{(j)}$.
+The Voronoi mesh smooths the nearest neighbor approximation by utilizing the Voronoi cells to define support via a generic basis function $v: \mathbb{R}^d \rightarrow \mathbb{R}_+$ given by
 
-$$ v^{x^{(i)}}(y) = \left(1 - \frac{||y - x^{(i)}||_2}{2 \ d(y - x^{(i)} \mid x^{(i)})} \right), $$
-
-where $d(\ \cdot \mid x^{(i)})$ is the distance between $x^{(i)}$ and the nearest boundary of the Voronoi cell along the vector $\cdot\ $.
-While all basis functions $v^{x^{(i)}}(x^{(i)}) = 1$, the $2$ in the denominator causes all basis functions to go linearly to $0$ at the boundary of the twice-expanded Voronoi cell.
-Now we have
+$$ v^{x^{(i)}}(y) = \left(1 - \frac{||y - x^{(i)}||_2}{2 \ d(y - x^{(i)} \mid x^{(i)})} \right)_+, $$
+where $d(w \mid x^{(i)})$ is the distance between $x^{(i)}$ and the boundary of the Voronoi cell $C^{x^{(i)}}$ in the direction $w$.
+$v^{x^{(i)}}\bigl(x^{(j)}\bigr) = 0$ and $v^{x^{(i)}}$ has local support, giving the interpolated value
 
 $$ F_y = \frac{\sum_i v^{x^{(i)}}(y) F_{x^{(i)}}}{\sum_i v^{x^{(i)}}(y)}, $$
-
-where $v^{x^{(i)}}(y) \geq 0$ by definition and the denominator causes the sum of weights to be one.
-The computational complexity of interpolation via Voronoi mesh is $\mathcal{O}(n^2 d)$.
-The Voronoi mesh interpolates a function $F_y$ just as the Max Box mesh and Delaunay triangulation, by a convex combination of known functions $F_{x^{(i)}}$.
+where $0 \leq v^{x^{(i)}}(y) \leq 1$.
+The computational complexity of interpolation via the Voronoi mesh is $\mathcal{O}(n^2 d)$.
+All of the approximations are an interpolant involving a convex combination of known functions $F_{x^{(i)}}$.
 %% ===================================================================
@@
 
-Machine learning has demonstrated that an important procedure in any application of predictive methodologies is identifying those features of training data that are most relevant to making accurate predictions \cite{guyon2003introduction}.
+It is well-known that an important procedure in any application of predictive methodologies is identifying those features of the data that are most relevant to making accurate predictions \cite{guyon2003introduction}.
 Selection strategies such as the floating searches studied in \cite{pudil1994floating} or others compared in \cite{ferri1994comparative} can be too expensive for large approximation problems.
 Rather, this work poses feature selection as a continuous optimization problem.
-Let $X$ be an $n \times d$ matrix of $n$ known system configurations with $d$ parameters each normalized to be in the unit cube.
-Define an error function $err: \mathbb{R}^d \rightarrow \mathbb{R}_+$ that estimates the error of a predictive model trained on $X diag(w)$, $w \in \mathbb{R}^d$,  by performing 10-fold cross validation using 80\% of the rows of $X diag(w)$ as training and 20\% as testing.
-A minimal solution to this error function could be considered an optimal weighting of the features of $X$.
-Minimization is performed using the modified simulated annealing technique presented in \cite{lux2016convergence} because of its empirically fast convergence in the absence of a readily computable gradient.
+Let $X$ be an $n \times d$ matrix of $n$ known system configurations with $d$ parameters each normalized to be in $[0,1]$.
+Define an error function that computes the error of a predictive model trained on $X\hbox{ diag }w$, $w \in \mathbb{R}^d$,  by performing ten random splits with 80\% of the rows of $X\hbox{ diag }w$ for training and 20\% for testing.
+A minimum of this error function could be considered an optimal weighting of the features of $X$.
+Minimization is performed using a zero order method in the absence of a readily computable gradient.
 
@@
 
-The two-sample KS test is a useful nonparametric test for declaring when two CDFs do not originate from the same underlying distribution while only assuming stationarity, finite mean, and finite variance.
+The two-sample KS test is a useful nonparametric test for comparing two CDFs while only assuming stationarity, finite mean, and finite variance.
 The null hypothesis (that two CDFs come from the same underlying distribution) is rejected at level $p \in [0,1]$ when
 
-$$ KS > \sqrt{-\frac{1}{2}ln\big(\frac{p}{2}\big)} \sqrt{\frac{1}{n_1} + \frac{1}{n_2}}, $$
-
+$$ KS > \sqrt{-\frac{1}{2}\ln\biggl(\frac{p}{2}\biggr)} \sqrt{\frac{1}{n_1} + \frac{1}{n_2}}, $$
 with distribution sample sizes $n_1,n_2 \in \mathcal{N}$.
@@
   \end{tabular}
-  \caption{A description of system parameters considered for IOzone. Record size must be less than file size during execution.
+  \caption{A description of system parameters considered for IOzone. Record size must be $\leq$ file size during execution.
     \vspace{-.5cm}}
@@
 While the CPU frequency varies depending on the test configuration, the I/O from IOzone is performed by an ext4 filesystem sitting above an Intel SSDSC2BA20 SSD drive.
-At the time of data collection, a Linux kernel version of 4.13.0 was used.
+At the time of data collection, Linux kernel Version 4.13.0 was used.
 The system performance data was collected over two weeks by executing IOzone 150 times for each of a select set of approximately 18K system configurations, for a total of approximately 2.7M executions of IOzone.
-A single IOzone execution reports the max I/O throughput in kilobytes per second seen for the selected test.
-The summary of the data generated in the experiments for this paper can be seen in Table \ref{tab:data_description}.
+A single IOzone execution reports the max I/O throughput in kilobytes per second seen for the selected test type.
+The summary of the data components in $x^{(i)}$ for the experiments for this paper can be seen in Table \ref{tab:data_description}.
 Distributions of raw throughput values being modeled can be seen in Figure \ref{fig:throughput_histogram}.
@@
 All 150 repeated trials for a system configuration were grouped with that configuration.
-The only non-ordinal feature in this data is the test.
-All tests were treated as different applications and were separated for modeling and analysis.
-(i.e.
-predictions for the ``readers'' test were made using known configurations for the ``readers'' test only.) The division of non-ordinal features into separate models is necessary for the presented methodology, however future improvements are discussed in Section \ref{sec:discussion}.
+The only nonordinal feature in this data is the test type.
+All test types were treated as different applications and were separated for modeling and analysis, i.e., predictions for the ``readers'' test type were made using only known configurations for the ``readers'' test type.
 
@@
   \end{tikzpicture}
-  \caption{Histograms of the prediction error averaged over 10 folds for each modeling algorithm when trained with 80\% of the data.
+  \caption{Histograms of the prediction error for each modeling algorithm from ten random splits when trained with 80\% of the data.
 The distributions show the KS statistics for the predicted throughput distribution versus the actual throughput distribution.
-The four vertical red lines represent commonly used p-values \{0.05, 0.01, 0.001, 1.0e-6\} respectively.
-All predictions to the right of a red line represent failed predictions that are significantly different (by respective p-value) from the actual distribution according to the KS test.
+The four vertical red lines represent commonly used $p$-values \{0.05, 0.01, 0.001, 1.0e-6\} respectively.
+All predictions to the right of a red line represent CDF predictions that are significantly different (by respective $p$-value) from the actual distribution according to the KS test.
   \vspace{-.1cm}}
@@
 All three interpolation techniques are used to predict the distribution of I/O throughput values at previously unseen system configurations.
-In order to improve robustness of the error analysis, 10-fold cross validation is used.
-10 random selections of 80\% of the IOzone data are used to train each model and the remaining 20\% provide estimations of approximation error for each model.
-The recorded errors are averaged over the 10 random folds and grouped by unique system configuration.
-The folds are identically seeded for each interpolation technique, ensuring consistency in the training and testing sets.
+In order to improve robustness of the error analysis, ten random selections of 80\% of the IOzone data are used to train each model and the remaining 20\% provide approximation error for each model.
+The recorded errors are grouped by unique system configuration and then averaged within each group.
+The samples are identical for each interpolation technique, ensuring consistency in the training and testing sets.
 
 The aggregation of errors across all IOzone tests given 80\% of the data as training can be seen in Figure \ref{fig:ks_histogram_80_20}.
-Agglomerate errors for each technique present similarly, each resembles a Gamma distribution.
-The percentages of failed predictions with varying p-values are on display in Table \ref{tab:p_value_failure_rate}.
-The primary p-value used for analyses in this work is 0.001.
-This p-value is chosen because close to 2K predictions are made for each test.
-Also, applications executed in cloud and HPC systems that could benefit from statistical modeling will be executed at least on the order of 1000's of times.
+Agglomerate errors for each technique resemble a Gamma distribution.
+The percentages of significant prediction errors with varying $p$-values are on display in Table \ref{tab:p_value_failure_rate}.
+The primary $p$-value used for analyses in this work is 0.001, chosen because close to 2K predictions are made for each test type.
+Also, applications executed in cloud and HPC systems that could benefit from statistical modeling will be executed at least thousands of times.
 In line with this knowledge, it is important to ensure that only a small fraction of interpretable results could occur solely under the influence of random chance.
-When considering the p=0.001 results for each technique, a little under half of the generated predictions fail.
-In a traditional application of machine learning techniques a failure rate of 45\% would be a poor result, however in this situation the complexity of the problem warrants a slightly different interpretation.
-These predictions are of a very \textit{precise} characterization of performance variability, in fact the strongest possible characterization of variability.
+When considering the $p=0.001$ results for each technique, a little under half of the predicted CDFs are significantly different from the measured (and presumed) correct CDFs.
+A rejection rate of 45\% would seem a poor result, however in this situation the complexity of the problem warrants a slightly different interpretation.
+These predictions are a very \textit{precise} characterization of performance variability, in fact the strongest possible characterization of variability.
 Globally, only a little under half of the predictions fail to capture \textit{all} of the characteristics of performance variability at new system configurations.
@@
     \hline
-    \textbf{Algorithm} & \textbf{P-Value} & \textbf{\% Failed Predictions} \\
+    \textbf{Algorithm} & \textbf{$P$-Value} & \textbf{\% N.H. Rejections} \\
     \hline
@@
   \end{tabular}
-  \caption{The prediction failure rate by the KS-test when provided different selections of p-values. These accompany the results from Figure \ref{fig:ks_histogram_80_20}. 
+  \caption{Percent of null hypothesis rejections rate by the KS-test when provided different selections of $p$-values.
+These accompany the percent of null hypothesis rejection results from Figure \ref{fig:ks_histogram_80_20}.
     \vspace{-.5cm}}
@@
 
-While interpreting failure rates of these interpolation techniques, it is important to consider at what rate the error reduces with increasing amounts of training data.
-Figure \ref{fig:ks_failure_by_training} displays the change in p=0.001 failure rate with increasing density of training data up to the maximum density allowed by this training set.
-Delaunay provides the best results with the least training data by about 5\%, but these low-density prediction error rates are unacceptably high (90\% failure).
-Increased density demonstrates a roughly $(1 - \sqrt{n})$ rate of reduction in error that will level off around 20-30\% with a training set twice the size of the one collected for this study.
-However, constructing a training set of such a size would likely require months of compute time and be infeasible for many applications.
-The rate of convergence seen also suggests that these models are successfully converging on the lowest possible error given the (presumed) Lipschitz constant of the data and any latent variables that may perfectly describe system I/O throughput variability.
+While interpreting null hypothesis rejection rates for these interpolation techniques, it is important to consider how the rejection rate reduces with increasing amounts of training data.
+Figure \ref{fig:ks_failure_by_training} displays the change in $p=0.001$ null hypothesis rejection rate with increasing density of training data up to the maximum density allowed by this set.
+Delaunay interpolation provides the best results with the least training data by about 5\%, but these low density rejection rates are unacceptably high (90\%).
+Figure \ref{fig:ks_failure_by_training} clearly shows that this data set and/or the system variables used in the models of performance variability is inadequate to capture the full variability map from system parameters to performance CDF.
+Which or both obtains is not clear.
+A few well chosen data points can significantly improve the interpolants, and thus a careful study of the rejection instances is warranted, besides enlarging the set of system variables being modeled.
+
+It may be misleading to consider the global performance of each prediction technique across all test types, as some test types are more difficult than others to predict and have more apparent latent variables.
+In Figure \ref{fig:ks_failure_by_training_and_test}, the relative difficulty of each IOzone test type can be compared.
+The I/O test types analyzing reads are typically approximated with lower error than those test types analyzing writes.
+Regardless of test type, in the aggregate results the KS statistics hover consistently around 0.15, demonstrating an impressively low KS statistic for predictions.
+In order to address the opacity of aggregate analysis, another case study and an application of the methodology from Section \ref{sec:feature_weighting} is presented in Table \ref{tab:optimized_p_value_failure_rate}.
 
@@
     \node[below=of img, node distance=1cm, yshift=1cm] {Percentage of Training Data};
-    \node[left=of img, node distance=0cm, rotate=90, anchor=center,yshift=-0.7cm] {Percentage of Failed Predictions};
+    \node[left=of img, node distance=0cm, rotate=90, anchor=center,yshift=-0.7cm] {\% Null Hypothesis Rejections};
   \end{tikzpicture}
-  \caption{The performance of each algorithm with increasing amounts of training data estimated using 10-fold cross validation and the KS test with p=0.001.
-The KS statistic for any given training percentage is achieved by averaging over the KS statistic for all tests and system configurations.
-The training percentages range from 5\% to 95\% by increments of 5\%.
+  \caption{The performance of each algorithm on the KS test ($p=0.001$) with increasing amounts of training data averaged over all IOzone test types and ten random splits of the data.
+The training percentages range from 5\% to 95\% in increments of 5\%.
 Delaunay is the best performer until 95\% of data is used for training, at which Max Box mesh becomes the best performer by a fraction of a percent.
@@
 
-It may be misleading to consider the global performance of each prediction technique across all tests, as some tests are more difficult than others to predict and have more apparent latent activity.
-In Figure \ref{fig:ks_failure_by_training_and_test}, the relative difficulty of each IOzone test can be compared.
-The I/O tests analyzing reads are typically approximated with lower error than those tests analyzing writes.
-Regardless of test, in the aggregate results the modal KS statistics hover consistently around 0.15, demonstrating an impressively low KS statistic for modal predictions.
-In order to address the opacity of aggregate analysis, another case study and an application of the methodology from Section \ref{sec:feature_weighting} is presented in Table \ref{tab:optimized_p_value_failure_rate}.
+The results presented in Table \ref{tab:optimized_p_value_failure_rate} are achieved by permitting each approximation technique 300 iterations of simulated annealing.
+In each iteration, the impact of potential weights on the average KS statistic were considered.
+All weights were kept in the range [0,2], and were applied to the normalized features for frequency, file size, record size, and number of threads.
+All three approximation techniques had similar optimal weights of approximately $(.001, 2, 1.7, 1.5)$ for frequency, file size, record size, and number of threads, respectively.
+Recall that each interpolation technique uses small distances to denote large influences on predictions, meaning that frequency was the most important feature when predicting variability for the ``readers'' test type, followed not-so-closely by number of threads, then record size.
 
-The results presented in Table \ref{tab:optimized_p_value_failure_rate} are achieved by permitting each approximation technique 300 iterations of modified simulated annealing.
-In each iteration, the impact of potential weights on average validation KS statistic were considered.
-All weights were kept in the range [0,2], and were applied to the normalized features for frequency, file size, record size, and number of threads.
-All three techniques converged on similar optimal weights for the validation sets of approximately \{.001, 2, 1.7, 1.5\} for each respective feature.
-Recall that each interpolation technique uses smaller distances to denote larger influence on predictions, meaning that frequency was the most important feature when predicting variability for the ``readers'' test, followed not-so-closely by number of threads, then record size.
-
-The ``readers'' test results demonstrate that the underlying prediction techniques work and are capable of seeing failure rates below 5\% when optimized for a given application.
-It is important to emphasize that the roughly 95\% of predictions that did not fail are predicting the \textit{precise} distribution of I/O throughput that will be witnessed at a previously unseen system configuration.
-To present knowledge, there is no existing methodology that is generically applicable to any system performance metric, agnostic of system architecture, and capable of making such powerful predictions.
+The ``readers'' test type results demonstrate that the underlying prediction techniques work and are capable of seeing rejection rates below 5\% when tuned for a given application.
+It is important to emphasize that the roughly 95\% of predictions for which the null hypothesis was not rejected are predicting the \textit{precise} distribution of I/O throughput that will be witnessed at a previously unseen system configuration.
+To the authors' knowledge, there is no existing methodology that is generically applicable to any system performance metric, agnostic of system architecture, and capable of making such powerful predictions.
 
@@
     \hline
-    \textbf{Algorithm} & \textbf{P-Value} & \multilinecell{\textbf{Unweighted}\\\textbf{\% Failure}} & \multilinecell{\textbf{Weighted}\\\textbf{\% Failure}}\\
+    \textbf{Algorithm} & \textbf{$P$-Value} & \multilinecell{\textbf{Unweighted}\\\textbf{\% N.H. Rejection}} & \multilinecell{\textbf{Weighted}\\\textbf{\% N.H. Rejection}}\\
     \hline
@@
   \end{tabular}
-  \caption{The prediction failure rates by various p-values for the KS-test.
-These results are strictly for the ``readers'' IOzone test and show unweighted results as well as the results with weights optimized for minimum error (KS statistic) by 300 iterations of modified simulated annealing.
-Notice that the weights identified for the Delaunay model cause overfitting and reduction in apparent performance.
+  \caption{The null hypothesis rejection rates for various $p$-values with the KS-test.
+These results are strictly for the ``readers'' IOzone test type and show unweighted results as well as the results with weights tuned for minimum error (KS statistic) by 300 iterations of modified simulated annealing.
+Notice that the weights identified for the Delaunay model cause data dependent tuning, reducing performance.
 MaxBoxMesh performance is improved by a negligible amount.
-VoronoiMesh predictions are notably improved.
+VoronoiMesh performance is notably improved.
     \vspace{-.5cm}}
@@
 The KS statistic captures the worst part of any prediction and hence provides a conservatively large estimate of approximation error.
-The average absolute error in the predicted CDFs are always lower than the KS statistics.
+The average absolute errors in the predicted CDFs are always lower than the KS statistics.
 However, the KS statistic was chosen because of the important statistical theory surrounding it as an error measure.
@@
 As mentioned in Section \ref{sec:introduction}, no prior work has attempted to model an arbitrary performance metric for a system to such a high degree of precision.
-All previous statistical modeling attempts capture few (\textless 3) ordinal measures of performance.
+All previous statistical modeling attempts capture a few ($<3$) ordinal performance metrics.
 Generating models that have such high degrees of accuracy allows system engineers to identify previously unused configurations that present desired characteristics.
-Service level agreements (SLAs) in cloud computing environments are cause for capital competition that is impacted heavily by system performance \cite{patel2009service}.
-Users prefer SLAs that allow the most computing power per capital expense, incentivizing service providers to guarantee the greatest possible performance.
+Service level agreements (SLAs) in cloud computing environments are cause for capital competition that is affected heavily by system performance \cite{patel2009service}.
+Users prefer SLAs that allow the most computing power per monetary unit, incentivizing service providers to guarantee the greatest possible performance.
 Overscheduling and irregular usage patterns force cloud service providers to occasionally overload machines, in which case precise models of system performance can be used to statistically minimize the probability of SLA violation.
@@
 
-A final application domain that may also be affected by this methodology is computer security.
-Collocated users on cloud systems are a source of recent attention \cite{ali2015security}.
-If a malicious collocated user is capable of achieving specific insight into the configuration of the system, or the activity of other collocated users by executing performance evaluation programs (i.e.
-IOzone), a new attack vector may present itself.
+A final application domain affected by this methodology is computer security.
+Collocated users on cloud systems have received attention recently \cite{ali2015security}.
+If a malicious collocated user is capable of achieving specific insight into the configuration of the system, or the activity of other collocated users by executing performance evaluation programs (i.e., IOzone), a new attack vector may present itself.
 Malicious users could be capable of identifying common performance distributions of vulnerable system configurations and vulnerable active user jobs.
@@
     \node[below=of img, node distance=1cm, yshift=1cm] {Percentage of Training Data};
-    \node[left=of img, node distance=0cm, rotate=90, anchor=center,yshift=-0.7cm] {Percentage of Failed Predictions};
+    \node[left=of img, node distance=0cm, rotate=90, anchor=center,yshift=-0.7cm] {\% Null Hypothesis Rejections};
   \end{tikzpicture}
-  \caption{The performance of each algorithm on different IOzone tests with increasing amounts of training data measured using 10-fold cross validation and the KS test with p=0.001.
-The KS statistic for any given training percentage is achieved by averaging over the KS statistic for all tests and system configurations.
-The training percentages range from 5\% to 95\% by increments of 5\%.
-The ``readers'' tests tend to allow lower failure rates than the ``writers'' tests.
+  \caption{The percentage of null hypothesis rejections for predictions made by each algorithm on the KS test ($p=0.001$) over different IOzone test types with increasing amounts of training data.
+Each percentage of null hypothesis rejections is an average over ten random splits of the data.
+The training percentages range from 5\% to 95\% in increments of 5\%.
+The read test types tend to allow lower rejection rates than the write test types.
   \vspace{-.1cm}}
@@
 
-It remains unclear what mathematical properties of each approximation technique contribute the most to overall accuracy.
-Results presented in Table \ref{tab:optimized_p_value_failure_rate} are particularly interesting, demonstrating that Delaunay appears most vulnerable to overfitting, Max Box mesh is largely invariant to feature scaling, and Voronoi mesh generalizes feature weights well to unseen configurations.
+Results presented in Table \ref{tab:optimized_p_value_failure_rate} are particularly interesting, demonstrating that Delaunay appears most vulnerable to data dependent tuning, Max Box mesh is largely insensitive to such tuning, and Voronoi mesh benefits (for this data set) from the tuning.
 
 There are many avenues for extending this modeling methodology.
-One of the most severe limitations of the present framework is the restriction to modeling ordinal system parameters.
-Many applications have nonordinal settings and individually modeling all combinations of those settings is not a tractable solution.
-Automatic processing of categorical settings into an appropriate set of ordinal features constitutes future work.
-Presently the failure rate of distribution predictions can only be reduced with large volumes of performance data, however future work may be able to effectively reduce the amount of training data required by introducing appropriate expectations of stochastic noise in the measured performance.
-Finally, more case studies need to be explored in order to test the robustness of the present modeling techniques to changes in domain and performance metric.
+One extension is to add categorical variables to the models.
+Presently the rejection rate of distribution predictions can only be reduced with large volumes of performance data, however the judicious choice (via experimental design, e.g.) of new data points may be able to effectively reduce the amount of training data required.
+Finally, more case studies need to be done to test the robustness of the present modeling techniques to changes in domain and performance metric.
 
@@
 
-The presented methodology and application is capable of providing new insights, extending existing analyses, and improving the management of computational performance variability.
-Each of Delaunay, Max Box, and Voronoi meshes are viable techniques for constructing approximations of performance cumulative distribution functions.
-A case study on I/O throughput demonstrated that models are capable of effectively predicting most unseen system configurations for any of the available I/O tests.
+The methodology presented is capable of providing new insights, extending existing analyses, and improving the management of computational performance variability.
+Delaunay, Max Box mesh, and Voronoi mesh interpolation are viable techniques for constructing approximations of performance cumulative distribution functions.
+A case study on I/O throughput demonstrated that the models are capable of effectively predicting CDFs for most unseen system configurations for any of the available I/O test types.
 The present methodology represents a notable increase in the ability to statistically model arbitrary system performance metrics involving the interaction of many ordinal system parameters.

### 2018-02-13 Paper_[2018-01-20]_IEEESE/Results/initial_processing.py

@@ -10,10 +10,11 @@ DIST_PREDICTION_FILE = "dist_prediction.html"
 TEST_HIST_FILE = "tests_by_machine_(aggregated).html"
 CLEAN_DATA_CSV = "clean_data.csv"
 
+PATH_TO_CLEAN_DATA = os.path.join("/Users","thomaslux","Git","LargeFiles")
 INITIAL_DATA_READ = not os.path.exists(DATA_FILE_PKL)
 PLOT_AGGREGATE_HISTOGRAMS = False
-VALUES_SUMMARY = False
-SHOW_COUNT_SUMMARY = False
+VALUES_SUMMARY = True
+SHOW_COUNT_SUMMARY = True
 SAMPLE_FUNCTION_PREDICTION = False
 SAVE_CLEAN_DATA_CSV = False
 
@@ -22,11 +23,11 @@ SAVE_CLEAN_DATA_CSV = False
 #      Read in new data     
 if INITIAL_DATA_READ:
     print("Reading data 1...")
-    data1 = read_struct("2018-1.csv")
+    data1 = read_struct(os.path.join(PATH_TO_CLEAN_DATA,"2018-1.csv"))
     print("Reading data 2...")
-    data2 = read_struct("2018-2.csv")
+    data2 = read_struct(os.path.join(PATH_TO_CLEAN_DATA,"2018-2.csv"))
     print("Reading data 3...")
-    data3 = read_struct("2018-3.csv")
+    data3 = read_struct(os.path.join(PATH_TO_CLEAN_DATA,"2018-3.csv"))
     print("Saving all 3 to file...")    
     with open(DATA_FILE_PKL, "wb") as f:
         pickle.dump( (data1, data2, data3), f )
@@ -58,6 +59,17 @@ else:
     print("Loading '%s'..."%(COUNTS_FILE_PKL))
     with open(COUNTS_FILE_PKL, "rb") as f:
         counts = pickle.load(f)
+
+print("="*70)
+print(sorted(set(list(map(len, [v for v in counts.values()])))))
+exit()
+with open("output.txt", "w") as f:
+    for config in sorted(counts):
+        print("%30s,"%(str(config)[1:-1]),len(counts[config]), file=f)
+
+        # if len(counts[config]) == 2:
+        #     print()
+exit()
 
 # ====================================================================
 # Generate the cleaned data (for prediction and analysis)
@@ -145,10 +157,6 @@ if SHOW_COUNT_SUMMARY:
             if (num_printed >= 3): 
[…]

### 2018-02-13 Report_[2018-01-15]_System_Throughput_Sample/small_throughput.py

@@ -1,0 +1,74 @@
+import os, pickle
+import numpy as np
+from util.data import read_struct
+from util.plotly import Plot
+from util.stats import cdf_fit_func
+
+CLEAN_DATA_CSV = "clean_data.csv"
+CLEAN_DATA_CSV_PKL = "clean_data.csv.pkl"
+
+if not os.path.exists(CLEAN_DATA_CSV_PKL):
+    data = read_struct(CLEAN_DATA_CSV)
+    print("Saving '%s'..."%(CLEAN_DATA_CSV_PKL))
+    with open(CLEAN_DATA_CSV_PKL, "wb") as f:
+        pickle.dump(data,f)
+else:
+    print("Loading '%s'..."%(CLEAN_DATA_CSV_PKL))
+    with open(CLEAN_DATA_CSV_PKL, "rb") as f: 
+        data = pickle.load(f)
+
+header = list(data.dtype.names)
+h_config = header[:4]
+h_test = header[4]
+h_values = header[5:]
+tests = sorted(np.unique(data[h_test]))
+test_len = max(map(len, tests))
+# Extract a subset of data, only the "readers" test
+# data = data[data["Test"] == "readers"]
+
+# ============================================================
+
+count = 0
+failures = 0
+slow_configs = []
+throughput_threshold = 40000
+total_threshold = 80
+
+p = Plot("Low throughput configurations", "Sample number", "Throughput")
+for row in data:
+    total = sum(1 for h in h_values if row[h] < throughput_threshold)
+    if total > 0:
+        slow_configs.append((tuple(row[h] for h in h_config), total))
+        if total >= total_threshold:
+            raw_values = np.array([row[h] for h in h_values])
+            name = str(slow_configs[-1][0] + (row["Test"],))
+            # p.add("",np.arange(len(raw_values)), raw_values, 
+            #       frame=name, group="")
+            p.add_histogram("", raw_values, frame=name, group="")
+    if (row["File Size"] == 4 and row["Record Size"] == 4): 
+        count += 1
+        if total > 0:
+            failures += total
+slow_configs = sorted(slow_configs,key=lambda r: r[1])[-16:]
+for (c,t) in slow_configs:
+    print(c,t)
+
+p.plot(show_slider_labels=False, frame_label="System Configuration: ",
+       data_easing=False, show_legend=False, bounce=False,
+       redraw=True, show_play_pause=False)
+exit()
+
+print()
+print("Number of slow runs: ", failures)
[…]

### 2018-03-08 Paper_[2017-12-15]_HPC/paper.tex
Changed some folder names, made final modifications to papers for submission.

--- 2018-01-08 Paper_[2017-11]_HPC/paper.tex
@@
 \vspace{-10pt}
+
 In order to evaluate the viability of multivariate models for
@@
 four-dimensional dataset produced by executing the IOzone benchmark
-from \shortciteN{iozone} on a homogeneous cluster of computers. The
-system performance data was collected by executing IOzone 40 times for
-each of a select set of system configurations. A single IOzone
-execution reports the max I/O throughput seen for the selected
-test. The 40 executions for each system configuration are converted
-into the mean and variance, both values in $\mathbb{R}$ capable of
-being modeled individually by the multivariate approximation
-techniques presented in Section \ref{sec:multivariate}. The summary of
-data used in the experiments for this paper can be seen in Table
-\ref{tab:data_type}.  Distributions of raw throughput values being
-modeled can be seen in Figure \ref{fig:raw_throughput}.
+from \shortciteN{iozone} on a homogeneous cluster of computers. All
+experiments were performed on parallel shared-memory nodes common to
+HPC systems. Each system had a lone guest Linux operating system
+(Ubuntu 14.04 LTS//XEN 4.0) on a dedicated 2TB HDD on a 2 socket, 4
+core (2 hyperthreads per core) Intel Xeon E5-2623 v3 (Haswell)
+platform with 32 GB DDR4. The system performance data was collected by
+executing IOzone 40 times for each of a select set of system
+configurations. A single IOzone execution reports the max I/O
+throughput seen for the selected test in kilobytes per second. The 40
+executions for each system configuration are converted into the mean
+and variance, both values in $\mathbb{R}$ capable of being modeled
+individually by the multivariate approximation techniques presented in
+Section \ref{sec:multivariate}. The summary of data used in the
+experiments for this paper can be seen in Table \ref{tab:data_type}.
+Distributions of raw throughput values being modeled can be seen in
+Figure \ref{fig:raw_throughput}.
 
@@
 achieving relative errors strictly less than $10^3$ for throughput
-mean and $10^6$ for throughput variance. Ideally, models will
-yield relative errors significantly smaller than 1.
+mean and $10^6$ for throughput variance. Ideally, models will yield
+relative errors significantly smaller than 1. The time required to
+compute thousands of models involved in processing the IOzone data
+through MDA was approximately five hours on a CentOS workstation with
+an Intel i7-3770 CPU at 3.40GHz. In four dimensions for example, each
+of the models could be constructed and evaluated over hundreds of
+points in less than a few seconds.
 
@@
 specification, although qualitatively correct predictions can be
-achieved with mush less (10\%, say) effort.
+achieved with much less (10\%, say) effort.
 
@@
 The most severe limitation to the present work is the restriction to
-modeling strictly ordinal (not categorical) system
-parameters. Existing statistical approaches for including categorical
-variables are inadequate for nonlinear interactions in high
-dimensions. Future work could attempt to identify the viability of
-different techniques for making predictions including categorical
-system parameters.
+modeling strictly ordinal (not categorical) system parameters.
+Existing statistical approaches for including categorical variables
+are inadequate for nonlinear interactions in high dimensions. Future
+work could attempt to identify the viability of different techniques
+for making predictions including categorical system parameters.
 
@@
 
+%     Author Biography Block     
+%================================
+\section*{Author Biographies}
+
+\textbf{\uppercase{Thomas C. H. Lux}} is a Ph.D. student at Virginia Tech
+studying computer science under Dr. Layne Watson.
+
+\textbf{\uppercase{Layne T. Watson}} (Ph.D., Michigan, 1974) has interests
+in numerical analysis, mathematical programming, bioinformatics, and data
+science.
+He has been involved with the organization of HPCS since 2000.
+
+\textbf{\uppercase{Tyler H. Chang}} is a Ph.D. student at Virginia Tech
+studying computer science under Dr. Layne Watson.
+
+\textbf{\uppercase{Jon Bernard}} is a Ph.D. student at Virginia Tech
+studying computer science under Dr. Kirk Cameron.
+
+\textbf{\uppercase{Bo Li}} is a senior Ph.D. student at Virginia Tech
+studying computer science under Dr. Kirk Cameron.
+
+\textbf{\uppercase{Li Xu}} is a Ph.D. student at Virginia Tech studying
+statistics under Dr. Yili Hong.
+
+\textbf{\uppercase{Godmar Back}} (Ph.D., University of Utah, 2002) has
+broad interests in computer systems, with a focus on performance and
+reliability aspects of operating systems and virtual machines.
+
+\textbf{\uppercase{Ali R. Butt}} (Ph.D., Purdue, 2006) has interests in cloud computing, distributed computing, and operating system induced variability.
+
+\textbf{\uppercase{Kirk W. Cameron}} (Ph.D., Louisiana State, 2000) has
+interests in computer systems design, performance analysis, and power and
+energy efficiency.
+
+\textbf{\uppercase{Yili Hong}} (Ph.D., Iowa State, 2009) has interests in
+engineering statistics, statistical modeling, and data analysis.
+
 \end{document}

### 2018-03-08 Paper_[2018-02-16]_IEEESE/paper.tex
Last push didn't include the renamed folders. This has it.

--- 2018-02-13 Paper_[2018-01-20]_IEEESE/paper.tex
@@
 The sources of variability, however, are distributed throughout the system stack and often difficult to identify.
-The methodology presented in this work is applicable to modeling the expected variability of useful computer system performance metrics without any prior knowledge of system architecture.
-Some examples of interesting performance metrics that could be modeled with the techniques in this work include computational throughput, power consumption, processor idle time, number of context switches, and RAM usage, as well as any other ordinal performance metric.
+The methodology presented in this work is applicable to modeling the expected variability of useful computer system performance measures without any prior knowledge of system architecture.
+Some examples of interesting performance measures that could be modeled with the techniques in this work include computational throughput, power consumption, processor idle time, number of context switches, and RAM usage, as well as any other ordinal measure of performance.
 
@@
 
-Many previous works attempting to model system performance have used simulated environments to estimate the performance of a system \cite{grobelny2007fase,wang2009simulation,wang2013towards}.
-Some refer to statistical models as being oversimplified and not capable of capturing the true complexity of the underlying system.
-This claim is partially correct, noting that a large portion of predictive statistical models rely on simplifying the machine to one or two parameters \cite{snavely2002framework,bailey2005performance,barker2009using,ye2010analyzing}.
-These limited statistical models have provided satisfactory performance in very narrow application settings.
-Many of the aforementioned statistical modeling techniques claim to generalize, while simultaneously requiring additional code annotations, hardware abstractions, or additional application level understandings in order to generate models.
-The approach presented here requires no modifications of the application, no architectural abstractions, nor any structural descriptions of the input data being modeled.
-The techniques used are purely mathematical and only need performance data as input.
+Many previous works attempting to model system performance have used simulated environments \cite{grobelny2007fase,wang2009simulation,wang2013towards}.
+Grobelny et al.
+refer to statistical models as being oversimplified and not capable of capturing the true complexity of the underlying system.
+Historically, statistical models have been limited to modelling at most one or two system parameters and have therefore not been capable of modeling the complexity of the underlying system.
+\cite{snavely2002framework,bailey2005performance,barker2009using,ye2010analyzing}.
+These limited statistical models have provided satisfactory performance in narrow application settings.
+However these techniques require additional code annotations, hardware abstractions, or additional assumptions about the behavior of the application in order to generate models.
+In contrast, the approach presented here requires no modifications of the application, no architectural abstractions, nor any structural descriptions of the input data being modeled.
+The techniques used in this work only require performance data as input.
 
-Among the statistical models presented in prior works, \cite{bailey2005performance} specifically mentions that it is difficult for the simplified models to capture variability introduced by I/O.
-System variability in general has become a problem of increasing interest to the cloud and HPC systems communities, however most existing work has focused on operating system (OS) induced variability \cite{beckman2008benchmarking,de2007identifying}.
-The work that has focused on managing I/O variability does not use any sophisticated modeling techniques \cite{lofstead2010managing}.
+Most existing work on performance variability has focused on operating system (OS) induced variability \cite{beckman2008benchmarking,de2007identifying}.
+Yet, system I/O variability has been particularly difficult for statistical models to capture \cite{bailey2005performance}.
+The prior work attempting to model I/O variability has been similarly limited to one or two system parameters \cite{lofstead2010managing}.
 
-This paper presents an application of advanced statistical modeling to the domain of computer system I/O throughput.
-The models are used to predict the cumulative distribution function (CDF) of the expected I/O throughput for a system at previously unseen configurations.
-The techniques in this paper can tractably model tens of interacting system parameters with tens of thousands of unique configurations.
+This paper presents and evaluates an application of multivariate interpolation to the domain of computer system I/O throughput.
+The interpolatory models are used to predict the cumulative distribution functions of the expected I/O throughput for a system at previously unseen configurations.
+Beyond this I/O case study, the techniques in this paper can tractably model tens of interacting system parameters with tens of thousands of known configurations.
+The primary contribution of this work is a modeling framework using multivariate interpolation to capture precise characteristics (via cumulative distribution functions) of an arbitrary performance measure on any type of computer system.
 
@@
 Mean, range, standard deviation, variance, and interquartile range are a few summary statistics that describe performance and variability.
-However, the most precise characterization of any ordinal performance metric is the cumulative distribution function (CDF), or its derivative the probability density function (PDF).
+However, the most precise characterization of any performance measure is the cumulative distribution function (CDF), or its derivative the probability density function (PDF).
 Previous techniques for predicting system performance have strictly modeled real-valued summary statistics because there exists a large base of mathematical techniques capable of approximating functions of the form $f: \mathbb{R}^d \rightarrow \mathbb{R}$.
-There is little systems work in approximating functions of the form $f: \mathbb{R}^d \rightarrow \{g \mid g: \mathbb{R} \rightarrow \mathbb{R}\}$.
+However, there is little systems work approximating functions $f: \mathbb{R}^d \rightarrow \{g \mid g: \mathbb{R} \rightarrow \mathbb{R}\}$.
 
@@
 It is necessary that a PDF $f: \mathbb{R} \rightarrow \mathbb{R}$ have the properties $f(x) \geq 0$ and $\int_{-\infty}^{\infty}f(x)dx = 1$.
-However, for a CDF $F: \mathbb{R} \rightarrow \mathbb{R}$ the properties are $F(x) \in [0,1]$ and $F(x)$ is monotone increasing.
+However, for a CDF $F: \mathbb{R} \rightarrow \mathbb{R}$ the properties are $F(x) \in [0,1]$ and $F(x)$ is right continuous and nondecreasing.
 For the purpose of function approximation, the properties of the CDF are more straightforward to maintain.
@@
 A demonstration of how this is applied can be seen in Figure \ref{fig:prediction_example}.
-Next, the three techniques used to model CDFs are summarized, a mechanism for identifying important features is proposed, and the error measure used throughout this work is presented.
+Next, the three techniques used to model CDFs are summarized, a mechanism for identifying important features is proposed, and the error metric used throughout this work is presented.
 For each of the following, $n \in \mathcal{N}$ refers to the number of training samples (known system configurations), $d \in \mathcal{N}$ refers to the dimension of the data (ordinal system parameters), $x^{(i)} \in \mathbb{R}^d$ is a known system configuration with known performance CDF $F_{x^{(i)}}$, and $y \in \mathbb{R}^d$ refers to a previously unseen system configuration with performance CDF $F_y$.
@@
     \node[below=of img, node distance=1cm, yshift=1cm] {KS Statistic for Predicted vs. Actual};
-    \node[left=of img, node distance=0cm, rotate=90, anchor=center,yshift=-0.7cm] {Count};
+    \node[left=of img, node distance=0cm, rotate=90, anchor=center,yshift=-0.7cm] {Count of Predictions with Given KS Statistic};
   \end{tikzpicture}
-  \caption{Histograms of the prediction error for each modeling algorithm from ten random splits when trained with 80\% of the data.
+  \caption{Histograms of the prediction error for each modeling algorithm from ten random splits when trained with 80\% of the data aggregated over all different test types.
 The distributions show the KS statistics for the predicted throughput distribution versus the actual throughput distribution.
@@
 A rejection rate of 45\% would seem a poor result, however in this situation the complexity of the problem warrants a slightly different interpretation.
-These predictions are a very \textit{precise} characterization of performance variability, in fact the strongest possible characterization of variability.
+These predictions are a very \textit{precise} characterization of performance variability, in fact the cumulative distribution function of a random variable is the strongest possible characterization of variability that can be predicted.
 Globally, only a little under half of the predictions fail to capture \textit{all} of the characteristics of performance variability at new system configurations.
@@
 All weights were kept in the range [0,2], and were applied to the normalized features for frequency, file size, record size, and number of threads.
-All three approximation techniques had similar optimal weights of approximately $(.001, 2, 1.7, 1.5)$ for frequency, file size, record size, and number of threads, respectively.
+All three approximation techniques had similar optimal weights achieved by simulated annealing of approximately $(.001, 2, 1.7, 1.5)$ for frequency, file size, record size, and number of threads, respectively.
 Recall that each interpolation technique uses small distances to denote large influences on predictions, meaning that frequency was the most important feature when predicting variability for the ``readers'' test type, followed not-so-closely by number of threads, then record size.
@@
 It is important to emphasize that the roughly 95\% of predictions for which the null hypothesis was not rejected are predicting the \textit{precise} distribution of I/O throughput that will be witnessed at a previously unseen system configuration.
-To the authors' knowledge, there is no existing methodology that is generically applicable to any system performance metric, agnostic of system architecture, and capable of making such powerful predictions.
+To the authors' knowledge, there is no existing methodology that is generically applicable to any system performance measure, agnostic of system architecture, and capable of making such powerful predictions.
 
@@
   \caption{The null hypothesis rejection rates for various $p$-values with the KS-test.
-These results are strictly for the ``readers'' IOzone test type and show unweighted results as well as the results with weights tuned for minimum error (KS statistic) by 300 iterations of modified simulated annealing.
+These results are strictly for the ``readers'' IOzone test type and show unweighted results as well as the results with weights tuned for minimum error (KS statistic) by 300 iterations of simulated annealing.
 Notice that the weights identified for the Delaunay model cause data dependent tuning, reducing performance.
@@
 
-As mentioned in Section \ref{sec:introduction}, no prior work has attempted to model an arbitrary performance metric for a system to such a high degree of precision.
-All previous statistical modeling attempts capture a few ($<3$) ordinal performance metrics.
+As mentioned in Section \ref{sec:introduction}, no prior work has attempted to model an arbitrary performance measure for a system to such a high degree of precision.
+All previous statistical modeling attempts capture a few ($<3$) ordinal performance measures.
 Generating models that have such high degrees of accuracy allows system engineers to identify previously unused configurations that present desired characteristics.
@@
 Presently the rejection rate of distribution predictions can only be reduced with large volumes of performance data, however the judicious choice (via experimental design, e.g.) of new data points may be able to effectively reduce the amount of training data required.
-Finally, more case studies need to be done to test the robustness of the present modeling techniques to changes in domain and performance metric.
+Finally, more case studies need to be done to test the robustness of the present modeling techniques to changes in domain and performance measure.
 
@@
 A case study on I/O throughput demonstrated that the models are capable of effectively predicting CDFs for most unseen system configurations for any of the available I/O test types.
-The present methodology represents a notable increase in the ability to statistically model arbitrary system performance metrics involving the interaction of many ordinal system parameters.
+The present methodology represents a notable increase in the ability to statistically model arbitrary system performance measures involving the interaction of many ordinal system parameters.
 %% \begin{itemize}

### 2019-02-08 Documents_[2019]/approximation/Old/paper_old.tex
Ready to run approximation experiments.

\documentclass[sigconf]{acmart}

\author{Thomas C. H. Lux}
%\authornote{Dr.~Trovato insisted his name be first.}
%\orcid{1234-5678-9012}
\affiliation{%
  \department{Dept. of Computer Science}
  \institution{Virginia Polytechnic Institute and State University}
  \city{Blacksburg} 
  \state{Virginia} 
  \postcode{24060}
}
\email{tchlux@vt.edu}

\author{Layne T. Watson}
%\authornote{Dr.~Trovato insisted his name be first.}
%\orcid{1234-5678-9012}
\affiliation{%
  \department{Depts. of Computer Science, Mathematics, and Aerospace \& Ocean Engineering}
  \institution{Virginia Polytechnic Institute and State University}
}

\author{Tyler H. Chang}
\affiliation{%
  \department{Dept. of Computer Science}
  \institution{Virginia Polytechnic Institute and State University}
}

% The default list of authors is too long for headers.
\renewcommand{\shortauthors}{T. Lux et al.}
\def\mn#1{\left\Vert #1\right\Vert_\infty}

\def\Bigl{\mathopen\Big}
\def\Bigr{\mathclose\Big}

\title{Interpolation of Sparse High-Dimensional Data*}

%% \documentclass[letterpaper, 10 pt, conference]{ieeeconf}  % Comment this line out
%%                                                           % if you need a4paper
%% %\documentclass[a4paper, 10pt, conference]{ieeeconf}      % Use this line for a4 paper
%% \IEEEoverridecommandlockouts                              % This command is only needed 
%% %% % copyright info
%% %% \IEEEpubid{\makebox[\columnwidth]{978-1-5386-6133-8/18/\$31.00
%% %%         ~\copyright~2018 IEEE \hfill}\hspace{\columnsep}
%% %%         \makebox[\columnwidth]{ }}

%% % if you want to use the \thanks command
%% \overrideIEEEmargins
%% % See the \addtolength command later in the file to balance the column lengths on the last page of the document

%% \title{\LARGE \bf Interpolants Scalable by Dimension* }

%% \author{Thomas C. H. Lux$^{1}$, Layne T. Watson$^{2}$, Tyler H. Chang$^{1}$% <-this % stops a space
%% \thanks{*This work was supported by the National Science Foundation Grant CNS-1565314.}%
%% \thanks{$^{1}$Doctoral student, Department of Computer Science, Virginia Polytechnic Institute and State University, Blacksburg, Virginia 24060, USA {\tt\small (tchlux at vt.edu)}}%
%% \thanks{$^{2}$Department of Computer Science, Virginia Polytechnic Institute and State University, Blacksburg, Virginia 24060, USA}%
%% }

% The following packages can be found on http:\\www.ctan.org
\usepackage{tikz}     % For adding axes to plots
\usetikzlibrary{positioning}
\usepackage{graphicx} % For including PDFs
\usepackage{amsmath} % assumes amsmath package installed
\usepackage{amssymb}  % assumes amsmath package installed
\usepackage{booktabs} % For formal tables
\usepackage{hyphenat} % custom hyphenation rules
\hyphenation{op-tical net-works semi-conduc-tor}

%% \usepackage{color}   % Needed to color text for drafts
%% \usepackage{algorithm}     % needed for the algorithm block
%% \usepackage{algpseudocode} % algorithmicx psuedo-code styling
%% \usepackage[pdftex,colorlinks=true,urlcolor=blue,citecolor=black,anchorcolor=black,linkcolor=black]{hyperref}
%% \usepackage[dvips,colorlinks=true,urlcolor=blue,citecolor=black,anchorcolor=black,linkcolor=black]{hyperref}

%% Command for creating multiplie-line cells inside a table environment
\newcommand{\multilinecell}[2][c]{%
  \begin{tabular}[#1]{@{}c@{}}#2\end{tabular}}

\begin{document}

\maketitle
\thispagestyle{empty}
\pagestyle{empty}
%%====================================================================
%%                    BODY OF PAPER
%%====================================================================

\section*{Abstract}

A rapid increase in the quantity of data available is allowing all fields of science to generate more accurate models of multivariate phenomena. Regression and interpolation become challenging when the dimension of data is large, especially while maintaining tractable computational complexity. While regression is a popular approach to solving approximation problems with high dimension, there are often advantages to interpolation. This paper presents error bounds for interpolants on moderately high dimensional problems and contrasts their performance with that of popular regression techniques. Empirical results demonstrate the viability of interpolants for high dimensional approximation problems, suggesting that these techniques are capable of effectively modeling multivariate phenomena while maintaining flexibility in different application domains.

\textbf{Keywords:} Approximation, regression, interpolation, high dimension, error bound

%     Introduction     
%======================

\section{Introduction}
\label{sec:introduction}

Regression and interpolation are problems of considerable importance that find applications across many fields of science. Pollution and air quality analysis \cite{de2008field}, energy consumption management \cite{lazos2014optimisation}, and student performance prediction \cite{cortez2008using,lux2016applications} are a few of many interdisciplinary applications of multivariate regression for predictive analysis. As discussed later, these techniques can also be applied to prediction problems related to high performance computing (HPC) file input/output (I/O) \cite{lux2018nonparametric}, Parkinson's patient clinical evaluations \cite{tsanas2010accurate}, and forest fire risk assessment \cite{cortez2007data}.

Regression and interpolation have a considerable theoretical base in one dimension \cite{cheney2009course}. Splines in particular are well understood as an interpolation technique in one dimension \cite{de1978practical}, particularly B-splines. Tensor products of B-splines \cite{unther1996interpolating} or other basis functions have an unfortunate exponential scaling with increasing dimension. Exponential scaling prohibits tensor products from being reasonably applied beyond three-dimensional data. In order to address this dimensional scaling challenge, C. de Boor and others proposed box splines \cite{de2013box}, of which one of the approximation techniques in this work is composed \cite{lux2018novel}.

The theoretical foundation of low dimensional interpolation allows the construction of strong error bounds that are absent from high dimensional problems. This work extends some known results regarding the secant method \cite{dennis1996numerical} to construct an interpolation error bound for problems of arbitrary dimension. These error bounds are useful, considering the same cannot be said for regression algorithms in general. The maximum complexity of an interpolant is bounded by the amount of data available, while the maximum complexity of a regressor is bounded by both the amount of data and the chosen parametric form. For this reason, generic uniform bounds are largely unobtainable for regression techniques on arbitrary approximation problems, even when the approximation domain is bounded.

Aside from theoretical motivation for the use of interpolants, there are often computational advantages as well. Interpolants do not have the need for \textit{fitting} data, or minimizing error with respect to model parameters. In applications where the amount of data is large and the relative number of predictions that need to be made for a given collection of data is small, the direct application of an interpolant is much less computationally expensive.

In this work, multivariate interpolation is defined given a closed convex subset $Y$ of a metrizable topological vector space with metric $s$, some function $f:\mathbb{R}^d \rightarrow Y$ and a set of points $X = \bigl\{x^{(1)}$,$\ldots$,$x^{(n)}\bigr\} \subset \mathbb{R}^d,$ along with associated function values $f\bigl(x^{(i)}\bigr)$. The goal is to construct an approximation $\hat f: \mathbb{R}^d \rightarrow Y$ such that $\hat f\bigl(x^{(i)}\bigr) = f\bigl(x^{(i)}\bigr)$ for all $i = 1$,$\ldots$,$n$. It is often the case that the form of the true underlying function $f$ is unknown, however it is still desirable to construct an approximation $\hat f$ with small approximation error at $y \notin X$. The two metric spaces that will be discussed in this work are the real numbers with metric $s(x,y) = |x-y|$, and the set of cumulative distribution functions (CDFs) with the Kolmogorov-Smirnov (KS) statistic as metric.

Multivariate regression is often used when the underlying function is presumed to be stochastic, or stochastic error is introduced in the evaluation of $f$. Hence, multivariate regression relaxes the conditions of interpolation by choosing parameters $P$ defining $\hat f(x;P)$ to minimize the error vector $\Bigl( \bigl | \hat f \bigl(x^{(1)};P\bigr) - f\bigl(x^{(1)}\bigr) \bigr|$, $\ldots$, $\bigl | \hat f \bigl(x^{(n)}; P\bigr) - f\bigl(x^{(n)}\bigr) \bigr | \Bigr)$ in some norm. The difficult question in the case of regression is often what parametric form to adopt for any given application.

The most challenging problem when scaling in dimension is that the number of possible interactions between dimensions grows exponentially. Quantifying all possible interactions becomes intractable and hence beyond three-dimensional data, mostly linear models are used. That is not to say nonlinear models are absent, but nonlinearities are often either preconceived or model pairwise interactions between dimensions at most. Even globally nonlinear approximations such as neural networks are constructed from compositions of summed low-interaction functions \cite{clevert2015fast}.

Provided the theoretical and practical motivations for exploring interpolants, this work aims to study the empirical performance differences between a set of scalable (moderately) interpolation techniques and a set of common regression techniques. These techniques are applied to a collection of moderately high dimensional problems ($5 \le d \le 50$) and the empirical results are discussed.

The remainder of this paper is organized as follows. Section \ref{sec:multivariate} presents the multivariate models. Section \textbf{???} presents the error measuring methodology that is used for collecting empirical results. Section \textbf{???} presents the theoretical error bounds for interpolation. Section \textbf{???} presents the data sets for empirical approximation analysis and presents the approximation results for all models on these data sets. Section \textbf{???} analyzes and discusses the results, their implications, and their limitations. Finally, Section \textbf{???} concludes.

% Art is a subjective form of personal expression and computers are
% just robots that are going to take over the world. \textit{*deep
% breath*} In order to communicate my frustration with the
% technological and information revolution, I'm going to write the
% rest of this paper in poetry.

% Star light, star bright \ldots

\section{Multivariate Regression}
Multivariate regressors are capable of accurately modeling a complex dependence of a response (in $Y$) on multiple variables (represented as a points in $\mathbb{R}^{d}$). The approximations to some (unknown) underlying function $f: \mathbb{R}^d \rightarrow Y$ are chosen to minimize some error measure related to data samples $f\bigl(x^{(i)}\bigr)$. For example, least squares regression uses the sum of squared differences between modeled function values and true function values as an error measure. In this section and the next, some techniques are limited to approximating real valued functions ($Y \subset \Re$). These techniques are extended to real vector-valued functions by repeating the construction for each component of the vector outputs. Throughout the following, $x$ denotes a $d$-tuple, $x_i$ the $i$th component of $x$, and $x^{(i)}$ the $i$th $d$-tuple data point. Different symbols are used to represent the approximation function $\hat f$.

\subsection{Multivariate Adaptive Regression Splines}
This approximation was introduced in \cite{friedman1991multivariate} and subsequently improved to its current version in \cite{stanford1993fast}, called fast multivariate adaptive regression splines (Fast MARS).In Fast MARS, a least squares fit model is iteratively built by beginning with a single constant valued function and adding two new basis functions at each iteration of the form
\begin{align*}
  B_{2j-1}(x) &= B_l(x) \bigl(x_i-x^{(p)}_i\bigr)_+, \\
  B_{2j}(x) &= B_k(x) \bigl(x_i-x^{(p)}_i\bigr)_- ,
\end{align*}
where $j$ is the iteration number, $1 \le p \le n$, and $B_l(x)$, $B_k(x)$ are basis functions from the previous iteration,
$$w_+ = \begin{cases} w, & w \geq 0 \\ 0, & w < 0 \end{cases},$$
and $w_- = (-w)_+$. After iteratively constructing a model, MARS then iteratively removes basis functions that do not contribute to goodness of fit. In effect, MARS creates a locally component-wise tensor product approximation of the data. The overall computational complexity of Fast MARS is $\mathcal{O}(n d m^3)$ where $m$ is the maximum number of underlying basis functions. This paper uses an implementation of Fast MARS \cite{rudy2017pyearth} with $m = 200$.

\subsection{Multilayer Perceptron Regressor}
The neural network is a well studied and widely used method for both regression and classification tasks \cite{hornik1989multilayer}. When using the rectified linear unit (ReLU) activation function \cite{dahl2013improving} and training with the BFGS minimization technique \cite{moller1993scaled}, the model built by a multilayer perceptron uses layers $l : \mathbb{R}^{i} \rightarrow \mathbb{R}^{j}$ defined by
$$ l(u) = \big( u^t W_l \big)_+ ,$$
where $W_l$ is the $i$ by $j$ weight matrix for layer $l$. In this form, the multilayer perceptron (MLP) produces a piecewise linear model of the input data. The computational complexity of training a multilayer perceptron is $\mathcal{O}(n d m)$, where $m$ is determined by the sizes of the layers of the network and the stopping criterion of the BFGS minimization used for finding weights. This paper uses the scikit-learn MLP regressor \cite{scikit-learn}, a single hidden layer with 100 nodes, ReLU activation, and BFGS for training.

\subsection{Support Vector Regressor}
Support vector machines are a common method used in machine learning classification tasks that can be adapted for the purpose of regression \cite{basak2007support}. How to build a support vector regressor (SVR) is beyond the scope of this summary, but the resulting functional fit $p : \mathbb{R}^d \rightarrow \mathbb{R}$ has the form
$$ p(x)  = \sum_{i=1}^{n}a_i K(x,x^{(i)}) + b ,$$
where $K$ is the selected kernel function, $a \in \mathbb{R}^n$, $b \in \mathbb{R}$ are coefficients to be solved for simultaneously. The computational complexity of the SVR is $\mathcal{O}(n^2dm)$, with $m$ being determined by the minimization convergence criterion. This paper uses the scikit-learn SVR \cite{scikit-learn} with a polynomial kernel function.

\section{Multivariate Interpolation}
The following interpolation techniques demonstrate a reasonable variety of approaches to interpolation. All of the presented interpolants produce approximations that are continuous in value, which is often a desirable property in applied approximation problems.

\subsection{Linear Shepard}
The linear Shepard method (LSHEP) is a blending function using local linear interpolants, a special case of the general Shepard algorithm \cite{thacker2010algorithm}. The interpolant has the form
$$ p(x) = \frac{\sum\limits_{k=1}^{n}W_k(x)P_k(x)}{\sum\limits_{k=1}^{n}W_k(x)} ,$$
where $W_k(x)$ is a locally supported weighting function and $P_k(x)$ is a local linear approximation to the data satisfying $P_k\big(x^{(k)}\big) = f\big(x^{(x)}\big)$. The computational complexity of LSHEP is $\mathcal{O}(n^2d^3)$. This paper uses the Fortran 95 implementation of LSHEP in SHEPPACK \cite{thacker2010algorithm}.

%% ===================================================================
\subsection{Delaunay}
\label{sec:delaunay}

The Delaunay triangulation is a well-studied geometric technique for producing an interpolant \cite{lee1980two}. The Delaunay triangulation of a set of data points into simplices is such that there are no data points inside the sphere defined by the vertices of each simplex. For a $d$-simplex S with vertices $x^{(0)}$, $x^{(1)}$, $\ldots$, $x^{(d)}$, and function values $f(x^{(i)})$, $i=0$, $\ldots$, $d$, $y \in S$ is a unique convex combination of the vertices,
$$ y = \sum_{i=0}^{d} w_i x^{(i)}, \quad \sum_{i=0}^{d} w_i = 1, \quad w_i \geq 0, \quad i=0,\ldots,d, $$
and the Delaunay interpolant $\hat f(y)$ at $y$ is given by
$$ \hat f(y) = \sum_{i=0}^{d} w_i f(x^{(i)}). $$
The computational complexity of Delaunay interpolation (for the implementation used) is $\mathcal{O}(n d^4 \log d)$, which is capable of scaling reasonably to $d \leq 50$. In the present application, a Delaunay simplex $S$ containing $y$ is found, then the $d+1$ vertices (points in $X$) of $S$ are used to assign weights to each vertex and produce the predicted function value for point $y$.

%% ===================================================================
\subsection{Iterative Box Mesh}
\label{sec:max_box_mesh}

The Iterative Box mesh is an interpolation technique that utilizes overlapping box splines \cite{de2013box} as basis functions that are shifted and scaled to have support over box shaped regions. The boxes are constructed in a way to guarantee a covering of the domain. Given a set of box splines $\bigl\{b^{x^{(i)}}\bigr\}$ with the iterative box properties anchored at interior points $\bigl\{x^{(i)}\bigr\}$,
$$ \hat f(y) = \frac{\sum\limits_{i=1}^n b^{x^{(i)}}(y) f(x^{(i)})}{\sum\limits_{i=1}^n b^{x^{(i)}}(y)}. $$
Note that the box splines always satisfy $b^{x^{(i)}}\bigl(x^{(j)}\bigr) = \delta_{ij}$ and $b^{x^{(i)}}(y) \geq 0$. The computational complexity of interpolation via the Iterative Box mesh is $\mathcal{O}(n^2 d)$.

%% ===================================================================
\subsection{Voronoi Mesh}
\label{sec:voronoi_mesh}

A well-studied technique for classification and approximation is the nearest neighbor algorithm \cite{cover1967nearest}. Nearest neighbor inherently utilizes the convex region $C^{x^{(i)}}$ (Voronoi cell \cite{dirichlet1850reduction}) consisting of all points closer to $x^{(i)}$ than to any other point $x^{(j)}$. The Voronoi mesh smooths the nearest neighbor approximation by utilizing the Voronoi cells to define support via a generic basis function $v: \mathbb{R}^d \rightarrow \mathbb{R}_+$ given by
$$ v^{x^{(i)}}(y) = \left(1 - \frac{||y - x^{(i)}||_2}{2 \ h(y - x^{(i)} \mid x^{(i)})} \right)_+, $$
where $h(w \mid x^{(i)})$ is the distance between $x^{(i)}$ and the boundary of the Voronoi cell $C^{x^{(i)}}$ in the direction $w$. $v^{x^{(i)}}\bigl(x^{(j)}\bigr) = \delta_{ij}$ and $v^{x^{(i)}}$ has local support, giving the interpolated value
$$ f(y) = \frac{\sum_{i=1}^n v^{x^{(i)}}(y) f(x^{(i)})}{\sum_{i=1}^n v^{x^{(i)}}(y)}, $$
where $0 \leq v^{x^{(i)}}(y) \leq 1$. The computational complexity of interpolation via the Voronoi mesh is $\mathcal{O}(n^2 d)$. All of the approximations are an interpolant involving a convex combination of known function values $f(x^{(i)})$.

\begin{figure}[htb]
  \begin{tikzpicture}
    \node (img)  {\includegraphics[width=0.4\textwidth,]{plot-delaunay-example-prediction.pdf}};
    \node[below=of img, node distance=1cm, yshift=1cm] {I/O Throughput};
    \node[left=of img, node distance=0cm, rotate=90, anchor=center,yshift=-0.7cm] {CDF Value};
  \end{tikzpicture}
  \vspace{-0.3cm}
  \caption{In this HPC I/O example, the general methodology for predicting a CDF and evaluating error can be seen. The Delaunay method chose three source distributions (dotted lines) and assigned weights \{.3, .4, .3\} (top to bottom at arrow). The weighted sum of the three known CDFs produces the predicted CDF (dashed line). The KS Statistic (arrow) computed between the true CDF (solid line) and predicted CDF (dashed line) is 0.2 for this example. The KS test null hypothesis is rejected at $p$-value 0.01, however it is not rejected at $p$-value 0.001.
  \vspace{-.1cm}}
  \label{fig:prediction_example}
\end{figure}

%% ===================================================================
\subsection{Measuring Error}

When the range of an approximation is the real numbers, error is reported with summary statistics including: min absolute error, max absolute error, absolute error quartiles, and mean absolute error. In general, all errors are recorded as signed relative error and any figures depicting distributions of errors will present these values.

A hurdle when modeling function-valued outputs such as cumulative distribution functions (CDFs) or probability density functions (PDFs) is that certain properties must be maintained. It is necessary that a PDF $f: \mathbb{R} \rightarrow \mathbb{R}$ have the properties $f(x) \geq 0$ and $\int_{-\infty}^{\infty}f(x)dx = 1$. However, for a CDF $F: \mathbb{R} \rightarrow \mathbb{R}$ the properties are $F(x) \in [0,1]$ and $F(x)$ is absolutely continuous and nondecreasing. This work utilizes the fact that a convex combination of CDFs (or PDFs) results in a valid CDF. Given $G(x) = \sum_{i}w_i F_i(x)$, $\sum_{i} w_i = 1$, $w_i \geq 0$, and each $F_i$ is a valid CDF, $G$ must also be a valid CDF. A demonstration of how this is applied can be seen in Figure \ref{fig:prediction_example}.

%% %% ===================================================================
%% \subsection{Feature Weighting}
%% \label{sec:feature_weighting}

%% It is well-known that an important procedure in any application of predictive methodologies is identifying those features of the data that are most relevant to making accurate predictions \cite{guyon2003introduction}. Selection strategies such as the floating searches studied in \cite{pudil1994floating} or others compared in \cite{ferri1994comparative} can be too expensive for large approximation problems. Rather, this work poses feature selection as a continuous optimization problem. Let $X$ be an $n \times d$ matrix of $n$ known system configurations with $d$ parameters each normalized to be in $[0,1]$. Define an error function that computes the error of a predictive model trained on $X\hbox{ diag }w$, $w \in \mathbb{R}^d$,  by performing ten random splits with 80\% of the rows of $X\hbox{ diag }w$ for training and 20\% for testing. A minimum of this error function could be considered an optimal weighting of the features of $X$. Minimization is performed using a zero order method in the absence of a readily computable gradient.

%% ===================================================================

The performance of approximation techniques that predict probability functions can be analyzed through a variety of summary statistics. Mean absolute error, mean squared error, and the max absolute error are all popular measures. This work uses the max absolute error, also known as the Kolmogorov-Smirnov (KS) statistic \cite{lilliefors1967kolmogorov} for its compatibility with the KS test.

The two-sample KS test is a useful nonparametric test for comparing two CDFs while only assuming stationarity, finite mean, and finite variance. The null hypothesis (that two CDFs come from the same underlying distribution) is rejected at level $p \in [0,1]$ when

$$ KS > \sqrt{-\frac{1}{2}\ln\biggl(\frac{p}{2}\biggr)} \sqrt{\frac{1}{n_1} + \frac{1}{n_2}}, $$

with distribution sample sizes $n_1,n_2 \in \mathcal{N}$. For all applications of the KS test presented in this work $n_1 = n_2$. An example of the round-trip prediction methodology from known and predicted distributions to the calculation of error can be seen in Figure \ref{fig:prediction_example}.

\section{Theoretical Error Bound}

This section presents the theoretical results bounding the error of Delaunay interpolation. The error analysis relies on the Delaunay interpolation for two reasons: (1) second order results can be obtained utilizing a Lipschitz constant on the gradient of a function, rather than standard Lipschitz bounds and (2) multiple other interpolants analyzed compute predictions as convex combinations of observed function values, which may allow for straight forward extensions of this error bound.

$ $

\noindent \textbf{\textit{Lemma 1.}} Let $S \subset \mathbb{R}^d$ be open and convex, $\ f: \mathbb{R}^d \rightarrow \mathbb{R},$ and $\nabla f \in Lip_{(\gamma,\|\cdot\|_2)}(S)$, the set of $\gamma$-Lipschitz continuous functions in the $2$-norm. Then for all $x,y \in S$

$$\big|\ f(y) - f(x) - \langle \nabla f(x), y - x \rangle \big| \leq \frac{\gamma \|y - x\|_2^2}{2}.$$

\noindent \textit{Proof.} Consider the function $g(t) = f \big((1-t) x + t y \big),$ $0 \leq t \leq 1,$ whose derivative $g'(t) = \big\langle \nabla f \big((1-t) x + t y \big), y - x \big\rangle$ is the directional derivative of $f$ in the direction $(y - x).$

\begin{align*}
  \big|\ f(y) - f&(x) - \langle \nabla f(x), y - x \rangle \big| &\\ 
     &= \big|\ g(1) - g(0) - g'(0) \big| & \\
     &= \bigg| \int_0^1 g'(t) - g'(0)\ dt \bigg| \\
     &\leq \int_0^1 \big|g'(t) - g'(0)\big|\ dt \\
     &= \int_0^1 \bigg| \big \langle \nabla f\big((1-t)x + ty\big) - \nabla f(x), y - x \big \rangle \bigg|\ dt \\
     &\leq \int_0^1 \big \| \nabla f\big((1-t)x + ty\big) - \nabla f(x) \big \|_2\ \| y - x \|_2\ dt \\
     &\leq \int_0^1 \big ( \gamma\ \|y-x\|_2 \big) \ \big( \|y-x\|_2 \big) t\ dt \\
     &= \frac{\gamma \|y - x\|_2^2}{2}.
\end{align*}

$\qed$

\noindent \textbf{\textit{Lemma 2.}} Let $x, y, v_i \in \mathbb{R}^d,$ $c_i \in \mathbb{R},$ and $|\langle y - x, v_i \rangle| \leq c_i$ for $i = 1,$ $\ldots,$ $d.$ If $M = (v_1,$ $\ldots,$ $v_d)$ is nonsingular, then

$$\|y - x\|_2^2 \leq \frac{1}{\sigma_d^2} \sum_{i=1}^d c_i^2,$$

where $\sigma_d$ is the smallest singular value of $M.$

\noindent \textit{Proof.} Using the facts that $M$ and $M^t$ have the same singular values, and $\|M^tw\|_2 \geq \sigma_d \|w\|_2,$ gives

\begin{align*}
     \|y - x\|_2^2 &\leq \frac{\|M^t (y - x)\|_2^2}{\sigma_d^2} \\
                   &=    \frac{1}{\sigma_d^2} \sum_{i=1}^d \langle y - x, v_i \rangle^2 \\
                   &\leq \frac{1}{\sigma_d^2} \sum_{i=1}^d c_i^2.
\end{align*}

$\qed$

\noindent \textbf{\textit{Lemma 3.}} Given $\ f,$ $\gamma,$ $S$ as in {\it Lemma 1}, let $X = \{x_0,$ $x_1,$ $\ldots,$ $x_d\}$ $\subset S$ be the vertices of a $d$-simplex, and let $\hat f(x) = \langle c, x - x_0 \rangle + f(x_0),$ $c \in \mathbb{R}^d$ be the linear function interpolating $\ f$ on $X.$ Let $\sigma_d$ be the smallest singular value of the matrix $M = (x_1 - x_0,$ $\ldots,$ $x_d - x_0),$ and $k = \max\limits_{1\ \leq\ j\ \leq\ d} \|x_j - x_0\|_2.$ Then

$$\big\|\nabla f(x_0) - c\big\|_2 \leq \sqrt{d} \frac{\gamma k^2}{\sigma_d}.$$

\noindent \textit{Proof.} Consider $f(x) - \hat f(x)$ along the line segment $z(t) = (1-t)x_0 + t x_j,$ $0 \leq t \leq 1.$ By Rolle's Theorem, for some $0 < \hat t < 1,$ $\big\langle \nabla f\big(z(\ \hat t\ )\big) - c, x_j - x_0 \big\rangle = 0.$ Now

\begin{align*}
       \big| \big< \nabla f(x_0) &- c, x_j - x_0 \big \rangle \big| \\
        &= \big| \big< \nabla f(x_0) - \nabla f\big(z(\ \hat t\ )\big) + \nabla f\big(z(\ \hat t\ )\big) - c, x_j - x_0 \big \rangle \big| \\
	&= \big| \big\langle \nabla f(x_0) - \nabla f \big(z(\ \hat t\ )\big), x_j - x_0 \big \rangle \big| \\
	&\leq \big \| \nabla f(x_0) - \nabla f\big(z(\ \hat t\ )\big) \big \|_2 \|x_j - x_0\|_2 \\
	&\leq \gamma \|x_0 - z(\ \hat t\ )\|_2\ \|x_j - x_0\|_2 \\
	&\leq \gamma \|x_j - x_0\|_2^2 \leq \gamma k^2, \\
\end{align*}

for all $1 \leq j \leq d.$ Using {\it Lemma 2},

$$\big \| \nabla f(x_i) - c \big \|_2^2 \leq \frac{d}{\sigma_d^2} \big( \gamma k^2\big)^2 \implies \big \| \nabla f(x_i) - c \big \|_2 \leq \sqrt{d} \frac{\gamma k^2}{\sigma_d}.$$

$\qed$

\noindent \textbf{\textit{Theorem.}} Under the assumptions of {\it Lemma 1} and {\it Lemma 3}, for $z \in S,$

$$ \big|\ f(z) - \hat f(z)\big| \leq \frac{\gamma \|z - x_0\|_2^2}{2} + \sqrt{d} \frac{\gamma k^2}{\sigma_d} \|z - x_0\|_2.$$

\noindent \textit{Proof.} Let $v = \nabla f(x_0) - c,$ where $\|v\|_2 \leq \sqrt{d} \gamma k^2 / \sigma_d$ by {\it Lemma 3}. Now

\begin{align*}
       \big|\ f(z) - \hat f(z)\ \big| &= \big|\ f(z) - f(x_0) - \langle c, z - x_0 \rangle \big| \\ 
       &= \big|\ f(z) - f(x_0) - \langle \nabla f(x_0) - v, z - x_0 \rangle \big| \\
       &= \big|\ f(z) - f(x_0) - \langle \nabla f(x_0) , z - x_0 \rangle + \langle v , z - x_0 \rangle \big| \\
       &\leq \big|\ f(z) - f(x_0) - \langle \nabla f(x_0) , z - x_0 \rangle \big| + \big| \langle v , z - x_0 \rangle \big| \\
       &\leq \big|\ f(z) - f(x_0) - \langle \nabla f(x_0) , z - x_0 \rangle \big| + \|v\|_2 \ \|z - x_0\|_2 \\
       &\leq \big|\ f(z) - f(x_0) - \langle \nabla f(x_0), z - x_0 \rangle \big| + \textstyle{\frac{\gamma k^2 \sqrt{d}}{\sigma_d}} \|z - x_0\|_2 \\
       &\leq \frac{\gamma \|z - x_0\|_2^2}{2} + \sqrt{d}\frac{\gamma k^2}{\sigma_d} \|z - x_0\|_2,
\end{align*}

where the last inequality follows from {\it Lemma 1}.

$\qed$

\section{Data and Analysis}
In order to evaluate the proposed interpolation and approximation techniques, this paper utilizes three data sets of varying dimension and application. In the following subsections the sources and targets of each data set are described, as well as challenges and limitations related to interpolating and approximating these data sets. The distributions of response values being modeled can be seen in Figure \ref{fig_response_hists}. The preprocessing and approximation processes are described in Section \ref{sec_performance_analysis}.

\begin{figure}
  \includegraphics[width=.5\textwidth,height=2cm]{p-hist.pdf}
  \includegraphics[width=.5\textwidth,height=2cm]{f-hist.pdf}
  \includegraphics[width=.5\textwidth,height=2cm]{h-hist.pdf}
  \caption{Histograms of Parkinsons (total UPDRS), forest fire (area), and HPC I/O (mean throughput) response values respectively. Notice that both the forest fire and HPC I/O data sets are heavily skewed.
  \vspace{-.5cm}}
  \label{fig_response_hists}
\end{figure}

\subsection{Data Summary}

\subsubsection{High Performance Computing I/O ($n = 532, d = 4$)}
The first of three data sets is a four-dimensional data set produced by executing the IOzone benchmark from \cite{iozone} on a homogeneous cluster of computers. The system performance data was collected by executing IOzone 40 times for each of a select set of system configurations. A single IOzone execution reports the max I/O file-read throughput seen. The 40 executions for each system configuration are converted to their mean, which is capable of being modeled by each of the multivariate approximation techniques presented in Section \ref{sec_mesh_construction}. The four dimensions being modeled to predict throughput mean are file size, record size, thread count, and CPU frequency.

\begin{figure*}[htb]
  \begin{tikzpicture}
    \node (img)  {\includegraphics[width=0.95\textwidth,height=4cm]{eval-times.pdf}};
    \node[below=of img, node distance=0cm, yshift=1cm] {Relative Error Tolerance while Bootstrapping};
    \node[left=of img, node distance=0cm, rotate=90, anchor=center,yshift=-0.7cm] {Fit Time};
  \end{tikzpicture}
  \caption{Time required to generate model fits for each technique with varying relative error tolerance during bootstrapping.
    \vspace{-.3cm}}
  \label{fig_eval_times}
\end{figure*}

\subsubsection{Forest Fire ($n = 517, d = 12$)}
The forest fire data set \cite{cortez2007data} describes the area of Montesinho park burned on specific days and months of the year in terms of the environmental conditions. The twelve dimensions being used to model burn area are the $x$ and $y$ spatial coordinates of burns in the park, month and day of year, the FFMC, DMC, DC, and ISI indices (see source for details), the temperature in Celsius, relative humidity, wind speed, and outdoor rain. The original analysis of this data set demonstrated it to be difficult to model, likely due to the skew in response values.

\subsubsection{Parkinson's Telemonitoring ($n = 468, d = 16$)}
The final data set for evaluation \cite{tsanas2010accurate} is derived from a speech monitoring study with the intent to automatically estimate Parkinson's disease symptom development in Parkinson's patients. The function to be predicted is a time-consuming clinical evaluation measure referred to as the UPDRS score. The total UPDRS score given by a clinical evaluation is estimated through 16 real numbers generated from biomedical voice measures of in-home sound recordings.

\subsection{Performance Analysis}
\label{sec_performance_analysis}

The performance of the approximation techniques varies considerably across the three evaluation data sets. Relative errors for the most na\"{\i}ve approximators such as nearest neighbor can range from zero to $\displaystyle \big(\max_x f(x) - \min_x f(x)\big) / \min_x f(x)$ when modeling a positive function $f(x)$ from data. Each of the approximation techniques presented remain within these bounds and all errors are presented in signed relative form $(\hat f(x) - f(x)) / f(x)$. Before the models are constructed all data values (components $x^{(i)}_r$ of $x^{(i)} \in X$) are shifted and scaled to be in the unit cube $[0,1]^d$, while the response values are taken in their original form. All models are evaluated with $10$ random $80/20$ splits of the data.

\begin{table}
  \centering
  \begin{tabular}{c|c|c|c}
    \hline
    \textbf{Data Set} & \textbf{Technique} & \textbf{Tolerance} & \textbf{Average Error}\\
    \hline
    HPC I/O & MBM & 1.2 & 0.597\\
    Forest Fire & MBM & 1.8 & 3.517\\
    Parkinson's & MBM & 0.6 & 0.114\\
    \hline
    HPC I/O & IBM & 0.4 & 0.419\\
    Forest Fire & IBM & 1.8 & 3.615\\
    Parkinson's & IBM & 1.8 & 0.121\\
    \hline
    HPC I/O & VM & 0.2 & 0.382\\
    Forest Fire & VM & 1.0 & 4.783\\
    Parkinson's & VM & 2.0 & 1.824\\
    \hline
  \end{tabular}
  \caption{The optimal error tolerance bootstrapping parameters for each technique and each data set as well as the average absolute relative errors achieved by that tolerance. Notice that large relative error tolerances occasionally yield even lower evaluation errors, demonstrating the benefits of approximation over interpolation for noisy data sets.
  \vspace{-.5cm}}
  \label{tab_optimal_tolerance}
\end{table}

Each of the approximation techniques presented incorporates bootstrapping based on an allowable error tolerance $t$. An analysis of the effects of bootstrapping error tolerances on validation accuracy can be seen in Figure \ref{fig_all_performance}. The approximation meshes perform best on the forest fire and Parkinson's data sets when the error tolerance used for fitting is large (smoothing rather than interpolating), while near-interpolation generally produces the most accurate models for HPC I/O. Another performance result of note is that the $MBM$ and $IBM$ have very similar basis functions with largely different outputs.

\begin{figure*}
  \begin{tikzpicture}
    \node (img)  {\includegraphics[width=0.95\textwidth,height=9.5cm]{all-performance.pdf}};
    \node[below=of img, node distance=0cm, yshift=1cm] {Relative Error Tolerance while Bootstrapping};
    \node[left=of img, node distance=0cm, rotate=90, anchor=center,yshift=-0.7cm] {Signed Relative Error};
  \end{tikzpicture}
  \caption{The performance of all three techniques with varied relative error tolerance for the bootstrapping parameter. The columns are for Max Box Mesh, Iterative Box Mesh, and Voronoi Mesh, respectively. The rows are for HPC I/O, Forest Fire, and Parkinson's respectively. Notice the techniques' behavior on the Parkinson's and Forest Fire data sets, performance increases with larger error tolerance.}
  \label{fig_all_performance}
\end{figure*}

\begin{figure*}
  \begin{tikzpicture}
    \node (img)  {\includegraphics[width=0.95\textwidth,height=9.5cm]{perf-sample.pdf}};
    \node[below=of img, node distance=0cm, yshift=1cm] {Signed Relative Error in Prediction};
    \node[left=of img, node distance=0cm, rotate=90, anchor=center,yshift=-0.7cm] {Count};
  \end{tikzpicture}
  \caption{A sample of relative errors for all three techniques with optimal selections of error tolerance. The columns are for Max Box Mesh, Iterative Box Mesh, and Voronoi Mesh, respectively. The rows are for HPC I/O, Forest Fire, and Parkinson's respectively.
    \vspace{-.3cm}}
  \label{fig_perf_sample}
\end{figure*}

The selection of bootstrapping error tolerance also effects the computation time required to fit each of the models to data. Figure \ref{fig_eval_times} presents the time required to construct approximations for each model and each data set with varying $t$. The rapid reduction in computation time required for the forest fire and HPC I/O data sets suggests that large reductions in error can be achieved with relatively few basis functions. The Parkinson's data set however presents a more noisy response, with increasing number of basis functions reducing error much less quickly.

The distributions of errors experienced by each approximation technique when the optimal bootstrapping relative error tolerance is selected can be seen in Figure \ref{fig_perf_sample}. HPC I/O exhibits the most normal approximation errors, which suggests that the models are converging on the random noise of the response for the data set. The worst relative approximation errors are produced by the Voronoi mesh on the forest fire data set. The small magnitude true response values contribute to the larger relative errors. Regardless, the $VM$ errors are unacceptably large.

\section{Discussion}

The bootstrapping procedure presented for each approximation technique still has much room for improvement. Initial analysis suggests that the appropriate relative error tolerance needs to be discovered empirically for each application of a modeling technique. Further analytic studies could arrive at methods for determining optimal error tolerances at runtime, however increases in runtime complexity may not be afforded in many applications.

The box-shaped basis functions and the construction algorithms used for the $MBM$ and $IBM$ could become a source of error when $d$ (the dimension of the data $X$) is comparable to $n$ (the number of known points). The blending regions in which multiple basis functions overlap are always axis aligned and in applications such as image analysis, any single dimension may be unsuitable for approximating the true underlying function. The Voronoi mesh attempts to address this problem by utilizing boundaries between points in multiple dimensions simultaneously. However, it is empirically unclear whether the true benefits of the $VM$ are seen in applications where $d \ll n$.

Each of the case studies presented have fewer than $1000$ points. The complexities of the presented approximation techniques are suitable for large dimension, but the increased complexity associated with brute-force bootstrapping currently prohibits their use on larger data sets. The Voronoi mesh in particular has a large complexity with respect to $n$ which could be significantly improved via more greedy bootstrapping. While each technique requires less than ten seconds on average to produce a fit in the presented case studies, the fit time required quickly grows into minutes around $1000$ points. While these initial results appear somewhat limiting, they demonstrate the viability of each mesh and leave room for further theoretical exploration of techniques to reduce the runtime complexity while maintaining the approximation power and flexibility.

\section{Conclusion}

The Max Box Mesh, Iterative Box Mesh, and Voronoi Mesh each provide novel strategies for effectively approximating multivariate phenomonon. The underlying constructions are theoretically straightforward, yet powerful and flexible. The computational complexities of each make them particularly suitable for applications in many dimensions, while the bootstrapping error tolerance parameter allows a balance between smoothing and interpolation to be explored empirically with each application.

\subsection{Future Work}

A thorough comparison with constituent multivariate approximation techniques including but not limited to, linear Shepard interpolation, multivariate adaptive regression splines, multilayer perceptron regression, and Delaunay triangulation constitutes future work. A more detailed study of alternative bootstrapping techniques may also provide valuable insight.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%% ===================================================================
\section{DATA}
\label{sec:data}

\begin{table}
  \centering
  \begin{tabular}{c|c}
    \hline
    \textbf{System Parameter} & \textbf{Values}\\
    \hline
    File Size (KB) & 4, 16, 64, 256, 1024, 4096, 8192, 16384\\
    \hline
    Record Size (KB) & \multilinecell{4, 8, 16, 32, 64, 128, 256, 512,\\ 1024, 2048, 4096, 8192, 16384}\\
    \hline
    Thread Count & 1, 8, 16, 24, 32, 40, 48, 56, 64\\
    \hline
    Frequency (GHz) & 1.2, 1.6, 2, 2.3, 2.8, 3.2, 3.5\\
    \hline
    Test Type & \multilinecell{Readers, Rereaders, Random Readers, \\ Initial Writers, Rewriters, Random Writers}\\
    \hline
  \end{tabular}
  \caption{A description of system parameters considered for IOzone. Record size must be $\leq$ file size during execution.
    \vspace{-.5cm}}
  \label{tab:data_description}
\end{table}

\begin{figure}
  \includegraphics[width=.5\textwidth]{plot-histogram-throughput.pdf}
  \caption{Histogram of the raw throughput values recorded during all IOzone tests across all system configurations. The distribution is skewed right, with few tests having significantly higher throughput than most others.
  \vspace{-.5cm}}
  \label{fig:throughput_histogram}
\end{figure}

This paper presents a variability modeling case study with a five-dimensional dataset produced by executing the IOzone benchmark \cite{iozone} on a homogeneous cluster of computers. Each node contains two Intel Xeon E5-2637 CPUs offering a total of 16 CPU cores with 16GB of DRAM. While the CPU frequency varies depending on the test configuration, the I/O from IOzone is performed by an ext4 filesystem sitting above an Intel SSDSC2BA20 SSD drive. At the time of data collection, Linux kernel Version 4.13.0 was used. The system performance data was collected over two weeks by executing IOzone 150 times for each of a select set of approximately 18K system configurations, for a total of approximately 2.7M executions of IOzone. A single IOzone execution reports the max I/O throughput in kilobytes per second seen for the selected test type. The summary of the data components in $x^{(i)}$ for the experiments for this paper can be seen in Table \ref{tab:data_description}. Distributions of raw throughput values being modeled can be seen in Figure \ref{fig:throughput_histogram}.

Some mild preprocessing was necessary to prepare the data for modeling and analysis. All features were shifted by their minimum value and scaled by their range, mapping each feature independently into $[0,1]$. This normalization ensures each feature is treated equally by the interpolation techniques and should be performed on all data before building models and making predictions regardless of application. All 150 repeated trials for a system configuration were grouped with that configuration. The only nonordinal feature in this data is the test type. All test types were treated as different applications and were separated for modeling and analysis, i.e., predictions for the ``readers'' test type were made using only known configurations for the ``readers'' test type.

%% ===================================================================

\begin{figure}
  \begin{tikzpicture}
    \node (img)  {\includegraphics[width=0.5\textwidth,trim={1.5cm 0 0 0}]{plot-histogram-80_20-KS.pdf}};
    \node[below=of img, node distance=1cm, yshift=1cm] {KS Statistic for Predicted vs. Actual};
    \node[left=of img, node distance=0cm, rotate=90, anchor=center,yshift=-0.7cm] {Count of Predictions with Given KS Statistic};
  \end{tikzpicture}
  \caption{Histograms of the prediction error for each modeling algorithm from ten random splits when trained with 80\% of the data aggregated over all different test types. The distributions show the KS statistics for the predicted throughput distribution versus the actual throughput distribution. The four vertical red lines represent commonly used $p$-values \{0.05, 0.01, 0.001, 1.0e-6\} respectively. All predictions to the right of a red line represent CDF predictions that are significantly different (by respective $p$-value) from the actual distribution according to the KS test.
  \vspace{-.1cm}}
  \label{fig:ks_histogram_80_20}
\end{figure}

\section{RESULTS}
\label{sec:results}

All three interpolation techniques are used to predict the distribution of I/O throughput values at previously unseen system configurations. In order to improve robustness of the error analysis, ten random selections of 80\% of the IOzone data are used to train each model and the remaining 20\% provide approximation error for each model. The recorded errors are grouped by unique system configuration and then averaged within each group. The samples are identical for each interpolation technique, ensuring consistency in the training and testing sets.

The aggregation of errors across all IOzone tests given 80\% of the data as training can be seen in Figure \ref{fig:ks_histogram_80_20}. Agglomerate errors for each technique resemble a Gamma distribution. The percentages of significant prediction errors with varying $p$-values are on display in Table \ref{tab:p_value_failure_rate}. The primary $p$-value used for analyses in this work is 0.001, chosen because close to 2K predictions are made for each test type. Also, applications executed in cloud and HPC systems that could benefit from statistical modeling will be executed at least thousands of times. In line with this knowledge, it is important to ensure that only a small fraction of interpretable results could occur solely under the influence of random chance. When considering the $p=0.001$ results for each technique, a little under half of the predicted CDFs are significantly different from the measured (and presumed) correct CDFs. A rejection rate of 45\% would seem a poor result, however in this situation the complexity of the problem warrants a slightly different interpretation. These predictions are a very \textit{precise} characterization of performance variability, in fact the cumulative distribution function of a random variable is the strongest possible characterization of variability that can be predicted. Globally, only a little under half of the predictions fail to capture \textit{all} of the characteristics of performance variability at new system configurations. It is also demonstrated later in this Section that this result can likely be improved.

\begin{table}
  \centering
  \begin{tabular}{c|c|c}
    \hline
    \textbf{Algorithm} & \textbf{$P$-Value} & \textbf{\% N.H. Rejections} \\
    \hline
    \multilinecell{Delaunay\\Max Box Mesh\\Voronoi Mesh} & .05 & \multilinecell{58.4\\69.3\\61.9}\\
    \hline
    \multilinecell{Delaunay\\Max Box Mesh\\Voronoi Mesh} & .01 & \multilinecell{51.1\\58.4\\53.4}\\
    \hline
    \multilinecell{Delaunay\\Max Box Mesh\\Voronoi Mesh} & .001 & \multilinecell{44.1\\46.9\\45.1}\\
    \hline
    \multilinecell{Delaunay\\Max Box Mesh\\Voronoi Mesh} & 1.0e-6 & \multilinecell{31.4\\26.6\\28.7}\\
    \hline
  \end{tabular}
  \caption{Percent of null hypothesis rejections rate by the KS-test when provided different selections of $p$-values. These accompany the percent of null hypothesis rejection results from Figure \ref{fig:ks_histogram_80_20}. 
    \vspace{-.5cm}}
  \label{tab:p_value_failure_rate}
\end{table}

While interpreting null hypothesis rejection rates for these interpolation techniques, it is important to consider how the rejection rate reduces with increasing amounts of training data. Figure \ref{fig:ks_failure_by_training} displays the change in $p=0.001$ null hypothesis rejection rate with increasing density of training data up to the maximum density allowed by this set. Delaunay interpolation provides the best results with the least training data by about 5\%, but these low density rejection rates are unacceptably high (90\%). Figure \ref{fig:ks_failure_by_training} clearly shows that this data set and/or the system variables used in the models of performance variability is inadequate to capture the full variability map from system parameters to performance CDF. Which or both obtains is not clear. A few well chosen data points can significantly improve the interpolants, and thus a careful study of the rejection instances is warranted, besides enlarging the set of system variables being modeled.

It may be misleading to consider the global performance of each prediction technique across all test types, as some test types are more difficult than others to predict and have more apparent latent variables. In Figure \ref{fig:ks_failure_by_training_and_test}, the relative difficulty of each IOzone test type can be compared. The I/O test types analyzing reads are typically approximated with lower error than those test types analyzing writes. Regardless of test type, in the aggregate results the KS statistics hover consistently around 0.15, demonstrating an impressively low KS statistic for predictions. In order to address the opacity of aggregate analysis, another case study and an application of the methodology from Section \ref{sec:feature_weighting} is presented in Table \ref{tab:optimized_p_value_failure_rate}.

\begin{figure}
  \begin{tikzpicture}
    \node (img)  {\includegraphics[width=0.4\textwidth,]{plot-KS-Failure-by-Training.pdf}};
    \node[below=of img, node distance=1cm, yshift=1cm] {Percentage of Training Data};
    \node[left=of img, node distance=0cm, rotate=90, anchor=center,yshift=-0.7cm] {\% Null Hypothesis Rejections};
  \end{tikzpicture}
  \caption{The performance of each algorithm on the KS test ($p=0.001$) with increasing amounts of training data averaged over all IOzone test types and ten random splits of the data. The training percentages range from 5\% to 95\% in increments of 5\%. Delaunay is the best performer until 95\% of data is used for training, at which Max Box mesh becomes the best performer by a fraction of a percent.
  \vspace{-.1cm}}
  \label{fig:ks_failure_by_training}
\end{figure}

The results presented in Table \ref{tab:optimized_p_value_failure_rate} are achieved by permitting each approximation technique 300 iterations of simulated annealing. In each iteration, the impact of potential weights on the average KS statistic were considered. All weights were kept in the range [0,2], and were applied to the normalized features for frequency, file size, record size, and number of threads. All three approximation techniques had similar optimal weights achieved by simulated annealing of approximately $(.001, 2, 1.7, 1.5)$ for frequency, file size, record size, and number of threads, respectively. Recall that each interpolation technique uses small distances to denote large influences on predictions, meaning that frequency was the most important feature when predicting variability for the ``readers'' test type, followed not-so-closely by number of threads, then record size.

The ``readers'' test type results demonstrate that the underlying prediction techniques work and are capable of seeing rejection rates below 5\% when tuned for a given application. It is important to emphasize that the roughly 95\% of predictions for which the null hypothesis was not rejected are predicting the \textit{precise} distribution of I/O throughput that will be witnessed at a previously unseen system configuration. To the authors' knowledge, there is no existing methodology that is generically applicable to any system performance measure, agnostic of system architecture, and capable of making such powerful predictions.

\begin{table}
  \centering
  \begin{tabular}{c|c|c|c}
    \hline
    \textbf{Algorithm} & \textbf{$P$-Value} & \multilinecell{\textbf{Unweighted}\\\textbf{\% N.H. Rejection}} & \multilinecell{\textbf{Weighted}\\\textbf{\% N.H. Rejection}}\\
    \hline
    \multilinecell{Delaunay\\Max Box Mesh\\Voronoi Mesh} & .05 & \multilinecell{24.9\\21.3\\18.7} & \multilinecell{30.2\\21.2\\11.3}\\
    \hline
    \multilinecell{Delaunay\\Max Box Mesh\\Voronoi Mesh} & .01 & \multilinecell{21.6\\16.4\\14.9} & \multilinecell{27.4\\16.4\\7.0}\\
    \hline
    \multilinecell{Delaunay\\Max Box Mesh\\Voronoi Mesh} & .001 & \multilinecell{19.7\\13.1\\12.3} & \multilinecell{25.4\\13.1\\4.6}\\
    \hline
    \multilinecell{Delaunay\\Max Box Mesh\\Voronoi Mesh} & 1.0e-6 & \multilinecell{17.9\\11.3\\8.5} & \multilinecell{23.4\\11.3\\2.3}\\
    \hline
  \end{tabular}
  \caption{The null hypothesis rejection rates for various $p$-values with the KS-test. These results are strictly for the ``readers'' IOzone test type and show unweighted results as well as the results with weights tuned for minimum error (KS statistic) by 300 iterations of simulated annealing. Notice that the weights identified for the Delaunay model cause data dependent tuning, reducing performance. MaxBoxMesh performance is improved by a negligible amount. VoronoiMesh performance is notably improved.
    \vspace{-.5cm}}
  \label{tab:optimized_p_value_failure_rate}
\end{table}

%% \begin{itemize}
%% \item Case studies of good and bad predictions of distributions made, present 3 (best, median, worst)
%% \item Table / Figure showing error for each of the tests (hardest test to predict)
%% \item Presentation of the top most difficult configurations to predict for each test (and the bad performance at those points)
%% %% \item Comparison to predictions made using the "test" numerical column
%% \end{itemize}

\section{DISCUSSION}
\label{sec:discussion}

The results of the IOzone case study indicate that predicting the CDF of I/O throughput at previously unseen system configurations is a challenging problem. The KS statistic captures the worst part of any prediction and hence provides a conservatively large estimate of approximation error. The average absolute errors in the predicted CDFs are always lower than the KS statistics. However, the KS statistic was chosen because of the important statistical theory surrounding it as an error measure. Considering this circumstance, a nonnegligible volume of predictions provide impressively low levels of error. Powerful predictive tools such as those presented in this work allow for more in-depth analysis of system performance variability. For example, system configurations that are most difficult to predict in these tests are likely ``outlier'' configurations that do not resemble those configurations that share many similar parameters. Analysis of these configurations may provide valuable insight into effective application specific operation of computer systems.

As mentioned in Section \ref{sec:introduction}, no prior work has attempted to model an arbitrary performance measure for a system to such a high degree of precision. All previous statistical modeling attempts capture a few ($<3$) ordinal performance measures. Generating models that have such high degrees of accuracy allows system engineers to identify previously unused configurations that present desired characteristics. Service level agreements (SLAs) in cloud computing environments are cause for capital competition that is affected heavily by system performance \cite{patel2009service}. Users prefer SLAs that allow the most computing power per monetary unit, incentivizing service providers to guarantee the greatest possible performance. Overscheduling and irregular usage patterns force cloud service providers to occasionally overload machines, in which case precise models of system performance can be used to statistically minimize the probability of SLA violation. Similar targeted performance tuning techniques can be applied to HPC system configuration to maximize application throughput or minimize system power consumption.

A final application domain affected by this methodology is computer security. Collocated users on cloud systems have received attention recently \cite{ali2015security}. If a malicious collocated user is capable of achieving specific insight into the configuration of the system, or the activity of other collocated users by executing performance evaluation programs (i.e., IOzone), a new attack vector may present itself. Malicious users could be capable of identifying common performance distributions of vulnerable system configurations and vulnerable active user jobs. This knowledge may allow targeted exploits to be executed. Light inspection of raw IOzone I/O throughputs provides substantial evidence that distinct performance distributions coincide closely with specific system configuration parameters. Conversely, a service provider may defend against such attacks by deliberately obfuscating the performance of the machine. Models such as those presented in this paper could identify optimal staggering and time-delay whose introduction into the system would prevent malicious users from identifying system configurations and active jobs.

\begin{figure}
  \begin{tikzpicture}
    \node (img)  {\includegraphics[width=0.4\textwidth,]{plot-KS-failure-by-training-and-test.pdf}};
    \node[below=of img, node distance=1cm, yshift=1cm] {Percentage of Training Data};
    \node[left=of img, node distance=0cm, rotate=90, anchor=center,yshift=-0.7cm] {\% Null Hypothesis Rejections};
  \end{tikzpicture}
  \caption{The percentage of null hypothesis rejections for predictions made by each algorithm on the KS test ($p=0.001$) over different IOzone test types with increasing amounts of training data. Each percentage of null hypothesis rejections is an average over ten random splits of the data. The training percentages range from 5\% to 95\% in increments of 5\%. The read test types tend to allow lower rejection rates than the write test types.
  \vspace{-.1cm}}
  \label{fig:ks_failure_by_training_and_test}
\end{figure}

Results presented in Table \ref{tab:optimized_p_value_failure_rate} are particularly interesting, demonstrating that Delaunay appears most vulnerable to data dependent tuning, Max Box mesh is largely insensitive to such tuning, and Voronoi mesh benefits (for this data set) from the tuning.

There are many avenues for extending this modeling methodology. One extension is to add categorical variables to the models. Presently the rejection rate of distribution predictions can only be reduced with large volumes of performance data, however the judicious choice (via experimental design, e.g.) of new data points may be able to effectively reduce the amount of training data required. Finally, more case studies need to be done to test the robustness of the present modeling techniques to changes in domain and performance measure.

\section{CONCLUSION}
\label{sec:conclusion}

The methodology presented is capable of providing new insights, extending existing analyses, and improving the management of computational performance variability. Delaunay, Max Box mesh, and Voronoi mesh interpolation are viable techniques for constructing approximations of performance cumulative distribution functions. A case study on I/O throughput demonstrated that the models are capable of effectively predicting CDFs for most unseen system configurations for any of the available I/O test types. The present methodology represents a notable increase in the ability to statistically model arbitrary system performance measures involving the interaction of many ordinal system parameters.
%% \begin{itemize}
%% \item Modeling the distributions of system performance is valuable.
%% \item This work presents a steep increase in the power of predictions made by statistical modeling techniques.
%% \end{itemize}

%%====================================================================
%%====================================================================
\bibliographystyle{IEEEtran}
\bibliography{paper}

\end{document}

### 2019-02-08 Documents_[2019]/approximation/paper.tex
Ready to run approximation experiments.

--- 2019-02-08 Documents_[2019]/approximation/Old/paper_old.tex
@@
-\documentclass[sigconf]{acmart}
-
-\author{Thomas C. H. Lux}
-%\authornote{Dr.~Trovato insisted his name be first.}
-%\orcid{1234-5678-9012}
-\affiliation{%
-  \department{Dept. of Computer Science}
-  \institution{Virginia Polytechnic Institute and State University}
-  \city{Blacksburg} 
-  \state{Virginia} 
-  \postcode{24060}
-}
-\email{tchlux@vt.edu}
-
-\author{Layne T. Watson}
-%\authornote{Dr.~Trovato insisted his name be first.}
-%\orcid{1234-5678-9012}
-\affiliation{%
-  \department{Depts. of Computer Science, Mathematics, and Aerospace \& Ocean Engineering}
-  \institution{Virginia Polytechnic Institute and State University}
-}
-
-\author{Tyler H. Chang}
-\affiliation{%
-  \department{Dept. of Computer Science}
-  \institution{Virginia Polytechnic Institute and State University}
-}
-
-% The default list of authors is too long for headers.
-\renewcommand{\shortauthors}{T. Lux et al.}
-\def\mn#1{\left\Vert #1\right\Vert_\infty}
-
-\def\Bigl{\mathopen\Big}
-\def\Bigr{\mathclose\Big}
-
-\title{Interpolation of Sparse High-Dimensional Data*}
-
-%% \documentclass[letterpaper, 10 pt, conference]{ieeeconf}  % Comment this line out
-%%                                                           % if you need a4paper
-%% %\documentclass[a4paper, 10pt, conference]{ieeeconf}      % Use this line for a4 paper
-%% \IEEEoverridecommandlockouts                              % This command is only needed 
-%% %% % copyright info
-%% %% \IEEEpubid{\makebox[\columnwidth]{978-1-5386-6133-8/18/\$31.00
-%% %%         ~\copyright~2018 IEEE \hfill}\hspace{\columnsep}
-%% %%         \makebox[\columnwidth]{ }}
-
-%% % if you want to use the \thanks command
-%% \overrideIEEEmargins
-%% % See the \addtolength command later in the file to balance the column lengths on the last page of the document
-
-%% \title{\LARGE \bf Interpolants Scalable by Dimension* }
-
-%% \author{Thomas C. H. Lux$^{1}$, Layne T. Watson$^{2}$, Tyler H. Chang$^{1}$% <-this % stops a space
-%% \thanks{*This work was supported by the National Science Foundation Grant CNS-1565314.}%
-%% \thanks{$^{1}$Doctoral student, Department of Computer Science, Virginia Polytechnic Institute and State University, Blacksburg, Virginia 24060, USA {\tt\small (tchlux at vt.edu)}}%
-%% \thanks{$^{2}$Department of Computer Science, Virginia Polytechnic Institute and State University, Blacksburg, Virginia 24060, USA}%
-%% }
-
-% The following packages can be found on http:\\www.ctan.org
+\RequirePackage{fix-cm}
+\documentclass[smallextended]{svjour3}       % onecolumn (second format)
+\smartqed  % flush right qed marks, e.g. at end of proof
+\usepackage{graphicx}
+%
+% \usepackage{mathptmx}      % use Times fonts if available on your TeX system
+%
+% insert here the call for the packages your document requires
+%\usepackage{latexsym}
 \usepackage{tikz}     % For adding axes to plots
@@
 \usepackage{booktabs} % For formal tables
-\usepackage{hyphenat} % custom hyphenation rules
-\hyphenation{op-tical net-works semi-conduc-tor}
 
-%% \usepackage{color}   % Needed to color text for drafts
-%% \usepackage{algorithm}     % needed for the algorithm block
-%% \usepackage{algpseudocode} % algorithmicx psuedo-code styling
-%% \usepackage[pdftex,colorlinks=true,urlcolor=blue,citecolor=black,anchorcolor=black,linkcolor=black]{hyperref}
-%% \usepackage[dvips,colorlinks=true,urlcolor=blue,citecolor=black,anchorcolor=black,linkcolor=black]{hyperref}
+%
+% please place your own definitions here and don't use \def but
+% \newcommand{}{}
 
@@
 
+%% \newcommand{\Bigl}{\mathopen\Big}
+%% \newcommand{\Bigr}{\mathclose\Big}
+%
+% Insert the name of "your journal" with
+\journalname{Numerical Algorithms}
+%
 \begin{document}
 
+\title{Interpolation of Sparse High-Dimensional Data\thanks{This
+    projected was supported by NSF grant}
+}
+% \subtitle{Do you have a subtitle?\\ If so, write it here}
+
+%\titlerunning{Short form of title}        % if too long for running head
+
+\author{Thomas C. H. Lux   \and
+  Layne T. Watson          \and
+  Tyler H. Chang           \and
+  Yili Hong                \and
+  Kirk Cameron
+}
+
+%\authorrunning{Short form of author list} % if too long for running head
+
+\institute{Thomas Lux \at
+  Virginia Polytechnic Institute \& State University (VPI\&SU) \\
+  Blacksburg, VA 24061 \\
+  \email{tchlux@vt.edu}
+}
+
+\date{Received: date / Accepted: date}
+% The correct dates will be entered by the editor
+
 \maketitle
-\thispagestyle{empty}
-\pagestyle{empty}
-%%====================================================================
-%%                    BODY OF PAPER
-%%====================================================================
 
-\section*{Abstract}
-
+\begin{abstract}
 A rapid increase in the quantity of data available is allowing all fields of science to generate more accurate models of multivariate phenomena.
@@
 
-\textbf{Keywords:} Approximation, regression, interpolation, high dimension, error bound
+\keywords{approximation \and regression \and interpolation \and high dimension \and error bound}
+% \PACS{PACS code1 \and PACS code2 \and more}
+% \subclass{MSC code1 \and MSC code2 \and more}
+\end{abstract}
 
@@
 For example, least squares regression uses the sum of squared differences between modeled function values and true function values as an error measure.
-In this section and the next, some techniques are limited to approximating real valued functions ($Y \subset \Re$).
+In this section and the next, some techniques are limited to approximating real valued functions ($Y \subset \mathbb{R}$).
 These techniques are extended to real vector-valued functions by repeating the construction for each component of the vector outputs.
@@
 $v^{x^{(i)}}\bigl(x^{(j)}\bigr) = \delta_{ij}$ and $v^{x^{(i)}}$ has local support, giving the interpolated value
-$$ f(y) = \frac{\sum_{i=1}^n v^{x^{(i)}}(y) f(x^{(i)})}{\sum_{i=1}^n v^{x^{(i)}}(y)}, $$
+$$ f(y) = \frac{\sum\limits_{i=1}^n v^{x^{(i)}}(y) f(x^{(i)})}{\sum\limits_{i=1}^n v^{x^{(i)}}(y)}, $$
 where $0 \leq v^{x^{(i)}}(y) \leq 1$.
@@
   \begin{tikzpicture}
-    \node (img)  {\includegraphics[width=0.4\textwidth,]{plot-delaunay-example-prediction.pdf}};
+    \node (img)  {\includegraphics[width=0.4\textwidth,]{Figures/plot-delaunay-example-prediction.pdf}};
     \node[below=of img, node distance=1cm, yshift=1cm] {I/O Throughput};
@@
 The performance of approximation techniques that predict probability functions can be analyzed through a variety of summary statistics.
-Mean absolute error, mean squared error, and the max absolute error are all popular measures.
 This work uses the max absolute error, also known as the Kolmogorov-Smirnov (KS) statistic \cite{lilliefors1967kolmogorov} for its compatibility with the KS test.
@@
 The null hypothesis (that two CDFs come from the same underlying distribution) is rejected at level $p \in [0,1]$ when
-
 $$ KS > \sqrt{-\frac{1}{2}\ln\biggl(\frac{p}{2}\biggr)} \sqrt{\frac{1}{n_1} + \frac{1}{n_2}}, $$
-
 with distribution sample sizes $n_1,n_2 \in \mathcal{N}$.
@@
 
-\noindent \textbf{\textit{Lemma 1.}} Let $S \subset \mathbb{R}^d$ be open and convex, $\ f: \mathbb{R}^d \rightarrow \mathbb{R},$ and $\nabla f \in Lip_{(\gamma,\|\cdot\|_2)}(S)$, the set of $\gamma$-Lipschitz continuous functions in the $2$-norm.
+\noindent \textbf{\textit{Lemma 1.}} Let $S \subset \mathbb{R}^d$ be open and convex, $f: \mathbb{R}^d \rightarrow \mathbb{R},$ and $\nabla f \in Lip_{(\gamma,\|\cdot\|_2)}(S)$, the set of $\gamma$-Lipschitz continuous functions in the $2$-norm.
 Then for all $x,y \in S$
-
-$$\big|\ f(y) - f(x) - \langle \nabla f(x), y - x \rangle \big| \leq \frac{\gamma \|y - x\|_2^2}{2}.$$
+%
+$$\big|f(y) - f(x) - \langle \nabla f(x), y - x \rangle \big| \leq \frac{\gamma \|y - x\|_2^2}{2}.$$
 
 \noindent \textit{Proof.} Consider the function $g(t) = f \big((1-t) x + t y \big),$ $0 \leq t \leq 1,$ whose derivative $g'(t) = \big\langle \nabla f \big((1-t) x + t y \big), y - x \big\rangle$ is the directional derivative of $f$ in the direction $(y - x).$
-
+%
 \begin{align*}
-  \big|\ f(y) - f&(x) - \langle \nabla f(x), y - x \rangle \big| &\\ 
-     &= \big|\ g(1) - g(0) - g'(0) \big| & \\
+  \big|f(y) - f(x) - \langle \nabla f(x), y - x \rangle \big|
+     &= \big|g(1) - g(0) - g'(0) \big| & \\
      &= \bigg| \int_0^1 g'(t) - g'(0)\ dt \bigg| \\
@@
 \end{align*}
-
-$\qed$
+%
+\qed
 
 \noindent \textbf{\textit{Lemma 2.}} Let $x, y, v_i \in \mathbb{R}^d,$ $c_i \in \mathbb{R},$ and $|\langle y - x, v_i \rangle| \leq c_i$ for $i = 1,$ $\ldots,$ $d.$ If $M = (v_1,$ $\ldots,$ $v_d)$ is nonsingular, then
-
+%
 $$\|y - x\|_2^2 \leq \frac{1}{\sigma_d^2} \sum_{i=1}^d c_i^2,$$
-
+%
 where $\sigma_d$ is the smallest singular value of $M.$
@@
 \noindent \textit{Proof.} Using the facts that $M$ and $M^t$ have the same singular values, and $\|M^tw\|_2 \geq \sigma_d \|w\|_2,$ gives
-
+%
 \begin{align*}
@@
 \end{align*}
+%
+\qed
 
-$\qed$
-
-\noindent \textbf{\textit{Lemma 3.}} Given $\ f,$ $\gamma,$ $S$ as in {\it Lemma 1}, let $X = \{x_0,$ $x_1,$ $\ldots,$ $x_d\}$ $\subset S$ be the vertices of a $d$-simplex, and let $\hat f(x) = \langle c, x - x_0 \rangle + f(x_0),$ $c \in \mathbb{R}^d$ be the linear function interpolating $\ f$ on $X.$ Let $\sigma_d$ be the smallest singular value of the matrix $M = (x_1 - x_0,$ $\ldots,$ $x_d - x_0),$ and $k = \max\limits_{1\ \leq\ j\ \leq\ d} \|x_j - x_0\|_2.$ Then
-
+\noindent \textbf{\textit{Lemma 3.}} Given $f,$ $\gamma,$ $S$ as in {\it Lemma 1}, let $X = \{x_0,$ $x_1,$ $\ldots,$ $x_d\}$ $\subset S$ be the vertices of a $d$-simplex, and let $\hat f(x) = \langle c, x - x_0 \rangle + f(x_0),$ $c \in \mathbb{R}^d$ be the linear function interpolating $f$ on $X.$ Let $\sigma_d$ be the smallest singular value of the matrix $M = (x_1 - x_0,$ $\ldots,$ $x_d - x_0),$ and $k = \max\limits_{1\ \leq\ j\ \leq\ d} \|x_j - x_0\|_2.$ Then
+%
 $$\big\|\nabla f(x_0) - c\big\|_2 \leq \sqrt{d} \frac{\gamma k^2}{\sigma_d}.$$
@@
 \noindent \textit{Proof.} Consider $f(x) - \hat f(x)$ along the line segment $z(t) = (1-t)x_0 + t x_j,$ $0 \leq t \leq 1.$ By Rolle's Theorem, for some $0 < \hat t < 1,$ $\big\langle \nabla f\big(z(\ \hat t\ )\big) - c, x_j - x_0 \big\rangle = 0.$ Now
-
+%
 \begin{align*}
-       \big| \big< \nabla f(x_0) &- c, x_j - x_0 \big \rangle \big| \\
+  \big| \big< \nabla f(x_0) - c, x_j - x_0 \big \rangle \big|
         &= \big| \big< \nabla f(x_0) - \nabla f\big(z(\ \hat t\ )\big) + \nabla f\big(z(\ \hat t\ )\big) - c, x_j - x_0 \big \rangle \big| \\
@@
 \end{align*}
-
+%
 for all $1 \leq j \leq d.$ Using {\it Lemma 2},
-
+%
 $$\big \| \nabla f(x_i) - c \big \|_2^2 \leq \frac{d}{\sigma_d^2} \big( \gamma k^2\big)^2 \implies \big \| \nabla f(x_i) - c \big \|_2 \leq \sqrt{d} \frac{\gamma k^2}{\sigma_d}.$$
-
-$\qed$
+%
+\qed
 
 \noindent \textbf{\textit{Theorem.}} Under the assumptions of {\it Lemma 1} and {\it Lemma 3}, for $z \in S,$
-
-$$ \big|\ f(z) - \hat f(z)\big| \leq \frac{\gamma \|z - x_0\|_2^2}{2} + \sqrt{d} \frac{\gamma k^2}{\sigma_d} \|z - x_0\|_2.$$
+%
+$$ \big|f(z) - \hat f(z)\big| \leq \frac{\gamma \|z - x_0\|_2^2}{2} + \sqrt{d} \frac{\gamma k^2}{\sigma_d} \|z - x_0\|_2.$$
 
 \noindent \textit{Proof.} Let $v = \nabla f(x_0) - c,$ where $\|v\|_2 \leq \sqrt{d} \gamma k^2 / \sigma_d$ by {\it Lemma 3}. Now
-
+%
 \begin{align*}
-       \big|\ f(z) - \hat f(z)\ \big| &= \big|\ f(z) - f(x_0) - \langle c, z - x_0 \rangle \big| \\ 
-       &= \big|\ f(z) - f(x_0) - \langle \nabla f(x_0) - v, z - x_0 \rangle \big| \\
-       &= \big|\ f(z) - f(x_0) - \langle \nabla f(x_0) , z - x_0 \rangle + \langle v , z - x_0 \rangle \big| \\
-       &\leq \big|\ f(z) - f(x_0) - \langle \nabla f(x_0) , z - x_0 \rangle \big| + \big| \langle v , z - x_0 \rangle \big| \\
-       &\leq \big|\ f(z) - f(x_0) - \langle \nabla f(x_0) , z - x_0 \rangle \big| + \|v\|_2 \ \|z - x_0\|_2 \\
-       &\leq \big|\ f(z) - f(x_0) - \langle \nabla f(x_0), z - x_0 \rangle \big| + \textstyle{\frac{\gamma k^2 \sqrt{d}}{\sigma_d}} \|z - x_0\|_2 \\
+  \big|f(z) - \hat f(z)\big|
+       &= \big|f(z) - f(x_0) - \langle c, z - x_0 \rangle \big| \\ 
+       &= \big|f(z) - f(x_0) - \langle \nabla f(x_0) - v, z - x_0 \rangle \big| \\
+       &= \big|f(z) - f(x_0) - \langle \nabla f(x_0) , z - x_0 \rangle + \langle v , z - x_0 \rangle \big| \\
+       &\leq \big|f(z) - f(x_0) - \langle \nabla f(x_0) , z - x_0 \rangle \big| + \big| \langle v , z - x_0 \rangle \big| \\
+       &\leq \big|f(z) - f(x_0) - \langle \nabla f(x_0) , z - x_0 \rangle \big| + \|v\|_2 \|z - x_0\|_2 \\
+       &\leq \big|f(z) - f(x_0) - \langle \nabla f(x_0), z - x_0 \rangle \big| + \textstyle{\frac{\gamma k^2 \sqrt{d}}{\sigma_d}} \|z - x_0\|_2 \\
        &\leq \frac{\gamma \|z - x_0\|_2^2}{2} + \sqrt{d}\frac{\gamma k^2}{\sigma_d} \|z - x_0\|_2,
 \end{align*}
+%
+where the last inequality follows from {\it Lemma 1}.
+%
+\qed
 
-where the last inequality follows from {\it Lemma 1}.
+\section{Planning for Paper}
 
-$\qed$
+\begin{itemize}
+\item Construct analytic function to test theoretical claims made about Delaunay.
+\item Demonstrate the empirical performance of Delaunay, discuss alignment with theory.
+\item Demonstrate similar empirical performance is achieved by other interpolants.
+\item Argue that interpolation is not as commonly applied, but that it works well in practice.
+\item Present the data for predicting real-valued outputs, an empirical study.
+  \begin{itemize}
+  \item Forest fire data
+  \item Parkinsons
+  \item https://www.kaggle.com/jsphyg/weather-dataset-rattle-package
+  \end{itemize}
+\item Present prediction results on each of these data sets for each algorithm.
+\item Discuss the results between the regressors and interpolants.
+\item Present the data for predicting function-valued outputs, an empirical study.
+  \begin{itemize}
+  \item VarSys large scale system configuration prediction
+  \item \textit{galaxy data from Carnegie Melon?}
+  \end{itemize}
+\item Present prediction results on each of these data sets for each algorithm.
+\item Discuss the results between the regressors and interpolants.
+\item Present a summary table showing the performance differences between regerssors and interpolants.
+\item Discuss the overall applicability of interpolants and encourage further application. Restate the benefits of interpolants
+\item Conclude
+\end{itemize}
 
@@
 \begin{figure}
-  \includegraphics[width=.5\textwidth,height=2cm]{p-hist.pdf}
-  \includegraphics[width=.5\textwidth,height=2cm]{f-hist.pdf}
-  \includegraphics[width=.5\textwidth,height=2cm]{h-hist.pdf}
+  \includegraphics[width=.5\textwidth,height=2cm]{Figures/p-hist.pdf}
+  \includegraphics[width=.5\textwidth,height=2cm]{Figures/f-hist.pdf}
+  \includegraphics[width=.5\textwidth,height=2cm]{Figures/h-hist.pdf}
   \caption{Histograms of Parkinsons (total UPDRS), forest fire (area), and HPC I/O (mean throughput) response values respectively.
@@
   \begin{tikzpicture}
-    \node (img)  {\includegraphics[width=0.95\textwidth,height=4cm]{eval-times.pdf}};
+    \node (img)  {\includegraphics[width=0.95\textwidth,height=4cm]{Figures/eval-times.pdf}};
     \node[below=of img, node distance=0cm, yshift=1cm] {Relative Error Tolerance while Bootstrapping};
@@
   \begin{tikzpicture}
-    \node (img)  {\includegraphics[width=0.95\textwidth,height=9.5cm]{all-performance.pdf}};
+    \node (img)  {\includegraphics[width=0.95\textwidth,height=9.5cm]{Figures/all-performance.pdf}};
     \node[below=of img, node distance=0cm, yshift=1cm] {Relative Error Tolerance while Bootstrapping};
@@
   \begin{tikzpicture}
-    \node (img)  {\includegraphics[width=0.95\textwidth,height=9.5cm]{perf-sample.pdf}};
+    \node (img)  {\includegraphics[width=0.95\textwidth,height=9.5cm]{Figures/perf-sample.pdf}};
     \node[below=of img, node distance=0cm, yshift=1cm] {Signed Relative Error in Prediction};
@@
 \begin{figure}
-  \includegraphics[width=.5\textwidth]{plot-histogram-throughput.pdf}
+  \includegraphics[width=.5\textwidth]{Figures/plot-histogram-throughput.pdf}
   \caption{Histogram of the raw throughput values recorded during all IOzone tests across all system configurations.
@@
   \begin{tikzpicture}
-    \node (img)  {\includegraphics[width=0.5\textwidth,trim={1.5cm 0 0 0}]{plot-histogram-80_20-KS.pdf}};
+    \node (img)  {\includegraphics[width=0.5\textwidth,trim={1.5cm 0 0 0}]{Figures/plot-histogram-80_20-KS.pdf}};
     \node[below=of img, node distance=1cm, yshift=1cm] {KS Statistic for Predicted vs. Actual};
@@
   \begin{tikzpicture}
-    \node (img)  {\includegraphics[width=0.4\textwidth,]{plot-KS-Failure-by-Training.pdf}};
+    \node (img)  {\includegraphics[width=0.4\textwidth,]{Figures/plot-KS-Failure-by-Training.pdf}};
     \node[below=of img, node distance=1cm, yshift=1cm] {Percentage of Training Data};
@@
   \begin{tikzpicture}
-    \node (img)  {\includegraphics[width=0.4\textwidth,]{plot-KS-failure-by-training-and-test.pdf}};
+    \node (img)  {\includegraphics[width=0.4\textwidth,]{Figures/plot-KS-failure-by-training-and-test.pdf}};
     \node[below=of img, node distance=1cm, yshift=1cm] {Percentage of Training Data};
@@
 
+%\begin{acknowledgements}
+%If you'd like to thank anyone, place your comments here
+%and remove the percent signs.
+%\end{acknowledgements}
+
 %%====================================================================
 %%====================================================================
-\bibliographystyle{IEEEtran}
+
+\bibliographystyle{spmpsci}      % mathematics and physical sciences
 \bibliography{paper}

### 2019-02-08 Documents_[2019]/approximation/Results/make_predictions.py

@@ -1,0 +1,151 @@
+import os
+import numpy as np
+
+from util.data import Data
+from util.system import Timer
+from util.misc import pairwise_distance
+from util.approximate import Delaunay, Voronoi, BoxMesh, ShepMod, \
+    LSHEP, BFGS1000, MARS, SVR, class_name
+
+# Make sure all of the prepared and pre-processed data files exist.
+from preprocess_data import cwd, raw_dir, data_dir
+
+seed = 0
+folds = 10
+
+data_name = lambda p: os.path.basename(p).replace(".dill","").replace(".csv","")
+
+models = [ShepMod, BoxMesh, Voronoi, Delaunay,
+          BFGS1000, LSHEP, SVR, MARS,]
+
+
+# Recording data for each set and for each algorithm that looks like:
+# | Data name | Dimension | Fold Size | Fold Number | Model Name | Model fit time |
+# | x | ... | y | ... | Model y | ... | Model nearest | Model max edge | Model furthest | Model num contributors | Model prediction time |
+
+fit_data = Data(names=["Data name", "Dimension", "Fold size", "Fold number", "Model name", "Fit time"],
+                types=[str,         int,         int,         int,           str,          float     ])
+
+print()
+for raw_file in sorted(os.listdir(raw_dir)):
+    raw_path = os.path.join(raw_dir, raw_file)
+    data_path = os.path.join(data_dir, raw_file + '.dill')
+    intermediate_results_file = os.path.join(".", "Predictions",
+                    f"intermediate_results_{data_name(raw_file)}.dill")
+    final_results_file = os.path.join(".", "Predictions",
+                    f"results_{data_name(raw_file)}.dill")
+    print('-'*70)
+    print(data_name(raw_path))
+    print(raw_path)
+    print(final_results_file)
+    print()
+    d = Data.load(data_path)
+    target = d.names[-1]
+    num_train_columns = len(d.names) - 1
+    d["Indices"] = range(len(d))
+    print(f"Data shape: {d.shape} predicting '{target}'")
+    for i,(train, test) in enumerate(d.k_fold(k=folds, seed=seed)):
+        print(f"  Fold {i+1} of {folds}..")
+        # Get the numeric representation of this data based on training,
+        # ignore extra added columns and the "target" column.
+        num = train[:,:num_train_columns].to_matrix()
+        train_x = num.data
+        train_y = list(train[target])
+        # Generate the numeric testing set basesd on training.
[…]

### 2019-02-27 Paper_[2019-02]_Numerical_Algorithms/Data/make_visuals.py
Updated code for running neural network.

@@ -1,14 +1,21 @@
 import os, math
+import numpy as np
 from util.plot import Plot
 from util.data import Data
 from util.stats import cdf_fit
+from util.misc.paper import latex_table
 
 results_folder = os.path.join(os.path.dirname(os.path.abspath(__file__)), "Predictions")
+percentiles = [0, 25, 50, 75, 100]
+
 for f in os.listdir(results_folder):
     if ("results" != f[:len("results")]): continue
     path = os.path.join(results_folder, f)
     d = Data.load(path)
+    print()
     print(f, d.shape)
+    print(d)
+    print()
     # Get the target column and the list of models that made predictions.
     target_index = d.names.index("Indices") - 1
     target = d.names[target_index]
@@ -16,16 +23,37 @@ for f in os.listdir(results_folder):
     # Generate a plot of the model errors.
     p = Plot(f"{f} Absolute Error Distributions")
     max_error = -float('inf')
+    table = [["Algorithm", "Min"] + [f"${v}^{{th}}$" for v in percentiles[1:-1]] + ["Max"]]
     for m_name in models:
         errors = [truth - guess
                   for (truth,guess) in zip(d[target], d[f"{m_name} {target}"])
                   if ((truth != None) and (guess != None))]
-        errors = list(map(abs, errors))
-        max_error = max(max_error, max(errors))
-        p.add_box(m_name, errors)
+        print()
+        # Skip bad models.
+        if 'nan' in map(str,errors): errors = [v for v in errors if str(v) != 'nan']
+        if (len(errors) == 0):
+            print(f"Skipping {m_name} because no predictions are found..")
+            continue
+        abs_errors = list(map(abs,errors))
+        max_error = max(max_error, max(abs_errors))
+        print("errors: ",len(errors), errors[:100])
+        print("m_name: ",m_name)
+        print("max_error: ",max_error)
+        p.add_box(m_name, abs_errors)
+        table.append([m_name] + [np.percentile(abs_errors, p) for p in percentiles])
         # print(f"{m_name} absolute relative error (min,max):", (min(errors), max(errors)))
         # fit = cdf_fit(errors)
         # p.add_func(m_name, fit, fit())
+    # Fix the table to have bold entries for the minimal values at each percentile.
+    col_mins = [min(row[i] for row in table[1:]) for i in range(1+len(percentiles))]
+    for row in range(1,len(table)):
+        for col in range(1,len(table[row])):
[…]

### 2019-03-04 Paper_[2019-02]_Numerical_Algorithms/tchlux-NA-2019.tex
Finished the draft of the NA journal paper.

\RequirePackage{fix-cm}
\documentclass[smallextended,final]{svjour3}       % onecolumn (second format)
\smartqed  % flush right qed marks, e.g. at end of proof
\usepackage{graphicx}
%
% \usepackage{mathptmx}      % use Times fonts if available on your TeX system
%
% insert here the call for the packages your document requires
%\usepackage{latexsym}
\usepackage{tikz}     % For adding axes to plots
\usetikzlibrary{positioning}
\usepackage{graphicx} % For including PDFs
\usepackage{amsmath} % assumes amsmath package installed
\usepackage{amssymb}  % assumes amsmath package installed
\usepackage{booktabs} % For formal tables
\usepackage[colorlinks,allcolors=blue]{hyperref}

%
% please place your own definitions here and don't use \def but
% \newcommand{}{}

%% Command for creating multiplie-line cells inside a table environment
\newcommand{\multilinecell}[2][c]{%
  \begin{tabular}[#1]{@{}c@{}}#2\end{tabular}}

%% Create a ``Lemma'' environment that does not have italics font.
\spnewtheorem{plainlemma}{Lemma}{\bf}{}
%% Create a ``Theorem'' environment that is not numbered.
\spnewtheorem*{onetheorem}{Theorem}{\bf}{}
\spnewtheorem*{proofdot}{Proof.}{\it}{}

%% \newcommand{\Bigl}{\mathopen\Big}
%% \newcommand{\Bigr}{\mathclose\Big}
%
% Insert the name of "your journal" with
\journalname{Numerical Algorithms}
%
\begin{document}

\title{Interpolation of Sparse High-Dimensional Data\thanks
  {This work was supported by the National Science Foundation Grant CNS-1565314.}
}
% \subtitle{Do you have a subtitle?\\ If so, write it here}

%\titlerunning{Short form of title}        % if too long for running head

\author{Thomas C. H. Lux   \and
  Layne T. Watson          \and
  Tyler H. Chang           \and
  Yili Hong                \and
  Kirk Cameron
}

%\authorrunning{Short form of author list} % if too long for running head

\institute{Thomas Lux \at
  Virginia Polytechnic Institute \& State University (VPI\&SU) \\
  Blacksburg, VA 24061 \\
  \email{tchlux@vt.edu}
}

\date{Received: date / Accepted: date}
% The correct dates will be entered by the editor

\maketitle

\begin{abstract}
A rapid increase in the quantity of data available is allowing all fields of science to generate more accurate models of multivariate phenomena. Regression and interpolation become challenging when the dimension of data is large, especially while maintaining tractable computational complexity. While regression is a popular approach to solving approximation problems with high dimension, there are often advantages to interpolation. This paper presents error bounds for interpolants on moderately high dimensional problems and contrasts their performance with that of popular regression techniques. Empirical results demonstrate the viability of interpolants for high dimensional approximation problems, suggesting that these techniques are capable of effectively modeling multivariate phenomena while maintaining flexibility in different application domains.

\keywords{approximation \and regression \and interpolation \and high dimension \and error bound}
% \PACS{PACS code1 \and PACS code2 \and more}
% \subclass{MSC code1 \and MSC code2 \and more}
\end{abstract}

%     Introduction     
%======================

%% ===================================================================
\section{Introduction}
\label{sec:introduction}

Regression and interpolation are problems of considerable importance that find applications across many fields of science. Pollution and air quality analysis \cite{de2008field}, energy consumption management \cite{lazos2014optimisation}, and student performance prediction \cite{cortez2008using,lux2016applications} are a few of many interdisciplinary applications of multivariate regression for predictive analysis. As discussed later, these techniques can also be applied to prediction problems related to forest fire risk assessment \cite{cortez2007data}, Parkinson's patient clinical evaluations \cite{tsanas2010accurate}, local rainfall and weather \cite{williams2009rattle}, credit card transactions \cite{pozzolo2015calibrating}, and high performance computing (HPC) file input/output (I/O) \cite{lux2018nonparametric}.

Regression and interpolation have a considerable theoretical base in one dimension \cite{cheney2009course}. Splines in particular are well understood as an interpolation technique in one dimension \cite{de1978practical}, particularly B-splines. Tensor products of B-splines \cite{unther1996interpolating} or other basis functions have an unfortunate exponential scaling with increasing dimension. Exponential scaling prohibits tensor products from being reasonably applied beyond three-dimensional data. In order to address this dimensional scaling challenge, C. de Boor and others proposed box splines \cite{de2013box}, of which one of the approximation techniques in this work is composed \cite{lux2018novel}.

The theoretical foundation of low dimensional interpolation allows the construction of strong error bounds that are absent from high dimensional problems. This work extends some known results regarding the secant method \cite{dennis1996numerical} to construct an interpolation error bound for problems of arbitrary dimension. These error bounds are useful, considering the same cannot be said for regression algorithms in general. The maximum complexity of an interpolant is bounded by the amount of data available, while the maximum complexity of a regressor is bounded by both the amount of data and the chosen parametric form. For this reason, generic uniform bounds are largely unobtainable for regression techniques on arbitrary approximation problems, even when the approximation domain is bounded.

Aside from theoretical motivation for the use of interpolants, there are often computational advantages as well. Interpolants do not have the need for \textit{fitting} data, or minimizing error with respect to model parameters. In applications where the amount of data is large and the relative number of predictions that need to be made for a given collection of data is small, the direct application of an interpolant is much less computationally expensive.

In this work, multivariate interpolation is defined given a closed convex subset $Y$ of a metrizable topological vector space with metric $s$, some function $f:\mathbb{R}^d \rightarrow Y$ and a set of points $X = \bigl\{x^{(1)}$,$\ldots$,$x^{(n)}\bigr\} \subset \mathbb{R}^d,$ along with associated function values $f\bigl(x^{(i)}\bigr)$. The goal is to construct an approximation $\hat f: \mathbb{R}^d \rightarrow Y$ such that $\hat f\bigl(x^{(i)}\bigr) = f\bigl(x^{(i)}\bigr)$ for all $i = 1$,$\ldots$,$n$. It is often the case that the form of the true underlying function $f$ is unknown, however it is still desirable to construct an approximation $\hat f$ with small approximation error at $y \notin X$. The two metric spaces that will be discussed in this work are the real numbers with metric $s(x,y) = |x-y|$, and the set of cumulative distribution functions (CDFs) with the Kolmogorov-Smirnov (KS) statistic as metric.

Multivariate regression is often used when the underlying function is presumed to be stochastic, or stochastic error is introduced in the evaluation of $f$. Hence, multivariate regression relaxes the conditions of interpolation by choosing parameters $P$ defining $\hat f(x;P)$ to minimize the error vector $\Bigl( \bigl | \hat f \bigl(x^{(1)};P\bigr) - f\bigl(x^{(1)}\bigr) \bigr|$, $\ldots$, $\bigl | \hat f \bigl(x^{(n)}; P\bigr) - f\bigl(x^{(n)}\bigr) \bigr | \Bigr)$ in some norm. The difficult question in the case of regression is often what parametric form to adopt for any given application.

The most challenging problem when scaling in dimension is that the number of possible interactions between dimensions grows exponentially. Quantifying all possible interactions becomes intractable and hence beyond three-dimensional data, mostly linear models are used. That is not to say nonlinear models are absent, but nonlinearities are often either preconceived or model pairwise interactions between dimensions at most. Even globally nonlinear approximations such as neural networks are constructed from compositions of summed low-interaction functions \cite{clevert2015fast}.

Provided the theoretical and practical motivations for exploring interpolants, this work aims to study the empirical performance differences between a set of scalable (moderately) interpolation techniques and a set of common regression techniques. These techniques are applied to a collection of moderately high dimensional problems ($5 \le d \le 30$) and the empirical results are discussed.

The remainder of this paper is organized as follows. Sections \ref{sec:regression} and \ref{sec:interpolation} present the multivariate models. Section \ref{sec:error} presents the error measuring methodology that is used for collecting empirical results. Section \ref{sec:theory} presents the theoretical error bounds for interpolation. Section \ref{sec:data} presents the data sets for empirical approximation analysis and presents the approximation results for all models on these data sets. Section \ref{sec:discussion} analyzes and discusses the results, their implications, and their limitations. Finally, Section \ref{sec:conclusion} concludes.

% Art is a subjective form of personal expression and computers are
% just robots that are going to take over the world. \textit{*deep
% breath*} In order to communicate my frustration with the
% technological and information revolution, I'm going to write the
% rest of this paper in poetry.

% Star light, star bright \ldots

%% ===================================================================
\section{Multivariate Regression}
\label{sec:regression}
Multivariate regressors are capable of accurately modeling a complex dependence of a response (in $Y$) on multiple variables (represented as a points in $\mathbb{R}^{d}$). The approximations to some (unknown) underlying function $f: \mathbb{R}^d \rightarrow Y$ are chosen to minimize some error measure related to data samples $f\bigl(x^{(i)}\bigr)$. For example, least squares regression uses the sum of squared differences between modeled function values and true function values as an error measure. In this section and the next, some techniques are limited to approximating real valued functions ($Y \subset \mathbb{R}$). These techniques can be extended to real vector-valued ranges by repeating the construction for each component of the vector output. Throughout the following, $x$ denotes a $d$-tuple, $x_i$ the $i$th component of $x$, and $x^{(i)}$ the $i$th $d$-tuple data point. Different symbols are used to represent the approximation function $\hat f$.

\subsection{Multivariate Adaptive Regression Splines}
\label{sec:mars}
This approximation was introduced in \cite{friedman1991multivariate} and subsequently improved to its current version in \cite{stanford1993fast}, called fast multivariate adaptive regression splines (Fast MARS).In Fast MARS, a least squares fit model is iteratively built by beginning with a single constant valued function and adding two new basis functions at each iteration of the form
\begin{align*}
  B_{2j-1}(x) &= B_l(x) \bigl(x_i-x^{(p)}_i\bigr)_+, \\
  B_{2j}(x) &= B_k(x) \bigl(x_i-x^{(p)}_i\bigr)_- ,
\end{align*}
where $j$ is the iteration number, $1 \le p \le n$, and $B_l(x)$, $B_k(x)$ are basis functions from the previous iteration,
$$w_+ = \begin{cases} w, & w \geq 0 \\ 0, & w < 0 \end{cases},$$
and $w_- = (-w)_+$. After iteratively constructing a model, MARS then iteratively removes basis functions that do not contribute to goodness of fit. In effect, MARS creates a locally component-wise tensor product approximation of the data. The overall computational complexity of Fast MARS is $\mathcal{O}(n d m^3)$ where $m$ is the maximum number of underlying basis functions. This paper uses an implementation of Fast MARS \cite{rudy2017pyearth} with $m = 200$ throughout all experiments.

\subsection{Support Vector Regressor}
\label{sec:svr}
Support vector machines are a common method used in machine learning classification tasks that can be adapted for the purpose of regression \cite{basak2007support}. How to build a support vector regressor (SVR) is beyond the scope of this summary, but the resulting functional fit $p : \mathbb{R}^d \rightarrow \mathbb{R}$ has the form
$$ p(x)  = \sum_{i=1}^{n}a_i K\bigl(x,x^{(i)}\bigr) + b ,$$
where $K$ is the selected kernel function, $a_i \in \mathbb{R}^n$, $b \in \mathbb{R}$ are coefficients to be solved for simultaneously. The computational complexity of the SVR is $\mathcal{O}(n^2dm)$, with $m$ being determined by the minimization convergence criterion. This paper uses the scikit-learn SVR \cite{scikit-learn} with a polynomial kernel function.

\subsection{Multilayer Perceptron Regressor}
\label{sec:mlp}
The neural network is a well studied and widely used method for both regression and classification tasks \cite{rumelhart1988learning,hornik1989multilayer}. When using the rectified linear unit (ReLU) activation function \cite{dahl2013improving} and fitting the model with stochastic gradient descent (SGD) or BFGS minimization techniques \cite{goh2017why,moller1993scaled,robbins1951stochastic}, the model built by a multilayer perceptron uses layers $l : \mathbb{R}^{i} \rightarrow \mathbb{R}^{j}$ defined by
$$ l(u) = \bigl( u^t W_l \bigr)_+ ,$$
where $W_l$ is the $i$ by $j$ weight matrix for layer $l$. In this form, the multilayer perceptron (MLP) produces a piecewise linear model of the input data. The computational complexity of training a multilayer perceptron is $\mathcal{O}(n d m)$, where $m$ is determined by the sizes of the layers of the network and the stopping criterion of the minimization used for finding weights. This paper uses a MLP built from Keras and Tensorflow to perform regression \cite{chollet2015keras,tensorflow2015-whitepaper} with ten hidden layers each having thirty two nodes (a total of approximately ten thousand parameters), ReLU activation, and five thousand epochs of SGD for training.

%% ===================================================================
\section{Multivariate Interpolation}
\label{sec:interpolation}
The following interpolation techniques demonstrate a reasonable variety of approaches to interpolation. All of the presented interpolants produce approximations that are continuous in value, which is often a desirable property in applied approximation problems.

\subsection{Delaunay}
\label{sec:delaunay}

The Delaunay triangulation is a well-studied geometric technique for producing an interpolant \cite{lee1980two}. The Delaunay triangulation of a set of data points into simplices is such that there are no data points inside the sphere defined by the vertices of each simplex. For a $d$-simplex S with vertices $x^{(0)}$, $x^{(1)}$, $\ldots$, $x^{(d)}$, and function values $f\bigl(x^{(i)}\bigr)$, $i=0$, $\ldots$, $d$, $y \in S$ is a unique convex combination of the vertices,
$$ y = \sum_{i=0}^{d} w_i x^{(i)}, \quad \sum_{i=0}^{d} w_i = 1, \quad w_i \geq 0, \quad i=0,\ldots,d, $$
and the Delaunay interpolant $\hat f(y)$ at $y$ is given by
$$ \hat f(y) = \sum_{i=0}^{d} w_i f\bigl(x^{(i)}\bigr). $$
The computational complexity of Delaunay interpolation (for the implementation used) is $\mathcal{O}(n d^4 \log d)$, which is capable of scaling reasonably to $d \leq 50$. In the present application, a Delaunay simplex $S$ containing $y$ is found, then the $d+1$ vertices (points in $X$) of $S$ are used to assign weights to each vertex and produce the predicted function value for point $y$.

\newpage
\subsection{Modified Shepard}
\label{sec:modified-shepard}

The modified Shepard method used here (ShepMod) generates a continuous approximation based on Euclidean distance and resembles a nearest neighbor interpolant \cite{cover1967nearest}. This model is a type of \textit{metric interpolation}, also called a Shepard method \cite{shepard1968two,gordon1978shepard}. The interpolant has the form
$$ p(x) = \frac{\sum\limits_{k=1}^{n}W_k(x)f\bigl(x^{(k)}\bigr)}{\sum\limits_{k=1}^{n}W_k(x)} ,$$
where $W_k(x) \propto \bigl\|x - x^{(k)}\bigr\|_2^{-2}$ and decreases to zero at a radius $r_k \in \mathbb{R}$ such that exactly $d+1$ other points $x^{(j)}$, $j \not = k$ are inside the closed Euclidean ball of radius $r_k$ about $x^{(k)}$. The computational complexity of ShepMod is $\mathcal{O}(n^2d)$. This paper uses a Fortran 95 implementation of ShepMod derived from SHEPPACK \cite{thacker2010algorithm}.

\subsection{Linear Shepard}
The linear Shepard method (LSHEP) is a blending function using local linear interpolants, a special case of the general Shepard algorithm \cite{thacker2010algorithm}. The interpolant has the form
$$ p(x) = \frac{\sum\limits_{k=1}^{n}W_k(x)P_k(x)}{\sum\limits_{k=1}^{n}W_k(x)} ,$$
where $W_k(x)$ is a locally supported weighting function and $P_k(x)$ is a local linear approximation to the data satisfying $P_k\bigl(x^{(k)}\bigr) = f\bigl(x^{(x)}\bigr)$. The computational complexity of LSHEP is $\mathcal{O}(n^2d^3)$. This paper uses the Fortran 95 implementation of LSHEP in SHEPPACK \cite{thacker2010algorithm}.

\subsection{Box Splines}
\label{sec:box-splines}

The box spline model used here is an interpolation technique built from overlapping box splines \cite{de2013box}. The box splines serve as basis functions that are shifted and scaled to have support over box shaped regions. The boxes are constructed in a way to guarantee a covering of the domain \cite{lux2018novel}. Given a set of box splines $\bigl\{b^{x^{(i)}}\bigr\}$ with the iterative box properties outlined in \cite{lux2018novel} and anchored at interior points $\bigl\{x^{(i)}\bigr\}$,
$$ \hat f(y) = \frac{\sum\limits_{i=1}^n b^{x^{(i)}}(y) f\bigl(x^{(i)}\bigr)}{\sum\limits_{i=1}^n b^{x^{(i)}}(y)}. $$
Note that the box splines always satisfy $b^{x^{(i)}}\bigl(x^{(j)}\bigr) = \delta_{ij}$ and $b^{x^{(i)}}(y) \geq 0$. The computational complexity of interpolation via the box spline model is $\mathcal{O}(n^2 d)$.

\subsection{Voronoi}
\label{sec:voronoi}

A well-studied technique for classification and approximation is the nearest neighbor algorithm \cite{cover1967nearest}. Nearest neighbor inherently utilizes the convex region $C^{x^{(i)}}$ (Voronoi cell \cite{dirichlet1850reduction}) consisting of all points closer to $x^{(i)}$ than to any other point $x^{(j)}$. The Voronoi model smooths the nearest neighbor approximation by utilizing the Voronoi cells to define support via a generic basis function $v: \mathbb{R}^d \rightarrow \mathbb{R}_+$ given by
$$ v^{x^{(i)}}(y) = \left(1 - \frac{\bigl\|y - x^{(i)}\bigr\|_2}{2 \ h\bigl(y - x^{(i)} \mid x^{(i)}\bigr)} \right)_+, $$
where $h\bigl(w \mid x^{(i)}\bigr)$ is the distance between $x^{(i)}$ and the boundary of the Voronoi cell $C^{x^{(i)}}$ in the direction $w$. $v^{x^{(i)}}\bigl(x^{(j)}\bigr) = \delta_{ij}$ and $v^{x^{(i)}}$ has local support, giving the interpolated value
$$ f(y) = \frac{\sum\limits_{i=1}^n v^{x^{(i)}}(y) f\bigl(x^{(i)}\bigr)}{\sum\limits_{i=1}^n v^{x^{(i)}}(y)}, $$
where $0 \leq v^{x^{(i)}}(y) \leq 1$. The computational complexity of interpolation via this Voronoi model is $\mathcal{O}(n^2 d)$. All of the approximations are an interpolant involving a convex combination of known function values $f\bigl(x^{(i)}\bigr)$.

%% ===================================================================
\section{Measuring Error}
\label{sec:error}

\begin{figure}[htb]
  \centering
  \begin{tikzpicture}
    \node (img)  {\includegraphics[width=0.75\textwidth,]{Figures/plot-delaunay-example-prediction.pdf}};
    \node[below=of img, node distance=1cm, yshift=1cm] {I/O Throughput};
    \node[left=of img, node distance=0cm, rotate=90, anchor=center,yshift=-0.7cm] {CDF Value};
  \end{tikzpicture}
  \vspace{-0.3cm}
  \caption{In this HPC I/O example, the general methodology for predicting a CDF and evaluating error can be seen. The Delaunay method chose three source distributions (dotted lines) and assigned weights \{.3, .4, .3\} (top to bottom at arrow). The weighted sum of the three known CDFs produces the predicted CDF (dashed line). The KS Statistic (arrow) computed between the true CDF (solid line) and predicted CDF (dashed line) is 0.2 for this example. The KS test null hypothesis is rejected at $p$-value 0.01, however it is not rejected at $p$-value 0.001.
  \vspace{-.1cm}}
  \label{fig:prediction-example}
\end{figure}

When the range of an approximation is the real numbers, error is reported with summary statistics including: min absolute error, max absolute error, and absolute error quartiles. When the range of an approximation is the space of cumulative distribution functions, the Kolmogorov-Smirnov statistic (max-norm difference between the functions) is used.

A hurdle when modeling function-valued outputs such as cumulative distribution functions (CDFs) or probability density functions (PDFs) is that certain properties must be maintained. It is necessary that a PDF $f: \mathbb{R} \rightarrow \mathbb{R}$ have the properties $f(x) \geq 0$ and $\int_{-\infty}^{\infty}f(x)dx = 1$. However, for a CDF $F: \mathbb{R} \rightarrow \mathbb{R}$ the properties are $F(x) \in [0,1]$ and $F(x)$ is absolutely continuous and nondecreasing. This work utilizes the fact that a convex combination of CDFs (or PDFs) results in a valid CDF. Given $G(x) = \sum_{i}w_i F_i(x)$, $\sum_{i} w_i = 1$, $w_i \geq 0$, and each $F_i$ is a valid CDF, $G$ must also be a valid CDF. A demonstration of how this is applied can be seen in Figure \ref{fig:prediction-example}.

The performance of approximation techniques that predict probability functions can be analyzed through a variety of summary statistics. This work uses the max absolute difference, also known as the Kolmogorov-Smirnov (KS) statistic \cite{lilliefors1967kolmogorov} for its compatibility with the KS test.

The two-sample KS test is a useful nonparametric test for comparing two CDFs while only assuming stationarity, finite mean, and finite variance. The null hypothesis (that two CDFs come from the same underlying distribution) is rejected at level $p \in [0,1]$ when
$$ KS > \sqrt{-\frac{1}{2}\ln\biggl(\frac{p}{2}\biggr)} \sqrt{\frac{1}{n_1} + \frac{1}{n_2}}, $$
with distribution sample sizes $n_1,n_2 \in \mathcal{N}$. For all applications of the KS test presented in this work $n_1 = n_2$. An example of the round-trip prediction methodology from known and predicted distributions to the calculation of error can be seen in Figure \ref{fig:prediction-example}.

\section{Theoretical Error Bound}
\label{sec:theory}

This section presents the theoretical results bounding the error of Delaunay interpolation. The error analysis relies on the Delaunay interpolation for two reasons: (1) second order results can be obtained utilizing a Lipschitz constant on the gradient of a function, rather than standard Lipschitz bounds and (2) multiple other interpolants analyzed compute predictions as convex combinations of observed function values, which may allow for straight forward extensions of this error bound.

\begin{plainlemma}
  \label{lemma:1}
  Let $S \subset \mathbb{R}^d$ be open and convex, $f: \mathbb{R}^d \rightarrow \mathbb{R}$, and $\nabla f \in Lip_{(\gamma,\|\cdot\|_2)}(S)$, the set of $\gamma$-Lipschitz continuous functions in the $2$-norm. Then for all $x,y \in S$
  $$\big|f(y) - f(x) - \langle \nabla f(x), y - x \rangle \big| \leq \frac{\gamma \|y - x\|_2^2}{2}.$$
\end{plainlemma}

\begin{proofdot}
  Consider the function $g(t) = f \big((1-t) x + t y \big)$, $0 \leq t \leq 1$, whose derivative $g'(t) = \big\langle \nabla f \big((1-t) x + t y \big), y - x \big\rangle$ is the directional derivative of $f$ in the direction $(y - x).$
  \begin{align*}
    \big|f(y) - f(x) - \langle \nabla f(x), y - x \rangle \big|
       &= \big|g(1) - g(0) - g'(0) \big| & \\
       &= \bigg| \int_0^1 g'(t) - g'(0)\ dt \bigg| \\
       &\leq \int_0^1 \big|g'(t) - g'(0)\big|\ dt \\
       &= \int_0^1 \bigg| \big \langle \nabla f\big((1-t)x + ty\big) - \nabla f(x), y - x \big \rangle \bigg|\ dt \\
       &\leq \int_0^1 \big \| \nabla f\big((1-t)x + ty\big) - \nabla f(x) \big \|_2\ \| y - x \|_2\ dt \\
       &\leq \int_0^1 \big ( \gamma\ \|y-x\|_2 \big) \ \big( \|y-x\|_2 \big) t\ dt \\
       &= \frac{\gamma \|y - x\|_2^2}{2}.
  \end{align*}
  \qed
\end{proofdot}

\begin{plainlemma}
  \label{lemma:2}
  Let $x, y, v_i \in \mathbb{R}^d$, $c_i \in \mathbb{R}$, and $|\langle y - x, v_i \rangle| \leq c_i$ for $i = 1$, $\ldots$, $d.$ If $M = (v_1$, $\ldots$, $v_d)$ is nonsingular, then
  $$\|y - x\|_2^2 \leq \frac{1}{\sigma_d^2} \sum_{i=1}^d c_i^2,$$
  where $\sigma_d$ is the smallest singular value of $M.$
\end{plainlemma}

\begin{proofdot}
  Using the facts that $M$ and $M^t$ have the same singular values, and $\|M^tw\|_2 \geq \sigma_d \|w\|_2$, gives
  \begin{align*}
    \|y - x\|_2^2 &\leq \frac{\|M^t (y - x)\|_2^2}{\sigma_d^2} \\
                  &=    \frac{1}{\sigma_d^2} \sum_{i=1}^d \langle y - x, v_i \rangle^2 \\
                  &\leq \frac{1}{\sigma_d^2} \sum_{i=1}^d c_i^2.
  \end{align*}
  \qed
\end{proofdot}

\begin{plainlemma}
  \label{lemma:3}
  Given $f$, $\gamma$, $S$ as in {\it Lemma \ref{lemma:1}}, let $X = \{x_0$, $x_1$, $\ldots$, $x_d\}$ $\subset S$ be the vertices of a $d$-simplex, and let $\hat f(x) = \langle c, x - x_0 \rangle + f(x_0)$, $c \in \mathbb{R}^d$ be the linear function interpolating $f$ on $X.$ Let $\sigma_d$ be the smallest singular value of the matrix $M = (x_1 - x_0$, $\ldots$, $x_d - x_0)$, and $k = \max\limits_{1\ \leq\ j\ \leq\ d} \|x_j - x_0\|_2.$ Then
  $$\big\|\nabla f(x_0) - c\big\|_2 \leq \sqrt{d} \frac{\gamma k^2}{\sigma_d}.$$
\end{plainlemma}

\begin{proofdot}
  Consider $f(x) - \hat f(x)$ along the line segment $z(t) = (1-t)x_0 + t x_j$, $0 \leq t \leq 1.$ By Rolle's Theorem, for some $0 < \hat t < 1$, $\big\langle \nabla f\big(z(\ \hat t\ )\big) - c, x_j - x_0 \big\rangle = 0.$ Now
  \begin{align*}
    \big| \big< \nabla f(x_0) - c, x_j - x_0 \big \rangle \big|
        &= \big| \big< \nabla f(x_0) - \nabla f\big(z(\ \hat t\ )\big) + \nabla f\big(z(\ \hat t\ )\big) - c, x_j - x_0 \big \rangle \big| \\
	&= \big| \big\langle \nabla f(x_0) - \nabla f \big(z(\ \hat t\ )\big), x_j - x_0 \big \rangle \big| \\
	&\leq \big \| \nabla f(x_0) - \nabla f\big(z(\ \hat t\ )\big) \big \|_2 \|x_j - x_0\|_2 \\
	&\leq \gamma \|x_0 - z(\ \hat t\ )\|_2\ \|x_j - x_0\|_2 \\
	&\leq \gamma \|x_j - x_0\|_2^2 \leq \gamma k^2, \\
  \end{align*}
  for all $1 \leq j \leq d.$ Using {\it Lemma \ref{lemma:2}},
  $$\big \| \nabla f(x_i) - c \big \|_2^2 \leq \frac{d}{\sigma_d^2} \big( \gamma k^2\big)^2 \implies \big \| \nabla f(x_i) - c \big \|_2 \leq \sqrt{d} \frac{\gamma k^2}{\sigma_d}.$$
  \qed
\end{proofdot}

\begin{onetheorem}
  Under the assumptions of {\it Lemma \ref{lemma:1}} and {\it Lemma \ref{lemma:3}}, for $z \in S$,
  $$ \big|f(z) - \hat f(z)\big| \leq \frac{\gamma \|z - x_0\|_2^2}{2} + \sqrt{d} \frac{\gamma k^2}{\sigma_d} \|z - x_0\|_2.$$
\end{onetheorem}

\begin{proofdot}
  Let $v = \nabla f(x_0) - c$, where $\|v\|_2 \leq \sqrt{d} \gamma k^2 / \sigma_d$ by {\it Lemma \ref{lemma:3}}. Now
  \begin{align*}
    \big|f(z) - \hat f(z)\big|
       &= \big|f(z) - f(x_0) - \langle c, z - x_0 \rangle \big| \\ 
       &= \big|f(z) - f(x_0) - \langle \nabla f(x_0) - v, z - x_0 \rangle \big| \\
       &= \big|f(z) - f(x_0) - \langle \nabla f(x_0) , z - x_0 \rangle + \langle v , z - x_0 \rangle \big| \\
       &\leq \big|f(z) - f(x_0) - \langle \nabla f(x_0) , z - x_0 \rangle \big| + \big| \langle v , z - x_0 \rangle \big| \\
       &\leq \big|f(z) - f(x_0) - \langle \nabla f(x_0) , z - x_0 \rangle \big| + \|v\|_2 \|z - x_0\|_2 \\
       &\leq \big|f(z) - f(x_0) - \langle \nabla f(x_0), z - x_0 \rangle \big| + \textstyle{\frac{\gamma k^2 \sqrt{d}}{\sigma_d}} \|z - x_0\|_2 \\
       &\leq \frac{\gamma \|z - x_0\|_2^2}{2} + \sqrt{d}\frac{\gamma k^2}{\sigma_d} \|z - x_0\|_2,
  \end{align*}
  where the last inequality follows from {\it Lemma \ref{lemma:1}}.
  %
  \qed
\end{proofdot}

In summary, the approximation error of a Delaunay interpolant tends quadratically towards zero when approaching observed data only when the diameter of the interpolating simplex is also reduced at a proportional rate. Without the incorporation of additional observations, only linear convergence to the true function is be achieved. Notice that the approximation error is largely determined by the spacing of observed data. Predictions made by simplices whose vertices are not well-spaced (i.e. have large diameter, or are nearly contained in a hyperplane) have higher error. In light of this error bound, an empirical evaluation of presented algorithms follows.

%% ===================================================================
\section{Data and Empirical Analysis}
\label{sec:data}

The theoretical results constructed in Section \ref{sec:theory} for Delaunay interpolation are promising, however they are difficult to interpret in context with the other approximation algorithms that do not have similar known uniform error bounds. For that reason, this paper utilizes five different data sets of varying dimension and application to construct approximations and compare the accuracy of different techniques.

In the following five subsections the sources and targets of each test data set are described, as well as challenges and limitations related to approximating these data. The distribution of response values being modeled is presented followed by the distribution of approximation errors for each algorithm.

Before approximations are constructed, all five data sets are rescaled such that the domain of approximation is the unit hypercube. The range of the first four data sets is the real numbers, while the range of the fifth data set is the space of cumulative distribution functions. All approximation techniques are applied to the first four data sets, while only those interpolants whose approximations are convex combinations of observed data are applied to the final data set.

All approximations are constructed using $10$-fold cross validation as described in \cite{kohavi1995study}. This approach requires the division of data randomly into $10$ partitions (the same partitions are used by each algorithm). An algorithm is then evaluated by constructing an approximation over each set of $9$ partitions, predicting the values of the final partition. As a result, each observed data point is used in the construction of $9$ different approximations and is approximated exactly once. The $k$-fold cross validation method is data-efficient and provides an unbiased estimate of the expected prediction error \cite{kohavi1995study}, however it should be noted that neither this method nor others can provide a universally unbiased estimator for the variance of prediction error \cite{bengio2004no}.

In addition to the figures displaying approximation results for each data set, tables of accompanying numerical results are located in the Appendix. All of the test data sets capture underlying functions that are almost certainly stochastic. As described in Section \ref{sec:introduction}, regression techniques appear most appropriate for these problems. However, in high dimension data grows exponentially more sparse. Given sparse data, regressors tend towards interpolation.

%% -------------------------------------------------------------------
\subsection{Forest Fire ($n = 504, d = 12$)}

\begin{figure}
  \centering
  \begin{tikzpicture}
    \node (img)  {\includegraphics[width=0.75\textwidth,]{Figures/raw-histogram-forest-fire.pdf}};
    \node[below=of img, node distance=1cm, yshift=1cm] {Forest First Area Burned};
    \node[left=of img, node distance=0cm, rotate=90, anchor=center,yshift=-0.7cm] {Count};
  \end{tikzpicture}
  \caption{Above is a histogram of forest fire area burned under recorded weather conditions. The data is presented on a natural log scale because most values are small with exponentially fewer fires on record that burn large areas.}
  \label{fig:hist-forest-fire}
\end{figure}

\begin{figure}
  \centering
  \begin{tikzpicture}
    \node (img)  {\includegraphics[width=0.8\textwidth,]{Figures/error-box-forest-fire.pdf}};
    \node[below=of img, node distance=1cm, yshift=1cm] {Area Burned Error};
  \end{tikzpicture}
  \caption{All models are applied to approximate the amount of area that would be burned given environment conditions. The data is broken up into $10$ partitions and each set of $9$ partitions is used to construct a model that approximates the remaining partition. This results in exactly one prediction from each algorithm for each data point. These boxes depict the median (middle bar), median $95\%$ confidence interval (cones), quartiles (box edges), fences at $3/2$ interquartile range (whiskers), and outliers (dots) of absolute prediction error for each model. Similar to figure \ref{fig:hist-forest-fire}, the errors are presented on a natural log scale. The complementing numerical data to this figure is provided in table \ref{table:error-forest-fire} in the Appendix.}
  \label{fig:error-forest-fire}
\end{figure}

The forest fire data set \cite{cortez2007data} describes the area of Montesinho park burned over months of the year along with environmental conditions. The twelve dimensions being used to model burn area are the $x$ and $y$ spatial coordinates of burns in the park, month of year (mapped to a 2D circle), the FFMC, DMC, DC, and ISI indices (see source for details), the temperature, relative humidity, wind speed, and outdoor rain. The original analysis of this data set demonstrated it to be difficult to model, likely due to the skew in response values.

As suggested by figure \ref{fig:error-forest-fire}, the SVR has the lowest absolute prediction errors for 80\% of the data, with ShepMod and BoxSpline being the nearest overall competitors. The effectiveness of SVR on this data suggests the underlying function can be defined by relatively few parameters, as well as the importance of capturing the low-burn-area data points.
%% -------------------------------------------------------------------

%% -------------------------------------------------------------------
\subsection{Parkinson's Telemonitoring ($n = 5875, d = 19$)}

\begin{figure}
  \centering
  \begin{tikzpicture}
    \node (img)  {\includegraphics[width=0.75\textwidth,]{Figures/raw-histogram-parkinsons.pdf}};
    \node[below=of img, node distance=1cm, yshift=1cm] {Parkinson's Total UPDRS Score};
    \node[left=of img, node distance=0cm, rotate=90, anchor=center,yshift=-0.7cm] {Count};
  \end{tikzpicture}
  \caption{Above is a histogram of the Parkinson's patient total UPDRS clinical scores that will be approximated by each algorithm.}
  \label{fig:hist-parkinsons}
\end{figure}

\begin{figure}
  \centering
  \begin{tikzpicture}
    \node (img)  {\includegraphics[width=0.8\textwidth,]{Figures/error-box-parkinsons.pdf}};
    \node[below=of img, node distance=1cm, yshift=1cm] {Total UPDRS Score Error};
  \end{tikzpicture}
  \caption{All models are applied to approximate the total UPDRS score given audio features from patients' home life. The data is broken up into $10$ partitions and each set of $9$ partitions is used to construct a model that approximates the remaining partition. This results in exactly one prediction from each algorithm for each data point. These boxes depict the median (middle bar), median $95\%$ confidence interval (cones), quartiles (box edges), fences at $3/2$ interquartile range (whiskers), and outliers (dots) of absolute prediction error for each model. The complementing numerical data to this is provided in table \ref{table:error-parkinsons} in the Appendix.}
  \label{fig:error-parkinsons}
\end{figure}

The second data set for evaluation \cite{tsanas2010accurate} is derived from a speech monitoring study with the intent to automatically estimate Parkinson's disease symptom development in Parkinson's patients. The function to be approximated is a time-consuming clinical evaluation measure referred to as the UPDRS score. The total UPDRS score given by a clinical evaluation is estimated through 16 real numbers generated from biomedical voice measures of in-home sound recordings.

Figure \ref{fig:error-parkinsons} shows the ShepMod algorithm has the lowest minimum, first quartile, and median of absolute errors for this problem, while providing the best prediction 66\% of the time. The MLP has the lowest third quartile and provides the best prediction for 32\% of approximations. The dominance of ShepMod may be in part to regular-interval total UPDRS scores provided by clinicians, favoring a nearest-neighbor strategy of prediction.
%% -------------------------------------------------------------------

%% -------------------------------------------------------------------
\subsection{Australian Daily Rainfall Volume ($n = 2609, d = 23$)}

\begin{figure}
  \centering
  \begin{tikzpicture}
    \node (img)  {\includegraphics[width=0.75\textwidth,]{Figures/raw-histogram-weather.pdf}};
    \node[below=of img, node distance=1cm, yshift=1cm] {Daily Rainfall in Sydney};
    \node[left=of img, node distance=0cm, rotate=90, anchor=center,yshift=-0.7cm] {Count};
  \end{tikzpicture}
  \caption{Above is a histogram of daily rainfall in Sydney, Australia. The data is presented on a natural log scale because the frequency of larger amounts of rainfall is significantly less. There is a peak in occurrence of the value $0$, which has a notable effect on the resulting model performance.}
  \label{fig:hist-weather}
\end{figure}

\begin{figure}
  \centering
  \begin{tikzpicture}
    \node (img)  {\includegraphics[width=0.8\textwidth,]{Figures/error-box-weather.pdf}};
    \node[below=of img, node distance=1cm, yshift=1cm] {Sydney Rainfall Tomorrow Error};
  \end{tikzpicture}
  \caption{All models are applied to approximate the amount of rainfall expected on the next calendar day given various sources of local meteorological data. The data is broken up into $10$ partitions and each set of $9$ partitions is used to construct models that approximates the remaining partition. This results in exactly one prediction from each algorithm for each data point. These boxes depict the median (middle bar), median $95\%$ confidence interval (cones), quartiles (box edges), fences at $3/2$ interquartile range (whiskers), and outliers (dots) of absolute prediction error for each model. The errors are presented on a natural log scale, mimicking the presentation in figure \ref{fig:hist-weather}. The complementing numerical data to this figure is provided in table \ref{table:error-weather} in the Appendix.}
  \label{fig:error-weather}
\end{figure}

The third data regarding the total daily rainfall in Sydney, Australia \cite{williams2009rattle} provides a slightly higher dimension challenge for the interpolants and regressors. This data is composed of many meteorological readings from the region on a day including: min and max temperatures, sunshine, wind speed directions (converted to coordinates on circle), wind speeds, and humidities throughout the day, day of the year (converted to coordinates on circle), and the model must predict the amount of rainfall tomorrow.

While figure \ref{fig:hist-weather} makes MARS look far better than other techniques, it only provides the best prediction for 11\% of points. The MLP has the lowest absolute error for 56\% of points and LSHEP is best for 28\%. MARS likely achieves such a low first quartile due to the high occurrence of the value 0 in the data.
%% -------------------------------------------------------------------

%% -------------------------------------------------------------------
\subsection{Credit Card Transaction Amount ($n = 5562, d = 28$)}

\begin{figure}
  \centering
  \begin{tikzpicture}
    \node (img)  {\includegraphics[width=0.75\textwidth,]{Figures/raw-histogram-credit-card.pdf}};
    \node[below=of img, node distance=1cm, yshift=1cm] {Credit Card Transaction Amount};
    \node[left=of img, node distance=0cm, rotate=90, anchor=center,yshift=-0.7cm] {Count};
  \end{tikzpicture}
  \caption{Above is a histogram of credit card transaction amounts. The data is presented on a natural log scale. The data contains a notable frequency peak around $\$1$ transactions. Fewer large purchases are made, but some large purchases are in excess of five orders of magnitude greater than the smallest purchases.}
  \label{fig:hist-credit-card}
\end{figure}

\begin{figure}
  \centering
  \begin{tikzpicture}
    \node (img)  {\includegraphics[width=0.8\textwidth,]{Figures/error-box-credit-card.pdf}};
    \node[below=of img, node distance=1cm, yshift=1cm] {Transaction Amount Error};
  \end{tikzpicture}
  \caption{All models are applied to approximate the expected transaction amount given transformed (and obfuscated) vendor and customer-descriptive features. The data is broken up into $10$ partitions and each set of $9$ partitions is used to construct a model that approximates the remaining partition. This results in exactly one prediction from each algorithm for each data point. These boxes depict the median (middle bar), median $95\%$ confidence interval (cones), quartiles (box edges), fences at $3/2$ interquartile range (whiskers), and outliers (dots) of absolute prediction error for each model. The absolute errors in transaction amount predictions are presented on a natural log scale, just as in figure \ref{fig:hist-credit-card}. The complementing numerical data to this figure is provided in table \ref{table:error-credit-card} in the Appendix.}
  \label{fig:error-credit-card}
\end{figure}

The fourth test data set, and the final with a real-valued range, is a collection of credit card transactions \cite{pozzolo2015calibrating}. The provided data carries no direct real-world meaning, it is the output of a principle component analysis on the original hidden source data. This obfuscation is done to protect the information of the credit card users. This data has the largest dimension of all considered, at 28. A model over this data predicts the transaction amount given the vector of principle component information.

As can be seen in figure \ref{fig:error-credit-card}, the MLP outperforms all other algorithms at the first, second, third, and fourth quartiles. The MLP produces the lowest absolute error prediction for 80\% of transactions, Delaunay bests the remaining 20\%. It is likely that with less data, Delaunay would be the best performer.
%% -------------------------------------------------------------------

%% -------------------------------------------------------------------
\subsection{High Performance Computing I/O ($n = 3016, d = 4$)}

\begin{figure}
  \centering
  \begin{tikzpicture}
    \node (img)  {\includegraphics[width=0.75\textwidth,]{Figures/raw-histogram-throughput.pdf}};
    \node[below=of img, node distance=1cm, yshift=1cm] {I/O Read Throughtput};
    \node[left=of img, node distance=0cm, rotate=90, anchor=center,yshift=-0.7cm] {Count};
  \end{tikzpicture}
  \caption{Histogram of the raw throughput values recorded during all IOzone tests across all system configurations. The distribution is skewed right, with few tests having significantly higher throughput than most others. The data is presented on a natural log scale, however the resulting left-skew reveals the data is not precisely distributed according to an exponential.}
  \label{fig:hist-throughput}
\end{figure}

\begin{figure}
  \centering
  \begin{tikzpicture}
    \node (img)  {\includegraphics[width=0.8\textwidth,]{Figures/error-box-throughput.pdf}};
    \node[below=of img, node distance=1cm, yshift=1cm] {Read Throughtput KS Statistic};
  \end{tikzpicture}
  \caption{Above, the models directly capable of predicting distributions are applied to predicting the expected CDF of I/O throughput at a previously unseen system configuration. The KS statistic between the observed distribution at each system configuration and the predicted distribution is recorded and presented above. Note that the above plot is \textit{not} log-scaled like the histogram. The complimentary numerical data for this figure is provided in Appendix table \ref{table:error-throughput}.}
  \label{fig:error-throughput}
\end{figure}

The final of five data sets is four-dimensional and is produced by executing the IOzone benchmark from \cite{iozone} on a homogeneous cluster of computers. The system performance data was collected by executing IOzone 150 times for each of a select set of system configurations. A single IOzone execution reports the max I/O file-read throughput seen. The 150 executions for each system configuration are converted
to their empirical distribution function point, which are fit with a piecewise cubic Hermite interpolating polynomial \cite{fritsch1980monotone} to approximate the CDF. These CDFs are capable of being approximated by all interpolants except LSHEP. The four dimensions being modeled to predict throughput mean are file size, record size, thread count, and CPU frequency.

Delaunay achieves the lowest KS statistic for 62\% of approximations, while Voronoi is best for the remaining 38\%. Figure \ref{fig:error-throughput} shows that while Delaunay may have more best predictions, the behavior of Voronoi may be preferable.
%% -------------------------------------------------------------------

%% -------------------------------------------------------------------
\begin{figure}
  \centering
  \begin{tikzpicture}
    \node (img)  {\includegraphics[width=0.8\textwidth]{Figures/plot-throughput-ks.pdf}};
    \node[below=of img, node distance=1cm, yshift=1cm] {KS Statistic for Predicted vs. Actual};
    \node[left=of img, node distance=0cm, rotate=90, anchor=center,yshift=-0.7cm] {Count of KS Statistic};
  \end{tikzpicture}
  \caption{Histograms of the prediction error for each interpolant that produces predictions as convex combinations of observed data. The data is broken up into $10$ partitions and each set of $9$ partitions is used to construct a model that approximates the remaining partition. This results in exactly one prediction from each algorithm for each data point. The distributions show the KS statistics for the predicted throughput distribution versus the actual throughput distribution. The four vertical red lines represent commonly used $p$-values \{0.05, 0.01, 0.001, 1.0e-6\} respectively. All predictions to the right of a red line represent CDF predictions that are significantly different (by respective $p$-value) from the actual distribution according to the KS test. The numerical counterpart to this figure is presented in table \ref{table:null-hypothesis-results}.}
  \label{fig:throughput-prediction}
\end{figure}

\begin{table}
  \renewcommand{\arraystretch}{1.3}
  \centering
  \begin{tabular}{c|c|c|c|c}
     & $p$ = .05 & $p$ = .01 & $p$ = .001 & $p$ = 1.0e-6\\
    \hline
    Delaunay & $\mathbf{50.3}\%$ & $\mathit{43.5}\%$ & $\mathit{36.2}\%$ & $\mathit{24.7}\%$\\
    ShepMod & $\mathit{51.4}\%$ & $44.8\%$ & $38.1\%$ & $27.7\%$\\
    Voronoi & $52.6\%$ & $\mathbf{43.4}\%$ & $\mathbf{34.4}\%$ & $\mathbf{19.1}\%$\\
    BoxMesh & $99.5\%$ & $98.5\%$ & $96.6\%$ & $89.5\%$\\
  \end{tabular}
  \caption{This is the numerical counterpart to the histogram data presented in figure \ref{fig:throughput-prediction}. The columns display the percent of null hypothesis rejections by the KS-test when provided different selections of $p$-values for each algorithm. The algorithm with the lowest rejection rate at each $p$ is boldface, while the second lowest is italicized.}
  \label{table:null-hypothesis-results}
\end{table}

%% ===================================================================

Figure \ref{fig:throughput-prediction} expands on the KS statistic results presented in figure \ref{fig:error-throughput}. Agglomerate errors for each algorithm resemble a Gamma distribution. The percentages of significant prediction errors with varying $p$-values are on display in Table \ref{table:null-hypothesis-results}. When considering the $p=0.001$ results for each technique, a little over one third of the predicted CDFs are significantly different from the measured (and presumed) correct CDFs. However, it should be noted that with 150 samples, the error of an empirical distribution function (EDF) can reasonably be upwards of $.1$, which serves as a rough estimate for the lower limit of achievable error. Globally, only a third of Voronoi predictions fail to capture \textit{all} of the characteristics of the CDFs at new system configurations.

\section{Discussion}
\label{sec:discussion}

\begin{table}
  \centering
  \begin{tabular}{c|c|c|c}
    \hline
    Algorithm & Avg. \% Best & Avg. Fit Time & Avg. App. Time (1)\\
    \hline
    MARS & $4.6\%$ & $22.3$s & $1.37 \times 10^{-3}$s\\
    SVR & $\mathit{21.1}\%$ & $0.503$s & $\mathit{1.37 \times 10^{-4}}$s\\
    MLP & $\mathbf{42.1}\%$ & $210.0$s & $1.02 \times 10^{-3}$s\\
    Delaunay & $6.0\%$ & $\mathbf{1.43 \times 10^{-4}}$s & $3.03$s\\
    ShepMod & $18.4\%$ & $0.711$s & $1.49 \times 10^{-4}$s\\
    LSHEP & $8.4\%$ & $1.69$s & $\mathbf{1.16 \times 10^{-4}}$s\\
    Voronoi & $0.5\%$ & $\mathit{0.127}$s & $0.0375$s\\
    BoxSplines & $0.2\%$ & $0.816$s & $5.26 \times 10^{-4}$s\\
    \hline
  \end{tabular}
  \caption{This data is the average of Appendix tables \ref{table:error-forest-fire}, \ref{table:error-parkinsons}, \ref{table:error-weather}, \ref{table:error-credit-card}, and provides a gross summary of overall results. The columns display (weighted equally by data set, \textit{not} points) the average frequency with which any algorithm provided the lowest absolute error approximation, the average time-to-fit, and the average time required to approximated one point. Interpolants provide the lowest error approximation one third of the time across all data, while regressors occupy the other two thirds. This result is obtained without any customized tuning or preprocessing to maximize the performance of any given algorithm. In practice, tuning and preprocessing may have large effects on approximation performance.}
  \label{table:avg-performance}
\end{table}

Table \ref{table:avg-performance} summarizes results across the four test data sets with real-valued ranges. The interpolants discussed in this paper produce the \textit{best} approximations roughly one third of the time, and produce competitive approximations for almost all data sets. These test problems are almost certainly \textit{stochastic} in nature, but the high dimension yields to greater sparsity and makes interpolation a competitive approximation technique.

The major advantages to interpolation lie in the near absence of \textit{fit} time. Delaunay has no preparations required, LSHEP and ShepMod require pairwise distance calculations to determine the radii of influence for points in each model. At least hundreds, and sometimes hundreds of thousands of predictions can be made by the interpolants before the most widely used regressor (the MLP) finishes fitting these relatively small data sets. However, the computational complexities of all interpolants presented are greater than linear in either dimension or number of points, meaning large enough data (in dimension and number of observations) would cause interpolation to be slower.

The theoretical results presented in Section \ref{sec:theory} directly apply to Delaunay interpolation, however the performance of Delaunay does not appear significantly better than other algorithms on these data sets. This observation may be due to the stochastic nature of the data, but it also speaks to the alignment of the geometry of the underlying functions with the approximations generated by the different interpolation methods. The strong performance of other interpolants suggests that theoretical results similar to those presented in this work can be achieved for the other interpolants under reasonable assumptions.

Finally, most of the interpolants presented in this work benefit from the ability to model any function over a real vector space with a range that is closed under convex combinations. The results of the distribution prediction case study indicate that interpolants can effectively predict CDFs. The error analysis for that work relies on the KS statistic, which captures the worst part of any prediction and hence provides a conservatively large estimate of approximation error. The average absolute errors in the predicted CDFs are always lower than the KS statistics. However, the KS statistic was chosen because of the important statistical theory surrounding it as an error measure. Considering this circumstance, a nonnegligible volume of predictions provide impressively low levels of error in that final case study.

\section{Conclusion}
\label{sec:conclusion}

This work presents new theoretical results for interpolation in arbitrary dimension with simplicial approximations. Coinciding with the theoretical results, an empirical evaluation across real-world problems demonstrates that interpolants produce competitively accurate models of multivariate phenomenon when compared with common regressors for sparse, moderately high dimensional problems.

The various interpolants discussed in this paper have been demonstrated to effectively approximate multivariate phenomena up to $30$ dimensions. The underlying constructions are theoretically straightforward, yet powerful and flexible. Most of the interpolants' computational complexities make them particularly suitable for applications in even higher dimension, where the major benefits are seen when only a small number of approximations are to be made from any given data ($\leq 1000$).

%\begin{acknowledgements}
%If you'd like to thank anyone, place your comments here
%and remove the percent signs.
%\end{acknowledgements}

%%====================================================================
%%====================================================================

\bibliographystyle{spmpsci}      % mathematics and physical sciences
\bibliography{paper}

\begin{appendix}

\section*{Appendix}

\begin{table}
  \centering
  \begin{tabular}{c|c|c|c|c|c}
    \hline
    Algorithm & Min & $25^{th}$ & $50^{th}$ & $75^{th}$ & Max\\
    \hline
    MARS & $0.00984$ & $3.11$ & $7.01$ & $\mathit{11.7}$ & $1090.0$\\
    SVR & $\mathit{0.00402}$ & $\mathbf{0.243}$ & $\mathbf{0.416}$ & $\mathbf{6.19}$ & $1090.0$\\
    MLP & $0.0426$ & $2.63$ & $\mathit{5.27}$ & $14.0$ & $1090.0$\\
    Delaunay & $\mathbf{0.00}$ & $1.98$ & $5.37$ & $13.1$ & $\mathit{1080.0}$\\
    ShepMod & $\mathbf{0.00}$ & $\mathit{1.93}$ & $6.27$ & $16.0$ & $1090.0$\\
    LSHEP & $0.0400$ & $2.17$ & $8.87$ & $19.1$ & $\mathbf{1070.0}$\\
    Voronoi & $0.00982$ & $3.65$ & $7.56$ & $15.6$ & $1090.0$\\
    BoxSpline & $\mathbf{0.00}$ & $2.27$ & $5.91$ & $16.4$ & $1090.0$\\
    \hline
  \end{tabular}
  \caption{This numerical data accompanies the visual provided in figure \ref{fig:error-forest-fire}. The columns of absolute error percentiles correspond to the minimum, first quartile, median, second quartile, and maximum absolute errors respectively. The minimum of each column is boldface, while the second lowest value is italicized. All values are rounded to three significant digits.}
  \label{table:error-forest-fire}
\end{table}

\begin{table}
  \centering
  \begin{tabular}{|c|c| c |c|c|c|}
    \cline{1-2}\cline{4-6}
    Algorithm & \% Best &  & Fit Time & App. Time (1) & Total App. Time\\
    \cline{1-2}\cline{4-6}
    MARS & $\mathit{7.7}\%$ &  & $29.1$s & $1.37 \times 10^{-3}$s & $0.0686$s\\
    SVR & $\mathbf{80.2}\%$ &  & $5.84 \times 10^{-3}$s & $\mathit{6.20 \times 10^{-5}}$s & $\mathit{3.10 \times 10^{-3}}$s\\
    MLP & $0.0\%$ &  & $32.8$s & $8.71 \times 10^{-4}$s & $0.0436$s\\
    Delaunay & $3.5\%$ &  & $\mathbf{2.49 \times 10^{-5}}$s & $0.0385$s & $1.93$s\\
    ShepMod & $3.7\%$ &  & $6.34 \times 10^{-3}$s & $6.44 \times 10^{-5}$s & $3.22 \times 10^{-3}$s\\
    LSHEP & $5.1\%$ &  & $0.0275$s & $\mathbf{4.63 \times 10^{-5}}$s & $\mathbf{2.31 \times 10^{-3}}$s\\
    Voronoi & $0.0\%$ &  & $\mathit{1.82 \times 10^{-4}}$s & $3.96 \times 10^{-4}$s & $0.0198$s\\
    BoxSpline & $0.4\%$ &  & $7.24 \times 10^{-3}$s & $9.78 \times 10^{-5}$s & $4.89 \times 10^{-3}$s\\
    \cline{1-2}\cline{4-6}
  \end{tabular}
  \caption{The left above shows how often each algorithm had the lowest absolute error approximating forest fire data. On the right columns are median fit time of 454 points, median time for one approximation, and median time approximating 50 points.}
  \label{table:best-forest-fire}
\end{table}

\begin{table}
  \centering
  \begin{tabular}{c|c|c|c|c|c}
    \hline
    Algorithm & Min & $25^{th}$ & $50^{th}$ & $75^{th}$ & Max\\
    \hline
    MARS & $0.00948$ & $9.98$ & $20.4$ & $32.9$ & $863.0$\\
    SVR & $0.00138$ & $3.17$ & $7.31$ & $12.6$ & $\mathbf{27.5}$\\
    MLP & $2.39 \times 10^{-5}$ & $\mathit{0.533}$ & $\mathit{1.25}$ & $\mathbf{2.84}$ & $39.3$\\
    Delaunay & $\mathit{3.72 \times 10^{-12}}$ & $1.20$ & $3.50$ & $7.67$ & $30.7$\\
    ShepMod & $\mathbf{0.00}$ & $\mathbf{0.255}$ & $\mathbf{0.908}$ & $\mathit{3.43}$ & $34.5$\\
    LSHEP & $0.00254$ & $2.93$ & $7.16$ & $13.1$ & $\mathit{29.0}$\\
    Voronoi & $\mathbf{0.00}$ & $1.29$ & $3.52$ & $6.87$ & $30.1$\\
    BoxSpline & $0.00287$ & $4.39$ & $9.10$ & $14.8$ & $41.3$\\
    \hline
  \end{tabular}
  \caption{This numerical data accompanies the visual provided in figure \ref{fig:error-parkinsons}. The columns of absolute error percentiles correspond to the minimum, first quartile, median, second quartile, and maximum absolute errors respectively. The minimum of each column is boldface, while the second lowest value is italicized. All values are rounded to three significant digits.}
  \label{table:error-parkinsons}
\end{table}

\begin{table}
  \centering
  \begin{tabular}{|c|c| c |c|c|c|}
    \cline{1-2}\cline{4-6}
    Algorithm & \% Best &  & Fit Time & App. Time (1) & Total App. Time\\
    \cline{1-2}\cline{4-6}
    MARS & $0.0\%$ &  & $37.9$s & $2.53 \times 10^{-3}$s & $1.48$s\\
    SVR & $0.1\%$ &  & $0.859$s & $\mathit{1.81 \times 10^{-4}}$s & $\mathit{0.106}$s\\
    MLP & $\mathit{32.0}\%$ &  & $348.0$s & $1.11 \times 10^{-3}$s & $0.653$s\\
    Delaunay & $0.0\%$ &  & $\mathbf{1.90 \times 10^{-4}}$s & $3.69$s & $2.17 \times 10^{3}$s\\
    ShepMod & $\mathbf{66.4}\%$ &  & $1.13$s & $1.82 \times 10^{-4}$s & $0.107$s\\
    LSHEP & $0.0\%$ &  & $2.39$s & $\mathbf{1.44 \times 10^{-4}}$s & $\mathbf{0.0845}$s\\
    Voronoi & $1.6\%$ &  & $\mathit{0.298}$s & $0.0274$s & $16.1$s\\
    BoxSpline & $0.0\%$ &  & $1.26$s & $6.43 \times 10^{-4}$s & $0.377$s\\
    \cline{1-2}\cline{4-6}
  \end{tabular}
  \caption{The left above shows how often each algorithm had the lowest absolute error approximating parkinsons data. On the right columns are median fit time of 5288 points, median time for one approximation, and median time approximating 587 points.}
  \label{table:best-parkinsons}
\end{table}

\begin{table}
  \centering
  \begin{tabular}{c|c|c|c|c|c}
    \hline
    Algorithm & Min & $25^{th}$ & $50^{th}$ & $75^{th}$ & Max\\
    \hline
    MARS & $\mathit{6.45 \times 10^{-15}}$ & $\mathbf{2.70 \times 10^{-14}}$ & $1.66$ & $1.96$ & $\mathit{53.3}$\\
    SVR & $0.0420$ & $0.143$ & $0.148$ & $\mathit{0.860}$ & $119.0$\\
    MLP & $6.89 \times 10^{-5}$ & $0.0337$ & $\mathbf{0.0795}$ & $\mathbf{0.264}$ & $\mathbf{5.31}$\\
    Delaunay & $\mathbf{0.00}$ & $0.187$ & $0.919$ & $2.72$ & $56.3$\\
    ShepMod & $\mathbf{0.00}$ & $0.0957$ & $0.685$ & $2.90$ & $73.2$\\
    LSHEP & $\mathbf{0.00}$ & $\mathit{0.0153}$ & $\mathit{0.106}$ & $1.17$ & $113.0$\\
    Voronoi & $\mathbf{0.00}$ & $0.430$ & $1.28$ & $2.94$ & $83.8$\\
    BoxSpline & $\mathbf{0.00}$ & $0.789$ & $2.06$ & $5.25$ & $119.0$\\
    \hline
  \end{tabular}
  \caption{This numerical data accompanies the visual provided in figure \ref{fig:error-weather}. The columns of absolute error percentiles correspond to the minimum, first quartile, median, second quartile, and maximum absolute errors respectively. The minimum value of each column is boldface, while the second lowest is italicized. All values are rounded to three significant digits.}
  \label{table:error-weather}
\end{table}

\begin{table}
  \centering
  \begin{tabular}{|c|c| c |c|c|c|}
    \cline{1-2}\cline{4-6}
    Algorithm & \% Best &  & Fit Time & App. Time (1) & Total App. Time\\
    \cline{1-2}\cline{4-6}
    MARS & $10.7\%$ &  & $0.151$s & $1.17 \times 10^{-4}$s & $0.0304$s\\
    SVR & $4.1\%$ &  & $0.133$s & $\mathbf{9.47 \times 10^{-5}}$s & $\mathbf{0.0246}$s\\
    MLP & $\mathbf{56.8}\%$ &  & $169.0$s & $1.37 \times 10^{-3}$s & $0.356$s\\
    Delaunay & $0.1\%$ &  & $\mathbf{1.16 \times 10^{-4}}$s & $1.55$s & $403.0$s\\
    ShepMod & $3.5\%$ &  & $0.265$s & $1.28 \times 10^{-4}$s & $0.0333$s\\
    LSHEP & $\mathit{28.4}\%$ &  & $0.874$s & $\mathit{9.75 \times 10^{-5}}$s & $\mathit{0.0254}$s\\
    Voronoi & $0.2\%$ &  & $\mathit{0.0112}$s & $0.0270$s & $7.01$s\\
    BoxSpline & $0.3\%$ &  & $0.330$s & $4.06 \times 10^{-4}$s & $0.106$s\\
    \cline{1-2}\cline{4-6}
  \end{tabular}
  \caption{The left above shows how often each algorithm had the lowest absolute error approximating Sydney rainfall data. On the right columns are median fit time of 2349 points, median time for one approximation, and median time approximating 260 points.}
  \label{table:best-weather}
\end{table}

\begin{table}
  \centering
  \begin{tabular}{c|c|c|c|c|c}
    \hline
    Algorithm & Min & $25^{th}$ & $50^{th}$ & $75^{th}$ & Max\\
    \hline
    MARS & $5.36$ & $3610.0$ & $4580.0$ & $5450.0$ & $13400.0$\\
    SVR & $0.0138$ & $8.47$ & $15.4$ & $41.9$ & $7700.0$\\
    MLP & $0.00151$ & $\mathbf{1.60}$ & $\mathbf{3.86}$ & $\mathbf{8.34}$ & $\mathbf{604.0}$\\
    Delaunay & $\mathbf{0.00}$ & $\mathit{2.69}$ & $\mathit{13.8}$ & $\mathit{35.0}$ & $\mathit{4840.0}$\\
    ShepMod & $\mathbf{0.00}$ & $4.21$ & $17.3$ & $51.6$ & $6510.0$\\
    LSHEP & $1.27$ & $199.0$ & $260.0$ & $343.0$ & $6530.0$\\
    Voronoi & $\mathit{2.89 \times 10^{-10}}$ & $14.7$ & $32.8$ & $52.9$ & $4860.0$\\
    BoxSpline & $0.00740$ & $20.8$ & $40.9$ & $79.6$ & $7640.0$\\
    \hline
  \end{tabular}
  \caption{This numerical data accompanies the visual provided in figure \ref{fig:error-credit-card}. The columns of absolute error percentiles correspond to the minimum, first quartile, median, second quartile, and maximum absolute errors respectively. The minimum value of each column is boldface, while the second lowest is italicized. All values are rounded to three significant digits.}
  \label{table:error-credit-card}
\end{table}

\begin{table}
  \centering
  \begin{tabular}{|c|c| c |c|c|c|}
    \cline{1-2}\cline{4-6}
    Algorithm & \% Best &  & Fit Time & App. Time (1) & Total App. Time\\
    \cline{1-2}\cline{4-6}
    MARS & $0.0\%$ &  & $22.0$s & $1.48 \times 10^{-3}$s & $0.820$s\\
    SVR & $0.0\%$ &  & $1.01$s & $\mathit{2.10 \times 10^{-4}}$s & $\mathit{0.117}$s\\
    MLP & $\mathbf{79.5}\%$ &  & $290.0$s & $7.14 \times 10^{-4}$s & $0.397$s\\
    Delaunay & $\mathit{20.5}\%$ &  & $\mathbf{2.43 \times 10^{-4}}$s & $6.83$s & $3.80 \times 10^{3}$s\\
    ShepMod & $0.1\%$ &  & $1.45$s & $2.20 \times 10^{-4}$s & $0.122$s\\
    LSHEP & $0.0\%$ &  & $3.47$s & $\mathbf{1.76 \times 10^{-4}}$s & $\mathbf{0.0981}$s\\
    Voronoi & $0.0\%$ &  & $\mathit{0.197}$s & $0.0950$s & $52.8$s\\
    BoxSpline & $0.0\%$ &  & $1.66$s & $9.56 \times 10^{-4}$s & $0.532$s\\
    \cline{1-2}\cline{4-6}
  \end{tabular}
  \caption{The left above shows how often each algorithm had the lowest absolute error approximating credit card transaction data. On the right columns are median fit time of 5006 points, median time for one approximation, and median time approximating 556 points.}
  \label{table:best-credit-card}
\end{table}

\begin{table}
  \centering
  \begin{tabular}{c|c|c|c|c|c}
    \hline
    Algorithm & Min & $25^{th}$ & $50^{th}$ & $75^{th}$ & Max\\
    \hline
    Delaunay & $\mathbf{0.0287}$ & $\mathbf{0.0914}$ & $\mathbf{0.158}$ & $\mathit{0.308}$ & $\mathit{0.897}$\\
    ShepMod & $0.0307$ & $\mathit{0.0983}$ & $\mathit{0.164}$ & $0.335$ & $1.00$\\
    Voronoi & $\mathit{0.0303}$ & $0.100$ & $0.165$ & $\mathbf{0.274}$ & $\mathbf{0.762}$\\
    BoxSpline & $0.0703$ & $0.417$ & $0.561$ & $0.745$ & $1.00$\\
    \hline
  \end{tabular}
  \caption{This numerical data accompanies the visual provided in figure \ref{fig:error-throughput}. The columns of absolute error percentiles correspond to the minimum, first quartile, median, second quartile, and maximum KS statistics respectively between truth and guess for models predicting the distribution of I/O throughput that will be observed at previously unseen system configurations. The minimum value of each column is boldface, while the second lowest is italicized. All values are rounded to three significant digits.}
  \label{table:error-throughput}
\end{table}

\begin{table}
  \centering
  \begin{tabular}{|c|c| c |c|c|c|}
    \cline{1-2}\cline{4-6}
    Algorithm & \% Best &  & Fit Time & App. Time (1) & Total App. Time\\
    \cline{1-2}\cline{4-6}
    Delaunay & $\mathbf{62.0}\%$ &  & $\mathbf{5.17 \times 10^{-4}}$s & $0.353$s & $106.0$s\\
    ShepMod & $0.0\%$ &  & $0.0884$s & $\mathbf{1.45 \times 10^{-4}}$s & $\mathbf{0.0436}$s\\
    Voronoi & $\mathit{38.0}\%$ &  & $\mathit{0.0173}$s & $2.53 \times 10^{-3}$s & $0.762$s\\
    BoxSpline & $0.0\%$ &  & $0.0972$s & $\mathit{2.10 \times 10^{-4}}$s & $\mathit{0.0633}$s\\
    \cline{1-2}\cline{4-6}
  \end{tabular}
  \caption{The left above shows how often each algorithm had the lowest KS statistic on the I/O throughput distribution data. On the right columns are median fit time of 2715 points, median time for one approximation, and median time approximating 301 points.}
  \label{table:best-throughput}
\end{table}

\end{appendix}

\end{document}

### 2019-03-07 Paper_[2019-02]_Numerical_Algorithms/NA2019b.tex
Before first round of revisions.e

--- 2019-03-04 Paper_[2019-02]_Numerical_Algorithms/tchlux-NA-2019.tex
@@
 
-In this work, multivariate interpolation is defined given a closed convex subset $Y$ of a metrizable topological vector space with metric $s$, some function $f:\mathbb{R}^d \rightarrow Y$ and a set of points $X = \bigl\{x^{(1)}$,$\ldots$,$x^{(n)}\bigr\} \subset \mathbb{R}^d,$ along with associated function values $f\bigl(x^{(i)}\bigr)$.
-The goal is to construct an approximation $\hat f: \mathbb{R}^d \rightarrow Y$ such that $\hat f\bigl(x^{(i)}\bigr) = f\bigl(x^{(i)}\bigr)$ for all $i = 1$,$\ldots$,$n$.
-It is often the case that the form of the true underlying function $f$ is unknown, however it is still desirable to construct an approximation $\hat f$ with small approximation error at $y \notin X$.
-The two metric spaces that will be discussed in this work are the real numbers with metric $s(x,y) = |x-y|$, and the set of cumulative distribution functions (CDFs) with the Kolmogorov-Smirnov (KS) statistic as metric.
+In this work, multivariate interpolation is defined given a closed convex
+subset $Y$ of a metrizable topological vector space with metric $s$, some
+function $f:\mathbb{R}^d \rightarrow Y$ and a set of points $X =
+\bigl\{x^{(1)}$, $\ldots$, $x^{(n)}\bigr\} \subset \mathbb{R}^d,$ along with
+associated function values $f\bigl(x^{(i)}\bigr)$. The goal is to construct
+an approximation $\hat f: \mathbb{R}^d \rightarrow Y$ such that $\hat
+f\bigl(x^{(i)}\bigr) = f\bigl(x^{(i)}\bigr)$ for all $i = 1$, $\ldots$, $n$.
+It is often the case that the form of the true underlying function $f$ is
+unknown, however it is still desirable to construct an approximation $\hat
+f$ with small approximation error at $y \notin X$. The two metric spaces
+that will be discussed in this work are the real numbers with metric $s(x,y)
+= |x-y|$, and the set of cumulative distribution functions (CDFs) with the
+Kolmogorov-Smirnov (KS) statistic as metric.
 
@@
 
-Provided the theoretical and practical motivations for exploring interpolants, this work aims to study the empirical performance differences between a set of scalable (moderately) interpolation techniques and a set of common regression techniques.
-These techniques are applied to a collection of moderately high dimensional problems ($5 \le d \le 30$) and the empirical results are discussed.
-
-The remainder of this paper is organized as follows.
-Sections \ref{sec:regression} and \ref{sec:interpolation} present the multivariate models.
-Section \ref{sec:error} presents the error measuring methodology that is used for collecting empirical results.
-Section \ref{sec:theory} presents the theoretical error bounds for interpolation.
-Section \ref{sec:data} presents the data sets for empirical approximation analysis and presents the approximation results for all models on these data sets.
-Section \ref{sec:discussion} analyzes and discusses the results, their implications, and their limitations.
-Finally, Section \ref{sec:conclusion} concludes.
+Provided the theoretical and practical motivations for exploring
+interpolants, this work aims to study the empirical performance differences
+between a set of scalable (moderately) interpolation techniques and a set
+of common regression techniques. These techniques are applied to a collection
+of moderately high dimensional problems ($5 \le d \le 30$) and the empirical
+results are discussed.
+
+The remainder of this paper is organized as follows. Sections
+\ref{sec:regression} and \ref{sec:interpolation} present the multivariate
+models. Section \ref{sec:error} gives the error measuring methodology
+that is used for collecting empirical results. Section \ref{sec:theory}
+contains the theoretical error bounds for interpolation. Section
+\ref{sec:data} presents the data sets for empirical approximation analysis
+and the approximation results for all models on these data sets.
+Section \ref{sec:discussion} analyzes and discusses the results, their
+implications, and their limitations. Finally, Section \ref{sec:conclusion}
+concludes.
 
@@
 \label{sec:svr}
-Support vector machines are a common method used in machine learning classification tasks that can be adapted for the purpose of regression \cite{basak2007support}.
-How to build a support vector regressor (SVR) is beyond the scope of this summary, but the resulting functional fit $p : \mathbb{R}^d \rightarrow \mathbb{R}$ has the form
+Support vector machines are a common method used in machine learning
+classification tasks that can be adapted for the purpose of regression
+\cite{basak2007support}. How to build a support vector regressor (SVR) is
+beyond the scope of this summary, but the resulting functional fit $p :
+\mathbb{R}^d \rightarrow \mathbb{R}$ has the form
  $$ p(x)  = \sum_{i=1}^{n}a_i K\bigl(x,x^{(i)}\bigr) + b ,$$
-where $K$ is the selected kernel function, $a_i \in \mathbb{R}^n$, $b \in \mathbb{R}$ are coefficients to be solved for simultaneously.
-The computational complexity of the SVR is $\mathcal{O}(n^2dm)$, with $m$ being determined by the minimization convergence criterion.
-This paper uses the scikit-learn SVR \cite{scikit-learn} with a polynomial kernel function.
+where $K$ is the selected kernel function, $a \in \mathbb{R}^n$, $b \in
+\mathbb{R}$ are coefficients to be solved for simultaneously. The
+computational complexity of the SVR is $\mathcal{O}(n^2dm)$, with $m$ being
+determined by the minimization convergence criterion. This paper uses the
+scikit-learn SVR \cite{scikit-learn} with a polynomial kernel function.
 
@@
 \label{sec:mlp}
-The neural network is a well studied and widely used method for both regression and classification tasks \cite{rumelhart1988learning,hornik1989multilayer}.
-When using the rectified linear unit (ReLU) activation function \cite{dahl2013improving} and fitting the model with stochastic gradient descent (SGD) or BFGS minimization techniques \cite{goh2017why,moller1993scaled,robbins1951stochastic}, the model built by a multilayer perceptron uses layers $l : \mathbb{R}^{i} \rightarrow \mathbb{R}^{j}$ defined by
+The neural network is a well studied and widely used method for both
+regression and classification tasks
+\cite{rumelhart1988learning,hornik1989multilayer}. When using the rectified
+linear unit (ReLU) activation function \cite{dahl2013improving} and fitting
+the model with stochastic gradient descent (SGD) or BFGS minimization
+techniques \cite{goh2017why,moller1993scaled,robbins1951stochastic}, the
+model built by a multilayer perceptron uses layers $l : \mathbb{R}^{i}
+\rightarrow \mathbb{R}^{j}$ defined by
  $$ l(u) = \bigl( u^t W_l \bigr)_+ ,$$
-where $W_l$ is the $i$ by $j$ weight matrix for layer $l$.
-In this form, the multilayer perceptron (MLP) produces a piecewise linear model of the input data.
-The computational complexity of training a multilayer perceptron is $\mathcal{O}(n d m)$, where $m$ is determined by the sizes of the layers of the network and the stopping criterion of the minimization used for finding weights.
-This paper uses a MLP built from Keras and Tensorflow to perform regression \cite{chollet2015keras,tensorflow2015-whitepaper} with ten hidden layers each having thirty two nodes (a total of approximately ten thousand parameters), ReLU activation, and five thousand epochs of SGD for training.
+where $W_l$ is the $i\times j$ weight matrix for layer $l$. In this form,
+the multilayer perceptron (MLP) produces a piecewise linear model of the
+input data. The computational complexity of training a multilayer perceptron
+is $\mathcal{O}(n d m)$, where $m$ is determined by the sizes of the layers
+of the network and the stopping criterion of the minimization used for
+finding weights. This paper uses a MLP built from Keras and Tensorflow to
+perform regression \cite{chollet2015keras,tensorflow2015-whitepaper} with
+ten hidden layers each having thirty two nodes (a total of approximately
+ten thousand parameters), ReLU activation, and five thousand epochs of SGD
+for training.
 
@@
 
-The modified Shepard method used here (ShepMod) generates a continuous approximation based on Euclidean distance and resembles a nearest neighbor interpolant \cite{cover1967nearest}.
-This model is a type of \textit{metric interpolation}, also called a Shepard method \cite{shepard1968two,gordon1978shepard}.
-The interpolant has the form
-$$ p(x) = \frac{\sum\limits_{k=1}^{n}W_k(x)f\bigl(x^{(k)}\bigr)}{\sum\limits_{k=1}^{n}W_k(x)} ,$$
-where $W_k(x) \propto \bigl\|x - x^{(k)}\bigr\|_2^{-2}$ and decreases to zero at a radius $r_k \in \mathbb{R}$ such that exactly $d+1$ other points $x^{(j)}$, $j \not = k$ are inside the closed Euclidean ball of radius $r_k$ about $x^{(k)}$.
-The computational complexity of ShepMod is $\mathcal{O}(n^2d)$.
-This paper uses a Fortran 95 implementation of ShepMod derived from SHEPPACK \cite{thacker2010algorithm}.
+The modified Shepard method used here (ShepMod) generates a continuous
+approximation based on Euclidean distance and resembles a nearest neighbor
+interpolant \cite{cover1967nearest}. This model is a type of \textit{metric
+interpolation}, also called a Shepard method
+\cite{shepard1968two,gordon1978shepard}. The interpolant has the form
+ $$ p(x) = \frac{\sum\limits_{k=1}^{n}W_k(x)f\bigl(x^{(k)}\bigr)}
+ {\sum\limits_{k=1}^{n}W_k(x)} ,$$
+where $W_k(x) = \bigl(r_k - \bigl\|x - x^{(k)}\bigr\|\bigr)_+ \big/
+\bigl\|x - x^{(k)}\bigr\|_2^2$ and $r_k \in \mathbb{R}$ is the smallest
+radius such that at least $d+1$ other points $x^{(j)}$, $j \not = k$, are
+inside the closed Euclidean ball of radius $r_k$ about $x^{(k)}$. The
+computational complexity of ShepMod is $\mathcal{O}(n^2d)$. This paper
+uses a Fortran 95 implementation of ShepMod derived from SHEPPACK
+\cite{thacker2010algorithm}.
 
 \subsection{Linear Shepard}
-The linear Shepard method (LSHEP) is a blending function using local linear interpolants, a special case of the general Shepard algorithm \cite{thacker2010algorithm}. The interpolant has the form
-$$ p(x) = \frac{\sum\limits_{k=1}^{n}W_k(x)P_k(x)}{\sum\limits_{k=1}^{n}W_k(x)} ,$$
-where $W_k(x)$ is a locally supported weighting function and $P_k(x)$ is a local linear approximation to the data satisfying $P_k\bigl(x^{(k)}\bigr) = f\bigl(x^{(x)}\bigr)$.
-The computational complexity of LSHEP is $\mathcal{O}(n^2d^3)$.
-This paper uses the Fortran 95 implementation of LSHEP in SHEPPACK \cite{thacker2010algorithm}.
+The linear Shepard method (LSHEP) is a blending function using local linear
+interpolants, a special case of the general Shepard algorithm
+\cite{thacker2010algorithm}. The interpolant has the form
+ $$ p(x) = \frac{\sum\limits_{k=1}^{n}W_k(x)P_k(x)}
+ {\sum\limits_{k=1}^{n}W_k(x)} ,$$
+where $W_k(x)$ is a locally supported weighting function and $P_k(x)$ is
+a local linear approximation to the data satisfying $P_k\bigl(x^{(k)}\bigr)
+= f\bigl(x^{(x)}\bigr)$. The computational complexity of LSHEP is
+$\mathcal{O}(n^2d^3)$. This paper uses the Fortran 95 implementation of
+LSHEP in SHEPPACK \cite{thacker2010algorithm}.
 
@@
 
-A well-studied technique for classification and approximation is the nearest neighbor algorithm \cite{cover1967nearest}.
-Nearest neighbor inherently utilizes the convex region $C^{x^{(i)}}$ (Voronoi cell \cite{dirichlet1850reduction}) consisting of all points closer to $x^{(i)}$ than to any other point $x^{(j)}$.
-The Voronoi model smooths the nearest neighbor approximation by utilizing the Voronoi cells to define support via a generic basis function $v: \mathbb{R}^d \rightarrow \mathbb{R}_+$ given by
-$$ v^{x^{(i)}}(y) = \left(1 - \frac{\bigl\|y - x^{(i)}\bigr\|_2}{2 \ h\bigl(y - x^{(i)} \mid x^{(i)}\bigr)} \right)_+, $$
-where $h\bigl(w \mid x^{(i)}\bigr)$ is the distance between $x^{(i)}$ and the boundary of the Voronoi cell $C^{x^{(i)}}$ in the direction $w$.
-$v^{x^{(i)}}\bigl(x^{(j)}\bigr) = \delta_{ij}$ and $v^{x^{(i)}}$ has local support, giving the interpolated value
-$$ f(y) = \frac{\sum\limits_{i=1}^n v^{x^{(i)}}(y) f\bigl(x^{(i)}\bigr)}{\sum\limits_{i=1}^n v^{x^{(i)}}(y)}, $$
-where $0 \leq v^{x^{(i)}}(y) \leq 1$.
-The computational complexity of interpolation via this Voronoi model is $\mathcal{O}(n^2 d)$.
-All of the approximations are an interpolant involving a convex combination of known function values $f\bigl(x^{(i)}\bigr)$.
+A well-studied technique for classification and approximation is the nearest
+neighbor algorithm \cite{cover1967nearest}. Nearest neighbor inherently
+utilizes the convex region $C^{x^{(i)}}$ (Voronoi cell
+\cite{dirichlet1850reduction}) consisting of all points closer to $x^{(i)}$
+than to any other point $x^{(j)}$. The Voronoi model smooths the nearest
+neighbor approximation by utilizing the Voronoi cells to define support
+via a generic basis function $v: \mathbb{R}^d \rightarrow \mathbb{R}_+$
+given by
+ $$ v^{x^{(i)}}(y) = \left(1 - \frac{\bigl\|y - x^{(i)}\bigr\|_2} {2 \
+ h\bigl(y - x^{(i)} \mid x^{(i)}\bigr)} \right)_+, $$
+where $h\bigl(w \mid x^{(i)}\bigr)$ is the distance between $x^{(i)}$ and
+the boundary of the Voronoi cell $C^{x^{(i)}}$ in the direction $w$.
+$v^{x^{(i)}}\bigl(x^{(j)}\bigr) = \delta_{ij}$ and $v^{x^{(i)}}$ has local
+support, giving the interpolated value
+ $$ f(y) = \frac{\sum\limits_{i=1}^n v^{x^{(i)}}(y)
+ f\bigl(x^{(i)}\bigr)}{\sum\limits_{i=1}^n v^{x^{(i)}}(y)}, $$
+where $0 \leq v^{x^{(i)}}(y) \leq 1$. The computational complexity of
+interpolation via this Voronoi model is $\mathcal{O}(n^2 d)$. All of the
+approximations are an interpolant involving a convex combination of known
+function values $f\bigl(x^{(i)}\bigr)$.
 
@@
 
-When the range of an approximation is the real numbers, error is reported with summary statistics including: min absolute error, max absolute error, and absolute error quartiles.
-When the range of an approximation is the space of cumulative distribution functions, the Kolmogorov-Smirnov statistic (max-norm difference between the functions) is used.
-
-A hurdle when modeling function-valued outputs such as cumulative distribution functions (CDFs) or probability density functions (PDFs) is that certain properties must be maintained.
-It is necessary that a PDF $f: \mathbb{R} \rightarrow \mathbb{R}$ have the properties $f(x) \geq 0$ and $\int_{-\infty}^{\infty}f(x)dx = 1$.
-However, for a CDF $F: \mathbb{R} \rightarrow \mathbb{R}$ the properties are $F(x) \in [0,1]$ and $F(x)$ is absolutely continuous and nondecreasing.
-This work utilizes the fact that a convex combination of CDFs (or PDFs) results in a valid CDF.
-Given $G(x) = \sum_{i}w_i F_i(x)$, $\sum_{i} w_i = 1$, $w_i \geq 0$, and each $F_i$ is a valid CDF, $G$ must also be a valid CDF.
-A demonstration of how this is applied can be seen in Figure \ref{fig:prediction-example}.
+When the range of an approximation is the real numbers, error is reported
+with summary statistics including: min absolute error, max absolute error,
+and absolute error quartiles. When the range of an approximation is the
+space of cumulative distribution functions, the Kolmogorov-Smirnov statistic
+(max-norm difference between the functions) is used.
+
+A hurdle when modeling function-valued outputs such as cumulative
+distribution functions (CDFs) or probability density functions (PDFs) is
+that certain properties must be maintained. It is necessary that a PDF $f:
+\mathbb{R} \rightarrow \mathbb{R}$ have the properties $f(x) \geq 0$ and
+$\int_{-\infty}^{\infty}f(x)dx = 1$. However, for a CDF $F: \mathbb{R}
+\rightarrow \mathbb{R}$ the properties are $F(x) \in [0,1]$ and $F(x)$ is
+absolutely continuous and nondecreasing. This work utilizes the fact that
+a convex combination of CDFs (or PDFs) results in a valid CDF (or PDF).
+Given $G(x) = \sum_{i}w_i F_i(x)$, $\sum_{i} w_i = 1$, $w_i \geq 0$, and
+each $F_i$ is a valid CDF, $G$ must also be a valid CDF. A demonstration
+of how this is applied can be seen in Figure \ref{fig:prediction-example}.
 
@@
 
-This section presents the theoretical results bounding the error of Delaunay interpolation.
-The error analysis relies on the Delaunay interpolation for two reasons: (1) second order results can be obtained utilizing a Lipschitz constant on the gradient of a function, rather than standard Lipschitz bounds and (2) multiple other interpolants analyzed compute predictions as convex combinations of observed function values, which may allow for straight forward extensions of this error bound.
+This section presents the theoretical results bounding the error of Delaunay
+interpolation. The error analysis relies on the Delaunay interpolation for
+two reasons: (1) second order results can be obtained utilizing a Lipschitz
+constant on the gradient of a function, rather than standard Lipschitz
+bounds, and (2) multiple other interpolants analyzed compute predictions
+as convex combinations of observed function values, which may allow for
+straightforward extensions of this error bound.
 
@@
 
-In summary, the approximation error of a Delaunay interpolant tends quadratically towards zero when approaching observed data only when the diameter of the interpolating simplex is also reduced at a proportional rate.
-Without the incorporation of additional observations, only linear convergence to the true function is be achieved.
-Notice that the approximation error is largely determined by the spacing of observed data.
-Predictions made by simplices whose vertices are not well-spaced (i.e.
-have large diameter, or are nearly contained in a hyperplane) have higher error.
-In light of this error bound, an empirical evaluation of presented algorithms follows.
+In summary, the approximation error of a Delaunay interpolant tends
+quadratically towards zero when approaching observed data only when the
+diameter of the interpolating simplex is also reduced at a proportional
+rate. Without the incorporation of additional observations, only linear
+convergence to the true function is be achieved. Notice that the
+approximation error is largely determined by the spacing of observed data.
+Predictions made by simplices whose vertices are not well-spaced (i.e.,
+have large diameter, or are nearly contained in a hyperplane) have higher
+error. In light of this error bound, an empirical evaluation of the
+presented algorithms follows.
 
@@
 
-The theoretical results constructed in Section \ref{sec:theory} for Delaunay interpolation are promising, however they are difficult to interpret in context with the other approximation algorithms that do not have similar known uniform error bounds.
-For that reason, this paper utilizes five different data sets of varying dimension and application to construct approximations and compare the accuracy of different techniques.
-
-In the following five subsections the sources and targets of each test data set are described, as well as challenges and limitations related to approximating these data.
-The distribution of response values being modeled is presented followed by the distribution of approximation errors for each algorithm.
-
-Before approximations are constructed, all five data sets are rescaled such that the domain of approximation is the unit hypercube.
-The range of the first four data sets is the real numbers, while the range of the fifth data set is the space of cumulative distribution functions.
-All approximation techniques are applied to the first four data sets, while only those interpolants whose approximations are convex combinations of observed data are applied to the final data set.
-
-All approximations are constructed using $10$-fold cross validation as described in \cite{kohavi1995study}.
-This approach requires the division of data randomly into $10$ partitions (the same partitions are used by each algorithm).
-An algorithm is then evaluated by constructing an approximation over each set of $9$ partitions, predicting the values of the final partition.
-As a result, each observed data point is used in the construction of $9$ different approximations and is approximated exactly once.
-The $k$-fold cross validation method is data-efficient and provides an unbiased estimate of the expected prediction error \cite{kohavi1995study}, however it should be noted that neither this method nor others can provide a universally unbiased estimator for the variance of prediction error \cite{bengio2004no}.
-
-In addition to the figures displaying approximation results for each data set, tables of accompanying numerical results are located in the Appendix.
-All of the test data sets capture underlying functions that are almost certainly stochastic.
-As described in Section \ref{sec:introduction}, regression techniques appear most appropriate for these problems.
-However, in high dimension data grows exponentially more sparse.
-Given sparse data, regressors tend towards interpolation.
+The theoretical results constructed in Section \ref{sec:theory} for Delaunay
+interpolation are promising, however they are difficult to interpret in
+context with the other approximation algorithms that do not have similar
+known uniform error bounds. For that reason, this paper utilizes five
+different application data sets of varying dimension to construct
+approximations and compare the accuracy of different techniques.
+
+In the following five subsections the sources and targets of each test
+data set are described, as well as challenges and limitations related to
+approximating these data. The distribution of response values being modeled
+is presented followed by the distribution of approximation errors for each
+algorithm. The plots for all five data sets have the same statistical format.
+
+Before approximations are constructed, all five data sets are rescaled
+such that the domain of approximation is the unit hypercube. The range of
+the first four data sets is the real numbers, while the range of the fifth
+data set is the space of cumulative distribution functions. All approximation
+techniques are applied to the first four data sets, while only those
+interpolants whose approximations are convex combinations of observed data
+are applied to the final data set.
+
+All approximations are constructed using 10-fold cross validation as
+described in \cite{kohavi1995study}. This approach requires the partition
+of data randomly into ten subsets (the same subsets are used by
+each algorithm). An algorithm is then evaluated by constructing an
+approximation over each set of nine subsets, predicting the values of
+the omitted subset. As a result, each observed data point is used in the
+construction of nine different approximations and is approximated exactly
+once. The $k$-fold cross validation method is data-efficient and provides
+an unbiased estimate of the expected prediction error \cite{kohavi1995study},
+however it should be noted that neither this method nor others can provide
+a universally unbiased estimator for the variance of prediction error
+\cite{bengio2004no}.
+
+In addition to the figures displaying approximation results for each data
+set, tables of accompanying numerical results are located in the Appendix.
+All of the test data sets capture underlying functions that are almost
+certainly stochastic. As described in Section \ref{sec:introduction},
+regression techniques appear most appropriate for these problems. However,
+typically data grows exponentially more sparse with increasing
+dimension. Given sparse data, regressors tend towards interpolation.
 
 %% -------------------------------------------------------------------
-\subsection{Forest Fire ($n = 504, d = 12$)}
+\subsection{Forest Fire ($n = 504$, $d = 12$)}
 
@@
     \node (img)  {\includegraphics[width=0.75\textwidth,]{Figures/raw-histogram-forest-fire.pdf}};
-    \node[below=of img, node distance=1cm, yshift=1cm] {Forest First Area Burned};
+    \node[below=of img, node distance=1cm, yshift=1cm] {Forest Fire Area Burned};
     \node[left=of img, node distance=0cm, rotate=90, anchor=center,yshift=-0.7cm] {Count};
   \end{tikzpicture}
-  \caption{Above is a histogram of forest fire area burned under recorded weather conditions.
-The data is presented on a natural log scale because most values are small with exponentially fewer fires on record that burn large areas.}
+  \caption{Histogram of forest fire area burned under recorded weather
+conditions. The data is presented on a $\log_{10}$ scale because most values
+are small with exponentially fewer fires on record that burn large areas.}
   \label{fig:hist-forest-fire}
@@
   \end{tikzpicture}
-  \caption{All models are applied to approximate the amount of area that would be burned given environment conditions.
-The data is broken up into $10$ partitions and each set of $9$ partitions is used to construct a model that approximates the remaining partition.
-This results in exactly one prediction from each algorithm for each data point.
-These boxes depict the median (middle bar), median $95\%$ confidence interval (cones), quartiles (box edges), fences at $3/2$ interquartile range (whiskers), and outliers (dots) of absolute prediction error for each model.
-Similar to figure \ref{fig:hist-forest-fire}, the errors are presented on a natural log scale.
-The complementing numerical data to this figure is provided in table \ref{table:error-forest-fire} in the Appendix.}
-  \label{fig:error-forest-fire}
-\end{figure}
-
-The forest fire data set \cite{cortez2007data} describes the area of Montesinho park burned over months of the year along with environmental conditions.
-The twelve dimensions being used to model burn area are the $x$ and $y$ spatial coordinates of burns in the park, month of year (mapped to a 2D circle), the FFMC, DMC, DC, and ISI indices (see source for details), the temperature, relative humidity, wind speed, and outdoor rain.
-The original analysis of this data set demonstrated it to be difficult to model, likely due to the skew in response values.
-
-As suggested by figure \ref{fig:error-forest-fire}, the SVR has the lowest absolute prediction errors for 80\% of the data, with ShepMod and BoxSpline being the nearest overall competitors.
-The effectiveness of SVR on this data suggests the underlying function can be defined by relatively few parameters, as well as the importance of capturing the low-burn-area data points.
+  \caption{All models are applied to approximate the amount of area that
+  would be burned given environmental conditions. The data is partitioned
+  into ten subsets and each set of nine subsets is used to construct a
+  model that approximates the omitted subset. This results in exactly one
+  prediction from each algorithm for each data point. These boxes depict
+  the median (middle bar), median $95\%$ confidence interval (cones),
+  quartiles (box edges), fences at $3/2$ interquartile range (whiskers),
+  and outliers (dots) of absolute prediction error for each model. Similar
+  to Figure \ref{fig:hist-forest-fire}, the errors are presented on a
+  $\log_{10}$ scale. The numerical data corresponding to this figure is
+  provided in Table \ref{table:error-forest-fire} in the Appendix.}
+\label{fig:error-forest-fire} \end{figure}
+
+The forest fire data set \cite{cortez2007data} describes the area of
+Montesinho park burned over months of the year along with environmental
+conditions. The twelve dimensions being used to model burn area are the
+$x$ and $y$ spatial coordinates of burns in the park, month of year (mapped
+to a 2D circle), the FFMC, DMC, DC, and ISI indices (see source for details),
+the temperature, relative humidity, wind speed, and outdoor rain. The
+original analysis of this data set demonstrated it to be difficult to
+model, likely due to the skew in response values.  ***Missing 12th
+variable above*****
+
+As suggested by Figure \ref{fig:error-forest-fire}, the SVR has the lowest
+absolute prediction errors for 80\% of the data, with MLP and Delaunay
+being the nearest overall competitors. The effectiveness of SVR on this
+data suggests the underlying function can be defined by relatively few
+parameters, as well as the importance of capturing the low-burn-area data
+points. 
 %% -------------------------------------------------------------------
@@
 %% -------------------------------------------------------------------
-\subsection{Parkinson's Telemonitoring ($n = 5875, d = 19$)}
+\subsection{Parkinson's Telemonitoring ($n = 5875$, $d = 19$)}
 
@@
   \end{tikzpicture}
-  \caption{Above is a histogram of the Parkinson's patient total UPDRS clinical scores that will be approximated by each algorithm.}
+  \caption{Histogram of the Parkinson's patient total UPDRS clinical
+scores that will be approximated by each algorithm.}
   \label{fig:hist-parkinsons}
@@
   \end{tikzpicture}
-  \caption{All models are applied to approximate the total UPDRS score given audio features from patients' home life.
-The data is broken up into $10$ partitions and each set of $9$ partitions is used to construct a model that approximates the remaining partition.
-This results in exactly one prediction from each algorithm for each data point.
-These boxes depict the median (middle bar), median $95\%$ confidence interval (cones), quartiles (box edges), fences at $3/2$ interquartile range (whiskers), and outliers (dots) of absolute prediction error for each model.
-The complementing numerical data to this is provided in table \ref{table:error-parkinsons} in the Appendix.}
+  \caption{All models are applied to approximate the total UPDRS score
+  given audio features from patients' home life, using 10-fold cross
+  validation.  These boxes depict the median (middle bar), median $95\%$
+  confidence interval (cones), quartiles (box edges), fences at $3/2$
+  interquartile range (whiskers), and outliers (dots) of absolute prediction
+  error for each model. The numerical data corresponding to this figure
+  is provided in Table \ref{table:error-parkinsons} in the Appendix.}
  \label{fig:error-parkinsons}
@@
 
-The second data set for evaluation \cite{tsanas2010accurate} is derived from a speech monitoring study with the intent to automatically estimate Parkinson's disease symptom development in Parkinson's patients.
-The function to be approximated is a time-consuming clinical evaluation measure referred to as the UPDRS score.
-The total UPDRS score given by a clinical evaluation is estimated through 16 real numbers generated from biomedical voice measures of in-home sound recordings.
-
-Figure \ref{fig:error-parkinsons} shows the ShepMod algorithm has the lowest minimum, first quartile, and median of absolute errors for this problem, while providing the best prediction 66\% of the time.
-The MLP has the lowest third quartile and provides the best prediction for 32\% of approximations.
-The dominance of ShepMod may be in part to regular-interval total UPDRS scores provided by clinicians, favoring a nearest-neighbor strategy of prediction.
+The second data set for evaluation \cite{tsanas2010accurate} is derived
+from a speech monitoring study with the intent to automatically estimate
+Parkinson's disease symptom development in Parkinson's patients. The
+function to be approximated is a time-consuming clinical evaluation measure
+referred to as the UPDRS score. The total UPDRS score given by a clinical
+evaluation is estimated through 19 real numbers generated from biomedical
+voice measures of in-home sound recordings.
+
+Figure \ref{fig:error-parkinsons} shows the ShepMod algorithm has the
+lowest minimum, first quartile, and median of absolute errors for this
+problem, while providing the best prediction 66\% of the time. The MLP has
+the lowest third quartile and provides the best prediction for 32\% of
+approximations. The dominance of ShepMod may be due in part to
+regular-interval total UPDRS scores provided by clinicians, favoring a
+nearest-neighbor strategy of prediction.
 %% -------------------------------------------------------------------
@@
 %% -------------------------------------------------------------------
-\subsection{Australian Daily Rainfall Volume ($n = 2609, d = 23$)}
+\subsection{Australian Daily Rainfall Volume ($n = 2609$, $d = 23$)}
 
@@
   \end{tikzpicture}
-  \caption{Above is a histogram of daily rainfall in Sydney, Australia.
-The data is presented on a natural log scale because the frequency of larger amounts of rainfall is significantly less.
-There is a peak in occurrence of the value $0$, which has a notable effect on the resulting model performance.}
+  \caption{Histogram of daily rainfall in Sydney, Australia, presented on a
+$\log_{10}$ scale because the frequency of larger amounts of rainfall
+is significantly less. There is a peak in occurrence of the value $0$,
+which has a notable effect on the resulting model performance.}
   \label{fig:hist-weather}
@@
   \end{tikzpicture}
-  \caption{All models are applied to approximate the amount of rainfall expected on the next calendar day given various sources of local meteorological data.
-The data is broken up into $10$ partitions and each set of $9$ partitions is used to construct models that approximates the remaining partition.
-This results in exactly one prediction from each algorithm for each data point.
-These boxes depict the median (middle bar), median $95\%$ confidence interval (cones), quartiles (box edges), fences at $3/2$ interquartile range (whiskers), and outliers (dots) of absolute prediction error for each model.
-The errors are presented on a natural log scale, mimicking the presentation in figure \ref{fig:hist-weather}.
-The complementing numerical data to this figure is provided in table \ref{table:error-weather} in the Appendix.}
+  \caption{All models are applied to approximate the amount of rainfall
+  expected on the next calendar day given various sources of local
+  meteorological data, using 10-fold cross validation.
+  These boxes depict the median (middle bar), median $95\%$ confidence
+  interval (cones), quartiles (box edges), fences at $3/2$ interquartile
+  range (whiskers), and outliers (dots) of absolute prediction error for
+  each model. The errors are presented on a $\log_{10}$ scale, mimicking
+  the presentation in Figure \ref{fig:hist-weather}. The 
+  numerical data corresponding to this figure is provided in Table
+  \ref{table:error-weather} in the Appendix.}
 \label{fig:error-weather}
@@
 
-The third data regarding the total daily rainfall in Sydney, Australia \cite{williams2009rattle} provides a slightly higher dimension challenge for the interpolants and regressors.
-This data is composed of many meteorological readings from the region on a day including: min and max temperatures, sunshine, wind speed directions (converted to coordinates on circle), wind speeds, and humidities throughout the day, day of the year (converted to coordinates on circle), and the model must predict the amount of rainfall tomorrow.
-
-While figure \ref{fig:hist-weather} makes MARS look far better than other techniques, it only provides the best prediction for 11\% of points.
-The MLP has the lowest absolute error for 56\% of points and LSHEP is best for 28\%.
-MARS likely achieves such a low first quartile due to the high occurrence of the value 0 in the data.
+The third data set for the total daily rainfall in Sydney, Australia
+\cite{williams2009rattle} provides a slightly higher dimensional challenge
+for the interpolants and regressors. This data is composed of many
+meteorological readings from the region in a day including: min and max
+temperatures, sunshine, wind speed directions (converted to coordinates
+on a circle), wind speeds, and humidities throughout the day, day of the
+year (converted to coordinates on a circle), and the model must predict the
+amount of rainfall tomorrow.
+
+While Figure \ref{fig:hist-weather} makes MARS look far better than other
+techniques, it only provides the best prediction for 11\% of points. The
+MLP has the lowest absolute error for 56\% of points and LSHEP is best for
+28\%.  MARS likely achieves such a low first quartile due to the high
+occurrence of the value zero in the data.
 %% -------------------------------------------------------------------
@@
 %% -------------------------------------------------------------------
-\subsection{Credit Card Transaction Amount ($n = 5562, d = 28$)}
+\subsection{Credit Card Transaction Amount ($n = 5562$, $d = 28$)}
 
@@
   \end{tikzpicture}
-  \caption{Above is a histogram of credit card transaction amounts.
-The data is presented on a natural log scale.
-The data contains a notable frequency peak around $\$1$ transactions.
-Fewer large purchases are made, but some large purchases are in excess of five orders of magnitude greater than the smallest purchases.}
+  \caption{Histogram of credit card transaction amounts, presented on a
+  $\log_{10}$ scale. The data contains a notable frequency peak around
+  $\$1$ transactions. Fewer large purchases are made, but some large
+  purchases are in excess of five orders of magnitude greater than the
+  smallest purchases.}
+
  \label{fig:hist-credit-card}
@@
   \end{tikzpicture}
-  \caption{All models are applied to approximate the expected transaction amount given transformed (and obfuscated) vendor and customer-descriptive features.
-The data is broken up into $10$ partitions and each set of $9$ partitions is used to construct a model that approximates the remaining partition.
-This results in exactly one prediction from each algorithm for each data point.
-These boxes depict the median (middle bar), median $95\%$ confidence interval (cones), quartiles (box edges), fences at $3/2$ interquartile range (whiskers), and outliers (dots) of absolute prediction error for each model.
-The absolute errors in transaction amount predictions are presented on a natural log scale, just as in figure \ref{fig:hist-credit-card}.
-The complementing numerical data to this figure is provided in table \ref{table:error-credit-card} in the Appendix.}
+  \caption{All models are applied to approximate the expected transaction
+  amount given transformed (and obfuscated) vendor and customer-descriptive
+  features, using 10-fold cross validation.
+  These boxes depict the median (middle bar), median
+  $95\%$ confidence interval (cones), quartiles (box edges), fences at
+  $3/2$ interquartile range (whiskers), and outliers (dots) of absolute
+  prediction error for each model. The absolute errors in transaction
+  amount predictions are presented on a $\log_{10}$ scale, just as in
+  Figure \ref{fig:hist-credit-card}. The numerical data corresponding to
+  this figure is provided in Table \ref{table:error-credit-card} in the
+  Appendix.}
+
  \label{fig:error-credit-card}
@@
 
-The fourth test data set, and the final with a real-valued range, is a collection of credit card transactions \cite{pozzolo2015calibrating}.
-The provided data carries no direct real-world meaning, it is the output of a principle component analysis on the original hidden source data.
-This obfuscation is done to protect the information of the credit card users.
-This data has the largest dimension of all considered, at 28.
-A model over this data predicts the transaction amount given the vector of principle component information.
-
-As can be seen in figure \ref{fig:error-credit-card}, the MLP outperforms all other algorithms at the first, second, third, and fourth quartiles.
-The MLP produces the lowest absolute error prediction for 80\% of transactions, Delaunay bests the remaining 20\%.
-It is likely that with less data, Delaunay would be the best performer.
+The fourth test data set, and the final with a real-valued range, is a
+collection of credit card transactions \cite{pozzolo2015calibrating}. The
+provided data carries no direct real-world meaning, being the output of a
+principle component analysis on the original hidden source data. This
+obfuscation is done to protect the information of the credit card users.
+This data has the largest dimension of all considered, at 28. A model for
+this data predicts the transaction amount given the vector of principle
+component information.
+
+As can be seen in Figure \ref{fig:error-credit-card}, the MLP outperforms
+all other algorithms at the first, second, third, and fourth quartiles.
+The MLP produces the lowest absolute error prediction for 80\% of
+transactions, Delaunay bests the remaining 20\%. It is likely that with
+less data, Delaunay would be the best performer.
+
 %% -------------------------------------------------------------------
@@
 %% -------------------------------------------------------------------
-\subsection{High Performance Computing I/O ($n = 3016, d = 4$)}
+\subsection{High Performance Computing I/O ($n = 3016$, $d = 4$)}
 
@@
   \end{tikzpicture}
-  \caption{Histogram of the raw throughput values recorded during all IOzone tests across all system configurations.
-The distribution is skewed right, with few tests having significantly higher throughput than most others.
-The data is presented on a natural log scale, however the resulting left-skew reveals the data is not precisely distributed according to an exponential.}
+  \caption{Histogram of the raw throughput values recorded during all
+  IOzone tests across all system configurations. The distribution is skewed
+  right, with few tests having significantly lower throughput than most
+  others.}
 \label{fig:hist-throughput}
@@
   \end{tikzpicture}
-  \caption{Above, the models directly capable of predicting distributions are applied to predicting the expected CDF of I/O throughput at a previously unseen system configuration.
-The KS statistic between the observed distribution at each system configuration and the predicted distribution is recorded and presented above.
-Note that the above plot is \textit{not} log-scaled like the histogram.
-The complimentary numerical data for this figure is provided in Appendix table \ref{table:error-throughput}.}
+  \caption{The models directly capable of predicting distributions are
+  applied to predicting the expected CDF of I/O throughput at a previously
+  unseen system configuration, using 10-fold cross validation. The KS
+  statistic between the observed distribution at each system configuration
+  and the predicted distribution is recorded and presented above. Note
+  that the above plot is \textit{not} log-scaled like the histogram. The
+  numerical data corresponding to this figure is provided in Table
+  \ref{table:error-throughput} in the Appendix.}
 \label{fig:error-throughput}
@@
 
-The final of five data sets is four-dimensional and is produced by executing the IOzone benchmark from \cite{iozone} on a homogeneous cluster of computers.
-The system performance data was collected by executing IOzone 150 times for each of a select set of system configurations.
-A single IOzone execution reports the max I/O file-read throughput seen.
-The 150 executions for each system configuration are converted
-to their empirical distribution function point, which are fit with a piecewise cubic Hermite interpolating polynomial \cite{fritsch1980monotone} to approximate the CDF.
-These CDFs are capable of being approximated by all interpolants except LSHEP.
-The four dimensions being modeled to predict throughput mean are file size, record size, thread count, and CPU frequency.
-
-Delaunay achieves the lowest KS statistic for 62\% of approximations, while Voronoi is best for the remaining 38\%.
-Figure \ref{fig:error-throughput} shows that while Delaunay may have more best predictions, the behavior of Voronoi may be preferable.
+The final of five data sets is four-dimensional and is produced by executing
+the IOzone benchmark from \cite{iozone} on a homogeneous cluster of
+computers. The system performance data was collected by executing IOzone
+150 times for each of a select set of system configurations. A single
+IOzone execution reports the max I/O file-read throughput seen. The 150
+executions for each system configuration are converted to their empirical
+distribution step function, which is fit with a monotone piecewise Hermite
+cubic interpolating polynomial \cite{fritsch1980monotone} to approximate
+the CDF. These CDFs are capable of being approximated by all interpolants
+except LSHEP. The four dimensions being modeled to predict the throughput
+CDF are file size, record size, thread count, and CPU frequency.
+
+Delaunay achieves the lowest KS statistic for 62\% of approximations, while
+Voronoi is best for the remaining 38\%. Figure \ref{fig:error-throughput}
+shows that while Delaunay may have more best predictions, the behavior of
+Voronoi may be preferable.
+
 %% -------------------------------------------------------------------
@@
   \end{tikzpicture}
-  \caption{Histograms of the prediction error for each interpolant that produces predictions as convex combinations of observed data.
-The data is broken up into $10$ partitions and each set of $9$ partitions is used to construct a model that approximates the remaining partition.
-This results in exactly one prediction from each algorithm for each data point.
-The distributions show the KS statistics for the predicted throughput distribution versus the actual throughput distribution.
-The four vertical red lines represent commonly used $p$-values \{0.05, 0.01, 0.001, 1.0e-6\} respectively.
-All predictions to the right of a red line represent CDF predictions that are significantly different (by respective $p$-value) from the actual distribution according to the KS test.
-The numerical counterpart to this figure is presented in table \ref{table:null-hypothesis-results}.}
+  \caption{Histograms of the prediction error for each interpolant that
+  produces predictions as convex combinations of observed data, using
+  10-fold cross validation.  The histograms show the KS statistics for the
+  predicted throughput distribution versus the actual throughput
+  distribution. The four vertical lines represent commonly used $p$-values
+  0.05, 0.01, 0.001, $10^{-6}$, respectively. All predictions to the right
+  of a vertical line represent CDF predictions that are significantly
+  different (by respective $p$-value) from the actual distribution according
+  to the KS test. The numerical counterpart to this figure is presented
+  in Table \ref{table:null-hypothesis-results}.}
+
  \label{fig:throughput-prediction}
@@
   \begin{tabular}{c|c|c|c|c}
-     & $p$ = .05 & $p$ = .01 & $p$ = .001 & $p$ = 1.0e-6\\
+     & $p = .05$ & $p = .01$ & $p = .001$ & $p = 10^{-6}$\\
     \hline
@@
   \end{tabular}
-  \caption{This is the numerical counterpart to the histogram data presented in figure \ref{fig:throughput-prediction}.
-The columns display the percent of null hypothesis rejections by the KS-test when provided different selections of $p$-values for each algorithm.
-The algorithm with the lowest rejection rate at each $p$ is boldface, while the second lowest is italicized.}
+  \caption{Numerical counterpart of the histogram data presented in Figure
+  \ref{fig:throughput-prediction}. The columns display the percent of null
+  hypothesis rejections by the KS-test when provided different selections
+  of $p$-values for each algorithm. The algorithm with the lowest rejection
+  rate at each $p$ is boldface, while the second lowest is italicized.}
+
  \label{table:null-hypothesis-results}
@@
 
-Figure \ref{fig:throughput-prediction} expands on the KS statistic results presented in figure \ref{fig:error-throughput}.
-Agglomerate errors for each algorithm resemble a Gamma distribution.
-The percentages of significant prediction errors with varying $p$-values are on display in Table \ref{table:null-hypothesis-results}.
-When considering the $p=0.001$ results for each technique, a little over one third of the predicted CDFs are significantly different from the measured (and presumed) correct CDFs.
-However, it should be noted that with 150 samples, the error of an empirical distribution function (EDF) can reasonably be upwards of $.1$, which serves as a rough estimate for the lower limit of achievable error.
-Globally, only a third of Voronoi predictions fail to capture \textit{all} of the characteristics of the CDFs at new system configurations.
+Figure \ref{fig:throughput-prediction} expands on the KS statistic results
+presented in Figure \ref{fig:error-throughput}. Agglomerate errors for
+each algorithm resemble a Gamma distribution. The percentages of significant
+prediction errors with varying $p$-values are on display in Table
+\ref{table:null-hypothesis-results}. When considering the $p=0.001$ results
+for each technique, a little over one third of the predicted CDFs are
+significantly different from the measured (and presumed) correct CDFs.
+However, it should be noted that with 150 samples, the error of an empirical
+distribution function (EDF) can reasonably be upwards of $.1$, which serves
+as a rough estimate for the lower limit of achievable error. Globally,
+only a third of Voronoi predictions fail to capture \textit{all} of the
+characteristics of the CDFs at new system configurations.
 
@@
   \end{tabular}
-  \caption{This data is the average of Appendix tables \ref{table:error-forest-fire}, \ref{table:error-parkinsons}, \ref{table:error-weather}, \ref{table:error-credit-card}, and provides a gross summary of overall results.
-The columns display (weighted equally by data set, \textit{not} points) the average frequency with which any algorithm provided the lowest absolute error approximation, the average time-to-fit, and the average time required to approximated one point.
-Interpolants provide the lowest error approximation one third of the time across all data, while regressors occupy the other two thirds.
-This result is obtained without any customized tuning or preprocessing to maximize the performance of any given algorithm.
-In practice, tuning and preprocessing may have large effects on approximation performance.}
+  \caption{This average of Appendix Tables
+  \ref{table:error-forest-fire}, \ref{table:error-parkinsons},
+  \ref{table:error-weather}, and \ref{table:error-credit-card} provides
+  a gross summary of overall results. The columns display (weighted equally
+  by data set, \textit{not} points) the average frequency with which any
+  algorithm provided the lowest absolute error approximation, the average
+  time-to-fit, and the average time required to approximated one point.
+  Interpolants provide the lowest error approximation one third of the
+  time across all data, while regressors occupy the other two thirds. This
+  result is obtained without any customized tuning or preprocessing to
+  maximize the performance of any given algorithm. In practice, tuning and
+  preprocessing may have large effects on approximation performance.}
+
  \label{table:avg-performance}
@@
 
-Table \ref{table:avg-performance} summarizes results across the four test data sets with real-valued ranges.
-The interpolants discussed in this paper produce the \textit{best} approximations roughly one third of the time, and produce competitive approximations for almost all data sets.
-These test problems are almost certainly \textit{stochastic} in nature, but the high dimension yields to greater sparsity and makes interpolation a competitive approximation technique.
+Table \ref{table:avg-performance} summarizes results across the four test
+data sets with real-valued ranges. The interpolants discussed in this paper
+produce the \textit{best} approximations roughly one third of the time,
+and produce competitive approximations for almost all data sets. These
+test problems are almost certainly \textit{stochastic} in nature, but the
+high dimension leads to greater data sparsity and model construction cost,
+making interpolation more competitive.
 

### 2019-03-07 Paper_[2019-02]_Numerical_Algorithms/paper.tex
Updated code for running neural network.
Before first round of revisions.e

--- 2019-02-08 Documents_[2019]/approximation/paper.tex
@@
 \RequirePackage{fix-cm}
-\documentclass[smallextended]{svjour3}       % onecolumn (second format)
+\documentclass[smallextended,final]{svjour3}       % onecolumn (second format)
 \smartqed  % flush right qed marks, e.g. at end of proof
@@
 \usepackage{booktabs} % For formal tables
+\usepackage[colorlinks,allcolors=blue]{hyperref}
 
@@
   \begin{tabular}[#1]{@{}c@{}}#2\end{tabular}}
+
+%% Create a ``Lemma'' environment that does not have italics font.
+\spnewtheorem{plainlemma}{Lemma}{\bf}{}
+%% Create a ``Theorem'' environment that is not numbered.
+\spnewtheorem*{onetheorem}{Theorem}{\bf}{}
+\spnewtheorem*{proofdot}{Proof.}{\it}{}
 
@@
 
-\title{Interpolation of Sparse High-Dimensional Data\thanks{This
-    projected was supported by NSF grant}
+\title{Interpolation of Sparse High-Dimensional Data\thanks
+  {This work was supported by the National Science Foundation Grant CNS-1565314.}
 }
@@
 \begin{abstract}
-A rapid increase in the quantity of data available is allowing all fields of science to generate more accurate models of multivariate phenomena.
-Regression and interpolation become challenging when the dimension of data is large, especially while maintaining tractable computational complexity.
-While regression is a popular approach to solving approximation problems with high dimension, there are often advantages to interpolation.
-This paper presents error bounds for interpolants on moderately high dimensional problems and contrasts their performance with that of popular regression techniques.
-Empirical results demonstrate the viability of interpolants for high dimensional approximation problems, suggesting that these techniques are capable of effectively modeling multivariate phenomena while maintaining flexibility in different application domains.
-
-\keywords{approximation \and regression \and interpolation \and high dimension \and error bound}
+A rapid increase in the quantity of data available is allowing all
+fields of science to generate more accurate models of multivariate
+phenomena. Regression and interpolation become challenging when the
+dimension of data is large, especially while maintaining tractable
+computational complexity. While regression is a popular approach to
+solving approximation problems with high dimension, there are often
+advantages to interpolation. This paper presents error bounds for
+interpolants on moderately high dimensional problems and contrasts
+their performance with that of popular regression
+techniques. Empirical results demonstrate the viability of
+interpolants for high dimensional approximation problems, suggesting
+that these techniques are capable of effectively modeling multivariate
+phenomena while maintaining flexibility in different application
+domains.
+
+\keywords{approximation \and regression \and interpolation \and high
+  dimension \and error bound}
 % \PACS{PACS code1 \and PACS code2 \and more}
@@
 
+%% ===================================================================
 \section{Introduction}
@@
 
-Regression and interpolation are problems of considerable importance that find applications across many fields of science.
-Pollution and air quality analysis \cite{de2008field}, energy consumption management \cite{lazos2014optimisation}, and student performance prediction \cite{cortez2008using,lux2016applications} are a few of many interdisciplinary applications of multivariate regression for predictive analysis.
-As discussed later, these techniques can also be applied to prediction problems related to high performance computing (HPC) file input/output (I/O) \cite{lux2018nonparametric}, Parkinson's patient clinical evaluations \cite{tsanas2010accurate}, and forest fire risk assessment \cite{cortez2007data}.
-
-Regression and interpolation have a considerable theoretical base in one dimension \cite{cheney2009course}.
-Splines in particular are well understood as an interpolation technique in one dimension \cite{de1978practical}, particularly B-splines.
-Tensor products of B-splines \cite{unther1996interpolating} or other basis functions have an unfortunate exponential scaling with increasing dimension.
-Exponential scaling prohibits tensor products from being reasonably applied beyond three-dimensional data.
-In order to address this dimensional scaling challenge, C.
-de Boor and others proposed box splines \cite{de2013box}, of which one of the approximation techniques in this work is composed \cite{lux2018novel}.
-
-The theoretical foundation of low dimensional interpolation allows the construction of strong error bounds that are absent from high dimensional problems.
-This work extends some known results regarding the secant method \cite{dennis1996numerical} to construct an interpolation error bound for problems of arbitrary dimension.
-These error bounds are useful, considering the same cannot be said for regression algorithms in general.
-The maximum complexity of an interpolant is bounded by the amount of data available, while the maximum complexity of a regressor is bounded by both the amount of data and the chosen parametric form.
-For this reason, generic uniform bounds are largely unobtainable for regression techniques on arbitrary approximation problems, even when the approximation domain is bounded.
-
-Aside from theoretical motivation for the use of interpolants, there are often computational advantages as well.
-Interpolants do not have the need for \textit{fitting} data, or minimizing error with respect to model parameters.
-In applications where the amount of data is large and the relative number of predictions that need to be made for a given collection of data is small, the direct application of an interpolant is much less computationally expensive.
-
-In this work, multivariate interpolation is defined given a closed convex subset $Y$ of a metrizable topological vector space with metric $s$, some function $f:\mathbb{R}^d \rightarrow Y$ and a set of points $X = \bigl\{x^{(1)}$,$\ldots$,$x^{(n)}\bigr\} \subset \mathbb{R}^d,$ along with associated function values $f\bigl(x^{(i)}\bigr)$.
-The goal is to construct an approximation $\hat f: \mathbb{R}^d \rightarrow Y$ such that $\hat f\bigl(x^{(i)}\bigr) = f\bigl(x^{(i)}\bigr)$ for all $i = 1$,$\ldots$,$n$.
-It is often the case that the form of the true underlying function $f$ is unknown, however it is still desirable to construct an approximation $\hat f$ with small approximation error at $y \notin X$.
-The two metric spaces that will be discussed in this work are the real numbers with metric $s(x,y) = |x-y|$, and the set of cumulative distribution functions (CDFs) with the Kolmogorov-Smirnov (KS) statistic as metric.
-
-Multivariate regression is often used when the underlying function is presumed to be stochastic, or stochastic error is introduced in the evaluation of $f$.
-Hence, multivariate regression relaxes the conditions of interpolation by choosing parameters $P$ defining $\hat f(x;P)$ to minimize the error vector $\Bigl( \bigl | \hat f \bigl(x^{(1)};P\bigr) - f\bigl(x^{(1)}\bigr) \bigr|$, $\ldots$, $\bigl | \hat f \bigl(x^{(n)}; P\bigr) - f\bigl(x^{(n)}\bigr) \bigr | \Bigr)$ in some norm.
-The difficult question in the case of regression is often what parametric form to adopt for any given application.
-
-The most challenging problem when scaling in dimension is that the number of possible interactions between dimensions grows exponentially.
-Quantifying all possible interactions becomes intractable and hence beyond three-dimensional data, mostly linear models are used.
-That is not to say nonlinear models are absent, but nonlinearities are often either preconceived or model pairwise interactions between dimensions at most.
-Even globally nonlinear approximations such as neural networks are constructed from compositions of summed low-interaction functions \cite{clevert2015fast}.
-
-Provided the theoretical and practical motivations for exploring interpolants, this work aims to study the empirical performance differences between a set of scalable (moderately) interpolation techniques and a set of common regression techniques.
-These techniques are applied to a collection of moderately high dimensional problems ($5 \le d \le 50$) and the empirical results are discussed.
-
-The remainder of this paper is organized as follows.
-Section \ref{sec:multivariate} presents the multivariate models.
-Section \textbf{???} presents the error measuring methodology that is used for collecting empirical results.
-Section \textbf{???} presents the theoretical error bounds for interpolation.
-Section \textbf{???} presents the data sets for empirical approximation analysis and presents the approximation results for all models on these data sets.
-Section \textbf{???} analyzes and discusses the results, their implications, and their limitations.
-Finally, Section \textbf{???} concludes.
+Regression and interpolation are problems of considerable importance
+that find applications across many fields of science. Pollution and
+air quality analysis \cite{de2008field}, energy consumption management
+\cite{lazos2014optimisation}, and student performance prediction
+\cite{cortez2008using,lux2016applications} are a few of many
+interdisciplinary applications of multivariate regression for
+predictive analysis. As discussed later, these techniques can also be
+applied to prediction problems related to forest fire risk assessment
+\cite{cortez2007data}, Parkinson's patient clinical evaluations
+\cite{tsanas2010accurate}, local rainfall and weather
+\cite{williams2009rattle}, credit card transactions
+\cite{pozzolo2015calibrating}, and high performance computing (HPC)
+file input/output (I/O) \cite{lux2018nonparametric}.
+
+Regression and interpolation have a considerable theoretical base in
+one dimension \cite{cheney2009course}. Splines in particular are well
+understood as an interpolation technique in one dimension
+\cite{de1978practical}, particularly B-splines. Tensor products of
+B-splines \cite{unther1996interpolating} or other basis functions have
+an unfortunate exponential scaling with increasing
+dimension. Exponential scaling prohibits tensor products from being
+reasonably applied beyond three-dimensional data. In order to address
+this dimensional scaling challenge, C. de Boor and others proposed box
+splines \cite{de2013box}, of which one of the approximation techniques
+in this work is composed \cite{lux2018novel}.
+
+The theoretical foundation of low dimensional interpolation allows the
+construction of strong error bounds that are absent from high
+dimensional problems. This work extends some known results regarding
+the secant method \cite{dennis1996numerical} to construct an
+interpolation error bound for problems of arbitrary dimension. These
+error bounds are useful, considering the same cannot be said for
+regression algorithms in general. The maximum complexity of an
+interpolant is bounded by the amount of data available, while the
+maximum complexity of a regressor is bounded by both the amount of
+data and the chosen parametric form. For this reason, generic uniform
+bounds are largely unobtainable for regression techniques on arbitrary
+approximation problems, even when the approximation domain is bounded.
+
+Aside from theoretical motivation for the use of interpolants, there
+are often computational advantages as well. Interpolants do not have
+the need for \textit{fitting} data, or minimizing error with respect
+to model parameters. In applications where the amount of data is large
+and the relative number of predictions that need to be made for a
+given collection of data is small, the direct application of an
+interpolant is much less computationally expensive.
+
+In this work, multivariate interpolation is defined given a closed
+convex subset $Y$ of a metrizable topological vector space with metric
+$s$, some function $f:\mathbb{R}^d \rightarrow Y$ and a set of points
+$X = \bigl\{x^{(1)}$,$\ldots$,$x^{(n)}\bigr\} \subset \mathbb{R}^d,$
+along with associated function values $f\bigl(x^{(i)}\bigr)$. The goal
+is to construct an approximation $\hat f: \mathbb{R}^d \rightarrow Y$
+such that $\hat f\bigl(x^{(i)}\bigr) = f\bigl(x^{(i)}\bigr)$ for all
+$i = 1$,$\ldots$,$n$. It is often the case that the form of the true
+underlying function $f$ is unknown, however it is still desirable to
+construct an approximation $\hat f$ with small approximation error at
+$y \notin X$. The two metric spaces that will be discussed in this
+work are the real numbers with metric $s(x,y) = |x-y|$, and the set of
+cumulative distribution functions (CDFs) with the Kolmogorov-Smirnov
+(KS) statistic as metric.
+
+Multivariate regression is often used when the underlying function is
+presumed to be stochastic, or stochastic error is introduced in the
+evaluation of $f$. Hence, multivariate regression relaxes the
+conditions of interpolation by choosing parameters $P$ defining $\hat
+f(x;P)$ to minimize the error vector $\Bigl( \bigl | \hat f
+\bigl(x^{(1)};P\bigr) - f\bigl(x^{(1)}\bigr) \bigr|$, $\ldots$, $\bigl
+| \hat f \bigl(x^{(n)}; P\bigr) - f\bigl(x^{(n)}\bigr) \bigr | \Bigr)$
+in some norm. The difficult question in the case of regression is
+often what parametric form to adopt for any given application.
+
+The most challenging problem when scaling in dimension is that the
+number of possible interactions between dimensions grows
+exponentially. Quantifying all possible interactions becomes
+intractable and hence beyond three-dimensional data, mostly linear
+models are used. That is not to say nonlinear models are absent, but
+nonlinearities are often either preconceived or model pairwise
+interactions between dimensions at most. Even globally nonlinear
+approximations such as neural networks are constructed from
+compositions of summed low-interaction functions
+\cite{clevert2015fast}.
+
+Provided the theoretical and practical motivations for exploring
+interpolants, this work aims to study the empirical performance
+differences between a set of scalable (moderately) interpolation
+techniques and a set of common regression techniques. These techniques
+are applied to a collection of moderately high dimensional problems
+($5 \le d \le 30$) and the empirical results are discussed.
+
+The remainder of this paper is organized as follows. Sections
+\ref{sec:regression} and \ref{sec:interpolation} present the
+multivariate models. Section \ref{sec:error} presents the error
+measuring methodology that is used for collecting empirical
+results. Section \ref{sec:theory} presents the theoretical error
+bounds for interpolation. Section \ref{sec:data} presents the data
+sets for empirical approximation analysis and presents the
+approximation results for all models on these data sets. Section
+\ref{sec:discussion} analyzes and discusses the results, their
+implications, and their limitations. Finally, Section
+\ref{sec:conclusion} concludes.
 
@@
 
+%% ===================================================================
 \section{Multivariate Regression}
-Multivariate regressors are capable of accurately modeling a complex dependence of a response (in $Y$) on multiple variables (represented as a points in $\mathbb{R}^{d}$).
-The approximations to some (unknown) underlying function $f: \mathbb{R}^d \rightarrow Y$ are chosen to minimize some error measure related to data samples $f\bigl(x^{(i)}\bigr)$.
-For example, least squares regression uses the sum of squared differences between modeled function values and true function values as an error measure.
-In this section and the next, some techniques are limited to approximating real valued functions ($Y \subset \mathbb{R}$).
-These techniques are extended to real vector-valued functions by repeating the construction for each component of the vector outputs.
-Throughout the following, $x$ denotes a $d$-tuple, $x_i$ the $i$th component of $x$, and $x^{(i)}$ the $i$th $d$-tuple data point.
-Different symbols are used to represent the approximation function $\hat f$.
+\label{sec:regression}
+Multivariate regressors are capable of accurately modeling a complex
+dependence of a response (in $Y$) on multiple variables (represented
+as a points in $\mathbb{R}^{d}$). The approximations to some (unknown)
+underlying function $f: \mathbb{R}^d \rightarrow Y$ are chosen to
+minimize some error measure related to data samples
+$f\bigl(x^{(i)}\bigr)$. For example, least squares regression uses the
+sum of squared differences between modeled function values and true
+function values as an error measure. In this section and the next,
+some techniques are limited to approximating real valued functions ($Y
+\subset \mathbb{R}$). These techniques can be extended to real
+vector-valued ranges by repeating the construction for each component
+of the vector output. Throughout the following, $x$ denotes a
+$d$-tuple, $x_i$ the $i$th component of $x$, and $x^{(i)}$ the $i$th
+$d$-tuple data point. Different symbols are used to represent the
+approximation function $\hat f$.
 
 \subsection{Multivariate Adaptive Regression Splines}
-This approximation was introduced in \cite{friedman1991multivariate} and subsequently improved to its current version in \cite{stanford1993fast}, called fast multivariate adaptive regression splines (Fast MARS).In Fast MARS, a least squares fit model is iteratively built by beginning with a single constant valued function and adding two new basis functions at each iteration of the form
+\label{sec:mars}
+This approximation was introduced in \cite{friedman1991multivariate}
+and subsequently improved to its current version in
+\cite{stanford1993fast}, called fast multivariate adaptive regression
+splines (Fast MARS).In Fast MARS, a least squares fit model is
+iteratively built by beginning with a single constant valued function
+and adding two new basis functions at each iteration of the form
 \begin{align*}
@@
 \end{align*}
-where $j$ is the iteration number, $1 \le p \le n$, and $B_l(x)$, $B_k(x)$ are basis functions from the previous iteration,
+where $j$ is the iteration number, $1 \le p \le n$, and $B_l(x)$,
+$B_k(x)$ are basis functions from the previous iteration,
 $$w_+ = \begin{cases} w, & w \geq 0 \\ 0, & w < 0 \end{cases},$$
-and $w_- = (-w)_+$.
-After iteratively constructing a model, MARS then iteratively removes basis functions that do not contribute to goodness of fit.
-In effect, MARS creates a locally component-wise tensor product approximation of the data.
-The overall computational complexity of Fast MARS is $\mathcal{O}(n d m^3)$ where $m$ is the maximum number of underlying basis functions.
-This paper uses an implementation of Fast MARS \cite{rudy2017pyearth} with $m = 200$.
+and $w_- = (-w)_+$. After iteratively constructing a model, MARS then
+iteratively removes basis functions that do not contribute to goodness
+of fit. In effect, MARS creates a locally component-wise tensor
+product approximation of the data. The overall computational
+complexity of Fast MARS is $\mathcal{O}(n d m^3)$ where $m$ is the
+maximum number of underlying basis functions. This paper uses an
+implementation of Fast MARS \cite{rudy2017pyearth} with $m = 200$
+throughout all experiments.
+
+\subsection{Support Vector Regressor}
+\label{sec:svr}
+Support vector machines are a common method used in machine learning
+classification tasks that can be adapted for the purpose of regression
+\cite{basak2007support}. How to build a support vector regressor (SVR)
+is beyond the scope of this summary, but the resulting functional fit
+$p : \mathbb{R}^d \rightarrow \mathbb{R}$ has the form
+$$ p(x)  = \sum_{i=1}^{n}a_i K\bigl(x,x^{(i)}\bigr) + b ,$$
+where $K$ is the selected kernel function, $a_i \in \mathbb{R}^n$, 
+$b \in \mathbb{R}$ are coefficients to be solved for simultaneously. The
+computational complexity of the SVR is $\mathcal{O}(n^2dm)$, with $m$
+being determined by the minimization convergence criterion. This paper
+uses the scikit-learn SVR \cite{scikit-learn} with a polynomial kernel
+function.
 
 \subsection{Multilayer Perceptron Regressor}
-The neural network is a well studied and widely used method for both regression and classification tasks \cite{hornik1989multilayer}.
-When using the rectified linear unit (ReLU) activation function \cite{dahl2013improving} and training with the BFGS minimization technique \cite{moller1993scaled}, the model built by a multilayer perceptron uses layers $l : \mathbb{R}^{i} \rightarrow \mathbb{R}^{j}$ defined by
-$$ l(u) = \big( u^t W_l \big)_+ ,$$
-where $W_l$ is the $i$ by $j$ weight matrix for layer $l$.
-In this form, the multilayer perceptron (MLP) produces a piecewise linear model of the input data.
-The computational complexity of training a multilayer perceptron is $\mathcal{O}(n d m)$, where $m$ is determined by the sizes of the layers of the network and the stopping criterion of the BFGS minimization used for finding weights.
-This paper uses the scikit-learn MLP regressor \cite{scikit-learn}, a single hidden layer with 100 nodes, ReLU activation, and BFGS for training.
-
-\subsection{Support Vector Regressor}
-Support vector machines are a common method used in machine learning classification tasks that can be adapted for the purpose of regression \cite{basak2007support}.
-How to build a support vector regressor (SVR) is beyond the scope of this summary, but the resulting functional fit $p : \mathbb{R}^d \rightarrow \mathbb{R}$ has the form
-$$ p(x)  = \sum_{i=1}^{n}a_i K(x,x^{(i)}) + b ,$$
-where $K$ is the selected kernel function, $a \in \mathbb{R}^n$, $b \in \mathbb{R}$ are coefficients to be solved for simultaneously.
-The computational complexity of the SVR is $\mathcal{O}(n^2dm)$, with $m$ being determined by the minimization convergence criterion.
-This paper uses the scikit-learn SVR \cite{scikit-learn} with a polynomial kernel function.
-
+\label{sec:mlp}
+The neural network is a well studied and widely used method for both
+regression and classification tasks
+\cite{rumelhart1988learning,hornik1989multilayer}. When using the
+rectified linear unit (ReLU) activation function
+\cite{dahl2013improving} and fitting the model with stochastic
+gradient descent (SGD) or BFGS minimization techniques
+\cite{goh2017why,moller1993scaled,robbins1951stochastic}, the model
+built by a multilayer perceptron uses layers $l : \mathbb{R}^{i}
+\rightarrow \mathbb{R}^{j}$ defined by
+$$ l(u) = \bigl( u^t W_l \bigr)_+ ,$$
+where $W_l$ is the $i$ by $j$ weight matrix for layer $l$. In this
+form, the multilayer perceptron (MLP) produces a piecewise linear
+model of the input data. The computational complexity of training a
+multilayer perceptron is $\mathcal{O}(n d m)$, where $m$ is determined
+by the sizes of the layers of the network and the stopping criterion
+of the minimization used for finding weights. This paper uses a MLP
+built from Keras and Tensorflow to perform regression
+\cite{chollet2015keras,tensorflow2015-whitepaper} with ten hidden
+layers each having thirty two nodes (a total of approximately ten
+thousand parameters), ReLU activation, and five thousand epochs of SGD
+for training.
+
+%% ===================================================================
 \section{Multivariate Interpolation}
-The following interpolation techniques demonstrate a reasonable variety of approaches to interpolation.
-All of the presented interpolants produce approximations that are continuous in value, which is often a desirable property in applied approximation problems.
-
-\subsection{Linear Shepard}
-The linear Shepard method (LSHEP) is a blending function using local linear interpolants, a special case of the general Shepard algorithm \cite{thacker2010algorithm}. The interpolant has the form
-$$ p(x) = \frac{\sum\limits_{k=1}^{n}W_k(x)P_k(x)}{\sum\limits_{k=1}^{n}W_k(x)} ,$$
-where $W_k(x)$ is a locally supported weighting function and $P_k(x)$ is a local linear approximation to the data satisfying $P_k\big(x^{(k)}\big) = f\big(x^{(x)}\big)$.
-The computational complexity of LSHEP is $\mathcal{O}(n^2d^3)$.
-This paper uses the Fortran 95 implementation of LSHEP in SHEPPACK \cite{thacker2010algorithm}.
-
-%% ===================================================================
+\label{sec:interpolation}
+The following interpolation techniques demonstrate a reasonable
+variety of approaches to interpolation. All of the presented
+interpolants produce approximations that are continuous in value,
+which is often a desirable property in applied approximation problems.
+
 \subsection{Delaunay}
@@
 
-The Delaunay triangulation is a well-studied geometric technique for producing an interpolant \cite{lee1980two}.
-The Delaunay triangulation of a set of data points into simplices is such that there are no data points inside the sphere defined by the vertices of each simplex.
-For a $d$-simplex S with vertices $x^{(0)}$, $x^{(1)}$, $\ldots$, $x^{(d)}$, and function values $f(x^{(i)})$, $i=0$, $\ldots$, $d$, $y \in S$ is a unique convex combination of the vertices,
+The Delaunay triangulation is a well-studied geometric technique for
+producing an interpolant \cite{lee1980two}. The Delaunay triangulation
+of a set of data points into simplices is such that there are no data
+points inside the sphere defined by the vertices of each simplex. For
+a $d$-simplex S with vertices $x^{(0)}$, $x^{(1)}$, $\ldots$,
+$x^{(d)}$, and function values $f\bigl(x^{(i)}\bigr)$, $i=0$,
+$\ldots$, $d$, $y \in S$ is a unique convex combination of the
+vertices,
 $$ y = \sum_{i=0}^{d} w_i x^{(i)}, \quad \sum_{i=0}^{d} w_i = 1, \quad w_i \geq 0, \quad i=0,\ldots,d, $$
 and the Delaunay interpolant $\hat f(y)$ at $y$ is given by
-$$ \hat f(y) = \sum_{i=0}^{d} w_i f(x^{(i)}). $$
-The computational complexity of Delaunay interpolation (for the implementation used) is $\mathcal{O}(n d^4 \log d)$, which is capable of scaling reasonably to $d \leq 50$.
-In the present application, a Delaunay simplex $S$ containing $y$ is found, then the $d+1$ vertices (points in $X$) of $S$ are used to assign weights to each vertex and produce the predicted function value for point $y$.
+$$ \hat f(y) = \sum_{i=0}^{d} w_i f\bigl(x^{(i)}\bigr). $$
+The computational complexity of Delaunay interpolation (for the
+implementation used \cite{chang2018polynomial}) is $\mathcal{O}(n d^4
+\log d)$, which is capable of scaling reasonably to $d \leq 50$. In
+the present application, a Delaunay simplex $S$ containing $y$ is
+found, then the $d+1$ vertices (points in $X$) of $S$ are used to
+assign weights to each vertex and produce the predicted function value
+for point $y$.
+
+\subsection{Modified Shepard}
+\label{sec:modified-shepard}
+
+The modified Shepard method used here (ShepMod) generates a continuous
+approximation based on Euclidean distance and resembles a nearest
+neighbor interpolant \cite{cover1967nearest}. This model is a type of
+\textit{metric interpolation}, also called a Shepard method
+\cite{shepard1968two,gordon1978shepard}. The interpolant has the form
+$$ p(x) = \frac{\sum\limits_{k=1}^{n}W_k(x)f\bigl(x^{(k)}\bigr)}{\sum\limits_{k=1}^{n}W_k(x)},$$
+where $x \not = x^{(k)}$, $k = 1$, $\ldots$, $n$, for weighting functions
+$$W_k(x) = \left( \frac{\big(r_k - \bigl\|x - x^{(k)}\bigr\|_2\big)_+}{r_k \bigl\|x - x^{(k)}\bigr\|_2} \right)^2,$$ 
+where $r_k \in \mathbb{R}$ are chosen such that at least $d+1$ other
+points $x^{(j)}$, $j \not = k$ are inside the closed Euclidean ball of
+radius $r_k$ about $x^{(k)}$. The computational complexity of ShepMod
+is $\mathcal{O}(n^2d)$. This paper uses a Fortran 95 implementation of
+ShepMod derived from SHEPPACK \cite{thacker2010algorithm}.
+
+\subsection{Linear Shepard}
+The linear Shepard method (LSHEP) is a blending function using local
+linear interpolants, a special case of the general Shepard algorithm
+\cite{thacker2010algorithm}. The interpolant has the form
+$$ p(x) = \frac{\sum\limits_{k=1}^{n}W_k(x)P_k(x)}{\sum\limits_{k=1}^{n}W_k(x)} ,$$
+where $W_k(x)$ is a locally supported weighting function and $P_k(x)$
+is a local linear approximation to the data satisfying
+$P_k\bigl(x^{(k)}\bigr) = f\bigl(x^{(x)}\bigr)$. The computational
+complexity of LSHEP is $\mathcal{O}(n^2d^3)$. This paper uses the
+Fortran 95 implementation of LSHEP in SHEPPACK
+\cite{thacker2010algorithm}.
+
+\subsection{Box Splines}
+\label{sec:box-splines}
+
+The box spline model used here is an interpolation technique built
+from overlapping box splines \cite{de2013box}. The box splines serve
+as basis functions that are shifted and scaled to have support over
+box shaped regions. The boxes are constructed in a way to guarantee a
+covering of the domain \cite{lux2018novel}. Given a set of box splines
+$\bigl\{b^{x^{(i)}}\bigr\}$ with the iterative box properties outlined
+in \cite{lux2018novel} and anchored at interior points
+$\bigl\{x^{(i)}\bigr\}$,
+$$ \hat f(y) = \frac{\sum\limits_{i=1}^n b^{x^{(i)}}(y) f\bigl(x^{(i)}\bigr)}{\sum\limits_{i=1}^n b^{x^{(i)}}(y)}. $$
+Note that the box splines always satisfy
+$b^{x^{(i)}}\bigl(x^{(j)}\bigr) = \delta_{ij}$ and $b^{x^{(i)}}(y)
+\geq 0$. The computational complexity of interpolation via the box
+spline model is $\mathcal{O}(n^2 d)$.
+
+\subsection{Voronoi}
+\label{sec:voronoi}
+
+A well-studied technique for classification and approximation is the
+nearest neighbor algorithm \cite{cover1967nearest}. Nearest neighbor
+inherently utilizes the convex region $C^{x^{(i)}}$ (Voronoi cell
+\cite{dirichlet1850reduction}) consisting of all points closer to
+$x^{(i)}$ than to any other point $x^{(j)}$. The Voronoi model smooths
+the nearest neighbor approximation by utilizing the Voronoi cells to
+define support via a generic basis function $v: \mathbb{R}^d
+\rightarrow \mathbb{R}_+$ given by
+$$ v^{x^{(i)}}(y) = \left(1 - \frac{\bigl\|y - x^{(i)}\bigr\|_2}{2 \ h\bigl(y - x^{(i)} \mid x^{(i)}\bigr)} \right)_+, $$
+where $h\bigl(w \mid x^{(i)}\bigr)$ is the distance between $x^{(i)}$
+and the boundary of the Voronoi cell $C^{x^{(i)}}$ in the direction
+$w$. $v^{x^{(i)}}\bigl(x^{(j)}\bigr) = \delta_{ij}$ and $v^{x^{(i)}}$
+has local support, giving the interpolated value
+$$ f(y) = \frac{\sum\limits_{i=1}^n v^{x^{(i)}}(y) f\bigl(x^{(i)}\bigr)}{\sum\limits_{i=1}^n v^{x^{(i)}}(y)}, $$
+where $0 \leq v^{x^{(i)}}(y) \leq 1$. The computational complexity of
+interpolation via this Voronoi model is $\mathcal{O}(n^2 d)$. All of
+the approximations are an interpolant involving a convex combination
+of known function values $f\bigl(x^{(i)}\bigr)$.
 
 %% ===================================================================
-\subsection{Iterative Box Mesh}
-\label{sec:max_box_mesh}
-
-The Iterative Box mesh is an interpolation technique that utilizes overlapping box splines \cite{de2013box} as basis functions that are shifted and scaled to have support over box shaped regions.
-The boxes are constructed in a way to guarantee a covering of the domain.
-Given a set of box splines $\bigl\{b^{x^{(i)}}\bigr\}$ with the iterative box properties anchored at interior points $\bigl\{x^{(i)}\bigr\}$,
-$$ \hat f(y) = \frac{\sum\limits_{i=1}^n b^{x^{(i)}}(y) f(x^{(i)})}{\sum\limits_{i=1}^n b^{x^{(i)}}(y)}. $$
-Note that the box splines always satisfy $b^{x^{(i)}}\bigl(x^{(j)}\bigr) = \delta_{ij}$ and $b^{x^{(i)}}(y) \geq 0$.
-The computational complexity of interpolation via the Iterative Box mesh is $\mathcal{O}(n^2 d)$.
-
-%% ===================================================================
-\subsection{Voronoi Mesh}
-\label{sec:voronoi_mesh}
-
-A well-studied technique for classification and approximation is the nearest neighbor algorithm \cite{cover1967nearest}.
-Nearest neighbor inherently utilizes the convex region $C^{x^{(i)}}$ (Voronoi cell \cite{dirichlet1850reduction}) consisting of all points closer to $x^{(i)}$ than to any other point $x^{(j)}$.
-The Voronoi mesh smooths the nearest neighbor approximation by utilizing the Voronoi cells to define support via a generic basis function $v: \mathbb{R}^d \rightarrow \mathbb{R}_+$ given by
-$$ v^{x^{(i)}}(y) = \left(1 - \frac{||y - x^{(i)}||_2}{2 \ h(y - x^{(i)} \mid x^{(i)})} \right)_+, $$
-where $h(w \mid x^{(i)})$ is the distance between $x^{(i)}$ and the boundary of the Voronoi cell $C^{x^{(i)}}$ in the direction $w$.
-$v^{x^{(i)}}\bigl(x^{(j)}\bigr) = \delta_{ij}$ and $v^{x^{(i)}}$ has local support, giving the interpolated value
-$$ f(y) = \frac{\sum\limits_{i=1}^n v^{x^{(i)}}(y) f(x^{(i)})}{\sum\limits_{i=1}^n v^{x^{(i)}}(y)}, $$
-where $0 \leq v^{x^{(i)}}(y) \leq 1$.
-The computational complexity of interpolation via the Voronoi mesh is $\mathcal{O}(n^2 d)$.
-All of the approximations are an interpolant involving a convex combination of known function values $f(x^{(i)})$.
+\section{Measuring Error}
+\label{sec:error}
 
 \begin{figure}[htb]
+  \centering
   \begin{tikzpicture}
-    \node (img)  {\includegraphics[width=0.4\textwidth,]{Figures/plot-delaunay-example-prediction.pdf}};
+    \node (img)  {\includegraphics[width=0.75\textwidth,]{Figures/plot-delaunay-example-prediction.pdf}};
     \node[below=of img, node distance=1cm, yshift=1cm] {I/O Throughput};
@@
   \vspace{-0.3cm}
-  \caption{In this HPC I/O example, the general methodology for predicting a CDF and evaluating error can be seen.
-The Delaunay method chose three source distributions (dotted lines) and assigned weights \{.3, .4, .3\} (top to bottom at arrow).
-The weighted sum of the three known CDFs produces the predicted CDF (dashed line).
-The KS Statistic (arrow) computed between the true CDF (solid line) and predicted CDF (dashed line) is 0.2 for this example.
-The KS test null hypothesis is rejected at $p$-value 0.01, however it is not rejected at $p$-value 0.001.
+  \caption{In this HPC I/O example, the general methodology for
+    predicting a CDF and evaluating error can be seen. The Delaunay
+    method chose three source distributions (dotted lines) and
+    assigned weights \{.3, .4, .3\} (top to bottom at arrow). The
+    weighted sum of the three known CDFs produces the predicted CDF
+    (dashed line). The KS Statistic (arrow) computed between the true
+    CDF (solid line) and predicted CDF (dashed line) is 0.2 for this
+    example. The KS test null hypothesis is rejected at $p$-value
+    0.01, however it is not rejected at $p$-value 0.001.
   \vspace{-.1cm}}
-  \label{fig:prediction_example}
+  \label{fig:prediction-example}
 \end{figure}
 
-%% ===================================================================
-\subsection{Measuring Error}
-
-When the range of an approximation is the real numbers, error is reported with summary statistics including: min absolute error, max absolute error, absolute error quartiles, and mean absolute error.
-In general, all errors are recorded as signed relative error and any figures depicting distributions of errors will present these values.
-
-A hurdle when modeling function-valued outputs such as cumulative distribution functions (CDFs) or probability density functions (PDFs) is that certain properties must be maintained.
-It is necessary that a PDF $f: \mathbb{R} \rightarrow \mathbb{R}$ have the properties $f(x) \geq 0$ and $\int_{-\infty}^{\infty}f(x)dx = 1$.
-However, for a CDF $F: \mathbb{R} \rightarrow \mathbb{R}$ the properties are $F(x) \in [0,1]$ and $F(x)$ is absolutely continuous and nondecreasing.
-This work utilizes the fact that a convex combination of CDFs (or PDFs) results in a valid CDF.
-Given $G(x) = \sum_{i}w_i F_i(x)$, $\sum_{i} w_i = 1$, $w_i \geq 0$, and each $F_i$ is a valid CDF, $G$ must also be a valid CDF.
-A demonstration of how this is applied can be seen in Figure \ref{fig:prediction_example}.
-
-%% %% ===================================================================
-%% \subsection{Feature Weighting}
-%% \label{sec:feature_weighting}
-
-%% It is well-known that an important procedure in any application of predictive methodologies is identifying those features of the data that are most relevant to making accurate predictions \cite{guyon2003introduction}.
-Selection strategies such as the floating searches studied in \cite{pudil1994floating} or others compared in \cite{ferri1994comparative} can be too expensive for large approximation problems.
-Rather, this work poses feature selection as a continuous optimization problem.
-Let $X$ be an $n \times d$ matrix of $n$ known system configurations with $d$ parameters each normalized to be in $[0,1]$.
-Define an error function that computes the error of a predictive model trained on $X\hbox{ diag }w$, $w \in \mathbb{R}^d$,  by performing ten random splits with 80\% of the rows of $X\hbox{ diag }w$ for training and 20\% for testing.
-A minimum of this error function could be considered an optimal weighting of the features of $X$.
-Minimization is performed using a zero order method in the absence of a readily computable gradient.
-
-%% ===================================================================
-
-The performance of approximation techniques that predict probability functions can be analyzed through a variety of summary statistics.
-This work uses the max absolute error, also known as the Kolmogorov-Smirnov (KS) statistic \cite{lilliefors1967kolmogorov} for its compatibility with the KS test.
-
-The two-sample KS test is a useful nonparametric test for comparing two CDFs while only assuming stationarity, finite mean, and finite variance.
-The null hypothesis (that two CDFs come from the same underlying distribution) is rejected at level $p \in [0,1]$ when
+When the range of an approximation is the real numbers, error is
+reported with summary statistics including: min absolute error, max
+absolute error, and absolute error quartiles. When the range of an
+approximation is the space of cumulative distribution functions, the
+Kolmogorov-Smirnov statistic (max-norm difference between the
+functions) is used.
+
+A hurdle when modeling function-valued outputs such as cumulative
+distribution functions (CDFs) or probability density functions (PDFs)
+is that certain properties must be maintained. It is necessary that a
+PDF $f: \mathbb{R} \rightarrow \mathbb{R}$ have the properties $f(x)
+\geq 0$ and $\int_{-\infty}^{\infty}f(x)dx = 1$. However, for a CDF
+$F: \mathbb{R} \rightarrow \mathbb{R}$ the properties are
+$F(x) \in [0,1]$ and $F(x)$ is absolutely continuous and
+nondecreasing. This work utilizes the fact that a convex combination
+of CDFs (or PDFs) results in a valid CDF. Given $G(x) = \sum_{i}w_i
+F_i(x)$, $\sum_{i} w_i = 1$, $w_i \geq 0$, and each $F_i$ is a valid
+CDF, $G$ must also be a valid CDF. A demonstration of how this is
+applied can be seen in Figure \ref{fig:prediction-example}.
+
+The performance of approximation techniques that predict probability
+functions can be analyzed through a variety of summary
+statistics. This work uses the max absolute difference, also known as
+the Kolmogorov-Smirnov (KS) statistic \cite{lilliefors1967kolmogorov}
+for its compatibility with the KS test.
+
+The two-sample KS test is a useful nonparametric test for comparing
+two CDFs while only assuming stationarity, finite mean, and finite
+variance. The null hypothesis (that two CDFs come from the same
+underlying distribution) is rejected at level $p \in [0,1]$ when
 $$ KS > \sqrt{-\frac{1}{2}\ln\biggl(\frac{p}{2}\biggr)} \sqrt{\frac{1}{n_1} + \frac{1}{n_2}}, $$
-with distribution sample sizes $n_1,n_2 \in \mathcal{N}$.
-For all applications of the KS test presented in this work $n_1 = n_2$.
-An example of the round-trip prediction methodology from known and predicted distributions to the calculation of error can be seen in Figure \ref{fig:prediction_example}.
+with distribution sample sizes $n_1,n_2 \in \mathcal{N}$. For all
+applications of the KS test presented in this work $n_1 = n_2$. An
+example of the round-trip prediction methodology from known and
+predicted distributions to the calculation of error can be seen in
+Figure \ref{fig:prediction-example}.
 
 \section{Theoretical Error Bound}
-
-This section presents the theoretical results bounding the error of Delaunay interpolation.
-The error analysis relies on the Delaunay interpolation for two reasons: (1) second order results can be obtained utilizing a Lipschitz constant on the gradient of a function, rather than standard Lipschitz bounds and (2) multiple other interpolants analyzed compute predictions as convex combinations of observed function values, which may allow for straight forward extensions of this error bound.
-
-$ $
-
-\noindent \textbf{\textit{Lemma 1.}} Let $S \subset \mathbb{R}^d$ be open and convex, $f: \mathbb{R}^d \rightarrow \mathbb{R},$ and $\nabla f \in Lip_{(\gamma,\|\cdot\|_2)}(S)$, the set of $\gamma$-Lipschitz continuous functions in the $2$-norm.
-Then for all $x,y \in S$
-%
+\label{sec:theory}
+
+This section presents the theoretical results bounding the error of
+simplicial interpolation. The error analysis relies on simplicial
+interpolation for three reasons: (1) second order results can be
+obtained utilizing a Lipschitz constant on the gradient of a function,
+rather than standard Lipschitz bounds; (2) the results directly
+apply to Delaunay; and (3) multiple other interpolants in this paper
+compute predictions as convex combinations of observed function
+values, which may allow for straight forward extensions of this error
+bound.
+
+\begin{plainlemma}
+  \label{lemma:1}
+  Let $S \subset \mathbb{R}^d$ be open and convex, $f: \mathbb{R}^d
+  \rightarrow \mathbb{R}$, and $\nabla f \in
+  Lip_{(\gamma,\|\cdot\|_2)}(S)$, the set of $\gamma$-Lipschitz
+  continuous functions in the $2$-norm. Then for all $x,y \in S$
   $$\big|f(y) - f(x) - \langle \nabla f(x), y - x \rangle \big| \leq \frac{\gamma \|y - x\|_2^2}{2}.$$
-
-\noindent \textit{Proof.} Consider the function $g(t) = f \big((1-t) x + t y \big),$ $0 \leq t \leq 1,$ whose derivative $g'(t) = \big\langle \nabla f \big((1-t) x + t y \big), y - x \big\rangle$ is the directional derivative of $f$ in the direction $(y - x).$
-%
+\end{plainlemma}
+
+\begin{proofdot}
+  Consider the function $g(t) = f \big((1-t) x + t y \big)$, $0 \leq t
+  \leq 1$, whose derivative $g'(t) = \big\langle \nabla f \big((1-t) x
+  + t y \big), y - x \big\rangle$ is the directional derivative of $f$
+  in the direction $(y - x).$
   \begin{align*}
@@
   \end{align*}
-%
   \qed
-
-\noindent \textbf{\textit{Lemma 2.}} Let $x, y, v_i \in \mathbb{R}^d,$ $c_i \in \mathbb{R},$ and $|\langle y - x, v_i \rangle| \leq c_i$ for $i = 1,$ $\ldots,$ $d.$ If $M = (v_1,$ $\ldots,$ $v_d)$ is nonsingular, then
-%
+\end{proofdot}
+
+\begin{plainlemma}
+  \label{lemma:2}
+  Let $x, y, v_i \in \mathbb{R}^d$, $c_i \in \mathbb{R}$, and
+  $|\langle y - x, v_i \rangle| \leq c_i$ for $i = 1$, $\ldots$, $d.$
+  If $M = (v_1$, $\ldots$, $v_d)$ is nonsingular, then
   $$\|y - x\|_2^2 \leq \frac{1}{\sigma_d^2} \sum_{i=1}^d c_i^2,$$
-%
   where $\sigma_d$ is the smallest singular value of $M.$
-
-\noindent \textit{Proof.} Using the facts that $M$ and $M^t$ have the same singular values, and $\|M^tw\|_2 \geq \sigma_d \|w\|_2,$ gives
-%
+\end{plainlemma}
+
+\begin{proofdot}
+  Using the facts that $M$ and $M^t$ have the same singular values,
+  and $\|M^tw\|_2 \geq \sigma_d \|w\|_2$, gives
   \begin{align*}
@@
   \end{align*}
-%
   \qed
-
-\noindent \textbf{\textit{Lemma 3.}} Given $f,$ $\gamma,$ $S$ as in {\it Lemma 1}, let $X = \{x_0,$ $x_1,$ $\ldots,$ $x_d\}$ $\subset S$ be the vertices of a $d$-simplex, and let $\hat f(x) = \langle c, x - x_0 \rangle + f(x_0),$ $c \in \mathbb{R}^d$ be the linear function interpolating $f$ on $X.$ Let $\sigma_d$ be the smallest singular value of the matrix $M = (x_1 - x_0,$ $\ldots,$ $x_d - x_0),$ and $k = \max\limits_{1\ \leq\ j\ \leq\ d} \|x_j - x_0\|_2.$ Then
-%
+\end{proofdot}
+
+\begin{plainlemma}
+  \label{lemma:3}
+  Given $f$, $\gamma$, $S$ as in {\it Lemma \ref{lemma:1}}, let $X =
+  \{x_0$, $x_1$, $\ldots$, $x_d\}$ $\subset S$ be the vertices of a
+  $d$-simplex, and let $\hat f(x) = \langle c, x - x_0 \rangle +
+  f(x_0)$, $c \in \mathbb{R}^d$ be the linear function interpolating
+  $f$ on $X.$ Let $\sigma_d$ be the smallest singular value of the
+  matrix $M = (x_1 - x_0$, $\ldots$, $x_d - x_0)$, and $k =
+  \max\limits_{1\ \leq\ j\ \leq\ d} \|x_j - x_0\|_2.$ Then
   $$\big\|\nabla f(x_0) - c\big\|_2 \leq \sqrt{d} \frac{\gamma k^2}{\sigma_d}.$$
-
-\noindent \textit{Proof.} Consider $f(x) - \hat f(x)$ along the line segment $z(t) = (1-t)x_0 + t x_j,$ $0 \leq t \leq 1.$ By Rolle's Theorem, for some $0 < \hat t < 1,$ $\big\langle \nabla f\big(z(\ \hat t\ )\big) - c, x_j - x_0 \big\rangle = 0.$ Now
-%
+\end{plainlemma}
+
+\begin{proofdot}
+  Consider $f(x) - \hat f(x)$ along the line segment $z(t) = (1-t)x_0
+  + t x_j$, $0 \leq t \leq 1.$ By Rolle's Theorem, for some $0 < \hat
+  t < 1$, $\big\langle \nabla f\big(z(\ \hat t\ )\big) - c, x_j - x_0
+  \big\rangle = 0.$ Now
   \begin{align*}
@@
   \end{align*}
-%
-for all $1 \leq j \leq d.$ Using {\it Lemma 2},
-%
-$$\big \| \nabla f(x_i) - c \big \|_2^2 \leq \frac{d}{\sigma_d^2} \big( \gamma k^2\big)^2 \implies \big \| \nabla f(x_i) - c \big \|_2 \leq \sqrt{d} \frac{\gamma k^2}{\sigma_d}.$$
-%
-\qed
-
-\noindent \textbf{\textit{Theorem.}} Under the assumptions of {\it Lemma 1} and {\it Lemma 3}, for $z \in S,$
-%
+  for all $1 \leq j \leq d.$ Using {\it Lemma \ref{lemma:2}},
+  $$\big \| \nabla f(x_i) - c \big \|_2^2 \leq \frac{d}{\sigma_d^2}
+  \big( \gamma k^2\big)^2 \implies \big \| \nabla f(x_i) - c \big \|_2
+  \leq \sqrt{d} \frac{\gamma k^2}{\sigma_d}.$$ \qed
+\end{proofdot}
+
+\begin{onetheorem}
+  Under the assumptions of {\it Lemma \ref{lemma:1}} and {\it Lemma \ref{lemma:3}}, for $z \in S$,
   $$ \big|f(z) - \hat f(z)\big| \leq \frac{\gamma \|z - x_0\|_2^2}{2} + \sqrt{d} \frac{\gamma k^2}{\sigma_d} \|z - x_0\|_2.$$
-
-\noindent \textit{Proof.} Let $v = \nabla f(x_0) - c,$ where $\|v\|_2 \leq \sqrt{d} \gamma k^2 / \sigma_d$ by {\it Lemma 3}. Now
-%
+\end{onetheorem}
+
+\begin{proofdot}
+  Let $v = \nabla f(x_0) - c$, where $\|v\|_2 \leq \sqrt{d} \gamma k^2 / \sigma_d$ by {\it Lemma \ref{lemma:3}}. Now
   \begin{align*}
@@
   \end{align*}
-%
-where the last inequality follows from {\it Lemma 1}.
+  where the last inequality follows from {\it Lemma \ref{lemma:1}}.
   %
   \qed
-
-\section{Planning for Paper}
-
-\begin{itemize}
-\item Construct analytic function to test theoretical claims made about Delaunay.
-\item Demonstrate the empirical performance of Delaunay, discuss alignment with theory.
-\item Demonstrate similar empirical performance is achieved by other interpolants.
-\item Argue that interpolation is not as commonly applied, but that it works well in practice.
-\item Present the data for predicting real-valued outputs, an empirical study.
-  \begin{itemize}
-  \item Forest fire data
-  \item Parkinsons
-  \item https://www.kaggle.com/jsphyg/weather-dataset-rattle-package
-  \end{itemize}
-\item Present prediction results on each of these data sets for each algorithm.
-\item Discuss the results between the regressors and interpolants.
-\item Present the data for predicting function-valued outputs, an empirical study.
-  \begin{itemize}
-  \item VarSys large scale system configuration prediction
-  \item \textit{galaxy data from Carnegie Melon?}
-  \end{itemize}
-\item Present prediction results on each of these data sets for each algorithm.
-\item Discuss the results between the regressors and interpolants.
-\item Present a summary table showing the performance differences between regerssors and interpolants.
-\item Discuss the overall applicability of interpolants and encourage further application. Restate the benefits of interpolants
-\item Conclude
-\end{itemize}
-
-\section{Data and Analysis}
-In order to evaluate the proposed interpolation and approximation techniques, this paper utilizes three data sets of varying dimension and application.
-In the following subsections the sources and targets of each data set are described, as well as challenges and limitations related to interpolating and approximating these data sets.
-The distributions of response values being modeled can be seen in Figure \ref{fig_response_hists}.
-The preprocessing and approximation processes are described in Section \ref{sec_performance_analysis}.
+\end{proofdot}
+
+In summary, the approximation error of a simplicial interpolant tends
+quadratically towards zero when approaching observed data only when
+the diameter of the simplex is also reduced at a proportional rate.
+Only linear convergence to the true function can be achieved in
+practice, without the incorporation of additional observations. Notice
+that the approximation error is largely determined by the spacing of
+observed data. Predictions made by simplices whose vertices are not
+well-spaced (i.e. have large diameter, or are nearly contained in a
+hyperplane) have higher error. In light of this error bound, an
+empirical evaluation of presented algorithms follows.
+
+%% ===================================================================
+\section{Data and Empirical Analysis}
+\label{sec:data}
+
+The theoretical results constructed in Section \ref{sec:theory} for
+simplicial interpolation are promising, however they are difficult to
+interpret in context with approximation algorithms that do not have
+similar known uniform error bounds. For that reason, this paper
+utilizes five different data sets of varying dimension and application
+to construct approximations and compare the accuracy of different
+techniques.
+
+In the following five subsections the sources and targets of each test
+data set are described, as well as challenges and limitations related
+to approximating these data. The distribution of response values being
+modeled is presented followed by the distribution of approximation
+errors for each algorithm.
+
+All five data sets are rescaled such that the domain of approximation
+is the unit hypercube. The range of the first four data sets is the
+real numbers, while the range of the fifth data set is the space of
+cumulative distribution functions. All approximation techniques are
+applied to the first four data sets, while only those interpolants
+whose approximations are convex combinations of observed data are
+applied to the final data set.
+
+All approximations are constructed using $k$-fold cross validation as
+described in \cite{kohavi1995study} with $k=10$. This approach
+randomly partitions data into $k$ nearly equal sized sets. Each
+algorithm is then evaluated by constructing an approximation over each
+unique union of $k-1$ elements of the partition, making predictions
+for points in the remaining element. As a result, each observed data
+point is used in the construction of $k-1$ different approximations
+and is approximated exactly once. The $k$-fold cross validation method
+is data-efficient and provides an unbiased estimate of the expected
+prediction error \cite{kohavi1995study}, however it should be noted
+that neither this method nor others can provide a universally unbiased
+estimator for the variance of prediction error \cite{bengio2004no}.
+
+In addition to the figures displaying approximation results for each
+data set, tables of accompanying numerical results are located in the
+Appendix. All of the test data sets capture underlying functions that
+are almost certainly stochastic. As described in Section
+\ref{sec:introduction}, regression techniques appear most appropriate
+for these problems. However, in high dimension data grows
+exponentially more sparse. Given sparse data, regressors tend towards
+interpolation.
+
+%% -------------------------------------------------------------------
+\subsection{Forest Fire ($n = 504, d = 12$)}
 
 \begin{figure}
-  \includegraphics[width=.5\textwidth,height=2cm]{Figures/p-hist.pdf}
-  \includegraphics[width=.5\textwidth,height=2cm]{Figures/f-hist.pdf}
-  \includegraphics[width=.5\textwidth,height=2cm]{Figures/h-hist.pdf}
-  \caption{Histograms of Parkinsons (total UPDRS), forest fire (area), and HPC I/O (mean throughput) response values respectively.
-Notice that both the forest fire and HPC I/O data sets are heavily skewed.
-  \vspace{-.5cm}}
-  \label{fig_response_hists}
-\end{figure}
-
-\subsection{Data Summary}
-
-\subsubsection{High Performance Computing I/O ($n = 532, d = 4$)}
-The first of three data sets is a four-dimensional data set produced by executing the IOzone benchmark from \cite{iozone} on a homogeneous cluster of computers.
-The system performance data was collected by executing IOzone 40 times for each of a select set of system configurations.
-A single IOzone execution reports the max I/O file-read throughput seen.
-The 40 executions for each system configuration are converted to their mean, which is capable of being modeled by each of the multivariate approximation techniques presented in Section \ref{sec_mesh_construction}.
-The four dimensions being modeled to predict throughput mean are file size, record size, thread count, and CPU frequency.
-
-\begin{figure*}[htb]
+  \centering
   \begin{tikzpicture}
-    \node (img)  {\includegraphics[width=0.95\textwidth,height=4cm]{Figures/eval-times.pdf}};
-    \node[below=of img, node distance=0cm, yshift=1cm] {Relative Error Tolerance while Bootstrapping};
-    \node[left=of img, node distance=0cm, rotate=90, anchor=center,yshift=-0.7cm] {Fit Time};
-  \end{tikzpicture}
-  \caption{Time required to generate model fits for each technique with varying relative error tolerance during bootstrapping.
-    \vspace{-.3cm}}
-  \label{fig_eval_times}
-\end{figure*}
-
-\subsubsection{Forest Fire ($n = 517, d = 12$)}
-The forest fire data set \cite{cortez2007data} describes the area of Montesinho park burned on specific days and months of the year in terms of the environmental conditions.
-The twelve dimensions being used to model burn area are the $x$ and $y$ spatial coordinates of burns in the park, month and day of year, the FFMC, DMC, DC, and ISI indices (see source for details), the temperature in Celsius, relative humidity, wind speed, and outdoor rain.
-The original analysis of this data set demonstrated it to be difficult to model, likely due to the skew in response values.
-
-\subsubsection{Parkinson's Telemonitoring ($n = 468, d = 16$)}
-The final data set for evaluation \cite{tsanas2010accurate} is derived from a speech monitoring study with the intent to automatically estimate Parkinson's disease symptom development in Parkinson's patients.
-The function to be predicted is a time-consuming clinical evaluation measure referred to as the UPDRS score.
-The total UPDRS score given by a clinical evaluation is estimated through 16 real numbers generated from biomedical voice measures of in-home sound recordings.
-
-\subsection{Performance Analysis}
-\label{sec_performance_analysis}
-
-The performance of the approximation techniques varies considerably across the three evaluation data sets.
-Relative errors for the most na\"{\i}ve approximators such as nearest neighbor can range from zero to $\displaystyle \big(\max_x f(x) - \min_x f(x)\big) / \min_x f(x)$ when modeling a positive function $f(x)$ from data.
-Each of the approximation techniques presented remain within these bounds and all errors are presented in signed relative form $(\hat f(x) - f(x)) / f(x)$.
-Before the models are constructed all data values (components $x^{(i)}_r$ of $x^{(i)} \in X$) are shifted and scaled to be in the unit cube $[0,1]^d$, while the response values are taken in their original form.
-All models are evaluated with $10$ random $80/20$ splits of the data.
-
-\begin{table}
-  \centering
-  \begin{tabular}{c|c|c|c}
-    \hline
-    \textbf{Data Set} & \textbf{Technique} & \textbf{Tolerance} & \textbf{Average Error}\\
-    \hline
-    HPC I/O & MBM & 1.2 & 0.597\\
-    Forest Fire & MBM & 1.8 & 3.517\\
-    Parkinson's & MBM & 0.6 & 0.114\\
-    \hline
-    HPC I/O & IBM & 0.4 & 0.419\\
-    Forest Fire & IBM & 1.8 & 3.615\\
-    Parkinson's & IBM & 1.8 & 0.121\\
-    \hline
-    HPC I/O & VM & 0.2 & 0.382\\
-    Forest Fire & VM & 1.0 & 4.783\\
-    Parkinson's & VM & 2.0 & 1.824\\
-    \hline
-  \end{tabular}
-  \caption{The optimal error tolerance bootstrapping parameters for each technique and each data set as well as the average absolute relative errors achieved by that tolerance.
-Notice that large relative error tolerances occasionally yield even lower evaluation errors, demonstrating the benefits of approximation over interpolation for noisy data sets.
-  \vspace{-.5cm}}
-  \label{tab_optimal_tolerance}
-\end{table}
-
-Each of the approximation techniques presented incorporates bootstrapping based on an allowable error tolerance $t$.
-An analysis of the effects of bootstrapping error tolerances on validation accuracy can be seen in Figure \ref{fig_all_performance}.
-The approximation meshes perform best on the forest fire and Parkinson's data sets when the error tolerance used for fitting is large (smoothing rather than interpolating), while near-interpolation generally produces the most accurate models for HPC I/O.
-Another performance result of note is that the $MBM$ and $IBM$ have very similar basis functions with largely different outputs.
-
-\begin{figure*}
-  \begin{tikzpicture}
-    \node (img)  {\includegraphics[width=0.95\textwidth,height=9.5cm]{Figures/all-performance.pdf}};
-    \node[below=of img, node distance=0cm, yshift=1cm] {Relative Error Tolerance while Bootstrapping};
-    \node[left=of img, node distance=0cm, rotate=90, anchor=center,yshift=-0.7cm] {Signed Relative Error};
-  \end{tikzpicture}
-  \caption{The performance of all three techniques with varied relative error tolerance for the bootstrapping parameter.
-The columns are for Max Box Mesh, Iterative Box Mesh, and Voronoi Mesh, respectively.
-The rows are for HPC I/O, Forest Fire, and Parkinson's respectively.
-Notice the techniques' behavior on the Parkinson's and Forest Fire data sets, performance increases with larger error tolerance.}
-  \label{fig_all_performance}
-\end{figure*}
-
-\begin{figure*}
-  \begin{tikzpicture}
-    \node (img)  {\includegraphics[width=0.95\textwidth,height=9.5cm]{Figures/perf-sample.pdf}};
-    \node[below=of img, node distance=0cm, yshift=1cm] {Signed Relative Error in Prediction};
+    \node (img)  {\includegraphics[width=0.75\textwidth,]{Figures/raw-histogram-forest-fire.pdf}};
+    \node[below=of img, node distance=1cm, yshift=1cm] {Forest First Area Burned};
     \node[left=of img, node distance=0cm, rotate=90, anchor=center,yshift=-0.7cm] {Count};
   \end{tikzpicture}
-  \caption{A sample of relative errors for all three techniques with optimal selections of error tolerance.
-The columns are for Max Box Mesh, Iterative Box Mesh, and Voronoi Mesh, respectively.
-The rows are for HPC I/O, Forest Fire, and Parkinson's respectively.
-    \vspace{-.3cm}}
-  \label{fig_perf_sample}
-\end{figure*}
-
-The selection of bootstrapping error tolerance also effects the computation time required to fit each of the models to data.
-Figure \ref{fig_eval_times} presents the time required to construct approximations for each model and each data set with varying $t$.
-The rapid reduction in computation time required for the forest fire and HPC I/O data sets suggests that large reductions in error can be achieved with relatively few basis functions.
-The Parkinson's data set however presents a more noisy response, with increasing number of basis functions reducing error much less quickly.
-
-The distributions of errors experienced by each approximation technique when the optimal bootstrapping relative error tolerance is selected can be seen in Figure \ref{fig_perf_sample}.
-HPC I/O exhibits the most normal approximation errors, which suggests that the models are converging on the random noise of the response for the data set.
-The worst relative approximation errors are produced by the Voronoi mesh on the forest fire data set.
-The small magnitude true response values contribute to the larger relative errors.
-Regardless, the $VM$ errors are unacceptably large.
+  \caption{Above is a histogram of forest fire area burned under
+    recorded weather conditions. The data is presented on a natural
+    log scale because most values are small with exponentially fewer
+    fires on record that burn large areas.}
+  \label{fig:hist-forest-fire}
+\end{figure}
+
+\begin{figure}
+  \centering
+  \begin{tikzpicture}
+    \node (img)  {\includegraphics[width=0.8\textwidth,]{Figures/error-box-forest-fire.pdf}};
+    \node[below=of img, node distance=1cm, yshift=1cm] {Area Burned Error};
+  \end{tikzpicture}
+  \caption{All models are applied to approximate the amount of area
+    that would be burned given environment conditions. $10$-fold
+    cross validation as described in the beginning of Section
+    \ref{sec:data} is used to evaluated each algorithm. This results
+    in exactly one prediction from each algorithm for each data
+    point. These boxes depict the median (middle bar), median $95\%$
+    confidence interval (cones), quartiles (box edges), fences at
+    $3/2$ interquartile range (whiskers), and outliers (dots) of
+    absolute prediction error for each model. Similar to figure
+    \ref{fig:hist-forest-fire}, the errors are presented on a natural
+    log scale. The complementing numerical data to this figure is
+    provided in Table \ref{table:error-forest-fire} in the Appendix.}
+  \label{fig:error-forest-fire}
+\end{figure}
+
+The forest fire data set \cite{cortez2007data} describes the area of
+Montesinho park burned over months of the year along with
+environmental conditions. The twelve dimensions being used to model
+burn area are the $x$ and $y$ spatial coordinates of burns in the
+park, month of year (mapped to a 2D circle), the FFMC, DMC, DC, and
+ISI indices (see source for details), the temperature, relative
+humidity, wind speed, and outdoor rain. The original analysis of this
+data set demonstrated it to be difficult to model, likely due to the
+skew in response values.
+
+As suggested by Figure \ref{fig:error-forest-fire}, the SVR has the
+lowest absolute prediction errors for 80\% of the data, with ShepMod
+and BoxSpline being the nearest overall competitors. The effectiveness
+of SVR on this data suggests the underlying function can be defined by
+relatively few parameters, as well as the importance of capturing the
+low-burn-area data points.
+%% -------------------------------------------------------------------
+
+%% -------------------------------------------------------------------
+\subsection{Parkinson's Telemonitoring ($n = 5875, d = 19$)}
+
+\begin{figure}
+  \centering
+  \begin{tikzpicture}
+    \node (img)  {\includegraphics[width=0.75\textwidth,]{Figures/raw-histogram-parkinsons.pdf}};
+    \node[below=of img, node distance=1cm, yshift=1cm] {Parkinson's Total UPDRS Score};
+    \node[left=of img, node distance=0cm, rotate=90, anchor=center,yshift=-0.7cm] {Count};
+  \end{tikzpicture}
+  \caption{Above is a histogram of the Parkinson's patient total UPDRS
+    clinical scores that will be approximated by each algorithm.}
+  \label{fig:hist-parkinsons}
+\end{figure}
+
+\begin{figure}
+  \centering
+  \begin{tikzpicture}
+    \node (img)  {\includegraphics[width=0.8\textwidth,]{Figures/error-box-parkinsons.pdf}};
+    \node[below=of img, node distance=1cm, yshift=1cm] {Total UPDRS Score Error};
+  \end{tikzpicture}
+  \caption{All models are applied to approximate the total UPDRS score
+    given audio features from patients' home life. $10$-fold
+    cross validation as described in the beginning of Section
+    \ref{sec:data} is used to evaluated each algorithm. This results
+    in exactly one prediction from each algorithm for each data
+    point. These boxes depict the median (middle bar), median $95\%$
+    confidence interval (cones), quartiles (box edges), fences at
+    $3/2$ interquartile range (whiskers), and outliers (dots) of
+    absolute prediction error for each model. The complementing
+    numerical data to this is provided in table
+    \ref{table:error-parkinsons} in the Appendix.}
+  \label{fig:error-parkinsons}
+\end{figure}
+
+The second data set for evaluation \cite{tsanas2010accurate} is
+derived from a speech monitoring study with the intent to
+automatically estimate Parkinson's disease symptom development in
+Parkinson's patients. The function to be approximated is a
+time-consuming clinical evaluation measure referred to as the UPDRS
+score. The total UPDRS score given by a clinical evaluation is
+estimated through 16 real numbers generated from biomedical voice
+measures of in-home sound recordings.
+
+Figure \ref{fig:error-parkinsons} shows the ShepMod algorithm has the
+lowest minimum, first quartile, and median of absolute errors for this
+problem, while providing the best prediction 66\% of the time. The MLP
+has the lowest third quartile and provides the best prediction for
+32\% of approximations. The dominance of ShepMod may be in part to
+regular-interval total UPDRS scores provided by clinicians, favoring a
+nearest-neighbor strategy of prediction.
+%% -------------------------------------------------------------------
+
+%% -------------------------------------------------------------------
+\subsection{Australian Daily Rainfall Volume ($n = 2609, d = 23$)}
+
+\begin{figure}
+  \centering
+  \begin{tikzpicture}
+    \node (img)  {\includegraphics[width=0.75\textwidth,]{Figures/raw-histogram-weather.pdf}};
+    \node[below=of img, node distance=1cm, yshift=1cm] {Daily Rainfall in Sydney};
+    \node[left=of img, node distance=0cm, rotate=90, anchor=center,yshift=-0.7cm] {Count};
+  \end{tikzpicture}
+  \caption{Above is a histogram of daily rainfall in Sydney,
+    Australia. The data is presented on a natural log scale because
+    the frequency of larger amounts of rainfall is significantly
+    less. There is a peak in occurrence of the value $0$, which has a
+    notable effect on the resulting model performance.}
+  \label{fig:hist-weather}
+\end{figure}
+
+\begin{figure}
+  \centering
+  \begin{tikzpicture}
+    \node (img)  {\includegraphics[width=0.8\textwidth,]{Figures/error-box-weather.pdf}};
+    \node[below=of img, node distance=1cm, yshift=1cm] {Sydney Rainfall Tomorrow Error};
+  \end{tikzpicture}
+  \caption{All models are applied to approximate the amount of
+    rainfall expected on the next calendar day given various sources
+    of local meteorological data. $10$-fold cross validation as
+    described in the beginning of Section \ref{sec:data} is used to
+    evaluated each algorithm. This results in exactly one prediction
+    from each algorithm for each data point. These boxes depict the
+    median (middle bar), median $95\%$ confidence interval (cones),
+    quartiles (box edges), fences at $3/2$ interquartile range
+    (whiskers), and outliers (dots) of absolute prediction error for
+    each model. The errors are presented on a natural log scale,
+    mimicking the presentation in Figure \ref{fig:hist-weather}. The
+    complementing numerical data to this figure is provided in table
+    \ref{table:error-weather} in the Appendix.}
+  \label{fig:error-weather}
+\end{figure}
+
+The third data regarding the total daily rainfall in Sydney, Australia
+\cite{williams2009rattle} provides a slightly higher dimension
+challenge for the interpolants and regressors. This data is composed
+of many meteorological readings from the region on a day including:
+min and max temperatures, sunshine, wind speed directions (converted
+to coordinates on circle), wind speeds, and humidities throughout the
+day, day of the year (converted to coordinates on circle), and the
+model must predict the amount of rainfall tomorrow.
+
+While Figure \ref{fig:hist-weather} makes MARS look far better than
+other techniques, it only provides the best prediction for 11\% of
+points. The MLP has the lowest absolute error for 56\% of points and
+LSHEP is best for 28\%. MARS likely achieves such a low first quartile
+due to the high occurrence of the value 0 in the data.
+%% -------------------------------------------------------------------
+
+%% -------------------------------------------------------------------
+\subsection{Credit Card Transaction Amount ($n = 5562, d = 28$)}
+
+\begin{figure}
+  \centering
+  \begin{tikzpicture}
+    \node (img)  {\includegraphics[width=0.75\textwidth,]{Figures/raw-histogram-credit-card.pdf}};
+    \node[below=of img, node distance=1cm, yshift=1cm] {Credit Card Transaction Amount};
+    \node[left=of img, node distance=0cm, rotate=90, anchor=center,yshift=-0.7cm] {Count};
+  \end{tikzpicture}
+  \caption{Above is a histogram of credit card transaction
+    amounts. The data is presented on a natural log scale. The data
+    contains a notable frequency peak around $\$1$ transactions. Fewer
+    large purchases are made, but some large purchases are in excess
+    of five orders of magnitude greater than the smallest purchases.}
+  \label{fig:hist-credit-card}
+\end{figure}
+
+\begin{figure}
+  \centering
+  \begin{tikzpicture}
+    \node (img)  {\includegraphics[width=0.8\textwidth,]{Figures/error-box-credit-card.pdf}};
+    \node[below=of img, node distance=1cm, yshift=1cm] {Transaction Amount Error};
+  \end{tikzpicture}
+  \caption{All models are applied to approximate the expected
+    transaction amount given transformed (and obfuscated) vendor and
+    customer-descriptive features. $10$-fold cross validation as
+    described in the beginning of Section \ref{sec:data} is used to
+    evaluated each algorithm. This results in exactly one prediction
+    from each algorithm for each data point. These boxes depict the
+    median (middle bar), median $95\%$ confidence interval (cones),
+    quartiles (box edges), fences at $3/2$ interquartile range
+    (whiskers), and outliers (dots) of absolute prediction error for
+    each model. The absolute errors in transaction amount predictions
+    are presented on a natural log scale, just as in figure
+    \ref{fig:hist-credit-card}. The complementing numerical data to
+    this figure is provided in Table \ref{table:error-credit-card} in
+    the Appendix.}
+  \label{fig:error-credit-card}
+\end{figure}
+
+The fourth test data set, and the final with a real-valued range, is a
+collection of credit card transactions
+\cite{pozzolo2015calibrating}. The provided data carries no direct
+real-world meaning, it is the output of a principle component analysis
+on the original hidden source data. This obfuscation is done to
+protect the information of the credit card users. This data has the
+largest dimension of all considered, at 28. A model over this data
+predicts the transaction amount given the vector of principle
+component information.
+
+As can be seen in Figure \ref{fig:error-credit-card}, the MLP
+outperforms all other algorithms at the first, second, third, and
+fourth quartiles. The MLP produces the lowest absolute error
+prediction for 80\% of transactions, Delaunay bests the remaining
+20\%. It is likely that with less data, Delaunay would be the best
+performer.
+%% -------------------------------------------------------------------
+
+%% -------------------------------------------------------------------
+\subsection{High Performance Computing I/O ($n = 3016, d = 4$)}
+
+\begin{figure}
+  \centering
+  \begin{tikzpicture}
+    \node (img)  {\includegraphics[width=0.75\textwidth,]{Figures/raw-histogram-throughput.pdf}};
+    \node[below=of img, node distance=1cm, yshift=1cm] {I/O Read Throughtput};
+    \node[left=of img, node distance=0cm, rotate=90, anchor=center,yshift=-0.7cm] {Count};
+  \end{tikzpicture}
+  \caption{Histogram of the raw throughput values recorded during all
+    IOzone tests across all system configurations. The distribution is
+    skewed right, with few tests having significantly higher
+    throughput than most others. The data is presented on a natural
+    log scale, however the resulting left-skew reveals the data is not
+    precisely distributed according to an exponential.}
+  \label{fig:hist-throughput}
+\end{figure}
+
+\begin{figure}
+  \centering
+  \begin{tikzpicture}
+    \node (img)  {\includegraphics[width=0.8\textwidth,]{Figures/error-box-throughput.pdf}};
+    \node[below=of img, node distance=1cm, yshift=1cm] {Read Throughtput KS Statistic};
+  \end{tikzpicture}
+  \caption{Above, the models directly capable of predicting
+    distributions are applied to predicting the expected CDF of I/O
+    throughput at previously unseen system configurations. These
+    boxes depict the median (middle bar), median $95\%$ confidence
+    interval (cones), quartiles (box edges), fences at $3/2$
+    interquartile range (whiskers), and outliers (dots) of KS
+    statistic between the observed distribution at each system
+    configuration and the predicted distribution. The above figure is
+    \textit{not} log-scaled like Figure \ref{fig:hist-throughput}. The
+    complimentary numerical data for this figure is provided in
+    Appendix Table \ref{table:error-throughput}.}
+  \label{fig:error-throughput}
+\end{figure}
+
+The final of five data sets is derived from \cite{cameron2019moana},
+which provides four-dimensional distribution data by executing the
+IOzone benchmark \cite{iozone} on a computer system and varying the
+system's file size, record size, thread count, and CPU frequency. At
+each configuration, IOzone samples the I/O file-read throughput (in
+bytes per second) 150 times. Empirical distribution function points
+are computed from each set of 150 executions, which are interpolated
+with a piecewise cubic Hermite interpolating polynomial
+\cite{fritsch1980monotone} to approximate the CDF. All interpolation
+algorithms with the exception of LSHEP are used to approximate these
+CDFs from system configurations.
+
+Delaunay achieves the lowest KS statistic for 62\% of approximations,
+while Voronoi is best for the remaining 38\%. Figure
+\ref{fig:error-throughput} shows that while Delaunay may have more
+best predictions, the behavior of Voronoi may be preferable.
+%% -------------------------------------------------------------------
+
+%% -------------------------------------------------------------------
+\begin{figure}
+  \centering
+  \begin{tikzpicture}
+    \node (img)  {\includegraphics[width=0.8\textwidth]{Figures/plot-throughput-ks.pdf}};
+    \node[below=of img, node distance=1cm, yshift=1cm] {KS Statistic for Predicted vs. Actual};
+    \node[left=of img, node distance=0cm, rotate=90, anchor=center,yshift=-0.7cm] {Count of KS Statistic};
+  \end{tikzpicture}
+  \caption{Histograms of the prediction error for each interpolant
+    that produces predictions as convex combinations of observed
+    data. $10$-fold cross validation as described in the beginning of
+    Section \ref{sec:data} is used to evaluated each algorithm. This
+    results in exactly one prediction from each algorithm for each
+    data point. The distributions show the KS statistics for the
+    predicted throughput distribution versus the actual throughput
+    distribution. The four vertical red lines represent commonly used
+    $p$-values \{0.05, 0.01, 0.001, 1.0e-6\} respectively. All
+    predictions to the right of a red line represent CDF predictions
+    that are significantly different (by respective $p$-value) from
+    the actual distribution according to the KS test. The numerical
+    counterpart to this figure is presented in table
+    \ref{table:null-hypothesis-results}.}
+  \label{fig:throughput-prediction}
+\end{figure}
+
+\begin{table}
+  \renewcommand{\arraystretch}{1.3}
+  \centering
+  \begin{tabular}{c|c|c|c|c}
+     & $p$ = .05 & $p$ = .01 & $p$ = .001 & $p$ = 1.0e-6\\
+    \hline
+    Delaunay & $\mathbf{50.3}\%$ & $\mathit{43.5}\%$ & $\mathit{36.2}\%$ & $\mathit{24.7}\%$\\
+    ShepMod & $\mathit{51.4}\%$ & $44.8\%$ & $38.1\%$ & $27.7\%$\\
+    Voronoi & $52.6\%$ & $\mathbf{43.4}\%$ & $\mathbf{34.4}\%$ & $\mathbf{19.1}\%$\\
+    BoxMesh & $99.5\%$ & $98.5\%$ & $96.6\%$ & $89.5\%$\\
+  \end{tabular}
+  \caption{This is the numerical counterpart to the histogram data
+    presented in Figure \ref{fig:throughput-prediction}. The columns
+    display the percent of null hypothesis rejections by the KS-test
+    when provided different selections of $p$-values for each
+    algorithm. The algorithm with the lowest rejection rate at each
+    $p$ is boldface, while the second lowest is italicized.}
+  \label{table:null-hypothesis-results}
+\end{table}
+
+%% ===================================================================
+
+Figure \ref{fig:throughput-prediction} expands on the KS statistic
+results presented in Figure \ref{fig:error-throughput}. Agglomerate
+errors for each algorithm resemble a Gamma distribution. The
+percentages of significant prediction errors with varying $p$-values
+are on display in Table \ref{table:null-hypothesis-results}. When
+considering the $p=0.001$ results for each technique, a little over
+one third of the predicted CDFs are significantly different from the
+measured (and presumed) correct CDFs. However, it should be noted that
+with 150 samples, the error of an empirical distribution function
+(EDF) can reasonably be upwards of $.1$, which serves as a rough
+estimate for the lower limit of achievable error. Globally, only a
+third of Voronoi predictions fail to capture \textit{all} of the
+characteristics of the CDFs at new system configurations.
 
 \section{Discussion}
-
-The bootstrapping procedure presented for each approximation technique still has much room for improvement.
-Initial analysis suggests that the appropriate relative error tolerance needs to be discovered empirically for each application of a modeling technique.
-Further analytic studies could arrive at methods for determining optimal error tolerances at runtime, however increases in runtime complexity may not be afforded in many applications.
-
-The box-shaped basis functions and the construction algorithms used for the $MBM$ and $IBM$ could become a source of error when $d$ (the dimension of the data $X$) is comparable to $n$ (the number of known points).
-The blending regions in which multiple basis functions overlap are always axis aligned and in applications such as image analysis, any single dimension may be unsuitable for approximating the true underlying function.
-The Voronoi mesh attempts to address this problem by utilizing boundaries between points in multiple dimensions simultaneously.
-However, it is empirically unclear whether the true benefits of the $VM$ are seen in applications where $d \ll n$.
-
-Each of the case studies presented have fewer than $1000$ points.
-The complexities of the presented approximation techniques are suitable for large dimension, but the increased complexity associated with brute-force bootstrapping currently prohibits their use on larger data sets.
-The Voronoi mesh in particular has a large complexity with respect to $n$ which could be significantly improved via more greedy bootstrapping.
-While each technique requires less than ten seconds on average to produce a fit in the presented case studies, the fit time required quickly grows into minutes around $1000$ points.
-While these initial results appear somewhat limiting, they demonstrate the viability of each mesh and leave room for further theoretical exploration of techniques to reduce the runtime complexity while maintaining the approximation power and flexibility.
+\label{sec:discussion}
+
+\begin{table}
+  \centering
+  \begin{tabular}{c|c|c|c}
+    \hline
+    Algorithm & Avg. \% Best & Avg. Fit Time & Avg. App. Time (1)\\
+    \hline
+    MARS & $4.6\%$ & $22.3$s & $1.37 \times 10^{-3}$s\\
+    SVR & $\mathit{21.1}\%$ & $0.503$s & $\mathit{1.37 \times 10^{-4}}$s\\
+    MLP & $\mathbf{42.1}\%$ & $210.0$s & $1.02 \times 10^{-3}$s\\
+    Delaunay & $6.0\%$ & $\mathbf{1.43 \times 10^{-4}}$s & $3.03$s\\
+    ShepMod & $18.4\%$ & $0.711$s & $1.49 \times 10^{-4}$s\\
+    LSHEP & $8.4\%$ & $1.69$s & $\mathbf{1.16 \times 10^{-4}}$s\\
+    Voronoi & $0.5\%$ & $\mathit{0.127}$s & $0.0375$s\\
+    BoxSplines & $0.2\%$ & $0.816$s & $5.26 \times 10^{-4}$s\\
+    \hline
+  \end{tabular}
+  \caption{This data is the average of Appendix tables
+    \ref{table:error-forest-fire}, \ref{table:error-parkinsons},
+    \ref{table:error-weather}, \ref{table:error-credit-card}, and
+    provides a gross summary of overall results. The columns display
+    (weighted equally by data set, \textit{not} points) the average
+    frequency with which any algorithm provided the lowest absolute
+    error approximation, the average time-to-fit, and the average time
+    required to approximate one point. Interpolants provide the
+    lowest error approximation one third of the time across all data,
+    while regressors occupy the other two thirds. This result is
+    obtained without any customized tuning or preprocessing to
+    maximize the performance of any given algorithm. In practice,
+    tuning and preprocessing may have large effects on approximation
+    performance.}
+  \label{table:avg-performance}
+\end{table}
+
+Table \ref{table:avg-performance} summarizes results across the four
+test data sets with real-valued ranges. The interpolants discussed in
+this paper produce the \textit{best} approximations roughly one third
+of the time, and produce competitive approximations for almost all
+data sets. These test problems are almost certainly
+\textit{stochastic} in nature, but the high dimension yields to
+greater sparsity and makes interpolation a competitive approximation
+technique.
+
+The major advantages to interpolation lie in the near absence of
+\textit{fit} time. Delaunay has no preparations required, LSHEP and
+ShepMod require pairwise distance calculations to determine the radii
+of influence for points in each model. At least hundreds, and
+sometimes hundreds of thousands of predictions can be made by the
+interpolants before the most widely used regressor (the MLP) finishes
+fitting these relatively small data sets. However, the computational
+complexities of all interpolants presented are greater than linear in
+either dimension or number of points, meaning large enough data (in
+dimension and number of observations) would cause interpolation to be
+slower.
+
+The theoretical results presented in Section \ref{sec:theory} directly
+apply to Delaunay interpolation, however the performance of Delaunay
+does not appear significantly better than other algorithms on these
+data sets. This observation may be due to the stochastic nature of the
+data, but it also speaks to the alignment of the geometry of the
+underlying functions with the approximations generated by the
+different interpolation methods. The strong performance of other
+interpolants suggests that theoretical results similar to those
+presented in this work can be achieved for the other interpolants
+under reasonable assumptions.
+
+Finally, most of the interpolants presented in this work benefit from
+the ability to model any function over a real vector space with a
+range that is closed under convex combinations. The results of the
+distribution prediction case study indicate that interpolants can
+effectively predict CDFs. The error analysis for that work relies on
+the KS statistic, which captures the worst part of any prediction and
+hence provides a conservatively large estimate of approximation
+error. The average absolute errors in the predicted CDFs are always
+lower than the KS statistics. However, the KS statistic was chosen
+because of the important statistical theory surrounding it as an error
+measure. Considering this circumstance, a nonnegligible volume of
+predictions provide impressively low levels of error in that final
+case study.
 
 \section{Conclusion}
-
-The Max Box Mesh, Iterative Box Mesh, and Voronoi Mesh each provide novel strategies for effectively approximating multivariate phenomonon.
-The underlying constructions are theoretically straightforward, yet powerful and flexible.
-The computational complexities of each make them particularly suitable for applications in many dimensions, while the bootstrapping error tolerance parameter allows a balance between smoothing and interpolation to be explored empirically with each application.
-
-\subsection{Future Work}
-
-A thorough comparison with constituent multivariate approximation techniques including but not limited to, linear Shepard interpolation, multivariate adaptive regression splines, multilayer perceptron regression, and Delaunay triangulation constitutes future work.
-A more detailed study of alternative bootstrapping techniques may also provide valuable insight.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-%% ===================================================================
-\section{DATA}
-\label{sec:data}
-
-\begin{table}
-  \centering
-  \begin{tabular}{c|c}
-    \hline
-    \textbf{System Parameter} & \textbf{Values}\\
-    \hline
-    File Size (KB) & 4, 16, 64, 256, 1024, 4096, 8192, 16384\\
-    \hline
-    Record Size (KB) & \multilinecell{4, 8, 16, 32, 64, 128, 256, 512,\\ 1024, 2048, 4096, 8192, 16384}\\
-    \hline
-    Thread Count & 1, 8, 16, 24, 32, 40, 48, 56, 64\\
-    \hline
-    Frequency (GHz) & 1.2, 1.6, 2, 2.3, 2.8, 3.2, 3.5\\
-    \hline
-    Test Type & \multilinecell{Readers, Rereaders, Random Readers, \\ Initial Writers, Rewriters, Random Writers}\\
-    \hline
-  \end{tabular}
-  \caption{A description of system parameters considered for IOzone. Record size must be $\leq$ file size during execution.
-    \vspace{-.5cm}}
-  \label{tab:data_description}
-\end{table}
-
-\begin{figure}
-  \includegraphics[width=.5\textwidth]{Figures/plot-histogram-throughput.pdf}
-  \caption{Histogram of the raw throughput values recorded during all IOzone tests across all system configurations.
-The distribution is skewed right, with few tests having significantly higher throughput than most others.
-  \vspace{-.5cm}}
-  \label{fig:throughput_histogram}
-\end{figure}
-
-This paper presents a variability modeling case study with a five-dimensional dataset produced by executing the IOzone benchmark \cite{iozone} on a homogeneous cluster of computers.
-Each node contains two Intel Xeon E5-2637 CPUs offering a total of 16 CPU cores with 16GB of DRAM.
-While the CPU frequency varies depending on the test configuration, the I/O from IOzone is performed by an ext4 filesystem sitting above an Intel SSDSC2BA20 SSD drive.
-At the time of data collection, Linux kernel Version 4.13.0 was used.
-The system performance data was collected over two weeks by executing IOzone 150 times for each of a select set of approximately 18K system configurations, for a total of approximately 2.7M executions of IOzone.
-A single IOzone execution reports the max I/O throughput in kilobytes per second seen for the selected test type.
-The summary of the data components in $x^{(i)}$ for the experiments for this paper can be seen in Table \ref{tab:data_description}.
-Distributions of raw throughput values being modeled can be seen in Figure \ref{fig:throughput_histogram}.
-
-Some mild preprocessing was necessary to prepare the data for modeling and analysis.
-All features were shifted by their minimum value and scaled by their range, mapping each feature independently into $[0,1]$.
-This normalization ensures each feature is treated equally by the interpolation techniques and should be performed on all data before building models and making predictions regardless of application.
-All 150 repeated trials for a system configuration were grouped with that configuration.
-The only nonordinal feature in this data is the test type.
-All test types were treated as different applications and were separated for modeling and analysis, i.e., predictions for the ``readers'' test type were made using only known configurations for the ``readers'' test type.
-
-%% ===================================================================
-
-\begin{figure}
-  \begin{tikzpicture}
-    \node (img)  {\includegraphics[width=0.5\textwidth,trim={1.5cm 0 0 0}]{Figures/plot-histogram-80_20-KS.pdf}};
-    \node[below=of img, node distance=1cm, yshift=1cm] {KS Statistic for Predicted vs. Actual};
-    \node[left=of img, node distance=0cm, rotate=90, anchor=center,yshift=-0.7cm] {Count of Predictions with Given KS Statistic};
-  \end{tikzpicture}
-  \caption{Histograms of the prediction error for each modeling algorithm from ten random splits when trained with 80\% of the data aggregated over all different test types.
-The distributions show the KS statistics for the predicted throughput distribution versus the actual throughput distribution.
-The four vertical red lines represent commonly used $p$-values \{0.05, 0.01, 0.001, 1.0e-6\} respectively.
-All predictions to the right of a red line represent CDF predictions that are significantly different (by respective $p$-value) from the actual distribution according to the KS test.
-  \vspace{-.1cm}}
-  \label{fig:ks_histogram_80_20}
-\end{figure}
-
-\section{RESULTS}
-\label{sec:results}
-
-All three interpolation techniques are used to predict the distribution of I/O throughput values at previously unseen system configurations.
-In order to improve robustness of the error analysis, ten random selections of 80\% of the IOzone data are used to train each model and the remaining 20\% provide approximation error for each model.
-The recorded errors are grouped by unique system configuration and then averaged within each group.
-The samples are identical for each interpolation technique, ensuring consistency in the training and testing sets.
-
-The aggregation of errors across all IOzone tests given 80\% of the data as training can be seen in Figure \ref{fig:ks_histogram_80_20}.
-Agglomerate errors for each technique resemble a Gamma distribution.
-The percentages of significant prediction errors with varying $p$-values are on display in Table \ref{tab:p_value_failure_rate}.
-The primary $p$-value used for analyses in this work is 0.001, chosen because close to 2K predictions are made for each test type.
-Also, applications executed in cloud and HPC systems that could benefit from statistical modeling will be executed at least thousands of times.
-In line with this knowledge, it is important to ensure that only a small fraction of interpretable results could occur solely under the influence of random chance.
-When considering the $p=0.001$ results for each technique, a little under half of the predicted CDFs are significantly different from the measured (and presumed) correct CDFs.
-A rejection rate of 45\% would seem a poor result, however in this situation the complexity of the problem warrants a slightly different interpretation.
-These predictions are a very \textit{precise} characterization of performance variability, in fact the cumulative distribution function of a random variable is the strongest possible characterization of variability that can be predicted.
-Globally, only a little under half of the predictions fail to capture \textit{all} of the characteristics of performance variability at new system configurations.
-It is also demonstrated later in this Section that this result can likely be improved.
-
-\begin{table}
-  \centering
-  \begin{tabular}{c|c|c}
-    \hline
-    \textbf{Algorithm} & \textbf{$P$-Value} & \textbf{\% N.H. Rejections} \\
-    \hline
-    \multilinecell{Delaunay\\Max Box Mesh\\Voronoi Mesh} & .05 & \multilinecell{58.4\\69.3\\61.9}\\
-    \hline
-    \multilinecell{Delaunay\\Max Box Mesh\\Voronoi Mesh} & .01 & \multilinecell{51.1\\58.4\\53.4}\\
-    \hline
-    \multilinecell{Delaunay\\Max Box Mesh\\Voronoi Mesh} & .001 & \multilinecell{44.1\\46.9\\45.1}\\
-    \hline
-    \multilinecell{Delaunay\\Max Box Mesh\\Voronoi Mesh} & 1.0e-6 & \multilinecell{31.4\\26.6\\28.7}\\
-    \hline
-  \end{tabular}
-  \caption{Percent of null hypothesis rejections rate by the KS-test when provided different selections of $p$-values.
-These accompany the percent of null hypothesis rejection results from Figure \ref{fig:ks_histogram_80_20}.
-    \vspace{-.5cm}}
-  \label{tab:p_value_failure_rate}
-\end{table}
-
-While interpreting null hypothesis rejection rates for these interpolation techniques, it is important to consider how the rejection rate reduces with increasing amounts of training data.
-Figure \ref{fig:ks_failure_by_training} displays the change in $p=0.001$ null hypothesis rejection rate with increasing density of training data up to the maximum density allowed by this set.
-Delaunay interpolation provides the best results with the least training data by about 5\%, but these low density rejection rates are unacceptably high (90\%).
-Figure \ref{fig:ks_failure_by_training} clearly shows that this data set and/or the system variables used in the models of performance variability is inadequate to capture the full variability map from system parameters to performance CDF.
-Which or both obtains is not clear.
-A few well chosen data points can significantly improve the interpolants, and thus a careful study of the rejection instances is warranted, besides enlarging the set of system variables being modeled.
-
-It may be misleading to consider the global performance of each prediction technique across all test types, as some test types are more difficult than others to predict and have more apparent latent variables.
-In Figure \ref{fig:ks_failure_by_training_and_test}, the relative difficulty of each IOzone test type can be compared.
-The I/O test types analyzing reads are typically approximated with lower error than those test types analyzing writes.
-Regardless of test type, in the aggregate results the KS statistics hover consistently around 0.15, demonstrating an impressively low KS statistic for predictions.
-In order to address the opacity of aggregate analysis, another case study and an application of the methodology from Section \ref{sec:feature_weighting} is presented in Table \ref{tab:optimized_p_value_failure_rate}.
-
-\begin{figure}
-  \begin{tikzpicture}
-    \node (img)  {\includegraphics[width=0.4\textwidth,]{Figures/plot-KS-Failure-by-Training.pdf}};
-    \node[below=of img, node distance=1cm, yshift=1cm] {Percentage of Training Data};
-    \node[left=of img, node distance=0cm, rotate=90, anchor=center,yshift=-0.7cm] {\% Null Hypothesis Rejections};
-  \end{tikzpicture}
-  \caption{The performance of each algorithm on the KS test ($p=0.001$) with increasing amounts of training data averaged over all IOzone test types and ten random splits of the data.
-The training percentages range from 5\% to 95\% in increments of 5\%.
-Delaunay is the best performer until 95\% of data is used for training, at which Max Box mesh becomes the best performer by a fraction of a percent.
-  \vspace{-.1cm}}
-  \label{fig:ks_failure_by_training}
-\end{figure}
-
-The results presented in Table \ref{tab:optimized_p_value_failure_rate} are achieved by permitting each approximation technique 300 iterations of simulated annealing.
-In each iteration, the impact of potential weights on the average KS statistic were considered.
-All weights were kept in the range [0,2], and were applied to the normalized features for frequency, file size, record size, and number of threads.
-All three approximation techniques had similar optimal weights achieved by simulated annealing of approximately $(.001, 2, 1.7, 1.5)$ for frequency, file size, record size, and number of threads, respectively.
-Recall that each interpolation technique uses small distances to denote large influences on predictions, meaning that frequency was the most important feature when predicting variability for the ``readers'' test type, followed not-so-closely by number of threads, then record size.
-
-The ``readers'' test type results demonstrate that the underlying prediction techniques work and are capable of seeing rejection rates below 5\% when tuned for a given application.
-It is important to emphasize that the roughly 95\% of predictions for which the null hypothesis was not rejected are predicting the \textit{precise} distribution of I/O throughput that will be witnessed at a previously unseen system configuration.
-To the authors' knowledge, there is no existing methodology that is generically applicable to any system performance measure, agnostic of system architecture, and capable of making such powerful predictions.
-
-\begin{table}
-  \centering
-  \begin{tabular}{c|c|c|c}
-    \hline
-    \textbf{Algorithm} & \textbf{$P$-Value} & \multilinecell{\textbf{Unweighted}\\\textbf{\% N.H. Rejection}} & \multilinecell{\textbf{Weighted}\\\textbf{\% N.H. Rejection}}\\
-    \hline
-    \multilinecell{Delaunay\\Max Box Mesh\\Voronoi Mesh} & .05 & \multilinecell{24.9\\21.3\\18.7} & \multilinecell{30.2\\21.2\\11.3}\\
-    \hline
-    \multilinecell{Delaunay\\Max Box Mesh\\Voronoi Mesh} & .01 & \multilinecell{21.6\\16.4\\14.9} & \multilinecell{27.4\\16.4\\7.0}\\
-    \hline
-    \multilinecell{Delaunay\\Max Box Mesh\\Voronoi Mesh} & .001 & \multilinecell{19.7\\13.1\\12.3} & \multilinecell{25.4\\13.1\\4.6}\\
-    \hline
-    \multilinecell{Delaunay\\Max Box Mesh\\Voronoi Mesh} & 1.0e-6 & \multilinecell{17.9\\11.3\\8.5} & \multilinecell{23.4\\11.3\\2.3}\\
-    \hline
-  \end{tabular}
-  \caption{The null hypothesis rejection rates for various $p$-values with the KS-test.
-These results are strictly for the ``readers'' IOzone test type and show unweighted results as well as the results with weights tuned for minimum error (KS statistic) by 300 iterations of simulated annealing.
-Notice that the weights identified for the Delaunay model cause data dependent tuning, reducing performance.
-MaxBoxMesh performance is improved by a negligible amount.
-VoronoiMesh performance is notably improved.
-    \vspace{-.5cm}}
-  \label{tab:optimized_p_value_failure_rate}
-\end{table}
-
-%% \begin{itemize}
-%% \item Case studies of good and bad predictions of distributions made, present 3 (best, median, worst)
-%% \item Table / Figure showing error for each of the tests (hardest test to predict)
-%% \item Presentation of the top most difficult configurations to predict for each test (and the bad performance at those points)
-%% %% \item Comparison to predictions made using the "test" numerical column
-%% \end{itemize}
-
-\section{DISCUSSION}
-\label{sec:discussion}
-
-The results of the IOzone case study indicate that predicting the CDF of I/O throughput at previously unseen system configurations is a challenging problem.
-The KS statistic captures the worst part of any prediction and hence provides a conservatively large estimate of approximation error.
-The average absolute errors in the predicted CDFs are always lower than the KS statistics.
-However, the KS statistic was chosen because of the important statistical theory surrounding it as an error measure.
-Considering this circumstance, a nonnegligible volume of predictions provide impressively low levels of error.
-Powerful predictive tools such as those presented in this work allow for more in-depth analysis of system performance variability.
-For example, system configurations that are most difficult to predict in these tests are likely ``outlier'' configurations that do not resemble those configurations that share many similar parameters.
-Analysis of these configurations may provide valuable insight into effective application specific operation of computer systems.
-
-As mentioned in Section \ref{sec:introduction}, no prior work has attempted to model an arbitrary performance measure for a system to such a high degree of precision.
-All previous statistical modeling attempts capture a few ($<3$) ordinal performance measures.
-Generating models that have such high degrees of accuracy allows system engineers to identify previously unused configurations that present desired characteristics.
-Service level agreements (SLAs) in cloud computing environments are cause for capital competition that is affected heavily by system performance \cite{patel2009service}.
-Users prefer SLAs that allow the most computing power per monetary unit, incentivizing service providers to guarantee the greatest possible performance.
-Overscheduling and irregular usage patterns force cloud service providers to occasionally overload machines, in which case precise models of system performance can be used to statistically minimize the probability of SLA violation.
-Similar targeted performance tuning techniques can be applied to HPC system configuration to maximize application throughput or minimize system power consumption.
-
-A final application domain affected by this methodology is computer security.
-Collocated users on cloud systems have received attention recently \cite{ali2015security}.
-If a malicious collocated user is capable of achieving specific insight into the configuration of the system, or the activity of other collocated users by executing performance evaluation programs (i.e., IOzone), a new attack vector may present itself.
-Malicious users could be capable of identifying common performance distributions of vulnerable system configurations and vulnerable active user jobs.
-This knowledge may allow targeted exploits to be executed.
-Light inspection of raw IOzone I/O throughputs provides substantial evidence that distinct performance distributions coincide closely with specific system configuration parameters.
-Conversely, a service provider may defend against such attacks by deliberately obfuscating the performance of the machine.
-Models such as those presented in this paper could identify optimal staggering and time-delay whose introduction into the system would prevent malicious users from identifying system configurations and active jobs.
-
-\begin{figure}
-  \begin{tikzpicture}
-    \node (img)  {\includegraphics[width=0.4\textwidth,]{Figures/plot-KS-failure-by-training-and-test.pdf}};
-    \node[below=of img, node distance=1cm, yshift=1cm] {Percentage of Training Data};
-    \node[left=of img, node distance=0cm, rotate=90, anchor=center,yshift=-0.7cm] {\% Null Hypothesis Rejections};
-  \end{tikzpicture}
-  \caption{The percentage of null hypothesis rejections for predictions made by each algorithm on the KS test ($p=0.001$) over different IOzone test types with increasing amounts of training data.
-Each percentage of null hypothesis rejections is an average over ten random splits of the data.
-The training percentages range from 5\% to 95\% in increments of 5\%.
-The read test types tend to allow lower rejection rates than the write test types.
-  \vspace{-.1cm}}
-  \label{fig:ks_failure_by_training_and_test}
-\end{figure}
-
-Results presented in Table \ref{tab:optimized_p_value_failure_rate} are particularly interesting, demonstrating that Delaunay appears most vulnerable to data dependent tuning, Max Box mesh is largely insensitive to such tuning, and Voronoi mesh benefits (for this data set) from the tuning.
-
-There are many avenues for extending this modeling methodology.
-One extension is to add categorical variables to the models.
-Presently the rejection rate of distribution predictions can only be reduced with large volumes of performance data, however the judicious choice (via experimental design, e.g.) of new data points may be able to effectively reduce the amount of training data required.
-Finally, more case studies need to be done to test the robustness of the present modeling techniques to changes in domain and performance measure.
-
-\section{CONCLUSION}
 \label{sec:conclusion}
 
-The methodology presented is capable of providing new insights, extending existing analyses, and improving the management of computational performance variability.
-Delaunay, Max Box mesh, and Voronoi mesh interpolation are viable techniques for constructing approximations of performance cumulative distribution functions.
-A case study on I/O throughput demonstrated that the models are capable of effectively predicting CDFs for most unseen system configurations for any of the available I/O test types.
-The present methodology represents a notable increase in the ability to statistically model arbitrary system performance measures involving the interaction of many ordinal system parameters.
-%% \begin{itemize}
-%% \item Modeling the distributions of system performance is valuable.
-%% \item This work presents a steep increase in the power of predictions made by statistical modeling techniques.
-%% \end{itemize}
+This work presents new theoretical results for interpolation in
+arbitrary dimension with simplicial approximations. Coinciding with
+the theoretical results, an empirical evaluation across real-world
+problems demonstrates that interpolants produce competitively accurate
+models of multivariate phenomenon when compared with common regressors
+for sparse, moderately high dimensional problems.
+
+The various interpolants discussed in this paper have been
+demonstrated to effectively approximate multivariate phenomena in up
+to $30$ dimensions. The underlying constructions are theoretically
+straightforward, yet powerful and flexible. Most of the interpolants'
+computational complexities make them particularly suitable for
+applications in even higher dimension. The major benefits of
+interpolation are seen when only a small number of approximations
+($\leq 1000$) are made from data and when there are relatively few
+data points for the dimension (for empirical results presented, $\log_d n \leq 5$).
 
@@
 
+\newpage
+
+\begin{appendix}
+
+\section*{Appendix}
+
+\begin{table}
+  \centering
+  \begin{tabular}{c|c|c|c|c|c}
+    \hline
+    Algorithm & Min & $25^{th}$ & $50^{th}$ & $75^{th}$ & Max\\
+    \hline
+    MARS & $0.00984$ & $3.11$ & $7.01$ & $\mathit{11.7}$ & $1090.0$\\
+    SVR & $\mathit{0.00402}$ & $\mathbf{0.243}$ & $\mathbf{0.416}$ & $\mathbf{6.19}$ & $1090.0$\\
+    MLP & $0.0426$ & $2.63$ & $\mathit{5.27}$ & $14.0$ & $1090.0$\\
+    Delaunay & $\mathbf{0.00}$ & $1.98$ & $5.37$ & $13.1$ & $\mathit{1080.0}$\\
+    ShepMod & $\mathbf{0.00}$ & $\mathit{1.93}$ & $6.27$ & $16.0$ & $1090.0$\\
+    LSHEP & $0.0400$ & $2.17$ & $8.87$ & $19.1$ & $\mathbf{1070.0}$\\
+    Voronoi & $0.00982$ & $3.65$ & $7.56$ & $15.6$ & $1090.0$\\
+    BoxSpline & $\mathbf{0.00}$ & $2.27$ & $5.91$ & $16.4$ & $1090.0$\\
+    \hline
+  \end{tabular}
+  \caption{This numerical data accompanies the visual provided in
+    Figure \ref{fig:error-forest-fire}. The columns of absolute error
+    percentiles correspond to the minimum, first quartile, median,
+    second quartile, and maximum absolute errors respectively. The
+    minimum of each column is boldface, while the second lowest value
+    is italicized. All values are rounded to three significant
+    digits.}
+  \label{table:error-forest-fire}
+\end{table}
+
+\begin{table}
+  \centering
+  \begin{tabular}{|c|c| c |c|c|c|}
+    \cline{1-2}\cline{4-6}
+    Algorithm & \% Best &  & Fit Time & App. Time (1) & Total App. Time\\
+    \cline{1-2}\cline{4-6}
+    MARS & $\mathit{7.7}\%$ &  & $29.1$s & $1.37 \times 10^{-3}$s & $0.0686$s\\
+    SVR & $\mathbf{80.2}\%$ &  & $5.84 \times 10^{-3}$s & $\mathit{6.20 \times 10^{-5}}$s & $\mathit{3.10 \times 10^{-3}}$s\\
+    MLP & $0.0\%$ &  & $32.8$s & $8.71 \times 10^{-4}$s & $0.0436$s\\
+    Delaunay & $3.5\%$ &  & $\mathbf{2.49 \times 10^{-5}}$s & $0.0385$s & $1.93$s\\
+    ShepMod & $3.7\%$ &  & $6.34 \times 10^{-3}$s & $6.44 \times 10^{-5}$s & $3.22 \times 10^{-3}$s\\
+    LSHEP & $5.1\%$ &  & $0.0275$s & $\mathbf{4.63 \times 10^{-5}}$s & $\mathbf{2.31 \times 10^{-3}}$s\\
+    Voronoi & $0.0\%$ &  & $\mathit{1.82 \times 10^{-4}}$s & $3.96 \times 10^{-4}$s & $0.0198$s\\
+    BoxSpline & $0.4\%$ &  & $7.24 \times 10^{-3}$s & $9.78 \times 10^{-5}$s & $4.89 \times 10^{-3}$s\\
+    \cline{1-2}\cline{4-6}
+  \end{tabular}
+  \caption{The left above shows how often each algorithm had the
+    lowest absolute error approximating forest fire data in table
+    \ref{table:error-forest-fire}. On the right columns are median fit
+    time of 454 points, median time for one approximation, and median
+    time approximating 50 points.}
+  \label{table:best-forest-fire}
+\end{table}
+
+\begin{table}
+  \centering
+  \begin{tabular}{c|c|c|c|c|c}
+    \hline
+    Algorithm & Min & $25^{th}$ & $50^{th}$ & $75^{th}$ & Max\\
+    \hline
+    MARS & $0.00948$ & $9.98$ & $20.4$ & $32.9$ & $863.0$\\
+    SVR & $0.00138$ & $3.17$ & $7.31$ & $12.6$ & $\mathbf{27.5}$\\
+    MLP & $2.39 \times 10^{-5}$ & $\mathit{0.533}$ & $\mathit{1.25}$ & $\mathbf{2.84}$ & $39.3$\\
+    Delaunay & $\mathit{3.72 \times 10^{-12}}$ & $1.20$ & $3.50$ & $7.67$ & $30.7$\\
+    ShepMod & $\mathbf{0.00}$ & $\mathbf{0.255}$ & $\mathbf{0.908}$ & $\mathit{3.43}$ & $34.5$\\
+    LSHEP & $0.00254$ & $2.93$ & $7.16$ & $13.1$ & $\mathit{29.0}$\\
+    Voronoi & $\mathbf{0.00}$ & $1.29$ & $3.52$ & $6.87$ & $30.1$\\
+    BoxSpline & $0.00287$ & $4.39$ & $9.10$ & $14.8$ & $41.3$\\
+    \hline
+  \end{tabular}
+  \caption{This numerical data accompanies the visual provided in
+    Figure \ref{fig:error-parkinsons}. The columns of absolute error
+    percentiles correspond to the minimum, first quartile, median,
+    second quartile, and maximum absolute errors respectively. The
+    minimum of each column is boldface, while the second lowest value
+    is italicized. All values are rounded to three significant
+    digits.}
+  \label{table:error-parkinsons}
+\end{table}
+
+\begin{table}
+  \centering
+  \begin{tabular}{|c|c| c |c|c|c|}
+    \cline{1-2}\cline{4-6}
+    Algorithm & \% Best &  & Fit Time & App. Time (1) & Total App. Time\\
+    \cline{1-2}\cline{4-6}
+    MARS & $0.0\%$ &  & $37.9$s & $2.53 \times 10^{-3}$s & $1.48$s\\
+    SVR & $0.1\%$ &  & $0.859$s & $\mathit{1.81 \times 10^{-4}}$s & $\mathit{0.106}$s\\
+    MLP & $\mathit{32.0}\%$ &  & $348.0$s & $1.11 \times 10^{-3}$s & $0.653$s\\
+    Delaunay & $0.0\%$ &  & $\mathbf{1.90 \times 10^{-4}}$s & $3.69$s & $2.17 \times 10^{3}$s\\
+    ShepMod & $\mathbf{66.4}\%$ &  & $1.13$s & $1.82 \times 10^{-4}$s & $0.107$s\\
+    LSHEP & $0.0\%$ &  & $2.39$s & $\mathbf{1.44 \times 10^{-4}}$s & $\mathbf{0.0845}$s\\
+    Voronoi & $1.6\%$ &  & $\mathit{0.298}$s & $0.0274$s & $16.1$s\\
+    BoxSpline & $0.0\%$ &  & $1.26$s & $6.43 \times 10^{-4}$s & $0.377$s\\
+    \cline{1-2}\cline{4-6}
+  \end{tabular}
+  \caption{The left above shows how often each algorithm had the
+    lowest absolute error approximating parkinsons data in table
+    \ref{table:error-parkinsons}. On the right columns are median fit
+    time of 5288 points, median time for one approximation, and median
+    time approximating 587 points.}
+  \label{table:best-parkinsons}
+\end{table}
+
+\begin{table}
+  \centering
+  \begin{tabular}{c|c|c|c|c|c}
+    \hline
+    Algorithm & Min & $25^{th}$ & $50^{th}$ & $75^{th}$ & Max\\
+    \hline
+    MARS & $\mathit{6.45 \times 10^{-15}}$ & $\mathbf{2.70 \times 10^{-14}}$ & $1.66$ & $1.96$ & $\mathit{53.3}$\\
+    SVR & $0.0420$ & $0.143$ & $0.148$ & $\mathit{0.860}$ & $119.0$\\
+    MLP & $6.89 \times 10^{-5}$ & $0.0337$ & $\mathbf{0.0795}$ & $\mathbf{0.264}$ & $\mathbf{5.31}$\\
+    Delaunay & $\mathbf{0.00}$ & $0.187$ & $0.919$ & $2.72$ & $56.3$\\
+    ShepMod & $\mathbf{0.00}$ & $0.0957$ & $0.685$ & $2.90$ & $73.2$\\
+    LSHEP & $\mathbf{0.00}$ & $\mathit{0.0153}$ & $\mathit{0.106}$ & $1.17$ & $113.0$\\
+    Voronoi & $\mathbf{0.00}$ & $0.430$ & $1.28$ & $2.94$ & $83.8$\\
+    BoxSpline & $\mathbf{0.00}$ & $0.789$ & $2.06$ & $5.25$ & $119.0$\\
+    \hline
+  \end{tabular}
+  \caption{This numerical data accompanies the visual provided in
+    Figure \ref{fig:error-weather}. The columns of absolute error
+    percentiles correspond to the minimum, first quartile, median,
+    second quartile, and maximum absolute errors respectively. The
+    minimum value of each column is boldface, while the second lowest
+    is italicized. All values are rounded to three significant
+    digits.}
+  \label{table:error-weather}
+\end{table}
+
+\begin{table}
+  \centering
+  \begin{tabular}{|c|c| c |c|c|c|}
+    \cline{1-2}\cline{4-6}
+    Algorithm & \% Best &  & Fit Time & App. Time (1) & Total App. Time\\
+    \cline{1-2}\cline{4-6}
+    MARS & $10.7\%$ &  & $0.151$s & $1.17 \times 10^{-4}$s & $0.0304$s\\
+    SVR & $4.1\%$ &  & $0.133$s & $\mathbf{9.47 \times 10^{-5}}$s & $\mathbf{0.0246}$s\\
+    MLP & $\mathbf{56.8}\%$ &  & $169.0$s & $1.37 \times 10^{-3}$s & $0.356$s\\
+    Delaunay & $0.1\%$ &  & $\mathbf{1.16 \times 10^{-4}}$s & $1.55$s & $403.0$s\\
+    ShepMod & $3.5\%$ &  & $0.265$s & $1.28 \times 10^{-4}$s & $0.0333$s\\
+    LSHEP & $\mathit{28.4}\%$ &  & $0.874$s & $\mathit{9.75 \times 10^{-5}}$s & $\mathit{0.0254}$s\\
+    Voronoi & $0.2\%$ &  & $\mathit{0.0112}$s & $0.0270$s & $7.01$s\\
+    BoxSpline & $0.3\%$ &  & $0.330$s & $4.06 \times 10^{-4}$s & $0.106$s\\
+    \cline{1-2}\cline{4-6}
+  \end{tabular}
+  \caption{The left above shows how often each algorithm had the
+    lowest absolute error approximating Sydney rainfall data in table
+    \ref{table:error-weather}. On the right columns are median fit
+    time of 2349 points, median time for one approximation, and median
+    time approximating 260 points.}
+  \label{table:best-weather}
+\end{table}
+
+\begin{table}
+  \centering
+  \begin{tabular}{c|c|c|c|c|c}
+    \hline
+    Algorithm & Min & $25^{th}$ & $50^{th}$ & $75^{th}$ & Max\\
+    \hline
+    MARS & $5.36$ & $3610.0$ & $4580.0$ & $5450.0$ & $13400.0$\\
+    SVR & $0.0138$ & $8.47$ & $15.4$ & $41.9$ & $7700.0$\\
+    MLP & $0.00151$ & $\mathbf{1.60}$ & $\mathbf{3.86}$ & $\mathbf{8.34}$ & $\mathbf{604.0}$\\
+    Delaunay & $\mathbf{0.00}$ & $\mathit{2.69}$ & $\mathit{13.8}$ & $\mathit{35.0}$ & $\mathit{4840.0}$\\
+    ShepMod & $\mathbf{0.00}$ & $4.21$ & $17.3$ & $51.6$ & $6510.0$\\
+    LSHEP & $1.27$ & $199.0$ & $260.0$ & $343.0$ & $6530.0$\\
+    Voronoi & $\mathit{2.89 \times 10^{-10}}$ & $14.7$ & $32.8$ & $52.9$ & $4860.0$\\
+    BoxSpline & $0.00740$ & $20.8$ & $40.9$ & $79.6$ & $7640.0$\\
+    \hline
+  \end{tabular}
+  \caption{This numerical data accompanies the visual provided in
+    Figure \ref{fig:error-credit-card}. The columns of absolute error
+    percentiles correspond to the minimum, first quartile, median,
+    second quartile, and maximum absolute errors respectively. The
+    minimum value of each column is boldface, while the second lowest
+    is italicized. All values are rounded to three significant
+    digits.}
+  \label{table:error-credit-card}
+\end{table}
+
+\begin{table}
+  \centering
+  \begin{tabular}{|c|c| c |c|c|c|}
+    \cline{1-2}\cline{4-6}
+    Algorithm & \% Best &  & Fit Time & App. Time (1) & Total App. Time\\
+    \cline{1-2}\cline{4-6}
+    MARS & $0.0\%$ &  & $22.0$s & $1.48 \times 10^{-3}$s & $0.820$s\\
+    SVR & $0.0\%$ &  & $1.01$s & $\mathit{2.10 \times 10^{-4}}$s & $\mathit{0.117}$s\\
+    MLP & $\mathbf{79.5}\%$ &  & $290.0$s & $7.14 \times 10^{-4}$s & $0.397$s\\
+    Delaunay & $\mathit{20.5}\%$ &  & $\mathbf{2.43 \times 10^{-4}}$s & $6.83$s & $3.80 \times 10^{3}$s\\
+    ShepMod & $0.1\%$ &  & $1.45$s & $2.20 \times 10^{-4}$s & $0.122$s\\
+    LSHEP & $0.0\%$ &  & $3.47$s & $\mathbf{1.76 \times 10^{-4}}$s & $\mathbf{0.0981}$s\\
+    Voronoi & $0.0\%$ &  & $\mathit{0.197}$s & $0.0950$s & $52.8$s\\
+    BoxSpline & $0.0\%$ &  & $1.66$s & $9.56 \times 10^{-4}$s & $0.532$s\\
+    \cline{1-2}\cline{4-6}
+  \end{tabular}
+  \caption{The left above shows how often each algorithm had the
+    lowest absolute error approximating credit card transaction data
+    in Table \ref{table:error-credit-card}. On the right columns are
+    median fit time of 5006 points, median time for one approximation,
+    and median time approximating 556 points.}
+  \label{table:best-credit-card}
+\end{table}
+
+\begin{table}
+  \centering
+  \begin{tabular}{c|c|c|c|c|c}
+    \hline
+    Algorithm & Min & $25^{th}$ & $50^{th}$ & $75^{th}$ & Max\\
+    \hline
+    Delaunay & $\mathbf{0.0287}$ & $\mathbf{0.0914}$ & $\mathbf{0.158}$ & $\mathit{0.308}$ & $\mathit{0.897}$\\
+    ShepMod & $0.0307$ & $\mathit{0.0983}$ & $\mathit{0.164}$ & $0.335$ & $1.00$\\
+    Voronoi & $\mathit{0.0303}$ & $0.100$ & $0.165$ & $\mathbf{0.274}$ & $\mathbf{0.762}$\\
+    BoxSpline & $0.0703$ & $0.417$ & $0.561$ & $0.745$ & $1.00$\\
+    \hline
+  \end{tabular}
+  \caption{This numerical data accompanies the visual provided in
+    Figure \ref{fig:error-throughput}. The columns of absolute error
+    percentiles correspond to the minimum, first quartile, median,
+    second quartile, and maximum KS statistics respectively between
+    truth and guess for models predicting the distribution of I/O
+    throughput that will be observed at previously unseen system
+    configurations. The minimum value of each column is boldface,
+    while the second lowest is italicized. All values are rounded to
+    three significant digits.}
+  \label{table:error-throughput}
+\end{table}
+
+\begin{table}
+  \centering
+  \begin{tabular}{|c|c| c |c|c|c|}
+    \cline{1-2}\cline{4-6}
+    Algorithm & \% Best &  & Fit Time & App. Time (1) & Total App. Time\\
+    \cline{1-2}\cline{4-6}
+    Delaunay & $\mathbf{62.0}\%$ &  & $\mathbf{5.17 \times 10^{-4}}$s & $0.353$s & $106.0$s\\
+    ShepMod & $0.0\%$ &  & $0.0884$s & $\mathbf{1.45 \times 10^{-4}}$s & $\mathbf{0.0436}$s\\
+    Voronoi & $\mathit{38.0}\%$ &  & $\mathit{0.0173}$s & $2.53 \times 10^{-3}$s & $0.762$s\\
+    BoxSpline & $0.0\%$ &  & $0.0972$s & $\mathit{2.10 \times 10^{-4}}$s & $\mathit{0.0633}$s\\
+    \cline{1-2}\cline{4-6}
+  \end{tabular}
+  \caption{The left above shows how often each algorithm had the
+    lowest KS statistic on the I/O throughput distribution data in
+    Table \ref{table:error-throughput}. On the right columns are
+    median fit time of 2715 points, median time for one approximation,
+    and median time approximating 301 points.}
+  \label{table:best-throughput}
+\end{table}
+
+\end{appendix}
+
 \end{document}

### 2019-03-07 Paper_[2019-02]_Numerical_Algorithms/tchlux-NA-2019.tex
Before first round of revisions.e

@@
 
-Regression and interpolation are problems of considerable importance that find applications across many fields of science.
-Pollution and air quality analysis \cite{de2008field}, energy consumption management \cite{lazos2014optimisation}, and student performance prediction \cite{cortez2008using,lux2016applications} are a few of many interdisciplinary applications of multivariate regression for predictive analysis.
-As discussed later, these techniques can also be applied to prediction problems related to forest fire risk assessment \cite{cortez2007data}, Parkinson's patient clinical evaluations \cite{tsanas2010accurate}, local rainfall and weather \cite{williams2009rattle}, credit card transactions \cite{pozzolo2015calibrating}, and high performance computing (HPC) file input/output (I/O) \cite{lux2018nonparametric}.
-
-Regression and interpolation have a considerable theoretical base in one dimension \cite{cheney2009course}.
-Splines in particular are well understood as an interpolation technique in one dimension \cite{de1978practical}, particularly B-splines.
-Tensor products of B-splines \cite{unther1996interpolating} or other basis functions have an unfortunate exponential scaling with increasing dimension.
-Exponential scaling prohibits tensor products from being reasonably applied beyond three-dimensional data.
-In order to address this dimensional scaling challenge, C.
-de Boor and others proposed box splines \cite{de2013box}, of which one of the approximation techniques in this work is composed \cite{lux2018novel}.
-
-The theoretical foundation of low dimensional interpolation allows the construction of strong error bounds that are absent from high dimensional problems.
-This work extends some known results regarding the secant method \cite{dennis1996numerical} to construct an interpolation error bound for problems of arbitrary dimension.
-These error bounds are useful, considering the same cannot be said for regression algorithms in general.
-The maximum complexity of an interpolant is bounded by the amount of data available, while the maximum complexity of a regressor is bounded by both the amount of data and the chosen parametric form.
-For this reason, generic uniform bounds are largely unobtainable for regression techniques on arbitrary approximation problems, even when the approximation domain is bounded.
-
-Aside from theoretical motivation for the use of interpolants, there are often computational advantages as well.
-Interpolants do not have the need for \textit{fitting} data, or minimizing error with respect to model parameters.
-In applications where the amount of data is large and the relative number of predictions that need to be made for a given collection of data is small, the direct application of an interpolant is much less computationally expensive.
-
-In this work, multivariate interpolation is defined given a closed convex subset $Y$ of a metrizable topological vector space with metric $s$, some function $f:\mathbb{R}^d \rightarrow Y$ and a set of points $X = \bigl\{x^{(1)}$,$\ldots$,$x^{(n)}\bigr\} \subset \mathbb{R}^d,$ along with associated function values $f\bigl(x^{(i)}\bigr)$.
-The goal is to construct an approximation $\hat f: \mathbb{R}^d \rightarrow Y$ such that $\hat f\bigl(x^{(i)}\bigr) = f\bigl(x^{(i)}\bigr)$ for all $i = 1$,$\ldots$,$n$.
-It is often the case that the form of the true underlying function $f$ is unknown, however it is still desirable to construct an approximation $\hat f$ with small approximation error at $y \notin X$.
-The two metric spaces that will be discussed in this work are the real numbers with metric $s(x,y) = |x-y|$, and the set of cumulative distribution functions (CDFs) with the Kolmogorov-Smirnov (KS) statistic as metric.
-
-Multivariate regression is often used when the underlying function is presumed to be stochastic, or stochastic error is introduced in the evaluation of $f$.
-Hence, multivariate regression relaxes the conditions of interpolation by choosing parameters $P$ defining $\hat f(x;P)$ to minimize the error vector $\Bigl( \bigl | \hat f \bigl(x^{(1)};P\bigr) - f\bigl(x^{(1)}\bigr) \bigr|$, $\ldots$, $\bigl | \hat f \bigl(x^{(n)}; P\bigr) - f\bigl(x^{(n)}\bigr) \bigr | \Bigr)$ in some norm.
-The difficult question in the case of regression is often what parametric form to adopt for any given application.
-
-The most challenging problem when scaling in dimension is that the number of possible interactions between dimensions grows exponentially.
-Quantifying all possible interactions becomes intractable and hence beyond three-dimensional data, mostly linear models are used.
-That is not to say nonlinear models are absent, but nonlinearities are often either preconceived or model pairwise interactions between dimensions at most.
-Even globally nonlinear approximations such as neural networks are constructed from compositions of summed low-interaction functions \cite{clevert2015fast}.
-
-Provided the theoretical and practical motivations for exploring interpolants, this work aims to study the empirical performance differences between a set of scalable (moderately) interpolation techniques and a set of common regression techniques.
-These techniques are applied to a collection of moderately high dimensional problems ($5 \le d \le 30$) and the empirical results are discussed.
-
-The remainder of this paper is organized as follows.
-Sections \ref{sec:regression} and \ref{sec:interpolation} present the multivariate models.
-Section \ref{sec:error} presents the error measuring methodology that is used for collecting empirical results.
-Section \ref{sec:theory} presents the theoretical error bounds for interpolation.
-Section \ref{sec:data} presents the data sets for empirical approximation analysis and presents the approximation results for all models on these data sets.
-Section \ref{sec:discussion} analyzes and discusses the results, their implications, and their limitations.
-Finally, Section \ref{sec:conclusion} concludes.
+Regression and interpolation are problems of considerable importance
+that find applications across many fields of science. Pollution and
+air quality analysis \cite{de2008field}, energy consumption management
+\cite{lazos2014optimisation}, and student performance prediction
+\cite{cortez2008using,lux2016applications} are a few of many
+interdisciplinary applications of multivariate regression for
+predictive analysis. As discussed later, these techniques can also be
+applied to prediction problems related to forest fire risk assessment
+\cite{cortez2007data}, Parkinson's patient clinical evaluations
+\cite{tsanas2010accurate}, local rainfall and weather
+\cite{williams2009rattle}, credit card transactions
+\cite{pozzolo2015calibrating}, and high performance computing (HPC)
+file input/output (I/O) \cite{lux2018nonparametric}.
+
+Regression and interpolation have a considerable theoretical base in
+one dimension \cite{cheney2009course}. Splines in particular are well
+understood as an interpolation technique in one dimension
+\cite{de1978practical}, particularly B-splines. Tensor products of
+B-splines \cite{unther1996interpolating} or other basis functions have
+an unfortunate exponential scaling with increasing
+dimension. Exponential scaling prohibits tensor products from being
+reasonably applied beyond three-dimensional data. In order to address
+this dimensional scaling challenge, C. de Boor and others proposed box
+splines \cite{de2013box}, of which one of the approximation techniques
+in this work is composed \cite{lux2018novel}.
+
+The theoretical foundation of low dimensional interpolation allows the
+construction of strong error bounds that are absent from high
+dimensional problems. This work extends some known results regarding
+the secant method \cite{dennis1996numerical} to construct an
+interpolation error bound for problems of arbitrary dimension. These
+error bounds are useful, considering the same cannot be said for
+regression algorithms in general. The maximum complexity of an
+interpolant is bounded by the amount of data available, while the
+maximum complexity of a regressor is bounded by both the amount of
+data and the chosen parametric form. For this reason, generic uniform
+bounds are largely unobtainable for regression techniques on arbitrary
+approximation problems, even when the approximation domain is bounded.
+
+Aside from theoretical motivation for the use of interpolants, there
+are often computational advantages as well. Interpolants do not have
+the need for \textit{fitting} data, or minimizing error with respect
+to model parameters. In applications where the amount of data is large
+and the relative number of predictions that need to be made for a
+given collection of data is small, the direct application of an
+interpolant is much less computationally expensive.
+
+In this work, multivariate interpolation is defined given a closed
+convex subset $Y$ of a metrizable topological vector space with metric
+$s$, some function $f:\mathbb{R}^d \rightarrow Y$ and a set of points
+$X = \bigl\{x^{(1)}$,$\ldots$,$x^{(n)}\bigr\} \subset \mathbb{R}^d,$
+along with associated function values $f\bigl(x^{(i)}\bigr)$. The goal
+is to construct an approximation $\hat f: \mathbb{R}^d \rightarrow Y$
+such that $\hat f\bigl(x^{(i)}\bigr) = f\bigl(x^{(i)}\bigr)$ for all
+$i = 1$,$\ldots$,$n$. It is often the case that the form of the true
+underlying function $f$ is unknown, however it is still desirable to
+construct an approximation $\hat f$ with small approximation error at
+$y \notin X$. The two metric spaces that will be discussed in this
+work are the real numbers with metric $s(x,y) = |x-y|$, and the set of
+cumulative distribution functions (CDFs) with the Kolmogorov-Smirnov
+(KS) statistic as metric.
+
+Multivariate regression is often used when the underlying function is
+presumed to be stochastic, or stochastic error is introduced in the
+evaluation of $f$. Hence, multivariate regression relaxes the
+conditions of interpolation by choosing parameters $P$ defining $\hat
+f(x;P)$ to minimize the error vector $\Bigl( \bigl | \hat f
+\bigl(x^{(1)};P\bigr) - f\bigl(x^{(1)}\bigr) \bigr|$, $\ldots$, $\bigl
+| \hat f \bigl(x^{(n)}; P\bigr) - f\bigl(x^{(n)}\bigr) \bigr | \Bigr)$
+in some norm. The difficult question in the case of regression is
+often what parametric form to adopt for any given application.
+
+The most challenging problem when scaling in dimension is that the
+number of possible interactions between dimensions grows
+exponentially. Quantifying all possible interactions becomes
+intractable and hence beyond three-dimensional data, mostly linear
+models are used. That is not to say nonlinear models are absent, but
+nonlinearities are often either preconceived or model pairwise
+interactions between dimensions at most. Even globally nonlinear
+approximations such as neural networks are constructed from
+compositions of summed low-interaction functions
+\cite{clevert2015fast}.
+
+Provided the theoretical and practical motivations for exploring
+interpolants, this work aims to study the empirical performance
+differences between a set of scalable (moderately) interpolation
+techniques and a set of common regression techniques. These techniques
+are applied to a collection of moderately high dimensional problems
+($5 \le d \le 30$) and the empirical results are discussed.
+
+The remainder of this paper is organized as follows. Sections
+\ref{sec:regression} and \ref{sec:interpolation} present the
+multivariate models. Section \ref{sec:error} presents the error
+measuring methodology that is used for collecting empirical
+results. Section \ref{sec:theory} presents the theoretical error
+bounds for interpolation. Section \ref{sec:data} presents the data
+sets for empirical approximation analysis and presents the
+approximation results for all models on these data sets. Section
+\ref{sec:discussion} analyzes and discusses the results, their
+implications, and their limitations. Finally, Section
+\ref{sec:conclusion} concludes.
 
@@
 \label{sec:regression}
-Multivariate regressors are capable of accurately modeling a complex dependence of a response (in $Y$) on multiple variables (represented as a points in $\mathbb{R}^{d}$).
-The approximations to some (unknown) underlying function $f: \mathbb{R}^d \rightarrow Y$ are chosen to minimize some error measure related to data samples $f\bigl(x^{(i)}\bigr)$.
-For example, least squares regression uses the sum of squared differences between modeled function values and true function values as an error measure.
-In this section and the next, some techniques are limited to approximating real valued functions ($Y \subset \mathbb{R}$).
-These techniques can be extended to real vector-valued ranges by repeating the construction for each component of the vector output.
-Throughout the following, $x$ denotes a $d$-tuple, $x_i$ the $i$th component of $x$, and $x^{(i)}$ the $i$th $d$-tuple data point.
-Different symbols are used to represent the approximation function $\hat f$.
+Multivariate regressors are capable of accurately modeling a complex
+dependence of a response (in $Y$) on multiple variables (represented
+as a points in $\mathbb{R}^{d}$). The approximations to some (unknown)
+underlying function $f: \mathbb{R}^d \rightarrow Y$ are chosen to
+minimize some error measure related to data samples
+$f\bigl(x^{(i)}\bigr)$. For example, least squares regression uses the
+sum of squared differences between modeled function values and true
+function values as an error measure. In this section and the next,
+some techniques are limited to approximating real valued functions ($Y
+\subset \mathbb{R}$). These techniques can be extended to real
+vector-valued ranges by repeating the construction for each component
+of the vector output. Throughout the following, $x$ denotes a
+$d$-tuple, $x_i$ the $i$th component of $x$, and $x^{(i)}$ the $i$th
+$d$-tuple data point. Different symbols are used to represent the
+approximation function $\hat f$.
 
@@
 \label{sec:mars}
-This approximation was introduced in \cite{friedman1991multivariate} and subsequently improved to its current version in \cite{stanford1993fast}, called fast multivariate adaptive regression splines (Fast MARS).In Fast MARS, a least squares fit model is iteratively built by beginning with a single constant valued function and adding two new basis functions at each iteration of the form
+This approximation was introduced in \cite{friedman1991multivariate}
+and subsequently improved to its current version in
+\cite{stanford1993fast}, called fast multivariate adaptive regression
+splines (Fast MARS).In Fast MARS, a least squares fit model is
+iteratively built by beginning with a single constant valued function
+and adding two new basis functions at each iteration of the form
 \begin{align*}
@@
 \end{align*}
-where $j$ is the iteration number, $1 \le p \le n$, and $B_l(x)$, $B_k(x)$ are basis functions from the previous iteration,
+where $j$ is the iteration number, $1 \le p \le n$, and $B_l(x)$,
+$B_k(x)$ are basis functions from the previous iteration,
 $$w_+ = \begin{cases} w, & w \geq 0 \\ 0, & w < 0 \end{cases},$$
-and $w_- = (-w)_+$.
-After iteratively constructing a model, MARS then iteratively removes basis functions that do not contribute to goodness of fit.
-In effect, MARS creates a locally component-wise tensor product approximation of the data.
-The overall computational complexity of Fast MARS is $\mathcal{O}(n d m^3)$ where $m$ is the maximum number of underlying basis functions.
-This paper uses an implementation of Fast MARS \cite{rudy2017pyearth} with $m = 200$ throughout all experiments.
+and $w_- = (-w)_+$. After iteratively constructing a model, MARS then
+iteratively removes basis functions that do not contribute to goodness
+of fit. In effect, MARS creates a locally component-wise tensor
+product approximation of the data. The overall computational
+complexity of Fast MARS is $\mathcal{O}(n d m^3)$ where $m$ is the
+maximum number of underlying basis functions. This paper uses an
+implementation of Fast MARS \cite{rudy2017pyearth} with $m = 200$
+throughout all experiments.
 
@@
 \label{sec:svr}
-Support vector machines are a common method used in machine learning classification tasks that can be adapted for the purpose of regression \cite{basak2007support}.
-How to build a support vector regressor (SVR) is beyond the scope of this summary, but the resulting functional fit $p : \mathbb{R}^d \rightarrow \mathbb{R}$ has the form
+Support vector machines are a common method used in machine learning
+classification tasks that can be adapted for the purpose of regression
+\cite{basak2007support}. How to build a support vector regressor (SVR)
+is beyond the scope of this summary, but the resulting functional fit
+$p : \mathbb{R}^d \rightarrow \mathbb{R}$ has the form
 $$ p(x)  = \sum_{i=1}^{n}a_i K\bigl(x,x^{(i)}\bigr) + b ,$$
-where $K$ is the selected kernel function, $a_i \in \mathbb{R}^n$, $b \in \mathbb{R}$ are coefficients to be solved for simultaneously.
-The computational complexity of the SVR is $\mathcal{O}(n^2dm)$, with $m$ being determined by the minimization convergence criterion.
-This paper uses the scikit-learn SVR \cite{scikit-learn} with a polynomial kernel function.
+where $K$ is the selected kernel function, $a_i \in \mathbb{R}^n$, 
+$b \in \mathbb{R}$ are coefficients to be solved for simultaneously. The
+computational complexity of the SVR is $\mathcal{O}(n^2dm)$, with $m$
+being determined by the minimization convergence criterion. This paper
+uses the scikit-learn SVR \cite{scikit-learn} with a polynomial kernel
+function.
 
@@
 \label{sec:mlp}
-The neural network is a well studied and widely used method for both regression and classification tasks \cite{rumelhart1988learning,hornik1989multilayer}.
-When using the rectified linear unit (ReLU) activation function \cite{dahl2013improving} and fitting the model with stochastic gradient descent (SGD) or BFGS minimization techniques \cite{goh2017why,moller1993scaled,robbins1951stochastic}, the model built by a multilayer perceptron uses layers $l : \mathbb{R}^{i} \rightarrow \mathbb{R}^{j}$ defined by
+The neural network is a well studied and widely used method for both
+regression and classification tasks
+\cite{rumelhart1988learning,hornik1989multilayer}. When using the
+rectified linear unit (ReLU) activation function
+\cite{dahl2013improving} and fitting the model with stochastic
+gradient descent (SGD) or BFGS minimization techniques
+\cite{goh2017why,moller1993scaled,robbins1951stochastic}, the model
+built by a multilayer perceptron uses layers $l : \mathbb{R}^{i}
+\rightarrow \mathbb{R}^{j}$ defined by
 $$ l(u) = \bigl( u^t W_l \bigr)_+ ,$$
-where $W_l$ is the $i$ by $j$ weight matrix for layer $l$.
-In this form, the multilayer perceptron (MLP) produces a piecewise linear model of the input data.
-The computational complexity of training a multilayer perceptron is $\mathcal{O}(n d m)$, where $m$ is determined by the sizes of the layers of the network and the stopping criterion of the minimization used for finding weights.
-This paper uses a MLP built from Keras and Tensorflow to perform regression \cite{chollet2015keras,tensorflow2015-whitepaper} with ten hidden layers each having thirty two nodes (a total of approximately ten thousand parameters), ReLU activation, and five thousand epochs of SGD for training.
+where $W_l$ is the $i$ by $j$ weight matrix for layer $l$. In this
+form, the multilayer perceptron (MLP) produces a piecewise linear
+model of the input data. The computational complexity of training a
+multilayer perceptron is $\mathcal{O}(n d m)$, where $m$ is determined
+by the sizes of the layers of the network and the stopping criterion
+of the minimization used for finding weights. This paper uses a MLP
+built from Keras and Tensorflow to perform regression
+\cite{chollet2015keras,tensorflow2015-whitepaper} with ten hidden
+layers each having thirty two nodes (a total of approximately ten
+thousand parameters), ReLU activation, and five thousand epochs of SGD
+for training.
 
@@
 \label{sec:interpolation}
-The following interpolation techniques demonstrate a reasonable variety of approaches to interpolation.
-All of the presented interpolants produce approximations that are continuous in value, which is often a desirable property in applied approximation problems.
+The following interpolation techniques demonstrate a reasonable
+variety of approaches to interpolation. All of the presented
+interpolants produce approximations that are continuous in value,
+which is often a desirable property in applied approximation problems.
 
@@
 
-The Delaunay triangulation is a well-studied geometric technique for producing an interpolant \cite{lee1980two}.
-The Delaunay triangulation of a set of data points into simplices is such that there are no data points inside the sphere defined by the vertices of each simplex.
-For a $d$-simplex S with vertices $x^{(0)}$, $x^{(1)}$, $\ldots$, $x^{(d)}$, and function values $f\bigl(x^{(i)}\bigr)$, $i=0$, $\ldots$, $d$, $y \in S$ is a unique convex combination of the vertices,
+The Delaunay triangulation is a well-studied geometric technique for
+producing an interpolant \cite{lee1980two}. The Delaunay triangulation
+of a set of data points into simplices is such that there are no data
+points inside the sphere defined by the vertices of each simplex. For
+a $d$-simplex S with vertices $x^{(0)}$, $x^{(1)}$, $\ldots$,
+$x^{(d)}$, and function values $f\bigl(x^{(i)}\bigr)$, $i=0$,
+$\ldots$, $d$, $y \in S$ is a unique convex combination of the
+vertices,
 $$ y = \sum_{i=0}^{d} w_i x^{(i)}, \quad \sum_{i=0}^{d} w_i = 1, \quad w_i \geq 0, \quad i=0,\ldots,d, $$
@@
 $$ \hat f(y) = \sum_{i=0}^{d} w_i f\bigl(x^{(i)}\bigr). $$
-The computational complexity of Delaunay interpolation (for the implementation used) is $\mathcal{O}(n d^4 \log d)$, which is capable of scaling reasonably to $d \leq 50$.
-In the present application, a Delaunay simplex $S$ containing $y$ is found, then the $d+1$ vertices (points in $X$) of $S$ are used to assign weights to each vertex and produce the predicted function value for point $y$.
-
-\newpage
+The computational complexity of Delaunay interpolation (for the
+implementation used \cite{chang2018polynomial}) is $\mathcal{O}(n d^4
+\log d)$, which is capable of scaling reasonably to $d \leq 50$. In
+the present application, a Delaunay simplex $S$ containing $y$ is
+found, then the $d+1$ vertices (points in $X$) of $S$ are used to
+assign weights to each vertex and produce the predicted function value
+for point $y$.
+
 \subsection{Modified Shepard}
@@
 
-The modified Shepard method used here (ShepMod) generates a continuous approximation based on Euclidean distance and resembles a nearest neighbor interpolant \cite{cover1967nearest}.
-This model is a type of \textit{metric interpolation}, also called a Shepard method \cite{shepard1968two,gordon1978shepard}.
-The interpolant has the form
-$$ p(x) = \frac{\sum\limits_{k=1}^{n}W_k(x)f\bigl(x^{(k)}\bigr)}{\sum\limits_{k=1}^{n}W_k(x)} ,$$
-where $W_k(x) \propto \bigl\|x - x^{(k)}\bigr\|_2^{-2}$ and decreases to zero at a radius $r_k \in \mathbb{R}$ such that exactly $d+1$ other points $x^{(j)}$, $j \not = k$ are inside the closed Euclidean ball of radius $r_k$ about $x^{(k)}$.
-The computational complexity of ShepMod is $\mathcal{O}(n^2d)$.
-This paper uses a Fortran 95 implementation of ShepMod derived from SHEPPACK \cite{thacker2010algorithm}.
+The modified Shepard method used here (ShepMod) generates a continuous
+approximation based on Euclidean distance and resembles a nearest
+neighbor interpolant \cite{cover1967nearest}. This model is a type of
+\textit{metric interpolation}, also called a Shepard method
+\cite{shepard1968two,gordon1978shepard}. The interpolant has the form
+$$ p(x) = \frac{\sum\limits_{k=1}^{n}W_k(x)f\bigl(x^{(k)}\bigr)}{\sum\limits_{k=1}^{n}W_k(x)},$$
+where $x \not = x^{(k)}$, $k = 1$, $\ldots$, $n$, for weighting functions
+$$W_k(x) = \left( \frac{\big(r_k - \bigl\|x - x^{(k)}\bigr\|_2\big)_+}{r_k \bigl\|x - x^{(k)}\bigr\|_2} \right)^2,$$ 
+where $r_k \in \mathbb{R}$ are chosen such that at least $d+1$ other
+points $x^{(j)}$, $j \not = k$ are inside the closed Euclidean ball of
+radius $r_k$ about $x^{(k)}$. The computational complexity of ShepMod
+is $\mathcal{O}(n^2d)$. This paper uses a Fortran 95 implementation of
+ShepMod derived from SHEPPACK \cite{thacker2010algorithm}.
 
 \subsection{Linear Shepard}
-The linear Shepard method (LSHEP) is a blending function using local linear interpolants, a special case of the general Shepard algorithm \cite{thacker2010algorithm}. The interpolant has the form
+The linear Shepard method (LSHEP) is a blending function using local
+linear interpolants, a special case of the general Shepard algorithm
+\cite{thacker2010algorithm}. The interpolant has the form
 $$ p(x) = \frac{\sum\limits_{k=1}^{n}W_k(x)P_k(x)}{\sum\limits_{k=1}^{n}W_k(x)} ,$$
-where $W_k(x)$ is a locally supported weighting function and $P_k(x)$ is a local linear approximation to the data satisfying $P_k\bigl(x^{(k)}\bigr) = f\bigl(x^{(x)}\bigr)$.
-The computational complexity of LSHEP is $\mathcal{O}(n^2d^3)$.
-This paper uses the Fortran 95 implementation of LSHEP in SHEPPACK \cite{thacker2010algorithm}.
+where $W_k(x)$ is a locally supported weighting function and $P_k(x)$
+is a local linear approximation to the data satisfying
+$P_k\bigl(x^{(k)}\bigr) = f\bigl(x^{(x)}\bigr)$. The computational
+complexity of LSHEP is $\mathcal{O}(n^2d^3)$. This paper uses the
+Fortran 95 implementation of LSHEP in SHEPPACK
+\cite{thacker2010algorithm}.
 
@@
 
-The box spline model used here is an interpolation technique built from overlapping box splines \cite{de2013box}.
-The box splines serve as basis functions that are shifted and scaled to have support over box shaped regions.
-The boxes are constructed in a way to guarantee a covering of the domain \cite{lux2018novel}.
-Given a set of box splines $\bigl\{b^{x^{(i)}}\bigr\}$ with the iterative box properties outlined in \cite{lux2018novel} and anchored at interior points $\bigl\{x^{(i)}\bigr\}$,
+The box spline model used here is an interpolation technique built
+from overlapping box splines \cite{de2013box}. The box splines serve
+as basis functions that are shifted and scaled to have support over
+box shaped regions. The boxes are constructed in a way to guarantee a
+covering of the domain \cite{lux2018novel}. Given a set of box splines
+$\bigl\{b^{x^{(i)}}\bigr\}$ with the iterative box properties outlined
+in \cite{lux2018novel} and anchored at interior points
+$\bigl\{x^{(i)}\bigr\}$,
 $$ \hat f(y) = \frac{\sum\limits_{i=1}^n b^{x^{(i)}}(y) f\bigl(x^{(i)}\bigr)}{\sum\limits_{i=1}^n b^{x^{(i)}}(y)}. $$
-Note that the box splines always satisfy $b^{x^{(i)}}\bigl(x^{(j)}\bigr) = \delta_{ij}$ and $b^{x^{(i)}}(y) \geq 0$.
-The computational complexity of interpolation via the box spline model is $\mathcal{O}(n^2 d)$.
+Note that the box splines always satisfy
+$b^{x^{(i)}}\bigl(x^{(j)}\bigr) = \delta_{ij}$ and $b^{x^{(i)}}(y)
+\geq 0$. The computational complexity of interpolation via the box
+spline model is $\mathcal{O}(n^2 d)$.
 
@@
 
-A well-studied technique for classification and approximation is the nearest neighbor algorithm \cite{cover1967nearest}.
-Nearest neighbor inherently utilizes the convex region $C^{x^{(i)}}$ (Voronoi cell \cite{dirichlet1850reduction}) consisting of all points closer to $x^{(i)}$ than to any other point $x^{(j)}$.
-The Voronoi model smooths the nearest neighbor approximation by utilizing the Voronoi cells to define support via a generic basis function $v: \mathbb{R}^d \rightarrow \mathbb{R}_+$ given by
+A well-studied technique for classification and approximation is the
+nearest neighbor algorithm \cite{cover1967nearest}. Nearest neighbor
+inherently utilizes the convex region $C^{x^{(i)}}$ (Voronoi cell
+\cite{dirichlet1850reduction}) consisting of all points closer to
+$x^{(i)}$ than to any other point $x^{(j)}$. The Voronoi model smooths
+the nearest neighbor approximation by utilizing the Voronoi cells to
+define support via a generic basis function $v: \mathbb{R}^d
+\rightarrow \mathbb{R}_+$ given by
 $$ v^{x^{(i)}}(y) = \left(1 - \frac{\bigl\|y - x^{(i)}\bigr\|_2}{2 \ h\bigl(y - x^{(i)} \mid x^{(i)}\bigr)} \right)_+, $$
-where $h\bigl(w \mid x^{(i)}\bigr)$ is the distance between $x^{(i)}$ and the boundary of the Voronoi cell $C^{x^{(i)}}$ in the direction $w$.
-$v^{x^{(i)}}\bigl(x^{(j)}\bigr) = \delta_{ij}$ and $v^{x^{(i)}}$ has local support, giving the interpolated value
+where $h\bigl(w \mid x^{(i)}\bigr)$ is the distance between $x^{(i)}$
+and the boundary of the Voronoi cell $C^{x^{(i)}}$ in the direction
+$w$. $v^{x^{(i)}}\bigl(x^{(j)}\bigr) = \delta_{ij}$ and $v^{x^{(i)}}$
+has local support, giving the interpolated value
 $$ f(y) = \frac{\sum\limits_{i=1}^n v^{x^{(i)}}(y) f\bigl(x^{(i)}\bigr)}{\sum\limits_{i=1}^n v^{x^{(i)}}(y)}, $$
-where $0 \leq v^{x^{(i)}}(y) \leq 1$.
-The computational complexity of interpolation via this Voronoi model is $\mathcal{O}(n^2 d)$.
-All of the approximations are an interpolant involving a convex combination of known function values $f\bigl(x^{(i)}\bigr)$.
+where $0 \leq v^{x^{(i)}}(y) \leq 1$. The computational complexity of
+interpolation via this Voronoi model is $\mathcal{O}(n^2 d)$. All of
+the approximations are an interpolant involving a convex combination
+of known function values $f\bigl(x^{(i)}\bigr)$.
 
@@
   \vspace{-0.3cm}
-  \caption{In this HPC I/O example, the general methodology for predicting a CDF and evaluating error can be seen.
-The Delaunay method chose three source distributions (dotted lines) and assigned weights \{.3, .4, .3\} (top to bottom at arrow).
-The weighted sum of the three known CDFs produces the predicted CDF (dashed line).
-The KS Statistic (arrow) computed between the true CDF (solid line) and predicted CDF (dashed line) is 0.2 for this example.
-The KS test null hypothesis is rejected at $p$-value 0.01, however it is not rejected at $p$-value 0.001.
+  \caption{In this HPC I/O example, the general methodology for
+    predicting a CDF and evaluating error can be seen. The Delaunay
+    method chose three source distributions (dotted lines) and
+    assigned weights \{.3, .4, .3\} (top to bottom at arrow). The
+    weighted sum of the three known CDFs produces the predicted CDF
+    (dashed line). The KS Statistic (arrow) computed between the true
+    CDF (solid line) and predicted CDF (dashed line) is 0.2 for this
+    example. The KS test null hypothesis is rejected at $p$-value
+    0.01, however it is not rejected at $p$-value 0.001.
   \vspace{-.1cm}}
@@
 
-When the range of an approximation is the real numbers, error is reported with summary statistics including: min absolute error, max absolute error, and absolute error quartiles.
-When the range of an approximation is the space of cumulative distribution functions, the Kolmogorov-Smirnov statistic (max-norm difference between the functions) is used.
-
-A hurdle when modeling function-valued outputs such as cumulative distribution functions (CDFs) or probability density functions (PDFs) is that certain properties must be maintained.
-It is necessary that a PDF $f: \mathbb{R} \rightarrow \mathbb{R}$ have the properties $f(x) \geq 0$ and $\int_{-\infty}^{\infty}f(x)dx = 1$.
-However, for a CDF $F: \mathbb{R} \rightarrow \mathbb{R}$ the properties are $F(x) \in [0,1]$ and $F(x)$ is absolutely continuous and nondecreasing.
-This work utilizes the fact that a convex combination of CDFs (or PDFs) results in a valid CDF.
-Given $G(x) = \sum_{i}w_i F_i(x)$, $\sum_{i} w_i = 1$, $w_i \geq 0$, and each $F_i$ is a valid CDF, $G$ must also be a valid CDF.
-A demonstration of how this is applied can be seen in Figure \ref{fig:prediction-example}.
-
-The performance of approximation techniques that predict probability functions can be analyzed through a variety of summary statistics.
-This work uses the max absolute difference, also known as the Kolmogorov-Smirnov (KS) statistic \cite{lilliefors1967kolmogorov} for its compatibility with the KS test.
-
-The two-sample KS test is a useful nonparametric test for comparing two CDFs while only assuming stationarity, finite mean, and finite variance.
-The null hypothesis (that two CDFs come from the same underlying distribution) is rejected at level $p \in [0,1]$ when
+When the range of an approximation is the real numbers, error is
+reported with summary statistics including: min absolute error, max
+absolute error, and absolute error quartiles. When the range of an
+approximation is the space of cumulative distribution functions, the
+Kolmogorov-Smirnov statistic (max-norm difference between the
+functions) is used.
+
+A hurdle when modeling function-valued outputs such as cumulative
+distribution functions (CDFs) or probability density functions (PDFs)
+is that certain properties must be maintained. It is necessary that a
+PDF $f: \mathbb{R} \rightarrow \mathbb{R}$ have the properties $f(x)
+\geq 0$ and $\int_{-\infty}^{\infty}f(x)dx = 1$. However, for a CDF
+$F: \mathbb{R} \rightarrow \mathbb{R}$ the properties are
+$F(x) \in [0,1]$ and $F(x)$ is absolutely continuous and
+nondecreasing. This work utilizes the fact that a convex combination
+of CDFs (or PDFs) results in a valid CDF. Given $G(x) = \sum_{i}w_i
+F_i(x)$, $\sum_{i} w_i = 1$, $w_i \geq 0$, and each $F_i$ is a valid
+CDF, $G$ must also be a valid CDF. A demonstration of how this is
+applied can be seen in Figure \ref{fig:prediction-example}.
+
+The performance of approximation techniques that predict probability
+functions can be analyzed through a variety of summary
+statistics. This work uses the max absolute difference, also known as
+the Kolmogorov-Smirnov (KS) statistic \cite{lilliefors1967kolmogorov}
+for its compatibility with the KS test.
+
+The two-sample KS test is a useful nonparametric test for comparing
+two CDFs while only assuming stationarity, finite mean, and finite
+variance. The null hypothesis (that two CDFs come from the same
+underlying distribution) is rejected at level $p \in [0,1]$ when
 $$ KS > \sqrt{-\frac{1}{2}\ln\biggl(\frac{p}{2}\biggr)} \sqrt{\frac{1}{n_1} + \frac{1}{n_2}}, $$
-with distribution sample sizes $n_1,n_2 \in \mathcal{N}$.
-For all applications of the KS test presented in this work $n_1 = n_2$.
-An example of the round-trip prediction methodology from known and predicted distributions to the calculation of error can be seen in Figure \ref{fig:prediction-example}.
+with distribution sample sizes $n_1,n_2 \in \mathcal{N}$. For all
+applications of the KS test presented in this work $n_1 = n_2$. An
+example of the round-trip prediction methodology from known and
+predicted distributions to the calculation of error can be seen in
+Figure \ref{fig:prediction-example}.
 
@@
 
-This section presents the theoretical results bounding the error of Delaunay interpolation.
-The error analysis relies on the Delaunay interpolation for two reasons: (1) second order results can be obtained utilizing a Lipschitz constant on the gradient of a function, rather than standard Lipschitz bounds and (2) multiple other interpolants analyzed compute predictions as convex combinations of observed function values, which may allow for straight forward extensions of this error bound.
+This section presents the theoretical results bounding the error of
+Delaunay interpolation. The error analysis relies on the Delaunay
+interpolation for two reasons: (1) second order results can be
+obtained utilizing a Lipschitz constant on the gradient of a function,
+rather than standard Lipschitz bounds and (2) multiple other
+interpolants analyzed compute predictions as convex combinations of
+observed function values, which may allow for straight forward
+extensions of this error bound.
 
@@
   \label{lemma:1}
-  Let $S \subset \mathbb{R}^d$ be open and convex, $f: \mathbb{R}^d \rightarrow \mathbb{R}$, and $\nabla f \in Lip_{(\gamma,\|\cdot\|_2)}(S)$, the set of $\gamma$-Lipschitz continuous functions in the $2$-norm.
-Then for all $x,y \in S$
+  Let $S \subset \mathbb{R}^d$ be open and convex, $f: \mathbb{R}^d
+  \rightarrow \mathbb{R}$, and $\nabla f \in
+  Lip_{(\gamma,\|\cdot\|_2)}(S)$, the set of $\gamma$-Lipschitz
+  continuous functions in the $2$-norm. Then for all $x,y \in S$
   $$\big|f(y) - f(x) - \langle \nabla f(x), y - x \rangle \big| \leq \frac{\gamma \|y - x\|_2^2}{2}.$$
@@
 \begin{proofdot}
-  Consider the function $g(t) = f \big((1-t) x + t y \big)$, $0 \leq t \leq 1$, whose derivative $g'(t) = \big\langle \nabla f \big((1-t) x + t y \big), y - x \big\rangle$ is the directional derivative of $f$ in the direction $(y - x).$
+  Consider the function $g(t) = f \big((1-t) x + t y \big)$, $0 \leq t
+  \leq 1$, whose derivative $g'(t) = \big\langle \nabla f \big((1-t) x
+  + t y \big), y - x \big\rangle$ is the directional derivative of $f$
+  in the direction $(y - x).$
   \begin{align*}
@@
   \label{lemma:2}
-  Let $x, y, v_i \in \mathbb{R}^d$, $c_i \in \mathbb{R}$, and $|\langle y - x, v_i \rangle| \leq c_i$ for $i = 1$, $\ldots$, $d.$ If $M = (v_1$, $\ldots$, $v_d)$ is nonsingular, then
+  Let $x, y, v_i \in \mathbb{R}^d$, $c_i \in \mathbb{R}$, and
+  $|\langle y - x, v_i \rangle| \leq c_i$ for $i = 1$, $\ldots$, $d.$
+  If $M = (v_1$, $\ldots$, $v_d)$ is nonsingular, then
   $$\|y - x\|_2^2 \leq \frac{1}{\sigma_d^2} \sum_{i=1}^d c_i^2,$$
@@
 \begin{proofdot}
-  Using the facts that $M$ and $M^t$ have the same singular values, and $\|M^tw\|_2 \geq \sigma_d \|w\|_2$, gives
+  Using the facts that $M$ and $M^t$ have the same singular values,
+  and $\|M^tw\|_2 \geq \sigma_d \|w\|_2$, gives
   \begin{align*}
@@
   \label{lemma:3}
-  Given $f$, $\gamma$, $S$ as in {\it Lemma \ref{lemma:1}}, let $X = \{x_0$, $x_1$, $\ldots$, $x_d\}$ $\subset S$ be the vertices of a $d$-simplex, and let $\hat f(x) = \langle c, x - x_0 \rangle + f(x_0)$, $c \in \mathbb{R}^d$ be the linear function interpolating $f$ on $X.$ Let $\sigma_d$ be the smallest singular value of the matrix $M = (x_1 - x_0$, $\ldots$, $x_d - x_0)$, and $k = \max\limits_{1\ \leq\ j\ \leq\ d} \|x_j - x_0\|_2.$ Then
+  Given $f$, $\gamma$, $S$ as in {\it Lemma \ref{lemma:1}}, let $X =
+  \{x_0$, $x_1$, $\ldots$, $x_d\}$ $\subset S$ be the vertices of a
+  $d$-simplex, and let $\hat f(x) = \langle c, x - x_0 \rangle +
+  f(x_0)$, $c \in \mathbb{R}^d$ be the linear function interpolating
+  $f$ on $X.$ Let $\sigma_d$ be the smallest singular value of the
+  matrix $M = (x_1 - x_0$, $\ldots$, $x_d - x_0)$, and $k =
+  \max\limits_{1\ \leq\ j\ \leq\ d} \|x_j - x_0\|_2.$ Then
   $$\big\|\nabla f(x_0) - c\big\|_2 \leq \sqrt{d} \frac{\gamma k^2}{\sigma_d}.$$
@@
 \begin{proofdot}
-  Consider $f(x) - \hat f(x)$ along the line segment $z(t) = (1-t)x_0 + t x_j$, $0 \leq t \leq 1.$ By Rolle's Theorem, for some $0 < \hat t < 1$, $\big\langle \nabla f\big(z(\ \hat t\ )\big) - c, x_j - x_0 \big\rangle = 0.$ Now
+  Consider $f(x) - \hat f(x)$ along the line segment $z(t) = (1-t)x_0
+  + t x_j$, $0 \leq t \leq 1.$ By Rolle's Theorem, for some $0 < \hat
+  t < 1$, $\big\langle \nabla f\big(z(\ \hat t\ )\big) - c, x_j - x_0
+  \big\rangle = 0.$ Now
   \begin{align*}
@@
   for all $1 \leq j \leq d.$ Using {\it Lemma \ref{lemma:2}},
-  $$\big \| \nabla f(x_i) - c \big \|_2^2 \leq \frac{d}{\sigma_d^2} \big( \gamma k^2\big)^2 \implies \big \| \nabla f(x_i) - c \big \|_2 \leq \sqrt{d} \frac{\gamma k^2}{\sigma_d}.$$
-  \qed
+  $$\big \| \nabla f(x_i) - c \big \|_2^2 \leq \frac{d}{\sigma_d^2}
+  \big( \gamma k^2\big)^2 \implies \big \| \nabla f(x_i) - c \big \|_2
+  \leq \sqrt{d} \frac{\gamma k^2}{\sigma_d}.$$ \qed
 \end{proofdot}
@@
 
-In summary, the approximation error of a Delaunay interpolant tends quadratically towards zero when approaching observed data only when the diameter of the interpolating simplex is also reduced at a proportional rate.
-Without the incorporation of additional observations, only linear convergence to the true function is be achieved.
-Notice that the approximation error is largely determined by the spacing of observed data.
-Predictions made by simplices whose vertices are not well-spaced (i.e.
-have large diameter, or are nearly contained in a hyperplane) have higher error.
-In light of this error bound, an empirical evaluation of presented algorithms follows.
+In summary, the approximation error of a Delaunay interpolant tends
+quadratically towards zero when approaching observed data only when
+the diameter of the interpolating simplex is also reduced at a
+proportional rate. Without the incorporation of additional
+observations, only linear convergence to the true function is be
+achieved. Notice that the approximation error is largely determined by
+the spacing of observed data. Predictions made by simplices whose
+vertices are not well-spaced (i.e. have large diameter, or are nearly
+contained in a hyperplane) have higher error. In light of this error
+bound, an empirical evaluation of presented algorithms follows.
 
@@
 
-The theoretical results constructed in Section \ref{sec:theory} for Delaunay interpolation are promising, however they are difficult to interpret in context with the other approximation algorithms that do not have similar known uniform error bounds.
-For that reason, this paper utilizes five different data sets of varying dimension and application to construct approximations and compare the accuracy of different techniques.
-
-In the following five subsections the sources and targets of each test data set are described, as well as challenges and limitations related to approximating these data.
-The distribution of response values being modeled is presented followed by the distribution of approximation errors for each algorithm.
-
-Before approximations are constructed, all five data sets are rescaled such that the domain of approximation is the unit hypercube.
-The range of the first four data sets is the real numbers, while the range of the fifth data set is the space of cumulative distribution functions.
-All approximation techniques are applied to the first four data sets, while only those interpolants whose approximations are convex combinations of observed data are applied to the final data set.
-
-All approximations are constructed using $10$-fold cross validation as described in \cite{kohavi1995study}.
-This approach requires the division of data randomly into $10$ partitions (the same partitions are used by each algorithm).
-An algorithm is then evaluated by constructing an approximation over each set of $9$ partitions, predicting the values of the final partition.
-As a result, each observed data point is used in the construction of $9$ different approximations and is approximated exactly once.
-The $k$-fold cross validation method is data-efficient and provides an unbiased estimate of the expected prediction error \cite{kohavi1995study}, however it should be noted that neither this method nor others can provide a universally unbiased estimator for the variance of prediction error \cite{bengio2004no}.
-
-In addition to the figures displaying approximation results for each data set, tables of accompanying numerical results are located in the Appendix.
-All of the test data sets capture underlying functions that are almost certainly stochastic.
-As described in Section \ref{sec:introduction}, regression techniques appear most appropriate for these problems.
-However, in high dimension data grows exponentially more sparse.
-Given sparse data, regressors tend towards interpolation.
+The theoretical results constructed in Section \ref{sec:theory} for
+Delaunay interpolation are promising, however they are difficult to
+interpret in context with the other approximation algorithms that do
+not have similar known uniform error bounds. For that reason, this
+paper utilizes five different data sets of varying dimension and
+application to construct approximations and compare the accuracy of
+different techniques.
+
+In the following five subsections the sources and targets of each test
+data set are described, as well as challenges and limitations related
+to approximating these data. The distribution of response values being
+modeled is presented followed by the distribution of approximation
+errors for each algorithm.
+
+All five data sets are rescaled such that the domain of approximation
+is the unit hypercube. The range of the first four data sets is the
+real numbers, while the range of the fifth data set is the space of
+cumulative distribution functions. All approximation techniques are
+applied to the first four data sets, while only those interpolants
+whose approximations are convex combinations of observed data are
+applied to the final data set.
+
+All approximations are constructed using $k$-fold cross validation as
+described in \cite{kohavi1995study} with $k=10$. This approach
+randomly partitions data into $k$ nearly equal sized sets. Each
+algorithm is then evaluated by constructing an approximation over each
+unique union of $k-1$ elements of the partition, making predictions
+for points in the remaining element. As a result, each observed data
+point is used in the construction of $k-1$ different approximations
+and is approximated exactly once. The $k$-fold cross validation method
+is data-efficient and provides an unbiased estimate of the expected
+prediction error \cite{kohavi1995study}, however it should be noted
+that neither this method nor others can provide a universally unbiased
+estimator for the variance of prediction error \cite{bengio2004no}.
+
+In addition to the figures displaying approximation results for each
+data set, tables of accompanying numerical results are located in the
+Appendix. All of the test data sets capture underlying functions that
+are almost certainly stochastic. As described in Section
+\ref{sec:introduction}, regression techniques appear most appropriate
+for these problems. However, in high dimension data grows
+exponentially more sparse. Given sparse data, regressors tend towards
+interpolation.
 
@@
   \end{tikzpicture}
-  \caption{Above is a histogram of forest fire area burned under recorded weather conditions.
-The data is presented on a natural log scale because most values are small with exponentially fewer fires on record that burn large areas.}
+  \caption{Above is a histogram of forest fire area burned under
+    recorded weather conditions. The data is presented on a natural
+    log scale because most values are small with exponentially fewer
+    fires on record that burn large areas.}
   \label{fig:hist-forest-fire}
@@
   \end{tikzpicture}
-  \caption{All models are applied to approximate the amount of area that would be burned given environment conditions.
-The data is broken up into $10$ partitions and each set of $9$ partitions is used to construct a model that approximates the remaining partition.
-This results in exactly one prediction from each algorithm for each data point.
-These boxes depict the median (middle bar), median $95\%$ confidence interval (cones), quartiles (box edges), fences at $3/2$ interquartile range (whiskers), and outliers (dots) of absolute prediction error for each model.
-Similar to figure \ref{fig:hist-forest-fire}, the errors are presented on a natural log scale.
-The complementing numerical data to this figure is provided in table \ref{table:error-forest-fire} in the Appendix.}
+  \caption{All models are applied to approximate the amount of area
+    that would be burned given environment conditions. $10$-fold
+    cross validation as described in the beginning of Section
+    \ref{sec:data} is used to evaluated each algorithm. This results
+    in exactly one prediction from each algorithm for each data
+    point. These boxes depict the median (middle bar), median $95\%$
+    confidence interval (cones), quartiles (box edges), fences at
+    $3/2$ interquartile range (whiskers), and outliers (dots) of
+    absolute prediction error for each model. Similar to figure
+    \ref{fig:hist-forest-fire}, the errors are presented on a natural
+    log scale. The complementing numerical data to this figure is
+    provided in table \ref{table:error-forest-fire} in the Appendix.}
   \label{fig:error-forest-fire}
@@
 
-The forest fire data set \cite{cortez2007data} describes the area of Montesinho park burned over months of the year along with environmental conditions.
-The twelve dimensions being used to model burn area are the $x$ and $y$ spatial coordinates of burns in the park, month of year (mapped to a 2D circle), the FFMC, DMC, DC, and ISI indices (see source for details), the temperature, relative humidity, wind speed, and outdoor rain.
-The original analysis of this data set demonstrated it to be difficult to model, likely due to the skew in response values.
-
-As suggested by figure \ref{fig:error-forest-fire}, the SVR has the lowest absolute prediction errors for 80\% of the data, with ShepMod and BoxSpline being the nearest overall competitors.
-The effectiveness of SVR on this data suggests the underlying function can be defined by relatively few parameters, as well as the importance of capturing the low-burn-area data points.
+The forest fire data set \cite{cortez2007data} describes the area of
+Montesinho park burned over months of the year along with
+environmental conditions. The twelve dimensions being used to model
+burn area are the $x$ and $y$ spatial coordinates of burns in the
+park, month of year (mapped to a 2D circle), the FFMC, DMC, DC, and
+ISI indices (see source for details), the temperature, relative
+humidity, wind speed, and outdoor rain. The original analysis of this
+data set demonstrated it to be difficult to model, likely due to the
+skew in response values.
+
+As suggested by figure \ref{fig:error-forest-fire}, the SVR has the
+lowest absolute prediction errors for 80\% of the data, with ShepMod
+and BoxSpline being the nearest overall competitors. The effectiveness
+of SVR on this data suggests the underlying function can be defined by
+relatively few parameters, as well as the importance of capturing the
+low-burn-area data points.
 %% -------------------------------------------------------------------
@@
   \end{tikzpicture}
-  \caption{Above is a histogram of the Parkinson's patient total UPDRS clinical scores that will be approximated by each algorithm.}
+  \caption{Above is a histogram of the Parkinson's patient total UPDRS
+    clinical scores that will be approximated by each algorithm.}
   \label{fig:hist-parkinsons}
@@
   \end{tikzpicture}
-  \caption{All models are applied to approximate the total UPDRS score given audio features from patients' home life.
-The data is broken up into $10$ partitions and each set of $9$ partitions is used to construct a model that approximates the remaining partition.
-This results in exactly one prediction from each algorithm for each data point.
-These boxes depict the median (middle bar), median $95\%$ confidence interval (cones), quartiles (box edges), fences at $3/2$ interquartile range (whiskers), and outliers (dots) of absolute prediction error for each model.
-The complementing numerical data to this is provided in table \ref{table:error-parkinsons} in the Appendix.}
+  \caption{All models are applied to approximate the total UPDRS score
+    given audio features from patients' home life. $10$-fold
+    cross validation as described in the beginning of Section
+    \ref{sec:data} is used to evaluated each algorithm. This results
+    in exactly one prediction from each algorithm for each data
+    point. These boxes depict the median (middle bar), median $95\%$
+    confidence interval (cones), quartiles (box edges), fences at
+    $3/2$ interquartile range (whiskers), and outliers (dots) of
+    absolute prediction error for each model. The complementing
+    numerical data to this is provided in table
+    \ref{table:error-parkinsons} in the Appendix.}
   \label{fig:error-parkinsons}
@@
 
-The second data set for evaluation \cite{tsanas2010accurate} is derived from a speech monitoring study with the intent to automatically estimate Parkinson's disease symptom development in Parkinson's patients.
-The function to be approximated is a time-consuming clinical evaluation measure referred to as the UPDRS score.
-The total UPDRS score given by a clinical evaluation is estimated through 16 real numbers generated from biomedical voice measures of in-home sound recordings.
-
-Figure \ref{fig:error-parkinsons} shows the ShepMod algorithm has the lowest minimum, first quartile, and median of absolute errors for this problem, while providing the best prediction 66\% of the time.
-The MLP has the lowest third quartile and provides the best prediction for 32\% of approximations.
-The dominance of ShepMod may be in part to regular-interval total UPDRS scores provided by clinicians, favoring a nearest-neighbor strategy of prediction.
+The second data set for evaluation \cite{tsanas2010accurate} is
+derived from a speech monitoring study with the intent to
+automatically estimate Parkinson's disease symptom development in
+Parkinson's patients. The function to be approximated is a
+time-consuming clinical evaluation measure referred to as the UPDRS
+score. The total UPDRS score given by a clinical evaluation is
+estimated through 16 real numbers generated from biomedical voice
+measures of in-home sound recordings.
+
+Figure \ref{fig:error-parkinsons} shows the ShepMod algorithm has the
+lowest minimum, first quartile, and median of absolute errors for this
+problem, while providing the best prediction 66\% of the time. The MLP
+has the lowest third quartile and provides the best prediction for
+32\% of approximations. The dominance of ShepMod may be in part to
+regular-interval total UPDRS scores provided by clinicians, favoring a
+nearest-neighbor strategy of prediction.
 %% -------------------------------------------------------------------
@@
   \end{tikzpicture}
-  \caption{Above is a histogram of daily rainfall in Sydney, Australia.
-The data is presented on a natural log scale because the frequency of larger amounts of rainfall is significantly less.
-There is a peak in occurrence of the value $0$, which has a notable effect on the resulting model performance.}
+  \caption{Above is a histogram of daily rainfall in Sydney,
+    Australia. The data is presented on a natural log scale because
+    the frequency of larger amounts of rainfall is significantly
+    less. There is a peak in occurrence of the value $0$, which has a
+    notable effect on the resulting model performance.}
   \label{fig:hist-weather}
@@
   \end{tikzpicture}
-  \caption{All models are applied to approximate the amount of rainfall expected on the next calendar day given various sources of local meteorological data.
-The data is broken up into $10$ partitions and each set of $9$ partitions is used to construct models that approximates the remaining partition.
-This results in exactly one prediction from each algorithm for each data point.
-These boxes depict the median (middle bar), median $95\%$ confidence interval (cones), quartiles (box edges), fences at $3/2$ interquartile range (whiskers), and outliers (dots) of absolute prediction error for each model.
-The errors are presented on a natural log scale, mimicking the presentation in figure \ref{fig:hist-weather}.
-The complementing numerical data to this figure is provided in table \ref{table:error-weather} in the Appendix.}
+  \caption{All models are applied to approximate the amount of
+    rainfall expected on the next calendar day given various sources
+    of local meteorological data. $10$-fold cross validation as
+    described in the beginning of Section \ref{sec:data} is used to
+    evaluated each algorithm. This results in exactly one prediction
+    from each algorithm for each data point. These boxes depict the
+    median (middle bar), median $95\%$ confidence interval (cones),
+    quartiles (box edges), fences at $3/2$ interquartile range
+    (whiskers), and outliers (dots) of absolute prediction error for
+    each model. The errors are presented on a natural log scale,
+    mimicking the presentation in figure \ref{fig:hist-weather}. The
+    complementing numerical data to this figure is provided in table
+    \ref{table:error-weather} in the Appendix.}
   \label{fig:error-weather}
@@
 
-The third data regarding the total daily rainfall in Sydney, Australia \cite{williams2009rattle} provides a slightly higher dimension challenge for the interpolants and regressors.
-This data is composed of many meteorological readings from the region on a day including: min and max temperatures, sunshine, wind speed directions (converted to coordinates on circle), wind speeds, and humidities throughout the day, day of the year (converted to coordinates on circle), and the model must predict the amount of rainfall tomorrow.
-
-While figure \ref{fig:hist-weather} makes MARS look far better than other techniques, it only provides the best prediction for 11\% of points.
-The MLP has the lowest absolute error for 56\% of points and LSHEP is best for 28\%.
-MARS likely achieves such a low first quartile due to the high occurrence of the value 0 in the data.
+The third data regarding the total daily rainfall in Sydney, Australia
+\cite{williams2009rattle} provides a slightly higher dimension
+challenge for the interpolants and regressors. This data is composed
+of many meteorological readings from the region on a day including:
+min and max temperatures, sunshine, wind speed directions (converted
+to coordinates on circle), wind speeds, and humidities throughout the
+day, day of the year (converted to coordinates on circle), and the
+model must predict the amount of rainfall tomorrow.
+
+While figure \ref{fig:hist-weather} makes MARS look far better than
+other techniques, it only provides the best prediction for 11\% of
+points. The MLP has the lowest absolute error for 56\% of points and
+LSHEP is best for 28\%. MARS likely achieves such a low first quartile
+due to the high occurrence of the value 0 in the data.
 %% -------------------------------------------------------------------
@@
   \end{tikzpicture}
-  \caption{Above is a histogram of credit card transaction amounts.
-The data is presented on a natural log scale.
-The data contains a notable frequency peak around $\$1$ transactions.
-Fewer large purchases are made, but some large purchases are in excess of five orders of magnitude greater than the smallest purchases.}
+  \caption{Above is a histogram of credit card transaction
+    amounts. The data is presented on a natural log scale. The data
+    contains a notable frequency peak around $\$1$ transactions. Fewer
+    large purchases are made, but some large purchases are in excess
+    of five orders of magnitude greater than the smallest purchases.}
   \label{fig:hist-credit-card}
@@
   \end{tikzpicture}
-  \caption{All models are applied to approximate the expected transaction amount given transformed (and obfuscated) vendor and customer-descriptive features.
-The data is broken up into $10$ partitions and each set of $9$ partitions is used to construct a model that approximates the remaining partition.
-This results in exactly one prediction from each algorithm for each data point.
-These boxes depict the median (middle bar), median $95\%$ confidence interval (cones), quartiles (box edges), fences at $3/2$ interquartile range (whiskers), and outliers (dots) of absolute prediction error for each model.
-The absolute errors in transaction amount predictions are presented on a natural log scale, just as in figure \ref{fig:hist-credit-card}.
-The complementing numerical data to this figure is provided in table \ref{table:error-credit-card} in the Appendix.}
+  \caption{All models are applied to approximate the expected
+    transaction amount given transformed (and obfuscated) vendor and
+    customer-descriptive features. $10$-fold cross validation as
+    described in the beginning of Section \ref{sec:data} is used to
+    evaluated each algorithm. This results in exactly one prediction
+    from each algorithm for each data point. These boxes depict the
+    median (middle bar), median $95\%$ confidence interval (cones),
+    quartiles (box edges), fences at $3/2$ interquartile range
+    (whiskers), and outliers (dots) of absolute prediction error for
+    each model. The absolute errors in transaction amount predictions
+    are presented on a natural log scale, just as in figure
+    \ref{fig:hist-credit-card}. The complementing numerical data to
+    this figure is provided in table \ref{table:error-credit-card} in
+    the Appendix.}
   \label{fig:error-credit-card}
@@
 
-The fourth test data set, and the final with a real-valued range, is a collection of credit card transactions \cite{pozzolo2015calibrating}.
-The provided data carries no direct real-world meaning, it is the output of a principle component analysis on the original hidden source data.
-This obfuscation is done to protect the information of the credit card users.
-This data has the largest dimension of all considered, at 28.
-A model over this data predicts the transaction amount given the vector of principle component information.
-
-As can be seen in figure \ref{fig:error-credit-card}, the MLP outperforms all other algorithms at the first, second, third, and fourth quartiles.
-The MLP produces the lowest absolute error prediction for 80\% of transactions, Delaunay bests the remaining 20\%.
-It is likely that with less data, Delaunay would be the best performer.
+The fourth test data set, and the final with a real-valued range, is a
+collection of credit card transactions
+\cite{pozzolo2015calibrating}. The provided data carries no direct
+real-world meaning, it is the output of a principle component analysis
+on the original hidden source data. This obfuscation is done to
+protect the information of the credit card users. This data has the
+largest dimension of all considered, at 28. A model over this data
+predicts the transaction amount given the vector of principle
+component information.
+
+As can be seen in figure \ref{fig:error-credit-card}, the MLP
+outperforms all other algorithms at the first, second, third, and
+fourth quartiles. The MLP produces the lowest absolute error
+prediction for 80\% of transactions, Delaunay bests the remaining
+20\%. It is likely that with less data, Delaunay would be the best
+performer.
 %% -------------------------------------------------------------------
@@
   \end{tikzpicture}
-  \caption{Histogram of the raw throughput values recorded during all IOzone tests across all system configurations.
-The distribution is skewed right, with few tests having significantly higher throughput than most others.
-The data is presented on a natural log scale, however the resulting left-skew reveals the data is not precisely distributed according to an exponential.}
+  \caption{Histogram of the raw throughput values recorded during all
+    IOzone tests across all system configurations. The distribution is
+    skewed right, with few tests having significantly higher
+    throughput than most others. The data is presented on a natural
+    log scale, however the resulting left-skew reveals the data is not
+    precisely distributed according to an exponential.}
   \label{fig:hist-throughput}
@@
   \end{tikzpicture}
-  \caption{Above, the models directly capable of predicting distributions are applied to predicting the expected CDF of I/O throughput at a previously unseen system configuration.
-The KS statistic between the observed distribution at each system configuration and the predicted distribution is recorded and presented above.
-Note that the above plot is \textit{not} log-scaled like the histogram.
-The complimentary numerical data for this figure is provided in Appendix table \ref{table:error-throughput}.}
+  \caption{Above, the models directly capable of predicting
+    distributions are applied to predicting the expected CDF of I/O
+    throughput at previously unseen system configurations. These
+    boxes depict the median (middle bar), median $95\%$ confidence
+    interval (cones), quartiles (box edges), fences at $3/2$
+    interquartile range (whiskers), and outliers (dots) of KS
+    statistic between the observed distribution at each system
+    configuration and the predicted distribution. The above figure is
+    \textit{not} log-scaled like figure \ref{fig:hist-throughput}. The
+    complimentary numerical data for this figure is provided in
+    Appendix table \ref{table:error-throughput}.}
   \label{fig:error-throughput}
@@
 
-The final of five data sets is four-dimensional and is produced by executing the IOzone benchmark from \cite{iozone} on a homogeneous cluster of computers.
-The system performance data was collected by executing IOzone 150 times for each of a select set of system configurations.
-A single IOzone execution reports the max I/O file-read throughput seen.
-The 150 executions for each system configuration are converted
-to their empirical distribution function point, which are fit with a piecewise cubic Hermite interpolating polynomial \cite{fritsch1980monotone} to approximate the CDF.
-These CDFs are capable of being approximated by all interpolants except LSHEP.
-The four dimensions being modeled to predict throughput mean are file size, record size, thread count, and CPU frequency.
-
-Delaunay achieves the lowest KS statistic for 62\% of approximations, while Voronoi is best for the remaining 38\%.
-Figure \ref{fig:error-throughput} shows that while Delaunay may have more best predictions, the behavior of Voronoi may be preferable.
+The final of five data sets is derived from \cite{cameron2019moana},
+which provides four-dimensional distribution data by executing the
+IOzone benchmark \cite{iozone} on a computer system and varying the
+system's file size, record size, thread count, and CPU frequency. At
+each configuration, IOzone samples the I/O file-read throughput (in
+bytes per second) 150 times. Empirical distribution function points
+are computed from each set of 150 executions, which are interpolated
+with a piecewise cubic Hermite interpolating polynomial
+\cite{fritsch1980monotone} to approximate the CDF. All interpolation
+algorithms with the exception of LSHEP are used to approximate these
+CDFs from system configurations.
+
+Delaunay achieves the lowest KS statistic for 62\% of approximations,
+while Voronoi is best for the remaining 38\%. Figure
+\ref{fig:error-throughput} shows that while Delaunay may have more
+best predictions, the behavior of Voronoi may be preferable.
 %% -------------------------------------------------------------------
@@
   \end{tikzpicture}
-  \caption{Histograms of the prediction error for each interpolant that produces predictions as convex combinations of observed data.
-The data is broken up into $10$ partitions and each set of $9$ partitions is used to construct a model that approximates the remaining partition.
-This results in exactly one prediction from each algorithm for each data point.
-The distributions show the KS statistics for the predicted throughput distribution versus the actual throughput distribution.
-The four vertical red lines represent commonly used $p$-values \{0.05, 0.01, 0.001, 1.0e-6\} respectively.
-All predictions to the right of a red line represent CDF predictions that are significantly different (by respective $p$-value) from the actual distribution according to the KS test.
-The numerical counterpart to this figure is presented in table \ref{table:null-hypothesis-results}.}
+  \caption{Histograms of the prediction error for each interpolant
+    that produces predictions as convex combinations of observed
+    data. $10$-fold cross validation as described in the beginning of
+    Section \ref{sec:data} is used to evaluated each algorithm. This
+    results in exactly one prediction from each algorithm for each
+    data point. The distributions show the KS statistics for the
+    predicted throughput distribution versus the actual throughput
+    distribution. The four vertical red lines represent commonly used
+    $p$-values \{0.05, 0.01, 0.001, 1.0e-6\} respectively. All
+    predictions to the right of a red line represent CDF predictions
+    that are significantly different (by respective $p$-value) from
+    the actual distribution according to the KS test. The numerical
+    counterpart to this figure is presented in table
+    \ref{table:null-hypothesis-results}.}
   \label{fig:throughput-prediction}
@@
   \end{tabular}
-  \caption{This is the numerical counterpart to the histogram data presented in figure \ref{fig:throughput-prediction}.
-The columns display the percent of null hypothesis rejections by the KS-test when provided different selections of $p$-values for each algorithm.
-The algorithm with the lowest rejection rate at each $p$ is boldface, while the second lowest is italicized.}
+  \caption{This is the numerical counterpart to the histogram data
+    presented in figure \ref{fig:throughput-prediction}. The columns
+    display the percent of null hypothesis rejections by the KS-test
+    when provided different selections of $p$-values for each
+    algorithm. The algorithm with the lowest rejection rate at each
+    $p$ is boldface, while the second lowest is italicized.}
   \label{table:null-hypothesis-results}
@@
 
-Figure \ref{fig:throughput-prediction} expands on the KS statistic results presented in figure \ref{fig:error-throughput}.
-Agglomerate errors for each algorithm resemble a Gamma distribution.
-The percentages of significant prediction errors with varying $p$-values are on display in Table \ref{table:null-hypothesis-results}.
-When considering the $p=0.001$ results for each technique, a little over one third of the predicted CDFs are significantly different from the measured (and presumed) correct CDFs.
-However, it should be noted that with 150 samples, the error of an empirical distribution function (EDF) can reasonably be upwards of $.1$, which serves as a rough estimate for the lower limit of achievable error.
-Globally, only a third of Voronoi predictions fail to capture \textit{all} of the characteristics of the CDFs at new system configurations.
+Figure \ref{fig:throughput-prediction} expands on the KS statistic
+results presented in figure \ref{fig:error-throughput}. Agglomerate
+errors for each algorithm resemble a Gamma distribution. The
+percentages of significant prediction errors with varying $p$-values
+are on display in Table \ref{table:null-hypothesis-results}. When
+considering the $p=0.001$ results for each technique, a little over
+one third of the predicted CDFs are significantly different from the
+measured (and presumed) correct CDFs. However, it should be noted that
+with 150 samples, the error of an empirical distribution function
+(EDF) can reasonably be upwards of $.1$, which serves as a rough
+estimate for the lower limit of achievable error. Globally, only a
+third of Voronoi predictions fail to capture \textit{all} of the
+characteristics of the CDFs at new system configurations.
 
@@
   \end{tabular}
-  \caption{This data is the average of Appendix tables \ref{table:error-forest-fire}, \ref{table:error-parkinsons}, \ref{table:error-weather}, \ref{table:error-credit-card}, and provides a gross summary of overall results.
-The columns display (weighted equally by data set, \textit{not} points) the average frequency with which any algorithm provided the lowest absolute error approximation, the average time-to-fit, and the average time required to approximated one point.
-Interpolants provide the lowest error approximation one third of the time across all data, while regressors occupy the other two thirds.
-This result is obtained without any customized tuning or preprocessing to maximize the performance of any given algorithm.
-In practice, tuning and preprocessing may have large effects on approximation performance.}
+  \caption{This data is the average of Appendix tables
+    \ref{table:error-forest-fire}, \ref{table:error-parkinsons},
+    \ref{table:error-weather}, \ref{table:error-credit-card}, and
+    provides a gross summary of overall results. The columns display
+    (weighted equally by data set, \textit{not} points) the average
+    frequency with which any algorithm provided the lowest absolute
+    error approximation, the average time-to-fit, and the average time
+    required to approximated one point. Interpolants provide the
+    lowest error approximation one third of the time across all data,
+    while regressors occupy the other two thirds. This result is
+    obtained without any customized tuning or preprocessing to
+    maximize the performance of any given algorithm. In practice,
+    tuning and preprocessing may have large effects on approximation
+    performance.}
   \label{table:avg-performance}
@@
 
-Table \ref{table:avg-performance} summarizes results across the four test data sets with real-valued ranges.
-The interpolants discussed in this paper produce the \textit{best} approximations roughly one third of the time, and produce competitive approximations for almost all data sets.
-These test problems are almost certainly \textit{stochastic} in nature, but the high dimension yields to greater sparsity and makes interpolation a competitive approximation technique.
-
-The major advantages to interpolation lie in the near absence of \textit{fit} time.
-Delaunay has no preparations required, LSHEP and ShepMod require pairwise distance calculations to determine the radii of influence for points in each model.
-At least hundreds, and sometimes hundreds of thousands of predictions can be made by the interpolants before the most widely used regressor (the MLP) finishes fitting these relatively small data sets.
-However, the computational complexities of all interpolants presented are greater than linear in either dimension or number of points, meaning large enough data (in dimension and number of observations) would cause interpolation to be slower.
-
-The theoretical results presented in Section \ref{sec:theory} directly apply to Delaunay interpolation, however the performance of Delaunay does not appear significantly better than other algorithms on these data sets.
-This observation may be due to the stochastic nature of the data, but it also speaks to the alignment of the geometry of the underlying functions with the approximations generated by the different interpolation methods.
-The strong performance of other interpolants suggests that theoretical results similar to those presented in this work can be achieved for the other interpolants under reasonable assumptions.
-
-Finally, most of the interpolants presented in this work benefit from the ability to model any function over a real vector space with a range that is closed under convex combinations.
-The results of the distribution prediction case study indicate that interpolants can effectively predict CDFs.
-The error analysis for that work relies on the KS statistic, which captures the worst part of any prediction and hence provides a conservatively large estimate of approximation error.
-The average absolute errors in the predicted CDFs are always lower than the KS statistics.
-However, the KS statistic was chosen because of the important statistical theory surrounding it as an error measure.
-Considering this circumstance, a nonnegligible volume of predictions provide impressively low levels of error in that final case study.
+Table \ref{table:avg-performance} summarizes results across the four
+test data sets with real-valued ranges. The interpolants discussed in
+this paper produce the \textit{best} approximations roughly one third
+of the time, and produce competitive approximations for almost all
+data sets. These test problems are almost certainly
+\textit{stochastic} in nature, but the high dimension yields to
+greater sparsity and makes interpolation a competitive approximation
+technique.
+
+The major advantages to interpolation lie in the near absence of
+\textit{fit} time. Delaunay has no preparations required, LSHEP and
+ShepMod require pairwise distance calculations to determine the radii
+of influence for points in each model. At least hundreds, and
+sometimes hundreds of thousands of predictions can be made by the
+interpolants before the most widely used regressor (the MLP) finishes
+fitting these relatively small data sets. However, the computational
+complexities of all interpolants presented are greater than linear in
+either dimension or number of points, meaning large enough data (in
+dimension and number of observations) would cause interpolation to be
+slower.
+
+The theoretical results presented in Section \ref{sec:theory} directly
+apply to Delaunay interpolation, however the performance of Delaunay
+does not appear significantly better than other algorithms on these
+data sets. This observation may be due to the stochastic nature of the
+data, but it also speaks to the alignment of the geometry of the
+underlying functions with the approximations generated by the
+different interpolation methods. The strong performance of other
+interpolants suggests that theoretical results similar to those
+presented in this work can be achieved for the other interpolants
+under reasonable assumptions.
+
+Finally, most of the interpolants presented in this work benefit from
+the ability to model any function over a real vector space with a
+range that is closed under convex combinations. The results of the
+distribution prediction case study indicate that interpolants can
+effectively predict CDFs. The error analysis for that work relies on
+the KS statistic, which captures the worst part of any prediction and
+hence provides a conservatively large estimate of approximation
+error. The average absolute errors in the predicted CDFs are always
+lower than the KS statistics. However, the KS statistic was chosen
+because of the important statistical theory surrounding it as an error
+measure. Considering this circumstance, a nonnegligible volume of
+predictions provide impressively low levels of error in that final
+case study.
 
@@
 
-This work presents new theoretical results for interpolation in arbitrary dimension with simplicial approximations.
-Coinciding with the theoretical results, an empirical evaluation across real-world problems demonstrates that interpolants produce competitively accurate models of multivariate phenomenon when compared with common regressors for sparse, moderately high dimensional problems.
-
-The various interpolants discussed in this paper have been demonstrated to effectively approximate multivariate phenomena up to $30$ dimensions.
-The underlying constructions are theoretically straightforward, yet powerful and flexible.
-Most of the interpolants' computational complexities make them particularly suitable for applications in even higher dimension, where the major benefits are seen when only a small number of approximations are to be made from any given data ($\leq 1000$).
+This work presents new theoretical results for interpolation in
+arbitrary dimension with simplicial approximations. Coinciding with
+the theoretical results, an empirical evaluation across real-world
+problems demonstrates that interpolants produce competitively accurate
+models of multivariate phenomenon when compared with common regressors
+for sparse, moderately high dimensional problems.
+
+The various interpolants discussed in this paper have been
+demonstrated to effectively approximate multivariate phenomena up to
+$30$ dimensions. The underlying constructions are theoretically
+straightforward, yet powerful and flexible. Most of the interpolants'
+computational complexities make them particularly suitable for
+applications in even higher dimension, where the major benefits are
+seen when only a small number of approximations are to be made from
+any given data ($\leq 1000$).
 
@@
   \end{tabular}
-  \caption{This numerical data accompanies the visual provided in figure \ref{fig:error-forest-fire}.
-The columns of absolute error percentiles correspond to the minimum, first quartile, median, second quartile, and maximum absolute errors respectively.
-The minimum of each column is boldface, while the second lowest value is italicized.
-All values are rounded to three significant digits.}
+  \caption{This numerical data accompanies the visual provided in
+    figure \ref{fig:error-forest-fire}. The columns of absolute error
+    percentiles correspond to the minimum, first quartile, median,
+    second quartile, and maximum absolute errors respectively. The
+    minimum of each column is boldface, while the second lowest value
+    is italicized. All values are rounded to three significant
+    digits.}
   \label{table:error-forest-fire}
@@
   \end{tabular}
-  \caption{The left above shows how often each algorithm had the lowest absolute error approximating forest fire data.
-On the right columns are median fit time of 454 points, median time for one approximation, and median time approximating 50 points.}
+  \caption{The left above shows how often each algorithm had the
+    lowest absolute error approximating forest fire data in table
+    \ref{table:error-forest-fire}. On the right columns are median fit
+    time of 454 points, median time for one approximation, and median
+    time approximating 50 points.}
   \label{table:best-forest-fire}
@@
   \end{tabular}
-  \caption{This numerical data accompanies the visual provided in figure \ref{fig:error-parkinsons}.
-The columns of absolute error percentiles correspond to the minimum, first quartile, median, second quartile, and maximum absolute errors respectively.
-The minimum of each column is boldface, while the second lowest value is italicized.
-All values are rounded to three significant digits.}
+  \caption{This numerical data accompanies the visual provided in
+    figure \ref{fig:error-parkinsons}. The columns of absolute error
+    percentiles correspond to the minimum, first quartile, median,
+    second quartile, and maximum absolute errors respectively. The
+    minimum of each column is boldface, while the second lowest value
+    is italicized. All values are rounded to three significant
+    digits.}
   \label{table:error-parkinsons}
@@
   \end{tabular}
-  \caption{The left above shows how often each algorithm had the lowest absolute error approximating parkinsons data.
-On the right columns are median fit time of 5288 points, median time for one approximation, and median time approximating 587 points.}
+  \caption{The left above shows how often each algorithm had the
+    lowest absolute error approximating parkinsons data in table
+    \ref{table:error-parkinsons}. On the right columns are median fit
+    time of 5288 points, median time for one approximation, and median
+    time approximating 587 points.}
   \label{table:best-parkinsons}
@@
   \end{tabular}
-  \caption{This numerical data accompanies the visual provided in figure \ref{fig:error-weather}.
-The columns of absolute error percentiles correspond to the minimum, first quartile, median, second quartile, and maximum absolute errors respectively.
-The minimum value of each column is boldface, while the second lowest is italicized.
-All values are rounded to three significant digits.}
+  \caption{This numerical data accompanies the visual provided in
+    figure \ref{fig:error-weather}. The columns of absolute error
+    percentiles correspond to the minimum, first quartile, median,
+    second quartile, and maximum absolute errors respectively. The
+    minimum value of each column is boldface, while the second lowest
+    is italicized. All values are rounded to three significant
+    digits.}
   \label{table:error-weather}
@@
   \end{tabular}
-  \caption{The left above shows how often each algorithm had the lowest absolute error approximating Sydney rainfall data.
-On the right columns are median fit time of 2349 points, median time for one approximation, and median time approximating 260 points.}
+  \caption{The left above shows how often each algorithm had the
+    lowest absolute error approximating Sydney rainfall data in table
+    \ref{table:error-weather}. On the right columns are median fit
+    time of 2349 points, median time for one approximation, and median
+    time approximating 260 points.}
   \label{table:best-weather}
@@
   \end{tabular}
-  \caption{This numerical data accompanies the visual provided in figure \ref{fig:error-credit-card}.
-The columns of absolute error percentiles correspond to the minimum, first quartile, median, second quartile, and maximum absolute errors respectively.
-The minimum value of each column is boldface, while the second lowest is italicized.
-All values are rounded to three significant digits.}
+  \caption{This numerical data accompanies the visual provided in
+    figure \ref{fig:error-credit-card}. The columns of absolute error
+    percentiles correspond to the minimum, first quartile, median,
+    second quartile, and maximum absolute errors respectively. The
+    minimum value of each column is boldface, while the second lowest
+    is italicized. All values are rounded to three significant
+    digits.}
   \label{table:error-credit-card}
@@
   \end{tabular}
-  \caption{The left above shows how often each algorithm had the lowest absolute error approximating credit card transaction data.
-On the right columns are median fit time of 5006 points, median time for one approximation, and median time approximating 556 points.}
+  \caption{The left above shows how often each algorithm had the
+    lowest absolute error approximating credit card transaction data
+    in table \ref{table:error-credit-card}. On the right columns are
+    median fit time of 5006 points, median time for one approximation,
+    and median time approximating 556 points.}
   \label{table:best-credit-card}
@@
   \end{tabular}
-  \caption{This numerical data accompanies the visual provided in figure \ref{fig:error-throughput}.
-The columns of absolute error percentiles correspond to the minimum, first quartile, median, second quartile, and maximum KS statistics respectively between truth and guess for models predicting the distribution of I/O throughput that will be observed at previously unseen system configurations.
-The minimum value of each column is boldface, while the second lowest is italicized.
-All values are rounded to three significant digits.}
+  \caption{This numerical data accompanies the visual provided in
+    figure \ref{fig:error-throughput}. The columns of absolute error
+    percentiles correspond to the minimum, first quartile, median,
+    second quartile, and maximum KS statistics respectively between
+    truth and guess for models predicting the distribution of I/O
+    throughput that will be observed at previously unseen system
+    configurations. The minimum value of each column is boldface,
+    while the second lowest is italicized. All values are rounded to
+    three significant digits.}
   \label{table:error-throughput}
@@
   \end{tabular}
-  \caption{The left above shows how often each algorithm had the lowest KS statistic on the I/O throughput distribution data.
-On the right columns are median fit time of 2715 points, median time for one approximation, and median time approximating 301 points.}
+  \caption{The left above shows how often each algorithm had the
+    lowest KS statistic on the I/O throughput distribution data in
+    table \ref{table:error-throughput}. On the right columns are
+    median fit time of 2715 points, median time for one approximation,
+    and median time approximating 301 points.}
   \label{table:best-throughput}

### 2019-03-20 Paper_[2019-02]_Numerical_Algorithms/paper.tex
created an affine weights visual. finished NA paper draft. wrote report on Xiaodong clock cycle data.

@@
 \smartqed  % flush right qed marks, e.g. at end of proof
+
+% Insert here the call for the packages your document requires
 \usepackage{graphicx}
-%
-% \usepackage{mathptmx}      % use Times fonts if available on your TeX system
-%
-% insert here the call for the packages your document requires
-%\usepackage{latexsym}
 \usepackage{tikz}     % For adding axes to plots
@@
 \usepackage{booktabs} % For formal tables
+\usepackage{tabularx} % For tables with nice column widths.
 \usepackage[colorlinks,allcolors=blue]{hyperref}
+\usepackage{float}    % For controlling the placement of tables.
 
@@
   \begin{tabular}[#1]{@{}c@{}}#2\end{tabular}}
+
+\newcolumntype{d}[1]{D{.}{.}{#1}}
 
@@
 \spnewtheorem*{onetheorem}{Theorem}{\bf}{}
+%% Create a ``Proof'' environment that starts with 'Proof.'
 \spnewtheorem*{proofdot}{Proof.}{\it}{}
@@
 % \subtitle{Do you have a subtitle?\\ If so, write it here}
-
 %\titlerunning{Short form of title}        % if too long for running head
@@
 \begin{abstract}
-A rapid increase in the quantity of data available is allowing all
-fields of science to generate more accurate models of multivariate
-phenomena. Regression and interpolation become challenging when the
-dimension of data is large, especially while maintaining tractable
-computational complexity. While regression is a popular approach to
-solving approximation problems with high dimension, there are often
-advantages to interpolation. This paper presents error bounds for
-interpolants on moderately high dimensional problems and contrasts
-their performance with that of popular regression
-techniques. Empirical results demonstrate the viability of
-interpolants for high dimensional approximation problems, suggesting
-that these techniques are capable of effectively modeling multivariate
-phenomena while maintaining flexibility in different application
-domains.
+Increases in the quantity of available data have allowed all fields of
+science to generate more accurate models of multivariate phenomena.
+Regression and interpolation become challenging when the dimension of
+data is large, especially while maintaining tractable computational
+complexity. Regression is a popular approach to solving approximation
+problems with high dimension, however there are often advantages to
+interpolation. This paper presents a novel and insightful error bound
+for (piecewise) linear interpolation in arbitrary dimension and
+contrasts the performance of some interpolation techniques with
+popular regression techniques. Empirical results demonstrate the
+viability of interpolation for moderately high dimensional
+approximation problems, and encourage broader application of
+interpolants to multivariate approximation in science.
 
@@
 $s$, some function $f:\mathbb{R}^d \rightarrow Y$ and a set of points
-$X = \bigl\{x^{(1)}$,$\ldots$,$x^{(n)}\bigr\} \subset \mathbb{R}^d,$
+$X = \bigl\{x^{(1)}$, $\ldots$, $x^{(n)}\bigr\} \subset \mathbb{R}^d$,
 along with associated function values $f\bigl(x^{(i)}\bigr)$. The goal
@@
 such that $\hat f\bigl(x^{(i)}\bigr) = f\bigl(x^{(i)}\bigr)$ for all
-$i = 1$,$\ldots$,$n$. It is often the case that the form of the true
+$i = 1$, $\ldots$, $n$. It is often the case that the form of the true
 underlying function $f$ is unknown, however it is still desirable to
@@
 \ref{sec:regression} and \ref{sec:interpolation} present the
-multivariate models. Section \ref{sec:error} presents the error
-measuring methodology that is used for collecting empirical
-results. Section \ref{sec:theory} presents the theoretical error
-bounds for interpolation. Section \ref{sec:data} presents the data
-sets for empirical approximation analysis and presents the
-approximation results for all models on these data sets. Section
-\ref{sec:discussion} analyzes and discusses the results, their
-implications, and their limitations. Finally, Section
-\ref{sec:conclusion} concludes.
-
-% Art is a subjective form of personal expression and computers are
-% just robots that are going to take over the world. \textit{*deep
-% breath*} In order to communicate my frustration with the
-% technological and information revolution, I'm going to write the
-% rest of this paper in poetry.
-
-% Star light, star bright \ldots
+multivariate models. Section \ref{sec:error} gives the error measuring
+methodology that is used for collecting empirical results. Section
+\ref{sec:theory} contains the theoretical error bounds for
+interpolation. Section \ref{sec:data} presents the data sets for
+empirical approximation analysis and the approximation results for all
+models on these data sets. Section \ref{sec:discussion} analyzes and
+discusses the results, their implications, and their
+limitations. Finally, Section \ref{sec:conclusion} concludes.
 
@@
 where $K$ is the selected kernel function, $a_i \in \mathbb{R}^n$, 
-$b \in \mathbb{R}$ are coefficients to be solved for simultaneously. The
-computational complexity of the SVR is $\mathcal{O}(n^2dm)$, with $m$
-being determined by the minimization convergence criterion. This paper
-uses the scikit-learn SVR \cite{scikit-learn} with a polynomial kernel
-function.
+$b \in \mathbb{R}$ are coefficients to be solved for simultaneously.
+The computational complexity of the SVR is $\mathcal{O}(n^2dm)$, with
+$m$ being determined by the minimization convergence criterion. This
+paper uses the scikit-learn SVR \cite{scikit-learn} with a polynomial
+kernel function.
 
@@
  $$ l(u) = \bigl( u^t W_l \bigr)_+ ,$$
-where $W_l$ is the $i$ by $j$ weight matrix for layer $l$. In this
+where $W_l$ is the $i \times j$ weight matrix for layer $l$. In this
 form, the multilayer perceptron (MLP) produces a piecewise linear
@@
 
+\newpage
+
 \subsection{Modified Shepard}
@@
 \cite{shepard1968two,gordon1978shepard}. The interpolant has the form
-$$ p(x) = \frac{\sum\limits_{k=1}^{n}W_k(x)f\bigl(x^{(k)}\bigr)}{\sum\limits_{k=1}^{n}W_k(x)},$$
-where $x \not = x^{(k)}$, $k = 1$, $\ldots$, $n$, for weighting functions
-$$W_k(x) = \left( \frac{\big(r_k - \bigl\|x - x^{(k)}\bigr\|_2\big)_+}{r_k \bigl\|x - x^{(k)}\bigr\|_2} \right)^2,$$ 
-where $r_k \in \mathbb{R}$ are chosen such that at least $d+1$ other
-points $x^{(j)}$, $j \not = k$ are inside the closed Euclidean ball of
-radius $r_k$ about $x^{(k)}$. The computational complexity of ShepMod
-is $\mathcal{O}(n^2d)$. This paper uses a Fortran 95 implementation of
+ $$ p(x) = \frac{\sum\limits_{k=1}^{n}W_k(x)f\bigl(x^{(k)}\bigr)}
+     {\sum\limits_{k=1}^{n}W_k(x)} ,$$
+where $W_k(x) = \bigl(r_k - \bigl\|x - x^{(k)}\bigr\|\bigr)_+^2 \big/
+\bigl\|x - x^{(k)}\bigr\|_2^2$ and $r_k \in \mathbb{R}$ is the
+smallest radius such that at least $d+1$ other points $x^{(j)}$, $j
+\not = k$, are inside the closed Euclidean ball of radius $r_k$ about
+$x^{(k)}$. The computational complexity of ShepMod is
+$\mathcal{O}(n^2d)$. This paper uses a Fortran 95 implementation of
 ShepMod derived from SHEPPACK \cite{thacker2010algorithm}.
@@
 \cite{thacker2010algorithm}. The interpolant has the form
-$$ p(x) = \frac{\sum\limits_{k=1}^{n}W_k(x)P_k(x)}{\sum\limits_{k=1}^{n}W_k(x)} ,$$
+ $$ p(x) = \frac{\sum\limits_{k=1}^{n}W_k(x)P_k(x)}
+ {\sum\limits_{k=1}^{n}W_k(x)} ,$$
 where $W_k(x)$ is a locally supported weighting function and $P_k(x)$
@@
 \rightarrow \mathbb{R}_+$ given by
-$$ v^{x^{(i)}}(y) = \left(1 - \frac{\bigl\|y - x^{(i)}\bigr\|_2}{2 \ h\bigl(y - x^{(i)} \mid x^{(i)}\bigr)} \right)_+, $$
+ $$ v^{x^{(i)}}(y) = \left(1 - \frac{\bigl\|y - x^{(i)}\bigr\|_2}
+ {2 \ h \bigl(y - x^{(i)} \mid x^{(i)}\bigr)} \right)_+, $$
 where $h\bigl(w \mid x^{(i)}\bigr)$ is the distance between $x^{(i)}$
@@
 has local support, giving the interpolated value
-$$ f(y) = \frac{\sum\limits_{i=1}^n v^{x^{(i)}}(y) f\bigl(x^{(i)}\bigr)}{\sum\limits_{i=1}^n v^{x^{(i)}}(y)}, $$
+ $$ f(y) = \frac{\sum\limits_{i=1}^n v^{x^{(i)}}(y)
+  f\bigl(x^{(i)}\bigr)}{\sum\limits_{i=1}^n v^{x^{(i)}}(y)}, $$
 where $0 \leq v^{x^{(i)}}(y) \leq 1$. The computational complexity of
@@
 nondecreasing. This work utilizes the fact that a convex combination
-of CDFs (or PDFs) results in a valid CDF. Given $G(x) = \sum_{i}w_i
-F_i(x)$, $\sum_{i} w_i = 1$, $w_i \geq 0$, and each $F_i$ is a valid
-CDF, $G$ must also be a valid CDF. A demonstration of how this is
-applied can be seen in Figure \ref{fig:prediction-example}.
+of CDFs (or PDFs) results in a valid CDF (or PDF). Given $G(x) =
+\sum_{i}w_i F_i(x)$, $\sum_{i} w_i = 1$, $w_i \geq 0$, and each $F_i$
+is a valid CDF, $G$ must also be a valid CDF. A demonstration of how
+this is applied can be seen in Figure \ref{fig:prediction-example}.
 
@@
 This section presents the theoretical results bounding the error of
-simplicial interpolation. The error analysis relies on simplicial
+(piecewise) linear interpolation. The error analysis relies on linear
 interpolation for three reasons: (1) second order results can be
 obtained utilizing a Lipschitz constant on the gradient of a function,
-rather than standard Lipschitz bounds; (2) the results directly
-apply to Delaunay; and (3) multiple other interpolants in this paper
-compute predictions as convex combinations of observed function
+rather than standard Lipschitz bounds; (2) the results directly apply
+to Delaunay interpolation; and (3) multiple other interpolants in this
+paper compute predictions as convex combinations of observed function
 values, which may allow for straight forward extensions of this error
@@
   \big( \gamma k^2\big)^2 \implies \big \| \nabla f(x_i) - c \big \|_2
-  \leq \sqrt{d} \frac{\gamma k^2}{\sigma_d}.$$ \qed
+  \leq \sqrt{d} \frac{\gamma k^2}{\sigma_d}.$$
+  \qed
 \end{proofdot}
@@
 \begin{onetheorem}
-  Under the assumptions of {\it Lemma \ref{lemma:1}} and {\it Lemma \ref{lemma:3}}, for $z \in S$,
+  Under the assumptions of {\it Lemma \ref{lemma:1}} and {\it Lemma
+    \ref{lemma:3}}, for $z \in S$,
   $$ \big|f(z) - \hat f(z)\big| \leq \frac{\gamma \|z - x_0\|_2^2}{2} + \sqrt{d} \frac{\gamma k^2}{\sigma_d} \|z - x_0\|_2.$$
@@
 
-In summary, the approximation error of a simplicial interpolant tends
-quadratically towards zero when approaching observed data only when
-the diameter of the simplex is also reduced at a proportional rate.
-Only linear convergence to the true function can be achieved in
-practice, without the incorporation of additional observations. Notice
-that the approximation error is largely determined by the spacing of
-observed data. Predictions made by simplices whose vertices are not
-well-spaced (i.e. have large diameter, or are nearly contained in a
-hyperplane) have higher error. In light of this error bound, an
-empirical evaluation of presented algorithms follows.
+In summary, the approximation error of a linear (simplicial)
+interpolant tends quadratically towards zero when approaching observed
+data only when the diameter of the simplex is also reduced at a
+proportional rate.  Only linear convergence to the true function can
+be achieved in practice, without the incorporation of additional
+observations. Notice that the approximation error is largely
+determined by the spacing of observed data. Predictions made by
+simplices whose vertices are not well-spaced (i.e. have large
+diameter, or are nearly contained in a hyperplane) have higher
+error. In light of this error bound, an empirical evaluation of
+presented algorithms follows.
 
 %% ===================================================================
+\vspace{-2mm}
 \section{Data and Empirical Analysis}
@@
 The theoretical results constructed in Section \ref{sec:theory} for
-simplicial interpolation are promising, however they are difficult to
-interpret in context with approximation algorithms that do not have
-similar known uniform error bounds. For that reason, this paper
-utilizes five different data sets of varying dimension and application
-to construct approximations and compare the accuracy of different
-techniques.
+(piecewise) linear interpolation are promising and apply directly to
+Delaunay interpolation, however they are difficult to interpret in
+context with approximation algorithms that do not have similar known
+uniform error bounds. For that reason, this paper utilizes five
+different data sets of varying dimension and application to construct
+approximations and compare the accuracy of different techniques.
 
@@
 modeled is presented followed by the distribution of approximation
-errors for each algorithm.
+errors for each algorithm. The plots for all five data sets have the
+same statistical format.
 
@@
 described in \cite{kohavi1995study} with $k=10$. This approach
-randomly partitions data into $k$ nearly equal sized sets. Each
+randomly partitions data into $k$ equal sized (nearly) sets. Each
 algorithm is then evaluated by constructing an approximation over each
@@
 \ref{sec:introduction}, regression techniques appear most appropriate
-for these problems. However, in high dimension data grows
-exponentially more sparse. Given sparse data, regressors tend towards
-interpolation.
+for these problems. However, typically data grows exponentially more
+sparse with increasing dimension. Given sparse data, regressors tend
+towards interpolation.
 
 %% -------------------------------------------------------------------
-\subsection{Forest Fire ($n = 504, d = 12$)}
+\subsection{Forest Fire ($n = 504$, $d = 12$)}
 
@@
     \node (img)  {\includegraphics[width=0.75\textwidth,]{Figures/raw-histogram-forest-fire.pdf}};
-    \node[below=of img, node distance=1cm, yshift=1cm] {Forest First Area Burned};
+    \node[below=of img, node distance=1cm, yshift=1cm] {Forest Fire Area Burned};
     \node[left=of img, node distance=0cm, rotate=90, anchor=center,yshift=-0.7cm] {Count};
   \end{tikzpicture}
-  \caption{Above is a histogram of forest fire area burned under
-    recorded weather conditions. The data is presented on a natural
-    log scale because most values are small with exponentially fewer
-    fires on record that burn large areas.}
+  \caption{Histogram of forest fire area burned under recorded weather
+    conditions. The data is presented on a $\ln$ scale because
+    most values are small with exponentially fewer fires on record
+    that burn large areas.}
   \label{fig:hist-forest-fire}
@@
   \caption{All models are applied to approximate the amount of area
-    that would be burned given environment conditions. $10$-fold
-    cross validation as described in the beginning of Section
-    \ref{sec:data} is used to evaluated each algorithm. This results
-    in exactly one prediction from each algorithm for each data
-    point. These boxes depict the median (middle bar), median $95\%$
-    confidence interval (cones), quartiles (box edges), fences at
-    $3/2$ interquartile range (whiskers), and outliers (dots) of
-    absolute prediction error for each model. Similar to figure
-    \ref{fig:hist-forest-fire}, the errors are presented on a natural
-    log scale. The complementing numerical data to this figure is
-    provided in Table \ref{table:error-forest-fire} in the Appendix.}
+    that would be burned given environment conditions. $10$-fold cross
+    validation as described in the beginning of Section \ref{sec:data}
+    is used to evaluated each algorithm. This results in exactly one
+    prediction from each algorithm for each data point. These boxes
+    depict the median (middle bar), median $95\%$ confidence interval
+    (cones), quartiles (box edges), fences at $3/2$ interquartile
+    range (whiskers), and outliers (dots) of absolute prediction error
+    for each model. Similar to Figure \ref{fig:hist-forest-fire}, the
+    errors are presented on a $\ln$ scale. The numerical data
+    corresponding to this figure is provided in Table
+    \ref{table:error-forest-fire} in the Appendix.}
   \label{fig:error-forest-fire}
@@
 As suggested by Figure \ref{fig:error-forest-fire}, the SVR has the
-lowest absolute prediction errors for 80\% of the data, with ShepMod
-and BoxSpline being the nearest overall competitors. The effectiveness
-of SVR on this data suggests the underlying function can be defined by
+lowest absolute prediction errors for $80\%$ of the data, with MLP and
+Delaunay being the nearest overall competitors. The effectiveness of
+SVR on this data suggests the underlying function can be defined by
 relatively few parameters, as well as the importance of capturing the
@@
   \end{tikzpicture}
-  \caption{Above is a histogram of the Parkinson's patient total UPDRS
-    clinical scores that will be approximated by each algorithm.}
+  \caption{Histogram of the Parkinson's patient total UPDRS clinical
+    scores that will be approximated by each algorithm.}
   \label{fig:hist-parkinsons}
@@
   \caption{All models are applied to approximate the total UPDRS score
-    given audio features from patients' home life. $10$-fold
-    cross validation as described in the beginning of Section
-    \ref{sec:data} is used to evaluated each algorithm. This results
-    in exactly one prediction from each algorithm for each data
-    point. These boxes depict the median (middle bar), median $95\%$
-    confidence interval (cones), quartiles (box edges), fences at
-    $3/2$ interquartile range (whiskers), and outliers (dots) of
-    absolute prediction error for each model. The complementing
-    numerical data to this is provided in table
+    given audio features from patients' home life, using $10$-fold
+    cross validation. These boxes depict the median (middle bar),
+    median $95\%$ confidence interval (cones), quartiles (box edges),
+    fences at $3/2$ interquartile range (whiskers), and outliers
+    (dots) of absolute prediction error for each model. The numerical
+    data corresponding to this figure is provided in Table
     \ref{table:error-parkinsons} in the Appendix.}
@@
 score. The total UPDRS score given by a clinical evaluation is
-estimated through 16 real numbers generated from biomedical voice
+estimated through 19 real numbers generated from biomedical voice
 measures of in-home sound recordings.
@@
 lowest minimum, first quartile, and median of absolute errors for this
-problem, while providing the best prediction 66\% of the time. The MLP
-has the lowest third quartile and provides the best prediction for
-32\% of approximations. The dominance of ShepMod may be in part to
-regular-interval total UPDRS scores provided by clinicians, favoring a
-nearest-neighbor strategy of prediction.
+problem, while providing the best prediction $66\%$ of the time. The
+MLP has the lowest third quartile and provides the best prediction for
+$32\%$ of approximations. The dominance of ShepMod may be due in part
+to regular-interval total UPDRS scores provided by clinicians,
+favoring a nearest-neighbor strategy of prediction.
 %% -------------------------------------------------------------------
@@
   \end{tikzpicture}
-  \caption{Above is a histogram of daily rainfall in Sydney,
-    Australia. The data is presented on a natural log scale because
-    the frequency of larger amounts of rainfall is significantly
-    less. There is a peak in occurrence of the value $0$, which has a
-    notable effect on the resulting model performance.}
+  \caption{Histogram of daily rainfall in Sydney, Australia, presented
+    on a $\ln$ scale because the frequency of larger amounts of
+    rainfall is significantly less. There is a peak in occurrence of
+    the value $0$, which has a notable effect on the resulting model
+    performance.}
   \label{fig:hist-weather}
@@
     rainfall expected on the next calendar day given various sources
-    of local meteorological data. $10$-fold cross validation as
-    described in the beginning of Section \ref{sec:data} is used to
-    evaluated each algorithm. This results in exactly one prediction
-    from each algorithm for each data point. These boxes depict the
-    median (middle bar), median $95\%$ confidence interval (cones),
-    quartiles (box edges), fences at $3/2$ interquartile range
-    (whiskers), and outliers (dots) of absolute prediction error for
-    each model. The errors are presented on a natural log scale,
-    mimicking the presentation in Figure \ref{fig:hist-weather}. The
-    complementing numerical data to this figure is provided in table
-    \ref{table:error-weather} in the Appendix.}
+    of local meteorological data, using $10$-fold cross validation.
+    These boxes depict the median (middle bar), median $95\%$
+    confidence interval (cones), quartiles (box edges), fences at
+    $3/2$ interquartile range (whiskers), and outliers (dots) of
+    absolute prediction error for each model. The errors are presented
+    on a $\ln$ scale, mimicking the presentation in Figure
+    \ref{fig:hist-weather}. The numerical data corresponding to this
+    figure is provided in Table \ref{table:error-weather} in the
+    Appendix.}
   \label{fig:error-weather}
@@
 
-The third data regarding the total daily rainfall in Sydney, Australia
-\cite{williams2009rattle} provides a slightly higher dimension
+The third data set for the total daily rainfall in Sydney, Australia
+\cite{williams2009rattle} provides a slightly higher dimensional
 challenge for the interpolants and regressors. This data is composed
-of many meteorological readings from the region on a day including:
+of many meteorological readings from the region in a day including:
 min and max temperatures, sunshine, wind speed directions (converted
-to coordinates on circle), wind speeds, and humidities throughout the
-day, day of the year (converted to coordinates on circle), and the
-model must predict the amount of rainfall tomorrow.
+to coordinates on a circle), wind speeds, and humidities throughout
+the day, day of the year (converted to coordinates on a circle), and
+the model must predict the amount of rainfall tomorrow.
 
 While Figure \ref{fig:hist-weather} makes MARS look far better than
-other techniques, it only provides the best prediction for 11\% of
-points. The MLP has the lowest absolute error for 56\% of points and
-LSHEP is best for 28\%. MARS likely achieves such a low first quartile
-due to the high occurrence of the value 0 in the data.
+other techniques, it only provides the best prediction for $11\%$ of
+points. The MLP has the lowest absolute error for $56\%$ of points and
+LSHEP is best for $28\%$. MARS likely achieves such a low first
+quartile due to the high occurrence of the value zero in the data.
 %% -------------------------------------------------------------------
@@
   \end{tikzpicture}
-  \caption{Above is a histogram of credit card transaction
-    amounts. The data is presented on a natural log scale. The data
-    contains a notable frequency peak around $\$1$ transactions. Fewer
-    large purchases are made, but some large purchases are in excess
-    of five orders of magnitude greater than the smallest purchases.}
+  \caption{Histogram of credit card transaction amounts, presented on
+    a $\ln$ scale. The data contains a notable frequency peak around
+    $\$1$ transactions. Fewer large purchases are made, but some large
+    purchases are in excess of five orders of magnitude greater than
+    the smallest purchases.}
   \label{fig:hist-credit-card}
@@
     transaction amount given transformed (and obfuscated) vendor and
-    customer-descriptive features. $10$-fold cross validation as
-    described in the beginning of Section \ref{sec:data} is used to
-    evaluated each algorithm. This results in exactly one prediction
-    from each algorithm for each data point. These boxes depict the
-    median (middle bar), median $95\%$ confidence interval (cones),
-    quartiles (box edges), fences at $3/2$ interquartile range
-    (whiskers), and outliers (dots) of absolute prediction error for
-    each model. The absolute errors in transaction amount predictions
-    are presented on a natural log scale, just as in figure
-    \ref{fig:hist-credit-card}. The complementing numerical data to
-    this figure is provided in Table \ref{table:error-credit-card} in
-    the Appendix.}
+    customer-descriptive features, using $10$-fold cross validation.
+    These boxes depict the median (middle bar), median $95\%$
+    confidence interval (cones), quartiles (box edges), fences at
+    $3/2$ interquartile range (whiskers), and outliers (dots) of
+    absolute prediction error for each model. The absolute errors in
+    transaction amount predictions are presented on a $\ln$ scale,
+    just as in Figure \ref{fig:hist-credit-card}. The numerical data
+    corresponding to this figure is provided in Table
+    \ref{table:error-credit-card} in the Appendix.}
   \label{fig:error-credit-card}
@@
 \cite{pozzolo2015calibrating}. The provided data carries no direct
-real-world meaning, it is the output of a principle component analysis
+real-world meaning, being the output of a principle component analysis
 on the original hidden source data. This obfuscation is done to
 protect the information of the credit card users. This data has the
-largest dimension of all considered, at 28. A model over this data
+largest dimension of all considered, at 28. A model for this data
 predicts the transaction amount given the vector of principle
@@
 fourth quartiles. The MLP produces the lowest absolute error
-prediction for 80\% of transactions, Delaunay bests the remaining
-20\%. It is likely that with less data, Delaunay would be the best
+prediction for $80\%$ of transactions, Delaunay bests the remaining
+$20\%$. It is likely that with less data, Delaunay would be the best
 performer.
@@
     skewed right, with few tests having significantly higher
-    throughput than most others. The data is presented on a natural
-    log scale, however the resulting left-skew reveals the data is not
+    throughput than most others. The data is presented on a $\ln$
+    scale, however the resulting left-skew reveals the data is not
     precisely distributed according to an exponential.}
@@
   \end{tikzpicture}
-  \caption{Above, the models directly capable of predicting
-    distributions are applied to predicting the expected CDF of I/O
-    throughput at previously unseen system configurations. These
-    boxes depict the median (middle bar), median $95\%$ confidence
-    interval (cones), quartiles (box edges), fences at $3/2$
-    interquartile range (whiskers), and outliers (dots) of KS
-    statistic between the observed distribution at each system
-    configuration and the predicted distribution. The above figure is
+  \caption{The models directly capable of predicting distributions are
+    applied to predicting the expected CDF of I/O throughput at a
+    previously unseen system configuration, using $10$-fold cross
+    validation. The KS statistic between the observed distribution at
+    each system configuration and the predicted distribution is
+    recorded and presented above. Note that the above figure is
     \textit{not} log-scaled like Figure \ref{fig:hist-throughput}. The
-    complimentary numerical data for this figure is provided in
-    Appendix Table \ref{table:error-throughput}.}
+    numerical data corresponding to this figure is provided in Table
+    \ref{table:error-throughput} in the Appendix.}
   \label{fig:error-throughput}
@@
 
-Delaunay achieves the lowest KS statistic for 62\% of approximations,
-while Voronoi is best for the remaining 38\%. Figure
+Delaunay achieves the lowest KS statistic for $62\%$ of
+approximations, while Voronoi is best for the remaining $38\%$. Figure
 \ref{fig:error-throughput} shows that while Delaunay may have more
@@
   \caption{Histograms of the prediction error for each interpolant
-    that produces predictions as convex combinations of observed
-    data. $10$-fold cross validation as described in the beginning of
-    Section \ref{sec:data} is used to evaluated each algorithm. This
-    results in exactly one prediction from each algorithm for each
-    data point. The distributions show the KS statistics for the
-    predicted throughput distribution versus the actual throughput
-    distribution. The four vertical red lines represent commonly used
-    $p$-values \{0.05, 0.01, 0.001, 1.0e-6\} respectively. All
-    predictions to the right of a red line represent CDF predictions
-    that are significantly different (by respective $p$-value) from
-    the actual distribution according to the KS test. The numerical
-    counterpart to this figure is presented in table
-    \ref{table:null-hypothesis-results}.}
+    that produces predictions as convex combinations of observed data,
+    using $10$-fold cross validation. The histograms show the KS
+    statistics for the predicted throughput distribution versus the
+    actual throughput distribution. The four vertical lines represent
+    commonly used $p$-values 0.05, 0.01, 0.001, $10^{-6}$,
+    respectively. All predictions to the right of a vertical line
+    represent CDF predictions that are significantly different (by
+    respective $p$-value) from the actual distribution according to
+    the KS test. The numerical counterpart to this figure is presented
+    in Table \ref{table:null-hypothesis-results}.}
   \label{fig:throughput-prediction}
@@
   \end{tabular}
-  \caption{This is the numerical counterpart to the histogram data
-    presented in Figure \ref{fig:throughput-prediction}. The columns
-    display the percent of null hypothesis rejections by the KS-test
-    when provided different selections of $p$-values for each
-    algorithm. The algorithm with the lowest rejection rate at each
-    $p$ is boldface, while the second lowest is italicized.}
+  \caption{Numerical counterpart of the histogram data presented in
+    Figure \ref{fig:throughput-prediction}. The columns display the
+    percent of null hypothesis rejections by the KS-test when provided
+    different selections of $p$-values for each algorithm. The
+    algorithm with the lowest rejection rate at each $p$ is boldface,
+    while the second lowest is italicized.}
   \label{table:null-hypothesis-results}
@@
   \centering
-  \begin{tabular}{c|c|c|c}
-    \hline
-    Algorithm & Avg. \% Best & Avg. Fit Time & Avg. App. Time (1)\\
-    \hline
-    MARS & $4.6\%$ & $22.3$s & $1.37 \times 10^{-3}$s\\
-    SVR & $\mathit{21.1}\%$ & $0.503$s & $\mathit{1.37 \times 10^{-4}}$s\\
-    MLP & $\mathbf{42.1}\%$ & $210.0$s & $1.02 \times 10^{-3}$s\\
-    Delaunay & $6.0\%$ & $\mathbf{1.43 \times 10^{-4}}$s & $3.03$s\\
-    ShepMod & $18.4\%$ & $0.711$s & $1.49 \times 10^{-4}$s\\
-    LSHEP & $8.4\%$ & $1.69$s & $\mathbf{1.16 \times 10^{-4}}$s\\
-    Voronoi & $0.5\%$ & $\mathit{0.127}$s & $0.0375$s\\
-    BoxSplines & $0.2\%$ & $0.816$s & $5.26 \times 10^{-4}$s\\
-    \hline
+  \begin{tabular}{c|r@{.}l|r@{.}l|r@{.}l}
+    Algorithm & \multicolumn{2}{c|}{Avg. \% Best} & \multicolumn{2}{c|}{\multilinecell{Avg. Fit or \\ Prep. Time (s)}} & \multicolumn{2}{c}{Avg. App. Time (s)}\\
+    \hline
+    MARS & \quad\quad$4$&$6$ & \,\quad$22$&$3$ & \quad\quad$0$&$00137$\\
+    SVR & $\mathit{21}$&$\mathit{1}$ & $0$&$503$ & $\mathit{0}$&$\mathit{000137}$\\
+    MLP & $\mathbf{42}$&$\mathbf{1}$ & $210$&$0$ & $0$&$00102$\\
+    Delaunay & $6$&$0$ & $\mathbf{0}$&$\mathbf{000143}$ & $3$&$03$\\
+    ShepMod & $18$&$4$ & $0$&$711$ & $0$&$000149$\\
+    LSHEP & $8$&$4$ & $1$&$69$ & $\mathbf{0}$&$\mathbf{000116}$\\
+    Voronoi & $0$&$5$ & $\mathit{0}$&$\mathit{127}$ & $0$&$0375$\\
+    BoxSplines & $0$&$2$ & $0$&$816$ & $0$&$000526$\\
   \end{tabular}
-  \caption{This data is the average of Appendix tables
-    \ref{table:error-forest-fire}, \ref{table:error-parkinsons},
-    \ref{table:error-weather}, \ref{table:error-credit-card}, and
+  \caption{This average of Appendix Tables
+    \ref{table:best-forest-fire}, \ref{table:best-parkinsons},
+    \ref{table:best-weather}, and \ref{table:best-credit-card}
     provides a gross summary of overall results. The columns display
@@
     frequency with which any algorithm provided the lowest absolute
-    error approximation, the average time-to-fit, and the average time
-    required to approximate one point. Interpolants provide the
-    lowest error approximation one third of the time across all data,
-    while regressors occupy the other two thirds. This result is
-    obtained without any customized tuning or preprocessing to
-    maximize the performance of any given algorithm. In practice,
-    tuning and preprocessing may have large effects on approximation
-    performance.}
+    error approximation, the average time to fit / prepare, and the
+    average time required to approximate one point. Interpolants
+    provide the lowest error approximation one third of the time
+    across all data, while regressors occupy the other two
+    thirds. This result is obtained without any customized tuning or
+    preprocessing to maximize the performance of any given
+    algorithm. In practice, tuning and preprocessing may have large
+    effects on approximation performance.}
   \label{table:avg-performance}
@@
 data sets. These test problems are almost certainly
-\textit{stochastic} in nature, but the high dimension yields to
-greater sparsity and makes interpolation a competitive approximation
-technique.
+\textit{stochastic} in nature, but the high dimension leads to greater
+data sparsity and model construction cost, making interpolation more
+competitive.
 
 The major advantages to interpolation lie in the near absence of
-\textit{fit} time. Delaunay has no preparations required, LSHEP and
-ShepMod require pairwise distance calculations to determine the radii
-of influence for points in each model. At least hundreds, and
-sometimes hundreds of thousands of predictions can be made by the
-interpolants before the most widely used regressor (the MLP) finishes
-fitting these relatively small data sets. However, the computational
-complexities of all interpolants presented are greater than linear in
-either dimension or number of points, meaning large enough data (in
-dimension and number of observations) would cause interpolation to be
-slower.
-
-The theoretical results presented in Section \ref{sec:theory} directly
-apply to Delaunay interpolation, however the performance of Delaunay
-does not appear significantly better than other algorithms on these
-data sets. This observation may be due to the stochastic nature of the
-data, but it also speaks to the alignment of the geometry of the
-underlying functions with the approximations generated by the
-different interpolation methods. The strong performance of other
-interpolants suggests that theoretical results similar to those
-presented in this work can be achieved for the other interpolants
-under reasonable assumptions.
+\textit{fit} time. Delaunay, LSHEP, and ShepMod all require pairwise
+distance calculations, for numerical robustness (Delaunay) and to
+determine the radii of influence for data points (LSHEP and
+ShepMod). At least hundreds, and sometimes hundreds of thousands of
+predictions can be made by the interpolants before the most widely
+used regressor (MLP) finishes fitting these relatively small data
+sets.  However, the computational complexities of all interpolants
+presented are greater than linear in either dimension or number of
+points, whereas the regressors' nonlinear complexity in dimension
+generally comes from the model fitting optimization.
+
+The new theoretical results presented in Section \ref{sec:theory}
+directly apply to Delaunay interpolation, however the performance of
+Delaunay does not appear significantly better than other algorithms on
+these data sets.  This observation may be due to the stochastic nature
+of the data, but it also speaks to the power of the approximations
+generated by the different interpolation methods.  The strong
+performance of other interpolants suggests that theoretical results
+similar to those presented in this work can be achieved for the other
+interpolants under reasonable assumptions.
 
 Finally, most of the interpolants presented in this work benefit from
-the ability to model any function over a real vector space with a
-range that is closed under convex combinations. The results of the
+the ability to model any function over real $d$-tuples with a range
+that is closed under convex combinations. The results of the
 distribution prediction case study indicate that interpolants can
@@
 error. The average absolute errors in the predicted CDFs are always
-lower than the KS statistics. However, the KS statistic was chosen
-because of the important statistical theory surrounding it as an error
-measure. Considering this circumstance, a nonnegligible volume of
-predictions provide impressively low levels of error in that final
-case study.
+lower than the KS statistics. However, the KS statistic was chosen as
+a metric because of the important surrounding statistical theory. A
+nonnegligible volume of predictions provide impressively low levels of
+average absolute error in that final case study.
 
@@
 
-This work presents new theoretical results for interpolation in
-arbitrary dimension with simplicial approximations. Coinciding with
-the theoretical results, an empirical evaluation across real-world
-problems demonstrates that interpolants produce competitively accurate
-models of multivariate phenomenon when compared with common regressors
-for sparse, moderately high dimensional problems.
+The major contributions of this work are: 1) new uniform theoretical
+error bounds for piecewise linear interpolation in arbitrary dimension
+(Section \ref{sec:theory}); and 2) an empirical evaluation across
+real-world problems (Section \ref{sec:data}) that demonstrates
+interpolants produce competitively accurate models of multivariate
+phenomenon when compared with common regressors for sparse, moderately
+high dimensional problems.
 
 The various interpolants discussed in this paper have been
-demonstrated to effectively approximate multivariate phenomena in up
-to $30$ dimensions. The underlying constructions are theoretically
-straightforward, yet powerful and flexible. Most of the interpolants'
-computational complexities make them particularly suitable for
-applications in even higher dimension. The major benefits of
-interpolation are seen when only a small number of approximations
-($\leq 1000$) are made from data and when there are relatively few
-data points for the dimension (for empirical results presented, $\log_d n \leq 5$).
+demonstrated to effectively approximate multivariate phenomena up to
+$30$ dimensions. The underlying constructions are theoretically
+straightforward, interpretable, and yield reasonably accurate
+predictions. Most of the interpolants' computational complexities make
+them particularly suitable for applications in even higher
+dimension. The major benefits of interpolation are seen when only a
+small number of approximations ($\leq 1000$) are made from data and
+when there are relatively few data points for the dimension (for
+empirical results presented, $\log_d n \leq 5$).
 
@@
 \begin{appendix}
-
 \section*{Appendix}
-
-\begin{table}
-  \centering
-  \begin{tabular}{c|c|c|c|c|c}
-    \hline
-    Algorithm & Min & $25^{th}$ & $50^{th}$ & $75^{th}$ & Max\\
-    \hline
-    MARS & $0.00984$ & $3.11$ & $7.01$ & $\mathit{11.7}$ & $1090.0$\\
-    SVR & $\mathit{0.00402}$ & $\mathbf{0.243}$ & $\mathbf{0.416}$ & $\mathbf{6.19}$ & $1090.0$\\
-    MLP & $0.0426$ & $2.63$ & $\mathit{5.27}$ & $14.0$ & $1090.0$\\
-    Delaunay & $\mathbf{0.00}$ & $1.98$ & $5.37$ & $13.1$ & $\mathit{1080.0}$\\
-    ShepMod & $\mathbf{0.00}$ & $\mathit{1.93}$ & $6.27$ & $16.0$ & $1090.0$\\
-    LSHEP & $0.0400$ & $2.17$ & $8.87$ & $19.1$ & $\mathbf{1070.0}$\\
-    Voronoi & $0.00982$ & $3.65$ & $7.56$ & $15.6$ & $1090.0$\\
-    BoxSpline & $\mathbf{0.00}$ & $2.27$ & $5.91$ & $16.4$ & $1090.0$\\
+\label{sec:appendix}
+
+\vspace*{\fill}
+
+\begin{table}[H]
+  \centering
+  \begin{tabular}{c|r@{.}l|r@{.}l|r@{.}l|r@{.}l|r@{.}l}
+    \hline
+    Algorithm & \multicolumn{2}{c|}{Min} & \multicolumn{2}{c|}{$25^{th}$} & \multicolumn{2}{c|}{$50^{th}$} & \multicolumn{2}{c|}{$75^{th}$} & \multicolumn{2}{c}{Max}\\
+    \hline
+    MARS & $0$&$00984$ & $3$&$11$ & $7$&$01$ & $\mathit{11}$&$\mathit{7}$ & $1090$&$0$\\
+    SVR & $\mathit{0}$&$\mathit{00402}$ & $\mathbf{0}$&$\mathbf{243}$ & $\mathbf{0}$&$\mathbf{416}$ & $\mathbf{6}$&$\mathbf{19}$ & $1090$&$0$\\
+    MLP & $0$&$0426$ & $2$&$63$ & $\mathit{5}$&$\mathit{27}$ & $14$&$0$ & $1090$&$0$\\
+    Delaunay & $\mathbf{0}$&$\mathbf{00}$ & $1$&$98$ & $5$&$37$ & $13$&$1$ & $\mathit{1080}$&$\mathit{0}$\\
+    ShepMod & $\mathbf{0}$&$\mathbf{00}$ & $\mathit{1}$&$\mathit{93}$ & $6$&$27$ & $16$&$0$ & $1090$&$0$\\
+    LSHEP & $0$&$0400$ & $2$&$17$ & $8$&$87$ & $19$&$1$ & $\mathbf{1070}$&$\mathbf{0}$\\
+    Voronoi & $0$&$00982$ & $3$&$65$ & $7$&$56$ & $15$&$6$ & $1090$&$0$\\
+    BoxSpline & $\mathbf{0}$&$\mathbf{00}$ & $2$&$27$ & $5$&$91$ & $16$&$4$ & $1090$&$0$\\
     \hline
@@
     percentiles correspond to the minimum, first quartile, median,
-    second quartile, and maximum absolute errors respectively. The
+    third quartile, and maximum absolute errors respectively. The
     minimum of each column is boldface, while the second lowest value
@@
 
-\begin{table}
-  \centering
-  \begin{tabular}{|c|c| c |c|c|c|}
-    \cline{1-2}\cline{4-6}
-    Algorithm & \% Best &  & Fit Time & App. Time (1) & Total App. Time\\
-    \cline{1-2}\cline{4-6}
-    MARS & $\mathit{7.7}\%$ &  & $29.1$s & $1.37 \times 10^{-3}$s & $0.0686$s\\
-    SVR & $\mathbf{80.2}\%$ &  & $5.84 \times 10^{-3}$s & $\mathit{6.20 \times 10^{-5}}$s & $\mathit{3.10 \times 10^{-3}}$s\\
-    MLP & $0.0\%$ &  & $32.8$s & $8.71 \times 10^{-4}$s & $0.0436$s\\
-    Delaunay & $3.5\%$ &  & $\mathbf{2.49 \times 10^{-5}}$s & $0.0385$s & $1.93$s\\
-    ShepMod & $3.7\%$ &  & $6.34 \times 10^{-3}$s & $6.44 \times 10^{-5}$s & $3.22 \times 10^{-3}$s\\
-    LSHEP & $5.1\%$ &  & $0.0275$s & $\mathbf{4.63 \times 10^{-5}}$s & $\mathbf{2.31 \times 10^{-3}}$s\\
-    Voronoi & $0.0\%$ &  & $\mathit{1.82 \times 10^{-4}}$s & $3.96 \times 10^{-4}$s & $0.0198$s\\
-    BoxSpline & $0.4\%$ &  & $7.24 \times 10^{-3}$s & $9.78 \times 10^{-5}$s & $4.89 \times 10^{-3}$s\\
-    \cline{1-2}\cline{4-6}
+\begin{table}[H]
+  \centering
+  \begin{tabular}{|c|r@{.}l| c |r@{.}l|r@{.}l|r@{.}l|}
+    \cline{1-3}\cline{5-10}
+    Algorithm & \multicolumn{2}{c|}{\% Best} &  & \multicolumn{2}{c|}{Fit/Prep. Time (s)} & \multicolumn{2}{c|}{App. Time (s)} & \multicolumn{2}{c|}{Total App. Time (s)}\\
+    \cline{1-3}\cline{5-10}
+    MARS & \quad$\mathit{7}$&$\mathit{7}$ &  & \quad\,\,\,$29$&$1$ & \quad$0$&$00137$ & \quad\quad\quad$0$&$0686$\\
+    SVR & $\mathbf{80}$&$\mathbf{2}$ &  & $0$&$00584$ & $\mathit{0}$&$\mathit{0000620}$ & $\mathit{0}$&$\mathit{00310}$\\
+    MLP & $0$&$0$ &  & $32$&$8$ & $0$&$000871$ & $0$&$0436$\\
+    Delaunay & $3$&$5$ &  & $\mathbf{0}$&$\mathbf{0000249}$ & $0$&$0385$ & $1$&$93$\\
+    ShepMod & $3$&$7$ &  & $0$&$00634$ & $0$&$0000644$ & $0$&$00322$\\
+    LSHEP & $5$&$1$ &  & $0$&$0275$ & $\mathbf{0}$&$\mathbf{0000463}$ & $\mathbf{0}$&$\mathbf{00231}$\\
+    Voronoi & $0$&$0$ &  & $\mathit{0}$&$\mathit{000182}$ & $0$&$000396$ & $0$&$0198$\\
+    BoxSpline & $0$&$4$ &  & $0$&$00724$ & $0$&$0000978$ & $0$&$00489$\\
+    \cline{1-3}\cline{5-10}
   \end{tabular}
   \caption{The left above shows how often each algorithm had the
-    lowest absolute error approximating forest fire data in table
+    lowest absolute error approximating forest fire data in Table
     \ref{table:error-forest-fire}. On the right columns are median fit
@@
 
+\vspace*{\fill}
+\newpage
+
 \begin{table}
   \centering
-  \begin{tabular}{c|c|c|c|c|c}
-    \hline
-    Algorithm & Min & $25^{th}$ & $50^{th}$ & $75^{th}$ & Max\\
-    \hline
-    MARS & $0.00948$ & $9.98$ & $20.4$ & $32.9$ & $863.0$\\
-    SVR & $0.00138$ & $3.17$ & $7.31$ & $12.6$ & $\mathbf{27.5}$\\
-    MLP & $2.39 \times 10^{-5}$ & $\mathit{0.533}$ & $\mathit{1.25}$ & $\mathbf{2.84}$ & $39.3$\\
-    Delaunay & $\mathit{3.72 \times 10^{-12}}$ & $1.20$ & $3.50$ & $7.67$ & $30.7$\\
-    ShepMod & $\mathbf{0.00}$ & $\mathbf{0.255}$ & $\mathbf{0.908}$ & $\mathit{3.43}$ & $34.5$\\
-    LSHEP & $0.00254$ & $2.93$ & $7.16$ & $13.1$ & $\mathit{29.0}$\\
-    Voronoi & $\mathbf{0.00}$ & $1.29$ & $3.52$ & $6.87$ & $30.1$\\
-    BoxSpline & $0.00287$ & $4.39$ & $9.10$ & $14.8$ & $41.3$\\
+  \begin{tabular}{c|r@{.}l|r@{.}l|r@{.}l|r@{.}l|r@{.}l}
+    \hline
+    Algorithm & \multicolumn{2}{c|}{Min} & \multicolumn{2}{c|}{$25^{th}$} & \multicolumn{2}{c|}{$50^{th}$} & \multicolumn{2}{c|}{$75^{th}$} & \multicolumn{2}{c}{Max}\\
+    \hline
+    MARS & $0$&$00948$ & $9$&$98$ & $20$&$4$ & $32$&$9$ & $863$&$0$\\
+    SVR & $0$&$00138$ & $3$&$17$ & $7$&$31$ & $12$&$6$ & $\mathbf{27}$&$\mathbf{5}$\\
+    MLP & $0$&$0000239$ & $\mathit{0}$&$\mathit{533}$ & $\mathit{1}$&$\mathit{25}$ & $\mathbf{2}$&$\mathbf{84}$ & $39$&$3$\\
+    Delaunay & $\mathit{3}$&$\mathit{72 \times 10^{-12}}$ & $1$&$20$ & $3$&$50$ & $7$&$67$ & $30$&$7$\\
+    ShepMod & $\mathbf{0}$&$\mathbf{00}$ & $\mathbf{0}$&$\mathbf{255}$ & $\mathbf{0}$&$\mathbf{908}$ & $\mathit{3}$&$\mathit{43}$ & $34$&$5$\\
+    LSHEP & $0$&$00254$ & $2$&$93$ & $7$&$16$ & $13$&$1$ & $\mathit{29}$&$\mathit{0}$\\
+    Voronoi & $\mathbf{0}$&$\mathbf{00}$ & $1$&$29$ & $3$&$52$ & $6$&$87$ & $30$&$1$\\
+    BoxSpline & $0$&$00287$ & $4$&$39$ & $9$&$10$ & $14$&$8$ & $41$&$3$\\
     \hline
@@
   \centering
-  \begin{tabular}{|c|c| c |c|c|c|}
-    \cline{1-2}\cline{4-6}
-    Algorithm & \% Best &  & Fit Time & App. Time (1) & Total App. Time\\
-    \cline{1-2}\cline{4-6}
-    MARS & $0.0\%$ &  & $37.9$s & $2.53 \times 10^{-3}$s & $1.48$s\\
-    SVR & $0.1\%$ &  & $0.859$s & $\mathit{1.81 \times 10^{-4}}$s & $\mathit{0.106}$s\\
-    MLP & $\mathit{32.0}\%$ &  & $348.0$s & $1.11 \times 10^{-3}$s & $0.653$s\\
-    Delaunay & $0.0\%$ &  & $\mathbf{1.90 \times 10^{-4}}$s & $3.69$s & $2.17 \times 10^{3}$s\\
-    ShepMod & $\mathbf{66.4}\%$ &  & $1.13$s & $1.82 \times 10^{-4}$s & $0.107$s\\
-    LSHEP & $0.0\%$ &  & $2.39$s & $\mathbf{1.44 \times 10^{-4}}$s & $\mathbf{0.0845}$s\\
-    Voronoi & $1.6\%$ &  & $\mathit{0.298}$s & $0.0274$s & $16.1$s\\
-    BoxSpline & $0.0\%$ &  & $1.26$s & $6.43 \times 10^{-4}$s & $0.377$s\\
-    \cline{1-2}\cline{4-6}
+  \begin{tabular}{|c|r@{.}l| c |r@{.}l|r@{.}l|r@{.}l|}
+    \cline{1-3}\cline{5-10}
+    Algorithm & \multicolumn{2}{c|}{\% Best} &  & \multicolumn{2}{c|}{Fit/Prep. Time (s)} & \multicolumn{2}{c|}{App. Time (s)} & \multicolumn{2}{c|}{Total App. Time (s)}\\
+    \cline{1-3}\cline{5-10}
+    MARS & \quad\,$0$&$0$ &  & \quad\quad$37$&$9$ & \quad\,\,$0$&$00253$ & \quad\quad\quad\,\,$1$&$48$\\
+    SVR & $0$&$1$ &  & $0$&$859$ & $\mathit{0}$&$\mathit{000181}$ & $\mathit{0}$&$\mathit{106}$\\
+    MLP & $\mathit{32}$&$\mathit{0}$ &  & $348$&$0$ & $0$&$00111$ & $0$&$653$\\
+    Delaunay & $0$&$0$ &  & $\mathbf{0}$&$\mathbf{000190}$ & $3$&$69$ & $2170$&\\
+    ShepMod & $\mathbf{66}$&$\mathbf{4}$ &  & $1$&$13$ & $0$&$000182$ & $0$&$107$\\
+    LSHEP & $0$&$0$ &  & $2$&$39$ & $\mathbf{0}$&$\mathbf{000144}$ & $\mathbf{0}$&$\mathbf{0845}$\\
+    Voronoi & $1$&$6$ &  & $\mathit{0}$&$\mathit{298}$ & $0$&$0274$ & $16$&$1$\\
+    BoxSpline & $0$&$0$ &  & $1$&$26$ & $0$&$000643$ & $0$&$377$\\
+    \cline{1-3}\cline{5-10}
   \end{tabular}
   \caption{The left above shows how often each algorithm had the
-    lowest absolute error approximating parkinsons data in table
+    lowest absolute error approximating Parkinson's data in Table
     \ref{table:error-parkinsons}. On the right columns are median fit
@@
   \centering
-  \begin{tabular}{c|c|c|c|c|c}
-    \hline
-    Algorithm & Min & $25^{th}$ & $50^{th}$ & $75^{th}$ & Max\\
-    \hline
-    MARS & $\mathit{6.45 \times 10^{-15}}$ & $\mathbf{2.70 \times 10^{-14}}$ & $1.66$ & $1.96$ & $\mathit{53.3}$\\
-    SVR & $0.0420$ & $0.143$ & $0.148$ & $\mathit{0.860}$ & $119.0$\\
-    MLP & $6.89 \times 10^{-5}$ & $0.0337$ & $\mathbf{0.0795}$ & $\mathbf{0.264}$ & $\mathbf{5.31}$\\
-    Delaunay & $\mathbf{0.00}$ & $0.187$ & $0.919$ & $2.72$ & $56.3$\\
-    ShepMod & $\mathbf{0.00}$ & $0.0957$ & $0.685$ & $2.90$ & $73.2$\\
-    LSHEP & $\mathbf{0.00}$ & $\mathit{0.0153}$ & $\mathit{0.106}$ & $1.17$ & $113.0$\\
-    Voronoi & $\mathbf{0.00}$ & $0.430$ & $1.28$ & $2.94$ & $83.8$\\
-    BoxSpline & $\mathbf{0.00}$ & $0.789$ & $2.06$ & $5.25$ & $119.0$\\
+  \begin{tabular}{c|r@{.}l|r@{.}l|r@{.}l|r@{.}l|r@{.}l}
+    \hline
+    Algorithm & \multicolumn{2}{c|}{Min} & \multicolumn{2}{c|}{$25^{th}$} & \multicolumn{2}{c|}{$50^{th}$} & \multicolumn{2}{c|}{$75^{th}$} & \multicolumn{2}{c}{Max}\\
+    \hline
+    MARS & $\mathit{6}$&$\mathit{45 \times 10^{-15}}$ & $\mathbf{2}$&$\mathbf{70 \times 10^{-14}}$ & $1$&$66$ & $1$&$96$ & $\mathit{53}$&$\mathit{3}$\\
+    SVR & $0$&$0420$ & $0$&$143$ & $0$&$148$ & $\mathit{0}$&$\mathit{860}$ & $119$&$0$\\
+    MLP & $0$&$0000689$ & $0$&$0337$ & $\mathbf{0}$&$\mathbf{0795}$ & $\mathbf{0}$&$\mathbf{264}$ & $\mathbf{5}$&$\mathbf{31}$\\
+    Delaunay & $\mathbf{0}$&$\mathbf{00}$ & $0$&$187$ & $0$&$919$ & $2$&$72$ & $56$&$3$\\
+    ShepMod & $\mathbf{0}$&$\mathbf{00}$ & $0$&$0957$ & $0$&$685$ & $2$&$90$ & $73$&$2$\\
+    LSHEP & $\mathbf{0}$&$\mathbf{00}$ & $\mathit{0}$&$\mathit{0153}$ & $\mathit{0}$&$\mathit{106}$ & $1$&$17$ & $113$&$0$\\
+    Voronoi & $\mathbf{0}$&$\mathbf{00}$ & $0$&$430$ & $1$&$28$ & $2$&$94$ & $83$&$8$\\
+    BoxSpline & $\mathbf{0}$&$\mathbf{00}$ & $0$&$789$ & $2$&$06$ & $5$&$25$ & $119$&$0$\\
     \hline
@@
     percentiles correspond to the minimum, first quartile, median,
-    second quartile, and maximum absolute errors respectively. The
+    third quartile, and maximum absolute errors respectively. The
     minimum value of each column is boldface, while the second lowest
@@
   \centering
-  \begin{tabular}{|c|c| c |c|c|c|}
-    \cline{1-2}\cline{4-6}
-    Algorithm & \% Best &  & Fit Time & App. Time (1) & Total App. Time\\
-    \cline{1-2}\cline{4-6}
-    MARS & $10.7\%$ &  & $0.151$s & $1.17 \times 10^{-4}$s & $0.0304$s\\
-    SVR & $4.1\%$ &  & $0.133$s & $\mathbf{9.47 \times 10^{-5}}$s & $\mathbf{0.0246}$s\\
-    MLP & $\mathbf{56.8}\%$ &  & $169.0$s & $1.37 \times 10^{-3}$s & $0.356$s\\
-    Delaunay & $0.1\%$ &  & $\mathbf{1.16 \times 10^{-4}}$s & $1.55$s & $403.0$s\\
-    ShepMod & $3.5\%$ &  & $0.265$s & $1.28 \times 10^{-4}$s & $0.0333$s\\
-    LSHEP & $\mathit{28.4}\%$ &  & $0.874$s & $\mathit{9.75 \times 10^{-5}}$s & $\mathit{0.0254}$s\\
-    Voronoi & $0.2\%$ &  & $\mathit{0.0112}$s & $0.0270$s & $7.01$s\\
-    BoxSpline & $0.3\%$ &  & $0.330$s & $4.06 \times 10^{-4}$s & $0.106$s\\
-    \cline{1-2}\cline{4-6}
+  \begin{tabular}{|c|r@{.}l| c |r@{.}l|r@{.}l|r@{.}l|}
+    \cline{1-3}\cline{5-10}
+    Algorithm & \multicolumn{2}{c|}{\% Best} &  & \multicolumn{2}{c|}{Fit/Prep. Time (s)} & \multicolumn{2}{c|}{App. Time (s)} & \multicolumn{2}{c|}{Total App. Time (s)}\\
+    \cline{1-3}\cline{5-10}
+    MARS & \,\,\,\,$10$&$7$ &  & \quad\quad\,\,$0$&$151$ & \quad$0$&$000117$ & \quad\quad\quad\,\,$0$&$0304$\\
+    SVR & $4$&$1$ &  & $0$&$133$ & $\mathbf{0}$&$\mathbf{0000947}$ & $\mathbf{0}$&$\mathbf{0246}$\\
+    MLP & $\mathbf{56}$&$\mathbf{8}$ &  & $169$&$0$ & $0$&$00137$ & $0$&$356$\\
+    Delaunay & $0$&$1$ &  & $\mathbf{0}$&$\mathbf{000116}$ & $1$&$55$ & $403$&$0$\\
+    ShepMod & $3$&$5$ &  & $0$&$265$ & $0$&$000128$ & $0$&$0333$\\
+    LSHEP & $\mathit{28}$&$\mathit{4}$ &  & $0$&$874$ & $\mathit{0}$&$\mathit{0000975}$ & $\mathit{0}$&$\mathit{0254}$\\
+    Voronoi & $0$&$2$ &  & $\mathit{0}$&$\mathit{0112}$ & $0$&$0270$ & $7$&$01$\\
+    BoxSpline & $0$&$3$ &  & $0$&$330$ & $0$&$000406$ & $0$&$106$\\
+    \cline{1-3}\cline{5-10}
   \end{tabular}
-  \caption{The left above shows how often each algorithm had the
-    lowest absolute error approximating Sydney rainfall data in table
+  \caption{Left table shows how often each algorithm had the lowest
+    absolute error approximating Sydney rainfall data in Table
     \ref{table:error-weather}. On the right columns are median fit
@@
   \centering
-  \begin{tabular}{c|c|c|c|c|c}
-    \hline
-    Algorithm & Min & $25^{th}$ & $50^{th}$ & $75^{th}$ & Max\\
-    \hline
-    MARS & $5.36$ & $3610.0$ & $4580.0$ & $5450.0$ & $13400.0$\\
-    SVR & $0.0138$ & $8.47$ & $15.4$ & $41.9$ & $7700.0$\\
-    MLP & $0.00151$ & $\mathbf{1.60}$ & $\mathbf{3.86}$ & $\mathbf{8.34}$ & $\mathbf{604.0}$\\
-    Delaunay & $\mathbf{0.00}$ & $\mathit{2.69}$ & $\mathit{13.8}$ & $\mathit{35.0}$ & $\mathit{4840.0}$\\
-    ShepMod & $\mathbf{0.00}$ & $4.21$ & $17.3$ & $51.6$ & $6510.0$\\
-    LSHEP & $1.27$ & $199.0$ & $260.0$ & $343.0$ & $6530.0$\\
-    Voronoi & $\mathit{2.89 \times 10^{-10}}$ & $14.7$ & $32.8$ & $52.9$ & $4860.0$\\
-    BoxSpline & $0.00740$ & $20.8$ & $40.9$ & $79.6$ & $7640.0$\\
+  \begin{tabular}{c|r@{.}l|r@{.}l|r@{.}l|r@{.}l|r@{.}l}
+    \hline
+    Algorithm & \multicolumn{2}{c|}{Min} & \multicolumn{2}{c|}{$25^{th}$} & \multicolumn{2}{c|}{$50^{th}$} & \multicolumn{2}{c|}{$75^{th}$} & \multicolumn{2}{c}{Max}\\
+    \hline
+    MARS & $5$&$36$ & $3610$&$0$ & $4580$&$0$ & $5450$&$0$ & $13400$&$0$\\
+    SVR & $0$&$0138$ & $8$&$47$ & $15$&$4$ & $41$&$9$ & $7700$&$0$\\
+    MLP & $0$&$00151$ & $\mathbf{1}$&$\mathbf{60}$ & $\mathbf{3}$&$\mathbf{86}$ & $\mathbf{8}$&$\mathbf{34}$ & $\mathbf{604}$&$\mathbf{0}$\\
+    Delaunay & $\mathbf{0}$&$\mathbf{00}$ & $\mathit{2}$&$\mathit{69}$ & $\mathit{13}$&$\mathit{8}$ & $\mathit{35}$&$\mathit{0}$ & $\mathit{4840}$&$\mathit{0}$\\
+    ShepMod & $\mathbf{0}$&$\mathbf{00}$ & $4$&$21$ & $17$&$3$ & $51$&$6$ & $6510$&$0$\\
+    LSHEP & $1$&$27$ & $199$&$0$ & $260$&$0$ & $343$&$0$ & $6530$&$0$\\
+    Voronoi & $\mathit{2}$&$\mathit{89 \times 10^{-10}}$ & $14$&$7$ & $32$&$8$ & $52$&$9$ & $4860$&$0$\\
+    BoxSpline & $0$&$00740$ & $20$&$8$ & $40$&$9$ & $79$&$6$ & $7640$&$0$\\
     \hline
@@
   \centering
-  \begin{tabular}{|c|c| c |c|c|c|}
-    \cline{1-2}\cline{4-6}
-    Algorithm & \% Best &  & Fit Time & App. Time (1) & Total App. Time\\
-    \cline{1-2}\cline{4-6}
-    MARS & $0.0\%$ &  & $22.0$s & $1.48 \times 10^{-3}$s & $0.820$s\\
-    SVR & $0.0\%$ &  & $1.01$s & $\mathit{2.10 \times 10^{-4}}$s & $\mathit{0.117}$s\\
-    MLP & $\mathbf{79.5}\%$ &  & $290.0$s & $7.14 \times 10^{-4}$s & $0.397$s\\
-    Delaunay & $\mathit{20.5}\%$ &  & $\mathbf{2.43 \times 10^{-4}}$s & $6.83$s & $3.80 \times 10^{3}$s\\
-    ShepMod & $0.1\%$ &  & $1.45$s & $2.20 \times 10^{-4}$s & $0.122$s\\
-    LSHEP & $0.0\%$ &  & $3.47$s & $\mathbf{1.76 \times 10^{-4}}$s & $\mathbf{0.0981}$s\\
-    Voronoi & $0.0\%$ &  & $\mathit{0.197}$s & $0.0950$s & $52.8$s\\
-    BoxSpline & $0.0\%$ &  & $1.66$s & $9.56 \times 10^{-4}$s & $0.532$s\\
-    \cline{1-2}\cline{4-6}
+  \begin{tabular}{|c|r@{.}l| c |r@{.}l|r@{.}l|r@{.}l|}
+    \cline{1-3}\cline{5-10}
+    Algorithm & \multicolumn{2}{c|}{\% Best} &  & \multicolumn{2}{c|}{Fit/Prep. Time (s)} & \multicolumn{2}{c|}{App. Time (s)} & \multicolumn{2}{c|}{Total App. Time (s)}\\
+    \cline{1-3}\cline{5-10}
+    MARS & \quad\,$0$&$0$ &  & \quad\quad\,\,$22$&$0$ & \quad$0$&$00148$ & \quad\quad\quad$0$&$820$\\
+    SVR & $0$&$0$ &  & $1$&$01$ & $\mathit{0}$&$\mathit{000210}$ & $\mathit{0}$&$\mathit{117}$\\
+    MLP & $\mathbf{79}$&$\mathbf{5}$ &  & $290$&$0$ & $0$&$000714$ & $0$&$397$\\
+    Delaunay & $\mathit{20}$&$\mathit{5}$ &  & $\mathbf{0}$&$\mathbf{000243}$ & $6$&$83$ & $3800$&$$\\
+    ShepMod & $0$&$1$ &  & $1$&$45$ & $0$&$000220$ & $0$&$122$\\
+    LSHEP & $0$&$0$ &  & $3$&$47$ & $\mathbf{0}$&$\mathbf{000176}$ & $\mathbf{0}$&$\mathbf{0981}$\\
+    Voronoi & $0$&$0$ &  & $\mathit{0}$&$\mathit{197}$ & $0$&$0950$ & $52$&$8$\\
+    BoxSpline & $0$&$0$ &  & $1$&$66$ & $0$&$000956$ & $0$&$532$\\
+    \cline{1-3}\cline{5-10}
   \end{tabular}
@@
   \centering
-  \begin{tabular}{c|c|c|c|c|c}
-    \hline
-    Algorithm & Min & $25^{th}$ & $50^{th}$ & $75^{th}$ & Max\\
-    \hline
-    Delaunay & $\mathbf{0.0287}$ & $\mathbf{0.0914}$ & $\mathbf{0.158}$ & $\mathit{0.308}$ & $\mathit{0.897}$\\
-    ShepMod & $0.0307$ & $\mathit{0.0983}$ & $\mathit{0.164}$ & $0.335$ & $1.00$\\
-    Voronoi & $\mathit{0.0303}$ & $0.100$ & $0.165$ & $\mathbf{0.274}$ & $\mathbf{0.762}$\\
-    BoxSpline & $0.0703$ & $0.417$ & $0.561$ & $0.745$ & $1.00$\\
+  \begin{tabular}{c|r@{.}l|r@{.}l|r@{.}l|r@{.}l|r@{.}l}
+    \hline
+    Algorithm & \multicolumn{2}{c|}{Min} & \multicolumn{2}{c|}{$25^{th}$} & \multicolumn{2}{c|}{$50^{th}$} & \multicolumn{2}{c|}{$75^{th}$} & \multicolumn{2}{c}{Max}\\
+    \hline
+    Delaunay & $\mathbf{0}$&$\mathbf{0287}$ & $\mathbf{0}$&$\mathbf{0914}$ & $\mathbf{0}$&$\mathbf{158}$ & $\mathit{0}$&$\mathit{308}$ & $\mathit{0}$&$\mathit{897}$\\
+    ShepMod & $0$&$0307$ & $\mathit{0}$&$\mathit{0983}$ & $\mathit{0}$&$\mathit{164}$ & $0$&$335$ & $1$&$00$\\
+    Voronoi & $\mathit{0}$&$\mathit{0303}$ & $0$&$100$ & $0$&$165$ & $\mathbf{0}$&$\mathbf{274}$ & $\mathbf{0}$&$\mathbf{762}$\\
+    BoxSpline & $0$&$0703$ & $0$&$417$ & $0$&$561$ & $0$&$745$ & $1$&$00$\\
     \hline
@@
     percentiles correspond to the minimum, first quartile, median,
-    second quartile, and maximum KS statistics respectively between
+    third quartile, and maximum KS statistics respectively between
     truth and guess for models predicting the distribution of I/O
@@
   \centering
-  \begin{tabular}{|c|c| c |c|c|c|}
-    \cline{1-2}\cline{4-6}
-    Algorithm & \% Best &  & Fit Time & App. Time (1) & Total App. Time\\
-    \cline{1-2}\cline{4-6}
-    Delaunay & $\mathbf{62.0}\%$ &  & $\mathbf{5.17 \times 10^{-4}}$s & $0.353$s & $106.0$s\\
-    ShepMod & $0.0\%$ &  & $0.0884$s & $\mathbf{1.45 \times 10^{-4}}$s & $\mathbf{0.0436}$s\\
-    Voronoi & $\mathit{38.0}\%$ &  & $\mathit{0.0173}$s & $2.53 \times 10^{-3}$s & $0.762$s\\
-    BoxSpline & $0.0\%$ &  & $0.0972$s & $\mathit{2.10 \times 10^{-4}}$s & $\mathit{0.0633}$s\\
-    \cline{1-2}\cline{4-6}
+  \begin{tabular}{|c|r@{.}l| c |r@{.}l|r@{.}l|r@{.}l|}
+    \cline{1-3}\cline{5-10}
+    Algorithm & \multicolumn{2}{c|}{\% Best} &  & \multicolumn{2}{c|}{Fit/Prep. Time (s)} & \multicolumn{2}{c|}{App. Time (s)} & \multicolumn{2}{c|}{Total App. Time (s)}\\
+    \cline{1-3}\cline{5-10}
+    Delaunay & \,\,$\mathbf{62}$&$\mathbf{0}$ &  & \quad\quad$\mathbf{0}$&$\mathbf{000517}$ & \quad$0$&$353$ & \quad\quad\,\,$106$&$0$\\
+    ShepMod & $0$&$0$ &  & $0$&$0884$ & $\mathbf{0}$&$\mathbf{000145}$ & $\mathbf{0}$&$\mathbf{0436}$\\
+    Voronoi & $\mathit{38}$&$\mathit{0}$ &  & $\mathit{0}$&$\mathit{0173}$ & $0$&$00253$ & $0$&$762$\\
+    BoxSpline & $0$&$0$ &  & $0$&$0972$ & $\mathit{0}$&$\mathit{000210}$ & $\mathit{0}$&$\mathit{0633}$\\
+    \cline{1-3}\cline{5-10}
   \end{tabular}

### 2019-03-20 Paper_[2019-02]_Numerical_Algorithms/revision-one.tex
created an affine weights visual. finished NA paper draft. wrote report on Xiaodong clock cycle data.

--- 2019-03-20 Paper_[2019-02]_Numerical_Algorithms/paper.tex
@@
 \usepackage{booktabs} % For formal tables
-\usepackage{tabularx} % For tables with nice column widths.
+\usepackage{siunitx}  % For table aligned by decimal point
 \usepackage[colorlinks,allcolors=blue]{hyperref}
-\usepackage{float}    % For controlling the placement of tables.
 
@@
   \begin{tabular}[#1]{@{}c@{}}#2\end{tabular}}
-
-\newcolumntype{d}[1]{D{.}{.}{#1}}
 
@@
 \begin{abstract}
-Increases in the quantity of available data have allowed all fields of
-science to generate more accurate models of multivariate phenomena.
-Regression and interpolation become challenging when the dimension of
-data is large, especially while maintaining tractable computational
-complexity. Regression is a popular approach to solving approximation
-problems with high dimension, however there are often advantages to
-interpolation. This paper presents a novel and insightful error bound
-for (piecewise) linear interpolation in arbitrary dimension and
-contrasts the performance of some interpolation techniques with
-popular regression techniques. Empirical results demonstrate the
-viability of interpolation for moderately high dimensional
-approximation problems, and encourage broader application of
-interpolants to multivariate approximation in science.
+A rapid increase in the quantity of data available is allowing all
+fields of science to generate more accurate models of multivariate
+phenomena. Regression and interpolation become challenging when the
+dimension of data is large, especially while maintaining tractable
+computational complexity. While regression is a popular approach to
+solving approximation problems with high dimension, there are often
+advantages to interpolation. This paper presents error bounds for
+interpolants on moderately high dimensional problems and contrasts
+their performance with that of popular regression techniques.
+Empirical results demonstrate the viability of interpolants for high
+dimensional approximation problems, suggesting that these techniques
+are capable of effectively modeling multivariate phenomena while
+maintaining flexibility in different application domains.
 
@@
 
-\newpage
-
 \subsection{Modified Shepard}
@@
 
-This section presents the theoretical results bounding the error of
-(piecewise) linear interpolation. The error analysis relies on linear
+\color{red} This section presents the theoretical results bounding the error of
+simplicial interpolation. The error analysis relies on simplicial
 interpolation for three reasons: (1) second order results can be
@@
 values, which may allow for straight forward extensions of this error
-bound.
+bound. \color{black}
 
@@
 
-In summary, the approximation error of a linear (simplicial)
-interpolant tends quadratically towards zero when approaching observed
-data only when the diameter of the simplex is also reduced at a
-proportional rate.  Only linear convergence to the true function can
-be achieved in practice, without the incorporation of additional
-observations. Notice that the approximation error is largely
-determined by the spacing of observed data. Predictions made by
-simplices whose vertices are not well-spaced (i.e. have large
-diameter, or are nearly contained in a hyperplane) have higher
-error. In light of this error bound, an empirical evaluation of
-presented algorithms follows.
+In summary, the approximation error of a simplicial interpolant tends
+quadratically towards zero when approaching observed data only when
+the diameter of the simplex is also reduced at a proportional rate.
+Only linear convergence to the true function can be achieved in
+practice, without the incorporation of additional observations. Notice
+that the approximation error is largely determined by the spacing of
+observed data. Predictions made by simplices whose vertices are not
+well-spaced (i.e. have large diameter, or are nearly contained in a
+hyperplane) have higher error. In light of this error bound, an
+empirical evaluation of presented algorithms follows.
 
 %% ===================================================================
-\vspace{-2mm}
 \section{Data and Empirical Analysis}
@@
 The theoretical results constructed in Section \ref{sec:theory} for
-(piecewise) linear interpolation are promising and apply directly to
-Delaunay interpolation, however they are difficult to interpret in
-context with approximation algorithms that do not have similar known
-uniform error bounds. For that reason, this paper utilizes five
-different data sets of varying dimension and application to construct
-approximations and compare the accuracy of different techniques.
+simplicial interpolation are promising, however they are difficult to
+interpret in context with approximation algorithms that do not have
+similar known uniform error bounds. For that reason, this paper
+utilizes five different data sets of varying dimension and application
+to construct approximations and compare the accuracy of different
+techniques.
 
@@
 described in \cite{kohavi1995study} with $k=10$. This approach
-randomly partitions data into $k$ equal sized (nearly) sets. Each
+randomly partitions data into $k$ nearly equal sized sets. Each
 algorithm is then evaluated by constructing an approximation over each
@@
   \centering
-  \begin{tabular}{c|r@{.}l|r@{.}l|r@{.}l}
-    Algorithm & \multicolumn{2}{c|}{Avg. \% Best} & \multicolumn{2}{c|}{\multilinecell{Avg. Fit or \\ Prep. Time (s)}} & \multicolumn{2}{c}{Avg. App. Time (s)}\\
-    \hline
-    MARS & \quad\quad$4$&$6$ & \,\quad$22$&$3$ & \quad\quad$0$&$00137$\\
-    SVR & $\mathit{21}$&$\mathit{1}$ & $0$&$503$ & $\mathit{0}$&$\mathit{000137}$\\
-    MLP & $\mathbf{42}$&$\mathbf{1}$ & $210$&$0$ & $0$&$00102$\\
-    Delaunay & $6$&$0$ & $\mathbf{0}$&$\mathbf{000143}$ & $3$&$03$\\
-    ShepMod & $18$&$4$ & $0$&$711$ & $0$&$000149$\\
-    LSHEP & $8$&$4$ & $1$&$69$ & $\mathbf{0}$&$\mathbf{000116}$\\
-    Voronoi & $0$&$5$ & $\mathit{0}$&$\mathit{127}$ & $0$&$0375$\\
-    BoxSplines & $0$&$2$ & $0$&$816$ & $0$&$000526$\\
+  \begin{tabular}{c|c|c|c}
+    \hline
+    Algorithm & Avg. \% Best & Avg. Fit / Prep. Time (s) & Avg. App. Time (s)\\
+    \hline
+    MARS & $4.6$ & $22.3$ & $1.37 \times 10^{-3}$\\
+    SVR & $\mathit{21.1}$ & $0.503$ & $\mathit{1.37 \times 10^{-4}}$\\
+    MLP & $\mathbf{42.1}$ & $210.0$ & $1.02 \times 10^{-3}$\\
+    Delaunay & $6.0$ & $\mathbf{1.43 \times 10^{-4}}$ & $3.03$\\
+    ShepMod & $18.4$ & $0.711$ & $1.49 \times 10^{-4}$\\
+    LSHEP & $8.4$ & $1.69$ & $\mathbf{1.16 \times 10^{-4}}$\\
+    Voronoi & $0.5$ & $\mathit{0.127}$ & $0.0375$\\
+    BoxSplines & $0.2$ & $0.816$ & $5.26 \times 10^{-4}$\\
+    \hline
   \end{tabular}
@@
 The major advantages to interpolation lie in the near absence of
-\textit{fit} time. Delaunay, LSHEP, and ShepMod all require pairwise
-distance calculations, for numerical robustness (Delaunay) and to
-determine the radii of influence for data points (LSHEP and
-ShepMod). At least hundreds, and sometimes hundreds of thousands of
-predictions can be made by the interpolants before the most widely
-used regressor (MLP) finishes fitting these relatively small data
-sets.  However, the computational complexities of all interpolants
-presented are greater than linear in either dimension or number of
-points, whereas the regressors' nonlinear complexity in dimension
-generally comes from the model fitting optimization.
-
-The new theoretical results presented in Section \ref{sec:theory}
-directly apply to Delaunay interpolation, however the performance of
-Delaunay does not appear significantly better than other algorithms on
-these data sets.  This observation may be due to the stochastic nature
-of the data, but it also speaks to the power of the approximations
-generated by the different interpolation methods.  The strong
-performance of other interpolants suggests that theoretical results
-similar to those presented in this work can be achieved for the other
-interpolants under reasonable assumptions.
+\textit{fit} time. Delaunay has no preparations required, LSHEP and
+ShepMod require pairwise distance calculations to determine the radii
+of influence for points in each model. At least hundreds, and
+sometimes hundreds of thousands of predictions can be made by the
+interpolants before the most widely used regressor (the MLP) finishes
+fitting these relatively small data sets. However, the computational
+complexities of all interpolants presented are greater than linear in
+either dimension or number of points, meaning large enough data (in
+dimension and number of observations) would cause interpolation to be
+slower.
+
+The theoretical results presented in Section \ref{sec:theory} directly
+apply to Delaunay interpolation, however the performance of Delaunay
+does not appear significantly better than other algorithms on these
+data sets. This observation may be due to the stochastic nature of the
+data, but it also speaks to the alignment of the geometry of the
+underlying functions with the approximations generated by the
+different interpolation methods. The strong performance of other
+interpolants suggests that theoretical results similar to those
+presented in this work can be achieved for the other interpolants
+under reasonable assumptions.
 
 Finally, most of the interpolants presented in this work benefit from
-the ability to model any function over real $d$-tuples with a range
-that is closed under convex combinations. The results of the
+the ability to model any function over a real vector space with a
+range that is closed under convex combinations. The results of the
 distribution prediction case study indicate that interpolants can
@@
 error. The average absolute errors in the predicted CDFs are always
-lower than the KS statistics. However, the KS statistic was chosen as
-a metric because of the important surrounding statistical theory. A
-nonnegligible volume of predictions provide impressively low levels of
-average absolute error in that final case study.
+lower than the KS statistics. However, the KS statistic was chosen
+because of the important statistical theory surrounding it as an error
+measure. Considering this circumstance, a nonnegligible volume of
+predictions provide impressively low levels of error in that final
+case study.
 
@@
 
-The major contributions of this work are: 1) new uniform theoretical
-error bounds for piecewise linear interpolation in arbitrary dimension
-(Section \ref{sec:theory}); and 2) an empirical evaluation across
-real-world problems (Section \ref{sec:data}) that demonstrates
-interpolants produce competitively accurate models of multivariate
-phenomenon when compared with common regressors for sparse, moderately
-high dimensional problems.
+This work presents new theoretical results for interpolation in
+arbitrary dimension with simplicial approximations. Coinciding with
+the theoretical results, an empirical evaluation across real-world
+problems demonstrates that interpolants produce competitively accurate
+models of multivariate phenomenon when compared with common regressors
+for sparse, moderately high dimensional problems.
 
@@
 $30$ dimensions. The underlying constructions are theoretically
-straightforward, interpretable, and yield reasonably accurate
-predictions. Most of the interpolants' computational complexities make
-them particularly suitable for applications in even higher
-dimension. The major benefits of interpolation are seen when only a
-small number of approximations ($\leq 1000$) are made from data and
-when there are relatively few data points for the dimension (for
-empirical results presented, $\log_d n \leq 5$).
+straightforward, yet powerful and flexible. Most of the interpolants'
+computational complexities make them particularly suitable for
+applications in even higher dimension. The major benefits of
+interpolation are seen when only a small number of approximations
+($\leq 1000$) are made from data and when there are relatively few
+data points for the dimension (for empirical results presented,
+$\log_d n \leq 5$).
 
@@
 \begin{appendix}
+
 \section*{Appendix}
-\label{sec:appendix}
-
-\vspace*{\fill}
-
-\begin{table}[H]
-  \centering
-  \begin{tabular}{c|r@{.}l|r@{.}l|r@{.}l|r@{.}l|r@{.}l}
-    \hline
-    Algorithm & \multicolumn{2}{c|}{Min} & \multicolumn{2}{c|}{$25^{th}$} & \multicolumn{2}{c|}{$50^{th}$} & \multicolumn{2}{c|}{$75^{th}$} & \multicolumn{2}{c}{Max}\\
-    \hline
-    MARS & $0$&$00984$ & $3$&$11$ & $7$&$01$ & $\mathit{11}$&$\mathit{7}$ & $1090$&$0$\\
-    SVR & $\mathit{0}$&$\mathit{00402}$ & $\mathbf{0}$&$\mathbf{243}$ & $\mathbf{0}$&$\mathbf{416}$ & $\mathbf{6}$&$\mathbf{19}$ & $1090$&$0$\\
-    MLP & $0$&$0426$ & $2$&$63$ & $\mathit{5}$&$\mathit{27}$ & $14$&$0$ & $1090$&$0$\\
-    Delaunay & $\mathbf{0}$&$\mathbf{00}$ & $1$&$98$ & $5$&$37$ & $13$&$1$ & $\mathit{1080}$&$\mathit{0}$\\
-    ShepMod & $\mathbf{0}$&$\mathbf{00}$ & $\mathit{1}$&$\mathit{93}$ & $6$&$27$ & $16$&$0$ & $1090$&$0$\\
-    LSHEP & $0$&$0400$ & $2$&$17$ & $8$&$87$ & $19$&$1$ & $\mathbf{1070}$&$\mathbf{0}$\\
-    Voronoi & $0$&$00982$ & $3$&$65$ & $7$&$56$ & $15$&$6$ & $1090$&$0$\\
-    BoxSpline & $\mathbf{0}$&$\mathbf{00}$ & $2$&$27$ & $5$&$91$ & $16$&$4$ & $1090$&$0$\\
+
+\begin{table}
+  \centering
+  \begin{tabular}{c|c|c|c|c|c}
+    \hline
+    Algorithm & Min & $25^{th}$ & $50^{th}$ & $75^{th}$ & Max\\
+    \hline
+    MARS & $0.00984$ & $3.11$ & $7.01$ & $\mathit{11.7}$ & $1090.0$\\
+    SVR & $\mathit{0.00402}$ & $\mathbf{0.243}$ & $\mathbf{0.416}$ & $\mathbf{6.19}$ & $1090.0$\\
+    MLP & $0.0426$ & $2.63$ & $\mathit{5.27}$ & $14.0$ & $1090.0$\\
+    Delaunay & $\mathbf{0.00}$ & $1.98$ & $5.37$ & $13.1$ & $\mathit{1080.0}$\\
+    ShepMod & $\mathbf{0.00}$ & $\mathit{1.93}$ & $6.27$ & $16.0$ & $1090.0$\\
+    LSHEP & $0.0400$ & $2.17$ & $8.87$ & $19.1$ & $\mathbf{1070.0}$\\
+    Voronoi & $0.00982$ & $3.65$ & $7.56$ & $15.6$ & $1090.0$\\
+    BoxSpline & $\mathbf{0.00}$ & $2.27$ & $5.91$ & $16.4$ & $1090.0$\\
     \hline
@@
 
-\begin{table}[H]
-  \centering
-  \begin{tabular}{|c|r@{.}l| c |r@{.}l|r@{.}l|r@{.}l|}
-    \cline{1-3}\cline{5-10}
-    Algorithm & \multicolumn{2}{c|}{\% Best} &  & \multicolumn{2}{c|}{Fit/Prep. Time (s)} & \multicolumn{2}{c|}{App. Time (s)} & \multicolumn{2}{c|}{Total App. Time (s)}\\
-    \cline{1-3}\cline{5-10}
-    MARS & \quad$\mathit{7}$&$\mathit{7}$ &  & \quad\,\,\,$29$&$1$ & \quad$0$&$00137$ & \quad\quad\quad$0$&$0686$\\
-    SVR & $\mathbf{80}$&$\mathbf{2}$ &  & $0$&$00584$ & $\mathit{0}$&$\mathit{0000620}$ & $\mathit{0}$&$\mathit{00310}$\\
-    MLP & $0$&$0$ &  & $32$&$8$ & $0$&$000871$ & $0$&$0436$\\
-    Delaunay & $3$&$5$ &  & $\mathbf{0}$&$\mathbf{0000249}$ & $0$&$0385$ & $1$&$93$\\
-    ShepMod & $3$&$7$ &  & $0$&$00634$ & $0$&$0000644$ & $0$&$00322$\\
-    LSHEP & $5$&$1$ &  & $0$&$0275$ & $\mathbf{0}$&$\mathbf{0000463}$ & $\mathbf{0}$&$\mathbf{00231}$\\
-    Voronoi & $0$&$0$ &  & $\mathit{0}$&$\mathit{000182}$ & $0$&$000396$ & $0$&$0198$\\
-    BoxSpline & $0$&$4$ &  & $0$&$00724$ & $0$&$0000978$ & $0$&$00489$\\
-    \cline{1-3}\cline{5-10}
+\begin{table}
+  \centering
+  \begin{tabular}{|c|c| c |c|c|c|}
+    \cline{1-2}\cline{4-6}
+    Algorithm & \% Best &  & Fit / Prep. Time (s) & App. Time (s) & Total App. Time (s)\\
+    \cline{1-2}\cline{4-6}
+    MARS & $\mathit{7.7}$ &  & $29.1$ & $1.37 \times 10^{-3}$ & $0.0686$\\
+    SVR & $\mathbf{80.2}$ &  & $5.84 \times 10^{-3}$ & $\mathit{6.20 \times 10^{-5}}$ & $\mathit{3.10 \times 10^{-3}}$\\
+    MLP & $0.0$ &  & $32.8$ & $8.71 \times 10^{-4}$ & $0.0436$\\
+    Delaunay & $3.5$ &  & $\mathbf{2.49 \times 10^{-5}}$ & $0.0385$ & $1.93$\\
+    ShepMod & $3.7$ &  & $6.34 \times 10^{-3}$ & $6.44 \times 10^{-5}$ & $3.22 \times 10^{-3}$\\
+    LSHEP & $5.1$ &  & $0.0275$ & $\mathbf{4.63 \times 10^{-5}}$ & $\mathbf{2.31 \times 10^{-3}}$\\
+    Voronoi & $0.0$ &  & $\mathit{1.82 \times 10^{-4}}$ & $3.96 \times 10^{-4}$ & $0.0198$\\
+    BoxSpline & $0.4$ &  & $7.24 \times 10^{-3}$ & $9.78 \times 10^{-5}$ & $4.89 \times 10^{-3}$\\
+    \cline{1-2}\cline{4-6}
   \end{tabular}
@@
 
-\vspace*{\fill}
-\newpage
-
 \begin{table}
   \centering
-  \begin{tabular}{c|r@{.}l|r@{.}l|r@{.}l|r@{.}l|r@{.}l}
-    \hline
-    Algorithm & \multicolumn{2}{c|}{Min} & \multicolumn{2}{c|}{$25^{th}$} & \multicolumn{2}{c|}{$50^{th}$} & \multicolumn{2}{c|}{$75^{th}$} & \multicolumn{2}{c}{Max}\\
-    \hline
-    MARS & $0$&$00948$ & $9$&$98$ & $20$&$4$ & $32$&$9$ & $863$&$0$\\
-    SVR & $0$&$00138$ & $3$&$17$ & $7$&$31$ & $12$&$6$ & $\mathbf{27}$&$\mathbf{5}$\\
-    MLP & $0$&$0000239$ & $\mathit{0}$&$\mathit{533}$ & $\mathit{1}$&$\mathit{25}$ & $\mathbf{2}$&$\mathbf{84}$ & $39$&$3$\\
-    Delaunay & $\mathit{3}$&$\mathit{72 \times 10^{-12}}$ & $1$&$20$ & $3$&$50$ & $7$&$67$ & $30$&$7$\\
-    ShepMod & $\mathbf{0}$&$\mathbf{00}$ & $\mathbf{0}$&$\mathbf{255}$ & $\mathbf{0}$&$\mathbf{908}$ & $\mathit{3}$&$\mathit{43}$ & $34$&$5$\\
-    LSHEP & $0$&$00254$ & $2$&$93$ & $7$&$16$ & $13$&$1$ & $\mathit{29}$&$\mathit{0}$\\
-    Voronoi & $\mathbf{0}$&$\mathbf{00}$ & $1$&$29$ & $3$&$52$ & $6$&$87$ & $30$&$1$\\
-    BoxSpline & $0$&$00287$ & $4$&$39$ & $9$&$10$ & $14$&$8$ & $41$&$3$\\
+  \begin{tabular}{c|c|c|c|c|c}
+    \hline
+    Algorithm & Min & $25^{th}$ & $50^{th}$ & $75^{th}$ & Max\\
+    \hline
+    MARS & $0.00948$ & $9.98$ & $20.4$ & $32.9$ & $863.0$\\
+    SVR & $0.00138$ & $3.17$ & $7.31$ & $12.6$ & $\mathbf{27.5}$\\
+    MLP & $2.39 \times 10^{-5}$ & $\mathit{0.533}$ & $\mathit{1.25}$ & $\mathbf{2.84}$ & $39.3$\\
+    Delaunay & $\mathit{3.72 \times 10^{-12}}$ & $1.20$ & $3.50$ & $7.67$ & $30.7$\\
+    ShepMod & $\mathbf{0.00}$ & $\mathbf{0.255}$ & $\mathbf{0.908}$ & $\mathit{3.43}$ & $34.5$\\
+    LSHEP & $0.00254$ & $2.93$ & $7.16$ & $13.1$ & $\mathit{29.0}$\\
+    Voronoi & $\mathbf{0.00}$ & $1.29$ & $3.52$ & $6.87$ & $30.1$\\
+    BoxSpline & $0.00287$ & $4.39$ & $9.10$ & $14.8$ & $41.3$\\
     \hline
@@
   \centering
-  \begin{tabular}{|c|r@{.}l| c |r@{.}l|r@{.}l|r@{.}l|}
-    \cline{1-3}\cline{5-10}
-    Algorithm & \multicolumn{2}{c|}{\% Best} &  & \multicolumn{2}{c|}{Fit/Prep. Time (s)} & \multicolumn{2}{c|}{App. Time (s)} & \multicolumn{2}{c|}{Total App. Time (s)}\\
-    \cline{1-3}\cline{5-10}
-    MARS & \quad\,$0$&$0$ &  & \quad\quad$37$&$9$ & \quad\,\,$0$&$00253$ & \quad\quad\quad\,\,$1$&$48$\\
-    SVR & $0$&$1$ &  & $0$&$859$ & $\mathit{0}$&$\mathit{000181}$ & $\mathit{0}$&$\mathit{106}$\\
-    MLP & $\mathit{32}$&$\mathit{0}$ &  & $348$&$0$ & $0$&$00111$ & $0$&$653$\\
-    Delaunay & $0$&$0$ &  & $\mathbf{0}$&$\mathbf{000190}$ & $3$&$69$ & $2170$&\\
-    ShepMod & $\mathbf{66}$&$\mathbf{4}$ &  & $1$&$13$ & $0$&$000182$ & $0$&$107$\\
-    LSHEP & $0$&$0$ &  & $2$&$39$ & $\mathbf{0}$&$\mathbf{000144}$ & $\mathbf{0}$&$\mathbf{0845}$\\
-    Voronoi & $1$&$6$ &  & $\mathit{0}$&$\mathit{298}$ & $0$&$0274$ & $16$&$1$\\
-    BoxSpline & $0$&$0$ &  & $1$&$26$ & $0$&$000643$ & $0$&$377$\\
-    \cline{1-3}\cline{5-10}
+  \begin{tabular}{|c|c| c |c|c|c|}
+    \cline{1-2}\cline{4-6}
+    Algorithm & \% Best &  & Fit / Prep. Time (s) & App. Time (s) & Total App. Time (s)\\
+    \cline{1-2}\cline{4-6}
+    MARS & $0.0$ &  & $37.9$ & $2.53 \times 10^{-3}$ & $1.48$\\
+    SVR & $0.1$ &  & $0.859$ & $\mathit{1.81 \times 10^{-4}}$ & $\mathit{0.106}$\\
+    MLP & $\mathit{32.0}$ &  & $348.0$ & $1.11 \times 10^{-3}$ & $0.653$\\
+    Delaunay & $0.0$ &  & $\mathbf{1.90 \times 10^{-4}}$ & $3.69$ & $2.17 \times 10^{3}$\\
+    ShepMod & $\mathbf{66.4}$ &  & $1.13$ & $1.82 \times 10^{-4}$ & $0.107$\\
+    LSHEP & $0.0$ &  & $2.39$ & $\mathbf{1.44 \times 10^{-4}}$ & $\mathbf{0.0845}$\\
+    Voronoi & $1.6$ &  & $\mathit{0.298}$ & $0.0274$ & $16.1$\\
+    BoxSpline & $0.0$ &  & $1.26$ & $6.43 \times 10^{-4}$ & $0.377$\\
+    \cline{1-2}\cline{4-6}
   \end{tabular}
@@
   \centering
-  \begin{tabular}{c|r@{.}l|r@{.}l|r@{.}l|r@{.}l|r@{.}l}
-    \hline
-    Algorithm & \multicolumn{2}{c|}{Min} & \multicolumn{2}{c|}{$25^{th}$} & \multicolumn{2}{c|}{$50^{th}$} & \multicolumn{2}{c|}{$75^{th}$} & \multicolumn{2}{c}{Max}\\
-    \hline
-    MARS & $\mathit{6}$&$\mathit{45 \times 10^{-15}}$ & $\mathbf{2}$&$\mathbf{70 \times 10^{-14}}$ & $1$&$66$ & $1$&$96$ & $\mathit{53}$&$\mathit{3}$\\
-    SVR & $0$&$0420$ & $0$&$143$ & $0$&$148$ & $\mathit{0}$&$\mathit{860}$ & $119$&$0$\\
-    MLP & $0$&$0000689$ & $0$&$0337$ & $\mathbf{0}$&$\mathbf{0795}$ & $\mathbf{0}$&$\mathbf{264}$ & $\mathbf{5}$&$\mathbf{31}$\\
-    Delaunay & $\mathbf{0}$&$\mathbf{00}$ & $0$&$187$ & $0$&$919$ & $2$&$72$ & $56$&$3$\\
-    ShepMod & $\mathbf{0}$&$\mathbf{00}$ & $0$&$0957$ & $0$&$685$ & $2$&$90$ & $73$&$2$\\
-    LSHEP & $\mathbf{0}$&$\mathbf{00}$ & $\mathit{0}$&$\mathit{0153}$ & $\mathit{0}$&$\mathit{106}$ & $1$&$17$ & $113$&$0$\\
-    Voronoi & $\mathbf{0}$&$\mathbf{00}$ & $0$&$430$ & $1$&$28$ & $2$&$94$ & $83$&$8$\\
-    BoxSpline & $\mathbf{0}$&$\mathbf{00}$ & $0$&$789$ & $2$&$06$ & $5$&$25$ & $119$&$0$\\
+  \begin{tabular}{c|c|c|c|c|c}
+    \hline
+    Algorithm & Min & $25^{th}$ & $50^{th}$ & $75^{th}$ & Max\\
+    \hline
+    MARS & $\mathit{6.45 \times 10^{-15}}$ & $\mathbf{2.70 \times 10^{-14}}$ & $1.66$ & $1.96$ & $\mathit{53.3}$\\
+    SVR & $0.0420$ & $0.143$ & $0.148$ & $\mathit{0.860}$ & $119.0$\\
+    MLP & $6.89 \times 10^{-5}$ & $0.0337$ & $\mathbf{0.0795}$ & $\mathbf{0.264}$ & $\mathbf{5.31}$\\
+    Delaunay & $\mathbf{0.00}$ & $0.187$ & $0.919$ & $2.72$ & $56.3$\\
+    ShepMod & $\mathbf{0.00}$ & $0.0957$ & $0.685$ & $2.90$ & $73.2$\\
+    LSHEP & $\mathbf{0.00}$ & $\mathit{0.0153}$ & $\mathit{0.106}$ & $1.17$ & $113.0$\\
+    Voronoi & $\mathbf{0.00}$ & $0.430$ & $1.28$ & $2.94$ & $83.8$\\
+    BoxSpline & $\mathbf{0.00}$ & $0.789$ & $2.06$ & $5.25$ & $119.0$\\
     \hline
@@
   \centering
-  \begin{tabular}{|c|r@{.}l| c |r@{.}l|r@{.}l|r@{.}l|}
-    \cline{1-3}\cline{5-10}
-    Algorithm & \multicolumn{2}{c|}{\% Best} &  & \multicolumn{2}{c|}{Fit/Prep. Time (s)} & \multicolumn{2}{c|}{App. Time (s)} & \multicolumn{2}{c|}{Total App. Time (s)}\\
-    \cline{1-3}\cline{5-10}
-    MARS & \,\,\,\,$10$&$7$ &  & \quad\quad\,\,$0$&$151$ & \quad$0$&$000117$ & \quad\quad\quad\,\,$0$&$0304$\\
-    SVR & $4$&$1$ &  & $0$&$133$ & $\mathbf{0}$&$\mathbf{0000947}$ & $\mathbf{0}$&$\mathbf{0246}$\\
-    MLP & $\mathbf{56}$&$\mathbf{8}$ &  & $169$&$0$ & $0$&$00137$ & $0$&$356$\\
-    Delaunay & $0$&$1$ &  & $\mathbf{0}$&$\mathbf{000116}$ & $1$&$55$ & $403$&$0$\\
-    ShepMod & $3$&$5$ &  & $0$&$265$ & $0$&$000128$ & $0$&$0333$\\
-    LSHEP & $\mathit{28}$&$\mathit{4}$ &  & $0$&$874$ & $\mathit{0}$&$\mathit{0000975}$ & $\mathit{0}$&$\mathit{0254}$\\
-    Voronoi & $0$&$2$ &  & $\mathit{0}$&$\mathit{0112}$ & $0$&$0270$ & $7$&$01$\\
-    BoxSpline & $0$&$3$ &  & $0$&$330$ & $0$&$000406$ & $0$&$106$\\
-    \cline{1-3}\cline{5-10}
+  \begin{tabular}{|c|c| c |c|c|c|}
+    \cline{1-2}\cline{4-6}
+    Algorithm & \% Best &  & Fit / Prep. Time (s) & App. Time (s) & Total App. Time (s)\\
+    \cline{1-2}\cline{4-6}
+    MARS & $10.7$ &  & $0.151$ & $1.17 \times 10^{-4}$ & $0.0304$\\
+    SVR & $4.1$ &  & $0.133$ & $\mathbf{9.47 \times 10^{-5}}$ & $\mathbf{0.0246}$\\
+    MLP & $\mathbf{56.8}$ &  & $169.0$ & $1.37 \times 10^{-3}$ & $0.356$\\
+    Delaunay & $0.1$ &  & $\mathbf{1.16 \times 10^{-4}}$ & $1.55$ & $403.0$\\
+    ShepMod & $3.5$ &  & $0.265$ & $1.28 \times 10^{-4}$ & $0.0333$\\
+    LSHEP & $\mathit{28.4}$ &  & $0.874$ & $\mathit{9.75 \times 10^{-5}}$ & $\mathit{0.0254}$\\
+    Voronoi & $0.2$ &  & $\mathit{0.0112}$ & $0.0270$ & $7.01$\\
+    BoxSpline & $0.3$ &  & $0.330$ & $4.06 \times 10^{-4}$ & $0.106$\\
+    \cline{1-2}\cline{4-6}
   \end{tabular}
-  \caption{Left table shows how often each algorithm had the lowest
-    absolute error approximating Sydney rainfall data in Table
+  \caption{The left above shows how often each algorithm had the
+    lowest absolute error approximating Sydney rainfall data in Table
     \ref{table:error-weather}. On the right columns are median fit
@@
   \centering
-  \begin{tabular}{c|r@{.}l|r@{.}l|r@{.}l|r@{.}l|r@{.}l}
-    \hline
-    Algorithm & \multicolumn{2}{c|}{Min} & \multicolumn{2}{c|}{$25^{th}$} & \multicolumn{2}{c|}{$50^{th}$} & \multicolumn{2}{c|}{$75^{th}$} & \multicolumn{2}{c}{Max}\\
-    \hline
-    MARS & $5$&$36$ & $3610$&$0$ & $4580$&$0$ & $5450$&$0$ & $13400$&$0$\\
-    SVR & $0$&$0138$ & $8$&$47$ & $15$&$4$ & $41$&$9$ & $7700$&$0$\\
-    MLP & $0$&$00151$ & $\mathbf{1}$&$\mathbf{60}$ & $\mathbf{3}$&$\mathbf{86}$ & $\mathbf{8}$&$\mathbf{34}$ & $\mathbf{604}$&$\mathbf{0}$\\
-    Delaunay & $\mathbf{0}$&$\mathbf{00}$ & $\mathit{2}$&$\mathit{69}$ & $\mathit{13}$&$\mathit{8}$ & $\mathit{35}$&$\mathit{0}$ & $\mathit{4840}$&$\mathit{0}$\\
-    ShepMod & $\mathbf{0}$&$\mathbf{00}$ & $4$&$21$ & $17$&$3$ & $51$&$6$ & $6510$&$0$\\
-    LSHEP & $1$&$27$ & $199$&$0$ & $260$&$0$ & $343$&$0$ & $6530$&$0$\\
-    Voronoi & $\mathit{2}$&$\mathit{89 \times 10^{-10}}$ & $14$&$7$ & $32$&$8$ & $52$&$9$ & $4860$&$0$\\
-    BoxSpline & $0$&$00740$ & $20$&$8$ & $40$&$9$ & $79$&$6$ & $7640$&$0$\\
+  \begin{tabular}{c|c|c|c|c|c}
+    \hline
+    Algorithm & Min & $25^{th}$ & $50^{th}$ & $75^{th}$ & Max\\
+    \hline
+    MARS & $5.36$ & $3610.0$ & $4580.0$ & $5450.0$ & $13400.0$\\
+    SVR & $0.0138$ & $8.47$ & $15.4$ & $41.9$ & $7700.0$\\
+    MLP & $0.00151$ & $\mathbf{1.60}$ & $\mathbf{3.86}$ & $\mathbf{8.34}$ & $\mathbf{604.0}$\\
+    Delaunay & $\mathbf{0.00}$ & $\mathit{2.69}$ & $\mathit{13.8}$ & $\mathit{35.0}$ & $\mathit{4840.0}$\\
+    ShepMod & $\mathbf{0.00}$ & $4.21$ & $17.3$ & $51.6$ & $6510.0$\\
+    LSHEP & $1.27$ & $199.0$ & $260.0$ & $343.0$ & $6530.0$\\
+    Voronoi & $\mathit{2.89 \times 10^{-10}}$ & $14.7$ & $32.8$ & $52.9$ & $4860.0$\\
+    BoxSpline & $0.00740$ & $20.8$ & $40.9$ & $79.6$ & $7640.0$\\
     \hline
@@
   \centering
-  \begin{tabular}{|c|r@{.}l| c |r@{.}l|r@{.}l|r@{.}l|}
-    \cline{1-3}\cline{5-10}
-    Algorithm & \multicolumn{2}{c|}{\% Best} &  & \multicolumn{2}{c|}{Fit/Prep. Time (s)} & \multicolumn{2}{c|}{App. Time (s)} & \multicolumn{2}{c|}{Total App. Time (s)}\\
-    \cline{1-3}\cline{5-10}
-    MARS & \quad\,$0$&$0$ &  & \quad\quad\,\,$22$&$0$ & \quad$0$&$00148$ & \quad\quad\quad$0$&$820$\\
-    SVR & $0$&$0$ &  & $1$&$01$ & $\mathit{0}$&$\mathit{000210}$ & $\mathit{0}$&$\mathit{117}$\\
-    MLP & $\mathbf{79}$&$\mathbf{5}$ &  & $290$&$0$ & $0$&$000714$ & $0$&$397$\\
-    Delaunay & $\mathit{20}$&$\mathit{5}$ &  & $\mathbf{0}$&$\mathbf{000243}$ & $6$&$83$ & $3800$&$$\\
-    ShepMod & $0$&$1$ &  & $1$&$45$ & $0$&$000220$ & $0$&$122$\\
-    LSHEP & $0$&$0$ &  & $3$&$47$ & $\mathbf{0}$&$\mathbf{000176}$ & $\mathbf{0}$&$\mathbf{0981}$\\
-    Voronoi & $0$&$0$ &  & $\mathit{0}$&$\mathit{197}$ & $0$&$0950$ & $52$&$8$\\
-    BoxSpline & $0$&$0$ &  & $1$&$66$ & $0$&$000956$ & $0$&$532$\\
-    \cline{1-3}\cline{5-10}
+  \begin{tabular}{|c|c| c |c|c|c|}
+    \cline{1-2}\cline{4-6}
+    Algorithm & \% Best &  & Fit / Prep. Time (s) & App. Time (s) & Total App. Time (s)\\
+    \cline{1-2}\cline{4-6}
+    MARS & $0.0$ &  & $22.0$ & $1.48 \times 10^{-3}$ & $0.820$\\
+    SVR & $0.0$ &  & $1.01$ & $\mathit{2.10 \times 10^{-4}}$ & $\mathit{0.117}$\\
+    MLP & $\mathbf{79.5}$ &  & $290.0$ & $7.14 \times 10^{-4}$ & $0.397$\\
+    Delaunay & $\mathit{20.5}$ &  & $\mathbf{2.43 \times 10^{-4}}$ & $6.83$ & $3.80 \times 10^{3}$\\
+    ShepMod & $0.1$ &  & $1.45$ & $2.20 \times 10^{-4}$ & $0.122$\\
+    LSHEP & $0.0$ &  & $3.47$ & $\mathbf{1.76 \times 10^{-4}}$ & $\mathbf{0.0981}$\\
+    Voronoi & $0.0$ &  & $\mathit{0.197}$ & $0.0950$ & $52.8$\\
+    BoxSpline & $0.0$ &  & $1.66$ & $9.56 \times 10^{-4}$ & $0.532$\\
+    \cline{1-2}\cline{4-6}
   \end{tabular}
@@
   \centering
-  \begin{tabular}{c|r@{.}l|r@{.}l|r@{.}l|r@{.}l|r@{.}l}
-    \hline
-    Algorithm & \multicolumn{2}{c|}{Min} & \multicolumn{2}{c|}{$25^{th}$} & \multicolumn{2}{c|}{$50^{th}$} & \multicolumn{2}{c|}{$75^{th}$} & \multicolumn{2}{c}{Max}\\
-    \hline
-    Delaunay & $\mathbf{0}$&$\mathbf{0287}$ & $\mathbf{0}$&$\mathbf{0914}$ & $\mathbf{0}$&$\mathbf{158}$ & $\mathit{0}$&$\mathit{308}$ & $\mathit{0}$&$\mathit{897}$\\
-    ShepMod & $0$&$0307$ & $\mathit{0}$&$\mathit{0983}$ & $\mathit{0}$&$\mathit{164}$ & $0$&$335$ & $1$&$00$\\
-    Voronoi & $\mathit{0}$&$\mathit{0303}$ & $0$&$100$ & $0$&$165$ & $\mathbf{0}$&$\mathbf{274}$ & $\mathbf{0}$&$\mathbf{762}$\\
-    BoxSpline & $0$&$0703$ & $0$&$417$ & $0$&$561$ & $0$&$745$ & $1$&$00$\\
+  \begin{tabular}{c|c|c|c|c|c}
+    \hline
+    Algorithm & Min & $25^{th}$ & $50^{th}$ & $75^{th}$ & Max\\
+    \hline
+    Delaunay & $\mathbf{0.0287}$ & $\mathbf{0.0914}$ & $\mathbf{0.158}$ & $\mathit{0.308}$ & $\mathit{0.897}$\\
+    ShepMod & $0.0307$ & $\mathit{0.0983}$ & $\mathit{0.164}$ & $0.335$ & $1.00$\\
+    Voronoi & $\mathit{0.0303}$ & $0.100$ & $0.165$ & $\mathbf{0.274}$ & $\mathbf{0.762}$\\
+    BoxSpline & $0.0703$ & $0.417$ & $0.561$ & $0.745$ & $1.00$\\
     \hline
@@
   \centering
-  \begin{tabular}{|c|r@{.}l| c |r@{.}l|r@{.}l|r@{.}l|}
-    \cline{1-3}\cline{5-10}
-    Algorithm & \multicolumn{2}{c|}{\% Best} &  & \multicolumn{2}{c|}{Fit/Prep. Time (s)} & \multicolumn{2}{c|}{App. Time (s)} & \multicolumn{2}{c|}{Total App. Time (s)}\\
-    \cline{1-3}\cline{5-10}
-    Delaunay & \,\,$\mathbf{62}$&$\mathbf{0}$ &  & \quad\quad$\mathbf{0}$&$\mathbf{000517}$ & \quad$0$&$353$ & \quad\quad\,\,$106$&$0$\\
-    ShepMod & $0$&$0$ &  & $0$&$0884$ & $\mathbf{0}$&$\mathbf{000145}$ & $\mathbf{0}$&$\mathbf{0436}$\\
-    Voronoi & $\mathit{38}$&$\mathit{0}$ &  & $\mathit{0}$&$\mathit{0173}$ & $0$&$00253$ & $0$&$762$\\
-    BoxSpline & $0$&$0$ &  & $0$&$0972$ & $\mathit{0}$&$\mathit{000210}$ & $\mathit{0}$&$\mathit{0633}$\\
-    \cline{1-3}\cline{5-10}
+  \begin{tabular}{|c|c| c |c|c|c|}
+    \cline{1-2}\cline{4-6}
+    Algorithm & \% Best &  & Fit / Prep. Time (s) & App. Time (s) & Total App. Time (s)\\
+    \cline{1-2}\cline{4-6}
+    Delaunay & $\mathbf{62.0}$ &  & $\mathbf{5.17 \times 10^{-4}}$ & $0.353$ & $106.0$\\
+    ShepMod & $0.0$ &  & $0.0884$ & $\mathbf{1.45 \times 10^{-4}}$ & $\mathbf{0.0436}$\\
+    Voronoi & $\mathit{38.0}$ &  & $\mathit{0.0173}$ & $2.53 \times 10^{-3}$ & $0.762$\\
+    BoxSpline & $0.0$ &  & $0.0972$ & $\mathit{2.10 \times 10^{-4}}$ & $\mathit{0.0633}$\\
+    \cline{1-2}\cline{4-6}
   \end{tabular}

### 2019-03-20 Paper_[2019-02]_Numerical_Algorithms/tchlux-NA-2019.tex
created an affine weights visual. finished NA paper draft. wrote report on Xiaodong clock cycle data.

@@
 \ref{sec:regression} and \ref{sec:interpolation} present the
-multivariate models. Section \ref{sec:error} presents the error
-measuring methodology that is used for collecting empirical
-results. Section \ref{sec:theory} presents the theoretical error
-bounds for interpolation. Section \ref{sec:data} presents the data
-sets for empirical approximation analysis and presents the
-approximation results for all models on these data sets. Section
-\ref{sec:discussion} analyzes and discusses the results, their
-implications, and their limitations. Finally, Section
-\ref{sec:conclusion} concludes.
+multivariate models. Section \ref{sec:error} gives the error measuring
+methodology that is used for collecting empirical results. Section
+\ref{sec:theory} presents the theoretical error bounds for
+interpolation. Section \ref{sec:data} presents the data sets for
+empirical approximation analysis and the approximation results for all
+models on these data sets. Section \ref{sec:discussion} analyzes and
+discusses the results, their implications, and their
+limitations. Finally, Section \ref{sec:conclusion} concludes.
 
@@
 This section presents the theoretical results bounding the error of
-Delaunay interpolation. The error analysis relies on the Delaunay
-interpolation for two reasons: (1) second order results can be
+simplicial interpolation. The error analysis relies on simplicial
+interpolation for three reasons: (1) second order results can be
 obtained utilizing a Lipschitz constant on the gradient of a function,
-rather than standard Lipschitz bounds and (2) multiple other
-interpolants analyzed compute predictions as convex combinations of
-observed function values, which may allow for straight forward
-extensions of this error bound.
+rather than standard Lipschitz bounds; (2) the results directly
+apply to Delaunay; and (3) multiple other interpolants in this paper
+compute predictions as convex combinations of observed function
+values, which may allow for straight forward extensions of this error
+bound.
 
@@
 
-In summary, the approximation error of a Delaunay interpolant tends
+In summary, the approximation error of a simplicial interpolant tends
 quadratically towards zero when approaching observed data only when
-the diameter of the interpolating simplex is also reduced at a
-proportional rate. Without the incorporation of additional
-observations, only linear convergence to the true function is be
-achieved. Notice that the approximation error is largely determined by
-the spacing of observed data. Predictions made by simplices whose
-vertices are not well-spaced (i.e. have large diameter, or are nearly
-contained in a hyperplane) have higher error. In light of this error
-bound, an empirical evaluation of presented algorithms follows.
+the diameter of the simplex is also reduced at a proportional rate.
+Only linear convergence to the true function can be achieved in
+practice, without the incorporation of additional observations. Notice
+that the approximation error is largely determined by the spacing of
+observed data. Predictions made by simplices whose vertices are not
+well-spaced (i.e. have large diameter, or are nearly contained in a
+hyperplane) have higher error. In light of this error bound, an
+empirical evaluation of presented algorithms follows.
 
@@
 The theoretical results constructed in Section \ref{sec:theory} for
-Delaunay interpolation are promising, however they are difficult to
-interpret in context with the other approximation algorithms that do
-not have similar known uniform error bounds. For that reason, this
-paper utilizes five different data sets of varying dimension and
-application to construct approximations and compare the accuracy of
-different techniques.
+simplicial interpolation are promising, however they are difficult to
+interpret in context with approximation algorithms that do not have
+similar known uniform error bounds. For that reason, this paper
+utilizes five different data sets of varying dimension and application
+to construct approximations and compare the accuracy of different
+techniques.
 
@@
     log scale. The complementing numerical data to this figure is
-    provided in table \ref{table:error-forest-fire} in the Appendix.}
+    provided in Table \ref{table:error-forest-fire} in the Appendix.}
   \label{fig:error-forest-fire}
@@
 
-As suggested by figure \ref{fig:error-forest-fire}, the SVR has the
+As suggested by Figure \ref{fig:error-forest-fire}, the SVR has the
 lowest absolute prediction errors for 80\% of the data, with ShepMod
@@
     each model. The errors are presented on a natural log scale,
-    mimicking the presentation in figure \ref{fig:hist-weather}. The
+    mimicking the presentation in Figure \ref{fig:hist-weather}. The
     complementing numerical data to this figure is provided in table
@@
 
-While figure \ref{fig:hist-weather} makes MARS look far better than
+While Figure \ref{fig:hist-weather} makes MARS look far better than
 other techniques, it only provides the best prediction for 11\% of
@@
     \ref{fig:hist-credit-card}. The complementing numerical data to
-    this figure is provided in table \ref{table:error-credit-card} in
+    this figure is provided in Table \ref{table:error-credit-card} in
     the Appendix.}
@@
 
-As can be seen in figure \ref{fig:error-credit-card}, the MLP
+As can be seen in Figure \ref{fig:error-credit-card}, the MLP
 outperforms all other algorithms at the first, second, third, and
@@
     configuration and the predicted distribution. The above figure is
-    \textit{not} log-scaled like figure \ref{fig:hist-throughput}. The
+    \textit{not} log-scaled like Figure \ref{fig:hist-throughput}. The
     complimentary numerical data for this figure is provided in
-    Appendix table \ref{table:error-throughput}.}
+    Appendix Table \ref{table:error-throughput}.}
   \label{fig:error-throughput}
@@
   \caption{This is the numerical counterpart to the histogram data
-    presented in figure \ref{fig:throughput-prediction}. The columns
+    presented in Figure \ref{fig:throughput-prediction}. The columns
     display the percent of null hypothesis rejections by the KS-test
@@
 Figure \ref{fig:throughput-prediction} expands on the KS statistic
-results presented in figure \ref{fig:error-throughput}. Agglomerate
+results presented in Figure \ref{fig:error-throughput}. Agglomerate
 errors for each algorithm resemble a Gamma distribution. The
@@
     error approximation, the average time-to-fit, and the average time
-    required to approximated one point. Interpolants provide the
+    required to approximate one point. Interpolants provide the
     lowest error approximation one third of the time across all data,
@@
 The various interpolants discussed in this paper have been
-demonstrated to effectively approximate multivariate phenomena up to
-$30$ dimensions. The underlying constructions are theoretically
+demonstrated to effectively approximate multivariate phenomena in up
+to $30$ dimensions. The underlying constructions are theoretically
 straightforward, yet powerful and flexible. Most of the interpolants'
 computational complexities make them particularly suitable for
-applications in even higher dimension, where the major benefits are
-seen when only a small number of approximations are to be made from
-any given data ($\leq 1000$).
+applications in even higher dimension. The major benefits of
+interpolation are seen when only a small number of approximations
+($\leq 1000$) are made from data and when there are relatively few
+data points for the dimension (for empirical results presented, $\log_d n \leq 5$).
 
@@
 \bibliography{paper}
+
+\newpage
 
@@
   \caption{This numerical data accompanies the visual provided in
-    figure \ref{fig:error-forest-fire}. The columns of absolute error
+    Figure \ref{fig:error-forest-fire}. The columns of absolute error
     percentiles correspond to the minimum, first quartile, median,
@@
   \caption{This numerical data accompanies the visual provided in
-    figure \ref{fig:error-parkinsons}. The columns of absolute error
+    Figure \ref{fig:error-parkinsons}. The columns of absolute error
     percentiles correspond to the minimum, first quartile, median,
@@
   \caption{This numerical data accompanies the visual provided in
-    figure \ref{fig:error-weather}. The columns of absolute error
+    Figure \ref{fig:error-weather}. The columns of absolute error
     percentiles correspond to the minimum, first quartile, median,
@@
   \caption{This numerical data accompanies the visual provided in
-    figure \ref{fig:error-credit-card}. The columns of absolute error
+    Figure \ref{fig:error-credit-card}. The columns of absolute error
     percentiles correspond to the minimum, first quartile, median,
@@
     lowest absolute error approximating credit card transaction data
-    in table \ref{table:error-credit-card}. On the right columns are
+    in Table \ref{table:error-credit-card}. On the right columns are
     median fit time of 5006 points, median time for one approximation,
@@
   \caption{This numerical data accompanies the visual provided in
-    figure \ref{fig:error-throughput}. The columns of absolute error
+    Figure \ref{fig:error-throughput}. The columns of absolute error
     percentiles correspond to the minimum, first quartile, median,
@@
     lowest KS statistic on the I/O throughput distribution data in
-    table \ref{table:error-throughput}. On the right columns are
+    Table \ref{table:error-throughput}. On the right columns are
     median fit time of 2715 points, median time for one approximation,

### 2019-03-20 Documents_[2019]/Miscellaneous/affine_weights.py

@@ -1,0 +1,59 @@
+
+import numpy as np
+from util.misc.grid import grid as grid_func
+from util.plot import Plot
+from util.math import transpose
+
+np.random.seed(0)
+# x = np.random.random((3,2))
+
+names = ["equilateral", "right", "skinny"]
+xs = [
+    np.array([
+        [.45,.7],
+        [.6, .3],
+        [.3,.3],
+    ]),
+    np.array([
+        [.45, .7],
+        [.45,.3],
+        [.3,.3],
+    ]),
+    np.array([
+        [.45, .7],
+        [.45,.3],
+        [.4,.5],
+    ])
+]
+
+for (name, x) in zip(names, xs):
+
+    space = .2
+    grid = grid_func(np.min(x)-space, np.max(x)+space, dim=2, num=10000)
+    # grid += np.random.random(grid.shape) * .05
+    # grid = np.random.random((100,2))
+    ones = np.ones((x.shape[0],1))
+    labels = []
+
+    # print("x.shape: ",x.shape)
+    # print("ones:    \n",ones)
+
+    for pt in grid:
+        system = np.concatenate((x, ones), axis=1).T
+        x_pt = np.concatenate((pt,[1]))
+        # print("system:  ",system)
+        # print("x_pt:    ",x_pt)
+        weights = np.linalg.solve(system, x_pt)
+        # print("weights: ",weights, sum(weights))
+        labels.append( "  ".join([f"{w:.2f}" for w in weights]) )
+        # print(labels[-1])
+        # exit()
+
+    p = Plot(f"Affine weights for {name} triangle")
+    p.add("grid", *(grid.T), text=labels, color=p.color(color=(255,255,255)),
+          opacity=0, hoverinfo="text")
+    p.add("Points", *transpose(list(x) + [x[0]]), marker_size=20,
+          mode="markers+lines+text", text=["1", "2", "3", ""],
+          line_color="rgba(0,0,0,.5)", shade=False)
+    p.graph(hovermode="closest", file_name="affine_weights.html",
+            show=(name==names[-1]), append=(name!=names[0]))

### 2019-03-20 Report_[2019-02-03]_Sample_Size_Determination/CI_function_verify.R

@@ -1,0 +1,254 @@
+MixtureNormInfoM_2comp_verify=function(x,mean,sd,lambda)
+{
+  normDu=function(mean,sd,x)
+  {
+    z=(x-mean)/sd
+    f=dnorm(x,mean,sd)
+    return(z*f/sd)
+  }
+  normDuDu=function(mean,sd,x)
+  {
+    z=(x-mean)/sd
+    f=dnorm(x,mean,sd)
+    f*(z/sd)^2-1/sd^2*f
+  }
+  normDsigma=function(mean,sd,x)
+  {
+    z=(x-mean)/sd
+    f=dnorm(x,mean,sd)
+    -1/sd*f+f*z^2/sd
+  }
+  normDsigmaDsigma=function(mean,sd,x)
+  {
+    z=(x-mean)/sd
+    f=dnorm(x,mean,sd)
+    1/sd^2*f-normDsigma(mean,sd,x)/sd+normDsigma(mean,sd,x)*z^2/sd-3*f*z^2/sd^2
+  }
+  normDuDsiama=function(mean,sd,x)
+  {
+    z=(x-mean)/sd
+    f=dnorm(x,mean,sd)
+    -1/sd*normDu(mean,sd,x)+z^2/sd*normDu(mean,sd,x)-2*f*z/sd^2
+  }
+  
+  #calculat the likelihood
+  normMix_2comp_lik=function(mean,sd,lambda,x)
+  {
+    val=lambda*dnorm(x,mean = mean[1],sd = sd[1])+
+      (1-lambda)*dnorm(x,mean = mean[2],sd = sd[2])
+    val
+  }
+  
+  h=matrix(ncol = 5, nrow = 5)
+  f1=dnorm(x,mean[1],sd[1])
+  f2=dnorm(x,mean[2],sd[2])
+  Likeli=f1*lambda+(1-lambda)*f2
+  #LogLikeli=log(Likeli)
+  
+  Ivec=rep(0,5)
+  Ivec[1]=1/Likeli*(f1-f2)
+  Ivec[2]=1/Likeli*lambda*normDu(mean[1],sd[1],x)
+  Ivec[3]=1/Likeli*lambda*normDsigma(mean[1],sd[1],x)
+  Ivec[4]=1/Likeli*(1-lambda)*normDu(mean[2],sd[2],x)
[…]

### 2019-03-20 Report_[2019-03-19]_Mean_Clock_Cycle_Prediction/code.py

@@ -1,0 +1,51 @@
+from util.data import Data
+
+print("Loading original data..")
+d = Data.load("2019-03-19_Xiaodong-data.csv")
+# Collect by unique configurations.
+print("Collecting data by four parameters..")
+d = d[:,:-1].unique().collect(d)
+d["Samples"] = map(len, d["Clock Cycles"])
+d["Mean Clock Cycles"] = map(lambda l: sum(l) / len(l), d["Clock Cycles"])
+d.reorder(d.names[:-3] + ["Mean Clock Cycles", "Samples"])
+d["Index"] = range(len(d))
+d["Predicted Mean Clock Cycles"] = (0. for row in d)
+# Remove the unused columns.
+d.pop("Samples")
+d.pop("Clock Cycles")
+# Print the data for verification.
+print(d)
+print()
+print("Making predictions..")
+
+
+from util.approximate import Delaunay as Algorithm
+import numpy as np
+for i,(train, test) in enumerate(d.k_fold(k=len(d))):
+    print(f" fold {i+1}..", train.shape, test.shape)
+    matrix = train[:,:4].to_matrix()
+    train_x = matrix.data
+    train_y = np.array(list(train["Mean Clock Cycles"]))
+    # Assume leave-one-out prediction.
+    test_x = np.array(test[0][:4], dtype=float)
+    # normalize the train and test by the train
+    train_x -= matrix.shift
+    train_x /= matrix.scale
+    test_x -= matrix.shift
+    test_x /= matrix.scale
+    model = Algorithm()
+    model.fit(train_x, train_y)
+    # Store the output predictions.
+    d[test[0,"Index"], "Predicted Mean Clock Cycles"] = float(model.predict(test_x))
+
+# Remove the additional index column that was added.
+d.pop("Index")
+# Compute the prediction errors.
+d["Error"] = (g-t for (t,g) in zip(d["Mean Clock Cycles"], d["Predicted Mean Clock Cycles"]))
+d["Absolute Error"] = (abs(g-t) for (t,g) in zip(d["Mean Clock Cycles"], d["Predicted Mean Clock Cycles"]))
+print()
+print(d)
+
+d.summarize()
+
+d.save("predictions.csv")

### 2019-03-20 Paper_[2019-02]_Numerical_Algorithms/paper.tex
Added code to time Delaunay for fit time fix. Updated some report.

@@
 burn area are the $x$ and $y$ spatial coordinates of burns in the
-park, month of year (mapped to a 2D circle), the FFMC, DMC, DC, and
-ISI indices (see source for details), the temperature, relative
-humidity, wind speed, and outdoor rain. The original analysis of this
-data set demonstrated it to be difficult to model, likely due to the
-skew in response values.
+park, month of year (mapped to $x$, $y$ coordinates on a unit circle),
+the FFMC, DMC, DC, and ISI indices (see source for details), the
+temperature, relative humidity, wind speed, and outdoor rain. The
+original analysis of this data set demonstrated it to be difficult to
+model, likely due to the skew in response values.
 
@@
     throughput than most others. The data is presented on a $\ln$
-    scale, however the resulting left-skew reveals the data is not
-    precisely distributed according to an exponential.}
+    scale.}
   \label{fig:hist-throughput}
@@
     frequency with which any algorithm provided the lowest absolute
-    error approximation, the average time to fit / prepare, and the
+    error approximation, the average time to fit/prepare, and the
     average time required to approximate one point. Interpolants
@@
     Delaunay & $\mathit{3}$&$\mathit{72 \times 10^{-12}}$ & $1$&$20$ & $3$&$50$ & $7$&$67$ & $30$&$7$\\
-    ShepMod & $\mathbf{0}$&$\mathbf{00}$ & $\mathbf{0}$&$\mathbf{255}$ & $\mathbf{0}$&$\mathbf{908}$ & $\mathit{3}$&$\mathit{43}$ & $34$&$5$\\
+    ShepMod & $\mathbf{0}$&$\mathbf{0}$ & $\mathbf{0}$&$\mathbf{255}$ & $\mathbf{0}$&$\mathbf{908}$ & $\mathit{3}$&$\mathit{43}$ & $34$&$5$\\
     LSHEP & $0$&$00254$ & $2$&$93$ & $7$&$16$ & $13$&$1$ & $\mathit{29}$&$\mathit{0}$\\
-    Voronoi & $\mathbf{0}$&$\mathbf{00}$ & $1$&$29$ & $3$&$52$ & $6$&$87$ & $30$&$1$\\
+    Voronoi & $\mathbf{0}$&$\mathbf{0}$ & $1$&$29$ & $3$&$52$ & $6$&$87$ & $30$&$1$\\
     BoxSpline & $0$&$00287$ & $4$&$39$ & $9$&$10$ & $14$&$8$ & $41$&$3$\\
@@
     MLP & $\mathit{32}$&$\mathit{0}$ &  & $348$&$0$ & $0$&$00111$ & $0$&$653$\\
-    Delaunay & $0$&$0$ &  & $\mathbf{0}$&$\mathbf{000190}$ & $3$&$69$ & $2170$&\\
+    Delaunay & $0$&$0$ &  & $\mathbf{0}$&$\mathbf{000190}$ & $3$&$69$ & $2170$&$0$\\
     ShepMod & $\mathbf{66}$&$\mathbf{4}$ &  & $1$&$13$ & $0$&$000182$ & $0$&$107$\\
@@
     MLP & $0$&$0000689$ & $0$&$0337$ & $\mathbf{0}$&$\mathbf{0795}$ & $\mathbf{0}$&$\mathbf{264}$ & $\mathbf{5}$&$\mathbf{31}$\\
-    Delaunay & $\mathbf{0}$&$\mathbf{00}$ & $0$&$187$ & $0$&$919$ & $2$&$72$ & $56$&$3$\\
-    ShepMod & $\mathbf{0}$&$\mathbf{00}$ & $0$&$0957$ & $0$&$685$ & $2$&$90$ & $73$&$2$\\
-    LSHEP & $\mathbf{0}$&$\mathbf{00}$ & $\mathit{0}$&$\mathit{0153}$ & $\mathit{0}$&$\mathit{106}$ & $1$&$17$ & $113$&$0$\\
-    Voronoi & $\mathbf{0}$&$\mathbf{00}$ & $0$&$430$ & $1$&$28$ & $2$&$94$ & $83$&$8$\\
-    BoxSpline & $\mathbf{0}$&$\mathbf{00}$ & $0$&$789$ & $2$&$06$ & $5$&$25$ & $119$&$0$\\
+    Delaunay & $\mathbf{0}$&$\mathbf{0}$ & $0$&$187$ & $0$&$919$ & $2$&$72$ & $56$&$3$\\
+    ShepMod & $\mathbf{0}$&$\mathbf{0}$ & $0$&$0957$ & $0$&$685$ & $2$&$90$ & $73$&$2$\\
+    LSHEP & $\mathbf{0}$&$\mathbf{0}$ & $\mathit{0}$&$\mathit{0153}$ & $\mathit{0}$&$\mathit{106}$ & $1$&$17$ & $113$&$0$\\
+    Voronoi & $\mathbf{0}$&$\mathbf{0}$ & $0$&$430$ & $1$&$28$ & $2$&$94$ & $83$&$8$\\
+    BoxSpline & $\mathbf{0}$&$\mathbf{0}$ & $0$&$789$ & $2$&$06$ & $5$&$25$ & $119$&$0$\\
     \hline
@@
     MLP & $0$&$00151$ & $\mathbf{1}$&$\mathbf{60}$ & $\mathbf{3}$&$\mathbf{86}$ & $\mathbf{8}$&$\mathbf{34}$ & $\mathbf{604}$&$\mathbf{0}$\\
-    Delaunay & $\mathbf{0}$&$\mathbf{00}$ & $\mathit{2}$&$\mathit{69}$ & $\mathit{13}$&$\mathit{8}$ & $\mathit{35}$&$\mathit{0}$ & $\mathit{4840}$&$\mathit{0}$\\
-    ShepMod & $\mathbf{0}$&$\mathbf{00}$ & $4$&$21$ & $17$&$3$ & $51$&$6$ & $6510$&$0$\\
+    Delaunay & $\mathbf{0}$&$\mathbf{0}$ & $\mathit{2}$&$\mathit{69}$ & $\mathit{13}$&$\mathit{8}$ & $\mathit{35}$&$\mathit{0}$ & $\mathit{4840}$&$\mathit{0}$\\
+    ShepMod & $\mathbf{0}$&$\mathbf{0}$ & $4$&$21$ & $17$&$3$ & $51$&$6$ & $6510$&$0$\\
     LSHEP & $1$&$27$ & $199$&$0$ & $260$&$0$ & $343$&$0$ & $6530$&$0$\\
@@
     MLP & $\mathbf{79}$&$\mathbf{5}$ &  & $290$&$0$ & $0$&$000714$ & $0$&$397$\\
-    Delaunay & $\mathit{20}$&$\mathit{5}$ &  & $\mathbf{0}$&$\mathbf{000243}$ & $6$&$83$ & $3800$&$$\\
+    Delaunay & $\mathit{20}$&$\mathit{5}$ &  & $\mathbf{0}$&$\mathbf{000243}$ & $6$&$83$ & $3800$&$0$\\
     ShepMod & $0$&$1$ &  & $1$&$45$ & $0$&$000220$ & $0$&$122$\\

### 2019-03-20 Paper_[2019-02]_Numerical_Algorithms/Data/time_delaunay.py

@@ -1,0 +1,87 @@
+import os
+import numpy as np
+
+from util.data import Data
+from util.system import Timer
+from util.approximate import Delaunay, class_name
+
+# Make sure all of the prepared and pre-processed data files exist.
+from preprocess_data import cwd, raw_dir, data_dir
+
+to_test = 2
+
+
+# Overwrite the print function so that it forces prints to flush.
+def print(*args, og_print=print, **kwargs):
+    kwargs["flush"] = True
+    return og_print(*args, **kwargs)
+
+
+seed = 0
+folds = 10
+
+data_name = lambda p: os.path.basename(p).replace(".dill","").replace(".csv","")
+
+# Recording data for each set and for each algorithm that looks like:
+# | Data name | Dimension | Fold Size | Fold Number | Model Name | Model fit time |
+# | x | ... | y | ... | Model y | ... | Model nearest | Model max edge | Model furthest | Model num contributors | Model prediction time |
+
+for raw_file in sorted(os.listdir(raw_dir)):
+    raw_file = raw_file.replace(".gz","")
+    raw_path = os.path.join(raw_dir, raw_file)
+    data_path = os.path.join(data_dir, raw_file + '.dill')
+    intermediate_results_file = os.path.join(".", "Predictions",
+                    f"intermediate_results_{data_name(raw_file)}.dill")
+    final_results_file = os.path.join(".", "Predictions",
+                    f"results_{data_name(raw_file)}.dill")
+    print('-'*70)
+    print(data_name(raw_path))
+    print(raw_path)
+    print(final_results_file)
+    d = Data.load(data_path)
+    target = d.names[-1]
+    # Declare the indices..
+    d["Indices"] = range(len(d))
+    # Compute the number of training columns.
+    num_train_columns = d.names.index(target)
+    print(f"Data shape: {d.shape} predicting '{target}'")
+    # Start the folds.
+    for i,(train, test) in enumerate(d.k_fold(k=folds, seed=seed)):
+        # Get the numeric representation of this data based on training,
+        # ignore extra added columns and the "target" column.
+        num = train[:,:num_train_columns].to_matrix()
+        train_x = num.data
+        train_y = list(train[target])
+        # Generate the numeric testing set basesd on training.
+        test_x = np.array([num.to_real(row[:num_train_columns]) for row in test])
+        test_y = list(test[target])
+        # Normalize the training and testing data to the unit hypercube.
+        train_x = (train_x - num.shift) / num.scale
[…]

### 2019-03-30 Paper_[2019-02]_Numerical_Algorithms/paper.tex
Final modifications based on reviewer feedback.

@@
 cumulative distribution functions (CDFs) with the Kolmogorov-Smirnov
-(KS) statistic as metric.
+(KS) statistic \cite{lilliefors1967kolmogorov} as metric.
 
@@
 exponentially. Quantifying all possible interactions becomes
-intractable and hence beyond three-dimensional data, mostly linear
+intractable, and hence beyond three-dimensional data mostly linear
 models are used. That is not to say nonlinear models are absent, but
@@
 \cite{stanford1993fast}, called fast multivariate adaptive regression
-splines (Fast MARS).In Fast MARS, a least squares fit model is
+splines (Fast MARS). In Fast MARS, a least squares fit model is
 iteratively built by beginning with a single constant valued function
@@
 \rightarrow \mathbb{R}^{j}$ defined by
- $$ l(u) = \bigl( u^t W_l \bigr)_+ ,$$
-where $W_l$ is the $i \times j$ weight matrix for layer $l$. In this
-form, the multilayer perceptron (MLP) produces a piecewise linear
-model of the input data. The computational complexity of training a
-multilayer perceptron is $\mathcal{O}(n d m)$, where $m$ is determined
-by the sizes of the layers of the network and the stopping criterion
-of the minimization used for finding weights. This paper uses a MLP
-built from Keras and Tensorflow to perform regression
-\cite{chollet2015keras,tensorflow2015-whitepaper} with ten hidden
-layers each having thirty two nodes (a total of approximately ten
-thousand parameters), ReLU activation, and five thousand epochs of SGD
-for training.
+ $$ l(u) = \bigl( u^t W_l + c_l \bigr)_+ ,$$
+where $c_l \in \mathbb{R}^j$ and $W_l$ is the $i \times j$ weight
+matrix for layer $l$. In this form, the multilayer perceptron (MLP)
+produces a piecewise linear model of the input data. The computational
+complexity of training a multilayer perceptron is $\mathcal{O}(n d
+m)$, where $m$ is determined by the sizes of the layers of the network
+and the stopping criterion of the minimization used for finding
+weights. This paper uses an MLP built from Keras and Tensorflow to
+perform regression \cite{tensorflow2015-whitepaper,chollet2015keras}
+with ten hidden layers each having thirty two nodes (a total of
+approximately ten thousand parameters), ReLU activation, and five
+thousand epochs of SGD for training.
 
@@
 and the Delaunay interpolant $\hat f(y)$ at $y$ is given by
- $$ \hat f(y) = \sum_{i=0}^{d} w_i f\bigl(x^{(i)}\bigr). $$
-The computational complexity of Delaunay interpolation (for the
-implementation used \cite{chang2018polynomial}) is $\mathcal{O}(n d^4
-\log d)$, which is capable of scaling reasonably to $d \leq 50$. In
+ $$ \hat f(y) = \sum_{i=0}^{d} w_i f\bigl(x^{(i)}\bigr). $$ The
+computational complexity of Delaunay interpolation (for the
+implementation used \cite{chang2018polynomial}) is $\mathcal{O}(k n
+d^3)$. Given pathological data the entire triangulation could be
+computed with $k = n^{\lceil d / 2 \rceil}$, but the average case yields
+$k = d \log d$ and is capable of scaling reasonably to $d \leq 50$. In
 the present application, a Delaunay simplex $S$ containing $y$ is
@@
 for point $y$.
-
-\newpage
 
@@
 
-A well-studied technique for classification and approximation is the
-nearest neighbor algorithm \cite{cover1967nearest}. Nearest neighbor
-inherently utilizes the convex region $C^{x^{(i)}}$ (Voronoi cell
-\cite{dirichlet1850reduction}) consisting of all points closer to
+The nearest neighbor algorithm \cite{cover1967nearest} is a
+well-studied technique for classification and approximation. Nearest
+neighbor inherently utilizes the convex region $C^{x^{(i)}}$ (Voronoi
+cell \cite{dirichlet1850reduction}) consisting of all points closer to
 $x^{(i)}$ than to any other point $x^{(j)}$. The Voronoi model smooths
@@
 determined by the spacing of observed data. Predictions made by
-simplices whose vertices are not well-spaced (i.e. have large
+simplices whose vertices are not well-spaced (i.e., have large
 diameter, or are nearly contained in a hyperplane) have higher
@@
 errors for each algorithm. The plots for all five data sets have the
-same statistical format.
+same format.
 
@@
 described in \cite{kohavi1995study} with $k=10$. This approach
-randomly partitions data into $k$ equal sized (nearly) sets. Each
+randomly partitions data into $k$ (nearly) equal sized sets. Each
 algorithm is then evaluated by constructing an approximation over each
@@
     actual throughput distribution. The four vertical lines represent
-    commonly used $p$-values 0.05, 0.01, 0.001, $10^{-6}$,
-    respectively. All predictions to the right of a vertical line
-    represent CDF predictions that are significantly different (by
-    respective $p$-value) from the actual distribution according to
-    the KS test. The numerical counterpart to this figure is presented
-    in Table \ref{table:null-hypothesis-results}.}
+    cutoff KS statistics given $150$ samples for commonly used
+    $p$-values 0.05, 0.01, 0.001, $10^{-6}$, respectively. All
+    predictions to the right of a vertical line represent CDF
+    predictions that are significantly different (by respective
+    $p$-value) from the actual distribution according to the KS
+    test. The numerical counterpart to this figure is presented in
+    Table \ref{table:null-hypothesis-results}.}
   \label{fig:throughput-prediction}
@@
     Voronoi & $52.6\%$ & $\mathbf{43.4}\%$ & $\mathbf{34.4}\%$ & $\mathbf{19.1}\%$\\
-    BoxMesh & $99.5\%$ & $98.5\%$ & $96.6\%$ & $89.5\%$\\
+    BoxSpline & $99.5\%$ & $98.5\%$ & $96.6\%$ & $89.5\%$\\
   \end{tabular}
@@
     \hline
-    MARS & \quad\quad$4$&$6$ & \,\quad$22$&$3$ & \quad\quad$0$&$00137$\\
-    SVR & $\mathit{21}$&$\mathit{1}$ & $0$&$503$ & $\mathit{0}$&$\mathit{000137}$\\
-    MLP & $\mathbf{42}$&$\mathbf{1}$ & $210$&$0$ & $0$&$00102$\\
-    Delaunay & $6$&$0$ & $\mathbf{0}$&$\mathbf{000143}$ & $3$&$03$\\
-    ShepMod & $18$&$4$ & $0$&$711$ & $0$&$000149$\\
-    LSHEP & $8$&$4$ & $1$&$69$ & $\mathbf{0}$&$\mathbf{000116}$\\
-    Voronoi & $0$&$5$ & $\mathit{0}$&$\mathit{127}$ & $0$&$0375$\\
-    BoxSplines & $0$&$2$ & $0$&$816$ & $0$&$000526$\\
+    MARS & \quad\quad$4$&$6$ & \,\quad$20$&$0$ & \quad\quad$0$&$001$\\
+    SVR & $\mathit{21}$&$\mathit{1}$ & $\mathit{0}$&$\mathit{5}$ & $\mathbf{0}$&$\mathbf{0001}$\\
+    MLP & $\mathbf{42}$&$\mathbf{1}$ & $200$&$0$ & $0$&$001$\\
+    Delaunay & $6$&$0$ & $1$&$0$ & $1$&$0$\\
+    ShepMod & $18$&$4$ & $0$&$7$ & $\mathbf{0}$&$\mathbf{0001}$\\
+    LSHEP & $8$&$4$ & $2$&$0$ & $\mathbf{0}$&$\mathbf{0001}$\\
+    Voronoi & $0$&$5$ & $\mathbf{0}$&$\mathbf{1}$ & $0$&$04$\\
+    BoxSpline & $0$&$2$ & $0$&$8$ & $\mathit{0}$&$\mathit{0005}$\\
   \end{tabular}
@@
     error approximation, the average time to fit/prepare, and the
-    average time required to approximate one point. Interpolants
-    provide the lowest error approximation one third of the time
-    across all data, while regressors occupy the other two
+    average time required to approximate one point. The times have
+    been rounded to one significant digit, as reasonably large
+    fluctuations may be observed due to implementation hardware.
+    Interpolants provide the lowest error approximation for nearly one
+    third of all data, while regressors occupy the other two
     thirds. This result is obtained without any customized tuning or
@@
 error bounds for piecewise linear interpolation in arbitrary dimension
-(Section \ref{sec:theory}); and 2) an empirical evaluation across
+(Section \ref{sec:theory}); 2) an empirical evaluation across
 real-world problems (Section \ref{sec:data}) that demonstrates
@@
 phenomenon when compared with common regressors for sparse, moderately
-high dimensional problems.
+high dimensional problems; and 3) a demonstration that some
+interpolants generalize to interpolation in function spaces,
+preserving monotonicity (with CDFs, e.g.), neither of which common
+regressors can do.
 
@@
     \cline{1-3}\cline{5-10}
-    MARS & \quad$\mathit{7}$&$\mathit{7}$ &  & \quad\,\,\,$29$&$1$ & \quad$0$&$00137$ & \quad\quad\quad$0$&$0686$\\
-    SVR & $\mathbf{80}$&$\mathbf{2}$ &  & $0$&$00584$ & $\mathit{0}$&$\mathit{0000620}$ & $\mathit{0}$&$\mathit{00310}$\\
+    MARS & \quad$\mathit{7}$&$\mathit{7}$ &  & \quad\quad$29$&$1$ & \quad$0$&$00137$ & \quad\quad\quad$0$&$0686$\\
+    SVR & $\mathbf{80}$&$\mathbf{2}$ &  & $\mathit{0}$&$\mathit{00584}$ & $\mathit{0}$&$\mathit{0000620}$ & $\mathit{0}$&$\mathit{00310}$\\
     MLP & $0$&$0$ &  & $32$&$8$ & $0$&$000871$ & $0$&$0436$\\
-    Delaunay & $3$&$5$ &  & $\mathbf{0}$&$\mathbf{0000249}$ & $0$&$0385$ & $1$&$93$\\
+    Delaunay & $3$&$5$ &  & $0$&$0151$ & $0$&$0234$ & $1$&$18$\\
     ShepMod & $3$&$7$ &  & $0$&$00634$ & $0$&$0000644$ & $0$&$00322$\\
     LSHEP & $5$&$1$ &  & $0$&$0275$ & $\mathbf{0}$&$\mathbf{0000463}$ & $\mathbf{0}$&$\mathbf{00231}$\\
-    Voronoi & $0$&$0$ &  & $\mathit{0}$&$\mathit{000182}$ & $0$&$000396$ & $0$&$0198$\\
+    Voronoi & $0$&$0$ &  & $\mathbf{0}$&$\mathbf{000182}$ & $0$&$000396$ & $0$&$0198$\\
     BoxSpline & $0$&$4$ &  & $0$&$00724$ & $0$&$0000978$ & $0$&$00489$\\
@@
     \cline{1-3}\cline{5-10}
-    MARS & \quad\,$0$&$0$ &  & \quad\quad$37$&$9$ & \quad\,\,$0$&$00253$ & \quad\quad\quad\,\,$1$&$48$\\
-    SVR & $0$&$1$ &  & $0$&$859$ & $\mathit{0}$&$\mathit{000181}$ & $\mathit{0}$&$\mathit{106}$\\
+    MARS & \quad\,$0$&$0$ &  & \quad\quad\quad$37$&$9$ & \quad\,\,$0$&$00253$ & \quad\quad\quad\,\,$1$&$48$\\
+    SVR & $0$&$1$ &  & $\mathit{0}$&$\mathit{859}$ & $\mathit{0}$&$\mathit{000181}$ & $\mathit{0}$&$\mathit{106}$\\
     MLP & $\mathit{32}$&$\mathit{0}$ &  & $348$&$0$ & $0$&$00111$ & $0$&$653$\\
-    Delaunay & $0$&$0$ &  & $\mathbf{0}$&$\mathbf{000190}$ & $3$&$69$ & $2170$&$0$\\
+    Delaunay & $0$&$0$ &  & $2$&$47$ & $1$&$22$ & $720$&$0$\\
     ShepMod & $\mathbf{66}$&$\mathbf{4}$ &  & $1$&$13$ & $0$&$000182$ & $0$&$107$\\
     LSHEP & $0$&$0$ &  & $2$&$39$ & $\mathbf{0}$&$\mathbf{000144}$ & $\mathbf{0}$&$\mathbf{0845}$\\
-    Voronoi & $1$&$6$ &  & $\mathit{0}$&$\mathit{298}$ & $0$&$0274$ & $16$&$1$\\
+    Voronoi & $1$&$6$ &  & $\mathbf{0}$&$\mathbf{298}$ & $0$&$0274$ & $16$&$1$\\
     BoxSpline & $0$&$0$ &  & $1$&$26$ & $0$&$000643$ & $0$&$377$\\
@@
     \cline{1-3}\cline{5-10}
-    MARS & \,\,\,\,$10$&$7$ &  & \quad\quad\,\,$0$&$151$ & \quad$0$&$000117$ & \quad\quad\quad\,\,$0$&$0304$\\
-    SVR & $4$&$1$ &  & $0$&$133$ & $\mathbf{0}$&$\mathbf{0000947}$ & $\mathbf{0}$&$\mathbf{0246}$\\
+    MARS & \,\,\,\,$10$&$7$ &  & \quad\quad\quad$0$&$151$ & \quad$0$&$000117$ & \quad\quad\quad\,\,$0$&$0304$\\
+    SVR & $4$&$1$ &  & $\mathit{0}$&$\mathit{133}$ & $\mathbf{0}$&$\mathbf{0000947}$ & $\mathbf{0}$&$\mathbf{0246}$\\
     MLP & $\mathbf{56}$&$\mathbf{8}$ &  & $169$&$0$ & $0$&$00137$ & $0$&$356$\\
-    Delaunay & $0$&$1$ &  & $\mathbf{0}$&$\mathbf{000116}$ & $1$&$55$ & $403$&$0$\\
+    Delaunay & $0$&$1$ &  & $0$&$664$ & $0$&$886$ & $230$&$0$\\
     ShepMod & $3$&$5$ &  & $0$&$265$ & $0$&$000128$ & $0$&$0333$\\
     LSHEP & $\mathit{28}$&$\mathit{4}$ &  & $0$&$874$ & $\mathit{0}$&$\mathit{0000975}$ & $\mathit{0}$&$\mathit{0254}$\\
-    Voronoi & $0$&$2$ &  & $\mathit{0}$&$\mathit{0112}$ & $0$&$0270$ & $7$&$01$\\
+    Voronoi & $0$&$2$ &  & $\mathbf{0}$&$\mathbf{012}$ & $0$&$0270$ & $7$&$01$\\
     BoxSpline & $0$&$3$ &  & $0$&$330$ & $0$&$000406$ & $0$&$106$\\
@@
     \cline{1-3}\cline{5-10}
-    MARS & \quad\,$0$&$0$ &  & \quad\quad\,\,$22$&$0$ & \quad$0$&$00148$ & \quad\quad\quad$0$&$820$\\
-    SVR & $0$&$0$ &  & $1$&$01$ & $\mathit{0}$&$\mathit{000210}$ & $\mathit{0}$&$\mathit{117}$\\
+    MARS & \quad\,$0$&$0$ &  & \quad\quad\quad$22$&$0$ & \quad$0$&$00148$ & \quad\quad\quad\,\,$0$&$820$\\
+    SVR & $0$&$0$ &  & $\mathit{1}$&$\mathit{01}$ & $\mathit{0}$&$\mathit{000210}$ & $\mathit{0}$&$\mathit{117}$\\
     MLP & $\mathbf{79}$&$\mathbf{5}$ &  & $290$&$0$ & $0$&$000714$ & $0$&$397$\\
-    Delaunay & $\mathit{20}$&$\mathit{5}$ &  & $\mathbf{0}$&$\mathbf{000243}$ & $6$&$83$ & $3800$&$0$\\
+    Delaunay & $\mathit{20}$&$\mathit{5}$ &  & $3$&$12$ & $3$&$71$ & $2070$&$0$\\
     ShepMod & $0$&$1$ &  & $1$&$45$ & $0$&$000220$ & $0$&$122$\\
     LSHEP & $0$&$0$ &  & $3$&$47$ & $\mathbf{0}$&$\mathbf{000176}$ & $\mathbf{0}$&$\mathbf{0981}$\\
-    Voronoi & $0$&$0$ &  & $\mathit{0}$&$\mathit{197}$ & $0$&$0950$ & $52$&$8$\\
+    Voronoi & $0$&$0$ &  & $\mathbf{0}$&$\mathbf{197}$ & $0$&$0950$ & $52$&$8$\\
     BoxSpline & $0$&$0$ &  & $1$&$66$ & $0$&$000956$ & $0$&$532$\\
@@
     \cline{1-3}\cline{5-10}
-    Delaunay & \,\,$\mathbf{62}$&$\mathbf{0}$ &  & \quad\quad$\mathbf{0}$&$\mathbf{000517}$ & \quad$0$&$353$ & \quad\quad\,\,$106$&$0$\\
-    ShepMod & $0$&$0$ &  & $0$&$0884$ & $\mathbf{0}$&$\mathbf{000145}$ & $\mathbf{0}$&$\mathbf{0436}$\\
-    Voronoi & $\mathit{38}$&$\mathit{0}$ &  & $\mathit{0}$&$\mathit{0173}$ & $0$&$00253$ & $0$&$762$\\
+    Delaunay & \,\,$\mathbf{62}$&$\mathbf{0}$ &  & \quad\quad\quad$0$&$344$ & \quad$0$&$00984$ & \quad\quad\,\,$2$&$71$\\
+    ShepMod & $0$&$0$ &  & $\mathbf{0}$&$\mathbf{0884}$ & $\mathbf{0}$&$\mathbf{000145}$ & $\mathbf{0}$&$\mathbf{0436}$\\
+    Voronoi & $\mathit{38}$&$\mathit{0}$ &  & $1$&$0$ & $0$&$00253$ & $0$&$762$\\
     BoxSpline & $0$&$0$ &  & $0$&$0972$ & $\mathit{0}$&$\mathit{000210}$ & $\mathit{0}$&$\mathit{0633}$\\

### 2019-03-30 Paper_[2019-02]_Numerical_Algorithms/paper.tex
Final update before submission.

@@
     MARS & \quad\quad$4$&$6$ & \,\quad$20$&$0$ & \quad\quad$0$&$001$\\
-    SVR & $\mathit{21}$&$\mathit{1}$ & $\mathit{0}$&$\mathit{5}$ & $\mathbf{0}$&$\mathbf{0001}$\\
+    SVR & $\mathit{21}$&$\mathit{1}$ & $\mathbf{0}$&$\mathbf{5}$ & $\mathbf{0}$&$\mathbf{0001}$\\
     MLP & $\mathbf{42}$&$\mathbf{1}$ & $200$&$0$ & $0$&$001$\\
     Delaunay & $6$&$0$ & $1$&$0$ & $1$&$0$\\
-    ShepMod & $18$&$4$ & $0$&$7$ & $\mathbf{0}$&$\mathbf{0001}$\\
+    ShepMod & $18$&$4$ & $\mathit{0}$&$\mathit{7}$ & $\mathbf{0}$&$\mathbf{0001}$\\
     LSHEP & $8$&$4$ & $2$&$0$ & $\mathbf{0}$&$\mathbf{0001}$\\
-    Voronoi & $0$&$5$ & $\mathbf{0}$&$\mathbf{1}$ & $0$&$04$\\
+    Voronoi & $0$&$5$ & $1$&$0$ & $0$&$04$\\
     BoxSpline & $0$&$2$ & $0$&$8$ & $\mathit{0}$&$\mathit{0005}$\\
@@
 (Section \ref{sec:theory}); 2) an empirical evaluation across
-real-world problems (Section \ref{sec:data}) that demonstrates
-interpolants produce competitively accurate models of multivariate
-phenomenon when compared with common regressors for sparse, moderately
-high dimensional problems; and 3) a demonstration that some
-interpolants generalize to interpolation in function spaces,
-preserving monotonicity (with CDFs, e.g.), neither of which common
-regressors can do.
+real-world problems that demonstrates interpolants produce
+competitively accurate models of multivariate phenomenon when compared
+with common regressors for sparse, moderately high dimensional
+problems (Section \ref{sec:data}); and 3) a demonstration that some
+interpolants generalize to interpolation in function spaces (Section
+\ref{sec:error}), preserving monotonicity (with CDFs, e.g.), neither
+of which common regressors can do.
 
@@
 when there are relatively few data points for the dimension (for
-empirical results presented, $\log_d n \leq 5$).
+empirical results presented, $\log_d n \leq 5$). These findings
+encourage broader application of interpolants to multivariate
+approximation in science
 
@@
 \label{sec:appendix}
+
+The tables that follow show the precise experimental results for all
+data sets presented in Section \ref{sec:data}. The tests were all run
+serially on an otherwise idle machine with a CentOS 6.10 operating
+system and an Intel i7-3770 CPU operating at 3.4 GHz. The detailed
+performance results in the tables that follow are very much dependent
+on the problem and the algorithm implementation (e.g., some codes are
+TOMS software, some industry distributions, and others are from
+conference paper venues). Different typeface is used to show best
+performers, however not much significance should be attached to
+ranking algorithms based on small time (millisecond) differences. The
+results serve as a demonstration of conceptual validity.
 
@@
     MARS & \quad$\mathit{7}$&$\mathit{7}$ &  & \quad\quad$29$&$1$ & \quad$0$&$00137$ & \quad\quad\quad$0$&$0686$\\
-    SVR & $\mathbf{80}$&$\mathbf{2}$ &  & $\mathit{0}$&$\mathit{00584}$ & $\mathit{0}$&$\mathit{0000620}$ & $\mathit{0}$&$\mathit{00310}$\\
+    SVR & $\mathbf{80}$&$\mathbf{2}$ &  & $\mathbf{0}$&$\mathbf{00584}$ & $\mathit{0}$&$\mathit{0000620}$ & $\mathit{0}$&$\mathit{00310}$\\
     MLP & $0$&$0$ &  & $32$&$8$ & $0$&$000871$ & $0$&$0436$\\
     Delaunay & $3$&$5$ &  & $0$&$0151$ & $0$&$0234$ & $1$&$18$\\
-    ShepMod & $3$&$7$ &  & $0$&$00634$ & $0$&$0000644$ & $0$&$00322$\\
+    ShepMod & $3$&$7$ &  & $\mathit{0}$&$\mathit{00634}$ & $0$&$0000644$ & $0$&$00322$\\
     LSHEP & $5$&$1$ &  & $0$&$0275$ & $\mathbf{0}$&$\mathbf{0000463}$ & $\mathbf{0}$&$\mathbf{00231}$\\
-    Voronoi & $0$&$0$ &  & $\mathbf{0}$&$\mathbf{000182}$ & $0$&$000396$ & $0$&$0198$\\
+    Voronoi & $0$&$0$ &  & $0$&$0152$ & $0$&$000396$ & $0$&$0198$\\
     BoxSpline & $0$&$4$ &  & $0$&$00724$ & $0$&$0000978$ & $0$&$00489$\\
@@
     MARS & \quad\,$0$&$0$ &  & \quad\quad\quad$37$&$9$ & \quad\,\,$0$&$00253$ & \quad\quad\quad\,\,$1$&$48$\\
-    SVR & $0$&$1$ &  & $\mathit{0}$&$\mathit{859}$ & $\mathit{0}$&$\mathit{000181}$ & $\mathit{0}$&$\mathit{106}$\\
+    SVR & $0$&$1$ &  & $\mathbf{0}$&$\mathbf{859}$ & $\mathit{0}$&$\mathit{000181}$ & $\mathit{0}$&$\mathit{106}$\\
     MLP & $\mathit{32}$&$\mathit{0}$ &  & $348$&$0$ & $0$&$00111$ & $0$&$653$\\
     Delaunay & $0$&$0$ &  & $2$&$47$ & $1$&$22$ & $720$&$0$\\
-    ShepMod & $\mathbf{66}$&$\mathbf{4}$ &  & $1$&$13$ & $0$&$000182$ & $0$&$107$\\
+    ShepMod & $\mathbf{66}$&$\mathbf{4}$ &  & $\mathit{1}$&$\mathit{13}$ & $0$&$000182$ & $0$&$107$\\
     LSHEP & $0$&$0$ &  & $2$&$39$ & $\mathbf{0}$&$\mathbf{000144}$ & $\mathbf{0}$&$\mathbf{0845}$\\
-    Voronoi & $1$&$6$ &  & $\mathbf{0}$&$\mathbf{298}$ & $0$&$0274$ & $16$&$1$\\
+    Voronoi & $1$&$6$ &  & $2$&$77$ & $0$&$0274$ & $16$&$1$\\
     BoxSpline & $0$&$0$ &  & $1$&$26$ & $0$&$000643$ & $0$&$377$\\
@@
     \cline{1-3}\cline{5-10}
-    MARS & \,\,\,\,$10$&$7$ &  & \quad\quad\quad$0$&$151$ & \quad$0$&$000117$ & \quad\quad\quad\,\,$0$&$0304$\\
-    SVR & $4$&$1$ &  & $\mathit{0}$&$\mathit{133}$ & $\mathbf{0}$&$\mathbf{0000947}$ & $\mathbf{0}$&$\mathbf{0246}$\\
+    MARS & \,\,\,\,$10$&$7$ &  & \quad\quad\quad$\mathit{0}$&$\mathit{151}$ & \quad$0$&$000117$ & \quad\quad\quad\,\,$0$&$0304$\\
+    SVR & $4$&$1$ &  & $\mathbf{0}$&$\mathbf{133}$ & $\mathbf{0}$&$\mathbf{0000947}$ & $\mathbf{0}$&$\mathbf{0246}$\\
     MLP & $\mathbf{56}$&$\mathbf{8}$ &  & $169$&$0$ & $0$&$00137$ & $0$&$356$\\
@@
     LSHEP & $\mathit{28}$&$\mathit{4}$ &  & $0$&$874$ & $\mathit{0}$&$\mathit{0000975}$ & $\mathit{0}$&$\mathit{0254}$\\
-    Voronoi & $0$&$2$ &  & $\mathbf{0}$&$\mathbf{012}$ & $0$&$0270$ & $7$&$01$\\
+    Voronoi & $0$&$2$ &  & $0$&$675$ & $0$&$0270$ & $7$&$01$\\
     BoxSpline & $0$&$3$ &  & $0$&$330$ & $0$&$000406$ & $0$&$106$\\
@@
     MARS & \quad\,$0$&$0$ &  & \quad\quad\quad$22$&$0$ & \quad$0$&$00148$ & \quad\quad\quad\,\,$0$&$820$\\
-    SVR & $0$&$0$ &  & $\mathit{1}$&$\mathit{01}$ & $\mathit{0}$&$\mathit{000210}$ & $\mathit{0}$&$\mathit{117}$\\
+    SVR & $0$&$0$ &  & $\mathbf{1}$&$\mathbf{01}$ & $\mathit{0}$&$\mathit{000210}$ & $\mathit{0}$&$\mathit{117}$\\
     MLP & $\mathbf{79}$&$\mathbf{5}$ &  & $290$&$0$ & $0$&$000714$ & $0$&$397$\\
     Delaunay & $\mathit{20}$&$\mathit{5}$ &  & $3$&$12$ & $3$&$71$ & $2070$&$0$\\
-    ShepMod & $0$&$1$ &  & $1$&$45$ & $0$&$000220$ & $0$&$122$\\
+    ShepMod & $0$&$1$ &  & $\mathit{1}$&$\mathit{4}5$ & $0$&$000220$ & $0$&$122$\\
     LSHEP & $0$&$0$ &  & $3$&$47$ & $\mathbf{0}$&$\mathbf{000176}$ & $\mathbf{0}$&$\mathbf{0981}$\\
-    Voronoi & $0$&$0$ &  & $\mathbf{0}$&$\mathbf{197}$ & $0$&$0950$ & $52$&$8$\\
+    Voronoi & $0$&$0$ &  & $3$&$32$ & $0$&$0950$ & $52$&$8$\\
     BoxSpline & $0$&$0$ &  & $1$&$66$ & $0$&$000956$ & $0$&$532$\\
@@
     ShepMod & $0$&$0$ &  & $\mathbf{0}$&$\mathbf{0884}$ & $\mathbf{0}$&$\mathbf{000145}$ & $\mathbf{0}$&$\mathbf{0436}$\\
-    Voronoi & $\mathit{38}$&$\mathit{0}$ &  & $1$&$0$ & $0$&$00253$ & $0$&$762$\\
-    BoxSpline & $0$&$0$ &  & $0$&$0972$ & $\mathit{0}$&$\mathit{000210}$ & $\mathit{0}$&$\mathit{0633}$\\
+    Voronoi & $\mathit{38}$&$\mathit{0}$ &  & $0$&$360$ & $0$&$00253$ & $0$&$762$\\
+    BoxSpline & $0$&$0$ &  & $\mathit{0}$&$\mathit{0972}$ & $\mathit{0}$&$\mathit{000210}$ & $\mathit{0}$&$\mathit{0633}$\\
     \cline{1-3}\cline{5-10}

### 2019-03-31 Paper_[2019-02]_Numerical_Algorithms/paper.tex
Finished paper, renamed some files for submission. Submitted.

@@
 regression and classification tasks
-\cite{rumelhart1988learning,hornik1989multilayer}. When using the
+\cite{hornik1989multilayer,rumelhart1988learning}. When using the
 rectified linear unit (ReLU) activation function
@@
 \textit{metric interpolation}, also called a Shepard method
-\cite{shepard1968two,gordon1978shepard}. The interpolant has the form
+\cite{gordon1978shepard,shepard1968two}. The interpolant has the form
  $$ p(x) = \frac{\sum\limits_{k=1}^{n}W_k(x)f\bigl(x^{(k)}\bigr)}
@@
   \begin{tikzpicture}
-    \node (img)  {\includegraphics[width=0.75\textwidth,]{Figures/plot-delaunay-example-prediction.pdf}};
+    \node (img)  {\includegraphics[width=0.75\textwidth,]{Fig1.pdf}};
     \node[below=of img, node distance=1cm, yshift=1cm] {I/O Throughput};
@@
   \begin{tikzpicture}
-    \node (img)  {\includegraphics[width=0.75\textwidth,]{Figures/raw-histogram-forest-fire.pdf}};
+    \node (img)  {\includegraphics[width=0.75\textwidth,]{Fig2.pdf}};
     \node[below=of img, node distance=1cm, yshift=1cm] {Forest Fire Area Burned};
@@
   \begin{tikzpicture}
-    \node (img)  {\includegraphics[width=0.8\textwidth,]{Figures/error-box-forest-fire.pdf}};
+    \node (img)  {\includegraphics[width=0.8\textwidth,]{Fig3.pdf}};
     \node[below=of img, node distance=1cm, yshift=1cm] {Area Burned Error};
@@
   \begin{tikzpicture}
-    \node (img)  {\includegraphics[width=0.75\textwidth,]{Figures/raw-histogram-parkinsons.pdf}};
+    \node (img)  {\includegraphics[width=0.75\textwidth,]{Fig4.pdf}};
     \node[below=of img, node distance=1cm, yshift=1cm] {Parkinson's Total UPDRS Score};
@@
   \begin{tikzpicture}
-    \node (img)  {\includegraphics[width=0.8\textwidth,]{Figures/error-box-parkinsons.pdf}};
+    \node (img)  {\includegraphics[width=0.8\textwidth,]{Fig5.pdf}};
     \node[below=of img, node distance=1cm, yshift=1cm] {Total UPDRS Score Error};
@@
   \begin{tikzpicture}
-    \node (img)  {\includegraphics[width=0.75\textwidth,]{Figures/raw-histogram-weather.pdf}};
+    \node (img)  {\includegraphics[width=0.75\textwidth,]{Fig6.pdf}};
     \node[below=of img, node distance=1cm, yshift=1cm] {Daily Rainfall in Sydney};
@@
   \begin{tikzpicture}
-    \node (img)  {\includegraphics[width=0.8\textwidth,]{Figures/error-box-weather.pdf}};
+    \node (img)  {\includegraphics[width=0.8\textwidth,]{Fig7.pdf}};
     \node[below=of img, node distance=1cm, yshift=1cm] {Sydney Rainfall Tomorrow Error};
@@
   \begin{tikzpicture}
-    \node (img)  {\includegraphics[width=0.75\textwidth,]{Figures/raw-histogram-credit-card.pdf}};
+    \node (img)  {\includegraphics[width=0.75\textwidth,]{Fig8.pdf}};
     \node[below=of img, node distance=1cm, yshift=1cm] {Credit Card Transaction Amount};
@@
   \begin{tikzpicture}
-    \node (img)  {\includegraphics[width=0.8\textwidth,]{Figures/error-box-credit-card.pdf}};
+    \node (img)  {\includegraphics[width=0.8\textwidth,]{Fig9.pdf}};
     \node[below=of img, node distance=1cm, yshift=1cm] {Transaction Amount Error};
@@
   \begin{tikzpicture}
-    \node (img)  {\includegraphics[width=0.75\textwidth,]{Figures/raw-histogram-throughput.pdf}};
-    \node[below=of img, node distance=1cm, yshift=1cm] {I/O Read Throughtput};
+    \node (img)  {\includegraphics[width=0.75\textwidth,]{Fig10.pdf}};
+    \node[below=of img, node distance=1cm, yshift=1cm] {I/O Read Throughput};
     \node[left=of img, node distance=0cm, rotate=90, anchor=center,yshift=-0.7cm] {Count};
@@
   \begin{tikzpicture}
-    \node (img)  {\includegraphics[width=0.8\textwidth,]{Figures/error-box-throughput.pdf}};
-    \node[below=of img, node distance=1cm, yshift=1cm] {Read Throughtput KS Statistic};
+    \node (img)  {\includegraphics[width=0.8\textwidth,]{Fig11.pdf}};
+    \node[below=of img, node distance=1cm, yshift=1cm] {Read Throughput KS Statistic};
   \end{tikzpicture}
@@
   \begin{tikzpicture}
-    \node (img)  {\includegraphics[width=0.8\textwidth]{Figures/plot-throughput-ks.pdf}};
+    \node (img)  {\includegraphics[width=0.8\textwidth]{Fig12.pdf}};
     \node[below=of img, node distance=1cm, yshift=1cm] {KS Statistic for Predicted vs. Actual};

### 2019-05-03 Disseration/algs.tex
Finished first full draft of preliminary defense document. Ready to share with committee.

--- 2019-03-30 Paper_[2019-02]_Numerical_Algorithms/paper.tex
@@
-\RequirePackage{fix-cm}
-\documentclass[smallextended,final]{svjour3}  % onecolumn (second format)
-\smartqed  % flush right qed marks, e.g. at end of proof
-
-% Insert here the call for the packages your document requires
-\usepackage{graphicx}
-\usepackage{tikz}     % For adding axes to plots
-\usetikzlibrary{positioning}
-\usepackage{graphicx} % For including PDFs
-\usepackage{amsmath}  % assumes amsmath package installed
-\usepackage{amssymb}  % assumes amsmath package installed
-\usepackage{booktabs} % For formal tables
-\usepackage{tabularx} % For tables with nice column widths.
-\usepackage[colorlinks,allcolors=blue]{hyperref}
-\usepackage{float}    % For controlling the placement of tables.
-
-%
-% please place your own definitions here and don't use \def but
-% \newcommand{}{}
-
-%% Command for creating multiplie-line cells inside a table environment
-\newcommand{\multilinecell}[2][c]{%
-  \begin{tabular}[#1]{@{}c@{}}#2\end{tabular}}
-
-\newcolumntype{d}[1]{D{.}{.}{#1}}
-
-%% Create a ``Lemma'' environment that does not have italics font.
-\spnewtheorem{plainlemma}{Lemma}{\bf}{}
-%% Create a ``Theorem'' environment that is not numbered.
-\spnewtheorem*{onetheorem}{Theorem}{\bf}{}
-%% Create a ``Proof'' environment that starts with 'Proof.'
-\spnewtheorem*{proofdot}{Proof.}{\it}{}
-
-%% \newcommand{\Bigl}{\mathopen\Big}
-%% \newcommand{\Bigr}{\mathclose\Big}
-%
-% Insert the name of "your journal" with
-\journalname{Numerical Algorithms}
-%
-\begin{document}
-
-\title{Interpolation of Sparse High-Dimensional Data\thanks
-  {This work was supported by the National Science Foundation Grant CNS-1565314.}
-}
-% \subtitle{Do you have a subtitle?\\ If so, write it here}
-%\titlerunning{Short form of title}        % if too long for running head
-
-\author{Thomas C. H. Lux   \and
-  Layne T. Watson          \and
-  Tyler H. Chang           \and
-  Yili Hong                \and
-  Kirk Cameron
-}
-
-%\authorrunning{Short form of author list} % if too long for running head
-
-\institute{Thomas Lux \at
-  Virginia Polytechnic Institute \& State University (VPI\&SU) \\
-  Blacksburg, VA 24061 \\
-  \email{tchlux@vt.edu}
-}
-
-\date{Received: date / Accepted: date}
-% The correct dates will be entered by the editor
-
-\maketitle
-
-\begin{abstract}
-Increases in the quantity of available data have allowed all fields of
-science to generate more accurate models of multivariate phenomena.
-Regression and interpolation become challenging when the dimension of
-data is large, especially while maintaining tractable computational
-complexity. Regression is a popular approach to solving approximation
-problems with high dimension, however there are often advantages to
-interpolation. This paper presents a novel and insightful error bound
-for (piecewise) linear interpolation in arbitrary dimension and
-contrasts the performance of some interpolation techniques with
-popular regression techniques. Empirical results demonstrate the
-viability of interpolation for moderately high dimensional
-approximation problems, and encourage broader application of
-interpolants to multivariate approximation in science.
-
-\keywords{approximation \and regression \and interpolation \and high
-  dimension \and error bound}
-% \PACS{PACS code1 \and PACS code2 \and more}
-% \subclass{MSC code1 \and MSC code2 \and more}
-\end{abstract}
-
-%     Introduction     
-%======================
-
-%% ===================================================================
-\section{Introduction}
-\label{sec:introduction}
-
-Regression and interpolation are problems of considerable importance
-that find applications across many fields of science. Pollution and
-air quality analysis \cite{de2008field}, energy consumption management
-\cite{lazos2014optimisation}, and student performance prediction
-\cite{cortez2008using,lux2016applications} are a few of many
-interdisciplinary applications of multivariate regression for
-predictive analysis. As discussed later, these techniques can also be
-applied to prediction problems related to forest fire risk assessment
-\cite{cortez2007data}, Parkinson's patient clinical evaluations
-\cite{tsanas2010accurate}, local rainfall and weather
-\cite{williams2009rattle}, credit card transactions
-\cite{pozzolo2015calibrating}, and high performance computing (HPC)
-file input/output (I/O) \cite{lux2018nonparametric}.
-
-Regression and interpolation have a considerable theoretical base in
-one dimension \cite{cheney2009course}. Splines in particular are well
-understood as an interpolation technique in one dimension
-\cite{de1978practical}, particularly B-splines. Tensor products of
-B-splines \cite{unther1996interpolating} or other basis functions have
-an unfortunate exponential scaling with increasing
-dimension. Exponential scaling prohibits tensor products from being
-reasonably applied beyond three-dimensional data. In order to address
-this dimensional scaling challenge, C. de Boor and others proposed box
-splines \cite{de2013box}, of which one of the approximation techniques
-in this work is composed \cite{lux2018novel}.
-
-The theoretical foundation of low dimensional interpolation allows the
-construction of strong error bounds that are absent from high
-dimensional problems. This work extends some known results regarding
-the secant method \cite{dennis1996numerical} to construct an
-interpolation error bound for problems of arbitrary dimension. These
-error bounds are useful, considering the same cannot be said for
-regression algorithms in general. The maximum complexity of an
-interpolant is bounded by the amount of data available, while the
-maximum complexity of a regressor is bounded by both the amount of
-data and the chosen parametric form. For this reason, generic uniform
-bounds are largely unobtainable for regression techniques on arbitrary
-approximation problems, even when the approximation domain is bounded.
-
-Aside from theoretical motivation for the use of interpolants, there
-are often computational advantages as well. Interpolants do not have
-the need for \textit{fitting} data, or minimizing error with respect
-to model parameters. In applications where the amount of data is large
-and the relative number of predictions that need to be made for a
-given collection of data is small, the direct application of an
-interpolant is much less computationally expensive.
-
-In this work, multivariate interpolation is defined given a closed
-convex subset $Y$ of a metrizable topological vector space with metric
-$s$, some function $f:\mathbb{R}^d \rightarrow Y$ and a set of points
-$X = \bigl\{x^{(1)}$, $\ldots$, $x^{(n)}\bigr\} \subset \mathbb{R}^d$,
-along with associated function values $f\bigl(x^{(i)}\bigr)$. The goal
-is to construct an approximation $\hat f: \mathbb{R}^d \rightarrow Y$
-such that $\hat f\bigl(x^{(i)}\bigr) = f\bigl(x^{(i)}\bigr)$ for all
-$i = 1$, $\ldots$, $n$. It is often the case that the form of the true
-underlying function $f$ is unknown, however it is still desirable to
-construct an approximation $\hat f$ with small approximation error at
-$y \notin X$. The two metric spaces that will be discussed in this
-work are the real numbers with metric $s(x,y) = |x-y|$, and the set of
-cumulative distribution functions (CDFs) with the Kolmogorov-Smirnov
-(KS) statistic \cite{lilliefors1967kolmogorov} as metric.
-
-Multivariate regression is often used when the underlying function is
-presumed to be stochastic, or stochastic error is introduced in the
-evaluation of $f$. Hence, multivariate regression relaxes the
-conditions of interpolation by choosing parameters $P$ defining $\hat
-f(x;P)$ to minimize the error vector $\Bigl( \bigl | \hat f
-\bigl(x^{(1)};P\bigr) - f\bigl(x^{(1)}\bigr) \bigr|$, $\ldots$, $\bigl
-| \hat f \bigl(x^{(n)}; P\bigr) - f\bigl(x^{(n)}\bigr) \bigr | \Bigr)$
-in some norm. The difficult question in the case of regression is
-often what parametric form to adopt for any given application.
-
-The most challenging problem when scaling in dimension is that the
-number of possible interactions between dimensions grows
-exponentially. Quantifying all possible interactions becomes
-intractable, and hence beyond three-dimensional data mostly linear
-models are used. That is not to say nonlinear models are absent, but
-nonlinearities are often either preconceived or model pairwise
-interactions between dimensions at most. Even globally nonlinear
-approximations such as neural networks are constructed from
-compositions of summed low-interaction functions
-\cite{clevert2015fast}.
-
-Provided the theoretical and practical motivations for exploring
-interpolants, this work aims to study the empirical performance
-differences between a set of scalable (moderately) interpolation
-techniques and a set of common regression techniques. These techniques
-are applied to a collection of moderately high dimensional problems
-($5 \le d \le 30$) and the empirical results are discussed.
-
-The remainder of this paper is organized as follows. Sections
-\ref{sec:regression} and \ref{sec:interpolation} present the
-multivariate models. Section \ref{sec:error} gives the error measuring
-methodology that is used for collecting empirical results. Section
-\ref{sec:theory} contains the theoretical error bounds for
-interpolation. Section \ref{sec:data} presents the data sets for
-empirical approximation analysis and the approximation results for all
-models on these data sets. Section \ref{sec:discussion} analyzes and
-discusses the results, their implications, and their
-limitations. Finally, Section \ref{sec:conclusion} concludes.
-
 %% ===================================================================
@@
 \cite{thacker2010algorithm}.
-
-\subsection{Box Splines}
-\label{sec:box-splines}
-
-The box spline model used here is an interpolation technique built
-from overlapping box splines \cite{de2013box}. The box splines serve
-as basis functions that are shifted and scaled to have support over
-box shaped regions. The boxes are constructed in a way to guarantee a
-covering of the domain \cite{lux2018novel}. Given a set of box splines
-$\bigl\{b^{x^{(i)}}\bigr\}$ with the iterative box properties outlined
-in \cite{lux2018novel} and anchored at interior points
-$\bigl\{x^{(i)}\bigr\}$,
- $$ \hat f(y) = \frac{\sum\limits_{i=1}^n b^{x^{(i)}}(y) f\bigl(x^{(i)}\bigr)}{\sum\limits_{i=1}^n b^{x^{(i)}}(y)}. $$
-Note that the box splines always satisfy
-$b^{x^{(i)}}\bigl(x^{(j)}\bigr) = \delta_{ij}$ and $b^{x^{(i)}}(y)
-\geq 0$. The computational complexity of interpolation via the box
-spline model is $\mathcal{O}(n^2 d)$.
-
-\subsection{Voronoi}
-\label{sec:voronoi}
-
-The nearest neighbor algorithm \cite{cover1967nearest} is a
-well-studied technique for classification and approximation. Nearest
-neighbor inherently utilizes the convex region $C^{x^{(i)}}$ (Voronoi
-cell \cite{dirichlet1850reduction}) consisting of all points closer to
-$x^{(i)}$ than to any other point $x^{(j)}$. The Voronoi model smooths
-the nearest neighbor approximation by utilizing the Voronoi cells to
-define support via a generic basis function $v: \mathbb{R}^d
-\rightarrow \mathbb{R}_+$ given by
- $$ v^{x^{(i)}}(y) = \left(1 - \frac{\bigl\|y - x^{(i)}\bigr\|_2}
- {2 \ h \bigl(y - x^{(i)} \mid x^{(i)}\bigr)} \right)_+, $$
-where $h\bigl(w \mid x^{(i)}\bigr)$ is the distance between $x^{(i)}$
-and the boundary of the Voronoi cell $C^{x^{(i)}}$ in the direction
-$w$. $v^{x^{(i)}}\bigl(x^{(j)}\bigr) = \delta_{ij}$ and $v^{x^{(i)}}$
-has local support, giving the interpolated value
- $$ f(y) = \frac{\sum\limits_{i=1}^n v^{x^{(i)}}(y)
-  f\bigl(x^{(i)}\bigr)}{\sum\limits_{i=1}^n v^{x^{(i)}}(y)}, $$
-where $0 \leq v^{x^{(i)}}(y) \leq 1$. The computational complexity of
-interpolation via this Voronoi model is $\mathcal{O}(n^2 d)$. All of
-the approximations are an interpolant involving a convex combination
-of known function values $f\bigl(x^{(i)}\bigr)$.
-
-%% ===================================================================
-\section{Measuring Error}
-\label{sec:error}
-
-\begin{figure}[htb]
-  \centering
-  \begin{tikzpicture}
-    \node (img)  {\includegraphics[width=0.75\textwidth,]{Figures/plot-delaunay-example-prediction.pdf}};
-    \node[below=of img, node distance=1cm, yshift=1cm] {I/O Throughput};
-    \node[left=of img, node distance=0cm, rotate=90, anchor=center,yshift=-0.7cm] {CDF Value};
-  \end{tikzpicture}
-  \vspace{-0.3cm}
-  \caption{In this HPC I/O example, the general methodology for
-    predicting a CDF and evaluating error can be seen. The Delaunay
-    method chose three source distributions (dotted lines) and
-    assigned weights \{.3, .4, .3\} (top to bottom at arrow). The
-    weighted sum of the three known CDFs produces the predicted CDF
-    (dashed line). The KS Statistic (arrow) computed between the true
-    CDF (solid line) and predicted CDF (dashed line) is 0.2 for this
-    example. The KS test null hypothesis is rejected at $p$-value
-    0.01, however it is not rejected at $p$-value 0.001.
-  \vspace{-.1cm}}
-  \label{fig:prediction-example}
-\end{figure}
-
-When the range of an approximation is the real numbers, error is
-reported with summary statistics including: min absolute error, max
-absolute error, and absolute error quartiles. When the range of an
-approximation is the space of cumulative distribution functions, the
-Kolmogorov-Smirnov statistic (max-norm difference between the
-functions) is used.
-
-A hurdle when modeling function-valued outputs such as cumulative
-distribution functions (CDFs) or probability density functions (PDFs)
-is that certain properties must be maintained. It is necessary that a
-PDF $f: \mathbb{R} \rightarrow \mathbb{R}$ have the properties $f(x)
-\geq 0$ and $\int_{-\infty}^{\infty}f(x)dx = 1$. However, for a CDF
-$F: \mathbb{R} \rightarrow \mathbb{R}$ the properties are
-$F(x) \in [0,1]$ and $F(x)$ is absolutely continuous and
-nondecreasing. This work utilizes the fact that a convex combination
-of CDFs (or PDFs) results in a valid CDF (or PDF). Given $G(x) =
-\sum_{i}w_i F_i(x)$, $\sum_{i} w_i = 1$, $w_i \geq 0$, and each $F_i$
-is a valid CDF, $G$ must also be a valid CDF. A demonstration of how
-this is applied can be seen in Figure \ref{fig:prediction-example}.
-
-The performance of approximation techniques that predict probability
-functions can be analyzed through a variety of summary
-statistics. This work uses the max absolute difference, also known as
-the Kolmogorov-Smirnov (KS) statistic \cite{lilliefors1967kolmogorov}
-for its compatibility with the KS test.
-
-The two-sample KS test is a useful nonparametric test for comparing
-two CDFs while only assuming stationarity, finite mean, and finite
-variance. The null hypothesis (that two CDFs come from the same
-underlying distribution) is rejected at level $p \in [0,1]$ when
- $$ KS > \sqrt{-\frac{1}{2}\ln\biggl(\frac{p}{2}\biggr)} \sqrt{\frac{1}{n_1} + \frac{1}{n_2}}, $$
-with distribution sample sizes $n_1,n_2 \in \mathcal{N}$. For all
-applications of the KS test presented in this work $n_1 = n_2$. An
-example of the round-trip prediction methodology from known and
-predicted distributions to the calculation of error can be seen in
-Figure \ref{fig:prediction-example}.
-
-\section{Theoretical Error Bound}
-\label{sec:theory}
-
-This section presents the theoretical results bounding the error of
-(piecewise) linear interpolation. The error analysis relies on linear
-interpolation for three reasons: (1) second order results can be
-obtained utilizing a Lipschitz constant on the gradient of a function,
-rather than standard Lipschitz bounds; (2) the results directly apply
-to Delaunay interpolation; and (3) multiple other interpolants in this
-paper compute predictions as convex combinations of observed function
-values, which may allow for straight forward extensions of this error
-bound.
-
-\begin{plainlemma}
-  \label{lemma:1}
-  Let $S \subset \mathbb{R}^d$ be open and convex, $f: \mathbb{R}^d
-  \rightarrow \mathbb{R}$, and $\nabla f \in
-  Lip_{(\gamma,\|\cdot\|_2)}(S)$, the set of $\gamma$-Lipschitz
-  continuous functions in the $2$-norm. Then for all $x,y \in S$
-  $$\big|f(y) - f(x) - \langle \nabla f(x), y - x \rangle \big| \leq \frac{\gamma \|y - x\|_2^2}{2}.$$
-\end{plainlemma}
-
-\begin{proofdot}
-  Consider the function $g(t) = f \big((1-t) x + t y \big)$, $0 \leq t
-  \leq 1$, whose derivative $g'(t) = \big\langle \nabla f \big((1-t) x
-  + t y \big), y - x \big\rangle$ is the directional derivative of $f$
-  in the direction $(y - x).$
-  \begin{align*}
-    \big|f(y) - f(x) - \langle \nabla f(x), y - x \rangle \big|
-       &= \big|g(1) - g(0) - g'(0) \big| & \\
-       &= \bigg| \int_0^1 g'(t) - g'(0)\ dt \bigg| \\
-       &\leq \int_0^1 \big|g'(t) - g'(0)\big|\ dt \\
-       &= \int_0^1 \bigg| \big \langle \nabla f\big((1-t)x + ty\big) - \nabla f(x), y - x \big \rangle \bigg|\ dt \\
-       &\leq \int_0^1 \big \| \nabla f\big((1-t)x + ty\big) - \nabla f(x) \big \|_2\ \| y - x \|_2\ dt \\
-       &\leq \int_0^1 \big ( \gamma\ \|y-x\|_2 \big) \ \big( \|y-x\|_2 \big) t\ dt \\
-       &= \frac{\gamma \|y - x\|_2^2}{2}.
-  \end{align*}
-  \qed
-\end{proofdot}
-
-\begin{plainlemma}
-  \label{lemma:2}
-  Let $x, y, v_i \in \mathbb{R}^d$, $c_i \in \mathbb{R}$, and
-  $|\langle y - x, v_i \rangle| \leq c_i$ for $i = 1$, $\ldots$, $d.$
-  If $M = (v_1$, $\ldots$, $v_d)$ is nonsingular, then
-  $$\|y - x\|_2^2 \leq \frac{1}{\sigma_d^2} \sum_{i=1}^d c_i^2,$$
-  where $\sigma_d$ is the smallest singular value of $M.$
-\end{plainlemma}
-
-\begin{proofdot}
-  Using the facts that $M$ and $M^t$ have the same singular values,
-  and $\|M^tw\|_2 \geq \sigma_d \|w\|_2$, gives
-  \begin{align*}
-    \|y - x\|_2^2 &\leq \frac{\|M^t (y - x)\|_2^2}{\sigma_d^2} \\
-                  &=    \frac{1}{\sigma_d^2} \sum_{i=1}^d \langle y - x, v_i \rangle^2 \\
-                  &\leq \frac{1}{\sigma_d^2} \sum_{i=1}^d c_i^2.
-  \end{align*}
-  \qed
-\end{proofdot}
-
-\begin{plainlemma}
-  \label{lemma:3}
-  Given $f$, $\gamma$, $S$ as in {\it Lemma \ref{lemma:1}}, let $X =
-  \{x_0$, $x_1$, $\ldots$, $x_d\}$ $\subset S$ be the vertices of a
-  $d$-simplex, and let $\hat f(x) = \langle c, x - x_0 \rangle +
-  f(x_0)$, $c \in \mathbb{R}^d$ be the linear function interpolating
-  $f$ on $X.$ Let $\sigma_d$ be the smallest singular value of the
-  matrix $M = (x_1 - x_0$, $\ldots$, $x_d - x_0)$, and $k =
-  \max\limits_{1\ \leq\ j\ \leq\ d} \|x_j - x_0\|_2.$ Then
-  $$\big\|\nabla f(x_0) - c\big\|_2 \leq \sqrt{d} \frac{\gamma k^2}{\sigma_d}.$$
-\end{plainlemma}
-
-\begin{proofdot}
-  Consider $f(x) - \hat f(x)$ along the line segment $z(t) = (1-t)x_0
-  + t x_j$, $0 \leq t \leq 1.$ By Rolle's Theorem, for some $0 < \hat
-  t < 1$, $\big\langle \nabla f\big(z(\ \hat t\ )\big) - c, x_j - x_0
-  \big\rangle = 0.$ Now
-  \begin{align*}
-    \big| \big< \nabla f(x_0) - c, x_j - x_0 \big \rangle \big|
-        &= \big| \big< \nabla f(x_0) - \nabla f\big(z(\ \hat t\ )\big) + \nabla f\big(z(\ \hat t\ )\big) - c, x_j - x_0 \big \rangle \big| \\
-	&= \big| \big\langle \nabla f(x_0) - \nabla f \big(z(\ \hat t\ )\big), x_j - x_0 \big \rangle \big| \\
-	&\leq \big \| \nabla f(x_0) - \nabla f\big(z(\ \hat t\ )\big) \big \|_2 \|x_j - x_0\|_2 \\
-	&\leq \gamma \|x_0 - z(\ \hat t\ )\|_2\ \|x_j - x_0\|_2 \\
-	&\leq \gamma \|x_j - x_0\|_2^2 \leq \gamma k^2, \\
-  \end{align*}
-  for all $1 \leq j \leq d.$ Using {\it Lemma \ref{lemma:2}},
-  $$\big \| \nabla f(x_i) - c \big \|_2^2 \leq \frac{d}{\sigma_d^2}
-  \big( \gamma k^2\big)^2 \implies \big \| \nabla f(x_i) - c \big \|_2
-  \leq \sqrt{d} \frac{\gamma k^2}{\sigma_d}.$$
-  \qed
-\end{proofdot}
-
-\begin{onetheorem}
-  Under the assumptions of {\it Lemma \ref{lemma:1}} and {\it Lemma
-    \ref{lemma:3}}, for $z \in S$,
-  $$ \big|f(z) - \hat f(z)\big| \leq \frac{\gamma \|z - x_0\|_2^2}{2} + \sqrt{d} \frac{\gamma k^2}{\sigma_d} \|z - x_0\|_2.$$
-\end{onetheorem}
-
-\begin{proofdot}
-  Let $v = \nabla f(x_0) - c$, where $\|v\|_2 \leq \sqrt{d} \gamma k^2 / \sigma_d$ by {\it Lemma \ref{lemma:3}}. Now
-  \begin{align*}
-    \big|f(z) - \hat f(z)\big|
-       &= \big|f(z) - f(x_0) - \langle c, z - x_0 \rangle \big| \\ 
-       &= \big|f(z) - f(x_0) - \langle \nabla f(x_0) - v, z - x_0 \rangle \big| \\
-       &= \big|f(z) - f(x_0) - \langle \nabla f(x_0) , z - x_0 \rangle + \langle v , z - x_0 \rangle \big| \\
-       &\leq \big|f(z) - f(x_0) - \langle \nabla f(x_0) , z - x_0 \rangle \big| + \big| \langle v , z - x_0 \rangle \big| \\
-       &\leq \big|f(z) - f(x_0) - \langle \nabla f(x_0) , z - x_0 \rangle \big| + \|v\|_2 \|z - x_0\|_2 \\
-       &\leq \big|f(z) - f(x_0) - \langle \nabla f(x_0), z - x_0 \rangle \big| + \textstyle{\frac{\gamma k^2 \sqrt{d}}{\sigma_d}} \|z - x_0\|_2 \\
-       &\leq \frac{\gamma \|z - x_0\|_2^2}{2} + \sqrt{d}\frac{\gamma k^2}{\sigma_d} \|z - x_0\|_2,
-  \end{align*}
-  where the last inequality follows from {\it Lemma \ref{lemma:1}}.
-  %
-  \qed
-\end{proofdot}
-
-In summary, the approximation error of a linear (simplicial)
-interpolant tends quadratically towards zero when approaching observed
-data only when the diameter of the simplex is also reduced at a
-proportional rate.  Only linear convergence to the true function can
-be achieved in practice, without the incorporation of additional
-observations. Notice that the approximation error is largely
-determined by the spacing of observed data. Predictions made by
-simplices whose vertices are not well-spaced (i.e., have large
-diameter, or are nearly contained in a hyperplane) have higher
-error. In light of this error bound, an empirical evaluation of
-presented algorithms follows.
-
-%% ===================================================================
-\vspace{-2mm}
-\section{Data and Empirical Analysis}
-\label{sec:data}
-
-The theoretical results constructed in Section \ref{sec:theory} for
-(piecewise) linear interpolation are promising and apply directly to
-Delaunay interpolation, however they are difficult to interpret in
-context with approximation algorithms that do not have similar known
-uniform error bounds. For that reason, this paper utilizes five
-different data sets of varying dimension and application to construct
-approximations and compare the accuracy of different techniques.
-
-In the following five subsections the sources and targets of each test
-data set are described, as well as challenges and limitations related
-to approximating these data. The distribution of response values being
-modeled is presented followed by the distribution of approximation
-errors for each algorithm. The plots for all five data sets have the
-same format.
-
-All five data sets are rescaled such that the domain of approximation
-is the unit hypercube. The range of the first four data sets is the
-real numbers, while the range of the fifth data set is the space of
-cumulative distribution functions. All approximation techniques are
-applied to the first four data sets, while only those interpolants
-whose approximations are convex combinations of observed data are
-applied to the final data set.
-
-All approximations are constructed using $k$-fold cross validation as
-described in \cite{kohavi1995study} with $k=10$. This approach
-randomly partitions data into $k$ (nearly) equal sized sets. Each
-algorithm is then evaluated by constructing an approximation over each
-unique union of $k-1$ elements of the partition, making predictions
-for points in the remaining element. As a result, each observed data
-point is used in the construction of $k-1$ different approximations
-and is approximated exactly once. The $k$-fold cross validation method
-is data-efficient and provides an unbiased estimate of the expected
-prediction error \cite{kohavi1995study}, however it should be noted
-that neither this method nor others can provide a universally unbiased
-estimator for the variance of prediction error \cite{bengio2004no}.
-
-In addition to the figures displaying approximation results for each
-data set, tables of accompanying numerical results are located in the
-Appendix. All of the test data sets capture underlying functions that
-are almost certainly stochastic. As described in Section
-\ref{sec:introduction}, regression techniques appear most appropriate
-for these problems. However, typically data grows exponentially more
-sparse with increasing dimension. Given sparse data, regressors tend
-towards interpolation.
-
-%% -------------------------------------------------------------------
-\subsection{Forest Fire ($n = 504$, $d = 12$)}
-
-\begin{figure}
-  \centering
-  \begin{tikzpicture}
-    \node (img)  {\includegraphics[width=0.75\textwidth,]{Figures/raw-histogram-forest-fire.pdf}};
-    \node[below=of img, node distance=1cm, yshift=1cm] {Forest Fire Area Burned};
-    \node[left=of img, node distance=0cm, rotate=90, anchor=center,yshift=-0.7cm] {Count};
-  \end{tikzpicture}
-  \caption{Histogram of forest fire area burned under recorded weather
-    conditions. The data is presented on a $\ln$ scale because
-    most values are small with exponentially fewer fires on record
-    that burn large areas.}
-  \label{fig:hist-forest-fire}
-\end{figure}
-
-\begin{figure}
-  \centering
-  \begin{tikzpicture}
-    \node (img)  {\includegraphics[width=0.8\textwidth,]{Figures/error-box-forest-fire.pdf}};
-    \node[below=of img, node distance=1cm, yshift=1cm] {Area Burned Error};
-  \end{tikzpicture}
-  \caption{All models are applied to approximate the amount of area
-    that would be burned given environment conditions. $10$-fold cross
-    validation as described in the beginning of Section \ref{sec:data}
-    is used to evaluated each algorithm. This results in exactly one
-    prediction from each algorithm for each data point. These boxes
-    depict the median (middle bar), median $95\%$ confidence interval
-    (cones), quartiles (box edges), fences at $3/2$ interquartile
-    range (whiskers), and outliers (dots) of absolute prediction error
-    for each model. Similar to Figure \ref{fig:hist-forest-fire}, the
-    errors are presented on a $\ln$ scale. The numerical data
-    corresponding to this figure is provided in Table
-    \ref{table:error-forest-fire} in the Appendix.}
-  \label{fig:error-forest-fire}
-\end{figure}
-
-The forest fire data set \cite{cortez2007data} describes the area of
-Montesinho park burned over months of the year along with
-environmental conditions. The twelve dimensions being used to model
-burn area are the $x$ and $y$ spatial coordinates of burns in the
-park, month of year (mapped to $x$, $y$ coordinates on a unit circle),
-the FFMC, DMC, DC, and ISI indices (see source for details), the
-temperature, relative humidity, wind speed, and outdoor rain. The
-original analysis of this data set demonstrated it to be difficult to
-model, likely due to the skew in response values.
-
-As suggested by Figure \ref{fig:error-forest-fire}, the SVR has the
-lowest absolute prediction errors for $80\%$ of the data, with MLP and
-Delaunay being the nearest overall competitors. The effectiveness of
-SVR on this data suggests the underlying function can be defined by
-relatively few parameters, as well as the importance of capturing the
-low-burn-area data points.
-%% -------------------------------------------------------------------
-
-%% -------------------------------------------------------------------
-\subsection{Parkinson's Telemonitoring ($n = 5875$, $d = 19$)}
-
-\begin{figure}
-  \centering
-  \begin{tikzpicture}
-    \node (img)  {\includegraphics[width=0.75\textwidth,]{Figures/raw-histogram-parkinsons.pdf}};
-    \node[below=of img, node distance=1cm, yshift=1cm] {Parkinson's Total UPDRS Score};
-    \node[left=of img, node distance=0cm, rotate=90, anchor=center,yshift=-0.7cm] {Count};
-  \end{tikzpicture}
-  \caption{Histogram of the Parkinson's patient total UPDRS clinical
-    scores that will be approximated by each algorithm.}
-  \label{fig:hist-parkinsons}
-\end{figure}
-
-\begin{figure}
-  \centering
-  \begin{tikzpicture}
-    \node (img)  {\includegraphics[width=0.8\textwidth,]{Figures/error-box-parkinsons.pdf}};
-    \node[below=of img, node distance=1cm, yshift=1cm] {Total UPDRS Score Error};
-  \end{tikzpicture}
-  \caption{All models are applied to approximate the total UPDRS score
-    given audio features from patients' home life, using $10$-fold
-    cross validation. These boxes depict the median (middle bar),
-    median $95\%$ confidence interval (cones), quartiles (box edges),
-    fences at $3/2$ interquartile range (whiskers), and outliers
-    (dots) of absolute prediction error for each model. The numerical
-    data corresponding to this figure is provided in Table
-    \ref{table:error-parkinsons} in the Appendix.}
-  \label{fig:error-parkinsons}
-\end{figure}
-
-The second data set for evaluation \cite{tsanas2010accurate} is
-derived from a speech monitoring study with the intent to
-automatically estimate Parkinson's disease symptom development in
-Parkinson's patients. The function to be approximated is a
-time-consuming clinical evaluation measure referred to as the UPDRS
-score. The total UPDRS score given by a clinical evaluation is
-estimated through 19 real numbers generated from biomedical voice
-measures of in-home sound recordings.
-
-Figure \ref{fig:error-parkinsons} shows the ShepMod algorithm has the
-lowest minimum, first quartile, and median of absolute errors for this
-problem, while providing the best prediction $66\%$ of the time. The
-MLP has the lowest third quartile and provides the best prediction for
-$32\%$ of approximations. The dominance of ShepMod may be due in part
-to regular-interval total UPDRS scores provided by clinicians,
-favoring a nearest-neighbor strategy of prediction.
-%% -------------------------------------------------------------------
-
-%% -------------------------------------------------------------------
-\subsection{Australian Daily Rainfall Volume ($n = 2609$, $d = 23$)}
-
-\begin{figure}
-  \centering
-  \begin{tikzpicture}
-    \node (img)  {\includegraphics[width=0.75\textwidth,]{Figures/raw-histogram-weather.pdf}};
-    \node[below=of img, node distance=1cm, yshift=1cm] {Daily Rainfall in Sydney};
-    \node[left=of img, node distance=0cm, rotate=90, anchor=center,yshift=-0.7cm] {Count};
-  \end{tikzpicture}
-  \caption{Histogram of daily rainfall in Sydney, Australia, presented
-    on a $\ln$ scale because the frequency of larger amounts of
-    rainfall is significantly less. There is a peak in occurrence of
-    the value $0$, which has a notable effect on the resulting model
-    performance.}
-  \label{fig:hist-weather}
-\end{figure}
-
-\begin{figure}
-  \centering
-  \begin{tikzpicture}
-    \node (img)  {\includegraphics[width=0.8\textwidth,]{Figures/error-box-weather.pdf}};
-    \node[below=of img, node distance=1cm, yshift=1cm] {Sydney Rainfall Tomorrow Error};
-  \end{tikzpicture}
-  \caption{All models are applied to approximate the amount of
-    rainfall expected on the next calendar day given various sources
-    of local meteorological data, using $10$-fold cross validation.
-    These boxes depict the median (middle bar), median $95\%$
-    confidence interval (cones), quartiles (box edges), fences at
-    $3/2$ interquartile range (whiskers), and outliers (dots) of
-    absolute prediction error for each model. The errors are presented
-    on a $\ln$ scale, mimicking the presentation in Figure
-    \ref{fig:hist-weather}. The numerical data corresponding to this
-    figure is provided in Table \ref{table:error-weather} in the
-    Appendix.}
-  \label{fig:error-weather}
-\end{figure}
-
-The third data set for the total daily rainfall in Sydney, Australia
-\cite{williams2009rattle} provides a slightly higher dimensional
-challenge for the interpolants and regressors. This data is composed
-of many meteorological readings from the region in a day including:
-min and max temperatures, sunshine, wind speed directions (converted
-to coordinates on a circle), wind speeds, and humidities throughout
-the day, day of the year (converted to coordinates on a circle), and
-the model must predict the amount of rainfall tomorrow.
-
-While Figure \ref{fig:hist-weather} makes MARS look far better than
-other techniques, it only provides the best prediction for $11\%$ of
-points. The MLP has the lowest absolute error for $56\%$ of points and
-LSHEP is best for $28\%$. MARS likely achieves such a low first
-quartile due to the high occurrence of the value zero in the data.
-%% -------------------------------------------------------------------
-
-%% -------------------------------------------------------------------
-\subsection{Credit Card Transaction Amount ($n = 5562$, $d = 28$)}
-
-\begin{figure}
-  \centering
-  \begin{tikzpicture}
-    \node (img)  {\includegraphics[width=0.75\textwidth,]{Figures/raw-histogram-credit-card.pdf}};
-    \node[below=of img, node distance=1cm, yshift=1cm] {Credit Card Transaction Amount};
-    \node[left=of img, node distance=0cm, rotate=90, anchor=center,yshift=-0.7cm] {Count};
-  \end{tikzpicture}
-  \caption{Histogram of credit card transaction amounts, presented on
-    a $\ln$ scale. The data contains a notable frequency peak around
-    $\$1$ transactions. Fewer large purchases are made, but some large
-    purchases are in excess of five orders of magnitude greater than
-    the smallest purchases.}
-  \label{fig:hist-credit-card}
-\end{figure}
-
-\begin{figure}
-  \centering
-  \begin{tikzpicture}
-    \node (img)  {\includegraphics[width=0.8\textwidth,]{Figures/error-box-credit-card.pdf}};
-    \node[below=of img, node distance=1cm, yshift=1cm] {Transaction Amount Error};
-  \end{tikzpicture}
-  \caption{All models are applied to approximate the expected
-    transaction amount given transformed (and obfuscated) vendor and
-    customer-descriptive features, using $10$-fold cross validation.
-    These boxes depict the median (middle bar), median $95\%$
-    confidence interval (cones), quartiles (box edges), fences at
-    $3/2$ interquartile range (whiskers), and outliers (dots) of
-    absolute prediction error for each model. The absolute errors in
-    transaction amount predictions are presented on a $\ln$ scale,
-    just as in Figure \ref{fig:hist-credit-card}. The numerical data
-    corresponding to this figure is provided in Table
-    \ref{table:error-credit-card} in the Appendix.}
-  \label{fig:error-credit-card}
-\end{figure}
-
-The fourth test data set, and the final with a real-valued range, is a
-collection of credit card transactions
-\cite{pozzolo2015calibrating}. The provided data carries no direct
-real-world meaning, being the output of a principle component analysis
-on the original hidden source data. This obfuscation is done to
-protect the information of the credit card users. This data has the
-largest dimension of all considered, at 28. A model for this data
-predicts the transaction amount given the vector of principle
-component information.
-
-As can be seen in Figure \ref{fig:error-credit-card}, the MLP
-outperforms all other algorithms at the first, second, third, and
-fourth quartiles. The MLP produces the lowest absolute error
-prediction for $80\%$ of transactions, Delaunay bests the remaining
-$20\%$. It is likely that with less data, Delaunay would be the best
-performer.
-%% -------------------------------------------------------------------
-
-%% -------------------------------------------------------------------
-\subsection{High Performance Computing I/O ($n = 3016$, $d = 4$)}
-
-\begin{figure}
-  \centering
-  \begin{tikzpicture}
-    \node (img)  {\includegraphics[width=0.75\textwidth,]{Figures/raw-histogram-throughput.pdf}};
-    \node[below=of img, node distance=1cm, yshift=1cm] {I/O Read Throughtput};
-    \node[left=of img, node distance=0cm, rotate=90, anchor=center,yshift=-0.7cm] {Count};
-  \end{tikzpicture}
-  \caption{Histogram of the raw throughput values recorded during all
-    IOzone tests across all system configurations. The distribution is
-    skewed right, with few tests having significantly higher
-    throughput than most others. The data is presented on a $\ln$
-    scale.}
-  \label{fig:hist-throughput}
-\end{figure}
-
-\begin{figure}
-  \centering
-  \begin{tikzpicture}
-    \node (img)  {\includegraphics[width=0.8\textwidth,]{Figures/error-box-throughput.pdf}};
-    \node[below=of img, node distance=1cm, yshift=1cm] {Read Throughtput KS Statistic};
-  \end{tikzpicture}
-  \caption{The models directly capable of predicting distributions are
-    applied to predicting the expected CDF of I/O throughput at a
-    previously unseen system configuration, using $10$-fold cross
-    validation. The KS statistic between the observed distribution at
-    each system configuration and the predicted distribution is
-    recorded and presented above. Note that the above figure is
-    \textit{not} log-scaled like Figure \ref{fig:hist-throughput}. The
-    numerical data corresponding to this figure is provided in Table
-    \ref{table:error-throughput} in the Appendix.}
-  \label{fig:error-throughput}
-\end{figure}
-
-The final of five data sets is derived from \cite{cameron2019moana},
-which provides four-dimensional distribution data by executing the
-IOzone benchmark \cite{iozone} on a computer system and varying the
-system's file size, record size, thread count, and CPU frequency. At
-each configuration, IOzone samples the I/O file-read throughput (in
-bytes per second) 150 times. Empirical distribution function points
-are computed from each set of 150 executions, which are interpolated
-with a piecewise cubic Hermite interpolating polynomial
-\cite{fritsch1980monotone} to approximate the CDF. All interpolation
-algorithms with the exception of LSHEP are used to approximate these
-CDFs from system configurations.
-
-Delaunay achieves the lowest KS statistic for $62\%$ of
-approximations, while Voronoi is best for the remaining $38\%$. Figure
-\ref{fig:error-throughput} shows that while Delaunay may have more
-best predictions, the behavior of Voronoi may be preferable.
-%% -------------------------------------------------------------------
-
-%% -------------------------------------------------------------------
-\begin{figure}
-  \centering
-  \begin{tikzpicture}
-    \node (img)  {\includegraphics[width=0.8\textwidth]{Figures/plot-throughput-ks.pdf}};
-    \node[below=of img, node distance=1cm, yshift=1cm] {KS Statistic for Predicted vs. Actual};
-    \node[left=of img, node distance=0cm, rotate=90, anchor=center,yshift=-0.7cm] {Count of KS Statistic};
-  \end{tikzpicture}
-  \caption{Histograms of the prediction error for each interpolant
-    that produces predictions as convex combinations of observed data,
-    using $10$-fold cross validation. The histograms show the KS
-    statistics for the predicted throughput distribution versus the
-    actual throughput distribution. The four vertical lines represent
-    cutoff KS statistics given $150$ samples for commonly used
-    $p$-values 0.05, 0.01, 0.001, $10^{-6}$, respectively. All
-    predictions to the right of a vertical line represent CDF
-    predictions that are significantly different (by respective
-    $p$-value) from the actual distribution according to the KS
-    test. The numerical counterpart to this figure is presented in
-    Table \ref{table:null-hypothesis-results}.}
-  \label{fig:throughput-prediction}
-\end{figure}
-
-\begin{table}
-  \renewcommand{\arraystretch}{1.3}
-  \centering
-  \begin{tabular}{c|c|c|c|c}
-     & $p = .05$ & $p = .01$ & $p = .001$ & $p = 10^{-6}$\\
-    \hline
-    Delaunay & $\mathbf{50.3}\%$ & $\mathit{43.5}\%$ & $\mathit{36.2}\%$ & $\mathit{24.7}\%$\\
-    ShepMod & $\mathit{51.4}\%$ & $44.8\%$ & $38.1\%$ & $27.7\%$\\
-    Voronoi & $52.6\%$ & $\mathbf{43.4}\%$ & $\mathbf{34.4}\%$ & $\mathbf{19.1}\%$\\
-    BoxSpline & $99.5\%$ & $98.5\%$ & $96.6\%$ & $89.5\%$\\
-  \end{tabular}
-  \caption{Numerical counterpart of the histogram data presented in
-    Figure \ref{fig:throughput-prediction}. The columns display the
-    percent of null hypothesis rejections by the KS-test when provided
-    different selections of $p$-values for each algorithm. The
-    algorithm with the lowest rejection rate at each $p$ is boldface,
-    while the second lowest is italicized.}
-  \label{table:null-hypothesis-results}
-\end{table}
-
-%% ===================================================================
-
-Figure \ref{fig:throughput-prediction} expands on the KS statistic
-results presented in Figure \ref{fig:error-throughput}. Agglomerate
-errors for each algorithm resemble a Gamma distribution. The
-percentages of significant prediction errors with varying $p$-values
-are on display in Table \ref{table:null-hypothesis-results}. When
-considering the $p=0.001$ results for each technique, a little over
-one third of the predicted CDFs are significantly different from the
-measured (and presumed) correct CDFs. However, it should be noted that
-with 150 samples, the error of an empirical distribution function
-(EDF) can reasonably be upwards of $.1$, which serves as a rough
-estimate for the lower limit of achievable error. Globally, only a
-third of Voronoi predictions fail to capture \textit{all} of the
-characteristics of the CDFs at new system configurations.
-
-\section{Discussion}
-\label{sec:discussion}
-
-\begin{table}
-  \centering
-  \begin{tabular}{c|r@{.}l|r@{.}l|r@{.}l}
-    Algorithm & \multicolumn{2}{c|}{Avg. \% Best} & \multicolumn{2}{c|}{\multilinecell{Avg. Fit or \\ Prep. Time (s)}} & \multicolumn{2}{c}{Avg. App. Time (s)}\\
-    \hline
-    MARS & \quad\quad$4$&$6$ & \,\quad$20$&$0$ & \quad\quad$0$&$001$\\
-    SVR & $\mathit{21}$&$\mathit{1}$ & $\mathit{0}$&$\mathit{5}$ & $\mathbf{0}$&$\mathbf{0001}$\\
-    MLP & $\mathbf{42}$&$\mathbf{1}$ & $200$&$0$ & $0$&$001$\\
-    Delaunay & $6$&$0$ & $1$&$0$ & $1$&$0$\\
-    ShepMod & $18$&$4$ & $0$&$7$ & $\mathbf{0}$&$\mathbf{0001}$\\
-    LSHEP & $8$&$4$ & $2$&$0$ & $\mathbf{0}$&$\mathbf{0001}$\\
-    Voronoi & $0$&$5$ & $\mathbf{0}$&$\mathbf{1}$ & $0$&$04$\\
-    BoxSpline & $0$&$2$ & $0$&$8$ & $\mathit{0}$&$\mathit{0005}$\\
-  \end{tabular}
-  \caption{This average of Appendix Tables
-    \ref{table:best-forest-fire}, \ref{table:best-parkinsons},
-    \ref{table:best-weather}, and \ref{table:best-credit-card}
-    provides a gross summary of overall results. The columns display
-    (weighted equally by data set, \textit{not} points) the average
-    frequency with which any algorithm provided the lowest absolute
-    error approximation, the average time to fit/prepare, and the
-    average time required to approximate one point. The times have
-    been rounded to one significant digit, as reasonably large
-    fluctuations may be observed due to implementation hardware.
-    Interpolants provide the lowest error approximation for nearly one
-    third of all data, while regressors occupy the other two
-    thirds. This result is obtained without any customized tuning or
-    preprocessing to maximize the performance of any given
-    algorithm. In practice, tuning and preprocessing may have large
-    effects on approximation performance.}
-  \label{table:avg-performance}
-\end{table}
-
-Table \ref{table:avg-performance} summarizes results across the four
-test data sets with real-valued ranges. The interpolants discussed in
-this paper produce the \textit{best} approximations roughly one third
-of the time, and produce competitive approximations for almost all
-data sets. These test problems are almost certainly
-\textit{stochastic} in nature, but the high dimension leads to greater
-data sparsity and model construction cost, making interpolation more
-competitive.
-
-The major advantages to interpolation lie in the near absence of
-\textit{fit} time. Delaunay, LSHEP, and ShepMod all require pairwise
-distance calculations, for numerical robustness (Delaunay) and to
-determine the radii of influence for data points (LSHEP and
-ShepMod). At least hundreds, and sometimes hundreds of thousands of
-predictions can be made by the interpolants before the most widely
-used regressor (MLP) finishes fitting these relatively small data
-sets.  However, the computational complexities of all interpolants
-presented are greater than linear in either dimension or number of
-points, whereas the regressors' nonlinear complexity in dimension
-generally comes from the model fitting optimization.
-
-The new theoretical results presented in Section \ref{sec:theory}
-directly apply to Delaunay interpolation, however the performance of
-Delaunay does not appear significantly better than other algorithms on
-these data sets.  This observation may be due to the stochastic nature
-of the data, but it also speaks to the power of the approximations
-generated by the different interpolation methods.  The strong
-performance of other interpolants suggests that theoretical results
-similar to those presented in this work can be achieved for the other
-interpolants under reasonable assumptions.
-
-Finally, most of the interpolants presented in this work benefit from
-the ability to model any function over real $d$-tuples with a range
-that is closed under convex combinations. The results of the
-distribution prediction case study indicate that interpolants can
-effectively predict CDFs. The error analysis for that work relies on
-the KS statistic, which captures the worst part of any prediction and
-hence provides a conservatively large estimate of approximation
-error. The average absolute errors in the predicted CDFs are always
-lower than the KS statistics. However, the KS statistic was chosen as
-a metric because of the important surrounding statistical theory. A
-nonnegligible volume of predictions provide impressively low levels of
-average absolute error in that final case study.
-
-\section{Conclusion}
-\label{sec:conclusion}
-
-The major contributions of this work are: 1) new uniform theoretical
-error bounds for piecewise linear interpolation in arbitrary dimension
-(Section \ref{sec:theory}); 2) an empirical evaluation across
-real-world problems (Section \ref{sec:data}) that demonstrates
-interpolants produce competitively accurate models of multivariate
-phenomenon when compared with common regressors for sparse, moderately
-high dimensional problems; and 3) a demonstration that some
-interpolants generalize to interpolation in function spaces,
-preserving monotonicity (with CDFs, e.g.), neither of which common
-regressors can do.
-
-The various interpolants discussed in this paper have been
-demonstrated to effectively approximate multivariate phenomena up to
-$30$ dimensions. The underlying constructions are theoretically
-straightforward, interpretable, and yield reasonably accurate
-predictions. Most of the interpolants' computational complexities make
-them particularly suitable for applications in even higher
-dimension. The major benefits of interpolation are seen when only a
-small number of approximations ($\leq 1000$) are made from data and
-when there are relatively few data points for the dimension (for
-empirical results presented, $\log_d n \leq 5$).
-
-%\begin{acknowledgements}
-%If you'd like to thank anyone, place your comments here
-%and remove the percent signs.
-%\end{acknowledgements}
-
-%%====================================================================
-%%====================================================================
-
-\bibliographystyle{spmpsci}      % mathematics and physical sciences
-\bibliography{paper}
-
-\newpage
-
-\begin{appendix}
-\section*{Appendix}
-\label{sec:appendix}
-
-\vspace*{\fill}
-
-\begin{table}[H]
-  \centering
-  \begin{tabular}{c|r@{.}l|r@{.}l|r@{.}l|r@{.}l|r@{.}l}
-    \hline
-    Algorithm & \multicolumn{2}{c|}{Min} & \multicolumn{2}{c|}{$25^{th}$} & \multicolumn{2}{c|}{$50^{th}$} & \multicolumn{2}{c|}{$75^{th}$} & \multicolumn{2}{c}{Max}\\
-    \hline
-    MARS & $0$&$00984$ & $3$&$11$ & $7$&$01$ & $\mathit{11}$&$\mathit{7}$ & $1090$&$0$\\
-    SVR & $\mathit{0}$&$\mathit{00402}$ & $\mathbf{0}$&$\mathbf{243}$ & $\mathbf{0}$&$\mathbf{416}$ & $\mathbf{6}$&$\mathbf{19}$ & $1090$&$0$\\
-    MLP & $0$&$0426$ & $2$&$63$ & $\mathit{5}$&$\mathit{27}$ & $14$&$0$ & $1090$&$0$\\
-    Delaunay & $\mathbf{0}$&$\mathbf{00}$ & $1$&$98$ & $5$&$37$ & $13$&$1$ & $\mathit{1080}$&$\mathit{0}$\\
-    ShepMod & $\mathbf{0}$&$\mathbf{00}$ & $\mathit{1}$&$\mathit{93}$ & $6$&$27$ & $16$&$0$ & $1090$&$0$\\
-    LSHEP & $0$&$0400$ & $2$&$17$ & $8$&$87$ & $19$&$1$ & $\mathbf{1070}$&$\mathbf{0}$\\
-    Voronoi & $0$&$00982$ & $3$&$65$ & $7$&$56$ & $15$&$6$ & $1090$&$0$\\
-    BoxSpline & $\mathbf{0}$&$\mathbf{00}$ & $2$&$27$ & $5$&$91$ & $16$&$4$ & $1090$&$0$\\
-    \hline
-  \end{tabular}
-  \caption{This numerical data accompanies the visual provided in
-    Figure \ref{fig:error-forest-fire}. The columns of absolute error
-    percentiles correspond to the minimum, first quartile, median,
-    third quartile, and maximum absolute errors respectively. The
-    minimum of each column is boldface, while the second lowest value
-    is italicized. All values are rounded to three significant
-    digits.}
-  \label{table:error-forest-fire}
-\end{table}
-
-\begin{table}[H]
-  \centering
-  \begin{tabular}{|c|r@{.}l| c |r@{.}l|r@{.}l|r@{.}l|}
-    \cline{1-3}\cline{5-10}
-    Algorithm & \multicolumn{2}{c|}{\% Best} &  & \multicolumn{2}{c|}{Fit/Prep. Time (s)} & \multicolumn{2}{c|}{App. Time (s)} & \multicolumn{2}{c|}{Total App. Time (s)}\\
-    \cline{1-3}\cline{5-10}
-    MARS & \quad$\mathit{7}$&$\mathit{7}$ &  & \quad\quad$29$&$1$ & \quad$0$&$00137$ & \quad\quad\quad$0$&$0686$\\
-    SVR & $\mathbf{80}$&$\mathbf{2}$ &  & $\mathit{0}$&$\mathit{00584}$ & $\mathit{0}$&$\mathit{0000620}$ & $\mathit{0}$&$\mathit{00310}$\\
-    MLP & $0$&$0$ &  & $32$&$8$ & $0$&$000871$ & $0$&$0436$\\
-    Delaunay & $3$&$5$ &  & $0$&$0151$ & $0$&$0234$ & $1$&$18$\\
-    ShepMod & $3$&$7$ &  & $0$&$00634$ & $0$&$0000644$ & $0$&$00322$\\
-    LSHEP & $5$&$1$ &  & $0$&$0275$ & $\mathbf{0}$&$\mathbf{0000463}$ & $\mathbf{0}$&$\mathbf{00231}$\\
-    Voronoi & $0$&$0$ &  & $\mathbf{0}$&$\mathbf{000182}$ & $0$&$000396$ & $0$&$0198$\\
-    BoxSpline & $0$&$4$ &  & $0$&$00724$ & $0$&$0000978$ & $0$&$00489$\\
-    \cline{1-3}\cline{5-10}
-  \end{tabular}
-  \caption{The left above shows how often each algorithm had the
-    lowest absolute error approximating forest fire data in Table
-    \ref{table:error-forest-fire}. On the right columns are median fit
-    time of 454 points, median time for one approximation, and median
-    time approximating 50 points.}
-  \label{table:best-forest-fire}
-\end{table}
-
-\vspace*{\fill}
-\newpage
-
-\begin{table}
-  \centering
-  \begin{tabular}{c|r@{.}l|r@{.}l|r@{.}l|r@{.}l|r@{.}l}
-    \hline
-    Algorithm & \multicolumn{2}{c|}{Min} & \multicolumn{2}{c|}{$25^{th}$} & \multicolumn{2}{c|}{$50^{th}$} & \multicolumn{2}{c|}{$75^{th}$} & \multicolumn{2}{c}{Max}\\
-    \hline
-    MARS & $0$&$00948$ & $9$&$98$ & $20$&$4$ & $32$&$9$ & $863$&$0$\\
-    SVR & $0$&$00138$ & $3$&$17$ & $7$&$31$ & $12$&$6$ & $\mathbf{27}$&$\mathbf{5}$\\
-    MLP & $0$&$0000239$ & $\mathit{0}$&$\mathit{533}$ & $\mathit{1}$&$\mathit{25}$ & $\mathbf{2}$&$\mathbf{84}$ & $39$&$3$\\
-    Delaunay & $\mathit{3}$&$\mathit{72 \times 10^{-12}}$ & $1$&$20$ & $3$&$50$ & $7$&$67$ & $30$&$7$\\
-    ShepMod & $\mathbf{0}$&$\mathbf{0}$ & $\mathbf{0}$&$\mathbf{255}$ & $\mathbf{0}$&$\mathbf{908}$ & $\mathit{3}$&$\mathit{43}$ & $34$&$5$\\
-    LSHEP & $0$&$00254$ & $2$&$93$ & $7$&$16$ & $13$&$1$ & $\mathit{29}$&$\mathit{0}$\\
-    Voronoi & $\mathbf{0}$&$\mathbf{0}$ & $1$&$29$ & $3$&$52$ & $6$&$87$ & $30$&$1$\\
-    BoxSpline & $0$&$00287$ & $4$&$39$ & $9$&$10$ & $14$&$8$ & $41$&$3$\\
-    \hline
-  \end{tabular}
-  \caption{This numerical data accompanies the visual provided in
-    Figure \ref{fig:error-parkinsons}. The columns of absolute error
-    percentiles correspond to the minimum, first quartile, median,
-    third quartile, and maximum absolute errors respectively. The
-    minimum of each column is boldface, while the second lowest value
-    is italicized. All values are rounded to three significant
-    digits.}
-  \label{table:error-parkinsons}
-\end{table}
-
-\begin{table}
-  \centering
-  \begin{tabular}{|c|r@{.}l| c |r@{.}l|r@{.}l|r@{.}l|}
-    \cline{1-3}\cline{5-10}
-    Algorithm & \multicolumn{2}{c|}{\% Best} &  & \multicolumn{2}{c|}{Fit/Prep. Time (s)} & \multicolumn{2}{c|}{App. Time (s)} & \multicolumn{2}{c|}{Total App. Time (s)}\\
-    \cline{1-3}\cline{5-10}
-    MARS & \quad\,$0$&$0$ &  & \quad\quad\quad$37$&$9$ & \quad\,\,$0$&$00253$ & \quad\quad\quad\,\,$1$&$48$\\
-    SVR & $0$&$1$ &  & $\mathit{0}$&$\mathit{859}$ & $\mathit{0}$&$\mathit{000181}$ & $\mathit{0}$&$\mathit{106}$\\
-    MLP & $\mathit{32}$&$\mathit{0}$ &  & $348$&$0$ & $0$&$00111$ & $0$&$653$\\
-    Delaunay & $0$&$0$ &  & $2$&$47$ & $1$&$22$ & $720$&$0$\\
-    ShepMod & $\mathbf{66}$&$\mathbf{4}$ &  & $1$&$13$ & $0$&$000182$ & $0$&$107$\\
-    LSHEP & $0$&$0$ &  & $2$&$39$ & $\mathbf{0}$&$\mathbf{000144}$ & $\mathbf{0}$&$\mathbf{0845}$\\
-    Voronoi & $1$&$6$ &  & $\mathbf{0}$&$\mathbf{298}$ & $0$&$0274$ & $16$&$1$\\
-    BoxSpline & $0$&$0$ &  & $1$&$26$ & $0$&$000643$ & $0$&$377$\\
-    \cline{1-3}\cline{5-10}
-  \end{tabular}
-  \caption{The left above shows how often each algorithm had the
-    lowest absolute error approximating Parkinson's data in Table
-    \ref{table:error-parkinsons}. On the right columns are median fit
-    time of 5288 points, median time for one approximation, and median
-    time approximating 587 points.}
-  \label{table:best-parkinsons}
-\end{table}
-
-\begin{table}
-  \centering
-  \begin{tabular}{c|r@{.}l|r@{.}l|r@{.}l|r@{.}l|r@{.}l}
-    \hline
-    Algorithm & \multicolumn{2}{c|}{Min} & \multicolumn{2}{c|}{$25^{th}$} & \multicolumn{2}{c|}{$50^{th}$} & \multicolumn{2}{c|}{$75^{th}$} & \multicolumn{2}{c}{Max}\\
-    \hline
-    MARS & $\mathit{6}$&$\mathit{45 \times 10^{-15}}$ & $\mathbf{2}$&$\mathbf{70 \times 10^{-14}}$ & $1$&$66$ & $1$&$96$ & $\mathit{53}$&$\mathit{3}$\\
-    SVR & $0$&$0420$ & $0$&$143$ & $0$&$148$ & $\mathit{0}$&$\mathit{860}$ & $119$&$0$\\
-    MLP & $0$&$0000689$ & $0$&$0337$ & $\mathbf{0}$&$\mathbf{0795}$ & $\mathbf{0}$&$\mathbf{264}$ & $\mathbf{5}$&$\mathbf{31}$\\
-    Delaunay & $\mathbf{0}$&$\mathbf{0}$ & $0$&$187$ & $0$&$919$ & $2$&$72$ & $56$&$3$\\
-    ShepMod & $\mathbf{0}$&$\mathbf{0}$ & $0$&$0957$ & $0$&$685$ & $2$&$90$ & $73$&$2$\\
-    LSHEP & $\mathbf{0}$&$\mathbf{0}$ & $\mathit{0}$&$\mathit{0153}$ & $\mathit{0}$&$\mathit{106}$ & $1$&$17$ & $113$&$0$\\
-    Voronoi & $\mathbf{0}$&$\mathbf{0}$ & $0$&$430$ & $1$&$28$ & $2$&$94$ & $83$&$8$\\
-    BoxSpline & $\mathbf{0}$&$\mathbf{0}$ & $0$&$789$ & $2$&$06$ & $5$&$25$ & $119$&$0$\\
-    \hline
-  \end{tabular}
-  \caption{This numerical data accompanies the visual provided in
-    Figure \ref{fig:error-weather}. The columns of absolute error
-    percentiles correspond to the minimum, first quartile, median,
-    third quartile, and maximum absolute errors respectively. The
-    minimum value of each column is boldface, while the second lowest
-    is italicized. All values are rounded to three significant
-    digits.}
-  \label{table:error-weather}
-\end{table}
-
-\begin{table}
-  \centering
-  \begin{tabular}{|c|r@{.}l| c |r@{.}l|r@{.}l|r@{.}l|}
-    \cline{1-3}\cline{5-10}
-    Algorithm & \multicolumn{2}{c|}{\% Best} &  & \multicolumn{2}{c|}{Fit/Prep. Time (s)} & \multicolumn{2}{c|}{App. Time (s)} & \multicolumn{2}{c|}{Total App. Time (s)}\\
-    \cline{1-3}\cline{5-10}
-    MARS & \,\,\,\,$10$&$7$ &  & \quad\quad\quad$0$&$151$ & \quad$0$&$000117$ & \quad\quad\quad\,\,$0$&$0304$\\
-    SVR & $4$&$1$ &  & $\mathit{0}$&$\mathit{133}$ & $\mathbf{0}$&$\mathbf{0000947}$ & $\mathbf{0}$&$\mathbf{0246}$\\
-    MLP & $\mathbf{56}$&$\mathbf{8}$ &  & $169$&$0$ & $0$&$00137$ & $0$&$356$\\
-    Delaunay & $0$&$1$ &  & $0$&$664$ & $0$&$886$ & $230$&$0$\\
-    ShepMod & $3$&$5$ &  & $0$&$265$ & $0$&$000128$ & $0$&$0333$\\
-    LSHEP & $\mathit{28}$&$\mathit{4}$ &  & $0$&$874$ & $\mathit{0}$&$\mathit{0000975}$ & $\mathit{0}$&$\mathit{0254}$\\
-    Voronoi & $0$&$2$ &  & $\mathbf{0}$&$\mathbf{012}$ & $0$&$0270$ & $7$&$01$\\
-    BoxSpline & $0$&$3$ &  & $0$&$330$ & $0$&$000406$ & $0$&$106$\\
-    \cline{1-3}\cline{5-10}
-  \end{tabular}
-  \caption{Left table shows how often each algorithm had the lowest
-    absolute error approximating Sydney rainfall data in Table
-    \ref{table:error-weather}. On the right columns are median fit
-    time of 2349 points, median time for one approximation, and median
-    time approximating 260 points.}
-  \label{table:best-weather}
-\end{table}
-
-\begin{table}
-  \centering
-  \begin{tabular}{c|r@{.}l|r@{.}l|r@{.}l|r@{.}l|r@{.}l}
-    \hline
-    Algorithm & \multicolumn{2}{c|}{Min} & \multicolumn{2}{c|}{$25^{th}$} & \multicolumn{2}{c|}{$50^{th}$} & \multicolumn{2}{c|}{$75^{th}$} & \multicolumn{2}{c}{Max}\\
-    \hline
-    MARS & $5$&$36$ & $3610$&$0$ & $4580$&$0$ & $5450$&$0$ & $13400$&$0$\\
-    SVR & $0$&$0138$ & $8$&$47$ & $15$&$4$ & $41$&$9$ & $7700$&$0$\\
-    MLP & $0$&$00151$ & $\mathbf{1}$&$\mathbf{60}$ & $\mathbf{3}$&$\mathbf{86}$ & $\mathbf{8}$&$\mathbf{34}$ & $\mathbf{604}$&$\mathbf{0}$\\
-    Delaunay & $\mathbf{0}$&$\mathbf{0}$ & $\mathit{2}$&$\mathit{69}$ & $\mathit{13}$&$\mathit{8}$ & $\mathit{35}$&$\mathit{0}$ & $\mathit{4840}$&$\mathit{0}$\\
-    ShepMod & $\mathbf{0}$&$\mathbf{0}$ & $4$&$21$ & $17$&$3$ & $51$&$6$ & $6510$&$0$\\
-    LSHEP & $1$&$27$ & $199$&$0$ & $260$&$0$ & $343$&$0$ & $6530$&$0$\\
-    Voronoi & $\mathit{2}$&$\mathit{89 \times 10^{-10}}$ & $14$&$7$ & $32$&$8$ & $52$&$9$ & $4860$&$0$\\
-    BoxSpline & $0$&$00740$ & $20$&$8$ & $40$&$9$ & $79$&$6$ & $7640$&$0$\\
-    \hline
-  \end{tabular}
-  \caption{This numerical data accompanies the visual provided in
-    Figure \ref{fig:error-credit-card}. The columns of absolute error
-    percentiles correspond to the minimum, first quartile, median,
-    third quartile, and maximum absolute errors respectively. The
-    minimum value of each column is boldface, while the second lowest
-    is italicized. All values are rounded to three significant
-    digits.}
-  \label{table:error-credit-card}
-\end{table}
-
-\begin{table}
-  \centering
-  \begin{tabular}{|c|r@{.}l| c |r@{.}l|r@{.}l|r@{.}l|}
-    \cline{1-3}\cline{5-10}
-    Algorithm & \multicolumn{2}{c|}{\% Best} &  & \multicolumn{2}{c|}{Fit/Prep. Time (s)} & \multicolumn{2}{c|}{App. Time (s)} & \multicolumn{2}{c|}{Total App. Time (s)}\\
-    \cline{1-3}\cline{5-10}
-    MARS & \quad\,$0$&$0$ &  & \quad\quad\quad$22$&$0$ & \quad$0$&$00148$ & \quad\quad\quad\,\,$0$&$820$\\
-    SVR & $0$&$0$ &  & $\mathit{1}$&$\mathit{01}$ & $\mathit{0}$&$\mathit{000210}$ & $\mathit{0}$&$\mathit{117}$\\
-    MLP & $\mathbf{79}$&$\mathbf{5}$ &  & $290$&$0$ & $0$&$000714$ & $0$&$397$\\
-    Delaunay & $\mathit{20}$&$\mathit{5}$ &  & $3$&$12$ & $3$&$71$ & $2070$&$0$\\
-    ShepMod & $0$&$1$ &  & $1$&$45$ & $0$&$000220$ & $0$&$122$\\
-    LSHEP & $0$&$0$ &  & $3$&$47$ & $\mathbf{0}$&$\mathbf{000176}$ & $\mathbf{0}$&$\mathbf{0981}$\\
-    Voronoi & $0$&$0$ &  & $\mathbf{0}$&$\mathbf{197}$ & $0$&$0950$ & $52$&$8$\\
-    BoxSpline & $0$&$0$ &  & $1$&$66$ & $0$&$000956$ & $0$&$532$\\
-    \cline{1-3}\cline{5-10}
-  \end{tabular}
-  \caption{The left above shows how often each algorithm had the
-    lowest absolute error approximating credit card transaction data
-    in Table \ref{table:error-credit-card}. On the right columns are
-    median fit time of 5006 points, median time for one approximation,
-    and median time approximating 556 points.}
-  \label{table:best-credit-card}
-\end{table}
-
-\begin{table}
-  \centering
-  \begin{tabular}{c|r@{.}l|r@{.}l|r@{.}l|r@{.}l|r@{.}l}
-    \hline
-    Algorithm & \multicolumn{2}{c|}{Min} & \multicolumn{2}{c|}{$25^{th}$} & \multicolumn{2}{c|}{$50^{th}$} & \multicolumn{2}{c|}{$75^{th}$} & \multicolumn{2}{c}{Max}\\
-    \hline
-    Delaunay & $\mathbf{0}$&$\mathbf{0287}$ & $\mathbf{0}$&$\mathbf{0914}$ & $\mathbf{0}$&$\mathbf{158}$ & $\mathit{0}$&$\mathit{308}$ & $\mathit{0}$&$\mathit{897}$\\
-    ShepMod & $0$&$0307$ & $\mathit{0}$&$\mathit{0983}$ & $\mathit{0}$&$\mathit{164}$ & $0$&$335$ & $1$&$00$\\
-    Voronoi & $\mathit{0}$&$\mathit{0303}$ & $0$&$100$ & $0$&$165$ & $\mathbf{0}$&$\mathbf{274}$ & $\mathbf{0}$&$\mathbf{762}$\\
-    BoxSpline & $0$&$0703$ & $0$&$417$ & $0$&$561$ & $0$&$745$ & $1$&$00$\\
-    \hline
-  \end{tabular}
-  \caption{This numerical data accompanies the visual provided in
-    Figure \ref{fig:error-throughput}. The columns of absolute error
-    percentiles correspond to the minimum, first quartile, median,
-    third quartile, and maximum KS statistics respectively between
-    truth and guess for models predicting the distribution of I/O
-    throughput that will be observed at previously unseen system
-    configurations. The minimum value of each column is boldface,
-    while the second lowest is italicized. All values are rounded to
-    three significant digits.}
-  \label{table:error-throughput}
-\end{table}
-
-\begin{table}
-  \centering
-  \begin{tabular}{|c|r@{.}l| c |r@{.}l|r@{.}l|r@{.}l|}
-    \cline{1-3}\cline{5-10}
-    Algorithm & \multicolumn{2}{c|}{\% Best} &  & \multicolumn{2}{c|}{Fit/Prep. Time (s)} & \multicolumn{2}{c|}{App. Time (s)} & \multicolumn{2}{c|}{Total App. Time (s)}\\
-    \cline{1-3}\cline{5-10}
-    Delaunay & \,\,$\mathbf{62}$&$\mathbf{0}$ &  & \quad\quad\quad$0$&$344$ & \quad$0$&$00984$ & \quad\quad\,\,$2$&$71$\\
-    ShepMod & $0$&$0$ &  & $\mathbf{0}$&$\mathbf{0884}$ & $\mathbf{0}$&$\mathbf{000145}$ & $\mathbf{0}$&$\mathbf{0436}$\\
-    Voronoi & $\mathit{38}$&$\mathit{0}$ &  & $1$&$0$ & $0$&$00253$ & $0$&$762$\\
-    BoxSpline & $0$&$0$ &  & $0$&$0972$ & $\mathit{0}$&$\mathit{000210}$ & $\mathit{0}$&$\mathit{0633}$\\
-    \cline{1-3}\cline{5-10}
-  \end{tabular}
-  \caption{The left above shows how often each algorithm had the
-    lowest KS statistic on the I/O throughput distribution data in
-    Table \ref{table:error-throughput}. On the right columns are
-    median fit time of 2715 points, median time for one approximation,
-    and median time approximating 301 points.}
-  \label{table:best-throughput}
-\end{table}
-
-\end{appendix}
-
-\end{document}

### 2019-05-03 Disseration/naive.tex
Finished first full draft of preliminary defense document. Ready to share with committee.

--- 2018-03-08 Paper_[2017-12-15]_HPC/paper.tex
@@
-%**************************************************************************
-%* SpringSim 2017 Author Kit
-%*
-%* Word Processing System: TeXnicCenter and MiKTeX
-%*
-%**************************************************************************
-
-\documentclass{scspaperproc}
-
-\usepackage{latexsym}
-\usepackage{graphicx}
-\usepackage{mathptmx}
-\usepackage{amsmath}
-\usepackage{amsfonts}
-\usepackage{amssymb}
-\usepackage{amsbsy}
-\usepackage{amsthm}
-%% \usepackage{color}   % Needed to color text for drafts
-
-%% \usepackage{algorithm}     % needed for the algorithm block
-%% \usepackage{algpseudocode} % algorithmicx psuedo-code styling
-%\usepackage[pdftex,colorlinks=true,urlcolor=blue,citecolor=black,anchorcolor=black,linkcolor=black]{hyperref}
-%% \usepackage[dvips,colorlinks=true,urlcolor=blue,citecolor=black,%
-%% anchorcolor=black,linkcolor=black]{hyperref}
-
-% custom hyphenation rules
-\usepackage{hyphenat}
-\hyphenation{op-tical net-works semi-conduc-tor}
-
-% theorem style
-\newtheoremstyle{scsthe}% hnamei
-{8pt}% hSpace abovei
-{8pt}% hSpace belowi
-{\it}% hBody fonti
-{}% hIndent amounti1
-{\bf}% hTheorem head fontbf
-{.}% hPunctuation after theorem headi
-{.5em}% hSpace after theorem headi2
-{}% hTheorem head spec (can be left empty, meaning `normal')i
-\theoremstyle{scsthe}
-\newtheorem{theorem}{Theorem}
-\renewcommand{\thetheorem}{\arabic{theorem}}
-\newtheorem{corollary}[theorem]{Corollary}
-\renewcommand{\thecorollary}{\arabic{corollary}}
-\newtheorem{definition}{Definition}
-\renewcommand{\thedefinition}{\arabic{definition}}
-\newcommand{\rpm}{\raisebox{.2ex}{$\scriptstyle\pm$}}
-
-% avoid overrunning the right margin
-\sloppy
-
-%% ***** NOTE *****
-
-%% The use of the long citation format (e.g. "Brown and Edwards
-%% (1993)" rather than "[5]") and at the same time using the hyperref
-%% package can lead to hard to trace bugs in case the citation is
-%% broken accross the line (usually this will mark the entire
-%% paragraph as a hyperlink (clickable) which is easily noticeable and
-%% fixed if using colorlinks, but not if the color is black -- as it
-%% is now). Worse yet, if a citation spans page boundary, LaTeX
-%% compilation can fail, with an obscure error message. Since this
-%% depends a lot on the flow of the text and wording, these bugs come
-%% and go and can be extremely hard for a beginner to trace. The error
-%% message can look like this:
-%%
-%%    ! pdfTeX error (ext4): \pdfendlink ended up in different nesting
-%%    level than \pdfstartlink.  \AtBegShi@Output ...ipout \box
-%%    \AtBeginShipoutBox \fi \fi
-%%    l.174 
-%%    ! ==> Fatal error occurred, no output PDF file produced!
-%%
-%% and can be universally fixed by putting an \mbox{} around the
-%% citation in question (in this case, at line 174) and maybe adapting
-%% the wording a little bit to improve the paragraph typesetting,
-%% which is perhaps not immediately obvious.
-%****************************************************************************
-
-% begin document
-\begin{document}
-
-% Page header (author list)
-\SCSpagesetup{Lux, Watson, Chang, Bernard, Li, Xu, Back, Butt, Cameron, and Hong}
-
-% Conference info
-\def\SCSconferenceacro{SpringSim}
-\def\SCSpublicationyear{2018}
-\def\SCSconferencedates{April 15-18}
-\def\SCSconferencevenue{Baltimore, MD, USA}
-\def\SCSsymposiumacro{HPC} % High Performance Computing Symposium
-
-% title
-\title{Predictive Modeling of I/O Characteristics \\ in High
-  Performance Computing Systems}
-
-% AUTHOR LIST
-% *** NOTE: May need to adjust titlevboxsize in the preamble
-\author{Thomas C. H. Lux \\ [8pt]
-Dept. of Computer Science \\
-Virginia Polytechnic Institute\\
-\& State University \\
-Blacksburg, VA 24061 \\
-tchlux@vt.edu \\
-\and
-Layne T. Watson \\[8pt]
-Dept. of Computer Science\\
-Dept. of Mathematics\\
-Dept. of Aerospace \& Ocean Eng.\\ 
-Virginia Polytechnic Institute\\
-\& State University \\
-\and
-\vspace{-30pt}\\
-Tyler H. Chang\\
-Jon Bernard\\
-Bo Li\\[8pt]
-Dept. of Computer Science\\ 
-Virginia Polytechnic Institute\\
-\& State University \\
-\and
-\vspace{-30pt}\\
-Li Xu\\[8pt]
-Dept. of Statistics\\ 
-Virginia Polytechnic Institute\\
-\& State University \\
-\and
-\vspace{-30pt}\\
-Godmar Back\\
-Ali R. Butt\\
-Kirk W. Cameron\\[8pt]
-Dept. of Computer Science\\ 
-Virginia Polytechnic Institute\\
-\& State University \\
-\and
-\vspace{-30pt}\\
-Yili Hong\\[8pt]
-Dept. of Statistics\\ 
-Virginia Polytechnic Institute\\
-\& State University \\
-}
-
-\maketitle
-
-\vspace{-40pt}
-
-\section*{Abstract}
-
-Each of high performance computing, cloud computing, and computer
-security have their own interests in modeling and predicting the
-performance of computers with respect to how they are configured. An
-effective model might infer internal mechanics, minimize power
-consumption, or maximize computational throughput of a given
-system. This paper analyzes a four-dimensional dataset measuring the
-input/output (I/O) characteristics of a cluster of identical computers
-using the benchmark IOzone. The I/O performance characteristics are
-modeled with respect to system configuration using multivariate
-interpolation and approximation techniques. The analysis reveals that
-accurate models of I/O characteristics for a computer system may be
-created from a small fraction of possible configurations, and that
-some modeling techniques will continue to perform well as the number
-of system parameters being modeled increases. These results have
-strong implications for future predictive analyses based on more
-comprehensive sets of system parameters.
-
-\textbf{Keywords:} Regression, approximation, interpolation,
-performance modeling
-
-%     Introduction     
-%======================
-\section{Introduction and related work}
-\label{sec:introduction}
-
-Performance tuning is often an experimentally complex and time
-intensive chore necessary for configuring HPC systems. The procedures
-for this tuning vary largely from system to system and are often
-subjectively guided by the system engineer(s). Once a desired level of
-performance is achieved, an HPC system may only be incrementally
-reconfigured as required by updates or specific jobs. In the case that
-a system has changing workloads or nonstationary performance
-objectives that range from maximizing computational throughput to
-minimizing power consumption and system variability, it becomes clear
-that a more effective and automated tool is needed for configuring
-systems. This scenario presents a challenging and important
-application of multivariate approximation and interpolation
-techniques.
-
-Predicting the performance of an HPC system is a challenging problem
-that is primarily attempted in one of two ways: (1) build a
-statistical model of the performance by running experiments on the
-system at select settings, or (2) run artificial experiments using a
-simplified simulation of the target system to estimate architecture
-and application bottlenecks. In this paper the proposed multivariate
-modeling techniques rest in the first category, and they represent a
-notable increase in the ability to model complex interactions between
-system parameters.
-
-Many previous works attempting to model system performance have used
-simulated environments to estimate the performance of a system
-\shortcite{grobelny2007fase,wang2009simulation,wang2013towards}. Some
-of these works refer to statistical models as being oversimplified and
-not capable of capturing the true complexity of the underlying
-system. This claim is partially correct, noting that a large portion
-of predictive statistical models rely on simplifying the model to one
-or two parameters
-\shortcite{snavely2002framework,bailey2005performance,barker2009using,ye2010analyzing}.
-These limited statistical models have provided satisfactory
-performance in very narrow application settings. Many of the
-aforementioned statistical modeling techniques claim to generalize,
-while simultaneously requiring additional code annotations, hardware
-abstractions, or additional application level understandings in order
-to generate models. The approach presented here requires no
-modifications of the application, no architectural abstractions, nor
-any structural descriptions of the input data being modeled. The
-techniques used are purely mathematical and only need performance data
-as input.
-
-Among the statistical models presented in prior works,
-\shortciteN{bailey2005performance} specifically mention that it is
-difficult for the simplified models to capture variability introduced
-by I/O. System variability in general has become a problem of
-increasing interest to the HPC and systems communities, however most
-of the work has focused on operating system (OS) induced variability
-\shortcite{beckman2008benchmarking,de2007identifying}. The work that
-has focused on managing I/O variability does not use any sophisticated
-modeling techniques \shortcite{lofstead2010managing}. Hence, this
-paper presents a case study applying advanced mathematical and
-statistical modeling techniques to the domain of HPC I/O
-characteristics. The models are used to predict the mean throughput of
-a system and the variance in throughput of a system. The discussion
-section outlines how the techniques presented can be applied to any
-performance metric and any system.
-
-%% Multivariate models for an HPC system would be a function of the
-%% tunable parameters built to accurately model some desired performance
-%% metric.
-
-%% \begin{enumerate}
-%% \item The value of multivariate Modeling
-%% \item The data context
-%% \item The proposed method for using multivariate models
-%% \item The impact of effective models
-%% \end{enumerate}
-
-In general, this paper compares five multivariate approximation
+This chapter compares five of the multivariate approximation
 techniques that operate on inputs in $\mathbb{R}^d$ ($d$-tuples of
-real numbers) and produce predicted responses in $\mathbb{R}$. In
-order to provide coverage of the varied mathematical strategies that
-can be employed to solve the continuous modeling problem, three of the
-techniques are regression based and the remaining two are
-interpolants. The sections below outline the mathematical formulation
-of each technique and provide computational complexity bounds with
-respect to the size (number of points and dimension) of input
-data. Throughout, $d$ will refer to the dimension of the input data,
-$n$ is the number of points in the input data, $x^{(i)} \in
-\mathbb{R}^d$ is the $i$-th input data point, $x^{(i)}_j$ is the
-$j$-th component of $x^{(i)}$, and $f(x^{(i)})$ is the response value
-of the $i$-th input data point.
-
-The remainder of the paper is broken up into five major parts. Section
-\ref{sec:multivariate} provides an overview of the multivariate
-modeling techniques, Section \ref{sec:methodology} outlines the
-methodology for comparing and evaluating the performance of the
-models, Section \ref{sec:results} presents the IOzone predictions,
-Section \ref{sec:discussion} discusses the obvious and subtle
-implications of the models' performance, and Section
-\ref{sec:conclusion} concludes and offers directions for future work.
-
-\section{Multivariate Models}
-\label{sec:multivariate}
-
-\subsection{Regression}
-\vspace{-10pt}
-Multivariate approximations are capable of accurately modeling a
-complex dependence of a response (in $\mathbb{R}$) on multiple
-variables (represented as a points in $\mathbb{R}^{d}$). The
-approximations to some (unknown) underlying function $f: \mathbb{R}^d
-\rightarrow \mathbb{R}$ are chosen to minimize some error measure
-related to data samples $f(x^{(i)})$. For example, least squares
-regression uses the sum of squared differences between modeled
-response values and true response values as an error measure.
-
-\vspace{-10pt}
-\subsubsection{Multivariate Adaptive Regression Splines}
-\vspace{-10pt}
-This approximation was introduced in
-\shortciteN{friedman1991multivariate} and subsequently improved to its
-current version in \shortciteN{stanford1993fast}, called fast
-multivariate adaptive regression splines (Fast MARS). In Fast MARS, a
-least squares fit model is iteratively built by beginning with a
-single constant valued function and adding two new basis functions at
-each iteration of the form
-
-$$ B_{2s-1}(x) = B_l(x) [c(x_i-v)]_+ ,$$
-$$ B_{2s}(x) = B_k(x) [c(x_i-v)]_- ,$$
-
-where $s$ is the iteration number, $B_l(x)$ and $B_k(x)$ are basis
-functions from the previous iteration, $c, v \in \mathbb{R}$,
-
-$$w_+ = \begin{cases} w, & w \geq 0 \\ 0, & w < 0 \end{cases},$$
-
-and $w_- = (-w)_+$. After iteratively constructing a model, MARS then
-iteratively removes basis functions that do not contribute to goodness
-of fit. In effect, MARS creates a locally component-wise tensor
-product approximation of the data. The overall computational
-complexity of Fast MARS is $\mathcal{O}(n d m^3)$ where $m$ is the
-maximum number of underlying basis functions. This paper uses an
-implementation of Fast MARS \shortcite{rudy2017pyearth} with $m =
-200$.
-
-\vspace{-10pt}
-\subsubsection{Multilayer Perceptron Regressor}
-\vspace{-10pt}
-The neural network is a well studied and widely used method for both
-regression and classification tasks
-\shortcite{hornik1989multilayer}. When using the rectified linear unit
-(ReLU) activation function \shortcite{dahl2013improving} and training
-with the BFGS minimization technique \shortcite{moller1993scaled}, the
-model built by a multilayer perceptron uses layers $l : \mathbb{R}^{i}
-\rightarrow \mathbb{R}^{j}$ defined by
-
-$$ l(u) = \big( u^t W_l \big)_+ ,$$
-
-where $W_l$ is the $i$ by $j$ weight matrix for layer $l$. In this
-form, the multilayer perceptron (MLP) produces a piecewise linear
-model of the input data. The computational complexity of training a
-multilayer perceptron is $\mathcal{O}(n d m)$, where $m$ is
-determined by the sizes of the layers of the network and the stopping
-criterion of the BFGS minimization used for finding weights. This
-paper uses the scikit-learn MLP regressor \shortcite{scikit-learn}, a
-single hidden layer with 100 nodes, ReLU activation, and BFGS for
-training.
-
-\subsubsection{Support Vector Regressor}
-\vspace{-10pt}
-Support vector machines are a common method used in machine learning
-classification tasks that can be adapted for the purpose of regression
-\shortcite{basak2007support}. How to build a support vector regressor
-(SVR) is beyond the scope of this summary, but the resulting
-functional fit $p : \mathbb{R}^d \rightarrow \mathbb{R}$ has the form
-
-$$ p(x)  = \sum_{i=1}^{n}a_i K(x,x^{(i)}) + b ,$$
-
-%% $$ \text{Minimize } \bigg\{ \frac{1}{2}\sum_{i=1}^{n}\sum_{j=1}^{n}
-%% a_i a_j K(x_i, x_j) + \epsilon \sum_{i=1}^{n} a_i - \sum_{i=1}^{n} y_i
-%% a_i \bigg\} $$
-
-%% $$ \text{Subject to } \sum_{i=1}^{n}a_i = 0 \text{ } \text{ and }
-%% \text{ } a_i \in [0,C] $$
-
-where $K$ is the selected kernel function, $a \in \mathbb{R}^n$, $b
-\in \mathbb{R}$ are coefficients to be solved for simultaneously. The
-computational complexity of the SVR is $\mathcal{O}(n^2dm)$, with $m$
-being determined by the minimization convergence criterion. This paper
-uses the scikit-learn SVR \shortcite{scikit-learn} with a polynomial
-kernel function.
-
-\vspace{-10pt}
-\subsection{Interpolation}
-\vspace{-10pt}
-In some cases it is desirable to have a model that can recreate the
-input data exactly. This is especially the case when the confidence in
-the response values for known data is high. Both interpolation models
-analyzed in this paper are based on linear functions.
-
-\vspace{-10pt}
-\subsubsection{Delaunay}
-\vspace{-10pt}
-The Delaunay method of interpolation is a well studied geometric
-technique for producing an interpolant \shortcite{lee1980two}. The
-Delaunay triangulation of a set of data points into simplices is such
-that the sphere defined by the vertices of each simplex contains no
-data points in the sphere's interior. For a $d$-simplex S with
-vertices $v^{(0)}$, $v^{(1)}$, $\ldots$, $v^{(d)}$, $x \in S$, and
-data values $f(v^{(i)})$, $i=0$, $\ldots$, $d$, $x$ is a unique convex
-combination of the vertices,
-
-$$ x = \sum_{i=0}^{d} w_i v^{(i)}, \quad \sum_{i=0}^{d} w_i = 1, \quad
-w_i \geq 0, \quad i=0,\ldots,d ,$$
-
-and the Delaunay interpolant to $f$ at $x$ is given by
-
-$$ p(x) = \sum_{i=0}^{d} w_i f(v^{(i)}). $$
-
-The computational complexity of the Delaunay triangulation (for the
-implementation used here) is $\mathcal{O}(n^{\lceil d/2 \rceil})$,
-which is not scalable to $d > 10$ \shortcite{sartipizadeh2016computing}.
-The scipy interface \shortcite{scipy} to the QuickHull implementation
-\shortcite{barber1996qhull} of the Delaunay triangulation is used here.
-
-\subsubsection{Linear Shepard}
-\vspace{-10pt}
-The linear Shepard method (LSHEP) is a blending function using local
-linear interpolants, a special case of the general Shepard algorithm
-\shortcite{thacker2010algorithm}. The interpolant has the form
-
-$$ p(x) = \frac{\sum_{k=1}^{n}W_k(x)P_k(x)}{\sum_{k=1}^{n}W_k(x)} ,$$
-
-where $W_k(x)$ is a locally supported weighting function and $P_k(x)$
-is a local linear approximation to the data satisfying
-$P_k\big(x^{(k)}\big) = f\big(x^{(x)}\big)$. The computational
-complexity of LSHEP is $\mathcal{O}(n^2d^3)$. This paper uses the
-Fortran\#95 implementation of LSHEP in SHEPPACK
-\shortcite{thacker2010algorithm}.
-
-%% %     Related Work     
-%% %======================
-%% \section{Related Work}
-%% \begin{enumerate}
-%% \item Not sure how much to include here? Shooting for thoroughness or
-%%   simply necessary coverage? How much background should I expect the
-%%   readers of this paper to have in the ``multivariate modeling of
-%%   systems'' area?
-%% \end{enumerate}
+real numbers) and produce predicted responses in $\mathbb{R}$. Three
+of the chosen techniques are regression based and the remaining two
+are interpolants, providing reasonable coverage of the varied
+mathematical strategies that can be employed to solve continuous
+modeling problems.
 
@@
 %=====================
-\section{Methodology}
-\label{sec:methodology}
-\subsection{Data}
-\vspace{-10pt}
+\section{I/O Data}
 
@@
 four-dimensional dataset produced by executing the IOzone benchmark
-from \shortciteN{iozone} on a homogeneous cluster of computers. All
+from \citet{iozone} on a homogeneous cluster of computers. Multiple
+I/O Zone data sets will be used throughout the work, however this
+chapter relies on this specific four-dimensional dataset. All
 experiments were performed on parallel shared-memory nodes common to
@@
 individually by the multivariate approximation techniques presented in
-Section \ref{sec:multivariate}. The summary of data used in the
-experiments for this paper can be seen in Table \ref{tab:data_type}.
-Distributions of raw throughput values being modeled can be seen in
-Figure \ref{fig:raw_throughput}.
+Chapter \ref{ch:algs}. The summary of data used in the experiments for
+this chapter can be seen in Table \ref{tab:data_type}.  Distributions
+of raw throughput values being modeled can be seen in Figure
+\ref{fig:raw_throughput}.
 
@@
   \centering
-  \includegraphics[width=\textwidth,trim={0 .5in 0 .4in}]{Raw_Throughput.pdf}
+  \includegraphics[width=\textwidth,trim={0 .5in 0 .4in}]{Figures/HPC/Raw_Throughput.pdf}
   \caption{Histograms of 100-bin reductions of the PMF of I/O
@@
 
-\vspace{-10pt}
 \subsection{Dimensional Analysis}
-\vspace{-10pt}
-This work utilizes an extension to standard $k$-fold cross validation
-that allows for a more thorough investigation of the expected model
-performance in a variety of real-world situations. Alongside
-randomized splits, two extra components are considered: the amount of
-training data provided, and the dimension of the input data. It is
-important to consider that algorithms that perform well with less
-training input also require less experimentation. Although, the amount
-of training data required may change as a function of the dimension of
-the input and this needs to be studied as well. The framework used
-here will be referred to as a multidimensional analysis (MDA) of the
-IOzone data.
 
-%% \color{red}
+This analysis utilizes an extension to standard $k$-fold cross
+validation that allows for a more thorough investigation of the
+expected model performance in a variety of real-world
+situations. Alongside randomized splits, two extra components are
+considered: the amount of training data provided, and the dimension of
+the input data. It is important to consider that algorithms that
+perform well with less training input also require less
+experimentation. Although, the amount of training data required may
+change as a function of the dimension of the input and this needs to
+be studied as well. The framework used here will be referred to as a
+multidimensional analysis (MDA) of the IOzone data.
 
 \subsubsection{Multidimensional Analysis}
-\vspace{-10pt}
+
 This procedure combines random selection of training and testing
@@
 ensure that training points are well spaced, a statistical method for
-picking points from \shortciteN{amos2014algorithm} is used:
+picking points from \citet{amos2014algorithm} is used:
 \begin{enumerate}
@@
 
-%% is open source, and is freely available at the url
-%% \textit{https://github.com/tchlux/VarSys/HPC\_Paper/Code/multi\_dim\_analysis.py}
-
-%% \begin{enumerate}
-%% \item Cycling the categorical settings
-%% \item Selecting subsets of 1,2,3 up to 4 dimensions
-%% \item Cycling different training : testing ratios (5:95 $\rightarrow$ 95:5)
-%% \item Generating 200 random training : testing splits
-%% \item Ensuring the testing points are on/inside the convex hull of the training.
-%% \item Ensuring the training points are well-spaced.
-%% \end{enumerate}
-
-%% \subsection{Prediction}
-%% \begin{enumerate}
-%% \item For each file generated from the dimensional analysis, train on
-%%   the training data, evaluate at the testing data points
-%% \end{enumerate}
-
-%     Results     
-%=================
-\section{Results}
-\label{sec:results}
+\section{Na\"{\i}ve Variability Modeling Results}
 
@@
 
-\vspace{-10pt}
 \subsection{I/O Throughput Mean}
-\vspace{-10pt}
+
 Almost all multivariate models analyzed make predictions with a
@@
   \centering
-  \includegraphics[width=\textwidth,trim={0 .5in 0 .3in}]{Mean_Dim.pdf}
+  \includegraphics[width=\textwidth,trim={0 .5in 0 .3in}]{Figures/HPC/Mean_Dim.pdf}
   \caption{These box plots show the prediction error of mean with
@@
   \centering
-  \includegraphics[width=\textwidth,trim={0 .5in 0 .3in}]{Mean_TT_Ratio.pdf}
+  \includegraphics[width=\textwidth,trim={0 .5in 0 .3in}]{Figures/HPC/Mean_TT_Ratio.pdf}
   \caption{These box plots show the prediction error of mean with
@@
   \centering
-  \includegraphics[width=\textwidth,trim={0 .5in 0 .3in}]{Var_TT_Ratio.pdf}
+  \includegraphics[width=\textwidth,trim={0 .5in 0 .3in}]{Figures/HPC/Var_TT_Ratio.pdf}
   \caption{These box plots show the prediction error of variance with
@@
 
-\vspace{-10pt}
 \subsection{I/O Throughput Variance}
-\vspace{-10pt} The prediction results for variance resemble those for
+
+The prediction results for variance resemble those for
 predicting mean. Delaunay remains the best overall predictor
@@
 
-\vspace{-10pt}
 \subsection{Increasing Dimension and Decreasing Training Data}
-\vspace{-10pt}
+
 As can be seen in Figure \ref{fig:mean_dim}, all of the models suffer
@@
 %====================
-\section{Discussion}
-\label{sec:discussion}
+\section{Discussion of Na\"{\i}ve Approximations}
 
-The present results demonstrate that a straightforward application of
-multivariate modeling techniques can be used to effectively predict
-HPC system performance. Some modeling effort on the part of a systems
-engineer combined with a significant amount of experimentation (days
-of CPU time for the IOzone data used here) can yield a model capable
-of accurately tuning an HPC system to the desired performance
-specification, although qualitatively correct predictions can be
-achieved with much less (10\%, say) effort.
+The results presented above demonstrate that a straightforward
+application of multivariate modeling techniques can be used to
+effectively predict HPC system performance. Some modeling effort on
+the part of a systems engineer combined with a significant amount of
+experimentation (days of CPU time for the IOzone data used here) can
+yield a model capable of accurately tuning an HPC system to the
+desired performance specification, although qualitatively correct
+predictions can be achieved with much less (10\%, say) effort.
 
-\vspace{-10pt}
 \subsection{Modeling the System}
-\vspace{-10pt}
+
 The modeling techniques generated estimates of drastically different
@@
 
-While the Delaunay method appears to be the best predictor in the
-present IOzone case study, it is important to note that the Delaunay
-computational complexity increases with the dimension of the input
-more rapidly than other techniques. The implementation of Delaunay
-(QuickHull) used would experience unacceptably large training times
-beyond ten-dimensional input. This leaves much room for other
-techniques to perform best in higher dimension unless a more efficient
-implementation of Delaunay can be used.
+The Delaunay method appears to be the best predictor in the present
+IOzone case study. Particularly a piecewise linear interpolant like
+Delaunay appears well-suited for prediction when relatively small
+amounts of data are available to mode la function. It is important to
+note that the Delaunay computational complexity in the dimension of
+the input is worse than other techniques.
 
@@
 
-\vspace{-10pt}
 \subsection{Extending the Analysis}
-\vspace{-10pt}
+
 System I/O throughput mean and variance are simple and useful system
-characteristics to model. The process presented in this work is
+characteristics to model. The process presented in this chapter is
 equally applicable to predicting other useful performance
@@
 or any other ordinal performance metric. For each of these there is
-the potential to model system variability as well. This work has
-chosen variance as a measure of variability, but the techniques used
-in this paper could be applied to more precise measures of variability
-such as the percentiles of the performance distribution or the entire
-distribution itself. A thorough exploration of HPC systems
-applications of multivariate modeling constitutes future work.
+the potential to model system variability as well. This chapter uses
+variance as a measure of variability, but the techniques are applied
+to more precise measures of variability (the entire distribution
+itself) in Chapter \ref{ch:strong}.
 
-%% \begin{enumerate}
-%% \item Measuring different system characteristics other than I/O
-%% \item Recording system performance constantly to maintain model accuracy
-%% \end{enumerate}
+\vspace{-10pt}
+\section{Takeaway From Na\"{\i}ve Approximation}
+\label{sec:conclusion}
 
-\section{Conclusion}
-\label{sec:conclusion}
 Multivariate models of HPC system performance can effectively predict
@@
 acceptable as the number of system parameters being modeled
-increases. These multivariate techniques can and should be applied to
-HPC systems with more than four tunable parameters in order to
-identify optimal system configurations that may not be discoverable
-via previous methods nor by manual performance tuning.
-
-%     Future Work     
-%=====================
-\vspace{-10pt}
-\subsection{Future Work}
-\vspace{-10pt}
-The most severe limitation to the present work is the restriction to
-modeling strictly ordinal (not categorical) system parameters.
-Existing statistical approaches for including categorical variables
-are inadequate for nonlinear interactions in high dimensions. Future
-work could attempt to identify the viability of different techniques
-for making predictions including categorical system parameters.
-
-There remain many other multivariate modeling techniques not included
-in this work that should be evaluated and applied to HPC performance
-prediction. For I/O alone, there are far more than the four tunable
-parameters studied in this work. Alongside experimentation with more
-models, there is room for a theoretical characterization of the
-combined model and data properties that allow for the greatest
-predictive power.
-
-%% \color{black}
-
-%% \newpage
-%% \ \\
-
-%% \vspace{3in}
-
-%% \texttt{\textbf{L}oc\textbf{A}ll\textbf{Y} li\textbf{NE}ar
-%%   \textbf{W}eighted \textbf{A}pproxima\textbf{T}ion\textbf{S}
-%%   \textbf{O}f fu\textbf{N}tions}\\
-%% %% \texttt{L\ \ A\ \ Y \ \ NE\ \ \  W\ \ \ \ \ \ \ \ A\ \ \ \ \ \ \ \ T\ \ \ S O\ \ \ \ N\ \ \ \ \ \ }
-%% \newpage
-
-\bibliographystyle{scsproc}
-\bibliography{paper}
-
-%     Author Biography Block     
-%================================
-\section*{Author Biographies}
-
-\textbf{\uppercase{Thomas C. H. Lux}} is a Ph.D. student at Virginia Tech
-studying computer science under Dr. Layne Watson.
-
-\textbf{\uppercase{Layne T. Watson}} (Ph.D., Michigan, 1974) has interests
-in numerical analysis, mathematical programming, bioinformatics, and data
-science.
-He has been involved with the organization of HPCS since 2000.
-
-\textbf{\uppercase{Tyler H. Chang}} is a Ph.D. student at Virginia Tech
-studying computer science under Dr. Layne Watson.
-
-\textbf{\uppercase{Jon Bernard}} is a Ph.D. student at Virginia Tech
-studying computer science under Dr. Kirk Cameron.
-
-\textbf{\uppercase{Bo Li}} is a senior Ph.D. student at Virginia Tech
-studying computer science under Dr. Kirk Cameron.
-
-\textbf{\uppercase{Li Xu}} is a Ph.D. student at Virginia Tech studying
-statistics under Dr. Yili Hong.
-
-\textbf{\uppercase{Godmar Back}} (Ph.D., University of Utah, 2002) has
-broad interests in computer systems, with a focus on performance and
-reliability aspects of operating systems and virtual machines.
-
-\textbf{\uppercase{Ali R. Butt}} (Ph.D., Purdue, 2006) has interests in cloud computing, distributed computing, and operating system induced variability.
-
-\textbf{\uppercase{Kirk W. Cameron}} (Ph.D., Louisiana State, 2000) has
-interests in computer systems design, performance analysis, and power and
-energy efficiency.
-
-\textbf{\uppercase{Yili Hong}} (Ph.D., Iowa State, 2009) has interests in
-engineering statistics, statistical modeling, and data analysis.
-
-\end{document}
+increases. These multivariate techniques should be applied to HPC
+systems with more than four tunable parameters in order to identify
+optimal system configurations that may not be discoverable via
+previous methods nor by manual performance tuning, which will be
+explored in later chapters.

### 2019-05-03 Disseration/prelim.tex
Finished first full draft of preliminary defense document. Ready to share with committee.

--- 2019-03-31 Paper_[2019-02]_Numerical_Algorithms/paper.tex
@@
-\RequirePackage{fix-cm}
-\documentclass[smallextended,final]{svjour3}  % onecolumn (second format)
-\smartqed  % flush right qed marks, e.g. at end of proof
+\documentclass[doublespace,nopageskip]{VTthesis} % nopageskip = Removes arbitrary blank pages.
 
-% Insert here the call for the packages your document requires
+%% HPC
+\usepackage{latexsym}
 \usepackage{graphicx}
-\usepackage{tikz}     % For adding axes to plots
+\usepackage{mathptmx}
+\usepackage{amsmath}
+\usepackage{amsfonts}
+\usepackage{amssymb}
+\usepackage{amsbsy}
+\usepackage{amsthm}
+\usepackage{float}    % For controlling the placement of tables.
+%% 
+\usepackage{booktabs} % For formal tables
+\usepackage{tikz}     % For drawing example boxes
 \usetikzlibrary{positioning}
-\usepackage{graphicx} % For including PDFs
-\usepackage{amsmath}  % assumes amsmath package installed
-\usepackage{amssymb}  % assumes amsmath package installed
-\usepackage{booktabs} % For formal tables
-\usepackage{tabularx} % For tables with nice column widths.
-\usepackage[colorlinks,allcolors=blue]{hyperref}
-\usepackage{float}    % For controlling the placement of tables.
 
-%
-% please place your own definitions here and don't use \def but
-% \newcommand{}{}
+\usepackage{hyphenat}
+\hyphenation{op-tical net-works semi-conduc-tor}
 
@@
 
-\newcolumntype{d}[1]{D{.}{.}{#1}}
+% Using the following header instead will create a draft copy of your thesis
+%\documentclass[doublespace,draft]{VTthesis}
 
-%% Create a ``Lemma'' environment that does not have italics font.
-\spnewtheorem{plainlemma}{Lemma}{\bf}{}
-%% Create a ``Theorem'' environment that is not numbered.
-\spnewtheorem*{onetheorem}{Theorem}{\bf}{}
-%% Create a ``Proof'' environment that starts with 'Proof.'
-\spnewtheorem*{proofdot}{Proof.}{\it}{}
+% Title of your thesis
+\title{Interpolation for Modeling and Predicting Variability}
 
-%% \newcommand{\Bigl}{\mathopen\Big}
-%% \newcommand{\Bigr}{\mathclose\Big}
-%
-% Insert the name of "your journal" with
-\journalname{Numerical Algorithms}
-%
+% You should include 3-5 keywords, separated by commas
+\keywords{Approximation Theory, Numerical Analysis, High Performance Computing, Computer Security, Nonparametric Statistics, Mathematical Software}
+
+% Your name, including middle initial(s)
+\author{Thomas C. H. Lux}
+
+% Change this to your program, e.g. Physics, Civil Engineering, etc.
+\program{Computer Science}
+
+% Change this to your degree, e.g. Master of Science, Master of Art, etc.
+\degree{Doctor of Philosophy}
+
+% This should be your defense date:
+\submitdate{May 16, 2019}
+
+% Committee members. Only have five readers and one chair available.
+% Only use the ones you need and don't include the ones you don't need.
+% You can also declare a Co-advisor. If you do, the principal and co-advisors
+% will be listed as co-advisors on the title page.  Per the VT ETD standards, 
+% you should not include titles or educational qualifications such as PhD or Dr.
+% You should, however, include middle initials if possible.
+\principaladvisor{Layne Watson}
+\coadvisor{Yili Hong}
+\firstreader{Kirk Cameron}
+\secondreader{Bert Huang}
+\thirdreader{Gang Wang}
+\fourthreader{Yang Cao}
+
+% The dedication and acknowledgement pages are optional. Comment them out to remove them.
+%% \dedication{...}
+%% \acknowledge{...}
+
+% The abstract is required and should be <=250 words for thesis, <=350 words for dissertation.
+\abstract{Approximation is an important problem.
+This work presents applications of interpolants to modeling random variables.
+Specifically, this work studies the prediction of distributions of random variables applied to computer system throughput variability.
+New methods for interpolating data with many dimensions are presented and novel theoretical error bounds are constructed for piecewise linear interpolation.
+Finally, a mathematical software for approximating distributions from data samples via monotone quintic splines is proposed.}
+
+% The general audience abstract is required. There are currently no word limits.
+\abstractgenaud{It is common for scientists to collect data on something they are studying.
+Often scientists want to create a (predictive) model of that phenomenon based on the data, but the choice of \textit{how} to model the data is a difficult one to answer.
+This work proposes methods for modeling data that operate under very few assumptions that are broadly applicable across science.
+Finally, a software package is proposed that would allow scientists to better understand the \textit{true} distribution of their data given relatively few observations.}
+
 \begin{document}
+% The following lines set up the front matter of your thesis or dissertation and is required to ensure proper formatting per the VT ETD standards. 
+\frontmatter
+\maketitle
+\tableofcontents
 
-\title{Interpolation of Sparse High-Dimensional Data\thanks
-  {This work was supported by the National Science Foundation Grant CNS-1565314.}
-}
-% \subtitle{Do you have a subtitle?\\ If so, write it here}
-%\titlerunning{Short form of title}        % if too long for running head
+% The list of figures and tables are now optional per the official ETD standards.
+Unless you have a very good reason for removing them, you should leave these lists in the document.
+Comment them out to remove them.
+\listoffigures
+\listoftables
 
-\author{Thomas C. H. Lux   \and
-  Layne T. Watson          \and
-  Tyler H. Chang           \and
-  Yili Hong                \and
-  Kirk Cameron
-}
+%% \printnomenclature %creates a list of abbreviations. Comment out to remove it. 
+%% ABBR is described here.
+%% \nomenclature{ABBR}{abbreviation}
 
-%\authorrunning{Short form of author list} % if too long for running head
+% The following sets up the document for the main part of the thesis or dissertation. Do not comment out or remove this line.
+\mainmatter
 
-\institute{Thomas Lux \at
-  Virginia Polytechnic Institute \& State University (VPI\&SU) \\
-  Blacksburg, VA 24061 \\
-  \email{tchlux@vt.edu}
-}
+%% Introduction of variability modeling, the importance of modeling in science in general, then dive into the specific application of modeling techniques to systems (and security).
+\chapter{The Importance and Applications of Variability} \label{ch:apps}
+\input{apps}
 
-\date{Received: date / Accepted: date}
-% The correct dates will be entered by the editor
+%% Describe all of the interpolants and regressors that will be used throughout this work. (except meshes)
+\chapter{Algorithms for Constructing Approximations} \label{ch:algs}
+\input{algs}
 
-\maketitle
+%% Demonstrate the naive application of interpolants and regressors to data. Discuss results.
+\chapter{Na\"{\i}ve Approximations of Variability} \label{ch:naive}
+\input{naive}
 
-\begin{abstract}
-Increases in the quantity of available data have allowed all fields of
-science to generate more accurate models of multivariate phenomena.
-Regression and interpolation become challenging when the dimension of
-data is large, especially while maintaining tractable computational
-complexity. Regression is a popular approach to solving approximation
-problems with high dimension, however there are often advantages to
-interpolation. This paper presents a novel and insightful error bound
-for (piecewise) linear interpolation in arbitrary dimension and
-contrasts the performance of some interpolation techniques with
-popular regression techniques. Empirical results demonstrate the
-viability of interpolation for moderately high dimensional
-approximation problems, and encourage broader application of
-interpolants to multivariate approximation in science.
+%% Introduce the theory on box splines. Introduce the algorithms that rely on box splines. Show the results of the code that produces box splines.
+%% Demonstrate the application of these meshes (and box splines particularly) to approximation problems. Discuss results.
+\chapter{Box Splines: Uses, Constructions, and Applications} \label{ch:boxes}
+\input{boxes}
 
-\keywords{approximation \and regression \and interpolation \and high
-  dimension \and error bound}
-% \PACS{PACS code1 \and PACS code2 \and more}
-% \subclass{MSC code1 \and MSC code2 \and more}
-\end{abstract}
+%% Introduce a measurement for error when predicting distribution outputs, then show what happens when system behavior is predicted as a distribution.
+\chapter{Stronger Approximations of Variability} \label{ch:strong}
+\input{strong}
 
-%     Introduction     
-%======================
+%% Construct error theory that works for all observed applications to this point.
+\chapter{An Error Bound on Piecewise Linear Interpolation} \label{ch:error}
+%% Apply approximation algorithms to a final set of problems. Discuss results.
+\input{error}
 
-%% ===================================================================
-\section{Introduction}
-\label{sec:introduction}
+%% Propose new method to improve predictions for distributions.
+\chapter{Improving Variability Estimates with Monotone $C^2$ Splines} \label{ch:splines}
+\input{splines}
 
-Regression and interpolation are problems of considerable importance
-that find applications across many fields of science. Pollution and
-air quality analysis \cite{de2008field}, energy consumption management
-\cite{lazos2014optimisation}, and student performance prediction
-\cite{cortez2008using,lux2016applications} are a few of many
-interdisciplinary applications of multivariate regression for
-predictive analysis. As discussed later, these techniques can also be
-applied to prediction problems related to forest fire risk assessment
-\cite{cortez2007data}, Parkinson's patient clinical evaluations
-\cite{tsanas2010accurate}, local rainfall and weather
-\cite{williams2009rattle}, credit card transactions
-\cite{pozzolo2015calibrating}, and high performance computing (HPC)
-file input/output (I/O) \cite{lux2018nonparametric}.
+%% Bibliography.
+\bibliography{prelim}
+\bibliographystyle{plainnat}
 
-Regression and interpolation have a considerable theoretical base in
-one dimension \cite{cheney2009course}. Splines in particular are well
-understood as an interpolation technique in one dimension
-\cite{de1978practical}, particularly B-splines. Tensor products of
-B-splines \cite{unther1996interpolating} or other basis functions have
-an unfortunate exponential scaling with increasing
-dimension. Exponential scaling prohibits tensor products from being
-reasonably applied beyond three-dimensional data. In order to address
-this dimensional scaling challenge, C. de Boor and others proposed box
-splines \cite{de2013box}, of which one of the approximation techniques
-in this work is composed \cite{lux2018novel}.
+% This formats the chapter name to appendix to properly define the headers:
+\appendix
 
-The theoretical foundation of low dimensional interpolation allows the
-construction of strong error bounds that are absent from high
-dimensional problems. This work extends some known results regarding
-the secant method \cite{dennis1996numerical} to construct an
-interpolation error bound for problems of arbitrary dimension. These
-error bounds are useful, considering the same cannot be said for
-regression algorithms in general. The maximum complexity of an
-interpolant is bounded by the amount of data available, while the
-maximum complexity of a regressor is bounded by both the amount of
-data and the chosen parametric form. For this reason, generic uniform
-bounds are largely unobtainable for regression techniques on arbitrary
-approximation problems, even when the approximation domain is bounded.
-
-Aside from theoretical motivation for the use of interpolants, there
-are often computational advantages as well. Interpolants do not have
-the need for \textit{fitting} data, or minimizing error with respect
-to model parameters. In applications where the amount of data is large
-and the relative number of predictions that need to be made for a
-given collection of data is small, the direct application of an
-interpolant is much less computationally expensive.
-
-In this work, multivariate interpolation is defined given a closed
-convex subset $Y$ of a metrizable topological vector space with metric
-$s$, some function $f:\mathbb{R}^d \rightarrow Y$ and a set of points
-$X = \bigl\{x^{(1)}$, $\ldots$, $x^{(n)}\bigr\} \subset \mathbb{R}^d$,
-along with associated function values $f\bigl(x^{(i)}\bigr)$. The goal
-is to construct an approximation $\hat f: \mathbb{R}^d \rightarrow Y$
-such that $\hat f\bigl(x^{(i)}\bigr) = f\bigl(x^{(i)}\bigr)$ for all
-$i = 1$, $\ldots$, $n$. It is often the case that the form of the true
-underlying function $f$ is unknown, however it is still desirable to
-construct an approximation $\hat f$ with small approximation error at
-$y \notin X$. The two metric spaces that will be discussed in this
-work are the real numbers with metric $s(x,y) = |x-y|$, and the set of
-cumulative distribution functions (CDFs) with the Kolmogorov-Smirnov
-(KS) statistic \cite{lilliefors1967kolmogorov} as metric.
-
-Multivariate regression is often used when the underlying function is
-presumed to be stochastic, or stochastic error is introduced in the
-evaluation of $f$. Hence, multivariate regression relaxes the
-conditions of interpolation by choosing parameters $P$ defining $\hat
-f(x;P)$ to minimize the error vector $\Bigl( \bigl | \hat f
-\bigl(x^{(1)};P\bigr) - f\bigl(x^{(1)}\bigr) \bigr|$, $\ldots$, $\bigl
-| \hat f \bigl(x^{(n)}; P\bigr) - f\bigl(x^{(n)}\bigr) \bigr | \Bigr)$
-in some norm. The difficult question in the case of regression is
-often what parametric form to adopt for any given application.
-
-The most challenging problem when scaling in dimension is that the
-number of possible interactions between dimensions grows
-exponentially. Quantifying all possible interactions becomes
-intractable, and hence beyond three-dimensional data mostly linear
-models are used. That is not to say nonlinear models are absent, but
-nonlinearities are often either preconceived or model pairwise
-interactions between dimensions at most. Even globally nonlinear
-approximations such as neural networks are constructed from
-compositions of summed low-interaction functions
-\cite{clevert2015fast}.
-
-Provided the theoretical and practical motivations for exploring
-interpolants, this work aims to study the empirical performance
-differences between a set of scalable (moderately) interpolation
-techniques and a set of common regression techniques. These techniques
-are applied to a collection of moderately high dimensional problems
-($5 \le d \le 30$) and the empirical results are discussed.
-
-The remainder of this paper is organized as follows. Sections
-\ref{sec:regression} and \ref{sec:interpolation} present the
-multivariate models. Section \ref{sec:error} gives the error measuring
-methodology that is used for collecting empirical results. Section
-\ref{sec:theory} contains the theoretical error bounds for
-interpolation. Section \ref{sec:data} presents the data sets for
-empirical approximation analysis and the approximation results for all
-models on these data sets. Section \ref{sec:discussion} analyzes and
-discusses the results, their implications, and their
-limitations. Finally, Section \ref{sec:conclusion} concludes.
-
-%% ===================================================================
-\section{Multivariate Regression}
-\label{sec:regression}
-Multivariate regressors are capable of accurately modeling a complex
-dependence of a response (in $Y$) on multiple variables (represented
-as a points in $\mathbb{R}^{d}$). The approximations to some (unknown)
-underlying function $f: \mathbb{R}^d \rightarrow Y$ are chosen to
-minimize some error measure related to data samples
-$f\bigl(x^{(i)}\bigr)$. For example, least squares regression uses the
-sum of squared differences between modeled function values and true
-function values as an error measure. In this section and the next,
-some techniques are limited to approximating real valued functions ($Y
-\subset \mathbb{R}$). These techniques can be extended to real
-vector-valued ranges by repeating the construction for each component
-of the vector output. Throughout the following, $x$ denotes a
-$d$-tuple, $x_i$ the $i$th component of $x$, and $x^{(i)}$ the $i$th
-$d$-tuple data point. Different symbols are used to represent the
-approximation function $\hat f$.
-
-\subsection{Multivariate Adaptive Regression Splines}
-\label{sec:mars}
-This approximation was introduced in \cite{friedman1991multivariate}
-and subsequently improved to its current version in
-\cite{stanford1993fast}, called fast multivariate adaptive regression
-splines (Fast MARS). In Fast MARS, a least squares fit model is
-iteratively built by beginning with a single constant valued function
-and adding two new basis functions at each iteration of the form
-\begin{align*}
-  B_{2j-1}(x) &= B_l(x) \bigl(x_i-x^{(p)}_i\bigr)_+, \\
-  B_{2j}(x) &= B_k(x) \bigl(x_i-x^{(p)}_i\bigr)_- ,
-\end{align*}
-where $j$ is the iteration number, $1 \le p \le n$, and $B_l(x)$,
-$B_k(x)$ are basis functions from the previous iteration,
- $$w_+ = \begin{cases} w, & w \geq 0 \\ 0, & w < 0 \end{cases},$$
-and $w_- = (-w)_+$. After iteratively constructing a model, MARS then
-iteratively removes basis functions that do not contribute to goodness
-of fit. In effect, MARS creates a locally component-wise tensor
-product approximation of the data. The overall computational
-complexity of Fast MARS is $\mathcal{O}(n d m^3)$ where $m$ is the
-maximum number of underlying basis functions. This paper uses an
-implementation of Fast MARS \cite{rudy2017pyearth} with $m = 200$
-throughout all experiments.
-
-\subsection{Support Vector Regressor}
-\label{sec:svr}
-Support vector machines are a common method used in machine learning
-classification tasks that can be adapted for the purpose of regression
-\cite{basak2007support}. How to build a support vector regressor (SVR)
-is beyond the scope of this summary, but the resulting functional fit
-$p : \mathbb{R}^d \rightarrow \mathbb{R}$ has the form
- $$ p(x)  = \sum_{i=1}^{n}a_i K\bigl(x,x^{(i)}\bigr) + b ,$$
-where $K$ is the selected kernel function, $a_i \in \mathbb{R}^n$, 
-$b \in \mathbb{R}$ are coefficients to be solved for simultaneously.
-The computational complexity of the SVR is $\mathcal{O}(n^2dm)$, with
-$m$ being determined by the minimization convergence criterion. This
-paper uses the scikit-learn SVR \cite{scikit-learn} with a polynomial
-kernel function.
-
-\subsection{Multilayer Perceptron Regressor}
-\label{sec:mlp}
-The neural network is a well studied and widely used method for both
-regression and classification tasks
-\cite{hornik1989multilayer,rumelhart1988learning}. When using the
-rectified linear unit (ReLU) activation function
-\cite{dahl2013improving} and fitting the model with stochastic
-gradient descent (SGD) or BFGS minimization techniques
-\cite{goh2017why,moller1993scaled,robbins1951stochastic}, the model
-built by a multilayer perceptron uses layers $l : \mathbb{R}^{i}
-\rightarrow \mathbb{R}^{j}$ defined by
- $$ l(u) = \bigl( u^t W_l + c_l \bigr)_+ ,$$
-where $c_l \in \mathbb{R}^j$ and $W_l$ is the $i \times j$ weight
-matrix for layer $l$. In this form, the multilayer perceptron (MLP)
-produces a piecewise linear model of the input data. The computational
-complexity of training a multilayer perceptron is $\mathcal{O}(n d
-m)$, where $m$ is determined by the sizes of the layers of the network
-and the stopping criterion of the minimization used for finding
-weights. This paper uses an MLP built from Keras and Tensorflow to
-perform regression \cite{tensorflow2015-whitepaper,chollet2015keras}
-with ten hidden layers each having thirty two nodes (a total of
-approximately ten thousand parameters), ReLU activation, and five
-thousand epochs of SGD for training.
-
-%% ===================================================================
-\section{Multivariate Interpolation}
-\label{sec:interpolation}
-The following interpolation techniques demonstrate a reasonable
-variety of approaches to interpolation. All of the presented
-interpolants produce approximations that are continuous in value,
-which is often a desirable property in applied approximation problems.
-
-\subsection{Delaunay}
-\label{sec:delaunay}
-
-The Delaunay triangulation is a well-studied geometric technique for
-producing an interpolant \cite{lee1980two}. The Delaunay triangulation
-of a set of data points into simplices is such that there are no data
-points inside the sphere defined by the vertices of each simplex. For
-a $d$-simplex S with vertices $x^{(0)}$, $x^{(1)}$, $\ldots$,
-$x^{(d)}$, and function values $f\bigl(x^{(i)}\bigr)$, $i=0$,
-$\ldots$, $d$, $y \in S$ is a unique convex combination of the
-vertices,
- $$ y = \sum_{i=0}^{d} w_i x^{(i)}, \quad \sum_{i=0}^{d} w_i = 1, \quad w_i \geq 0, \quad i=0,\ldots,d, $$
-and the Delaunay interpolant $\hat f(y)$ at $y$ is given by
- $$ \hat f(y) = \sum_{i=0}^{d} w_i f\bigl(x^{(i)}\bigr). $$ The
-computational complexity of Delaunay interpolation (for the
-implementation used \cite{chang2018polynomial}) is $\mathcal{O}(k n
-d^3)$. Given pathological data the entire triangulation could be
-computed with $k = n^{\lceil d / 2 \rceil}$, but the average case yields
-$k = d \log d$ and is capable of scaling reasonably to $d \leq 50$. In
-the present application, a Delaunay simplex $S$ containing $y$ is
-found, then the $d+1$ vertices (points in $X$) of $S$ are used to
-assign weights to each vertex and produce the predicted function value
-for point $y$.
-
-\subsection{Modified Shepard}
-\label{sec:modified-shepard}
-
-The modified Shepard method used here (ShepMod) generates a continuous
-approximation based on Euclidean distance and resembles a nearest
-neighbor interpolant \cite{cover1967nearest}. This model is a type of
-\textit{metric interpolation}, also called a Shepard method
-\cite{gordon1978shepard,shepard1968two}. The interpolant has the form
- $$ p(x) = \frac{\sum\limits_{k=1}^{n}W_k(x)f\bigl(x^{(k)}\bigr)}
-     {\sum\limits_{k=1}^{n}W_k(x)} ,$$
-where $W_k(x) = \bigl(r_k - \bigl\|x - x^{(k)}\bigr\|\bigr)_+^2 \big/
-\bigl\|x - x^{(k)}\bigr\|_2^2$ and $r_k \in \mathbb{R}$ is the
-smallest radius such that at least $d+1$ other points $x^{(j)}$, $j
-\not = k$, are inside the closed Euclidean ball of radius $r_k$ about
-$x^{(k)}$. The computational complexity of ShepMod is
-$\mathcal{O}(n^2d)$. This paper uses a Fortran 95 implementation of
-ShepMod derived from SHEPPACK \cite{thacker2010algorithm}.
-
-\subsection{Linear Shepard}
-The linear Shepard method (LSHEP) is a blending function using local
-linear interpolants, a special case of the general Shepard algorithm
-\cite{thacker2010algorithm}. The interpolant has the form
- $$ p(x) = \frac{\sum\limits_{k=1}^{n}W_k(x)P_k(x)}
- {\sum\limits_{k=1}^{n}W_k(x)} ,$$
-where $W_k(x)$ is a locally supported weighting function and $P_k(x)$
-is a local linear approximation to the data satisfying
-$P_k\bigl(x^{(k)}\bigr) = f\bigl(x^{(x)}\bigr)$. The computational
-complexity of LSHEP is $\mathcal{O}(n^2d^3)$. This paper uses the
-Fortran 95 implementation of LSHEP in SHEPPACK
-\cite{thacker2010algorithm}.
-
-\subsection{Box Splines}
-\label{sec:box-splines}
-
-The box spline model used here is an interpolation technique built
-from overlapping box splines \cite{de2013box}. The box splines serve
-as basis functions that are shifted and scaled to have support over
-box shaped regions. The boxes are constructed in a way to guarantee a
-covering of the domain \cite{lux2018novel}. Given a set of box splines
-$\bigl\{b^{x^{(i)}}\bigr\}$ with the iterative box properties outlined
-in \cite{lux2018novel} and anchored at interior points
-$\bigl\{x^{(i)}\bigr\}$,
- $$ \hat f(y) = \frac{\sum\limits_{i=1}^n b^{x^{(i)}}(y) f\bigl(x^{(i)}\bigr)}{\sum\limits_{i=1}^n b^{x^{(i)}}(y)}. $$
-Note that the box splines always satisfy
-$b^{x^{(i)}}\bigl(x^{(j)}\bigr) = \delta_{ij}$ and $b^{x^{(i)}}(y)
-\geq 0$. The computational complexity of interpolation via the box
-spline model is $\mathcal{O}(n^2 d)$.
-
-\subsection{Voronoi}
-\label{sec:voronoi}
-
-The nearest neighbor algorithm \cite{cover1967nearest} is a
-well-studied technique for classification and approximation. Nearest
-neighbor inherently utilizes the convex region $C^{x^{(i)}}$ (Voronoi
-cell \cite{dirichlet1850reduction}) consisting of all points closer to
-$x^{(i)}$ than to any other point $x^{(j)}$. The Voronoi model smooths
-the nearest neighbor approximation by utilizing the Voronoi cells to
-define support via a generic basis function $v: \mathbb{R}^d
-\rightarrow \mathbb{R}_+$ given by
- $$ v^{x^{(i)}}(y) = \left(1 - \frac{\bigl\|y - x^{(i)}\bigr\|_2}
- {2 \ h \bigl(y - x^{(i)} \mid x^{(i)}\bigr)} \right)_+, $$
-where $h\bigl(w \mid x^{(i)}\bigr)$ is the distance between $x^{(i)}$
-and the boundary of the Voronoi cell $C^{x^{(i)}}$ in the direction
-$w$. $v^{x^{(i)}}\bigl(x^{(j)}\bigr) = \delta_{ij}$ and $v^{x^{(i)}}$
-has local support, giving the interpolated value
- $$ f(y) = \frac{\sum\limits_{i=1}^n v^{x^{(i)}}(y)
-  f\bigl(x^{(i)}\bigr)}{\sum\limits_{i=1}^n v^{x^{(i)}}(y)}, $$
-where $0 \leq v^{x^{(i)}}(y) \leq 1$. The computational complexity of
-interpolation via this Voronoi model is $\mathcal{O}(n^2 d)$. All of
-the approximations are an interpolant involving a convex combination
-of known function values $f\bigl(x^{(i)}\bigr)$.
-
-%% ===================================================================
-\section{Measuring Error}
-\label{sec:error}
-
-\begin{figure}[htb]
-  \centering
-  \begin{tikzpicture}
-    \node (img)  {\includegraphics[width=0.75\textwidth,]{Fig1.pdf}};
-    \node[below=of img, node distance=1cm, yshift=1cm] {I/O Throughput};
-    \node[left=of img, node distance=0cm, rotate=90, anchor=center,yshift=-0.7cm] {CDF Value};
-  \end{tikzpicture}
-  \vspace{-0.3cm}
-  \caption{In this HPC I/O example, the general methodology for
-    predicting a CDF and evaluating error can be seen. The Delaunay
-    method chose three source distributions (dotted lines) and
-    assigned weights \{.3, .4, .3\} (top to bottom at arrow). The
-    weighted sum of the three known CDFs produces the predicted CDF
-    (dashed line). The KS Statistic (arrow) computed between the true
-    CDF (solid line) and predicted CDF (dashed line) is 0.2 for this
-    example. The KS test null hypothesis is rejected at $p$-value
-    0.01, however it is not rejected at $p$-value 0.001.
-  \vspace{-.1cm}}
-  \label{fig:prediction-example}
-\end{figure}
-
-When the range of an approximation is the real numbers, error is
-reported with summary statistics including: min absolute error, max
-absolute error, and absolute error quartiles. When the range of an
-approximation is the space of cumulative distribution functions, the
-Kolmogorov-Smirnov statistic (max-norm difference between the
-functions) is used.
-
-A hurdle when modeling function-valued outputs such as cumulative
-distribution functions (CDFs) or probability density functions (PDFs)
-is that certain properties must be maintained. It is necessary that a
-PDF $f: \mathbb{R} \rightarrow \mathbb{R}$ have the properties $f(x)
-\geq 0$ and $\int_{-\infty}^{\infty}f(x)dx = 1$. However, for a CDF
-$F: \mathbb{R} \rightarrow \mathbb{R}$ the properties are
-$F(x) \in [0,1]$ and $F(x)$ is absolutely continuous and
-nondecreasing. This work utilizes the fact that a convex combination
-of CDFs (or PDFs) results in a valid CDF (or PDF). Given $G(x) =
-\sum_{i}w_i F_i(x)$, $\sum_{i} w_i = 1$, $w_i \geq 0$, and each $F_i$
-is a valid CDF, $G$ must also be a valid CDF. A demonstration of how
-this is applied can be seen in Figure \ref{fig:prediction-example}.
-
-The performance of approximation techniques that predict probability
-functions can be analyzed through a variety of summary
-statistics. This work uses the max absolute difference, also known as
-the Kolmogorov-Smirnov (KS) statistic \cite{lilliefors1967kolmogorov}
-for its compatibility with the KS test.
-
-The two-sample KS test is a useful nonparametric test for comparing
-two CDFs while only assuming stationarity, finite mean, and finite
-variance. The null hypothesis (that two CDFs come from the same
-underlying distribution) is rejected at level $p \in [0,1]$ when
- $$ KS > \sqrt{-\frac{1}{2}\ln\biggl(\frac{p}{2}\biggr)} \sqrt{\frac{1}{n_1} + \frac{1}{n_2}}, $$
-with distribution sample sizes $n_1,n_2 \in \mathcal{N}$. For all
-applications of the KS test presented in this work $n_1 = n_2$. An
-example of the round-trip prediction methodology from known and
-predicted distributions to the calculation of error can be seen in
-Figure \ref{fig:prediction-example}.
-
-\section{Theoretical Error Bound}
-\label{sec:theory}
-
-This section presents the theoretical results bounding the error of
-(piecewise) linear interpolation. The error analysis relies on linear
-interpolation for three reasons: (1) second order results can be
-obtained utilizing a Lipschitz constant on the gradient of a function,
-rather than standard Lipschitz bounds; (2) the results directly apply
-to Delaunay interpolation; and (3) multiple other interpolants in this
-paper compute predictions as convex combinations of observed function
-values, which may allow for straight forward extensions of this error
-bound.
-
-\begin{plainlemma}
-  \label{lemma:1}
-  Let $S \subset \mathbb{R}^d$ be open and convex, $f: \mathbb{R}^d
-  \rightarrow \mathbb{R}$, and $\nabla f \in
-  Lip_{(\gamma,\|\cdot\|_2)}(S)$, the set of $\gamma$-Lipschitz
-  continuous functions in the $2$-norm. Then for all $x,y \in S$
-  $$\big|f(y) - f(x) - \langle \nabla f(x), y - x \rangle \big| \leq \frac{\gamma \|y - x\|_2^2}{2}.$$
-\end{plainlemma}
-
-\begin{proofdot}
-  Consider the function $g(t) = f \big((1-t) x + t y \big)$, $0 \leq t
-  \leq 1$, whose derivative $g'(t) = \big\langle \nabla f \big((1-t) x
-  + t y \big), y - x \big\rangle$ is the directional derivative of $f$
-  in the direction $(y - x).$
-  \begin{align*}
-    \big|f(y) - f(x) - \langle \nabla f(x), y - x \rangle \big|
-       &= \big|g(1) - g(0) - g'(0) \big| & \\
-       &= \bigg| \int_0^1 g'(t) - g'(0)\ dt \bigg| \\
-       &\leq \int_0^1 \big|g'(t) - g'(0)\big|\ dt \\
-       &= \int_0^1 \bigg| \big \langle \nabla f\big((1-t)x + ty\big) - \nabla f(x), y - x \big \rangle \bigg|\ dt \\
-       &\leq \int_0^1 \big \| \nabla f\big((1-t)x + ty\big) - \nabla f(x) \big \|_2\ \| y - x \|_2\ dt \\
-       &\leq \int_0^1 \big ( \gamma\ \|y-x\|_2 \big) \ \big( \|y-x\|_2 \big) t\ dt \\
-       &= \frac{\gamma \|y - x\|_2^2}{2}.
-  \end{align*}
-  \qed
-\end{proofdot}
-
-\begin{plainlemma}
-  \label{lemma:2}
-  Let $x, y, v_i \in \mathbb{R}^d$, $c_i \in \mathbb{R}$, and
-  $|\langle y - x, v_i \rangle| \leq c_i$ for $i = 1$, $\ldots$, $d.$
-  If $M = (v_1$, $\ldots$, $v_d)$ is nonsingular, then
-  $$\|y - x\|_2^2 \leq \frac{1}{\sigma_d^2} \sum_{i=1}^d c_i^2,$$
-  where $\sigma_d$ is the smallest singular value of $M.$
-\end{plainlemma}
-
-\begin{proofdot}
-  Using the facts that $M$ and $M^t$ have the same singular values,
-  and $\|M^tw\|_2 \geq \sigma_d \|w\|_2$, gives
-  \begin{align*}
-    \|y - x\|_2^2 &\leq \frac{\|M^t (y - x)\|_2^2}{\sigma_d^2} \\
-                  &=    \frac{1}{\sigma_d^2} \sum_{i=1}^d \langle y - x, v_i \rangle^2 \\
-                  &\leq \frac{1}{\sigma_d^2} \sum_{i=1}^d c_i^2.
-  \end{align*}
-  \qed
-\end{proofdot}
-
-\begin{plainlemma}
-  \label{lemma:3}
-  Given $f$, $\gamma$, $S$ as in {\it Lemma \ref{lemma:1}}, let $X =
-  \{x_0$, $x_1$, $\ldots$, $x_d\}$ $\subset S$ be the vertices of a
-  $d$-simplex, and let $\hat f(x) = \langle c, x - x_0 \rangle +
-  f(x_0)$, $c \in \mathbb{R}^d$ be the linear function interpolating
-  $f$ on $X.$ Let $\sigma_d$ be the smallest singular value of the
-  matrix $M = (x_1 - x_0$, $\ldots$, $x_d - x_0)$, and $k =
-  \max\limits_{1\ \leq\ j\ \leq\ d} \|x_j - x_0\|_2.$ Then
-  $$\big\|\nabla f(x_0) - c\big\|_2 \leq \sqrt{d} \frac{\gamma k^2}{\sigma_d}.$$
-\end{plainlemma}
-
-\begin{proofdot}
-  Consider $f(x) - \hat f(x)$ along the line segment $z(t) = (1-t)x_0
-  + t x_j$, $0 \leq t \leq 1.$ By Rolle's Theorem, for some $0 < \hat
-  t < 1$, $\big\langle \nabla f\big(z(\ \hat t\ )\big) - c, x_j - x_0
-  \big\rangle = 0.$ Now
-  \begin{align*}
-    \big| \big< \nabla f(x_0) - c, x_j - x_0 \big \rangle \big|
-        &= \big| \big< \nabla f(x_0) - \nabla f\big(z(\ \hat t\ )\big) + \nabla f\big(z(\ \hat t\ )\big) - c, x_j - x_0 \big \rangle \big| \\
-	&= \big| \big\langle \nabla f(x_0) - \nabla f \big(z(\ \hat t\ )\big), x_j - x_0 \big \rangle \big| \\
-	&\leq \big \| \nabla f(x_0) - \nabla f\big(z(\ \hat t\ )\big) \big \|_2 \|x_j - x_0\|_2 \\
-	&\leq \gamma \|x_0 - z(\ \hat t\ )\|_2\ \|x_j - x_0\|_2 \\
-	&\leq \gamma \|x_j - x_0\|_2^2 \leq \gamma k^2, \\
-  \end{align*}
-  for all $1 \leq j \leq d.$ Using {\it Lemma \ref{lemma:2}},
-  $$\big \| \nabla f(x_i) - c \big \|_2^2 \leq \frac{d}{\sigma_d^2}
-  \big( \gamma k^2\big)^2 \implies \big \| \nabla f(x_i) - c \big \|_2
-  \leq \sqrt{d} \frac{\gamma k^2}{\sigma_d}.$$
-  \qed
-\end{proofdot}
-
-\begin{onetheorem}
-  Under the assumptions of {\it Lemma \ref{lemma:1}} and {\it Lemma
-    \ref{lemma:3}}, for $z \in S$,
-  $$ \big|f(z) - \hat f(z)\big| \leq \frac{\gamma \|z - x_0\|_2^2}{2} + \sqrt{d} \frac{\gamma k^2}{\sigma_d} \|z - x_0\|_2.$$
-\end{onetheorem}
-
-\begin{proofdot}
-  Let $v = \nabla f(x_0) - c$, where $\|v\|_2 \leq \sqrt{d} \gamma k^2 / \sigma_d$ by {\it Lemma \ref{lemma:3}}. Now
-  \begin{align*}
-    \big|f(z) - \hat f(z)\big|
-       &= \big|f(z) - f(x_0) - \langle c, z - x_0 \rangle \big| \\ 
-       &= \big|f(z) - f(x_0) - \langle \nabla f(x_0) - v, z - x_0 \rangle \big| \\
-       &= \big|f(z) - f(x_0) - \langle \nabla f(x_0) , z - x_0 \rangle + \langle v , z - x_0 \rangle \big| \\
-       &\leq \big|f(z) - f(x_0) - \langle \nabla f(x_0) , z - x_0 \rangle \big| + \big| \langle v , z - x_0 \rangle \big| \\
-       &\leq \big|f(z) - f(x_0) - \langle \nabla f(x_0) , z - x_0 \rangle \big| + \|v\|_2 \|z - x_0\|_2 \\
-       &\leq \big|f(z) - f(x_0) - \langle \nabla f(x_0), z - x_0 \rangle \big| + \textstyle{\frac{\gamma k^2 \sqrt{d}}{\sigma_d}} \|z - x_0\|_2 \\
-       &\leq \frac{\gamma \|z - x_0\|_2^2}{2} + \sqrt{d}\frac{\gamma k^2}{\sigma_d} \|z - x_0\|_2,
-  \end{align*}
-  where the last inequality follows from {\it Lemma \ref{lemma:1}}.
-  %
-  \qed
-\end{proofdot}
-
-In summary, the approximation error of a linear (simplicial)
-interpolant tends quadratically towards zero when approaching observed
-data only when the diameter of the simplex is also reduced at a
-proportional rate.  Only linear convergence to the true function can
-be achieved in practice, without the incorporation of additional
-observations. Notice that the approximation error is largely
-determined by the spacing of observed data. Predictions made by
-simplices whose vertices are not well-spaced (i.e., have large
-diameter, or are nearly contained in a hyperplane) have higher
-error. In light of this error bound, an empirical evaluation of
-presented algorithms follows.
-
-%% ===================================================================
-\vspace{-2mm}
-\section{Data and Empirical Analysis}
-\label{sec:data}
-
-The theoretical results constructed in Section \ref{sec:theory} for
-(piecewise) linear interpolation are promising and apply directly to
-Delaunay interpolation, however they are difficult to interpret in
-context with approximation algorithms that do not have similar known
-uniform error bounds. For that reason, this paper utilizes five
-different data sets of varying dimension and application to construct
-approximations and compare the accuracy of different techniques.
-
-In the following five subsections the sources and targets of each test
-data set are described, as well as challenges and limitations related
-to approximating these data. The distribution of response values being
-modeled is presented followed by the distribution of approximation
-errors for each algorithm. The plots for all five data sets have the
-same format.
-
-All five data sets are rescaled such that the domain of approximation
-is the unit hypercube. The range of the first four data sets is the
-real numbers, while the range of the fifth data set is the space of
-cumulative distribution functions. All approximation techniques are
-applied to the first four data sets, while only those interpolants
-whose approximations are convex combinations of observed data are
-applied to the final data set.
-
-All approximations are constructed using $k$-fold cross validation as
-described in \cite{kohavi1995study} with $k=10$. This approach
-randomly partitions data into $k$ (nearly) equal sized sets. Each
-algorithm is then evaluated by constructing an approximation over each
-unique union of $k-1$ elements of the partition, making predictions
-for points in the remaining element. As a result, each observed data
-point is used in the construction of $k-1$ different approximations
-and is approximated exactly once. The $k$-fold cross validation method
-is data-efficient and provides an unbiased estimate of the expected
-prediction error \cite{kohavi1995study}, however it should be noted
-that neither this method nor others can provide a universally unbiased
-estimator for the variance of prediction error \cite{bengio2004no}.
-
-In addition to the figures displaying approximation results for each
-data set, tables of accompanying numerical results are located in the
-Appendix. All of the test data sets capture underlying functions that
-are almost certainly stochastic. As described in Section
-\ref{sec:introduction}, regression techniques appear most appropriate
-for these problems. However, typically data grows exponentially more
-sparse with increasing dimension. Given sparse data, regressors tend
-towards interpolation.
-
-%% -------------------------------------------------------------------
-\subsection{Forest Fire ($n = 504$, $d = 12$)}
-
-\begin{figure}
-  \centering
-  \begin{tikzpicture}
-    \node (img)  {\includegraphics[width=0.75\textwidth,]{Fig2.pdf}};
-    \node[below=of img, node distance=1cm, yshift=1cm] {Forest Fire Area Burned};
-    \node[left=of img, node distance=0cm, rotate=90, anchor=center,yshift=-0.7cm] {Count};
-  \end{tikzpicture}
-  \caption{Histogram of forest fire area burned under recorded weather
-    conditions. The data is presented on a $\ln$ scale because
-    most values are small with exponentially fewer fires on record
-    that burn large areas.}
-  \label{fig:hist-forest-fire}
-\end{figure}
-
-\begin{figure}
-  \centering
-  \begin{tikzpicture}
-    \node (img)  {\includegraphics[width=0.8\textwidth,]{Fig3.pdf}};
-    \node[below=of img, node distance=1cm, yshift=1cm] {Area Burned Error};
-  \end{tikzpicture}
-  \caption{All models are applied to approximate the amount of area
-    that would be burned given environment conditions. $10$-fold cross
-    validation as described in the beginning of Section \ref{sec:data}
-    is used to evaluated each algorithm. This results in exactly one
-    prediction from each algorithm for each data point. These boxes
-    depict the median (middle bar), median $95\%$ confidence interval
-    (cones), quartiles (box edges), fences at $3/2$ interquartile
-    range (whiskers), and outliers (dots) of absolute prediction error
-    for each model. Similar to Figure \ref{fig:hist-forest-fire}, the
-    errors are presented on a $\ln$ scale. The numerical data
-    corresponding to this figure is provided in Table
-    \ref{table:error-forest-fire} in the Appendix.}
-  \label{fig:error-forest-fire}
-\end{figure}
-
-The forest fire data set \cite{cortez2007data} describes the area of
-Montesinho park burned over months of the year along with
-environmental conditions. The twelve dimensions being used to model
-burn area are the $x$ and $y$ spatial coordinates of burns in the
-park, month of year (mapped to $x$, $y$ coordinates on a unit circle),
-the FFMC, DMC, DC, and ISI indices (see source for details), the
-temperature, relative humidity, wind speed, and outdoor rain. The
-original analysis of this data set demonstrated it to be difficult to
-model, likely due to the skew in response values.
-
-As suggested by Figure \ref{fig:error-forest-fire}, the SVR has the
-lowest absolute prediction errors for $80\%$ of the data, with MLP and
-Delaunay being the nearest overall competitors. The effectiveness of
-SVR on this data suggests the underlying function can be defined by
-relatively few parameters, as well as the importance of capturing the
-low-burn-area data points.
-%% -------------------------------------------------------------------
-
-%% -------------------------------------------------------------------
-\subsection{Parkinson's Telemonitoring ($n = 5875$, $d = 19$)}
-
-\begin{figure}
-  \centering
-  \begin{tikzpicture}
-    \node (img)  {\includegraphics[width=0.75\textwidth,]{Fig4.pdf}};
-    \node[below=of img, node distance=1cm, yshift=1cm] {Parkinson's Total UPDRS Score};
-    \node[left=of img, node distance=0cm, rotate=90, anchor=center,yshift=-0.7cm] {Count};
-  \end{tikzpicture}
-  \caption{Histogram of the Parkinson's patient total UPDRS clinical
-    scores that will be approximated by each algorithm.}
-  \label{fig:hist-parkinsons}
-\end{figure}
-
-\begin{figure}
-  \centering
-  \begin{tikzpicture}
-    \node (img)  {\includegraphics[width=0.8\textwidth,]{Fig5.pdf}};
-    \node[below=of img, node distance=1cm, yshift=1cm] {Total UPDRS Score Error};
-  \end{tikzpicture}
-  \caption{All models are applied to approximate the total UPDRS score
-    given audio features from patients' home life, using $10$-fold
-    cross validation. These boxes depict the median (middle bar),
-    median $95\%$ confidence interval (cones), quartiles (box edges),
-    fences at $3/2$ interquartile range (whiskers), and outliers
-    (dots) of absolute prediction error for each model. The numerical
-    data corresponding to this figure is provided in Table
-    \ref{table:error-parkinsons} in the Appendix.}
-  \label{fig:error-parkinsons}
-\end{figure}
-
-The second data set for evaluation \cite{tsanas2010accurate} is
-derived from a speech monitoring study with the intent to
-automatically estimate Parkinson's disease symptom development in
-Parkinson's patients. The function to be approximated is a
-time-consuming clinical evaluation measure referred to as the UPDRS
-score. The total UPDRS score given by a clinical evaluation is
-estimated through 19 real numbers generated from biomedical voice
-measures of in-home sound recordings.
-
-Figure \ref{fig:error-parkinsons} shows the ShepMod algorithm has the
-lowest minimum, first quartile, and median of absolute errors for this
-problem, while providing the best prediction $66\%$ of the time. The
-MLP has the lowest third quartile and provides the best prediction for
-$32\%$ of approximations. The dominance of ShepMod may be due in part
-to regular-interval total UPDRS scores provided by clinicians,
-favoring a nearest-neighbor strategy of prediction.
-%% -------------------------------------------------------------------
-
-%% -------------------------------------------------------------------
-\subsection{Australian Daily Rainfall Volume ($n = 2609$, $d = 23$)}
-
-\begin{figure}
-  \centering
-  \begin{tikzpicture}
-    \node (img)  {\includegraphics[width=0.75\textwidth,]{Fig6.pdf}};
-    \node[below=of img, node distance=1cm, yshift=1cm] {Daily Rainfall in Sydney};
-    \node[left=of img, node distance=0cm, rotate=90, anchor=center,yshift=-0.7cm] {Count};
-  \end{tikzpicture}
-  \caption{Histogram of daily rainfall in Sydney, Australia, presented
-    on a $\ln$ scale because the frequency of larger amounts of
-    rainfall is significantly less. There is a peak in occurrence of
-    the value $0$, which has a notable effect on the resulting model
-    performance.}
-  \label{fig:hist-weather}
-\end{figure}
-
-\begin{figure}
-  \centering
-  \begin{tikzpicture}
-    \node (img)  {\includegraphics[width=0.8\textwidth,]{Fig7.pdf}};
-    \node[below=of img, node distance=1cm, yshift=1cm] {Sydney Rainfall Tomorrow Error};
-  \end{tikzpicture}
-  \caption{All models are applied to approximate the amount of
-    rainfall expected on the next calendar day given various sources
-    of local meteorological data, using $10$-fold cross validation.
-    These boxes depict the median (middle bar), median $95\%$
-    confidence interval (cones), quartiles (box edges), fences at
-    $3/2$ interquartile range (whiskers), and outliers (dots) of
-    absolute prediction error for each model. The errors are presented
-    on a $\ln$ scale, mimicking the presentation in Figure
-    \ref{fig:hist-weather}. The numerical data corresponding to this
-    figure is provided in Table \ref{table:error-weather} in the
-    Appendix.}
-  \label{fig:error-weather}
-\end{figure}
-
-The third data set for the total daily rainfall in Sydney, Australia
-\cite{williams2009rattle} provides a slightly higher dimensional
-challenge for the interpolants and regressors. This data is composed
-of many meteorological readings from the region in a day including:
-min and max temperatures, sunshine, wind speed directions (converted
-to coordinates on a circle), wind speeds, and humidities throughout
-the day, day of the year (converted to coordinates on a circle), and
-the model must predict the amount of rainfall tomorrow.
-
-While Figure \ref{fig:hist-weather} makes MARS look far better than
-other techniques, it only provides the best prediction for $11\%$ of
-points. The MLP has the lowest absolute error for $56\%$ of points and
-LSHEP is best for $28\%$. MARS likely achieves such a low first
-quartile due to the high occurrence of the value zero in the data.
-%% -------------------------------------------------------------------
-
-%% -------------------------------------------------------------------
-\subsection{Credit Card Transaction Amount ($n = 5562$, $d = 28$)}
-
-\begin{figure}
-  \centering
-  \begin{tikzpicture}
-    \node (img)  {\includegraphics[width=0.75\textwidth,]{Fig8.pdf}};
-    \node[below=of img, node distance=1cm, yshift=1cm] {Credit Card Transaction Amount};
-    \node[left=of img, node distance=0cm, rotate=90, anchor=center,yshift=-0.7cm] {Count};
-  \end{tikzpicture}
-  \caption{Histogram of credit card transaction amounts, presented on
-    a $\ln$ scale. The data contains a notable frequency peak around
-    $\$1$ transactions. Fewer large purchases are made, but some large
-    purchases are in excess of five orders of magnitude greater than
-    the smallest purchases.}
-  \label{fig:hist-credit-card}
-\end{figure}
-
-\begin{figure}
-  \centering
-  \begin{tikzpicture}
-    \node (img)  {\includegraphics[width=0.8\textwidth,]{Fig9.pdf}};
-    \node[below=of img, node distance=1cm, yshift=1cm] {Transaction Amount Error};
-  \end{tikzpicture}
-  \caption{All models are applied to approximate the expected
-    transaction amount given transformed (and obfuscated) vendor and
-    customer-descriptive features, using $10$-fold cross validation.
-    These boxes depict the median (middle bar), median $95\%$
-    confidence interval (cones), quartiles (box edges), fences at
-    $3/2$ interquartile range (whiskers), and outliers (dots) of
-    absolute prediction error for each model. The absolute errors in
-    transaction amount predictions are presented on a $\ln$ scale,
-    just as in Figure \ref{fig:hist-credit-card}. The numerical data
-    corresponding to this figure is provided in Table
-    \ref{table:error-credit-card} in the Appendix.}
-  \label{fig:error-credit-card}
-\end{figure}
-
-The fourth test data set, and the final with a real-valued range, is a
-collection of credit card transactions
-\cite{pozzolo2015calibrating}. The provided data carries no direct
-real-world meaning, being the output of a principle component analysis
-on the original hidden source data. This obfuscation is done to
-protect the information of the credit card users. This data has the
-largest dimension of all considered, at 28. A model for this data
-predicts the transaction amount given the vector of principle
-component information.
-
-As can be seen in Figure \ref{fig:error-credit-card}, the MLP
-outperforms all other algorithms at the first, second, third, and
-fourth quartiles. The MLP produces the lowest absolute error
-prediction for $80\%$ of transactions, Delaunay bests the remaining
-$20\%$. It is likely that with less data, Delaunay would be the best
-performer.
-%% -------------------------------------------------------------------
-
-%% -------------------------------------------------------------------
-\subsection{High Performance Computing I/O ($n = 3016$, $d = 4$)}
-
-\begin{figure}
-  \centering
-  \begin{tikzpicture}
-    \node (img)  {\includegraphics[width=0.75\textwidth,]{Fig10.pdf}};
-    \node[below=of img, node distance=1cm, yshift=1cm] {I/O Read Throughput};
-    \node[left=of img, node distance=0cm, rotate=90, anchor=center,yshift=-0.7cm] {Count};
-  \end{tikzpicture}
-  \caption{Histogram of the raw throughput values recorded during all
-    IOzone tests across all system configurations. The distribution is
-    skewed right, with few tests having significantly higher
-    throughput than most others. The data is presented on a $\ln$
-    scale.}
-  \label{fig:hist-throughput}
-\end{figure}
-
-\begin{figure}
-  \centering
-  \begin{tikzpicture}
-    \node (img)  {\includegraphics[width=0.8\textwidth,]{Fig11.pdf}};
-    \node[below=of img, node distance=1cm, yshift=1cm] {Read Throughput KS Statistic};
-  \end{tikzpicture}
-  \caption{The models directly capable of predicting distributions are
-    applied to predicting the expected CDF of I/O throughput at a
-    previously unseen system configuration, using $10$-fold cross
-    validation. The KS statistic between the observed distribution at
-    each system configuration and the predicted distribution is
-    recorded and presented above. Note that the above figure is
-    \textit{not} log-scaled like Figure \ref{fig:hist-throughput}. The
-    numerical data corresponding to this figure is provided in Table
-    \ref{table:error-throughput} in the Appendix.}
-  \label{fig:error-throughput}
-\end{figure}
-
-The final of five data sets is derived from \cite{cameron2019moana},
-which provides four-dimensional distribution data by executing the
-IOzone benchmark \cite{iozone} on a computer system and varying the
-system's file size, record size, thread count, and CPU frequency. At
-each configuration, IOzone samples the I/O file-read throughput (in
-bytes per second) 150 times. Empirical distribution function points
-are computed from each set of 150 executions, which are interpolated
-with a piecewise cubic Hermite interpolating polynomial
-\cite{fritsch1980monotone} to approximate the CDF. All interpolation
-algorithms with the exception of LSHEP are used to approximate these
-CDFs from system configurations.
-
-Delaunay achieves the lowest KS statistic for $62\%$ of
-approximations, while Voronoi is best for the remaining $38\%$. Figure
-\ref{fig:error-throughput} shows that while Delaunay may have more
-best predictions, the behavior of Voronoi may be preferable.
-%% -------------------------------------------------------------------
-
-%% -------------------------------------------------------------------
-\begin{figure}
-  \centering
-  \begin{tikzpicture}
-    \node (img)  {\includegraphics[width=0.8\textwidth]{Fig12.pdf}};
-    \node[below=of img, node distance=1cm, yshift=1cm] {KS Statistic for Predicted vs. Actual};
-    \node[left=of img, node distance=0cm, rotate=90, anchor=center,yshift=-0.7cm] {Count of KS Statistic};
-  \end{tikzpicture}
-  \caption{Histograms of the prediction error for each interpolant
-    that produces predictions as convex combinations of observed data,
-    using $10$-fold cross validation. The histograms show the KS
-    statistics for the predicted throughput distribution versus the
-    actual throughput distribution. The four vertical lines represent
-    cutoff KS statistics given $150$ samples for commonly used
-    $p$-values 0.05, 0.01, 0.001, $10^{-6}$, respectively. All
-    predictions to the right of a vertical line represent CDF
-    predictions that are significantly different (by respective
-    $p$-value) from the actual distribution according to the KS
-    test. The numerical counterpart to this figure is presented in
-    Table \ref{table:null-hypothesis-results}.}
-  \label{fig:throughput-prediction}
-\end{figure}
-
-\begin{table}
-  \renewcommand{\arraystretch}{1.3}
-  \centering
-  \begin{tabular}{c|c|c|c|c}
-     & $p = .05$ & $p = .01$ & $p = .001$ & $p = 10^{-6}$\\
-    \hline
-    Delaunay & $\mathbf{50.3}\%$ & $\mathit{43.5}\%$ & $\mathit{36.2}\%$ & $\mathit{24.7}\%$\\
-    ShepMod & $\mathit{51.4}\%$ & $44.8\%$ & $38.1\%$ & $27.7\%$\\
-    Voronoi & $52.6\%$ & $\mathbf{43.4}\%$ & $\mathbf{34.4}\%$ & $\mathbf{19.1}\%$\\
-    BoxSpline & $99.5\%$ & $98.5\%$ & $96.6\%$ & $89.5\%$\\
-  \end{tabular}
-  \caption{Numerical counterpart of the histogram data presented in
-    Figure \ref{fig:throughput-prediction}. The columns display the
-    percent of null hypothesis rejections by the KS-test when provided
-    different selections of $p$-values for each algorithm. The
-    algorithm with the lowest rejection rate at each $p$ is boldface,
-    while the second lowest is italicized.}
-  \label{table:null-hypothesis-results}
-\end{table}
-
-%% ===================================================================
-
-Figure \ref{fig:throughput-prediction} expands on the KS statistic
-results presented in Figure \ref{fig:error-throughput}. Agglomerate
-errors for each algorithm resemble a Gamma distribution. The
-percentages of significant prediction errors with varying $p$-values
-are on display in Table \ref{table:null-hypothesis-results}. When
-considering the $p=0.001$ results for each technique, a little over
-one third of the predicted CDFs are significantly different from the
-measured (and presumed) correct CDFs. However, it should be noted that
-with 150 samples, the error of an empirical distribution function
-(EDF) can reasonably be upwards of $.1$, which serves as a rough
-estimate for the lower limit of achievable error. Globally, only a
-third of Voronoi predictions fail to capture \textit{all} of the
-characteristics of the CDFs at new system configurations.
-
-\section{Discussion}
-\label{sec:discussion}
-
-\begin{table}
-  \centering
-  \begin{tabular}{c|r@{.}l|r@{.}l|r@{.}l}
-    Algorithm & \multicolumn{2}{c|}{Avg. \% Best} & \multicolumn{2}{c|}{\multilinecell{Avg. Fit or \\ Prep. Time (s)}} & \multicolumn{2}{c}{Avg. App. Time (s)}\\
-    \hline
-    MARS & \quad\quad$4$&$6$ & \,\quad$20$&$0$ & \quad\quad$0$&$001$\\
-    SVR & $\mathit{21}$&$\mathit{1}$ & $\mathbf{0}$&$\mathbf{5}$ & $\mathbf{0}$&$\mathbf{0001}$\\
-    MLP & $\mathbf{42}$&$\mathbf{1}$ & $200$&$0$ & $0$&$001$\\
-    Delaunay & $6$&$0$ & $1$&$0$ & $1$&$0$\\
-    ShepMod & $18$&$4$ & $\mathit{0}$&$\mathit{7}$ & $\mathbf{0}$&$\mathbf{0001}$\\
-    LSHEP & $8$&$4$ & $2$&$0$ & $\mathbf{0}$&$\mathbf{0001}$\\
-    Voronoi & $0$&$5$ & $1$&$0$ & $0$&$04$\\
-    BoxSpline & $0$&$2$ & $0$&$8$ & $\mathit{0}$&$\mathit{0005}$\\
-  \end{tabular}
-  \caption{This average of Appendix Tables
-    \ref{table:best-forest-fire}, \ref{table:best-parkinsons},
-    \ref{table:best-weather}, and \ref{table:best-credit-card}
-    provides a gross summary of overall results. The columns display
-    (weighted equally by data set, \textit{not} points) the average
-    frequency with which any algorithm provided the lowest absolute
-    error approximation, the average time to fit/prepare, and the
-    average time required to approximate one point. The times have
-    been rounded to one significant digit, as reasonably large
-    fluctuations may be observed due to implementation hardware.
-    Interpolants provide the lowest error approximation for nearly one
-    third of all data, while regressors occupy the other two
-    thirds. This result is obtained without any customized tuning or
-    preprocessing to maximize the performance of any given
-    algorithm. In practice, tuning and preprocessing may have large
-    effects on approximation performance.}
-  \label{table:avg-performance}
-\end{table}
-
-Table \ref{table:avg-performance} summarizes results across the four
-test data sets with real-valued ranges. The interpolants discussed in
-this paper produce the \textit{best} approximations roughly one third
-of the time, and produce competitive approximations for almost all
-data sets. These test problems are almost certainly
-\textit{stochastic} in nature, but the high dimension leads to greater
-data sparsity and model construction cost, making interpolation more
-competitive.
-
-The major advantages to interpolation lie in the near absence of
-\textit{fit} time. Delaunay, LSHEP, and ShepMod all require pairwise
-distance calculations, for numerical robustness (Delaunay) and to
-determine the radii of influence for data points (LSHEP and
-ShepMod). At least hundreds, and sometimes hundreds of thousands of
-predictions can be made by the interpolants before the most widely
-used regressor (MLP) finishes fitting these relatively small data
-sets.  However, the computational complexities of all interpolants
-presented are greater than linear in either dimension or number of
-points, whereas the regressors' nonlinear complexity in dimension
-generally comes from the model fitting optimization.
-
-The new theoretical results presented in Section \ref{sec:theory}
-directly apply to Delaunay interpolation, however the performance of
-Delaunay does not appear significantly better than other algorithms on
-these data sets.  This observation may be due to the stochastic nature
-of the data, but it also speaks to the power of the approximations
-generated by the different interpolation methods.  The strong
-performance of other interpolants suggests that theoretical results
-similar to those presented in this work can be achieved for the other
-interpolants under reasonable assumptions.
-
-Finally, most of the interpolants presented in this work benefit from
-the ability to model any function over real $d$-tuples with a range
-that is closed under convex combinations. The results of the
-distribution prediction case study indicate that interpolants can
-effectively predict CDFs. The error analysis for that work relies on
-the KS statistic, which captures the worst part of any prediction and
-hence provides a conservatively large estimate of approximation
-error. The average absolute errors in the predicted CDFs are always
-lower than the KS statistics. However, the KS statistic was chosen as
-a metric because of the important surrounding statistical theory. A
-nonnegligible volume of predictions provide impressively low levels of
-average absolute error in that final case study.
-
-\section{Conclusion}
-\label{sec:conclusion}
-
-The major contributions of this work are: 1) new uniform theoretical
-error bounds for piecewise linear interpolation in arbitrary dimension
-(Section \ref{sec:theory}); 2) an empirical evaluation across
-real-world problems that demonstrates interpolants produce
-competitively accurate models of multivariate phenomenon when compared
-with common regressors for sparse, moderately high dimensional
-problems (Section \ref{sec:data}); and 3) a demonstration that some
-interpolants generalize to interpolation in function spaces (Section
-\ref{sec:error}), preserving monotonicity (with CDFs, e.g.), neither
-of which common regressors can do.
-
-The various interpolants discussed in this paper have been
-demonstrated to effectively approximate multivariate phenomena up to
-$30$ dimensions. The underlying constructions are theoretically
-straightforward, interpretable, and yield reasonably accurate
-predictions. Most of the interpolants' computational complexities make
-them particularly suitable for applications in even higher
-dimension. The major benefits of interpolation are seen when only a
-small number of approximations ($\leq 1000$) are made from data and
-when there are relatively few data points for the dimension (for
-empirical results presented, $\log_d n \leq 5$). These findings
-encourage broader application of interpolants to multivariate
-approximation in science
-
-%\begin{acknowledgements}
-%If you'd like to thank anyone, place your comments here
-%and remove the percent signs.
-%\end{acknowledgements}
-
-%%====================================================================
-%%====================================================================
-
-\bibliographystyle{spmpsci}      % mathematics and physical sciences
-\bibliography{paper}
-
-\newpage
-
-\begin{appendix}
-\section*{Appendix}
-\label{sec:appendix}
+% Add your appendices here. You must leave the appendices enclosed in the appendices environment in order for the table of contents to be correct.
+\begin{appendices}
+  \chapter{Error Bound Appendices} \label{app:error}
 
@@
   results serve as a demonstration of conceptual validity.
-
-\vspace*{\fill}
 
@@
   \end{table}
-
-\vspace*{\fill}
-\newpage
 
@@
 
-\end{appendix}
+\end{appendices}
 
 \end{document}
+
+%****************************************************************************
+% Below are some general suggestions for writing your dissertation:
+%
+% 1. Label everything with a meaningful prefix so that you
+%    can refer back to sections, tables, figures, equations, etc.
+%    Usage \label{<prefix>:<label_name>} where some suggested
+%    prefixes are:
+%			ch: Chapter
+%     		se: Section
+%     		ss: Subsection
+%     		sss: Sub-subsection
+%			app: Appendix
+%     		ase: Appendix section
+%     		tab: Tables
+%     		fig: Figures
+%     		sfig: Sub-figures
+%     		eq: Equations
+%
+% 2. The VTthesis class provides for natbib citations. You should upload
+%	 one or more *.bib bibtex files. Suppose you have two bib files: some_refs.bib and 
+%    other_refs.bib.  Then your bibliography line to include them
+%    will be:
+%      \bibliography{some_refs, other_refs}
+%    where multiple files are separated by commas. In the body of 
+%    your work, you can cite your references using natbib citations.
+%    Examples:
+%      Citation                     Output
+%      -------------------------------------------------------
+%      \cite{doe_title_2016}        [18]
+%      \citet{doe_title_2016}       Doe et al. [18]
+%      \citet*{doe_title_2016}      Doe, Jones, and Smith [18]
+%
+%    For a complete list of options, see
+%      https://www.ctan.org/pkg/natbib?lang=en
+%
+% 3. Here is a sample table. Notice that the caption is centered at the top. Also
+%    notice that we use booktabs formatting. You should not use vertical lines
+%    in your tables.
+% 
+%				\begin{table}[htb]
+%					\centering
+%					\caption{Approximate computation times in hh:mm:ss for full order 						versus reduced order models.}
+%					\begin{tabular}{ccc}
+%						\toprule
+%						& \multicolumn{2}{c}{Computation Time}\\
+%						\cmidrule(r){2-3}
+%						$\overline{U}_{in}$ m/s & Full Model & ROM \\
+%						\midrule
+%						0.90 & 2:00:00 & 2:08:00\\
+%						0.88 & 2:00:00 & 0:00:03\\
+%						0.92 & 2:00:00 & 0:00:03\\
+%						\midrule
+%						Total & 6:00:00 & 2:08:06\\
+%						\bottomrule
+%					\end{tabular}
+%					\label{tab:time_rom}
+%				\end{table}
+% 
+% 4. Below are some sample figures. Notice the caption is centered below the
+%    figure.
+%    a. Single centered figure:
+%					\begin{figure}[htb]
+%						\centering
+%						\includegraphics[scale=0.5]{my_figure.eps}
+%						\caption{Average outlet velocity magnitude given an average  
+%				        input velocity magnitude of 0.88 m/s.} 
+%						\label{fig:output_rom}
+%					\end{figure}
+%    b. Two by two grid of figures with subcaptions
+%					\begin{figure}[htb]
+%						\centering
+%						\begin{subfigure}[h]{0.45\textwidth}
+%							\centering
+%							\includegraphics[scale=0.4]{figure_1_1.eps}
+%							\caption{Subcaption number one}
+%							\label{sfig:first_subfig}
+%						\end{subfigure}
+%						\begin{subfigure}[h]{0.45\textwidth}
+%							\centering
+%							\includegraphics[scale=0.4]{figure_1_2.png}
+%							\caption{Subcaption number two}
+%							\label{sfig:second_subfig}
+%						\end{subfigure}
+%
+%						\begin{subfigure}[h]{0.45\textwidth}
+%							\centering
+%							\includegraphics[scale=0.4]{figure_2_1.pdf}
+%							\caption{Subcaption number three}
+%							\label{sfig:third_subfig}
+%						\end{subfigure}
+%						\begin{subfigure}[h]{0.45\textwidth}
+%							\centering
+%							\includegraphics[scale=0.4]{figure_2_2.eps}
+%							\caption{Subcaption number four}
+%							\label{sfig:fourth_subfig}
+%						\end{subfigure}
+%						\caption{Here is my main caption describing the relationship 							between the 4 subimages}
+%						\label{fig:main_figure}
+%					\end{figure}
+%
+%----------------------------------------------------------------------------
+%
+% The following is a list of definitions and packages provided by VTthesis:
+%
+% A. The following packages are provided by the VTthesis class:
+%      amsmath, amsthm, amssymb, enumerate, natbib, hyperref, graphicx, 
+%      tikz (with shapes and arrows libraries), caption, subcaption,
+%      listings, verbatim
+%
+% B. The following theorem environments are defined by VTthesis:
+%      theorem, proposition, lemma, corollary, conjecture
+% 
+% C. The following definition environments are defined by VTthesis:
+%      definition, example, remark, algorithm
+%
+%----------------------------------------------------------------------------
+%
+%  I hope this template file and the VTthesis class will keep you from having 
+%  to worry about the formatting and allow you to focus on the actual writing.
+%  Good luck, and happy writing.
+%    Alan Lattimer, VT, 2016
+%
+%****************************************************************************

### 2019-05-03 Code_[2018-06-11]_Stable_Arbitrary_Box_Splines/Testing_Fortran/analyze_results.py

@@ -1,148 +1,0 @@
-import os
-from util.data import Data
-
-all_data_file = "performance_data.pkl"
-if not os.path.exists(all_data_file):
-    d1 = Data.load("Pima.csv")
-    d2 = Data.load("Navajo.csv")
-    d1["Computer"] = ("Pima" for row in d1)
-    d2["Computer"] = ("Navajo" for row in d2)
-    d = d1[:]
-    d += d2
-    d.save("performance_data.csv")
-    d.save(all_data_file)
-
-# Load data from file and reduce it to the unique tests that were run.
-d = Data.load(all_data_file)
-tests = d.names[:-2]
-d = d[tests].unique().collect(d)
-d["Num Trials"] = map(len, d["Time"])
-d[d["Optimization"] == None, "Optimization"] = ""
-d["Computer"] = (c[0] for c in d["Computer"])
-d["Time"] = d.pop("Time")
-
-# Print out summary information about the data.
-d.summarize()
-print(d)
-print()
-
-# Check for nonuniform numbers of executions for different settings.
-if len(set(d["Num Trials"])) > 1:
-    print()
-    print("RUNS ARE STILL NEEDED..")
-    to_run = d[d["Num Trials"] < 20]
-    num_trials = list(to_run["Num Trials"])
-    to_run = to_run[to_run.names[:-3]]
-    to_run["Runs Needed"] = (20-v for v in num_trials)
-    # print(to_run)
-    c = to_run.names.index("Compiler")
-    o = to_run.names.index("Optimization")
-    v = to_run.names.index("Version")
-    to_run = to_run["Compiler", "Optimization", "Version"].unique().collect(to_run)
-    to_run["Runs Needed"] = (sorted(set(runs))[-1] for runs in to_run["Runs Needed"])
-    to_run.pop("Multiplicity")
-    to_run.pop("Element")
-    to_run.pop("Num Points")
-    print(to_run)
-    for (c,o,v,r) in to_run:
-        string = "%s -lblas -llapack %s -o test real_precision.f90 test_boxspline.f90 %s"
-        for i in range(r):
-            print(string%(c,o,v))
-    exit()
-
-# 
-from util.stats import cdf_fit_func
-from util.plot import Plot, multiplot
-import numpy as np
-
-legend_settings = dict(
-    xanchor = "center",
-    yanchor = "top",
-    x = .5,
-    y = -.18,
-    orientation = "h",
-)
-
-# Break up the data by compiler
-ifort = d[d["Compiler"] == "ifort"]
[…]

### 2019-05-03 Code_[2018-06-11]_Stable_Arbitrary_Box_Splines/Testing_Fortran/bs-automatic.f90

@@ -1,605 +1,0 @@
-! This file (boxspline.f90) contains the subroutine BOXSPLEV that can
-! be used to evaluate a box-spline, defined by its associated
-! direction vector set, at given evaluation points, as well as the
-! module REAL_PRECISION defining the real precision.
-! 
-! ====================================================================
-! MODULE REAL_PRECISION  ! HOMPACK90 module for 64-bit arithmetic.
-!   INTEGER, PARAMETER :: R8=SELECTED_REAL_KIND(13)
-! END MODULE REAL_PRECISION
-
-SUBROUTINE BOXSPLEV(UNIQUE_DVECS, DVEC_MULTS, EVAL_PTS, BOX_EVALS, ERROR)
-  ! BOXSPLEV evaluates a box-spline, defined by a direction vector set
-  ! in dimension S, at given evaluation points.
-  ! 
-  ! This implementation uses the numerically consistent algorithm for
-  ! evaluating box-splines originally presented in [1]. Most notably,
-  ! the evaluation of the box-spline near the boundaries of polynomial
-  ! pieces does not exhibit the random behavior that is seen in the
-  ! naive recursive implementation. As described in [1], the
-  ! computational complexity for direction vector sets with repeated
-  ! direction vectors is reduced from the naive recursive
-  ! implementation complexity
-  !     O( 2^{n - s} {n! / s!} ),
-  ! where "n" is the number of direction vectors and "s" is the
-  ! dimension, to
-  !     O( (k)^{n - s} ),
-  ! where "k" is the number of unique direction vectors, "n" is the
-  ! total number of direction vectors, and "s" is the dimension.
-  ! 
-  ! The memory complexity of the implementation provided here is
-  !     O( s k + n m s + n^2 ),
-  ! where "s" is the dimension, "k" is the number of unique direction
[…]

### 2019-05-03 Code_[2018-06-11]_Stable_Arbitrary_Box_Splines/Testing_Fortran/bs-manual.f90

@@ -1,634 +1,0 @@
-! This file (boxspline.f90) contains the subroutine BOXSPLEV that can
-! be used to evaluate a box-spline, defined by its associated
-! direction vector set, at given evaluation points, as well as the
-! module REAL_PRECISION defining the real precision.
-! 
-! ====================================================================
-MODULE REAL_PRECISION  ! HOMPACK90 module for 64-bit arithmetic.
-  INTEGER, PARAMETER :: R8=SELECTED_REAL_KIND(13)
-END MODULE REAL_PRECISION
-
-SUBROUTINE BOXSPLEV(UNIQUE_DVECS, DVEC_MULTS, EVAL_PTS, BOX_EVALS, ERROR)
-  ! BOXSPLEV evaluates a box-spline, defined by a direction vector set
-  ! in dimension S, at given evaluation points.
-  ! 
-  ! This implementation uses the numerically consistent algorithm for
-  ! evaluating box-splines originally presented in [1]. Most notably,
-  ! the evaluation of the box-spline near the boundaries of polynomial
-  ! pieces does not exhibit the random behavior that is seen in the
-  ! naive recursive implementation. As described in [1], the
-  ! computational complexity for direction vector sets with repeated
-  ! direction vectors is reduced from the naive recursive
-  ! implementation complexity
-  !     O( 2^{n - s} {n! / s!} ),
-  ! where "n" is the number of direction vectors and "s" is the
-  ! dimension, to
-  !     O( (k)^{n - s} ),
-  ! where "k" is the number of unique direction vectors, "n" is the
-  ! total number of direction vectors, and "s" is the dimension.
-  ! 
-  ! The memory complexity of the implementation provided here is
-  !     O( s k + n m s + n^2 ),
-  ! where "s" is the dimension, "k" is the number of unique direction
[…]

### 2019-05-03 Disseration/Code/cubic.py

@@ -1,0 +1,583 @@
+
+# Custom exception for when there are data errors.
+class BadUsage(Exception): pass
+class BadData(Exception): pass
+class BadValues(Exception): pass
+class UnexpectedError(Exception): pass
+
+# Given three points, this will solve the equation for the quadratic
+# function which interpolates all 3 values. Returns coefficient 3-tuple.
+def solve_quadratic(x, y):
+    if len(x) != len(y): raise(BadUsage("X and Y must be the same length."))
+    if len(x) != 3:      raise(BadUsage(f"Exactly 3 (x,y) coordinates must be given, received '{x}'."))
+    x1, x2, x3 = x
+    y1, y2, y3 = y
+    a = -((-x2 * y1 + x3 * y1 + x1 * y2 - x3 * y2 - x1 * y3 + x2 * y3)/((-x1 + x2) * (x2 - x3) * (-x1 + x3)))
+    b = -(( x2**2 * y1 - x3**2 * y1 - x1**2 * y2 + x3**2 * y2 + x1**2 * y3 - x2**2 * y3)/((x1 - x2) * (x1 - x3) * (x2 - x3)))
+    c = -((-x2**2 * x3 * y1 + x2 * x3**2 * y1 + x1**2 * x3 * y2 - x1 * x3**2 * y2 - x1**2 * x2 * y3 + x1 * x2**2 * y3)/((x1 - x2) * (x1 - x3) * (x2 - x3)))
+    return (a,b,c)
+
+# Given the left and right slope ratios (alpha and beta in paper),
+# compute a new left and right slope ratio that rest in the feasible
+# region for a monotonic cubic interpolating polynomial.
+# 
+#   mode -- 1 (curvy)   Project onto the box inside of "3".
+#           2 (default) Project onto circle of radius "3".
+#           3 ( ~~ )    Project onto line "y = 3 - x".
+#           4 (linear)  Project onto max of "y = 3 - 2x" and "x = 3 - 2y".
+# 
+#   project -- 1 uses the intercept of a line with the region.
+#              2 uses the closest point on the region boundary.
+# 
+def compute_feasible(left_ratio, right_ratio, mode, project):
+    # First check to make sure we have valid ratios (no division by 0).
+    if (max(left_ratio, right_ratio) == 0):
+        pass
+    # Project the ratios onto the appropriate surface.
+    elif (mode == 1):
+        # Option 1: Project onto the square of side length 3.
+        if project == 1:
+            # Determine which line segment it will project onto.
[…]

### 2019-05-03 Report_[2018-07-27]_Box_Spline_Peformance/Testing_Fortran/bs-dynamic.f90

@@ -1,0 +1,662 @@
+! This file (boxspline.f90) contains the subroutine BOXSPLEV that can
+! be used to evaluate a box-spline, defined by its associated
+! direction vector set, at given evaluation points, as well as the
+! module REAL_PRECISION defining the real precision.
+! 
+! ====================================================================
+! MODULE REAL_PRECISION  ! HOMPACK90 module for 64-bit arithmetic.
+!   INTEGER, PARAMETER :: R8=SELECTED_REAL_KIND(13)
+! END MODULE REAL_PRECISION
+
+SUBROUTINE BOXSPLEV(DVECS, DVEC_MULTS, EVAL_PTS, BOX_EVALS, ERROR)
+  ! BOXSPLEV evaluates a box-spline, defined by a direction vector set
+  ! in dimension S, at given evaluation points.
+  ! 
+  ! This implementation uses the numerically consistent algorithm for
+  ! evaluating box-splines originally presented in [1]. Most notably,
+  ! the evaluation of the box-spline near the boundaries of polynomial
+  ! pieces does not exhibit the random behavior that is seen in the
+  ! naive recursive implementation. As described in [1], the
+  ! computational complexity for direction vector sets with repeated
+  ! direction vectors is reduced from the naive recursive
+  ! implementation complexity
+  !     O( 2^{n - s} {n! / s!} ),
+  ! where "n" is the number of direction vectors and "s" is the
+  ! dimension, to
+  !     O( (2k)^{n - s} ),
+  ! where "k" is the number of unique direction vectors, "n" is the
+  ! total number of direction vectors, and "s" is the dimension.
+  ! 
+  ! The memory complexity of the implementation provided here is
+  !     O( s k + n m s + n^2 ),
+  ! where "s" is the dimension, "k" is the number of unique direction
+  ! vectors, "n" is the number of direction vectors, and "m" is the
[…]

### 2019-05-03 Disseration/prelim.tex
Updated abstract to be more clear.

@@
 % The abstract is required and should be <=250 words for thesis, <=350 words for dissertation.
-\abstract{Approximation is an important problem.
+\abstract{Function approximation is an important problem.
 This work presents applications of interpolants to modeling random variables.
 Specifically, this work studies the prediction of distributions of random variables applied to computer system throughput variability.
-New methods for interpolating data with many dimensions are presented and novel theoretical error bounds are constructed for piecewise linear interpolation.
+New methods for interpolating data with moderately high dimension are presented and novel theoretical error bounds are constructed for piecewise linear interpolation.
 Finally, a mathematical software for approximating distributions from data samples via monotone quintic splines is proposed.}

### 2019-08-13 Paper_[2019-02]_NUMA/Data/make_visuals.py
Updated NUMA paper to fix an error in denominator. Working on dissertation coee still.

@@ -1,0 +1,275 @@
+import os, math
+import numpy as np
+from util.plot import Plot
+from util.data import Data
+from util.math import flatten, transpose
+from util.stats import cdf_fit, rank_probability
+from util.misc.paper import latex_table
+
+
+percentiles = [0, 25, 50, 75, 100]
+
+model_order = [
+    # "BFGS1000",
+    # "SGD10K",
+    "MARS",
+    "SVR",
+    "NeuralNetwork",
+    "Delaunay",
+    "ShepMod",
+    "LSHEP",
+    "Voronoi",
+    "BoxMesh",
+]
+
+data_order = ["results_forestfires.dill",
+              "results_parkinsons_updrs.dill",
+              "results_weatherAUS.dill",
+              "results_creditcard.dill",
+              "results_iozone_150.dill",
+]
+
+
+print("Importing preprocess data..",end="\r")
+
+# Make sure all of the prepared and pre-processed data files exist.
+from preprocess_data import cwd, raw_dir, data_dir, results_folder, processed_results_folder
+
+print()
+
+fit_time = Data.load(os.path.join(results_folder,"fit-final.dill"))
+
+# Create a container for all of the 'best' percents and 'timings'
+all_comparisons = {}
+
+# Cycle through and prepare data from results.
+for f in data_order:
+    if ("results" != f[:len("results")]): continue
+    # Work with the other data.
+    data_name = f[len("results_"):-len(".dill")]
+    path = os.path.join(results_folder, f)
+    print('\n' + '-'*70)
+    print(f"loading {data_name}..", end="\r")
+    d = Data.load(path)
+    print(f, data_name, d.shape)
+
+    print("extracting fit data..", end="\r")
+    # Get the "fit" data from the records.
+    sub_fit_data = fit_time[fit_time["Data name"] == data_name]
+    sub_fit_data.reorder([sub_fit_data.names[0],"Model name"])
+    # Replace the fit times with average fit time per training point.
+    timing_data = sub_fit_data[:,:2].unique().collect(sub_fit_data)
+    timing_data['Fit time'] = map(float,map(np.median, timing_data['Fit time']))
+    for n in timing_data.names[2:-1]:
+        timing_data.pop(n)
+    timing_data['Point Predict time'] = (None for row in timing_data)
+    timing_data['Total Predict time'] = (None for row in timing_data)
+    timing_data._max_display = len(timing_data)
+
+    print("extracting timing data..", end="\r")
[…]

### 2019-08-13 Paper_[2019-02]_Numerical_Algorithms/Data/NA_Tooling_Test/time_delaunay.py

@@ -1,87 +1,0 @@
-import os
-import numpy as np
-
-from util.data import Data
-from util.system import Timer
-from util.approximate import Delaunay, class_name
-
-# Make sure all of the prepared and pre-processed data files exist.
-from preprocess_data import cwd, raw_dir, data_dir
-
-to_test = 30
-
-
-# Overwrite the print function so that it forces prints to flush.
-def print(*args, og_print=print, **kwargs):
-    kwargs["flush"] = True
-    return og_print(*args, **kwargs)
-
-
-seed = 0
-folds = 10
-
-data_name = lambda p: os.path.basename(p).replace(".dill","").replace(".csv","")
-
-# Recording data for each set and for each algorithm that looks like:
-# | Data name | Dimension | Fold Size | Fold Number | Model Name | Model fit time |
-# | x | ... | y | ... | Model y | ... | Model nearest | Model max edge | Model furthest | Model num contributors | Model prediction time |
-
-for raw_file in sorted(os.listdir(raw_dir)):
-    raw_file = raw_file.replace(".gz","")
-    raw_path = os.path.join(raw_dir, raw_file)
-    data_path = os.path.join(data_dir, raw_file + '.dill')
-    intermediate_results_file = os.path.join(".", "Predictions",
-                    f"intermediate_results_{data_name(raw_file)}.dill")
-    final_results_file = os.path.join(".", "Predictions",
-                    f"results_{data_name(raw_file)}.dill")
-    print('-'*70)
-    print(data_name(raw_path))
-    print(raw_path)
-    print(final_results_file)
-    d = Data.load(data_path)
-    target = d.names[-1]
-    # Declare the indices..
-    d["Indices"] = range(len(d))
-    # Compute the number of training columns.
-    num_train_columns = d.names.index(target)
-    print(f"Data shape: {d.shape} predicting '{target}'")
-    # Start the folds.
-    for i,(train, test) in enumerate(d.k_fold(k=folds, seed=seed)):
-        # Get the numeric representation of this data based on training,
-        # ignore extra added columns and the "target" column.
-        num = train[:,:num_train_columns].to_matrix()
-        train_x = num.data
-        train_y = list(train[target])
-        # Generate the numeric testing set basesd on training.
-        test_x = np.array([num.to_real(row[:num_train_columns]) for row in test])
-        test_y = list(test[target])
-        # Normalize the training and testing data to the unit hypercube.
-        train_x = (train_x - num.shift) / num.scale
[…]

### 2019-08-13 Paper_[2019-02]_Numerical_Algorithms/Data/Predictions/fix_data.py

@@ -1,12 +1,0 @@
-from util.data import Data
-import os, dill
-
-for f in os.listdir("."):
-    if ("results_" != f[:len("results_")]): continue
-    print(f)
-    with open(f, "rb") as f:
-        d = dill.load(f)
-    print(d)
-    d = Data([list(r) for r in d])
-    print(d)
-    exit()

### 2019-08-13 Paper_[2019-02]_Numerical_Algorithms/Data/preprocess_data.py

@@ -1,189 +1,0 @@
-import os
-from util.data import Data
-from util.pairs import pairwise_distance
-from util.math import SMALL, flatten, transpose
-
-
-# Read in the data files.
-cwd = os.path.dirname(os.path.abspath(__file__))
-raw_dir = os.path.join( cwd, "Raw" )
-data_dir = os.path.join( cwd, "Data" )
-results_folder = os.path.join(os.path.dirname(os.path.abspath(__file__)), "Predictions")
-processed_results_folder = os.path.join(os.path.dirname(os.path.abspath(__file__)), "Results")
-
-# Given a count, return "count" evenly spaced points on the 
-# unit circle in 2 dimensions.
-def circle_points(n=12):
-    import numpy as np
-    radians = np.linspace(0, 2*np.pi, n+1)[:-1]
-    points = [(float(np.cos(r)), float(np.sin(r))) for r in radians]
-    return points
-
-# Given a string date "year-month-day" convert it to a triple
-#  (time, year_x, year_y) where "_x" and "_y" are coordinates on a circle.
-def date_to_triple(date):
-    import numpy as np
-    year, month, day = map(int, date.split("-"))
-    # Given three integers, return the percentage through the year.
-    def to_0_1(year, month, day):
-        # Code the number of days in each month.
-        days = {m:31 for m in range(12)}
-        days[4] = days[6] = days[9] = days[11] = 30
-        days[2] = 28 + (not year%4)
-        # Record the total number of days in the year.
-        total_days = 365 + (not year%4)
-        # Count how many days have elapsed.
-        elapsed = 0
-        for i in range(1, month): elapsed += days[i]
-        elapsed += day
-        return elapsed / total_days
-
-    # Get the time on a real line.
-    progress = to_0_1(year, month, day)
-    time = year + progress
-    x = np.cos( progress * 2 * np.pi )
-    y = np.sin( progress * 2 * np.pi )
-    return (time, float(x), float(y))
-
-# Given a cardinal direction string return unit circle coordinates.
-def cardinal_to_circle(card):
-    directions = ["E", "ENE", "NE", "NNE", "N", "NNW", "NW", "WNW", 
-                  "W", "WSW", "SW", "SSW", "S", "SSE", "SE", "ESE"]
-    try:    return circle_points(len(directions))[ directions.index(card) ]
-    except: return (None, None)
-
-
-for raw_file in sorted(os.listdir(raw_dir)):
[…]

### 2019-09-08 Disseration/Writeup/pqmip.tex
Nearly done with writeup for algorithm.

\documentclass{article}

%% Get access to mathematical symbols.
\usepackage{amssymb}
%% Get access to the \rightarrow command.
\usepackage{amsmath}
%% Use color links to make them clickable and easy to see.
\usepackage[colorlinks,allcolors=blue]{hyperref}
%% Set the font size of Section headers to be different.
\usepackage{sectsty}
\sectionfont{\fontsize{12}{15}\selectfont}
%%

\begin{document}

\title{Algorithm for Constructing Piecewise Quintic \\ Monotone Interpolating Splines}
\author{Thomas C.H. Lux}
\maketitle

When provided data that has no assigned first and second derivative values, the derivative data is filled by a linear fit of neighboring data points. End points are set to be the slope between the end and its nearest neighbor.

The method finding a maximal value on a line is the Golden Section search. This will be referred to in pseudo code as \texttt{line\_search($a$,$b$)} where $a$ and $b$ are $n$-tuples for integer $n$.

After assigning function values and derivative values, an interpolating function is constructed by solving for the unique weights of a set of quintic B-splines in a linear system.

\vspace{10pt}

%% ----------------------------------------------------------------------

\section{Verifying Monotonicity of a Quintic Polynomial}
\label{is_monotone}
Let $f$ be a quintic polynomial over a closed interval $[x_0, x_1]$ $\subset \mathbb{R}$. Now $f$ is uniquely defined by the evaluation tuples $\big(x_0,$ $f(x_0),$ $f'(x_0),$ $f''(x_0)\big)$ and $\big(x_1,$ $f(x_1),$ $f'(x_1),$ $f''(x_1)\big).$ Assume without loss of generality that $f(x_0) < f(x_1),$ where the case of monotonic decreasing $f$ uses the negated the function values. This algorithm will determine whether or not $f$ is monotone increasing on the interval $[x_0, x_1].$

\vspace{10pt}
\hrule
\vspace{3pt}
\noindent\textbf{\textit{Algorithm 1a:}} \texttt{is\_monotone}
\vspace{3pt}
\hrule

\begin{itemize}
  \itemsep0pt
  \parskip0pt

\item[0:] \texttt{if ($f'(x_0) = 0$ or $f'(x_1) = 0$); return is\_monotone\_simplified}
\item[1:] \texttt{if ($f'(x_0) < 0$ or $f'(x_1) < 0$); return FALSE}
  \begin{itemize}
    \item[] \textit{This can be seen clearly from the fact that $f$ is analytic and there will exist some $0 < \epsilon < x_1 - x_0$ such that either $f'(x_0 + \epsilon)$ or $f'(x_1 - \epsilon)$ will be negative.}
  \end{itemize}

\item[2:] $A = f'(x_0)\frac{x_1 - x_0}{f(x_1) - f(x_0)}$
\item[3:] $B = f'(x_1) \frac{x_1 - x_0}{f(x_1) - f(x_0)}$
  \begin{itemize}
    \item[] \textit{The variables $A$ and $B$ correspond directly to the theoretical foundation for positive quartic polynomials laid in \cite{ulrich1994positivity}.}
  \end{itemize}
\item[8:] $\gamma_0 = 4 \frac{f'(x_0)}{f'(x_1)} (B/A)^{3/4}$
\item[9:] $\gamma_1 = \frac{x_1 - x_0}{f'(x_1)} (B/A)^{3/4}$
\item[4:] $\alpha_0 = 4 (B/A)^{1/4}$
\item[5:] $\alpha_1 = -\frac{x_1 - x_0}{f'(x_1)} (B/A)^{1/4}$
\item[6:] $\beta_0 = 30 - \frac{12 \big(f'(x_0) + f'(x_1)\big) (x_1 - x_0)}{\big(f(x_1) - f(x_0)\big) \sqrt{A}\sqrt{B}}$
\item[7:] $\beta_1 = \frac{-3 (x_1 - x_0)^2}{2 \big(f(x_1) - f(x_0)\big) \sqrt{A} \sqrt{B}} $
  \begin{itemize}
    \item[] \textit{The $\gamma,$ $\alpha,$ and $\beta$ terms with subscripts $0$ and $1$ are algebraic reductions of the original variables from \cite{ulrich1994positivity} that give the computation of each corresponding variable the form  $v = v_0 + v_1 c,$ where $c$ is a term involving only the second derivative values.}
  \end{itemize}
\item[11:] $\gamma = \gamma_0 + \gamma_1 f''(x_0)$
\item[10:] $\alpha = \alpha_0 + \alpha_1 f''(x_1)$
\item[12:] $\beta = \beta_0 + \beta_1 \big(f''(x_0) - f''(x_1)\big)$
\item[13:] \texttt{if $(\beta \leq 6)$; return $\alpha > - (\beta + 2) / 2$}
\item[14:] \texttt{else; return $\gamma > -2 \sqrt{\beta - 2}$ }

\end{itemize}
%% ----------------------------------------------------------------------

%% ----------------------------------------------------------------------
\subsection{Verifying Monotonicity of a Simplified Quintic}

Consider the same initial conditions outlined in Section \ref{is_monotone}.

\vspace{10pt}
\hrule
\vspace{3pt}
\noindent\textbf{\textit{Algorithm 1b:}} \texttt{is\_monotone\_simplified}
\vspace{3pt}
\hrule

\begin{itemize}
  \itemsep0pt
  \parskip0pt

\item Compute
\end{itemize}
%% ----------------------------------------------------------------------

%% ----------------------------------------------------------------------
\section{Enforcing Monotonicity of a Quintic Polynomial}

\vspace{10pt}
\hrule
\vspace{3pt}
\noindent\textbf{\textit{Algorithm 2a:}} \texttt{make\_monotone}
\vspace{3pt}
\hrule

\begin{itemize}
  \itemsep0pt
  \parskip0pt

\item Compute
\end{itemize}
%% ----------------------------------------------------------------------

%% ----------------------------------------------------------------------
\section{Enforcing Monotonicity of a Simplified Quintic}

\vspace{10pt}
\hrule
\vspace{3pt}
\noindent\textbf{\textit{Algorithm 2b:}} \texttt{make\_monotone\_simplified}
\vspace{3pt}
\hrule

\begin{itemize}
  \itemsep0pt
  \parskip0pt

\item Compute

  \begin{itemize}
    \item[] This follows the simplified conditions outlined in proposition 2 of \cite{schmidt1988positivity}.
  \end{itemize}

\end{itemize}
%% ----------------------------------------------------------------------

%% ----------------------------------------------------------------------
\section{Constructing a Piecewise Quintic Monotone Spline}
\label{monotone_spline}

Let $f: \mathbb{R} \rightarrow \mathbb{R}$ be a function in $\mathcal{C}^2.$ Proceed given evaluation tuples $\big(x_i,$ $f(x_i),$ $f'(x_i),$ $f''(x_i)\big)$ for $i = 0,\ldots,N$ such that $x_i < x_{i+1}$ and (without loss of generality) $f(x_i) \leq f(x_{i+1})$ for $i = 1,$ $\ldots,$ $N-1$.

\vspace{10pt}
\hrule
\vspace{3pt}
\noindent\textbf{\textit{Algorithm 3:}} \texttt{monotone\_spline}
\vspace{3pt}
\hrule

\begin{itemize}
  \itemsep0pt
  \parskip0pt

\item Compute
\end{itemize}
%% ----------------------------------------------------------------------

\bibliographystyle{spmpsci}
\bibliography{pqmip}

\end{document}

### 2019-09-08 Disseration/Code/monotone.py

@@ -226,8 +226,8 @@ def monotone_quintic(x, y):
     # down until montonicity is achieved (using binary search).
     alpha_constant   = 4 * (B**(1/4) / A**(1/4))
     alpha_multiplier = ((U0-U1) / DX1) * B**(1/4) / A**(1/4)
-    gamma_constant   = 4 * ((DX0 * B**(3/4)) / (DX1 * A**(3/4)))
-    gamma_multiplier = ((U1-U0) / DX1) * B**(3/4) / A**(3/4)
+    gamma_constant   = 4 * (DX0 / DX1) * (B**(3/4) / A**(3/4))
+    gamma_multiplier = ((U1-U0) / DX1) * (B**(3/4) / A**(3/4))
     beta_constant    = (12 * (DX0+DX1) * (U1-U0) + 30 * (X0-X1)) / ((X0-X1) * A**(1/2) * B**(1/2))
     beta_multiplier  = (3 * (U0-U1)**2) / (2 * (X0-X1) * A**(1/2) * B**(1/2))
     def is_monotone():

### 2019-09-09 Disseration/Writeup/pqmip.tex
Nearly done with algorithm writeup. Only need to write the spline section.

@@
 
+\subsection*{Forward}
+
 When provided data that has no assigned first and second derivative values, the derivative data is filled by a linear fit of neighboring data points.
@@
 
-The method finding a maximal value on a line is the Golden Section search. This will be referred to in pseudo code as \texttt{line\_search($a$,$b$)} where $a$ and $b$ are $n$-tuples for integer $n$.
+The method for finding the transition point of a boolean function on a line is the Golden Section search.
+This will be referred to in pseudo code as \texttt{line\_search($g,$ $a,$ $b$)} where $a,$ $b$ $\in S$ for $S$ closed under convex combination, $g: S \rightarrow \{0,1\}$ is a boolean function, and $g(b) = 1$.
+If $g(a) = 1$ then $a$ is returned, otherwise the smallest $c \in [0,1]$ such that $g\big(a(1-c) + c b\big) = 1$ is returned.
 
-After assigning function values and derivative values, an interpolating function is constructed by solving for the unique weights of a set of quintic B-splines in a linear system.
+After assigning function values and derivative values, an interpolating function is constructed from a quintic B-spline basis.
 
@@
 Let $f$ be a quintic polynomial over a closed interval $[x_0, x_1]$ $\subset \mathbb{R}$.
-Now $f$ is uniquely defined by the evaluation tuples $\big(x_0,$ $f(x_0),$ $f'(x_0),$ $f''(x_0)\big)$ and $\big(x_1,$ $f(x_1),$ $f'(x_1),$ $f''(x_1)\big).$ Assume without loss of generality that $f(x_0) < f(x_1),$ where the case of monotonic decreasing $f$ uses the negated the function values.
-This algorithm will determine whether or not $f$ is monotone increasing on the interval $[x_0, x_1].$
+Now $f$ is uniquely defined by the evaluation tuples $\big(x_0,$ $f(x_0),$ $f'(x_0),$ $f''(x_0)\big)$ and $\big(x_1,$ $f(x_1),$ $f'(x_1),$ $f''(x_1)\big).$ Assume without loss of generality that $f(x_0) < f(x_1),$ where the case of monotonic decreasing $f$ would consider the negated the function values.
+The following algorithm will determine whether or not $f$ is monotone increasing on the interval $[x_0, x_1].$
 
@@
 
-\item[0:] \texttt{if ($f'(x_0) = 0$ or $f'(x_1) = 0$); return is\_monotone\_simplified}
-\item[1:] \texttt{if ($f'(x_0) < 0$ or $f'(x_1) < 0$); return FALSE}
+\item[0:] \texttt{if $\big(f'(x_0) = 0$ or $f'(x_1) = 0\big)$ return is\_monotone\_simplified}
+\item[1:] \texttt{if $\big(f'(x_0) < 0$ or $f'(x_1) < 0\big)$ return FALSE}
   \begin{itemize}
-    \item[] \textit{This can be seen clearly from the fact that $f$ is analytic and there will exist some $0 < \epsilon < x_1 - x_0$ such that either $f'(x_0 + \epsilon)$ or $f'(x_1 - \epsilon)$ will be negative.}
+    \item[] \textit{This can be seen clearly from the fact that $f$ is analytic; there will exist some nonempty ball about $x_0$ or $x_1$ on which $f$ is decreasing.}
   \end{itemize}
@@
   \begin{itemize}
-    \item[] \textit{The variables $A$ and $B$ correspond directly to the theoretical foundation for positive quartic polynomials laid in \cite{ulrich1994positivity}.}
+    \item[] \textit{The variables $A$ and $B$ correspond directly to the theoretical foundation for positive quartic polynomials established in \cite{ulrich1994positivity}, first defined after equation 18.}
   \end{itemize}
@@
   \begin{itemize}
-    \item[] \textit{The $\gamma,$ $\alpha,$ and $\beta$ terms with subscripts $0$ and $1$ are algebraic reductions of the original variables from \cite{ulrich1994positivity} that give the computation of each corresponding variable the form  $v = v_0 + v_1 c,$ where $c$ is a term involving only the second derivative values.}
+    \item[] \textit{The $\gamma,$ $\alpha,$ and $\beta$ terms with subscripts $0$ and $1$ are algebraic reductions of the simplified conditions for satisfying Theorem 2 in \cite{ulrich1994positivity} (equation 16).
+These terms with subscripts $0$ and $1$ give the computation of $\alpha,$ $\beta,$ and $\gamma$ the form seen in lines $10$-$12$ below.}
   \end{itemize}
-\item[11:] $\gamma = \gamma_0 + \gamma_1 f''(x_0)$
-\item[10:] $\alpha = \alpha_0 + \alpha_1 f''(x_1)$
+\item[10:] $\gamma = \gamma_0 + \gamma_1 f''(x_0)$
+\item[11:] $\alpha = \alpha_0 + \alpha_1 f''(x_1)$
 \item[12:] $\beta = \beta_0 + \beta_1 \big(f''(x_0) - f''(x_1)\big)$
-\item[13:] \texttt{if $(\beta \leq 6)$; return $\alpha > - (\beta + 2) / 2$}
-\item[14:] \texttt{else; return $\gamma > -2 \sqrt{\beta - 2}$ }
+\item[13:] \texttt{if $(\beta \leq 6)$ then return $\alpha > - (\beta + 2) / 2$}
+\item[14:] \texttt{else return $\gamma > -2 \sqrt{\beta - 2}$ }
 
 \end{itemize}
+\hrule
+\vspace{10pt}
 %% ----------------------------------------------------------------------
@@
 %% ----------------------------------------------------------------------
-\subsection{Verifying Monotonicity of a Simplified Quintic}
-
-Consider the same initial conditions outlined in Section \ref{is_monotone}.
+Given the same initial conditions there are special circumstances which allow for the usage of simpler monotonicity conditions.
+In this case, consider when the quintic function has either $f'(x_0) = 0$ or $f'(x_1) = 0$, which can be tested for monotonicity via the cubic positivity conditions established by \cite{schmidt1988positivity}.
 
@@
 \hrule
-
 \begin{itemize}
@@
 
-\item Compute
+\item[0:] $\alpha = 30 - \frac{(x_1 - x_0)\big( 14 f'(x_0) + 16 f'(x_1) - \big(f''(x_1) - f''(x_0) \big) (x_1 - x_0)\big)}{2\big(f(x_1) - f(x_0)\big)}$
+\item[1:] $\beta = 30 - \frac{(x_1 - x_0)\big( 2 f'(x_0) + 24 f'(x_1) - \big(f''(x_0) + 3 f''(x_1) \big) (x_1 - x_0)\big)}{2\big(f(x_1) - f(x_0)\big)}$
+\item[2:] $\gamma = \frac{(x_1 - x_0)\big( 7 f'(x_0) + f''(x_0) (x_1 - x_0) \big)}{f(x_1) - f(x_0)}$
+\item[3:] $\delta = \frac{f'(x_0) (x_1 - x_0)}{f(x_1) - f(x_0)}$
+
+  \begin{itemize}
+    \item[] \textit{The variables above are algebraic expansions of the coefficients for the cubic derivative function in \cite{schmidt1988positivity}.}
   \end{itemize}
+
+\item[4:] \texttt{if $\big($min$(\alpha, \delta) < 0\big)$ return FALSE}
+\item[5:] \texttt{else if $\big(\beta < \alpha - 2 \sqrt{\alpha \delta}\big)$ return FALSE}
+\item[6:] \texttt{else if $\big(\gamma < \delta - 2 \sqrt{\alpha \delta}\big)$ return FALSE}
+\item[7:] \texttt{else return TRUE}
+
+\end{itemize}
+\hrule
+\vspace{10pt}
 %% ----------------------------------------------------------------------
+
+Next the modification of a quintic spline to enforce monotonicity will be discussed.
 
@@
 
-\item Compute
+\item[0:] \texttt{if }$\big(f(x_1) - f(x_0) = 0\big)$ \texttt{ return } $f'(x_0) = f'(x_1) = f''(x_0) = f''(x_1) = 0$
+\item[1:] $f'(x_0) = \ $\texttt{median}$\big(0, f'(x_0), 14 \frac{f(x_1) - f(x_0)}{x_1 - x_0} \big)$
+\item[2:] $f'(x_1) = \ $\texttt{median}$\big(0, f'(x_1), 14 \frac{f(x_1) - f(x_0)}{x_1 - x_0}\big)$
+  \begin{itemize}
+  \item[] \textit{This selection of value for $f'(x_0)$ and $f'(x_1)$ is suggested by \cite{ulrich1994positivity} (originally from \cite{huynh1993accurate}), and quickly enforces upper and lower bounds on derivative values to allow for quintic monotonicity.}
   \end{itemize}
+\item[3:] $A = f'(x_0)\frac{x_1 - x_0}{f(x_1) - f(x_0)}$
+\item[4:] $B = f'(x_1) \frac{x_1 - x_0}{f(x_1) - f(x_0)}$
+\item[5:] \texttt{if }$AB \leq 0$\texttt{ return make\_monotone\_simplified}
+\item[6:] \texttt{if $\big($max}$(A,B) > 6\big)$\\$f'(x_0) = 6 f'(x_0)\  /$\texttt{ max}$(A,B)$\\$f'(x_1) = 6 f'(x_1)\  /$\texttt{ max}$(A,B)$
+
+  \begin{itemize}
+  \item[] \textit{This simple box bound ensures that $(A,B)$ remains within a viable region of monotonicity (satisfying Theorem 4, seen in Fig. 6 of \cite{ulrich1994positivity}).}
+  \end{itemize}
+
+\item[7:] $\hat f''(x_0) = - \sqrt{A} \big( 7 \sqrt{A} + 3 \sqrt{B} \big) \frac{f(x_1) - f(x_0)}{(x_1 - x_0)^2}$\\$\hat f''(x_1) = \sqrt{B} \big( 3 \sqrt{A} + 7 \sqrt{B} \big) \frac{f(x_1) - f(x_0)}{(x_1 - x_0)^2}$
+
+  \begin{itemize}
+    \item[] \textit{This selection of values of $f''$ is guaranteed to satisfy Theorem 4 from \cite{ulrich1994positivity} and is chosen because it is (reasonably) the average of the two endpoints of the interval of monotonicity for second derivative values.}
+  \end{itemize}
+
+\item[8:] $\eta = \big(f''(x_0), f''(x_1)\big)$\\$\eta_0 = \big(\hat f''(x_0), f''(x_1)\big)$\\$f''(x_0), f''(x_1) = $\texttt{ line\_search$\big($is\_monotone, $\eta,$ $\eta_0\big)$}
+\end{itemize}
+\hrule
+\vspace{10pt}
 %% ----------------------------------------------------------------------
 
+Similar to the simplified check for monotonicity, when the derivative value at one endpoint of the interval is $0$, a simplified set of steps can be taken to enforce monotonicity.
+
 %% ----------------------------------------------------------------------
-\section{Enforcing Monotonicity of a Simplified Quintic}
-
 \vspace{10pt}
@@
 
-\item Compute
+\item[0:] $f''(x_0) = $\texttt{max}$\bigg(f''(x_0), \frac{-6 f'(x_0)}{x_1 - x_0}\bigg)$
 
   \begin{itemize}
-    \item[] This follows the simplified conditions outlined in proposition 2 of \cite{schmidt1988positivity}.
+    \item[] \textit{Considering the $\alpha,$ $\gamma,$ $\beta,$ and $\delta$ defined in \cite{schmidt1988positivity}, this first step enforces $\gamma > \delta.$ It is already guaranteed that $\delta = \frac{f'(x_0)(x_1 - x_0)}{f(x_1) - f(x_0)} > 0.$ Only two conditions remain to guarantee monotonicity.}
   \end{itemize}
 
+\item[1:] $f''(x_1) = $max$\bigg( f''(x_1), f''(x_0) + \frac{14 f'(x_0) + 16 f'(x_1) + 60(f(x_1) - f(x_0)) / (x_1 - x_0)}{x_1 - x_0} \bigg)$
+
+  \begin{itemize}
+    \item[] \textit{Now it is guaranteed that $\alpha \geq 0.$}
   \end{itemize}
+
+\item[2:] $f''(x_1) = $max$\bigg( f''(x_1), 6 f'(x_0) - 4 f'(x_1) - f''(x_0) \bigg)$
+
+  \begin{itemize}
+    \item[] \textit{Lastly, this guarantees that $\beta \geq \alpha.$ All conditions are met to satisfy proposition 2 of \cite{schmidt1988positivity} and ensure monotonicity.}
+  \end{itemize}
+\end{itemize}
+\hrule
+\vspace{10pt}
 %% ----------------------------------------------------------------------
@@
 \label{monotone_spline}
+
+Finally, the construction of a full piecewise quintic spline is outlined using the above algorithms.
 

### 2019-09-27 Paper_[2019-02]_NUMA/paper.tex
Updated NUMA paper to fix an error in denominator. Working on dissertation coee still.
Updated with analytic example at end of section 5.

--- 2019-03-31 Paper_[2019-02]_Numerical_Algorithms/paper.tex
@@
   \max\limits_{1\ \leq\ j\ \leq\ d} \|x_j - x_0\|_2.$ Then
-  $$\big\|\nabla f(x_0) - c\big\|_2 \leq \sqrt{d} \frac{\gamma k^2}{\sigma_d}.$$
+  $$\big\|\nabla f(x_0) - c\big\|_2 \leq \sqrt{d} \frac{\gamma k^2}{2 \sigma_d}.$$
 \end{plainlemma}
@@
 	&\leq \gamma \|x_0 - z(\ \hat t\ )\|_2\ \|x_j - x_0\|_2 \\
-	&\leq \gamma \|x_j - x_0\|_2^2 \leq \gamma k^2, \\
+	&\leq \frac{\gamma}{2} \|x_j - x_0\|_2^2 \leq \frac{\gamma k^2}{2}, \\
   \end{align*}
@@
   $$\big \| \nabla f(x_i) - c \big \|_2^2 \leq \frac{d}{\sigma_d^2}
-  \big( \gamma k^2\big)^2 \implies \big \| \nabla f(x_i) - c \big \|_2
-  \leq \sqrt{d} \frac{\gamma k^2}{\sigma_d}.$$
+  \big( \gamma k^2 / 2 \big)^2 \implies \big \| \nabla f(x_i) - c \big \|_2
+  \leq \sqrt{d} \frac{\gamma k^2}{2 \sigma_d}.$$
   \qed
@@
     \ref{lemma:3}}, for $z \in S$,
-  $$ \big|f(z) - \hat f(z)\big| \leq \frac{\gamma \|z - x_0\|_2^2}{2} + \sqrt{d} \frac{\gamma k^2}{\sigma_d} \|z - x_0\|_2.$$
+  $$ \big|f(z) - \hat f(z)\big| \leq \frac{\gamma \|z - x_0\|_2^2}{2} + \sqrt{d} \frac{\gamma k^2}{2 \sigma_d} \|z - x_0\|_2.$$
 \end{onetheorem}
@@
 \begin{proofdot}
-  Let $v = \nabla f(x_0) - c$, where $\|v\|_2 \leq \sqrt{d} \gamma k^2 / \sigma_d$ by {\it Lemma \ref{lemma:3}}. Now
+  Let $v = \nabla f(x_0) - c$, where $\|v\|_2 \leq \sqrt{d} \gamma k^2 / (2 \sigma_d)$ by {\it Lemma \ref{lemma:3}}. Now
   \begin{align*}
@@
        &\leq \big|f(z) - f(x_0) - \langle \nabla f(x_0) , z - x_0 \rangle \big| + \|v\|_2 \|z - x_0\|_2 \\
-       &\leq \big|f(z) - f(x_0) - \langle \nabla f(x_0), z - x_0 \rangle \big| + \textstyle{\frac{\gamma k^2 \sqrt{d}}{\sigma_d}} \|z - x_0\|_2 \\
-       &\leq \frac{\gamma \|z - x_0\|_2^2}{2} + \sqrt{d}\frac{\gamma k^2}{\sigma_d} \|z - x_0\|_2,
+       &\leq \big|f(z) - f(x_0) - \langle \nabla f(x_0), z - x_0 \rangle \big| + \textstyle{\sqrt{d}\frac{\gamma k^2}{2 \sigma_d}} \|z - x_0\|_2 \\
+       &\leq \frac{\gamma \|z - x_0\|_2^2}{2} + \sqrt{d}\frac{\gamma k^2}{2 \sigma_d} \|z - x_0\|_2,
   \end{align*}
@@
 diameter, or are nearly contained in a hyperplane) have higher
-error. In light of this error bound, an empirical evaluation of
-presented algorithms follows.
+error.
+
+The error bound provided is tight and that is easily observed by
+considering the test function $q(x) = \|x\|_2^2,$ $x \in \mathbb{R}^d$
+for $d \in \mathcal{N},$ using the simplex $X = \{\mathbf{0}, I\},$
+where $I$ is the $d \times d$ identity matrix, and an approximation is
+made at $z = \mathbf{1/2}.$ Now this setup yields
+
+\vspace{-4mm}
+$$ \gamma = 1, \qquad \sigma_d = 1, \qquad k = 1, \qquad x_0 =
+\mathbf{0}, \qquad \hat q(x) = \langle\mathbf{1}, x - \mathbf{0}\rangle +
+q(\mathbf{0}). $$
+\vspace{-2mm}
+\noindent Plugging these values into the Theorem, we get
+\vspace{-2mm}
+
+\begin{align*}
+  \big|q(z) - \hat q(z)\big| \quad
+  =& \quad \big|\|z\|_2^2 - \langle\mathbf{1},z\rangle\big| \\
+  =& \quad \bigg| \frac{d}{4} - \frac{d}{2} \bigg| \quad = \quad \frac{d}{4} \\
+  \leq& \quad \frac{\gamma \|z - x_0\|_2^2}{2} + \sqrt{d} \frac{\gamma k^2}
+      {2 \sigma_d} \|z - x_0\|_2 \\
+  =& \quad \frac{\|z\|_2^2 + \sqrt{d} \|z\|_2}{2} \\
+  =& \quad \bigg(\frac{d}{4} + \frac{d}{4}\bigg)\ /\ 2 \quad = \quad \frac{d}{4}.
+\end{align*}
+
+In light of this error bound and analytic test, an empirical
+evaluation of presented algorithms follows.
 

### 2019-10-04 Paper_[2019-02]_NUMA/paper.tex
Nearly done with NUMA revisions. Ran new test for SVR and BoxSplines, integrated with old data.

@@
 
-The error bound provided is tight and that is easily observed by
-considering the test function $q(x) = \|x\|_2^2,$ $x \in \mathbb{R}^d$
-for $d \in \mathcal{N},$ using the simplex $X = \{\mathbf{0}, I\},$
-where $I$ is the $d \times d$ identity matrix, and an approximation is
-made at $z = \mathbf{1/2}.$ Now this setup yields
-
-\vspace{-4mm}
-$$ \gamma = 1, \qquad \sigma_d = 1, \qquad k = 1, \qquad x_0 =
+The theoretical error bound is tight and that can be observed with the
+test function $q(x) = \|x\|_2^2,$ $x \in \mathbb{R}^d$ for $d \in
+\mathcal{N},$ and the simplex defined by vertices $X = \{\mathbf{0}, e_1,
+\ldots, e_d\}$ where $e_i$ is $i$-th unit vector in $\mathbb{R}^d.$
+The relevant constants to the error bound are
+
+\vspace{-2mm}
+$$ \gamma = 2, \qquad \sigma_d = 1, \qquad k = 1, \qquad x_0 =
 \mathbf{0}, \qquad \hat q(x) = \langle\mathbf{1}, x - \mathbf{0}\rangle +
@@
 \vspace{-2mm}
-\noindent Plugging these values into the Theorem, we get
+
+\noindent Noting that $q(\mathbf{0}) = 0,$ the approximation error at $z = -\mathbf{1/2}$ is
+
 \vspace{-2mm}
-
 \begin{align*}
@@
   =& \quad \big|\|z\|_2^2 - \langle\mathbf{1},z\rangle\big| \\
-  =& \quad \bigg| \frac{d}{4} - \frac{d}{2} \bigg| \quad = \quad \frac{d}{4} \\
+  =& \quad \bigg| \frac{d}{4} + \frac{d}{2} \bigg| \quad = \quad \frac{3d}{4},
+\end{align*}
+\vspace{-2mm}
+
+\noindent while plugging the constants into the error bound from the theorem gives
+
+\vspace{-2mm}
+\begin{align*}
+  \big|q(z) - \hat q(z)\big| \quad
   \leq& \quad \frac{\gamma \|z - x_0\|_2^2}{2} + \sqrt{d} \frac{\gamma k^2}
       {2 \sigma_d} \|z - x_0\|_2 \\
-  =& \quad \frac{\|z\|_2^2 + \sqrt{d} \|z\|_2}{2} \\
-  =& \quad \bigg(\frac{d}{4} + \frac{d}{4}\bigg)\ /\ 2 \quad = \quad \frac{d}{4}.
+  =& \quad \|z\|_2^2 + \sqrt{d} \|z\|_2 \\
+  =& \quad \frac{d}{4} + \frac{d}{2} \quad = \quad \frac{3d}{4}.
 \end{align*}
-
-In light of this error bound and analytic test, an empirical
-evaluation of presented algorithms follows.
+\vspace{-2mm}
+
+\noindent In light of this error bound, an empirical evaluation of
+presented interpolation and regression algorithms follows.
+
+%% ===================================================================
+\vspace{-2mm}
+\section{Analytic Test of Convergence}
+\label{sec:analytic}
+
+The theoretical results constructed in Section \ref{sec:theory} for
+(piecewise) linear interpolation are promising and apply directly to
+Delaunay interpolation, however they are difficult to interpret in
+context with approximation algorithms that do not have similar known
+uniform error bounds. For that reason, an analytic function is used to
+test the error of interpolants and regressors given an increasing
+number of examples and increasing relative random error at data
+points.
 
@@
 
-The theoretical results constructed in Section \ref{sec:theory} for
-(piecewise) linear interpolation are promising and apply directly to
-Delaunay interpolation, however they are difficult to interpret in
-context with approximation algorithms that do not have similar known
-uniform error bounds. For that reason, this paper utilizes five
-different data sets of varying dimension and application to construct
+This section extends the comparison of interpolation and regression
+algorithms to a sample of real-world problems. Five different data
+sets of varying dimension and application are utilized to construct
 approximations and compare the accuracy of different techniques.
@@
     Voronoi    & $52.6\%$          & $\mathbf{43.4}\%$ & $\mathbf{34.4}\%$ & $\mathbf{19.1}\%$\\
-    BoxSpline & $99.5\%$ & $98.5\%$ & $96.6\%$ & $89.5\%$\\
+    BoxSplines & $99.4\%$          & $98.5\%$          & $96.6\%$          & $89.3\%$\\
   \end{tabular}
@@
     \hline
-    MARS & \quad\quad$4$&$6$ & \,\quad$20$&$0$ & \quad\quad$0$&$001$\\
-    SVR & $\mathit{21}$&$\mathit{1}$ & $\mathbf{0}$&$\mathbf{5}$ & $\mathbf{0}$&$\mathbf{0001}$\\
-    MLP & $\mathbf{42}$&$\mathbf{1}$ & $200$&$0$ & $0$&$001$\\
-    Delaunay & $6$&$0$ & $1$&$0$ & $1$&$0$\\
-    ShepMod & $18$&$4$ & $\mathit{0}$&$\mathit{7}$ & $\mathbf{0}$&$\mathbf{0001}$\\
-    LSHEP & $8$&$4$ & $2$&$0$ & $\mathbf{0}$&$\mathbf{0001}$\\
-    Voronoi & $0$&$5$ & $1$&$0$ & $0$&$04$\\
-    BoxSpline & $0$&$2$ & $0$&$8$ & $\mathit{0}$&$\mathit{0005}$\\
+    MARS        & \quad\quad$4$&$5$          & \quad\quad$20$&$0$s        & \qquad\quad $0$&$001$s\\
+    SVR         & $\mathit{19}$&$\mathit{5}$ & $\mathbf{0}$&$\mathbf{5}$s & $\mathbf{0}$&$\mathbf{0001}$s\\
+    MLP         & $\mathbf{43}$&$\mathbf{1}$ & $200$&$0$s                 & $0$&$001$s\\
+    Delaunay    & $5$&$2$                    & $1$&$0$s                   & $1$&$0$s\\
+    ShepMod     & $18$&$0$                   & $\textit{0}$&$\textit{7}$s & $\mathbf{0}$&$\mathbf{0001}$s\\
+    LSHEP       & $8$&$4$                    & $2$&$0$s                   & $\mathbf{0}$&$\mathbf{0001}$s\\
+    Voronoi     & $0$&$5$                    & $1$&$0$s                   & $0$&$04$s\\
+    BoxSplines  & $3$&$5$                    & $0$&$8$s                   & $\mathit{0}$&$\mathit{0005}$s\\
   \end{tabular}
@@
     \hline
-    Algorithm & \multicolumn{2}{c|}{Min} & \multicolumn{2}{c|}{$25^{th}$} & \multicolumn{2}{c|}{$50^{th}$} & \multicolumn{2}{c|}{$75^{th}$} & \multicolumn{2}{c}{Max}\\
+    Algorithm & \multicolumn{2}{c}{Min}       & \multicolumn{2}{c}{$25^{th}$} & \multicolumn{2}{c}{$50^{th}$} & \multicolumn{2}{c}{$75^{th}$} & \multicolumn{2}{c|}{Max}\\
     \hline
     MARS      & $0$&$00984$                   & $3$&$11$                      & $7$&$01$                      & $\mathit{11}$&$\mathit{7}$    & $1090$&$0$\\
-    SVR & $\mathit{0}$&$\mathit{00402}$ & $\mathbf{0}$&$\mathbf{243}$ & $\mathbf{0}$&$\mathbf{416}$ & $\mathbf{6}$&$\mathbf{19}$ & $1090$&$0$\\
-    MLP & $0$&$0426$ & $2$&$63$ & $\mathit{5}$&$\mathit{27}$ & $14$&$0$ & $1090$&$0$\\
-    Delaunay & $\mathbf{0}$&$\mathbf{00}$ & $1$&$98$ & $5$&$37$ & $13$&$1$ & $\mathit{1080}$&$\mathit{0}$\\
-    ShepMod & $\mathbf{0}$&$\mathbf{00}$ & $\mathit{1}$&$\mathit{93}$ & $6$&$27$ & $16$&$0$ & $1090$&$0$\\
-    LSHEP & $0$&$0400$ & $2$&$17$ & $8$&$87$ & $19$&$1$ & $\mathbf{1070}$&$\mathbf{0}$\\
-    Voronoi & $0$&$00982$ & $3$&$65$ & $7$&$56$ & $15$&$6$ & $1090$&$0$\\
-    BoxSpline & $\mathbf{0}$&$\mathbf{00}$ & $2$&$27$ & $5$&$91$ & $16$&$4$ & $1090$&$0$\\
+    SVR       & $0$&$0118$                    & $\mathbf{0}$&$\mathbf{615}$   & $\mathbf{0}$&$\mathbf{931}$   & $\mathbf{5}$&$\mathbf{89}$    & $1090$&$0$\\
+    MLP       & $0$&$0426$                    & $2$&$63$                      & $5$&$27$                      & $14$&$0$                      & $1090$&$0$\\
+    Delaunay  & $\mathbf{0}$&$\mathbf{0}$     & $1$&$98$                      & $5$&$37$                      & $13$&$1$                      & $\mathit{1080}$&$\mathit{0}$\\
+    ShepMod   & $\mathbf{0}$&$\mathbf{0}$     & $1$&$93$                      & $6$&$27$                      & $16$&$0$                      & $1090$&$0$\\
+    LSHEP     & $0$&$04$                      & $2$&$17$                      & $8$&$87$                      & $19$&$1$                      & $\mathbf{1070}$&$\mathbf{0}$\\
+    Voronoi   & $\mathit{0}$&$\mathit{00982}$ & $3$&$65$                      & $7$&$56$                      & $15$&$6$                      & $1090$&$0$\\
+    BoxSpline & $\mathbf{0}$&$\mathbf{0}$     & $\mathit{1}$&$\mathit{27}$    & $\mathit{4}$&$\mathit{61}$    & $12$&$6$                      & $1090$&$0$\\
     \hline
@@
     \cline{1-3}\cline{5-10}
-    MARS & \quad$\mathit{7}$&$\mathit{7}$ &  & \quad\quad$29$&$1$ & \quad$0$&$00137$ & \quad\quad\quad$0$&$0686$\\
-    SVR & $\mathbf{80}$&$\mathbf{2}$ &  & $\mathbf{0}$&$\mathbf{00584}$ & $\mathit{0}$&$\mathit{0000620}$ & $\mathit{0}$&$\mathit{00310}$\\
+    MARS      & \quad$7$&$3$                   &  & \quad\quad$29$&$1$                      & \quad$0$&$00137$                   & \quad\quad\quad$0$&$0686$\\
+    SVR       & $\mathbf{78}$&$\mathbf{0}$     &  & $\mathbf{0}$&$\mathbf{00584}$           & $\mathit{0}$&$\mathit{0000620}$    & $\mathit{0}$&$\mathit{00310}$\\
     MLP       & $0$&$0$                        &  & $32$&$8$                                & $0$&$000871$                       & $0$&$0436$\\
-    Delaunay & $3$&$5$ &  & $0$&$0151$ & $0$&$0234$ & $1$&$18$\\
-    ShepMod & $3$&$7$ &  & $\mathit{0}$&$\mathit{00634}$ & $0$&$0000644$ & $0$&$00322$\\
+    Delaunay  & $0$&$2$                        &  & $0$&$0151$                              & $0$&$0234$                         & $1$&$18$\\
+    ShepMod   & $2$&$0$                        &  & $\mathit{0}$&$\mathit{00634}$           & $0$&$0000644$                      & $0$&$00322$\\
     LSHEP     & $5$&$1$                        &  & $0$&$0275$                              & $\mathbf{0}$&$\mathbf{0000463}$    & $\mathbf{0}$&$\mathbf{00231}$\\
     Voronoi   & $0$&$0$                        &  & $0$&$0152$                              & $0$&$000396$                       & $0$&$0198$\\
-    BoxSpline & $0$&$4$ &  & $0$&$00724$ & $0$&$0000978$ & $0$&$00489$\\
+    BoxSpline & $\mathit{9}$&$\mathit{7}$      &  & $0$&$00724$                             & $0$&$0000978$                      & $0$&$00489$\\
     \cline{1-3}\cline{5-10}
@@
     \hline
-    Algorithm & \multicolumn{2}{c|}{Min} & \multicolumn{2}{c|}{$25^{th}$} & \multicolumn{2}{c|}{$50^{th}$} & \multicolumn{2}{c|}{$75^{th}$} & \multicolumn{2}{c}{Max}\\
+    Algorithm & \multicolumn{2}{c}{Min}                   & \multicolumn{2}{c}{$25^{th}$} & \multicolumn{2}{c}{$50^{th}$} & \multicolumn{2}{c}{$75^{th}$} & \multicolumn{2}{c|}{Max}\\
     \hline
     MARS      & $0$&$00948$                               & $9$&$98$                      & $20$&$4$                      & $32$&$9$                      & $863$&$0$\\
-    SVR & $0$&$00138$ & $3$&$17$ & $7$&$31$ & $12$&$6$ & $\mathbf{27}$&$\mathbf{5}$\\
+    SVR       & $0$&$00233$                               & $3$&$15$                      & $7$&$21$                      & $11$&$3$                      & $\mathbf{28}$&$\mathbf{6}$\\
     MLP       & $0$&$0000239$                             & $\mathit{0}$&$\mathit{533}$   & $\mathit{1}$&$\mathit{25}$    & $\mathbf{2}$&$\mathbf{84}$    & $39$&$3$\\
-    Delaunay & $\mathit{3}$&$\mathit{72 \times 10^{-12}}$ & $1$&$20$ & $3$&$50$ & $7$&$67$ & $30$&$7$\\
+    Delaunay  & $\mathit{3}$&$\mathit{72 \times 10^{-12}}$ & $1$&$2$                       & $3$&$5$                       & $7$&$67$                      & $30$&$7$\\
     ShepMod   & $\mathbf{0}$&$\mathbf{0}$                 & $\mathbf{0}$&$\mathbf{255}$   & $\mathbf{0}$&$\mathbf{908}$   & $\mathit{3}$&$\mathit{43}$    & $34$&$5$\\
@@
     Voronoi   & $\mathbf{0}$&$\mathbf{0}$                 & $1$&$29$                      & $3$&$52$                      & $6$&$87$                      & $30$&$1$\\
-    BoxSpline & $0$&$00287$ & $4$&$39$ & $9$&$10$ & $14$&$8$ & $41$&$3$\\
+    BoxSpline & $0$&$006$                                 & $4$&$3$                       & $8$&$91$                      & $15$&$1$                      & $45$&$3$\\
     \hline
@@
     \hline
-    Algorithm & \multicolumn{2}{c|}{Min} & \multicolumn{2}{c|}{$25^{th}$} & \multicolumn{2}{c|}{$50^{th}$} & \multicolumn{2}{c|}{$75^{th}$} & \multicolumn{2}{c}{Max}\\
+    Algorithm & \multicolumn{2}{c}{Min}                    & \multicolumn{2}{c}{$25^{th}$}              & \multicolumn{2}{c}{$50^{th}$}  & \multicolumn{2}{c}{$75^{th}$}  & \multicolumn{2}{c|}{Max}\\
     \hline
     MARS      & $\mathit{6}$&$\mathit{45 \times 10^{-15}}$ & $\mathbf{2}$&$\mathbf{70 \times 10^{-14}}$ & $1$&$66$                       & $1$&$96$                       & $\mathit{53}$&$\mathit{3}$\\
-    SVR & $0$&$0420$ & $0$&$143$ & $0$&$148$ & $\mathit{0}$&$\mathit{860}$ & $119$&$0$\\
+    SVR       & $0$&$0000915$                              & $0$&$0833$                                 & $0$&$263$                      & $\mathit{1}$&$\mathit{13}$     & $109$&$0$\\
     MLP       & $0$&$0000689$                              & $0$&$0337$                                 & $\mathbf{0}$&$\mathbf{0795}$   & $\mathbf{0}$&$\mathbf{264}$    & $\mathbf{5}$&$\mathbf{31}$\\
     Delaunay  & $\mathbf{0}$&$\mathbf{0}$                  & $0$&$187$                                  & $0$&$919$                      & $2$&$72$                       & $56$&$3$\\
-    ShepMod & $\mathbf{0}$&$\mathbf{0}$ & $0$&$0957$ & $0$&$685$ & $2$&$90$ & $73$&$2$\\
+    ShepMod   & $\mathbf{0}$&$\mathbf{0}$                  & $0$&$0957$                                 & $0$&$685$                      & $2$&$9$                        & $73$&$2$\\
     LSHEP     & $\mathbf{0}$&$\mathbf{0}$                  & $\mathit{0}$&$\mathit{0153}$               & $\mathit{0}$&$\mathit{106}$    & $1$&$17$                       & $113$&$0$\\
-    Voronoi & $\mathbf{0}$&$\mathbf{0}$ & $0$&$430$ & $1$&$28$ & $2$&$94$ & $83$&$8$\\
-    BoxSpline & $\mathbf{0}$&$\mathbf{0}$ & $0$&$789$ & $2$&$06$ & $5$&$25$ & $119$&$0$\\
+    Voronoi   & $\mathbf{0}$&$\mathbf{0}$                  & $0$&$43$                                   & $1$&$28$                       & $2$&$94$                       & $83$&$8$\\
+    BoxSpline & $\mathbf{0}$&$\mathbf{0}$                  & $0$&$342$                                  & $1$&$59$                       & $5$&$53$                       & $119$&$0$\\
     \hline
@@
     MARS      & \,\,\,\,$10$&$7$             &  & \quad\quad\quad$\mathit{0}$&$\mathit{151}$ & \quad$0$&$000117$                  & \quad\quad\quad\,\,$0$&$0304$\\
-    SVR & $4$&$1$ &  & $\mathbf{0}$&$\mathbf{133}$ & $\mathbf{0}$&$\mathbf{0000947}$ & $\mathbf{0}$&$\mathbf{0246}$\\
-    MLP & $\mathbf{56}$&$\mathbf{8}$ &  & $169$&$0$ & $0$&$00137$ & $0$&$356$\\
+    SVR       & $0$&$0$                      &  & $\mathbf{0}$&$\mathbf{133}$                & $\mathbf{0}$&$\mathbf{0000947}$    & $\mathbf{0}$&$\mathbf{0246}$\\
+    MLP       & $\mathbf{60}$&$\mathbf{9}$   &  & $169$&$0$                                  & $0$&$00137$                        & $0$&$356$\\
     Delaunay  & $0$&$1$                      &  & $0$&$664$                                  & $0$&$886$                          & $230$&$0$\\
@@
     Voronoi   & $0$&$2$                      &  & $0$&$675$                                  & $0$&$0270$                         & $7$&$01$\\
-    BoxSpline & $0$&$3$ &  & $0$&$330$ & $0$&$000406$ & $0$&$106$\\
+    BoxSpline & $4$&$1$                      &  & $0$&$330$                                  & $0$&$000406$                       & $0$&$106$\\
     \cline{1-3}\cline{5-10}
@@
     \hline
-    Algorithm & \multicolumn{2}{c|}{Min} & \multicolumn{2}{c|}{$25^{th}$} & \multicolumn{2}{c|}{$50^{th}$} & \multicolumn{2}{c|}{$75^{th}$} & \multicolumn{2}{c}{Max}\\
+    Algorithm & \multicolumn{2}{c}{Min}                    & \multicolumn{2}{c}{$25^{th}$} & \multicolumn{2}{c}{$50^{th}$} & \multicolumn{2}{c}{$75^{th}$} & \multicolumn{2}{c|}{Max}\\
     \hline
     MARS      & $5$&$36$                                   & $3610$&$0$                    & $4580$&$0$                    & $5450$&$0$                    & $13400$&$0$\\
-    SVR & $0$&$0138$ & $8$&$47$ & $15$&$4$ & $41$&$9$ & $7700$&$0$\\
-    MLP & $0$&$00151$ & $\mathbf{1}$&$\mathbf{60}$ & $\mathbf{3}$&$\mathbf{86}$ & $\mathbf{8}$&$\mathbf{34}$ & $\mathbf{604}$&$\mathbf{0}$\\
-    Delaunay & $\mathbf{0}$&$\mathbf{0}$ & $\mathit{2}$&$\mathit{69}$ & $\mathit{13}$&$\mathit{8}$ & $\mathit{35}$&$\mathit{0}$ & $\mathit{4840}$&$\mathit{0}$\\
+    SVR       & $0$&$00706$                                & $8$&$14$                      & $\mathit{13}$&$\mathit{0}$    & $41$&$6$                      & $7690$&$0$\\
+    MLP       & $0$&$00151$                                & $\mathbf{1}$&$\mathbf{6}$     & $\mathbf{3}$&$\mathbf{86}$    & $\mathbf{8}$&$\mathbf{34}$    & $\mathbf{604}$&$\mathbf{0}$\\
+    Delaunay  & $\mathbf{0}$&$\mathbf{0}$                  & $\mathit{2}$&$\mathit{69}$    & $13$&$8$                      & $\mathit{35}$&$\mathit{0}$    & $\mathit{4840}$&$\mathit{0}$\\
     ShepMod   & $\mathbf{0}$&$\mathbf{0}$                  & $4$&$21$                      & $17$&$3$                      & $51$&$6$                      & $6510$&$0$\\
@@
     Voronoi   & $\mathit{2}$&$\mathit{89 \times 10^{-10}}$ & $14$&$7$                      & $32$&$8$                      & $52$&$9$                      & $4860$&$0$\\
-    BoxSpline & $0$&$00740$ & $20$&$8$ & $40$&$9$ & $79$&$6$ & $7640$&$0$\\
+    BoxSpline & $\mathbf{0}$&$\mathbf{0}$                  & $12$&$4$                      & $35$&$0$                      & $90$&$1$                      & $7690$&$0$\\
     \hline
@@
     \hline
-    Algorithm & \multicolumn{2}{c|}{Min} & \multicolumn{2}{c|}{$25^{th}$} & \multicolumn{2}{c|}{$50^{th}$} & \multicolumn{2}{c|}{$75^{th}$} & \multicolumn{2}{c}{Max}\\
+    Algorithm & \multicolumn{2}{c}{Min}      & \multicolumn{2}{c}{$25^{th}$} & \multicolumn{2}{c}{$50^{th}$} & \multicolumn{2}{c}{$75^{th}$} & \multicolumn{2}{c|}{Max}\\
     \hline
     Delaunay  & $\mathbf{0}$&$\mathbf{0287}$ & $\mathbf{0}$&$\mathbf{0914}$  & $\mathbf{0}$&$\mathbf{158}$   & $\mathit{0}$&$\mathit{308}$   & $\mathit{0}$&$\mathit{897}$\\
-    ShepMod & $0$&$0307$ & $\mathit{0}$&$\mathit{0983}$ & $\mathit{0}$&$\mathit{164}$ & $0$&$335$ & $1$&$00$\\
-    Voronoi & $\mathit{0}$&$\mathit{0303}$ & $0$&$100$ & $0$&$165$ & $\mathbf{0}$&$\mathbf{274}$ & $\mathbf{0}$&$\mathbf{762}$\\
-    BoxSpline & $0$&$0703$ & $0$&$417$ & $0$&$561$ & $0$&$745$ & $1$&$00$\\
+    ShepMod   & $0$&$0307$                   & $\mathit{0}$&$\mathit{0983}$  & $\mathit{0}$&$\mathit{164}$   & $0$&$335$                     & $1$&$0$\\
+    Voronoi   & $\mathit{0}$&$\mathit{0303}$ & $0$&$1$                       & $0$&$165$                     & $\mathbf{0}$&$\mathbf{274}$   & $\mathbf{0}$&$\mathbf{762}$\\
+    BoxSpline & $0$&$0679$                   & $0$&$429$                     & $0$&$571$                     & $0$&$752$                     & $1$&$0$\\
     \hline

### 2019-10-04 Paper_[2019-02]_NUMA/Data/analytic/make_analytic_visuals.py

@@ -1,0 +1,49 @@
+from util.data import Data
+
+d = Data.load("2-analytic-results.pkl")
+
+# # Strip out the invalid tests (bad number of train points).
+# to_keep = [i for i in range(len(d)) if ((d[i,"Train"]-1)%10)]
+# # Make a copy so that we can do assignment etc.
+# d = d[to_keep][:]
+# d.save("1-analytic-results.pkl")
+
+# Generate interesting extra columns.
+d["Abs Errors"] = ([float(abs(v)) for v in l] for l in d["Errors"])
+d["Mean Abs Error"] = (sum(l) / len(l) for l in d["Abs Errors"])
+d["Min Abs Error"] = (min(l) for l in d["Abs Errors"])
+d["Max Abs Error"] = (max(l) for l in d["Abs Errors"])
+# d.summarize()
+
+config_cols = ["SNR", "Train"]
+
+configs = d[:,config_cols].unique()
+configs.sort()
+configs.max_display=float('inf')
+print(configs)
+
+from util.plot import Plot
+from util.stats import cdf_fit
+
+for conf in configs:
+    interest = d[d[:,config_cols] == conf]
+    interest.sort()
+    print(interest)
+    S, N = conf
+    p = Plot(f"{N} train with SNR {S}")
+    seen = {"NearestNeighbor"}
+    for row in interest:
+        name = row["Algorithm"]
+        # Skip algorithms that have already been plotted.
+        if name in seen: 
+            print(f"  skipping extra '{name}'")
+            continue
+        else:            seen.add(name)
+        # Plot the performance of this algorithm.
+        fit_time = row["Fit Time"]
+        pred_time = row["Predict Time"]
+        errors = row["Abs Errors"]
+        # f = cdf_fit(errors)
+        # p.add_func(name, f, f())
+        p.add_box(name, errors)
+    p.show(append=True, show=all(conf == configs[-1]))

### 2019-10-04 Paper_[2019-02]_NUMA/Data/analytic/run_analytic_test.py

@@ -1,0 +1,151 @@
+import os, time, random
+import numpy as np
+from util.data import Data
+from util.system import Timer, save, load
+from util.math import fekete_points
+from util.approximate import DelaunayP1, Voronoi, BoxMesh, ShepMod, \
+    LSHEP, NeuralNetwork, MARS, SVR, SVR_RBF, class_name
+
+seed = 0
+algorithms  = [DelaunayP1,
+               Voronoi,
+               BoxMesh, ShepMod, LSHEP,
+               NeuralNetwork,
+               MARS, SVR_RBF]
+train_sizes = [25, 50, 100, 500, 1000]
+snr_values  = [0, .1, .2, .3, .4, .5]
+test_size   = 1000
+
+# The oscillatory test problem from:
+#    'High Dimensional Polynomial Interpolation' by Barthelman et al. (2000)
+# In the default case, it is a componentwise cosine function.
+def oscillatory(x, c=1, w=0):
+    if len(x.shape) <= 1:
+        return np.cos(2*np.pi*w + np.sum(c*x))
+    elif len(x.shape) == 2:
+        return np.cos(2*np.pi*w + np.sum(c*x, axis=1))
+    else:
+        class UsageError(Exception): pass
+        raise(UsageError(f"This objective function only supports arrays and matrices, received shape {x.shape}."))
+
+# The quadratic test problem from. Uses the squared value of the 2-norm.
+def quadratic(x):
+    if len(x.shape) <= 1:
+        return np.linalg.norm(x)**2
+    elif len(x.shape) == 2:
+        return np.linalg.norm(x, axis=1)**2
+    else:
+        class UsageError(Exception): pass
+        raise(UsageError(f"This objective function only supports arrays and matrices, received shape {x.shape}."))
+
+
+# Overwrite the print function so that it forces prints to flush.
+def print(*args, og_print=print, **kwargs):
+    kwargs["flush"] = True
+    return og_print(*args, **kwargs)
+
+# Assign a function (that supports matrix evaluation) as the test function.
+test_function = quadratic
+test_func_name = str(test_function).split()[1]
+dimension = 10
+
+if (test_func_name == "oscillaatory"):
+    # Set up the test problem more specifically (after getting its name).
+    difficulty = 9.0
+    c = np.random.random(size=(dimension,))
+    c /= np.sum(c)
+    c *= 9.0
+    test_function = lambda x: oscillatory(x, c=c)
+
+
[…]

### 2019-10-04 Paper_[2019-02]_NUMA/Data/make_visuals.py

@@ -13,7 +13,7 @@ model_order = [
     # "BFGS1000",
     # "SGD10K",
     "MARS",
-    "SVR",
+    "SVR_RBF",
     "NeuralNetwork",
     "Delaunay",
     "ShepMod",
@@ -55,10 +55,10 @@ for f in data_order:
 
     print("extracting fit data..", end="\r")
     # Get the "fit" data from the records.
-    sub_fit_data = fit_time[fit_time["Data name"] == data_name]
+    sub_fit_data = fit_time[fit_time["Data name"] == data_name].copy()
     sub_fit_data.reorder([sub_fit_data.names[0],"Model name"])
     # Replace the fit times with average fit time per training point.
-    timing_data = sub_fit_data[:,:2].unique().collect(sub_fit_data)
+    timing_data = sub_fit_data[:,:2].unique().copy().collect(sub_fit_data)
     timing_data['Fit time'] = map(float,map(np.median, timing_data['Fit time']))
     for n in timing_data.names[2:-1]:
         timing_data.pop(n)
@@ -82,7 +82,7 @@ for f in data_order:
     for m in model_order:
         if m not in models: continue
         indices.append( list(timing_data["Model name"]).index(m) )
-    timing_data = timing_data[indices]
+    timing_data = timing_data[indices].copy()
 
     print("collecting error data..", end="\r")
     # Collect the error and timing data.
@@ -92,7 +92,9 @@ for f in data_order:
         errors = []
         prediction_times = []
         for row in range(len(d)):
-            if "nan" in set(map(str,d[row,[f"{m} {target}" for m in models]][0])): continue
+            if 'iozone' not in data_name:
+                to_clean = [f"{m} {target}" for m in models]
+                if "nan" in set(map(str,d[row,to_clean])): continue
             truth, guess = d[row, target], d[row, f"{m_name} {target}"]
             errors.append( truth - guess )
             prediction_times.append(d[row, f"{m_name} prediction time"])
@@ -108,7 +110,8 @@ for f in data_order:
             continue
         abs_errors = list(map(float,map(abs,errors)))
         error_data.append( [m_name.replace("NeuralNetwork","MLP")
-                            .replace("BoxMesh","BoxSplines")] + abs_errors )
+                            .replace("BoxMesh","BoxSplines")
[…]

### 2019-10-04 Paper_[2019-02]_NUMA/Data/analytic/run_analytic_test.py
Collected first half of results for analytic test function. Other half about to run on Pawnee.

@@ -7,26 +7,34 @@ from util.approximate import DelaunayP1, Voronoi, BoxMesh, ShepMod, \
     LSHEP, NeuralNetwork, MARS, SVR, SVR_RBF, class_name
 
 seed = 0
-algorithms  = [DelaunayP1,
-               Voronoi,
-               BoxMesh, ShepMod, LSHEP,
-               NeuralNetwork,
-               MARS, SVR_RBF]
-train_sizes = [25, 50, 100, 500, 1000]
-snr_values  = [0, .1, .2, .3, .4, .5]
+algorithms  = [
+    DelaunayP1,
+    Voronoi,
+    BoxMesh,
+    ShepMod,
+    LSHEP,
+    NeuralNetwork,
+    MARS,
+    SVR_RBF
+]
+train_sizes = [2 * 4**(i) for i in range(1,6+1)]
+snr_values  = [0, .1]
 test_size   = 1000
+dimension   = 2
 
-# The oscillatory test problem from:
+# A modified version of the oscillatory test problem from:
 #    'High Dimensional Polynomial Interpolation' by Barthelman et al. (2000)
-# In the default case, it is a componentwise cosine function.
-def oscillatory(x, c=1, w=0):
+# It is a cosine function of the norm of the input vectors. This
+# function has no componentwise structure to take advantage of.
+def oscillatory(x, c=1):
     if len(x.shape) <= 1:
-        return np.cos(2*np.pi*w + np.sum(c*x))
+        return np.cos(c*np.pi + np.linalg.norm(x))
     elif len(x.shape) == 2:
-        return np.cos(2*np.pi*w + np.sum(c*x, axis=1))
+        return np.cos(c*np.pi + np.linalg.norm(x, axis=1))
     else:
         class UsageError(Exception): pass
         raise(UsageError(f"This objective function only supports arrays and matrices, received shape {x.shape}."))
+
 
 # The quadratic test problem from. Uses the squared value of the 2-norm.
 def quadratic(x):
@@ -45,107 +53,106 @@ def print(*args, og_print=print, **kwargs):
     return og_print(*args, **kwargs)
 
 # Assign a function (that supports matrix evaluation) as the test function.
-test_function = quadratic
+test_function = oscillatory
 test_func_name = str(test_function).split()[1]
-dimension = 10
-
-if (test_func_name == "oscillaatory"):
-    # Set up the test problem more specifically (after getting its name).
-    difficulty = 9.0
[…]

### 2019-10-15 Paper_[2019-02]_NUMA/paper.tex
Collected first half of results for analytic test function. Other half about to run on Pawnee.
Finished implementing revisions to NUMA paper. Sending files to Dr. Watson for confirmation.

@@
 cumulative distribution functions (CDFs) with the Kolmogorov-Smirnov
-(KS) statistic \cite{lilliefors1967kolmogorov} as metric.
+(KS) statistic \cite{lilliefors1967kolmogorov} as a metric.
 
@@
 Provided the theoretical and practical motivations for exploring
-interpolants, this work aims to study the empirical performance
-differences between a set of scalable (moderately) interpolation
-techniques and a set of common regression techniques. These techniques
-are applied to a collection of moderately high dimensional problems
-($5 \le d \le 30$) and the empirical results are discussed.
+interpolants, the main objective of this work is to study the
+empirical performance differences in both accuracy and computational
+expense between a set of interpolation techniques and a set of common
+regression techniques. The main contributions of this work are a sharp
+theoretical error bound for (piecewise) linear interpolation in
+arbitrary dimension and multiple empirical comparisons that illustrate
+potential advantages of interpolation.
 
@@
 \ref{sec:theory} contains the theoretical error bounds for
-interpolation. Section \ref{sec:data} presents the data sets for
-empirical approximation analysis and the approximation results for all
-models on these data sets. Section \ref{sec:discussion} analyzes and
-discusses the results, their implications, and their
-limitations. Finally, Section \ref{sec:conclusion} concludes.
+interpolation as well as an analytic test for demonstration. Section
+\ref{sec:data} presents the data sets for empirical approximation
+analysis and the approximation results for all models on these data
+sets. Section \ref{sec:discussion} analyzes and discusses the results,
+their implications, and their limitations. Finally, Section
+\ref{sec:conclusion} concludes.
 
@@
 $m$ being determined by the minimization convergence criterion. This
-paper uses the scikit-learn SVR \cite{scikit-learn} with a polynomial
-kernel function.
+paper uses the scikit-learn SVR \cite{scikit-learn} with a radial
+basis function kernel.
 
@@
 
-The box spline model used here is an interpolation technique built
-from overlapping box splines \cite{de2013box}. The box splines serve
-as basis functions that are shifted and scaled to have support over
-box shaped regions. The boxes are constructed in a way to guarantee a
-covering of the domain \cite{lux2018novel}. Given a set of box splines
-$\bigl\{b^{x^{(i)}}\bigr\}$ with the iterative box properties outlined
-in \cite{lux2018novel} and anchored at interior points
-$\bigl\{x^{(i)}\bigr\}$,
+The box spline model used here is a sparse grid interpolation
+technique built from overlapping box splines \cite{de2013box}. The box
+splines serve as basis functions that are shifted and scaled to have
+support over box shaped regions. The boxes are constructed in a way to
+guarantee a covering of the domain \cite{lux2018novel}. Given a set of
+box splines $\bigl\{b^{x^{(i)}}\bigr\}$ with the iterative box
+properties outlined in \cite{lux2018novel} and anchored at interior
+points $\bigl\{x^{(i)}\bigr\}$,
  $$ \hat f(y) = \frac{\sum\limits_{i=1}^n b^{x^{(i)}}(y) f\bigl(x^{(i)}\bigr)}{\sum\limits_{i=1}^n b^{x^{(i)}}(y)}. $$
@@
     CDF (solid line) and predicted CDF (dashed line) is 0.2 for this
-    example. The KS test null hypothesis is rejected at $p$-value
-    0.01, however it is not rejected at $p$-value 0.001.
+    example. For this example the KS test null hypothesis is rejected
+    at $p$-value 0.01, however it is not rejected at $p$-value 0.001.
   \vspace{-.1cm}}
@@
 predicted distributions to the calculation of error can be seen in
-Figure \ref{fig:prediction-example}.
+Figure \ref{fig:prediction-example}. A brief listing of relevant
+statistical terms used throughout this work is provided in
+the Appendix (Section \ref{sec:appendix}).
 
@@
 paper compute predictions as convex combinations of observed function
-values, which may allow for straight forward extensions of this error
+values, which may allow for straightforward extensions of this error
 bound.
@@
 \begin{proofdot}
-  Consider $f(x) - \hat f(x)$ along the line segment $z(t) = (1-t)x_0
-  + t x_j$, $0 \leq t \leq 1.$ By Rolle's Theorem, for some $0 < \hat
-  t < 1$, $\big\langle \nabla f\big(z(\ \hat t\ )\big) - c, x_j - x_0
-  \big\rangle = 0.$ Now
+  \def\g{\gamma_g}
+  Consider $g(t)=f(x(t)) - \hat f(x(t))$ along
+  the line segment $x(t)=(1-t)x_0 + tx_j$, $0\le t\le1$, from $x_0$ to
+  $x_j$.  Observe that $g(0)=f(x_0) - \hat f(x_0) = 0$, $g(1)=f(x_j) -
+  \hat f(x_j) = 0$, and $g'(t)$ is $\g$-Lipschitz continuous with $\g
+  = \gamma \|x_j-x_0\|_2^2$.
+
+  Suppose $g'(0) > \g/2$. Then $|g'(0)-g'(t)| \le \g t \implies g'(0)
+  - \g t \le g'(t)$, and the line $w=g'(0)-\g t$ intersects the
+  $t$-axis at $\tilde t=g'(0)/\g > 1/2$. $\tilde t<1$ necessarily,
+  since by Rolle's Theorem there exists $0<z<1$ such that $g'(z)=0$
+  and so $g'(0) - \g z \le g'(z) = 0$. Now
+  $$\frac{g'(0)^2}{2\g} = \frac{g'(0)\tilde t}{2} = \int_0^{\tilde t}
+  (g'(0)-\g t)dt \le \int_0^{\tilde t} g'(t)dt = g(\tilde t),$$ and
   \begin{align*}
-    \big| \big< \nabla f(x_0) - c, x_j - x_0 \big \rangle \big|
-        &= \big| \big< \nabla f(x_0) - \nabla f\big(z(\ \hat t\ )\big) + \nabla f\big(z(\ \hat t\ )\big) - c, x_j - x_0 \big \rangle \big| \\
-	&= \big| \big\langle \nabla f(x_0) - \nabla f \big(z(\ \hat t\ )\big), x_j - x_0 \big \rangle \big| \\
-	&\leq \big \| \nabla f(x_0) - \nabla f\big(z(\ \hat t\ )\big) \big \|_2 \|x_j - x_0\|_2 \\
-	&\leq \gamma \|x_0 - z(\ \hat t\ )\|_2\ \|x_j - x_0\|_2 \\
-	&\leq \frac{\gamma}{2} \|x_j - x_0\|_2^2 \leq \frac{\gamma k^2}{2}, \\
+    \frac{-g'(0)^2}{2\g} < g'(0) - \frac{g'(0)^2}{2\g} - \g/2 =
+    \frac{(1-\tilde t)(g'(0)-\g)}{2} &= \int_{\tilde t}^1 (g'(0)-\g
+    t)dt \\ &\le \int_{\tilde t}^1 g'(t)dt = -g(\tilde t),
   \end{align*}
-  for all $1 \leq j \leq d.$ Using {\it Lemma \ref{lemma:2}},
-  $$\big \| \nabla f(x_i) - c \big \|_2^2 \leq \frac{d}{\sigma_d^2}
-  \big( \gamma k^2 / 2 \big)^2 \implies \big \| \nabla f(x_i) - c \big \|_2
-  \leq \sqrt{d} \frac{\gamma k^2}{2 \sigma_d}.$$
-  \qed
+  \noindent a contradiction for the value of $g(\tilde t)$. A similar
+  contradiction arises for $g'(0)<-\g/2$. Therefore $|g'(0)| \le
+  \g/2$. In terms of $f$,
+  $$\bigl|\langle \nabla f(x_0)-c,x_j-x_0\rangle\bigr| = |g'(0)| \le
+  \gamma \|x_j-x_0\|_2^2/2 \le \gamma k^2 / 2,$$ which holds for all
+  $1\le j\le d$.  Finally, using {\it Lemma \ref{lemma:2}},
+  $$\|\nabla f(x_0)-c\|_2^2 \le \frac{d}{\sigma_d^2}(\gamma k^2/2)^2
+  \implies \|\nabla f(x_0)-c\|_2 \le \sqrt{d}\frac{\gamma
+    k^2}{2\sigma_d}.$$ \qed
 \end{proofdot}
@@
 \begin{proofdot}
-  Let $v = \nabla f(x_0) - c$, where $\|v\|_2 \leq \sqrt{d} \gamma k^2 / (2 \sigma_d)$ by {\it Lemma \ref{lemma:3}}. Now
+  Let $v = \nabla f(x_0) - c$, where $\|v\|_2 \leq \sqrt{d} \gamma k / (2 \sigma_d)$ by {\it Lemma \ref{lemma:3}}. Now
   \begin{align*}
@@
 
-The theoretical error bound is tight and that can be observed with the
-test function $q(x) = \|x\|_2^2,$ $x \in \mathbb{R}^d$ for $d \in
-\mathcal{N},$ and the simplex defined by vertices $X = \{\mathbf{0}, e_1,
-\ldots, e_d\}$ where $e_i$ is $i$-th unit vector in $\mathbb{R}^d.$
-The relevant constants to the error bound are
+The theoretical error bound is sharp and that can be observed with the
+test function $q(x) = \|x\|_2^2,$ $x \in \mathbb{R}^d,$ and the
+simplex defined by vertices $X = \{0, e_1, \ldots, e_d\},$ where $e_i$
+is $i$-th standard basis vector in $\mathbb{R}^d,$ $e = (1,\ldots,1)
+\in \mathbb{R}^d,$ and $0$ denotes the zero vector in any dimension.
+The constants relevant to the error bound are
 
 \vspace{-2mm}
-$$ \gamma = 2, \qquad \sigma_d = 1, \qquad k = 1, \qquad x_0 =
-\mathbf{0}, \qquad \hat q(x) = \langle\mathbf{1}, x - \mathbf{0}\rangle +
-q(\mathbf{0}). $$
+$$ \gamma = 2, \qquad \sigma_d = 1, \qquad k = 1, \qquad x_0 = 0,
+\qquad \hat q(x) = \langle e, x - 0\rangle + q(0). $$
 \vspace{-2mm}
 
-\noindent Noting that $q(\mathbf{0}) = 0,$ the approximation error at $z = -\mathbf{1/2}$ is
+\noindent Noting that $q(0) = 0,$ the approximation error at $z =
+-(1/2)e$ is
 
@@
   \big|q(z) - \hat q(z)\big| \quad
-  =& \quad \big|\|z\|_2^2 - \langle\mathbf{1},z\rangle\big| \\
-  =& \quad \bigg| \frac{d}{4} + \frac{d}{2} \bigg| \quad = \quad \frac{3d}{4},
+  =& \quad \big|\|z\|_2^2 - \langle e,z \rangle\big| \\
+  =& \quad \big| d/4 + d/2 \big| \quad = \quad 3d/4,
 \end{align*}
@@
 
-\noindent while plugging the constants into the error bound from the theorem gives
+\noindent while the error bound from the theorem gives
 
@@
   \big|q(z) - \hat q(z)\big| \quad
-  \leq& \quad \frac{\gamma \|z - x_0\|_2^2}{2} + \sqrt{d} \frac{\gamma k^2}
+  \leq& \quad \frac{\gamma \|z - x_0\|_2^2}{2} + \sqrt{d} \frac{\gamma k}
       {2 \sigma_d} \|z - x_0\|_2 \\
   =& \quad \|z\|_2^2 + \sqrt{d} \|z\|_2 \\
-  =& \quad \frac{d}{4} + \frac{d}{2} \quad = \quad \frac{3d}{4}.
+  =& \quad d/4 + d/2 \quad = \quad 3d/4.
 \end{align*}
@@
 
-\noindent In light of this error bound, an empirical evaluation of
-presented interpolation and regression algorithms follows.
+\noindent Acknowledging that the error bound is sharp, it may be of
+interest to observe the error of piecewise linear approximation
+techniques on an analytic test function.
+
+\begin{figure}
+  \centering
+  \begin{tikzpicture}
+    \node (img)  {\includegraphics[width=0.9\textwidth,]{Fig2.pdf}};
+    \node[below=of img, node distance=1cm, yshift=1cm] {Number of Data Points};
+    \node[left=of img, node distance=0cm, rotate=90, anchor=center,yshift=-0.7cm] {Absolute Error};
+  \end{tikzpicture}
+  \caption{Delaunay and MLP approximations are constructed from Fekete
+    points over the unit cube evaluating the test function $f(x) =
+    \cos(\|x\|_2)$ for $x \in \mathbb{R}^2$. The figure shows the
+    first/third quartiles at the box bottom/top, the second quartile
+    (median) at the white bar, median 95\% confidence interval (cones,
+    barely visible in figure), and whiskers at 3/2 of the adjacent
+    interquartile ranges, for the absolute prediction error for each
+    model at $1000$ random evaluation points. The left plot observes a
+    perfect interpolation problem with exact evaluations of $f.$ The
+    right plot observes a regression problem with uniform random noise
+    giving values in $[.9 f(x),\ 1.1f(x)]$ for each $x.$ Both axes are
+    log scaled.}
+  \label{fig:convergence-2d}
+\end{figure}
 
@@
 \vspace{-2mm}
-\section{Analytic Test of Convergence}
+\subsection{Demonstration on an Analytic Test Function}
 \label{sec:analytic}
@@
 uniform error bounds. For that reason, an analytic function is used to
-test the error of interpolants and regressors given an increasing
-number of examples and increasing relative random error at data
-points.
+measure the error of a piecewise linear interpolant (Delaunay) and a
+piecewise linear regressor (the MLP) when provided an increasing
+amount of data with varying levels of random noise.
+
+The test function chosen here for analysis resembles the
+\textit{oscillatory} function used by \cite{barthelmann2000high}.
+However, a slight modification is made to remove the simple dot
+product structure (which is favorable for the MLP). Let $f(x)
+=\cos(\|x\|_2)$ for $x \in \mathbb{R}^d$ where $d$ is either $2$
+(Fig. \ref{fig:convergence-2d}) or $20$
+(Fig. \ref{fig:convergence-20d}). This function has a bounded change
+in gradient $2$-norm and hence meets the necessary Lipschitz condition
+for the error bound. Data points and approximation points for this
+test will be within the unit cube $[0,1]^d$.
+
+The error of a linear interpolant constructed from well spaced points
+is dominated by the distance to the nearest data point. In order to
+uniformly decrease the distance to the nearest data point across the
+unit cube, exponentially more data points must be available for
+approximation. For this experiment $N = 2(4^i),$ $i = 1,$ $\ldots,$
+$6,$ chosen points are kept well spaced by computing approximate
+Fekete points. Fekete points have a history in potential theory
+\cite{kovari_pommerenke_1968} and are most generally defined as those
+points that maximize the absolute value of a Vandermonde
+determinant. Here, the QR method outlined in \cite{bos2010computing}
+is used to identify approximate Fekete points from a Vandermonde
+matrix. A multivariate polynomial basis of size $2N$ is constructed
+containing all polynomials of degree $\leq n$ for the largest $n$ such
+that ${d+n \choose n} \leq 2N$ and $2N - {d+n \choose n}$
+arbitrarily selected polynomials of degree $n + 1.$ The Vandermonde
+matrix for these $2N$ polynomials is used to select $N$ approximate
+Fekete points.
+
+Finally an additional aspect is added to this test problem by
+incorporating random uniform noise into the evaluations of $f.$ For
+each test two experiments are executed, one with exact function
+evaluations (for interpolation) and one with a constant signal-to-noise
+(SNR) ratio of 10:1 (for regression).
+
+\begin{figure}
+  \centering
+  \begin{tikzpicture}
+    \node (img)  {\includegraphics[width=0.9\textwidth,]{Fig3.pdf}};
+    \node[below=of img, node distance=1cm, yshift=1cm] {Number of Data Points};
+    \node[left=of img, node distance=0cm, rotate=90, anchor=center,yshift=-0.7cm] {Absolute Error};
+  \end{tikzpicture}
+  \caption{Delaunay and MLP approximations are constructed from Fekete
+    points over the unit cube evaluating the test function $f(x) =
+    \cos(\|x\|_2)$ for $x \in \mathbb{R}^{20}$. The details are the
+    same as for Fig. \ref{fig:convergence-2d}.}
+  \label{fig:convergence-20d}
+\end{figure}
+
+The bound from the theorem suggests that exponentially increasing the
+number of data points should result in a decreasing error. The $d = 2$
+test seen on the left half of Figure \ref{fig:convergence-2d} shows a
+consistent decrease in error for Delaunay and also shows the eventual
+accuracy plateau obtained by a parametric regression form (the MLP, at
+roughly 500 points). On the right hand side of Figure
+\ref{fig:convergence-2d} the random noise clearly prohibits Delaunay
+from converging to $f$, while the MLP is able to improve its
+approximation with more data points on average.  The convergence
+result for a very low-dimensional problem like this is expected.
+However, intuition fails for higher dimensional problems.
+
+In Figure \ref{fig:convergence-20d} shows the test function with $d =
+20$ presents a significantly more challenging approximation problem
+than its counterpart in low dimension. The same increase in number of
+data points from 32 to 8192 causes no apparent improvement in
+approximation for either the noise-free or the noisy
+problems. Perhaps unexpectedly, the interpolation technique performs
+better than the regression technique on the noisy data (right) in
+Figure \ref{fig:convergence-20d}, and worse than the regression
+technique on the noise-free data (left). This result emphasizes the
+relevance of interpolation for problems in high dimension. It also
+reveals that the outcome of MLP regressions can get worse when adding
+more data points.
+
+In summary, the regime of interpolation is much greater for moderate
+to high dimensional problems, even for functions with random noise
+incorporated into their evaluation. Acknowledging the viability of
+interpolation for problems of moderate dimension, the next section
+will consider real-world problems of similar proportion (thousands of
+examples in tens of dimensions).
 
@@
 data set, tables of accompanying numerical results are located in the
-Appendix. All of the test data sets capture underlying functions that
-are almost certainly stochastic. As described in Section
-\ref{sec:introduction}, regression techniques appear most appropriate
-for these problems. However, typically data grows exponentially more
-sparse with increasing dimension. Given sparse data, regressors tend
-towards interpolation.
+Appendix (Section \ref{sec:appendix}). All of the test data sets
+capture underlying functions that are almost certainly stochastic. As
+described in Section \ref{sec:introduction}, regression techniques
+appear most appropriate for these problems. However, typically data
+grows exponentially more sparse with increasing dimension. Given that
+sparse data regressors tend towards interpolation and, as demonstrated
+in Section \ref{sec:analytic}, interpolants produce similar (if not
+identical) results, it is presumed that interpolants are equally
+viable approximation techniques on these problems.
 
@@
   \begin{tikzpicture}
-    \node (img)  {\includegraphics[width=0.75\textwidth,]{Fig2.pdf}};
+    \node (img)  {\includegraphics[width=0.75\textwidth,]{Fig4.pdf}};
     \node[below=of img, node distance=1cm, yshift=1cm] {Forest Fire Area Burned};
@@
   \begin{tikzpicture}
-    \node (img)  {\includegraphics[width=0.8\textwidth,]{Fig3.pdf}};
+    \node (img)  {\includegraphics[width=0.8\textwidth,]{Fig5.pdf}};
     \node[below=of img, node distance=1cm, yshift=1cm] {Area Burned Error};
@@
   \begin{tikzpicture}
-    \node (img)  {\includegraphics[width=0.75\textwidth,]{Fig4.pdf}};
+    \node (img)  {\includegraphics[width=0.75\textwidth,]{Fig6.pdf}};
     \node[below=of img, node distance=1cm, yshift=1cm] {Parkinson's Total UPDRS Score};
@@
   \begin{tikzpicture}
-    \node (img)  {\includegraphics[width=0.8\textwidth,]{Fig5.pdf}};
+    \node (img)  {\includegraphics[width=0.8\textwidth,]{Fig7.pdf}};
     \node[below=of img, node distance=1cm, yshift=1cm] {Total UPDRS Score Error};
@@
   \begin{tikzpicture}
-    \node (img)  {\includegraphics[width=0.75\textwidth,]{Fig6.pdf}};
+    \node (img)  {\includegraphics[width=0.75\textwidth,]{Fig8.pdf}};
     \node[below=of img, node distance=1cm, yshift=1cm] {Daily Rainfall in Sydney};
@@
   \begin{tikzpicture}
-    \node (img)  {\includegraphics[width=0.8\textwidth,]{Fig7.pdf}};
+    \node (img)  {\includegraphics[width=0.8\textwidth,]{Fig9.pdf}};
     \node[below=of img, node distance=1cm, yshift=1cm] {Sydney Rainfall Tomorrow Error};
@@
   \begin{tikzpicture}
-    \node (img)  {\includegraphics[width=0.75\textwidth,]{Fig8.pdf}};
+    \node (img)  {\includegraphics[width=0.75\textwidth,]{Fig10.pdf}};
     \node[below=of img, node distance=1cm, yshift=1cm] {Credit Card Transaction Amount};
@@
   \begin{tikzpicture}
-    \node (img)  {\includegraphics[width=0.8\textwidth,]{Fig9.pdf}};
+    \node (img)  {\includegraphics[width=0.8\textwidth,]{Fig11.pdf}};
     \node[below=of img, node distance=1cm, yshift=1cm] {Transaction Amount Error};
@@
   \begin{tikzpicture}
-    \node (img)  {\includegraphics[width=0.75\textwidth,]{Fig10.pdf}};
+    \node (img)  {\includegraphics[width=0.75\textwidth,]{Fig12.pdf}};
     \node[below=of img, node distance=1cm, yshift=1cm] {I/O Read Throughput};
@@
   \begin{tikzpicture}
-    \node (img)  {\includegraphics[width=0.8\textwidth,]{Fig11.pdf}};
+    \node (img)  {\includegraphics[width=0.8\textwidth,]{Fig13.pdf}};
     \node[below=of img, node distance=1cm, yshift=1cm] {Read Throughput KS Statistic};
@@
     previously unseen system configuration, using $10$-fold cross
-    validation. The KS statistic between the observed distribution at
-    each system configuration and the predicted distribution is
-    recorded and presented above. Note that the above figure is
-    \textit{not} log-scaled like Figure \ref{fig:hist-throughput}. The
-    numerical data corresponding to this figure is provided in Table
-    \ref{table:error-throughput} in the Appendix.}
+    validation. The KS statistic (max norm) between the observed
+    distribution at each system configuration and the predicted
+    distribution is recorded and presented above. Note that the above
+    figure is \textit{not} log-scaled like Figure
+    \ref{fig:hist-throughput}. The numerical data corresponding to
+    this figure is provided in Table \ref{table:error-throughput} in
+    the Appendix.}
   \label{fig:error-throughput}
@@
 
-Delaunay achieves the lowest KS statistic for $62\%$ of
-approximations, while Voronoi is best for the remaining $38\%$. Figure
-\ref{fig:error-throughput} shows that while Delaunay may have more
-best predictions, the behavior of Voronoi may be preferable.
+Delaunay achieves the lowest KS statistic (max norm difference) for
+$62\%$ of approximations, while Voronoi is best for the remaining
+$38\%$. Figure \ref{fig:error-throughput} shows that while Delaunay
+may have more best predictions, the behavior of Voronoi may be
+preferable.
 %% -------------------------------------------------------------------
@@
   \begin{tikzpicture}
-    \node (img)  {\includegraphics[width=0.8\textwidth]{Fig12.pdf}};
+    \node (img)  {\includegraphics[width=0.8\textwidth]{Fig14.pdf}};
     \node[below=of img, node distance=1cm, yshift=1cm] {KS Statistic for Predicted vs. Actual};
@@
 Finally, most of the interpolants presented in this work benefit from
-the ability to model any function over real $d$-tuples with a range
-that is closed under convex combinations. The results of the
-distribution prediction case study indicate that interpolants can
-effectively predict CDFs. The error analysis for that work relies on
-the KS statistic, which captures the worst part of any prediction and
-hence provides a conservatively large estimate of approximation
-error. The average absolute errors in the predicted CDFs are always
-lower than the KS statistics. However, the KS statistic was chosen as
-a metric because of the important surrounding statistical theory. A
-nonnegligible volume of predictions provide impressively low levels of
-average absolute error in that final case study.
+the ability to model \textit{any} function over real $d$-tuples with a
+range that is closed under convex combinations. In general, error can
+be quantified by any measure (particularly of interest may be
+$L^2$, $L^\infty$, etc.). The results of the distribution prediction
+case study indicate that interpolants can effectively predict
+CDFs. The error analysis for that work relies on the KS statistic,
+which captures the worst part of any prediction and hence provides a
+conservatively large estimate of approximation error. The average
+absolute errors in the predicted CDFs are always lower than the KS
+statistics. However, the KS statistic was chosen as a metric because
+of the important surrounding statistical theory. A nonnegligible
+volume of predictions provide impressively low levels of average
+absolute error in that final case study.
 
@@
 encourage broader application of interpolants to multivariate
-approximation in science
+approximation in science.
 
@@
 \begin{appendix}
-\section*{Appendix}
+\section{Appendix}
 \label{sec:appendix}
 
-The tables that follow show the precise experimental results for all
-data sets presented in Section \ref{sec:data}. The tests were all run
-serially on an otherwise idle machine with a CentOS 6.10 operating
-system and an Intel i7-3770 CPU operating at 3.4 GHz. The detailed
-performance results in the tables that follow are very much dependent
-on the problem and the algorithm implementation (e.g., some codes are
-TOMS software, some industry distributions, and others are from
-conference paper venues). Different typeface is used to show best
-performers, however not much significance should be attached to
-ranking algorithms based on small time (millisecond) differences. The
-results serve as a demonstration of conceptual validity.
+\begin{normalsize}
+\textbf{Statistical Terminology.} A random variable $X$ is precisely
+defined by its \textit{cumulative distribution function} (CDF) $F_X$
+and the derivative of the CDF, the \textit{probability density
+  function} (PDF) $f_X.$ For any possible value $x$ of $X$, the
+\textit{percentile} of $x$ is $100$ $F_X(x)$, the percentage values
+drawn from $X$ that would be less than or equal to $x$ as the number
+of samples tends towards infinity. The \textit{quartiles} of $X$ are
+its $25$-th, $50$-th (median), and $75$-th percentiles. The absolute
+difference between the median and an adjacent quartile is an
+\textit{interquartile range}. Given an independent and identically
+distributed sample from $X$ and presuming that $X$ has finite mean and
+variance, a \textit{confidence interval} can be drawn about any
+percentile estimated from the sample. A \textit{confidence interval}
+describes the probability that a value lies within an interval. The
+\textit{null hypothesis} is a statement (derived from some test
+statistic) that the expected value of the observed statistic is equal
+to an assumed population statistic. The $p$ value is the probability
+of observing a given statistic if the \textit{null hypothesis} is
+true.
+
+\vspace{4mm}
+\noindent \textbf{Raw Numerical Results.} The tables that follow show the
+precise experimental results for all data sets presented in Section
+\ref{sec:data}. The tests were all run serially on an otherwise idle
+machine with a CentOS 6.10 operating system and an Intel i7-3770 CPU
+operating at 3.4 GHz. The detailed performance results in the tables
+that follow are very much dependent on the problem and the algorithm
+implementation (e.g., some codes are TOMS software, some industry
+distributions, and others are from conference paper venues). Different
+typeface is used to show best performers, however not much
+significance should be attached to ranking algorithms based on small
+time (millisecond) differences. The results serve as a demonstration
+of conceptual validity.
+
+\end{normalsize}
+
+\newpage
 

### 2019-10-15 Paper_[2019-02]_NUMA/Data/test.py

@@ -1,167 +1,0 @@
-# ==========================================================
-#      Test a difficult problem for the box spline code     
-# ==========================================================
-
-import numpy as np
-from util.system import load, save
-from balltree import BallTree
-from util.approximate import BoxMesh
-from util.math import pairwise_distance
-
-# Load in the example (try for the minimal example, otherwise raw).
-try:
-    train_x, train_y, test_pt = load("small-dim-bad-example.pkl")
-except:
-    try:
-        train_x, train_y, test_pt = load("small-bad-example.pkl")    
-    except:
-        train_x, train_y, test_pt = load("bad_box_case.pkl")
-
-
-# First reduction is in the number of points.
-if (train_x.shape[0] > 10):
-    print()
-    print("train_x: ",train_x.shape)
-    print("train_y: ",np.array(train_y).shape)
-    print("test_pt: ",test_pt.shape)
-    print()
-
-    # Construct a ball tree to get point statistics
-    bt = BallTree(train_x)
-    d, i = bt.nearest(test_pt, k=20)
-    print()
-    print("nearest i: ",i)
-    print("nearest d: ",d)
-    print()
-    pd = pairwise_distance(train_x)
-    pd[np.diag_indices(len(pd))] = float('inf')
-    print(pd)
-    print(np.min(pd))
-    print()
-
-    # Perform an initial reduction, to a range that still causes the error.
-    train_x = train_x[649:975]
-
-    # Now perform a thorough reduction.
-    m = BoxMesh()
-    while True:
-        keep = np.arange(len(train_x))[1:]
-        print("remaining pts: ",len(keep)+1, end="\r")
-        for i in range(len(keep)):
-            m.fit(train_x[keep])
-            inds,_ = m(test_pt)
-            # Break the search if we know it still doesn't work.
-            if (len(inds) == 0):
-                train_x = train_x[keep]
-                break
-            # Otherwise, cycle to drop the next point.
-            else: keep[i] = i
-        # If we made it through the whole loop and none could be dropped,
-        # then break out of the `while` loop.
-        else:
-            m.fit(train_x[keep])
-            inds,_ = m(test_pt)
-            # Break the search if we know it still doesn't work.
-            if (len(inds) != 0): break
-            else: train_x = train_x[keep]
-    # Save the smallest possible bad example.
[…]

### 2019-12-09 Paper_[2019-11]_HPC/paper/paper.tex
Working on the draft the HPC paper, nearly done.

%**************************************************************************
%* SpringSim 2020 Author Kit
%*
%* Word Processing System: TeXnicCenter and MiKTeX
%*
%**************************************************************************

\documentclass{scspaperproc}

\usepackage{latexsym}
\usepackage{graphicx}
\usepackage{mathptmx}

%
%****************************************************************************
% AUTHOR: You may want to use some of these packages. (Optional)
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{amsbsy}
\usepackage{amsthm}
%****************************************************************************

%
%****************************************************************************
% AUTHOR: If you do not wish to use hyperlinks, then just comment
% out the hyperref usepackage commands below.

%% This version of the command is used if you use pdflatex. In this case you
%% cannot use ps or eps files for graphics, but pdf, jpeg, png etc are fine.

\usepackage[pdftex,colorlinks=true,urlcolor=blue,citecolor=black,anchorcolor=black,linkcolor=black,bookmarks=false]{hyperref}

%% The next versions of the hyperref command are used if you adopt the
%% outdated latex-dvips-ps2pdf route in generating your pdf file. In
%% this case you can use ps or eps files for graphics, but not pdf, jpeg, png etc.
%% However, the final pdf file should embed all fonts required which means that you have to use file
%% formats which can embed fonts. Please note that the final PDF file will not be generated on your computer!
%% If you are using WinEdt or PCTeX, then use the following. If you are using
%% Y&Y TeX then replace "dvips" with "dvipsone"

%% \usepackage[dvips,colorlinks=true,urlcolor=blue,citecolor=black,%
%% anchorcolor=black,linkcolor=black]{hyperref}

%% The use of the long citation format (e.g. "Brown and Edwards (1993)" rather than "[5]") and at the same
%% time using the hyperref package can lead to hard to trace bugs in case the citation is broken accross the
%% line (usually this will mark the entire paragraph as a hyperlink (clickable) which is easily noticeable and fixed
%% if using colorlinks, but not if the color is black -- as it is now). Worse yet, if a citation spans page boundary,
%% LaTeX compilation can fail, with an obscure error message. Since this depends a lot on the flow of the text
%% and wording, these bugs come and go and can be extremely hard for a beginner to trace. The error
%% message can look like this:
%%
%%    ! pdfTeX error (ext4): \pdfendlink ended up in different nesting level than \pdfstartlink.
%%    \AtBegShi@Output ...ipout \box \AtBeginShipoutBox 
%%    \fi \fi 
%%    l.174 
%%    ! ==> Fatal error occurred, no output PDF file produced!
%%
%% and can be universally fixed by putting an \mbox{} around the citation in question (in this case, at line 174)
%% and maybe adapting the wording a little bit to improve the paragraph typesetting, which is perhaps not
%% immediately obvious.
%****************************************************************************

%****************************************************************************
%*
%* AUTHOR: YOUR CALL!  Document-specific macros can come here.
%*
%****************************************************************************

% add custom hyphenation rules here
\usepackage{hyphenat}
\hyphenation{op-tical net-works semi-conduc-tor}

% If you use theorems
\newtheoremstyle{scsthe}% hnamei
{8pt}% hSpace abovei
{8pt}% hSpace belowi
{\it}% hBody fonti
{}% hIndent amounti1
{\bf}% hTheorem head fontbf
{.}% hPunctuation after theorem headi
{.5em}% hSpace after theorem headi2
{}% hTheorem head spec (can be left empty, meaning `normal')i

\theoremstyle{scsthe}
\newtheorem{theorem}{Theorem}
\renewcommand{\thetheorem}{\arabic{theorem}}
\newtheorem{corollary}[theorem]{Corollary}
\renewcommand{\thecorollary}{\arabic{corollary}}
\newtheorem{definition}{Definition}
\renewcommand{\thedefinition}{\arabic{definition}}

% avoid overrunning the right margin; you are welcome to remove this, provided that you take care not to overrun the right margin anywhere in your paper
\sloppy

%#########################################################
%*
%*  The Document.
%*
\begin{document}

%***************************************************************************
% AUTHOR: AUTHOR NAMES GO HERE
% FORMAT AUTHORS NAMES Like: Author1, Author2 and Author3 (last names)
%
%		You need to change the author listing below!
%               Please list ALL authors using last name only, separate by a comma except
%               for the last author, separate with "and"

\SCSpagesetup{Lux, Watson, Chang, Xu, Wang, and Hong}

% AUTHOR: Uncomment ONE of these correct conference names.
\def\SCSconferenceacro{SpringSim'20}
%\def\SCSconferenceacro{SummerSim}
%\def\SCSconferenceacro{AutumnSim}
%\def\SCSconferenceacro{PowerPlantSim}

% AUTHOR: Set the correct year of the conference.
\def\SCSpublicationyear{2020}

% AUTHOR: Set the correct month and dates; the dates are separated by a single minus sign
% with no spaces and no leading zeros, the month is a full name (e.g. April) with the first letter
% capitalized. For example, "April 8-13".
\def\SCSconferencedates{May 19-May 21}

% AUTHOR: Set the correct venue in the form "City, State, Country", for example "Los Angeles, CA, USA".
\def\SCSconferencevenue{Fairfax, VA, USA}

% AUTHOR: Enter the title, all letters in upper case
\title{An Algorithm for Constructing Monotone Piecewise \\ Quintic Hermite Interpolating Polynomials}

% AUTHOR: Enter the authors of the article, see end of the example document for further examples
\author{
Thomas C. H. Lux\\
Layne T. Watson\\
Tyler H. Chang\\ [12pt]
Department of Computer Science\\
Virginia Polytechnic Institute and State University $\,$\\
2000 Torgersen Hall, Blacksburg, VA, USA\\
\{tchlux,ltwatson,thchang\}@vt.edu\\
\and
Li Xu\\%
Yueyao Wang\\%
Yili Hong\\ [12pt]
Department of Statistics\\
$\,$ Virginia Polytechnic Institute and State University\\
213 Hutcheson Hall, Blacksburg, VA, USA\\
yilihong@vt.edu\\
}

\maketitle

%% Examples and notes.
%% 
%% \begin{table}[htb]
%%   \centering
%%   \caption{Caption above the table!}\label{tab:example}
%%   \begin{tabular}{r|l}
%%     First & row \\ \hline
%%     second & row \\
%%     third & row \\
%%   \end{tabular}
%% \end{table}
%% 
%% \begin{figure}[htb]
%%   \centering
%%   \includegraphics[width=0.50\textwidth]{path_without_extension}
%%   \caption{Caption below the figure!}\label{fig:example}
%% \end{figure}
%% 
%% \cite{ref}  - Make a citation that is in parenthesis
%% \citeN{ref} - Make a citation out of parenthesis (for noun usage)
%% \shortcite{ref}  - Make a citation for references with >3 authors.
%% \shortciteN{ref} - Make a noun-form citation for references with >3 authors.
%% 
%% 
%% 5 - 12 page paper mandated
%% 150 word abstract
%% 5 keywords maximum

\section*{Abstract}

An algorithm for computing monotone piecewise quintic interpolating
polynomials is proposed. Algebraic constraints for enforcing
monotonicity are provided that align with quintic monotonicity
theory. The algorithm is implemented, tested, and applied to several
sample problems to demonstrate the improved accuracy of piecewise
quintic monotone interpolation compared to the existing piecewise
cubic solution.

\textbf{Keywords:} monotone, spline, Hermite interpolation, quintic polynomial.

\section{Introduction and Motivation}
\label{sec:introduction}

%% \begin{figure}[htb]
%%   \centering
%%   \includegraphics[width=0.50\textwidth]{demo_projection}
%%   \caption{These are the projection mechanisms that make a cubic polynomial piece monotone.}\label{fig:projection}
%% \end{figure}

%% The ability to interpolate or regress a function with continuous value and derivatives has proven fundamentally useful. In many applications, there are known shapes or characteristics that are expected of approximations. For example, a cumulative distribution function in statistics cannot decrease, the entropy of a closed thermodynamic system can never increase, an electric motor should not suddenly change direction when slowing down or speeding up. In each of these cases it is important that any constructed spline approximation is monotone.

Many domains of science rely on smooth approximations to real-valued functions over a closed interval. Piecewise polynomial functions (splines) provide the smooth approximations for animation in graphics \cite{herman2015techniques,quint2003scalable}, aesthetic structural support in architecture, efficient aerodynamic surfaces in automotive and aerospace engineering \cite{brennan2019measure}, prolonged effective operation of electric motors \cite{berglund2009planning}, and accurate non parametric approximations in statistics \cite{knott2012interpolating}. While polynomial interpolants or regressors apply broadly, splines are often a good choice because they can approximate globally complex functions while minimizing the local complexity of an approximation.

It is often the case that the true underlying function or phenomenon being modeled has known properties e.g., convexity, positivity, various levels of continuity, or monotonicity. Given a reasonably large amount of data, it can be impossible to maintain these properties with a single polynomial function. In general, the maintenance of function properties through interpolation / regression is usually referred to as \textit{shape preserving} \cite{fritsch1980monotone,gregory1985shape}. The specific shapes this work will focus on are monotonicity and multiple levels of continuity for a function. These properties are chiefly important to the approximation of cumulative distribution functions and subsequently the effective generation of random numbers from a specified distribution.

In statistics especially, the construction of a monotone interpolating spline that is continuous in second derivative is meaningfully useful \cite{ramsay1988monotone}. A function with these properties could approximate random variables to a high level of accuracy with relatively few intervals. A continuously twice differentiable approximation to a cumulative distribution function (CDF) would also produce a corresponding probability density function (PDF) that is continuously differentiable, which is a property many standard parametric distributions maintain.

The currently available software for monotone piecewise polynomial interpolation includes quadratic \cite{he1998monotone}, cubic \cite{fritsch1980monotone}, and quartic (with limited application) \cite{wang2004rational,piah2011improved,yao2018unconditionally} cases. Even though theory has been provided for the quintic case \cite{ulrich1994positivity,hess1994positive}, this theory has not yet been used to construct a piecewise quintic monotone spline interpolation routine. Recent work suggests that the lack of quintic software may be due to a general unawareness of the theory \cite{xie2018semiparametric}.

\begin{figure}
  \centering
  \includegraphics[width=.9\textwidth]{spline-demonstration}
  \caption{Example polynomials that interpolate function values at the ends of the interval $[0,1]$. The first only interpolates the function values $f(0) = 1$ and $f(1) = 0$, making it the order two polynomial $f(x) = 1 - x$. For the second plot $f(x) = 8x^3 - 12x^2 + 4x + 1/2$, which is order four and interpolates the values $f(0) = 1/2$, $f'(0) = 4$, $f(1) = 1/2$, $f'(1) = 4$. Finally the third plot shows the order six polynomial $f(x) = - 64x^5 + 160x^4 - 140x^3 + 50x^2 - 7x + 1$ interpolating the function values $f(0) = 1$, $f'(0) = -7$, $f''(0) = 100$, $f(1) = 0$, $f'(1) = -7$, $f''(1) = -100$. Notice that interpolating the same fixed number of function values at each endpoint will always result in an even order interpolating polynomial.
  }\label{fig:spline-demonstration}
\end{figure}

The importance of piecewise quintic interpolation over lower order approximations can be simply demonstrated. In general, the order of a polynomial determines the number of function values it can interpolate, and the growth rate of error away from the interpolated function values. As demonstrated in Figure \ref{fig:spline-demonstration}, it can be seen that matching a value at either end of the interval requires an order two (linear) approximation and each additional derivative at the ends of the interval raises the necessary polynomial order by two. The body of this work is composed of a novel algorithm for enforcing monotonicity on quintic polynomial pieces, then extending that solution to work on generic splines.

%% Software exists that does monotone piecewise quintic regression \cite{}, but the available routines do not computationally take advantage of the quintic monotonicity theory. ???

The major contribution of this work is an algorithm for constructing piecewise quintic Hermite interpolating polynomials that utilizes existing quintic monotonicity theory. The remainder of this paper is structured as follows: ...

\subsection{Computing a Monotone Cubic Interpolant}

The current state of the art monotone interpolating spline with a mathematical software implementation is piecewise cubic, continuously differentiable, and was first proposed in \cite{fritsch1980monotone} then expanded upon in \cite{carlson1985monotone}. Let $\pi: x_0 = k_1 < k_2 < \cdots < k_n = x_1$ be a partition of the interval $[x_0,x_1]$. Let $f: \mathbb{R} \rightarrow \mathbb{R}$ and $\{f(k_i) : i = 1,2,\ldots,n\}$ be a given set of data values at the partition points for a monotone function $f$, meaning $f(k_i) \leq f(k_{i+1})$ for $i = 1, \ldots, n-1$ or $f(k_i) \geq f(k_{i+1})$ for $i = 1, \ldots, n-1$. Let $\hat f$ be a piecewise cubic spline defined in each sub-interval $I_i = [k_i, k_{i+1}]$ by

\begin{align*}
h_i =& \ k_{i+1} - k_{i} \\
u(t) =& \ 3t^2 - 2t^3 \\
p(t) =& \ t^3 - t^2 \\
\hat f(x) =& \ f(k_i)\ u\big((k_{i+1} - x) / h_i\big) + f(k_{i+1})\ u\big((x - k_i) / h_i\big) \\
& - \hat f^{\ \prime}(k_i)\ p\big((k_{i+1}-x)/h_i\big) + \hat f^{\ \prime}(k_{i+1})\ p\big((x-k_i)/h_i \big).
\end{align*}

Notice that a trivially monotone spline results when $\hat f^{\ \prime}(k_i) = 0$, for $i = 1, \ldots, n$. However, such a spline has too many \textit{wiggles} for most applications. Fritsch and Carlson show that simple conditions on the derivative values can guarantee monotonicity, and that these conditions can be enforced in a way that ensures modifications on one interval will not break the monotonicity of cubic polynomials over any neighboring intervals. Consider the terms $\alpha = \big(\hat f^{\ \prime}(k_i) (k_{i+1}-k_i)\big) / \big(f(k_{i+1}) - f(k_i)\big)$ and $\beta = \big(\hat f^{\ \prime}(k_{i+1}) (k_{i+1}-k_i)\big) / \big(f(k_{i+1}) - f(k_i)\big)$, now monotonicity of a cubic polynomial over a sub-interval can be maintained by ensuring that $\alpha$ and $\beta$ reside in any of the regions depicted in Figure \ref{fig:projection}.

\begin{figure}
  \centering
  \includegraphics[width=0.50\textwidth]{demo_projection}
  \caption{These are the feasible regions of monotonicity for cubic splines and the projections that make a cubic polynomial piece monotone.}\label{fig:projection}
\end{figure}

The actual region of monotonicity for a cubic polynomial is larger, but projection of $(\alpha, \beta)$ into one of these regions ensures that monotonicity will be achieved and not violated for neighboring regions. The user must decide which region is most for the projections based on the application, Fritsch and Carlson recommend using region 2.

While the cubic monotonicity case presents a straight-forward solution, the region of monotonicity is not so simple in the quintic case. In the next section, an algorithm for performing a projection similar to those for cubic polynomials is proposed.

%% - show picture of how the cubic is computed

%% - show the algorithm for computing a cubic (and that it is trivial)

%% - quintic interpolation is more accurate when data is available, or higher order information can be inferred

\section{Monotone Quintic Interpolation}

When provided data that has no assigned first and second derivative values, the derivative data is filled by a linear fit of neighboring data points. End points are set to be the slope between the end and its nearest neighbor.

The method for finding the transition point of a boolean function on a line is the Golden Section search. This will be referred to in pseudo code as \texttt{line\_search($g,$ $a,$ $b$)} where $a,$ $b$ $\in S$ for $S$ closed under convex combination, $g: S \rightarrow \{0,1\}$ is a boolean function, and $g(b) = 1$. If $g(a) = 1$ then $a$ is returned, otherwise the smallest $c \in [0,1]$ such that $g\big(a(1-c) + c b\big) = 1$ is returned.

After assigning function values and derivative values, an interpolating function is constructed from a quintic B-spline basis.

\vspace{10pt}

%% ----------------------------------------------------------------------

\subsection{Verifying Monotonicity of a Quintic Polynomial}
\label{is_monotone}
Let $f$ be a quintic polynomial over a closed interval $[x_0, x_1]$ $\subset \mathbb{R}$. Now $f$ is uniquely defined by the evaluation tuples $\big(x_0,$ $f(x_0),$ $f'(x_0),$ $f''(x_0)\big)$ and $\big(x_1,$ $f(x_1),$ $f'(x_1),$ $f''(x_1)\big).$ Assume without loss of generality that $f(x_0) < f(x_1),$ where the case of monotonic decreasing $f$ would consider the negated the function values. The following algorithm will determine whether or not $f$ is monotone increasing on the interval $[x_0, x_1].$

\vspace{10pt}%
\hrule%
\vspace{3pt}%
\noindent\textbf{\textit{Algorithm 1:}} \texttt{is\_monotone}%
\vspace{3pt}%
\hrule%

\begin{itemize}
  \itemsep0pt
  \parskip0pt

\item[0:] \texttt{if $\big(f'(x_0) = 0$ or $f'(x_1) = 0\big)$ return is\_monotone\_simplified}
\item[1:] \texttt{if $\big(f'(x_0) < 0$ or $f'(x_1) < 0\big)$ return FALSE}
  \begin{itemize}
    \item[] \textit{This can be seen clearly from the fact that $f$ is analytic; there will exist some nonempty interval about $x_0$ or $x_1$ for which $f'$ is negative.}
  \end{itemize}

\item[2:] $A = f'(x_0)\frac{x_1 - x_0}{f(x_1) - f(x_0)}$
\item[3:] $B = f'(x_1) \frac{x_1 - x_0}{f(x_1) - f(x_0)}$
  \begin{itemize}
    \item[] \textit{The variables $A$ and $B$ correspond directly to the theoretical foundation for positive quartic polynomials established in \cite{ulrich1994positivity}, first defined after equation 18.}
  \end{itemize}
\item[8:] $\gamma_0 = 4 \frac{f'(x_0)}{f'(x_1)} (B/A)^{3/4}$
\item[9:] $\gamma_1 = \frac{x_1 - x_0}{f'(x_1)} (B/A)^{3/4}$
\item[4:] $\alpha_0 = 4 (B/A)^{1/4}$
\item[5:] $\alpha_1 = -\frac{x_1 - x_0}{f'(x_1)} (B/A)^{1/4}$
\item[6:] $\beta_0 = 30 - \frac{12 \big(f'(x_0) + f'(x_1)\big) (x_1 - x_0)}{\big(f(x_1) - f(x_0)\big) \sqrt{A}\sqrt{B}}$
\item[7:] $\beta_1 = \frac{-3 (x_1 - x_0)^2}{2 \big(f(x_1) - f(x_0)\big) \sqrt{A} \sqrt{B}} $
  \begin{itemize}
    \item[] \textit{The $\gamma,$ $\alpha,$ and $\beta$ terms with subscripts $0$ and $1$ are algebraic reductions of the simplified conditions for satisfying Theorem 2 in \cite{ulrich1994positivity} (equation 16). These terms with subscripts $0$ and $1$ give the computation of $\alpha,$ $\beta,$ and $\gamma$ the form seen in lines $10$-$12$ below.}
  \end{itemize}
\item[10:] $\gamma = \gamma_0 + \gamma_1 f''(x_0)$
\item[11:] $\alpha = \alpha_0 + \alpha_1 f''(x_1)$
\item[12:] $\beta = \beta_0 + \beta_1 \big(f''(x_0) - f''(x_1)\big)$
\item[13:] \texttt{if $(\beta \leq 6)$ then return $\alpha > - (\beta + 2) / 2$}
\item[14:] \texttt{else return $\gamma > -2 \sqrt{\beta - 2}$ }

\end{itemize}
\hrule
\vspace{10pt}
%% ----------------------------------------------------------------------

%% ----------------------------------------------------------------------
Given the same initial conditions there are special circumstances which allow for the usage of simpler monotonicity conditions. In this case, consider when the quintic function has either $f'(x_0) = 0$ or $f'(x_1) = 0.$ This reduces the problem of verifying monotonicity to one of cubics established by \cite{schmidt1988positivity}.

\vspace{10pt}%
\hrule%
\vspace{3pt}%
\noindent\textbf{\textit{Algorithm 1b:}} \texttt{is\_monotone\_simplified}%
\vspace{3pt}%
\hrule%
\begin{itemize}
  \itemsep0pt
  \parskip0pt

\item[0:] $\alpha = 30 - \frac{(x_1 - x_0)\big( 14 f'(x_0) + 16 f'(x_1) - \big(f''(x_1) - f''(x_0) \big) (x_1 - x_0)\big)}{2\big(f(x_1) - f(x_0)\big)}$
\item[1:] $\beta = 30 - \frac{(x_1 - x_0)\big( 2 f'(x_0) + 24 f'(x_1) - \big(f''(x_0) + 3 f''(x_1) \big) (x_1 - x_0)\big)}{2\big(f(x_1) - f(x_0)\big)}$
\item[2:] $\gamma = \frac{(x_1 - x_0)\big( 7 f'(x_0) + f''(x_0) (x_1 - x_0) \big)}{f(x_1) - f(x_0)}$
\item[3:] $\delta = \frac{f'(x_0) (x_1 - x_0)}{f(x_1) - f(x_0)}$

  \begin{itemize}
    \item[] \textit{The variables above are algebraic expansions of the coefficients for the cubic derivative function in \cite{schmidt1988positivity}.}
  \end{itemize}

\item[4:] \texttt{if $\big($min$(\alpha, \delta) < 0\big)$ return FALSE}
\item[5:] \texttt{else if $\big(\beta < \alpha - 2 \sqrt{\alpha \delta}\big)$ return FALSE}
\item[6:] \texttt{else if $\big(\gamma < \delta - 2 \sqrt{\alpha \delta}\big)$ return FALSE}
\item[7:] \texttt{else return TRUE}

\end{itemize}
\hrule
\vspace{10pt}
%% ----------------------------------------------------------------------

Next the modification of a quintic spline to enforce monotonicity will be discussed.

%% ----------------------------------------------------------------------
\subsection{Enforcing Monotonicity of a Quintic Polynomial}

\vspace{10pt}%
\hrule%
\vspace{3pt}%
\noindent\textbf{\textit{Algorithm 2a:}} \texttt{make\_monotone}%
\vspace{3pt}%
\hrule%

\begin{itemize}
  \itemsep0pt
  \parskip0pt

\item[0:] \texttt{if }$\big(f(x_1) - f(x_0) = 0\big)$ \texttt{ return } $f'(x_0) = f'(x_1) = f''(x_0) = f''(x_1) = 0$
\item[1:] $f'(x_0) = \ $\texttt{median}$\big(0, f'(x_0), 14 \frac{f(x_1) - f(x_0)}{x_1 - x_0} \big)$
\item[2:] $f'(x_1) = \ $\texttt{median}$\big(0, f'(x_1), 14 \frac{f(x_1) - f(x_0)}{x_1 - x_0}\big)$
  \begin{itemize}
  \item[] \textit{This selection of values for $f'(x_0)$ and $f'(x_1)$ is suggested by \cite{ulrich1994positivity} (originally from \cite{huynh1993accurate}), and quickly enforces upper and lower bounds on derivative values to ensure quintic monotonicity is obtainable.}
  \end{itemize}
\item[3:] $A = f'(x_0)\frac{x_1 - x_0}{f(x_1) - f(x_0)}$
\item[4:] $B = f'(x_1) \frac{x_1 - x_0}{f(x_1) - f(x_0)}$
\item[5:] \texttt{if }$AB \leq 0$\texttt{ return make\_monotone\_simplified}
\item[6:] \texttt{if $\big($max}$(A,B) > 6\big)$\\$f'(x_0) = 6 f'(x_0)\  /$\texttt{ max}$(A,B)$\\$f'(x_1) = 6 f'(x_1)\  /$\texttt{ max}$(A,B)$

  \begin{itemize}
  \item[] \textit{This box bound ensures that $(A,B)$ remains within a viable region of monotonicity (satisfying Theorem 4, seen in Fig. 6 of \cite{ulrich1994positivity}).}
  \end{itemize}

\item[7:] $\hat f''(x_0) = - \sqrt{A} \big( 7 \sqrt{A} + 3 \sqrt{B} \big) \frac{f(x_1) - f(x_0)}{(x_1 - x_0)^2}$\\$\hat f''(x_1) = \sqrt{B} \big( 3 \sqrt{A} + 7 \sqrt{B} \big) \frac{f(x_1) - f(x_0)}{(x_1 - x_0)^2}$

  \begin{itemize}
    \item[] \textit{This selection of values of $f''$ is guaranteed to satisfy Theorem 4 from \cite{ulrich1994positivity} and is chosen because it is (reasonably) the average of the two endpoints of the interval of monotonicity for second derivative values.}
  \end{itemize}

\item[8:] $\eta = \big(f''(x_0), f''(x_1)\big)$\\$\eta_0 = \big(\hat f''(x_0), f''(x_1)\big)$\\$f''(x_0), f''(x_1) = $\texttt{ line\_search$\big($is\_monotone, $\eta,$ $\eta_0\big)$}
\end{itemize}
\hrule
\vspace{10pt}
%% ----------------------------------------------------------------------

Similar to the simplified check for monotonicity, when the derivative value at one endpoint of the interval is $0$, a simplified set of steps can be taken to enforce monotonicity.

%% ----------------------------------------------------------------------
\vspace{10pt}%
\hrule%
\vspace{3pt}%
\noindent\textbf{\textit{Algorithm 2b:}} \texttt{make\_monotone\_simplified}%
\vspace{3pt}%
\hrule%

\begin{itemize}
  \itemsep0pt
  \parskip0pt

\item[0:] $f''(x_0) = $\texttt{ max}$\bigg(f''(x_0), \frac{-6 f'(x_0)}{x_1 - x_0}\bigg)$

  \begin{itemize}
    \item[] \textit{Considering the $\alpha,$ $\gamma,$ $\beta,$ and $\delta$ defined in \cite{schmidt1988positivity}, this first step enforces $\gamma > \delta.$ It is already guaranteed that $\delta = \frac{f'(x_0)(x_1 - x_0)}{f(x_1) - f(x_0)} > 0.$ Only two conditions remain to guarantee monotonicity.}
  \end{itemize}

\item[1:] $f''(x_1) = $\texttt{ max}$\bigg( f''(x_1), f''(x_0) + \frac{14 f'(x_0) + 16 f'(x_1) + 60(f(x_1) - f(x_0)) / (x_1 - x_0)}{x_1 - x_0} \bigg)$

  \begin{itemize}
    \item[] \textit{Now it is guaranteed that $\alpha \geq 0.$}
  \end{itemize}

\item[2:] $f''(x_1) = $\texttt{ max}$\bigg( f''(x_1), 6 f'(x_0) - 4 f'(x_1) - f''(x_0) \bigg)$

  \begin{itemize}
    \item[] \textit{Lastly, this guarantees that $\beta \geq \alpha.$ All conditions are met to satisfy proposition 2 of \cite{schmidt1988positivity} and ensure monotonicity.}
  \end{itemize}
\end{itemize}
\hrule
\vspace{10pt}
%% ----------------------------------------------------------------------
Notice that all above algorithms (assuming a fixed level of precision is desired) have $\mathcal{O}(1)$ runtime. Only a finite number of operations are needed for monotonicity verification. A line search is performed for enforcement, however that search requires a fixed number of steps to achieve any predetermined relative precision on the line.

%% ----------------------------------------------------------------------
\subsection{Constructing a Piecewise Quintic Monotone Spline}
\label{monotone_spline}

Finally, the construction of a full piecewise quintic spline is outlined using the above algorithms. Let $f: \mathbb{R} \rightarrow \mathbb{R}$ be a function in $\mathcal{C}^2.$ Proceed given evaluation tuples $\big(x_i,$ $f(x_i),$ $f'(x_i),$ $f''(x_i)\big)$ for $i = 0,\ldots,N$ such that $x_i < x_{i+1}$ and (without loss of generality) $f(x_i) \leq f(x_{i+1})$ for $i = 1,$ $\ldots,$ $N-1$.

\vspace{10pt}%
\hrule%
\vspace{3pt}%
\noindent\textbf{\textit{Algorithm 3:}} \texttt{monotone\_spline}%
\vspace{3pt}%
\hrule%

\begin{itemize}
  \itemsep0pt
  \parskip0pt

\item[0:] \texttt{for }$i=0,\ldots,N-1$
\item[1:] $\quad$\texttt{if $\big($not is\_monotone$(i,i+1)\big)$  make\_monotone$(i,i+1)$}

  \begin{itemize}
    \item[] \textit{In the shorthand notation above, $i$ and $i+1$ refer to the associated tuples of the form $(x_i,$ $f(x_i),$ $f'(x_i),$ $f''(x_i)).$ This notation will be used through the remainder of this algorithm.}
  \end{itemize}

\item[2:] $\quad$\texttt{for $j=i-1,\ldots,0$}
\item[3:] $\quad\quad$\texttt{if $\big($not is\_monotone$(j,j+1)\big)$  make\_monotone$(j,j+1)$}
\item[4:] $\quad\quad$\texttt{else break}

  \begin{itemize}
    \item[] \textit{The above `for' loop will be referred to as the cascade fix. If the adjustment of second derivative values causes the previous interval to become nonmonotone, then the left-hand second derivative value must be updated. This may (abnormally) require adjustments across all previously visited intervals.}
  \end{itemize}

\item[5:] $\quad$\texttt{while $\big($not is\_monotone$(i,i+1)\big)$}
\item[6:] $\quad\quad f'(x_i) = f'(x_i) / k$
\item[7:] $\quad\quad f'(x_{i+1}) = f'(x_{i+1}) / k$
\item[8:] $\quad\quad$\texttt{make\_monotone$(i,i+1)$}

  \begin{itemize}
    \item[] \textit{In the case that the two corrections to neighboring intervals contradict, the first derivative values of the active interval are decreased to enlarge the overlap of regions I and II (from \cite{ulrich1994positivity}) of the two intervals.}
  \end{itemize}

\item[9:] $\quad\quad$\texttt{for $j=i-1,\ldots,0$}
\item[10:] $\quad\quad\quad$\texttt{if $\big($not is\_monotone$(j,j+1)\big)$  make\_monotone$(j,j+1)$}
\item[11:] $\quad\quad\quad$\texttt{else break}

  \begin{itemize}
    \item[] \textit{Finally an additional `cascade fix' is performed to ensure that all previous intervals are still monotone after shrinking the derivative values of the current interval.}
  \end{itemize}

\end{itemize}
\hrule
\vspace{10pt}
%% ----------------------------------------------------------------------

It is mentioned in \cite{ulrich1994positivity} that for sufficiently small $f'(x_i)$ and $f'(x_{i+1})$ the admissible solution interval of second derivative values becomes arbitrarily large. It can also be seen that decreasing the assigned derivative on right-hand side of an interval always allows the achievement of monotonicity because shrinking $f'(x_1)$ results in $\gamma$ growing faster than $\sqrt{\beta}$ for \textit{algorithm 1a}, for \textit{algorithm 1b} $\beta$ and $\gamma$ will grow faster $\alpha$ and $\delta$ respectively. In application, one needs only to pick some $k > 1$ to ensure successful termination.

Given the potential for successive cascade fixes, the runtime of \textit{algorithm 3} is $\mathcal{O}(N^2)$. In practice this has been observed to be incredibly unlikely and difficult to produce. While single cascade fixes can occur from hand crafted examples, the generously rounded values selected to ensure monotonicity frequently prevent successive cascade fixes.

%% - background is provided by these papers

%% - Table of relevant terms and what they mean

%% \subsection{Algorithm: Monotone Quintic Interpolation}
%% - correctness of algorithm
%% - runtime of algorithm
%% - weakness of algorithm

\section{Experimental Results}

- show a Figure of a simple quintic interpolant and its derivatives

\subsection{Approximating a Trigonometric Function}

For this experiment, the function $\sin(x) + x$ is considered over the
interval $[0,(5/2)\pi]$.

\begin{figure}[htb]
  \centering
  \includegraphics[width=.7\textwidth]{cubic-quintic-sin}
  \caption{Depicted above are the monotone cubic and quintic spline
    interpolant of the function $\sin(x) + x$ over the nodes $[0,
      (5/2) \pi]$.  Notice that the maximum error of the cubic
    interpolant is larger, because it only captures first derivative
    information at the nodes. The quintic interpolant captures both
    first and second derivative information at the nodes.
  }\label{fig:cubic-quintic-sin}
\end{figure}

- Pick a trig function, approximate it with cubic and quintic

- Show a Figure, and a Table of the maximum error and distribution of
  error across the segment

\subsection{Approximating a Cumulative Distribution Function}

- generate large sets of random monotone data, show Table of number
  of metrics growing with increasing number of nodes

- show visual of the \textit{error}

\subsection{Random Monotone Data}

- generate large sets of random monotone data, show Table of number
  of metrics growing with increasing number of nodes (how many binary
  searches, how many corrections were needed, steps were taken)

\section{Discussion}

- this algorithm has its quirks, but it works

- this is an improvement over the cubic, although the complexity of
  the algorithm is greatly increased as well

\section{Future Work}

- a method for finding regions of monotonicity that is less ad-hoc

- constructing monotone interpolants for higher order approximations

- using improved distribution estimates to improve distribution modeling

\section{Conclusion}

- this paper proposes and tests an algorithm for constructing quintic
  monotone interpolants

- the results show an improvement over the cubic usage

\section*{Acknowledgments}
The NSF is acknowledged for its support of this project.

% Please don't change the bibliographystyle style
\bibliographystyle{scsproc}
% AUTHOR: Include your bib file here
\bibliography{paper}

\section*{Author Biographies}

\textbf{\uppercase{THOMAS C. H. LUX}} is a Ph.D. candidate in Computer Science at Virginia Polytechnic Institute and State University. His research interests include Computational Science, Approximation Theory, Optimization, and Artificial Intelligence. His email address is \email{tchlux@vt.edu}.

\newpage

\appendix

\section{Author Checklist}
We strive for a consistent appearance in all papers published in the proceedings. If you used the template and styles within this author’s kit, then almost all of the requirements in this checklist will be automatically satisfied, and there is very little to check.

Please \textbf{print a hardcopy of your paper}, and go over your printed paper to make sure it adheres to the following requirements. \textit{Thank you!}

\begin{enumerate}
	\item Abstract
  \begin{enumerate}
	  \item 150 or fewer words.
	  \item Provide 3-5 keywords (mandatory). This set of keywords will identify your paper in indices and databases.
	\end{enumerate}
	\item Paper Length
  \begin{enumerate}
	  \item At least 5, but no more than 12 pages.
	  \item Page size is letter size (8.5’’ x 11’’, or 216 mm x 279 mm).
	\end{enumerate}
	\item All text is in 11-Point Times New Roman except title, header and footer (default in this template).
	\item Paper title is in 12-Point Times New Roman \textbf{BOLDFACE ALL CAPS} (default in this template).
	\item The paper has been spellchecked using U.S. English. 
	\item Spacing and Margins
  \begin{enumerate}
	  \item Single spaced.
	  \item Left and right margins are each 1 inch (default in this template).
	  \item Top and bottom margins are according to the template.
	  \item Title starts 1.25 inches from the top of the page (default in this template).
	\end{enumerate}
	\item Section Headings
  \begin{enumerate}
	  \item Left justified and set in \textbf{BOLDFACE ALL CAPS} (default in this template).
	  \item Numbered, except for the abstract, acknowledgments, references and author biographies.
	  \item Subsection headings are set in \textbf{Boldface Headline Style} (default in this template).
	\end{enumerate}
	\item No footnotes or page numbers.
	\item The copyright notice on the first page contains the name of the symposium or a track where you want to submit your paper, and the running head on subsequent pages is the surnames of all authors.
	\item Multiple authors are formatted correctly.
	\item Equations are centered and any equation numbers are in parentheses and right-justified (default).
	\item Figures and Tables
  \begin{enumerate}
	  \item All text in figures and tables is readable.
	  \item Table captions appear above the table.
	  \item Figure captions appear below the figure.
	\end{enumerate}
	\item Citations and References
  \begin{enumerate}
	  \item Citations are by author and year, and are enclosed in parentheses, not brackets (default \BibTeX). 
	  \item References are in the hangref style, and are listed alphabetically by the last names(s) of the author(s) (also default \BibTeX\ setting in this template). 
	\end{enumerate}
	\item Author biographies are one paragraph per author.
	\item Hyperlinks
  \begin{enumerate}
	  \item Be sure that hyperlinks will probably work in the future as well.
	  \item Live hyperlinks are blue. Nonlive hyperlinks are black (default in this template).
	\end{enumerate}
	\item All fonts must be embedded in the resulting \texttt{.pdf}. If you use \texttt{pdflatex}, this should already be true--unless you included figures in the \texttt{.eps} or \texttt{.pdf} format which may introduce additional font dependencies. You can use the \texttt{pdffonts} command to see if all the fonts are embedded (the column ``emb'' should say yes for all rows). You can use GhostScript to force embed the fonts, like so: {\texttt{gs -dNOPAUSE -dBATCH -sDEVICE=pdfwrite -dEmbedAllFonts=true -sOutputFile=out.pdf -f in.pdf}} (in Windows, use \texttt{gswin32} or \texttt{gswin32c} to invoke GhostScript).
\end{enumerate}

After verifying that your paper meets these requirements, please go to the final submission page at conference website and submit your paper. Be sure to complete the transfer of copyright form and upload the \texttt{.pdf} receipt. \textit{Thank you for contributing to the SCS conferences!}

\section{Most Observed Mistakes}

The following list comprises \textbf{the most common sources of error} that had to be corrected by previous editors. Please make sure to go through the following list and check that your paper is formatted correctly:
\begin{enumerate}
	\item	The paper is less than 5 or more than 12 pages long.
	\item	Paper title and section titles are in \textbf{BOLD ALL CAPS}, subsections are \textbf{Bold and Capitalize the First Letter of Important Words}. Please use the templates.
	\item	Paper is in A4 format, not letter format. Please use the required margins.
	\item	The copyright notice is incorrect.
	\item	The running heads are incorrect. Do not forget that the LastNameLastAuthor is preceded by ``, and ''.
	\item	The citation format is incorrect. Use \BibTeX\ for citations and do not change the format used by the template.
	\item	The biographies are missing. Do not forget the ``author biographies'' section.
	\item	Figures or tables are not referenced in the text or have the incorrect caption format.
	\item	The author section after the title is not formatted correctly, the number of organizations does not define the number of blocks, or the number of blocks does not define the layout.
	\item In the heading on the title page, country names are in all capitals.
	\item	Paragraphs are not indented.
	\item Some fonts in the resulting \texttt{.pdf} are not embedded.
\end{enumerate}

\end{document}

### 2019-12-09 Paper_[2019-11]_HPC/Code/__init__.py

@@ -1,0 +1,7 @@
+'''
+This is the code package for the HPC paper.
+
+All of the code necesasry to run 3 expirements involving monotone
+splines exists here.
+
+'''

### 2019-12-09 Paper_[2019-11]_HPC/paper/paper.tex
Working through the algorithm section of the paper.

@@
 Many domains of science rely on smooth approximations to real-valued functions over a closed interval.
-Piecewise polynomial functions (splines) provide the smooth approximations for animation in graphics \cite{herman2015techniques,quint2003scalable}, aesthetic structural support in architecture, efficient aerodynamic surfaces in automotive and aerospace engineering \cite{brennan2019measure}, prolonged effective operation of electric motors \cite{berglund2009planning}, and accurate non parametric approximations in statistics \cite{knott2012interpolating}.
+Piecewise polynomial functions (splines) provide the smooth approximations for animation in graphics \cite{herman2015techniques,quint2003scalable}, aesthetic structural support in architecture \cite{brennan2019measure}, efficient aerodynamic surfaces in automotive and aerospace engineering \cite{brennan2019measure}, prolonged effective operation of electric motors \shortcite{berglund2009planning}, and accurate non parametric approximations in statistics \cite{knott2012interpolating}.
 While polynomial interpolants or regressors apply broadly, splines are often a good choice because they can approximate globally complex functions while minimizing the local complexity of an approximation.
@@
 It is often the case that the true underlying function or phenomenon being modeled has known properties e.g., convexity, positivity, various levels of continuity, or monotonicity.
-Given a reasonably large amount of data, it can be impossible to maintain these properties with a single polynomial function.
-In general, the maintenance of function properties through interpolation / regression is usually referred to as \textit{shape preserving} \cite{fritsch1980monotone,gregory1985shape}.
-The specific shapes this work will focus on are monotonicity and multiple levels of continuity for a function.
+Given a reasonable amount of data, it quickly becomes difficult to achieve desirable properties in a single polynomial function.
+In general, the maintenance of function properties through interpolation / regression is referred to as \textit{shape preserving} \cite{fritsch1980monotone,gregory1985shape}.
+The specific shapes this work will achieve in approximations are monotonicity and two levels of continuity.
 These properties are chiefly important to the approximation of cumulative distribution functions and subsequently the effective generation of random numbers from a specified distribution.
@@
 
-The currently available software for monotone piecewise polynomial interpolation includes quadratic \cite{he1998monotone}, cubic \cite{fritsch1980monotone}, and quartic (with limited application) \cite{wang2004rational,piah2011improved,yao2018unconditionally} cases.
-Even though theory has been provided for the quintic case \cite{ulrich1994positivity,hess1994positive}, this theory has not yet been used to construct a piecewise quintic monotone spline interpolation routine.
-Recent work suggests that the lack of quintic software may be due to a general unawareness of the theory \cite{xie2018semiparametric}.
+The currently available software for monotone piecewise polynomial interpolation includes quadratic \cite{he1998monotone}, cubic \cite{fritsch1980monotone}, and (with limited application) quartic \cite{wang2004rational,piah2011improved,yao2018unconditionally} cases.
+In addition, a statistical method for bootstrapping the construction of an arbitrarily smooth monotone fit exists \cite{leitenstorfer2006generalized}, but the method does not take advantage of known analytic properties related to quintic polynomials.
+Theory has been provided for the quintic case \cite{ulrich1994positivity,hess1994positive}, however this theory has not yet been used to construct a piecewise quintic monotone spline interpolation routine.
+Recent work suggests that the lack of quintic software may be due to a general unawareness of the theory \shortcite{xie2018semiparametric}.
 
@@
 The actual region of monotonicity for a cubic polynomial is larger, but projection of $(\alpha, \beta)$ into one of these regions ensures that monotonicity will be achieved and not violated for neighboring regions.
-The user must decide which region is most for the projections based on the application, Fritsch and Carlson recommend using region 2.
+The user must decide which region is most appropriate for the projections based on the application, Fritsch and Carlson recommend using region 2.
 
-While the cubic monotonicity case presents a straight-forward solution, the region of monotonicity is not so simple in the quintic case.
+While the cubic monotonicity case affords such a concise solution, the region of monotonicity is not so simple in the quintic case.
 In the next section, an algorithm for performing a projection similar to those for cubic polynomials is proposed.
@@
 %% - show picture of how the cubic is computed
-
 %% - show the algorithm for computing a cubic (and that it is trivial)
-
 %% - quintic interpolation is more accurate when data is available, or higher order information can be inferred
@@
 
-When provided data that has no assigned first and second derivative values, the derivative data is filled by a linear fit of neighboring data points.
-End points are set to be the slope between the end and its nearest neighbor.
+The following section is composed of three algorithms that together are used to construct a piecewise quintic monotone interpolating spline.
+Section \ref{is_monotone} regards checking monotonicity, Section \ref{make_monotone} presents a procedure for enforcing monotonicity on a polynomial piece, and Section \ref{monotone_spline} combines the previous two algorithms into a routine for enforcing monotonicity over a quintic spline.
 
-The method for finding the transition point of a boolean function on a line is the Golden Section search.
-This will be referred to in pseudo code as \texttt{line\_search($g,$ $a,$ $b$)} where $a,$ $b$ $\in S$ for $S$ closed under convex combination, $g: S \rightarrow \{0,1\}$ is a boolean function, and $g(b) = 1$.
-If $g(a) = 1$ then $a$ is returned, otherwise the smallest $c \in [0,1]$ such that $g\big(a(1-c) + c b\big) = 1$ is returned.
+Similar to the cubic case, let $\pi: x_0 = k_1 < k_2 < \cdots < k_n = x_1$ be a partition of the interval $[x_0,x_1]$.
+Let $f: \mathbb{R} \rightarrow \mathbb{R}$ and $\{f(k_i) : i = 1,2,\ldots,n\}$ be a given set of values at the partition points for a monotone function $f$, meaning $f(k_i) \leq f(k_{i+1})$ or $f(k_i) \geq f(k_{i+1})$ for $i = 1, \ldots, n-1$.
 
-After assigning function values and derivative values, an interpolating function is constructed from a quintic B-spline basis.
+In pseudo code, a function is used that is referred to as \texttt{line\_search($g,$ $a,$ $b$)}, where $a,$ $b$ $\in S$ for $S$ closed under convex combination, $g: S \rightarrow \{0,1\}$ is a binary function, and $g(b) = 1$.
+The \texttt{line\_search} is performed by Binary search in the presented framework, when $g(a) = 1$, $a$ is returned, otherwise the smallest $c \in [0,1]$ such that $g\big(a(1-c) + c b\big) = 1$ is returned.
+
+In the case that first or second derivative information is not available at time of approximation, it is estimated with the central divided difference of the function values.
+The final quintic spline is represented with an exact rational arithmetic Newton Polynomial.
+% and evaluated with a B-spline basis.
 
@@
 \vspace{3pt}%
-\noindent\textbf{\textit{Algorithm 1:}} \texttt{is\_monotone}%
+\noindent\textbf{\textit{Algorithm 1:}} \texttt{is\_monotone$\big (x_0$,$x_1$,$f \big)$}%
 \vspace{3pt}%
@@
 
-\item[0:] \texttt{if $\big(f'(x_0) = 0$ or $f'(x_1) = 0\big)$ return is\_monotone\_simplified}
-\item[1:] \texttt{if $\big(f'(x_0) < 0$ or $f'(x_1) < 0\big)$ return FALSE}
+\item[1:] \texttt{if $\big(f(x_0) = f(x_1)\big)$ return $\big( Df(x_0) = Df(x_1) = D^2f(x_0) = D^2f(x_1) = 0 \big)$}
+\item[2:] \texttt{if $\big(Df(x_0) < 0$ or $Df(x_1) < 0\big)$ return FALSE}
   \begin{itemize}
-    \item[] \textit{This can be seen clearly from the fact that $f$ is analytic; there will exist some nonempty interval about $x_0$ or $x_1$ for which $f'$ is negative.}
+    \item[] \textit{This can be seen clearly from the fact that $f$ is analytic; there will exist some nonempty interval about $x_0$ or $x_1$ for which $Df$ is negative.}
   \end{itemize}
 
-\item[2:] $A = f'(x_0)\frac{x_1 - x_0}{f(x_1) - f(x_0)}$
-\item[3:] $B = f'(x_1) \frac{x_1 - x_0}{f(x_1) - f(x_0)}$
+\item[2:] \texttt{if $\big(Df(x_0) = 0$ or $Df(x_1) = 0\big)$}
+\item[3:] $\quad w = x_0 - x_1$
+\item[4:] $\quad g(y) = \big({-4} Df(x_1) / w \big) < \big( ({-f}(x_1)/w - 24 y - 32 Df(x_1)) / 5w \big)$
+\item[5:] $\quad$ \texttt{line\_search}$(g, 0, Df(x_0))$
+\item[6:] $A = Df(x_0)\frac{x_1 - x_0}{f(x_1) - f(x_0)}$
+\item[7:] $B = Df(x_1) \frac{x_1 - x_0}{f(x_1) - f(x_0)}$
   \begin{itemize}
@@
   \end{itemize}
-\item[8:] $\gamma_0 = 4 \frac{f'(x_0)}{f'(x_1)} (B/A)^{3/4}$
-\item[9:] $\gamma_1 = \frac{x_1 - x_0}{f'(x_1)} (B/A)^{3/4}$
+\item[8:] $\gamma_0 = 4 \frac{Df(x_0)}{Df(x_1)} (B/A)^{3/4}$
+\item[9:] $\gamma_1 = \frac{x_1 - x_0}{Df(x_1)} (B/A)^{3/4}$
 \item[4:] $\alpha_0 = 4 (B/A)^{1/4}$
-\item[5:] $\alpha_1 = -\frac{x_1 - x_0}{f'(x_1)} (B/A)^{1/4}$
-\item[6:] $\beta_0 = 30 - \frac{12 \big(f'(x_0) + f'(x_1)\big) (x_1 - x_0)}{\big(f(x_1) - f(x_0)\big) \sqrt{A}\sqrt{B}}$
+\item[5:] $\alpha_1 = -\frac{x_1 - x_0}{Df(x_1)} (B/A)^{1/4}$
+\item[6:] $\beta_0 = 30 - \frac{12 \big(Df(x_0) + Df(x_1)\big) (x_1 - x_0)}{\big(f(x_1) - f(x_0)\big) \sqrt{A}\sqrt{B}}$
 \item[7:] $\beta_1 = \frac{-3 (x_1 - x_0)^2}{2 \big(f(x_1) - f(x_0)\big) \sqrt{A} \sqrt{B}} $
@@
   \end{itemize}
-\item[10:] $\gamma = \gamma_0 + \gamma_1 f''(x_0)$
-\item[11:] $\alpha = \alpha_0 + \alpha_1 f''(x_1)$
-\item[12:] $\beta = \beta_0 + \beta_1 \big(f''(x_0) - f''(x_1)\big)$
+\item[10:] $\gamma = \gamma_0 + \gamma_1 D^2f(x_0)$
+\item[11:] $\alpha = \alpha_0 + \alpha_1 D^2f(x_1)$
+\item[12:] $\beta = \beta_0 + \beta_1 \big(D^2f(x_0) - D^2f(x_1)\big)$
 \item[13:] \texttt{if $(\beta \leq 6)$ then return $\alpha > - (\beta + 2) / 2$}
@@
 \subsection{Enforcing Monotonicity of a Quintic Polynomial}
+\label{make_monotone}
 
@@
 
-\item[7:] $\hat f''(x_0) = - \sqrt{A} \big( 7 \sqrt{A} + 3 \sqrt{B} \big) \frac{f(x_1) - f(x_0)}{(x_1 - x_0)^2}$\\$\hat f''(x_1) = \sqrt{B} \big( 3 \sqrt{A} + 7 \sqrt{B} \big) \frac{f(x_1) - f(x_0)}{(x_1 - x_0)^2}$
+\item[7:] $f''(x_0) = - \sqrt{A} \big( 7 \sqrt{A} + 3 \sqrt{B} \big) \frac{f(x_1) - f(x_0)}{(x_1 - x_0)^2}$\\$f''(x_1) = \sqrt{B} \big( 3 \sqrt{A} + 7 \sqrt{B} \big) \frac{f(x_1) - f(x_0)}{(x_1 - x_0)^2}$
 
@@
 
-\item[8:] $\eta = \big(f''(x_0), f''(x_1)\big)$\\$\eta_0 = \big(\hat f''(x_0), f''(x_1)\big)$\\$f''(x_0), f''(x_1) = $\texttt{ line\_search$\big($is\_monotone, $\eta,$ $\eta_0\big)$}
+\item[8:] $\eta = \big(f''(x_0), f''(x_1)\big)$\\$\eta_0 = \big(f''(x_0), f''(x_1)\big)$\\$f''(x_0), f''(x_1) = $\texttt{ line\_search$\big($is\_monotone, $\eta,$ $\eta_0\big)$}
 \end{itemize}

### 2019-12-15 Paper_[2019-11]_HPC/paper/paper.tex
Nearly done with HPC paper. Incorporated Dr. Watson revisions. Running third experiment more extensively for final figures.

@@
 
+%% Use a custom `algorithm` command for this paper.
+\input{lux-algorithm}
+
 % add custom hyphenation rules here
@@
 %***************************************************************************
-% AUTHOR: AUTHOR NAMES GO HERE
+% AUTHOR: Thomas C.H. Lux et al.
 % FORMAT AUTHORS NAMES Like: Author1, Author2 and Author3 (last names)
@@
 % AUTHOR: Enter the title, all letters in upper case
-\title{An Algorithm for Constructing Monotone Piecewise \\ Quintic Hermite Interpolating Polynomials}
+\title{An Algorithm for Constructing Monotone \\ Quintic Interpolating Splines}
 
@@
 Tyler H. Chang\\ [12pt]
-Department of Computer Science\\
-Virginia Polytechnic Institute and State University $\,$\\
+Departments of Computer Science, Mathematics,\\
+and Aerospace and Ocean Engineering\\
+Virginia Polytechnic Institute \& State University\\
 2000 Torgersen Hall, Blacksburg, VA, USA\\
-\{tchlux,ltwatson,thchang\}@vt.edu\\
+\{tchlux,ltw,thchang\}@vt.edu\\
 \and
@@
 Department of Statistics\\
-$\,$ Virginia Polytechnic Institute and State University\\
+Virginia Polytechnic Institute \& State University\\
 213 Hutcheson Hall, Blacksburg, VA, USA\\
@@
 
-An algorithm for computing monotone piecewise quintic interpolating
-polynomials is proposed. Algebraic constraints for enforcing
+An algorithm for computing monotone order six piecewise polynomial
+interpolants is proposed. Algebraic constraints for enforcing
 monotonicity are provided that align with quintic monotonicity
 theory. The algorithm is implemented, tested, and applied to several
-sample problems to demonstrate the improved accuracy of piecewise
-quintic monotone interpolation compared to the existing piecewise
-cubic solution.
+sample problems to demonstrate the improved accuracy of monotone
+quintic spline interpolants compared to that for monotone cubic spline
+interpolants.
 
-\textbf{Keywords:} monotone, spline, Hermite interpolation, quintic polynomial.
+\textbf{Keywords:} monotone, quintic spline, Hermite interpolation, sixth order polynomial.
 
@@
 Many domains of science rely on smooth approximations to real-valued functions over a closed interval.
-Piecewise polynomial functions (splines) provide the smooth approximations for animation in graphics \cite{herman2015techniques,quint2003scalable}, aesthetic structural support in architecture \cite{brennan2019measure}, efficient aerodynamic surfaces in automotive and aerospace engineering \cite{brennan2019measure}, prolonged effective operation of electric motors \shortcite{berglund2009planning}, and accurate non parametric approximations in statistics \cite{knott2012interpolating}.
+Piecewise polynomial functions (splines) provide the smooth approximations for animation in graphics \cite{herman2015techniques,quint2003scalable}, aesthetic structural support in architecture \cite{brennan2019measure}, efficient aerodynamic surfaces in automotive and aerospace engineering \cite{brennan2019measure}, prolonged effective operation of electric motors \shortcite{berglund2009planning}, and accurate nonparametric approximations in statistics \cite{knott2012interpolating}.
 While polynomial interpolants or regressors apply broadly, splines are often a good choice because they can approximate globally complex functions while minimizing the local complexity of an approximation.
@@
 Given a reasonable amount of data, it quickly becomes difficult to achieve desirable properties in a single polynomial function.
-In general, the maintenance of function properties through interpolation / regression is referred to as \textit{shape preserving} \cite{fritsch1980monotone,gregory1985shape}.
-The specific shapes this work will achieve in approximations are monotonicity and two levels of continuity.
+In general, the maintenance of function properties through interpolation/regression is referred to as \textit{shape preserving} \cite{fritsch1980monotone,gregory1985shape}.
+The specific shapes this work will achieve in approximations are monotonicity and $C^2$ continuity.
 These properties are chiefly important to the approximation of cumulative distribution functions and subsequently the effective generation of random numbers from a specified distribution.
 
-In statistics especially, the construction of a monotone interpolating spline that is continuous in second derivative is meaningfully useful \cite{ramsay1988monotone}.
-A function with these properties could approximate random variables to a high level of accuracy with relatively few intervals.
-A continuously twice differentiable approximation to a cumulative distribution function (CDF) would also produce a corresponding probability density function (PDF) that is continuously differentiable, which is a property many standard parametric distributions maintain.
+In statistics especially, the construction of a monotone interpolating spline that is $C^2$ continuous is meaningfully useful \cite{ramsay1988monotone}.
+A function with these properties could approximate a cumulative distribution function to a high level of accuracy with relatively few intervals.
+A twice continuously differentiable approximation to a cumulative distribution function (CDF) would also produce a corresponding probability density function (PDF) that is continuously differentiable, which is a property many standard parametric distributions maintain.
 
@@
 In addition, a statistical method for bootstrapping the construction of an arbitrarily smooth monotone fit exists \cite{leitenstorfer2006generalized}, but the method does not take advantage of known analytic properties related to quintic polynomials.
-Theory has been provided for the quintic case \cite{ulrich1994positivity,hess1994positive}, however this theory has not yet been used to construct a piecewise quintic monotone spline interpolation routine.
+Theory has been provided for the quintic case \cite{ulrich1994positivity,hess1994positive}, however this theory has not yet been used to construct a monotone quintic spline interpolation routine.
 Recent work suggests that the lack of quintic software may be due to a general unawareness of the theory \shortcite{xie2018semiparametric}.
@@
   \centering
-  \includegraphics[width=.9\textwidth]{spline-demonstration}
+  \includegraphics[width=.9\textwidth]{spline_demonstration}
   \caption{Example polynomials that interpolate function values at the ends of the interval $[0,1]$.
-The first only interpolates the function values $f(0) = 1$ and $f(1) = 0$, making it the order two polynomial $f(x) = 1 - x$.
-For the second plot $f(x) = 8x^3 - 12x^2 + 4x + 1/2$, which is order four and interpolates the values $f(0) = 1/2$, $f'(0) = 4$, $f(1) = 1/2$, $f'(1) = 4$.
-Finally the third plot shows the order six polynomial $f(x) = - 64x^5 + 160x^4 - 140x^3 + 50x^2 - 7x + 1$ interpolating the function values $f(0) = 1$, $f'(0) = -7$, $f''(0) = 100$, $f(1) = 0$, $f'(1) = -7$, $f''(1) = -100$.
+The first only interpolates the function values $g_2(0) = 1$ and $g_2(1) = 0$, making it the order two polynomial $g_2(x) = 1 - x$.
+For the second plot $g_4(x) = 8x^3 - 12x^2 + 4x + 1/2$, which is order four and interpolates the values $g_4(0) = 1/2$, $g_4'(0) = 4$, $g_4(1) = 1/2$, $g_4'(1) = 4$.
+Finally the third plot shows the order six polynomial $g_6(x) = - 64x^5 + 160x^4 - 140x^3 + 50x^2 - 7x + 1$ interpolating the function values $g_6(0) = 1$, $g_6'(0) = -7$, $g_6''(0) = 100$, $g_6(1) = 0$, $g_6'(1) = -7$, $g_6''(1) = -100$.
 Notice that interpolating the same fixed number of function values at each endpoint will always result in an even order interpolating polynomial.
-  }\label{fig:spline-demonstration}
+  }\label{fig:spline_demonstration}
 \end{figure}
@@
 In general, the order of a polynomial determines the number of function values it can interpolate, and the growth rate of error away from the interpolated function values.
-As demonstrated in Figure \ref{fig:spline-demonstration}, it can be seen that matching a value at either end of the interval requires an order two (linear) approximation and each additional derivative at the ends of the interval raises the necessary polynomial order by two.
-The body of this work is composed of a novel algorithm for enforcing monotonicity on quintic polynomial pieces, then extending that solution to work on generic splines.
+As demonstrated in Figure \ref{fig:spline_demonstration}, it can be seen that matching a value at either end of the interval requires an order two (linear) approximation and each additional derivative at the ends of the interval raises the necessary polynomial order by two.
+The body of this work is composed of a novel algorithm for enforcing monotonicity on quintic polynomial pieces, then extending that solution to work on quintic splines.
 
@@
 
-The major contribution of this work is an algorithm for constructing piecewise quintic Hermite interpolating polynomials that utilizes existing quintic monotonicity theory.
-The remainder of this paper is structured as follows: ...
+The major contribution of this work is an algorithm for constructing monotone quintic interpolating splines that utilizes existing quintic monotonicity theory.
+The remainder of this paper is structured as follows: {\bf ?????}
 
@@
 
-The current state of the art monotone interpolating spline with a mathematical software implementation is piecewise cubic, continuously differentiable, and was first proposed in \cite{fritsch1980monotone} then expanded upon in \cite{carlson1985monotone}.
+The current state-of-the-art monotone interpolating spline with a mathematical software implementation is piecewise cubic, continuously differentiable, and was first proposed in \cite{fritsch1980monotone} then expanded upon in \cite{carlson1985monotone}.
 Let $\pi: x_0 = k_1 < k_2 < \cdots < k_n = x_1$ be a partition of the interval $[x_0,x_1]$.
-Let $f: \mathbb{R} \rightarrow \mathbb{R}$ and $\{f(k_i) : i = 1,2,\ldots,n\}$ be a given set of data values at the partition points for a monotone function $f$, meaning $f(k_i) \leq f(k_{i+1})$ for $i = 1, \ldots, n-1$ or $f(k_i) \geq f(k_{i+1})$ for $i = 1, \ldots, n-1$.
-Let $\hat f$ be a piecewise cubic spline defined in each sub-interval $I_i = [k_i, k_{i+1}]$ by
+Let $f: \mathbb{R} \rightarrow \mathbb{R}$ be $C^2$, and $\bigl\{f(k_i)\bigr\}_{i=1}^n$ and $\bigl\{\Delta_i\bigr\}_{i=1}^n$ be given sets of function and derivative values at the partition points for a monotone function $f$.
+Either $f(k_i) \leq f(k_{i+1})$, $i=1$, $\ldots$, $n-1$, and $\Delta_i\ge0$, $i=1$, $\ldots$, $n$, or $f(k_i) \geq f(k_{i+1})$, $i=1$, $\ldots$, $n-1$, and $\Delta_i\le0$, $i=1$, $\ldots$, $n$.
+Let $\hat f$ be a piecewise cubic polynomial defined in each subinterval $I_i = [k_i, k_{i+1}]$ by
 
 \begin{align*}
-h_i =& \ k_{i+1} - k_{i} \\
-u(t) =& \ 3t^2 - 2t^3 \\
-p(t) =& \ t^3 - t^2 \\
-\hat f(x) =& \ f(k_i)\ u\big((k_{i+1} - x) / h_i\big) + f(k_{i+1})\ u\big((x - k_i) / h_i\big) \\
-& - \hat f^{\ \prime}(k_i)\ p\big((k_{i+1}-x)/h_i\big) + \hat f^{\ \prime}(k_{i+1})\ p\big((x-k_i)/h_i \big).
+h_i &= k_{i+1} - k_{i}, \quad
+u(t) = 3t^2 - 2t^3, \quad
+p(t) = t^3 - t^2,\\
+\hat f(x) &= f(k_i) u\big((k_{i+1} - x) / h_i\big) +
+f(k_{i+1}) u\big((x - k_i) / h_i\big) \\
+&{} - h_i\Delta_i p\big((k_{i+1}-x)/h_i\big) + h_i\Delta_{i+1}
+p\big((x-k_i)/h_i\big).
 \end{align*}
 
-Notice that a trivially monotone spline results when $\hat f^{\ \prime}(k_i) = 0$, for $i = 1, \ldots, n$.
+Notice that a trivially monotone spline results when $\Delta_i = 0$, for $i = 1$, $\ldots$, $n$.
 However, such a spline has too many \textit{wiggles} for most applications.
-Fritsch and Carlson show that simple conditions on the derivative values can guarantee monotonicity, and that these conditions can be enforced in a way that ensures modifications on one interval will not break the monotonicity of cubic polynomials over any neighboring intervals.
-Consider the terms $\alpha = \big(\hat f^{\ \prime}(k_i) (k_{i+1}-k_i)\big) / \big(f(k_{i+1}) - f(k_i)\big)$ and $\beta = \big(\hat f^{\ \prime}(k_{i+1}) (k_{i+1}-k_i)\big) / \big(f(k_{i+1}) - f(k_i)\big)$, now monotonicity of a cubic polynomial over a sub-interval can be maintained by ensuring that $\alpha$ and $\beta$ reside in any of the regions depicted in Figure \ref{fig:projection}.
+\citeN{carlson1985monotone} show that simple conditions on the derivative values can guarantee monotonicity, and that these conditions can be enforced in a way that ensures modifications on one interval will not break the monotonicity of cubic polynomials over any neighboring intervals.
+Consider the terms $\alpha = \big(\Delta_i (k_{i+1}-k_i)\big) / \big(f(k_{i+1}) - f(k_i)\big)$ and $\beta = \big(\Delta_{i+1}(k_{i+1}-k_i)\big) / \big(f(k_{i+1}) - f(k_i)\big)$.
+If $f(k_{i+1}) - f(k_i)=0$, take $\Delta_i = \Delta_{i+1} =0$ and $\alpha =\beta =1$.
+In general, monotonicity of a cubic polynomial over a subinterval can be maintained by ensuring that $\alpha$ and $\beta$ reside in any of the regions depicted in Figure \ref{fig:projection}.
 
@@
   \centering
-  \includegraphics[width=0.50\textwidth]{demo_projection}
-  \caption{These are the feasible regions of monotonicity for cubic splines and the projections that make a cubic polynomial piece monotone.}\label{fig:projection}
+  \includegraphics[width=0.45\textwidth]{cubic_projection_demonstration}
+  \caption{These are the feasible regions of monotonicity for cubic splines and the projections that make a cubic polynomial piece monotone.
+The regions themselves are numbered 1--4 corresponding to their original description in Fritsch and Carlson (1980).
+One point is projected onto each region as a demonstration.}\label{fig:projection}
 \end{figure}
@@
 The actual region of monotonicity for a cubic polynomial is larger, but projection of $(\alpha, \beta)$ into one of these regions ensures that monotonicity will be achieved and not violated for neighboring regions.
-The user must decide which region is most appropriate for the projections based on the application, Fritsch and Carlson recommend using region 2.
+The user must decide which region is most appropriate for the projections based on the application, Fritsch and Carlson recommend using Region 2.
 
@@
 
-%% - show picture of how the cubic is computed
-%% - show the algorithm for computing a cubic (and that it is trivial)
-%% - quintic interpolation is more accurate when data is available, or higher order information can be inferred
-
 \section{Monotone Quintic Interpolation}
 
-The following section is composed of three algorithms that together are used to construct a piecewise quintic monotone interpolating spline.
-Section \ref{is_monotone} regards checking monotonicity, Section \ref{make_monotone} presents a procedure for enforcing monotonicity on a polynomial piece, and Section \ref{monotone_spline} combines the previous two algorithms into a routine for enforcing monotonicity over a quintic spline.
+The following section is composed of three algorithms that together are used to construct a monotone quintic spline interpolant.
+Without loss of generality, the algorithms will only consider the monotone increasing (nondecreasing) case.
+The monotone decreasing (nonincreasing) case is handled similarly.
+Algorithm \ref{alg:check-monotone} checks monotonicity,
+Algorithm \ref{alg:make-monotone} enforces monotonicity on an order six
+polynomial piece of the quintic spline, and Algorithm
+\ref{alg:monotone-spline} uses the previous two algorithms to enforce
+monotonicity for the entire quintic spline.
 
-Similar to the cubic case, let $\pi: x_0 = k_1 < k_2 < \cdots < k_n = x_1$ be a partition of the interval $[x_0,x_1]$.
-Let $f: \mathbb{R} \rightarrow \mathbb{R}$ and $\{f(k_i) : i = 1,2,\ldots,n\}$ be a given set of values at the partition points for a monotone function $f$, meaning $f(k_i) \leq f(k_{i+1})$ or $f(k_i) \geq f(k_{i+1})$ for $i = 1, \ldots, n-1$.
+In pseudocode, a function \texttt{binary\_search($g,$ $a,$ $b$)} is used, where $a,b\in S\subset \mathbb{R}^p$ for convex $S$, $g: S \rightarrow \{$\texttt{FALSE}, \texttt{TRUE}$\}$ is a right continuous Boolean function, $g(b) = \hbox{\texttt{TRUE}}$, and for $\mu \in [0,1]$, $g\big((1-\mu)a + \mu b\big) =\hbox{\texttt{TRUE}} \Longrightarrow g\big((1-\nu)a + \nu b\big) =\hbox{\texttt{TRUE}}$ for $\mu \le \nu\le1$.
+The search returns $\big((1-c)a + c b\big)$ for the smallest $c \in [0,1]$ such that $g\big((1-c)a + c b\big) = \hbox{\texttt{TRUE}}$.
 
-In pseudo code, a function is used that is referred to as \texttt{line\_search($g,$ $a,$ $b$)}, where $a,$ $b$ $\in S$ for $S$ closed under convex combination, $g: S \rightarrow \{0,1\}$ is a binary function, and $g(b) = 1$.
-The \texttt{line\_search} is performed by Binary search in the presented framework, when $g(a) = 1$, $a$ is returned, otherwise the smallest $c \in [0,1]$ such that $g\big(a(1-c) + c b\big) = 1$ is returned.
+These algorithms make use of first and second derivatives of the approximated function.
+In the case that the first and second derivative information is not provided along with function values, the derivatives are estimated with finite differences of the function values.
+The final quintic spline is represented as a piecewise polynomial using the Newton form for each polynomial piece, and ultimately converted to a $B$-spline representation for evaluation.
 
-In the case that first or second derivative information is not available at time of approximation, it is estimated with the central divided difference of the function values.
-The final quintic spline is represented with an exact rational arithmetic Newton Polynomial.
-% and evaluated with a B-spline basis.
+\algorithm
+    {is\_monotone$\big(x_0, x_1, f \big)$}
+    {where $x_0$, $x_1 \in \mathbb{R}$, $x_0 < x_1$, and $f$ is an order six polynomial. Returns \texttt{TRUE} if $f$ is monotone increasing on $[x_0,x_1]$.}
+    {\label{alg:check-monotone}
+  \item if $\big(f(x_0) = f(x_1)\big)$ and not $\big( 0 = Df(x_0) = Df(x_1) = D^2f(x_0) = D^2f(x_1) \big)$
+    \indented{return FALSE}
+  \item if $\big(Df(x_0) < 0$ or $Df(x_1) < 0\big)$
+    \indented{return FALSE}
+    \comment{The necessity of these first two steps follows directly from the fact that $f$ is $C^2$.
+The next case that will be covered is in accordance with a simplified condition for quintic monotonicity that reduces to one of cubic positivity studied in \citeN{schmidt1988positivity}, where $\alpha$, $\beta$, $\gamma$, and $\delta$ are defined in terms of values and derivatives of $f$ at $x_0$ and $x_1$.
+Step \ref{alg1-alpha} checks for the necessary condition that $\alpha \ge 0$, Step \ref{alg1-beta} checks $\beta \geq \alpha$, and Step \ref{alg1-gamma} checks $\gamma \geq \delta$, all from \citeN{schmidt1988positivity}.
+If all necessary conditions are met, then the order six piece is monotone and Step \ref{alg1-simplified-true} concludes this check.}
+  \item if $\big(Df(x_0) = 0$ or $Df(x_1) = 0\big)$
+  \item \indented{$w := x_0 - x_1$\\$v := f(x_0) - f(x_1)$}
+  \item \indented{if $\big(D^2f(x_1) > {-4}Df(x_1) / w \big)$ return FALSE}
+    \label{alg1-alpha}
+  \item \indented{if $\big(D^2f(x_1) < (3w D^2f(x_0) - 24 Df(x_0) - 32
+    Df(x_1) + 60v/w) / (5w) \big)$ return FALSE} \label{alg1-beta}
+  \item \indented{if $\big(D^2f(x_0) < 3 Df(x_0) / w \big)$ return FALSE}
+    \label{alg1-gamma}
+  \item \indented{return TRUE} \label{alg1-simplified-true}
+    \comment{The following code considers the remaining case where $Df(x_0) \neq 0$ and $Df(x_1) \neq 0$.}
+  \item $\displaystyle A := Df(x_0)\frac{x_1 - x_0}{f(x_1) - f(x_0)}$
+    \\ $\displaystyle B := Df(x_1) \frac{x_1 - x_0}{f(x_1) - f(x_0)}$
+    \comment{The variables $A$ and $B$ correspond directly to the theoretical foundation for positive quartic polynomials established in \citeN{ulrich1994positivity}, first defined after Equation (18).}
+  \item $\displaystyle \gamma_0 := 4 \frac{Df(x_0)}{Df(x_1)} (B/A)^{3/4}$
+    \\ $\displaystyle \gamma_1 := \frac{x_1 - x_0}{Df(x_1)} (B/A)^{3/4}$
+    \\ $\alpha_0 := 4 (B/A)^{1/4}$
+    \\ $\displaystyle \alpha_1 := -\frac{x_1 - x_0}{Df(x_1)} (B/A)^{1/4}$
+    \\ $\displaystyle \beta_0 := 30 - \frac{12 \big(Df(x_0) + Df(x_1)\big) (x_1 - x_0)}{\big(f(x_1) - f(x_0)\big) \sqrt{A}\sqrt{B}}$
+    \\ $\displaystyle \beta_1 := \frac{-3 (x_1 - x_0)^2}{2 \big(f(x_1) - f(x_0)\big) \sqrt{A} \sqrt{B}} $ \label{alg1-linearized-abg}
+    \comment{The $\gamma,$ $\alpha,$ and $\beta$ terms with subscripts $0$ and $1$ are algebraic reductions of the simplified conditions for satisfying Theorem 2 in Equation (16) of \citeN{ulrich1994positivity}.
+These terms with subscripts $0$ and $1$ make the computation of $\alpha,$ $\beta,$ and $\gamma$ functions of the second derivative terms, as seen in Step \ref{alg1-abg} below.}
+  \item $\gamma := \gamma_0 + \gamma_1 D^2f(x_0)$
+    \\ $\alpha := \alpha_0 + \alpha_1 D^2f(x_1)$
+    \\ $\beta := \beta_0 + \beta_1 \big(D^2f(x_0) - D^2f(x_1)\big)$
+    \label{alg1-abg}
+  \item if $(\beta \leq 6)$ then return $\big(\alpha > - (\beta + 2) / 2\big)$
+  \item else return $\big(\gamma > -2 \sqrt{\beta - 2}\big)$ \label{alg1-last}
+}
+%% ----------------------------------------------------------------------
 
-\vspace{10pt}
+The reason for structuring the $\alpha$, $\beta$, and $\gamma$ computations in terms of the second derivative of $f$ (seen in Step \ref{alg1-abg} of Algorithm \ref{alg:check-monotone}) will become more apparent later.
+The next problem to consider is that of making a nonmonotone order six polynomial piece into a monotone one by modifying its first and second derivative values at the ends of an interval.
+Note that the actual value of the function at the ends of the interval is not modified, as the resulting polynomial needs to interpolate.
+
+%% ----------------------------------------------------------------------
+\algorithm
+    {make\_monotone$\big (x_0$,$x_1$,$f \big)$}
+    {where $x_0,\ x_1 \in \mathbb{R}$, $x_0 < x_1$, and $f$ is an order six polynomial. Returns monotone $f$.}
+    {\label{alg:make-monotone}
+    \item if $\big(f(x_1) - f(x_0) = 0\big)$
+      \indented{$Df(x_0) := Df(x_1) := D^2f(x_0) := D^2f(x_1) := 0$}
+      \indented{return $f$}
+    \item $Df(x_0) :=\ $median$\displaystyle \bigg(0,\ Df(x_0),\ 14 \frac{f(x_1) - f(x_0)}{x_1 - x_0} \bigg)$
+      \\ $Df(x_1) :=\ $median$\displaystyle \bigg(0,\ Df(x_1),\ 14 \frac{f(x_1) - f(x_0)}{x_1 - x_0}\bigg)$ \label{alg2-init-df}
+      \comment{The selection of values in Step \ref{alg2-init-df} for $Df(x_0)$ and $Df(x_1)$ is suggested by \citeN{ulrich1994positivity} and originally from \citeN{huynh1993accurate}.
+These rules quickly enforce upper and lower bounds on derivative values to ensure quintic monotonicity is obtainable.}
+      \comment{The next case that will be covered is in accordance with a simplified condition for quintic monotonicity that reduces to one of cubic positivity studied in \citeN{schmidt1988positivity}, where $\alpha$, $\beta$, $\gamma$, and $\delta$ are defined in terms of values and derivatives of $f$ at $x_0$ and $x_1$.
+Steps \ref{alg2-dx-conditions} and \ref{alg2-rescale-dx} ensure that the following bounds on $D^2f(x_0)$ are compatible.
+Step \ref{alg2-ddx0-upper} is the final preparatory step, ensuring that the conditions on $D^2f(x_1)$ are compatible.
+To conclude the simplified case, Step \ref{alg2-gamma} first enforces $\gamma \geq \delta$, then Steps \ref{alg2-alpha} and \ref{alg2-beta} enforce $\alpha > 0$ and $\beta \geq \alpha$ respectively.
+Combined, these three conditions derived from Equation (2.18) in \citeN{schmidt1988positivity} guarantee monotonicity of this polynomial piece.}
+    \item if $\big(Df(x_0) = 0$ or $Df(x_1) = 0\big)$
+    \item \indented{$w := x_0 - x_1$\\$v := f(x_0) - f(x_1)$}
+    \item \indented{if $\big(5 Df(x_0) + 4Df(x_1) > 20v / w)$} \label{alg2-dx-conditions}
+    \item \indented{\indented{$Df(x_0) := Df(x_0) \text{ max}\displaystyle \bigg(0, \frac{20v}{w \big(5 Df(x_0) + 4 Df(x_1)\big)}\bigg) $}}
+      \indented{\indented{$Df(x_1) := Df(x_1) \text{ max}\displaystyle \bigg(0, \frac{20v}{w \big(5 Df(x_0) + 4 Df(x_1)\big)}\bigg) $}} \label{alg2-rescale-dx}
+    \item \indented{$D^2f(x_0) :=\ $min$\displaystyle \bigg(D^2f(x_0), \frac{4 (2 Df(x_0)+Df(x_1))+20 v/w}{w} \bigg)$} \label{alg2-ddx0-upper}
+    \item  \indented{$D^2f(x_0) :=\ $max$\big(D^2f(x_0), 3 Df(x_0) / w \big)$} \label{alg2-gamma}
+    \item \indented{$D^2f(x_1) :=\ $min$\big(D^2f(x_1), {-4}Df(x_1) / w \big)$} \label{alg2-alpha}
+    \item  \indented{$D^2f(x_1) :=\ $max$\displaystyle \bigg(D^2f(x_1), \frac{3w D^2f(x_0) - 24 Df(x_0) - 32 Df(x_1) + 60v/w}{5w} \bigg)$} \label{alg2-beta}
+    \item
+      \indented{return $f$}
+      \comment{The following code considers the remaining case where $Df(x_0) \neq 0$ and $Df(x_1) \neq 0$.}
+    \item $\displaystyle A := Df(x_0)\frac{x_1 - x_0}{f(x_1) - f(x_0)}$
+      \\ $\displaystyle B := Df(x_1) \frac{x_1 - x_0}{f(x_1) - f(x_0)}$
+    \item if $\big($max$(A,B) > 6\big)$
+      \indented{$Df(x_0) := 6 Df(x_0)\ /\ $max$(A,B)$}
+      \indented{$Df(x_1) := 6 Df(x_1)\ /\ $max$(A,B)$}
+      \comment{This box bound ensures that $(A,B)$ remains within a viable region of monotonicity (satisfying Theorem 4, seen in Fig. 6 of \citeN{ulrich1994positivity}).}
+    \item $\eta := \big(D^2f(x_0), D^2f(x_1)\big)$
+      \\ $\displaystyle \eta_0 := \bigg({-\frac{\sqrt{A}}{4}}\big(7\sqrt{A} + 3\sqrt{B}\big),\ \frac{\sqrt{B}}{4}\big(3\sqrt{A} + 7\sqrt{B}\big)  \bigg)$
+    \item $D^2f(x_0),\ D^2f(x_1) := $ binary\_search$\big($is\_monotone$(x_0,x_1,f)$, $\eta,$ $\eta_0\big)$
+      \comment{This binary search identifies feasible values of $D^2f(x_0)$ and $D^2f(x_1)$ that are nearest to the current values.
+This search is valid because $\eta_0$ is guaranteed to produce a monotone function, which can be seen in Equation (23) of \citeN{ulrich1994positivity}.}
+    \item return $f$
+}
+%% ----------------------------------------------------------------------
+
+Notice that both Algorithms \ref{alg:check-monotone} and \ref{alg:make-monotone} have $\mathcal{O}(1)$ runtime  assuming a fixed level of relative precision.
+A constant number of operations are needed for verifying monotonicity (Algorithm \ref{alg:check-monotone}), while a constant set of operations and a single binary search are performed for enforcing monotonicity (Algorithm \ref{alg:make-monotone}).
+The search requires a fixed number of steps to achieve any predetermined relative precision.
+Also note that an efficient implementation of Algorithm \ref{alg:make-monotone} only needs to recalculate Steps \ref{alg1-abg} through \ref{alg1-last} of Algorithm \ref{alg:check-monotone} during the binary search.
+Next, an algorithm for constructing a monotone quintic interpolating spline is presented.
 
@@
 
-\subsection{Verifying Monotonicity of a Quintic Polynomial}
-\label{is_monotone}
-Let $f$ be a quintic polynomial over a closed interval $[x_0, x_1]$ $\subset \mathbb{R}$.
-Now $f$ is uniquely defined by the evaluation tuples $\big(x_0,$ $f(x_0),$ $f'(x_0),$ $f''(x_0)\big)$ and $\big(x_1,$ $f(x_1),$ $f'(x_1),$ $f''(x_1)\big).$ Assume without loss of generality that $f(x_0) < f(x_1),$ where the case of monotonic decreasing $f$ would consider the negated the function values.
-The following algorithm will determine whether or not $f$ is monotone increasing on the interval $[x_0, x_1].$
-
-\vspace{10pt}%
-\hrule%
-\vspace{3pt}%
-\noindent\textbf{\textit{Algorithm 1:}} \texttt{is\_monotone$\big (x_0$,$x_1$,$f \big)$}%
-\vspace{3pt}%
-\hrule%
-
-\begin{itemize}
-  \itemsep0pt
-  \parskip0pt
-
-\item[1:] \texttt{if $\big(f(x_0) = f(x_1)\big)$ return $\big( Df(x_0) = Df(x_1) = D^2f(x_0) = D^2f(x_1) = 0 \big)$}
-\item[2:] \texttt{if $\big(Df(x_0) < 0$ or $Df(x_1) < 0\big)$ return FALSE}
-  \begin{itemize}
-    \item[] \textit{This can be seen clearly from the fact that $f$ is analytic; there will exist some nonempty interval about $x_0$ or $x_1$ for which $Df$ is negative.}
-  \end{itemize}
-
-\item[2:] \texttt{if $\big(Df(x_0) = 0$ or $Df(x_1) = 0\big)$}
-\item[3:] $\quad w = x_0 - x_1$
-\item[4:] $\quad g(y) = \big({-4} Df(x_1) / w \big) < \big( ({-f}(x_1)/w - 24 y - 32 Df(x_1)) / 5w \big)$
-\item[5:] $\quad$ \texttt{line\_search}$(g, 0, Df(x_0))$
-\item[6:] $A = Df(x_0)\frac{x_1 - x_0}{f(x_1) - f(x_0)}$
-\item[7:] $B = Df(x_1) \frac{x_1 - x_0}{f(x_1) - f(x_0)}$
-  \begin{itemize}
-    \item[] \textit{The variables $A$ and $B$ correspond directly to the theoretical foundation for positive quartic polynomials established in \cite{ulrich1994positivity}, first defined after equation 18.}
-  \end{itemize}
-\item[8:] $\gamma_0 = 4 \frac{Df(x_0)}{Df(x_1)} (B/A)^{3/4}$
-\item[9:] $\gamma_1 = \frac{x_1 - x_0}{Df(x_1)} (B/A)^{3/4}$
-\item[4:] $\alpha_0 = 4 (B/A)^{1/4}$
-\item[5:] $\alpha_1 = -\frac{x_1 - x_0}{Df(x_1)} (B/A)^{1/4}$
-\item[6:] $\beta_0 = 30 - \frac{12 \big(Df(x_0) + Df(x_1)\big) (x_1 - x_0)}{\big(f(x_1) - f(x_0)\big) \sqrt{A}\sqrt{B}}$
-\item[7:] $\beta_1 = \frac{-3 (x_1 - x_0)^2}{2 \big(f(x_1) - f(x_0)\big) \sqrt{A} \sqrt{B}} $
-  \begin{itemize}
-    \item[] \textit{The $\gamma,$ $\alpha,$ and $\beta$ terms with subscripts $0$ and $1$ are algebraic reductions of the simplified conditions for satisfying Theorem 2 in \cite{ulrich1994positivity} (equation 16).
-These terms with subscripts $0$ and $1$ give the computation of $\alpha,$ $\beta,$ and $\gamma$ the form seen in lines $10$-$12$ below.}
-  \end{itemize}
-\item[10:] $\gamma = \gamma_0 + \gamma_1 D^2f(x_0)$
-\item[11:] $\alpha = \alpha_0 + \alpha_1 D^2f(x_1)$
-\item[12:] $\beta = \beta_0 + \beta_1 \big(D^2f(x_0) - D^2f(x_1)\big)$
-\item[13:] \texttt{if $(\beta \leq 6)$ then return $\alpha > - (\beta + 2) / 2$}
-\item[14:] \texttt{else return $\gamma > -2 \sqrt{\beta - 2}$ }
-
-\end{itemize}
-\hrule
-\vspace{10pt}
+\algorithm
+    {monotone\_spline$\big(\{k_1,\ldots,k_n\},\ f,\ s\big)$}
+    {where $\{k_1,\ldots,k_n\}$ is an increasing sequence of real numbers, $f$ is a piecewise order six polynomial defined over intervals $[k_i, k_{i+1}]$  for $i = 1, \ldots, n-1$, and integer $s \geq 1$.}
+    {\label{alg:monotone-spline}
+    \item create queue
+      \\ create changed$(1, \ldots, n) := $ FALSE
+      \\ create step\_size$(1, \ldots, n) := \{Df(k_i) / s$ for $i = 1, \ldots, n\}$\label{step:create}
+      \comment{The three variables defined in Step \ref{step:create} are used to ensure the eventual achievement of monotonicity.
+The \texttt{queue} is a standard first in first out (FIFO) queue with \texttt{enqueue} and \texttt{dequeue} operations.
+The array \texttt{changed} contains Booleans describing whether or not a partition point belongs to an interval that has been modified to enforce monotonicity.
+The \texttt{step\_size} array contains the real valued derivative step sizes to use in the search for a valid monotone spline.}
+    \item for $i := 1, \ldots, n-1$
+      \indented{enqueue $(k_i, k_{i+1})$}
+      \comment{Initially, all intervals must be checked for monotonicity. The following loop will continue until all intervals are verified as monotone without need for modification.}
+    \item while $($size$($queue$) > 0)$
+    \item \indented{$x_j,\ x_{j+1} := $ dequeue}
+    \item \indented{if $\big($not is\_monotone$(x_j,\ x_{j+1},\ f)\big)$}
+    \item \indented{\indented{if $\big($changed$(j)\big) \qquad \ Df(x_j) := Df(x_j) - $step\_size$(j)$}}
+      \indented{\indented{if $\big($changed$(j+1)\big) \quad Df(x_{j+1}) := Df(x_{j+1}) - $step\_size$(j+1)$}}
+      \indented{\indented{\comment{If a partition point belongs to an interval that has been previously modified, then the enforced monotonicity conditions on the second derivatives of adjacent intervals must contradict one another.
+In turn, the involved first derivatives are decreased towards zero by a predetermined step size.}}}
+    \item \indented{\indented{make\_monotone$(x_j,\ x_{j+1},\ f)$}}
+    \item \indented{\indented{changed$(j) :=$ TRUE}}
+      \indented{\indented{changed$(j+1) :=$ TRUE}} \label{step:changed}
+    \item \indented{\indented{if $\big(j > 1$ and $(x_{j-1}, x_j) \not \in$ queue$\big)$ enqueue $(x_{j-1}, x_j)$}}
+      \indented{\indented{if $\big((x_j, x_{j+1}) \not \in$ queue$\big)$ enqueue $(x_j, x_{j+1})$}}
+      \indented{\indented{if $\big(j+1 < n$ and $(x_{j+1}, x_{j+2}) \not \in$ queue$\big)$ enqueue $(x_{j+1}, x_{j+2})$}} \label{step:requeue}
+      \indented{\indented{\comment{Step \ref{step:changed} records the endpoints of the current interval as having been changed, while \ref{step:requeue} adds adjacent intervals to \texttt{queue} that may have inadvertently been made nonmonotone by the changes to the present interval.}}}
+}
 %% ----------------------------------------------------------------------
 
-%% ----------------------------------------------------------------------
-Given the same initial conditions there are special circumstances which allow for the usage of simpler monotonicity conditions.
-In this case, consider when the quintic function has either $f'(x_0) = 0$ or $f'(x_1) = 0.$ This reduces the problem of verifying monotonicity to one of cubics established by \cite{schmidt1988positivity}.
+It is mentioned in \citeN{ulrich1994positivity} that for sufficiently small $Df(k_i)$ and $Df(k_{i+1})$ the admissible solution interval of second derivative values becomes arbitrarily large.
+It can also be seen that decreasing the assigned derivative values towards zero will eventually cause Steps \ref{alg1-alpha} through \ref{alg1-gamma} of Algorithm \ref{alg:check-monotone} to all fail, resulting in Step \ref{alg1-simplified-true} returning \texttt{TRUE}.
 
-\vspace{10pt}%
-\hrule%
-\vspace{3pt}%
-\noindent\textbf{\textit{Algorithm 1b:}} \texttt{is\_monotone\_simplified}%
-\vspace{3pt}%
-\hrule%
-\begin{itemize}
-  \itemsep0pt
-  \parskip0pt
-
-\item[0:] $\alpha = 30 - \frac{(x_1 - x_0)\big( 14 f'(x_0) + 16 f'(x_1) - \big(f''(x_1) - f''(x_0) \big) (x_1 - x_0)\big)}{2\big(f(x_1) - f(x_0)\big)}$
-\item[1:] $\beta = 30 - \frac{(x_1 - x_0)\big( 2 f'(x_0) + 24 f'(x_1) - \big(f''(x_0) + 3 f''(x_1) \big) (x_1 - x_0)\big)}{2\big(f(x_1) - f(x_0)\big)}$
-\item[2:] $\gamma = \frac{(x_1 - x_0)\big( 7 f'(x_0) + f''(x_0) (x_1 - x_0) \big)}{f(x_1) - f(x_0)}$
-\item[3:] $\delta = \frac{f'(x_0) (x_1 - x_0)}{f(x_1) - f(x_0)}$
-
-  \begin{itemize}
-    \item[] \textit{The variables above are algebraic expansions of the coefficients for the cubic derivative function in \cite{schmidt1988positivity}.}
-  \end{itemize}
-
-\item[4:] \texttt{if $\big($min$(\alpha, \delta) < 0\big)$ return FALSE}
-\item[5:] \texttt{else if $\big(\beta < \alpha - 2 \sqrt{\alpha \delta}\big)$ return FALSE}
-\item[6:] \texttt{else if $\big(\gamma < \delta - 2 \sqrt{\alpha \delta}\big)$ return FALSE}
-\item[7:] \texttt{else return TRUE}
-
-\end{itemize}
-\hrule
-\vspace{10pt}
-%% ----------------------------------------------------------------------
-
-Next the modification of a quintic spline to enforce monotonicity will be discussed.
-
-%% ----------------------------------------------------------------------
-\subsection{Enforcing Monotonicity of a Quintic Polynomial}
-\label{make_monotone}
-
-\vspace{10pt}%
-\hrule%
-\vspace{3pt}%
-\noindent\textbf{\textit{Algorithm 2a:}} \texttt{make\_monotone}%
-\vspace{3pt}%
-\hrule%
-
-\begin{itemize}
-  \itemsep0pt
-  \parskip0pt
-
-\item[0:] \texttt{if }$\big(f(x_1) - f(x_0) = 0\big)$ \texttt{ return } $f'(x_0) = f'(x_1) = f''(x_0) = f''(x_1) = 0$
-\item[1:] $f'(x_0) = \ $\texttt{median}$\big(0, f'(x_0), 14 \frac{f(x_1) - f(x_0)}{x_1 - x_0} \big)$
-\item[2:] $f'(x_1) = \ $\texttt{median}$\big(0, f'(x_1), 14 \frac{f(x_1) - f(x_0)}{x_1 - x_0}\big)$
-  \begin{itemize}
-  \item[] \textit{This selection of values for $f'(x_0)$ and $f'(x_1)$ is suggested by \cite{ulrich1994positivity} (originally from \cite{huynh1993accurate}), and quickly enforces upper and lower bounds on derivative values to ensure quintic monotonicity is obtainable.}
-  \end{itemize}
-\item[3:] $A = f'(x_0)\frac{x_1 - x_0}{f(x_1) - f(x_0)}$
-\item[4:] $B = f'(x_1) \frac{x_1 - x_0}{f(x_1) - f(x_0)}$
-\item[5:] \texttt{if }$AB \leq 0$\texttt{ return make\_monotone\_simplified}
-\item[6:] \texttt{if $\big($max}$(A,B) > 6\big)$\\$f'(x_0) = 6 f'(x_0)\  /$\texttt{ max}$(A,B)$\\$f'(x_1) = 6 f'(x_1)\  /$\texttt{ max}$(A,B)$
-
-  \begin{itemize}
-  \item[] \textit{This box bound ensures that $(A,B)$ remains within a viable region of monotonicity (satisfying Theorem 4, seen in Fig. 6 of \cite{ulrich1994positivity}).}
-  \end{itemize}
-
-\item[7:] $f''(x_0) = - \sqrt{A} \big( 7 \sqrt{A} + 3 \sqrt{B} \big) \frac{f(x_1) - f(x_0)}{(x_1 - x_0)^2}$\\$f''(x_1) = \sqrt{B} \big( 3 \sqrt{A} + 7 \sqrt{B} \big) \frac{f(x_1) - f(x_0)}{(x_1 - x_0)^2}$
-
-  \begin{itemize}
-    \item[] \textit{This selection of values of $f''$ is guaranteed to satisfy Theorem 4 from \cite{ulrich1994positivity} and is chosen because it is (reasonably) the average of the two endpoints of the interval of monotonicity for second derivative values.}
-  \end{itemize}
-
-\item[8:] $\eta = \big(f''(x_0), f''(x_1)\big)$\\$\eta_0 = \big(f''(x_0), f''(x_1)\big)$\\$f''(x_0), f''(x_1) = $\texttt{ line\_search$\big($is\_monotone, $\eta,$ $\eta_0\big)$}
-\end{itemize}
-\hrule
-\vspace{10pt}
-%% ----------------------------------------------------------------------
-
-Similar to the simplified check for monotonicity, when the derivative value at one endpoint of the interval is $0$, a simplified set of steps can be taken to enforce monotonicity.
-
-%% ----------------------------------------------------------------------
-\vspace{10pt}%
-\hrule%
-\vspace{3pt}%
-\noindent\textbf{\textit{Algorithm 2b:}} \texttt{make\_monotone\_simplified}%
-\vspace{3pt}%
-\hrule%
-
-\begin{itemize}
-  \itemsep0pt
-  \parskip0pt
-
-\item[0:] $f''(x_0) = $\texttt{ max}$\bigg(f''(x_0), \frac{-6 f'(x_0)}{x_1 - x_0}\bigg)$
-
-  \begin{itemize}
-    \item[] \textit{Considering the $\alpha,$ $\gamma,$ $\beta,$ and $\delta$ defined in \cite{schmidt1988positivity}, this first step enforces $\gamma > \delta.$ It is already guaranteed that $\delta = \frac{f'(x_0)(x_1 - x_0)}{f(x_1) - f(x_0)} > 0.$ Only two conditions remain to guarantee monotonicity.}
-  \end{itemize}
-
-\item[1:] $f''(x_1) = $\texttt{ max}$\bigg( f''(x_1), f''(x_0) + \frac{14 f'(x_0) + 16 f'(x_1) + 60(f(x_1) - f(x_0)) / (x_1 - x_0)}{x_1 - x_0} \bigg)$
-
-  \begin{itemize}
-    \item[] \textit{Now it is guaranteed that $\alpha \geq 0.$}
-  \end{itemize}
-
-\item[2:] $f''(x_1) = $\texttt{ max}$\bigg( f''(x_1), 6 f'(x_0) - 4 f'(x_1) - f''(x_0) \bigg)$
-
-  \begin{itemize}
-    \item[] \textit{Lastly, this guarantees that $\beta \geq \alpha.$ All conditions are met to satisfy proposition 2 of \cite{schmidt1988positivity} and ensure monotonicity.}
-  \end{itemize}
-\end{itemize}
-\hrule
-\vspace{10pt}
-%% ----------------------------------------------------------------------
-Notice that all above algorithms (assuming a fixed level of precision is desired) have $\mathcal{O}(1)$ runtime.
-Only a finite number of operations are needed for monotonicity verification.
-A line search is performed for enforcement, however that search requires a fixed number of steps to achieve any predetermined relative precision on the line.
-
-%% ----------------------------------------------------------------------
-\subsection{Constructing a Piecewise Quintic Monotone Spline}
-\label{monotone_spline}
-
-Finally, the construction of a full piecewise quintic spline is outlined using the above algorithms.
-Let $f: \mathbb{R} \rightarrow \mathbb{R}$ be a function in $\mathcal{C}^2.$ Proceed given evaluation tuples $\big(x_i,$ $f(x_i),$ $f'(x_i),$ $f''(x_i)\big)$ for $i = 0,\ldots,N$ such that $x_i < x_{i+1}$ and (without loss of generality) $f(x_i) \leq f(x_{i+1})$ for $i = 1,$ $\ldots,$ $N-1$.
-
-\vspace{10pt}%
-\hrule%
-\vspace{3pt}%
-\noindent\textbf{\textit{Algorithm 3:}} \texttt{monotone\_spline}%
-\vspace{3pt}%
-\hrule%
-
-\begin{itemize}
-  \itemsep0pt
-  \parskip0pt
-
-\item[0:] \texttt{for }$i=0,\ldots,N-1$
-\item[1:] $\quad$\texttt{if $\big($not is\_monotone$(i,i+1)\big)$  make\_monotone$(i,i+1)$}
-
-  \begin{itemize}
-    \item[] \textit{In the shorthand notation above, $i$ and $i+1$ refer to the associated tuples of the form $(x_i,$ $f(x_i),$ $f'(x_i),$ $f''(x_i)).$ This notation will be used through the remainder of this algorithm.}
-  \end{itemize}
-
-\item[2:] $\quad$\texttt{for $j=i-1,\ldots,0$}
-\item[3:] $\quad\quad$\texttt{if $\big($not is\_monotone$(j,j+1)\big)$  make\_monotone$(j,j+1)$}
-\item[4:] $\quad\quad$\texttt{else break}
-
-  \begin{itemize}
-    \item[] \textit{The above `for' loop will be referred to as the cascade fix.
-If the adjustment of second derivative values causes the previous interval to become nonmonotone, then the left-hand second derivative value must be updated.
-This may (abnormally) require adjustments across all previously visited intervals.}
-  \end{itemize}
-
-\item[5:] $\quad$\texttt{while $\big($not is\_monotone$(i,i+1)\big)$}
-\item[6:] $\quad\quad f'(x_i) = f'(x_i) / k$
-\item[7:] $\quad\quad f'(x_{i+1}) = f'(x_{i+1}) / k$
-\item[8:] $\quad\quad$\texttt{make\_monotone$(i,i+1)$}
-
-  \begin{itemize}
-    \item[] \textit{In the case that the two corrections to neighboring intervals contradict, the first derivative values of the active interval are decreased to enlarge the overlap of regions I and II (from \cite{ulrich1994positivity}) of the two intervals.}
-  \end{itemize}
-
-\item[9:] $\quad\quad$\texttt{for $j=i-1,\ldots,0$}
-\item[10:] $\quad\quad\quad$\texttt{if $\big($not is\_monotone$(j,j+1)\big)$  make\_monotone$(j,j+1)$}
-\item[11:] $\quad\quad\quad$\texttt{else break}
-
-  \begin{itemize}
-    \item[] \textit{Finally an additional `cascade fix' is performed to ensure that all previous intervals are still monotone after shrinking the derivative values of the current interval.}
-  \end{itemize}
-
-\end{itemize}
-\hrule
-\vspace{10pt}
-%% ----------------------------------------------------------------------
-
-It is mentioned in \cite{ulrich1994positivity} that for sufficiently small $f'(x_i)$ and $f'(x_{i+1})$ the admissible solution interval of second derivative values becomes arbitrarily large.
-It can also be seen that decreasing the assigned derivative on right-hand side of an interval always allows the achievement of monotonicity because shrinking $f'(x_1)$ results in $\gamma$ growing faster than $\sqrt{\beta}$ for \textit{algorithm 1a}, for \textit{algorithm 1b} $\beta$ and $\gamma$ will grow faster $\alpha$ and $\delta$ respectively.
-In application, one needs only to pick some $k > 1$ to ensure successful termination.
-
-Given the potential for successive cascade fixes, the runtime of \textit{algorithm 3} is $\mathcal{O}(N^2)$.
-In practice this has been observed to be incredibly unlikely and difficult to produce.
-While single cascade fixes can occur from hand crafted examples, the generously rounded values selected to ensure monotonicity frequently prevent successive cascade fixes.
-
-%% - background is provided by these papers
-
-%% - Table of relevant terms and what they mean
-
-%% \subsection{Algorithm: Monotone Quintic Interpolation}
-%% - correctness of algorithm
-%% - runtime of algorithm
-%% - weakness of algorithm
+Given the potential for successive monotonicity fixes across all intervals, the worst case runtime of Algorithm \ref{alg:monotone-spline} is $\mathcal{O}(s n)$ for $n$ partition points and step size divisor $s$.
+In practice this worst case has been observed to be incredibly unlikely and difficult to produce.
 
@@
 
-- show a Figure of a simple quintic interpolant and its derivatives
+Given the increased order of approximation, it is naturally expected that some accuracy benefit could be gained by using a quintic spline over a cubic spline.
+Experiments 1 and 2 below test some accuracy differences between monotone cubic and quintic splines.
+Experiment 3 analyzes the number of operations performed by Algorithm \ref{alg:monotone-spline} with increasingly large samples of monotone data.
 
@@
 
-For this experiment, the function $\sin(x) + x$ is considered over the
-interval $[0,(5/2)\pi]$.
+For this experiment, the function $\sin(x) + x$ is considered over the interval $[0,(5/2)\pi]$ as seen in Figure \ref{fig:cubic_quintic_sin}.
+Given an increasing number of points, the maximum error of cubic and quintic approximations is shown in Figure \ref{fig:experiment_1}.
+The quintic approximation consistently has a maximum error that is roughly half that of the cubic approximation, given the same number of points.
+There is a (subtle) increase in the error gap between the two approximations as more points are added.
 
-\begin{figure}[htb]
+\begin{figure}
   \centering
   \includegraphics[width=.7\textwidth]{cubic-quintic-sin}
-  \caption{Depicted above are the monotone cubic and quintic spline
-    interpolant of the function $\sin(x) + x$ over the nodes $[0,
-      (5/2) \pi]$.  Notice that the maximum error of the cubic
-    interpolant is larger, because it only captures first derivative
-    information at the nodes. The quintic interpolant captures both
-    first and second derivative information at the nodes.
-  }\label{fig:cubic-quintic-sin}
+  \caption{Depicted above are the monotone cubic and quintic spline interpolants of the function $\sin(x) + x$ over the points $[0, (5/2) \pi]$.
+Notice that the maximum error of the cubic interpolant is larger, because it only captures first derivative information at the points.
+The quintic interpolant captures both first and second derivative information at the points.}
+  \label{fig:cubic_quintic_sin}
 \end{figure}
 
-- Pick a trig function, approximate it with cubic and quintic
+\begin{figure}
+  \centering
+  \includegraphics[width=.7\textwidth]{experiment_1_errors}
+  \caption{The error in the quintic and cubic approximations of $\sin(x) + x$ with increasing number of points.
+Both axes are log scaled.
+The quintic approximation generally provides an approximation with slightly more than half of the maximum error of the cubic approximation, while the relative gap between grows with increasing number of points.
+For larger absolute errors, this means the accuracy of the cubic is achieved with about ten fewer samples from the quintic.
+However, for smaller absolute errors the quintic spline requires hundreds fewer samples to achieve the same accuracy as the cubic.}
+  \label{fig:experiment_1}
+\end{figure}
 
-- Show a Figure, and a Table of the maximum error and distribution of
-  error across the segment
+\subsection{Random Number Generation}
 
-\subsection{Approximating a Cumulative Distribution Function}
+Consider the cumulative distribution function (CDF) defined by a mixture of $3$ Gaussian distributions and visualized in Figure \ref{fig:experiment_2_distribution}.
+The distributions have weights $(.3, .6, .1)$, means $(.2, .45, .85)$ and standard deviations $(.05, .08, .03)$.
+$200$ random samples are generated from this distribution with both cubic and quintic approximations from $4$ points, and over $100$ trials the resulting spread of empirical distribution functions (EDFs) are visualized in Figure \ref{fig:experiment_2_results}.
 
-- generate large sets of random monotone data, show Table of number
-  of metrics growing with increasing number of nodes
+While an increased number of points could be used to decrease the error in either approximation, the purpose of this experiment is to demonstrate the effect approximation error has on random number generation.
+The errors in the CDFs are magnified by the variance involved in randomly sampling from a distribution.
+As a result, while the cubic and quintic spline approximations have similar maximum error, the cubic distribution is significantly distorted around the first and second modes.
 
-- show visual of the \textit{error}
+\begin{figure}
+  \centering
+  \includegraphics[width=.8\textwidth]{experiment_2_distribution}
+  \caption{The cumulative distribution function (top) and probability density function (bottom) for the Gaussian mixture distribution in experiment 2.
+The distributions have weights $(.3, .6, .1)$, means $(.2, .45, .85)$ and standard deviations $(.05, .08, .03)$.
+Random samples are generated from this mixture distribution by evaluating the inverse of the CDF at uniformly random numbers in the range $[0,1]$.}
+  \label{fig:experiment_2_distribution}
+\end{figure}
+
+\begin{figure}
+  \centering
+  \includegraphics[width=.4\textwidth]{experiment_2_cubic_dist}
+  \includegraphics[width=.4\textwidth]{experiment_2_quintic_dist}
+  \caption{Both cubic and quintic splines are used to approximate the CDF of a Gaussian mixture distribution with three components.
+$4$ CDF points are used to build the approximation, $200$ random samples are generated with the approximate CDFs over $100$ trials.
+The resulting cloud of empirical distribution function (EDF) values is seen above.
+The maximum approximation error in the CDFs are similar, $.08$ for the cubic and $.05$ for the quintic.
+However, the maximum EDf differences observed are doubled by the variance of random samples, $.15$ for the cubic and $.1$ for the quintic.}
+  \label{fig:experiment_2_results}
+\end{figure}
+
+%% - generate large sets of random monotone data, show Table of number
+%%   of metrics growing with increasing number of points
+
+%% - show visual of the \textit{error}
 
@@
 
-- generate large sets of random monotone data, show Table of number
-  of metrics growing with increasing number of nodes (how many binary
-  searches, how many corrections were needed, steps were taken)
+This experiment studies the number of times the algorithms \texttt{is\_monotone} and \texttt{make\_monotone} are executed for increasingly large sequences of random monotone data.
+Statistics on the expected number of interval corrections are presented.
+
+\texttt{NEED STATISTICS TABLE}
 
@@
 
-- this algorithm has its quirks, but it works
+This is an improvement over the cubic, although the complexity of the algorithm is increased as well.
+In the present state, the algorithm for enforcing monotonicity on a spline is not as trivially parallel as the cubic algorithm, however it can still be parallelized.
 
-- this is an improvement over the cubic, although the complexity of
-  the algorithm is greatly increased as well
+Knowing this algorithm is slightly more complex than the cubic variant, it appears worth the computational effort because of gains in accuracy or a decreased necessary number of nodes.
 
-\section{Future Work}
+\section{Conclusion and Future Work}
 
-- a method for finding regions of monotonicity that is less ad-hoc
+This paper proposed and tested an algorithm for constructing monotone quintic interpolating splines.
+Experiments demonstrate an improvement in approximation accuracy over monotone cubic interpolating splines, as expected based on theory.
 
-- constructing monotone interpolants for higher order approximations
-
-- using improved distribution estimates to improve distribution modeling
-
-\section{Conclusion}
-
-- this paper proposes and tests an algorithm for constructing quintic
-  monotone interpolants
-
-- the results show an improvement over the cubic usage
+There are still open avenues of research going forward, such as an alternative sufficient condition for enforcing monotonicity or increased order monotone approximations.
+If the monotonicity conditions can be generalized to any order or made linear, the search for a monotone interpolating spline could potentially be formulated as a convex optimization problem.
+Finally, this work could be used to improve cumulative distribution function (CDF) estimates as well as predictive models that use CDFs.
 
@@
 
-\textbf{\uppercase{THOMAS C.
-H.
-LUX}} is a Ph.D.
-candidate in Computer Science at Virginia Polytechnic Institute and State University.
-His research interests include Computational Science, Approximation Theory, Optimization, and Artificial Intelligence.
-His email address is \email{tchlux@vt.edu}.
+\textbf{\uppercase{THOMAS C. H. LUX}} is a Ph.D. candidate in computer
+science at Virginia Polytechnic Institute and State University
+(VPI\&SU) studying under Dr. Layne Watson. His research interests
+include computational science, approximation theory, optimization, and
+artificial intelligence. His email address is \email{tchlux@vt.edu}.
 
-\newpage
+\textbf{\uppercase{Layne T. Watson}} (Ph.D., Michigan, 1974) has
+interests in numerical analysis, mathematical programming,
+bioinformatics, and data science.  He has been involved with the
+organization of HPCS since 2000.
 
-\appendix
+\textbf{\uppercase{Tyler H. Chang}} is a Ph.D. candidate at VPI\&SU
+studying computer science under Dr. Layne Watson.
 
-\section{Author Checklist}
-We strive for a consistent appearance in all papers published in the proceedings.
+\textbf{\uppercase{Li Xu}} is a Ph.D. candidate at VPI\&SU studying
+statistics under Dr. Yili Hong.
+
+\textbf{\uppercase{Yueyao Wang}} is a Ph.D. candidate at VPI\&SU
+studying statistics under Dr. Yili Hong.
+
+\textbf{\uppercase{Yili Hong}} (Ph.D., Iowa State, 2009) has interests in
+engineering statistics, statistical modeling, and data analysis.
+
+%% \newpage
+
+%% \appendix
+
+%% \section{Author Checklist}
+%% We strive for a consistent appearance in all papers published in the proceedings.
 If you used the template and styles within this author’s kit, then almost all of the requirements in this checklist will be automatically satisfied, and there is very little to check.
 
-Please \textbf{print a hardcopy of your paper}, and go over your printed paper to make sure it adheres to the following requirements. \textit{Thank you!}
+%% Please \textbf{print a hardcopy of your paper}, and go over your printed paper to make sure it adheres to the following requirements. \textit{Thank you!}
 
-\begin{enumerate}
-	\item Abstract
-  \begin{enumerate}
-	  \item 150 or fewer words.
-	  \item Provide 3-5 keywords (mandatory). This set of keywords will identify your paper in indices and databases.
-	\end{enumerate}
-	\item Paper Length
-  \begin{enumerate}
-	  \item At least 5, but no more than 12 pages.
-	  \item Page size is letter size (8.5’’ x 11’’, or 216 mm x 279 mm).
-	\end{enumerate}
-	\item All text is in 11-Point Times New Roman except title, header and footer (default in this template).
-	\item Paper title is in 12-Point Times New Roman \textbf{BOLDFACE ALL CAPS} (default in this template).
-	\item The paper has been spellchecked using U.S. English. 
-	\item Spacing and Margins
-  \begin{enumerate}
-	  \item Single spaced.
-	  \item Left and right margins are each 1 inch (default in this template).
-	  \item Top and bottom margins are according to the template.
-	  \item Title starts 1.25 inches from the top of the page (default in this template).
-	\end{enumerate}
-	\item Section Headings
-  \begin{enumerate}
-	  \item Left justified and set in \textbf{BOLDFACE ALL CAPS} (default in this template).
-	  \item Numbered, except for the abstract, acknowledgments, references and author biographies.
-	  \item Subsection headings are set in \textbf{Boldface Headline Style} (default in this template).
-	\end{enumerate}
-	\item No footnotes or page numbers.
-	\item The copyright notice on the first page contains the name of the symposium or a track where you want to submit your paper, and the running head on subsequent pages is the surnames of all authors.
-	\item Multiple authors are formatted correctly.
-	\item Equations are centered and any equation numbers are in parentheses and right-justified (default).
-	\item Figures and Tables
-  \begin{enumerate}
-	  \item All text in figures and tables is readable.
-	  \item Table captions appear above the table.
-	  \item Figure captions appear below the figure.
-	\end{enumerate}
-	\item Citations and References
-  \begin{enumerate}
-	  \item Citations are by author and year, and are enclosed in parentheses, not brackets (default \BibTeX). 
-	  \item References are in the hangref style, and are listed alphabetically by the last names(s) of the author(s) (also default \BibTeX\ setting in this template). 
-	\end{enumerate}
-	\item Author biographies are one paragraph per author.
-	\item Hyperlinks
-  \begin{enumerate}
-	  \item Be sure that hyperlinks will probably work in the future as well.
-	  \item Live hyperlinks are blue. Nonlive hyperlinks are black (default in this template).
-	\end{enumerate}
-	\item All fonts must be embedded in the resulting \texttt{.pdf}.
+%% \begin{enumerate}
+%% 	\item Abstract
+%%   \begin{enumerate}
+%% 	  \item 150 or fewer words.
+%% 	  \item Provide 3-5 keywords (mandatory). This set of keywords will identify your paper in indices and databases.
+%% 	\end{enumerate}
+%% 	\item Paper Length
+%%   \begin{enumerate}
+%% 	  \item At least 5, but no more than 12 pages.
+%% 	  \item Page size is letter size (8.5’’ x 11’’, or 216 mm x 279 mm).
+%% 	\end{enumerate}
+%% 	\item All text is in 11-Point Times New Roman except title, header and footer (default in this template).
+%% 	\item Paper title is in 12-Point Times New Roman \textbf{BOLDFACE ALL CAPS} (default in this template).
+%% 	\item The paper has been spellchecked using U.S. English. 
+%% 	\item Spacing and Margins
+%%   \begin{enumerate}
+%% 	  \item Single spaced.
+%% 	  \item Left and right margins are each 1 inch (default in this template).
+%% 	  \item Top and bottom margins are according to the template.
+%% 	  \item Title starts 1.25 inches from the top of the page (default in this template).
+%% 	\end{enumerate}
+%% 	\item Section Headings
+%%   \begin{enumerate}
+%% 	  \item Left justified and set in \textbf{BOLDFACE ALL CAPS} (default in this template).
+%% 	  \item Numbered, except for the abstract, acknowledgments, references and author biographies.
+%% 	  \item Subsection headings are set in \textbf{Boldface Headline Style} (default in this template).
+%% 	\end{enumerate}
+%% 	\item No footnotes or page numbers.
+%% 	\item The copyright notice on the first page contains the name of the symposium or a track where you want to submit your paper, and the running head on subsequent pages is the surnames of all authors.
+%% 	\item Multiple authors are formatted correctly.
+%% 	\item Equations are centered and any equation numbers are in parentheses and right-justified (default).
+%% 	\item Figures and Tables
+%%   \begin{enumerate}
+%% 	  \item All text in figures and tables is readable.
+%% 	  \item Table captions appear above the table.
+%% 	  \item Figure captions appear below the figure.
+%% 	\end{enumerate}
+%% 	\item Citations and References
+%%   \begin{enumerate}
+%% 	  \item Citations are by author and year, and are enclosed in parentheses, not brackets (default \BibTeX). 
+%% 	  \item References are in the hangref style, and are listed alphabetically by the last names(s) of the author(s) (also default \BibTeX\ setting in this template). 
+%% 	\end{enumerate}
+%% 	\item Author biographies are one paragraph per author.
+%% 	\item Hyperlinks
+%%   \begin{enumerate}
+%% 	  \item Be sure that hyperlinks will probably work in the future as well.
+%% 	  \item Live hyperlinks are blue. Nonlive hyperlinks are black (default in this template).
+%% 	\end{enumerate}
+%% 	\item All fonts must be embedded in the resulting \texttt{.pdf}.
 If you use \texttt{pdflatex}, this should already be true--unless you included figures in the \texttt{.eps} or \texttt{.pdf} format which may introduce additional font dependencies.
@@
 You can use GhostScript to force embed the fonts, like so: {\texttt{gs -dNOPAUSE -dBATCH -sDEVICE=pdfwrite -dEmbedAllFonts=true -sOutputFile=out.pdf -f in.pdf}} (in Windows, use \texttt{gswin32} or \texttt{gswin32c} to invoke GhostScript).
-\end{enumerate}
+%% \end{enumerate}
 
-After verifying that your paper meets these requirements, please go to the final submission page at conference website and submit your paper.
+%% After verifying that your paper meets these requirements, please go to the final submission page at conference website and submit your paper.
 Be sure to complete the transfer of copyright form and upload the \texttt{.pdf} receipt.
@@
 
-\section{Most Observed Mistakes}
+%% \section{Most Observed Mistakes}
 
-The following list comprises \textbf{the most common sources of error} that had to be corrected by previous editors.
+%% The following list comprises \textbf{the most common sources of error} that had to be corrected by previous editors.
 Please make sure to go through the following list and check that your paper is formatted correctly:
-\begin{enumerate}
-	\item	The paper is less than 5 or more than 12 pages long.
-	\item	Paper title and section titles are in \textbf{BOLD ALL CAPS}, subsections are \textbf{Bold and Capitalize the First Letter of Important Words}. Please use the templates.
-	\item	Paper is in A4 format, not letter format. Please use the required margins.
-	\item	The copyright notice is incorrect.
-	\item	The running heads are incorrect. Do not forget that the LastNameLastAuthor is preceded by ``, and ''.
-	\item	The citation format is incorrect. Use \BibTeX\ for citations and do not change the format used by the template.
-	\item	The biographies are missing. Do not forget the ``author biographies'' section.
-	\item	Figures or tables are not referenced in the text or have the incorrect caption format.
-	\item	The author section after the title is not formatted correctly, the number of organizations does not define the number of blocks, or the number of blocks does not define the layout.
-	\item In the heading on the title page, country names are in all capitals.
-	\item	Paragraphs are not indented.
-	\item Some fonts in the resulting \texttt{.pdf} are not embedded.
-\end{enumerate}
+%% \begin{enumerate}
+%% 	\item	The paper is less than 5 or more than 12 pages long.
+%% 	\item	Paper title and section titles are in \textbf{BOLD ALL CAPS}, subsections are \textbf{Bold and Capitalize the First Letter of Important Words}. Please use the templates.
+%% 	\item	Paper is in A4 format, not letter format. Please use the required margins.
+%% 	\item	The copyright notice is incorrect.
+%% 	\item	The running heads are incorrect. Do not forget that the LastNameLastAuthor is preceded by ``, and ''.
+%% 	\item	The citation format is incorrect. Use \BibTeX\ for citations and do not change the format used by the template.
+%% 	\item	The biographies are missing. Do not forget the ``author biographies'' section.
+%% 	\item	Figures or tables are not referenced in the text or have the incorrect caption format.
+%% 	\item	The author section after the title is not formatted correctly, the number of organizations does not define the number of blocks, or the number of blocks does not define the layout.
+%% 	\item In the heading on the title page, country names are in all capitals.
+%% 	\item	Paragraphs are not indented.
+%% 	\item Some fonts in the resulting \texttt{.pdf} are not embedded.
+%% \end{enumerate}
 

### 2019-12-15 Paper_[2019-11]_HPC/Code/__init__.py

@@ -1,7 +1,0 @@
-'''
-This is the code package for the HPC paper.
-
-All of the code necesasry to run 3 expirements involving monotone
-splines exists here.
-
-'''

### 2019-12-15 Paper_[2019-11]_HPC/experiments/experiment_3.py

@@ -1,0 +1,71 @@
+import numpy as np
+
+# Include the "code" directory into the path and import code custom to this project.
+import sys, os
+sys.path += [os.path.abspath("code")]
+from monotone import monotone_quintic_spline, monotone_cubic_spline
+import monotone
+from fraction import Fraction
+from data import Data
+from plot import Plot
+sys.path.pop(-1)
+
+
+SEED = 0
+POINTS = 10
+
+results_file = "e3_results.pkl"
+results = Data(names=["num points", "seed", "total checks", "total fixes", "search steps", "checks dict", "fixes dict"],
+               types=[int,          int,    int,            int,           int,            dict,          dict])
+
+# Try to load existing results if they exist.
+try: results.load(results_file)
+except: pass
+
+# Run the experiment and collect the data.
+for points in range(2, 101):
+    print("points: ",points)
+    for seed in range(100):
+        print("  seed: ",seed, end="\r")
+        # Skip already checked values.
+        if ((points,seed) in results["num points","seed"]): continue
+        # Generate random data to test the monotonic fit function.
+        np.random.seed(seed)
+        x = np.linspace(0, 1, points)
+        y = sorted(np.random.normal(size=(points,)))
+        x -= min(x); x /= max(x)
+        y -= min(y); y /= max(y); y -= max(y)
+        # Convert these arrays to exact arithmetic.
+        x = list(map(Fraction, x))
+        y = list(map(Fraction, y))
+        interval = [float(min(x)), float(max(x))]
+        # Compute the monotone quintic spline.
+        stdout = sys.stdout
+        # Block all print statements from the function.
+        with open(os.devnull, "w") as devnull:
+            sys.stdout = devnull
+            monotone.IS_MONOTONE_CALLS = 0
+            f, checks, changes = monotone_quintic_spline(x,y=y, verbose=True)
+        sys.stdout = stdout
+        # Store the results.
+        results.append(
+            [points, seed, sum(checks.values()), sum(changes.values()),
+             monotone.IS_MONOTONE_CALLS, checks, changes]
+        )
+    # Save the intermediate results.
+    results.sort()
+    results.save(results_file)
+
+print(" "*70)
+
+print(results[results["num points"] > 10])
+
+d = results
+d.stack(d.names[1:])
+print(d)
+
+p = Plot()
+p.add("Total Checks", d["num points"], list(map(np.mean, d["total checks"])))
[…]

### 2019-12-16 Paper_[2019-11]_HPC/paper/paper.tex
Preparing for final submission to HPC.

@@
 %****************************************************************************
+
+%% Use a package to allow for multi-row cells in a table.
+\usepackage{multirow}
 
@@
 213 Hutcheson Hall, Blacksburg, VA, USA\\
-yilihong@vt.edu\\
+\{lix1992,yueyao94,yilihong\}@vt.edu\\
 }
@@
     {is\_monotone$\big(x_0, x_1, f \big)$}
-    {where $x_0$, $x_1 \in \mathbb{R}$, $x_0 < x_1$, and $f$ is an order six polynomial. Returns \texttt{TRUE} if $f$ is monotone increasing on $[x_0,x_1]$.}
+    {where $x_0$, $x_1 \in \mathbb{R}$, $x_0 < x_1$, and $f$ is an order six polynomial defined by $f(x_0)$, $Df(x_0)$, $D^2f(x_0)$, $f(x_1)$,
+  $Df(x_1)$, $D^2f(x_1)$. Returns \texttt{TRUE} if $f$ is monotone increasing on $[x_0,x_1]$.}
     {\label{alg:check-monotone}
@@
     \indented{return FALSE}
+    end if
   \item if $\big(Df(x_0) < 0$ or $Df(x_1) < 0\big)$
     \indented{return FALSE}
+    end if
     \comment{The necessity of these first two steps follows directly from the fact that $f$ is $C^2$.
-The next case that will be covered is in accordance with a simplified condition for quintic monotonicity that reduces to one of cubic positivity studied in \citeN{schmidt1988positivity}, where $\alpha$, $\beta$, $\gamma$, and $\delta$ are defined in terms of values and derivatives of $f$ at $x_0$ and $x_1$.
+The next case is in accordance with a simplified condition for quintic monotonicity that reduces to one of cubic positivity studied in \citeN{schmidt1988positivity}, where $\alpha$, $\beta$, $\gamma$, and $\delta$ are defined in terms of values and derivatives of $f$ at $x_0$ and $x_1$.
 Step \ref{alg1-alpha} checks for the necessary condition that $\alpha \ge 0$, Step \ref{alg1-beta} checks $\beta \geq \alpha$, and Step \ref{alg1-gamma} checks $\gamma \geq \delta$, all from \citeN{schmidt1988positivity}.
@@
   \item \indented{return TRUE} \label{alg1-simplified-true}
+    end if
     \comment{The following code considers the remaining case where $Df(x_0) \neq 0$ and $Df(x_1) \neq 0$.}
@@
     \\ $\displaystyle \beta_1 := \frac{-3 (x_1 - x_0)^2}{2 \big(f(x_1) - f(x_0)\big) \sqrt{A} \sqrt{B}} $ \label{alg1-linearized-abg}
-    \comment{The $\gamma,$ $\alpha,$ and $\beta$ terms with subscripts $0$ and $1$ are algebraic reductions of the simplified conditions for satisfying Theorem 2 in Equation (16) of \citeN{ulrich1994positivity}.
-These terms with subscripts $0$ and $1$ make the computation of $\alpha,$ $\beta,$ and $\gamma$ functions of the second derivative terms, as seen in Step \ref{alg1-abg} below.}
+    \comment{The $\gamma$, $\alpha$, and $\beta$ terms with subscripts $0$ and $1$ are algebraic reductions of the simplified conditions for satisfying Theorem 2 in Equation (16) of \citeN{ulrich1994positivity}.
+These terms with subscripts $0$ and $1$ make the computation of $\alpha$, $\beta$, and $\gamma$ functions of the second derivative terms, as seen in Step \ref{alg1-abg} below.}
   \item $\gamma := \gamma_0 + \gamma_1 D^2f(x_0)$
@@
     \label{alg1-abg}
-  \item if $(\beta \leq 6)$ then return $\big(\alpha > - (\beta + 2) / 2\big)$
-  \item else return $\big(\gamma > -2 \sqrt{\beta - 2}\big)$ \label{alg1-last}
+  \item if $(\beta \leq 6)$ return $\big(\alpha > - (\beta + 2) / 2\big)$
+  \item else $\qquad$ return $\big(\gamma > -2 \sqrt{\beta - 2}\big)$
+    \\ end if \label{alg1-last}
 }
@@
     {make\_monotone$\big (x_0$,$x_1$,$f \big)$}
-    {where $x_0,\ x_1 \in \mathbb{R}$, $x_0 < x_1$, and $f$ is an order six polynomial. Returns monotone $f$.}
+    {where $x_0,\ x_1 \in \mathbb{R}$, $x_0 < x_1$, and $f$ is an order six polynomial defined by $f(x_0)$, $Df(x_0)$, $D^2f(x_0)$, $f(x_1)$, $Df(x_1)$, $D^2f(x_1)$.
+Returns $f$ monotone on $[x_0,x_1]$.}
     {\label{alg:make-monotone}
-    \item if $\big(f(x_1) - f(x_0) = 0\big)$
+    \item if $\big(f(x_1) = f(x_0)\big)$
       \indented{$Df(x_0) := Df(x_1) := D^2f(x_0) := D^2f(x_1) := 0$}
-      \indented{return $f$}
+      \indented{return}
+      end if
     \item $Df(x_0) :=\ $median$\displaystyle \bigg(0,\ Df(x_0),\ 14 \frac{f(x_1) - f(x_0)}{x_1 - x_0} \bigg)$
       \\ $Df(x_1) :=\ $median$\displaystyle \bigg(0,\ Df(x_1),\ 14 \frac{f(x_1) - f(x_0)}{x_1 - x_0}\bigg)$ \label{alg2-init-df}
-      \comment{The selection of values in Step \ref{alg2-init-df} for $Df(x_0)$ and $Df(x_1)$ is suggested by \citeN{ulrich1994positivity} and originally from \citeN{huynh1993accurate}.
+      \comment{The selection of values in Step \ref{alg2-init-df} for $Df(x_0)$ and $Df(x_1)$ is suggested by \citeN{ulrich1994positivity} and also by \citeN{huynh1993accurate}.
 These rules quickly enforce upper and lower bounds on derivative values to ensure quintic monotonicity is obtainable.}
       \comment{The next case that will be covered is in accordance with a simplified condition for quintic monotonicity that reduces to one of cubic positivity studied in \citeN{schmidt1988positivity}, where $\alpha$, $\beta$, $\gamma$, and $\delta$ are defined in terms of values and derivatives of $f$ at $x_0$ and $x_1$.
-Steps \ref{alg2-dx-conditions} and \ref{alg2-rescale-dx} ensure that the following bounds on $D^2f(x_0)$ are compatible.
-Step \ref{alg2-ddx0-upper} is the final preparatory step, ensuring that the conditions on $D^2f(x_1)$ are compatible.
-To conclude the simplified case, Step \ref{alg2-gamma} first enforces $\gamma \geq \delta$, then Steps \ref{alg2-alpha} and \ref{alg2-beta} enforce $\alpha > 0$ and $\beta \geq \alpha$ respectively.
+Steps \ref{alg2-dx-conditions} and \ref{alg2-rescale-dx} ensure that the following bounds on $D^2f(x_0)$ are compatible while Step \ref{alg2-ddx0-upper} ensures that the conditions on $D^2f(x_1)$ are compatible.
+To conclude the simplified case, Step \ref{alg2-gamma} first enforces $\gamma \geq \delta$, then Steps \ref{alg2-alpha} and \ref{alg2-beta} enforce $\alpha \geq 0$ and $\beta \geq \alpha$ respectively.
 Combined, these three conditions derived from Equation (2.18) in \citeN{schmidt1988positivity} guarantee monotonicity of this polynomial piece.}
@@
       \indented{\indented{$Df(x_1) := Df(x_1) \text{ max}\displaystyle \bigg(0, \frac{20v}{w \big(5 Df(x_0) + 4 Df(x_1)\big)}\bigg) $}} \label{alg2-rescale-dx}
+      \indented{end if}
     \item \indented{$D^2f(x_0) :=\ $min$\displaystyle \bigg(D^2f(x_0), \frac{4 (2 Df(x_0)+Df(x_1))+20 v/w}{w} \bigg)$} \label{alg2-ddx0-upper}
@@
     \item  \indented{$D^2f(x_1) :=\ $max$\displaystyle \bigg(D^2f(x_1), \frac{3w D^2f(x_0) - 24 Df(x_0) - 32 Df(x_1) + 60v/w}{5w} \bigg)$} \label{alg2-beta}
-    \item
-      \indented{return $f$}
-      \comment{The following code considers the remaining case where $Df(x_0) \neq 0$ and $Df(x_1) \neq 0$.}
+      \indented{return}
+      end if
+      \comment{The following code considers the case where $Df(x_0) \neq 0$ and $Df(x_1) \neq 0$.}
     \item $\displaystyle A := Df(x_0)\frac{x_1 - x_0}{f(x_1) - f(x_0)}$
@@
     \item if $\big($max$(A,B) > 6\big)$
-      \indented{$Df(x_0) := 6 Df(x_0)\ /\ $max$(A,B)$}
-      \indented{$Df(x_1) := 6 Df(x_1)\ /\ $max$(A,B)$}
-      \comment{This box bound ensures that $(A,B)$ remains within a viable region of monotonicity (satisfying Theorem 4, seen in Fig. 6 of \citeN{ulrich1994positivity}).}
+      \indented{$Df(x_0) := 6 Df(x_0) \big/ $max$(A,B)$}
+      \indented{$Df(x_1) := 6 Df(x_1) \big/ $max$(A,B)$}
+      end if
+      \comment{This ensures that $(A,B)$ remains within a viable region of monotonicity (satisfying Theorem 4, seen in Fig. 6 of \citeN{ulrich1994positivity}).}
     \item $\eta := \big(D^2f(x_0), D^2f(x_1)\big)$
-      \\ $\displaystyle \eta_0 := \bigg({-\frac{\sqrt{A}}{4}}\big(7\sqrt{A} + 3\sqrt{B}\big),\ \frac{\sqrt{B}}{4}\big(3\sqrt{A} + 7\sqrt{B}\big)  \bigg)$
-    \item $D^2f(x_0),\ D^2f(x_1) := $ binary\_search$\big($is\_monotone$(x_0,x_1,f)$, $\eta,$ $\eta_0\big)$
-      \comment{This binary search identifies feasible values of $D^2f(x_0)$ and $D^2f(x_1)$ that are nearest to the current values.
+      \\ $\displaystyle \eta_0 := \left({-\frac{\sqrt{A}}{4}}\big(7\sqrt{A} + 3\sqrt{B}\big),\ \frac{\sqrt{B}}{4}\big(3\sqrt{A} + 7\sqrt{B}\big) \right)$
+    \item $\bigl(D^2f(x_0)$, $D^2f(x_1)\bigr) :=$ binary\_search$\bigl(g,
+       \eta,\eta_0\bigr)$
+      \comment{Here $g(z) =$\texttt{ is\_monotone}$(x_0,x_1,f)$ where $x_0$, $x_1$, $f(x_0)$, $f(x_1)$, $Df(x_0)$, $Df(x_1)$ are fixed, and the variable $z=\bigl(D^2f(x_0),D^2f(x_1)\bigr)$.
+This binary search identifies feasible values of $D^2f(x_0)$ and $D^2f(x_1)$ that are nearest to the current values $\eta$.
 This search is valid because $\eta_0$ is guaranteed to produce a monotone function, which can be seen in Equation (23) of \citeN{ulrich1994positivity}.}
-    \item return $f$
+    return
 }
@@
 Also note that an efficient implementation of Algorithm \ref{alg:make-monotone} only needs to recalculate Steps \ref{alg1-abg} through \ref{alg1-last} of Algorithm \ref{alg:check-monotone} during the binary search.
-Next, an algorithm for constructing a monotone quintic interpolating spline is presented.
+Next, an algorithm for constructing a monotone quintic spline interpolant is presented.
 
@@
 \algorithm
-    {monotone\_spline$\big(\{k_1,\ldots,k_n\},\ f,\ s\big)$}
-    {where $\{k_1,\ldots,k_n\}$ is an increasing sequence of real numbers, $f$ is a piecewise order six polynomial defined over intervals $[k_i, k_{i+1}]$  for $i = 1, \ldots, n-1$, and integer $s \geq 1$.}
+    {monotone\_spline$\big((k_1$, $\ldots$, $k_n)$, $f$, $s\big)$}
+    {where $(k_1$, $\ldots$, $k_n)$ is an increasing sequence of real numbers, $f$ is an order six piecewise polynomial with breakpoints $k_1$, $\ldots$, $k_n$ defined by the data $\bigl\{f(k_i)\bigr\}_{i=1}^n$, $\bigl\{Df(k_i)\bigr\}_{i=1}^n$, $\bigl\{D^2f(k_i)\bigr\}_{i=1}^n$, and $s > 1$ is an integer shrink factor.}
     {\label{alg:monotone-spline}
     \item create queue
-      \\ create changed$(1, \ldots, n) := $ FALSE
-      \\ create step\_size$(1, \ldots, n) := \{Df(k_i) / s$ for $i = 1, \ldots, n\}$\label{step:create}
+      \\ create changed$(1$, $\ldots$, $n) := $ FALSE
+      \\ create step\_size$(1$, $\ldots$, $n) := \bigl(Df(k_1)/s$, $\ldots$, $Df(k_n)/s\bigr)$ \label{step:create}
       \comment{The three variables defined in Step \ref{step:create} are used to ensure the eventual achievement of monotonicity.
 The \texttt{queue} is a standard first in first out (FIFO) queue with \texttt{enqueue} and \texttt{dequeue} operations.
-The array \texttt{changed} contains Booleans describing whether or not a partition point belongs to an interval that has been modified to enforce monotonicity.
-The \texttt{step\_size} array contains the real valued derivative step sizes to use in the search for a valid monotone spline.}
-    \item for $i := 1, \ldots, n-1$
-      \indented{enqueue $(k_i, k_{i+1})$}
+The array \texttt{changed} contains Booleans describing whether or not a breakpoint belongs to an interval that has been modified to enforce monotonicity.
+The \texttt{step\_size} array contains the real-valued derivative decrement step sizes to use in the search for a valid monotone spline.}
+    \item for $i := 1$, $\ldots$, $n-1$
+      \indented{enqueue $\bigl((k_i, k_{i+1})\bigr)$}
       \comment{Initially, all intervals must be checked for monotonicity. The following loop will continue until all intervals are verified as monotone without need for modification.}
     \item while $($size$($queue$) > 0)$
-    \item \indented{$x_j,\ x_{j+1} := $ dequeue}
-    \item \indented{if $\big($not is\_monotone$(x_j,\ x_{j+1},\ f)\big)$}
-    \item \indented{\indented{if $\big($changed$(j)\big) \qquad \ Df(x_j) := Df(x_j) - $step\_size$(j)$}}
-      \indented{\indented{if $\big($changed$(j+1)\big) \quad Df(x_{j+1}) := Df(x_{j+1}) - $step\_size$(j+1)$}}
-      \indented{\indented{\comment{If a partition point belongs to an interval that has been previously modified, then the enforced monotonicity conditions on the second derivatives of adjacent intervals must contradict one another.
+    \item \indented{$(x_j$, $x_{j+1}) := $ dequeue}
+    \item \indented{if $\big($not is\_monotone$(x_j$, $x_{j+1}$, $f)\big)$}
+    \item \indented{\indented{if $\big($changed$(j)\big) \qquad \ Df(x_j) := Df(x_j) -{}$step\_size$(j) \qquad \ \ $ end if}}
+      \indented{\indented{if $\big($changed$(j+1)\big) \quad Df(x_{j+1}) := Df(x_{j+1}) -{}$step\_size$(j+1)$ end if}}
+      \indented{\indented{\comment{If a breakpoint belongs to an interval that has been previously modified, then the enforced monotonicity conditions on the second derivatives of adjacent intervals must contradict one another.
 In turn, the involved first derivatives are decreased towards zero by a predetermined step size.}}}
@@
       \indented{\indented{changed$(j+1) :=$ TRUE}} \label{step:changed}
-    \item \indented{\indented{if $\big(j > 1$ and $(x_{j-1}, x_j) \not \in$ queue$\big)$ enqueue $(x_{j-1}, x_j)$}}
-      \indented{\indented{if $\big((x_j, x_{j+1}) \not \in$ queue$\big)$ enqueue $(x_j, x_{j+1})$}}
-      \indented{\indented{if $\big(j+1 < n$ and $(x_{j+1}, x_{j+2}) \not \in$ queue$\big)$ enqueue $(x_{j+1}, x_{j+2})$}} \label{step:requeue}
+    \item \indented{\indented{if $\big(j > 1$ and $(x_{j-1}, x_j) \not\in$ queue$\big)$ enqueue $\bigl((x_{j-1}, x_j)\bigr)$ end if}}
+      \indented{\indented{enqueue $\bigl((x_j, x_{j+1})\bigr)$}}
+      \indented{\indented{if $\big(j+1 < n$ and $(x_{j+1}, x_{j+2}) \not\in$ queue$\big)$ enqueue $\bigl((x_{j+1}, x_{j+2})\bigr)$ end if}} \label{step:requeue}
       \indented{\indented{\comment{Step \ref{step:changed} records the endpoints of the current interval as having been changed, while \ref{step:requeue} adds adjacent intervals to \texttt{queue} that may have inadvertently been made nonmonotone by the changes to the present interval.}}}
+      \indented{end if}
+      end while
 }
@@
 
-Given the potential for successive monotonicity fixes across all intervals, the worst case runtime of Algorithm \ref{alg:monotone-spline} is $\mathcal{O}(s n)$ for $n$ partition points and step size divisor $s$.
-In practice this worst case has been observed to be incredibly unlikely and difficult to produce.
+Given the potential for successive monotonicity fixes across all intervals, the worst case runtime of Algorithm \ref{alg:monotone-spline} is $\mathcal{O}(s n)$ for $n$ breakpoints and shrink factor $s$.
+In practice this worst case has been observed to be unlikely and difficult to produce.
 
@@
 The quintic approximation consistently has a maximum error that is roughly half that of the cubic approximation, given the same number of points.
-There is a (subtle) increase in the error gap between the two approximations as more points are added.
+There is an increase in the error gap between the two approximations as more points are added.
+The quintic spline requires only $700$ points to approximate the function to the same accuracy achieved by the cubic with $1000$ points.
 
@@
 Both axes are log scaled.
-The quintic approximation generally provides an approximation with slightly more than half of the maximum error of the cubic approximation, while the relative gap between grows with increasing number of points.
-For larger absolute errors, this means the accuracy of the cubic is achieved with about ten fewer samples from the quintic.
-However, for smaller absolute errors the quintic spline requires hundreds fewer samples to achieve the same accuracy as the cubic.}
+The quintic approximation generally provides an approximation with about half of the maximum error of the cubic approximation, while the relative gap between grows with increasing number of points.
+In order to achieve the same absolute error the quintic spline requires ten fewer points at max error $10^{-4}$, thirty fewer points at max error $10^{-6}$, and $300$ fewer points at error $10^{-8}$.}
   \label{fig:experiment_1}
@@
 
-Consider the cumulative distribution function (CDF) defined by a mixture of $3$ Gaussian distributions and visualized in Figure \ref{fig:experiment_2_distribution}.
+Consider the cumulative distribution function (CDF) defined by a mixture of three Gaussian distributions and visualized in Figure \ref{fig:experiment_2_distribution}.
 The distributions have weights $(.3, .6, .1)$, means $(.2, .45, .85)$ and standard deviations $(.05, .08, .03)$.
-$200$ random samples are generated from this distribution with both cubic and quintic approximations from $4$ points, and over $100$ trials the resulting spread of empirical distribution functions (EDFs) are visualized in Figure \ref{fig:experiment_2_results}.
+$200$ random samples are generated from this distribution with both cubic and quintic approximations from four points, and over $100$ trials the resulting spread of empirical distribution functions (EDFs) are visualized in Figure \ref{fig:experiment_2_results}.
 
@@
   \includegraphics[width=.8\textwidth]{experiment_2_distribution}
-  \caption{The cumulative distribution function (top) and probability density function (bottom) for the Gaussian mixture distribution in experiment 2.
+  \caption{The cumulative distribution function (top) and probability density function (bottom) for the Gaussian mixture distribution in Experiment 2.
 The distributions have weights $(.3, .6, .1)$, means $(.2, .45, .85)$ and standard deviations $(.05, .08, .03)$.
-Random samples are generated from this mixture distribution by evaluating the inverse of the CDF at uniformly random numbers in the range $[0,1]$.}
+Random samples are generated from this mixture distribution by evaluating the inverse of the CDF at random numbers uniformly distributed in the range $[0,1]$.}
   \label{fig:experiment_2_distribution}
@@
   \caption{Both cubic and quintic splines are used to approximate the CDF of a Gaussian mixture distribution with three components.
-$4$ CDF points are used to build the approximation, $200$ random samples are generated with the approximate CDFs over $100$ trials.
+Four CDF points are used to build the approximation, $200$ random samples are generated with the approximate CDFs over $100$ trials.
 The resulting cloud of empirical distribution function (EDF) values is seen above.
@@
 This experiment studies the number of times the algorithms \texttt{is\_monotone} and \texttt{make\_monotone} are executed for increasingly large sequences of random monotone data.
-Statistics on the expected number of interval corrections are presented.
+The minimum, median, and maximum number of times that Algorithms \ref{alg:check-monotone} and \ref{alg:make-monotone} are executed is recorded, as well as the number of steps taken in all calls to \texttt{binary\_search}.
+The number of calls and steps grows linearly with $n$ as expected, requiring roughly one call to \texttt{make\_monotone} and $20$ binary search steps to create a monotone piece.
+Some (less common) problems require an average of three calls to \texttt{make\_monotone} per interval.
 
-\texttt{NEED STATISTICS TABLE}
+\begin{table}
+  \centering
+  \begin{tabular}{c|c c c|c c c|c c c}
+    \hline
+    \multirow{2}{*}{$n$}
+      & \multicolumn{3}{c|}{Checks} & \multicolumn{3}{c}{Fixes} & \multicolumn{3}{|c}{Search Steps} \\
+      & min & median & max & min & median & max & min & median & max \\
+    \hline
+    5 & 4 & 6 & 274 & 0 & 1 & 154 & 3 & 31 & 2006\\
+    25 & 28 & 47 & 378 & 2 & 14 & 225 & 80 & 348 & 3742\\
+    50 & 61 & 124 & 484 & 6 & 40 & 238 & 217 & 851 & 4445\\
+    75 & 102 & 216 & 842 & 14 & 74 & 411 & 463 & 1491 & 6333\\
+    100 & 142 & 306 & 776 & 21 & 107 & 380 & 629 & 2140 & 7414\\
+    \hline
+  \end{tabular}
+  \caption{Here we see a table of the minimum, median, and maximum number of checks, fixes, and binary search steps required in the execution of \texttt{make\_spline} for increasing size sequences, $n$, over $100$ randomly generated sets of monotone data.
+Notice the maximum for each counter is often significantly greater than the minimum and median, because the distribution of each counter is skewed right.}
+  \label{table:e3_results}
+\end{table}
 
@@
 
-This is an improvement over the cubic, although the complexity of the algorithm is increased as well.
+The monotone quintic spline interpolant provides a distinct increase in accuracy of the monotone cubic variant.
+Although the complexity of the algorithm is increased, the number of points required to achieve the same level of accuracy has decreased.
 In the present state, the algorithm for enforcing monotonicity on a spline is not as trivially parallel as the cubic algorithm, however it can still be parallelized.
 
-Knowing this algorithm is slightly more complex than the cubic variant, it appears worth the computational effort because of gains in accuracy or a decreased necessary number of nodes.
+Acknowledging that the algorithm for constructing a monotone quintic spline interpolant is slightly more complex than the cubic case, gains in accuracy or a decreased necessary number of points are often worth the computational effort.
 
@@
 
-This paper proposed and tested an algorithm for constructing monotone quintic interpolating splines.
-Experiments demonstrate an improvement in approximation accuracy over monotone cubic interpolating splines, as expected based on theory.
+This paper proposes and tests an algorithm for constructing monotone quintic spline interpolants.
+Experiments demonstrate an improvement in approximation accuracy over monotone cubic spline interpolants, as expected based on theory.
 
@@
 
-\section*{Acknowledgments}
-The NSF is acknowledged for its support of this project.
+%% \section*{Acknowledgments}
+%% The NSF is acknowledged for its support of this project.
 

### 2019-12-16 Paper_[2019-11]_HPC/experiments/example_2.py
Fixed some spelling errors and references.

@@ -3,6 +3,7 @@ import numpy as np
 import sys, os
 sys.path += [os.path.abspath("code")]
 from plot import Plot
+from polynomial import polynomial_piece
 sys.path.pop(-1)
 
 c = "rgba(50,50,50,.5)"
@@ -36,6 +37,33 @@ p.add_annotation(" 2 ", 1.55, 1.55, font_family="times", font_size=17,
 p.add_annotation(" 1 ", 2.3, 2.3, font_family="times", font_size=17,
                  show_arrow=False, border_width=1, x_anchor="left")
 
-
 p.show(file_name="cubic_projection.html", width=400, height=400, show_legend=False)
+
+
+
+VERIFY_CUBIC = False
+# --------------------------------------------------------------------
+#             VERIFY CORRECTNESS OF CUBIC INTERPOLANT
+if VERIFY_CUBIC:
+    # Knots
+    k0 = 0
+    k1 = 2
+    h0 = k1 - k0
+    # Truth function
+    f = lambda x: 0. if x < (k0+k1)/2 else 1.
+    df = lambda x: -1.
+    # Basis functions.
+    u = lambda t: 3*t**2 - 2*t**3
+    p = lambda t: t**3 - t**2
+    # Approximation function
+    fh = lambda x: f(k0) * u((k1-x)/h0) + f(k1) * u((x - k0)/h0) - \
+         h0 * df(k0) * p((k1-x)/h0) + h0 * df(k1) * p((x - k0)/h0)
+    pp = polynomial_piece([0,-1], [1,-1], (k0,k1))
+
+    # Plot this.
+    plot = Plot()
+    plot.add_func("F hat", fh, [k0, k1])
+    plot.add_func("Pp", pp, [k0, k1])
+    plot.show()
+# --------------------------------------------------------------------
 

### 2019-12-16 Paper_[2019-11]_HPC/paper/paper.tex
Fixed some spelling errors and references.
Final read complete. Submitting paper as is.

@@
 The major contribution of this work is an algorithm for constructing monotone quintic interpolating splines that utilizes existing quintic monotonicity theory.
-The remainder of this paper is structured as follows: {\bf ?????}
+The remainder of this paper is structured as follows: Section \ref{sec:monotone-cubic} summarizes the existing monotone cubic spline interpolation methodology, Section \ref{sec:monotone-quintic} presents an algorithm for constructing monotone quintic spline interpolants, Section \ref{sec:results} offers experiments and results with cubic and quintic methods, Sections \ref{sec:discussion} and \ref{sec:conclusion} discuss results and conclude.
 
 \subsection{Computing a Monotone Cubic Interpolant}
+\label{sec:monotone-cubic}
 
@@
 Consider the terms $\alpha = \big(\Delta_i (k_{i+1}-k_i)\big) / \big(f(k_{i+1}) - f(k_i)\big)$ and $\beta = \big(\Delta_{i+1}(k_{i+1}-k_i)\big) / \big(f(k_{i+1}) - f(k_i)\big)$.
-If $f(k_{i+1}) - f(k_i)=0$, take $\Delta_i = \Delta_{i+1} =0$ and $\alpha =\beta =1$.
-In general, monotonicity of a cubic polynomial over a subinterval can be maintained by ensuring that $\alpha$ and $\beta$ reside in any of the regions depicted in Figure \ref{fig:projection}.
+If $f(k_i) = f(k_{i+1})$, take $\Delta_i = \Delta_{i+1} =0$ and $\alpha =\beta =1$.
+Monotonicity of a cubic polynomial over a subinterval can be maintained by ensuring that $\alpha$ and $\beta$ reside in any of the regions depicted in Figure \ref{fig:projection}.
 
@@
 \section{Monotone Quintic Interpolation}
+\label{sec:monotone-quintic}
 
@@
   \item if $(\beta \leq 6)$ return $\big(\alpha > - (\beta + 2) / 2\big)$
-  \item else $\qquad$ return $\big(\gamma > -2 \sqrt{\beta - 2}\big)$
+    \\ else $\qquad$ return $\big(\gamma > -2 \sqrt{\beta - 2}\big)$
     \\ end if \label{alg1-last}
@@
        \eta,\eta_0\bigr)$
-      \comment{Here $g(z) =$\texttt{ is\_monotone}$(x_0,x_1,f)$ where $x_0$, $x_1$, $f(x_0)$, $f(x_1)$, $Df(x_0)$, $Df(x_1)$ are fixed, and the variable $z=\bigl(D^2f(x_0),D^2f(x_1)\bigr)$.
+      \comment{Here $g(z) =$\texttt{ is\_monotone}$(x_0,x_1,f)$ where $x_0$, $x_1$, $f(x_0)$, $f(x_1)$, $Df(x_0)$, $Df(x_1)$ are fixed, and the variable $z=\bigl(D^2f(x_0)$,$D^2f(x_1)\bigr)$.
 This binary search identifies feasible values of $D^2f(x_0)$ and $D^2f(x_1)$ that are nearest to the current values $\eta$.
@@
 The search requires a fixed number of steps to achieve any predetermined relative precision.
-Also note that an efficient implementation of Algorithm \ref{alg:make-monotone} only needs to recalculate Steps \ref{alg1-abg} through \ref{alg1-last} of Algorithm \ref{alg:check-monotone} during the binary search.
+Also note that an efficient implementation of Algorithm \ref{alg:make-monotone} only needs to recalculate Steps \ref{alg1-abg} and \ref{alg1-last} of Algorithm \ref{alg:check-monotone} during the binary search.
 Next, an algorithm for constructing a monotone quintic spline interpolant is presented.
@@
     \item \indented{if $\big($not is\_monotone$(x_j$, $x_{j+1}$, $f)\big)$}
-    \item \indented{\indented{if $\big($changed$(j)\big) \qquad \ Df(x_j) := Df(x_j) -{}$step\_size$(j) \qquad \ \ $ end if}}
-      \indented{\indented{if $\big($changed$(j+1)\big) \quad Df(x_{j+1}) := Df(x_{j+1}) -{}$step\_size$(j+1)$ end if}}
+    \item \indented{\indented{if $\big($changed$(j)\big) \qquad \ Df(x_j) := Df(x_j) -{}$step\_size$(j)$}}
+      \indented{\indented{if $\big($changed$(j+1)\big) \quad Df(x_{j+1}) := Df(x_{j+1}) -{}$step\_size$(j+1)$}}
       \indented{\indented{\comment{If a breakpoint belongs to an interval that has been previously modified, then the enforced monotonicity conditions on the second derivatives of adjacent intervals must contradict one another.
@@
       \indented{\indented{changed$(j+1) :=$ TRUE}} \label{step:changed}
-    \item \indented{\indented{if $\big(j > 1$ and $(x_{j-1}, x_j) \not\in$ queue$\big)$ enqueue $\bigl((x_{j-1}, x_j)\bigr)$ end if}}
+    \item \indented{\indented{if $\big(j > 1$ and $(x_{j-1}, x_j) \not\in$ queue$\big)$ enqueue $\bigl((x_{j-1}, x_j)\bigr)$}}
       \indented{\indented{enqueue $\bigl((x_j, x_{j+1})\bigr)$}}
-      \indented{\indented{if $\big(j+1 < n$ and $(x_{j+1}, x_{j+2}) \not\in$ queue$\big)$ enqueue $\bigl((x_{j+1}, x_{j+2})\bigr)$ end if}} \label{step:requeue}
+      \indented{\indented{if $\big(j+1 < n$ and $(x_{j+1}, x_{j+2}) \not\in$ queue$\big)$ enqueue $\bigl((x_{j+1}, x_{j+2})\bigr)$}} \label{step:requeue}
       \indented{\indented{\comment{Step \ref{step:changed} records the endpoints of the current interval as having been changed, while \ref{step:requeue} adds adjacent intervals to \texttt{queue} that may have inadvertently been made nonmonotone by the changes to the present interval.}}}
@@
 Given the potential for successive monotonicity fixes across all intervals, the worst case runtime of Algorithm \ref{alg:monotone-spline} is $\mathcal{O}(s n)$ for $n$ breakpoints and shrink factor $s$.
-In practice this worst case has been observed to be unlikely and difficult to produce.
+In practice this worst case has been observed to be unlikely.
 
 \section{Experimental Results}
+\label{sec:results}
 
@@
   \caption{The error in the quintic and cubic approximations of $\sin(x) + x$ with increasing number of points.
-Both axes are log scaled.
+Both axes are log scaled, where $\mu$ means $\times 10^{-6}$ and n means $\times 10^{-9}$ on the vertical axis.
 The quintic approximation generally provides an approximation with about half of the maximum error of the cubic approximation, while the relative gap between grows with increasing number of points.
-In order to achieve the same absolute error the quintic spline requires ten fewer points at max error $10^{-4}$, thirty fewer points at max error $10^{-6}$, and $300$ fewer points at error $10^{-8}$.}
+In order to achieve the same absolute error the quintic spline requires ten fewer points at max error $10^{-4}$, thirty fewer points at max error $10^{-6}$, and $300$ fewer points at max error $10^{-8}$.}
   \label{fig:experiment_1}
@@
 The maximum approximation error in the CDFs are similar, $.08$ for the cubic and $.05$ for the quintic.
-However, the maximum EDf differences observed are doubled by the variance of random samples, $.15$ for the cubic and $.1$ for the quintic.}
+However, the maximum EDF differences observed are doubled by the variance of random samples, $.15$ for the cubic and $.1$ for the quintic.}
   \label{fig:experiment_2_results}
@@
 \section{Discussion}
+\label{sec:discussion}
 
-The monotone quintic spline interpolant provides a distinct increase in accuracy of the monotone cubic variant.
+The monotone quintic spline interpolant provides a distinct increase in accuracy over the monotone cubic variant.
 Although the complexity of the algorithm is increased, the number of points required to achieve the same level of accuracy has decreased.
@@
 \section{Conclusion and Future Work}
+\label{sec:conclusion}
 

### 2019-12-16 Paper_[2019-11]_HPC/paper/paper.tex
Submitted the paper.

@@
 sample problems to demonstrate the improved accuracy of monotone
-quintic spline interpolants compared to that for monotone cubic spline
+quintic spline interpolants compared to that for existing monotone cubic spline
 interpolants.

### 2020-04-14 3-Disseration/abstracts.tex
Updated min wiggle fit tests, wrote report on the topic that was sent to Dr. Watson.

--- 2017-12-11 Paper_[2017-11]_HPC/paper.tex
@@
-%**************************************************************************
-%* SpringSim 2017 Author Kit
-%*
-%* Word Processing System: TeXnicCenter and MiKTeX
-%*
-%**************************************************************************
-
-\documentclass{scspaperproc}
-
-\usepackage{latexsym}
-\usepackage{graphicx}
-\usepackage{mathptmx}
-\usepackage{amsmath}
-\usepackage{amsfonts}
-\usepackage{amssymb}
-\usepackage{amsbsy}
-\usepackage{amsthm}
-%% \usepackage{color}   % Needed to color text for drafts
-
-%% \usepackage{algorithm}     % needed for the algorithm block
-%% \usepackage{algpseudocode} % algorithmicx psuedo-code styling
-%\usepackage[pdftex,colorlinks=true,urlcolor=blue,citecolor=black,anchorcolor=black,linkcolor=black]{hyperref}
-%% \usepackage[dvips,colorlinks=true,urlcolor=blue,citecolor=black,%
-%% anchorcolor=black,linkcolor=black]{hyperref}
-
-% custom hyphenation rules
-\usepackage{hyphenat}
-\hyphenation{op-tical net-works semi-conduc-tor}
-
-% theorem style
-\newtheoremstyle{scsthe}% hnamei
-{8pt}% hSpace abovei
-{8pt}% hSpace belowi
-{\it}% hBody fonti
-{}% hIndent amounti1
-{\bf}% hTheorem head fontbf
-{.}% hPunctuation after theorem headi
-{.5em}% hSpace after theorem headi2
-{}% hTheorem head spec (can be left empty, meaning `normal')i
-\theoremstyle{scsthe}
-\newtheorem{theorem}{Theorem}
-\renewcommand{\thetheorem}{\arabic{theorem}}
-\newtheorem{corollary}[theorem]{Corollary}
-\renewcommand{\thecorollary}{\arabic{corollary}}
-\newtheorem{definition}{Definition}
-\renewcommand{\thedefinition}{\arabic{definition}}
-\newcommand{\rpm}{\raisebox{.2ex}{$\scriptstyle\pm$}}
-
-% avoid overrunning the right margin
-\sloppy
-
-%% ***** NOTE *****
-
-%% The use of the long citation format (e.g. "Brown and Edwards
-%% (1993)" rather than "[5]") and at the same time using the hyperref
-%% package can lead to hard to trace bugs in case the citation is
-%% broken accross the line (usually this will mark the entire
-%% paragraph as a hyperlink (clickable) which is easily noticeable and
-%% fixed if using colorlinks, but not if the color is black -- as it
-%% is now). Worse yet, if a citation spans page boundary, LaTeX
-%% compilation can fail, with an obscure error message. Since this
-%% depends a lot on the flow of the text and wording, these bugs come
-%% and go and can be extremely hard for a beginner to trace. The error
-%% message can look like this:
-%%
-%%    ! pdfTeX error (ext4): \pdfendlink ended up in different nesting
-%%    level than \pdfstartlink.  \AtBegShi@Output ...ipout \box
-%%    \AtBeginShipoutBox \fi \fi
-%%    l.174 
-%%    ! ==> Fatal error occurred, no output PDF file produced!
-%%
-%% and can be universally fixed by putting an \mbox{} around the
-%% citation in question (in this case, at line 174) and maybe adapting
-%% the wording a little bit to improve the paragraph typesetting,
-%% which is perhaps not immediately obvious.
-%****************************************************************************
-
-% begin document
-\begin{document}
-
-% Page header (author list)
-\SCSpagesetup{Lux, Watson, Chang, Bernard, Li, Xu, Back, Butt, Cameron, and Hong}
-
-% Conference info
-\def\SCSconferenceacro{SpringSim}
-\def\SCSpublicationyear{2018}
-\def\SCSconferencedates{April 15-18}
-\def\SCSconferencevenue{Baltimore, MD, USA}
-\def\SCSsymposiumacro{HPC} % High Performance Computing Symposium
-
-% title
-\title{Predictive Modeling of I/O Characteristics \\ in High
-  Performance Computing Systems}
-
-% AUTHOR LIST
-% *** NOTE: May need to adjust titlevboxsize in the preamble
-\author{Thomas C. H. Lux \\ [8pt]
-Dept. of Computer Science \\
-Virginia Polytechnic Institute\\
-\& State University \\
-Blacksburg, VA 24061 \\
-tchlux@vt.edu \\
-\and
-Layne T. Watson \\[8pt]
-Dept. of Computer Science\\
-Dept. of Mathematics\\
-Dept. of Aerospace \& Ocean Eng.\\ 
-Virginia Polytechnic Institute\\
-\& State University \\
-\and
-\vspace{-30pt}\\
-Tyler H. Chang\\
-Jon Bernard\\
-Bo Li\\[8pt]
-Dept. of Computer Science\\ 
-Virginia Polytechnic Institute\\
-\& State University \\
-\and
-\vspace{-30pt}\\
-Li Xu\\[8pt]
-Dept. of Statistics\\ 
-Virginia Polytechnic Institute\\
-\& State University \\
-\and
-\vspace{-30pt}\\
-Godmar Back\\
-Ali R. Butt\\
-Kirk W. Cameron\\[8pt]
-Dept. of Computer Science\\ 
-Virginia Polytechnic Institute\\
-\& State University \\
-\and
-\vspace{-30pt}\\
-Yili Hong\\[8pt]
-Dept. of Statistics\\ 
-Virginia Polytechnic Institute\\
-\& State University \\
-}
-
-\maketitle
-
-\vspace{-40pt}
-
-\section*{Abstract}
-
+\section{Abstract of Naive}
 Each of high performance computing, cloud computing, and computer
@@
 
-\textbf{Keywords:} Regression, approximation, interpolation,
-performance modeling
-
-%     Introduction     
-%======================
-\section{Introduction and related work}
-\label{sec:introduction}
-
-Performance tuning is often an experimentally complex and time
-intensive chore necessary for configuring HPC systems. The procedures
-for this tuning vary largely from system to system and are often
-subjectively guided by the system engineer(s). Once a desired level of
-performance is achieved, an HPC system may only be incrementally
-reconfigured as required by updates or specific jobs. In the case that
-a system has changing workloads or nonstationary performance
-objectives that range from maximizing computational throughput to
-minimizing power consumption and system variability, it becomes clear
-that a more effective and automated tool is needed for configuring
-systems. This scenario presents a challenging and important
-application of multivariate approximation and interpolation
-techniques.
-
-Predicting the performance of an HPC system is a challenging problem
-that is primarily attempted in one of two ways: (1) build a
-statistical model of the performance by running experiments on the
-system at select settings, or (2) run artificial experiments using a
-simplified simulation of the target system to estimate architecture
-and application bottlenecks. In this paper the proposed multivariate
-modeling techniques rest in the first category, and they represent a
-notable increase in the ability to model complex interactions between
-system parameters.
-
-Many previous works attempting to model system performance have used
-simulated environments to estimate the performance of a system
-\shortcite{grobelny2007fase,wang2009simulation,wang2013towards}. Some
-of these works refer to statistical models as being oversimplified and
-not capable of capturing the true complexity of the underlying
-system. This claim is partially correct, noting that a large portion
-of predictive statistical models rely on simplifying the model to one
-or two parameters
-\shortcite{snavely2002framework,bailey2005performance,barker2009using,ye2010analyzing}.
-These limited statistical models have provided satisfactory
-performance in very narrow application settings. Many of the
-aforementioned statistical modeling techniques claim to generalize,
-while simultaneously requiring additional code annotations, hardware
-abstractions, or additional application level understandings in order
-to generate models. The approach presented here requires no
-modifications of the application, no architectural abstractions, nor
-any structural descriptions of the input data being modeled. The
-techniques used are purely mathematical and only need performance data
-as input.
-
-Among the statistical models presented in prior works,
-\shortciteN{bailey2005performance} specifically mention that it is
-difficult for the simplified models to capture variability introduced
-by I/O. System variability in general has become a problem of
-increasing interest to the HPC and systems communities, however most
-of the work has focused on operating system (OS) induced variability
-\shortcite{beckman2008benchmarking,de2007identifying}. The work that
-has focused on managing I/O variability does not use any sophisticated
-modeling techniques \shortcite{lofstead2010managing}. Hence, this
-paper presents a case study applying advanced mathematical and
-statistical modeling techniques to the domain of HPC I/O
-characteristics. The models are used to predict the mean throughput of
-a system and the variance in throughput of a system. The discussion
-section outlines how the techniques presented can be applied to any
-performance metric and any system.
-
-%% Multivariate models for an HPC system would be a function of the
-%% tunable parameters built to accurately model some desired performance
-%% metric.
-
-%% \begin{enumerate}
-%% \item The value of multivariate Modeling
-%% \item The data context
-%% \item The proposed method for using multivariate models
-%% \item The impact of effective models
-%% \end{enumerate}
-
-In general, this paper compares five multivariate approximation
-techniques that operate on inputs in $\mathbb{R}^d$ ($d$-tuples of
-real numbers) and produce predicted responses in $\mathbb{R}$. In
-order to provide coverage of the varied mathematical strategies that
-can be employed to solve the continuous modeling problem, three of the
-techniques are regression based and the remaining two are
-interpolants. The sections below outline the mathematical formulation
-of each technique and provide computational complexity bounds with
-respect to the size (number of points and dimension) of input
-data. Throughout, $d$ will refer to the dimension of the input data,
-$n$ is the number of points in the input data, $x^{(i)} \in
-\mathbb{R}^d$ is the $i$-th input data point, $x^{(i)}_j$ is the
-$j$-th component of $x^{(i)}$, and $f(x^{(i)})$ is the response value
-of the $i$-th input data point.
-
-The remainder of the paper is broken up into five major parts. Section
-\ref{sec:multivariate} provides an overview of the multivariate
-modeling techniques, Section \ref{sec:methodology} outlines the
-methodology for comparing and evaluating the performance of the
-models, Section \ref{sec:results} presents the IOzone predictions,
-Section \ref{sec:discussion} discusses the obvious and subtle
-implications of the models' performance, and Section
-\ref{sec:conclusion} concludes and offers directions for future work.
-
-\section{Multivariate Models}
-\label{sec:multivariate}
-
-\subsection{Regression}
-\vspace{-10pt}
-Multivariate approximations are capable of accurately modeling a
-complex dependence of a response (in $\mathbb{R}$) on multiple
-variables (represented as a points in $\mathbb{R}^{d}$). The
-approximations to some (unknown) underlying function $f: \mathbb{R}^d
-\rightarrow \mathbb{R}$ are chosen to minimize some error measure
-related to data samples $f(x^{(i)})$. For example, least squares
-regression uses the sum of squared differences between modeled
-response values and true response values as an error measure.
-
-\vspace{-10pt}
-\subsubsection{Multivariate Adaptive Regression Splines}
-\vspace{-10pt}
-This approximation was introduced in
-\shortciteN{friedman1991multivariate} and subsequently improved to its
-current version in \shortciteN{stanford1993fast}, called fast
-multivariate adaptive regression splines (Fast MARS). In Fast MARS, a
-least squares fit model is iteratively built by beginning with a
-single constant valued function and adding two new basis functions at
-each iteration of the form
-
-$$ B_{2s-1}(x) = B_l(x) [c(x_i-v)]_+ ,$$
-$$ B_{2s}(x) = B_k(x) [c(x_i-v)]_- ,$$
-
-where $s$ is the iteration number, $B_l(x)$ and $B_k(x)$ are basis
-functions from the previous iteration, $c, v \in \mathbb{R}$,
-
-$$w_+ = \begin{cases} w, & w \geq 0 \\ 0, & w < 0 \end{cases},$$
-
-and $w_- = (-w)_+$. After iteratively constructing a model, MARS then
-iteratively removes basis functions that do not contribute to goodness
-of fit. In effect, MARS creates a locally component-wise tensor
-product approximation of the data. The overall computational
-complexity of Fast MARS is $\mathcal{O}(n d m^3)$ where $m$ is the
-maximum number of underlying basis functions. This paper uses an
-implementation of Fast MARS \shortcite{rudy2017pyearth} with $m =
-200$.
-
-\vspace{-10pt}
-\subsubsection{Multilayer Perceptron Regressor}
-\vspace{-10pt}
-The neural network is a well studied and widely used method for both
-regression and classification tasks
-\shortcite{hornik1989multilayer}. When using the rectified linear unit
-(ReLU) activation function \shortcite{dahl2013improving} and training
-with the BFGS minimization technique \shortcite{moller1993scaled}, the
-model built by a multilayer perceptron uses layers $l : \mathbb{R}^{i}
-\rightarrow \mathbb{R}^{j}$ defined by
-
-$$ l(u) = \big( u^t W_l \big)_+ ,$$
-
-where $W_l$ is the $i$ by $j$ weight matrix for layer $l$. In this
-form, the multilayer perceptron (MLP) produces a piecewise linear
-model of the input data. The computational complexity of training a
-multilayer perceptron is $\mathcal{O}(n d m)$, where $m$ is
-determined by the sizes of the layers of the network and the stopping
-criterion of the BFGS minimization used for finding weights. This
-paper uses the scikit-learn MLP regressor \shortcite{scikit-learn}, a
-single hidden layer with 100 nodes, ReLU activation, and BFGS for
-training.
-
-\subsubsection{Support Vector Regressor}
-\vspace{-10pt}
-Support vector machines are a common method used in machine learning
-classification tasks that can be adapted for the purpose of regression
-\shortcite{basak2007support}. How to build a support vector regressor
-(SVR) is beyond the scope of this summary, but the resulting
-functional fit $p : \mathbb{R}^d \rightarrow \mathbb{R}$ has the form
-
-$$ p(x)  = \sum_{i=1}^{n}a_i K(x,x^{(i)}) + b ,$$
-
-%% $$ \text{Minimize } \bigg\{ \frac{1}{2}\sum_{i=1}^{n}\sum_{j=1}^{n}
-%% a_i a_j K(x_i, x_j) + \epsilon \sum_{i=1}^{n} a_i - \sum_{i=1}^{n} y_i
-%% a_i \bigg\} $$
-
-%% $$ \text{Subject to } \sum_{i=1}^{n}a_i = 0 \text{ } \text{ and }
-%% \text{ } a_i \in [0,C] $$
-
-where $K$ is the selected kernel function, $a \in \mathbb{R}^n$, $b
-\in \mathbb{R}$ are coefficients to be solved for simultaneously. The
-computational complexity of the SVR is $\mathcal{O}(n^2dm)$, with $m$
-being determined by the minimization convergence criterion. This paper
-uses the scikit-learn SVR \shortcite{scikit-learn} with a polynomial
-kernel function.
-
-\vspace{-10pt}
-\subsection{Interpolation}
-\vspace{-10pt}
-In some cases it is desirable to have a model that can recreate the
-input data exactly. This is especially the case when the confidence in
-the response values for known data is high. Both interpolation models
-analyzed in this paper are based on linear functions.
-
-\vspace{-10pt}
-\subsubsection{Delaunay}
-\vspace{-10pt}
-The Delaunay method of interpolation is a well studied geometric
-technique for producing an interpolant \shortcite{lee1980two}. The
-Delaunay triangulation of a set of data points into simplices is such
-that the sphere defined by the vertices of each simplex contains no
-data points in the sphere's interior. For a $d$-simplex S with
-vertices $v^{(0)}, v^{(1)}, \ldots, v^{(d)}$, $x \in S$, and data
-values $f(v^{(i)})$, $i=0$,$\ldots$,$d$, $x$ is a unique convex
-combination of the vertices,
-
-$$ x = \sum_{i=0}^{d} w_i v^{(i)}, \quad \sum_{i=0}^{d} w_i = 1, \quad
-w_i \geq 0, \quad i=0,\ldots,d ,$$
-
-and the Delaunay interpolant to $f$ at $x$ is given by
-
-$$ p(x) = \sum_{i=0}^{d} w_i f(v^{(i)}). $$
-
-The computational complexity of the Delaunay triangulation (for the
-implementation used here) is $\mathcal{O}(n^{\lceil d/2 \rceil})$,
-which is not scalable to $d > 10$ \shortcite{sartipizadeh2016computing}.
-The scipy interface \shortcite{scipy} to the QuickHull implementation
-\shortcite{barber1996qhull} of the Delaunay triangulation is used here.
-
-\subsubsection{Linear Shepard}
-\vspace{-10pt}
-The linear Shepard method (LSHEP) is a blending function using local
-linear interpolants, a special case of the general Shepard algorithm
-\shortcite{thacker2010algorithm}. The interpolant has the form
-
-$$ p(x) = \frac{\sum_{k=1}^{n}W_k(x)P_k(x)}{\sum_{k=1}^{n}W_k(x)} ,$$
-
-where $W_k(x)$ is a locally supported weighting function and $P_k(x)$
-is a local linear approximation to the data satisfying
-$P_k\big(x^{(k)}\big) = f\big(x^{(x)}\big)$. The computational
-complexity of LSHEP is $\mathcal{O}(n^2d^3)$. This paper uses the
-Fortran\#95 implementation of LSHEP in SHEPPACK
-\shortcite{thacker2010algorithm}.
-
-%% %     Related Work     
-%% %======================
-%% \section{Related Work}
-%% \begin{enumerate}
-%% \item Not sure how much to include here? Shooting for thoroughness or
-%%   simply necessary coverage? How much background should I expect the
-%%   readers of this paper to have in the ``multivariate modeling of
-%%   systems'' area?
-%% \end{enumerate}
-
-%     Methodology     
-%=====================
-\section{Methodology}
-\label{sec:methodology}
-\subsection{Data}
-\vspace{-10pt}
-In order to evaluate the viability of multivariate models for
-predicting system performance, this paper presents a case study of a
-four-dimensional dataset produced by executing the IOzone benchmark
-from \shortciteN{iozone} on a homogeneous cluster of computers. The
-system performance data was collected by executing IOzone 40 times for
-each of a select set of system configurations. A single IOzone
-execution reports the max I/O throughput seen for the selected
-test. The 40 executions for each system configuration are converted
-into the mean and variance, both values in $\mathbb{R}$ capable of
-being modeled individually by the multivariate approximation
-techniques presented in Section \ref{sec:multivariate}. The summary of
-data used in the experiments for this paper can be seen in Table
-\ref{tab:data_type}.  Distributions of raw throughput values being
-modeled can be seen in Figure \ref{fig:raw_throughput}.
-
-\begin{table}
-  \centering
-  \begin{tabular}{c|c}
-    \hline
-    \textbf{System Parameter} & \textbf{Values}\\
-    \hline
-    File Size & 64, 256, 1024\\
-    Record Size & 32, 128, 512\\
-    Thread Count & 1, 2, 4, 8, 16, 32, 64, 128, 256\\
-    Frequency & \{12, 14, 15, 16, 18, 19, 20, 21, 23, 24, 25, 27, 28, 29, 30, 30.01\} $\times 10^5$\\
-    \hline
-    \textbf{Response Values} & \\
-    \hline
-    Throughput Mean & [$2.6 \times 10^5$, $5.9 \times 10^8$]\\
-    Throughput Variance & [$5.9\times 10^{10} $, $4.7 \times 10^{16}$]\\
-    \hline
-  \end{tabular}
-  \caption{A description of the system parameters being considered in
-    the IOzone tests. Record size must not be greater than file size
-    and hence there are only six valid combinations of the two. In
-    total there are $6 \times 9 \times 16 = 864$ unique system
-    configurations.}
-  \label{tab:data_type}
-\end{table}
-
-\begin{figure}
-  \centering
-  \includegraphics[width=\textwidth,trim={0 .5in 0 .4in}]{Raw_Throughput.pdf}
-  \caption{Histograms of 100-bin reductions of the PMF of I/O
-    throughput mean (top) and I/O throughput variance (bottom). In the
-    mean plot, the first 1\% bin (truncated in plot) has a probability
-    mass of .45. In the variance plot, the second 1\% bin has a
-    probability mass of .58. It can be seen that the distributions of
-    throughputs are primarily of lower magnitude with occasional
-    extreme outliers.}
-  \label{fig:raw_throughput}
-\end{figure}
-
-\vspace{-10pt}
-\subsection{Dimensional Analysis}
-\vspace{-10pt}
-This work utilizes an extension to standard $k$-fold cross validation
-that allows for a more thorough investigation of the expected model
-performance in a variety of real-world situations. Alongside
-randomized splits, two extra components are considered: the amount of
-training data provided, and the dimension of the input data. It is
-important to consider that algorithms that perform well with less
-training input also require less experimentation. Although, the amount
-of training data required may change as a function of the dimension of
-the input and this needs to be studied as well. The framework used
-here will be referred to as a multidimensional analysis (MDA) of the
-IOzone data.
-
-%% \color{red}
-
-\subsubsection{Multidimensional Analysis}
-\vspace{-10pt}
-This procedure combines random selection of training and testing
-splits with changes in the input dimension and the ratio of training
-size to testing size. Given an input data matrix with $n$ rows
-(points) and $d$ columns (components), MDA proceeds as follows:
-\begin{enumerate}
-\item For all $k = 1$, $\ldots$, $d$ and for all nonempty subsets $F
-  \subset \{1, 2, \ldots, d\}$, reduce the input data to points $(z,
-  f_F(z))$ with $z \in \mathbb{R}^k$ and $f_F(z) = E[ \{ f(x^{(i)})
-    \mid (x^{(i)}_F = z) \} ]$, where $E[\cdot]$ denotes the mean and
-  $x^{(i)}_F$ is the subvector of $x^{(i)}$ indexed by $F$.
-\item For all $r$ in $\{5, 10, \ldots, 95\}$, generate $N$ random
-  splits $(train, test)$ of the reduced data with $r$ percentage for
-  training and $100 - r$ percentage for testing.
-\item When generating each of $N$ random $(train, test)$ splits,
-  ensure that all points from $test$ are in the convex hull of points
-  in $train$ (to prevent extrapolation); also ensure that the points
-  in $train$ are well spaced.
-\end{enumerate}
-
-In order to ensure that the testing points are in the convex hull of
-the training points, the convex hull vertices of each set of (reduced
-dimension) points are forcibly placed in the training set. In order to
-ensure that training points are well spaced, a statistical method for
-picking points from \shortciteN{amos2014algorithm} is used:
-\begin{enumerate}
-\item Generate a sequence of all pairs of points
-  ${(z^{(i_1)},z^{(j_1)}), (z^{(i_2)},z^{(j_2)}), \ldots}$ sorted by
-  ascending pairwise Euclidean distance between points, so that
-  $||z^{(i_k)}-z^{(j_k)}||_2 \leq ||z^{(i_{k+1})}-z^{(j_{k+1})}||_2$.
-\item Sequentially remove points from candidacy until only $|train|$
-  remain by randomly selecting one point from the pair $(z^{(i_m)},
-  z^{(j_m)})$ for $m = 1,\ldots$ if both $z^{(i_m)}$ and $z^{(j_m)}$
-  are still candidates for removal.
-\end{enumerate}
-
-Given the large number of constraints, level of reduction, and use of
-randomness in the MDA procedure, occasionally $N$ unique training /
-testing splits may not be created or may not exist. In these cases, if
-there are fewer than $N$ possible splits, then deterministically
-generated splits are used. Otherwise after $3N$ attempts, only the
-unique splits are kept for analysis. The MDA procedure has been
-implemented in Python\#3 while most regression and interpolation
-methods are Fortran wrapped with Python. All randomness has been
-seeded for repeatability.
-
-For any index subset $F$ (of size $k$) and selected value of $r$, MDA
-will generate up to $N$ multivariate models $f_F(z)$ and predictions
-$\hat{f}_F\big(z^{(i)}\big)$ for a point $z^{(i)} \in \mathbb{R}^k$.
-There may be fewer than $N$ predictions made for any given
-point. Extreme points of the convex hull for the selected index subset
-will always be used for training, never for testing. Points that do
-not have any close neighbors will often be used for training in order
-to ensure well-spacing. Finally, as mentioned before, some index
-subsets do not readily generate $N$ unique training and testing
-splits. The summary results presented in this work use the median of
-the ($N$ or fewer) values $\hat{f}_F(z)$ at each point as the model
-estimate for error analysis.
-
-%% is open source, and is freely available at the url
-%% \textit{https://github.com/tchlux/VarSys/HPC\_Paper/Code/multi\_dim\_analysis.py}
-
-%% \begin{enumerate}
-%% \item Cycling the categorical settings
-%% \item Selecting subsets of 1,2,3 up to 4 dimensions
-%% \item Cycling different training : testing ratios (5:95 $\rightarrow$ 95:5)
-%% \item Generating 200 random training : testing splits
-%% \item Ensuring the testing points are on/inside the convex hull of the training.
-%% \item Ensuring the training points are well-spaced.
-%% \end{enumerate}
-
-%% \subsection{Prediction}
-%% \begin{enumerate}
-%% \item For each file generated from the dimensional analysis, train on
-%%   the training data, evaluate at the testing data points
-%% \end{enumerate}
-
-%     Results     
-%=================
-\section{Results}
-\label{sec:results}
-
-A na\"{\i}ve multivariate prediction technique such as nearest
-neighbor could experience relative errors in the range $\displaystyle
-[0, \big(\max_x f(x) - \min_x f(x)\big) / \min_x f(x) ]$ when modeling
-a nonnegative function $f(x)$ from data. The IOzone mean data response
-values span three orders of magnitude (as can be seen in Table
-\ref{tab:data_type}) while variance data response values span six
-orders of magnitude. It is expected therefore, that all studied
-multivariate models perform better than a na\"{\i}ve approach,
-achieving relative errors strictly less than $10^3$ for throughput
-mean and $10^6$ for throughput sample variance. Ideally, models will
-yield relative errors significantly smaller than 1.
-
-\vspace{-10pt}
-\subsection{I/O Throughput Mean}
-\vspace{-10pt}
-Almost all multivariate models analyzed make predictions with a
-relative error less than 1 for most system configurations when
-predicting the mean I/O throughput of a system given varying amounts
-of training data. The overall best of the multivariate models,
-Delaunay, consistently makes predictions with relative error less than
-$.05$ (5\% error). In Figure \ref{fig:mean_tt_ratio} it can also be
-seen that the Delaunay model consistently makes good predictions even
-with as low as 5\% training data (43 of the 864 system configurations)
-regardless of the dimension of the data.
-
-\begin{figure}
-  \centering
-  \includegraphics[width=\textwidth,trim={0 .5in 0 .3in}]{Mean_Dim.pdf}
-  \caption{These box plots show the prediction error of mean with
-    increasing dimension. The top box whisker for SVR is 40, 80, 90
-    for dimensions 2, 3, and 4, respectively. Notice that each model
-    consistently experiences greater magnitude error with increasing
-    dimension. Results for all training percentages are aggregated.}
-  \label{fig:mean_dim}
-\end{figure}
-
-\begin{figure}
-  \centering
-  \includegraphics[width=\textwidth,trim={0 .5in 0 .3in}]{Mean_TT_Ratio.pdf}
-  \caption{These box plots show the prediction error of mean with
-    increasing amounts of training data provided to the models. Notice
-    that MARS is the only model whose primary spread of performance
-    increases with more training data. Recall that the response values
-    being predicted span three orders of magnitude and hence relative
-    errors should certainly remain within that range. For SVR the top
-    box whisker goes from around 100 to 50 from left to right and is
-    truncated in order to maintain focus on better models. Results for
-    all dimensions are aggregated. Max training percentage is 96\% due
-    to rounding training set size.}
-  \label{fig:mean_tt_ratio}
-\end{figure}
-
-\begin{figure}
-  \centering
-  \includegraphics[width=\textwidth,trim={0 .5in 0 .3in}]{Var_TT_Ratio.pdf}
-  \caption{These box plots show the prediction error of variance with
-    increasing amounts of training data provided to the models. The
-    response values being predicted span six orders of magnitude. For
-    SVR the top box whisker goes from around 6000 to 400 (decreasing
-    by factors of 2) from left to right and is truncated in order to
-    maintain focus on better models. Results for all dimensions are
-    aggregated. Max training percentage is 96\% due to rounding
-    training set size.}
-  \label{fig:var_tt_ratio}
-\end{figure}
-
-\vspace{-10pt}
-\subsection{I/O Throughput Variance}
-\vspace{-10pt} The prediction results for variance resemble those for
-predicting mean. Delaunay remains the best overall predictor
-(aggregated across training percentages and dimensions) with median
-relative error of .47 and LSHEP closely competes with Delaunay having
-a median signed relative error of -.92. Outliers in prediction error
-are much larger for all models. Delaunay produces relative errors as
-large as 78 and other models achieve relative errors around
-$10^3$. The relative errors for many models scaled proportional to the
-increased orders of magnitude spanned by the variance response
-compared with mean response. As can be seen in Figure
-\ref{fig:var_tt_ratio}, all models are more sensitive to the amount of
-training data provided than their counterparts for predicting mean.
-
-\vspace{-10pt}
-\subsection{Increasing Dimension and Decreasing Training Data}
-\vspace{-10pt}
-As can be seen in Figure \ref{fig:mean_dim}, all of the models suffer
-increasing error rates in higher dimension. This is expected, because
-the number of possible interactions to model grows
-exponentially. However, LSHEP and Delaunay maintain the slowest
-increase in relative error. The increase in error seen for Delaunay
-suggests that it is capable of making predictions with a range of
-relative errors that grows approximately linearly with increasing
-dimension input. This trend suggests that Delaunay would remain a
-viable technique for accurately modeling systems with 10's of
-parameters given only small amounts of training data. All models, with
-the exception of MARS, produce smaller errors given more training
-data. Increasing the amount of training data most notably reduces the
-number of prediction error outliers.
-
-%     Discussion     
-%====================
-\section{Discussion}
-\label{sec:discussion}
-
-The present results demonstrate that a straightforward application of
-multivariate modeling techniques can be used to effectively predict
-HPC system performance. Some modeling effort on the part of a systems
-engineer combined with a significant amount of experimentation (weeks
-of CPU time for the IOzone data used here) can yield a model capable
-of accurately tuning an HPC system to the desired performance
-specification, although qualitatively correct predictions can be
-achieved with mush less (10\%, say) effort.
-
-\vspace{-10pt}
-\subsection{Modeling the System}
-\vspace{-10pt}
-The modeling techniques generated estimates of drastically different
-quality when predicting I/O throughput mean and variance. A few
-observations: SVR has the largest range of errors for all selections
-of dimension and amounts of training data; MARS and LSHEP produce
-similar magnitude errors while the former consistently underestimates
-and the latter consistently overestimates; Delaunay has considerably
-fewer outliers than all other methods. SVR likely produces the poorest
-quality predictions because the underlying parametric representation
-is global and oversimplified (a single polynomial), making it unable
-to capture the complex local behaviors of system I/O. It is still
-unclear, however, what causes the behaviors of LSHEP, MARS, and
-Delaunay. An exploration of this topic is left to future work.
-
-While the Delaunay method appears to be the best predictor in the
-present IOzone case study, it is important to note that the Delaunay
-computational complexity increases with the dimension of the input
-more rapidly than other techniques. The implementation of Delaunay
-(QuickHull) used would experience unacceptably large training times
-beyond ten-dimensional input. This leaves much room for other
-techniques to perform best in higher dimension unless a more efficient
-implementation of Delaunay can be used.
-
-Finally, the ability of the models to predict variance was
-significantly worse than for the I/O mean. The larger scale in
-variance responses alone do not account for the increase in relative
-errors witnessed. This suggests that system variability has a greater
-underlying functional complexity than the system mean and that latent
-factors are reducing prediction performance.
-
-\vspace{-10pt}
-\subsection{Extending the Analysis}
-\vspace{-10pt}
-System I/O throughput mean and variance are simple and useful system
-characteristics to model. The process presented in this work is
-equally applicable to predicting other useful performance
-characteristics of HPC systems such as: computational throughput,
-power consumption, processor idle time, context switches, RAM usage,
-or any other ordinal performance metric. For each of these there is
-the potential to model system variability as well. This work has
-chosen variance as a measure of variability, but the techniques used
-in this paper could be applied to more precise measures of variability
-such as the percentiles of the performance distribution or the entire
-distribution itself. A thorough exploration of HPC systems
-applications of multivariate modeling constitutes future work.
-
-%% \begin{enumerate}
-%% \item Measuring different system characteristics other than I/O
-%% \item Recording system performance constantly to maintain model accuracy
-%% \end{enumerate}
-
-\section{Conclusion}
-\label{sec:conclusion}
-Multivariate models of HPC system performance can effectively predict
-I/O throughput mean and variance. These multivariate techniques
-significantly expand the scope and portability of statistical models
-for predicting computer system performance over previous work. In the
-IOzone case study presented, the Delaunay method produces the best
-overall results making predictions for 821 system configurations with
-less than 5\% error when trained on only 43 configurations. Analysis
-also suggests that the error in the Delaunay method will remain
-acceptable as the number of system parameters being modeled
-increases. These multivariate techniques can and should be applied to
-HPC systems with more than four tunable parameters in order to
-identify optimal system configurations that may not be discoverable
-via previous methods nor by manual performance tuning.
-
-%     Future Work     
-%=====================
-\vspace{-10pt}
-\subsection{Future Work}
-\vspace{-10pt}
-The most severe limitation to the present work is the restriction to
-modeling strictly ordinal (not categorical) system
-parameters. Existing statistical approaches for including categorical
-variables are inadequate for nonlinear interactions in high
-dimensions. Future work could attempt to identify the viability of
-different techniques for making predictions including categorical
-system parameters.
-
-There remain many other multivariate modeling techniques not included
-in this work that should be evaluated and applied to HPC performance
-prediction. For I/O alone, there are far more than the four tunable
-parameters studied in this work. Alongside experimentation with more
-models, there is room for a theoretical characterization of the
-combined model and data properties that allow for the greatest
-predictive power.
-
-%% \color{black}
-
-%% \newpage
-%% \ \\
-
-%% \vspace{3in}
-
-%% \texttt{\textbf{L}oc\textbf{A}ll\textbf{Y} li\textbf{NE}ar
-%%   \textbf{W}eighted \textbf{A}pproxima\textbf{T}ion\textbf{S}
-%%   \textbf{O}f fu\textbf{N}tions}\\
-%% %% \texttt{L\ \ A\ \ Y \ \ NE\ \ \  W\ \ \ \ \ \ \ \ A\ \ \ \ \ \ \ \ T\ \ \ S O\ \ \ \ N\ \ \ \ \ \ }
-%% \newpage
-
-\bibliographystyle{scsproc}
-\bibliography{paper}
-
-\end{document}
+\section{Abstract of Scalable}
+A rapid increase in the quantity of data available is allowing all fields of science to generate more accurate models of multivariate phenomena.
+Regression and interpolation become challenging when the dimension of data is large, especially while maintaining tractable computational complexity.
+This paper proposes three novel techniques for multivariate interpolation and regression that each have polynomial complexity with respect to number of instances (points) and number of attributes (dimension).
+Initial results suggest that these techniques are capable of effectively modeling multivariate phenomena while maintaining flexibility in different application domains.

### 2020-04-14 3-Disseration/apps.tex
Updated min wiggle fit tests, wrote report on the topic that was sent to Dr. Watson.

--- 2019-03-30 Paper_[2019-02]_Numerical_Algorithms/paper.tex
@@
-\RequirePackage{fix-cm}
-\documentclass[smallextended,final]{svjour3}  % onecolumn (second format)
-\smartqed  % flush right qed marks, e.g. at end of proof
+Computational variability presents itself in a variety of forms.
+Processes that are apparently identical in a cloud computing or high performance computing (HPC) environment may take different amounts of time to complete the same job.
+This variability can cause unintentional violations of service level agreements in cloud computing applications or indicate suboptimal performance in HPC applications.
+The sources of variability, however, are distributed throughout the system stack and often difficult to identify.
+The methodology presented in this work is applicable to modeling the expected variability of useful computer system performance measures without any prior knowledge of system architecture.
+Some examples of interesting performance measures that could be modeled with the techniques in this work include computational throughput, power consumption, processor idle time, number of context switches, and RAM usage, as well as any other numeric measure of performance.
 
-% Insert here the call for the packages your document requires
-\usepackage{graphicx}
-\usepackage{tikz}     % For adding axes to plots
-\usetikzlibrary{positioning}
-\usepackage{graphicx} % For including PDFs
-\usepackage{amsmath}  % assumes amsmath package installed
-\usepackage{amssymb}  % assumes amsmath package installed
-\usepackage{booktabs} % For formal tables
-\usepackage{tabularx} % For tables with nice column widths.
-\usepackage[colorlinks,allcolors=blue]{hyperref}
-\usepackage{float}    % For controlling the placement of tables.
+Predicting performance variability in a computer system is a challenging problem that has primarily been attempted in one of two ways: (1) build a statistical model of the performance data collected by running experiments on the system at select settings, or (2) run artificial experiments using a simplified simulation of the target system to estimate architecture and application bottlenecks.
+In this work, modeling techniques rest in the first category and represent a notable increase in the ability to model precise characteristics of variability.
 
-%
-% please place your own definitions here and don't use \def but
-% \newcommand{}{}
+Many previous works attempting to model system performance have used simulated environments \cite{grobelny2007fase,wang2009simulation,wang2013towards}.
+\citet{grobelny2007fase} refer to statistical models as being oversimplified and not capable of capturing the true complexity of the underlying system.
+Historically, statistical models have been limited to modelling at most one or two system parameters and have therefore not been capable of modeling the complexity of the underlying system.
+\cite{snavely2002framework,bailey2005performance,barker2009using,ye2010analyzing}.
+These limited statistical models have provided satisfactory performance in narrow application settings.
+However these techniques require additional code annotations, hardware abstractions, or additional assumptions about the behavior of the application in order to generate models.
+In contrast, the approaches that are presented here require no modifications of applications, no architectural abstractions, nor any structural descriptions of the input data being modeled.
+The techniques used in this work only require performance data as input.
 
-%% Command for creating multiplie-line cells inside a table environment
-\newcommand{\multilinecell}[2][c]{%
-  \begin{tabular}[#1]{@{}c@{}}#2\end{tabular}}
+Most existing work on performance variability has focused on operating system (OS) induced variability \cite{beckman2008benchmarking,de2007identifying}.
+Yet, system I/O variability has been particularly difficult for statistical models to capture \cite{bailey2005performance}.
+The prior work attempting to model I/O variability has been similarly limited to one or two system parameters \cite{lofstead2010managing}.
 
-\newcolumntype{d}[1]{D{.}{.}{#1}}
+Chapters \ref{ch:naive} and \ref{ch:strong} present and evaluate applications of multivariate interpolation to the domain of computer system I/O throughput.
+Interpolants and regressors are used to predict the different measures of variability (e.g., mean, variance, cumulative distribution functions) for the expected I/O throughput on a system with previously unseen configurations.
+Beyond these I/O case studies, the techniques discussed can tractably model tens of interacting system parameters with tens of thousands of known configurations.
+A major contribution of this work is a modeling framework that uses multivariate interpolation to capture precise characteristics (via cumulative distribution functions) of arbitrary performance measure on any type of computer system.
 
-%% Create a ``Lemma'' environment that does not have italics font.
-\spnewtheorem{plainlemma}{Lemma}{\bf}{}
-%% Create a ``Theorem'' environment that is not numbered.
-\spnewtheorem*{onetheorem}{Theorem}{\bf}{}
-%% Create a ``Proof'' environment that starts with 'Proof.'
-\spnewtheorem*{proofdot}{Proof.}{\it}{}
-
-%% \newcommand{\Bigl}{\mathopen\Big}
-%% \newcommand{\Bigr}{\mathclose\Big}
-%
-% Insert the name of "your journal" with
-\journalname{Numerical Algorithms}
-%
-\begin{document}
-
-\title{Interpolation of Sparse High-Dimensional Data\thanks
-  {This work was supported by the National Science Foundation Grant CNS-1565314.}
-}
-% \subtitle{Do you have a subtitle?\\ If so, write it here}
-%\titlerunning{Short form of title}        % if too long for running head
-
-\author{Thomas C. H. Lux   \and
-  Layne T. Watson          \and
-  Tyler H. Chang           \and
-  Yili Hong                \and
-  Kirk Cameron
-}
-
-%\authorrunning{Short form of author list} % if too long for running head
-
-\institute{Thomas Lux \at
-  Virginia Polytechnic Institute \& State University (VPI\&SU) \\
-  Blacksburg, VA 24061 \\
-  \email{tchlux@vt.edu}
-}
-
-\date{Received: date / Accepted: date}
-% The correct dates will be entered by the editor
-
-\maketitle
-
-\begin{abstract}
-Increases in the quantity of available data have allowed all fields of
-science to generate more accurate models of multivariate phenomena.
-Regression and interpolation become challenging when the dimension of
-data is large, especially while maintaining tractable computational
-complexity. Regression is a popular approach to solving approximation
-problems with high dimension, however there are often advantages to
-interpolation. This paper presents a novel and insightful error bound
-for (piecewise) linear interpolation in arbitrary dimension and
-contrasts the performance of some interpolation techniques with
-popular regression techniques. Empirical results demonstrate the
-viability of interpolation for moderately high dimensional
-approximation problems, and encourage broader application of
-interpolants to multivariate approximation in science.
-
-\keywords{approximation \and regression \and interpolation \and high
-  dimension \and error bound}
-% \PACS{PACS code1 \and PACS code2 \and more}
-% \subclass{MSC code1 \and MSC code2 \and more}
-\end{abstract}
-
-%     Introduction     
-%======================
-
-%% ===================================================================
-\section{Introduction}
-\label{sec:introduction}
+\section{Broader Applications of Approximation}
 
@@
 construction of strong error bounds that are absent from high
-dimensional problems. This work extends some known results regarding
-the secant method \cite{dennis1996numerical} to construct an
-interpolation error bound for problems of arbitrary dimension. These
-error bounds are useful, considering the same cannot be said for
-regression algorithms in general. The maximum complexity of an
-interpolant is bounded by the amount of data available, while the
-maximum complexity of a regressor is bounded by both the amount of
+dimensional problems. Chapter \ref{ch:error} extends some known
+results regarding the secant method \cite{dennis1996numerical} to
+construct an interpolation error bound for problems of arbitrary
+dimension. These error bounds are useful, considering the same cannot
+be said for regression algorithms in general. The maximum complexity
+of an interpolant is bounded by the amount of data available, while
+the maximum complexity of a regressor is bounded by both the amount of
 data and the chosen parametric form. For this reason, generic uniform
 bounds are largely unobtainable for regression techniques on arbitrary
-approximation problems, even when the approximation domain is bounded.
+approximation problems, even when the approximation domain is
+bounded. These generic bounds imply that interpolants may be suitable
+for a broader class of approximation problems than (heuristically
+chosen) regression techniques.
 
@@
 
-The remainder of this paper is organized as follows. Sections
-\ref{sec:regression} and \ref{sec:interpolation} present the
-multivariate models. Section \ref{sec:error} gives the error measuring
-methodology that is used for collecting empirical results. Section
-\ref{sec:theory} contains the theoretical error bounds for
-interpolation. Section \ref{sec:data} presents the data sets for
-empirical approximation analysis and the approximation results for all
-models on these data sets. Section \ref{sec:discussion} analyzes and
-discusses the results, their implications, and their
-limitations. Finally, Section \ref{sec:conclusion} concludes.
+\section{An Initial Application of Approximation}
 
-%% ===================================================================
-\section{Multivariate Regression}
-\label{sec:regression}
-Multivariate regressors are capable of accurately modeling a complex
-dependence of a response (in $Y$) on multiple variables (represented
-as a points in $\mathbb{R}^{d}$). The approximations to some (unknown)
-underlying function $f: \mathbb{R}^d \rightarrow Y$ are chosen to
-minimize some error measure related to data samples
-$f\bigl(x^{(i)}\bigr)$. For example, least squares regression uses the
-sum of squared differences between modeled function values and true
-function values as an error measure. In this section and the next,
-some techniques are limited to approximating real valued functions ($Y
-\subset \mathbb{R}$). These techniques can be extended to real
-vector-valued ranges by repeating the construction for each component
-of the vector output. Throughout the following, $x$ denotes a
-$d$-tuple, $x_i$ the $i$th component of $x$, and $x^{(i)}$ the $i$th
-$d$-tuple data point. Different symbols are used to represent the
-approximation function $\hat f$.
+Performance tuning is often an experimentally complex and time
+intensive chore necessary for configuring High Performance Computing
+(HPC) systems. The procedures for this tuning vary largely from system
+to system and are often subjectively guided by the system
+engineer(s). Once a desired level of performance is achieved, an HPC
+system may only be incrementally reconfigured as required by updates
+or specific jobs. In the case that a system has changing workloads or
+nonstationary performance objectives that range from maximizing
+computational throughput to minimizing power consumption and system
+variability, it becomes clear that a more effective and automated tool
+is needed for configuring systems. This scenario presents a
+challenging and important application of multivariate approximation
+and interpolation techniques.
 
-\subsection{Multivariate Adaptive Regression Splines}
-\label{sec:mars}
-This approximation was introduced in \cite{friedman1991multivariate}
-and subsequently improved to its current version in
-\cite{stanford1993fast}, called fast multivariate adaptive regression
-splines (Fast MARS). In Fast MARS, a least squares fit model is
-iteratively built by beginning with a single constant valued function
-and adding two new basis functions at each iteration of the form
-\begin{align*}
-  B_{2j-1}(x) &= B_l(x) \bigl(x_i-x^{(p)}_i\bigr)_+, \\
-  B_{2j}(x) &= B_k(x) \bigl(x_i-x^{(p)}_i\bigr)_- ,
-\end{align*}
-where $j$ is the iteration number, $1 \le p \le n$, and $B_l(x)$,
-$B_k(x)$ are basis functions from the previous iteration,
- $$w_+ = \begin{cases} w, & w \geq 0 \\ 0, & w < 0 \end{cases},$$
-and $w_- = (-w)_+$. After iteratively constructing a model, MARS then
-iteratively removes basis functions that do not contribute to goodness
-of fit. In effect, MARS creates a locally component-wise tensor
-product approximation of the data. The overall computational
-complexity of Fast MARS is $\mathcal{O}(n d m^3)$ where $m$ is the
-maximum number of underlying basis functions. This paper uses an
-implementation of Fast MARS \cite{rudy2017pyearth} with $m = 200$
-throughout all experiments.
+Predicting the performance of an HPC system is a challenging problem
+that is primarily attempted in one of two ways: (1) build a
+statistical model of the performance by running experiments on the
+system at select settings, or (2) run artificial experiments using a
+simplified simulation of the target system to estimate architecture
+and application bottlenecks. In this paper the proposed multivariate
+modeling techniques rest in the first category, and they represent a
+notable increase in the ability to model complex interactions between
+system parameters.
 
-\subsection{Support Vector Regressor}
-\label{sec:svr}
-Support vector machines are a common method used in machine learning
-classification tasks that can be adapted for the purpose of regression
-\cite{basak2007support}. How to build a support vector regressor (SVR)
-is beyond the scope of this summary, but the resulting functional fit
-$p : \mathbb{R}^d \rightarrow \mathbb{R}$ has the form
- $$ p(x)  = \sum_{i=1}^{n}a_i K\bigl(x,x^{(i)}\bigr) + b ,$$
-where $K$ is the selected kernel function, $a_i \in \mathbb{R}^n$, 
-$b \in \mathbb{R}$ are coefficients to be solved for simultaneously.
-The computational complexity of the SVR is $\mathcal{O}(n^2dm)$, with
-$m$ being determined by the minimization convergence criterion. This
-paper uses the scikit-learn SVR \cite{scikit-learn} with a polynomial
-kernel function.
+Many previous works attempting to model system performance have used
+simulated environments to estimate the performance of a system
+\cite{grobelny2007fase,wang2009simulation,wang2013towards}. Some
+of these works refer to statistical models as being oversimplified and
+not capable of capturing the true complexity of the underlying
+system. This claim is partially correct, noting that a large portion
+of predictive statistical models rely on simplifying the model to one
+or two parameters
+\cite{snavely2002framework,bailey2005performance,barker2009using,ye2010analyzing}.
+These limited statistical models have provided satisfactory
+performance in very narrow application settings. Many of the
+aforementioned statistical modeling techniques claim to generalize,
+while simultaneously requiring additional code annotations, hardware
+abstractions, or additional application level understandings in order
+to generate models. The approach presented here requires no
+modifications of the application, no architectural abstractions, nor
+any structural descriptions of the input data being modeled. The
+techniques used are purely mathematical and only need performance data
+as input.
 
-\subsection{Multilayer Perceptron Regressor}
-\label{sec:mlp}
-The neural network is a well studied and widely used method for both
-regression and classification tasks
-\cite{rumelhart1988learning,hornik1989multilayer}. When using the
-rectified linear unit (ReLU) activation function
-\cite{dahl2013improving} and fitting the model with stochastic
-gradient descent (SGD) or BFGS minimization techniques
-\cite{goh2017why,moller1993scaled,robbins1951stochastic}, the model
-built by a multilayer perceptron uses layers $l : \mathbb{R}^{i}
-\rightarrow \mathbb{R}^{j}$ defined by
- $$ l(u) = \bigl( u^t W_l + c_l \bigr)_+ ,$$
-where $c_l \in \mathbb{R}^j$ and $W_l$ is the $i \times j$ weight
-matrix for layer $l$. In this form, the multilayer perceptron (MLP)
-produces a piecewise linear model of the input data. The computational
-complexity of training a multilayer perceptron is $\mathcal{O}(n d
-m)$, where $m$ is determined by the sizes of the layers of the network
-and the stopping criterion of the minimization used for finding
-weights. This paper uses an MLP built from Keras and Tensorflow to
-perform regression \cite{tensorflow2015-whitepaper,chollet2015keras}
-with ten hidden layers each having thirty two nodes (a total of
-approximately ten thousand parameters), ReLU activation, and five
-thousand epochs of SGD for training.
-
-%% ===================================================================
-\section{Multivariate Interpolation}
-\label{sec:interpolation}
-The following interpolation techniques demonstrate a reasonable
-variety of approaches to interpolation. All of the presented
-interpolants produce approximations that are continuous in value,
-which is often a desirable property in applied approximation problems.
-
-\subsection{Delaunay}
-\label{sec:delaunay}
-
-The Delaunay triangulation is a well-studied geometric technique for
-producing an interpolant \cite{lee1980two}. The Delaunay triangulation
-of a set of data points into simplices is such that there are no data
-points inside the sphere defined by the vertices of each simplex. For
-a $d$-simplex S with vertices $x^{(0)}$, $x^{(1)}$, $\ldots$,
-$x^{(d)}$, and function values $f\bigl(x^{(i)}\bigr)$, $i=0$,
-$\ldots$, $d$, $y \in S$ is a unique convex combination of the
-vertices,
- $$ y = \sum_{i=0}^{d} w_i x^{(i)}, \quad \sum_{i=0}^{d} w_i = 1, \quad w_i \geq 0, \quad i=0,\ldots,d, $$
-and the Delaunay interpolant $\hat f(y)$ at $y$ is given by
- $$ \hat f(y) = \sum_{i=0}^{d} w_i f\bigl(x^{(i)}\bigr). $$ The
-computational complexity of Delaunay interpolation (for the
-implementation used \cite{chang2018polynomial}) is $\mathcal{O}(k n
-d^3)$. Given pathological data the entire triangulation could be
-computed with $k = n^{\lceil d / 2 \rceil}$, but the average case yields
-$k = d \log d$ and is capable of scaling reasonably to $d \leq 50$. In
-the present application, a Delaunay simplex $S$ containing $y$ is
-found, then the $d+1$ vertices (points in $X$) of $S$ are used to
-assign weights to each vertex and produce the predicted function value
-for point $y$.
-
-\subsection{Modified Shepard}
-\label{sec:modified-shepard}
-
-The modified Shepard method used here (ShepMod) generates a continuous
-approximation based on Euclidean distance and resembles a nearest
-neighbor interpolant \cite{cover1967nearest}. This model is a type of
-\textit{metric interpolation}, also called a Shepard method
-\cite{shepard1968two,gordon1978shepard}. The interpolant has the form
- $$ p(x) = \frac{\sum\limits_{k=1}^{n}W_k(x)f\bigl(x^{(k)}\bigr)}
-     {\sum\limits_{k=1}^{n}W_k(x)} ,$$
-where $W_k(x) = \bigl(r_k - \bigl\|x - x^{(k)}\bigr\|\bigr)_+^2 \big/
-\bigl\|x - x^{(k)}\bigr\|_2^2$ and $r_k \in \mathbb{R}$ is the
-smallest radius such that at least $d+1$ other points $x^{(j)}$, $j
-\not = k$, are inside the closed Euclidean ball of radius $r_k$ about
-$x^{(k)}$. The computational complexity of ShepMod is
-$\mathcal{O}(n^2d)$. This paper uses a Fortran 95 implementation of
-ShepMod derived from SHEPPACK \cite{thacker2010algorithm}.
-
-\subsection{Linear Shepard}
-The linear Shepard method (LSHEP) is a blending function using local
-linear interpolants, a special case of the general Shepard algorithm
-\cite{thacker2010algorithm}. The interpolant has the form
- $$ p(x) = \frac{\sum\limits_{k=1}^{n}W_k(x)P_k(x)}
- {\sum\limits_{k=1}^{n}W_k(x)} ,$$
-where $W_k(x)$ is a locally supported weighting function and $P_k(x)$
-is a local linear approximation to the data satisfying
-$P_k\bigl(x^{(k)}\bigr) = f\bigl(x^{(x)}\bigr)$. The computational
-complexity of LSHEP is $\mathcal{O}(n^2d^3)$. This paper uses the
-Fortran 95 implementation of LSHEP in SHEPPACK
-\cite{thacker2010algorithm}.
-
-\subsection{Box Splines}
-\label{sec:box-splines}
-
-The box spline model used here is an interpolation technique built
-from overlapping box splines \cite{de2013box}. The box splines serve
-as basis functions that are shifted and scaled to have support over
-box shaped regions. The boxes are constructed in a way to guarantee a
-covering of the domain \cite{lux2018novel}. Given a set of box splines
-$\bigl\{b^{x^{(i)}}\bigr\}$ with the iterative box properties outlined
-in \cite{lux2018novel} and anchored at interior points
-$\bigl\{x^{(i)}\bigr\}$,
- $$ \hat f(y) = \frac{\sum\limits_{i=1}^n b^{x^{(i)}}(y) f\bigl(x^{(i)}\bigr)}{\sum\limits_{i=1}^n b^{x^{(i)}}(y)}. $$
-Note that the box splines always satisfy
-$b^{x^{(i)}}\bigl(x^{(j)}\bigr) = \delta_{ij}$ and $b^{x^{(i)}}(y)
-\geq 0$. The computational complexity of interpolation via the box
-spline model is $\mathcal{O}(n^2 d)$.
-
-\subsection{Voronoi}
-\label{sec:voronoi}
-
-The nearest neighbor algorithm \cite{cover1967nearest} is a
-well-studied technique for classification and approximation. Nearest
-neighbor inherently utilizes the convex region $C^{x^{(i)}}$ (Voronoi
-cell \cite{dirichlet1850reduction}) consisting of all points closer to
-$x^{(i)}$ than to any other point $x^{(j)}$. The Voronoi model smooths
-the nearest neighbor approximation by utilizing the Voronoi cells to
-define support via a generic basis function $v: \mathbb{R}^d
-\rightarrow \mathbb{R}_+$ given by
- $$ v^{x^{(i)}}(y) = \left(1 - \frac{\bigl\|y - x^{(i)}\bigr\|_2}
- {2 \ h \bigl(y - x^{(i)} \mid x^{(i)}\bigr)} \right)_+, $$
-where $h\bigl(w \mid x^{(i)}\bigr)$ is the distance between $x^{(i)}$
-and the boundary of the Voronoi cell $C^{x^{(i)}}$ in the direction
-$w$. $v^{x^{(i)}}\bigl(x^{(j)}\bigr) = \delta_{ij}$ and $v^{x^{(i)}}$
-has local support, giving the interpolated value
- $$ f(y) = \frac{\sum\limits_{i=1}^n v^{x^{(i)}}(y)
-  f\bigl(x^{(i)}\bigr)}{\sum\limits_{i=1}^n v^{x^{(i)}}(y)}, $$
-where $0 \leq v^{x^{(i)}}(y) \leq 1$. The computational complexity of
-interpolation via this Voronoi model is $\mathcal{O}(n^2 d)$. All of
-the approximations are an interpolant involving a convex combination
-of known function values $f\bigl(x^{(i)}\bigr)$.
-
-%% ===================================================================
-\section{Measuring Error}
-\label{sec:error}
-
-\begin{figure}[htb]
-  \centering
-  \begin{tikzpicture}
-    \node (img)  {\includegraphics[width=0.75\textwidth,]{Figures/plot-delaunay-example-prediction.pdf}};
-    \node[below=of img, node distance=1cm, yshift=1cm] {I/O Throughput};
-    \node[left=of img, node distance=0cm, rotate=90, anchor=center,yshift=-0.7cm] {CDF Value};
-  \end{tikzpicture}
-  \vspace{-0.3cm}
-  \caption{In this HPC I/O example, the general methodology for
-    predicting a CDF and evaluating error can be seen. The Delaunay
-    method chose three source distributions (dotted lines) and
-    assigned weights \{.3, .4, .3\} (top to bottom at arrow). The
-    weighted sum of the three known CDFs produces the predicted CDF
-    (dashed line). The KS Statistic (arrow) computed between the true
-    CDF (solid line) and predicted CDF (dashed line) is 0.2 for this
-    example. The KS test null hypothesis is rejected at $p$-value
-    0.01, however it is not rejected at $p$-value 0.001.
-  \vspace{-.1cm}}
-  \label{fig:prediction-example}
-\end{figure}
-
-When the range of an approximation is the real numbers, error is
-reported with summary statistics including: min absolute error, max
-absolute error, and absolute error quartiles. When the range of an
-approximation is the space of cumulative distribution functions, the
-Kolmogorov-Smirnov statistic (max-norm difference between the
-functions) is used.
-
-A hurdle when modeling function-valued outputs such as cumulative
-distribution functions (CDFs) or probability density functions (PDFs)
-is that certain properties must be maintained. It is necessary that a
-PDF $f: \mathbb{R} \rightarrow \mathbb{R}$ have the properties $f(x)
-\geq 0$ and $\int_{-\infty}^{\infty}f(x)dx = 1$. However, for a CDF
-$F: \mathbb{R} \rightarrow \mathbb{R}$ the properties are
-$F(x) \in [0,1]$ and $F(x)$ is absolutely continuous and
-nondecreasing. This work utilizes the fact that a convex combination
-of CDFs (or PDFs) results in a valid CDF (or PDF). Given $G(x) =
-\sum_{i}w_i F_i(x)$, $\sum_{i} w_i = 1$, $w_i \geq 0$, and each $F_i$
-is a valid CDF, $G$ must also be a valid CDF. A demonstration of how
-this is applied can be seen in Figure \ref{fig:prediction-example}.
-
-The performance of approximation techniques that predict probability
-functions can be analyzed through a variety of summary
-statistics. This work uses the max absolute difference, also known as
-the Kolmogorov-Smirnov (KS) statistic \cite{lilliefors1967kolmogorov}
-for its compatibility with the KS test.
-
-The two-sample KS test is a useful nonparametric test for comparing
-two CDFs while only assuming stationarity, finite mean, and finite
-variance. The null hypothesis (that two CDFs come from the same
-underlying distribution) is rejected at level $p \in [0,1]$ when
- $$ KS > \sqrt{-\frac{1}{2}\ln\biggl(\frac{p}{2}\biggr)} \sqrt{\frac{1}{n_1} + \frac{1}{n_2}}, $$
-with distribution sample sizes $n_1,n_2 \in \mathcal{N}$. For all
-applications of the KS test presented in this work $n_1 = n_2$. An
-example of the round-trip prediction methodology from known and
-predicted distributions to the calculation of error can be seen in
-Figure \ref{fig:prediction-example}.
-
-\section{Theoretical Error Bound}
-\label{sec:theory}
-
-This section presents the theoretical results bounding the error of
-(piecewise) linear interpolation. The error analysis relies on linear
-interpolation for three reasons: (1) second order results can be
-obtained utilizing a Lipschitz constant on the gradient of a function,
-rather than standard Lipschitz bounds; (2) the results directly apply
-to Delaunay interpolation; and (3) multiple other interpolants in this
-paper compute predictions as convex combinations of observed function
-values, which may allow for straight forward extensions of this error
-bound.
-
-\begin{plainlemma}
-  \label{lemma:1}
-  Let $S \subset \mathbb{R}^d$ be open and convex, $f: \mathbb{R}^d
-  \rightarrow \mathbb{R}$, and $\nabla f \in
-  Lip_{(\gamma,\|\cdot\|_2)}(S)$, the set of $\gamma$-Lipschitz
-  continuous functions in the $2$-norm. Then for all $x,y \in S$
-  $$\big|f(y) - f(x) - \langle \nabla f(x), y - x \rangle \big| \leq \frac{\gamma \|y - x\|_2^2}{2}.$$
-\end{plainlemma}
-
-\begin{proofdot}
-  Consider the function $g(t) = f \big((1-t) x + t y \big)$, $0 \leq t
-  \leq 1$, whose derivative $g'(t) = \big\langle \nabla f \big((1-t) x
-  + t y \big), y - x \big\rangle$ is the directional derivative of $f$
-  in the direction $(y - x).$
-  \begin{align*}
-    \big|f(y) - f(x) - \langle \nabla f(x), y - x \rangle \big|
-       &= \big|g(1) - g(0) - g'(0) \big| & \\
-       &= \bigg| \int_0^1 g'(t) - g'(0)\ dt \bigg| \\
-       &\leq \int_0^1 \big|g'(t) - g'(0)\big|\ dt \\
-       &= \int_0^1 \bigg| \big \langle \nabla f\big((1-t)x + ty\big) - \nabla f(x), y - x \big \rangle \bigg|\ dt \\
-       &\leq \int_0^1 \big \| \nabla f\big((1-t)x + ty\big) - \nabla f(x) \big \|_2\ \| y - x \|_2\ dt \\
-       &\leq \int_0^1 \big ( \gamma\ \|y-x\|_2 \big) \ \big( \|y-x\|_2 \big) t\ dt \\
-       &= \frac{\gamma \|y - x\|_2^2}{2}.
-  \end{align*}
-  \qed
-\end{proofdot}
-
-\begin{plainlemma}
-  \label{lemma:2}
-  Let $x, y, v_i \in \mathbb{R}^d$, $c_i \in \mathbb{R}$, and
-  $|\langle y - x, v_i \rangle| \leq c_i$ for $i = 1$, $\ldots$, $d.$
-  If $M = (v_1$, $\ldots$, $v_d)$ is nonsingular, then
-  $$\|y - x\|_2^2 \leq \frac{1}{\sigma_d^2} \sum_{i=1}^d c_i^2,$$
-  where $\sigma_d$ is the smallest singular value of $M.$
-\end{plainlemma}
-
-\begin{proofdot}
-  Using the facts that $M$ and $M^t$ have the same singular values,
-  and $\|M^tw\|_2 \geq \sigma_d \|w\|_2$, gives
-  \begin{align*}
-    \|y - x\|_2^2 &\leq \frac{\|M^t (y - x)\|_2^2}{\sigma_d^2} \\
-                  &=    \frac{1}{\sigma_d^2} \sum_{i=1}^d \langle y - x, v_i \rangle^2 \\
-                  &\leq \frac{1}{\sigma_d^2} \sum_{i=1}^d c_i^2.
-  \end{align*}
-  \qed
-\end{proofdot}
-
-\begin{plainlemma}
-  \label{lemma:3}
-  Given $f$, $\gamma$, $S$ as in {\it Lemma \ref{lemma:1}}, let $X =
-  \{x_0$, $x_1$, $\ldots$, $x_d\}$ $\subset S$ be the vertices of a
-  $d$-simplex, and let $\hat f(x) = \langle c, x - x_0 \rangle +
-  f(x_0)$, $c \in \mathbb{R}^d$ be the linear function interpolating
-  $f$ on $X.$ Let $\sigma_d$ be the smallest singular value of the
-  matrix $M = (x_1 - x_0$, $\ldots$, $x_d - x_0)$, and $k =
-  \max\limits_{1\ \leq\ j\ \leq\ d} \|x_j - x_0\|_2.$ Then
-  $$\big\|\nabla f(x_0) - c\big\|_2 \leq \sqrt{d} \frac{\gamma k^2}{\sigma_d}.$$
-\end{plainlemma}
-
-\begin{proofdot}
-  Consider $f(x) - \hat f(x)$ along the line segment $z(t) = (1-t)x_0
-  + t x_j$, $0 \leq t \leq 1.$ By Rolle's Theorem, for some $0 < \hat
-  t < 1$, $\big\langle \nabla f\big(z(\ \hat t\ )\big) - c, x_j - x_0
-  \big\rangle = 0.$ Now
-  \begin{align*}
-    \big| \big< \nabla f(x_0) - c, x_j - x_0 \big \rangle \big|
-        &= \big| \big< \nabla f(x_0) - \nabla f\big(z(\ \hat t\ )\big) + \nabla f\big(z(\ \hat t\ )\big) - c, x_j - x_0 \big \rangle \big| \\
-	&= \big| \big\langle \nabla f(x_0) - \nabla f \big(z(\ \hat t\ )\big), x_j - x_0 \big \rangle \big| \\
-	&\leq \big \| \nabla f(x_0) - \nabla f\big(z(\ \hat t\ )\big) \big \|_2 \|x_j - x_0\|_2 \\
-	&\leq \gamma \|x_0 - z(\ \hat t\ )\|_2\ \|x_j - x_0\|_2 \\
-	&\leq \gamma \|x_j - x_0\|_2^2 \leq \gamma k^2, \\
-  \end{align*}
-  for all $1 \leq j \leq d.$ Using {\it Lemma \ref{lemma:2}},
-  $$\big \| \nabla f(x_i) - c \big \|_2^2 \leq \frac{d}{\sigma_d^2}
-  \big( \gamma k^2\big)^2 \implies \big \| \nabla f(x_i) - c \big \|_2
-  \leq \sqrt{d} \frac{\gamma k^2}{\sigma_d}.$$
-  \qed
-\end{proofdot}
-
-\begin{onetheorem}
-  Under the assumptions of {\it Lemma \ref{lemma:1}} and {\it Lemma
-    \ref{lemma:3}}, for $z \in S$,
-  $$ \big|f(z) - \hat f(z)\big| \leq \frac{\gamma \|z - x_0\|_2^2}{2} + \sqrt{d} \frac{\gamma k^2}{\sigma_d} \|z - x_0\|_2.$$
-\end{onetheorem}
-
-\begin{proofdot}
-  Let $v = \nabla f(x_0) - c$, where $\|v\|_2 \leq \sqrt{d} \gamma k^2 / \sigma_d$ by {\it Lemma \ref{lemma:3}}. Now
-  \begin{align*}
-    \big|f(z) - \hat f(z)\big|
-       &= \big|f(z) - f(x_0) - \langle c, z - x_0 \rangle \big| \\ 
-       &= \big|f(z) - f(x_0) - \langle \nabla f(x_0) - v, z - x_0 \rangle \big| \\
-       &= \big|f(z) - f(x_0) - \langle \nabla f(x_0) , z - x_0 \rangle + \langle v , z - x_0 \rangle \big| \\
-       &\leq \big|f(z) - f(x_0) - \langle \nabla f(x_0) , z - x_0 \rangle \big| + \big| \langle v , z - x_0 \rangle \big| \\
-       &\leq \big|f(z) - f(x_0) - \langle \nabla f(x_0) , z - x_0 \rangle \big| + \|v\|_2 \|z - x_0\|_2 \\
-       &\leq \big|f(z) - f(x_0) - \langle \nabla f(x_0), z - x_0 \rangle \big| + \textstyle{\frac{\gamma k^2 \sqrt{d}}{\sigma_d}} \|z - x_0\|_2 \\
-       &\leq \frac{\gamma \|z - x_0\|_2^2}{2} + \sqrt{d}\frac{\gamma k^2}{\sigma_d} \|z - x_0\|_2,
-  \end{align*}
-  where the last inequality follows from {\it Lemma \ref{lemma:1}}.
-  %
-  \qed
-\end{proofdot}
-
-In summary, the approximation error of a linear (simplicial)
-interpolant tends quadratically towards zero when approaching observed
-data only when the diameter of the simplex is also reduced at a
-proportional rate.  Only linear convergence to the true function can
-be achieved in practice, without the incorporation of additional
-observations. Notice that the approximation error is largely
-determined by the spacing of observed data. Predictions made by
-simplices whose vertices are not well-spaced (i.e., have large
-diameter, or are nearly contained in a hyperplane) have higher
-error. In light of this error bound, an empirical evaluation of
-presented algorithms follows.
-
-%% ===================================================================
-\vspace{-2mm}
-\section{Data and Empirical Analysis}
-\label{sec:data}
-
-The theoretical results constructed in Section \ref{sec:theory} for
-(piecewise) linear interpolation are promising and apply directly to
-Delaunay interpolation, however they are difficult to interpret in
-context with approximation algorithms that do not have similar known
-uniform error bounds. For that reason, this paper utilizes five
-different data sets of varying dimension and application to construct
-approximations and compare the accuracy of different techniques.
-
-In the following five subsections the sources and targets of each test
-data set are described, as well as challenges and limitations related
-to approximating these data. The distribution of response values being
-modeled is presented followed by the distribution of approximation
-errors for each algorithm. The plots for all five data sets have the
-same format.
-
-All five data sets are rescaled such that the domain of approximation
-is the unit hypercube. The range of the first four data sets is the
-real numbers, while the range of the fifth data set is the space of
-cumulative distribution functions. All approximation techniques are
-applied to the first four data sets, while only those interpolants
-whose approximations are convex combinations of observed data are
-applied to the final data set.
-
-All approximations are constructed using $k$-fold cross validation as
-described in \cite{kohavi1995study} with $k=10$. This approach
-randomly partitions data into $k$ (nearly) equal sized sets. Each
-algorithm is then evaluated by constructing an approximation over each
-unique union of $k-1$ elements of the partition, making predictions
-for points in the remaining element. As a result, each observed data
-point is used in the construction of $k-1$ different approximations
-and is approximated exactly once. The $k$-fold cross validation method
-is data-efficient and provides an unbiased estimate of the expected
-prediction error \cite{kohavi1995study}, however it should be noted
-that neither this method nor others can provide a universally unbiased
-estimator for the variance of prediction error \cite{bengio2004no}.
-
-In addition to the figures displaying approximation results for each
-data set, tables of accompanying numerical results are located in the
-Appendix. All of the test data sets capture underlying functions that
-are almost certainly stochastic. As described in Section
-\ref{sec:introduction}, regression techniques appear most appropriate
-for these problems. However, typically data grows exponentially more
-sparse with increasing dimension. Given sparse data, regressors tend
-towards interpolation.
-
-%% -------------------------------------------------------------------
-\subsection{Forest Fire ($n = 504$, $d = 12$)}
-
-\begin{figure}
-  \centering
-  \begin{tikzpicture}
-    \node (img)  {\includegraphics[width=0.75\textwidth,]{Figures/raw-histogram-forest-fire.pdf}};
-    \node[below=of img, node distance=1cm, yshift=1cm] {Forest Fire Area Burned};
-    \node[left=of img, node distance=0cm, rotate=90, anchor=center,yshift=-0.7cm] {Count};
-  \end{tikzpicture}
-  \caption{Histogram of forest fire area burned under recorded weather
-    conditions. The data is presented on a $\ln$ scale because
-    most values are small with exponentially fewer fires on record
-    that burn large areas.}
-  \label{fig:hist-forest-fire}
-\end{figure}
-
-\begin{figure}
-  \centering
-  \begin{tikzpicture}
-    \node (img)  {\includegraphics[width=0.8\textwidth,]{Figures/error-box-forest-fire.pdf}};
-    \node[below=of img, node distance=1cm, yshift=1cm] {Area Burned Error};
-  \end{tikzpicture}
-  \caption{All models are applied to approximate the amount of area
-    that would be burned given environment conditions. $10$-fold cross
-    validation as described in the beginning of Section \ref{sec:data}
-    is used to evaluated each algorithm. This results in exactly one
-    prediction from each algorithm for each data point. These boxes
-    depict the median (middle bar), median $95\%$ confidence interval
-    (cones), quartiles (box edges), fences at $3/2$ interquartile
-    range (whiskers), and outliers (dots) of absolute prediction error
-    for each model. Similar to Figure \ref{fig:hist-forest-fire}, the
-    errors are presented on a $\ln$ scale. The numerical data
-    corresponding to this figure is provided in Table
-    \ref{table:error-forest-fire} in the Appendix.}
-  \label{fig:error-forest-fire}
-\end{figure}
-
-The forest fire data set \cite{cortez2007data} describes the area of
-Montesinho park burned over months of the year along with
-environmental conditions. The twelve dimensions being used to model
-burn area are the $x$ and $y$ spatial coordinates of burns in the
-park, month of year (mapped to $x$, $y$ coordinates on a unit circle),
-the FFMC, DMC, DC, and ISI indices (see source for details), the
-temperature, relative humidity, wind speed, and outdoor rain. The
-original analysis of this data set demonstrated it to be difficult to
-model, likely due to the skew in response values.
-
-As suggested by Figure \ref{fig:error-forest-fire}, the SVR has the
-lowest absolute prediction errors for $80\%$ of the data, with MLP and
-Delaunay being the nearest overall competitors. The effectiveness of
-SVR on this data suggests the underlying function can be defined by
-relatively few parameters, as well as the importance of capturing the
-low-burn-area data points.
-%% -------------------------------------------------------------------
-
-%% -------------------------------------------------------------------
-\subsection{Parkinson's Telemonitoring ($n = 5875$, $d = 19$)}
-
-\begin{figure}
-  \centering
-  \begin{tikzpicture}
-    \node (img)  {\includegraphics[width=0.75\textwidth,]{Figures/raw-histogram-parkinsons.pdf}};
-    \node[below=of img, node distance=1cm, yshift=1cm] {Parkinson's Total UPDRS Score};
-    \node[left=of img, node distance=0cm, rotate=90, anchor=center,yshift=-0.7cm] {Count};
-  \end{tikzpicture}
-  \caption{Histogram of the Parkinson's patient total UPDRS clinical
-    scores that will be approximated by each algorithm.}
-  \label{fig:hist-parkinsons}
-\end{figure}
-
-\begin{figure}
-  \centering
-  \begin{tikzpicture}
-    \node (img)  {\includegraphics[width=0.8\textwidth,]{Figures/error-box-parkinsons.pdf}};
-    \node[below=of img, node distance=1cm, yshift=1cm] {Total UPDRS Score Error};
-  \end{tikzpicture}
-  \caption{All models are applied to approximate the total UPDRS score
-    given audio features from patients' home life, using $10$-fold
-    cross validation. These boxes depict the median (middle bar),
-    median $95\%$ confidence interval (cones), quartiles (box edges),
-    fences at $3/2$ interquartile range (whiskers), and outliers
-    (dots) of absolute prediction error for each model. The numerical
-    data corresponding to this figure is provided in Table
-    \ref{table:error-parkinsons} in the Appendix.}
-  \label{fig:error-parkinsons}
-\end{figure}
-
-The second data set for evaluation \cite{tsanas2010accurate} is
-derived from a speech monitoring study with the intent to
-automatically estimate Parkinson's disease symptom development in
-Parkinson's patients. The function to be approximated is a
-time-consuming clinical evaluation measure referred to as the UPDRS
-score. The total UPDRS score given by a clinical evaluation is
-estimated through 19 real numbers generated from biomedical voice
-measures of in-home sound recordings.
-
-Figure \ref{fig:error-parkinsons} shows the ShepMod algorithm has the
-lowest minimum, first quartile, and median of absolute errors for this
-problem, while providing the best prediction $66\%$ of the time. The
-MLP has the lowest third quartile and provides the best prediction for
-$32\%$ of approximations. The dominance of ShepMod may be due in part
-to regular-interval total UPDRS scores provided by clinicians,
-favoring a nearest-neighbor strategy of prediction.
-%% -------------------------------------------------------------------
-
-%% -------------------------------------------------------------------
-\subsection{Australian Daily Rainfall Volume ($n = 2609$, $d = 23$)}
-
-\begin{figure}
-  \centering
-  \begin{tikzpicture}
-    \node (img)  {\includegraphics[width=0.75\textwidth,]{Figures/raw-histogram-weather.pdf}};
-    \node[below=of img, node distance=1cm, yshift=1cm] {Daily Rainfall in Sydney};
-    \node[left=of img, node distance=0cm, rotate=90, anchor=center,yshift=-0.7cm] {Count};
-  \end{tikzpicture}
-  \caption{Histogram of daily rainfall in Sydney, Australia, presented
-    on a $\ln$ scale because the frequency of larger amounts of
-    rainfall is significantly less. There is a peak in occurrence of
-    the value $0$, which has a notable effect on the resulting model
-    performance.}
-  \label{fig:hist-weather}
-\end{figure}
-
-\begin{figure}
-  \centering
-  \begin{tikzpicture}
-    \node (img)  {\includegraphics[width=0.8\textwidth,]{Figures/error-box-weather.pdf}};
-    \node[below=of img, node distance=1cm, yshift=1cm] {Sydney Rainfall Tomorrow Error};
-  \end{tikzpicture}
-  \caption{All models are applied to approximate the amount of
-    rainfall expected on the next calendar day given various sources
-    of local meteorological data, using $10$-fold cross validation.
-    These boxes depict the median (middle bar), median $95\%$
-    confidence interval (cones), quartiles (box edges), fences at
-    $3/2$ interquartile range (whiskers), and outliers (dots) of
-    absolute prediction error for each model. The errors are presented
-    on a $\ln$ scale, mimicking the presentation in Figure
-    \ref{fig:hist-weather}. The numerical data corresponding to this
-    figure is provided in Table \ref{table:error-weather} in the
-    Appendix.}
-  \label{fig:error-weather}
-\end{figure}
-
-The third data set for the total daily rainfall in Sydney, Australia
-\cite{williams2009rattle} provides a slightly higher dimensional
-challenge for the interpolants and regressors. This data is composed
-of many meteorological readings from the region in a day including:
-min and max temperatures, sunshine, wind speed directions (converted
-to coordinates on a circle), wind speeds, and humidities throughout
-the day, day of the year (converted to coordinates on a circle), and
-the model must predict the amount of rainfall tomorrow.
-
-While Figure \ref{fig:hist-weather} makes MARS look far better than
-other techniques, it only provides the best prediction for $11\%$ of
-points. The MLP has the lowest absolute error for $56\%$ of points and
-LSHEP is best for $28\%$. MARS likely achieves such a low first
-quartile due to the high occurrence of the value zero in the data.
-%% -------------------------------------------------------------------
-
-%% -------------------------------------------------------------------
-\subsection{Credit Card Transaction Amount ($n = 5562$, $d = 28$)}
-
-\begin{figure}
-  \centering
-  \begin{tikzpicture}
-    \node (img)  {\includegraphics[width=0.75\textwidth,]{Figures/raw-histogram-credit-card.pdf}};
-    \node[below=of img, node distance=1cm, yshift=1cm] {Credit Card Transaction Amount};
-    \node[left=of img, node distance=0cm, rotate=90, anchor=center,yshift=-0.7cm] {Count};
-  \end{tikzpicture}
-  \caption{Histogram of credit card transaction amounts, presented on
-    a $\ln$ scale. The data contains a notable frequency peak around
-    $\$1$ transactions. Fewer large purchases are made, but some large
-    purchases are in excess of five orders of magnitude greater than
-    the smallest purchases.}
-  \label{fig:hist-credit-card}
-\end{figure}
-
-\begin{figure}
-  \centering
-  \begin{tikzpicture}
-    \node (img)  {\includegraphics[width=0.8\textwidth,]{Figures/error-box-credit-card.pdf}};
-    \node[below=of img, node distance=1cm, yshift=1cm] {Transaction Amount Error};
-  \end{tikzpicture}
-  \caption{All models are applied to approximate the expected
-    transaction amount given transformed (and obfuscated) vendor and
-    customer-descriptive features, using $10$-fold cross validation.
-    These boxes depict the median (middle bar), median $95\%$
-    confidence interval (cones), quartiles (box edges), fences at
-    $3/2$ interquartile range (whiskers), and outliers (dots) of
-    absolute prediction error for each model. The absolute errors in
-    transaction amount predictions are presented on a $\ln$ scale,
-    just as in Figure \ref{fig:hist-credit-card}. The numerical data
-    corresponding to this figure is provided in Table
-    \ref{table:error-credit-card} in the Appendix.}
-  \label{fig:error-credit-card}
-\end{figure}
-
-The fourth test data set, and the final with a real-valued range, is a
-collection of credit card transactions
-\cite{pozzolo2015calibrating}. The provided data carries no direct
-real-world meaning, being the output of a principle component analysis
-on the original hidden source data. This obfuscation is done to
-protect the information of the credit card users. This data has the
-largest dimension of all considered, at 28. A model for this data
-predicts the transaction amount given the vector of principle
-component information.
-
-As can be seen in Figure \ref{fig:error-credit-card}, the MLP
-outperforms all other algorithms at the first, second, third, and
-fourth quartiles. The MLP produces the lowest absolute error
-prediction for $80\%$ of transactions, Delaunay bests the remaining
-$20\%$. It is likely that with less data, Delaunay would be the best
-performer.
-%% -------------------------------------------------------------------
-
-%% -------------------------------------------------------------------
-\subsection{High Performance Computing I/O ($n = 3016$, $d = 4$)}
-
-\begin{figure}
-  \centering
-  \begin{tikzpicture}
-    \node (img)  {\includegraphics[width=0.75\textwidth,]{Figures/raw-histogram-throughput.pdf}};
-    \node[below=of img, node distance=1cm, yshift=1cm] {I/O Read Throughtput};
-    \node[left=of img, node distance=0cm, rotate=90, anchor=center,yshift=-0.7cm] {Count};
-  \end{tikzpicture}
-  \caption{Histogram of the raw throughput values recorded during all
-    IOzone tests across all system configurations. The distribution is
-    skewed right, with few tests having significantly higher
-    throughput than most others. The data is presented on a $\ln$
-    scale.}
-  \label{fig:hist-throughput}
-\end{figure}
-
-\begin{figure}
-  \centering
-  \begin{tikzpicture}
-    \node (img)  {\includegraphics[width=0.8\textwidth,]{Figures/error-box-throughput.pdf}};
-    \node[below=of img, node distance=1cm, yshift=1cm] {Read Throughtput KS Statistic};
-  \end{tikzpicture}
-  \caption{The models directly capable of predicting distributions are
-    applied to predicting the expected CDF of I/O throughput at a
-    previously unseen system configuration, using $10$-fold cross
-    validation. The KS statistic between the observed distribution at
-    each system configuration and the predicted distribution is
-    recorded and presented above. Note that the above figure is
-    \textit{not} log-scaled like Figure \ref{fig:hist-throughput}. The
-    numerical data corresponding to this figure is provided in Table
-    \ref{table:error-throughput} in the Appendix.}
-  \label{fig:error-throughput}
-\end{figure}
-
-The final of five data sets is derived from \cite{cameron2019moana},
-which provides four-dimensional distribution data by executing the
-IOzone benchmark \cite{iozone} on a computer system and varying the
-system's file size, record size, thread count, and CPU frequency. At
-each configuration, IOzone samples the I/O file-read throughput (in
-bytes per second) 150 times. Empirical distribution function points
-are computed from each set of 150 executions, which are interpolated
-with a piecewise cubic Hermite interpolating polynomial
-\cite{fritsch1980monotone} to approximate the CDF. All interpolation
-algorithms with the exception of LSHEP are used to approximate these
-CDFs from system configurations.
-
-Delaunay achieves the lowest KS statistic for $62\%$ of
-approximations, while Voronoi is best for the remaining $38\%$. Figure
-\ref{fig:error-throughput} shows that while Delaunay may have more
-best predictions, the behavior of Voronoi may be preferable.
-%% -------------------------------------------------------------------
-
-%% -------------------------------------------------------------------
-\begin{figure}
-  \centering
-  \begin{tikzpicture}
-    \node (img)  {\includegraphics[width=0.8\textwidth]{Figures/plot-throughput-ks.pdf}};
-    \node[below=of img, node distance=1cm, yshift=1cm] {KS Statistic for Predicted vs. Actual};
-    \node[left=of img, node distance=0cm, rotate=90, anchor=center,yshift=-0.7cm] {Count of KS Statistic};
-  \end{tikzpicture}
-  \caption{Histograms of the prediction error for each interpolant
-    that produces predictions as convex combinations of observed data,
-    using $10$-fold cross validation. The histograms show the KS
-    statistics for the predicted throughput distribution versus the
-    actual throughput distribution. The four vertical lines represent
-    cutoff KS statistics given $150$ samples for commonly used
-    $p$-values 0.05, 0.01, 0.001, $10^{-6}$, respectively. All
-    predictions to the right of a vertical line represent CDF
-    predictions that are significantly different (by respective
-    $p$-value) from the actual distribution according to the KS
-    test. The numerical counterpart to this figure is presented in
-    Table \ref{table:null-hypothesis-results}.}
-  \label{fig:throughput-prediction}
-\end{figure}
-
-\begin{table}
-  \renewcommand{\arraystretch}{1.3}
-  \centering
-  \begin{tabular}{c|c|c|c|c}
-     & $p = .05$ & $p = .01$ & $p = .001$ & $p = 10^{-6}$\\
-    \hline
-    Delaunay & $\mathbf{50.3}\%$ & $\mathit{43.5}\%$ & $\mathit{36.2}\%$ & $\mathit{24.7}\%$\\
-    ShepMod & $\mathit{51.4}\%$ & $44.8\%$ & $38.1\%$ & $27.7\%$\\
-    Voronoi & $52.6\%$ & $\mathbf{43.4}\%$ & $\mathbf{34.4}\%$ & $\mathbf{19.1}\%$\\
-    BoxSpline & $99.5\%$ & $98.5\%$ & $96.6\%$ & $89.5\%$\\
-  \end{tabular}
-  \caption{Numerical counterpart of the histogram data presented in
-    Figure \ref{fig:throughput-prediction}. The columns display the
-    percent of null hypothesis rejections by the KS-test when provided
-    different selections of $p$-values for each algorithm. The
-    algorithm with the lowest rejection rate at each $p$ is boldface,
-    while the second lowest is italicized.}
-  \label{table:null-hypothesis-results}
-\end{table}
-
-%% ===================================================================
-
-Figure \ref{fig:throughput-prediction} expands on the KS statistic
-results presented in Figure \ref{fig:error-throughput}. Agglomerate
-errors for each algorithm resemble a Gamma distribution. The
-percentages of significant prediction errors with varying $p$-values
-are on display in Table \ref{table:null-hypothesis-results}. When
-considering the $p=0.001$ results for each technique, a little over
-one third of the predicted CDFs are significantly different from the
-measured (and presumed) correct CDFs. However, it should be noted that
-with 150 samples, the error of an empirical distribution function
-(EDF) can reasonably be upwards of $.1$, which serves as a rough
-estimate for the lower limit of achievable error. Globally, only a
-third of Voronoi predictions fail to capture \textit{all} of the
-characteristics of the CDFs at new system configurations.
-
-\section{Discussion}
-\label{sec:discussion}
-
-\begin{table}
-  \centering
-  \begin{tabular}{c|r@{.}l|r@{.}l|r@{.}l}
-    Algorithm & \multicolumn{2}{c|}{Avg. \% Best} & \multicolumn{2}{c|}{\multilinecell{Avg. Fit or \\ Prep. Time (s)}} & \multicolumn{2}{c}{Avg. App. Time (s)}\\
-    \hline
-    MARS & \quad\quad$4$&$6$ & \,\quad$20$&$0$ & \quad\quad$0$&$001$\\
-    SVR & $\mathit{21}$&$\mathit{1}$ & $\mathit{0}$&$\mathit{5}$ & $\mathbf{0}$&$\mathbf{0001}$\\
-    MLP & $\mathbf{42}$&$\mathbf{1}$ & $200$&$0$ & $0$&$001$\\
-    Delaunay & $6$&$0$ & $1$&$0$ & $1$&$0$\\
-    ShepMod & $18$&$4$ & $0$&$7$ & $\mathbf{0}$&$\mathbf{0001}$\\
-    LSHEP & $8$&$4$ & $2$&$0$ & $\mathbf{0}$&$\mathbf{0001}$\\
-    Voronoi & $0$&$5$ & $\mathbf{0}$&$\mathbf{1}$ & $0$&$04$\\
-    BoxSpline & $0$&$2$ & $0$&$8$ & $\mathit{0}$&$\mathit{0005}$\\
-  \end{tabular}
-  \caption{This average of Appendix Tables
-    \ref{table:best-forest-fire}, \ref{table:best-parkinsons},
-    \ref{table:best-weather}, and \ref{table:best-credit-card}
-    provides a gross summary of overall results. The columns display
-    (weighted equally by data set, \textit{not} points) the average
-    frequency with which any algorithm provided the lowest absolute
-    error approximation, the average time to fit/prepare, and the
-    average time required to approximate one point. The times have
-    been rounded to one significant digit, as reasonably large
-    fluctuations may be observed due to implementation hardware.
-    Interpolants provide the lowest error approximation for nearly one
-    third of all data, while regressors occupy the other two
-    thirds. This result is obtained without any customized tuning or
-    preprocessing to maximize the performance of any given
-    algorithm. In practice, tuning and preprocessing may have large
-    effects on approximation performance.}
-  \label{table:avg-performance}
-\end{table}
-
-Table \ref{table:avg-performance} summarizes results across the four
-test data sets with real-valued ranges. The interpolants discussed in
-this paper produce the \textit{best} approximations roughly one third
-of the time, and produce competitive approximations for almost all
-data sets. These test problems are almost certainly
-\textit{stochastic} in nature, but the high dimension leads to greater
-data sparsity and model construction cost, making interpolation more
-competitive.
-
-The major advantages to interpolation lie in the near absence of
-\textit{fit} time. Delaunay, LSHEP, and ShepMod all require pairwise
-distance calculations, for numerical robustness (Delaunay) and to
-determine the radii of influence for data points (LSHEP and
-ShepMod). At least hundreds, and sometimes hundreds of thousands of
-predictions can be made by the interpolants before the most widely
-used regressor (MLP) finishes fitting these relatively small data
-sets.  However, the computational complexities of all interpolants
-presented are greater than linear in either dimension or number of
-points, whereas the regressors' nonlinear complexity in dimension
-generally comes from the model fitting optimization.
-
-The new theoretical results presented in Section \ref{sec:theory}
-directly apply to Delaunay interpolation, however the performance of
-Delaunay does not appear significantly better than other algorithms on
-these data sets.  This observation may be due to the stochastic nature
-of the data, but it also speaks to the power of the approximations
-generated by the different interpolation methods.  The strong
-performance of other interpolants suggests that theoretical results
-similar to those presented in this work can be achieved for the other
-interpolants under reasonable assumptions.
-
-Finally, most of the interpolants presented in this work benefit from
-the ability to model any function over real $d$-tuples with a range
-that is closed under convex combinations. The results of the
-distribution prediction case study indicate that interpolants can
-effectively predict CDFs. The error analysis for that work relies on
-the KS statistic, which captures the worst part of any prediction and
-hence provides a conservatively large estimate of approximation
-error. The average absolute errors in the predicted CDFs are always
-lower than the KS statistics. However, the KS statistic was chosen as
-a metric because of the important surrounding statistical theory. A
-nonnegligible volume of predictions provide impressively low levels of
-average absolute error in that final case study.
-
-\section{Conclusion}
-\label{sec:conclusion}
-
-The major contributions of this work are: 1) new uniform theoretical
-error bounds for piecewise linear interpolation in arbitrary dimension
-(Section \ref{sec:theory}); 2) an empirical evaluation across
-real-world problems (Section \ref{sec:data}) that demonstrates
-interpolants produce competitively accurate models of multivariate
-phenomenon when compared with common regressors for sparse, moderately
-high dimensional problems; and 3) a demonstration that some
-interpolants generalize to interpolation in function spaces,
-preserving monotonicity (with CDFs, e.g.), neither of which common
-regressors can do.
-
-The various interpolants discussed in this paper have been
-demonstrated to effectively approximate multivariate phenomena up to
-$30$ dimensions. The underlying constructions are theoretically
-straightforward, interpretable, and yield reasonably accurate
-predictions. Most of the interpolants' computational complexities make
-them particularly suitable for applications in even higher
-dimension. The major benefits of interpolation are seen when only a
-small number of approximations ($\leq 1000$) are made from data and
-when there are relatively few data points for the dimension (for
-empirical results presented, $\log_d n \leq 5$).
-
-%\begin{acknowledgements}
-%If you'd like to thank anyone, place your comments here
-%and remove the percent signs.
-%\end{acknowledgements}
-
-%%====================================================================
-%%====================================================================
-
-\bibliographystyle{spmpsci}      % mathematics and physical sciences
-\bibliography{paper}
-
-\newpage
-
-\begin{appendix}
-\section*{Appendix}
-\label{sec:appendix}
-
-\vspace*{\fill}
-
-\begin{table}[H]
-  \centering
-  \begin{tabular}{c|r@{.}l|r@{.}l|r@{.}l|r@{.}l|r@{.}l}
-    \hline
-    Algorithm & \multicolumn{2}{c|}{Min} & \multicolumn{2}{c|}{$25^{th}$} & \multicolumn{2}{c|}{$50^{th}$} & \multicolumn{2}{c|}{$75^{th}$} & \multicolumn{2}{c}{Max}\\
-    \hline
-    MARS & $0$&$00984$ & $3$&$11$ & $7$&$01$ & $\mathit{11}$&$\mathit{7}$ & $1090$&$0$\\
-    SVR & $\mathit{0}$&$\mathit{00402}$ & $\mathbf{0}$&$\mathbf{243}$ & $\mathbf{0}$&$\mathbf{416}$ & $\mathbf{6}$&$\mathbf{19}$ & $1090$&$0$\\
-    MLP & $0$&$0426$ & $2$&$63$ & $\mathit{5}$&$\mathit{27}$ & $14$&$0$ & $1090$&$0$\\
-    Delaunay & $\mathbf{0}$&$\mathbf{00}$ & $1$&$98$ & $5$&$37$ & $13$&$1$ & $\mathit{1080}$&$\mathit{0}$\\
-    ShepMod & $\mathbf{0}$&$\mathbf{00}$ & $\mathit{1}$&$\mathit{93}$ & $6$&$27$ & $16$&$0$ & $1090$&$0$\\
-    LSHEP & $0$&$0400$ & $2$&$17$ & $8$&$87$ & $19$&$1$ & $\mathbf{1070}$&$\mathbf{0}$\\
-    Voronoi & $0$&$00982$ & $3$&$65$ & $7$&$56$ & $15$&$6$ & $1090$&$0$\\
-    BoxSpline & $\mathbf{0}$&$\mathbf{00}$ & $2$&$27$ & $5$&$91$ & $16$&$4$ & $1090$&$0$\\
-    \hline
-  \end{tabular}
-  \caption{This numerical data accompanies the visual provided in
-    Figure \ref{fig:error-forest-fire}. The columns of absolute error
-    percentiles correspond to the minimum, first quartile, median,
-    third quartile, and maximum absolute errors respectively. The
-    minimum of each column is boldface, while the second lowest value
-    is italicized. All values are rounded to three significant
-    digits.}
-  \label{table:error-forest-fire}
-\end{table}
-
-\begin{table}[H]
-  \centering
-  \begin{tabular}{|c|r@{.}l| c |r@{.}l|r@{.}l|r@{.}l|}
-    \cline{1-3}\cline{5-10}
-    Algorithm & \multicolumn{2}{c|}{\% Best} &  & \multicolumn{2}{c|}{Fit/Prep. Time (s)} & \multicolumn{2}{c|}{App. Time (s)} & \multicolumn{2}{c|}{Total App. Time (s)}\\
-    \cline{1-3}\cline{5-10}
-    MARS & \quad$\mathit{7}$&$\mathit{7}$ &  & \quad\quad$29$&$1$ & \quad$0$&$00137$ & \quad\quad\quad$0$&$0686$\\
-    SVR & $\mathbf{80}$&$\mathbf{2}$ &  & $\mathit{0}$&$\mathit{00584}$ & $\mathit{0}$&$\mathit{0000620}$ & $\mathit{0}$&$\mathit{00310}$\\
-    MLP & $0$&$0$ &  & $32$&$8$ & $0$&$000871$ & $0$&$0436$\\
-    Delaunay & $3$&$5$ &  & $0$&$0151$ & $0$&$0234$ & $1$&$18$\\
-    ShepMod & $3$&$7$ &  & $0$&$00634$ & $0$&$0000644$ & $0$&$00322$\\
-    LSHEP & $5$&$1$ &  & $0$&$0275$ & $\mathbf{0}$&$\mathbf{0000463}$ & $\mathbf{0}$&$\mathbf{00231}$\\
-    Voronoi & $0$&$0$ &  & $\mathbf{0}$&$\mathbf{000182}$ & $0$&$000396$ & $0$&$0198$\\
-    BoxSpline & $0$&$4$ &  & $0$&$00724$ & $0$&$0000978$ & $0$&$00489$\\
-    \cline{1-3}\cline{5-10}
-  \end{tabular}
-  \caption{The left above shows how often each algorithm had the
-    lowest absolute error approximating forest fire data in Table
-    \ref{table:error-forest-fire}. On the right columns are median fit
-    time of 454 points, median time for one approximation, and median
-    time approximating 50 points.}
-  \label{table:best-forest-fire}
-\end{table}
-
-\vspace*{\fill}
-\newpage
-
-\begin{table}
-  \centering
-  \begin{tabular}{c|r@{.}l|r@{.}l|r@{.}l|r@{.}l|r@{.}l}
-    \hline
-    Algorithm & \multicolumn{2}{c|}{Min} & \multicolumn{2}{c|}{$25^{th}$} & \multicolumn{2}{c|}{$50^{th}$} & \multicolumn{2}{c|}{$75^{th}$} & \multicolumn{2}{c}{Max}\\
-    \hline
-    MARS & $0$&$00948$ & $9$&$98$ & $20$&$4$ & $32$&$9$ & $863$&$0$\\
-    SVR & $0$&$00138$ & $3$&$17$ & $7$&$31$ & $12$&$6$ & $\mathbf{27}$&$\mathbf{5}$\\
-    MLP & $0$&$0000239$ & $\mathit{0}$&$\mathit{533}$ & $\mathit{1}$&$\mathit{25}$ & $\mathbf{2}$&$\mathbf{84}$ & $39$&$3$\\
-    Delaunay & $\mathit{3}$&$\mathit{72 \times 10^{-12}}$ & $1$&$20$ & $3$&$50$ & $7$&$67$ & $30$&$7$\\
-    ShepMod & $\mathbf{0}$&$\mathbf{0}$ & $\mathbf{0}$&$\mathbf{255}$ & $\mathbf{0}$&$\mathbf{908}$ & $\mathit{3}$&$\mathit{43}$ & $34$&$5$\\
-    LSHEP & $0$&$00254$ & $2$&$93$ & $7$&$16$ & $13$&$1$ & $\mathit{29}$&$\mathit{0}$\\
-    Voronoi & $\mathbf{0}$&$\mathbf{0}$ & $1$&$29$ & $3$&$52$ & $6$&$87$ & $30$&$1$\\
-    BoxSpline & $0$&$00287$ & $4$&$39$ & $9$&$10$ & $14$&$8$ & $41$&$3$\\
-    \hline
-  \end{tabular}
-  \caption{This numerical data accompanies the visual provided in
-    Figure \ref{fig:error-parkinsons}. The columns of absolute error
-    percentiles correspond to the minimum, first quartile, median,
-    third quartile, and maximum absolute errors respectively. The
-    minimum of each column is boldface, while the second lowest value
-    is italicized. All values are rounded to three significant
-    digits.}
-  \label{table:error-parkinsons}
-\end{table}
-
-\begin{table}
-  \centering
-  \begin{tabular}{|c|r@{.}l| c |r@{.}l|r@{.}l|r@{.}l|}
-    \cline{1-3}\cline{5-10}
-    Algorithm & \multicolumn{2}{c|}{\% Best} &  & \multicolumn{2}{c|}{Fit/Prep. Time (s)} & \multicolumn{2}{c|}{App. Time (s)} & \multicolumn{2}{c|}{Total App. Time (s)}\\
-    \cline{1-3}\cline{5-10}
-    MARS & \quad\,$0$&$0$ &  & \quad\quad\quad$37$&$9$ & \quad\,\,$0$&$00253$ & \quad\quad\quad\,\,$1$&$48$\\
-    SVR & $0$&$1$ &  & $\mathit{0}$&$\mathit{859}$ & $\mathit{0}$&$\mathit{000181}$ & $\mathit{0}$&$\mathit{106}$\\
-    MLP & $\mathit{32}$&$\mathit{0}$ &  & $348$&$0$ & $0$&$00111$ & $0$&$653$\\
-    Delaunay & $0$&$0$ &  & $2$&$47$ & $1$&$22$ & $720$&$0$\\
-    ShepMod & $\mathbf{66}$&$\mathbf{4}$ &  & $1$&$13$ & $0$&$000182$ & $0$&$107$\\
-    LSHEP & $0$&$0$ &  & $2$&$39$ & $\mathbf{0}$&$\mathbf{000144}$ & $\mathbf{0}$&$\mathbf{0845}$\\
-    Voronoi & $1$&$6$ &  & $\mathbf{0}$&$\mathbf{298}$ & $0$&$0274$ & $16$&$1$\\
-    BoxSpline & $0$&$0$ &  & $1$&$26$ & $0$&$000643$ & $0$&$377$\\
-    \cline{1-3}\cline{5-10}
-  \end{tabular}
-  \caption{The left above shows how often each algorithm had the
-    lowest absolute error approximating Parkinson's data in Table
-    \ref{table:error-parkinsons}. On the right columns are median fit
-    time of 5288 points, median time for one approximation, and median
-    time approximating 587 points.}
-  \label{table:best-parkinsons}
-\end{table}
-
-\begin{table}
-  \centering
-  \begin{tabular}{c|r@{.}l|r@{.}l|r@{.}l|r@{.}l|r@{.}l}
-    \hline
-    Algorithm & \multicolumn{2}{c|}{Min} & \multicolumn{2}{c|}{$25^{th}$} & \multicolumn{2}{c|}{$50^{th}$} & \multicolumn{2}{c|}{$75^{th}$} & \multicolumn{2}{c}{Max}\\
-    \hline
-    MARS & $\mathit{6}$&$\mathit{45 \times 10^{-15}}$ & $\mathbf{2}$&$\mathbf{70 \times 10^{-14}}$ & $1$&$66$ & $1$&$96$ & $\mathit{53}$&$\mathit{3}$\\
-    SVR & $0$&$0420$ & $0$&$143$ & $0$&$148$ & $\mathit{0}$&$\mathit{860}$ & $119$&$0$\\
-    MLP & $0$&$0000689$ & $0$&$0337$ & $\mathbf{0}$&$\mathbf{0795}$ & $\mathbf{0}$&$\mathbf{264}$ & $\mathbf{5}$&$\mathbf{31}$\\
-    Delaunay & $\mathbf{0}$&$\mathbf{0}$ & $0$&$187$ & $0$&$919$ & $2$&$72$ & $56$&$3$\\
-    ShepMod & $\mathbf{0}$&$\mathbf{0}$ & $0$&$0957$ & $0$&$685$ & $2$&$90$ & $73$&$2$\\
-    LSHEP & $\mathbf{0}$&$\mathbf{0}$ & $\mathit{0}$&$\mathit{0153}$ & $\mathit{0}$&$\mathit{106}$ & $1$&$17$ & $113$&$0$\\
-    Voronoi & $\mathbf{0}$&$\mathbf{0}$ & $0$&$430$ & $1$&$28$ & $2$&$94$ & $83$&$8$\\
-    BoxSpline & $\mathbf{0}$&$\mathbf{0}$ & $0$&$789$ & $2$&$06$ & $5$&$25$ & $119$&$0$\\
-    \hline
-  \end{tabular}
-  \caption{This numerical data accompanies the visual provided in
-    Figure \ref{fig:error-weather}. The columns of absolute error
-    percentiles correspond to the minimum, first quartile, median,
-    third quartile, and maximum absolute errors respectively. The
-    minimum value of each column is boldface, while the second lowest
-    is italicized. All values are rounded to three significant
-    digits.}
-  \label{table:error-weather}
-\end{table}
-
-\begin{table}
-  \centering
-  \begin{tabular}{|c|r@{.}l| c |r@{.}l|r@{.}l|r@{.}l|}
-    \cline{1-3}\cline{5-10}
-    Algorithm & \multicolumn{2}{c|}{\% Best} &  & \multicolumn{2}{c|}{Fit/Prep. Time (s)} & \multicolumn{2}{c|}{App. Time (s)} & \multicolumn{2}{c|}{Total App. Time (s)}\\
-    \cline{1-3}\cline{5-10}
-    MARS & \,\,\,\,$10$&$7$ &  & \quad\quad\quad$0$&$151$ & \quad$0$&$000117$ & \quad\quad\quad\,\,$0$&$0304$\\
-    SVR & $4$&$1$ &  & $\mathit{0}$&$\mathit{133}$ & $\mathbf{0}$&$\mathbf{0000947}$ & $\mathbf{0}$&$\mathbf{0246}$\\
-    MLP & $\mathbf{56}$&$\mathbf{8}$ &  & $169$&$0$ & $0$&$00137$ & $0$&$356$\\
-    Delaunay & $0$&$1$ &  & $0$&$664$ & $0$&$886$ & $230$&$0$\\
-    ShepMod & $3$&$5$ &  & $0$&$265$ & $0$&$000128$ & $0$&$0333$\\
-    LSHEP & $\mathit{28}$&$\mathit{4}$ &  & $0$&$874$ & $\mathit{0}$&$\mathit{0000975}$ & $\mathit{0}$&$\mathit{0254}$\\
-    Voronoi & $0$&$2$ &  & $\mathbf{0}$&$\mathbf{012}$ & $0$&$0270$ & $7$&$01$\\
-    BoxSpline & $0$&$3$ &  & $0$&$330$ & $0$&$000406$ & $0$&$106$\\
-    \cline{1-3}\cline{5-10}
-  \end{tabular}
-  \caption{Left table shows how often each algorithm had the lowest
-    absolute error approximating Sydney rainfall data in Table
-    \ref{table:error-weather}. On the right columns are median fit
-    time of 2349 points, median time for one approximation, and median
-    time approximating 260 points.}
-  \label{table:best-weather}
-\end{table}
-
-\begin{table}
-  \centering
-  \begin{tabular}{c|r@{.}l|r@{.}l|r@{.}l|r@{.}l|r@{.}l}
-    \hline
-    Algorithm & \multicolumn{2}{c|}{Min} & \multicolumn{2}{c|}{$25^{th}$} & \multicolumn{2}{c|}{$50^{th}$} & \multicolumn{2}{c|}{$75^{th}$} & \multicolumn{2}{c}{Max}\\
-    \hline
-    MARS & $5$&$36$ & $3610$&$0$ & $4580$&$0$ & $5450$&$0$ & $13400$&$0$\\
-    SVR & $0$&$0138$ & $8$&$47$ & $15$&$4$ & $41$&$9$ & $7700$&$0$\\
-    MLP & $0$&$00151$ & $\mathbf{1}$&$\mathbf{60}$ & $\mathbf{3}$&$\mathbf{86}$ & $\mathbf{8}$&$\mathbf{34}$ & $\mathbf{604}$&$\mathbf{0}$\\
-    Delaunay & $\mathbf{0}$&$\mathbf{0}$ & $\mathit{2}$&$\mathit{69}$ & $\mathit{13}$&$\mathit{8}$ & $\mathit{35}$&$\mathit{0}$ & $\mathit{4840}$&$\mathit{0}$\\
-    ShepMod & $\mathbf{0}$&$\mathbf{0}$ & $4$&$21$ & $17$&$3$ & $51$&$6$ & $6510$&$0$\\
-    LSHEP & $1$&$27$ & $199$&$0$ & $260$&$0$ & $343$&$0$ & $6530$&$0$\\
-    Voronoi & $\mathit{2}$&$\mathit{89 \times 10^{-10}}$ & $14$&$7$ & $32$&$8$ & $52$&$9$ & $4860$&$0$\\
-    BoxSpline & $0$&$00740$ & $20$&$8$ & $40$&$9$ & $79$&$6$ & $7640$&$0$\\
-    \hline
-  \end{tabular}
-  \caption{This numerical data accompanies the visual provided in
-    Figure \ref{fig:error-credit-card}. The columns of absolute error
-    percentiles correspond to the minimum, first quartile, median,
-    third quartile, and maximum absolute errors respectively. The
-    minimum value of each column is boldface, while the second lowest
-    is italicized. All values are rounded to three significant
-    digits.}
-  \label{table:error-credit-card}
-\end{table}
-
-\begin{table}
-  \centering
-  \begin{tabular}{|c|r@{.}l| c |r@{.}l|r@{.}l|r@{.}l|}
-    \cline{1-3}\cline{5-10}
-    Algorithm & \multicolumn{2}{c|}{\% Best} &  & \multicolumn{2}{c|}{Fit/Prep. Time (s)} & \multicolumn{2}{c|}{App. Time (s)} & \multicolumn{2}{c|}{Total App. Time (s)}\\
-    \cline{1-3}\cline{5-10}
-    MARS & \quad\,$0$&$0$ &  & \quad\quad\quad$22$&$0$ & \quad$0$&$00148$ & \quad\quad\quad\,\,$0$&$820$\\
-    SVR & $0$&$0$ &  & $\mathit{1}$&$\mathit{01}$ & $\mathit{0}$&$\mathit{000210}$ & $\mathit{0}$&$\mathit{117}$\\
-    MLP & $\mathbf{79}$&$\mathbf{5}$ &  & $290$&$0$ & $0$&$000714$ & $0$&$397$\\
-    Delaunay & $\mathit{20}$&$\mathit{5}$ &  & $3$&$12$ & $3$&$71$ & $2070$&$0$\\
-    ShepMod & $0$&$1$ &  & $1$&$45$ & $0$&$000220$ & $0$&$122$\\
-    LSHEP & $0$&$0$ &  & $3$&$47$ & $\mathbf{0}$&$\mathbf{000176}$ & $\mathbf{0}$&$\mathbf{0981}$\\
-    Voronoi & $0$&$0$ &  & $\mathbf{0}$&$\mathbf{197}$ & $0$&$0950$ & $52$&$8$\\
-    BoxSpline & $0$&$0$ &  & $1$&$66$ & $0$&$000956$ & $0$&$532$\\
-    \cline{1-3}\cline{5-10}
-  \end{tabular}
-  \caption{The left above shows how often each algorithm had the
-    lowest absolute error approximating credit card transaction data
-    in Table \ref{table:error-credit-card}. On the right columns are
-    median fit time of 5006 points, median time for one approximation,
-    and median time approximating 556 points.}
-  \label{table:best-credit-card}
-\end{table}
-
-\begin{table}
-  \centering
-  \begin{tabular}{c|r@{.}l|r@{.}l|r@{.}l|r@{.}l|r@{.}l}
-    \hline
-    Algorithm & \multicolumn{2}{c|}{Min} & \multicolumn{2}{c|}{$25^{th}$} & \multicolumn{2}{c|}{$50^{th}$} & \multicolumn{2}{c|}{$75^{th}$} & \multicolumn{2}{c}{Max}\\
-    \hline
-    Delaunay & $\mathbf{0}$&$\mathbf{0287}$ & $\mathbf{0}$&$\mathbf{0914}$ & $\mathbf{0}$&$\mathbf{158}$ & $\mathit{0}$&$\mathit{308}$ & $\mathit{0}$&$\mathit{897}$\\
-    ShepMod & $0$&$0307$ & $\mathit{0}$&$\mathit{0983}$ & $\mathit{0}$&$\mathit{164}$ & $0$&$335$ & $1$&$00$\\
-    Voronoi & $\mathit{0}$&$\mathit{0303}$ & $0$&$100$ & $0$&$165$ & $\mathbf{0}$&$\mathbf{274}$ & $\mathbf{0}$&$\mathbf{762}$\\
-    BoxSpline & $0$&$0703$ & $0$&$417$ & $0$&$561$ & $0$&$745$ & $1$&$00$\\
-    \hline
-  \end{tabular}
-  \caption{This numerical data accompanies the visual provided in
-    Figure \ref{fig:error-throughput}. The columns of absolute error
-    percentiles correspond to the minimum, first quartile, median,
-    third quartile, and maximum KS statistics respectively between
-    truth and guess for models predicting the distribution of I/O
-    throughput that will be observed at previously unseen system
-    configurations. The minimum value of each column is boldface,
-    while the second lowest is italicized. All values are rounded to
-    three significant digits.}
-  \label{table:error-throughput}
-\end{table}
-
-\begin{table}
-  \centering
-  \begin{tabular}{|c|r@{.}l| c |r@{.}l|r@{.}l|r@{.}l|}
-    \cline{1-3}\cline{5-10}
-    Algorithm & \multicolumn{2}{c|}{\% Best} &  & \multicolumn{2}{c|}{Fit/Prep. Time (s)} & \multicolumn{2}{c|}{App. Time (s)} & \multicolumn{2}{c|}{Total App. Time (s)}\\
-    \cline{1-3}\cline{5-10}
-    Delaunay & \,\,$\mathbf{62}$&$\mathbf{0}$ &  & \quad\quad\quad$0$&$344$ & \quad$0$&$00984$ & \quad\quad\,\,$2$&$71$\\
-    ShepMod & $0$&$0$ &  & $\mathbf{0}$&$\mathbf{0884}$ & $\mathbf{0}$&$\mathbf{000145}$ & $\mathbf{0}$&$\mathbf{0436}$\\
-    Voronoi & $\mathit{38}$&$\mathit{0}$ &  & $1$&$0$ & $0$&$00253$ & $0$&$762$\\
-    BoxSpline & $0$&$0$ &  & $0$&$0972$ & $\mathit{0}$&$\mathit{000210}$ & $\mathit{0}$&$\mathit{0633}$\\
-    \cline{1-3}\cline{5-10}
-  \end{tabular}
-  \caption{The left above shows how often each algorithm had the
-    lowest KS statistic on the I/O throughput distribution data in
-    Table \ref{table:error-throughput}. On the right columns are
-    median fit time of 2715 points, median time for one approximation,
-    and median time approximating 301 points.}
-  \label{table:best-throughput}
-\end{table}
-
-\end{appendix}
-
-\end{document}
+Among the statistical models presented in prior works,
+\cite{bailey2005performance} specifically mention that it is difficult
+for the simplified models to capture variability introduced by
+I/O. System variability in general has become a problem of increasing
+interest to the HPC and systems communities, however most of the work
+has focused on operating system (OS) induced variability
+\cite{beckman2008benchmarking,de2007identifying}. The work that has
+focused on managing I/O variability does not use any sophisticated
+modeling techniques \cite{lofstead2010managing}. Hence, Chapter
+\ref{ch:naive} presents a case study applying advanced mathematical
+and statistical modeling techniques to the domain of HPC I/O
+characteristics. The models are used to predict the mean throughput of
+a system and the variance in throughput of a system. The discussion
+section that follows outlines how the techniques presented can be
+applied to any performance metric and any system.

### 2020-04-14 3-Disseration/boxes.tex
Updated min wiggle fit tests, wrote report on the topic that was sent to Dr. Watson.

--- 2018-02-13 Paper_[2018-01-09]_ACMSE/paper.tex
@@
-\def\mn#1{\left\Vert #1\right\Vert_\infty} 
-\section{Introduction}
-\label{sec_introduction}
+Chapter \ref{ch:naive} demonstrated that the modeling and
+approximation techniques described in Chapter \ref{ch:algs} can be
+used to model and predict numeric approximations of system
+variability. The case study was only for four-dimensional data, and
+the key weakness of the best predictor (Delaunay) is scaling with
+dimension. This Chapter discusses a modeling approach that may achieve
+better scaling with dimension, proposes a quasi-mesh construction, and
+analyzes the performance of this proposed technique.
 
-Regression and interpolation are problems of considerable importance that find applications across many fields of science.
-Pollution and air quality analysis \cite{de2008field}, energy consumption management \cite{lazos2014optimisation}, and student performance prediction \cite{cortez2008using} are a few examples of interdisciplinary applications of multivariate regression for predictive analysis.
-As discussed later, these techniques can also be applied to prediction problems related to high performance computing (HPC) file input/output (I/O), Parkinson's patient clinical evaluations \cite{tsanas2010accurate}, and forest fire risk assessment \cite{cortez2007data}.
-
-Multivariate interpolation is formally defined when there exists some function $f:\mathbb{R}^d \rightarrow \mathbb{R}$ and a set $X$ of $n$ points in $\mathbb{R}^d$ along with associated response values $f(x)$ for all $x \in X$.
-The problem is to construct an approximation $\hat f: \mathbb{R}^d \rightarrow \mathbb{R}$ such that $\hat f(x) = f(x)$ for all $x \in X$.
-It is often the case that the form of the true underlying function $f$ is unknown, however it is still desirable to construct an approximation $\hat f$ with minimum approximation error at $y \notin X$.
-
-Multivariate regression is often used when the underlying function is presumed to be stochastic, or stochastic error is introduced in the evaluation of $f$.
-Hence, multivariate regression relaxes the conditions of interpolation by minimizing the error in $\hat f$ at $x \in X$ while maintaining some parametric form with parameters $P$.
-This can be written as $\min_{P} ||\hat f(X) - f(X)||$, where $f(X)$ is a vector of $f(x)$ for all $x \in X$ and $||\cdot||$ is an appropriate measure.
-The difficult question in the case of regression is often what parametric form to adopt for any given application.
-This paper proposes basis functions with overlapping regions of support as the general parametric form.
-
-As the dimension of data increases, the number of possible interactions between dimensions grows exponentially.
-Quantifying all possible interactions becomes intractable and hence beyond three-dimensional data, mostly linear models are used.
-That is not to say nonlinear models are absent, but nonlinearities are often either preconceived or model pairwise interactions between dimensions at most.
-Similarly, higher order models of single variables can be generated from the techniques proposed in this paper, however only first-order interactions between variables are considered.
-
-Regression and interpolation have a considerable theoretical base in one dimension \cite{cheney2009course}.
-Splines in particular are well understood as an interpolation technique in one dimension \cite{de1978practical}, particularly B-splines.
-Fewer techniques have been discussed and evaluated for two dimensional problems.
-Though in two and three dimensions tensor product splines remain computable \cite{unther1996interpolating}, tensor products have an unfortunate exponential scaling in parameterization with increasing dimension.
-Exponential scaling prohibits tensor products from being reasonably applied beyond three-dimensional data.
-In order to address this dimensional scaling challenge, C.
-de Boor and others have recently proposed box splines \cite{de2013box}.
-Two of the three meshes introduced in this work involve box-shaped regions of support and use box splines as the underlying basis functions.
-
-% Art is a subjective form of personal expression and computers are just robots that are going to take over the world.
-\textit{*deep breath*} In order to communicate my frustration with the technological and information revolution, I'm going to write the rest of this paper in poetry.
-
-% Star light, star bright \ldots
-
-\subsection{Box Splines}
+\section{Box Splines}
 \label{sec_box_splines}
@@
 \begin{figure}
-  \includegraphics[width=0.23\textwidth,height=2.5cm]{1D-linear.pdf}
-  \includegraphics[width=0.23\textwidth,height=2.5cm]{1D-quadratic.pdf}
+  \centering
+  \includegraphics[width=0.45\textwidth]{Figures/ACM/1D-linear.pdf}
+  \includegraphics[width=0.45\textwidth]{Figures/ACM/1D-quadratic.pdf}
   \caption{1D linear (order 2) and quadratic (order 3) box splines with direction vector sets $\bigl( 1 \ 1 \bigr)$ and $\bigl( 1 \ 1 \ 1 \bigr)$ respectively.
-Notice that these direction vector sets form the B-Spline analogues, order 2 composed of two linear components and order 3 composed of 3 quadratic components (colored and styled in plot).
-  \vspace{-.5cm}}
+Notice that these direction vector sets form the B-Spline analogues, order 2 composed of two linear components and order 3 composed of 3 quadratic components (colored and styled in plot).}
   \label{fig_1D_boxes}
+\end{figure}
+
+\begin{figure}
+  \centering
+  \includegraphics[width=0.45\textwidth]{Figures/ACM/2D-linear.pdf}
+  \includegraphics[width=0.45\textwidth]{Figures/ACM/2D-quadratic.pdf}
+  \caption{2D linear (order 2) and quadratic (order 3) box splines with direction vector sets $\bigl( I \ I \bigr)$ and $\bigl( I \ I \ I \bigr)$ respectively, where $I$ is the identity matrix in two dimensions.
+Notice that these direction vector sets also produce boxes with $\text{order}^2$ subregions (colored in plot).}
+  \label{fig_2D_boxes}
 \end{figure}
@@
 
-The application of box splines presented in this paper always utilizes the $d$-dimensional identity matrix as $A_d$.
-This simplifies the computation in Equation \ref{eq_box_base} to be the indicator function over the unit cube.
+The application of box splines presented in this chapter always utilizes the $d$-dimensional identity matrix as $A_d$.
+This simplifies the computation in Equation \ref{eq_box_base} to be the characteristic function for the unit cube.
 Composing $A$ strictly out of $k$ repetitions of the identity matrix forms the $k$th order B-spline with knots located at $0$, $1$, $\ldots$, $k-1$, $k$ along each axis (see Figure \ref{fig_1D_boxes}).
@@
 For a more thorough introduction and exploration of box splines in their more general form, readers are referred to \cite{de2013box}.
-
-\begin{figure}
-  \vspace{-.5cm}
-  \includegraphics[width=0.23\textwidth,height=3cm]{2D-linear.pdf}
-  \includegraphics[width=0.23\textwidth,height=3cm]{2D-quadratic.pdf}
-  \caption{2D linear (order 2) and quadratic (order 3) box splines with direction vector sets $\bigl( I \ I \bigr)$ and $\bigl( I \ I \ I \bigr)$ respectively, where $I$ is the identity matrix in two dimensions.
-Notice that these direction vector sets also produce boxes with $\text{order}^2$ subregions (colored in plot).
-  \vspace{-0.1cm}}
-  \label{fig_2D_boxes}
-\end{figure}
-
-\section{Interpolation and Regression}
-\label{sec_mesh_construction}
-
-Throughout this section, the notation will be reused from Section \ref{sec_introduction}.
-$X \subset \mathbb{R}^d$ is a finite set of points with known response values $f(x)$ for all $x \in X$.
-Also let $L,U \in \mathbb{R}^d$ define a bounding box for $X$ such that $L < x < U$ for all $x \in X$ (and comparison is componentwise).
-
-Define a box $b^c = (c,l^c,u^c)$ in $d$ dimensions with anchor $c \in \mathbb{R}^d$, lower width vector $l^c \in \mathbb{R}^d_+$, and upper width vector $u^c \in \mathbb{R}^d_+$ (where $u^c_i$ refers to the $i$th component of $u^c$).
-A visual example of a box in two dimensions can be seen in Figure \ref{fig_example_box}.
-Now, define a componentwise rescaling function $g: \mathbb{R}^d \rightarrow \mathbb{R}^d$ at point $x \in \mathbb{R}^d$ to be
-\begin{equation}
-  \bigl(g^c(x)\bigr)_r = {k \over 2} \left( 1 - {(x_r - c_r)_- \over l^c_r} + {(x_r - c_r)_+ \over u^c_r} \right),
-\end{equation}
-where $y_+=\max\{y,0\}$, $y_-=(-y)_+$, $k$ is the order of the box spline as described in Section \ref{sec_box_splines}.
-Finally, each box spline in a box mesh can be evaluated as $B^c(x)$, where $B^c(x) = B(g^c(x) \mid A)$ presuming the order of approximation implies $A$.
-Both box meshes described in the following subsections use box spline basis functions of this form.
-
-A notable property of boxes defined with the linear rescaling function $g^c$, is that $C^0$ and $C^1$ continuity of the underlying box spline are maintained.
-$C^0$ continuity is maintained through scaling.
-$C^1$ continuity is maintained for all box splines with $C^1$ continuity (order $\geq 3$) because the scaling discontinuity is located at $c$, where all box splines (of the presented form) have first derivative zero.
-All continuity beyond the first derivative is lost through the rescaling function $g^c$.
 
@@
 
-\subsection{Max Box Mesh}
+Throughout this section, the notation will be reused from Chapter \ref{ch:algs}.
+$X \subset \mathbb{R}^d$ is a finite set of points with known response values $f(x)$ for all $x \in X$.
+Also let $L,U \in \mathbb{R}^d$ define a bounding box for $X$ such that $L < x < U$ for all $x \in X$.
+
+Define a box $b^c = (c,l^c,u^c)$ in $d$ dimensions with anchor $c \in \mathbb{R}^d$, lower width vector $l^c \in \mathbb{R}^d_+$, and upper width vector $u^c \in \mathbb{R}^d_+$ (where $u^c_i$ refers to the $i$th component of $u^c$).
+A visual example of a box in two dimensions can be seen in Figure \ref{fig_example_box}.
+Now, define a componentwise rescaling function $g: \mathbb{R}^d \rightarrow \mathbb{R}^d$ at point $x \in \mathbb{R}^d$ to be
+\begin{equation}
+  \bigl(g^c(x)\bigr)_r = {k \over 2} \left( 1 - {(x_r - c_r)_- \over l^c_r} + {(x_r - c_r)_+ \over u^c_r} \right),
+\end{equation}
+where $y_+=\max\{y,0\}$, $y_-=(-y)_+$, $k$ is the order of the box spline as described in Section \ref{sec_box_splines}.
+Finally, each box spline in a box mesh can be evaluated as $B^c(x) = B(g^c(x) \mid A)$ presuming the order of approximation implies $A$.
+Both box meshes described in the following sections use box spline basis functions of this form.
+
+A notable property of boxes defined with the linear rescaling function $g^c$, is that $C^0$ and $C^1$ continuity of the underlying box spline are maintained.
+$C^0$ continuity is maintained through scaling.
+$C^1$ continuity is maintained for all box splines with $C^1$ continuity (order $\geq 3$) because the scaling discontinuity is located at $c$, where all box splines (of the presented form) have first derivative zero.
+All continuity beyond the first derivative is lost through the rescaling function $g^c$.
+
+\section{Max Box Mesh}
 
@@
 The same process is used to construct boxes around all control points in $C$.
-In order to improve the generality of the approximation, a set of control points is initially chosen to be well-spaced using a statistical method from \cite{amos2014algorithm}:
+In order to improve the generality of the approximation, a set of control points is initially chosen to be well-spaced using a statistical method from \citet{amos2014algorithm}:
 
@@
 \begin{definition}
-A box $b^y$ is the upper \textit{boundary} of box $b^x$ along dimension $r$ if and only if $u^x_r = (y_r - x_r)$.
-A box $b^y$ is the lower \textit{boundary} of box $b^x$ along dimension $r$ if and only if $l^x_r = (x_r - y_r)$.
+The hyperplane $x_r = c_r + u^c_r$ is the upper boundary of box $b^c$ along dimension $r$, and similarly $x_r = c_r - l^c_r$ is the lower boundary of box $b^c$.
+When the anchor point $y$, for some box $b^y$, lies in the hyperplane (and facet of box $b^c$) defining either boundary along dimension $r$ of $b^c$ it is said that $b^y$ \textit{bounds} $b^c$ in dimension $r$ and is denoted $\bigl( b^c \bigm\vert_r b^y \bigr)$.
 \end{definition}
@@
 Throughout all experiments and all repeated trials conducted for this study, all tested interpolation points were covered by at least one box in the $MBM$.
-However, it is possible for the $MBM$ to not form a covering of $(L,U)$ when there are cyclic boundaries.
+However, it is possible for the $MBM$ to not form a covering of $[L,U]$ when there are cyclic boundaries.
 Consider the following example in three dimensions:
 \begin{align*}
-  C         &= \bigl\{(0,0,0), (1,0,2/3), (1,1,4/3)\bigr\} \\
-  b^{c^{(1)}} &= \bigl( (0,0,0),  \ (*,*,*),  \ (1,*,4/3) \bigr) \\
-  b^{c^{(2)}} &= \bigl( (1,0,2/3),\ (1,*,*),  \ (*,1,*)   \bigr) \\
-  b^{c^{(3)}} &= \bigl( (1,1,4/3),\ (*,1,4/3),\ (*,*,*)   \bigr)
+  C         &= \bigl\{(0,0,0), (1,0,2/3), (1,1,4/3)\bigr\}, \\
+  b^{c^{(1)}} &= \bigl( (0,0,0),  \ (*,*,*),  \ (1,*,4/3) \bigr), \\
+  b^{c^{(2)}} &= \bigl( (1,0,2/3),\ (1,*,*),  \ (*,1,*)   \bigr), \\
+  b^{c^{(3)}} &= \bigl( (1,1,4/3),\ (*,1,4/3),\ (*,*,*)   \bigr).
 \end{align*}
@@
 The point $(2,2,-3)$ is not in any of the max boxes defined above.
-In this case, there is a cycle in box boundaries that looks like $\bigl( b^{c^{(1)}} \bigm\vert_1 b^{c^{(2)}} \bigm\vert_2 b^{c^{(3)}} \bigm\vert_3 b^{c^{(1)}} \bigr)$, where $\bigl( b^x \bigm\vert_r b^y \bigr)$ is used to denote ``$b^x$ is bounded on dimension $r$ by $b^y$''.
+In this case, there is a cycle in box boundaries that looks like $\bigl( b^{c^{(1)}} \bigm\vert_1 b^{c^{(2)}} \bigm\vert_2 b^{c^{(3)}} \bigm\vert_3 b^{c^{(1)}} \bigr)$.
 This example demonstrates that it is geometrically possible for the max box mesh to fail to cover a space, however experiments demonstrate that it is empirically unlikely.
@@
 
-\subsection{Iterative Box Mesh}
+\section{Iterative Box Mesh}
 
-The iterative box mesh ($IBM$) composes box-shaped regions that each contain strictly one control point just as in the $MBM$.
-However, the mesh is a covering for $(L,U)$ by construction and places boxes in a way that reduces apparent error.
-The boxes are constructed via the following procedure given a finite set of points $X \subset \mathbb{R}^d$, where $C \subseteq X$ is the (initially empty) set of control points
+The iterative box mesh ($IBM$) comprises box-shaped regions that each contain exactly one control point in their interior just as in the $MBM$.
+However, the mesh is a covering for $[L,U]$ by construction and places boxes in a way that reduces apparent error.
+The boxes are constructed via the following procedure given a finite set of points $X \subset \mathbb{R}^d$, where $C \subseteq X$ is the (initially empty) set of control points.
 \begin{enumerate}
-\item Add a box that covers $(L,U)$ anchored at the most central point $x^{(1)} \in X$ to $C$ and fit the $IBM$ to all $X$.
-\item Add a new box that covers $(L,U)$ anchored at $x^{(i)} \notin C$ such that $x^{(i)} = \max_{x \in X \backslash C} \bigl| \hat f(x) - f(x) \bigr|$, reshaping all boxes $b^{x^{(j)}} \in C$ that contain $x^{(i)}$ by bounding the first dimension $r$ such that $\bigl| x^{(j)}_r - x^{(i)}_r \bigr| = \bigl\| x^{(j)} - x^{(i)} \bigr\|_\infty$ (also reshaping the box $b^{x^{(i)}}$ symmetrically) and then fit the mesh to all $X$.
+\item Add the box that covers $[L,U]$ anchored at the most central point $x^{(k)} \in X$, add $x^{(k)}$ to $C$, and least squares fit the $IBM$ model to all $x \in X$.
+\item Add a new box $[L,U]$ anchored at $x^{(i)} \notin C$ such that $\bigl| IBM(x^{(i)}) - f(x^{(i)}) \bigr| = \max_{x \in X \backslash C} \bigl| IBM(x) - f(x) \bigr|$, reshaping all boxes $b^{x^{(j)}}$ that contain $x^{(i)}$ by bounding the first dimension $r$ such that $\bigl| x^{(j)}_r - x^{(i)}_r \bigr| = \bigl\| x^{(j)} - x^{(i)} \bigr\|_\infty$ (also reshaping the box $b^{x^{(i)}}$ symmetrically), add $x^{(i)}$ to $C$, and then least squares fit the $IBM$ model to all $x \in X$.
 \label{step_add_box}
-\item Repeat Step \ref{step_add_box} until approximation error is below tolerance $t$.
+\item Repeat Step \ref{step_add_box} until model approximation error is below tolerance $t$.
 \end{enumerate}
 
-Just as the $MBM$, the parameters can be identified via a least squares solve.
+Just as for the $MBM$, the parameters can be identified via a least squares fit.
 The iterative box mesh is used to generate a $|X| \times |C|$ matrix $M$ of box spline function evaluations at all points in $X$.
-Now the set of coefficients for each box is the solution to the least squares problem $\min_P \bigl|\bigl| M\ P - f(X) \bigr|\bigr|_2$.
+Now the box spline coefficients are the solution to the least squares problem $\min_P \bigl|\bigl| M\ P - f(X) \bigr|\bigr|_2$.
 Also as for the $MBM$, $C = X$ causes $M$ to equal the $|X| \times |X|$ identity, making the iterative box mesh approximation $\hat f$ an interpolant.
@@
 The user is left to decide the most appropriate error tolerance, however a decision mechanism and analysis is presented in Section \ref{sec_performance_analysis}.
-As mentioned earlier, the iterative box mesh is a covering for $(L,U)$ by construction and this can be proved by an inductive argument.
+As mentioned earlier, the iterative box mesh is a covering for $[L,U]$ by construction and this can be proved by an inductive argument.
 
-After fitting for parameters, an $IBM$ can generate approximations for a set of new points $Z \subset \mathbb{R}^d$ by evaluating all box basis functions multiplied by the coefficients, $P$.
+An $IBM$ least squares fit $\hat f(z) = \sum_{j}P_j B^{c^{(i_j)}}(z)$ can generate approximations at new points $z \in Z \subset \mathbb{R}^d$ by evaluating $\hat f(z)$.
 The computational complexity for generating the mesh is $\mathcal{O}(c^2 n d)$ where $c$ is the number of control points determined by the minimum error threshold and $n = |X|$.
-The computational complexity of evaluating the mesh at a singple point is $\mathcal{O}(c d)$.
+The computational complexity of evaluating the mesh at a single point is $\mathcal{O}(c d)$.
 
-\subsection{Voronoi Mesh}
+\section{Voronoi Mesh}
 
-The final of the three meshes utilizes $\ell2$ norm distances to define boundaries rather than max norm distances.
+The final of the three meshes utilizes 2-norm distances to define boundaries rather than max norm distances.
+It also does not rely on box splines as the underlying basis function.
 A well-studied technique for classification and approximation is the nearest neighbor algorithm \cite{cover1967nearest}.
@@
 
-$$ V^{x^{(i)}}(y) = \left(1 - \frac{||y - x^{(i)}||_2}{2 \ d(y - x^{(i)} \mid x^{(i)})} \right)_+, $$
-where $x^{(i)}$ is the center of the Voronoi cell, $y \in \mathbb{R}^d$ is an interpolation point, and $d(w \mid x^{(i)})$ is the distance between $x^{(i)}$ and the boundary of the Voronoi cell $v^{x^{(i)}}$ in the direction $w$.
-$V^{x^{(i)}}\bigl(x^{(j)}\bigr) = 0$ and $V^{x^{(i)}}$ has local support.
-While all basis functions $V^{x^{(i)}}(x^{(i)}) = 1$, the $2$ in the denominator causes all basis functions to go linearly to $0$ at the boundary of the twice-expanded Voronoi cell.
+$$ V^{x^{(i)}}(y) = \left(1 - \frac{\bigl\|y - x^{(i)}\bigr\|_2}{2 \ d(y \mid x^{(i)})} \right)_+, $$
+where $x^{(i)}$ is the center of the Voronoi cell, $y \in \mathbb{R}^d$ is an interpolation point, and $d(y \mid x^{(i)})$ is the distance between $x^{(i)}$ and the boundary of the Voronoi cell $v^{x^{(i)}}$ in the direction $y - x^{(i)}$.
+$V^{x^{(i)}}\bigl(x^{(j)}\bigr) = \delta_{ij}$ and $V^{x^{(i)}}$ has local support.
+While $V^{x^{(i)}}(x^{(i)}) = 1$, the $2$ in the denominator causes all basis functions to go linearly to $0$ at the boundary of the twice-expanded Voronoi cell.
 Note that this basis function is $C^0$ because the boundaries of the Voronoi cell are $C^0$.
@@
 While the cost of computing the exact Voronoi cells for any given set of points grows exponentially \cite{dutour2009complexity}, the calculation of $d$ is linear with respect to the number of control points and dimensions.
-Given any center $x^{(i)} \in \mathbb{R}^d$, set of control points $C \subseteq X$, and interpolation point $y \in \mathbb{R}^d$, $d\bigl((y-x^{(i)}) \mid x^{(i)}\bigr)$ is the solution to
+Given any center $x^{(i)} \in \mathbb{R}^d$, set of control points $C \subseteq X$, and interpolation point $y \in \mathbb{R}^d$, $d\bigl(y \mid x^{(i)}\bigr)$ is the solution to
 \begin{equation}
-  \max_{\{c \in C\}\backslash x^{(i)}} \frac{\|y\|_2}{2} \ \frac{y \cdot \bigl(c - x^{(i)}\bigr) - x^{(i)} \cdot \bigl(c - x^{(i)}\bigr)}{c \cdot \bigl(c - x^{(i)}\bigr) - x^{(i)} \cdot \bigl(c - x^{(i)}\bigr)}.
+  \max_{c \in C\backslash\{x^{(i)}\}} \frac{\bigl\|y - x^{(i)}\bigr\|_2}{2} \ \frac{y \cdot \bigl(c - x^{(i)}\bigr) - x^{(i)} \cdot \bigl(c - x^{(i)}\bigr)}{c \cdot \bigl(c - x^{(i)}\bigr) - x^{(i)} \cdot \bigl(c - x^{(i)}\bigr)}.
 \end{equation}
@@
 The Voronoi mesh is used to generate a $|X| \times |C|$ matrix $M$ of basis function evaluations at all points in $X$.
-Now the set of coefficients for each basis function is the solution to the least squares problem $\min_P \bigl\| M \ P - f(X) \bigr\|_2$.
+Now the $VM$ coefficients are the solution to the least squares problem $\min_P \bigl\| M \ P - f(X) \bigr\|_2$.
 When $X = C$, $M$ is the identity making the mesh an interpolant.
-Bootstrapping can be performed with an identical procedure to that of the $IBM$.
+Bootstrapping can be performed with an identical procedure to that for the $IBM$.
 \begin{enumerate}
-\item Pick the most central point $x^{(1)} \in X$ to be the first control point in $C$ and fit the $VM$ to all $X$.
-\item Add a control point $x^{(i)} \notin C$ such that $x^{(i)} = \max_{x \in X \backslash C} \bigl| \hat f(x) - f(x) \bigr|$ and then fit the $VM$ to all $X$. \label{step_add_control}
+\item Pick the most central point $x^{(k)} \in X$ to be the first control point in $C$ and fit the $VM$ model to all $x \in X$.
+\item Identify a control point $x^{(i)} \notin C$ such that $\bigl| VM(x^{(i)}) - f(x^{(i)}) \bigr| = \max_{x \in X \backslash C} \bigl| VM(x) - f(x) \bigr|$, add $x^{(i)}$ to $C$, and then fit the $VM$ model to all $x \in X$.
+\label{step_add_control}
 \item Repeat Step \ref{step_add_control} until approximation error is below tolerance $t$.
@@
 
-Any $VM$ is na\"{\i}vely a covering for $(L,U)$, since any possible interpolation point will have a nearest neighbor control point.
+Any $VM$ is na\"{\i}vely a covering for $[L,U]$, since any possible interpolation point will have a nearest neighbor control point.
 The computational complexity of evaluating a parameterized Voronoi mesh with $c$ control points is $\mathcal{O}(c^2 d)$.
 Bootstrapping the generation of a Voronoi mesh requires $\mathcal{O}(c^2 n d)$ computations for a maximum number of basis functions $c$ determined by the error threshold.
+
+% Art is a subjective form of personal expression and computers are just robots that are going to take over the world.
+\textit{*deep breath*} In order to communicate my frustration with the technological and information revolution, I'm going to write the rest of this paper in poetry.
+
+% Star light, star bright \ldots
 
@@
 \section{Data and Analysis}
-In order to evaluate the proposed interpolation and approximation techniques, this paper utilizes three data sets of varying dimension and application.
+
+Some data sets will be used to evaluate the interpolation and regression meshes proposed above.
+This chapter utilizes three data sets of varying dimension and application.
 In the following subsections the sources and targets of each data set are described, as well as challenges and limitations related to interpolating and approximating these data sets.
@@
 \begin{figure}
-  \includegraphics[width=.5\textwidth,height=2cm]{p-hist.pdf}
-  \includegraphics[width=.5\textwidth,height=2cm]{f-hist.pdf}
-  \includegraphics[width=.5\textwidth,height=2cm]{h-hist.pdf}
+  \centering
+  \includegraphics[width=.8\textwidth]{Figures/ACM/p-hist.pdf}
+  \includegraphics[width=.8\textwidth]{Figures/ACM/f-hist.pdf}
+  \includegraphics[width=.8\textwidth]{Figures/ACM/h-hist.pdf}
   \caption{Histograms of Parkinsons (total UPDRS), forest fire (area), and HPC I/O (mean throughput) response values respectively.
@@
 
-\subsection{Data Summary}
-
-\subsubsection{High Performance Computing I/O}
-$$N = 532, D = 4$$
+\subsection{High Performance Computing I/O ($n = 532, d = 4$)}
 The first of three data sets is a four-dimensional data set produced by executing the IOzone benchmark from \cite{iozone} on a homogeneous cluster of computers.
@@
 A single IOzone execution reports the max I/O file-read throughput seen.
-The 40 executions for each system configuration are converted to their mean, which is capable of being modeled by each of the multivariate approximation techniques presented in Section \ref{sec_mesh_construction}.
-The four dimensions being modeled to predict throughput mean are file size, record size, thread count, and frequency.
+The 40 executions for each system configuration are converted to their mean, which is capable of being modeled by each of the multivariate approximation techniques presented earlier in this chapter.
+The four dimensions being modeled to predict throughput mean are file size, record size, thread count, and CPU frequency.
 
-\subsubsection{Forest Fire}
-$$N = 517, D = 12$$
+\begin{figure*}[htb]
+  \centering
+  \begin{tikzpicture}
+    \node (img)  {\includegraphics[width=0.95\textwidth,height=4cm]{Figures/ACM/eval-times.pdf}};
+    \node[below=of img, node distance=0cm, yshift=1cm] {Relative Error Tolerance while Bootstrapping};
+    \node[left=of img, node distance=0cm, rotate=90, anchor=center,yshift=-0.7cm] {Fit Time};
+  \end{tikzpicture}
+  \caption{Time required to generate model fits for each technique with varying relative error tolerance during bootstrapping.
+    \vspace{-.3cm}}
+  \label{fig_eval_times}
+\end{figure*}
 
-The forest fire data set provided by \cite{cortez2007data}, describes the area of Montesinho park burned on specific days and months of the year in terms of the environmental conditions.
+\subsection{Forest Fire ($n = 517, d = 12$)}
+The forest fire data set \cite{cortez2007data} describes the area of Montesinho park burned on specific days and months of the year in terms of the environmental conditions.
 The twelve dimensions being used to model burn area are the $x$ and $y$ spatial coordinates of burns in the park, month and day of year, the FFMC, DMC, DC, and ISI indices (see source for details), the temperature in Celsius, relative humidity, wind speed, and outdoor rain.
@@
 
-\subsubsection{Parkinson's Telemonitoring}
-$$N = 468, D = 16$$
-
-The final data set for evaluation is derived from a speech monitoring study with the intent to automatically estimate Parkinson's disease symptom development in Parkinson's patients from \cite{tsanas2010accurate}.
-Particularly, a time-consuming clinical evaluation measure referred to as the UPDRS score.
+\subsection{Parkinson's Telemonitoring ($n = 468, d = 16$)}
+The final data set for evaluation \cite{tsanas2010accurate} is derived from a speech monitoring study with the intent to automatically estimate Parkinson's disease symptom development in Parkinson's patients.
+The function to be predicted is a time-consuming clinical evaluation measure referred to as the UPDRS score.
 The total UPDRS score given by a clinical evaluation is estimated through 16 real numbers generated from biomedical voice measures of in-home sound recordings.
@@
 
-\begin{figure*}[htb]
-  \begin{tikzpicture}
-    \node (img)  {\includegraphics[width=0.95\textwidth,height=9.5cm]{all-performance.pdf}};
-    \node[below=of img, node distance=0cm, yshift=1cm] {Relative Error Tolerance while Bootstrapping};
-    \node[left=of img, node distance=0cm, rotate=90, anchor=center,yshift=-0.7cm] {Signed Relative Error};
-  \end{tikzpicture}
-  \caption{The performance of all three techniques with varied relative error tolerance for the bootstrapping parameter.
-The columns are for Max Box Mesh, Iterative Box Mesh, and Voronoi Mesh, respectively.
-The rows are for HPC I/O, Forest Fire, and Parkinson's respectively.
-Notice that the techniques exhibit counterintuitive behavior on the Parkinson's and Forest Fire data sets, performance increases with larger error tolerance.}
-  \label{fig_all_performance}
-\end{figure*}
-
 The performance of the approximation techniques varies considerably across the three evaluation data sets.
-Relative errors for the most na\"{\i}ve approximators such as nearest neighbor can range $\displaystyle [0, \big(\max_x f(x) - \min_x f(x)\big) / \min_x f(x) ]$ when modeling a nonnegative function $f(x)$ from data.
+Relative errors for the most na\"{\i}ve approximators such as nearest neighbor can range from zero to $\displaystyle \big(\max_x f(x) - \min_x f(x)\big) / \min_x f(x)$ when modeling a positive function $f(x)$ from data.
 Each of the approximation techniques presented remain within these bounds and all errors are presented in signed relative form $(\hat f(x) - f(x)) / f(x)$.
-Before the models are constructed all data $X$ values are shifted and scaled to be in the unit cube $[0,1]$, while the response values are taken in their original form.
+Before the models are constructed all data values (components $x^{(i)}_r$ of $x^{(i)} \in X$) are shifted and scaled to be in the unit cube $[0,1]^d$, while the response values are taken in their original form.
 All models are evaluated with $10$ random $80/20$ splits of the data.
@@
 Notice that large relative error tolerances occasionally yield even lower evaluation errors, demonstrating the benefits of approximation over interpolation for noisy data sets.
-  \vspace{-1cm}}
+  \vspace{-.5cm}}
   \label{tab_optimal_tolerance}
@@
 
-Each of the approximation techniques presented incorporates a bootstrapping based on an allowable error tolerance $t$.
+Each of the approximation techniques presented incorporates bootstrapping based on an allowable error tolerance $t$.
 An analysis of the effects of bootstrapping error tolerances on validation accuracy can be seen in Figure \ref{fig_all_performance}.
-The approximation meshes perform best on the forest fire and Parkinson's data sets when the error tolerance used for fitting is large (regressing rather than interpolating), while near-interpolation generally produces the most accurate models for HPC I/O.
+The approximation meshes perform best on the forest fire and Parkinson's data sets when the error tolerance used for fitting is large (smoothing rather than interpolating), while near-interpolation generally produces the most accurate models for HPC I/O.
 Another performance result of note is that the $MBM$ and $IBM$ have very similar basis functions with largely different outputs.
 
-\begin{figure*}[htb]
+\begin{figure*}
   \begin{tikzpicture}
-    \node (img)  {\includegraphics[width=0.95\textwidth,height=9.5cm]{perf-sample.pdf}};
+    \node (img)  {\includegraphics[width=0.95\textwidth,height=9.5cm]{Figures/ACM/all-performance.pdf}};
+    \node[below=of img, node distance=0cm, yshift=1cm] {Relative Error Tolerance while Bootstrapping};
+    \node[left=of img, node distance=0cm, rotate=90, anchor=center,yshift=-0.7cm] {Signed Relative Error};
+  \end{tikzpicture}
+  \caption{The performance of all three techniques with varied relative error tolerance for the bootstrapping parameter.
+The columns are for Max Box Mesh, Iterative Box Mesh, and Voronoi Mesh, respectively.
+The rows are for HPC I/O, Forest Fire, and Parkinson's respectively.
+Notice the techniques' behavior on the Parkinson's and Forest Fire data sets, performance increases with larger error tolerance.}
+  \label{fig_all_performance}
+\end{figure*}
+
+\begin{figure*}
+  \begin{tikzpicture}
+    \node (img)  {\includegraphics[width=0.95\textwidth,height=9.5cm]{Figures/ACM/perf-sample.pdf}};
     \node[below=of img, node distance=0cm, yshift=1cm] {Signed Relative Error in Prediction};
@@
 The rows are for HPC I/O, Forest Fire, and Parkinson's respectively.
-Notice that the Max Box Mesh exibits the least skewed distribution of errors across all three data sets.
     \vspace{-.3cm}}
@@
 
-\begin{figure*}
-  \begin{tikzpicture}
-    \node (img)  {\includegraphics[width=0.95\textwidth,height=4cm]{eval-times.pdf}};
-    \node[below=of img, node distance=0cm, yshift=1cm] {Relative Error Tolerance while Bootstrapping};
-    \node[left=of img, node distance=0cm, rotate=90, anchor=center,yshift=-0.7cm] {Fit Time};
-  \end{tikzpicture}
-  \caption{Time required to generate model fits for each technique with varying relative error tolerance during bootstrapping.
-    \vspace{-.3cm}}
-  \label{fig_eval_times}
-\end{figure*}
+\section{Discussion of Mesh Approximations}
 
-\section{Discussion}
-
-The bootstrapping procedure presented for each approximation technique still has much room for improvement.
-Initial analysis suggests that the appropriate relative error tolerance needs to be discovered empirically for each application of a modeling technique.
+The bootstrapping procedure presented for each quasi-mesh approximation is very computationally expensive and does not provide much improvement over the interpolatory approach.
+Analysis suggests that the appropriate relative error tolerance needs to be discovered empirically for each application of a modeling technique.
 Further analytic studies could arrive at methods for determining optimal error tolerances at runtime, however increases in runtime complexity may not be afforded in many applications.
 
-The box-shaped basis functions and the construction algorithms used for the $MBM$ and $IBM$ could become a source of error when $d$ (the dimension of the data $X$) is much larger than $n$ (the number of known points).
+The box-shaped basis functions and the construction algorithms used for the $MBM$ and $IBM$ could become a source of error when $d$ (the dimension of the data $X$) is comparable to $n$ (the number of known points).
 The blending regions in which multiple basis functions overlap are always axis aligned and in applications such as image analysis, any single dimension may be unsuitable for approximating the true underlying function.
 The Voronoi mesh attempts to address this problem by utilizing boundaries between points in multiple dimensions simultaneously.
-However, it is empirically unclear whether the true benefits of the $VM$ are seen in applications where $d < n$.
+However, it is empirically unclear whether the true benefits of the $VM$ are seen in applications where $d \ll n$.
 
 Each of the case studies presented have fewer than $1000$ points.
-The complexities of the presented approximation techniques are suitable for large dimension, but the increased complexity associated with brute-force bootstrapping currently prohibits their use on larger data sets.
-The Voronoi mesh in particular has a large complexity with respect to $n$ which could be significantly improved via more greedy bootstrapping.
+The complexity of the presented approximation techniques are suitable for large dimension, but the increased complexity associated with brute-force bootstrapping prohibits their use on larger data sets.
+The Voronoi mesh in particular has a large complexity with respect to $n$ which could be significantly improved by dropping bootstrapping, as is done in Chapter \ref{ch:strong}.
 While each technique requires less than ten seconds on average to produce a fit in the presented case studies, the fit time required quickly grows into minutes around $1000$ points.
-While these initial results appear somewhat limiting, they demonstrate the viability of each mesh and leave room for further theoretical exploration of techniques to reduce the runtime complexity while maintaining the approximation power and flexibility.
+These results demonstrate the limits of expensive bootstrapping.
+However, the viability of each mesh encourages the further applications explored in later chapters.
 
-\section{Conclusion}
+\section{Implications of Quasi-Mesh Results}
 
 The Max Box Mesh, Iterative Box Mesh, and Voronoi Mesh each provide novel strategies for effectively approximating multivariate phenomonon.
-The underlying constructions are theoretically straightforward, yet powerful and flexible.
-The computational complexities of each make them particularly suitable for applications in many dimensions, while the bootstrapping error tolerance parameter allows a balance between approximation and interplation to be explored empirically with each application.
-
-\subsection{Future Work}
-
-A thorough comparison with constituent multivariate approximation techniques including but not limited to, Linear Shepard interpolation, Multivariate Adaptive Regression Splines, Multi Layer Perceptron regression, and Delaunay Triangulation constitutes future work.
-A more detailed study of alternative bootstrapping techniques may also provide valuable insight.
+The underlying constructions are theoretically straightforward.
+The computational complexities of each make them particularly suitable for applications in many dimensions, while the bootstrapping error tolerance parameter allows a balance between smoothing and interpolation to be explored empirically with each application.
+However, the expense of bootstrapping prohibits its use on larger data sets.

### 2020-04-14 3-Disseration/error.tex
Updated min wiggle fit tests, wrote report on the topic that was sent to Dr. Watson.

--- 2019-03-31 Paper_[2019-02]_Numerical_Algorithms/paper.tex
@@
-\RequirePackage{fix-cm}
-\documentclass[smallextended,final]{svjour3}  % onecolumn (second format)
-\smartqed  % flush right qed marks, e.g. at end of proof
-
-% Insert here the call for the packages your document requires
-\usepackage{graphicx}
-\usepackage{tikz}     % For adding axes to plots
-\usetikzlibrary{positioning}
-\usepackage{graphicx} % For including PDFs
-\usepackage{amsmath}  % assumes amsmath package installed
-\usepackage{amssymb}  % assumes amsmath package installed
-\usepackage{booktabs} % For formal tables
-\usepackage{tabularx} % For tables with nice column widths.
-\usepackage[colorlinks,allcolors=blue]{hyperref}
-\usepackage{float}    % For controlling the placement of tables.
-
-%
-% please place your own definitions here and don't use \def but
-% \newcommand{}{}
-
-%% Command for creating multiplie-line cells inside a table environment
-\newcommand{\multilinecell}[2][c]{%
-  \begin{tabular}[#1]{@{}c@{}}#2\end{tabular}}
-
-\newcolumntype{d}[1]{D{.}{.}{#1}}
-
-%% Create a ``Lemma'' environment that does not have italics font.
-\spnewtheorem{plainlemma}{Lemma}{\bf}{}
-%% Create a ``Theorem'' environment that is not numbered.
-\spnewtheorem*{onetheorem}{Theorem}{\bf}{}
-%% Create a ``Proof'' environment that starts with 'Proof.'
-\spnewtheorem*{proofdot}{Proof.}{\it}{}
-
-%% \newcommand{\Bigl}{\mathopen\Big}
-%% \newcommand{\Bigr}{\mathclose\Big}
-%
-% Insert the name of "your journal" with
-\journalname{Numerical Algorithms}
-%
-\begin{document}
-
-\title{Interpolation of Sparse High-Dimensional Data\thanks
-  {This work was supported by the National Science Foundation Grant CNS-1565314.}
-}
-% \subtitle{Do you have a subtitle?\\ If so, write it here}
-%\titlerunning{Short form of title}        % if too long for running head
-
-\author{Thomas C. H. Lux   \and
-  Layne T. Watson          \and
-  Tyler H. Chang           \and
-  Yili Hong                \and
-  Kirk Cameron
-}
-
-%\authorrunning{Short form of author list} % if too long for running head
-
-\institute{Thomas Lux \at
-  Virginia Polytechnic Institute \& State University (VPI\&SU) \\
-  Blacksburg, VA 24061 \\
-  \email{tchlux@vt.edu}
-}
-
-\date{Received: date / Accepted: date}
-% The correct dates will be entered by the editor
-
-\maketitle
-
-\begin{abstract}
-Increases in the quantity of available data have allowed all fields of
-science to generate more accurate models of multivariate phenomena.
-Regression and interpolation become challenging when the dimension of
-data is large, especially while maintaining tractable computational
-complexity. Regression is a popular approach to solving approximation
-problems with high dimension, however there are often advantages to
-interpolation. This paper presents a novel and insightful error bound
-for (piecewise) linear interpolation in arbitrary dimension and
-contrasts the performance of some interpolation techniques with
-popular regression techniques. Empirical results demonstrate the
-viability of interpolation for moderately high dimensional
-approximation problems, and encourage broader application of
-interpolants to multivariate approximation in science.
-
-\keywords{approximation \and regression \and interpolation \and high
-  dimension \and error bound}
-% \PACS{PACS code1 \and PACS code2 \and more}
-% \subclass{MSC code1 \and MSC code2 \and more}
-\end{abstract}
-
-%     Introduction     
-%======================
-
-%% ===================================================================
-\section{Introduction}
-\label{sec:introduction}
-
-Regression and interpolation are problems of considerable importance
-that find applications across many fields of science. Pollution and
-air quality analysis \cite{de2008field}, energy consumption management
-\cite{lazos2014optimisation}, and student performance prediction
-\cite{cortez2008using,lux2016applications} are a few of many
-interdisciplinary applications of multivariate regression for
-predictive analysis. As discussed later, these techniques can also be
-applied to prediction problems related to forest fire risk assessment
-\cite{cortez2007data}, Parkinson's patient clinical evaluations
-\cite{tsanas2010accurate}, local rainfall and weather
-\cite{williams2009rattle}, credit card transactions
-\cite{pozzolo2015calibrating}, and high performance computing (HPC)
-file input/output (I/O) \cite{lux2018nonparametric}.
-
-Regression and interpolation have a considerable theoretical base in
-one dimension \cite{cheney2009course}. Splines in particular are well
-understood as an interpolation technique in one dimension
-\cite{de1978practical}, particularly B-splines. Tensor products of
-B-splines \cite{unther1996interpolating} or other basis functions have
-an unfortunate exponential scaling with increasing
-dimension. Exponential scaling prohibits tensor products from being
-reasonably applied beyond three-dimensional data. In order to address
-this dimensional scaling challenge, C. de Boor and others proposed box
-splines \cite{de2013box}, of which one of the approximation techniques
-in this work is composed \cite{lux2018novel}.
-
-The theoretical foundation of low dimensional interpolation allows the
-construction of strong error bounds that are absent from high
-dimensional problems. This work extends some known results regarding
-the secant method \cite{dennis1996numerical} to construct an
-interpolation error bound for problems of arbitrary dimension. These
-error bounds are useful, considering the same cannot be said for
-regression algorithms in general. The maximum complexity of an
-interpolant is bounded by the amount of data available, while the
-maximum complexity of a regressor is bounded by both the amount of
-data and the chosen parametric form. For this reason, generic uniform
-bounds are largely unobtainable for regression techniques on arbitrary
-approximation problems, even when the approximation domain is bounded.
-
-Aside from theoretical motivation for the use of interpolants, there
-are often computational advantages as well. Interpolants do not have
-the need for \textit{fitting} data, or minimizing error with respect
-to model parameters. In applications where the amount of data is large
-and the relative number of predictions that need to be made for a
-given collection of data is small, the direct application of an
-interpolant is much less computationally expensive.
-
-In this work, multivariate interpolation is defined given a closed
-convex subset $Y$ of a metrizable topological vector space with metric
-$s$, some function $f:\mathbb{R}^d \rightarrow Y$ and a set of points
-$X = \bigl\{x^{(1)}$, $\ldots$, $x^{(n)}\bigr\} \subset \mathbb{R}^d$,
-along with associated function values $f\bigl(x^{(i)}\bigr)$. The goal
-is to construct an approximation $\hat f: \mathbb{R}^d \rightarrow Y$
-such that $\hat f\bigl(x^{(i)}\bigr) = f\bigl(x^{(i)}\bigr)$ for all
-$i = 1$, $\ldots$, $n$. It is often the case that the form of the true
-underlying function $f$ is unknown, however it is still desirable to
-construct an approximation $\hat f$ with small approximation error at
-$y \notin X$. The two metric spaces that will be discussed in this
-work are the real numbers with metric $s(x,y) = |x-y|$, and the set of
-cumulative distribution functions (CDFs) with the Kolmogorov-Smirnov
-(KS) statistic \cite{lilliefors1967kolmogorov} as metric.
-
-Multivariate regression is often used when the underlying function is
-presumed to be stochastic, or stochastic error is introduced in the
-evaluation of $f$. Hence, multivariate regression relaxes the
-conditions of interpolation by choosing parameters $P$ defining $\hat
-f(x;P)$ to minimize the error vector $\Bigl( \bigl | \hat f
-\bigl(x^{(1)};P\bigr) - f\bigl(x^{(1)}\bigr) \bigr|$, $\ldots$, $\bigl
-| \hat f \bigl(x^{(n)}; P\bigr) - f\bigl(x^{(n)}\bigr) \bigr | \Bigr)$
-in some norm. The difficult question in the case of regression is
-often what parametric form to adopt for any given application.
-
-The most challenging problem when scaling in dimension is that the
-number of possible interactions between dimensions grows
-exponentially. Quantifying all possible interactions becomes
-intractable, and hence beyond three-dimensional data mostly linear
-models are used. That is not to say nonlinear models are absent, but
-nonlinearities are often either preconceived or model pairwise
-interactions between dimensions at most. Even globally nonlinear
-approximations such as neural networks are constructed from
-compositions of summed low-interaction functions
-\cite{clevert2015fast}.
-
-Provided the theoretical and practical motivations for exploring
-interpolants, this work aims to study the empirical performance
-differences between a set of scalable (moderately) interpolation
-techniques and a set of common regression techniques. These techniques
-are applied to a collection of moderately high dimensional problems
-($5 \le d \le 30$) and the empirical results are discussed.
-
-The remainder of this paper is organized as follows. Sections
-\ref{sec:regression} and \ref{sec:interpolation} present the
-multivariate models. Section \ref{sec:error} gives the error measuring
-methodology that is used for collecting empirical results. Section
-\ref{sec:theory} contains the theoretical error bounds for
-interpolation. Section \ref{sec:data} presents the data sets for
-empirical approximation analysis and the approximation results for all
-models on these data sets. Section \ref{sec:discussion} analyzes and
-discusses the results, their implications, and their
-limitations. Finally, Section \ref{sec:conclusion} concludes.
-
-%% ===================================================================
-\section{Multivariate Regression}
-\label{sec:regression}
-Multivariate regressors are capable of accurately modeling a complex
-dependence of a response (in $Y$) on multiple variables (represented
-as a points in $\mathbb{R}^{d}$). The approximations to some (unknown)
-underlying function $f: \mathbb{R}^d \rightarrow Y$ are chosen to
-minimize some error measure related to data samples
-$f\bigl(x^{(i)}\bigr)$. For example, least squares regression uses the
-sum of squared differences between modeled function values and true
-function values as an error measure. In this section and the next,
-some techniques are limited to approximating real valued functions ($Y
-\subset \mathbb{R}$). These techniques can be extended to real
-vector-valued ranges by repeating the construction for each component
-of the vector output. Throughout the following, $x$ denotes a
-$d$-tuple, $x_i$ the $i$th component of $x$, and $x^{(i)}$ the $i$th
-$d$-tuple data point. Different symbols are used to represent the
-approximation function $\hat f$.
-
-\subsection{Multivariate Adaptive Regression Splines}
-\label{sec:mars}
-This approximation was introduced in \cite{friedman1991multivariate}
-and subsequently improved to its current version in
-\cite{stanford1993fast}, called fast multivariate adaptive regression
-splines (Fast MARS). In Fast MARS, a least squares fit model is
-iteratively built by beginning with a single constant valued function
-and adding two new basis functions at each iteration of the form
-\begin{align*}
-  B_{2j-1}(x) &= B_l(x) \bigl(x_i-x^{(p)}_i\bigr)_+, \\
-  B_{2j}(x) &= B_k(x) \bigl(x_i-x^{(p)}_i\bigr)_- ,
-\end{align*}
-where $j$ is the iteration number, $1 \le p \le n$, and $B_l(x)$,
-$B_k(x)$ are basis functions from the previous iteration,
- $$w_+ = \begin{cases} w, & w \geq 0 \\ 0, & w < 0 \end{cases},$$
-and $w_- = (-w)_+$. After iteratively constructing a model, MARS then
-iteratively removes basis functions that do not contribute to goodness
-of fit. In effect, MARS creates a locally component-wise tensor
-product approximation of the data. The overall computational
-complexity of Fast MARS is $\mathcal{O}(n d m^3)$ where $m$ is the
-maximum number of underlying basis functions. This paper uses an
-implementation of Fast MARS \cite{rudy2017pyearth} with $m = 200$
-throughout all experiments.
-
-\subsection{Support Vector Regressor}
-\label{sec:svr}
-Support vector machines are a common method used in machine learning
-classification tasks that can be adapted for the purpose of regression
-\cite{basak2007support}. How to build a support vector regressor (SVR)
-is beyond the scope of this summary, but the resulting functional fit
-$p : \mathbb{R}^d \rightarrow \mathbb{R}$ has the form
- $$ p(x)  = \sum_{i=1}^{n}a_i K\bigl(x,x^{(i)}\bigr) + b ,$$
-where $K$ is the selected kernel function, $a_i \in \mathbb{R}^n$, 
-$b \in \mathbb{R}$ are coefficients to be solved for simultaneously.
-The computational complexity of the SVR is $\mathcal{O}(n^2dm)$, with
-$m$ being determined by the minimization convergence criterion. This
-paper uses the scikit-learn SVR \cite{scikit-learn} with a polynomial
-kernel function.
-
-\subsection{Multilayer Perceptron Regressor}
-\label{sec:mlp}
-The neural network is a well studied and widely used method for both
-regression and classification tasks
-\cite{hornik1989multilayer,rumelhart1988learning}. When using the
-rectified linear unit (ReLU) activation function
-\cite{dahl2013improving} and fitting the model with stochastic
-gradient descent (SGD) or BFGS minimization techniques
-\cite{goh2017why,moller1993scaled,robbins1951stochastic}, the model
-built by a multilayer perceptron uses layers $l : \mathbb{R}^{i}
-\rightarrow \mathbb{R}^{j}$ defined by
- $$ l(u) = \bigl( u^t W_l + c_l \bigr)_+ ,$$
-where $c_l \in \mathbb{R}^j$ and $W_l$ is the $i \times j$ weight
-matrix for layer $l$. In this form, the multilayer perceptron (MLP)
-produces a piecewise linear model of the input data. The computational
-complexity of training a multilayer perceptron is $\mathcal{O}(n d
-m)$, where $m$ is determined by the sizes of the layers of the network
-and the stopping criterion of the minimization used for finding
-weights. This paper uses an MLP built from Keras and Tensorflow to
-perform regression \cite{tensorflow2015-whitepaper,chollet2015keras}
-with ten hidden layers each having thirty two nodes (a total of
-approximately ten thousand parameters), ReLU activation, and five
-thousand epochs of SGD for training.
-
-%% ===================================================================
-\section{Multivariate Interpolation}
-\label{sec:interpolation}
-The following interpolation techniques demonstrate a reasonable
-variety of approaches to interpolation. All of the presented
-interpolants produce approximations that are continuous in value,
-which is often a desirable property in applied approximation problems.
-
-\subsection{Delaunay}
-\label{sec:delaunay}
-
-The Delaunay triangulation is a well-studied geometric technique for
-producing an interpolant \cite{lee1980two}. The Delaunay triangulation
-of a set of data points into simplices is such that there are no data
-points inside the sphere defined by the vertices of each simplex. For
-a $d$-simplex S with vertices $x^{(0)}$, $x^{(1)}$, $\ldots$,
-$x^{(d)}$, and function values $f\bigl(x^{(i)}\bigr)$, $i=0$,
-$\ldots$, $d$, $y \in S$ is a unique convex combination of the
-vertices,
- $$ y = \sum_{i=0}^{d} w_i x^{(i)}, \quad \sum_{i=0}^{d} w_i = 1, \quad w_i \geq 0, \quad i=0,\ldots,d, $$
-and the Delaunay interpolant $\hat f(y)$ at $y$ is given by
- $$ \hat f(y) = \sum_{i=0}^{d} w_i f\bigl(x^{(i)}\bigr). $$ The
-computational complexity of Delaunay interpolation (for the
-implementation used \cite{chang2018polynomial}) is $\mathcal{O}(k n
-d^3)$. Given pathological data the entire triangulation could be
-computed with $k = n^{\lceil d / 2 \rceil}$, but the average case yields
-$k = d \log d$ and is capable of scaling reasonably to $d \leq 50$. In
-the present application, a Delaunay simplex $S$ containing $y$ is
-found, then the $d+1$ vertices (points in $X$) of $S$ are used to
-assign weights to each vertex and produce the predicted function value
-for point $y$.
-
-\subsection{Modified Shepard}
-\label{sec:modified-shepard}
-
-The modified Shepard method used here (ShepMod) generates a continuous
-approximation based on Euclidean distance and resembles a nearest
-neighbor interpolant \cite{cover1967nearest}. This model is a type of
-\textit{metric interpolation}, also called a Shepard method
-\cite{gordon1978shepard,shepard1968two}. The interpolant has the form
- $$ p(x) = \frac{\sum\limits_{k=1}^{n}W_k(x)f\bigl(x^{(k)}\bigr)}
-     {\sum\limits_{k=1}^{n}W_k(x)} ,$$
-where $W_k(x) = \bigl(r_k - \bigl\|x - x^{(k)}\bigr\|\bigr)_+^2 \big/
-\bigl\|x - x^{(k)}\bigr\|_2^2$ and $r_k \in \mathbb{R}$ is the
-smallest radius such that at least $d+1$ other points $x^{(j)}$, $j
-\not = k$, are inside the closed Euclidean ball of radius $r_k$ about
-$x^{(k)}$. The computational complexity of ShepMod is
-$\mathcal{O}(n^2d)$. This paper uses a Fortran 95 implementation of
-ShepMod derived from SHEPPACK \cite{thacker2010algorithm}.
-
-\subsection{Linear Shepard}
-The linear Shepard method (LSHEP) is a blending function using local
-linear interpolants, a special case of the general Shepard algorithm
-\cite{thacker2010algorithm}. The interpolant has the form
- $$ p(x) = \frac{\sum\limits_{k=1}^{n}W_k(x)P_k(x)}
- {\sum\limits_{k=1}^{n}W_k(x)} ,$$
-where $W_k(x)$ is a locally supported weighting function and $P_k(x)$
-is a local linear approximation to the data satisfying
-$P_k\bigl(x^{(k)}\bigr) = f\bigl(x^{(x)}\bigr)$. The computational
-complexity of LSHEP is $\mathcal{O}(n^2d^3)$. This paper uses the
-Fortran 95 implementation of LSHEP in SHEPPACK
-\cite{thacker2010algorithm}.
-
-\subsection{Box Splines}
-\label{sec:box-splines}
-
-The box spline model used here is an interpolation technique built
-from overlapping box splines \cite{de2013box}. The box splines serve
-as basis functions that are shifted and scaled to have support over
-box shaped regions. The boxes are constructed in a way to guarantee a
-covering of the domain \cite{lux2018novel}. Given a set of box splines
-$\bigl\{b^{x^{(i)}}\bigr\}$ with the iterative box properties outlined
-in \cite{lux2018novel} and anchored at interior points
-$\bigl\{x^{(i)}\bigr\}$,
- $$ \hat f(y) = \frac{\sum\limits_{i=1}^n b^{x^{(i)}}(y) f\bigl(x^{(i)}\bigr)}{\sum\limits_{i=1}^n b^{x^{(i)}}(y)}. $$
-Note that the box splines always satisfy
-$b^{x^{(i)}}\bigl(x^{(j)}\bigr) = \delta_{ij}$ and $b^{x^{(i)}}(y)
-\geq 0$. The computational complexity of interpolation via the box
-spline model is $\mathcal{O}(n^2 d)$.
-
-\subsection{Voronoi}
-\label{sec:voronoi}
-
-The nearest neighbor algorithm \cite{cover1967nearest} is a
-well-studied technique for classification and approximation. Nearest
-neighbor inherently utilizes the convex region $C^{x^{(i)}}$ (Voronoi
-cell \cite{dirichlet1850reduction}) consisting of all points closer to
-$x^{(i)}$ than to any other point $x^{(j)}$. The Voronoi model smooths
-the nearest neighbor approximation by utilizing the Voronoi cells to
-define support via a generic basis function $v: \mathbb{R}^d
-\rightarrow \mathbb{R}_+$ given by
- $$ v^{x^{(i)}}(y) = \left(1 - \frac{\bigl\|y - x^{(i)}\bigr\|_2}
- {2 \ h \bigl(y - x^{(i)} \mid x^{(i)}\bigr)} \right)_+, $$
-where $h\bigl(w \mid x^{(i)}\bigr)$ is the distance between $x^{(i)}$
-and the boundary of the Voronoi cell $C^{x^{(i)}}$ in the direction
-$w$. $v^{x^{(i)}}\bigl(x^{(j)}\bigr) = \delta_{ij}$ and $v^{x^{(i)}}$
-has local support, giving the interpolated value
- $$ f(y) = \frac{\sum\limits_{i=1}^n v^{x^{(i)}}(y)
-  f\bigl(x^{(i)}\bigr)}{\sum\limits_{i=1}^n v^{x^{(i)}}(y)}, $$
-where $0 \leq v^{x^{(i)}}(y) \leq 1$. The computational complexity of
-interpolation via this Voronoi model is $\mathcal{O}(n^2 d)$. All of
-the approximations are an interpolant involving a convex combination
-of known function values $f\bigl(x^{(i)}\bigr)$.
-
-%% ===================================================================
-\section{Measuring Error}
-\label{sec:error}
-
-\begin{figure}[htb]
-  \centering
-  \begin{tikzpicture}
-    \node (img)  {\includegraphics[width=0.75\textwidth,]{Fig1.pdf}};
-    \node[below=of img, node distance=1cm, yshift=1cm] {I/O Throughput};
-    \node[left=of img, node distance=0cm, rotate=90, anchor=center,yshift=-0.7cm] {CDF Value};
-  \end{tikzpicture}
-  \vspace{-0.3cm}
-  \caption{In this HPC I/O example, the general methodology for
-    predicting a CDF and evaluating error can be seen. The Delaunay
-    method chose three source distributions (dotted lines) and
-    assigned weights \{.3, .4, .3\} (top to bottom at arrow). The
-    weighted sum of the three known CDFs produces the predicted CDF
-    (dashed line). The KS Statistic (arrow) computed between the true
-    CDF (solid line) and predicted CDF (dashed line) is 0.2 for this
-    example. The KS test null hypothesis is rejected at $p$-value
-    0.01, however it is not rejected at $p$-value 0.001.
-  \vspace{-.1cm}}
-  \label{fig:prediction-example}
-\end{figure}
-
-When the range of an approximation is the real numbers, error is
-reported with summary statistics including: min absolute error, max
-absolute error, and absolute error quartiles. When the range of an
-approximation is the space of cumulative distribution functions, the
-Kolmogorov-Smirnov statistic (max-norm difference between the
-functions) is used.
-
-A hurdle when modeling function-valued outputs such as cumulative
-distribution functions (CDFs) or probability density functions (PDFs)
-is that certain properties must be maintained. It is necessary that a
-PDF $f: \mathbb{R} \rightarrow \mathbb{R}$ have the properties $f(x)
-\geq 0$ and $\int_{-\infty}^{\infty}f(x)dx = 1$. However, for a CDF
-$F: \mathbb{R} \rightarrow \mathbb{R}$ the properties are
-$F(x) \in [0,1]$ and $F(x)$ is absolutely continuous and
-nondecreasing. This work utilizes the fact that a convex combination
-of CDFs (or PDFs) results in a valid CDF (or PDF). Given $G(x) =
-\sum_{i}w_i F_i(x)$, $\sum_{i} w_i = 1$, $w_i \geq 0$, and each $F_i$
-is a valid CDF, $G$ must also be a valid CDF. A demonstration of how
-this is applied can be seen in Figure \ref{fig:prediction-example}.
-
-The performance of approximation techniques that predict probability
-functions can be analyzed through a variety of summary
-statistics. This work uses the max absolute difference, also known as
-the Kolmogorov-Smirnov (KS) statistic \cite{lilliefors1967kolmogorov}
-for its compatibility with the KS test.
-
-The two-sample KS test is a useful nonparametric test for comparing
-two CDFs while only assuming stationarity, finite mean, and finite
-variance. The null hypothesis (that two CDFs come from the same
-underlying distribution) is rejected at level $p \in [0,1]$ when
- $$ KS > \sqrt{-\frac{1}{2}\ln\biggl(\frac{p}{2}\biggr)} \sqrt{\frac{1}{n_1} + \frac{1}{n_2}}, $$
-with distribution sample sizes $n_1,n_2 \in \mathcal{N}$. For all
-applications of the KS test presented in this work $n_1 = n_2$. An
-example of the round-trip prediction methodology from known and
-predicted distributions to the calculation of error can be seen in
-Figure \ref{fig:prediction-example}.
-
-\section{Theoretical Error Bound}
-\label{sec:theory}
-
-This section presents the theoretical results bounding the error of
+This chapter presents the theoretical results bounding the error of
 (piecewise) linear interpolation. The error analysis relies on linear
@@
 to Delaunay interpolation; and (3) multiple other interpolants in this
-paper compute predictions as convex combinations of observed function
+work compute predictions as convex combinations of observed function
 values, which may allow for straight forward extensions of this error
@@
 
-\begin{plainlemma}
+\begin{lemma}
   \label{lemma:1}
@@
   $$\big|f(y) - f(x) - \langle \nabla f(x), y - x \rangle \big| \leq \frac{\gamma \|y - x\|_2^2}{2}.$$
-\end{plainlemma}
+\end{lemma}
 
-\begin{proofdot}
+\begin{proof}
   Consider the function $g(t) = f \big((1-t) x + t y \big)$, $0 \leq t
@@
   \end{align*}
-  \qed
-\end{proofdot}
+\end{proof}
 
-\begin{plainlemma}
+\begin{lemma}
   \label{lemma:2}
@@
   where $\sigma_d$ is the smallest singular value of $M.$
-\end{plainlemma}
+\end{lemma}
 
-\begin{proofdot}
+\begin{proof}
   Using the facts that $M$ and $M^t$ have the same singular values,
@@
   \end{align*}
-  \qed
-\end{proofdot}
+\end{proof}
 
-\begin{plainlemma}
+\begin{lemma}
   \label{lemma:3}
@@
   $$\big\|\nabla f(x_0) - c\big\|_2 \leq \sqrt{d} \frac{\gamma k^2}{\sigma_d}.$$
-\end{plainlemma}
+\end{lemma}
 
-\begin{proofdot}
+\begin{proof}
   Consider $f(x) - \hat f(x)$ along the line segment $z(t) = (1-t)x_0
@@
   \leq \sqrt{d} \frac{\gamma k^2}{\sigma_d}.$$
-  \qed
-\end{proofdot}
+\end{proof}
 
-\begin{onetheorem}
+\begin{theorem}
   Under the assumptions of {\it Lemma \ref{lemma:1}} and {\it Lemma
@@
   $$ \big|f(z) - \hat f(z)\big| \leq \frac{\gamma \|z - x_0\|_2^2}{2} + \sqrt{d} \frac{\gamma k^2}{\sigma_d} \|z - x_0\|_2.$$
-\end{onetheorem}
+\end{theorem}
 
-\begin{proofdot}
+\begin{proof}
   Let $v = \nabla f(x_0) - c$, where $\|v\|_2 \leq \sqrt{d} \gamma k^2 / \sigma_d$ by {\it Lemma \ref{lemma:3}}. Now
@@
   where the last inequality follows from {\it Lemma \ref{lemma:1}}.
-  %
-  \qed
-\end{proofdot}
+\end{proof}
 
@@
 \section{Data and Empirical Analysis}
-\label{sec:data}
+\label{sec:error_data}
 
-The theoretical results constructed in Section \ref{sec:theory} for
+The theoretical results constructed at the start of this chapter for
 (piecewise) linear interpolation are promising and apply directly to
@@
 context with approximation algorithms that do not have similar known
-uniform error bounds. For that reason, this paper utilizes five
+uniform error bounds. For that reason, this chapter utilizes five
 different data sets of varying dimension and application to construct
@@
   \begin{tikzpicture}
-    \node (img)  {\includegraphics[width=0.75\textwidth,]{Fig2.pdf}};
+    \node (img)  {\includegraphics[width=0.75\textwidth,]{Figures/NA/Fig2.pdf}};
     \node[below=of img, node distance=1cm, yshift=1cm] {Forest Fire Area Burned};
@@
   \begin{tikzpicture}
-    \node (img)  {\includegraphics[width=0.8\textwidth,]{Fig3.pdf}};
+    \node (img)  {\includegraphics[width=0.8\textwidth,]{Figures/NA/Fig3.pdf}};
     \node[below=of img, node distance=1cm, yshift=1cm] {Area Burned Error};
@@
     that would be burned given environment conditions. $10$-fold cross
-    validation as described in the beginning of Section \ref{sec:data}
+    validation as described in the beginning of Section \ref{sec:error_data}
     is used to evaluated each algorithm. This results in exactly one
@@
   \begin{tikzpicture}
-    \node (img)  {\includegraphics[width=0.75\textwidth,]{Fig4.pdf}};
+    \node (img)  {\includegraphics[width=0.75\textwidth,]{Figures/NA/Fig4.pdf}};
     \node[below=of img, node distance=1cm, yshift=1cm] {Parkinson's Total UPDRS Score};
@@
   \begin{tikzpicture}
-    \node (img)  {\includegraphics[width=0.8\textwidth,]{Fig5.pdf}};
+    \node (img)  {\includegraphics[width=0.8\textwidth,]{Figures/NA/Fig5.pdf}};
     \node[below=of img, node distance=1cm, yshift=1cm] {Total UPDRS Score Error};
@@
   \begin{tikzpicture}
-    \node (img)  {\includegraphics[width=0.75\textwidth,]{Fig6.pdf}};
+    \node (img)  {\includegraphics[width=0.75\textwidth,]{Figures/NA/Fig6.pdf}};
     \node[below=of img, node distance=1cm, yshift=1cm] {Daily Rainfall in Sydney};
@@
   \begin{tikzpicture}
-    \node (img)  {\includegraphics[width=0.8\textwidth,]{Fig7.pdf}};
+    \node (img)  {\includegraphics[width=0.8\textwidth,]{Figures/NA/Fig7.pdf}};
     \node[below=of img, node distance=1cm, yshift=1cm] {Sydney Rainfall Tomorrow Error};
@@
   \begin{tikzpicture}
-    \node (img)  {\includegraphics[width=0.75\textwidth,]{Fig8.pdf}};
+    \node (img)  {\includegraphics[width=0.75\textwidth,]{Figures/NA/Fig8.pdf}};
     \node[below=of img, node distance=1cm, yshift=1cm] {Credit Card Transaction Amount};
@@
   \begin{tikzpicture}
-    \node (img)  {\includegraphics[width=0.8\textwidth,]{Fig9.pdf}};
+    \node (img)  {\includegraphics[width=0.8\textwidth,]{Figures/NA/Fig9.pdf}};
     \node[below=of img, node distance=1cm, yshift=1cm] {Transaction Amount Error};
@@
   \begin{tikzpicture}
-    \node (img)  {\includegraphics[width=0.75\textwidth,]{Fig10.pdf}};
+    \node (img)  {\includegraphics[width=0.75\textwidth,]{Figures/NA/Fig10.pdf}};
     \node[below=of img, node distance=1cm, yshift=1cm] {I/O Read Throughput};
@@
   \begin{tikzpicture}
-    \node (img)  {\includegraphics[width=0.8\textwidth,]{Fig11.pdf}};
+    \node (img)  {\includegraphics[width=0.8\textwidth,]{Figures/NA/Fig11.pdf}};
     \node[below=of img, node distance=1cm, yshift=1cm] {Read Throughput KS Statistic};
@@
   \begin{tikzpicture}
-    \node (img)  {\includegraphics[width=0.8\textwidth]{Fig12.pdf}};
+    \node (img)  {\includegraphics[width=0.8\textwidth]{Figures/NA/Fig12.pdf}};
     \node[below=of img, node distance=1cm, yshift=1cm] {KS Statistic for Predicted vs. Actual};
@@
 
-\section{Discussion}
-\label{sec:discussion}
+\section{Discussion of Empirical Results}
 
@@
 test data sets with real-valued ranges. The interpolants discussed in
-this paper produce the \textit{best} approximations roughly one third
+this chapter produce the \textit{best} approximations roughly one third
 of the time, and produce competitive approximations for almost all
@@
 
-The new theoretical results presented in Section \ref{sec:theory}
-directly apply to Delaunay interpolation, however the performance of
-Delaunay does not appear significantly better than other algorithms on
-these data sets.  This observation may be due to the stochastic nature
-of the data, but it also speaks to the power of the approximations
-generated by the different interpolation methods.  The strong
-performance of other interpolants suggests that theoretical results
-similar to those presented in this work can be achieved for the other
-interpolants under reasonable assumptions.
+The theoretical results presented in this chapter directly apply to
+Delaunay interpolation, however the performance of Delaunay does not
+appear significantly better than other algorithms on these data sets.
+This observation may be due to the stochastic nature of the data, but
+it also speaks to the power of the approximations generated by the
+different interpolation methods.  The strong performance of other
+interpolants suggests that theoretical results similar to those
+presented in this work can be achieved for the other interpolants
+under reasonable assumptions.
 
@@
 
-\section{Conclusion}
-\label{sec:conclusion}
+\section{Takeaway from Empirical Tests}
 
 The major contributions of this work are: 1) new uniform theoretical
-error bounds for piecewise linear interpolation in arbitrary dimension
-(Section \ref{sec:theory}); 2) an empirical evaluation across
-real-world problems that demonstrates interpolants produce
-competitively accurate models of multivariate phenomenon when compared
-with common regressors for sparse, moderately high dimensional
-problems (Section \ref{sec:data}); and 3) a demonstration that some
-interpolants generalize to interpolation in function spaces (Section
-\ref{sec:error}), preserving monotonicity (with CDFs, e.g.), neither
-of which common regressors can do.
+error bounds for piecewise linear interpolation in arbitrary
+dimension; 2) an empirical evaluation across real-world problems that
+demonstrates interpolants produce competitively accurate models of
+multivariate phenomenon when compared with common regressors for
+sparse, moderately high dimensional problems (Section
+\ref{sec:error_data}); and 3) a demonstration that some interpolants
+generalize to interpolation in function spaces, preserving
+monotonicity (with CDFs, e.g.), neither of which common regressors can
+do.
 
-The various interpolants discussed in this paper have been
+The various interpolants discussed in this work have been
 demonstrated to effectively approximate multivariate phenomena up to
@@
 approximation in science
-
-%\begin{acknowledgements}
-%If you'd like to thank anyone, place your comments here
-%and remove the percent signs.
-%\end{acknowledgements}
-
-%%====================================================================
-%%====================================================================
-
-\bibliographystyle{spmpsci}      % mathematics and physical sciences
-\bibliography{paper}
-
-\newpage
-
-\begin{appendix}
-\section*{Appendix}
-\label{sec:appendix}
-
-The tables that follow show the precise experimental results for all
-data sets presented in Section \ref{sec:data}. The tests were all run
-serially on an otherwise idle machine with a CentOS 6.10 operating
-system and an Intel i7-3770 CPU operating at 3.4 GHz. The detailed
-performance results in the tables that follow are very much dependent
-on the problem and the algorithm implementation (e.g., some codes are
-TOMS software, some industry distributions, and others are from
-conference paper venues). Different typeface is used to show best
-performers, however not much significance should be attached to
-ranking algorithms based on small time (millisecond) differences. The
-results serve as a demonstration of conceptual validity.
-
-\vspace*{\fill}
-
-\begin{table}[H]
-  \centering
-  \begin{tabular}{c|r@{.}l|r@{.}l|r@{.}l|r@{.}l|r@{.}l}
-    \hline
-    Algorithm & \multicolumn{2}{c|}{Min} & \multicolumn{2}{c|}{$25^{th}$} & \multicolumn{2}{c|}{$50^{th}$} & \multicolumn{2}{c|}{$75^{th}$} & \multicolumn{2}{c}{Max}\\
-    \hline
-    MARS & $0$&$00984$ & $3$&$11$ & $7$&$01$ & $\mathit{11}$&$\mathit{7}$ & $1090$&$0$\\
-    SVR & $\mathit{0}$&$\mathit{00402}$ & $\mathbf{0}$&$\mathbf{243}$ & $\mathbf{0}$&$\mathbf{416}$ & $\mathbf{6}$&$\mathbf{19}$ & $1090$&$0$\\
-    MLP & $0$&$0426$ & $2$&$63$ & $\mathit{5}$&$\mathit{27}$ & $14$&$0$ & $1090$&$0$\\
-    Delaunay & $\mathbf{0}$&$\mathbf{00}$ & $1$&$98$ & $5$&$37$ & $13$&$1$ & $\mathit{1080}$&$\mathit{0}$\\
-    ShepMod & $\mathbf{0}$&$\mathbf{00}$ & $\mathit{1}$&$\mathit{93}$ & $6$&$27$ & $16$&$0$ & $1090$&$0$\\
-    LSHEP & $0$&$0400$ & $2$&$17$ & $8$&$87$ & $19$&$1$ & $\mathbf{1070}$&$\mathbf{0}$\\
-    Voronoi & $0$&$00982$ & $3$&$65$ & $7$&$56$ & $15$&$6$ & $1090$&$0$\\
-    BoxSpline & $\mathbf{0}$&$\mathbf{00}$ & $2$&$27$ & $5$&$91$ & $16$&$4$ & $1090$&$0$\\
-    \hline
-  \end{tabular}
-  \caption{This numerical data accompanies the visual provided in
-    Figure \ref{fig:error-forest-fire}. The columns of absolute error
-    percentiles correspond to the minimum, first quartile, median,
-    third quartile, and maximum absolute errors respectively. The
-    minimum of each column is boldface, while the second lowest value
-    is italicized. All values are rounded to three significant
-    digits.}
-  \label{table:error-forest-fire}
-\end{table}
-
-\begin{table}[H]
-  \centering
-  \begin{tabular}{|c|r@{.}l| c |r@{.}l|r@{.}l|r@{.}l|}
-    \cline{1-3}\cline{5-10}
-    Algorithm & \multicolumn{2}{c|}{\% Best} &  & \multicolumn{2}{c|}{Fit/Prep. Time (s)} & \multicolumn{2}{c|}{App. Time (s)} & \multicolumn{2}{c|}{Total App. Time (s)}\\
-    \cline{1-3}\cline{5-10}
-    MARS & \quad$\mathit{7}$&$\mathit{7}$ &  & \quad\quad$29$&$1$ & \quad$0$&$00137$ & \quad\quad\quad$0$&$0686$\\
-    SVR & $\mathbf{80}$&$\mathbf{2}$ &  & $\mathbf{0}$&$\mathbf{00584}$ & $\mathit{0}$&$\mathit{0000620}$ & $\mathit{0}$&$\mathit{00310}$\\
-    MLP & $0$&$0$ &  & $32$&$8$ & $0$&$000871$ & $0$&$0436$\\
-    Delaunay & $3$&$5$ &  & $0$&$0151$ & $0$&$0234$ & $1$&$18$\\
-    ShepMod & $3$&$7$ &  & $\mathit{0}$&$\mathit{00634}$ & $0$&$0000644$ & $0$&$00322$\\
-    LSHEP & $5$&$1$ &  & $0$&$0275$ & $\mathbf{0}$&$\mathbf{0000463}$ & $\mathbf{0}$&$\mathbf{00231}$\\
-    Voronoi & $0$&$0$ &  & $0$&$0152$ & $0$&$000396$ & $0$&$0198$\\
-    BoxSpline & $0$&$4$ &  & $0$&$00724$ & $0$&$0000978$ & $0$&$00489$\\
-    \cline{1-3}\cline{5-10}
-  \end{tabular}
-  \caption{The left above shows how often each algorithm had the
-    lowest absolute error approximating forest fire data in Table
-    \ref{table:error-forest-fire}. On the right columns are median fit
-    time of 454 points, median time for one approximation, and median
-    time approximating 50 points.}
-  \label{table:best-forest-fire}
-\end{table}
-
-\vspace*{\fill}
-\newpage
-
-\begin{table}
-  \centering
-  \begin{tabular}{c|r@{.}l|r@{.}l|r@{.}l|r@{.}l|r@{.}l}
-    \hline
-    Algorithm & \multicolumn{2}{c|}{Min} & \multicolumn{2}{c|}{$25^{th}$} & \multicolumn{2}{c|}{$50^{th}$} & \multicolumn{2}{c|}{$75^{th}$} & \multicolumn{2}{c}{Max}\\
-    \hline
-    MARS & $0$&$00948$ & $9$&$98$ & $20$&$4$ & $32$&$9$ & $863$&$0$\\
-    SVR & $0$&$00138$ & $3$&$17$ & $7$&$31$ & $12$&$6$ & $\mathbf{27}$&$\mathbf{5}$\\
-    MLP & $0$&$0000239$ & $\mathit{0}$&$\mathit{533}$ & $\mathit{1}$&$\mathit{25}$ & $\mathbf{2}$&$\mathbf{84}$ & $39$&$3$\\
-    Delaunay & $\mathit{3}$&$\mathit{72 \times 10^{-12}}$ & $1$&$20$ & $3$&$50$ & $7$&$67$ & $30$&$7$\\
-    ShepMod & $\mathbf{0}$&$\mathbf{0}$ & $\mathbf{0}$&$\mathbf{255}$ & $\mathbf{0}$&$\mathbf{908}$ & $\mathit{3}$&$\mathit{43}$ & $34$&$5$\\
-    LSHEP & $0$&$00254$ & $2$&$93$ & $7$&$16$ & $13$&$1$ & $\mathit{29}$&$\mathit{0}$\\
-    Voronoi & $\mathbf{0}$&$\mathbf{0}$ & $1$&$29$ & $3$&$52$ & $6$&$87$ & $30$&$1$\\
-    BoxSpline & $0$&$00287$ & $4$&$39$ & $9$&$10$ & $14$&$8$ & $41$&$3$\\
-    \hline
-  \end{tabular}
-  \caption{This numerical data accompanies the visual provided in
-    Figure \ref{fig:error-parkinsons}. The columns of absolute error
-    percentiles correspond to the minimum, first quartile, median,
-    third quartile, and maximum absolute errors respectively. The
-    minimum of each column is boldface, while the second lowest value
-    is italicized. All values are rounded to three significant
-    digits.}
-  \label{table:error-parkinsons}
-\end{table}
-
-\begin{table}
-  \centering
-  \begin{tabular}{|c|r@{.}l| c |r@{.}l|r@{.}l|r@{.}l|}
-    \cline{1-3}\cline{5-10}
-    Algorithm & \multicolumn{2}{c|}{\% Best} &  & \multicolumn{2}{c|}{Fit/Prep. Time (s)} & \multicolumn{2}{c|}{App. Time (s)} & \multicolumn{2}{c|}{Total App. Time (s)}\\
-    \cline{1-3}\cline{5-10}
-    MARS & \quad\,$0$&$0$ &  & \quad\quad\quad$37$&$9$ & \quad\,\,$0$&$00253$ & \quad\quad\quad\,\,$1$&$48$\\
-    SVR & $0$&$1$ &  & $\mathbf{0}$&$\mathbf{859}$ & $\mathit{0}$&$\mathit{000181}$ & $\mathit{0}$&$\mathit{106}$\\
-    MLP & $\mathit{32}$&$\mathit{0}$ &  & $348$&$0$ & $0$&$00111$ & $0$&$653$\\
-    Delaunay & $0$&$0$ &  & $2$&$47$ & $1$&$22$ & $720$&$0$\\
-    ShepMod & $\mathbf{66}$&$\mathbf{4}$ &  & $\mathit{1}$&$\mathit{13}$ & $0$&$000182$ & $0$&$107$\\
-    LSHEP & $0$&$0$ &  & $2$&$39$ & $\mathbf{0}$&$\mathbf{000144}$ & $\mathbf{0}$&$\mathbf{0845}$\\
-    Voronoi & $1$&$6$ &  & $2$&$77$ & $0$&$0274$ & $16$&$1$\\
-    BoxSpline & $0$&$0$ &  & $1$&$26$ & $0$&$000643$ & $0$&$377$\\
-    \cline{1-3}\cline{5-10}
-  \end{tabular}
-  \caption{The left above shows how often each algorithm had the
-    lowest absolute error approximating Parkinson's data in Table
-    \ref{table:error-parkinsons}. On the right columns are median fit
-    time of 5288 points, median time for one approximation, and median
-    time approximating 587 points.}
-  \label{table:best-parkinsons}
-\end{table}
-
-\begin{table}
-  \centering
-  \begin{tabular}{c|r@{.}l|r@{.}l|r@{.}l|r@{.}l|r@{.}l}
-    \hline
-    Algorithm & \multicolumn{2}{c|}{Min} & \multicolumn{2}{c|}{$25^{th}$} & \multicolumn{2}{c|}{$50^{th}$} & \multicolumn{2}{c|}{$75^{th}$} & \multicolumn{2}{c}{Max}\\
-    \hline
-    MARS & $\mathit{6}$&$\mathit{45 \times 10^{-15}}$ & $\mathbf{2}$&$\mathbf{70 \times 10^{-14}}$ & $1$&$66$ & $1$&$96$ & $\mathit{53}$&$\mathit{3}$\\
-    SVR & $0$&$0420$ & $0$&$143$ & $0$&$148$ & $\mathit{0}$&$\mathit{860}$ & $119$&$0$\\
-    MLP & $0$&$0000689$ & $0$&$0337$ & $\mathbf{0}$&$\mathbf{0795}$ & $\mathbf{0}$&$\mathbf{264}$ & $\mathbf{5}$&$\mathbf{31}$\\
-    Delaunay & $\mathbf{0}$&$\mathbf{0}$ & $0$&$187$ & $0$&$919$ & $2$&$72$ & $56$&$3$\\
-    ShepMod & $\mathbf{0}$&$\mathbf{0}$ & $0$&$0957$ & $0$&$685$ & $2$&$90$ & $73$&$2$\\
-    LSHEP & $\mathbf{0}$&$\mathbf{0}$ & $\mathit{0}$&$\mathit{0153}$ & $\mathit{0}$&$\mathit{106}$ & $1$&$17$ & $113$&$0$\\
-    Voronoi & $\mathbf{0}$&$\mathbf{0}$ & $0$&$430$ & $1$&$28$ & $2$&$94$ & $83$&$8$\\
-    BoxSpline & $\mathbf{0}$&$\mathbf{0}$ & $0$&$789$ & $2$&$06$ & $5$&$25$ & $119$&$0$\\
-    \hline
-  \end{tabular}
-  \caption{This numerical data accompanies the visual provided in
-    Figure \ref{fig:error-weather}. The columns of absolute error
-    percentiles correspond to the minimum, first quartile, median,
-    third quartile, and maximum absolute errors respectively. The
-    minimum value of each column is boldface, while the second lowest
-    is italicized. All values are rounded to three significant
-    digits.}
-  \label{table:error-weather}
-\end{table}
-
-\begin{table}
-  \centering
-  \begin{tabular}{|c|r@{.}l| c |r@{.}l|r@{.}l|r@{.}l|}
-    \cline{1-3}\cline{5-10}
-    Algorithm & \multicolumn{2}{c|}{\% Best} &  & \multicolumn{2}{c|}{Fit/Prep. Time (s)} & \multicolumn{2}{c|}{App. Time (s)} & \multicolumn{2}{c|}{Total App. Time (s)}\\
-    \cline{1-3}\cline{5-10}
-    MARS & \,\,\,\,$10$&$7$ &  & \quad\quad\quad$\mathit{0}$&$\mathit{151}$ & \quad$0$&$000117$ & \quad\quad\quad\,\,$0$&$0304$\\
-    SVR & $4$&$1$ &  & $\mathbf{0}$&$\mathbf{133}$ & $\mathbf{0}$&$\mathbf{0000947}$ & $\mathbf{0}$&$\mathbf{0246}$\\
-    MLP & $\mathbf{56}$&$\mathbf{8}$ &  & $169$&$0$ & $0$&$00137$ & $0$&$356$\\
-    Delaunay & $0$&$1$ &  & $0$&$664$ & $0$&$886$ & $230$&$0$\\
-    ShepMod & $3$&$5$ &  & $0$&$265$ & $0$&$000128$ & $0$&$0333$\\
-    LSHEP & $\mathit{28}$&$\mathit{4}$ &  & $0$&$874$ & $\mathit{0}$&$\mathit{0000975}$ & $\mathit{0}$&$\mathit{0254}$\\
-    Voronoi & $0$&$2$ &  & $0$&$675$ & $0$&$0270$ & $7$&$01$\\
-    BoxSpline & $0$&$3$ &  & $0$&$330$ & $0$&$000406$ & $0$&$106$\\
-    \cline{1-3}\cline{5-10}
-  \end{tabular}
-  \caption{Left table shows how often each algorithm had the lowest
-    absolute error approximating Sydney rainfall data in Table
-    \ref{table:error-weather}. On the right columns are median fit
-    time of 2349 points, median time for one approximation, and median
-    time approximating 260 points.}
-  \label{table:best-weather}
-\end{table}
-
-\begin{table}
-  \centering
-  \begin{tabular}{c|r@{.}l|r@{.}l|r@{.}l|r@{.}l|r@{.}l}
-    \hline
-    Algorithm & \multicolumn{2}{c|}{Min} & \multicolumn{2}{c|}{$25^{th}$} & \multicolumn{2}{c|}{$50^{th}$} & \multicolumn{2}{c|}{$75^{th}$} & \multicolumn{2}{c}{Max}\\
-    \hline
-    MARS & $5$&$36$ & $3610$&$0$ & $4580$&$0$ & $5450$&$0$ & $13400$&$0$\\
-    SVR & $0$&$0138$ & $8$&$47$ & $15$&$4$ & $41$&$9$ & $7700$&$0$\\
-    MLP & $0$&$00151$ & $\mathbf{1}$&$\mathbf{60}$ & $\mathbf{3}$&$\mathbf{86}$ & $\mathbf{8}$&$\mathbf{34}$ & $\mathbf{604}$&$\mathbf{0}$\\
-    Delaunay & $\mathbf{0}$&$\mathbf{0}$ & $\mathit{2}$&$\mathit{69}$ & $\mathit{13}$&$\mathit{8}$ & $\mathit{35}$&$\mathit{0}$ & $\mathit{4840}$&$\mathit{0}$\\
-    ShepMod & $\mathbf{0}$&$\mathbf{0}$ & $4$&$21$ & $17$&$3$ & $51$&$6$ & $6510$&$0$\\
-    LSHEP & $1$&$27$ & $199$&$0$ & $260$&$0$ & $343$&$0$ & $6530$&$0$\\
-    Voronoi & $\mathit{2}$&$\mathit{89 \times 10^{-10}}$ & $14$&$7$ & $32$&$8$ & $52$&$9$ & $4860$&$0$\\
-    BoxSpline & $0$&$00740$ & $20$&$8$ & $40$&$9$ & $79$&$6$ & $7640$&$0$\\
-    \hline
-  \end{tabular}
-  \caption{This numerical data accompanies the visual provided in
-    Figure \ref{fig:error-credit-card}. The columns of absolute error
-    percentiles correspond to the minimum, first quartile, median,
-    third quartile, and maximum absolute errors respectively. The
-    minimum value of each column is boldface, while the second lowest
-    is italicized. All values are rounded to three significant
-    digits.}
-  \label{table:error-credit-card}
-\end{table}
-
-\begin{table}
-  \centering
-  \begin{tabular}{|c|r@{.}l| c |r@{.}l|r@{.}l|r@{.}l|}
-    \cline{1-3}\cline{5-10}
-    Algorithm & \multicolumn{2}{c|}{\% Best} &  & \multicolumn{2}{c|}{Fit/Prep. Time (s)} & \multicolumn{2}{c|}{App. Time (s)} & \multicolumn{2}{c|}{Total App. Time (s)}\\
-    \cline{1-3}\cline{5-10}
-    MARS & \quad\,$0$&$0$ &  & \quad\quad\quad$22$&$0$ & \quad$0$&$00148$ & \quad\quad\quad\,\,$0$&$820$\\
-    SVR & $0$&$0$ &  & $\mathbf{1}$&$\mathbf{01}$ & $\mathit{0}$&$\mathit{000210}$ & $\mathit{0}$&$\mathit{117}$\\
-    MLP & $\mathbf{79}$&$\mathbf{5}$ &  & $290$&$0$ & $0$&$000714$ & $0$&$397$\\
-    Delaunay & $\mathit{20}$&$\mathit{5}$ &  & $3$&$12$ & $3$&$71$ & $2070$&$0$\\
-    ShepMod & $0$&$1$ &  & $\mathit{1}$&$\mathit{4}5$ & $0$&$000220$ & $0$&$122$\\
-    LSHEP & $0$&$0$ &  & $3$&$47$ & $\mathbf{0}$&$\mathbf{000176}$ & $\mathbf{0}$&$\mathbf{0981}$\\
-    Voronoi & $0$&$0$ &  & $3$&$32$ & $0$&$0950$ & $52$&$8$\\
-    BoxSpline & $0$&$0$ &  & $1$&$66$ & $0$&$000956$ & $0$&$532$\\
-    \cline{1-3}\cline{5-10}
-  \end{tabular}
-  \caption{The left above shows how often each algorithm had the
-    lowest absolute error approximating credit card transaction data
-    in Table \ref{table:error-credit-card}. On the right columns are
-    median fit time of 5006 points, median time for one approximation,
-    and median time approximating 556 points.}
-  \label{table:best-credit-card}
-\end{table}
-
-\begin{table}
-  \centering
-  \begin{tabular}{c|r@{.}l|r@{.}l|r@{.}l|r@{.}l|r@{.}l}
-    \hline
-    Algorithm & \multicolumn{2}{c|}{Min} & \multicolumn{2}{c|}{$25^{th}$} & \multicolumn{2}{c|}{$50^{th}$} & \multicolumn{2}{c|}{$75^{th}$} & \multicolumn{2}{c}{Max}\\
-    \hline
-    Delaunay & $\mathbf{0}$&$\mathbf{0287}$ & $\mathbf{0}$&$\mathbf{0914}$ & $\mathbf{0}$&$\mathbf{158}$ & $\mathit{0}$&$\mathit{308}$ & $\mathit{0}$&$\mathit{897}$\\
-    ShepMod & $0$&$0307$ & $\mathit{0}$&$\mathit{0983}$ & $\mathit{0}$&$\mathit{164}$ & $0$&$335$ & $1$&$00$\\
-    Voronoi & $\mathit{0}$&$\mathit{0303}$ & $0$&$100$ & $0$&$165$ & $\mathbf{0}$&$\mathbf{274}$ & $\mathbf{0}$&$\mathbf{762}$\\
-    BoxSpline & $0$&$0703$ & $0$&$417$ & $0$&$561$ & $0$&$745$ & $1$&$00$\\
-    \hline
-  \end{tabular}
-  \caption{This numerical data accompanies the visual provided in
-    Figure \ref{fig:error-throughput}. The columns of absolute error
-    percentiles correspond to the minimum, first quartile, median,
-    third quartile, and maximum KS statistics respectively between
-    truth and guess for models predicting the distribution of I/O
-    throughput that will be observed at previously unseen system
-    configurations. The minimum value of each column is boldface,
-    while the second lowest is italicized. All values are rounded to
-    three significant digits.}
-  \label{table:error-throughput}
-\end{table}
-
-\begin{table}
-  \centering
-  \begin{tabular}{|c|r@{.}l| c |r@{.}l|r@{.}l|r@{.}l|}
-    \cline{1-3}\cline{5-10}
-    Algorithm & \multicolumn{2}{c|}{\% Best} &  & \multicolumn{2}{c|}{Fit/Prep. Time (s)} & \multicolumn{2}{c|}{App. Time (s)} & \multicolumn{2}{c|}{Total App. Time (s)}\\
-    \cline{1-3}\cline{5-10}
-    Delaunay & \,\,$\mathbf{62}$&$\mathbf{0}$ &  & \quad\quad\quad$0$&$344$ & \quad$0$&$00984$ & \quad\quad\,\,$2$&$71$\\
-    ShepMod & $0$&$0$ &  & $\mathbf{0}$&$\mathbf{0884}$ & $\mathbf{0}$&$\mathbf{000145}$ & $\mathbf{0}$&$\mathbf{0436}$\\
-    Voronoi & $\mathit{38}$&$\mathit{0}$ &  & $0$&$360$ & $0$&$00253$ & $0$&$762$\\
-    BoxSpline & $0$&$0$ &  & $\mathit{0}$&$\mathit{0972}$ & $\mathit{0}$&$\mathit{000210}$ & $\mathit{0}$&$\mathit{0633}$\\
-    \cline{1-3}\cline{5-10}
-  \end{tabular}
-  \caption{The left above shows how often each algorithm had the
-    lowest KS statistic on the I/O throughput distribution data in
-    Table \ref{table:error-throughput}. On the right columns are
-    median fit time of 2715 points, median time for one approximation,
-    and median time approximating 301 points.}
-  \label{table:best-throughput}
-\end{table}
-
-\end{appendix}
-
-\end{document}

### 2020-04-14 3-Disseration/splines.tex
Updated min wiggle fit tests, wrote report on the topic that was sent to Dr. Watson.

Many domains of science rely on smooth approximations to real-valued functions over a closed interval. These smooth approximations are regularly used in automotive and aerospace engineering, architecture, mathematics and especially statistics \cite{knott2012interpolating}. While polynomial interpolants or regressors apply broadly, piecewise polynomial functions (splines) are often a good choice because they can approximate globally complex functions while minimizing local complexity of the approximation. It is often the case that the true underlying function or phenomenon being modeled has known properties e.g., convexity, positivity, various levels of continuity, or monotonicity. Given a reasonably large amount of data, it can be impossible to maintain these properties with a single polynomial function.

In general, the maintenance of function properties through interpolation / regression is usually referred to as \textit{shape preserving} \cite{fritsch1980monotone,gregory1985shape}. The specific shapes this work will focus on are monotonicity and multiple levels of continuity for a function. These properties are chiefly important to the approximation of cumulative distribution functions and subsequently the effective generation of random numbers from a specified distribution.

In statistics especially, the construction of a monotone interpolating spline that is continuous in second derivative is meaningfully useful \cite{ramsay1988monotone}. A function with these properties could approximate random variables to a high level of accuracy with relatively few intervals (with even more accuracy given greater continuity). A continuously twice differentiable approximation to a cumulative distribution function would also produce a corresponding probability density function that is continuously differentiable, which is a property many commonly occurring parametric distributions maintain.

\section{Related Work}

The current state-of-the-art monotone interpolating spline with a mathematical software implementation is piecewise cubic, continuously differentiable, and was first proposed in \citet{fritsch1980monotone} then expanded upon in \citet{carlson1985monotone}. Let $\pi: a = x_1 < x_2 < \cdots < x_n = b$ be a partition of the interval $[a,b]$. Let $\{f(x_i) : i = 1,2,\ldots,n\}$ be a given set of data values at the partition points for a monotone function $f$, meaning $f(x_i) \leq f(x_{i+1})$ for $i = 1, \ldots, n-1$ or $f(x_i) \geq f(x_{i+1})$ for $i = 1, \ldots, n-1$. Let $\hat f$ be a piecewise cubic spline defined in each sub-interval $I_i = [x_i, x_{i+1}]$ by

\begin{align*}
h_i =& \ x_{i+1} - x_{i} \\
u(t) =& \ 3t^2 - 2t^3 \\
p(t) =& \ t^3 - t^2 \\
\hat f(x) =& \ f(x_i)\ u\big((x_{i+1} - x) / h_i\big) + f(x_{i+1})\ u\big((x - x_i) / h_i\big) \\
& - \hat f^{\ \prime}(x_i)\ p\big((x_{i+1}-x)/h_i\big) + \hat f^{\ \prime}(x_{i+1})\ p\big((x-x_i)/h_i \big).
\end{align*}

Notice that it is up to the user to choose values for $\hat f^{\ \prime}$and a viable monotonic cubic spline can be produced by choosing $\hat f^{\ \prime}(x_i) = 0$, $i = 1, \ldots, n$. However, such a spline has too many \textit{wiggles} for most applications. Fritsch and Carlson proceed to show that simple conditions on the derivative values can guarantee monotonicity, and that these conditions can be enforced in a way that ensures modifications on one interval will not break the monotonicity of cubic polynomials over any neighboring intervals. Consider the terms $\alpha = \frac{\hat f^{\ \prime}(x_i) (x_{i+1}-x_i)}{f(x_{i+1}) - f(x_i)}$ and $\beta = \frac{\hat f^{\ \prime}(x_{i+1}) (x_{i+1}-x_i)}{f(x_{i+1}) - f(x_i)}$, now monotonicity of a cubic polynomial over a sub-interval can be maintained by ensuring that $\alpha$ and $\beta$ reside in any of the following regions.

\begin{figure}[htb]
  \centering
  \includegraphics[scale=0.5]{Figures/splines/feasible_region.png}
  \caption{These are the feasible regions of monotonicity for cubic splines.}
  \label{fig:feasible_region}
\end{figure}

The actual region of monotonicity for a cubic polynomial is larger, but projection of $(\alpha, \beta)$ into one of these regions ensures that monotonicity will be achieved and not violated for neighboring regions. The user must decide which region is most suitable to project $(\alpha, \beta)$ onto, Fritsch and Carlson recommend using region 2.

This theoretical work has been extended in \cite{ulrich1994positivity,hess1994positive} to monotone quintic polynomials, but no mathematical software has been produced accordingly. That is where this proposed work comes in.

\section{Achieved Progress}

To expand my own understanding of the current state-of-the-art method for constructing a monotone spline interpolant, I \href{https://github.com/tchlux/VarSys/blob/master/Disseration/cubic.py}{fully implemented} the algorithm described in \cite{fritsch1980monotone} in Python. This code can be used to generate monotone cubic spline fits to data and allows for the user specification of a method((the means by which monotonicity is enforced)) and derivatives at any of the knots or endpoints. Clearly this code is not optimized for numerical robustness, nor speed, but it serves as a portable demonstration and educational tool.

Here we can see the recommended method of projection for moving cubic polynomials into the region of monotonicity.

\begin{figure}[htb]
  \centering
  \includegraphics[scale=0.5]{Figures/splines/demo_projection.png}
  \caption{A demonstration of projection onto the feasible region for cubic splines.}
  \label{fig:demo_projection}
\end{figure}

The resulting interpolating spline is $C^1$ and has a \textit{smooth} appearance, here is a demonstration when using Region 2.

\begin{figure}[htb]
  \centering
  \includegraphics[scale=0.5]{Figures/splines/demo_fit.png}
  \caption{A demonstration fit with a cubic spline.}
  \label{fig:demo_fit}
\end{figure}

And the derivative of the above spline is continuous.

\begin{figure}[htb]
  \centering
  \includegraphics[scale=0.5]{./Figures/splines/demo_fit_deriv.png}
  \caption{The derivative of Figure \ref{fig:demo_projection}.}
  \label{fig:demo_fit_deriv}
\end{figure}

\section{Research Goal}

{\centering \textbf{Produce numerically robust software that computes monotone quintic interpolating splines given data.}}

\section{Timeline}

The projected research timeline is shown in Table \ref{tab:timeline}. It is expected that a significant number of \textit{unpredictable changes} will occur as the code develops and the research progresses. I will update all committee members when appropriate.

\begin{table}
  \centering
  \renewcommand{\arraystretch}{1.2}
  \begin{tabular}{l l}
    \textbf{Date} & \textbf{Milestone} \\ \hline
    \textit{June 2019} & Implementation of monotone quintic polynomial \textit{interpolant}. \\
    \textit{August 2019} & Implementation of monotone quintic polynomial \textit{spline}. \\
    \textit{October 2019} & First draft of TOMS paper on algorithm. \\
    \textit{December 2019} & Research Defense of work. \\
    \textit{March 2020} & Submission of TOMS paper and code. \\
    \textit{April 2020} & Final Defense of Ph.D.
  \end{tabular}
  \caption{This table depicts my expected timeline going forward. For details on when I achieved previous milestones, please contact me and I can provide the list. I refrained from including it here for brevity.}
  \label{tab:timeline}
\end{table}

\section{Final Remarks}

This project has the potential to be very widely utilized. As described earlier, many statistical applications could greatly benefit from having a $C^2$ approximation to CDFs, because the resulting PDF will be both aesthetically pleasing and more accurate for continuous distributions.

### 2020-04-14 3-Disseration/strong.tex
Updated min wiggle fit tests, wrote report on the topic that was sent to Dr. Watson.

--- 2018-03-08 Paper_[2018-02-16]_IEEESE/paper.tex
@@
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%2345678901234567890123456789012345678901234567890123456789012345678901234567890
-%        1         2         3         4         5         6         7         8
-
-\documentclass[letterpaper, 10 pt, conference]{ieeeconf}  % Comment this line out
-                                                          % if you need a4paper
-%\documentclass[a4paper, 10pt, conference]{ieeeconf}      % Use this line for a4 paper
-\IEEEoverridecommandlockouts                              % This command is only needed if you want to use the \thanks command
-\overrideIEEEmargins
-% See the \addtolength command later in the file to balance the column lengths on the last page of the document
-
-% The following packages can be found on http:\\www.ctan.org
-\usepackage{tikz}     % For adding axes to plots
-\usepackage{graphicx} % For including PDFs
-\usetikzlibrary{positioning}
-
-%% \usepackage{graphics} % for pdf, bitmapped graphics files
-%\usepackage{epsfig} % for postscript graphics files
-%\usepackage{mathptmx} % assumes new font selection scheme installed
-%\usepackage{times} % assumes new font selection scheme installed
-\usepackage{amsmath} % assumes amsmath package installed
-\usepackage{amssymb}  % assumes amsmath package installed
-
-%% Command for creating multiplie-line cells inside a table environment
-\newcommand{\multilinecell}[2][c]{%
-  \begin{tabular}[#1]{@{}c@{}}#2\end{tabular}}
-
-\title{\LARGE \bf Nonparametric Distribution Models for Predicting and
-  \\ Managing Computational Performance Variability* }
-
-\author{Thomas C.
-H.
-Lux$^{1}$, Layne T.
-Watson$^{2}$, Jon Bernard$^{1}$, Tyler H.
-Chang$^{1}$, Bo Li$^{1}$, Xiaodong Yu$^{1}$,\\ Li Xu$^{3}$, Godmar Back$^{2}$, Ali R.
-Butt$^{2}$, Kirk W.
-Cameron$^{2}$, Yili Hong$^{4}$, Danfeng Yao$^{2}$% <-this % stops a space
-\thanks{*This work was supported by the National Science Foundation Grant CNS-1565314.}%
-\thanks{$^{1}$Doctoral student, Department of Computer Science, Virginia Polytechnic Institute and State University, Blacksburg, Virginia 24060, USA {\tt\small (tchlux at vt.edu)}}%
-\thanks{$^{2}$Department of Computer Science, Virginia Polytechnic Institute and State University, Blacksburg, Virginia 24060, USA}%
-\thanks{$^{3}$Doctoral student, Department of Statistics, Virginia Polytechnic Institute and State University, Blacksburg, Virginia 24060, USA}%
-\thanks{$^{4}$Department of Statistics, Virginia Polytechnic Institute and State University, Blacksburg, Virginia 24060, USA}%
-}
-
-\begin{document}
-
-\maketitle
-\thispagestyle{empty}
-\pagestyle{empty}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\begin{abstract}
-Performance variability can have a significant impact on many applications of computing.
-Cloud computing, high performance computing, and computer security communities each exert considerable effort managing and analyzing variability throughout the system stack.
-This work presents and evaluates a methodology for predicting precise characteristics of the computational performance variability of an input/output (I/O) application over varying system configurations.
-Results demonstrate that the presented methodology is capable of precisely modeling performance variability, which could allow applications that tighten service level agreements, maximize computational throughput, and obfuscate system configurations against malicious users.
-\end{abstract}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\section{INTRODUCTION}
-\label{sec:introduction}
-
-Computational variability presents itself in a variety of forms.
-Processes that are apparently identical in a cloud computing or high performance computing (HPC) environment may take different amounts of time to complete the same job.
-This variability can cause unintentional violations of service level agreements in cloud computing applications or indicate suboptimal performance in HPC applications.
-The sources of variability, however, are distributed throughout the system stack and often difficult to identify.
-The methodology presented in this work is applicable to modeling the expected variability of useful computer system performance measures without any prior knowledge of system architecture.
-Some examples of interesting performance measures that could be modeled with the techniques in this work include computational throughput, power consumption, processor idle time, number of context switches, and RAM usage, as well as any other ordinal measure of performance.
-
-Predicting performance variability in a computer system is a challenging problem that has primarily been attempted in one of two ways: (1) build a statistical model of the performance data collected by running experiments on the system at select settings, or (2) run artificial experiments using a simplified simulation of the target system to estimate architecture and application bottlenecks.
-In this paper, the proposed modeling techniques rest in the first category and represent a notable increase in the ability to model precise characteristics of variability.
-
-Many previous works attempting to model system performance have used simulated environments \cite{grobelny2007fase,wang2009simulation,wang2013towards}.
-Grobelny et al.
-refer to statistical models as being oversimplified and not capable of capturing the true complexity of the underlying system.
-Historically, statistical models have been limited to modelling at most one or two system parameters and have therefore not been capable of modeling the complexity of the underlying system.
-\cite{snavely2002framework,bailey2005performance,barker2009using,ye2010analyzing}.
-These limited statistical models have provided satisfactory performance in narrow application settings.
-However these techniques require additional code annotations, hardware abstractions, or additional assumptions about the behavior of the application in order to generate models.
-In contrast, the approach presented here requires no modifications of the application, no architectural abstractions, nor any structural descriptions of the input data being modeled.
-The techniques used in this work only require performance data as input.
-
-Most existing work on performance variability has focused on operating system (OS) induced variability \cite{beckman2008benchmarking,de2007identifying}.
-Yet, system I/O variability has been particularly difficult for statistical models to capture \cite{bailey2005performance}.
-The prior work attempting to model I/O variability has been similarly limited to one or two system parameters \cite{lofstead2010managing}.
-
-This paper presents and evaluates an application of multivariate interpolation to the domain of computer system I/O throughput.
-The interpolatory models are used to predict the cumulative distribution functions of the expected I/O throughput for a system at previously unseen configurations.
-Beyond this I/O case study, the techniques in this paper can tractably model tens of interacting system parameters with tens of thousands of known configurations.
-The primary contribution of this work is a modeling framework using multivariate interpolation to capture precise characteristics (via cumulative distribution functions) of an arbitrary performance measure on any type of computer system.
-
-Section \ref{sec:variability_modeling} details the mathematical techniques used to approximate variability, the chosen measurement of error, and an optimization strategy for improving model performance.
-Section \ref{sec:data} summarizes the data collection process as well as provides summary statistics of the data used for the case study.
-Section \ref{sec:results} presents the results of the I/O case study including analyses of approximation accuracy, model convergence, and system parameter importance.
-Section \ref{sec:discussion} discusses possible explanations for witnessed results, potential applications of the predictive methodology, and future improvements of the models.
-
-%% ===================================================================
-\section{VARIABILITY MODELING}
-\label{sec:variability_modeling}
-
 Performance and its variability can be summarized by a variety of statistics.
@@
 
-A major hurdle when modeling functions such as the CDF or PDF is that certain properties must be maintained.
-It is necessary that a PDF $f: \mathbb{R} \rightarrow \mathbb{R}$ have the properties $f(x) \geq 0$ and $\int_{-\infty}^{\infty}f(x)dx = 1$.
-However, for a CDF $F: \mathbb{R} \rightarrow \mathbb{R}$ the properties are $F(x) \in [0,1]$ and $F(x)$ is right continuous and nondecreasing.
-For the purpose of function approximation, the properties of the CDF are more straightforward to maintain.
-This work utilizes the fact that a convex combination of CDFs results in a valid CDF.
-Given $G(x) = \sum_{i}w_i F_i(x)$, $\sum_{i} w_i = 1$, $w_i \geq 0$, and each $F_i$ is a valid CDF, $G$ must also be a valid CDF.
-A demonstration of how this is applied can be seen in Figure \ref{fig:prediction_example}.
-Next, the three techniques used to model CDFs are summarized, a mechanism for identifying important features is proposed, and the error metric used throughout this work is presented.
-For each of the following, $n \in \mathcal{N}$ refers to the number of training samples (known system configurations), $d \in \mathcal{N}$ refers to the dimension of the data (ordinal system parameters), $x^{(i)} \in \mathbb{R}^d$ is a known system configuration with known performance CDF $F_{x^{(i)}}$, and $y \in \mathbb{R}^d$ refers to a previously unseen system configuration with performance CDF $F_y$.
-
 %% ===================================================================
-\subsection{Delaunay}
-\label{sec:delaunay}
-
-The Delaunay triangulation is a well-studied geometric technique for producing an interpolant \cite{lee1980two}.
-The Delaunay triangulation of a set of data points into simplices is such that there are no data points inside the sphere defined by the vertices of each simplex.
-For a $d$-simplex S with vertices $x^{(0)}$, $x^{(1)}$, $\ldots$, $x^{(d)}$, and functions $F_{x^{(i)}}$, $i=0$, $\ldots$, $d$, $y \in S$ is a unique convex combination of the vertices,
-
-$$ y = \sum_{i=0}^{d} w_i x^{(i)}, \quad \sum_{i=0}^{d} w_i = 1, \quad w_i \geq 0, \quad i=0,\ldots,d, $$
-
-and the Delaunay interpolant $F_y$ at $y$ is given by
-
-$$ F_y = \sum_{i=0}^{d} w_i F_{x^{(i)}}. $$
+\section{Measuring Error}
+\label{sec:error}
 
 \begin{figure}[htb]
+  \centering
   \begin{tikzpicture}
-    \node (img)  {\includegraphics[width=0.4\textwidth,]{plot-delaunay-example-prediction.pdf}};
+    \node (img)  {\includegraphics[width=0.75\textwidth,]{Figures/NA/Fig1.pdf}};
     \node[below=of img, node distance=1cm, yshift=1cm] {I/O Throughput};
@@
   \vspace{-0.3cm}
-  \caption{In this example, the general methodology for predicting a CDF and evaluating error can be seen.
-The Delaunay method chose three source system configurations (dotted lines) and assigned weights \{.3, .4, .3\} (top to bottom).
-The weighted sum of the three known CDFs produces the predicted CDF (dashed line).
-The KS Statistic (red arrow) computed between the true CDF (solid line) and predicted CDF (dashed line) is 0.2 for this example.
-The KS test null hypothesis is rejected by $p$-value 0.01, however it is not rejected by $p$-value 0.001.
-  \vspace{-.1cm}}
-  \label{fig:prediction_example}
+  \caption{In this HPC I/O example, the general methodology for
+    predicting a CDF and evaluating error can be seen. The Delaunay
+    method chose three source distributions (dotted lines) and
+    assigned weights \{.3, .4, .3\} (top to bottom at arrow). The
+    weighted sum of the three known CDFs produces the predicted CDF
+    (dashed line). The KS Statistic (arrow) computed between the true
+    CDF (solid line) and predicted CDF (dashed line) is 0.2 for this
+    example. The KS test null hypothesis is rejected at $p$-value
+    0.01, however it is not rejected at $p$-value 0.001.}
+  \label{fig:prediction-example}
 \end{figure}
 
-The computational complexity of Delaunay interpolation (for the implementation used) is $\mathcal{O}(n d^4 \log d)$, which is capable of scaling reasonably to $d \leq 50$.
-In the present application, a Delaunay simplex $S$ containing $y$ is found, then the $d+1$ vertices (system configurations) of $S$ are used to assign weights to each vertex and produce the predicted CDF for system configuration $y$.
+When the range of an approximation is the real numbers, error is
+reported with summary statistics including: min absolute error, max
+absolute error, and absolute error quartiles. When the range of an
+approximation is the space of cumulative distribution functions, the
+Kolmogorov-Smirnov statistic (max-norm difference between the
+functions) is used.
+
+A hurdle when modeling function-valued outputs such as cumulative
+distribution functions (CDFs) or probability density functions (PDFs)
+is that certain properties must be maintained. It is necessary that a
+PDF $f: \mathbb{R} \rightarrow \mathbb{R}$ have the properties $f(x)
+\geq 0$ and $\int_{-\infty}^{\infty}f(x)dx = 1$. However, for a CDF
+$F: \mathbb{R} \rightarrow \mathbb{R}$ the properties are
+$F(x) \in [0,1]$ and $F(x)$ is absolutely continuous and
+nondecreasing. This work utilizes the fact that a convex combination
+of CDFs (or PDFs) results in a valid CDF (or PDF). Given $G(x) =
+\sum_{i}w_i F_i(x)$, $\sum_{i} w_i = 1$, $w_i \geq 0$, and each $F_i$
+is a valid CDF, $G$ must also be a valid CDF. A demonstration of how
+this is applied can be seen in Figure \ref{fig:prediction-example}.
+
+The performance of approximation techniques that predict probability
+functions can be analyzed through a variety of summary
+statistics. This work uses the max absolute difference, also known as
+the Kolmogorov-Smirnov (KS) statistic \cite{lilliefors1967kolmogorov}
+for its compatibility with the KS test.
+
+The two-sample KS test is a useful nonparametric test for comparing
+two CDFs while only assuming stationarity, finite mean, and finite
+variance. The null hypothesis (that two CDFs come from the same
+underlying distribution) is rejected at level $p \in [0,1]$ when
+ $$ KS > \sqrt{-\frac{1}{2}\ln\biggl(\frac{p}{2}\biggr)} \sqrt{\frac{1}{n_1} + \frac{1}{n_2}}, $$
+with distribution sample sizes $n_1,n_2 \in \mathcal{N}$. For all
+applications of the KS test presented in this work $n_1 = n_2$. An
+example of the round-trip prediction methodology from known and
+predicted distributions to the calculation of error can be seen in
+Figure \ref{fig:prediction-example}.
 
 %% ===================================================================
-\subsection{Max Box Mesh}
-\label{sec:max_box_mesh}
 
-The Max Box mesh is an interpolation technique that utilizes overlapping box splines \cite{de2013box} as basis functions that are shifted and scaled to have support over box shaped regions.
-The boxes are constructed such that each box has exactly one point in its interior while maintaining the largest minimum distance between the interior point and the closest face of the box (a measure of centrality).
-Given a set of box splines $\bigl\{b^{x^{(i)}}\bigr\}$ with the max box properties anchored at interior points $\bigl\{x^{(i)}\bigr\}$,
-
-$$ F_y = \frac{\sum_i b^{x^{(i)}}(y) F_{x^{(i)}}}{\sum_i b^{x^{(i)}}(y)}. $$
-Note that the box splines always satisfy $b^{x^{(i)}}(y) \geq 0$. The computational complexity of interpolation via the Max Box mesh is $\mathcal{O}(n^2 d)$.
-
-%% ===================================================================
-\subsection{Voronoi Mesh}
-\label{sec:voronoi_mesh}
-
-A well-studied technique for classification and approximation is the nearest neighbor algorithm \cite{cover1967nearest}.
-Nearest neighbor inherently utilizes the convex region $C^{x^{(i)}}$ (Voronoi cell \cite{dirichlet1850reduction}) consisting of all points closer to $x^{(i)}$ than any other point $x^{(j)}$.
-The Voronoi mesh smooths the nearest neighbor approximation by utilizing the Voronoi cells to define support via a generic basis function $v: \mathbb{R}^d \rightarrow \mathbb{R}_+$ given by
-
-$$ v^{x^{(i)}}(y) = \left(1 - \frac{||y - x^{(i)}||_2}{2 \ d(y - x^{(i)} \mid x^{(i)})} \right)_+, $$
-where $d(w \mid x^{(i)})$ is the distance between $x^{(i)}$ and the boundary of the Voronoi cell $C^{x^{(i)}}$ in the direction $w$.
-$v^{x^{(i)}}\bigl(x^{(j)}\bigr) = 0$ and $v^{x^{(i)}}$ has local support, giving the interpolated value
-
-$$ F_y = \frac{\sum_i v^{x^{(i)}}(y) F_{x^{(i)}}}{\sum_i v^{x^{(i)}}(y)}, $$
-where $0 \leq v^{x^{(i)}}(y) \leq 1$.
-The computational complexity of interpolation via the Voronoi mesh is $\mathcal{O}(n^2 d)$.
-All of the approximations are an interpolant involving a convex combination of known functions $F_{x^{(i)}}$.
-%% ===================================================================
 \subsection{Feature Weighting}
@@
 %% ===================================================================
-\subsection{Measuring Error}
-
-The performance of approximation techniques that predict functions can be analyzed through a variety of summary statistics.
-Mean error, mean absolute error, mean squared error, and the max absolute error are all popular measures.
-This work uses the max absolute error, also known as the Kolmogorov-Smirnov (KS) statistic \cite{lilliefors1967kolmogorov} for its compatibility with the KS test.
-
-The two-sample KS test is a useful nonparametric test for comparing two CDFs while only assuming stationarity, finite mean, and finite variance.
-The null hypothesis (that two CDFs come from the same underlying distribution) is rejected at level $p \in [0,1]$ when
-
-$$ KS > \sqrt{-\frac{1}{2}\ln\biggl(\frac{p}{2}\biggr)} \sqrt{\frac{1}{n_1} + \frac{1}{n_2}}, $$
-with distribution sample sizes $n_1,n_2 \in \mathcal{N}$.
-For all applications of the KS test presented in this work $n_1 = n_2 = 150$.
-An example of the round-trip prediction methodology from known and predicted distributions to the calculation of error can be seen in Figure \ref{fig:prediction_example}.
-
-%% The majority of results presented in this work are absent any asymetric preprocessing operations on the training data, some findings related to automatic identification of important system parameters are presented in Table \ref{tab:optimized_p_value_failure_rate} and analyzed in \ref{sec:results}.
-
-%% ===================================================================
-\section{DATA}
+\section{Variability Data}
 \label{sec:data}
@@
   \end{tabular}
-  \caption{A description of system parameters considered for IOzone. Record size must be $\leq$ file size during execution.
-    \vspace{-.5cm}}
+  \caption{A description of system parameters considered for IOzone. Record size must be $\leq$ file size during execution.}
   \label{tab:data_description}
@@
 \begin{figure}
-  \includegraphics[width=.5\textwidth]{plot-histogram-throughput.pdf}
+  \centering
+  \includegraphics[width=.8\textwidth]{Figures/IEEE/plot-histogram-throughput.pdf}
   \caption{Histogram of the raw throughput values recorded during all IOzone tests across all system configurations.
-The distribution is skewed right, with few tests having significantly higher throughput than most others.
-  \vspace{-.5cm}}
+The distribution is skewed right, with few tests having significantly higher throughput than most others.}
   \label{fig:throughput_histogram}
@@
 
-This paper presents a variability modeling case study with a five-dimensional dataset produced by executing the IOzone benchmark \cite{iozone} on a homogeneous cluster of computers.
+This chapter utilizes a variability modeling case study with a five-dimensional dataset produced by executing the IOzone benchmark \cite{iozone} on a homogeneous cluster of computers.
 Each node contains two Intel Xeon E5-2637 CPUs offering a total of 16 CPU cores with 16GB of DRAM.
@@
 A single IOzone execution reports the max I/O throughput in kilobytes per second seen for the selected test type.
-The summary of the data components in $x^{(i)}$ for the experiments for this paper can be seen in Table \ref{tab:data_description}.
+The summary of the data components in $x^{(i)}$ for the experiments for this chapter can be seen in Table \ref{tab:data_description}.
 Distributions of raw throughput values being modeled can be seen in Figure \ref{fig:throughput_histogram}.
@@
 \begin{figure}
+  \centering
   \begin{tikzpicture}
-    \node (img)  {\includegraphics[width=0.5\textwidth,trim={1.5cm 0 0 0}]{plot-histogram-80_20-KS.pdf}};
+    \node (img)  {\includegraphics[width=0.8\textwidth,trim={1.5cm 0 0 0}]{Figures/IEEE/plot-histogram-80_20-KS.pdf}};
     \node[below=of img, node distance=1cm, yshift=1cm] {KS Statistic for Predicted vs. Actual};
@@
 The four vertical red lines represent commonly used $p$-values \{0.05, 0.01, 0.001, 1.0e-6\} respectively.
-All predictions to the right of a red line represent CDF predictions that are significantly different (by respective $p$-value) from the actual distribution according to the KS test.
-  \vspace{-.1cm}}
+All predictions to the right of a red line represent CDF predictions that are significantly different (by respective $p$-value) from the actual distribution according to the KS test.}
   \label{fig:ks_histogram_80_20}
@@
 
-\section{RESULTS}
-\label{sec:results}
+\section{Distribution Prediction Results}
 
@@
   \caption{Percent of null hypothesis rejections rate by the KS-test when provided different selections of $p$-values.
-These accompany the percent of null hypothesis rejection results from Figure \ref{fig:ks_histogram_80_20}.
-    \vspace{-.5cm}}
+These accompany the percent of null hypothesis rejection results from Figure \ref{fig:ks_histogram_80_20}.}
   \label{tab:p_value_failure_rate}
@@
 \begin{figure}
+  \centering
   \begin{tikzpicture}
-    \node (img)  {\includegraphics[width=0.4\textwidth,]{plot-KS-Failure-by-Training.pdf}};
+    \node (img)  {\includegraphics[width=0.8\textwidth,]{Figures/IEEE/plot-KS-Failure-by-Training.pdf}};
     \node[below=of img, node distance=1cm, yshift=1cm] {Percentage of Training Data};
@@
 The training percentages range from 5\% to 95\% in increments of 5\%.
-Delaunay is the best performer until 95\% of data is used for training, at which Max Box mesh becomes the best performer by a fraction of a percent.
-  \vspace{-.1cm}}
+Delaunay is the best performer until 95\% of data is used for training, at which Max Box mesh becomes the best performer by a fraction of a percent.}
   \label{fig:ks_failure_by_training}
@@
 MaxBoxMesh performance is improved by a negligible amount.
-VoronoiMesh performance is notably improved.
-    \vspace{-.5cm}}
+VoronoiMesh performance is notably improved.}
   \label{tab:optimized_p_value_failure_rate}
@@
 
-\section{DISCUSSION}
+\section{Discussion of Distribution Prediction}
 \label{sec:discussion}
@@
 
-As mentioned in Section \ref{sec:introduction}, no prior work has attempted to model an arbitrary performance measure for a system to such a high degree of precision.
+As mentioned at the beginning of this chapter, no prior work has attempted to model an arbitrary performance measure for a system to such a high degree of precision.
 All previous statistical modeling attempts capture a few ($<3$) ordinal performance measures.
@@
 Conversely, a service provider may defend against such attacks by deliberately obfuscating the performance of the machine.
-Models such as those presented in this paper could identify optimal staggering and time-delay whose introduction into the system would prevent malicious users from identifying system configurations and active jobs.
+Models such as those presented in this chapter could identify optimal staggering and time-delay whose introduction into the system would prevent malicious users from identifying system configurations and active jobs.
 
 \begin{figure}
+  \centering
   \begin{tikzpicture}
-    \node (img)  {\includegraphics[width=0.4\textwidth,]{plot-KS-failure-by-training-and-test.pdf}};
+    \node (img)  {\includegraphics[width=0.8\textwidth,]{Figures/IEEE/plot-KS-failure-by-training-and-test.pdf}};
     \node[below=of img, node distance=1cm, yshift=1cm] {Percentage of Training Data};
@@
 The training percentages range from 5\% to 95\% in increments of 5\%.
-The read test types tend to allow lower rejection rates than the write test types.
-  \vspace{-.1cm}}
+The read test types tend to allow lower rejection rates than the write test types.}
   \label{fig:ks_failure_by_training_and_test}
@@
 
-\section{CONCLUSION}
-\label{sec:conclusion}
+\section{The Power of Distribution Prediction}
 
@@
 The present methodology represents a notable increase in the ability to statistically model arbitrary system performance measures involving the interaction of many ordinal system parameters.
-%% \begin{itemize}
-%% \item Modeling the distributions of system performance is valuable.
-%% \item This work presents a steep increase in the power of predictions made by statistical modeling techniques.
-%% \end{itemize}
-
-\bibliographystyle{IEEEtran}
-\bibliography{paper}
-
-\end{document}

### 2020-04-14 Research_[2020-04-07]_Fit_Min_Wiggle/polynomial.py

@@ -21,6 +21,8 @@
 #   fit              -- Use a local polynomial interpolant to estimate
 #                       derivatives and construct an interpolating
 #                       Spline with a specified level of continuity.
+#   local_polynomial -- Construct a local polynomial interpolant about
+#                       a given point in a list of points.
 #   fill_derivative  -- Compute all derivatives at points to be
 #                       reasonable values using either a linear or a
 #                       quadratic fit over neighboring points. 
@@ -32,7 +34,6 @@
 #                       "index" in the set. Use Shepard weighting.
 #   inverse          -- Given a function, use the Newton method to
 #                       find the nearest inverse to a guess point.
-
 # 
 
 from fraction import Fraction
@@ -215,7 +216,7 @@ class Spline:
             k = knots[i]
             my_poly = self.function_at(k)
             other_poly = other.function_at(k)
-            order = len(my_poly.coefficients) + len(other_poly.coefficients) - 1
+            order = max(2,len(my_poly.coefficients) + len(other_poly.coefficients) - 1)
             # Evaluate the function at equally spaced "x" values, TODO:
             # this should be Chebyshev nodes for numerical stability.
             x = [(step / (order-1)) * (right - left) + left for step in range(order)]
@@ -410,21 +411,37 @@ def to_polynomial(coefficients, points):
 # Given unique "x" values and associated "y" values (of same length),
 # construct an interpolating polynomial with the Newton divided
 # difference method. Return that polynomial as a function.
-def polynomial(x, y):
+def polynomial(given_x, given_y, **derivs):
     # Sort the data by "x" value.
-    indices = sorted(range(len(x)), key=lambda i: x[i])
-    x = [x[i] for i in indices]
-    y = [y[i] for i in indices]
+    indices = sorted(range(len(given_x)), key=lambda i: given_x[i])
+    # Construct the initial x and y (with repetitions to match derivatives).
+    x = []; y = []; index_ranges = {}
+    for i in indices:
+        dxs = sorted(key for key in derivs if (i == int(key.split('x')[1])))
+        largest_d = ([key.count("d") for key in dxs] + [0])[0]
+        index_ranges[(len(x), len(x)+largest_d)] = i
+        x += [given_x[i]] * (largest_d+1)
+        y += [given_y[i]] * (largest_d+1)
     # Compute the divided difference table.
+    divisor = 1
     dd_values = [y]
     for d in range(1, len(x)):
+        divisor *= d
[…]

### 2020-04-14 Research_[2020-04-07]_Fit_Min_Wiggle/post_process_data.py

@@ -1,0 +1,85 @@
+import os
+from util.data import Data
+
+
+out_file = "all_data.csv"
+
+if os.path.exists(out_file): d = Data.load(out_file, sample=None)
+else:
+
+    # Load in all of the existing data by seed file.
+    d = Data()
+    for f in sorted(os.listdir()):
+        if not (("seed-" == f[:len("seed-")]) and (f[-4:] == ".csv")): continue
+        f_data = Data.load(f)
+        f_data["Seed"] = int(f[5:-4])
+        d += f_data
+    # Copy in the names.
+    d.reorder(d.names[:2] + ["Seed"])
+    d.names[1] = "Trial Number"
+
+    # Re-organize the accumulated seed data into one big data.
+    methods = sorted({" ".join(n.split()[:2]) for n in d.names[3:]})
+    names = d.names[:3] + ["Method"] + sorted(
+        {" ".join(n.split()[2:]) for n in d.names[3:]} )
+
+    # Get the indices of columns for each method.
+    method_columns = {
+        m : [i for i in range(len(d.names)) if m in d.names[i]]
+        for m in methods }
+
+    # Sort the indices of the method columns by the order in "names".
+    for m in method_columns:
+        method_columns[m].sort(key=lambda i: names.index(" ".join(d.names[i].split()[2:])))
+
+    # Initialize the new data.
+    old_d = d
+    new_d = Data(names=names)
+    # print("methods: ",methods)
+    # print("names: ",names)
+    # print("method_columns: ",method_columns)
+
+    # Parse the old rows into new rows.
+    for row in old_d:
+        n, t, s = row[:3]
+        for m in methods:
+            new_d.append( [n,t,s,m] + [row[i] for i in method_columns[m]] )
+
+
+    # Rename to the variable "d".
+    d = new_d
+    d.save("all_data.csv")
+
+
+print(d)
+methods = sorted(set(d["Method"]))
+measures = d.names[4:]
+max_len = max(map(len,methods))
+
+from util.stats import rank_probability
+for meth in methods:
+    meth_data = list(d[d["Method"] == meth]["L2"])
+    other_data = [list(d[d["Method"] == other_meth]["L2"]) for
+                  other_meth in methods if other_meth != meth]
+    rp = 100*rank_probability(meth_data, other_data, rank=0)
+    print(f"{meth:{max_len}s} rank 0 probability: {rp:.1f}%")
+exit()
+
+import numpy as np
[…]

### 2020-04-14 Research_[2020-04-07]_Fit_Min_Wiggle/test_abs_value.py

@@ -1,0 +1,258 @@
+import numpy as np
+from polynomial import Spline, fit, local_polynomial, local_quadratic, \
+    polynomial, Polynomial
+from monotone import monotone_quintic_spline
+from fraction import Fraction
+from util.plot import Plot
+
+
+
+# Make a plot of the points where the constraints 
+def example_points(x, y, ms=10, p=None):
+    # Find all "zero derivative" points.
+    d_pos = []
+    d_neg = []
+    d_zero = []
+    dd_zero = []
+    for i in range(len(x)):
+        left  = (y[i]-y[i-1]) if (i > 0)        else ((-1)**int(y[0] > y[1])   if len(y) > 1 else 0)
+        right = (y[i+1]-y[i]) if (i+1 < len(y)) else ((-1)**int(y[-2] > y[-1]) if len(y) > 1 else 0)
+        if   (left * right) < 0:             d_zero.append(i)
+        elif ((left == 0) or (right == 0)):  dd_zero.append(i)
+        elif (left > 0):                     d_pos.append(i)
+        elif (left < 0):                     d_neg.append(i)
+        else: print("This is unexpected..", left, right)
+    # Create the plot.
+    if (p is None): p = Plot()
+    p.add("Local Extrema (Df = 0)", [x[i] for i in d_zero], [y[i] for i in d_zero], color=0, symbol="circle", marker_size=ms, marker_line_width=2)
+    p.add("Flat slope (DDf = 0, Df=0)", [x[i] for i in dd_zero], [y[i] for i in dd_zero], color=1, symbol="square", marker_size=ms, marker_line_width=2)
+    p.add("Monotone increasing (Df ≥ 0)", [x[i] for i in d_pos], [y[i] for i in d_pos], color=2, symbol="triangle-up", marker_size=ms, marker_line_width=2)
+    p.add("Monotone decreasing (Df ≤ 0)", [x[i] for i in d_neg], [y[i] for i in d_neg], color=3, symbol="triangle-down", marker_size=ms, marker_line_width=2)
+    return p
+
+
+# Visualize multiple methods.
+def visualize_multiple_methods(x, y, name=["Local quintic"], title="",
+                               method=[6], color=[1], dash=[None],
+                               **kwargs):
+    # Initialize the plot.
+    p = Plot(title=title, font_family="times")
+    # Add visuals of all methods.
+    for (n,m,c,d) in zip(name, method, color, dash):
+        if m is not None:
[…]

### 2020-04-14 Research_[2020-04-07]_Fit_Min_Wiggle/test_monotone_wiggle.py

@@ -1,0 +1,134 @@
+import numpy as np
+from fraction import Fraction
+from monotone import monotone_quintic_spline
+from util.plot import Plot
+from util.data import Data
+
+# Compute the L2 of a function over a range.
+def L2(f, low, upp):
+    f2_int = (f.derivative(2)**2).integral()
+    return f2_int(upp) - f2_int(low)
+
+# Construct multiple different fits over x and y values
+def compare_fits(x_vals, y_vals, plot=False):
+    x_vals = list(map(Fraction, x_vals))
+    y_vals = list(map(Fraction, y_vals))
+
+    # Execute with different operational modes, (tests happen internally).
+    functions = dict(
+        quintic_interp = monotone_quintic_spline(x_vals, y_vals, method="poly"),
+        quadratic_regres = monotone_quintic_spline(x_vals, y_vals, method="quad"),
+        quadratic_fill = monotone_quintic_spline(x_vals, y_vals, method="fill", mids=2, ends=2),
+    )
+
+    # Compute the L2 of the second derivatives.
+    wiggles = {name+"_L2":L2(functions[name],min(x_vals),max(x_vals))
+               for name in functions}
+    # Compute the second derivative values. and percentiles of them.
+    ddf_pos = np.linspace(float(x_vals[0]), float(x_vals[-1]), num=100)
+    for name in functions:
+        ddf_vals = abs(np.array(functions[name].derivative(2)(ddf_pos), dtype=float))
+        for p in (0, 25, 50, 75, 100):
+            wiggles[name+f"_percentile_{p:03d}"] = np.percentile(ddf_vals,p)
+
+    # Plot the data if the user wanted that.
+    if plot:
+        p = Plot()
+        p.add("Points", x_vals, y_vals)
+        # Add the functions
+        for name in sorted(functions):
+            p.add_func(name, functions[name], [min(x_vals), max(x_vals)])
+        # Add the L2 functions.
+        for i,name in enumerate(sorted(functions)):
+            f = functions[name]
+            f2_int = (f.derivative(2)**2).integral()
+            p.add_func(name, f2_int, [min(x_vals), max(x_vals)],
+                       color=i+1, dash="dot", group=0)
+        p.show()
+
+    # Return the results.
+    return wiggles
+
+
+# Run the tests.
+def run_tests(seed=0, trials=10, Ns=(4,8,16,32,64)):
[…]

### 2020-06-08 Research_[2020-05-18]_Piecewise_Monotone_Quintic_Fortran/paper/del/delTOMS18.tex
Added paper directory, all of this work may be moved into the papers repository soon.

% Algorithm XXX: DELAUNAYSPARSE---Interpolation via a sparse subset of
% the Delaunay triangulation in medium to high dimensions,
% Tyler H. Chang, Layne T. Watson, Thomas C.H. Lux, Ali R. Butt,
% Kirk W. Cameron, and Yili Hong,
% ACM Trans. Math. Software, ?/?/2018.
% 
%
\magnification=1000 \hsize=30pc \vsize=47pc
\parindent=10pt \parskip=0pt plus 3pt
\baselineskip=13pt plus 2pt minus 1pt \lineskiplimit=2pt
\lineskip=2pt plus 2pt                                  %\raggedright
\tolerance=800
%\abovedisplayskip=6pt plus 6pt minus 3pt
%\abovedisplayshortskip=0pt plus 3pt
%\belowdisplayskip=6pt plus 6pt minus 3pt
%\belowdisplayshortskip=6pt plus 2pt minus 3pt

% Color definitions.
\input colordvi
\def\BEGINC{\textPeach} \def\ENDC{\textBlack}
\def\red#1{\BEGINC#1\ENDC}
\let\beginred=\BEGINC
\let\endred=\ENDC

% fonts
\font\itVIII=cmti8 \font\slVIII=cmsl8 \font\bfVIII=cmbx8
\font\ttVIII=cmtt8 \font\caps=cmcsc10 \font\capsVIII=cmcsc10 at 8pt
\font\ttXVIII=cmtt12 at 18pt \font\eightmi=cmmi8 \font\eightsy=cmsy8
\font\rmeight=cmr8  \def\rmVIII{\rmeight \baselineskip=10pt plus 2pt
minus 1pt \lineskiplimit=2pt \lineskip=2pt plus 1pt
\parskip=0pt \textfont0=\rmeight \textfont1=\eightmi \textfont2=\eightsy}
\font\hlvXVIII=phvr at 18pt \font\hlvVIII=phvr at 8pt \font\hlv=phvr

\input epsf %Postscript figures.
\def\lcaption#1#2{{\noindent\rmVIII Fig.~#1. #2\par}}
\def\ltabcaption#1#2{{\noindent\rmVIII Table #1. #2\par}}

% other definitions.
\def\h{\par\hangindent=10pt\hangafter=1\noindent}
\def\heading#1{\par\bigskip\leftline{\hlv#1}\nobreak\smallskip
\noindent\ignorespaces}
\newdimen\headdigit{\setbox11=\hbox{\hlv 8. }\global\headdigit=\wd11}
\def\footnoterule{\bigskip\kern-3pt \hrule \kern 2.6pt \smallskip}
\def\bullitem{\par\vskip 2pt plus 2pt\hangindent=15pt \hangafter=1
\noindent\hbox to 15pt{---\hfil}\ignorespaces}
\def\bulllitem{\par\vskip 2pt plus 2pt\hangindent=30pt
\hangafter=1 \noindent\hskip 15pt\hbox to
15pt{---\hfil}\ignorespaces}
\def\Th#1 #2\par{\par\medskip{\it{\caps#1}. #2}\par\medskip}
\def\btable#1{\topinsert \rmVIII \let\bf=\bfVIII \textfont0=\scriptfont0
\scriptfont0=\scriptscriptfont0 \textfont1=\scriptfont1
\scriptfont1=\scriptscriptfont1 \centerline{#1}}
\let\xpar=\par
\def\ds{\displaystyle}\def\ab{\allowbreak}

% math definitions
\def\tn#1{\left\Vert #1\right\Vert_2}
\def\mn#1{\left\Vert #1\right\Vert_\infty}
\def\lee{\mathrel{\vcenter{\hbox{$\scriptstyle\mathord<$}\nointerlineskip
\vskip 1pt\hbox{$\scriptstyle\mathord=$}}}}
\def\gee{\mathrel{\vcenter{\hbox{$\scriptstyle\mathord>$}\nointerlineskip
\vskip 1pt\hbox{$\scriptstyle\mathord=$}}}}
\def\real{\mathop{\rm I\!R}\nolimits}
\def\R{{\bf R}}
\def\Rd{{\bf R}^d}
\def\Rdd{{\bf R}^{d\times d}}
%\def\bull{\hfill \vrule height6pt width6pt depth0pt}
\def\:{\mathrel{:\mathord=}}

% OUTPUT
\def\rightheadline{\hfill\hlvVIII\rightrh\hskip20pt$\scriptstyle\bullet$\hskip
20pt\folio}
\def\leftheadline{\hlvVIII\folio\hskip20pt$\scriptstyle\bullet$\hskip
20pt\leftrh\hfill}
\footline={\hfill}%no page number on first page
\output{\plainoutput}
\def\plainoutput{\shipout\vbox{\makeheadline\pagebody\makefootline}
  \advancepageno \global\footline={\hfill}
  \ifodd\pageno
        \global\headline={\rightheadline}
  \else
        \global\headline={\leftheadline}
  \fi
  \ifnum\outputpenalty>-20000 \else\dosupereject\fi}
\def\pagebody{\vbox to\vsize{\boxmaxdepth\maxdepth \pagecontents}}
\def\makeheadline{\vbox to 0pt{\vskip-22.5pt
  \line{\vbox to8.5pt{}\the\headline}\vss}\nointerlineskip}
\def\makefootline{\baselineskip24pt\vskip-6pt\line{\the\footline}}
\def\dosupereject{\ifnum\insertpenalties>0pt
  \line{}\kern-\topskip\nobreak\vfill\supereject\fi}
%\def\footnoterule{\vskip10pt\kern-3pt \hrule width 3em \kern 2.6pt}

\def\leftrh{Chang et al.}
\def\rightrh{Algorithm XXX: DELAUNAYSPARSE}

{\baselineskip=24pt \leftline{\hlvXVIII Algorithm XXX: DELAUNAYSPARSE:}
\leftline{\hlvXVIII Interpolation via a Sparse Subset of the}
\leftline{\hlvXVIII Delaunay Triangulation in Medium to High}
\leftline{\hlvXVIII Dimensions}}
\bigskip\medskip
\leftline{\hlv TYLER H. CHANG, LAYNE T. WATSON, THOMAS C.H. LUX, ALI R. BUTT,}
\smallskip
\leftline{\hlv KIRK W. CAMERON, AND YILI HONG}
\smallskip
\leftline{\hlv Virginia Polytechnic Institute and State University}
\bigskip

\hrule\bigskip\smallskip \footnote{}{\hskip
-\parindent{\parindent=0pt\rmVIII Authors' addresses:
T. H. Chang, T. C. H. Lux, A. R. Butt, K. W. Cameron,
Department of Computer Science, L. T. Watson,
Departments of Computer Science, Mathematics, and Aerospace and Ocean
Engineering, Y. Hong, Department of Statistics, Virginia
Polytechnic Institute \& State University, Blacksburg, VA 24061;
e-mail: {\ttVIII thchang@vt.edu}. \xpar
% Permission to make digital/hard copy of
%part or all of this work for personal or classroom use is granted
%without fee provided that the copies are not made or distributed for
%profit or commercial advantage, the copyright notice, the title of
%the publication, and its date appear, and notice is given that
%copying is by permission of the ACM, Inc. To copy otherwise, to
%republish, to post on servers, or to redistribute to lists, requires
%specific permission and/or fee. \xpar\copyright\ 2018 by the
%Association for Computing Machinery, Inc. \xpar 
}}

% Abstract and Introduction
{\rmVIII\parindent=0pt DELAUNAYSPARSE contains both serial and
parallel codes written in Fortran 2003 (with OpenMP) for performing
medium- to high-dimensional interpolation via the Delaunay triangulation.
To accommodate the exponential growth in the size of the Delaunay
triangulation in high dimensions, DELAUNAYSPARSE computes only a
sparse subset of the complete Delaunay triangulation, as necessary for
performing interpolation at the user specified points. This paper includes
algorithm and implementation details, complexity and sensitivity analyses,
usage information, and a brief performance study.

\medskip
Categories and Subject Descriptors: G.1.1 [{\bfVIII Numerical Analysis}]:
Interpolation --- Spline and piecewise polynomial interpolation;
J.2 [{\bfVIII Computer Applications}]: Physical Science and Engineering
--- {\itVIII Mathematics};
G.4 [{\bfVIII Mathematics of Computing}]: Mathematical Software

\medskip
General Terms: Algorithms, Computational Geometry, Documentation

\medskip
Additional Key Words and Phrases: Delaunay triangulation, interpolation,
multivariate approximation, high-dimensional data

}
\bigskip\hrule\bigskip\medskip
\heading{1. INTRODUCTION}

The Delaunay triangulation is an unstructured simplicial mesh that
is widely studied in the field of computational geometry. Due to its
many favourable properties, the Delaunay triangulation finds wide use
as a mesh for multivariate interpolation in the fields of geographic
information systems (GIS), civil engineering, physics, and computer
graphics. See Section 9.6 of [de Berg et al.\ 2008] for a brief
discussion of the usefulness of Delaunay triangulations, and see
[Schaap et al.\ 2000] for a specific usage example from computational
physics. The viability of Delaunay triangulations as a means for
interpolating arbitrary nonlinear functions in the context of data science
and machine learning has been explored by Omohundro [1990]. However,
this particular usage of the Delaunay triangulation never gained
widespread popularity, most likely due to the exponential computational
complexity of high-dimensional Delaunay triangulations. More recent
works have shown Delaunay triangulations to be effective for
interpolating real-world computer system data [Chang et al.\ 2018a 
and Lux et al.\ 2018], outperforming several common multivariate
interpolation and approximation techniques.

In this work, the main problem of interest is the multivariate
interpolation problem. Interpolation via meshes such as triangulations
and other tessellations is a classic practice. Delaunay triangulations
are widely considered optimal simplicial meshes for many meshing
applications, including interpolation. See the first chapter of Cheng
et al.\ [2012] for an overview of Delaunay meshing applications and
theory, and see Rajan [1994] for specific theorems on the optimality of
Delaunay triangulations in arbitrary dimension.

In two dimensions, the Delaunay triangulation of $n$ points can be
efficiently computed in ${\cal O}\big( n \log n \big)$ time
[Su et al.\ 1995]. After the Delaunay triangulation has been computed,
the cost of evaluating each interpolation point is reduced to the cost
of point location. In two dimensions, point location can be performed
in ${\cal O}\big( n^{1\over 3} \big)$ time [M{\"u}cke et al.\ 1999],
so the total cost of interpolating at $m$ points in two dimensions is
${\cal O}\big( n\log n + n^{1\over 3}m \big)$. However, Klee [1980]
showed that in $\Rd$, the worst case size of the Delaunay triangulation
is ${\cal O}\big(n^{\lceil d/2 \rceil}\big)$. Even in the generic case,
the Delaunay triangulation still tends to grow exponentially with the
dimension, a phenomenon often associated with 
{\it the curse of dimensionality}.

Despite the complexity of high-dimensional Delaunay triangulations,
there are currently a wide variety of algorithms for computing them.
The first algorithm capable of computing Delaunay triangulations in
arbitrary dimension was proposed independently by both Bowyer [1981]
and Watson [1981]. Perhaps the most widely used algorithm for computing
Delaunay triangulations in arbitrary dimension is the Quickhull algorithm
proposed by Barber et al.\ [1996]. Quickhull is a time-efficient
algorithm running in $\Theta \big(n\log n + k\big)$ time, where $k$
denotes the size of the Delaunay triangulation. Quickhull also boasts
a highly optimized numerically stable implementation. An alternative
to Quickhull is the graph based algorithm proposed by Boissonnat et
al.\ [2009]. An implementation of this algorithm, contained in the
Computational Geometry Algorithms Library (CGAL), stores the
Delaunay triangulation in a memory efficient graph structure, at the
cost of a slightly greater compute time. One final algorithm of interest
is the DeWall algorithm, proposed by Cignoni et al.\ [1998]. The DeWall 
algorithm, though not in widespread use, features a unique
divide-and-conquer paradigm, and was a major inspiration behind this work.

Due to the exponential growth of Delaunay triangulations in high
dimensions, none of the above mentioned algorithms are intended to
scale past six or seven dimensions. In fact, this failure to scale
to high dimensions is suffered by nearly every mesh based approximation.
Consequently, high-dimensional approximation is generally dominated by
mesh free methods such as multivariate polynomials, radial basis
functions, low order splines, inverse distance weightings, kernel
methods, and machine learning techniques such as support vector
regressors and artificial neural networks (see [Cheney et al.\ 2009]).
By leveraging the sparse nature of the interpolation problem, the
DELAUNAYSPARSE package aims to add the high-dimensional Delaunay mesh
based interpolant to the numerical analyst's toolbox.

The rest of this paper is organized as follows. Section 2 describes
the Delaunay interpolant in greater detail and outlines a novel algorithm
for computing it. Section 3 details the computational aspects of the
algorithm, with an emphasis on numerical stability and efficiency.
Section 4 describes the serial implementation of the algorithm and its
additional features. Section 5 describes the parallel implementation,
which uses OpenMP in a shared memory paradigm. Section 6 contains usage
information and package organization details. Section 7 shows performance
statistics, demonstrating the scalability of the algorithm. For additional
information on the properties of Delaunay triangulations and algorithms for
computing them, two excellent references are [Aurenhammer et al.\ 2013]
and [Cheng et al.\ 2012].

\heading{2. INTERPOLATION VIA THE DELAUNAY TRIANGULATION}

Let $P$ be a set of $n>d$ data points in $\Rd$. A $d$-dimensional
triangulation is defined as a mesh of $d$-simplices that
(1) are disjoint except on their shared boundaries,
(2) whose set of vertices is $P$, and 
(3) whose union is the convex hull of $P$, denoted $CH(P)$.
The interpolation problem is: given values $f(p)$ for all points
$p \in P$ where $f : \Rd \rightarrow \R^m$, find an approximation
${\hat f} \approx f$ such that ${\hat f}(p) = f(p)$ for all $p\in P$,
where ${\hat f}$ has support in $CH(P)$.

Let $T(P)$ be a $d$-dimensional triangulation of $P$. To define an
interpolant in terms of $T(P)$, let $q \in CH(P)$ be an interpolation
point, and let $S$ be a simplex in $T(P)$ with vertices 
$s_1$, $\ldots$, $s_{d+1}$ such that $q \in S$. Then there exist weights
$w_1$, $\ldots$, $w_{d+1}$ such that $q = \sum_{i=1}^{d+1} w_i s_i$,
$\sum_{i=1}^{d+1} w_i = 1$, and $w_i \geq 0$ for $i=1$, $\ldots$, $d+1$,
and the interpolant ${\hat f}_T$ is given by
$$
{\hat f}_T(q) = w_1f(s_1) + w_2f(s_2) + \ldots + w_{d+1}f(s_{d+1}).
\eqno(1)
$$

In DELAUNAYSPARSE, the interpolant ${\hat f}_{DT}$ is computed, where
$DT(P)$ denotes a Delaunay triangulation of $P$. The Delaunay
triangulation is often defined as the geometric dual of the Voronoi
diagram, also called the Dirichlet tessellation. Here the following
equivalent definition is preferred. For a $d$-simplex $S$,
let $B_S$ denote the open ball whose center and radius are given
by the $(d-1)$-sphere circumscribing $S$. Then a Delaunay triangulation
$DT(P)$ of a finite set of points $P \subset \Rd$ is any triangulation
of $P$ such that for each $S \in DT(P)$, $B_S$ satisfies
$B_S \cap P = \emptyset$. Remarks 2.1--3 below describe several key
properties of a Delaunay triangulation.

\medskip{\sl Remark 2.1.}\enspace
Given a set of $d+1$ vertices in $P$ that define a
$d$-simplex $S$, the condition that $B_S \cap P = \emptyset$ is not
only necessary, but also sufficient to conclude that $S \in DT(P)$
for {\it some} Delaunay triangulation $DT(P)$.

\medskip{\sl Remark 2.2.}\enspace
Let $F$ be a facet of a simplex $S \in DT(P)$. Let 
$p_1$, $\ldots$, $p_\ell$ be a sequence of points that are in $P$
and are in the same halfspace $H$ with hyperplane boundary
containing $F$. Define the open circumballs $B_1$, $\ldots$, $B_\ell$
such that each $B_i$ circumscribes $F$ and $p_i$. Assume 
$p_1$, $\ldots$, $p_\ell$ satisfies $p_i \in B_{i+1}$ for all 
$1 \leq i < \ell$.
Then $B_1 \cap H \subset B_2 \cap H \subset \cdots \subset B_\ell \cap H$.

\medskip{\sl Remark 2.3.}\enspace
In randomly generated data, the cases where $DT(P)$ does not exist
or is not unique occur with probability zero. Therefore, for
algorithmic analysis, it is common to make the simplifying assumption
that $P$ is in {\it general position}, meaning $DT(P)$ exists and is
unique. Furthermore, note that the case where $DT(P)$ does not exist
occurs only if all the points in $P$ are contained in some 
lower-dimensional linear manifold. In the context of interpolation,
this corresponds to an overparameterization of the underlying function
and can be resolved with dimension reduction techniques.
The case where $DT(P)$ is not unique can occur in real-world problems
and will be addressed in the implementation, discussed in Section 3.
\medskip

Note that given a set of $m$ interpolation points $Q$, at most $m$
simplices in $DT(P)$ are needed to compute ${\hat f}_{DT}(q)$ for
all $q\in Q$. Therefore, for this particular problem, it is possible
to ``cheat'' the curse of dimensionality when $m$ is significantly
less than the size of $DT(P)$. To do so, it suffices to compute
a sparse subset of $DT(P)$ such that $Q$ is contained in the subset.

As previously mentioned, one of the major inspirations behind this
work was the DeWall algorithm proposed by Cignoni et al.\ [1998].
The DeWall algorithm features a divide-and-conquer paradigm where
construction of each Delaunay simplex is carefully guided so as to
construct a ``wall'' of simplices, bisecting the data set. Each
successive simplex is completed from a facet of a previously constructed
simplex, using the same methodology as the classic gift-wrapping
approach (described in Section 5.6 of [Cheng et al.\ 2012]).

The two key components of the DeWall and gift-wrapping algorithm
that are used in DELAUNAYSPARSE are the growth of the seed simplex
and the completion of an open Delaunay facet. After iteratively
constructing a seed simplex, the idea is to perform a 
{\it visibility walk} to the simplex containing each interpolation
point, as described by Devillers et al.\ [2002]. Since the complete
triangulation is never computed, each step of the walk is performed by
completing the Delaunay facet designated by the visibility walk protocol.
Once each interpolation point $q$ has been located, each response value
${\hat f}_{DT}(q)$ can be computed using (1). A detailed description
and analysis of this algorithm is in Section 3 of [Chang et al.\ 2018b].
Pseudocode for interpolating at a single point follows.

{\parindent =0pt \parskip= 0pt
\smallskip
\leftskip 20pt
{\sl Algorithm 1}
\smallskip
$P$ contains $n$ $d$-dimensional data points;\par
$q \in CH(P)$ is the interpolation point;\par
$S$ denotes the current $d$-simplex; \par
$F$ denotes the facet of $S$ from which $q$ is visible. \par
\smallskip
{\bf begin}
Grow an initial seed $d$-simplex $S$, as described in Section 3.1.\par
{\bf while} $q \not\in S$ {\bf do}\par
\leftskip 40pt
Select the facet $F$ of $S$ from which $q$ is visible as
described in Section 3.2;\par
complete a new $d$-simplex $S^*$ from the facet $F$ as described 
in Section 3.3;\par
update $S \leftarrow S^*$.\par
\leftskip 20pt
{\bf enddo}\par
Since the loop has terminated, $q\in S$.
Compute ${\hat f}_{DT}(q)$ using (1).\par
\smallskip}

The advantages of this technique are maximized when the interpolation
points are sparse with respect to the size of the triangulation. In
both expectation and practice, the number of simplices constructed
during the walk to each interpolation point is a polynomial function
of $d$ and often seemingly independent of $n$. Given the exponential
nature of the problem, this makes for an effective sparse solution,
particularly in high dimensions.

\heading{3. COMPUTATIONAL ASPECTS}

In this section, the computational operations referenced in Algorithm 1
will be fully detailed.
These operations are the growth of the seed simplex, the visibility walk,
and flipping across a Delaunay facet.

\heading{3.1 \enspace Growing the Seed Simplex}

The seed simplex is constructed through a greedy algorithm, as detailed
in Section 3.1 of Chang et al.\ [2018b]. The initial vertex $s_1$ is
chosen to be the closest point in the data set $P$ to the interpolation
point $q$, and ties are resolved by choosing the point in $P$ with the
lowest index.
The second vertex $s_2 \in P \setminus \{s_1\}$ is chosen such that
$$
\|s_2 - s_1\| = \min_{p \in P, \atop p\ne s_1} \|p - s_1\|.
$$
Each subsequent vertex is chosen to minimize the radius of the minimum
radius circumsphere for the resulting vertex list.

For $2 \leq j \leq d$ and $p \in P\setminus \{s_1$, $\ldots$, $s_j\}$,
define the $j \times d$ matrix 
$$
A^{(j,p)} = \left[ \matrix{
  \left(s_2 - s_1\right)^T \cr
  \vdots                   \cr
  \left(s_j - s_1\right)^T \cr
  \left( p - s_1 \right)^T } \right]
$$
and the $j$-vector
$$
b^{(j,p)} = \left[ \matrix{
  \displaystyle{\|s_2 - s_1\|^2 \over 2} \cr
  \vdots                    \cr
  \displaystyle{\|s_j - s_1\|^2 \over 2} \cr
  \displaystyle{\|p - s_1\|^2 \over 2}  } \right].
$$

If rank~$A^{(j,p)} = j$, then the minimum norm solution to the
underdetermined system
$$
A^{(j,p)} x = b^{(j,p)} \eqno{(2)}
$$
is $x^* = c - s_1$, where $c$ denotes the center of the
minimum radius circumsphere about $s_1$, $\ldots$, $s_j$, $p$.
So, each subsequent vertex $s_{j+1}$ is given by the 
$p^* \in P \setminus \{s_1$, $\ldots$, $s_j\}$ such
that solving (2) with $A^{(j,p^*)}$ and $b^{(j,p^*)}$ produces the
minimum 2-norm solution $x^*$.

If rank~$A^{(j,p)} < j$, then $s_1$, $\ldots$, $s_j$, $p$ are not
the vertices of a $j$-simplex, and $p$ cannot be a vertex of any
$d$-simplex with vertices $s_1$, $\ldots$, $s_j$.
Hence, $p$ can be skipped and need not be considered again when
constructing the seed simplex. Due to memory constraints, there is no
mechanism for marking a $p$ that can be skipped, and hence any such $p$
could be revisited in the future, though it will always be skipped.

If $P$ is in general position, then there will always exist a unique
point $p^*$ that minimizes $\|x^*\|$.
However, if $P$ lies in a $(j-1)$-dimensional linear manifold where
$j \leq d$, then any set of $j+1$ or more points in $P$ will be affinely
dependent.
Therefore rank~$A^{(j,p)} < j$ for {\it all}
$p \in P \setminus \{s_1$, $\ldots$, $s_j\}$, and no solution $p^*$
can exist. Indeed, $DT(P)$ does not exist as discussed in Remark 2.3.
Furthermore, $DT(P)$ is not unique if there exist $d+2$ or more
points in $P$ that lie on the same circumball, since there may be more
than one $p^* \in P$ that minimize $\|x^*\|$. For the purpose of
interpolation, all of these solutions are equally suitable, and the
decision between any number of such candidate solutions can be made
arbitrarily. In particular, these cases are resolved by choosing
the candidate $p^*$ with the lowest index in $P$. Pseudocode for this
process follows.

{\parindent =0pt \parskip= 0pt
\smallskip
\leftskip 20pt
{\sl Algorithm 2}
\smallskip
$P$ contains $n$ $d$-dimensional data points;\par
$q \in CH(P)$ is the interpolation point;\par
$\{s_1$, $\ldots$, $s_{d+1}\}$ denote the vertices of the seed simplex. \par
\smallskip
{\bf begin}\par
$s_1 \leftarrow {\arg\min \atop {p\in P}} \|q - p\|$;\par
$s_2 \leftarrow {\arg\min \atop {p\in P, \atop p\neq s_1}} \|s_1 - p\|$;\par
{\bf for} $j = 2$, $\ldots$, $d$ {\bf do}\par
\leftskip 40pt
{\bf initialize} $r_{min} \leftarrow \infty$; $p^* \leftarrow null$;\par
{\bf for all} $p \in P \setminus \{s_1$, $\ldots$, $s_j\}$ {\bf do}\par
\leftskip 60pt
Compute $A^{(j,p)}$ and $b^{(j,p)}$;\par
{\bf if} rank~$A^{(j,p)} < j$ {\bf then} \par
\leftskip 80pt
Skip this point; \par
\leftskip 60pt
{\bf else if} rank~$A^{(j,p)} = j$ {\bf then} \par
\leftskip 80pt
Compute the minimum norm solution $x^*$ to (2);\par
{\bf if} $\|x^*\| < r_{min}$ {\bf then}\par
\leftskip 100pt
$r_{min} \leftarrow \|x^*\|$;\par
$p^* \leftarrow p$;\par
\leftskip 80pt
{\bf endif}\par
\leftskip 60pt
{\bf endif}\par
\leftskip 40pt
{\bf enddo}\par
$s_{j+1} \leftarrow p^*$;\par
\leftskip 20pt
{\bf enddo}\par
{\bf if} $p^* \neq null$ {\bf then}\par
\leftskip 40pt
{\bf return} $\{s_1$, $\ldots$, $s_{d+1}\}$;\par
\leftskip 20pt
{\bf else}\par
\leftskip 40pt
{\bf return} error (points in a lower-dimensional linear manifold);\par
\leftskip 20pt
{\bf endif}
\smallskip}

The dominant costs of the above algorithm are determining the rank of
$A^{(j,p)}$ and finding the minimum norm solution $x^*$ to (2). Both of
these computations can be done using a $LQ$ factorization of $A^{(j,p)}$
with row pivoting. Such a factorization has a computational
complexity of at most ${\cal O}\big(d^3\big)$ (in the case where $j=d$),
so the total complexity of Algorithm 2 (for growing a seed $d$-simplex)
is ${\cal O}\big(nd^4\big)$.

\medskip{\sl Remark 3.1.1.}\enspace
To determine the rank of $A^{(j,p)}$ using a $LQ$ factorization with row
pivoting, consider the final term on the diagonal of $L$, $L_{jj}$. 
By construction, rank~$A^{(j,p)}$ is at least $j-1$, and therefore rank~$L$
must also be at least $j-1$. Therefore, $L_{jj}$ is the exact magnitude of
the smallest perturbation that could cause rank-deficiency in $L$, and
equivalently, in $A^{(j,p)}$. To allow for floating point error,
$A^{(j,p)}$ is considered singular if $|L_{jj}| < \varepsilon$, where
$\varepsilon$ is a scale/machine dependent constant.
\medskip

\heading{3.2 \enspace The Visibility Walk}

After constructing the seed simplex, DELAUNAYSPARSE advances on the
simplex containing an interpolation point $q$ by following a visibility
walk. A facet $F$ of a simplex $S$ is said to be {\it visible} to $q$ if
there exists a point $\rho \in$~int~$S$ such that the line segment drawn
from $\rho$ to $q$ intersects $F$. A visibility walk is a sequence
of ``flips'' that always occur across a facet from which $q$ is visible.
Note that each flip in a visibility walk is generally not unique, as it
is possible for $q$ to be visible from multiple facets, and a flip across
{\it any} visible facet constitutes a valid step in a visibility walk.
Edelsbrunner [1989] showed that in a Delaunay triangulation, every
visibility walk is acyclic and therefore must converge for any
$q \in CH(P)$. The mechanics of performing each flip in the visibility
walk will be detailed in Section 3.3. In this section, the processes
of identifying each visible facet and terminating the visibility walk
will be explored.

For a simplex $S$ with vertices $s_1$, $\ldots$, $s_{d+1}$, define
the $d \times d$ matrix
$$
A^{(S)} = \left[ \matrix{
  \left(s_2 - s_1\right) \enspace
  \cdots 
  \enspace \left(s_{d+1} - s_1\right) } \right].
$$
Let $x_i$ denote the $i$th entry in the $d$-vector $x$, given by
the solution to the linear system
$$
A^{(S)}x = q - s_1. \eqno{(3)}.
$$
Then the vector of affine weights $w = [w_1$, $\ldots$, $w_{d+1}]^T$
for generating $q$ as a combination of $s_1$, $\ldots$, $s_{d+1}$ is given
by
$$
w = \left[ \matrix{
  \left(1 - \sum_{i=1}^d x_i\right) \cr
  x_1 \cr
  \vdots \cr
  x_d } \right].
$$

If $w_i \geq 0$ for $i=1$, $\ldots$, $d+1$, then $q\in S$ and
$w$ contains the interpolation weights in (1). If any $w_i < 0$, then
dropping the corresponding vertex $s_i$ leaves the vertices of a
facet of $S$ from which $q$ is visible. If $S$ is a valid Delaunay
simplex, $A^{(S)}$ is nonsingular and (3) can be solved via $LU$
factorization.

\medskip{\sl Remark 3.2.1.}\enspace
To account for floating point errors, the condition that $w_i \geq 0$
for $i=1$, $\ldots$, $d+1$ should be replaced with $w_i \geq -\varepsilon$,
where $\varepsilon$ is a scale/machine dependent constant, similarly as in
Remark 3.1.1.
\medskip

The cost of a single $LU$ factorization for solving (3) is insignificant
compared to the cost of performing a flip, as described in the next
section, which requires up to $n$ $LU$ factorizations. However, the total
length of the visibility walk (in number of flips) will be important in
determining the computational complexity of the DELAUNAYSPARSE algorithm.
This length $k$ can only be analytically bounded by the total size of
$DT(P)$. However, Bowyer [1981] claimed without proof that when
starting a visibility walk from the center of a Delaunay triangulation,
$k$ is ${\cal O}\big(n^{1/d}\big)$.
M{\"u}cke et al.\ [1999] proved that in up to three-dimensions, a stronger
claim can be made for some variations of the standard visibility walk.
Chang et al.\ [2018b] showed empirically that for pseudo-randomly
generated data points, when starting from a simplex grown off the nearest
data point to $q$, $k$ tends to grow polynomially with dimension and has
no dependence on $n$ for large values of $n$ and $d$.

\heading{3.3 \enspace Flipping Across a Facet}

Let $F = CH(\{s_1$, $\ldots$, $s_d\})$ be a facet of a previously
constructed Delaunay simplex from which the interpolation point $q$
is visible. Let $H(F)$ denote the hyperplane containing $F$, and let
$H_q(F)$ denote the open halfspace (with respect to $H(F)$) that
contains $q$. The goal of this section is to ``flip toward'' $q$, by
constructing a new Delaunay $d$-simplex with vertices
$s_1$, $\ldots$, $s_{d+1}$, where $s_{d+1} \in P \cap H_q(F)$.

Recall from Remark 2.1 that a $d$-simplex $S$ is Delaunay if and only
if $B_S \cap P = \emptyset$. Since at least one Delaunay simplex
(of which $F$ is a facet) has already been constructed, $DT(P)$ exists.
Therefore, if $F$ is {\bf not} a facet of $CH(P)$, there must
be at least one point $p^*\in P \cap H_q(F)$ such that the simplex
$S^*$ with vertices $\{s_1$, $\ldots$, $s_d$, $p^*\}$ is Delaunay,
satisfying $B_{S^*} \cap P = \emptyset$. If no such $p^*$ exists,
then it can be inferred that $q \not\in CH(P)$.
So, to perform a ``flip'' to a new Delaunay simplex
closer to $q$, it suffices to check inside the circumball of the
simplex with vertices $s_1$, $\ldots$, $s_d$, $p$, for each
$p \in P \cap H_q(F)$. By exploiting the property described in
Remark 2.2, this can be done in a single pass over $P$.

For a facet $F$ with vertices $s_1$, $\ldots$, $s_d$, let the
$(d-1) \times d$ matrix
$$
A^{(F)} = \left[ \matrix{
  \left(s_2 - s_1\right)^T \cr
  \vdots \cr
  \left(s_d - s_1\right)^T } \right].
$$
To obtain a normal to $H(F)$, it suffices to take any nontrivial
vector in the nullspace of $A^{(F)}$.
Since $F$ is a Delaunay facet, rank~$A^{(F)} = d-1$. Therefore, using a
$LQ$ factorization with row pivoting, ${\cal P}A^{(F)} = LQ$, yields
the final row $v^T = Q_{d\cdot}$ of $Q$ as a unit normal to $H(F)$.

Given the normal vector $v$ for $H(F)$, consider the function
$$
\sigma_F(p) = \hbox{sgn}
\bigl((p - s_1)^T v\bigr). \eqno{(4)}
$$
For each $p\in P$, $p\in H_q(F)$ if and only if $\sigma_F(p) = \sigma_F(q)$.

\medskip{\sl Remark 3.3.1.}\enspace
To account for floating point errors, the additional condition that 
$\bigl|(p - s_1)^T v\bigr| > \varepsilon$ should be imposed, where
$\varepsilon$ is a scale/machine dependent constant, similarly as in
Remarks 3.1.1 and 3.2.1.
\medskip

Consider now those $p \in P \setminus \{s_1$, $\ldots$, $s_d\}$ such that
$\sigma_F(p) = \sigma_F(q)$, i.e., $p \in H_q(F)$.
Similarly as in Section 3.1, the center of the circumball about $F$ and
$p$ is given by $c = x^* + s_1$, and the radius of the circumball is given
by $r = \|x^*\|$, where $x^*$ is a solution to the system
$$
A^{(d,p)}x = b^{(d,p)}. \eqno{(5)}
$$
Since $F$ is a valid Delaunay facet, the first $d-1$ columns of $A^{(d,p)}$
must be linearly independent. Furthermore, if the condition described in
Remark 3.1.1 has been satisfied, then it is safe to assume that $A^{(d,p)}$
is full-rank, and (5) has a unique solution. Then $p^*$ is any point in
$P \cap H_q(F)$ that satisfies $B_{\|x^*\|}(c) \cap P = \emptyset$, where
$B_{\|x^*\|}(c)$ denotes the open ball centered at $c$ with radius
$\|x^*\|$. Pseudocode for this entire process follows.

\medskip{\sl Remark 3.3.2.}\enspace
If $P$ is in general position, then $p^*$ is unique. If there exist
$d+2$ or more cospherical points in $P$, then $p^*$ may not be unique.
However, any $p^*\in P \cap H_q(F)$ that satisfies
$B_{\|x^*\|}(c) \cap P = \emptyset$ can be chosen in union with
$\{s_1$, $\ldots$, $s_d\}$ to form the vertex set for a valid simplex
in {\it some} Delaunay triangulation. Such situations can be resolved
by choosing the $p^*$ with the greatest index in $P$.
\medskip

{\parindent =0pt \parskip= 0pt
\smallskip
\leftskip 20pt
{\sl Algorithm 3}
\smallskip
$P$ contains $n$ $d$-dimensional data points;\par
$q \in CH(P)$ is the interpolation point;\par
$\{s_1$, $\ldots$, $s_{d+1}\}$ denote the vertices of the new simplex;\par
$F$ is the current Delaunay facet;\par
\smallskip
{\bf begin}
$\{s_1$, $\ldots$, $s_d\}$ are given by the vertices of $F$;\par
Compute the $LQ$ factorization ${\cal P}A^{(F)}=LQ$
and set $v^T \leftarrow Q_{d\cdot}$; \par
Compute $\sigma_F(q)$ using (4); \par
{\bf initialize} $r_{min} \leftarrow \infty$;
$c_{min} \leftarrow {\vec 0}$;
$p^* \leftarrow null$;\par
{\bf for all} $p \in P \setminus \{s_1$, $\ldots$, $s_d\}$ {\bf do}\par
\leftskip 40pt
{\bf if} $\sigma_F(p) \neq \sigma_F(q)$ {\bf or} 
$\bigl|(p - s_1)^T v\bigr| < \varepsilon$ {\bf then}\par
\leftskip 60pt
Skip this point;\par
\leftskip 40pt
{\bf else if} $\|p - c_{min}\| < r_{min}$ {\bf then}\par
\leftskip 60pt
Update $A^{(d,p)}$ and $b^{(d,p)}$ and compute $x^*$, the solution to (5);\par
$r_{min} \leftarrow \|x^*\|$;\par
$c_{min} \leftarrow x^* + s_1$;\par
$p^* \leftarrow p$;\par
\leftskip 40pt
{\bf endif}\par
\leftskip 20pt
{\bf enddo}\par
{\bf if} $p^* = null$ {\bf then} \par
\leftskip 40pt
{\bf return} error, $q \not\in CH(P)$;\par
\leftskip 20pt
{\bf else}\par
\leftskip 40pt
{\bf return} $s_{d+1} \leftarrow p^*$;\par
\leftskip 20pt
{\bf endif}
\smallskip}

The dominant cost for Algorithm 3 is repeatedly solving (5). Since
$A^{(d,p)}$ is always full-rank, each instance of (5) can be solved
using an $LU$ factorization, with computational complexity
${\cal O}\big(d^3\big)$. So the worst-case complexity of Algorithm 3
is ${\cal O}\big(nd^3\big)$.
Note that Algorithm 3 is called once in each iteration of the simplex
walk. Therefore, from Section 3.1 and 3.2, the overall computational
complexity for locating a single interpolation point is
${\cal O}\big(nd^4 + knd^3\big)$, where $k$ is the length of the
visibility walk.
Since $k$ is typically much greater than $d$, this can be simplified
to ${\cal O}\big(knd^3\big)$.

\heading{4. SERIAL IMPLEMENTATION}

The serial subroutine {\tt DELAUNAYSPARSES} is implemented in ISO Fortran
2003. For efficient numerically stable linear algebra, {\tt DELAUNAYSPARSES}
uses LAPACK [Anderson et al.\ 1999].

\heading{4.1 \enspace Handling Multiple Interpolation Points}

The serial subroutine {\tt DELAUNAYSPARSES} performs interpolation at $m$
points $Q = \{q_1$, $\ldots$, $q_m\}$ using Algorithms 1--3, as described
in Sections 2 and 3. By default, {\tt DELAUNAYSPARSES} will perform these
interpolations sequentially with no modification to Algorithms 1--3.
However, an optional argument can be set to ``daisy chain'' the visibility
walks, i.e., for the $i$th interpolation point $q_i$ where $i > 1$, the
last constructed Delaunay simplex (typically the simplex containing 
$q_{i-1}$) is used as the first simplex for walking to $q_i$, replacing
Algorithm 2. In general, this behavior can greatly increase the length of
each visibility walk and is not recommended. However, if the interpolation
points $Q$ are tightly clustered in a relatively small region of
$DT(P)$ or if the size of $DT(P)$ is relatively small, this behavior can
slightly improve performance by avoiding the expense of Algorithm 2.

Additionally, note that after computing the $LU$ factorization of $A^{(S)}$
for solving (3), it requires relatively little additional computation to
check whether $S$ contains any future interpolation points. Thus if $S$
is a simplex with vertices $s_1$, $\ldots$, $s_{d+1}$ that has been
constructed during the visibility walk to an interpolation point $q_i$,
{\tt DELAUNAYSPARSES} will check whether $q_j \in S$, where
$i < j \leq n$ and $q_j$ has not already been found in some simplex
of $DT(P)$, by solving
$$
A^{(S)}x = q_j - s_1, \eqno{(6)}
$$
and using the same criterion described in Section 3.2.
If any $q_j \in S$, then $s_1$, $\ldots$, $s_{d+1}$ and the corresponding
interpolation weights for (1) are saved, and $q_j$ is marked as found and
will not be considered again. Because of its cost effectiveness, this
behavior is always active during an execution of {\tt DELAUNAYSPARSES}.
However, it is most effective for tightly clustered interpolation points.

\heading{4.2 \enspace Extrapolation}

Often, it is reasonable to make predictions for extrapolation points
that are {\it slightly} outside $CH(P)$. In these cases, the most
reasonable solution is to project each extrapolation point onto
$CH(P)$ and interpolate the projection, provided the residual is small.
Let $z$ be an extrapolation point, and let $W$ be a $d\times n$ matrix
whose columns are points in $P$. Then the projection ${\hat z}$ of $z$
onto $CH(P)$ is given by ${\hat z} = Wx^*$, where $x^*$ is the solution
to the linearly constrained least squares problem
$$
\min_{x\in\R^n} \|Wx - z\| \quad\hbox{subject to}\quad
x \ge 0 \quad\hbox{and}\quad \sum_{i=1}^n x_i = 1. \eqno{(7)}
$$
Hanson et al.\ [1982] provide an efficient solution to (7)
based on a slack variable formulation. The most recent version
of their subroutine {\tt DWNNLS} is available in the SLATEC
software package, and included with the DELAUNAYSPARSE files.

\medskip{\sl Remark 4.2.1.}\enspace
Note that the visibility walk described in Section 3.2 is only guaranteed
to converge for $q\in CH(P)$. In particular, if a projection $\hat z$
is left within floating point error of $CH(P)$, and the matrix $A^{(S)}$
for a nearby Delaunay simplex has smallest singular value
${\cal O}(\varepsilon)$, then it is
possible for a visibility walk to fail by repeatedly calling for a flip
that would lead outside of the convex hull. This is an extremely rare
situation. However, in these situations, {\tt DELAUNAYSPARSE} will first
try to flip in other potential directions (i.e., by dropping different
negatively weighted vertices). Then, if no ``good'' direction can be
found, the correct interpolation weights must ultimately be computed by
a second {\tt DWWNLS} projection onto the current simplex.

\medskip

Given the above solution, the residual is given by $r = \|z - {\hat z}\|$.
When $r$ is small with respect to the scale of the data, it is reasonable
to perform extrapolation at $z$ by interpolating at $\hat z$.  However,
when $r$ is large it is impossible to make any reasonable prediction for
$f(z)$. By default, when {\tt DELAUNAYSPARSES} encounters an extrapolation
point $z$, it computes the projection ${\hat z}$ and residual $r$ using
{\tt DWNNLS}. If $r$ is smaller than some percentage of the diameter of
$P$, {\tt DELAUNAYSPARSES} resumes interpolation using $q={\hat z}$. If
$r$ is greater than that percentage of the diameter, the extrapolation
point $z$ is skipped and an appropriate error is returned.

By default, the threshold for extrapolation is $10\%$ of the diameter of
$P$, but this percentage can be adjusted using an optional input argument.
Furthermore, setting this optional value to $0\%$ of the diameter of $P$
will short-circuit the extrapolation process, preventing ${\hat z}$
and $r$ from ever being computed and preventing any {\tt DWNNLS}
work arrays from being allocated. Note that the time and space demands
of {\tt DWNNLS} can be significantly greater than those of
{\tt DELAUNAYSPARSES}. So, for large problems or in cases where
computational resources are limited, it is often appropriate to turn off
extrapolation by setting the extrapolation threshold to $0\%$ of the
diameter of $P$.

\medskip{\sl Remark 4.2.2.}\enspace
For similar reasons as in Remark 4.2.1, it is possible that for poorly
spaced data points $P$, {\tt DELAUNAYSPARSES} may incorrectly call for
a projection of an interpolation point that is within floating point
error of the boundary of $CH(P)$. After unnecessarily performing the
projection and finding that $r=0$, such situations are easily
detected retrospectively. However, if the extrapolation threshold is
set to 0\% of the data diameter, the projection will short circuit
and an interpolation point that is within $\varepsilon$ of the convex
hull $CH(P)$ could be incorrectly skipped. The conditions that could
lead to such an error are pathological, but might still occur.
\medskip

\heading{4.3 \enspace Data Scaling and Sensitivity Analysis}

Recall from Remarks 3.1.1, 3.2.1, 3.3.1, and 4.2.1 that a small
scale/machine dependent constant $\varepsilon > 0$ is used to account
for floating point error. Affine operations do not affect the Delaunay
triangulation or interpolation results, so to account for scaling, 
{\tt DELAUNAYSPARSES} rescales and shifts the data points $P$ and
the interpolation points $Q$ on input. First, the points in $P$ are
shifted so that their barycenter is at the origin, then they are rescaled
so that they are contained in the unit hypersphere. This ensures
that $\varepsilon$ can be chosen without accounting for data scale.
The interpolation points $Q$ must then be shifted and scaled by the same
amounts to maintain relative positions.

After scaling, the default value $\varepsilon$ can be chosen based
only on machine precision. By default, $\varepsilon = \sqrt{\mu}$
where $\mu$ denotes the unit round-off. This is the minimum appropriate
value of $\varepsilon$ for most applications, and an optional argument
can be used to increase $\varepsilon$ where appropriate. For the rescaled
data, {\tt DELAUNAYSPARSES} is backward stable for perturbations less
than $\varepsilon$.

\heading{4.4 \enspace Memory Usage}

{\tt DELAUNAYSPARSES} uses assumed-shape arrays where appropriate.
To ensure expected behavior, the dimensions of each dummy array are
checked against user-specified values of $d$, $n$, and $m$ on input.
Due to the size of $DT(P)$ in high dimensions, the space complexity of
any Delaunay triangulation algorithm is equally as important as its
time complexity. The computational operations described in Section 3
do not require any work arrays larger than ${\cal O}\big(d^2\big)$,
making DELAUNAYSPARSE a space efficient algorithm.

However, to take full advantage of LAPACK code optimizations, one
larger work array is required. Therefore, {\tt DELAUNAYSPARSES}
uses one allocatable work array, whose size is determined at runtime
based on LAPACK querries. Other allocatable work arrays of size
${\cal O}\big(nd\big)$ are required by {\tt DWNNLS} for extrapolation,
but are only allocated if an extrapolation is performed.

\heading{4.5 \enspace The Cost of Robustness and Correctness}

{\tt DELAUNAYSPARSES} is designed to be robust for a wide variety of
use cases and usage errors. In particular, {\tt DELAUNAYSPARSES} uses
the diameter of $P$ to judge extrapolation residuals, as discussed in
Section 4.2. Also, the minimum pairwise distance between points in
$P$ is used to catch bad inputs, since after rescaling, any two points
that are closer than $\varepsilon$ will be indistinguishable from
the perspective of {\tt DELAUNAYSPARSES} and could cause hard to find
bugs. The computation of the diameter and minimum pairwise distance
is performed while the points are being rescaled, as discussed in
Section 4.3. The computational complexity of these distance computations
is ${\cal O}\big(n^2d\big)$. Recall that the computational complexity of
the DELAUNAYSPARSE algorithm is ${\cal O}\big(knd^3)$, where $k$ is
independent of $n$ for uniformly spaced $P$. Therefore, in situations
where $n^2d$ is significantly larger than $knd^3$, the
complexity of {\tt DELAUNAYSPARSES} can be dominated by nonessential
distance computations, used only for robustness and extrapolation
checks.

\heading{5. PARALLEL IMPLEMENTATION}

The parallel subroutine {\tt DELAUNAYSPARSEP} is based on the serial
subroutine {\tt DELAUNAYSPARSES} and shares the implementation details
discussed in Sections 4.1--5. {\tt DELAUNAYSPARSEP} uses OpenMP 4.5 
[OpenMP ARB, 2015]
to implement a shared memory paradigm. It is also possible to achieve
distributed parallelism by breaking up the interpolation points $Q$ into
separate batches $Q = Q_1 \cup \ldots \cup Q_B$, then distributing these
batches $Q_i$ across available nodes (along with copies of the data points 
$P$). Each batch can then be evaluated independently on its corresponding
node.

\medskip{\sl Remark 5.1.}\enspace
Recall from Section 4.1 that code optimizations for handling multiple
interpolation points are most effective when the points are clustered
in $DT(P)$. Therefore, optimal distributed memory performance is achieved
when each $Q_i$ represents a cluster of interpolation points from $Q$.
Note that clustering is an open problem, and the above described
distributed memory parallelism can be implemented trivially using separate
calls to {\tt DELAUNAYSPARSES} or {\tt DELAUNAYSPARSEP}. Therefore, a
distributed implementation with clustering of $Q$ is not provided, and
left entirely to the user.
\medskip

For the OpenMP shared memory implementation, two levels of parallelism
are targeted. The first level of parallelism is the loop over all $m$
interpolation points $Q$. The second level of parallelism applies to
the various loops over all $n$ data points $P$ and the loop over all
unresolved interpolation points, as computed by (6) and described in
Section 4.1.
There is also a loop over the $n$ data points for computing the scale
factor (as discussed in Section 4.3) and a pair of nested loops over
the $n$ data points for computing the diameter and minimum pairwise
distance of $P$ (as discussed in Section 4.5).
These loops can be parallelized independently of either level of
parallelism using a static scheduler.

If $m$ is small with respect to the number of available processors,
level one parallelism will not saturate the available computational
resources. In the extreme case where $m=1$, level one parallelism is
not available. However, when available, level one parallelism is
significantly more efficient than level two parallelism. Therefore, the
default behavior for {\tt DELAUNAYSPARSEP} is to exploit level one
parallelism whenever it is available (if $m > 1$), and to exploit level
two parallelism otherwise (if $m = 1$). If this is not the desired
behavior, the type of parallelism can be set manually via an optional
argument, and for advanced
users, both levels can be activated at the same time resulting in
nested parallelism.

\medskip{\sl Remark 5.2.}\enspace
In order for OpenMP to apply nested parallelism, the environment variable
{\tt OMP\_NESTED} must be set to {\tt TRUE}. Furthermore, note that if
a team of $t_1$ threads is deployed at the first level and a team of $t_2$
threads is deployed at the second level, then a total of $t_1 \cdot t_2$
threads could be active at any time. $t_1$ and $t_2$ can be set by
assigning the environment variable {\tt OMP\_NUM\_THREADS=}$t_1$,$t_2$.
\medskip

\heading{5.1 \enspace Level 1 Parallelism}

The first level of parallelism is the loop over $m$ interpolation points
in $Q$: {\bf for all} $q \in Q$ {\bf do}, from Algorithm 1. Since
the variation in the length $k$ of each visibility walk could be large,
{\tt DELAUNAYSPARSEP} uses a dynamic scheduler with a chunk size of one
to parallelize this $Q$ loop. The only dependencies between iterations of
the $Q$ loop occur when implementing the code optimizations described in
Section 4.1.

First, when ``checking ahead'' for future interpolation points
$q_j\in Q$, there is a possible race condition since another thread
could already be performing a visibility walk toward $q_j$. Since these
dependencies are minimal, an OpenMP {\tt CRITICAL} lock is used with
minimal modification to the serial code to ensure safe sequentially
consistent memory accesses. Once any thread of {\tt DELAUNAYSPARSEP} has
begun constructing the first simplex in the walk toward $q_j$, no other
threads will continue to test $q_j$ when ``checking ahead.''
Additionally, if daisy chaining is activated, each thread in the team
must maintain a private copy of the seed simplex. Therefore, only
previously constructed simplices of the active thread are considered
for seeding future visibility walks.

Other than the minor issues discussed above, level one
parallelism is dependency free and dynamically load balanced.
Therefore, under ideal conditions, {\tt DELAUNAYSPARSEP} is capable of
weak scaling with respect to the problem dimension $m$, with negligible
overhead.

\heading{5.2 \enspace Level 2 Parallelism}

The second level of parallelism applies primarily to the various loops
over $n$ data points in $P$: {\bf for all} $p\in P \setminus \{s_1$,
$\ldots$, $s_j(s_d)\}$, as appear in Algorithms 2 and 3. Note that
these $P$ loops are not totally free of dependencies. Therefore, to
parallelize the $P$ loops, private copies of certain variables (such as
$r_{min}$ and $p^*$) must be maintained. Then, after completing these loops
in parallel and producing private solutions, a reduction can be done to
determine the global solutions. Note that this will result in some
redundant computations. There is relatively little room for performance
variation between iterations of the $P$ loops, so {\tt DELAUNAYSPARSEP}
parallelizes them with a static scheduler and a fixed chunk size of
$\lceil n/t_2 \rceil$, where $t_2$ denotes the number of threads in
each level two team.

The one exception to the above methodology is the loop over all future
interpolation points, as described in Section 4.1. This loop is dependency
free and can be parallelized using a static scheduler with a fixed chunk
size of $\lceil(m-i)/t_2\rceil$, where $i$ is the index of the current
interpolation
point. However, if level one parallelism is active, parallelizing this
loop results in significant conflict. Therefore, in the case of nested
(level one and two) parallelism, this loop executes serially within
each level one thread.

\medskip{\sl Remark 5.2.1}\enspace
There is no true dependency between iterations of the above loop over
remaining interpolation points. However, the OpenMP {\tt CRITICAL}
directive used in Section 5.1 locks code segments, as opposed to
variable addresses and cannot distinguish between loop
iterations, inducing a ``false'' conflict. As mentioned above, when
level one parallelism is available, it is recommended that all
available threads be devoted to level one parallelism. Therefore, in
the recommended use case, this loop would not offer significant
parallelism, and serializing it is no significant loss.
\medskip

Due to interloop dependencies, exploiting level two parallelism can
significantly increase the total number of computations performed by
DELAUNAYSPARSE. Furthermore, there are significant regions of serial
code separating each level two parallel block. So, 
the parallel efficiency of {\tt DELAUNAYSPARSEP} with level two
parallelism can be poor.

\heading{6. ORGANIZATION AND USAGE INFORMATION}

The physical organization of the DELAUNAYSPARSE package is described
in its included README file. Most notably, DELAUNAYSPARSE is distributed
with a copy of the REAL\_PRECISION module (from HOMPACK90, ACM TOMS
Algorithm 777), for approximately 64-bit arithmetic on all known
machines. For completeness, all required LAPACK and BLAS subroutines
are included, along with {\tt DWNNLS} and all its dependencies from
the SLATEC library. The included copies of the SLATEC subroutines have
been updated in accordance with the Fortran 2003 standard. Additionally,
legacy implementations for two of {\tt DWNNLS}'s BLAS dependencies 
({\tt DROTM} and {\tt DROTMG}) have
been included under different names. Finally, sample main programs
for the serial and parallel versions illustrating the use of the
optional arguments have been included.
Sample data sets for these main programs are real data sets (with points
not in general position) from the VarSys project on high performance
computing system performance variability [Cameron et al.\ 2019].

The master module DELSPARSE\_MOD includes the REAL\_PRECISION module
and interface blocks for both {\tt DELAUNAYSPARSES} and
{\tt DELAUNAYSPARSEP}, as well as an interface block for the updated
subroutine {\tt DWNNLS}, which may be of separate interest. Note that by
default, {\tt DELAUNAYSPARSES} and {\tt DELAUNAYSPARSEP} do not actually
compute the Delaunay interpolant, but return the containing simplex and
weights for computing (1). This behavior was chosen to accommodate a wide
variety of use cases, including those where the function values $f(x)$
cannot be expressed as real-valued vectors (e.g., when $f(x)$ is a function
$g(\omega;x)$ parameterized by $x$). When the
values $f(x)$ {\it can} be expressed as real-valued vectors, an optional
input argument can be supplied with the response values, and then an
optional output argument must appear to collect interpolation results.
Additionally, recall that $P$ and $Q$ are shifted and rescaled on input.
So, if the original $P$ and $Q$ are needed, copies should
be made prior to execution.

\heading{7. PERFORMANCE}

The approximation accuracy of the Delaunay interpolant has already been
explored in numerous other publications (see Chang et al.\ [2018a],
Lux et al.\ [2018], Omohundro et al.\ [1990], and Schaap et al.\ [2000]).
So, this section will focus on the runtime of {\tt DELAUNAYSPARSES} and
{\tt DELAUNAYSPARSEP}. For reference, first consider Table I, which presents
performance data (as reported by Boissonnat et al.\ [2009]) in up to
six dimensions for computing the complete Delaunay triangulation of uniform
randomly distributed data points in the unit cube using
Quickhull [Barber et al.\ 1996] and the graph based algorithm proposed
by Boissonnat et al.\ [2009]. Boissonnat et al.\ gathered this data
using a 2.6 GHz Intel processor with 6MB of level 2 cache and 4 GB of
DDR2 RAM. The purpose of including Table I is not for direct comparison,
as the problem of computing the complete Delaunay triangulation
is significantly harder than that of locating a single interpolation
point. Indeed, recall that standard Delaunay triangulation algorithms
are not capable of scaling past six or seven dimensions for sufficiently
large problems. Rather, this data is intended to clarify the issues addressed
by DELAUNAYSPARSE, and inform users on when DELAUNAYSPARSE is an appropriate
choice over algorithms that compute the complete Delaunay triangulation.

\topinsert\rmVIII
{\narrower\noindent Table I.
Time and space requirements (as reported in Boissonnat et al.\ [2009])
for computing the complete Delaunay
triangulation using Quickhull and the Delaunay Graph algorithm.
Entries containing the word ``swap'' indicate that the process
exceeded RAM limitations.}
\tabskip=0pt
$$\vbox{\settabs \+ \quad X dimensions, XX,000 points\quad &
\quad space \quad & \quad Delaunay Graph \quad &
\quad Delaunay Graph \quad & \cr
\+ \hfil Problem Sizes ($d$ \& $n$): \hfil & & \hfil Algorithms: \hfil & \cr
\vskip 1pt \hrule \vskip 3pt
\+ & & \hfil Quickhull \hfil & \hfil Delaunay Graph \hfil & \cr
\vskip 1pt \hrule \vskip 3pt
\+ \hfil 5 dimensions, 2,000 points\hfil & \hfil time \hfil
& \hfil 3.2 sec.\ \hfil & \hfil 58 sec.\ \hfil & \cr
\+ & \hfil space \hfil & \hfil 52 MB \hfil & \hfil 10.1 MB \hfil & \cr
\vskip 1pt \hrule \vskip 3pt
\+ \hfil 5 dimensions, 32,000 points\hfil & \hfil time \hfil
& \hfil 76 sec.\ \hfil & \hfil 1463.46 sec.\ \hfil & \cr
\+ & \hfil space \hfil & \hfil 973 MB \hfil & \hfil 106 MB \hfil & \cr
\vskip 1pt \hrule \vskip 3pt
\+ \hfil 6 dimensions, 32,000 points\hfil & \hfil time \hfil
& \hfil swap \hfil & \hfil 28,296 sec.\ \hfil & \cr
\+ & \hfil space \hfil & \hfil swap \hfil & \hfil 267 MB \hfil & \cr
\vskip 1pt \hrule
}$$
\endinsert

To test the performance of {\tt DELAUNAYSPARSES}, runtimes have been
gathered on AMD CPUs @2.3 GHz.
Table II presents runtimes for interpolating at a single interpolation
point (the center of the unit hypercube) using {\tt DELAUNAYSPARSES}, for
various problem sizes ($n$) and dimensions ($d$) with uniform randomly
distributed data in the unit hypercube. To account for performance
variance, each runtime represents an average over 20 independent
runs of {\tt DELAUNAYSPARSES}, each with a different data set of the
same size and dimension. Note that in the higher
dimensions, the data points ($P$) are extremely sparse, even for large
values of $n$. For such problems, it is typical to employ some
intelligent experimental design. Therefore, in the higher dimensions,
the uniform randomly spaced data used for testing becomes increasingly
unrepresentative of real-world data. However, this data is sufficient
for discussing how the runtime of {\tt DELAUNAYSPARSES} scales with $n$
and $d$ and is comparable to the data used to generate Table I.
Since extrapolation presents additional computational complexities,
any data set that does not contain the interpolation
point ($q = [0.5$, $0.5$, $\ldots$, $0.5]^T$) in its convex hull is
discarded and regenerated (an unlikely occurence for the
problem sizes shown).
Note that the distance computations discussed in Section 4.5 cause
an overall computational complexity of ${\cal O}\big(n^2\big)$ in
the lower dimensions.
However, in higher dimensions, the cost of the DELAUNAYSPARSE
algorithm dominates, and the runtimes approach linear growth
with respect to the number of data points $n$, as predicted in 
Section 3.

\topinsert\rmVIII
{\narrower\noindent Table II. Serial runtimes (in seconds).
Values marked ``n/a'' represent problem dimensions for which
$DT(P)$ does not exist ($n < d+1$), and values marked ``slow''
were aborted due to excessive runtimes (greater than 3600 seconds).}
\tabskip=0pt
$$\vbox{\settabs \+ Problem size ($n$):\qquad\qquad& \quad 00.000 \quad
&\quad 00.000 \quad&\quad 00.000 \quad&\quad 00.000 \quad
&\quad 00.000 \quad&\cr
\+ & Problem dimension ($d$) \cr
\vskip 1pt
\+ Problem size ($n$) & \hfil 2 \hfil & \hfil 8 \hfil
& \hfil 32 \hfil & \hfil 64 \hfil & \hfil 128 \hfil &\cr
\vskip 3pt \hrule \vskip 5pt
\+ \hfill 100 \qquad\qquad & \hfil 0.001 \hfil & \hfil 0.007 \hfil
& \hfil 0.307 \hfil & \hfil 0.800 \hfil & \hfil n/a \hfil & \cr
\vskip 2pt
\+ \hfill 500 \qquad\qquad & \hfil 0.022 \hfil & \hfil 0.063 \hfil
& \hfil 1.822 \hfil & \hfil 25.563 \hfil & \hfil 316.566 \hfil & \cr
\vskip 2pt
\+ \hfill 2000 \qquad\qquad & \hfil 0.337 \hfil & \hfil 0.672 \hfil
& \hfil 8.326 \hfil & \hfil 109.754 \hfil & \hfil 1443.694 \hfil & \cr
\vskip 2pt
\+ \hfill 8000 \qquad\qquad & \hfil 5.219 \hfil & \hfil 9.448 \hfil 
& \hfil 50.067 \hfil & \hfil 481.307 \hfil & \hfil slow \hfil & \cr
\vskip 2pt
\+ \hfill 16,000 \qquad\qquad & \hfil 20.930 \hfil & \hfil 37.078 \hfil
& \hfil 157.123 \hfil & \hfil 1080.377 \hfil & \hfil slow \hfil & \cr
\vskip 2pt
\+ \hfill 32,000 \qquad\qquad & \hfil 83.394 \hfil & \hfil 148.078 \hfil
& \hfil 504.267 \hfil & \hfil 2443.431 \hfil & \hfil slow \hfil & \cr
\vskip 3pt \hrule
}$$
\endinsert

To compare the performance of {\tt DELAUNAYSPARSEP} at all
levels of parallelism against the performance of
{\tt DELAUNAYSPARSES}, runtimes have been gathered over a cluster of
eight NUMA nodes with four AMD cores per node @2.3 GHz (each core
identical to when timing {\tt DELUNAYSPARSES}).
Figures 1--3 plot the average elapsed wallclock time (in seconds) for
20 independent runs of {\tt DELAUNAYSPARSEP} against the number of
cores used by OpenMP. The data for these experiments was generated
using a randomized Latin hypercube design, as described in Amos et
al.\ [2014]. Then the interpolation points were generated from random
convex combinations of $d+1$ randomly selected points from the design.

Figure 1 presents runtimes for interpolating at 1024 points in a
10-dimensional design with 1000 data points, reflecting workloads
where $m$ is large.
Figure 2 presents runtimes for interpolating at 64 points in a
10-dimensional design with 10,000 data points, reflecting workloads
where $n$ is large.
Figure 3 presents runtimes for interpolating at 64 points in a
20-dimensional design with 200 data points, reflecting workloads
where $d$ is large.
Note that Figures 1--3 are presented at $\log_2$ scale.
For nested parallel runs with four, eight, 32 total active cores,
the number of level one (two) threads is two, four, eight
(two, four, four), respectively.

\topinsert
\centerline{\epsfxsize=4.5truein \epsffile{vt_logo.eps}}
{\narrower\noindent\rmVIII Fig.\ 1.
Average runtime in seconds plotted against number of active cores
for a 10-dimensional problem, with $n=1000$ and $m=1024$.
\par}
\endinsert

\midinsert
\centerline{\epsfxsize=4.5truein \epsffile{vt_logo.eps}}
{\narrower\noindent\rmVIII Fig.\ 2.
Average runtime in seconds plotted against number of active cores
for a 10-dimensional problem, with $n=10,000$ and $m=64$.
\par}
\endinsert

\midinsert
\centerline{\epsfxsize=4.5truein \epsffile{vt_logo.eps}}
{\narrower\noindent\rmVIII Fig.\ 3.
Average runtime in seconds plotted against number of active cores
for a 20-dimensional problem, with $n=200$ and $m=64$.
\par}
\endinsert

In all three figures, level one parallelism achieves the fastest average
runtimes, as expected. Furthermore, note that overapplying level two
parallelism can significantly increase overall run time, due to the
significant number of added computations as well as communication overhead.
However, one can see that level two parallelism is most applicable for
extremely large values of $n$. Furthermore, for situations where the
number of interpolation points is sufficient to achieve some level one
parallelism but insufficient to saturate the system,
nested parallelism is able to compete with level one parallelism.
It is somewhat surprising to observe that despite having the longest
overall runtime, the problem size used for Figure 1 ($d=10$, $n=1000$, 
$m=1024$) offered the least parallelism for all modes {\tt DELAUNAYSPARSEP}.
In particular, it may seem surprising that level one parallelism did not
perform better at this problem size, given that level one parallelism
is maximized for large values of $m$.

In fact, when the cost of performing each individual flip is relatively
small and the thread count per contention group is relatively high, level
one threads can be blocked to the point of serialization by {\tt CRITICAL}
locks during the loop to ``check ahead,'' as described in Section 5.1. This
presents a trade-off since checking ahead does not result in significant
conflict for relatively large values of $n$ and $d$ and could offer
significant performance benefits when $Q$ is clustered. In the cases
where the cost of each flip is small, it is recommended that users
partition the interpolation points into
$Q = Q_1 \cup \ldots \cup Q_B$. Then each $Q_i$ can be handled by a
separate call to {\tt DELAUNAYSPARSEP} with level one parallelism
and an appropriate thread count.
For example, in the case of Figure 1, for optimal performance over
32 active cores, {\tt DELAUNAYSPARSEP} could be called twice, with 16
threads and 512 interpolation points per call.
In general, the optimal choices for $B$, the thread count, and the
physical partition are highly problem dependent and beyond the scope
of this work.

\heading{\hlvVIII BIBLIOGRAPHY} {\let\caps=\capsVIII \let\sl=\slVIII
\def\ys{\hskip 1em minus .5em}\let\rm=\rmeight \rmVIII
\newcount\refnum \refnum=0 \parskip=2pt plus 2pt minus 2pt
\frenchspacing

%\refj is for regular journal articles
\def\refj#1#2#3#4#5{\noindent\hangindent=10pt\hangafter=1
%\global\advance\refnum by 1[\number\refnum]
{\caps #1}\ys #2.\ys {\rm #3}. {\sl #4}, #5.\par}

%\refb is for books
\def\refb#1#2#3#4{\noindent\hangindent=10pt\hangafter=1
%\global\advance\refnum by 1[\number\refnum]
{\caps #1}\ys #2.\ys{\sl #3}. {\rm #4}.\par}

%\refp is for private communication
\def\refp#1#2#3{\noindent\hangindent=10pt\hangafter=1
%\global\advance\refnum by 1[\number\refnum]
{\caps #1}\ys #2.\ys{\rm #3}.\par}

%\reft is for tech. reports, theses, and conferences
\def\reft#1#2#3#4{\noindent\hangindent=10pt\hangafter=1
%\global\advance\refnum by 1[\number\refnum]
{\caps #1}\ys #2.\ys{\rm #3}. {\rm #4}.\par}

%%%%% top of refs
\reft{Amos, B. D., Easterling, D. R., Watson, L. T., Thacker, W. I.,
Castle, B. S., Trosset, M. W.}{2014}{Algorithm XXX: QNSTOP: Quasi-Newton
algorithm for stochastic optimization}{Technical Report 2014-07,
Virginia Polytechnic Institute and State University, Blacksburg, VA}

\refb{Anderson, E., Bai, Z., Bischof, C., Blackford, S., Demmel, J.,
Dongarra, J., Du Croz, J., Greenbaum, A., Hammarling, S., McKenney, A.,
and Sorensen, D.}{1999}{LAPACK Users' Guide Third Edition}
{SIAM, Philidelphia, PA}

\refb{Aurenhammer, F., Klein, R., and Lee, D. T.}{2013}
{Voronoi diagrams and Delaunay triangulations}
{World Scientific Publishing Co., Hackensack, NJ}

\refj{Barber, C. B., Dobkin, D. P., and Huhdanpaa, H.}{1996}
{The Quickhull algorithm for convex hulls}{ACM Trans. Math.
Softw. 22}{469--483}

\reft{Boissonnat, J.-D., Devillers, O., and Hornus, S.}{2009}
{Incremental construction of the Delaunay triangulation and the
Delaunay graph in medium dimension}{In {\sl Proc. Twenty-fifth Annual Symp.
on Computational Geometry}, ACM, New York, NY, 208--216}

\refj{Bowyer, A.}{1981}{Computing Dirichlet tessellations}
{The Computer Journal 24}{162--166}

\refj{Cameron, K. W., Anwar, A., Cheng, Y., Xu, L., Li, B., Ananth, U.,
Bernard, J., Jearls, C., Lux, T., Hong, Y., Watson, L. T.,
and Butt, A. R.}{2019}{MOANA: modeling and analyzing I/O variability
in parallel system experimental design}
{IEEE Trans. Parallel Distrib. Systems}{to appear}

\reft{Chang, T. H., Watson, L. T., Lux, T. C. H., Bernard, J., Li, B.,
Xu, L., Back, G., Butt, A. R., Cameron, K. W., and Hong, Y.}{2018a}
{Predicting system performance by interpolation using a high-dimensional
Delaunay triangulation} {In {\sl Proc. 2018 Spring Simulation Multiconference,
26th High Performance Computing Symp.}, Soc. for Modelling and Simulation
Internat., Vista, CA, Article No. 2}

\reft{Chang, T. H., Watson, L. T., Lux, T. C. H., Li, B., Xu, L., 
Butt, A. R., Cameron, K. W., and Hong, Y.}{2018b}{A polynomial time
algorithm for multivariate interpolation in arbitrary dimension via
the Delaunay triangulation} {In {\sl Proc. ACM 2018 Southeast Conference
(ACMSE '18)}, ACM, New York, NY, Article No. 12}

\refb{Cheney, E. W. and Light, W. A.}{2009}
{A Course in Approximation Theory}{American Mathematical Soc., 
Providence, RI}

\refb{Cheng, S. W., Dey, T. K., and Shewchuk, J.}{2012}
{Delaunay Mesh Generation}{Computer and Information Science
Series, CRC Press, New York, NY}

\refj{Cignoni, P., Montani, C., and Scopigno, R.}{1998}
{DeWall: A fast divide \& conquer Delaunay triangulation algorithm in $E^d$}
{Computer-Aided Design 30}{333--341}

\refb{de Berg, M., Cheong, O., van Kreveld, M., and Overmars, M.}{2008}
{Computational Geometry: Algorithms and Applications}{Springer-Verlag,
Berlin, Germany}

\refj{Devillers, O., Pion, S., and Teillaud, M.}{2002}
{Walking in a triangulation}{Internat. Journal of Foundations of
Computer Science 13}{181--199}

\reft{Edelsbrunner, H.}{1989}{An acyclicity theorem for cell complexes in
d dimensions} {In {\sl Proc. Fifth Annual Symp. on Computational Geometry}, 
ACM, New York, 145--151}

\refj{Hanson, R. J. and Haskell, K. H.}{1982}
{Algorithm 587: Two algorithms for the linearly constrained least
squares problem}{ACM Trans. Math. Softw. 8}{323--333}

\refj{Klee, V.}{1980}{On the complexity of d-dimensional Voronoi diagrams}
{Archiv der Mathematik 34}{75--80}

\reft{Lux, T. C. H., Watson, L. T., Chang, T. H., Bernard, J., Li, B.,
Xu, L., Back, G., Butt, A. R., Cameron, K. W., and Hong, Y.}{2018}
{Predictive modeling of I/O characteristics in high performance computing
systems}{In {\sl Proc. 2018 Spring Simulation Multiconference, 26th High 
Performance Computing Symp.}, Soc. for Modelling and Simulation Internat.,
Vista, CA, Article No. 8}

\refj{M{\"u}cke, E. P., Saias, I., and Zhu, B.}{1999}
{Fast randomized point location without preprocessing in two- and 
three-dimensional Delaunay triangulations}
{Computational Geometry 12}{63--83}

\refj{Omohundro, S. M.}{1990}{Geometric learning algorithms}
{Physica D: Nonlinear Phenomena 42}{307--321}

\reft{OpenMP Architecture Review Board (ARB)}{2015}
{OpenMP Application Programming Interface (version 4.5)}
{OpenMP}

\refj{Rajan, V. T.}{1994}
{Optimality of the Delaunay triangulation in $R^d$}
{Discrete \& Computational Geometry 12}{189--202}

\refj{Schaap, W. E. and van de Weygaert, R.}{2000}
{Continuous fields and discrete samples: reconstruction through
Delaunay tessellations}{Astronomy and Astrophysics 363}{L29--L32}

\refj{Su, P. and Drysdale, R. L. S.}{1997}
{A comparison of sequential Delaunay triangulation algorithms}
{Computational Geometry 7}{361--385}

\refj{Watson, D. F.}{1981}{Computing the n-dimensional Delaunay tessellation
with application to Voronoi polytopes}
{The Computer Journal 24}{167--172}

}\vfill\eject\end
%%%%% bottom

### 2020-06-08 Research_[2020-05-18]_Piecewise_Monotone_Quintic_Fortran/paper/pmqsi20.tex
Added paper directory, all of this work may be moved into the papers repository soon.

--- 2020-06-08 Research_[2020-05-18]_Piecewise_Monotone_Quintic_Fortran/paper/del/delTOMS18.tex
@@
-% Algorithm XXX: DELAUNAYSPARSE---Interpolation via a sparse subset of
-% the Delaunay triangulation in medium to high dimensions,
-% Tyler H. Chang, Layne T. Watson, Thomas C.H. Lux, Ali R. Butt,
-% Kirk W. Cameron, and Yili Hong,
+% Algorithm XXX: PMQSI---Piecewise monotone quintic spline interpolation,
+% Thomas C.H. Lux, Layne T. Watson, Tyler H. Chang, Joshua Detwiler, 
+% Kirk W. Cameron, and Yili Hong
 % ACM Trans. Math. Software, ?/?/2018.
@@
 \def\leftrh{Chang et al.}
-\def\rightrh{Algorithm XXX: DELAUNAYSPARSE}
-
-{\baselineskip=24pt \leftline{\hlvXVIII Algorithm XXX: DELAUNAYSPARSE:}
-\leftline{\hlvXVIII Interpolation via a Sparse Subset of the}
-\leftline{\hlvXVIII Delaunay Triangulation in Medium to High}
-\leftline{\hlvXVIII Dimensions}}
+\def\rightrh{Algorithm XXX: PMQSI}
+
+{\baselineskip=24pt \leftline{\hlvXVIII Algorithm XXX: PMQSI: Piecewise Monotone}
+\leftline{\hlvXVIII Quintic Spline Interpolation}
 \bigskip\medskip
-\leftline{\hlv TYLER H. CHANG, LAYNE T. WATSON, THOMAS C.H. LUX, ALI R. BUTT,}
+\leftline{\hlv THOMAS C.H. LUX, LAYNE T. WATSON, TYLER H. CHANG, JOSHUA DETWILER,}
 \smallskip
@@
 -\parindent{\parindent=0pt\rmVIII Authors' addresses:
-T. H. Chang, T. C. H. Lux, A. R. Butt, K. W. Cameron,
+T. C. H. Lux, T. H. Chang, J. Ditwiler, K. W. Cameron,
 Department of Computer Science, L. T. Watson,
@@
 Polytechnic Institute \& State University, Blacksburg, VA 24061;
-e-mail: {\ttVIII thchang@vt.edu}. \xpar
+e-mail: {\ttVIII tchlux@vt.edu}. \xpar
 % Permission to make digital/hard copy of
@@
 % Abstract and Introduction
-{\rmVIII\parindent=0pt DELAUNAYSPARSE contains both serial and
-parallel codes written in Fortran 2003 (with OpenMP) for performing
-medium- to high-dimensional interpolation via the Delaunay triangulation.
-To accommodate the exponential growth in the size of the Delaunay
-triangulation in high dimensions, DELAUNAYSPARSE computes only a
-sparse subset of the complete Delaunay triangulation, as necessary for
-performing interpolation at the user specified points. This paper includes
-algorithm and implementation details, complexity and sensitivity analyses,
-usage information, and a brief performance study.
+{\rmVIII\parindent=0pt PMQSI contains a serial code written in Fortran
+  1995 for constructing piecewise monotone quintic spline interpolants
+  to data. Using sharp theoretical monotonicity constraints, first and
+  second derivative estimates at data provided by a quadratic facet
+  model are refined to produce a $C^2$ piecewise monotone interpolant.
+  This paper includes algorithm and implementation details, complexity
+  and sensitivity analyses, usage information, and a brief performance
+  study.
 
@@
 \medskip
-General Terms: Algorithms, Computational Geometry, Documentation
+General Terms: Algorithms, Monotonicity, Documentation
 
 \medskip
-Additional Key Words and Phrases: Delaunay triangulation, interpolation,
-multivariate approximation, high-dimensional data
+Additional Key Words and Phrases: Quintic Spline, interpolation
 

### 2020-06-08 Research_[2020-05-18]_Piecewise_Monotone_Quintic_Fortran/pmqsi/PMQSI.f90

@@ -89,17 +89,6 @@ estimate_derivatives : DO I = 1, ND
    ! Precompute these frequently used integers.
    J = I+1
    K = I-1
-   ! Determine the direction of change at the point I.
-   IF (GROWING(I) .OR. SHRINKING(I)) THEN ; DIRECTION = 0.0_R8
-   ELSE IF (I .EQ. 1) THEN
-      IF      (Y(I) .LT. Y(J)) THEN; DIRECTION =  1.0_R8
-      ELSE IF (Y(I) .GT. Y(J)) THEN; DIRECTION = -1.0_R8
-      END IF
-   ELSE
-      IF      (Y(K) .LT. Y(I)) THEN; DIRECTION =  1.0_R8
-      ELSE IF (Y(K) .GT. Y(I)) THEN; DIRECTION = -1.0_R8
-      END IF
-   END IF
    ! Initialize the curvature to be maximally large.
    FX(I,3) = REAL_MAX
    ! If this is a local flat, first and second derivatives are zero valued.
@@ -117,6 +106,16 @@ estimate_derivatives : DO I = 1, ND
       ! Compute the actual second derivative (instead of coefficient A).
       FX(I,3) = MAX(MIN(2.0_R8 * FX(I,3), REAL_MAX), -REAL_MAX)
    ELSE
+      ! Determine the direction of change at the point I.
+      IF (I .EQ. 1) THEN
+         IF      (Y(I) .LT. Y(J)) THEN; DIRECTION =  1.0_R8
+         ELSE IF (Y(I) .GT. Y(J)) THEN; DIRECTION = -1.0_R8
+         END IF
+      ELSE
+         IF      (Y(K) .LT. Y(I)) THEN; DIRECTION =  1.0_R8
+         ELSE IF (Y(K) .GT. Y(I)) THEN; DIRECTION = -1.0_R8
+         END IF
+      END IF
       ! --------------------
       ! Quadratic left of I.
       IF (K .GT. 1) THEN
@@ -367,11 +366,6 @@ SUBROUTINE QUADRATIC(I2)
   I3 = I2+1
   ! Compute the shared denominator.
   D = (X(I1) - X(I2)) * (X(I1) - X(I3)) * (X(I2) - X(I3))
-  ! Prevent division by zero.
-  IF (D .EQ. 0.0_R8) THEN
-     A = REAL_MAX
-     B = 0.0_R8
-  END IF
[…]

### 2020-06-08 Research_[2020-05-18]_Piecewise_Monotone_Quintic_Fortran/test_splines.py

@@ -28,10 +28,10 @@ if TEST_PMQSI:
     SHOW_DERIVATIVES = False
     VISUALIZE_TEST = True
     RANDOM = False
-    NUM_KNOTS = 30
+    NUM_KNOTS = 18
     # NUM_KNOTS *= 2
     np.random.seed(0)
-    NAME = sorted(TESTS)[4]
+    NAME = sorted(TESTS)[1]
     FUNC = TESTS[NAME]
     print('-'*70)
     print("Possible tests:")
@@ -51,13 +51,17 @@ if TEST_PMQSI:
 
     # Get the knots and values for the test.
     values = FUNC(knots)
+
+    knots = [i*(2**(-26) + 2**(-52)) for i in range(18)]
+    values = [0, 1, 1, 1, 0, 20, 19, 18, 17, 0, 0, 3, 0, 1, 6, 16, 16.1, 1 ]
+
     knots = np.asfortranarray(knots, dtype=float)
     values = np.asfortranarray(values, dtype=float)
 
     print()
-    print(f"knots ({len(knots)}):\n  {knots}")
+    print(f"knots ({len(knots)}):\n  {repr(knots)}")
     print()
-    print(f"values ({len(values)}):\n  {values}")
+    print(f"values ({len(values)}):\n  {repr(values)}")
     print("values:")
     print()
     # Compute the spline knots and spline coefficients.

### 2020-06-12 Research_[2020-05-18]_Piecewise_Monotone_Quintic_Fortran/old/pmqsi.tex
Working through a new version of MQSI. Introduced some kind of Heisenbug, still investigating.

--- 2019-12-16 Paper_[2019-11]_HPC/paper/paper.tex
@@
-%**************************************************************************
-%* SpringSim 2020 Author Kit
-%*
-%* Word Processing System: TeXnicCenter and MiKTeX
-%*
-%**************************************************************************
-
-\documentclass{scspaperproc}
-
-\usepackage{latexsym}
-\usepackage{graphicx}
-\usepackage{mathptmx}
-
-%
-%****************************************************************************
-% AUTHOR: You may want to use some of these packages. (Optional)
-\usepackage{amsmath}
-\usepackage{amsfonts}
-\usepackage{amssymb}
-\usepackage{amsbsy}
-\usepackage{amsthm}
-%****************************************************************************
-
-%
-%****************************************************************************
-% AUTHOR: If you do not wish to use hyperlinks, then just comment
-% out the hyperref usepackage commands below.
-
-%% This version of the command is used if you use pdflatex. In this case you
-%% cannot use ps or eps files for graphics, but pdf, jpeg, png etc are fine.
-
-\usepackage[pdftex,colorlinks=true,urlcolor=blue,citecolor=black,anchorcolor=black,linkcolor=black,bookmarks=false]{hyperref}
-
-%% The next versions of the hyperref command are used if you adopt the
-%% outdated latex-dvips-ps2pdf route in generating your pdf file. In
-%% this case you can use ps or eps files for graphics, but not pdf, jpeg, png etc.
-%% However, the final pdf file should embed all fonts required which means that you have to use file
-%% formats which can embed fonts. Please note that the final PDF file will not be generated on your computer!
-%% If you are using WinEdt or PCTeX, then use the following. If you are using
-%% Y&Y TeX then replace "dvips" with "dvipsone"
-
-%% \usepackage[dvips,colorlinks=true,urlcolor=blue,citecolor=black,%
-%% anchorcolor=black,linkcolor=black]{hyperref}
-
-%% The use of the long citation format (e.g. "Brown and Edwards (1993)" rather than "[5]") and at the same
-%% time using the hyperref package can lead to hard to trace bugs in case the citation is broken accross the
-%% line (usually this will mark the entire paragraph as a hyperlink (clickable) which is easily noticeable and fixed
-%% if using colorlinks, but not if the color is black -- as it is now). Worse yet, if a citation spans page boundary,
-%% LaTeX compilation can fail, with an obscure error message. Since this depends a lot on the flow of the text
-%% and wording, these bugs come and go and can be extremely hard for a beginner to trace. The error
-%% message can look like this:
-%%
-%%    ! pdfTeX error (ext4): \pdfendlink ended up in different nesting level than \pdfstartlink.
-%%    \AtBegShi@Output ...ipout \box \AtBeginShipoutBox 
-%%    \fi \fi 
-%%    l.174 
-%%    ! ==> Fatal error occurred, no output PDF file produced!
-%%
-%% and can be universally fixed by putting an \mbox{} around the citation in question (in this case, at line 174)
-%% and maybe adapting the wording a little bit to improve the paragraph typesetting, which is perhaps not
-%% immediately obvious.
-%****************************************************************************
-
-%****************************************************************************
-%*
-%* AUTHOR: YOUR CALL!  Document-specific macros can come here.
-%*
-%****************************************************************************
-
-%% Use a package to allow for multi-row cells in a table.
-\usepackage{multirow}
+\documentclass[12pt,letterpaper]{article}
+\usepackage{fullpage}  % Make the margins smaller
+\usepackage{parskip}   % Don't use indents in front of paragraphs
+\usepackage{xcolor}    % Color package for coloring text.
+\usepackage{fontenc}   % Use a font package.
+\usepackage{libertine} % Pick the font.
+\usepackage{hyperref} % Links 
+\usepackage{amsfonts}  % Math font "mathbb"
 
@@
 \input{lux-algorithm}
+\renewcommand*{\ttdefault}{courier}
 
-% add custom hyphenation rules here
-\usepackage{hyphenat}
-\hyphenation{op-tical net-works semi-conduc-tor}
+\newcommand{\TODO}[1]{{\color{red} #1}}
 
-% If you use theorems
-\newtheoremstyle{scsthe}% hnamei
-{8pt}% hSpace abovei
-{8pt}% hSpace belowi
-{\it}% hBody fonti
-{}% hIndent amounti1
-{\bf}% hTheorem head fontbf
-{.}% hPunctuation after theorem headi
-{.5em}% hSpace after theorem headi2
-{}% hTheorem head spec (can be left empty, meaning `normal')i
+\medskip
 
-\theoremstyle{scsthe}
-\newtheorem{theorem}{Theorem}
-\renewcommand{\thetheorem}{\arabic{theorem}}
-\newtheorem{corollary}[theorem]{Corollary}
-\renewcommand{\thecorollary}{\arabic{corollary}}
-\newtheorem{definition}{Definition}
-\renewcommand{\thedefinition}{\arabic{definition}}
+\title{Piecewise Monotone Quintic Spline Interpolants}
+\author{Thomas C. H. Lux}
+\date{}
 
-% avoid overrunning the right margin; you are welcome to remove this, provided that you take care not to overrun the right margin anywhere in your paper
-\sloppy
+\begin{document}
+\thispagestyle{empty}
+\begin{center}
+    {\LARGE \textbf{Piecewise Monotone Quintic Spline Interpolants}} \hfill
+    %% \href{https://tchlux.info}{\textbf{tchlux.info}} \hfill
+    %% \textbf{804 317 8339} \hfill
+    %% \href{thomas.ch.lux@gmail.com}{\textbf{thomas.ch.lux@gmail.com}}
+    \\ \medskip\href{tchlux@vt.edu}{Thomas C.H. Lux}
+    %% \rule{\textwidth}{1.5pt}
+    \bigskip
+\end{center}
 
-%#########################################################
-%*
-%*  The Document.
-%*
-\begin{document}
+%% \begin{tabular}{@{}l r@{}}
+%%     \begin{tabular}{@{}l@{}}
+%%       %% Hiring Committee \\
+%%       Business \\
+%%       City, State \\
+%%       \today
+%%     \end{tabular} &
+%%     %% \hfill IMAGE
+%% \end{tabular}
 
-%***************************************************************************
-% AUTHOR: Thomas C.H. Lux et al.
-% FORMAT AUTHORS NAMES Like: Author1, Author2 and Author3 (last names)
-%
-%		You need to change the author listing below!
-%               Please list ALL authors using last name only, separate by a comma except
-%               for the last author, separate with "and"
-
-\SCSpagesetup{Lux, Watson, Chang, Xu, Wang, and Hong}
-
-% AUTHOR: Uncomment ONE of these correct conference names.
-\def\SCSconferenceacro{SpringSim'20}
-%\def\SCSconferenceacro{SummerSim}
-%\def\SCSconferenceacro{AutumnSim}
-%\def\SCSconferenceacro{PowerPlantSim}
-
-% AUTHOR: Set the correct year of the conference.
-\def\SCSpublicationyear{2020}
-
-% AUTHOR: Set the correct month and dates; the dates are separated by a single minus sign
-% with no spaces and no leading zeros, the month is a full name (e.g. April) with the first letter
-% capitalized. For example, "April 8-13".
-\def\SCSconferencedates{May 19-May 21}
-
-% AUTHOR: Set the correct venue in the form "City, State, Country", for example "Los Angeles, CA, USA".
-\def\SCSconferencevenue{Fairfax, VA, USA}
-
-% AUTHOR: Enter the title, all letters in upper case
-\title{An Algorithm for Constructing Monotone \\ Quintic Interpolating Splines}
-
-% AUTHOR: Enter the authors of the article, see end of the example document for further examples
-\author{
-Thomas C. H. Lux\\
-Layne T. Watson\\
-Tyler H. Chang\\ [12pt]
-Departments of Computer Science, Mathematics,\\
-and Aerospace and Ocean Engineering\\
-Virginia Polytechnic Institute \& State University\\
-2000 Torgersen Hall, Blacksburg, VA, USA\\
-\{tchlux,ltw,thchang\}@vt.edu\\
-\and
-Li Xu\\%
-Yueyao Wang\\%
-Yili Hong\\ [12pt]
-Department of Statistics\\
-Virginia Polytechnic Institute \& State University\\
-213 Hutcheson Hall, Blacksburg, VA, USA\\
-\{lix1992,yueyao94,yilihong\}@vt.edu\\
-}
-
-\maketitle
-
-%% Examples and notes.
-%% 
-%% \begin{table}[htb]
-%%   \centering
-%%   \caption{Caption above the table!}\label{tab:example}
-%%   \begin{tabular}{r|l}
-%%     First & row \\ \hline
-%%     second & row \\
-%%     third & row \\
-%%   \end{tabular}
-%% \end{table}
-%% 
-%% \begin{figure}[htb]
-%%   \centering
-%%   \includegraphics[width=0.50\textwidth]{path_without_extension}
-%%   \caption{Caption below the figure!}\label{fig:example}
-%% \end{figure}
-%% 
-%% \cite{ref}  - Make a citation that is in parenthesis
-%% \citeN{ref} - Make a citation out of parenthesis (for noun usage)
-%% \shortcite{ref}  - Make a citation for references with >3 authors.
-%% \shortciteN{ref} - Make a noun-form citation for references with >3 authors.
-%% 
-%% 
-%% 5 - 12 page paper mandated
-%% 150 word abstract
-%% 5 keywords maximum
-
-\section*{Abstract}
-
-An algorithm for computing monotone order six piecewise polynomial
-interpolants is proposed. Algebraic constraints for enforcing
-monotonicity are provided that align with quintic monotonicity
-theory. The algorithm is implemented, tested, and applied to several
-sample problems to demonstrate the improved accuracy of monotone
-quintic spline interpolants compared to that for monotone cubic spline
-interpolants.
-
-\textbf{Keywords:} monotone, quintic spline, Hermite interpolation, sixth order polynomial.
-
-\section{Introduction and Motivation}
-\label{sec:introduction}
-
-%% \begin{figure}[htb]
-%%   \centering
-%%   \includegraphics[width=0.50\textwidth]{demo_projection}
-%%   \caption{These are the projection mechanisms that make a cubic polynomial piece monotone.}\label{fig:projection}
-%% \end{figure}
-
-%% The ability to interpolate or regress a function with continuous value and derivatives has proven fundamentally useful.
-In many applications, there are known shapes or characteristics that are expected of approximations.
-For example, a cumulative distribution function in statistics cannot decrease, the entropy of a closed thermodynamic system can never increase, an electric motor should not suddenly change direction when slowing down or speeding up.
-In each of these cases it is important that any constructed spline approximation is monotone.
-
-Many domains of science rely on smooth approximations to real-valued functions over a closed interval.
-Piecewise polynomial functions (splines) provide the smooth approximations for animation in graphics \cite{herman2015techniques,quint2003scalable}, aesthetic structural support in architecture \cite{brennan2019measure}, efficient aerodynamic surfaces in automotive and aerospace engineering \cite{brennan2019measure}, prolonged effective operation of electric motors \shortcite{berglund2009planning}, and accurate nonparametric approximations in statistics \cite{knott2012interpolating}.
-While polynomial interpolants or regressors apply broadly, splines are often a good choice because they can approximate globally complex functions while minimizing the local complexity of an approximation.
-
-It is often the case that the true underlying function or phenomenon being modeled has known properties e.g., convexity, positivity, various levels of continuity, or monotonicity.
-Given a reasonable amount of data, it quickly becomes difficult to achieve desirable properties in a single polynomial function.
-In general, the maintenance of function properties through interpolation/regression is referred to as \textit{shape preserving} \cite{fritsch1980monotone,gregory1985shape}.
-The specific shapes this work will achieve in approximations are monotonicity and $C^2$ continuity.
-These properties are chiefly important to the approximation of cumulative distribution functions and subsequently the effective generation of random numbers from a specified distribution.
-
-In statistics especially, the construction of a monotone interpolating spline that is $C^2$ continuous is meaningfully useful \cite{ramsay1988monotone}.
-A function with these properties could approximate a cumulative distribution function to a high level of accuracy with relatively few intervals.
-A twice continuously differentiable approximation to a cumulative distribution function (CDF) would also produce a corresponding probability density function (PDF) that is continuously differentiable, which is a property many standard parametric distributions maintain.
-
-The currently available software for monotone piecewise polynomial interpolation includes quadratic \cite{he1998monotone}, cubic \cite{fritsch1980monotone}, and (with limited application) quartic \cite{wang2004rational,piah2011improved,yao2018unconditionally} cases.
-In addition, a statistical method for bootstrapping the construction of an arbitrarily smooth monotone fit exists \cite{leitenstorfer2006generalized}, but the method does not take advantage of known analytic properties related to quintic polynomials.
-Theory has been provided for the quintic case \cite{ulrich1994positivity,hess1994positive}, however this theory has not yet been used to construct a monotone quintic spline interpolation routine.
-Recent work suggests that the lack of quintic software may be due to a general unawareness of the theory \shortcite{xie2018semiparametric}.
-
-\begin{figure}
-  \centering
-  \includegraphics[width=.9\textwidth]{spline_demonstration}
-  \caption{Example polynomials that interpolate function values at the ends of the interval $[0,1]$.
-The first only interpolates the function values $g_2(0) = 1$ and $g_2(1) = 0$, making it the order two polynomial $g_2(x) = 1 - x$.
-For the second plot $g_4(x) = 8x^3 - 12x^2 + 4x + 1/2$, which is order four and interpolates the values $g_4(0) = 1/2$, $g_4'(0) = 4$, $g_4(1) = 1/2$, $g_4'(1) = 4$.
-Finally the third plot shows the order six polynomial $g_6(x) = - 64x^5 + 160x^4 - 140x^3 + 50x^2 - 7x + 1$ interpolating the function values $g_6(0) = 1$, $g_6'(0) = -7$, $g_6''(0) = 100$, $g_6(1) = 0$, $g_6'(1) = -7$, $g_6''(1) = -100$.
-Notice that interpolating the same fixed number of function values at each endpoint will always result in an even order interpolating polynomial.
-  }\label{fig:spline_demonstration}
-\end{figure}
-
-The importance of piecewise quintic interpolation over lower order approximations can be simply demonstrated.
-In general, the order of a polynomial determines the number of function values it can interpolate, and the growth rate of error away from the interpolated function values.
-As demonstrated in Figure \ref{fig:spline_demonstration}, it can be seen that matching a value at either end of the interval requires an order two (linear) approximation and each additional derivative at the ends of the interval raises the necessary polynomial order by two.
-The body of this work is composed of a novel algorithm for enforcing monotonicity on quintic polynomial pieces, then extending that solution to work on quintic splines.
-
-%% Software exists that does monotone piecewise quintic regression \cite{}, but the available routines do not computationally take advantage of the quintic monotonicity theory. ???
-
-The major contribution of this work is an algorithm for constructing monotone quintic interpolating splines that utilizes existing quintic monotonicity theory.
-The remainder of this paper is structured as follows: {\bf ?????}
-
-\subsection{Computing a Monotone Cubic Interpolant}
-
-The current state-of-the-art monotone interpolating spline with a mathematical software implementation is piecewise cubic, continuously differentiable, and was first proposed in \cite{fritsch1980monotone} then expanded upon in \cite{carlson1985monotone}.
-Let $\pi: x_0 = k_1 < k_2 < \cdots < k_n = x_1$ be a partition of the interval $[x_0,x_1]$.
-Let $f: \mathbb{R} \rightarrow \mathbb{R}$ be $C^2$, and $\bigl\{f(k_i)\bigr\}_{i=1}^n$ and $\bigl\{\Delta_i\bigr\}_{i=1}^n$ be given sets of function and derivative values at the partition points for a monotone function $f$.
-Either $f(k_i) \leq f(k_{i+1})$, $i=1$, $\ldots$, $n-1$, and $\Delta_i\ge0$, $i=1$, $\ldots$, $n$, or $f(k_i) \geq f(k_{i+1})$, $i=1$, $\ldots$, $n-1$, and $\Delta_i\le0$, $i=1$, $\ldots$, $n$.
-Let $\hat f$ be a piecewise cubic polynomial defined in each subinterval $I_i = [k_i, k_{i+1}]$ by
-
-\begin{align*}
-h_i &= k_{i+1} - k_{i}, \quad
-u(t) = 3t^2 - 2t^3, \quad
-p(t) = t^3 - t^2,\\
-\hat f(x) &= f(k_i) u\big((k_{i+1} - x) / h_i\big) +
-f(k_{i+1}) u\big((x - k_i) / h_i\big) \\
-&{} - h_i\Delta_i p\big((k_{i+1}-x)/h_i\big) + h_i\Delta_{i+1}
-p\big((x-k_i)/h_i\big).
-\end{align*}
-
-Notice that a trivially monotone spline results when $\Delta_i = 0$, for $i = 1$, $\ldots$, $n$.
-However, such a spline has too many \textit{wiggles} for most applications.
-\citeN{carlson1985monotone} show that simple conditions on the derivative values can guarantee monotonicity, and that these conditions can be enforced in a way that ensures modifications on one interval will not break the monotonicity of cubic polynomials over any neighboring intervals.
-Consider the terms $\alpha = \big(\Delta_i (k_{i+1}-k_i)\big) / \big(f(k_{i+1}) - f(k_i)\big)$ and $\beta = \big(\Delta_{i+1}(k_{i+1}-k_i)\big) / \big(f(k_{i+1}) - f(k_i)\big)$.
-If $f(k_{i+1}) - f(k_i)=0$, take $\Delta_i = \Delta_{i+1} =0$ and $\alpha =\beta =1$.
-In general, monotonicity of a cubic polynomial over a subinterval can be maintained by ensuring that $\alpha$ and $\beta$ reside in any of the regions depicted in Figure \ref{fig:projection}.
-
-\begin{figure}
-  \centering
-  \includegraphics[width=0.45\textwidth]{cubic_projection_demonstration}
-  \caption{These are the feasible regions of monotonicity for cubic splines and the projections that make a cubic polynomial piece monotone.
-The regions themselves are numbered 1--4 corresponding to their original description in Fritsch and Carlson (1980).
-One point is projected onto each region as a demonstration.}\label{fig:projection}
-\end{figure}
-
-The actual region of monotonicity for a cubic polynomial is larger, but projection of $(\alpha, \beta)$ into one of these regions ensures that monotonicity will be achieved and not violated for neighboring regions.
-The user must decide which region is most appropriate for the projections based on the application, Fritsch and Carlson recommend using Region 2.
-
-While the cubic monotonicity case affords such a concise solution, the region of monotonicity is not so simple in the quintic case.
-In the next section, an algorithm for performing a projection similar to those for cubic polynomials is proposed.
-
-\section{Monotone Quintic Interpolation}
-
-The following section is composed of three algorithms that together are used to construct a monotone quintic spline interpolant.
-Without loss of generality, the algorithms will only consider the monotone increasing (nondecreasing) case.
-The monotone decreasing (nonincreasing) case is handled similarly.
-Algorithm \ref{alg:check-monotone} checks monotonicity,
-Algorithm \ref{alg:make-monotone} enforces monotonicity on an order six
-polynomial piece of the quintic spline, and Algorithm
-\ref{alg:monotone-spline} uses the previous two algorithms to enforce
-monotonicity for the entire quintic spline.
-
-In pseudocode, a function \texttt{binary\_search($g,$ $a,$ $b$)} is used, where $a,b\in S\subset \mathbb{R}^p$ for convex $S$, $g: S \rightarrow \{$\texttt{FALSE}, \texttt{TRUE}$\}$ is a right continuous Boolean function, $g(b) = \hbox{\texttt{TRUE}}$, and for $\mu \in [0,1]$, $g\big((1-\mu)a + \mu b\big) =\hbox{\texttt{TRUE}} \Longrightarrow g\big((1-\nu)a + \nu b\big) =\hbox{\texttt{TRUE}}$ for $\mu \le \nu\le1$.
-The search returns $\big((1-c)a + c b\big)$ for the smallest $c \in [0,1]$ such that $g\big((1-c)a + c b\big) = \hbox{\texttt{TRUE}}$.
-
-These algorithms make use of first and second derivatives of the approximated function.
-In the case that the first and second derivative information is not provided along with function values, the derivatives are estimated with finite differences of the function values.
-The final quintic spline is represented as a piecewise polynomial using the Newton form for each polynomial piece, and ultimately converted to a $B$-spline representation for evaluation.
+\medskip
 
@@
     {where $x_0$, $x_1 \in \mathbb{R}$, $x_0 < x_1$, and $f$ is an order six polynomial defined by $f(x_0)$, $Df(x_0)$, $D^2f(x_0)$, $f(x_1)$,
-  $Df(x_1)$, $D^2f(x_1)$. Returns \texttt{TRUE} if $f$ is monotone increasing on $[x_0,x_1]$.}
+  $Df(x_1)$, $D^2f(x_1)$. Returns \texttt{TRUE} if $f$ is monotone on $[x_0,x_1]$.}
     {\label{alg:check-monotone}
@@
     \indented{return FALSE}
-    end if
+    else if $\big(f(x_0) < f(x_1)\big)$ ; $s = 1$
+    \\else ; $s = -1$
+    \\end if
   \item if $\big(Df(x_0) < 0$ or $Df(x_1) < 0\big)$
@@
     \comment{The necessity of these first two steps follows directly from the fact that $f$ is $C^2$.
-The next case is in accordance with a simplified condition for quintic monotonicity that reduces to one of cubic positivity studied in \citeN{schmidt1988positivity}, where $\alpha$, $\beta$, $\gamma$, and $\delta$ are defined in terms of values and derivatives of $f$ at $x_0$ and $x_1$.
-Step \ref{alg1-alpha} checks for the necessary condition that $\alpha \ge 0$, Step \ref{alg1-beta} checks $\beta \geq \alpha$, and Step \ref{alg1-gamma} checks $\gamma \geq \delta$, all from \citeN{schmidt1988positivity}.
+The next case is in accordance with a simplified condition for quintic monotonicity that reduces to one of cubic positivity studied in \cite{schmidt1988positivity}, where $\alpha$, $\beta$, $\gamma$, and $\delta$ are defined in terms of values and derivatives of $f$ at $x_0$ and $x_1$.
+Step \ref{alg1-alpha} checks for the necessary condition that $\alpha \ge 0$, Step \ref{alg1-beta} checks $\beta \geq \alpha$, and Step \ref{alg1-gamma} checks $\gamma \geq \delta$, all from \cite{schmidt1988positivity}.
 If all necessary conditions are met, then the order six piece is monotone and Step \ref{alg1-simplified-true} concludes this check.}
@@
     \\ $\displaystyle B := Df(x_1) \frac{x_1 - x_0}{f(x_1) - f(x_0)}$
-    \comment{The variables $A$ and $B$ correspond directly to the theoretical foundation for positive quartic polynomials established in \citeN{ulrich1994positivity}, first defined after Equation (18).}
+    \comment{The variables $A$ and $B$ correspond directly to the theoretical foundation for positive quartic polynomials established in \cite{ulrich1994positivity}, first defined after Equation (18).}
   \item $\displaystyle \gamma_0 := 4 \frac{Df(x_0)}{Df(x_1)} (B/A)^{3/4}$
@@
     \\ $\displaystyle \beta_1 := \frac{-3 (x_1 - x_0)^2}{2 \big(f(x_1) - f(x_0)\big) \sqrt{A} \sqrt{B}} $ \label{alg1-linearized-abg}
-    \comment{The $\gamma$, $\alpha$, and $\beta$ terms with subscripts $0$ and $1$ are algebraic reductions of the simplified conditions for satisfying Theorem 2 in Equation (16) of \citeN{ulrich1994positivity}.
+    \comment{The $\gamma$, $\alpha$, and $\beta$ terms with subscripts $0$ and $1$ are algebraic reductions of the simplified conditions for satisfying Theorem 2 in Equation (16) of \cite{ulrich1994positivity}.
 These terms with subscripts $0$ and $1$ make the computation of $\alpha$, $\beta$, and $\gamma$ functions of the second derivative terms, as seen in Step \ref{alg1-abg} below.}
@@
   \item if $(\beta \leq 6)$ return $\big(\alpha > - (\beta + 2) / 2\big)$
-  \item else $\qquad$ return $\big(\gamma > -2 \sqrt{\beta - 2}\big)$
+    \\ else $\qquad$ return $\big(\gamma > -2 \sqrt{\beta - 2}\big)$
     \\ end if \label{alg1-last}
 }
-%% ----------------------------------------------------------------------
 
-The reason for structuring the $\alpha$, $\beta$, and $\gamma$ computations in terms of the second derivative of $f$ (seen in Step \ref{alg1-abg} of Algorithm \ref{alg:check-monotone}) will become more apparent later.
-The next problem to consider is that of making a nonmonotone order six polynomial piece into a monotone one by modifying its first and second derivative values at the ends of an interval.
-Note that the actual value of the function at the ends of the interval is not modified, as the resulting polynomial needs to interpolate.
+\medskip
 
-%% ----------------------------------------------------------------------
-\algorithm
-    {make\_monotone$\big (x_0$,$x_1$,$f \big)$}
-    {where $x_0,\ x_1 \in \mathbb{R}$, $x_0 < x_1$, and $f$ is an order six polynomial defined by $f(x_0)$, $Df(x_0)$, $D^2f(x_0)$, $f(x_1)$, $Df(x_1)$, $D^2f(x_1)$.
-Returns $f$ monotone on $[x_0,x_1]$.}
-    {\label{alg:make-monotone}
-    \item if $\big(f(x_1) = f(x_0)\big)$
-      \indented{$Df(x_0) := Df(x_1) := D^2f(x_0) := D^2f(x_1) := 0$}
-      \indented{return}
-      end if
-    \item $Df(x_0) :=\ $median$\displaystyle \bigg(0,\ Df(x_0),\ 14 \frac{f(x_1) - f(x_0)}{x_1 - x_0} \bigg)$
-      \\ $Df(x_1) :=\ $median$\displaystyle \bigg(0,\ Df(x_1),\ 14 \frac{f(x_1) - f(x_0)}{x_1 - x_0}\bigg)$ \label{alg2-init-df}
-      \comment{The selection of values in Step \ref{alg2-init-df} for $Df(x_0)$ and $Df(x_1)$ is suggested by \citeN{ulrich1994positivity} and also by \citeN{huynh1993accurate}.
-These rules quickly enforce upper and lower bounds on derivative values to ensure quintic monotonicity is obtainable.}
-      \comment{The next case that will be covered is in accordance with a simplified condition for quintic monotonicity that reduces to one of cubic positivity studied in \citeN{schmidt1988positivity}, where $\alpha$, $\beta$, $\gamma$, and $\delta$ are defined in terms of values and derivatives of $f$ at $x_0$ and $x_1$.
-Steps \ref{alg2-dx-conditions} and \ref{alg2-rescale-dx} ensure that the following bounds on $D^2f(x_0)$ are compatible while Step \ref{alg2-ddx0-upper} ensures that the conditions on $D^2f(x_1)$ are compatible.
-To conclude the simplified case, Step \ref{alg2-gamma} first enforces $\gamma \geq \delta$, then Steps \ref{alg2-alpha} and \ref{alg2-beta} enforce $\alpha \geq 0$ and $\beta \geq \alpha$ respectively.
-Combined, these three conditions derived from Equation (2.18) in \citeN{schmidt1988positivity} guarantee monotonicity of this polynomial piece.}
-    \item if $\big(Df(x_0) = 0$ or $Df(x_1) = 0\big)$
-    \item \indented{$w := x_0 - x_1$\\$v := f(x_0) - f(x_1)$}
-    \item \indented{if $\big(5 Df(x_0) + 4Df(x_1) > 20v / w)$} \label{alg2-dx-conditions}
-    \item \indented{\indented{$Df(x_0) := Df(x_0) \text{ max}\displaystyle \bigg(0, \frac{20v}{w \big(5 Df(x_0) + 4 Df(x_1)\big)}\bigg) $}}
-      \indented{\indented{$Df(x_1) := Df(x_1) \text{ max}\displaystyle \bigg(0, \frac{20v}{w \big(5 Df(x_0) + 4 Df(x_1)\big)}\bigg) $}} \label{alg2-rescale-dx}
-      \indented{end if}
-    \item \indented{$D^2f(x_0) :=\ $min$\displaystyle \bigg(D^2f(x_0), \frac{4 (2 Df(x_0)+Df(x_1))+20 v/w}{w} \bigg)$} \label{alg2-ddx0-upper}
-    \item  \indented{$D^2f(x_0) :=\ $max$\big(D^2f(x_0), 3 Df(x_0) / w \big)$} \label{alg2-gamma}
-    \item \indented{$D^2f(x_1) :=\ $min$\big(D^2f(x_1), {-4}Df(x_1) / w \big)$} \label{alg2-alpha}
-    \item  \indented{$D^2f(x_1) :=\ $max$\displaystyle \bigg(D^2f(x_1), \frac{3w D^2f(x_0) - 24 Df(x_0) - 32 Df(x_1) + 60v/w}{5w} \bigg)$} \label{alg2-beta}
-      \indented{return}
-      end if
-      \comment{The following code considers the case where $Df(x_0) \neq 0$ and $Df(x_1) \neq 0$.}
-    \item $\displaystyle A := Df(x_0)\frac{x_1 - x_0}{f(x_1) - f(x_0)}$
-      \\ $\displaystyle B := Df(x_1) \frac{x_1 - x_0}{f(x_1) - f(x_0)}$
-    \item if $\big($max$(A,B) > 6\big)$
-      \indented{$Df(x_0) := 6 Df(x_0) \big/ $max$(A,B)$}
-      \indented{$Df(x_1) := 6 Df(x_1) \big/ $max$(A,B)$}
-      end if
-      \comment{This ensures that $(A,B)$ remains within a viable region of monotonicity (satisfying Theorem 4, seen in Fig. 6 of \citeN{ulrich1994positivity}).}
-    \item $\eta := \big(D^2f(x_0), D^2f(x_1)\big)$
-      \\ $\displaystyle \eta_0 := \left({-\frac{\sqrt{A}}{4}}\big(7\sqrt{A} + 3\sqrt{B}\big),\ \frac{\sqrt{B}}{4}\big(3\sqrt{A} + 7\sqrt{B}\big) \right)$
-    \item $\bigl(D^2f(x_0)$, $D^2f(x_1)\bigr) :=$ binary\_search$\bigl(g,
-       \eta,\eta_0\bigr)$
-      \comment{Here $g(z) =$\texttt{ is\_monotone}$(x_0,x_1,f)$ where $x_0$, $x_1$, $f(x_0)$, $f(x_1)$, $Df(x_0)$, $Df(x_1)$ are fixed, and the variable $z=\bigl(D^2f(x_0),D^2f(x_1)\bigr)$.
-This binary search identifies feasible values of $D^2f(x_0)$ and $D^2f(x_1)$ that are nearest to the current values $\eta$.
-This search is valid because $\eta_0$ is guaranteed to produce a monotone function, which can be seen in Equation (23) of \citeN{ulrich1994positivity}.}
-    return
-}
-%% ----------------------------------------------------------------------
+\section*{Comments}
 
-Notice that both Algorithms \ref{alg:check-monotone} and \ref{alg:make-monotone} have $\mathcal{O}(1)$ runtime  assuming a fixed level of relative precision.
-A constant number of operations are needed for verifying monotonicity (Algorithm \ref{alg:check-monotone}), while a constant set of operations and a single binary search are performed for enforcing monotonicity (Algorithm \ref{alg:make-monotone}).
-The search requires a fixed number of steps to achieve any predetermined relative precision.
-Also note that an efficient implementation of Algorithm \ref{alg:make-monotone} only needs to recalculate Steps \ref{alg1-abg} through \ref{alg1-last} of Algorithm \ref{alg:check-monotone} during the binary search.
-Next, an algorithm for constructing a monotone quintic spline interpolant is presented.
+This is a final section discussing some of the finer points.
 
-%% ----------------------------------------------------------------------
+%% \medskip
 
-\algorithm
-    {monotone\_spline$\big((k_1$, $\ldots$, $k_n)$, $f$, $s\big)$}
-    {where $(k_1$, $\ldots$, $k_n)$ is an increasing sequence of real numbers, $f$ is an order six piecewise polynomial with breakpoints $k_1$, $\ldots$, $k_n$ defined by the data $\bigl\{f(k_i)\bigr\}_{i=1}^n$, $\bigl\{Df(k_i)\bigr\}_{i=1}^n$, $\bigl\{D^2f(k_i)\bigr\}_{i=1}^n$, and $s > 1$ is an integer shrink factor.}
-    {\label{alg:monotone-spline}
-    \item create queue
-      \\ create changed$(1$, $\ldots$, $n) := $ FALSE
-      \\ create step\_size$(1$, $\ldots$, $n) := \bigl(Df(k_1)/s$, $\ldots$, $Df(k_n)/s\bigr)$ \label{step:create}
-      \comment{The three variables defined in Step \ref{step:create} are used to ensure the eventual achievement of monotonicity.
-The \texttt{queue} is a standard first in first out (FIFO) queue with \texttt{enqueue} and \texttt{dequeue} operations.
-The array \texttt{changed} contains Booleans describing whether or not a breakpoint belongs to an interval that has been modified to enforce monotonicity.
-The \texttt{step\_size} array contains the real-valued derivative decrement step sizes to use in the search for a valid monotone spline.}
-    \item for $i := 1$, $\ldots$, $n-1$
-      \indented{enqueue $\bigl((k_i, k_{i+1})\bigr)$}
-      \comment{Initially, all intervals must be checked for monotonicity. The following loop will continue until all intervals are verified as monotone without need for modification.}
-    \item while $($size$($queue$) > 0)$
-    \item \indented{$(x_j$, $x_{j+1}) := $ dequeue}
-    \item \indented{if $\big($not is\_monotone$(x_j$, $x_{j+1}$, $f)\big)$}
-    \item \indented{\indented{if $\big($changed$(j)\big) \qquad \ Df(x_j) := Df(x_j) -{}$step\_size$(j) \qquad \ \ $ end if}}
-      \indented{\indented{if $\big($changed$(j+1)\big) \quad Df(x_{j+1}) := Df(x_{j+1}) -{}$step\_size$(j+1)$ end if}}
-      \indented{\indented{\comment{If a breakpoint belongs to an interval that has been previously modified, then the enforced monotonicity conditions on the second derivatives of adjacent intervals must contradict one another.
-In turn, the involved first derivatives are decreased towards zero by a predetermined step size.}}}
-    \item \indented{\indented{make\_monotone$(x_j,\ x_{j+1},\ f)$}}
-    \item \indented{\indented{changed$(j) :=$ TRUE}}
-      \indented{\indented{changed$(j+1) :=$ TRUE}} \label{step:changed}
-    \item \indented{\indented{if $\big(j > 1$ and $(x_{j-1}, x_j) \not\in$ queue$\big)$ enqueue $\bigl((x_{j-1}, x_j)\bigr)$ end if}}
-      \indented{\indented{enqueue $\bigl((x_j, x_{j+1})\bigr)$}}
-      \indented{\indented{if $\big(j+1 < n$ and $(x_{j+1}, x_{j+2}) \not\in$ queue$\big)$ enqueue $\bigl((x_{j+1}, x_{j+2})\bigr)$ end if}} \label{step:requeue}
-      \indented{\indented{\comment{Step \ref{step:changed} records the endpoints of the current interval as having been changed, while \ref{step:requeue} adds adjacent intervals to \texttt{queue} that may have inadvertently been made nonmonotone by the changes to the present interval.}}}
-      \indented{end if}
-      end while
-}
-%% ----------------------------------------------------------------------
+%% With respect,\\
+%% \bigskip\\
+%% \bigskip\\
+%% Thomas Lux
+%% \vfill
 
-It is mentioned in \citeN{ulrich1994positivity} that for sufficiently small $Df(k_i)$ and $Df(k_{i+1})$ the admissible solution interval of second derivative values becomes arbitrarily large.
-It can also be seen that decreasing the assigned derivative values towards zero will eventually cause Steps \ref{alg1-alpha} through \ref{alg1-gamma} of Algorithm \ref{alg:check-monotone} to all fail, resulting in Step \ref{alg1-simplified-true} returning \texttt{TRUE}.
-
-Given the potential for successive monotonicity fixes across all intervals, the worst case runtime of Algorithm \ref{alg:monotone-spline} is $\mathcal{O}(s n)$ for $n$ breakpoints and shrink factor $s$.
-In practice this worst case has been observed to be unlikely and difficult to produce.
-
-\section{Experimental Results}
-
-Given the increased order of approximation, it is naturally expected that some accuracy benefit could be gained by using a quintic spline over a cubic spline.
-Experiments 1 and 2 below test some accuracy differences between monotone cubic and quintic splines.
-Experiment 3 analyzes the number of operations performed by Algorithm \ref{alg:monotone-spline} with increasingly large samples of monotone data.
-
-\subsection{Approximating a Trigonometric Function}
-
-For this experiment, the function $\sin(x) + x$ is considered over the interval $[0,(5/2)\pi]$ as seen in Figure \ref{fig:cubic_quintic_sin}.
-Given an increasing number of points, the maximum error of cubic and quintic approximations is shown in Figure \ref{fig:experiment_1}.
-The quintic approximation consistently has a maximum error that is roughly half that of the cubic approximation, given the same number of points.
-There is an increase in the error gap between the two approximations as more points are added.
-The quintic spline requires only $700$ points to approximate the function to the same accuracy achieved by the cubic with $1000$ points.
-
-\begin{figure}
-  \centering
-  \includegraphics[width=.7\textwidth]{cubic-quintic-sin}
-  \caption{Depicted above are the monotone cubic and quintic spline interpolants of the function $\sin(x) + x$ over the points $[0, (5/2) \pi]$.
-Notice that the maximum error of the cubic interpolant is larger, because it only captures first derivative information at the points.
-The quintic interpolant captures both first and second derivative information at the points.}
-  \label{fig:cubic_quintic_sin}
-\end{figure}
-
-\begin{figure}
-  \centering
-  \includegraphics[width=.7\textwidth]{experiment_1_errors}
-  \caption{The error in the quintic and cubic approximations of $\sin(x) + x$ with increasing number of points.
-Both axes are log scaled.
-The quintic approximation generally provides an approximation with about half of the maximum error of the cubic approximation, while the relative gap between grows with increasing number of points.
-In order to achieve the same absolute error the quintic spline requires ten fewer points at max error $10^{-4}$, thirty fewer points at max error $10^{-6}$, and $300$ fewer points at error $10^{-8}$.}
-  \label{fig:experiment_1}
-\end{figure}
-
-\subsection{Random Number Generation}
-
-Consider the cumulative distribution function (CDF) defined by a mixture of three Gaussian distributions and visualized in Figure \ref{fig:experiment_2_distribution}.
-The distributions have weights $(.3, .6, .1)$, means $(.2, .45, .85)$ and standard deviations $(.05, .08, .03)$.
-$200$ random samples are generated from this distribution with both cubic and quintic approximations from four points, and over $100$ trials the resulting spread of empirical distribution functions (EDFs) are visualized in Figure \ref{fig:experiment_2_results}.
-
-While an increased number of points could be used to decrease the error in either approximation, the purpose of this experiment is to demonstrate the effect approximation error has on random number generation.
-The errors in the CDFs are magnified by the variance involved in randomly sampling from a distribution.
-As a result, while the cubic and quintic spline approximations have similar maximum error, the cubic distribution is significantly distorted around the first and second modes.
-
-\begin{figure}
-  \centering
-  \includegraphics[width=.8\textwidth]{experiment_2_distribution}
-  \caption{The cumulative distribution function (top) and probability density function (bottom) for the Gaussian mixture distribution in Experiment 2.
-The distributions have weights $(.3, .6, .1)$, means $(.2, .45, .85)$ and standard deviations $(.05, .08, .03)$.
-Random samples are generated from this mixture distribution by evaluating the inverse of the CDF at random numbers uniformly distributed in the range $[0,1]$.}
-  \label{fig:experiment_2_distribution}
-\end{figure}
-
-\begin{figure}
-  \centering
-  \includegraphics[width=.4\textwidth]{experiment_2_cubic_dist}
-  \includegraphics[width=.4\textwidth]{experiment_2_quintic_dist}
-  \caption{Both cubic and quintic splines are used to approximate the CDF of a Gaussian mixture distribution with three components.
-Four CDF points are used to build the approximation, $200$ random samples are generated with the approximate CDFs over $100$ trials.
-The resulting cloud of empirical distribution function (EDF) values is seen above.
-The maximum approximation error in the CDFs are similar, $.08$ for the cubic and $.05$ for the quintic.
-However, the maximum EDf differences observed are doubled by the variance of random samples, $.15$ for the cubic and $.1$ for the quintic.}
-  \label{fig:experiment_2_results}
-\end{figure}
-
-%% - generate large sets of random monotone data, show Table of number
-%%   of metrics growing with increasing number of points
-
-%% - show visual of the \textit{error}
-
-\subsection{Random Monotone Data}
-
-This experiment studies the number of times the algorithms \texttt{is\_monotone} and \texttt{make\_monotone} are executed for increasingly large sequences of random monotone data.
-The minimum, median, and maximum number of times that Algorithms \ref{alg:check-monotone} and \ref{alg:make-monotone} are executed is recorded, as well as the number of steps taken in all calls to \texttt{binary\_search}.
-The number of calls and steps grows linearly with $n$ as expected, requiring roughly one call to \texttt{make\_monotone} and $20$ binary search steps to create a monotone piece.
-Some (less common) problems require an average of three calls to \texttt{make\_monotone} per interval.
-
-\begin{table}
-  \centering
-  \begin{tabular}{c|c c c|c c c|c c c}
-    \hline
-    \multirow{2}{*}{$n$}
-      & \multicolumn{3}{c|}{Checks} & \multicolumn{3}{c}{Fixes} & \multicolumn{3}{|c}{Search Steps} \\
-      & min & median & max & min & median & max & min & median & max \\
-    \hline
-    5 & 4 & 6 & 274 & 0 & 1 & 154 & 3 & 31 & 2006\\
-    25 & 28 & 47 & 378 & 2 & 14 & 225 & 80 & 348 & 3742\\
-    50 & 61 & 124 & 484 & 6 & 40 & 238 & 217 & 851 & 4445\\
-    75 & 102 & 216 & 842 & 14 & 74 & 411 & 463 & 1491 & 6333\\
-    100 & 142 & 306 & 776 & 21 & 107 & 380 & 629 & 2140 & 7414\\
-    \hline
-  \end{tabular}
-  \caption{Here we see a table of the minimum, median, and maximum number of checks, fixes, and binary search steps required in the execution of \texttt{make\_spline} for increasing size sequences, $n$, over $100$ randomly generated sets of monotone data.
-Notice the maximum for each counter is often significantly greater than the minimum and median, because the distribution of each counter is skewed right.}
-  \label{table:e3_results}
-\end{table}
-
-\section{Discussion}
-
-The monotone quintic spline interpolant provides a distinct increase in accuracy of the monotone cubic variant.
-Although the complexity of the algorithm is increased, the number of points required to achieve the same level of accuracy has decreased.
-In the present state, the algorithm for enforcing monotonicity on a spline is not as trivially parallel as the cubic algorithm, however it can still be parallelized.
-
-Acknowledging that the algorithm for constructing a monotone quintic spline interpolant is slightly more complex than the cubic case, gains in accuracy or a decreased necessary number of points are often worth the computational effort.
-
-\section{Conclusion and Future Work}
-
-This paper proposes and tests an algorithm for constructing monotone quintic spline interpolants.
-Experiments demonstrate an improvement in approximation accuracy over monotone cubic spline interpolants, as expected based on theory.
-
-There are still open avenues of research going forward, such as an alternative sufficient condition for enforcing monotonicity or increased order monotone approximations.
-If the monotonicity conditions can be generalized to any order or made linear, the search for a monotone interpolating spline could potentially be formulated as a convex optimization problem.
-Finally, this work could be used to improve cumulative distribution function (CDF) estimates as well as predictive models that use CDFs.
-
-%% \section*{Acknowledgments}
-%% The NSF is acknowledged for its support of this project.
-
-% Please don't change the bibliographystyle style
-\bibliographystyle{scsproc}
-% AUTHOR: Include your bib file here
-\bibliography{paper}
-
-\section*{Author Biographies}
-
-\textbf{\uppercase{THOMAS C. H. LUX}} is a Ph.D. candidate in computer
-science at Virginia Polytechnic Institute and State University
-(VPI\&SU) studying under Dr. Layne Watson. His research interests
-include computational science, approximation theory, optimization, and
-artificial intelligence. His email address is \email{tchlux@vt.edu}.
-
-\textbf{\uppercase{Layne T. Watson}} (Ph.D., Michigan, 1974) has
-interests in numerical analysis, mathematical programming,
-bioinformatics, and data science.  He has been involved with the
-organization of HPCS since 2000.
-
-\textbf{\uppercase{Tyler H. Chang}} is a Ph.D. candidate at VPI\&SU
-studying computer science under Dr. Layne Watson.
-
-\textbf{\uppercase{Li Xu}} is a Ph.D. candidate at VPI\&SU studying
-statistics under Dr. Yili Hong.
-
-\textbf{\uppercase{Yueyao Wang}} is a Ph.D. candidate at VPI\&SU
-studying statistics under Dr. Yili Hong.
-
-\textbf{\uppercase{Yili Hong}} (Ph.D., Iowa State, 2009) has interests in
-engineering statistics, statistical modeling, and data analysis.
-
-%% \newpage
-
-%% \appendix
-
-%% \section{Author Checklist}
-%% We strive for a consistent appearance in all papers published in the proceedings.
-If you used the template and styles within this author’s kit, then almost all of the requirements in this checklist will be automatically satisfied, and there is very little to check.
-
-%% Please \textbf{print a hardcopy of your paper}, and go over your printed paper to make sure it adheres to the following requirements. \textit{Thank you!}
-
-%% \begin{enumerate}
-%% 	\item Abstract
-%%   \begin{enumerate}
-%% 	  \item 150 or fewer words.
-%% 	  \item Provide 3-5 keywords (mandatory). This set of keywords will identify your paper in indices and databases.
-%% 	\end{enumerate}
-%% 	\item Paper Length
-%%   \begin{enumerate}
-%% 	  \item At least 5, but no more than 12 pages.
-%% 	  \item Page size is letter size (8.5’’ x 11’’, or 216 mm x 279 mm).
-%% 	\end{enumerate}
-%% 	\item All text is in 11-Point Times New Roman except title, header and footer (default in this template).
-%% 	\item Paper title is in 12-Point Times New Roman \textbf{BOLDFACE ALL CAPS} (default in this template).
-%% 	\item The paper has been spellchecked using U.S. English. 
-%% 	\item Spacing and Margins
-%%   \begin{enumerate}
-%% 	  \item Single spaced.
-%% 	  \item Left and right margins are each 1 inch (default in this template).
-%% 	  \item Top and bottom margins are according to the template.
-%% 	  \item Title starts 1.25 inches from the top of the page (default in this template).
-%% 	\end{enumerate}
-%% 	\item Section Headings
-%%   \begin{enumerate}
-%% 	  \item Left justified and set in \textbf{BOLDFACE ALL CAPS} (default in this template).
-%% 	  \item Numbered, except for the abstract, acknowledgments, references and author biographies.
-%% 	  \item Subsection headings are set in \textbf{Boldface Headline Style} (default in this template).
-%% 	\end{enumerate}
-%% 	\item No footnotes or page numbers.
-%% 	\item The copyright notice on the first page contains the name of the symposium or a track where you want to submit your paper, and the running head on subsequent pages is the surnames of all authors.
-%% 	\item Multiple authors are formatted correctly.
-%% 	\item Equations are centered and any equation numbers are in parentheses and right-justified (default).
-%% 	\item Figures and Tables
-%%   \begin{enumerate}
-%% 	  \item All text in figures and tables is readable.
-%% 	  \item Table captions appear above the table.
-%% 	  \item Figure captions appear below the figure.
-%% 	\end{enumerate}
-%% 	\item Citations and References
-%%   \begin{enumerate}
-%% 	  \item Citations are by author and year, and are enclosed in parentheses, not brackets (default \BibTeX). 
-%% 	  \item References are in the hangref style, and are listed alphabetically by the last names(s) of the author(s) (also default \BibTeX\ setting in this template). 
-%% 	\end{enumerate}
-%% 	\item Author biographies are one paragraph per author.
-%% 	\item Hyperlinks
-%%   \begin{enumerate}
-%% 	  \item Be sure that hyperlinks will probably work in the future as well.
-%% 	  \item Live hyperlinks are blue. Nonlive hyperlinks are black (default in this template).
-%% 	\end{enumerate}
-%% 	\item All fonts must be embedded in the resulting \texttt{.pdf}.
-If you use \texttt{pdflatex}, this should already be true--unless you included figures in the \texttt{.eps} or \texttt{.pdf} format which may introduce additional font dependencies.
-You can use the \texttt{pdffonts} command to see if all the fonts are embedded (the column ``emb'' should say yes for all rows).
-You can use GhostScript to force embed the fonts, like so: {\texttt{gs -dNOPAUSE -dBATCH -sDEVICE=pdfwrite -dEmbedAllFonts=true -sOutputFile=out.pdf -f in.pdf}} (in Windows, use \texttt{gswin32} or \texttt{gswin32c} to invoke GhostScript).
-%% \end{enumerate}
-
-%% After verifying that your paper meets these requirements, please go to the final submission page at conference website and submit your paper.
-Be sure to complete the transfer of copyright form and upload the \texttt{.pdf} receipt.
-\textit{Thank you for contributing to the SCS conferences!}
-
-%% \section{Most Observed Mistakes}
-
-%% The following list comprises \textbf{the most common sources of error} that had to be corrected by previous editors.
-Please make sure to go through the following list and check that your paper is formatted correctly:
-%% \begin{enumerate}
-%% 	\item	The paper is less than 5 or more than 12 pages long.
-%% 	\item	Paper title and section titles are in \textbf{BOLD ALL CAPS}, subsections are \textbf{Bold and Capitalize the First Letter of Important Words}. Please use the templates.
-%% 	\item	Paper is in A4 format, not letter format. Please use the required margins.
-%% 	\item	The copyright notice is incorrect.
-%% 	\item	The running heads are incorrect. Do not forget that the LastNameLastAuthor is preceded by ``, and ''.
-%% 	\item	The citation format is incorrect. Use \BibTeX\ for citations and do not change the format used by the template.
-%% 	\item	The biographies are missing. Do not forget the ``author biographies'' section.
-%% 	\item	Figures or tables are not referenced in the text or have the incorrect caption format.
-%% 	\item	The author section after the title is not formatted correctly, the number of organizations does not define the number of blocks, or the number of blocks does not define the layout.
-%% 	\item In the heading on the title page, country names are in all capitals.
-%% 	\item	Paragraphs are not indented.
-%% 	\item Some fonts in the resulting \texttt{.pdf} are not embedded.
-%% \end{enumerate}
+\bibliographystyle{unsrt}
+\bibliography{pmqsi}
 

### 2020-06-12 Research_[2020-05-18]_Piecewise_Monotone_Quintic_Fortran/paper/mqsi20.tex
Working through a new version of MQSI. Introduced some kind of Heisenbug, still investigating.

% Algorithm XXX: MQSI---Monotone quintic spline interpolation,
% Thomas C.H. Lux, Layne T. Watson, Tyler H. Chang,
% William Thacker, Kirk W. Cameron, and Yili Hong
% ACM Trans. Math. Software, ?/?/2020.
% 
%
\magnification=1000 \hsize=30pc \vsize=47pc
\parindent=10pt \parskip=0pt plus 3pt
\baselineskip=13pt plus 2pt minus 1pt \lineskiplimit=2pt
\lineskip=2pt plus 2pt                                  %\raggedright
\tolerance=800
%\abovedisplayskip=6pt plus 6pt minus 3pt
%\abovedisplayshortskip=0pt plus 3pt
%\belowdisplayskip=6pt plus 6pt minus 3pt
%\belowdisplayshortskip=6pt plus 2pt minus 3pt

% eplain package, mainly for its bibliography management.
\input eplain

% Color definitions.
\input colordvi
\def\BEGINC{\textPeach} \def\ENDC{\textBlack}
\def\red#1{\BEGINC#1\ENDC}
\let\beginred=\BEGINC
\let\endred=\ENDC

% Commands for inserting an 'algorithm' section.
\def\indented#1{{\leftskip=1cm}{#1}{\leftskip=0cm}}
\def\comment#1{\vskip 10mm {\it #1} \vskip 10mm}%
\def\thickrule{\hrule\hrule\hrule}

% fonts
\font\itVIII=cmti8 \font\slVIII=cmsl8 \font\bfVIII=cmbx8
\font\ttVIII=cmtt8 \font\caps=cmcsc10 \font\capsVIII=cmcsc10 at 8pt
\font\ttXVIII=cmtt12 at 18pt \font\eightmi=cmmi8 \font\eightsy=cmsy8
\font\rmeight=cmr8  \def\rmVIII{\rmeight \baselineskip=10pt plus 2pt
minus 1pt \lineskiplimit=2pt \lineskip=2pt plus 1pt
\parskip=0pt \textfont0=\rmeight \textfont1=\eightmi \textfont2=\eightsy}
\font\hlvXVIII=phvr at 18pt \font\hlvVIII=phvr at 8pt \font\hlv=phvr

% math blackboard font
\font\tenamsb=msbm10 \font\sevenamsb=msbm7 \font\fiveamsb=msbm5
\newfam\bbfam
\textfont\bbfam=\tenamsb
\scriptfont\bbfam=\sevenamsb
\scriptscriptfont\bbfam=\fiveamsb
\def\bbb{\fam\bbfam}

\input epsf %Postscript figures.
\def\lcaption#1#2{{\noindent\rmVIII Fig.~#1. #2\par}}
\def\ltabcaption#1#2{{\noindent\rmVIII Table #1. #2\par}}

% other definitions.
\def\h{\par\hangindent=10pt\hangafter=1\noindent}
\def\heading#1{\par\bigskip\leftline{\hlv#1}\nobreak\smallskip
\noindent\ignorespaces}
\newdimen\headdigit{\setbox11=\hbox{\hlv 8. }\global\headdigit=\wd11}
\def\footnoterule{\bigskip\kern-3pt \hrule \kern 2.6pt \smallskip}
\def\bullitem{\par\vskip 2pt plus 2pt\hangindent=15pt \hangafter=1
\noindent\hbox to 15pt{---\hfil}\ignorespaces}
\def\bulllitem{\par\vskip 2pt plus 2pt\hangindent=30pt
\hangafter=1 \noindent\hskip 15pt\hbox to
15pt{---\hfil}\ignorespaces}
\def\Th#1 #2\par{\par\medskip{\it{\caps#1}. #2}\par\medskip}
\def\btable#1{\topinsert \rmVIII \let\bf=\bfVIII \textfont0=\scriptfont0
\scriptfont0=\scriptscriptfont0 \textfont1=\scriptfont1
\scriptfont1=\scriptscriptfont1 \centerline{#1}}
\let\xpar=\par
\def\ds{\displaystyle}\def\ab{\allowbreak}

% math definitions
\def\tn#1{\left\Vert #1\right\Vert_2}
\def\mn#1{\left\Vert #1\right\Vert_\infty}
\def\lee{\mathrel{\vcenter{\hbox{$\scriptstyle\mathord<$}\nointerlineskip
\vskip 1pt\hbox{$\scriptstyle\mathord=$}}}}
\def\gee{\mathrel{\vcenter{\hbox{$\scriptstyle\mathord>$}\nointerlineskip
\vskip 1pt\hbox{$\scriptstyle\mathord=$}}}}
\def\real{\mathop{\rm I\!R}\nolimits}
\def\R{{\bf R}}
\def\Rd{{\bf R}^d}
\def\Rdd{{\bf R}^{d\times d}}
%\def\bull{\hfill \vrule height6pt width6pt depth0pt}
\def\:{\mathrel{:\mathord=}}

% OUTPUT
\def\rightheadline{\hfill\hlvVIII\rightrh\hskip20pt$\scriptstyle\bullet$\hskip
20pt\folio}
\def\leftheadline{\hlvVIII\folio\hskip20pt$\scriptstyle\bullet$\hskip
20pt\leftrh\hfill}
\footline={\hfill}%no page number on first page
\output{\plainoutput}
\def\plainoutput{\shipout\vbox{\makeheadline\pagebody\makefootline}
  \advancepageno \global\footline={\hfill}
  \ifodd\pageno
        \global\headline={\rightheadline}
  \else
        \global\headline={\leftheadline}
  \fi
  \ifnum\outputpenalty>-20000 \else\dosupereject\fi}
\def\pagebody{\vbox to\vsize{\boxmaxdepth\maxdepth \pagecontents}}
\def\makeheadline{\vbox to 0pt{\vskip-22.5pt
  \line{\vbox to8.5pt{}\the\headline}\vss}\nointerlineskip}
\def\makefootline{\baselineskip24pt\vskip-6pt\line{\the\footline}}
\def\dosupereject{\ifnum\insertpenalties>0pt
  \line{}\kern-\topskip\nobreak\vfill\supereject\fi}
%\def\footnoterule{\vskip10pt\kern-3pt \hrule width 3em \kern 2.6pt}

\def\leftrh{Lux et al.}
\def\rightrh{Algorithm XXX: MQSI}

{\baselineskip=24pt \leftline{\hlvXVIII Algorithm XXX: MQSI: Monotone Quintic}
\leftline{\hlvXVIII Spline Interpolation}}
\bigskip\medskip
\leftline{\hlv THOMAS C.H. LUX, LAYNE T. WATSON, TYLER H. CHANG,}
\smallskip
\leftline{\hlv WILLIAM THACKER, KIRK W. CAMERON, AND YILI HONG}
\smallskip
\leftline{\hlv Virginia Polytechnic Institute and State University}
\bigskip

\hrule\bigskip\smallskip \footnote{}{\hskip
-\parindent{\parindent=0pt\rmVIII Authors' addresses:
T. C. H. Lux, T. H. Chang, K. W. Cameron,
Department of Computer Science, L. T. Watson,
Departments of Computer Science, Mathematics, and Aerospace and Ocean
Engineering, Y. Hong, Department of Statistics, Virginia
Polytechnic Institute \& State University, Blacksburg, VA 24061;
e-mail: {\ttVIII tchlux@vt.edu}. \xpar
% Permission to make digital/hard copy of
%part or all of this work for personal or classroom use is granted
%without fee provided that the copies are not made or distributed for
%profit or commercial advantage, the copyright notice, the title of
%the publication, and its date appear, and notice is given that
%copying is by permission of the ACM, Inc. To copy otherwise, to
%republish, to post on servers, or to redistribute to lists, requires
%specific permission and/or fee. \xpar\copyright\ 2018 by the
%Association for Computing Machinery, Inc. \xpar 
}}

% Abstract and Introduction
{\rmVIII\parindent=0pt MQSI contains a serial code written in Fortran
  1995 for constructing monotone quintic spline interpolants to
  data. Using sharp theoretical monotonicity constraints, first and
  second derivative estimates at data provided by a quadratic facet
  model are refined to produce a $C^2$ monotone interpolant. This
  paper includes algorithm and implementation details, complexity and
  sensitivity analyses, usage information, and a brief performance
  study.

\medskip
Categories and Subject Descriptors: G.1.1 [{\bfVIII Numerical Analysis}]:
Interpolation --- Spline and piecewise polynomial interpolation;
J.2 [{\bfVIII Computer Applications}]: Physical Science and Engineering
--- {\itVIII Mathematics};
G.4 [{\bfVIII Mathematics of Computing}]: Mathematical Software

\medskip
General Terms: Algorithms, Monotonicity, Documentation

\medskip
Additional Key Words and Phrases: Quintic Spline, interpolation

}

\bigskip\hrule\bigskip\medskip
\heading{1. INTRODUCTION}

Many domains of science rely on smooth approximations to real-valued
functions over a closed interval. Piecewise polynomial functions
(splines) provide the smooth approximations for animation in graphics
\cite{herman2015techniques,quint2003scalable}, aesthetic structural
support in architecture \cite{brennan2019measure}, efficient
aerodynamic surfaces in automotive and aerospace engineering
\cite{brennan2019measure}, prolonged effective operation of electric
motors \cite{berglund2009planning}, and accurate nonparametric
approximations in statistics \cite{knott2012interpolating}. While
polynomial interpolants and regressors apply broadly, splines are
often a good choice because they can approximate globally complex
functions while minimizing the local complexity of an approximation.

It is often the case that the true underlying function or phenomenon
being modeled has known properties e.g., convexity, positivity,
various levels of continuity, or monotonicity. Given a reasonable
amount of data, it quickly becomes difficult to achieve desirable
properties in a single polynomial function. In general, the
maintenance of function properties through interpolation/regression is
referred to as {\it shape preserving}
\cite{fritsch1980monotone,gregory1985shape}. The specific shapes this
work will achieve in approximations are monotonicity and $C^2$
continuity. These properties are chiefly important to the
approximation of cumulative distribution functions and subsequently
the effective generation of random numbers from a specified
distribution.

In statistics especially, the construction of a monotone interpolating
spline that is $C^2$ continuous is meaningfully useful
\cite{ramsay1988monotone}. A function with these properties could
approximate a cumulative distribution function to a high level of
accuracy with relatively few intervals. A twice continuously
differentiable approximation to a cumulative distribution function
(CDF) would also produce a corresponding probability density function
(PDF) that is continuously differentiable, which is a property many
standard parametric distributions maintain.

The currently available software for monotone piecewise polynomial
interpolation includes quadratic \cite{he1998monotone}, cubic
\cite{fritsch1980monotone}, and (with limited application) quartic
\cite{wang2004rational,piah2011improved,yao2018unconditionally}
cases. In addition, a statistical method for bootstrapping the
construction of an arbitrarily smooth monotone fit exists
\cite{leitenstorfer2006generalized}, but the method does not take
advantage of known analytic properties related to quintic
polynomials. Theory has been provided for the quintic case
\cite{ulrich1994positivity,hess1994positive}, however this theory has
not yet been used to construct a monotone quintic spline interpolation
routine. Recent work suggests that the lack of quintic software may be
due to a general unawareness of the theory
\cite{xie2018semiparametric}.

\topinsert
\centerline{\epsfxsize=4.5truein \epsffile{vt_logo.eps}}
{\narrower\noindent\rmVIII Fig.\ 1.
Example polynomials that interpolate function values at the ends of
the interval $[0,1]$. The first only interpolates the function values
$g_2(0) = 1$ and $g_2(1) = 0$, making it the order two polynomial
$g_2(x) = 1 - x$. For the second plot $g_4(x) = 8x^3 - 12x^2 + 4x +
1/2$, which is order four and interpolates the values $g_4(0) = 1/2$,
$g_4'(0) = 4$, $g_4(1) = 1/2$, $g_4'(1) = 4$. Finally the third plot
shows the order six polynomial $g_6(x) = - 64x^5 + 160x^4 - 140x^3 +
50x^2 - 7x + 1$ interpolating the function values $g_6(0) = 1$,
$g_6'(0) = -7$, $g_6''(0) = 100$, $g_6(1) = 0$, $g_6'(1) = -7$,
$g_6''(1) = -100$. Notice that interpolating the same fixed number of
function values at each endpoint will always result in an even order
interpolating polynomial.
\par}
\endinsert

The importance of piecewise quintic interpolation over lower order
approximations can be simply demonstrated. In general, the order of a
polynomial determines the number of function values it can
interpolate, and the growth rate of error away from the interpolated
function values. As demonstrated in Figure
\ref{fig:spline_demonstration}, it can be seen that matching a value
at either end of the interval requires an order two (linear)
approximation and each additional derivative at the ends of the
interval raises the necessary polynomial order by two. Hence, only
even order (odd degree) interpolating splines are broadly
applicable. The body of this work is composed of a novel algorithm for
enforcing monotonicity on quintic polynomial pieces, then extending
that solution to work on quintic splines.

The major contribution of this work is an algorithm for constructing
monotone quintic interpolating splines that utilizes existing quintic
monotonicity theory. The remainder of this paper is structured as
follows: Section \ref{sec:monotone-cubic} summarizes the existing
monotone cubic spline interpolation methodology, Section
\ref{sec:monotone-quintic} presents an algorithm for constructing
monotone quintic spline interpolants, Section \ref{sec:results} offers
experiments and results with cubic and quintic methods, Sections
\ref{sec:discussion} and \ref{sec:conclusion} discuss results and
conclude.

\heading{1.1 \enspace Computing a Monotone Cubic Interpolant}

The current state-of-the-art monotone interpolating spline with a
mathematical software implementation is piecewise cubic, continuously
differentiable, and was first proposed in \cite{fritsch1980monotone}
then expanded upon in \cite{carlson1985monotone}. Let $\pi: x_0 = k_1
< k_2 < \cdots < k_n = x_1$ be a partition of the interval
$[x_0,x_1]$. Let $f: {\bbb R} \rightarrow {\bbb R}$ be $C^2$, and
$\bigl\{f(k_i)\bigr\}_{i=1}^n$ and $\bigl\{\Delta_i\bigr\}_{i=1}^n$ be
given sets of function and derivative values at the partition points
for a monotone function $f$. Either $f(k_i) \leq f(k_{i+1})$, $i=1$,
$\ldots$, $n-1$, and $\Delta_i\ge0$, $i=1$, $\ldots$, $n$, or $f(k_i)
\geq f(k_{i+1})$, $i=1$, $\ldots$, $n-1$, and $\Delta_i\le0$, $i=1$,
$\ldots$, $n$. Let $\hat f$ be a piecewise cubic polynomial defined in
each subinterval $I_i = [k_i, k_{i+1}]$ by
$$ h_i = k_{i+1} - k_{i}, \quad u(t) = 3t^2 - 2t^3, \quad p(t) = t^3 - t^2,$$
$$\hat f(x) = f(k_i) u\big((k_{i+1} - x) / h_i\big) + f(k_{i+1}) u\big((x - k_i) / h_i\big) $$
$$ - h_i\Delta_i p\big((k_{i+1}-x)/h_i\big) + h_i\Delta_{i+1} p\big((x-k_i)/h_i\big).$$ 
Notice that a trivially monotone spline results when $\Delta_i = 0$,
for $i = 1$, $\ldots$, $n$. However, such a spline has too many
{\it wiggles} for most applications. \cite{carlson1985monotone}
show that simple conditions on the derivative values can guarantee
monotonicity, and that these conditions can be enforced in a way that
ensures modifications on one interval will not break the monotonicity
of cubic polynomials over any neighboring intervals. If $f(k_i) =
f(k_{i+1})$, take $\Delta_i = \Delta_{i+1} =0$ and $\alpha =\beta =1$,
otherwise let $\alpha = \big(\Delta_i (k_{i+1}-k_i)\big) /
\big(f(k_{i+1}) - f(k_i)\big)$ and $\beta =
\big(\Delta_{i+1}(k_{i+1}-k_i)\big) / \big(f(k_{i+1}) -
f(k_i)\big)$. Monotonicity of a cubic polynomial over a subinterval
can be maintained by ensuring that $\alpha$ and $\beta$ reside in any
of the regions depicted in Figure \ref{fig:projection}.

\topinsert
\centerline{\epsfxsize=4.5truein \epsffile{vt_logo.eps}}
{\narrower\noindent\rmVIII Fig.\ 2.
These are the feasible regions of monotonicity for cubic splines and
the projections that make a cubic polynomial piece monotone. The
regions themselves are numbered 1--4 corresponding to their original
description in Fritsch and Carlson (1980). One point is projected onto
each region as a demonstration.
\par}
\endinsert

The actual region of monotonicity for a cubic polynomial is larger,
but projection of $(\alpha, \beta)$ into one of these regions ensures
that monotonicity will be achieved and not violated for neighboring
regions. The user must decide which region is most appropriate for the
projections based on the application, Fritsch and Carlson recommend
using Region 2.

While the cubic monotonicity case affords such a concise solution, the
region of monotonicity is not so simple in the quintic case. In the
next section, an algorithm for performing a projection similar to
those for cubic polynomials is proposed.

\heading{2. MONOTONE QUINTIC INTERPOLATION}

The following section is composed of three algorithms that together
are used to construct a monotone quintic spline interpolant. Without
loss of generality, the algorithms will only consider the monotone
increasing (nondecreasing) case. The monotone decreasing case is
handled similarly. Algorithm \ref{alg:check-monotone} checks
monotonicity, Algorithm \ref{alg:make-monotone} enforces monotonicity
on an order six polynomial piece of the quintic spline, and Algorithm
\ref{alg:monotone-spline} uses the previous two algorithms to enforce
monotonicity for the entire quintic spline.

In pseudocode, a function {\tt binary\_search($g,$ $a,$ $b$)} is
used, where $a,b\in S\subset {\bbb R}^p$ for convex $S$, $g: S
\rightarrow \{${\tt FALSE}, {\tt TRUE}$\}$ is a right continuous
Boolean function, $g(b) = \hbox{{\tt TRUE}}$, and for $\mu \in
[0,1]$, $g\big((1-\mu)a + \mu b\big) =\hbox{{\tt TRUE}}
\Longrightarrow g\big((1-\nu)a + \nu b\big) =\hbox{{\tt TRUE}}$ for
$\mu \le \nu\le1$. The search returns $\big((1-c)a + c b\big)$ for the
smallest $c \in [0,1]$ such that $g\big((1-c)a + c b\big) =
\hbox{{\tt TRUE}}$.

These algorithms make use of first and second derivatives of the approximated function where $D$ denotes the differentiation operator. In the case that the first and second derivative information is not provided along with function values, the derivatives are estimated with finite differences of the function values. The final quintic spline is represented as a piecewise polynomial using the Newton form for each polynomial piece, and ultimately converted to a $B$-spline representation for evaluation.

\def\newline{\hfil\break}

\parindent=0pt
\vskip 5mm
\thickrule
{\bf Algorithm 1:} {\tt is\_monotone}$\big(x_0, x_1, f \big)$ \newline
where $x_0$, $x_1 \in {\bbb R}$, $x_0 < x_1$, and $f$ is an order six
polynomial defined by $f(x_0)$, $Df(x_0)$, $D^2f(x_0)$, $f(x_1)$,
$Df(x_1)$, $D^2f(x_1)$. Returns {\tt TRUE} if $f$ is monotone
increasing on $[x_0,x_1]$.
\hrule
\vskip 3mm
\numberedlist
{\tt
\li if $\big(f(x_0) = f(x_1)\big)$ and not $\big( 0 = Df(x_0) = Df(x_1) = D^2f(x_0) = D^2f(x_1) \big)$
\li \indented{return FALSE} \newline
end if \newline
\li if $\big(Df(x_0) < 0$ or $Df(x_1) < 0\big)$ \newline
    \indented{return FALSE} \newline
    end if \newline
    \comment{The necessity of these first two steps follows directly from the fact that $f$ is $C^2$. The next case is in accordance with a simplified condition for quintic monotonicity that reduces to one of cubic positivity studied in \cite{schmidt1988positivity}, where $\alpha$, $\beta$, $\gamma$, and $\delta$ are defined in terms of values and derivatives of $f$ at $x_0$ and $x_1$. Step \ref{alg1-alpha} checks for the necessary condition that $\alpha \ge 0$, Step \ref{alg1-beta} checks $\beta \geq \alpha$, and Step \ref{alg1-gamma} checks $\gamma \geq \delta$, all from \cite{schmidt1988positivity}. If all necessary conditions are met, then the order six piece is monotone and Step \ref{alg1-simplified-true} concludes this check.}
  if $\big(Df(x_0) = 0$ or $Df(x_1) = 0\big)$ \newline
  \indented{$w := x_0 - x_1$} \newline
  \indented{$v := f(x_0) - f(x_1)$} \newline
  \indented{if $\big(D^2f(x_1) > {-4}Df(x_1) / w \big)$ return FALSE} \newline
  \indented{if $\big(D^2f(x_1) < (3w D^2f(x_0) - 24 Df(x_0) - 32
    Df(x_1) + 60v/w) / (5w) \big)$ return FALSE} \newline
  \indented{if $\big(D^2f(x_0) < 3 Df(x_0) / w \big)$ return FALSE} \newline
  \indented{return TRUE} \newline
    end if \newline
    \comment{The following code considers the remaining case where $Df(x_0) \neq 0$ and $Df(x_1) \neq 0$.} \newline
  $\displaystyle A := Df(x_0){x_1 - x_0 \over f(x_1) - f(x_0)}$ \newline
  $\displaystyle B := Df(x_1) {x_1 - x_0 \over f(x_1) - f(x_0)}$ \newline
    \comment{The variables $A$ and $B$ correspond directly to the theoretical foundation for positive quartic polynomials established in \cite{ulrich1994positivity}, first defined after Equation (18).} \newline
  $\displaystyle \gamma_0 := 4 {Df(x_0) \over Df(x_1)} (B/A)^{3/4}$ \newline
  $\displaystyle \gamma_1 := {x_1 - x_0 \over Df(x_1)} (B/A)^{3/4}$ \newline
  $\alpha_0 := 4 (B/A)^{1/4}$ \newline
  $\displaystyle \alpha_1 := -{x_1 - x_0 \over Df(x_1)} (B/A)^{1/4}$ \newline 
  $\displaystyle \beta_0 := 30 - {12 \big(Df(x_0) + Df(x_1)\big) (x_1 - x_0) \over \big(f(x_1) - f(x_0)\big) \sqrt{A}\sqrt{B}}$ \newline
  $\displaystyle \beta_1 := {-3 (x_1 - x_0)^2 \over 2 \big(f(x_1) - f(x_0)\big) \sqrt{A} \sqrt{B}} $ \newline
    \comment{The $\gamma$, $\alpha$, and $\beta$ terms with subscripts $0$ and $1$ are algebraic reductions of the simplified conditions for satisfying Theorem 2 in Equation (16) of \cite{ulrich1994positivity}. These terms with subscripts $0$ and $1$ make the computation of $\alpha$, $\beta$, and $\gamma$ functions of the second derivative terms, as seen in Step \ref{alg1-abg} below.} \newline
  $\gamma := \gamma_0 + \gamma_1 D^2f(x_0)$ \newline
  $\alpha := \alpha_0 + \alpha_1 D^2f(x_1)$ \newline
  $\beta := \beta_0 + \beta_1 \big(D^2f(x_0) - D^2f(x_1)\big)$ \newline
  if $(\beta \leq 6)$ return $\big(\alpha > - (\beta + 2) / 2\big)$ \newline
  else $\qquad$ return $\big(\gamma > -2 \sqrt{\beta - 2}\big)$ \newline
  end if \newline
}
\endnumberedlist
\vskip 2mm
\hrule
\vskip 3mm
\parindent=10pt

%% ----------------------------------------------------------------------

The reason for structuring the $\alpha$, $\beta$, and $\gamma$ computations in terms of the second derivative of $f$ (seen in Step \ref{alg1-abg} of Algorithm \ref{alg:check-monotone}) will become more apparent later. The next problem to consider is that of making a nonmonotone order six polynomial piece into a monotone one by modifying its first and second derivative values at the ends of an interval. Note that the actual value of the function at the ends of the interval is not modified, as the resulting polynomial needs to interpolate.

Notice that both Algorithms \ref{alg:check-monotone} and \ref{alg:make-monotone} have ${\cal O}(1)$ runtime assuming a fixed level of precision is chosen beforehand. A constant number of operations are needed for verifying monotonicity (Algorithm \ref{alg:check-monotone}), while a constant set of operations and a single binary search are performed for enforcing monotonicity (Algorithm \ref{alg:make-monotone}). The search requires a fixed number of steps to achieve any predetermined relative precision, since its accuracy is predetermined and not a function of the problem at hand. Also note that an efficient implementation of Algorithm \ref{alg:make-monotone} only needs to recalculate Steps \ref{alg1-abg} and \ref{alg1-last} of Algorithm \ref{alg:check-monotone} during the binary search. Next, an algorithm for constructing a monotone quintic spline interpolant is presented.

%% ----------------------------------------------------------------------

\vskip 5mm
\thickrule
{\bf Algorithm 2:} {\tt monotone\_spline}$\big((k_1$, $\ldots$, $k_n)$, $f$, $s\big)$
where $(k_1$, $\ldots$, $k_n)$ is an increasing sequence of real
numbers, $f$ is an order six piecewise polynomial with breakpoints
$k_1$, $\ldots$, $k_n$ defined by the data
$\bigl\{f(k_i)\bigr\}_{i=1}^n$, $\bigl\{Df(k_i)\bigr\}_{i=1}^n$,
$\bigl\{D^2f(k_i)\bigr\}_{i=1}^n$, and $s > 1$ is an integer shrink
factor.
\hrule
\vskip 3mm
    \newline create queue
      \newline create changed$(1$, $\ldots$, $n) := $ FALSE
      \newline create step\_size$(1$, $\ldots$, $n) := \bigl(Df(k_1)/s$, $\ldots$, $Df(k_n)/s\bigr)$
      \comment{The three variables defined in Step \ref{step:create} are used to ensure the eventual achievement of monotonicity. The {\tt queue} is a standard first in first out (FIFO) queue with {\tt enqueue} and {\tt dequeue} operations. The array {\tt changed} contains Booleans describing whether or not a breakpoint belongs to an interval that has been modified to enforce monotonicity. The {\tt step\_size} array contains the real-valued derivative decrement step sizes to use in the search for a valid monotone spline.}
    \newline for $i := 1$, $\ldots$, $n-1$
      \indented{enqueue $\bigl((k_i, k_{i+1})\bigr)$}
      end for
      \comment{Initially, all intervals must be checked for monotonicity. The following loop will continue until all intervals are verified as monotone without need for modification.}
    \newline while $($size$($queue$) > 0)$
    \newline \indented{$(k_j$, $k_{j+1}) := $ dequeue}
    \newline \indented{if $\big($not is\_monotone$(k_j$, $k_{j+1}$, $f)\big)$}
    \newline \indented{\indented{if $\big($changed$(j)\big) \qquad \ Df(k_j) := Df(k_j) -{}$step\_size$(j)$}}
      \indented{\indented{if $\big($changed$(j+1)\big) \quad Df(k_{j+1}) := Df(k_{j+1}) -{}$step\_size$(j+1)$}}
      \indented{\indented{\comment{If a breakpoint belongs to an interval that has been previously modified, then the enforced monotonicity conditions on the second derivatives of adjacent intervals must contradict one another. In turn, the involved first derivatives are decreased towards zero by a predetermined step size.}}}
    \newline \indented{\indented{make\_monotone$(k_j,\ k_{j+1},\ f)$}}
    \newline \indented{\indented{changed$(j) :=$ TRUE}}
      \indented{\indented{changed$(j+1) :=$ TRUE}}
    \newline \indented{\indented{if $\big(j > 1$ and $(k_{j-1}, k_j) \not\in$ queue$\big)$ enqueue $\bigl((k_{j-1}, k_j)\bigr)$}}
      \indented{\indented{enqueue $\bigl((k_j, k_{j+1})\bigr)$}}
      \indented{\indented{if $\big(j+1 < n$ and $(k_{j+1}, k_{j+2}) \not\in$ queue$\big)$ enqueue $\bigl((k_{j+1}, k_{j+2})\bigr)$}}
      \indented{\indented{\comment{Step \ref{step:changed} records the endpoints of the current interval as having been changed, while \ref{step:requeue} adds adjacent intervals to {\tt queue} that may have inadvertently been made nonmonotone by the changes to the present interval.}}}      
      \indented{end if}
      end while
\vskip 2mm
\hrule
\vskip 3mm

%% ----------------------------------------------------------------------

It is mentioned in \cite{ulrich1994positivity} that for sufficiently small $Df(k_i)$ and $Df(k_{i+1})$ the admissible solution interval of second derivative values becomes arbitrarily large. It can also be seen that decreasing the assigned derivative values towards zero will eventually cause Steps \ref{alg1-alpha} through \ref{alg1-gamma} of Algorithm \ref{alg:check-monotone} to all fail, resulting in Step \ref{alg1-simplified-true} returning {\tt TRUE}.

If successive monotonicity fixes are required for all intervals, the worst case runtime of Algorithm \ref{alg:monotone-spline} is ${\cal O}(s n)$ for $n$ breakpoints and shrink factor $s$. In practice this worst case has been observed to be unlikely.

\heading{3. EXPERIMENTAL RESULTS}

Given the increased order of approximation, it is naturally expected that some accuracy benefit could be gained by using a quintic spline over a cubic spline. Experiments 1 and 2 below test some accuracy differences between monotone cubic and quintic splines. Experiment 3 analyzes the number of operations performed by Algorithm \ref{alg:monotone-spline} with increasingly large samples of monotone data.

\heading{3.1 \enspace Approximating a Trigonometric Function}

For this experiment, the function $\sin(x) + x$ is considered over the interval $[0,(5/2)\pi]$ as seen in Figure \ref{fig:cubic_quintic_sin}. Given an increasing number of points, the maximum error of cubic and quintic approximations is shown in Figure \ref{fig:experiment_1}. The quintic approximation consistently has a maximum error that is roughly half that of the cubic approximation, given the same number of points. There is an increase in the error gap between the two approximations as more points are added. The quintic spline requires only $700$ points to approximate the function to the same accuracy achieved by the cubic with $1000$ points.

%% \begin{figure}[h]
%%   \centering
%%   \includegraphics[width=.7\textwidth]{cubic-quintic-sin}
%%   \caption{Depicted above are the monotone cubic and quintic spline interpolants of the function $\sin(x) + x$ over the points $[0, (5/2) \pi]$.  Notice that the maximum error of the cubic interpolant is larger, because it only captures first derivative information at the points. The quintic interpolant captures both first and second derivative information at the points.}
%% \end{figure}

%% \begin{figure}
%%   \centering
%%   \includegraphics[width=.7\textwidth]{experiment_1_errors}
%%   \caption{The error in the quintic and cubic approximations of $\sin(x) + x$ with increasing number of points. Both axes are log scaled, where $\mu$ means $\times 10^{-6}$ and n means $\times 10^{-9}$ on the vertical axis. The quintic approximation generally provides an approximation with about half of the maximum error of the cubic approximation, while the relative gap between grows with increasing number of points. In order to achieve the same absolute error the quintic spline requires ten fewer points at max error $10^{-4}$, thirty fewer points at max error $10^{-6}$, and $300$ fewer points at max error $10^{-8}$.}
%% \end{figure}

\heading{3.2 \enspace Random Number Generation}

Consider the cumulative distribution function (CDF) defined by a mixture of three Gaussian distributions and visualized in Figure \ref{fig:experiment_2_distribution}. The distributions have weights $(.3, .6, .1)$, means $(.2, .45, .85)$ and standard deviations $(.05, .08, .03)$. $200$ random samples are generated from this distribution with both cubic and quintic approximations from four points, and over $100$ trials the resulting spread of empirical distribution functions (EDFs) are visualized in Figure \ref{fig:experiment_2_results}.

While an increased number of points could be used to decrease the error in either approximation, the purpose of this experiment is to demonstrate the effect approximation error has on random number generation. The errors in the CDFs are magnified by the variance involved in randomly sampling from a distribution. As a result, while the cubic and quintic spline approximations have similar maximum error, the cubic distribution is significantly distorted around the first and second modes.

%% \begin{figure}
%%   \centering
%%   \includegraphics[width=.8\textwidth]{experiment_2_distribution}
%%   \caption{The cumulative distribution function (top) and probability density function (bottom) for the Gaussian mixture distribution in Experiment 2. The distributions have weights $(.3, .6, .1)$, means $(.2, .45, .85)$ and standard deviations $(.05, .08, .03)$. Random samples are generated from this mixture distribution by evaluating the inverse of the CDF at random numbers uniformly distributed in the range $[0,1]$.}
%% \end{figure}

%% \begin{figure}
%%   \centering
%%   \includegraphics[width=.4\textwidth]{experiment_2_cubic_dist}
%%   \includegraphics[width=.4\textwidth]{experiment_2_quintic_dist}
%%   \caption{Both cubic and quintic splines are used to approximate the CDF of a Gaussian mixture distribution with three components. Four CDF points are used to build the approximation, $200$ random samples are generated with the approximate CDFs over $100$ trials. The resulting cloud of empirical distribution function (EDF) values is seen above. The maximum approximation error in the CDFs are similar, $.08$ for the cubic and $.05$ for the quintic. However, the maximum EDF differences observed are doubled by the variance of random samples, $.15$ for the cubic and $.1$ for the quintic.}
%% \end{figure}

%% - generate large sets of random monotone data, show Table of number
%%   of metrics growing with increasing number of points

%% - show visual of the \textit{error}

\heading{3.3 \enspace Random Monotone Data}

This experiment studies the number of times the algorithms {\tt is\_monotone} and {\tt make\_monotone} are executed for increasingly large sequences of random monotone data. The minimum, median, and maximum number of times that Algorithms \ref{alg:check-monotone} and \ref{alg:make-monotone} are executed is recorded, as well as the number of steps taken in all calls to {\tt binary\_search}. The number of calls and steps grows linearly with $n$ as expected, requiring roughly one call to {\tt make\_monotone} and $20$ binary search steps to create a monotone piece. Some (less common) problems require an average of three calls to {\tt make\_monotone} per interval.

%% \begin{table}
%%   \centering
%%   \caption{The minimum, median, and maximum number of checks, fixes, and binary search steps required in the execution of {\tt make\_spline} for increasing size sequences, $n$, over $100$ randomly generated sets of monotone data. Notice the maximum for each counter is often significantly greater than the minimum and median, because the distribution of each counter is skewed right.}

%%   \begin{tabular}{c|c c c|c c c|c c c}
%%     \hline
%%     \multirow{2}{*}{$n$}
%%       & \multicolumn{3}{c|}{Checks} & \multicolumn{3}{c}{Fixes} & \multicolumn{3}{|c}{Search Steps} \\
%%       & min & median & max & min & median & max & min & median & max \\
%%     \hline
%%     5 & 4 & 6 & 274 & 0 & 1 & 154 & 3 & 31 & 2006\\
%%     25 & 28 & 47 & 378 & 2 & 14 & 225 & 80 & 348 & 3742\\
%%     50 & 61 & 124 & 484 & 6 & 40 & 238 & 217 & 851 & 4445\\
%%     75 & 102 & 216 & 842 & 14 & 74 & 411 & 463 & 1491 & 6333\\
%%     100 & 142 & 306 & 776 & 21 & 107 & 380 & 629 & 2140 & 7414\\
%%     \hline
%%   \end{tabular}
%% \end{table}

\heading{4. DISCUSSION}

The monotone quintic spline interpolant provides a distinct increase in accuracy over the monotone cubic variant. The computational complexity of this algorithm is identical to the existing cubic method and the cost of evaluating the spline after construction is unchanged. While the number of computations required per interval has increased, the number of points required to achieve the same level of accuracy has decreased. In the present state, the algorithm for enforcing monotonicity on a spline is not as trivially parallel as the cubic algorithm, however it can still be parallelized. The checks for monotonicity across all intervals are independent and the monotonicity adjustments could be computed independently and intelligently merged with minor changes to the serial algorithm.

Acknowledging that the algorithm for constructing a monotone quintic spline interpolant is slightly more expensive than the cubic case, gains in accuracy or decreased necessary number of points are often worth the computational effort. In practice the costs of spline interpolation are often dominated by evaluation rather than construction, and the increased accuracy is afforded for a negligible increase in computational cost.

\heading{5. CONCLUSION AND FUTURE WORK}

This paper proposes and tests an algorithm for constructing monotone quintic spline interpolants. Experiments demonstrate an improvement in approximation accuracy over monotone cubic spline interpolants, as expected based on theory. There are still open avenues of research going forward, such as an alternative sufficient condition for enforcing monotonicity or increased order monotone approximations. If the monotonicity conditions can be generalized to any order or made linear, the search for a monotone interpolating spline could potentially be formulated as a convex optimization problem. Finally, this work could be used to improve cumulative distribution function (CDF) estimates as well as predictive models that use CDFs.

\heading{Acknowledgments}
This work was supported by the National Science Foundation Grants CNS-1565314 and CNS-1838271.

\heading{\hlvVIII BIBLIOGRAPHY} %% {\let\caps=\capsVIII \let\sl=\slVIII
%% \def\ys{\hskip 1em minus .5em}\let\rm=\rmeight \rmVIII
%% \newcount\refnum \refnum=0 \parskip=2pt plus 2pt minus 2pt
%% \frenchspacing

\bibliography{mqsi20}
\bibliographystyle{plain}

\vfill\eject\end

%% %\refj is for regular journal articles
%% \def\refj#1#2#3#4#5{\noindent\hangindent=10pt\hangafter=1
%% %\global\advance\refnum by 1[\number\refnum]
%% {\caps #1}\ys #2.\ys {\rm #3}. {\sl #4}, #5.\par}

%% %\refb is for books
%% \def\refb#1#2#3#4{\noindent\hangindent=10pt\hangafter=1
%% %\global\advance\refnum by 1[\number\refnum]
%% {\caps #1}\ys #2.\ys{\sl #3}. {\rm #4}.\par}

%% %\refp is for private communication
%% \def\refp#1#2#3{\noindent\hangindent=10pt\hangafter=1
%% %\global\advance\refnum by 1[\number\refnum]
%% {\caps #1}\ys #2.\ys{\rm #3}.\par}

%% %\reft is for tech. reports, theses, and conferences
%% \def\reft#1#2#3#4{\noindent\hangindent=10pt\hangafter=1
%% %\global\advance\refnum by 1[\number\refnum]
%% {\caps #1}\ys #2.\ys{\rm #3}. {\rm #4}.\par}

%% %%%%% top of refs
%% \reft{Amos, B. D., Easterling, D. R., Watson, L. T., Thacker, W. I.,
%% Castle, B. S., Trosset, M. W.}{2014}{Algorithm XXX: QNSTOP: Quasi-Newton
%% algorithm for stochastic optimization}{Technical Report 2014-07,
%% Virginia Polytechnic Institute and State University, Blacksburg, VA}

%% \refb{Anderson, E., Bai, Z., Bischof, C., Blackford, S., Demmel, J.,
%% Dongarra, J., Du Croz, J., Greenbaum, A., Hammarling, S., McKenney, A.,
%% and Sorensen, D.}{1999}{LAPACK Users' Guide Third Edition}
%% {SIAM, Philidelphia, PA}

%% \refb{Aurenhammer, F., Klein, R., and Lee, D. T.}{2013}
%% {Voronoi diagrams and Delaunay triangulations}
%% {World Scientific Publishing Co., Hackensack, NJ}

%% \refj{Barber, C. B., Dobkin, D. P., and Huhdanpaa, H.}{1996}
%% {The Quickhull algorithm for convex hulls}{ACM Trans. Math.
%% Softw. 22}{469--483}

%% \reft{Boissonnat, J.-D., Devillers, O., and Hornus, S.}{2009}
%% {Incremental construction of the Delaunay triangulation and the
%% Delaunay graph in medium dimension}{In {\sl Proc. Twenty-fifth Annual Symp.
%% on Computational Geometry}, ACM, New York, NY, 208--216}

%% \refj{Bowyer, A.}{1981}{Computing Dirichlet tessellations}
%% {The Computer Journal 24}{162--166}

%% \refj{Cameron, K. W., Anwar, A., Cheng, Y., Xu, L., Li, B., Ananth, U.,
%% Bernard, J., Jearls, C., Lux, T., Hong, Y., Watson, L. T.,
%% and Butt, A. R.}{2019}{MOANA: modeling and analyzing I/O variability
%% in parallel system experimental design}
%% {IEEE Trans. Parallel Distrib. Systems}{to appear}

%% \reft{Chang, T. H., Watson, L. T., Lux, T. C. H., Bernard, J., Li, B.,
%% Xu, L., Back, G., Butt, A. R., Cameron, K. W., and Hong, Y.}{2018a}
%% {Predicting system performance by interpolation using a high-dimensional
%% Delaunay triangulation} {In {\sl Proc. 2018 Spring Simulation Multiconference,
%% 26th High Performance Computing Symp.}, Soc. for Modelling and Simulation
%% Internat., Vista, CA, Article No. 2}

%% \reft{Chang, T. H., Watson, L. T., Lux, T. C. H., Li, B., Xu, L., 
%% Butt, A. R., Cameron, K. W., and Hong, Y.}{2018b}{A polynomial time
%% algorithm for multivariate interpolation in arbitrary dimension via
%% the Delaunay triangulation} {In {\sl Proc. ACM 2018 Southeast Conference
%% (ACMSE '18)}, ACM, New York, NY, Article No. 12}

%% \refb{Cheney, E. W. and Light, W. A.}{2009}
%% {A Course in Approximation Theory}{American Mathematical Soc., 
%% Providence, RI}

%% \refb{Cheng, S. W., Dey, T. K., and Shewchuk, J.}{2012}
%% {Delaunay Mesh Generation}{Computer and Information Science
%% Series, CRC Press, New York, NY}

%% \refj{Cignoni, P., Montani, C., and Scopigno, R.}{1998}
%% {DeWall: A fast divide \& conquer Delaunay triangulation algorithm in $E^d$}
%% {Computer-Aided Design 30}{333--341}

%% \refb{de Berg, M., Cheong, O., van Kreveld, M., and Overmars, M.}{2008}
%% {Computational Geometry: Algorithms and Applications}{Springer-Verlag,
%% Berlin, Germany}

%% \refj{Devillers, O., Pion, S., and Teillaud, M.}{2002}
%% {Walking in a triangulation}{Internat. Journal of Foundations of
%% Computer Science 13}{181--199}

%% \reft{Edelsbrunner, H.}{1989}{An acyclicity theorem for cell complexes in
%% d dimensions} {In {\sl Proc. Fifth Annual Symp. on Computational Geometry}, 
%% ACM, New York, 145--151}

%% \refj{Hanson, R. J. and Haskell, K. H.}{1982}
%% {Algorithm 587: Two algorithms for the linearly constrained least
%% squares problem}{ACM Trans. Math. Softw. 8}{323--333}

%% \refj{Klee, V.}{1980}{On the complexity of d-dimensional Voronoi diagrams}
%% {Archiv der Mathematik 34}{75--80}

%% \reft{Lux, T. C. H., Watson, L. T., Chang, T. H., Bernard, J., Li, B.,
%% Xu, L., Back, G., Butt, A. R., Cameron, K. W., and Hong, Y.}{2018}
%% {Predictive modeling of I/O characteristics in high performance computing
%% systems}{In {\sl Proc. 2018 Spring Simulation Multiconference, 26th High 
%% Performance Computing Symp.}, Soc. for Modelling and Simulation Internat.,
%% Vista, CA, Article No. 8}

%% \refj{M{\"u}cke, E. P., Saias, I., and Zhu, B.}{1999}
%% {Fast randomized point location without preprocessing in two- and 
%% three-dimensional Delaunay triangulations}
%% {Computational Geometry 12}{63--83}

%% \refj{Omohundro, S. M.}{1990}{Geometric learning algorithms}
%% {Physica D: Nonlinear Phenomena 42}{307--321}

%% \reft{OpenMP Architecture Review Board (ARB)}{2015}
%% {OpenMP Application Programming Interface (version 4.5)}
%% {OpenMP}

%% \refj{Rajan, V. T.}{1994}
%% {Optimality of the Delaunay triangulation in $R^d$}
%% {Discrete \& Computational Geometry 12}{189--202}

%% \refj{Schaap, W. E. and van de Weygaert, R.}{2000}
%% {Continuous fields and discrete samples: reconstruction through
%% Delaunay tessellations}{Astronomy and Astrophysics 363}{L29--L32}

%% \refj{Su, P. and Drysdale, R. L. S.}{1997}
%% {A comparison of sequential Delaunay triangulation algorithms}
%% {Computational Geometry 7}{361--385}

%% \refj{Watson, D. F.}{1981}{Computing the n-dimensional Delaunay tessellation
%% with application to Voronoi polytopes}
%% {The Computer Journal 24}{167--172}

%% }\vfill\eject\end
%% %%%%% bottom

### 2020-06-12 Research_[2020-05-18]_Piecewise_Monotone_Quintic_Fortran/monotone_quintic.py

@@ -1,21 +1,15 @@
+from fraction import Fraction
 from polynomial import Spline, polynomial
 
 # Convert a value or list of values into a list of Fraction objects.
 def make_exact(value):
-    from fraction import Fraction
     try:    return list(map(make_exact, value))
     except: return Fraction(value)
 
 # Compute a monotone quintic spline that interpolates given data.
-def monotone_quintic_spline(x, y, exact=True, accuracy=2**(-26),
-                            verbose=False, local_fits=None,
-                            free=False, monotone=True):
-    # Initialize dicationary of local fits.
-    if local_fits is None: local_fits = {}
+def monotone_quintic_spline(x, y, accuracy=2**(-26), verbose=False, monotone=True):
     # Make the values exact.
-    if exact:
-        x = make_exact(x)
-        y = make_exact(y)
+    x, y = make_exact(x), make_exact(y)
     # Identify the local extreme points.
     extremes = {i for i in range(1, len(x)-1)
                       if ((y[i]-y[i-1]) * (y[i+1]-y[i]) < 0)}
@@ -24,56 +18,40 @@ def monotone_quintic_spline(x, y, exact=True, accuracy=2**(-26),
     # Compute "ideal" values everywhere.
     values = [[yi] for yi in y]
     # Compute first and second derivatives.
-    for i in range(len(x)): values[i] += estimate_derivative(
-            x, y, i, extremes, flats, free=free)
-    # Store local fits.
-    for i in range(len(x)):
-        local_fits[i] = (
-            polynomial([x[i]], [y[i]], dx0=values[i][1], ddx0=values[i][2]),
-            [x[max(0,i-1)], x[min(i+1,len(x)-1)]] )
+    for i in range(len(x)): values[i] += quadratic_facet(
+            x, y, i, extremes, flats)
     # Find values nearest to the ideal values that are monotone.
     if monotone:
         # Store the current "step size" that was used for all values.
         best_values = [vi.copy() for vi in values]
         # Make step size and values exact.
-        if exact: best_values, values = make_exact(best_values), make_exact(values)
-        # Track which ones have been shrank.
-        did_shrink = {}
+        best_values, values = make_exact(best_values), make_exact(values)
         # Walk as close to the ideal values as possible.
-        step = 0
[…]

### 2020-06-12 Research_[2020-05-18]_Piecewise_Monotone_Quintic_Fortran/mqsi/MQSI.f90

@@ -11,13 +11,14 @@ SUBROUTINE MQSI(X, Y, T, BCOEF, INFO)
 ! 
 ! OUTPUT:
 !   T(1:3*ND+6) -- The knots for the MQSI B-spline basis.
-!   BCOEF(1:3*ND) -- The coefficients for the MQSI B-spline basis.
+!   BCOEF(1:3*ND+1) -- The coefficients for the MQSI B-spline basis.
 !   INFO -- Integer representing the subroutine return status.
 !    0  Normal return.
 !    1  There are fewer than three data points, so there is nothing to do.
 !    2  X(:) and Y(:) have different sizes.
 !    3  The size of T(:) must be at least the number of knots 3*ND+6.
-!    4  The size of BCOEF(:) must be at least the spline space dimension 3*ND.
+!    4  The size of BCOEF(:) must be at least the spline space
+!       dimension 3*ND plus one (for the value offset).
 !    5  The values in X(:) are not increasing or not separated by at least
 !       their size times the square root of machine precision.
 !    6  The magnitude or spacing of the data (X(:), Y(:)) could lead to
@@ -44,8 +45,8 @@ REAL(KIND=R8), DIMENSION(SIZE(X),3) :: FX ! Spline function values.
 LOGICAL, DIMENSION(SIZE(X)) :: CHECKING, GROWING, SHRINKING ! Execution
 INTEGER, DIMENSION(SIZE(X)) :: TO_CHECK, TO_GROW, TO_SHRINK !   queues.
 REAL(KIND=R8) :: A, ACCURACY, B, DIRECTION, DX, STEP_SIZE
-REAL(KIND=R8), PARAMETER :: TM38 = 10.0_R8**(-38), TP38 = 10.0_R8**38, &
-   TM76 = 10.0_R8**(-76), TP76 = 10.0_R8**76
+REAL(KIND=R8), PARAMETER :: TM54 = 10.0_R8**(-54), &
+     TP38 = 10.0_R8**38, TP76 = 10.0_R8**76
 INTEGER :: I, J, K, NC, ND, NG, NS
 LOGICAL :: SEARCHING
 INTERFACE
[…]

### 2020-06-12 Research_[2020-05-18]_Piecewise_Monotone_Quintic_Fortran/mqsi/SPLINE.f90

@@ -12,8 +12,8 @@ SUBROUTINE FIT_SPLINE(XI, FX, T, BCOEF, INFO)
 ! OUTPUT:
 !   T(1:NB*NCC+2*NCC) -- The nondecreasing real-valued locations
 !                        of the knots for the B-spline basis.
-!   BCOEF(1:NB*NCC) -- The coefficients for the B-splines
-!                      that define the interpolating spline.
+!   BCOEF(1:NB*NCC+1) -- The coefficients for the B-splines
+!                        that define the interpolating spline.
 !   INFO -- Integer representing the subroutine execution status:
 !       0    Successful execution.
 !       1    SIZE(XI) is less than 1.
@@ -45,7 +45,6 @@ SUBROUTINE FIT_SPLINE(XI, FX, T, BCOEF, INFO)
 !   of the linear system defining BCOEF is large, the computed interpolant
 !   may fail to accurately reproduce the data, indicated by INFO = 7.
 !   
-! 
 USE REAL_PRECISION, ONLY: R8
 IMPLICIT NONE
 REAL(KIND=R8), INTENT(IN),  DIMENSION(:)   :: XI
@@ -53,7 +52,7 @@ REAL(KIND=R8), INTENT(IN),  DIMENSION(:,:) :: FX
 REAL(KIND=R8), INTENT(OUT), DIMENSION(:)   :: T, BCOEF
 INTEGER, INTENT(OUT) :: INFO
 ! Local variables.
-INTEGER, DIMENSION(SIZE(BCOEF)) :: IPIV
+INTEGER, DIMENSION(SIZE(BCOEF)-1) :: IPIV
 REAL(KIND=R8), DIMENSION(1 + 3*(2*SIZE(FX,2)-1), SIZE(FX)) :: AB
 REAL(KIND=R8) :: MAX_ERROR
 INTEGER :: DEGREE, & ! Degree of B-spline.
@@ -93,11 +92,11 @@ DEGREE = K - 1
 INFO = 0
 
 ! Check the shape of incoming arrays.
-IF      (NB .LT. 1)             THEN; INFO = 1; RETURN
-ELSE IF (NSPL .LT. 1)           THEN; INFO = 2; RETURN
-ELSE IF (SIZE(FX,1) .NE. NB)    THEN; INFO = 3; RETURN
-ELSE IF (SIZE(T) .LT. NK)       THEN; INFO = 4; RETURN
[…]

### 2020-06-12 Research_[2020-05-18]_Piecewise_Monotone_Quintic_Fortran/mqsi/sample_main.f90

@@ -30,14 +30,14 @@ REAL(KIND=R8), DIMENSION(ND) :: &
             19.0_R8, 18.0_R8, 17.0_R8, 0.0_R8, 0.0_R8, 3.0_R8, &
             0.0_R8, 1.0_R8, 6.0_R8, 16.0_R8, 16.1_R8, 1.0_R8 /)
 ! Definition of holder for spline knots, coefficients, and temporary data.
-REAL(KIND=R8) :: SC(1:3*ND), SK(1:3*ND+6), U(ND)
+REAL(KIND=R8) :: SC(1:3*ND+1), SK(1:3*ND+6), U(ND)
 ! Loop index and info integer for checking exeuction status.
 INTEGER :: I, INFO
 ! Initialize "X" values.
 DO I = 1, ND ; X(I) = REAL(I, KIND=R8) ; END DO
 ! ------------------------------------------------------------------------------
 !                    Optional rescaling operations 
-! X(:) = X(:) * (2.0 * SQRT(EPSILON(1.0_R8))) ! Low separation X values.
+X(:) = X(:) * (2.0 * SQRT(EPSILON(1.0_R8))) ! Low separation X values.
 ! X(:) = X(:) * (10.0_R8**38 / (MAXVAL(X(:)) * 2.0_R8)) ! High separation X values.
 ! Y(:) = Y(:) * (10.0_R8**(-38)) ! Low separation Y values.
 ! Y(:) = Y(:) * (10.0_R8**38 / (MAXVAL(Y(:)) * 2.0_R8)) ! High separation Y values.
@@ -51,7 +51,7 @@ IF (INFO .NE. 0) THEN; WRITE (*,"('MQSI error ',I3)") INFO; END IF
 ! Copy the "evaluation points" into the input/output buffer.
 U(:) = X(:)
 ! Evaluate the spline at all points (result is updated in-place in U).
-CALL EVAL_SPLINE(SK, SC, U, INFO)
+CALL EVAL_SPLINE(SK, SC, U, INFO, D=0)
 IF (INFO .NE. 0) THEN; WRITE (*,"('EVAL_SPLINE error ',I3)") INFO; END IF
 ! Show the evaluations.
 WRITE (*,"('  Q(X): ',100ES10.2)") U(:)
[…]

### 2020-06-12 Research_[2020-05-18]_Piecewise_Monotone_Quintic_Fortran/mqsi/test_all.f90

@@ -21,6 +21,29 @@ PROGRAM TEST_ALL
 USE REAL_PRECISION, ONLY: R8
 IMPLICIT NONE
 ! ------------------------------------------------------------------
+INTERFACE
+ SUBROUTINE MQSI(X, Y, T, BCOEF, INFO)
+   USE REAL_PRECISION, ONLY: R8
+   REAL(KIND=R8), INTENT(IN),  DIMENSION(:) :: X, Y
+   REAL(KIND=R8), INTENT(OUT), DIMENSION(:) :: T, BCOEF
+   INTEGER, INTENT(OUT) :: INFO
+ END SUBROUTINE MQSI
+ SUBROUTINE EVAL_SPLINE(T, BCOEF, XY, INFO, D)
+   USE REAL_PRECISION, ONLY: R8
+   REAL(KIND=R8), INTENT(IN), DIMENSION(:) :: T, BCOEF
+   REAL(KIND=R8), INTENT(INOUT), DIMENSION(:) :: XY
+   INTEGER, INTENT(OUT) :: INFO
+   INTEGER, INTENT(IN), OPTIONAL :: D
+ END SUBROUTINE EVAL_SPLINE
+ SUBROUTINE FIT_SPLINE(XI, FX, T, BCOEF, INFO)
+   USE REAL_PRECISION, ONLY: R8
+   REAL(KIND=R8), INTENT(IN),  DIMENSION(:)   :: XI
+   REAL(KIND=R8), INTENT(IN),  DIMENSION(:,:) :: FX
+   REAL(KIND=R8), INTENT(OUT), DIMENSION(:)   :: T, BCOEF
+   INTEGER, INTENT(OUT) :: INFO
+ END SUBROUTINE FIT_SPLINE
+END INTERFACE
+! ------------------------------------------------------------------
 INTEGER :: TRIALS
 REAL(KIND=R8) :: ERROR_TOLERANCE
 INTEGER :: NS(6)
@@ -61,6 +84,7 @@ main_loop : DO I = 1, SIZE(NS)
         ALL_PASSED = .FALSE.
         EXIT main_loop
   END IF
+
   ! Run a test with maximally spaced points.
   WRITE (*,"('  Huge values')")
   IF (.NOT. HUGE_TEST(N)) THEN
[…]

### 2020-06-12 Research_[2020-05-18]_Piecewise_Monotone_Quintic_Fortran/polynomial.py

@@ -9,8 +9,6 @@
 #                       evaluation, derivative, and a string method.
 #   NewtonPolynomial -- A Newton polynomial with stored coefficients,
 #                       offsets, evaluation, derivative, and a string method.
-#   Vector -- A limited-usage pure Python vector class, used for 
-#             convenience, but WARNING it does check for usage errors.
 # 
 # The following functions are provided:
 # 
@@ -20,36 +18,11 @@
 #   polynomial       -- Given x and y values, this produces a minimum
 #                       degree Newton polynomial that interpolates the
 #                       provided points (this is a *single* polynomial).
-#   local_polynomial -- Construct a local polynomial interpolant about
-#                       a given point in a list of points.
-#   local_quadratic  -- Given a set of points and an index, construct
-#                       a weighted quadratic fit about the point at
-#                       "index" in the set. Use Shepard weighting.
-#   local_facet      -- Given a set of opints and an index, construct
-#                       local linear fits near the index and return
-#                       the local linear regression with lowest error.
-#   inverse          -- Given a function, use the Newton method to
-#                       find the nearest inverse to a guess point.
 # 
 
 
 # This general purpose exception will be raised during user errors.
 class UsageError(Exception): pass
-
-# This class-method wrapper function ensures that the class method
-# recieves a fraction and returns a "float" if a fraction was not
-# provided as input (ensuring internal methods only recieve fractions).
-def float_fallback(class_method):
-    from fraction import Fraction
-    def wrapped_method(obj, x):
-        # Check for vector usage.
-        try: return [wrapped_method(obj, v) for v in x]
-        except:
-            # Return perfect precision if that was provided.
-            if (type(x) == Fraction): return class_method(obj, x)
-            # Otherwise, use exact arithmetic internally and return as float.
-            else: return float(class_method(obj, Fraction(x)))
-    return wrapped_method
 
 # A piecewise polynomial function that supports evaluation,
 # differentiation, and stitches together an arbitrary sequence of
@@ -412,35 +385,6 @@ class NewtonPolynomial(Polynomial):
         return s
 
 
-# A "Vector" subclass of a "list" for convenience, allows component-
[…]

### 2020-06-12 Research_[2020-05-18]_Piecewise_Monotone_Quintic_Fortran/test_integration.py

@@ -25,7 +25,7 @@ def exact(l):
 # f = Polynomial([Fraction(1),Fraction(0),Fraction(0),Fraction(0)])
 # f = Polynomial([Fraction(1),Fraction(0)])
 interval = [0,1]
-f = Spline(interval, exact([[1],[1]]))
+f = Spline(interval, exact([[1,-1],[1,1]]))
 g1 = f.integral(c=0)
 g2 = g1.integral(c=0)
 g3 = g2.integral(c=0)
@@ -49,7 +49,7 @@ print('y: ', y.shape)
 x = np.array(x, dtype=float)
 y = np.array(y, dtype=float, order='F')
 sk = np.ones(nb*ncc + 2*ncc, dtype=float)
-sc = np.ones(nb*ncc, dtype=float)
+sc = np.ones(nb*ncc+1, dtype=float)
 sk, sc, info = splines.fit_spline(x, y, sk, sc)
 
 print("sk:   ", list(sk))

### 2020-06-21 Research_[2020-05-18]_Piecewise_Monotone_Quintic_Fortran/old/example_tex/old_files/0-cubic.tex
Working through revisions given by Thacker. Sent next draft to Watson asking for some guidance.

--- 2020-06-12 Research_[2020-05-18]_Piecewise_Monotone_Quintic_Fortran/paper/mqsi20.tex
@@
-% Algorithm XXX: MQSI---Monotone quintic spline interpolation,
-% Thomas C.H. Lux, Layne T. Watson, Tyler H. Chang,
-% William Thacker, Kirk W. Cameron, and Yili Hong
-% ACM Trans. Math. Software, ?/?/2020.
-% 
-%
-\magnification=1000 \hsize=30pc \vsize=47pc
-\parindent=10pt \parskip=0pt plus 3pt
-\baselineskip=13pt plus 2pt minus 1pt \lineskiplimit=2pt
-\lineskip=2pt plus 2pt                                  %\raggedright
-\tolerance=800
-%\abovedisplayskip=6pt plus 6pt minus 3pt
-%\abovedisplayshortskip=0pt plus 3pt
-%\belowdisplayskip=6pt plus 6pt minus 3pt
-%\belowdisplayshortskip=6pt plus 2pt minus 3pt
-
-% eplain package, mainly for its bibliography management.
-\input eplain
-
-% Color definitions.
-\input colordvi
-\def\BEGINC{\textPeach} \def\ENDC{\textBlack}
-\def\red#1{\BEGINC#1\ENDC}
-\let\beginred=\BEGINC
-\let\endred=\ENDC
-
-% Commands for inserting an 'algorithm' section.
-\def\indented#1{{\leftskip=1cm}{#1}{\leftskip=0cm}}
-\def\comment#1{\vskip 10mm {\it #1} \vskip 10mm}%
-\def\thickrule{\hrule\hrule\hrule}
-
-% fonts
-\font\itVIII=cmti8 \font\slVIII=cmsl8 \font\bfVIII=cmbx8
-\font\ttVIII=cmtt8 \font\caps=cmcsc10 \font\capsVIII=cmcsc10 at 8pt
-\font\ttXVIII=cmtt12 at 18pt \font\eightmi=cmmi8 \font\eightsy=cmsy8
-\font\rmeight=cmr8  \def\rmVIII{\rmeight \baselineskip=10pt plus 2pt
-minus 1pt \lineskiplimit=2pt \lineskip=2pt plus 1pt
-\parskip=0pt \textfont0=\rmeight \textfont1=\eightmi \textfont2=\eightsy}
-\font\hlvXVIII=phvr at 18pt \font\hlvVIII=phvr at 8pt \font\hlv=phvr
-
-% math blackboard font
-\font\tenamsb=msbm10 \font\sevenamsb=msbm7 \font\fiveamsb=msbm5
-\newfam\bbfam
-\textfont\bbfam=\tenamsb
-\scriptfont\bbfam=\sevenamsb
-\scriptscriptfont\bbfam=\fiveamsb
-\def\bbb{\fam\bbfam}
-
-\input epsf %Postscript figures.
-\def\lcaption#1#2{{\noindent\rmVIII Fig.~#1. #2\par}}
-\def\ltabcaption#1#2{{\noindent\rmVIII Table #1. #2\par}}
-
-% other definitions.
-\def\h{\par\hangindent=10pt\hangafter=1\noindent}
-\def\heading#1{\par\bigskip\leftline{\hlv#1}\nobreak\smallskip
-\noindent\ignorespaces}
-\newdimen\headdigit{\setbox11=\hbox{\hlv 8. }\global\headdigit=\wd11}
-\def\footnoterule{\bigskip\kern-3pt \hrule \kern 2.6pt \smallskip}
-\def\bullitem{\par\vskip 2pt plus 2pt\hangindent=15pt \hangafter=1
-\noindent\hbox to 15pt{---\hfil}\ignorespaces}
-\def\bulllitem{\par\vskip 2pt plus 2pt\hangindent=30pt
-\hangafter=1 \noindent\hskip 15pt\hbox to
-15pt{---\hfil}\ignorespaces}
-\def\Th#1 #2\par{\par\medskip{\it{\caps#1}. #2}\par\medskip}
-\def\btable#1{\topinsert \rmVIII \let\bf=\bfVIII \textfont0=\scriptfont0
-\scriptfont0=\scriptscriptfont0 \textfont1=\scriptfont1
-\scriptfont1=\scriptscriptfont1 \centerline{#1}}
-\let\xpar=\par
-\def\ds{\displaystyle}\def\ab{\allowbreak}
-
-% math definitions
-\def\tn#1{\left\Vert #1\right\Vert_2}
-\def\mn#1{\left\Vert #1\right\Vert_\infty}
-\def\lee{\mathrel{\vcenter{\hbox{$\scriptstyle\mathord<$}\nointerlineskip
-\vskip 1pt\hbox{$\scriptstyle\mathord=$}}}}
-\def\gee{\mathrel{\vcenter{\hbox{$\scriptstyle\mathord>$}\nointerlineskip
-\vskip 1pt\hbox{$\scriptstyle\mathord=$}}}}
-\def\real{\mathop{\rm I\!R}\nolimits}
-\def\R{{\bf R}}
-\def\Rd{{\bf R}^d}
-\def\Rdd{{\bf R}^{d\times d}}
-%\def\bull{\hfill \vrule height6pt width6pt depth0pt}
-\def\:{\mathrel{:\mathord=}}
-
-% OUTPUT
-\def\rightheadline{\hfill\hlvVIII\rightrh\hskip20pt$\scriptstyle\bullet$\hskip
-20pt\folio}
-\def\leftheadline{\hlvVIII\folio\hskip20pt$\scriptstyle\bullet$\hskip
-20pt\leftrh\hfill}
-\footline={\hfill}%no page number on first page
-\output{\plainoutput}
-\def\plainoutput{\shipout\vbox{\makeheadline\pagebody\makefootline}
-  \advancepageno \global\footline={\hfill}
-  \ifodd\pageno
-        \global\headline={\rightheadline}
-  \else
-        \global\headline={\leftheadline}
-  \fi
-  \ifnum\outputpenalty>-20000 \else\dosupereject\fi}
-\def\pagebody{\vbox to\vsize{\boxmaxdepth\maxdepth \pagecontents}}
-\def\makeheadline{\vbox to 0pt{\vskip-22.5pt
-  \line{\vbox to8.5pt{}\the\headline}\vss}\nointerlineskip}
-\def\makefootline{\baselineskip24pt\vskip-6pt\line{\the\footline}}
-\def\dosupereject{\ifnum\insertpenalties>0pt
-  \line{}\kern-\topskip\nobreak\vfill\supereject\fi}
-%\def\footnoterule{\vskip10pt\kern-3pt \hrule width 3em \kern 2.6pt}
-
-\def\leftrh{Lux et al.}
-\def\rightrh{Algorithm XXX: MQSI}
-
-{\baselineskip=24pt \leftline{\hlvXVIII Algorithm XXX: MQSI: Monotone Quintic}
-\leftline{\hlvXVIII Spline Interpolation}}
-\bigskip\medskip
-\leftline{\hlv THOMAS C.H. LUX, LAYNE T. WATSON, TYLER H. CHANG,}
-\smallskip
-\leftline{\hlv WILLIAM THACKER, KIRK W. CAMERON, AND YILI HONG}
-\smallskip
-\leftline{\hlv Virginia Polytechnic Institute and State University}
-\bigskip
-
-\hrule\bigskip\smallskip \footnote{}{\hskip
--\parindent{\parindent=0pt\rmVIII Authors' addresses:
-T. C. H. Lux, T. H. Chang, K. W. Cameron,
-Department of Computer Science, L. T. Watson,
-Departments of Computer Science, Mathematics, and Aerospace and Ocean
-Engineering, Y. Hong, Department of Statistics, Virginia
-Polytechnic Institute \& State University, Blacksburg, VA 24061;
-e-mail: {\ttVIII tchlux@vt.edu}. \xpar
-% Permission to make digital/hard copy of
-%part or all of this work for personal or classroom use is granted
-%without fee provided that the copies are not made or distributed for
-%profit or commercial advantage, the copyright notice, the title of
-%the publication, and its date appear, and notice is given that
-%copying is by permission of the ACM, Inc. To copy otherwise, to
-%republish, to post on servers, or to redistribute to lists, requires
-%specific permission and/or fee. \xpar\copyright\ 2018 by the
-%Association for Computing Machinery, Inc. \xpar 
-}}
-
-% Abstract and Introduction
-{\rmVIII\parindent=0pt MQSI contains a serial code written in Fortran
-  1995 for constructing monotone quintic spline interpolants to
-  data. Using sharp theoretical monotonicity constraints, first and
-  second derivative estimates at data provided by a quadratic facet
-  model are refined to produce a $C^2$ monotone interpolant. This
-  paper includes algorithm and implementation details, complexity and
-  sensitivity analyses, usage information, and a brief performance
-  study.
-
-\medskip
-Categories and Subject Descriptors: G.1.1 [{\bfVIII Numerical Analysis}]:
-Interpolation --- Spline and piecewise polynomial interpolation;
-J.2 [{\bfVIII Computer Applications}]: Physical Science and Engineering
---- {\itVIII Mathematics};
-G.4 [{\bfVIII Mathematics of Computing}]: Mathematical Software
-
-\medskip
-General Terms: Algorithms, Monotonicity, Documentation
-
-\medskip
-Additional Key Words and Phrases: Quintic Spline, interpolation
-
-}
-
-\bigskip\hrule\bigskip\medskip
-\heading{1. INTRODUCTION}
-
-Many domains of science rely on smooth approximations to real-valued
-functions over a closed interval. Piecewise polynomial functions
-(splines) provide the smooth approximations for animation in graphics
-\cite{herman2015techniques,quint2003scalable}, aesthetic structural
-support in architecture \cite{brennan2019measure}, efficient
-aerodynamic surfaces in automotive and aerospace engineering
-\cite{brennan2019measure}, prolonged effective operation of electric
-motors \cite{berglund2009planning}, and accurate nonparametric
-approximations in statistics \cite{knott2012interpolating}. While
-polynomial interpolants and regressors apply broadly, splines are
-often a good choice because they can approximate globally complex
-functions while minimizing the local complexity of an approximation.
-
-It is often the case that the true underlying function or phenomenon
-being modeled has known properties e.g., convexity, positivity,
-various levels of continuity, or monotonicity. Given a reasonable
-amount of data, it quickly becomes difficult to achieve desirable
-properties in a single polynomial function. In general, the
-maintenance of function properties through interpolation/regression is
-referred to as {\it shape preserving}
-\cite{fritsch1980monotone,gregory1985shape}. The specific shapes this
-work will achieve in approximations are monotonicity and $C^2$
-continuity. These properties are chiefly important to the
-approximation of cumulative distribution functions and subsequently
-the effective generation of random numbers from a specified
-distribution.
-
-In statistics especially, the construction of a monotone interpolating
-spline that is $C^2$ continuous is meaningfully useful
-\cite{ramsay1988monotone}. A function with these properties could
-approximate a cumulative distribution function to a high level of
-accuracy with relatively few intervals. A twice continuously
-differentiable approximation to a cumulative distribution function
-(CDF) would also produce a corresponding probability density function
-(PDF) that is continuously differentiable, which is a property many
-standard parametric distributions maintain.
-
-The currently available software for monotone piecewise polynomial
-interpolation includes quadratic \cite{he1998monotone}, cubic
-\cite{fritsch1980monotone}, and (with limited application) quartic
-\cite{wang2004rational,piah2011improved,yao2018unconditionally}
-cases. In addition, a statistical method for bootstrapping the
-construction of an arbitrarily smooth monotone fit exists
-\cite{leitenstorfer2006generalized}, but the method does not take
-advantage of known analytic properties related to quintic
-polynomials. Theory has been provided for the quintic case
-\cite{ulrich1994positivity,hess1994positive}, however this theory has
-not yet been used to construct a monotone quintic spline interpolation
-routine. Recent work suggests that the lack of quintic software may be
-due to a general unawareness of the theory
-\cite{xie2018semiparametric}.
-
-\topinsert
-\centerline{\epsfxsize=4.5truein \epsffile{vt_logo.eps}}
-{\narrower\noindent\rmVIII Fig.\ 1.
-Example polynomials that interpolate function values at the ends of
-the interval $[0,1]$. The first only interpolates the function values
-$g_2(0) = 1$ and $g_2(1) = 0$, making it the order two polynomial
-$g_2(x) = 1 - x$. For the second plot $g_4(x) = 8x^3 - 12x^2 + 4x +
-1/2$, which is order four and interpolates the values $g_4(0) = 1/2$,
-$g_4'(0) = 4$, $g_4(1) = 1/2$, $g_4'(1) = 4$. Finally the third plot
-shows the order six polynomial $g_6(x) = - 64x^5 + 160x^4 - 140x^3 +
-50x^2 - 7x + 1$ interpolating the function values $g_6(0) = 1$,
-$g_6'(0) = -7$, $g_6''(0) = 100$, $g_6(1) = 0$, $g_6'(1) = -7$,
-$g_6''(1) = -100$. Notice that interpolating the same fixed number of
-function values at each endpoint will always result in an even order
-interpolating polynomial.
-\par}
-\endinsert
-
-The importance of piecewise quintic interpolation over lower order
-approximations can be simply demonstrated. In general, the order of a
-polynomial determines the number of function values it can
-interpolate, and the growth rate of error away from the interpolated
-function values. As demonstrated in Figure
-\ref{fig:spline_demonstration}, it can be seen that matching a value
-at either end of the interval requires an order two (linear)
-approximation and each additional derivative at the ends of the
-interval raises the necessary polynomial order by two. Hence, only
-even order (odd degree) interpolating splines are broadly
-applicable. The body of this work is composed of a novel algorithm for
-enforcing monotonicity on quintic polynomial pieces, then extending
-that solution to work on quintic splines.
-
-The major contribution of this work is an algorithm for constructing
-monotone quintic interpolating splines that utilizes existing quintic
-monotonicity theory. The remainder of this paper is structured as
-follows: Section \ref{sec:monotone-cubic} summarizes the existing
-monotone cubic spline interpolation methodology, Section
-\ref{sec:monotone-quintic} presents an algorithm for constructing
-monotone quintic spline interpolants, Section \ref{sec:results} offers
-experiments and results with cubic and quintic methods, Sections
-\ref{sec:discussion} and \ref{sec:conclusion} discuss results and
-conclude.
-
 \heading{1.1 \enspace Computing a Monotone Cubic Interpolant}
@@
 those for cubic polynomials is proposed.
-
-\heading{2. MONOTONE QUINTIC INTERPOLATION}
-
-The following section is composed of three algorithms that together
-are used to construct a monotone quintic spline interpolant. Without
-loss of generality, the algorithms will only consider the monotone
-increasing (nondecreasing) case. The monotone decreasing case is
-handled similarly. Algorithm \ref{alg:check-monotone} checks
-monotonicity, Algorithm \ref{alg:make-monotone} enforces monotonicity
-on an order six polynomial piece of the quintic spline, and Algorithm
-\ref{alg:monotone-spline} uses the previous two algorithms to enforce
-monotonicity for the entire quintic spline.
-
-In pseudocode, a function {\tt binary\_search($g,$ $a,$ $b$)} is
-used, where $a,b\in S\subset {\bbb R}^p$ for convex $S$, $g: S
-\rightarrow \{${\tt FALSE}, {\tt TRUE}$\}$ is a right continuous
-Boolean function, $g(b) = \hbox{{\tt TRUE}}$, and for $\mu \in
-[0,1]$, $g\big((1-\mu)a + \mu b\big) =\hbox{{\tt TRUE}}
-\Longrightarrow g\big((1-\nu)a + \nu b\big) =\hbox{{\tt TRUE}}$ for
-$\mu \le \nu\le1$. The search returns $\big((1-c)a + c b\big)$ for the
-smallest $c \in [0,1]$ such that $g\big((1-c)a + c b\big) =
-\hbox{{\tt TRUE}}$.
-
-These algorithms make use of first and second derivatives of the approximated function where $D$ denotes the differentiation operator.
-In the case that the first and second derivative information is not provided along with function values, the derivatives are estimated with finite differences of the function values.
-The final quintic spline is represented as a piecewise polynomial using the Newton form for each polynomial piece, and ultimately converted to a $B$-spline representation for evaluation.
-
-\def\newline{\hfil\break}
-
-\parindent=0pt
-\vskip 5mm
-\thickrule
-{\bf Algorithm 1:} {\tt is\_monotone}$\big(x_0, x_1, f \big)$ \newline
-where $x_0$, $x_1 \in {\bbb R}$, $x_0 < x_1$, and $f$ is an order six
-polynomial defined by $f(x_0)$, $Df(x_0)$, $D^2f(x_0)$, $f(x_1)$,
-$Df(x_1)$, $D^2f(x_1)$. Returns {\tt TRUE} if $f$ is monotone
-increasing on $[x_0,x_1]$.
-\hrule
-\vskip 3mm
-\numberedlist
-{\tt
-\li if $\big(f(x_0) = f(x_1)\big)$ and not $\big( 0 = Df(x_0) = Df(x_1) = D^2f(x_0) = D^2f(x_1) \big)$
-\li \indented{return FALSE} \newline
-end if \newline
-\li if $\big(Df(x_0) < 0$ or $Df(x_1) < 0\big)$ \newline
-    \indented{return FALSE} \newline
-    end if \newline
-    \comment{The necessity of these first two steps follows directly from the fact that $f$ is $C^2$.
-The next case is in accordance with a simplified condition for quintic monotonicity that reduces to one of cubic positivity studied in \cite{schmidt1988positivity}, where $\alpha$, $\beta$, $\gamma$, and $\delta$ are defined in terms of values and derivatives of $f$ at $x_0$ and $x_1$.
-Step \ref{alg1-alpha} checks for the necessary condition that $\alpha \ge 0$, Step \ref{alg1-beta} checks $\beta \geq \alpha$, and Step \ref{alg1-gamma} checks $\gamma \geq \delta$, all from \cite{schmidt1988positivity}.
-If all necessary conditions are met, then the order six piece is monotone and Step \ref{alg1-simplified-true} concludes this check.}
-  if $\big(Df(x_0) = 0$ or $Df(x_1) = 0\big)$ \newline
-  \indented{$w := x_0 - x_1$} \newline
-  \indented{$v := f(x_0) - f(x_1)$} \newline
-  \indented{if $\big(D^2f(x_1) > {-4}Df(x_1) / w \big)$ return FALSE} \newline
-  \indented{if $\big(D^2f(x_1) < (3w D^2f(x_0) - 24 Df(x_0) - 32
-    Df(x_1) + 60v/w) / (5w) \big)$ return FALSE} \newline
-  \indented{if $\big(D^2f(x_0) < 3 Df(x_0) / w \big)$ return FALSE} \newline
-  \indented{return TRUE} \newline
-    end if \newline
-    \comment{The following code considers the remaining case where $Df(x_0) \neq 0$ and $Df(x_1) \neq 0$.} \newline
-  $\displaystyle A := Df(x_0){x_1 - x_0 \over f(x_1) - f(x_0)}$ \newline
-  $\displaystyle B := Df(x_1) {x_1 - x_0 \over f(x_1) - f(x_0)}$ \newline
-    \comment{The variables $A$ and $B$ correspond directly to the theoretical foundation for positive quartic polynomials established in \cite{ulrich1994positivity}, first defined after Equation (18).} \newline
-  $\displaystyle \gamma_0 := 4 {Df(x_0) \over Df(x_1)} (B/A)^{3/4}$ \newline
-  $\displaystyle \gamma_1 := {x_1 - x_0 \over Df(x_1)} (B/A)^{3/4}$ \newline
-  $\alpha_0 := 4 (B/A)^{1/4}$ \newline
-  $\displaystyle \alpha_1 := -{x_1 - x_0 \over Df(x_1)} (B/A)^{1/4}$ \newline 
-  $\displaystyle \beta_0 := 30 - {12 \big(Df(x_0) + Df(x_1)\big) (x_1 - x_0) \over \big(f(x_1) - f(x_0)\big) \sqrt{A}\sqrt{B}}$ \newline
-  $\displaystyle \beta_1 := {-3 (x_1 - x_0)^2 \over 2 \big(f(x_1) - f(x_0)\big) \sqrt{A} \sqrt{B}} $ \newline
-    \comment{The $\gamma$, $\alpha$, and $\beta$ terms with subscripts $0$ and $1$ are algebraic reductions of the simplified conditions for satisfying Theorem 2 in Equation (16) of \cite{ulrich1994positivity}.
-These terms with subscripts $0$ and $1$ make the computation of $\alpha$, $\beta$, and $\gamma$ functions of the second derivative terms, as seen in Step \ref{alg1-abg} below.} \newline
-  $\gamma := \gamma_0 + \gamma_1 D^2f(x_0)$ \newline
-  $\alpha := \alpha_0 + \alpha_1 D^2f(x_1)$ \newline
-  $\beta := \beta_0 + \beta_1 \big(D^2f(x_0) - D^2f(x_1)\big)$ \newline
-  if $(\beta \leq 6)$ return $\big(\alpha > - (\beta + 2) / 2\big)$ \newline
-  else $\qquad$ return $\big(\gamma > -2 \sqrt{\beta - 2}\big)$ \newline
-  end if \newline
-}
-\endnumberedlist
-\vskip 2mm
-\hrule
-\vskip 3mm
-\parindent=10pt
-
-%% ----------------------------------------------------------------------
-
-The reason for structuring the $\alpha$, $\beta$, and $\gamma$ computations in terms of the second derivative of $f$ (seen in Step \ref{alg1-abg} of Algorithm \ref{alg:check-monotone}) will become more apparent later.
-The next problem to consider is that of making a nonmonotone order six polynomial piece into a monotone one by modifying its first and second derivative values at the ends of an interval.
-Note that the actual value of the function at the ends of the interval is not modified, as the resulting polynomial needs to interpolate.
-
-Notice that both Algorithms \ref{alg:check-monotone} and \ref{alg:make-monotone} have ${\cal O}(1)$ runtime assuming a fixed level of precision is chosen beforehand.
-A constant number of operations are needed for verifying monotonicity (Algorithm \ref{alg:check-monotone}), while a constant set of operations and a single binary search are performed for enforcing monotonicity (Algorithm \ref{alg:make-monotone}).
-The search requires a fixed number of steps to achieve any predetermined relative precision, since its accuracy is predetermined and not a function of the problem at hand.
-Also note that an efficient implementation of Algorithm \ref{alg:make-monotone} only needs to recalculate Steps \ref{alg1-abg} and \ref{alg1-last} of Algorithm \ref{alg:check-monotone} during the binary search.
-Next, an algorithm for constructing a monotone quintic spline interpolant is presented.
-
-%% ----------------------------------------------------------------------
-
-\vskip 5mm
-\thickrule
-{\bf Algorithm 2:} {\tt monotone\_spline}$\big((k_1$, $\ldots$, $k_n)$, $f$, $s\big)$
-where $(k_1$, $\ldots$, $k_n)$ is an increasing sequence of real
-numbers, $f$ is an order six piecewise polynomial with breakpoints
-$k_1$, $\ldots$, $k_n$ defined by the data
-$\bigl\{f(k_i)\bigr\}_{i=1}^n$, $\bigl\{Df(k_i)\bigr\}_{i=1}^n$,
-$\bigl\{D^2f(k_i)\bigr\}_{i=1}^n$, and $s > 1$ is an integer shrink
-factor.
-\hrule
-\vskip 3mm
-    \newline create queue
-      \newline create changed$(1$, $\ldots$, $n) := $ FALSE
-      \newline create step\_size$(1$, $\ldots$, $n) := \bigl(Df(k_1)/s$, $\ldots$, $Df(k_n)/s\bigr)$
-      \comment{The three variables defined in Step \ref{step:create} are used to ensure the eventual achievement of monotonicity.
-The {\tt queue} is a standard first in first out (FIFO) queue with {\tt enqueue} and {\tt dequeue} operations.
-The array {\tt changed} contains Booleans describing whether or not a breakpoint belongs to an interval that has been modified to enforce monotonicity.
-The {\tt step\_size} array contains the real-valued derivative decrement step sizes to use in the search for a valid monotone spline.}
-    \newline for $i := 1$, $\ldots$, $n-1$
-      \indented{enqueue $\bigl((k_i, k_{i+1})\bigr)$}
-      end for
-      \comment{Initially, all intervals must be checked for monotonicity. The following loop will continue until all intervals are verified as monotone without need for modification.}
-    \newline while $($size$($queue$) > 0)$
-    \newline \indented{$(k_j$, $k_{j+1}) := $ dequeue}
-    \newline \indented{if $\big($not is\_monotone$(k_j$, $k_{j+1}$, $f)\big)$}
-    \newline \indented{\indented{if $\big($changed$(j)\big) \qquad \ Df(k_j) := Df(k_j) -{}$step\_size$(j)$}}
-      \indented{\indented{if $\big($changed$(j+1)\big) \quad Df(k_{j+1}) := Df(k_{j+1}) -{}$step\_size$(j+1)$}}
-      \indented{\indented{\comment{If a breakpoint belongs to an interval that has been previously modified, then the enforced monotonicity conditions on the second derivatives of adjacent intervals must contradict one another.
-In turn, the involved first derivatives are decreased towards zero by a predetermined step size.}}}
-    \newline \indented{\indented{make\_monotone$(k_j,\ k_{j+1},\ f)$}}
-    \newline \indented{\indented{changed$(j) :=$ TRUE}}
-      \indented{\indented{changed$(j+1) :=$ TRUE}}
-    \newline \indented{\indented{if $\big(j > 1$ and $(k_{j-1}, k_j) \not\in$ queue$\big)$ enqueue $\bigl((k_{j-1}, k_j)\bigr)$}}
-      \indented{\indented{enqueue $\bigl((k_j, k_{j+1})\bigr)$}}
-      \indented{\indented{if $\big(j+1 < n$ and $(k_{j+1}, k_{j+2}) \not\in$ queue$\big)$ enqueue $\bigl((k_{j+1}, k_{j+2})\bigr)$}}
-      \indented{\indented{\comment{Step \ref{step:changed} records the endpoints of the current interval as having been changed, while \ref{step:requeue} adds adjacent intervals to {\tt queue} that may have inadvertently been made nonmonotone by the changes to the present interval.}}}
-      \indented{end if}
-      end while
-\vskip 2mm
-\hrule
-\vskip 3mm
-
-%% ----------------------------------------------------------------------
-
-It is mentioned in \cite{ulrich1994positivity} that for sufficiently small $Df(k_i)$ and $Df(k_{i+1})$ the admissible solution interval of second derivative values becomes arbitrarily large.
-It can also be seen that decreasing the assigned derivative values towards zero will eventually cause Steps \ref{alg1-alpha} through \ref{alg1-gamma} of Algorithm \ref{alg:check-monotone} to all fail, resulting in Step \ref{alg1-simplified-true} returning {\tt TRUE}.
-
-If successive monotonicity fixes are required for all intervals, the worst case runtime of Algorithm \ref{alg:monotone-spline} is ${\cal O}(s n)$ for $n$ breakpoints and shrink factor $s$.
-In practice this worst case has been observed to be unlikely.
-
-\heading{3. EXPERIMENTAL RESULTS}
-
-Given the increased order of approximation, it is naturally expected that some accuracy benefit could be gained by using a quintic spline over a cubic spline.
-Experiments 1 and 2 below test some accuracy differences between monotone cubic and quintic splines.
-Experiment 3 analyzes the number of operations performed by Algorithm \ref{alg:monotone-spline} with increasingly large samples of monotone data.
-
-\heading{3.1 \enspace Approximating a Trigonometric Function}
-
-For this experiment, the function $\sin(x) + x$ is considered over the interval $[0,(5/2)\pi]$ as seen in Figure \ref{fig:cubic_quintic_sin}.
-Given an increasing number of points, the maximum error of cubic and quintic approximations is shown in Figure \ref{fig:experiment_1}.
-The quintic approximation consistently has a maximum error that is roughly half that of the cubic approximation, given the same number of points.
-There is an increase in the error gap between the two approximations as more points are added.
-The quintic spline requires only $700$ points to approximate the function to the same accuracy achieved by the cubic with $1000$ points.
-
-%% \begin{figure}[h]
-%%   \centering
-%%   \includegraphics[width=.7\textwidth]{cubic-quintic-sin}
-%%   \caption{Depicted above are the monotone cubic and quintic spline interpolants of the function $\sin(x) + x$ over the points $[0, (5/2) \pi]$.
-Notice that the maximum error of the cubic interpolant is larger, because it only captures first derivative information at the points.
-The quintic interpolant captures both first and second derivative information at the points.}
-%% \end{figure}
-
-%% \begin{figure}
-%%   \centering
-%%   \includegraphics[width=.7\textwidth]{experiment_1_errors}
-%%   \caption{The error in the quintic and cubic approximations of $\sin(x) + x$ with increasing number of points.
-Both axes are log scaled, where $\mu$ means $\times 10^{-6}$ and n means $\times 10^{-9}$ on the vertical axis.
-The quintic approximation generally provides an approximation with about half of the maximum error of the cubic approximation, while the relative gap between grows with increasing number of points.
-In order to achieve the same absolute error the quintic spline requires ten fewer points at max error $10^{-4}$, thirty fewer points at max error $10^{-6}$, and $300$ fewer points at max error $10^{-8}$.}
-%% \end{figure}
-
-\heading{3.2 \enspace Random Number Generation}
-
-Consider the cumulative distribution function (CDF) defined by a mixture of three Gaussian distributions and visualized in Figure \ref{fig:experiment_2_distribution}.
-The distributions have weights $(.3, .6, .1)$, means $(.2, .45, .85)$ and standard deviations $(.05, .08, .03)$.
-$200$ random samples are generated from this distribution with both cubic and quintic approximations from four points, and over $100$ trials the resulting spread of empirical distribution functions (EDFs) are visualized in Figure \ref{fig:experiment_2_results}.
-
-While an increased number of points could be used to decrease the error in either approximation, the purpose of this experiment is to demonstrate the effect approximation error has on random number generation.
-The errors in the CDFs are magnified by the variance involved in randomly sampling from a distribution.
-As a result, while the cubic and quintic spline approximations have similar maximum error, the cubic distribution is significantly distorted around the first and second modes.
-
-%% \begin{figure}
-%%   \centering
-%%   \includegraphics[width=.8\textwidth]{experiment_2_distribution}
-%%   \caption{The cumulative distribution function (top) and probability density function (bottom) for the Gaussian mixture distribution in Experiment 2.
-The distributions have weights $(.3, .6, .1)$, means $(.2, .45, .85)$ and standard deviations $(.05, .08, .03)$.
-Random samples are generated from this mixture distribution by evaluating the inverse of the CDF at random numbers uniformly distributed in the range $[0,1]$.}
-%% \end{figure}
-
-%% \begin{figure}
-%%   \centering
-%%   \includegraphics[width=.4\textwidth]{experiment_2_cubic_dist}
-%%   \includegraphics[width=.4\textwidth]{experiment_2_quintic_dist}
-%%   \caption{Both cubic and quintic splines are used to approximate the CDF of a Gaussian mixture distribution with three components.
-Four CDF points are used to build the approximation, $200$ random samples are generated with the approximate CDFs over $100$ trials.
-The resulting cloud of empirical distribution function (EDF) values is seen above.
-The maximum approximation error in the CDFs are similar, $.08$ for the cubic and $.05$ for the quintic.
-However, the maximum EDF differences observed are doubled by the variance of random samples, $.15$ for the cubic and $.1$ for the quintic.}
-%% \end{figure}
-
-%% - generate large sets of random monotone data, show Table of number
-%%   of metrics growing with increasing number of points
-
-%% - show visual of the \textit{error}
-
-\heading{3.3 \enspace Random Monotone Data}
-
-This experiment studies the number of times the algorithms {\tt is\_monotone} and {\tt make\_monotone} are executed for increasingly large sequences of random monotone data.
-The minimum, median, and maximum number of times that Algorithms \ref{alg:check-monotone} and \ref{alg:make-monotone} are executed is recorded, as well as the number of steps taken in all calls to {\tt binary\_search}.
-The number of calls and steps grows linearly with $n$ as expected, requiring roughly one call to {\tt make\_monotone} and $20$ binary search steps to create a monotone piece.
-Some (less common) problems require an average of three calls to {\tt make\_monotone} per interval.
-
-%% \begin{table}
-%%   \centering
-%%   \caption{The minimum, median, and maximum number of checks, fixes, and binary search steps required in the execution of {\tt make\_spline} for increasing size sequences, $n$, over $100$ randomly generated sets of monotone data.
-Notice the maximum for each counter is often significantly greater than the minimum and median, because the distribution of each counter is skewed right.}
-
-%%   \begin{tabular}{c|c c c|c c c|c c c}
-%%     \hline
-%%     \multirow{2}{*}{$n$}
-%%       & \multicolumn{3}{c|}{Checks} & \multicolumn{3}{c}{Fixes} & \multicolumn{3}{|c}{Search Steps} \\
-%%       & min & median & max & min & median & max & min & median & max \\
-%%     \hline
-%%     5 & 4 & 6 & 274 & 0 & 1 & 154 & 3 & 31 & 2006\\
-%%     25 & 28 & 47 & 378 & 2 & 14 & 225 & 80 & 348 & 3742\\
-%%     50 & 61 & 124 & 484 & 6 & 40 & 238 & 217 & 851 & 4445\\
-%%     75 & 102 & 216 & 842 & 14 & 74 & 411 & 463 & 1491 & 6333\\
-%%     100 & 142 & 306 & 776 & 21 & 107 & 380 & 629 & 2140 & 7414\\
-%%     \hline
-%%   \end{tabular}
-%% \end{table}
-
-\heading{4. DISCUSSION}
-
-The monotone quintic spline interpolant provides a distinct increase in accuracy over the monotone cubic variant.
-The computational complexity of this algorithm is identical to the existing cubic method and the cost of evaluating the spline after construction is unchanged.
-While the number of computations required per interval has increased, the number of points required to achieve the same level of accuracy has decreased.
-In the present state, the algorithm for enforcing monotonicity on a spline is not as trivially parallel as the cubic algorithm, however it can still be parallelized.
-The checks for monotonicity across all intervals are independent and the monotonicity adjustments could be computed independently and intelligently merged with minor changes to the serial algorithm.
-
-Acknowledging that the algorithm for constructing a monotone quintic spline interpolant is slightly more expensive than the cubic case, gains in accuracy or decreased necessary number of points are often worth the computational effort.
-In practice the costs of spline interpolation are often dominated by evaluation rather than construction, and the increased accuracy is afforded for a negligible increase in computational cost.
-
-\heading{5. CONCLUSION AND FUTURE WORK}
-
-This paper proposes and tests an algorithm for constructing monotone quintic spline interpolants.
-Experiments demonstrate an improvement in approximation accuracy over monotone cubic spline interpolants, as expected based on theory.
-There are still open avenues of research going forward, such as an alternative sufficient condition for enforcing monotonicity or increased order monotone approximations.
-If the monotonicity conditions can be generalized to any order or made linear, the search for a monotone interpolating spline could potentially be formulated as a convex optimization problem.
-Finally, this work could be used to improve cumulative distribution function (CDF) estimates as well as predictive models that use CDFs.
-
-\heading{Acknowledgments}
-This work was supported by the National Science Foundation Grants CNS-1565314 and CNS-1838271.
-
-\heading{\hlvVIII BIBLIOGRAPHY} %% {\let\caps=\capsVIII \let\sl=\slVIII
-%% \def\ys{\hskip 1em minus .5em}\let\rm=\rmeight \rmVIII
-%% \newcount\refnum \refnum=0 \parskip=2pt plus 2pt minus 2pt
-%% \frenchspacing
-
-\bibliography{mqsi20}
-\bibliographystyle{plain}
-
-\vfill\eject\end
-
-%% %\refj is for regular journal articles
-%% \def\refj#1#2#3#4#5{\noindent\hangindent=10pt\hangafter=1
-%% %\global\advance\refnum by 1[\number\refnum]
-%% {\caps #1}\ys #2.\ys {\rm #3}. {\sl #4}, #5.\par}
-
-%% %\refb is for books
-%% \def\refb#1#2#3#4{\noindent\hangindent=10pt\hangafter=1
-%% %\global\advance\refnum by 1[\number\refnum]
-%% {\caps #1}\ys #2.\ys{\sl #3}. {\rm #4}.\par}
-
-%% %\refp is for private communication
-%% \def\refp#1#2#3{\noindent\hangindent=10pt\hangafter=1
-%% %\global\advance\refnum by 1[\number\refnum]
-%% {\caps #1}\ys #2.\ys{\rm #3}.\par}
-
-%% %\reft is for tech. reports, theses, and conferences
-%% \def\reft#1#2#3#4{\noindent\hangindent=10pt\hangafter=1
-%% %\global\advance\refnum by 1[\number\refnum]
-%% {\caps #1}\ys #2.\ys{\rm #3}. {\rm #4}.\par}
-
-%% %%%%% top of refs
-%% \reft{Amos, B. D., Easterling, D. R., Watson, L. T., Thacker, W. I.,
-%% Castle, B. S., Trosset, M. W.}{2014}{Algorithm XXX: QNSTOP: Quasi-Newton
-%% algorithm for stochastic optimization}{Technical Report 2014-07,
-%% Virginia Polytechnic Institute and State University, Blacksburg, VA}
-
-%% \refb{Anderson, E., Bai, Z., Bischof, C., Blackford, S., Demmel, J.,
-%% Dongarra, J., Du Croz, J., Greenbaum, A., Hammarling, S., McKenney, A.,
-%% and Sorensen, D.}{1999}{LAPACK Users' Guide Third Edition}
-%% {SIAM, Philidelphia, PA}
-
-%% \refb{Aurenhammer, F., Klein, R., and Lee, D. T.}{2013}
-%% {Voronoi diagrams and Delaunay triangulations}
-%% {World Scientific Publishing Co., Hackensack, NJ}
-
-%% \refj{Barber, C. B., Dobkin, D. P., and Huhdanpaa, H.}{1996}
-%% {The Quickhull algorithm for convex hulls}{ACM Trans. Math.
-%% Softw. 22}{469--483}
-
-%% \reft{Boissonnat, J.-D., Devillers, O., and Hornus, S.}{2009}
-%% {Incremental construction of the Delaunay triangulation and the
-%% Delaunay graph in medium dimension}{In {\sl Proc. Twenty-fifth Annual Symp.
-%% on Computational Geometry}, ACM, New York, NY, 208--216}
-
-%% \refj{Bowyer, A.}{1981}{Computing Dirichlet tessellations}
-%% {The Computer Journal 24}{162--166}
-
-%% \refj{Cameron, K. W., Anwar, A., Cheng, Y., Xu, L., Li, B., Ananth, U.,
-%% Bernard, J., Jearls, C., Lux, T., Hong, Y., Watson, L. T.,
-%% and Butt, A. R.}{2019}{MOANA: modeling and analyzing I/O variability
-%% in parallel system experimental design}
-%% {IEEE Trans. Parallel Distrib. Systems}{to appear}
-
-%% \reft{Chang, T. H., Watson, L. T., Lux, T. C. H., Bernard, J., Li, B.,
-%% Xu, L., Back, G., Butt, A. R., Cameron, K. W., and Hong, Y.}{2018a}
-%% {Predicting system performance by interpolation using a high-dimensional
-%% Delaunay triangulation} {In {\sl Proc. 2018 Spring Simulation Multiconference,
-%% 26th High Performance Computing Symp.}, Soc. for Modelling and Simulation
-%% Internat., Vista, CA, Article No. 2}
-
-%% \reft{Chang, T. H., Watson, L. T., Lux, T. C. H., Li, B., Xu, L., 
-%% Butt, A. R., Cameron, K. W., and Hong, Y.}{2018b}{A polynomial time
-%% algorithm for multivariate interpolation in arbitrary dimension via
-%% the Delaunay triangulation} {In {\sl Proc. ACM 2018 Southeast Conference
-%% (ACMSE '18)}, ACM, New York, NY, Article No. 12}
-
-%% \refb{Cheney, E. W. and Light, W. A.}{2009}
-%% {A Course in Approximation Theory}{American Mathematical Soc., 
-%% Providence, RI}
-
-%% \refb{Cheng, S. W., Dey, T. K., and Shewchuk, J.}{2012}
-%% {Delaunay Mesh Generation}{Computer and Information Science
-%% Series, CRC Press, New York, NY}
-
-%% \refj{Cignoni, P., Montani, C., and Scopigno, R.}{1998}
-%% {DeWall: A fast divide \& conquer Delaunay triangulation algorithm in $E^d$}
-%% {Computer-Aided Design 30}{333--341}
-
-%% \refb{de Berg, M., Cheong, O., van Kreveld, M., and Overmars, M.}{2008}
-%% {Computational Geometry: Algorithms and Applications}{Springer-Verlag,
-%% Berlin, Germany}
-
-%% \refj{Devillers, O., Pion, S., and Teillaud, M.}{2002}
-%% {Walking in a triangulation}{Internat. Journal of Foundations of
-%% Computer Science 13}{181--199}
-
-%% \reft{Edelsbrunner, H.}{1989}{An acyclicity theorem for cell complexes in
-%% d dimensions} {In {\sl Proc. Fifth Annual Symp. on Computational Geometry}, 
-%% ACM, New York, 145--151}
-
-%% \refj{Hanson, R. J. and Haskell, K. H.}{1982}
-%% {Algorithm 587: Two algorithms for the linearly constrained least
-%% squares problem}{ACM Trans. Math. Softw. 8}{323--333}
-
-%% \refj{Klee, V.}{1980}{On the complexity of d-dimensional Voronoi diagrams}
-%% {Archiv der Mathematik 34}{75--80}
-
-%% \reft{Lux, T. C. H., Watson, L. T., Chang, T. H., Bernard, J., Li, B.,
-%% Xu, L., Back, G., Butt, A. R., Cameron, K. W., and Hong, Y.}{2018}
-%% {Predictive modeling of I/O characteristics in high performance computing
-%% systems}{In {\sl Proc. 2018 Spring Simulation Multiconference, 26th High 
-%% Performance Computing Symp.}, Soc. for Modelling and Simulation Internat.,
-%% Vista, CA, Article No. 8}
-
-%% \refj{M{\"u}cke, E. P., Saias, I., and Zhu, B.}{1999}
-%% {Fast randomized point location without preprocessing in two- and 
-%% three-dimensional Delaunay triangulations}
-%% {Computational Geometry 12}{63--83}
-
-%% \refj{Omohundro, S. M.}{1990}{Geometric learning algorithms}
-%% {Physica D: Nonlinear Phenomena 42}{307--321}
-
-%% \reft{OpenMP Architecture Review Board (ARB)}{2015}
-%% {OpenMP Application Programming Interface (version 4.5)}
-%% {OpenMP}
-
-%% \refj{Rajan, V. T.}{1994}
-%% {Optimality of the Delaunay triangulation in $R^d$}
-%% {Discrete \& Computational Geometry 12}{189--202}
-
-%% \refj{Schaap, W. E. and van de Weygaert, R.}{2000}
-%% {Continuous fields and discrete samples: reconstruction through
-%% Delaunay tessellations}{Astronomy and Astrophysics 363}{L29--L32}
-
-%% \refj{Su, P. and Drysdale, R. L. S.}{1997}
-%% {A comparison of sequential Delaunay triangulation algorithms}
-%% {Computational Geometry 7}{361--385}
-
-%% \refj{Watson, D. F.}{1981}{Computing the n-dimensional Delaunay tessellation
-%% with application to Voronoi polytopes}
-%% {The Computer Journal 24}{167--172}
-
-%% }\vfill\eject\end
-%% %%%%% bottom

### 2020-06-21 Research_[2020-05-18]_Piecewise_Monotone_Quintic_Fortran/old/example_tex/old_files/0-experimental.tex
Working through revisions given by Thacker. Sent next draft to Watson asking for some guidance.

--- 2020-06-12 Research_[2020-05-18]_Piecewise_Monotone_Quintic_Fortran/paper/mqsi20.tex
@@
-% Algorithm XXX: MQSI---Monotone quintic spline interpolation,
-% Thomas C.H. Lux, Layne T. Watson, Tyler H. Chang,
-% William Thacker, Kirk W. Cameron, and Yili Hong
-% ACM Trans. Math. Software, ?/?/2020.
-% 
-%
-\magnification=1000 \hsize=30pc \vsize=47pc
-\parindent=10pt \parskip=0pt plus 3pt
-\baselineskip=13pt plus 2pt minus 1pt \lineskiplimit=2pt
-\lineskip=2pt plus 2pt                                  %\raggedright
-\tolerance=800
-%\abovedisplayskip=6pt plus 6pt minus 3pt
-%\abovedisplayshortskip=0pt plus 3pt
-%\belowdisplayskip=6pt plus 6pt minus 3pt
-%\belowdisplayshortskip=6pt plus 2pt minus 3pt
-
-% eplain package, mainly for its bibliography management.
-\input eplain
-
-% Color definitions.
-\input colordvi
-\def\BEGINC{\textPeach} \def\ENDC{\textBlack}
-\def\red#1{\BEGINC#1\ENDC}
-\let\beginred=\BEGINC
-\let\endred=\ENDC
-
-% Commands for inserting an 'algorithm' section.
-\def\indented#1{{\leftskip=1cm}{#1}{\leftskip=0cm}}
-\def\comment#1{\vskip 10mm {\it #1} \vskip 10mm}%
-\def\thickrule{\hrule\hrule\hrule}
-
-% fonts
-\font\itVIII=cmti8 \font\slVIII=cmsl8 \font\bfVIII=cmbx8
-\font\ttVIII=cmtt8 \font\caps=cmcsc10 \font\capsVIII=cmcsc10 at 8pt
-\font\ttXVIII=cmtt12 at 18pt \font\eightmi=cmmi8 \font\eightsy=cmsy8
-\font\rmeight=cmr8  \def\rmVIII{\rmeight \baselineskip=10pt plus 2pt
-minus 1pt \lineskiplimit=2pt \lineskip=2pt plus 1pt
-\parskip=0pt \textfont0=\rmeight \textfont1=\eightmi \textfont2=\eightsy}
-\font\hlvXVIII=phvr at 18pt \font\hlvVIII=phvr at 8pt \font\hlv=phvr
-
-% math blackboard font
-\font\tenamsb=msbm10 \font\sevenamsb=msbm7 \font\fiveamsb=msbm5
-\newfam\bbfam
-\textfont\bbfam=\tenamsb
-\scriptfont\bbfam=\sevenamsb
-\scriptscriptfont\bbfam=\fiveamsb
-\def\bbb{\fam\bbfam}
-
-\input epsf %Postscript figures.
-\def\lcaption#1#2{{\noindent\rmVIII Fig.~#1. #2\par}}
-\def\ltabcaption#1#2{{\noindent\rmVIII Table #1. #2\par}}
-
-% other definitions.
-\def\h{\par\hangindent=10pt\hangafter=1\noindent}
-\def\heading#1{\par\bigskip\leftline{\hlv#1}\nobreak\smallskip
-\noindent\ignorespaces}
-\newdimen\headdigit{\setbox11=\hbox{\hlv 8. }\global\headdigit=\wd11}
-\def\footnoterule{\bigskip\kern-3pt \hrule \kern 2.6pt \smallskip}
-\def\bullitem{\par\vskip 2pt plus 2pt\hangindent=15pt \hangafter=1
-\noindent\hbox to 15pt{---\hfil}\ignorespaces}
-\def\bulllitem{\par\vskip 2pt plus 2pt\hangindent=30pt
-\hangafter=1 \noindent\hskip 15pt\hbox to
-15pt{---\hfil}\ignorespaces}
-\def\Th#1 #2\par{\par\medskip{\it{\caps#1}. #2}\par\medskip}
-\def\btable#1{\topinsert \rmVIII \let\bf=\bfVIII \textfont0=\scriptfont0
-\scriptfont0=\scriptscriptfont0 \textfont1=\scriptfont1
-\scriptfont1=\scriptscriptfont1 \centerline{#1}}
-\let\xpar=\par
-\def\ds{\displaystyle}\def\ab{\allowbreak}
-
-% math definitions
-\def\tn#1{\left\Vert #1\right\Vert_2}
-\def\mn#1{\left\Vert #1\right\Vert_\infty}
-\def\lee{\mathrel{\vcenter{\hbox{$\scriptstyle\mathord<$}\nointerlineskip
-\vskip 1pt\hbox{$\scriptstyle\mathord=$}}}}
-\def\gee{\mathrel{\vcenter{\hbox{$\scriptstyle\mathord>$}\nointerlineskip
-\vskip 1pt\hbox{$\scriptstyle\mathord=$}}}}
-\def\real{\mathop{\rm I\!R}\nolimits}
-\def\R{{\bf R}}
-\def\Rd{{\bf R}^d}
-\def\Rdd{{\bf R}^{d\times d}}
-%\def\bull{\hfill \vrule height6pt width6pt depth0pt}
-\def\:{\mathrel{:\mathord=}}
-
-% OUTPUT
-\def\rightheadline{\hfill\hlvVIII\rightrh\hskip20pt$\scriptstyle\bullet$\hskip
-20pt\folio}
-\def\leftheadline{\hlvVIII\folio\hskip20pt$\scriptstyle\bullet$\hskip
-20pt\leftrh\hfill}
-\footline={\hfill}%no page number on first page
-\output{\plainoutput}
-\def\plainoutput{\shipout\vbox{\makeheadline\pagebody\makefootline}
-  \advancepageno \global\footline={\hfill}
-  \ifodd\pageno
-        \global\headline={\rightheadline}
-  \else
-        \global\headline={\leftheadline}
-  \fi
-  \ifnum\outputpenalty>-20000 \else\dosupereject\fi}
-\def\pagebody{\vbox to\vsize{\boxmaxdepth\maxdepth \pagecontents}}
-\def\makeheadline{\vbox to 0pt{\vskip-22.5pt
-  \line{\vbox to8.5pt{}\the\headline}\vss}\nointerlineskip}
-\def\makefootline{\baselineskip24pt\vskip-6pt\line{\the\footline}}
-\def\dosupereject{\ifnum\insertpenalties>0pt
-  \line{}\kern-\topskip\nobreak\vfill\supereject\fi}
-%\def\footnoterule{\vskip10pt\kern-3pt \hrule width 3em \kern 2.6pt}
-
-\def\leftrh{Lux et al.}
-\def\rightrh{Algorithm XXX: MQSI}
-
-{\baselineskip=24pt \leftline{\hlvXVIII Algorithm XXX: MQSI: Monotone Quintic}
-\leftline{\hlvXVIII Spline Interpolation}}
-\bigskip\medskip
-\leftline{\hlv THOMAS C.H. LUX, LAYNE T. WATSON, TYLER H. CHANG,}
-\smallskip
-\leftline{\hlv WILLIAM THACKER, KIRK W. CAMERON, AND YILI HONG}
-\smallskip
-\leftline{\hlv Virginia Polytechnic Institute and State University}
-\bigskip
-
-\hrule\bigskip\smallskip \footnote{}{\hskip
--\parindent{\parindent=0pt\rmVIII Authors' addresses:
-T. C. H. Lux, T. H. Chang, K. W. Cameron,
-Department of Computer Science, L. T. Watson,
-Departments of Computer Science, Mathematics, and Aerospace and Ocean
-Engineering, Y. Hong, Department of Statistics, Virginia
-Polytechnic Institute \& State University, Blacksburg, VA 24061;
-e-mail: {\ttVIII tchlux@vt.edu}. \xpar
-% Permission to make digital/hard copy of
-%part or all of this work for personal or classroom use is granted
-%without fee provided that the copies are not made or distributed for
-%profit or commercial advantage, the copyright notice, the title of
-%the publication, and its date appear, and notice is given that
-%copying is by permission of the ACM, Inc. To copy otherwise, to
-%republish, to post on servers, or to redistribute to lists, requires
-%specific permission and/or fee. \xpar\copyright\ 2018 by the
-%Association for Computing Machinery, Inc. \xpar 
-}}
-
-% Abstract and Introduction
-{\rmVIII\parindent=0pt MQSI contains a serial code written in Fortran
-  1995 for constructing monotone quintic spline interpolants to
-  data. Using sharp theoretical monotonicity constraints, first and
-  second derivative estimates at data provided by a quadratic facet
-  model are refined to produce a $C^2$ monotone interpolant. This
-  paper includes algorithm and implementation details, complexity and
-  sensitivity analyses, usage information, and a brief performance
-  study.
-
-\medskip
-Categories and Subject Descriptors: G.1.1 [{\bfVIII Numerical Analysis}]:
-Interpolation --- Spline and piecewise polynomial interpolation;
-J.2 [{\bfVIII Computer Applications}]: Physical Science and Engineering
---- {\itVIII Mathematics};
-G.4 [{\bfVIII Mathematics of Computing}]: Mathematical Software
-
-\medskip
-General Terms: Algorithms, Monotonicity, Documentation
-
-\medskip
-Additional Key Words and Phrases: Quintic Spline, interpolation
-
-}
-
-\bigskip\hrule\bigskip\medskip
-\heading{1. INTRODUCTION}
-
-Many domains of science rely on smooth approximations to real-valued
-functions over a closed interval. Piecewise polynomial functions
-(splines) provide the smooth approximations for animation in graphics
-\cite{herman2015techniques,quint2003scalable}, aesthetic structural
-support in architecture \cite{brennan2019measure}, efficient
-aerodynamic surfaces in automotive and aerospace engineering
-\cite{brennan2019measure}, prolonged effective operation of electric
-motors \cite{berglund2009planning}, and accurate nonparametric
-approximations in statistics \cite{knott2012interpolating}. While
-polynomial interpolants and regressors apply broadly, splines are
-often a good choice because they can approximate globally complex
-functions while minimizing the local complexity of an approximation.
-
-It is often the case that the true underlying function or phenomenon
-being modeled has known properties e.g., convexity, positivity,
-various levels of continuity, or monotonicity. Given a reasonable
-amount of data, it quickly becomes difficult to achieve desirable
-properties in a single polynomial function. In general, the
-maintenance of function properties through interpolation/regression is
-referred to as {\it shape preserving}
-\cite{fritsch1980monotone,gregory1985shape}. The specific shapes this
-work will achieve in approximations are monotonicity and $C^2$
-continuity. These properties are chiefly important to the
-approximation of cumulative distribution functions and subsequently
-the effective generation of random numbers from a specified
-distribution.
-
-In statistics especially, the construction of a monotone interpolating
-spline that is $C^2$ continuous is meaningfully useful
-\cite{ramsay1988monotone}. A function with these properties could
-approximate a cumulative distribution function to a high level of
-accuracy with relatively few intervals. A twice continuously
-differentiable approximation to a cumulative distribution function
-(CDF) would also produce a corresponding probability density function
-(PDF) that is continuously differentiable, which is a property many
-standard parametric distributions maintain.
-
-The currently available software for monotone piecewise polynomial
-interpolation includes quadratic \cite{he1998monotone}, cubic
-\cite{fritsch1980monotone}, and (with limited application) quartic
-\cite{wang2004rational,piah2011improved,yao2018unconditionally}
-cases. In addition, a statistical method for bootstrapping the
-construction of an arbitrarily smooth monotone fit exists
-\cite{leitenstorfer2006generalized}, but the method does not take
-advantage of known analytic properties related to quintic
-polynomials. Theory has been provided for the quintic case
-\cite{ulrich1994positivity,hess1994positive}, however this theory has
-not yet been used to construct a monotone quintic spline interpolation
-routine. Recent work suggests that the lack of quintic software may be
-due to a general unawareness of the theory
-\cite{xie2018semiparametric}.
-
-\topinsert
-\centerline{\epsfxsize=4.5truein \epsffile{vt_logo.eps}}
-{\narrower\noindent\rmVIII Fig.\ 1.
-Example polynomials that interpolate function values at the ends of
-the interval $[0,1]$. The first only interpolates the function values
-$g_2(0) = 1$ and $g_2(1) = 0$, making it the order two polynomial
-$g_2(x) = 1 - x$. For the second plot $g_4(x) = 8x^3 - 12x^2 + 4x +
-1/2$, which is order four and interpolates the values $g_4(0) = 1/2$,
-$g_4'(0) = 4$, $g_4(1) = 1/2$, $g_4'(1) = 4$. Finally the third plot
-shows the order six polynomial $g_6(x) = - 64x^5 + 160x^4 - 140x^3 +
-50x^2 - 7x + 1$ interpolating the function values $g_6(0) = 1$,
-$g_6'(0) = -7$, $g_6''(0) = 100$, $g_6(1) = 0$, $g_6'(1) = -7$,
-$g_6''(1) = -100$. Notice that interpolating the same fixed number of
-function values at each endpoint will always result in an even order
-interpolating polynomial.
-\par}
-\endinsert
-
-The importance of piecewise quintic interpolation over lower order
-approximations can be simply demonstrated. In general, the order of a
-polynomial determines the number of function values it can
-interpolate, and the growth rate of error away from the interpolated
-function values. As demonstrated in Figure
-\ref{fig:spline_demonstration}, it can be seen that matching a value
-at either end of the interval requires an order two (linear)
-approximation and each additional derivative at the ends of the
-interval raises the necessary polynomial order by two. Hence, only
-even order (odd degree) interpolating splines are broadly
-applicable. The body of this work is composed of a novel algorithm for
-enforcing monotonicity on quintic polynomial pieces, then extending
-that solution to work on quintic splines.
-
-The major contribution of this work is an algorithm for constructing
-monotone quintic interpolating splines that utilizes existing quintic
-monotonicity theory. The remainder of this paper is structured as
-follows: Section \ref{sec:monotone-cubic} summarizes the existing
-monotone cubic spline interpolation methodology, Section
-\ref{sec:monotone-quintic} presents an algorithm for constructing
-monotone quintic spline interpolants, Section \ref{sec:results} offers
-experiments and results with cubic and quintic methods, Sections
-\ref{sec:discussion} and \ref{sec:conclusion} discuss results and
-conclude.
-
-\heading{1.1 \enspace Computing a Monotone Cubic Interpolant}
-
-The current state-of-the-art monotone interpolating spline with a
-mathematical software implementation is piecewise cubic, continuously
-differentiable, and was first proposed in \cite{fritsch1980monotone}
-then expanded upon in \cite{carlson1985monotone}. Let $\pi: x_0 = k_1
-< k_2 < \cdots < k_n = x_1$ be a partition of the interval
-$[x_0,x_1]$. Let $f: {\bbb R} \rightarrow {\bbb R}$ be $C^2$, and
-$\bigl\{f(k_i)\bigr\}_{i=1}^n$ and $\bigl\{\Delta_i\bigr\}_{i=1}^n$ be
-given sets of function and derivative values at the partition points
-for a monotone function $f$. Either $f(k_i) \leq f(k_{i+1})$, $i=1$,
-$\ldots$, $n-1$, and $\Delta_i\ge0$, $i=1$, $\ldots$, $n$, or $f(k_i)
-\geq f(k_{i+1})$, $i=1$, $\ldots$, $n-1$, and $\Delta_i\le0$, $i=1$,
-$\ldots$, $n$. Let $\hat f$ be a piecewise cubic polynomial defined in
-each subinterval $I_i = [k_i, k_{i+1}]$ by
-$$ h_i = k_{i+1} - k_{i}, \quad u(t) = 3t^2 - 2t^3, \quad p(t) = t^3 - t^2,$$
-$$\hat f(x) = f(k_i) u\big((k_{i+1} - x) / h_i\big) + f(k_{i+1}) u\big((x - k_i) / h_i\big) $$
-$$ - h_i\Delta_i p\big((k_{i+1}-x)/h_i\big) + h_i\Delta_{i+1} p\big((x-k_i)/h_i\big).$$ 
-Notice that a trivially monotone spline results when $\Delta_i = 0$,
-for $i = 1$, $\ldots$, $n$. However, such a spline has too many
-{\it wiggles} for most applications. \cite{carlson1985monotone}
-show that simple conditions on the derivative values can guarantee
-monotonicity, and that these conditions can be enforced in a way that
-ensures modifications on one interval will not break the monotonicity
-of cubic polynomials over any neighboring intervals. If $f(k_i) =
-f(k_{i+1})$, take $\Delta_i = \Delta_{i+1} =0$ and $\alpha =\beta =1$,
-otherwise let $\alpha = \big(\Delta_i (k_{i+1}-k_i)\big) /
-\big(f(k_{i+1}) - f(k_i)\big)$ and $\beta =
-\big(\Delta_{i+1}(k_{i+1}-k_i)\big) / \big(f(k_{i+1}) -
-f(k_i)\big)$. Monotonicity of a cubic polynomial over a subinterval
-can be maintained by ensuring that $\alpha$ and $\beta$ reside in any
-of the regions depicted in Figure \ref{fig:projection}.
-
-\topinsert
-\centerline{\epsfxsize=4.5truein \epsffile{vt_logo.eps}}
-{\narrower\noindent\rmVIII Fig.\ 2.
-These are the feasible regions of monotonicity for cubic splines and
-the projections that make a cubic polynomial piece monotone. The
-regions themselves are numbered 1--4 corresponding to their original
-description in Fritsch and Carlson (1980). One point is projected onto
-each region as a demonstration.
-\par}
-\endinsert
-
-The actual region of monotonicity for a cubic polynomial is larger,
-but projection of $(\alpha, \beta)$ into one of these regions ensures
-that monotonicity will be achieved and not violated for neighboring
-regions. The user must decide which region is most appropriate for the
-projections based on the application, Fritsch and Carlson recommend
-using Region 2.
-
-While the cubic monotonicity case affords such a concise solution, the
-region of monotonicity is not so simple in the quintic case. In the
-next section, an algorithm for performing a projection similar to
-those for cubic polynomials is proposed.
-
-\heading{2. MONOTONE QUINTIC INTERPOLATION}
-
-The following section is composed of three algorithms that together
-are used to construct a monotone quintic spline interpolant. Without
-loss of generality, the algorithms will only consider the monotone
-increasing (nondecreasing) case. The monotone decreasing case is
-handled similarly. Algorithm \ref{alg:check-monotone} checks
-monotonicity, Algorithm \ref{alg:make-monotone} enforces monotonicity
-on an order six polynomial piece of the quintic spline, and Algorithm
-\ref{alg:monotone-spline} uses the previous two algorithms to enforce
-monotonicity for the entire quintic spline.
-
-In pseudocode, a function {\tt binary\_search($g,$ $a,$ $b$)} is
-used, where $a,b\in S\subset {\bbb R}^p$ for convex $S$, $g: S
-\rightarrow \{${\tt FALSE}, {\tt TRUE}$\}$ is a right continuous
-Boolean function, $g(b) = \hbox{{\tt TRUE}}$, and for $\mu \in
-[0,1]$, $g\big((1-\mu)a + \mu b\big) =\hbox{{\tt TRUE}}
-\Longrightarrow g\big((1-\nu)a + \nu b\big) =\hbox{{\tt TRUE}}$ for
-$\mu \le \nu\le1$. The search returns $\big((1-c)a + c b\big)$ for the
-smallest $c \in [0,1]$ such that $g\big((1-c)a + c b\big) =
-\hbox{{\tt TRUE}}$.
-
-These algorithms make use of first and second derivatives of the approximated function where $D$ denotes the differentiation operator.
-In the case that the first and second derivative information is not provided along with function values, the derivatives are estimated with finite differences of the function values.
-The final quintic spline is represented as a piecewise polynomial using the Newton form for each polynomial piece, and ultimately converted to a $B$-spline representation for evaluation.
-
-\def\newline{\hfil\break}
-
-\parindent=0pt
-\vskip 5mm
-\thickrule
-{\bf Algorithm 1:} {\tt is\_monotone}$\big(x_0, x_1, f \big)$ \newline
-where $x_0$, $x_1 \in {\bbb R}$, $x_0 < x_1$, and $f$ is an order six
-polynomial defined by $f(x_0)$, $Df(x_0)$, $D^2f(x_0)$, $f(x_1)$,
-$Df(x_1)$, $D^2f(x_1)$. Returns {\tt TRUE} if $f$ is monotone
-increasing on $[x_0,x_1]$.
-\hrule
-\vskip 3mm
-\numberedlist
-{\tt
-\li if $\big(f(x_0) = f(x_1)\big)$ and not $\big( 0 = Df(x_0) = Df(x_1) = D^2f(x_0) = D^2f(x_1) \big)$
-\li \indented{return FALSE} \newline
-end if \newline
-\li if $\big(Df(x_0) < 0$ or $Df(x_1) < 0\big)$ \newline
-    \indented{return FALSE} \newline
-    end if \newline
-    \comment{The necessity of these first two steps follows directly from the fact that $f$ is $C^2$.
-The next case is in accordance with a simplified condition for quintic monotonicity that reduces to one of cubic positivity studied in \cite{schmidt1988positivity}, where $\alpha$, $\beta$, $\gamma$, and $\delta$ are defined in terms of values and derivatives of $f$ at $x_0$ and $x_1$.
-Step \ref{alg1-alpha} checks for the necessary condition that $\alpha \ge 0$, Step \ref{alg1-beta} checks $\beta \geq \alpha$, and Step \ref{alg1-gamma} checks $\gamma \geq \delta$, all from \cite{schmidt1988positivity}.
-If all necessary conditions are met, then the order six piece is monotone and Step \ref{alg1-simplified-true} concludes this check.}
-  if $\big(Df(x_0) = 0$ or $Df(x_1) = 0\big)$ \newline
-  \indented{$w := x_0 - x_1$} \newline
-  \indented{$v := f(x_0) - f(x_1)$} \newline
-  \indented{if $\big(D^2f(x_1) > {-4}Df(x_1) / w \big)$ return FALSE} \newline
-  \indented{if $\big(D^2f(x_1) < (3w D^2f(x_0) - 24 Df(x_0) - 32
-    Df(x_1) + 60v/w) / (5w) \big)$ return FALSE} \newline
-  \indented{if $\big(D^2f(x_0) < 3 Df(x_0) / w \big)$ return FALSE} \newline
-  \indented{return TRUE} \newline
-    end if \newline
-    \comment{The following code considers the remaining case where $Df(x_0) \neq 0$ and $Df(x_1) \neq 0$.} \newline
-  $\displaystyle A := Df(x_0){x_1 - x_0 \over f(x_1) - f(x_0)}$ \newline
-  $\displaystyle B := Df(x_1) {x_1 - x_0 \over f(x_1) - f(x_0)}$ \newline
-    \comment{The variables $A$ and $B$ correspond directly to the theoretical foundation for positive quartic polynomials established in \cite{ulrich1994positivity}, first defined after Equation (18).} \newline
-  $\displaystyle \gamma_0 := 4 {Df(x_0) \over Df(x_1)} (B/A)^{3/4}$ \newline
-  $\displaystyle \gamma_1 := {x_1 - x_0 \over Df(x_1)} (B/A)^{3/4}$ \newline
-  $\alpha_0 := 4 (B/A)^{1/4}$ \newline
-  $\displaystyle \alpha_1 := -{x_1 - x_0 \over Df(x_1)} (B/A)^{1/4}$ \newline 
-  $\displaystyle \beta_0 := 30 - {12 \big(Df(x_0) + Df(x_1)\big) (x_1 - x_0) \over \big(f(x_1) - f(x_0)\big) \sqrt{A}\sqrt{B}}$ \newline
-  $\displaystyle \beta_1 := {-3 (x_1 - x_0)^2 \over 2 \big(f(x_1) - f(x_0)\big) \sqrt{A} \sqrt{B}} $ \newline
-    \comment{The $\gamma$, $\alpha$, and $\beta$ terms with subscripts $0$ and $1$ are algebraic reductions of the simplified conditions for satisfying Theorem 2 in Equation (16) of \cite{ulrich1994positivity}.
-These terms with subscripts $0$ and $1$ make the computation of $\alpha$, $\beta$, and $\gamma$ functions of the second derivative terms, as seen in Step \ref{alg1-abg} below.} \newline
-  $\gamma := \gamma_0 + \gamma_1 D^2f(x_0)$ \newline
-  $\alpha := \alpha_0 + \alpha_1 D^2f(x_1)$ \newline
-  $\beta := \beta_0 + \beta_1 \big(D^2f(x_0) - D^2f(x_1)\big)$ \newline
-  if $(\beta \leq 6)$ return $\big(\alpha > - (\beta + 2) / 2\big)$ \newline
-  else $\qquad$ return $\big(\gamma > -2 \sqrt{\beta - 2}\big)$ \newline
-  end if \newline
-}
-\endnumberedlist
-\vskip 2mm
-\hrule
-\vskip 3mm
-\parindent=10pt
-
-%% ----------------------------------------------------------------------
-
-The reason for structuring the $\alpha$, $\beta$, and $\gamma$ computations in terms of the second derivative of $f$ (seen in Step \ref{alg1-abg} of Algorithm \ref{alg:check-monotone}) will become more apparent later.
-The next problem to consider is that of making a nonmonotone order six polynomial piece into a monotone one by modifying its first and second derivative values at the ends of an interval.
-Note that the actual value of the function at the ends of the interval is not modified, as the resulting polynomial needs to interpolate.
-
-Notice that both Algorithms \ref{alg:check-monotone} and \ref{alg:make-monotone} have ${\cal O}(1)$ runtime assuming a fixed level of precision is chosen beforehand.
-A constant number of operations are needed for verifying monotonicity (Algorithm \ref{alg:check-monotone}), while a constant set of operations and a single binary search are performed for enforcing monotonicity (Algorithm \ref{alg:make-monotone}).
-The search requires a fixed number of steps to achieve any predetermined relative precision, since its accuracy is predetermined and not a function of the problem at hand.
-Also note that an efficient implementation of Algorithm \ref{alg:make-monotone} only needs to recalculate Steps \ref{alg1-abg} and \ref{alg1-last} of Algorithm \ref{alg:check-monotone} during the binary search.
-Next, an algorithm for constructing a monotone quintic spline interpolant is presented.
-
-%% ----------------------------------------------------------------------
-
-\vskip 5mm
-\thickrule
-{\bf Algorithm 2:} {\tt monotone\_spline}$\big((k_1$, $\ldots$, $k_n)$, $f$, $s\big)$
-where $(k_1$, $\ldots$, $k_n)$ is an increasing sequence of real
-numbers, $f$ is an order six piecewise polynomial with breakpoints
-$k_1$, $\ldots$, $k_n$ defined by the data
-$\bigl\{f(k_i)\bigr\}_{i=1}^n$, $\bigl\{Df(k_i)\bigr\}_{i=1}^n$,
-$\bigl\{D^2f(k_i)\bigr\}_{i=1}^n$, and $s > 1$ is an integer shrink
-factor.
-\hrule
-\vskip 3mm
-    \newline create queue
-      \newline create changed$(1$, $\ldots$, $n) := $ FALSE
-      \newline create step\_size$(1$, $\ldots$, $n) := \bigl(Df(k_1)/s$, $\ldots$, $Df(k_n)/s\bigr)$
-      \comment{The three variables defined in Step \ref{step:create} are used to ensure the eventual achievement of monotonicity.
-The {\tt queue} is a standard first in first out (FIFO) queue with {\tt enqueue} and {\tt dequeue} operations.
-The array {\tt changed} contains Booleans describing whether or not a breakpoint belongs to an interval that has been modified to enforce monotonicity.
-The {\tt step\_size} array contains the real-valued derivative decrement step sizes to use in the search for a valid monotone spline.}
-    \newline for $i := 1$, $\ldots$, $n-1$
-      \indented{enqueue $\bigl((k_i, k_{i+1})\bigr)$}
-      end for
-      \comment{Initially, all intervals must be checked for monotonicity. The following loop will continue until all intervals are verified as monotone without need for modification.}
-    \newline while $($size$($queue$) > 0)$
-    \newline \indented{$(k_j$, $k_{j+1}) := $ dequeue}
-    \newline \indented{if $\big($not is\_monotone$(k_j$, $k_{j+1}$, $f)\big)$}
-    \newline \indented{\indented{if $\big($changed$(j)\big) \qquad \ Df(k_j) := Df(k_j) -{}$step\_size$(j)$}}
-      \indented{\indented{if $\big($changed$(j+1)\big) \quad Df(k_{j+1}) := Df(k_{j+1}) -{}$step\_size$(j+1)$}}
-      \indented{\indented{\comment{If a breakpoint belongs to an interval that has been previously modified, then the enforced monotonicity conditions on the second derivatives of adjacent intervals must contradict one another.
-In turn, the involved first derivatives are decreased towards zero by a predetermined step size.}}}
-    \newline \indented{\indented{make\_monotone$(k_j,\ k_{j+1},\ f)$}}
-    \newline \indented{\indented{changed$(j) :=$ TRUE}}
-      \indented{\indented{changed$(j+1) :=$ TRUE}}
-    \newline \indented{\indented{if $\big(j > 1$ and $(k_{j-1}, k_j) \not\in$ queue$\big)$ enqueue $\bigl((k_{j-1}, k_j)\bigr)$}}
-      \indented{\indented{enqueue $\bigl((k_j, k_{j+1})\bigr)$}}
-      \indented{\indented{if $\big(j+1 < n$ and $(k_{j+1}, k_{j+2}) \not\in$ queue$\big)$ enqueue $\bigl((k_{j+1}, k_{j+2})\bigr)$}}
-      \indented{\indented{\comment{Step \ref{step:changed} records the endpoints of the current interval as having been changed, while \ref{step:requeue} adds adjacent intervals to {\tt queue} that may have inadvertently been made nonmonotone by the changes to the present interval.}}}
-      \indented{end if}
-      end while
-\vskip 2mm
-\hrule
-\vskip 3mm
-
-%% ----------------------------------------------------------------------
-
-It is mentioned in \cite{ulrich1994positivity} that for sufficiently small $Df(k_i)$ and $Df(k_{i+1})$ the admissible solution interval of second derivative values becomes arbitrarily large.
-It can also be seen that decreasing the assigned derivative values towards zero will eventually cause Steps \ref{alg1-alpha} through \ref{alg1-gamma} of Algorithm \ref{alg:check-monotone} to all fail, resulting in Step \ref{alg1-simplified-true} returning {\tt TRUE}.
-
-If successive monotonicity fixes are required for all intervals, the worst case runtime of Algorithm \ref{alg:monotone-spline} is ${\cal O}(s n)$ for $n$ breakpoints and shrink factor $s$.
-In practice this worst case has been observed to be unlikely.
-
 \heading{3. EXPERIMENTAL RESULTS}
 
-Given the increased order of approximation, it is naturally expected that some accuracy benefit could be gained by using a quintic spline over a cubic spline.
-Experiments 1 and 2 below test some accuracy differences between monotone cubic and quintic splines.
-Experiment 3 analyzes the number of operations performed by Algorithm \ref{alg:monotone-spline} with increasingly large samples of monotone data.
+Given the increased order of approximation, it is naturally expected
+that some accuracy benefit could be gained by using a quintic spline
+over a cubic spline. Experiments 1 and 2 below test some accuracy
+differences between monotone cubic and quintic splines. Experiment 3
+analyzes the number of operations performed by Algorithm
+\ref{alg:monotone-spline} with increasingly large samples of monotone
+data.
 
@@
 
-For this experiment, the function $\sin(x) + x$ is considered over the interval $[0,(5/2)\pi]$ as seen in Figure \ref{fig:cubic_quintic_sin}.
-Given an increasing number of points, the maximum error of cubic and quintic approximations is shown in Figure \ref{fig:experiment_1}.
-The quintic approximation consistently has a maximum error that is roughly half that of the cubic approximation, given the same number of points.
-There is an increase in the error gap between the two approximations as more points are added.
-The quintic spline requires only $700$ points to approximate the function to the same accuracy achieved by the cubic with $1000$ points.
+For this experiment, the function $\sin(x) + x$ is considered over the
+interval $[0,(5/2)\pi]$ as seen in Figure
+\ref{fig:cubic_quintic_sin}. Given an increasing number of points, the
+maximum error of cubic and quintic approximations is shown in Figure
+\ref{fig:experiment_1}. The quintic approximation consistently has a
+maximum error that is roughly half that of the cubic approximation,
+given the same number of points. There is an increase in the error gap
+between the two approximations as more points are added. The quintic
+spline requires only $700$ points to approximate the function to the
+same accuracy achieved by the cubic with $1000$ points.
 
@@
 %%   \includegraphics[width=.7\textwidth]{cubic-quintic-sin}
-%%   \caption{Depicted above are the monotone cubic and quintic spline interpolants of the function $\sin(x) + x$ over the points $[0, (5/2) \pi]$.
-Notice that the maximum error of the cubic interpolant is larger, because it only captures first derivative information at the points.
-The quintic interpolant captures both first and second derivative information at the points.}
+%%   \caption{Depicted above are the monotone cubic and quintic spline
+%%   interpolants of the function $\sin(x) + x$ over the points $[0,
+%%   (5/2) \pi]$.  Notice that the maximum error of the cubic
+%%   interpolant is larger, because it only captures first derivative
+%%   information at the points. The quintic interpolant captures both
+%%   first and second derivative information at the points.}
 %% \end{figure}
@@
 %%   \includegraphics[width=.7\textwidth]{experiment_1_errors}
-%%   \caption{The error in the quintic and cubic approximations of $\sin(x) + x$ with increasing number of points.
-Both axes are log scaled, where $\mu$ means $\times 10^{-6}$ and n means $\times 10^{-9}$ on the vertical axis.
-The quintic approximation generally provides an approximation with about half of the maximum error of the cubic approximation, while the relative gap between grows with increasing number of points.
-In order to achieve the same absolute error the quintic spline requires ten fewer points at max error $10^{-4}$, thirty fewer points at max error $10^{-6}$, and $300$ fewer points at max error $10^{-8}$.}
+%%   \caption{The error in the quintic and cubic approximations of
+%%   $\sin(x) + x$ with increasing number of points. Both axes are log
+%%   scaled, where $\mu$ means $\times 10^{-6}$ and n means $\times
+%%   10^{-9}$ on the vertical axis. The quintic approximation
+%%   generally provides an approximation with about half of the
+%%   maximum error of the cubic approximation, while the relative gap
+%%   between grows with increasing number of points. In order to
+%%   achieve the same absolute error the quintic spline requires ten
+%%   fewer points at max error $10^{-4}$, thirty fewer points at max
+%%   error $10^{-6}$, and $300$ fewer points at max error $10^{-8}$.}
 %% \end{figure}
@@
 
-Consider the cumulative distribution function (CDF) defined by a mixture of three Gaussian distributions and visualized in Figure \ref{fig:experiment_2_distribution}.
-The distributions have weights $(.3, .6, .1)$, means $(.2, .45, .85)$ and standard deviations $(.05, .08, .03)$.
-$200$ random samples are generated from this distribution with both cubic and quintic approximations from four points, and over $100$ trials the resulting spread of empirical distribution functions (EDFs) are visualized in Figure \ref{fig:experiment_2_results}.
+Consider the cumulative distribution function (CDF) defined by a
+mixture of three Gaussian distributions and visualized in Figure
+\ref{fig:experiment_2_distribution}. The distributions have weights
+$(.3, .6, .1)$, means $(.2, .45, .85)$ and standard deviations $(.05,
+.08, .03)$. $200$ random samples are generated from this distribution
+with both cubic and quintic approximations from four points, and over
+$100$ trials the resulting spread of empirical distribution functions
+(EDFs) are visualized in Figure \ref{fig:experiment_2_results}.
 
-While an increased number of points could be used to decrease the error in either approximation, the purpose of this experiment is to demonstrate the effect approximation error has on random number generation.
-The errors in the CDFs are magnified by the variance involved in randomly sampling from a distribution.
-As a result, while the cubic and quintic spline approximations have similar maximum error, the cubic distribution is significantly distorted around the first and second modes.
+While an increased number of points could be used to decrease the
+error in either approximation, the purpose of this experiment is to
+demonstrate the effect approximation error has on random number
+generation. The errors in the CDFs are magnified by the variance
+involved in randomly sampling from a distribution. As a result, while
+the cubic and quintic spline approximations have similar maximum
+error, the cubic distribution is significantly distorted around the
+first and second modes.
 
@@
 %%   \includegraphics[width=.8\textwidth]{experiment_2_distribution}
-%%   \caption{The cumulative distribution function (top) and probability density function (bottom) for the Gaussian mixture distribution in Experiment 2.
-The distributions have weights $(.3, .6, .1)$, means $(.2, .45, .85)$ and standard deviations $(.05, .08, .03)$.
-Random samples are generated from this mixture distribution by evaluating the inverse of the CDF at random numbers uniformly distributed in the range $[0,1]$.}
+%%   \caption{The cumulative distribution function (top) and
+%%   probability density function (bottom) for the Gaussian mixture
+%%   distribution in Experiment 2. The distributions have weights
+%%   $(.3, .6, .1)$, means $(.2, .45, .85)$ and standard deviations
+%%   $(.05, .08, .03)$. Random samples are generated from this mixture
+%%   distribution by evaluating the inverse of the CDF at random
+%%   numbers uniformly distributed in the range $[0,1]$.}
 %% \end{figure}
@@
 %%   \includegraphics[width=.4\textwidth]{experiment_2_quintic_dist}
-%%   \caption{Both cubic and quintic splines are used to approximate the CDF of a Gaussian mixture distribution with three components.
-Four CDF points are used to build the approximation, $200$ random samples are generated with the approximate CDFs over $100$ trials.
-The resulting cloud of empirical distribution function (EDF) values is seen above.
-The maximum approximation error in the CDFs are similar, $.08$ for the cubic and $.05$ for the quintic.
-However, the maximum EDF differences observed are doubled by the variance of random samples, $.15$ for the cubic and $.1$ for the quintic.}
+%%   \caption{Both cubic and quintic splines are used to approximate
+%%   the CDF of a Gaussian mixture distribution with three
+%%   components. Four CDF points are used to build the approximation,
+%%   $200$ random samples are generated with the approximate CDFs over
+%%   $100$ trials. The resulting cloud of empirical distribution
+%%   function (EDF) values is seen above. The maximum approximation
+%%   error in the CDFs are similar, $.08$ for the cubic and $.05$ for
+%%   the quintic. However, the maximum EDF differences observed are
+%%   doubled by the variance of random samples, $.15$ for the cubic
+%%   and $.1$ for the quintic.}
 %% \end{figure}
@@
 
-This experiment studies the number of times the algorithms {\tt is\_monotone} and {\tt make\_monotone} are executed for increasingly large sequences of random monotone data.
-The minimum, median, and maximum number of times that Algorithms \ref{alg:check-monotone} and \ref{alg:make-monotone} are executed is recorded, as well as the number of steps taken in all calls to {\tt binary\_search}.
-The number of calls and steps grows linearly with $n$ as expected, requiring roughly one call to {\tt make\_monotone} and $20$ binary search steps to create a monotone piece.
-Some (less common) problems require an average of three calls to {\tt make\_monotone} per interval.
+This experiment studies the number of times the algorithms {\tt
+  is\_monotone} and {\tt make\_monotone} are executed for increasingly
+large sequences of random monotone data. The minimum, median, and
+maximum number of times that Algorithms \ref{alg:check-monotone} and
+\ref{alg:make-monotone} are executed is recorded, as well as the
+number of steps taken in all calls to {\tt binary\_search}. The number
+of calls and steps grows linearly with $n$ as expected, requiring
+roughly one call to {\tt make\_monotone} and $20$ binary search steps
+to create a monotone piece. Some (less common) problems require an
+average of three calls to {\tt make\_monotone} per interval.
 
@@
 %%   \centering
-%%   \caption{The minimum, median, and maximum number of checks, fixes, and binary search steps required in the execution of {\tt make\_spline} for increasing size sequences, $n$, over $100$ randomly generated sets of monotone data.
-Notice the maximum for each counter is often significantly greater than the minimum and median, because the distribution of each counter is skewed right.}
-
+%%   \caption{The minimum, median, and maximum number of checks,
+%%   fixes, and binary search steps required in the execution of {\tt
+%%   make\_spline} for increasing size sequences, $n$, over $100$
+%%   randomly generated sets of monotone data. Notice the maximum for
+%%   each counter is often significantly greater than the minimum and
+%%   median, because the distribution of each counter is skewed
+%%   right.}
 %%   \begin{tabular}{c|c c c|c c c|c c c}
@@
 %% \end{table}
-
-\heading{4. DISCUSSION}
-
-The monotone quintic spline interpolant provides a distinct increase in accuracy over the monotone cubic variant.
-The computational complexity of this algorithm is identical to the existing cubic method and the cost of evaluating the spline after construction is unchanged.
-While the number of computations required per interval has increased, the number of points required to achieve the same level of accuracy has decreased.
-In the present state, the algorithm for enforcing monotonicity on a spline is not as trivially parallel as the cubic algorithm, however it can still be parallelized.
-The checks for monotonicity across all intervals are independent and the monotonicity adjustments could be computed independently and intelligently merged with minor changes to the serial algorithm.
-
-Acknowledging that the algorithm for constructing a monotone quintic spline interpolant is slightly more expensive than the cubic case, gains in accuracy or decreased necessary number of points are often worth the computational effort.
-In practice the costs of spline interpolation are often dominated by evaluation rather than construction, and the increased accuracy is afforded for a negligible increase in computational cost.
-
-\heading{5. CONCLUSION AND FUTURE WORK}
-
-This paper proposes and tests an algorithm for constructing monotone quintic spline interpolants.
-Experiments demonstrate an improvement in approximation accuracy over monotone cubic spline interpolants, as expected based on theory.
-There are still open avenues of research going forward, such as an alternative sufficient condition for enforcing monotonicity or increased order monotone approximations.
-If the monotonicity conditions can be generalized to any order or made linear, the search for a monotone interpolating spline could potentially be formulated as a convex optimization problem.
-Finally, this work could be used to improve cumulative distribution function (CDF) estimates as well as predictive models that use CDFs.
-
-\heading{Acknowledgments}
-This work was supported by the National Science Foundation Grants CNS-1565314 and CNS-1838271.
-
-\heading{\hlvVIII BIBLIOGRAPHY} %% {\let\caps=\capsVIII \let\sl=\slVIII
-%% \def\ys{\hskip 1em minus .5em}\let\rm=\rmeight \rmVIII
-%% \newcount\refnum \refnum=0 \parskip=2pt plus 2pt minus 2pt
-%% \frenchspacing
-
-\bibliography{mqsi20}
-\bibliographystyle{plain}
-
-\vfill\eject\end
-
-%% %\refj is for regular journal articles
-%% \def\refj#1#2#3#4#5{\noindent\hangindent=10pt\hangafter=1
-%% %\global\advance\refnum by 1[\number\refnum]
-%% {\caps #1}\ys #2.\ys {\rm #3}. {\sl #4}, #5.\par}
-
-%% %\refb is for books
-%% \def\refb#1#2#3#4{\noindent\hangindent=10pt\hangafter=1
-%% %\global\advance\refnum by 1[\number\refnum]
-%% {\caps #1}\ys #2.\ys{\sl #3}. {\rm #4}.\par}
-
-%% %\refp is for private communication
-%% \def\refp#1#2#3{\noindent\hangindent=10pt\hangafter=1
-%% %\global\advance\refnum by 1[\number\refnum]
-%% {\caps #1}\ys #2.\ys{\rm #3}.\par}
-
-%% %\reft is for tech. reports, theses, and conferences
-%% \def\reft#1#2#3#4{\noindent\hangindent=10pt\hangafter=1
-%% %\global\advance\refnum by 1[\number\refnum]
-%% {\caps #1}\ys #2.\ys{\rm #3}. {\rm #4}.\par}
-
-%% %%%%% top of refs
-%% \reft{Amos, B. D., Easterling, D. R., Watson, L. T., Thacker, W. I.,
-%% Castle, B. S., Trosset, M. W.}{2014}{Algorithm XXX: QNSTOP: Quasi-Newton
-%% algorithm for stochastic optimization}{Technical Report 2014-07,
-%% Virginia Polytechnic Institute and State University, Blacksburg, VA}
-
-%% \refb{Anderson, E., Bai, Z., Bischof, C., Blackford, S., Demmel, J.,
-%% Dongarra, J., Du Croz, J., Greenbaum, A., Hammarling, S., McKenney, A.,
-%% and Sorensen, D.}{1999}{LAPACK Users' Guide Third Edition}
-%% {SIAM, Philidelphia, PA}
-
-%% \refb{Aurenhammer, F., Klein, R., and Lee, D. T.}{2013}
-%% {Voronoi diagrams and Delaunay triangulations}
-%% {World Scientific Publishing Co., Hackensack, NJ}
-
-%% \refj{Barber, C. B., Dobkin, D. P., and Huhdanpaa, H.}{1996}
-%% {The Quickhull algorithm for convex hulls}{ACM Trans. Math.
-%% Softw. 22}{469--483}
-
-%% \reft{Boissonnat, J.-D., Devillers, O., and Hornus, S.}{2009}
-%% {Incremental construction of the Delaunay triangulation and the
-%% Delaunay graph in medium dimension}{In {\sl Proc. Twenty-fifth Annual Symp.
-%% on Computational Geometry}, ACM, New York, NY, 208--216}
-
-%% \refj{Bowyer, A.}{1981}{Computing Dirichlet tessellations}
-%% {The Computer Journal 24}{162--166}
-
-%% \refj{Cameron, K. W., Anwar, A., Cheng, Y., Xu, L., Li, B., Ananth, U.,
-%% Bernard, J., Jearls, C., Lux, T., Hong, Y., Watson, L. T.,
-%% and Butt, A. R.}{2019}{MOANA: modeling and analyzing I/O variability
-%% in parallel system experimental design}
-%% {IEEE Trans. Parallel Distrib. Systems}{to appear}
-
-%% \reft{Chang, T. H., Watson, L. T., Lux, T. C. H., Bernard, J., Li, B.,
-%% Xu, L., Back, G., Butt, A. R., Cameron, K. W., and Hong, Y.}{2018a}
-%% {Predicting system performance by interpolation using a high-dimensional
-%% Delaunay triangulation} {In {\sl Proc. 2018 Spring Simulation Multiconference,
-%% 26th High Performance Computing Symp.}, Soc. for Modelling and Simulation
-%% Internat., Vista, CA, Article No. 2}
-
-%% \reft{Chang, T. H., Watson, L. T., Lux, T. C. H., Li, B., Xu, L., 
-%% Butt, A. R., Cameron, K. W., and Hong, Y.}{2018b}{A polynomial time
-%% algorithm for multivariate interpolation in arbitrary dimension via
-%% the Delaunay triangulation} {In {\sl Proc. ACM 2018 Southeast Conference
-%% (ACMSE '18)}, ACM, New York, NY, Article No. 12}
-
-%% \refb{Cheney, E. W. and Light, W. A.}{2009}
-%% {A Course in Approximation Theory}{American Mathematical Soc., 
-%% Providence, RI}
-
-%% \refb{Cheng, S. W., Dey, T. K., and Shewchuk, J.}{2012}
-%% {Delaunay Mesh Generation}{Computer and Information Science
-%% Series, CRC Press, New York, NY}
-
-%% \refj{Cignoni, P., Montani, C., and Scopigno, R.}{1998}
-%% {DeWall: A fast divide \& conquer Delaunay triangulation algorithm in $E^d$}
-%% {Computer-Aided Design 30}{333--341}
-
-%% \refb{de Berg, M., Cheong, O., van Kreveld, M., and Overmars, M.}{2008}
-%% {Computational Geometry: Algorithms and Applications}{Springer-Verlag,
-%% Berlin, Germany}
-
-%% \refj{Devillers, O., Pion, S., and Teillaud, M.}{2002}
-%% {Walking in a triangulation}{Internat. Journal of Foundations of
-%% Computer Science 13}{181--199}
-
-%% \reft{Edelsbrunner, H.}{1989}{An acyclicity theorem for cell complexes in
-%% d dimensions} {In {\sl Proc. Fifth Annual Symp. on Computational Geometry}, 
-%% ACM, New York, 145--151}
-
-%% \refj{Hanson, R. J. and Haskell, K. H.}{1982}
-%% {Algorithm 587: Two algorithms for the linearly constrained least
-%% squares problem}{ACM Trans. Math. Softw. 8}{323--333}
-
-%% \refj{Klee, V.}{1980}{On the complexity of d-dimensional Voronoi diagrams}
-%% {Archiv der Mathematik 34}{75--80}
-
-%% \reft{Lux, T. C. H., Watson, L. T., Chang, T. H., Bernard, J., Li, B.,
-%% Xu, L., Back, G., Butt, A. R., Cameron, K. W., and Hong, Y.}{2018}
-%% {Predictive modeling of I/O characteristics in high performance computing
-%% systems}{In {\sl Proc. 2018 Spring Simulation Multiconference, 26th High 
-%% Performance Computing Symp.}, Soc. for Modelling and Simulation Internat.,
-%% Vista, CA, Article No. 8}
-
-%% \refj{M{\"u}cke, E. P., Saias, I., and Zhu, B.}{1999}
-%% {Fast randomized point location without preprocessing in two- and 
-%% three-dimensional Delaunay triangulations}
-%% {Computational Geometry 12}{63--83}
-
-%% \refj{Omohundro, S. M.}{1990}{Geometric learning algorithms}
-%% {Physica D: Nonlinear Phenomena 42}{307--321}
-
-%% \reft{OpenMP Architecture Review Board (ARB)}{2015}
-%% {OpenMP Application Programming Interface (version 4.5)}
-%% {OpenMP}
-
-%% \refj{Rajan, V. T.}{1994}
-%% {Optimality of the Delaunay triangulation in $R^d$}
-%% {Discrete \& Computational Geometry 12}{189--202}
-
-%% \refj{Schaap, W. E. and van de Weygaert, R.}{2000}
-%% {Continuous fields and discrete samples: reconstruction through
-%% Delaunay tessellations}{Astronomy and Astrophysics 363}{L29--L32}
-
-%% \refj{Su, P. and Drysdale, R. L. S.}{1997}
-%% {A comparison of sequential Delaunay triangulation algorithms}
-%% {Computational Geometry 7}{361--385}
-
-%% \refj{Watson, D. F.}{1981}{Computing the n-dimensional Delaunay tessellation
-%% with application to Voronoi polytopes}
-%% {The Computer Journal 24}{167--172}
-
-%% }\vfill\eject\end
-%% %%%%% bottom

### 2020-06-21 Research_[2020-05-18]_Piecewise_Monotone_Quintic_Fortran/test_interval.py

@@ -25,7 +25,7 @@ F0   =    3.4295043674554457E+021
 DF0  =   -7.3547132703516163E-002
 DDF0 =   -4.7637407608192277E-023
 F1   =    3.3049635425868652E+021
-DF1  =  -0.14222222222201797     
+DF1  =   -0.14222222222201797     
 DDF1 =   -4.1453609064447915E-023
 
 # ====================================================================
@@ -42,10 +42,8 @@ DDF1 =   -4.1453609064447915E-023
 # ====================================================================
 #                        Normal value cases.
 
-
-
 # Use Python to construct an exact quintic piece that matches this data.
-from polynomial import Spline
+from polynomial import Spline, Polynomial
 from fraction import Fraction
 def exact(l):
     try: return [exact(v) for v in l]
@@ -55,6 +53,7 @@ x = [U0, U1]
 fx = [[F0, DF0, DDF0], [F1, DF1, DDF1]]
 f = Spline(exact(x), exact(fx))
 df = f.derivative()
+print(Polynomial(f._functions[0]))
 
 # Use Fortran to construct a quintic piece thata matches this data.
 
@@ -68,14 +67,16 @@ class splines:
 
 nb, ncc = 2, 3
 sk = np.ones(nb*ncc + 2*ncc)
-sc = np.ones(nb*ncc+1)
+sc = np.ones(nb*ncc)
 x = np.asarray(x, dtype=float, order='f')
 fx = np.asarray(fx, dtype=float, order='f')
+shift = np.min(fx[:,0])
+fx[:,0] -= shift
 sk, sc, info = splines.fit_spline(x, fx, sk, sc)
+sc += shift
 print('info:', info)
 print('sk:', sk)
 print('sc:', sc)
-
 # MAXPT =     1.2116159427162417E+031
 # MINPT =     2.8397248657411912E+030
 # print("max df:", float(df(MAXPT)))
@@ -89,9 +90,9 @@ print(max(abs(np.array(df(x.copy()),dtype=float) -
               splines.eval_spline(sk,sc,x.copy(),d=1)[0])))
 # exit()
 
-p.add("Py f", x, f(x.copy()), mode='lines', group=0)
-p.add("Py df", x, df(x.copy()), mode='lines', group=1)
-p.add("Py ddf", x, df.derivative()(x.copy()), mode='lines', group=2)
+p.add("Exact f", x, f(x.copy()), mode='lines', group=0)
[…]

### 2020-06-23 Research_[2020-05-18]_Piecewise_Monotone_Quintic_Fortran/paper/1-introduction.tex
Updated code package sent off last night. New tex revisions from Dr. Watson received, implementing.

\bigskip\hrule\bigskip\medskip
\heading{1. INTRODUCTION}

Many domains of science rely on smooth approximations to real-valued
functions over a closed interval. Piecewise polynomial functions
(splines) provide the smooth approximations for animation in graphics
[Herman et al. 2015; Quint 2003], aesthetic structural support in
architecture [Brennan 2020], efficient aerodynamic surfaces in
automotive and aerospace engineering [Brennan 2020], prolonged
effective operation of electric motors [Berglund et al 2009], and
accurate nonparametric approximations in statistics [Knott
  2012]. While polynomial interpolants and regressors apply broadly,
splines are often a good choice because they can approximate globally
complex functions while minimizing the local complexity of an
approximation.

It is often the case that the true underlying function or phenomenon
being modeled has known properties e.g., convexity, positivity,
various levels of continuity, or monotonicity. Given a reasonable
amount of data, it quickly becomes difficult to achieve desirable
properties in a single polynomial function. In general, the
maintenance of function properties through interpolation/regression is
referred to as {\it shape preserving} [Fritsch and Carlson 1980;
  Gregory 1985]. The specific properties the present algorithm will
preserve in approximations are monotonicity and $C^2$ continuity. In
addition to previously mentioned applications, these properties are
crucially important in statistics to the approximation of a cumulative
distribution function and subsequently the effective generation of
random numbers from a specified distribution [Ramsay 1988].  A spline
function with these properties could approximate a cumulative
distribution function to a high level of accuracy with relatively few
intervals. A twice continuously differentiable approximation to a
cumulative distribution function (CDF) would produce a corresponding
probability density function (PDF) that is continuously
differentiable, which is desirable.

The currently available software for monotone piecewise polynomial
interpolation includes quadratic [He and Shi 1998], cubic [Fritsch and
  Carlson 1980], and (with limited application) quartic [Wang and Tan
  2004; Piah and Unsworth 2011; Yao and Nelson 2018] cases. In
addition, a statistical method for bootstrapping the construction of
an arbitrarily smooth monotone fit exists [Leitenstorfer and Tutz
  2006], but the method does not take advantage of known analytic
properties of quintic polynomials. The code by Fritsch [1982] for
$C^1$ cubic spline interpolation is the predominantly utilized code
for constructing monotone interpolants at present. Theory has been
provided for the quintic case [Ulrich and Watson 1994; He{\ss} and
  Schmidt 1994] and that theory was recently utilized in a proposed
algorithm [Lux 2020] for monotone quintic spline construction, however
no published mathematical software exists.

The importance of piecewise quintic interpolation over lower order
approximations can be simply observed. In general, the order of a
polynomial determines the number of function (and derivative) values
it can interpolate, which in turn determines the growth rate of error
away from interpolated values. This work constructs quintic splines in
order to match the function value and two derivatives at either end of
an interval, which requires six degrees of freedom in an interpolating
polynomial. This work provides a Fortran 2003 subroutine {\tt MQSI}
based on the necessary and sufficient conditions in Ulrich and Watson
[1994] for the construction of monotone quintic spline interpolants of
monotone data.

The remainder of this paper is structured as follows: Section 2 provides
the algorithms for constructing a $C^2$ monotone quintic spline interpolant
to monotone data, Section 3 outlines the method of spline representation
($B$-spline basis) and evaluation, Section 4 analyzes the complexity and
sensitivity of the algorithms in {\tt MQSI}, and Section 5 presents an
empirical performance study and some graphs of constructed interpolants.

### 2020-06-23 Research_[2020-05-18]_Piecewise_Monotone_Quintic_Fortran/paper/2-monotone-quintic.tex
Updated code package sent off last night. New tex revisions from Dr. Watson received, implementing.

\heading{2. MONOTONE QUINTIC INTERPOLATION}

In order to construct a monotone quintic interpolating spline, two
primary problems must be solved. First, reasonable derivative values
at data points need to be estimated. Second, the estimated derivative
values need to be modified to enforce monotonicity on all polynomial
pieces.

% The remainder of this section outlines three algorithms for
% constructing a quintic spline interpolant.  Algorithm 1 uses a
% quadratic facet model to construct initial first and second
% derivative estimates for the interpolating spline, Algorithm 2
% checks monotonicity of an interval, and Algorithm 3 defines a
% procedure for stepping initial first and second derivative estimates
% towards values which result in a monotone quintic spline.

Fritsch and Carlson [1980] originally proposed the use of central
differences to estimate derivatives, however this often leads to extra
and unnecessary {\it wiggles} in the spline when used to approximate
second derivatives. In an attempt to capture the local shape of the
data, this package uses a facet model [Haralick and Watson 1981] to
estimate first and second derivatives at data. Rather than picking a
local linear regressor with minimal residual, this work uses a
quadratic facet model that selects the local quadratic interpolant
with minimum magnitude curvature.

%% ===================================================================
%%                   Algorithm 1: QUADRATIC_FACET
\vskip 5mm
{\parindent 0mm
{\bf Algorithm 1:} {
\tt QUADRATIC\_FACET}$\big(X(1{:}n)$, $Y(1{:}n)$, $i \big)$ \vskip 1mm
where $X_j$,$Y_j$ $\in \bbb{R}$ for $j = 1$, $\ldots$, $n$, and $1 < i
< n$. Returns the slope and curvature at $X_i$ of the local quadratic
interpolant with minimum magnitude curvature.
\vskip 2mm}
{\parindent=6mm
%% -------------------------------------------------------------------
\item{} {\tt if} $\big((Y_i \approx Y_{i-1})$ {\tt or} $(Y_i \approx
  Y_{i+1})\big)$ {\tt then; return $(0,0)$}
\item{} {\tt else if} $\big((Y_{i+1} - Y_i)(Y_i - Y_{i-1}) < 0\big)$ {\tt then}
\itemitem{} The point $(X_i$, $Y_i)$ is an extreme point. The
quadratic with minimum curvature that has slope $0$ at $X_i$ will be returned.
\item{} \codent $f_1 := $ interpolate $(X_{i-1}$,$Y_{i-1})$, $(X_i$,$Y_i)$,
  and $Df_1(X_i) = 0$.
\item{} \codent $f_2 := $ interpolate $(X_i$,$Y_i)$, $(X_{i+1}$,$Y_{i+1})$,
  and $Df_2(X_i) = 0$.
\item{} \codent {\tt if} $\big(|D^2f_1| \leq |D^2f_2|\big)$ {\tt then; return} $\big(Df_1$, $D^2f_1\big)$
\item{} \codent {\tt else;} \hskip 34.25mm {\tt return} $\big(Df_2$, $D^2f_2\big)$
\item{} \codent {\tt endif}
\item{} {\tt else}
\itemitem{} {The point $(X_i$, $Y_i)$ is in a monotone segment of
data. In the following, it is possible that either $f_1$ or $f_3$ does
not exist because $i = 2$ or $i = n-1$. In those cases, the minimum
curvature among existing quadratics is chosen.}
\item{} \codent $f_1 := $ interpolate $(X_{i-2}$,$Y_{i-2})$,
  $(X_{i-1}$,$Y_{i-1})$, and $(X_i$,$Y_i)$.
\item{} \codent $f_2 := $ interpolate $(X_{i-1}$,$Y_{i-1})$,
  $(X_i$,$Y_i)$, and $(X_{i+1}$,$Y_{i+1})$.
\item{} \codent $f_3 := $ interpolate $(X_i$,$Y_i)$,
  $(X_{i+1}$,$Y_{i+1})$, and $(X_{i+2}$,$Y_{i+2})$.
\item{} \codent {\tt if} \hskip 9mm $\big(|D^2f_1| \leq |D^2f_2|, |D^2f_3|\big)$
  {\tt then; return} $\big(Df_1$, $D^2f_1\big)$
\item{} \codent {\tt else if} $\big(|D^2f_2| \leq |D^2f_1|, |D^2f_3|\big)$
  {\tt then; return} $\big(Df_2$, $D^2f_2\big)$
\item{} \codent {\tt else;} \hskip 55mm {\tt return} $\big(Df_3$, $D^2f_3\big)$
\item{} \codent {\tt endif}
\item{} {\tt endif}
}
\vskip 5mm
%% ----------------------------------------------------------------------

The estimated derivative values by the quadratic facet model are not
guaranteed to produce monotone quintic polynomial segments. Ulrich and
Watson [1994] established tight constraints on the monotonicity of a
quintic polynomial piece, while deferring to He{\ss} and Schmidt
[1994] for a relevant simplified case. The following algorithm
implements a sharp check for monotonicity by considering the
nondecreasing case. The nonincreasing case is handled similarly.

%% ===================================================================
%%                    Algorithm 2: IS_MONOTONE
\vskip 5mm
{\parindent 0mm
{\bf Algorithm 2:} {\tt IS\_MONOTONE}$\big(x_0, x_1, f \big)$ \vskip 1mm
where $x_0$, $x_1 \in {\bbb R}$, $x_0 < x_1$, and $f$ is an order six
polynomial defined by $f(x_0)$, $Df(x_0)$, $D^2f(x_0)$, $f(x_1)$,
$Df(x_1)$, $D^2f(x_1)$. Returns {\tt TRUE} if $f$ is monotone
increasing on $[x_0,x_1]$.
\vskip 2mm}
{\parindent=6mm
%% -------------------------------------------------------------------
\item{1.} {\tt if} $\big(f(x_0) \approx f(x_1)\big)$ {\tt then}
\item{2.} \codent {\tt return} $\big( 0 = Df(x_0) = Df(x_1)
  = D^2f(x_0) = D^2f(x_1) \big)$
\item{3.} {\tt endif}
\item{4.} {\tt if} $\big(Df(x_0) < 0$ {\tt or} $Df(x_1) < 0\big)$ {\tt
  then; return FALSE; endif}
\item{5.} $w := x_1 - x_0$
\item{6.} $v := f(x_1) - f(x_0)$ \vskip 1mm
\item{} {The necessity of steps 2 and 4 follows directly from
  the fact that $f$ is $C^2$. The following steps 7--13 coincide with
  a simplified condition for quintic monotonicity that reduces to one
  of cubic positivity studied by Schmidt and He{\ss} [1988]. Given
  $\alpha$, $\beta$, $\gamma$, and $\delta$ as defined by Schmidt and
  He{\ss}, monotonicity results when $\alpha \geq 0$, $\delta \geq 0$,
  $\beta \geq \alpha - 2 \sqrt{\alpha \delta}$, and $\gamma \geq
  \delta - 2 \sqrt{\alpha \delta}$.  Step 4 checked for $\delta < 0$,
  step 8 checks $\alpha < 0$, step 10 checks $\beta < \alpha - 2
  \sqrt{\alpha \delta}$, and step 11 checks $\gamma < \delta - 2
  \sqrt{\alpha \delta}$. If none of the monotonicity conditions were
  violated, then the order six piece is monotone and step 12
  concludes.} \vskip 1mm
\item{7.} {\tt if} $\big(Df(x_0) \approx 0$ {\tt or} $Df(x_1) \approx 0\big)$ {\tt then}
\item{8.} \codent {\tt if} $\big(D^2f(x_1)w > 4Df(x_1)$ {\tt then; return FALSE; endif}
\item{9.} \codent $t := 2 \sqrt{Df(x_0) \left(4Df(x_1) - D^2f(x_1) w \right)}$
\item{10.} \codent {\tt if} $\big(t + 3Df(x_0) + D^2f(x_0)w < 0 \big)$
  {\tt then; return FALSE; endif}
\item{11.} \codent {\tt if} $\big(60v - w\big(24Df(x_0) + 32Df(x_1) - 2t
  + w(3D^2f(x_0) - 5D^2f(x_1))\big) < 0\big)$
\item{}   \codent \codent {\tt then; return FALSE; endif}
\item{12.} \codent {\tt return TRUE}
\item{13.} {\tt endif} \vskip 1mm
\item{} {The following code considers the full quintic monotonicity
  case studied by Ulrich and Watson [1994]. Given $\tau_1$, $\alpha$,
  $\beta$, and $\gamma$ as defined by Ulrich and Watson, a quintic
  piece is proven to be monotone when $\tau_1 > 0$ and $\alpha, \gamma
  \geq -(\beta+2)/2$ when $\beta \leq 6$, and $\alpha, \gamma \geq -2
  \sqrt{\beta-2}$ when $\beta > 6$. Step 12 checks $\tau_1 <= 0$,
  steps 18 and 19 determine monotonicity based on $\alpha$, $\beta$,
  and $\gamma$.} \vskip 1mm
\item{12.} {\tt if} $\big( w\big(2\sqrt{Df(x_0) Df(x_1)} - 3(Df(x_0) +
  Df(x_1))\big) - 24v \leq 0 \big)$
\item{} \codent {\tt then; return FALSE; endif}
\item{14.} $t := \root{3/4}\of{Df(x_0) Df(x_1)}$
\item{15.} $\alpha := (4 Df(x_1) - D^2f(x_1)w) \sqrt{Df(x_0)} / t$
\item{16.} $\gamma := (4 Df(x_0) - D^2f(x_0)w) \sqrt{Df(x_1)} / t$
\item{17.} $\displaystyle \beta := {60v/w + 3\big(w(D^2f(x_1) -
  D^2f(x_0)) - 8(Df(x_0) + Df(x_1))\big) \over 2 \sqrt{Df(x_0) Df(x_1)}}$
\item{18.} {\tt if} $(\beta \leq 6)$ {\tt then; return} $\min(\alpha,\gamma) > - (\beta + 2) / 2$
\item{19.} {\tt else;} \hskip 18mm {\tt return} $\min(\alpha,\gamma) > -2 \sqrt{\beta - 2}$
\item{20.} {\tt endif}
}
\vskip 5mm
%% ----------------------------------------------------------------------

It is shown by Ulrich and Watson [1994] that when $0 = DQ(X_i)$ $=
DQ(X_{i+1})$ $= D^2Q(X_i)$ $= D^2Q(X_{i+1})$, the quintic polynomial
over $[X_i$, $X_{i+1}]$ is guaranteed to be monotone. Using this fact,
the following algorithm shrinks initial derivative estimates until a
monotone spline is achieved and outlines the core routine in the
accompanying package.

%% ===================================================================
%%       Algorithm 3: MQSI - Monotone quintic spline interpolant
\vskip 5mm
{\parindent 0mm
{\bf Algorithm 3:} {\tt MQSI}$\big(X(1{:}n), Y(1{:}n) \big)$ \vskip 1mm
where $(X_i,Y_i) \in (\bbb{R},\bbb{R})$ $i = 1,\ldots,n$ are data
points. Returns monotone quintic spline interpolant $Q(x)$ such that
$Q(X_i) = Y_i$ and is monotone increasing (decreasing) on all
intervals that $Y_i$ is increasing (decreasing).
\vskip 2mm}
{\parindent=6mm
\item{} Approximate first and second derivatives at all $X_i$ with {\tt QUADRATIC\_FACET}.
\item{} {\tt do} $i = 1, \ldots n-1$
\item{} \codent $D Q(X_i)$, $D^2 Q(X_i) :=$ {\tt QUADRATIC\_FACET}$(X$, $Y$, $i)$
\item{} {\tt enddo}
\item{} Identify and store all nonmonotone intervals in a {\tt queue}.
\item{} {\tt do} $i = 1, \ldots n-1$
\item{} \codent {\tt if not IS\_MONOTONE}$(X_i$, $X_{i+1}$, $Q)$ {\tt then}
\item{} \codent \codent Add interval $\big(X_i$, $X_{i+1}\big)$ to {\tt queue}.
\item{} \codent {\tt endif}
\item{} {\tt enddo}
\item{} {\tt do while} $\big($ {\tt queue} of nonmonotone intervals is nonempty $\big)$
\itemitem{} Shrink $DQ$ and $D^2Q$ that border nonmonotone intervals.
\itemitem{} Identify and store all remaining nonmonotone intervals in {\tt queue}.
\item{} {\tt enddo}
}
\vskip 5mm
%% ----------------------------------------------------------------------

Given the minimum curvature nature of the initial derivative
estimates, it is desirable to make the smallest necessary changes to
the initial interpolating spline $Q$ while enforcing monotonicity. In
practice a quasi-bisection search is used in place of solely shrinking
$DQ$ and $D^2Q$, adding an additional step that increases previously
shrunken derivative values by a smaller step. As long as a step size
schedule is set that allows the value $0$ to be obtained in a fixed
number of computations, this has no effect on computational
complexity. Notably, Algorithm 3 is equally capable of ensuring a {\it
  piecewise} monotone quintic spline interpolates piecewise monotone
data.

### 2020-06-23 Research_[2020-05-18]_Piecewise_Monotone_Quintic_Fortran/paper/3-spline-rep.tex
Updated code package sent off last night. New tex revisions from Dr. Watson received, implementing.

\heading{3. SPLINE REPRESENTATION}

The monotone quintic spline interpolant is represented in terms of a
B-spline basis. Two routines in this package are provided, {\tt
  FIT\_SPLINE} and {\tt EVAL\_SPLINE} that compute the B-spline
coefficients to match the piecewise quintic polynomial values and
derivatives, and evaluate a spline represented in terms of a B-spline
basis. A Fortran 95 implementation of the B-spline recurrence relation
evaluation code by C. deBoor for the value, derivatives, and integral
of a B-spline is also provided.

% This function uses the recurrence relation defining a B-spline:
%
% B_{I,1}(X) = {1, if T(I) <= X < T(I+1),
%              {0, otherwise,
%
% where I is the spline index, J = 2, ..., N-MAX{D,0}-1 is the order, and
%
%                X-T(I)                      T(I+J)-X
% B_{I,J}(X) = ------------- B_{I,J-1}(X) + ------------- B_{I+1,J-1}(X).
%             T(I+J-1)-T(I)                T(I+J)-T(I+1)
%                                                                
% All of the intermediate steps (J) are stored in a single block
% of memory that is reused for each step.
%
% The computation of the integral of the B-spline proceeds from
% the above formula one integration step at a time by adding a
% duplicate of the last knot, raising the order of all
% intermediate B-splines, summing their values, and rescaling
% each sum by the width of the supported interval divided by the
% degree plus the integration coefficient.
%
% For the computation of the derivative of the B-spline, the divided
% difference definition of B_{I,J}(X) is used, building from J = N-D,
% ..., N-1 via
%
%                      (J-1) B_{I,J-1}(X)     (J-1) B_{I+1,J-1}(X)  
% D/DX[B_{I,J}(X)] =  ------------------  -  --------------------.
%                       T(I+J-1) - T(I)         T(I+J) - T(I+1)     
%
% The final B-spline is right continuous and has support over the
% interval [T(1), T(N)).

### 2020-06-23 Research_[2020-05-18]_Piecewise_Monotone_Quintic_Fortran/paper/4-analysis.tex
Updated code package sent off last night. New tex revisions from Dr. Watson received, implementing.

\heading{4. COMPLEXITY AND SENSITIVITY}

Both Algorithms 1 and 2 have ${\cal O}(1)$ runtime. Given a fixed
schedule for shrinking derivative values, Algorithm 3 has a ${\cal
  O}(n)$ runtime for $n$ data points. In execution, the majority of
time is spent solving the banded linear system of equations for
B-spline coefficients. The quadratic facet model produces a unique
sensitivity to input perturbation, as small changes in input may cause
data points to be associated with different local quadratic
interpolants. While this makes the technique susceptible to random
perturbations, it generally produces high quality monotone quintic
splines and is preferred for that reason.

### 2020-06-23 Research_[2020-05-18]_Piecewise_Monotone_Quintic_Fortran/paper/mqsi20.tex
Updated code package sent off last night. New tex revisions from Dr. Watson received, implementing.

@@
-% Algorithm XXX: MQSI---Monotone quintic spline interpolation,
+% Algorithm XXXX: MQSI---Monotone quintic spline interpolation,
 % Thomas C.H. Lux, Layne T. Watson, Tyler H. Chang,
-% William Thacker, Kirk W. Cameron, and Yili Hong
+% and William Thacker
 % ACM Trans. Math. Software, ?/?/2020.
@@
 
-% eplain package, mainly for its bibliography management.
-\input eplain
+% \cite is a placeholder to use for references while writing.
+\def\cite#1{[{\tt REF}]}
 
@@
 % Commands for inserting an 'algorithm' section.
-\def\indented#1{{\leftskip=1cm}{#1}{\leftskip=0cm}}
-\def\comment#1{\vskip 10mm {\it #1} \vskip 10mm}%
 \def\thickrule{\hrule\hrule\hrule}
+\def\codent{\hskip 1.8em} % <- 'tab' for indentation.
 
@@
 
-\input epsf %Postscript figures.
+%Postscript figures.
+\input epsf
 \def\lcaption#1#2{{\noindent\rmVIII Fig.~#1. #2\par}}
@@
 \def\leftrh{Lux et al.}
-\def\rightrh{Algorithm XXX: MQSI}
+\def\rightrh{Algorithm XXXX: MQSI}
 
-{\baselineskip=24pt \leftline{\hlvXVIII Algorithm XXX: MQSI: Monotone Quintic}
+{\baselineskip=24pt
+\leftline{\hlvXVIII Algorithm XXXX: MQSI---Monotone Quintic}
 \leftline{\hlvXVIII Spline Interpolation}}
 \bigskip\medskip
-\leftline{\hlv THOMAS C.H. LUX, LAYNE T. WATSON, TYLER H. CHANG,}
-\smallskip
-\leftline{\hlv WILLIAM THACKER, KIRK W. CAMERON, AND YILI HONG}
+\leftline{\hlv THOMAS C. H. LUX and LAYNE T. WATSON}
 \smallskip
 \leftline{\hlv Virginia Polytechnic Institute and State University}
+\smallskip
+\leftline{\hlv TYLER H. CHANG}
+\smallskip
+\leftline{\hlv Argonne National Laboratory}
+\smallskip
+\leftline{\hlv WILLIAM I. THACKER}
+\smallskip
+\leftline{\hlv Winthrop University}
 \bigskip
@@
 -\parindent{\parindent=0pt\rmVIII Authors' addresses:
-T. C. H. Lux, T. H. Chang, K. W. Cameron,
-Department of Computer Science, L. T. Watson,
-Departments of Computer Science, Mathematics, and Aerospace and Ocean
-Engineering, Y. Hong, Department of Statistics, Virginia
-Polytechnic Institute \& State University, Blacksburg, VA 24061;
-e-mail: {\ttVIII tchlux@vt.edu}. \xpar
+T. C. H. Lux, L. T. Watson, Departments of Computer Science, Mathematics,
+and Aerospace and Ocean Engineering, Virginia Polytechnic Institute \&
+State University, Blacksburg, VA 24061; e-mails: {\ttVIII tchlux@vt.edu,
+ltwatson@computer.org}; T. H. Chang, Mathematics and Computer Science
+Division, Argonne National Laboratory, 9700 South Cass Avenue, Bldg. 240,
+Lemont, IL 60439; e-mail: {\ttVIII thchang@vt.edu}; W. I. Thacker, Winthrop
+University, Rock Hill, SC 47405; {\ttVIII thackerw@winthrop.edu}.
+
+\xpar
 % Permission to make digital/hard copy of
@@
 % Abstract and Introduction
-{\rmVIII\parindent=0pt MQSI contains a serial code written in Fortran
-  1995 for constructing monotone quintic spline interpolants to
-  data. Using sharp theoretical monotonicity constraints, first and
-  second derivative estimates at data provided by a quadratic facet
-  model are refined to produce a $C^2$ monotone interpolant. This
-  paper includes algorithm and implementation details, complexity and
-  sensitivity analyses, usage information, and a brief performance
-  study.
+{\rmVIII\parindent=0pt MQSI is a Fortran 2003 subroutine for constructing
+monotone quintic spline interpolants to monotone data. Using sharp
+theoretical monotonicity constraints, first and second derivative estimates
+at data provided by a quadratic facet model are refined to produce a $C^2$
+monotone interpolant.  Algorithm and implementation details, complexity
+and sensitivity analyses, usage information, and a brief performance study
+are included.
 
@@
 
-\bigskip\hrule\bigskip\medskip
-\heading{1. INTRODUCTION}
+%% Section 1
+\input 1-introduction
 
-Many domains of science rely on smooth approximations to real-valued
-functions over a closed interval. Piecewise polynomial functions
-(splines) provide the smooth approximations for animation in graphics
-\cite{herman2015techniques,quint2003scalable}, aesthetic structural
-support in architecture \cite{brennan2019measure}, efficient
-aerodynamic surfaces in automotive and aerospace engineering
-\cite{brennan2019measure}, prolonged effective operation of electric
-motors \cite{berglund2009planning}, and accurate nonparametric
-approximations in statistics \cite{knott2012interpolating}. While
-polynomial interpolants and regressors apply broadly, splines are
-often a good choice because they can approximate globally complex
-functions while minimizing the local complexity of an approximation.
+%% Section 2
+\input 2-monotone-quintic
 
-It is often the case that the true underlying function or phenomenon
-being modeled has known properties e.g., convexity, positivity,
-various levels of continuity, or monotonicity. Given a reasonable
-amount of data, it quickly becomes difficult to achieve desirable
-properties in a single polynomial function. In general, the
-maintenance of function properties through interpolation/regression is
-referred to as {\it shape preserving}
-\cite{fritsch1980monotone,gregory1985shape}. The specific shapes this
-work will achieve in approximations are monotonicity and $C^2$
-continuity. These properties are chiefly important to the
-approximation of cumulative distribution functions and subsequently
-the effective generation of random numbers from a specified
-distribution.
+%% Section 3
+\input 3-spline-rep
 
-In statistics especially, the construction of a monotone interpolating
-spline that is $C^2$ continuous is meaningfully useful
-\cite{ramsay1988monotone}. A function with these properties could
-approximate a cumulative distribution function to a high level of
-accuracy with relatively few intervals. A twice continuously
-differentiable approximation to a cumulative distribution function
-(CDF) would also produce a corresponding probability density function
-(PDF) that is continuously differentiable, which is a property many
-standard parametric distributions maintain.
+%% Section 4
+\input 4-analysis
 
-The currently available software for monotone piecewise polynomial
-interpolation includes quadratic \cite{he1998monotone}, cubic
-\cite{fritsch1980monotone}, and (with limited application) quartic
-\cite{wang2004rational,piah2011improved,yao2018unconditionally}
-cases. In addition, a statistical method for bootstrapping the
-construction of an arbitrarily smooth monotone fit exists
-\cite{leitenstorfer2006generalized}, but the method does not take
-advantage of known analytic properties related to quintic
-polynomials. Theory has been provided for the quintic case
-\cite{ulrich1994positivity,hess1994positive}, however this theory has
-not yet been used to construct a monotone quintic spline interpolation
-routine. Recent work suggests that the lack of quintic software may be
-due to a general unawareness of the theory
-\cite{xie2018semiparametric}.
+%% Section 5
+\input 5-applications
 
-\topinsert
-\centerline{\epsfxsize=4.5truein \epsffile{vt_logo.eps}}
-{\narrower\noindent\rmVIII Fig.\ 1.
-Example polynomials that interpolate function values at the ends of
-the interval $[0,1]$. The first only interpolates the function values
-$g_2(0) = 1$ and $g_2(1) = 0$, making it the order two polynomial
-$g_2(x) = 1 - x$. For the second plot $g_4(x) = 8x^3 - 12x^2 + 4x +
-1/2$, which is order four and interpolates the values $g_4(0) = 1/2$,
-$g_4'(0) = 4$, $g_4(1) = 1/2$, $g_4'(1) = 4$. Finally the third plot
-shows the order six polynomial $g_6(x) = - 64x^5 + 160x^4 - 140x^3 +
-50x^2 - 7x + 1$ interpolating the function values $g_6(0) = 1$,
-$g_6'(0) = -7$, $g_6''(0) = 100$, $g_6(1) = 0$, $g_6'(1) = -7$,
-$g_6''(1) = -100$. Notice that interpolating the same fixed number of
-function values at each endpoint will always result in an even order
-interpolating polynomial.
-\par}
-\endinsert
+%% Acknowledgements
+%% \heading{Acknowledgments}
+%% This work was supported by the National Science Foundation Grants CNS-1565314 and CNS-1838271.
 
-The importance of piecewise quintic interpolation over lower order
-approximations can be simply demonstrated. In general, the order of a
-polynomial determines the number of function values it can
-interpolate, and the growth rate of error away from the interpolated
-function values. As demonstrated in Figure
-\ref{fig:spline_demonstration}, it can be seen that matching a value
-at either end of the interval requires an order two (linear)
-approximation and each additional derivative at the ends of the
-interval raises the necessary polynomial order by two. Hence, only
-even order (odd degree) interpolating splines are broadly
-applicable. The body of this work is composed of a novel algorithm for
-enforcing monotonicity on quintic polynomial pieces, then extending
-that solution to work on quintic splines.
+% ====================================================================
+% Bibliography
+\heading{\hlvVIII BIBLIOGRAPHY} {
+\baselineskip=2pt \parskip=100mm
+\frenchspacing
+% Temporary 'caps', 'sl', and 'rm'.
+\let\caps=\capsVIII
+\let\sl=\slVIII
+\let\rm=\rmeight \rmVIII
 
-The major contribution of this work is an algorithm for constructing
-monotone quintic interpolating splines that utilizes existing quintic
-monotonicity theory. The remainder of this paper is structured as
-follows: Section \ref{sec:monotone-cubic} summarizes the existing
-monotone cubic spline interpolation methodology, Section
-\ref{sec:monotone-quintic} presents an algorithm for constructing
-monotone quintic spline interpolants, Section \ref{sec:results} offers
-experiments and results with cubic and quintic methods, Sections
-\ref{sec:discussion} and \ref{sec:conclusion} discuss results and
-conclude.
+% \ys makes a half-em horizontal space.
+\def\ys{\hskip 1em minus .5em}
 
-\heading{1.1 \enspace Computing a Monotone Cubic Interpolant}
+%\refj is for regular journal articles
+%  #1 -- Author(s)
+%  #2 -- Year
+%  #3 -- Title
+%  #4 -- Journal
+%  #5 -- Volume, number, and pages
+\def\refj#1#2#3#4#5{\noindent\hangindent=10pt\hangafter=1
+{\caps #1}\ys #2.\ys {\rm #3}. {\sl #4}, #5.\vskip 6pt}
 
-The current state-of-the-art monotone interpolating spline with a
-mathematical software implementation is piecewise cubic, continuously
-differentiable, and was first proposed in \cite{fritsch1980monotone}
-then expanded upon in \cite{carlson1985monotone}. Let $\pi: x_0 = k_1
-< k_2 < \cdots < k_n = x_1$ be a partition of the interval
-$[x_0,x_1]$. Let $f: {\bbb R} \rightarrow {\bbb R}$ be $C^2$, and
-$\bigl\{f(k_i)\bigr\}_{i=1}^n$ and $\bigl\{\Delta_i\bigr\}_{i=1}^n$ be
-given sets of function and derivative values at the partition points
-for a monotone function $f$. Either $f(k_i) \leq f(k_{i+1})$, $i=1$,
-$\ldots$, $n-1$, and $\Delta_i\ge0$, $i=1$, $\ldots$, $n$, or $f(k_i)
-\geq f(k_{i+1})$, $i=1$, $\ldots$, $n-1$, and $\Delta_i\le0$, $i=1$,
-$\ldots$, $n$. Let $\hat f$ be a piecewise cubic polynomial defined in
-each subinterval $I_i = [k_i, k_{i+1}]$ by
-$$ h_i = k_{i+1} - k_{i}, \quad u(t) = 3t^2 - 2t^3, \quad p(t) = t^3 - t^2,$$
-$$\hat f(x) = f(k_i) u\big((k_{i+1} - x) / h_i\big) + f(k_{i+1}) u\big((x - k_i) / h_i\big) $$
-$$ - h_i\Delta_i p\big((k_{i+1}-x)/h_i\big) + h_i\Delta_{i+1} p\big((x-k_i)/h_i\big).$$ 
-Notice that a trivially monotone spline results when $\Delta_i = 0$,
-for $i = 1$, $\ldots$, $n$. However, such a spline has too many
-{\it wiggles} for most applications. \cite{carlson1985monotone}
-show that simple conditions on the derivative values can guarantee
-monotonicity, and that these conditions can be enforced in a way that
-ensures modifications on one interval will not break the monotonicity
-of cubic polynomials over any neighboring intervals. If $f(k_i) =
-f(k_{i+1})$, take $\Delta_i = \Delta_{i+1} =0$ and $\alpha =\beta =1$,
-otherwise let $\alpha = \big(\Delta_i (k_{i+1}-k_i)\big) /
-\big(f(k_{i+1}) - f(k_i)\big)$ and $\beta =
-\big(\Delta_{i+1}(k_{i+1}-k_i)\big) / \big(f(k_{i+1}) -
-f(k_i)\big)$. Monotonicity of a cubic polynomial over a subinterval
-can be maintained by ensuring that $\alpha$ and $\beta$ reside in any
-of the regions depicted in Figure \ref{fig:projection}.
+%\refb is for books
+%  #1 -- Author(s)
+%  #2 -- Year
+%  #3 -- Title
+%  #4 -- Publisher, City, State
+\def\refb#1#2#3#4{\noindent\hangindent=10pt\hangafter=1
+{\caps #1}\ys #2.\ys{\sl #3}. {\rm #4}.\vskip 6pt}
 
-\topinsert
-\centerline{\epsfxsize=4.5truein \epsffile{vt_logo.eps}}
-{\narrower\noindent\rmVIII Fig.\ 2.
-These are the feasible regions of monotonicity for cubic splines and
-the projections that make a cubic polynomial piece monotone. The
-regions themselves are numbered 1--4 corresponding to their original
-description in Fritsch and Carlson (1980). One point is projected onto
-each region as a demonstration.
-\par}
-\endinsert
+%\reft is for tech. reports, theses, and conferences
+%  #1 -- Author(s)
+%  #2 -- Year
+%  #3 -- Title
+%  #4 -- Technical report ID, Publisher, City, State
+\def\reft#1#2#3#4{\noindent\hangindent=10pt\hangafter=1
+{\caps #1}\ys #2.\ys{\rm #3}. {\rm #4}.\vskip 6pt}
 
-The actual region of monotonicity for a cubic polynomial is larger,
-but projection of $(\alpha, \beta)$ into one of these regions ensures
-that monotonicity will be achieved and not violated for neighboring
-regions. The user must decide which region is most appropriate for the
-projections based on the application, Fritsch and Carlson recommend
-using Region 2.
+%% -------------------------------------------------------------------
 
-While the cubic monotonicity case affords such a concise solution, the
-region of monotonicity is not so simple in the quintic case. In the
-next section, an algorithm for performing a projection similar to
-those for cubic polynomials is proposed.
+% berglund2009planning
+\refj{Berglund, T., Brodnik, A., Jonsson, H., Staffanson, M., and
+  Soderkvist, I.}{2009}{Planning smooth and obstacle-avoiding B-spline
+  paths for autonomous mining vehicles}{IEEE Transactions on
+  Automation Science and Engineering}{7(1):167--172}
 
-\heading{2. MONOTONE QUINTIC INTERPOLATION}
+% brennan2019measure
+\refj{Brennan, A.}{2020}{Measure, Modulation and Metadesign: NC
+  Fabrication in Industrial Design and Architecture}{Journal of Design
+  History}{33(1):66--82}
 
-The following section is composed of three algorithms that together
-are used to construct a monotone quintic spline interpolant. Without
-loss of generality, the algorithms will only consider the monotone
-increasing (nondecreasing) case. The monotone decreasing case is
-handled similarly. Algorithm \ref{alg:check-monotone} checks
-monotonicity, Algorithm \ref{alg:make-monotone} enforces monotonicity
-on an order six polynomial piece of the quintic spline, and Algorithm
-\ref{alg:monotone-spline} uses the previous two algorithms to enforce
-monotonicity for the entire quintic spline.
+% fritsch1980monotone
+\refj{Fritsch., F and Carlson., R.}{1980}{Monotone piecewise cubic
+  interpolation}{SIAM Journal on Numerical Analysis}{17(2):238–-246}
 
-In pseudocode, a function {\tt binary\_search($g,$ $a,$ $b$)} is
-used, where $a,b\in S\subset {\bbb R}^p$ for convex $S$, $g: S
-\rightarrow \{${\tt FALSE}, {\tt TRUE}$\}$ is a right continuous
-Boolean function, $g(b) = \hbox{{\tt TRUE}}$, and for $\mu \in
-[0,1]$, $g\big((1-\mu)a + \mu b\big) =\hbox{{\tt TRUE}}
-\Longrightarrow g\big((1-\nu)a + \nu b\big) =\hbox{{\tt TRUE}}$ for
-$\mu \le \nu\le1$. The search returns $\big((1-c)a + c b\big)$ for the
-smallest $c \in [0,1]$ such that $g\big((1-c)a + c b\big) =
-\hbox{{\tt TRUE}}$.
+% fritsch1982LLNL
+\reft{Fritsch, F. N.}{1982}{Piecewise cubic Hermite interpolation
+  package, final specifications}{Computer Documentation UCID-30194,
+  Lawrence Livermore National Laboratory}
 
-These algorithms make use of first and second derivatives of the approximated function where $D$ denotes the differentiation operator.
-In the case that the first and second derivative information is not provided along with function values, the derivatives are estimated with finite differences of the function values.
-The final quintic spline is represented as a piecewise polynomial using the Newton form for each polynomial piece, and ultimately converted to a $B$-spline representation for evaluation.
+% gregory1985shape
+\reft{Gregory, J. A.}{1985}{Shape preserving spline
+  interpolation}{19850020252, Brunel University}
 
-\def\newline{\hfil\break}
+% haralick1981facet
+\refj{Haralick, R. M., and Watson, L. T.}{1981}{A facet model for
+  image data.}{Computer Graphics and Image Processing}{15(2):113--129}
 
-\parindent=0pt
-\vskip 5mm
-\thickrule
-{\bf Algorithm 1:} {\tt is\_monotone}$\big(x_0, x_1, f \big)$ \newline
-where $x_0$, $x_1 \in {\bbb R}$, $x_0 < x_1$, and $f$ is an order six
-polynomial defined by $f(x_0)$, $Df(x_0)$, $D^2f(x_0)$, $f(x_1)$,
-$Df(x_1)$, $D^2f(x_1)$. Returns {\tt TRUE} if $f$ is monotone
-increasing on $[x_0,x_1]$.
-\hrule
-\vskip 3mm
-\numberedlist
-{\tt
-\li if $\big(f(x_0) = f(x_1)\big)$ and not $\big( 0 = Df(x_0) = Df(x_1) = D^2f(x_0) = D^2f(x_1) \big)$
-\li \indented{return FALSE} \newline
-end if \newline
-\li if $\big(Df(x_0) < 0$ or $Df(x_1) < 0\big)$ \newline
-    \indented{return FALSE} \newline
-    end if \newline
-    \comment{The necessity of these first two steps follows directly from the fact that $f$ is $C^2$.
-The next case is in accordance with a simplified condition for quintic monotonicity that reduces to one of cubic positivity studied in \cite{schmidt1988positivity}, where $\alpha$, $\beta$, $\gamma$, and $\delta$ are defined in terms of values and derivatives of $f$ at $x_0$ and $x_1$.
-Step \ref{alg1-alpha} checks for the necessary condition that $\alpha \ge 0$, Step \ref{alg1-beta} checks $\beta \geq \alpha$, and Step \ref{alg1-gamma} checks $\gamma \geq \delta$, all from \cite{schmidt1988positivity}.
-If all necessary conditions are met, then the order six piece is monotone and Step \ref{alg1-simplified-true} concludes this check.}
-  if $\big(Df(x_0) = 0$ or $Df(x_1) = 0\big)$ \newline
-  \indented{$w := x_0 - x_1$} \newline
-  \indented{$v := f(x_0) - f(x_1)$} \newline
-  \indented{if $\big(D^2f(x_1) > {-4}Df(x_1) / w \big)$ return FALSE} \newline
-  \indented{if $\big(D^2f(x_1) < (3w D^2f(x_0) - 24 Df(x_0) - 32
-    Df(x_1) + 60v/w) / (5w) \big)$ return FALSE} \newline
-  \indented{if $\big(D^2f(x_0) < 3 Df(x_0) / w \big)$ return FALSE} \newline
-  \indented{return TRUE} \newline
-    end if \newline
-    \comment{The following code considers the remaining case where $Df(x_0) \neq 0$ and $Df(x_1) \neq 0$.} \newline
-  $\displaystyle A := Df(x_0){x_1 - x_0 \over f(x_1) - f(x_0)}$ \newline
-  $\displaystyle B := Df(x_1) {x_1 - x_0 \over f(x_1) - f(x_0)}$ \newline
-    \comment{The variables $A$ and $B$ correspond directly to the theoretical foundation for positive quartic polynomials established in \cite{ulrich1994positivity}, first defined after Equation (18).} \newline
-  $\displaystyle \gamma_0 := 4 {Df(x_0) \over Df(x_1)} (B/A)^{3/4}$ \newline
-  $\displaystyle \gamma_1 := {x_1 - x_0 \over Df(x_1)} (B/A)^{3/4}$ \newline
-  $\alpha_0 := 4 (B/A)^{1/4}$ \newline
-  $\displaystyle \alpha_1 := -{x_1 - x_0 \over Df(x_1)} (B/A)^{1/4}$ \newline 
-  $\displaystyle \beta_0 := 30 - {12 \big(Df(x_0) + Df(x_1)\big) (x_1 - x_0) \over \big(f(x_1) - f(x_0)\big) \sqrt{A}\sqrt{B}}$ \newline
-  $\displaystyle \beta_1 := {-3 (x_1 - x_0)^2 \over 2 \big(f(x_1) - f(x_0)\big) \sqrt{A} \sqrt{B}} $ \newline
-    \comment{The $\gamma$, $\alpha$, and $\beta$ terms with subscripts $0$ and $1$ are algebraic reductions of the simplified conditions for satisfying Theorem 2 in Equation (16) of \cite{ulrich1994positivity}.
-These terms with subscripts $0$ and $1$ make the computation of $\alpha$, $\beta$, and $\gamma$ functions of the second derivative terms, as seen in Step \ref{alg1-abg} below.} \newline
-  $\gamma := \gamma_0 + \gamma_1 D^2f(x_0)$ \newline
-  $\alpha := \alpha_0 + \alpha_1 D^2f(x_1)$ \newline
-  $\beta := \beta_0 + \beta_1 \big(D^2f(x_0) - D^2f(x_1)\big)$ \newline
-  if $(\beta \leq 6)$ return $\big(\alpha > - (\beta + 2) / 2\big)$ \newline
-  else $\qquad$ return $\big(\gamma > -2 \sqrt{\beta - 2}\big)$ \newline
-  end if \newline
-}
-\endnumberedlist
-\vskip 2mm
-\hrule
-\vskip 3mm
-\parindent=10pt
+% he1998monotone
+\refj{He, X. and Shi, P.}{1998}{Monotone B-spline smoothing}{Journal
+  of the American Statistical Association}{93(442):643--650}
 
-%% ----------------------------------------------------------------------
+% herman2015techniques
+\reft{Herman, D. L., and Oftedal, M. J.}{2015}{Techniques and
+  workflows for computer graphics animation systems}{US Patent
+  9,216,351}
 
-The reason for structuring the $\alpha$, $\beta$, and $\gamma$ computations in terms of the second derivative of $f$ (seen in Step \ref{alg1-abg} of Algorithm \ref{alg:check-monotone}) will become more apparent later.
-The next problem to consider is that of making a nonmonotone order six polynomial piece into a monotone one by modifying its first and second derivative values at the ends of an interval.
-Note that the actual value of the function at the ends of the interval is not modified, as the resulting polynomial needs to interpolate.
+% hess1994positive
+\refj{He{\ss}, W. and Schmidt, J. W.}{1994}{Positive quartic, monotone
+  quintic $C^2$-spline interpolation in one and two dimensions}{Journal
+  of Computational and Applied Mathematics}{55(1):51–-67}
 
-Notice that both Algorithms \ref{alg:check-monotone} and \ref{alg:make-monotone} have ${\cal O}(1)$ runtime assuming a fixed level of precision is chosen beforehand.
-A constant number of operations are needed for verifying monotonicity (Algorithm \ref{alg:check-monotone}), while a constant set of operations and a single binary search are performed for enforcing monotonicity (Algorithm \ref{alg:make-monotone}).
-The search requires a fixed number of steps to achieve any predetermined relative precision, since its accuracy is predetermined and not a function of the problem at hand.
-Also note that an efficient implementation of Algorithm \ref{alg:make-monotone} only needs to recalculate Steps \ref{alg1-abg} and \ref{alg1-last} of Algorithm \ref{alg:check-monotone} during the binary search.
-Next, an algorithm for constructing a monotone quintic spline interpolant is presented.
+% knott2012interpolating
+\reft{Knott, G. D.}{2012}{Interpolating cubic splines}{Springer Science
+  \& Business Media, Volume 18}
 
-%% ----------------------------------------------------------------------
+% leitenstorfer2006generalized
+\refj{Leitenstorfer, F. and Tutz, G.}{2006}{Generalized monotonic
+  regression based on B-splines with an application to air pollution
+  data}{Biostatistics}{8(3):654–-673}
 
-\vskip 5mm
-\thickrule
-{\bf Algorithm 2:} {\tt monotone\_spline}$\big((k_1$, $\ldots$, $k_n)$, $f$, $s\big)$
-where $(k_1$, $\ldots$, $k_n)$ is an increasing sequence of real
-numbers, $f$ is an order six piecewise polynomial with breakpoints
-$k_1$, $\ldots$, $k_n$ defined by the data
-$\bigl\{f(k_i)\bigr\}_{i=1}^n$, $\bigl\{Df(k_i)\bigr\}_{i=1}^n$,
-$\bigl\{D^2f(k_i)\bigr\}_{i=1}^n$, and $s > 1$ is an integer shrink
-factor.
-\hrule
-\vskip 3mm
-    \newline create queue
-      \newline create changed$(1$, $\ldots$, $n) := $ FALSE
-      \newline create step\_size$(1$, $\ldots$, $n) := \bigl(Df(k_1)/s$, $\ldots$, $Df(k_n)/s\bigr)$
-      \comment{The three variables defined in Step \ref{step:create} are used to ensure the eventual achievement of monotonicity.
-The {\tt queue} is a standard first in first out (FIFO) queue with {\tt enqueue} and {\tt dequeue} operations.
-The array {\tt changed} contains Booleans describing whether or not a breakpoint belongs to an interval that has been modified to enforce monotonicity.
-The {\tt step\_size} array contains the real-valued derivative decrement step sizes to use in the search for a valid monotone spline.}
-    \newline for $i := 1$, $\ldots$, $n-1$
-      \indented{enqueue $\bigl((k_i, k_{i+1})\bigr)$}
-      end for
-      \comment{Initially, all intervals must be checked for monotonicity. The following loop will continue until all intervals are verified as monotone without need for modification.}
-    \newline while $($size$($queue$) > 0)$
-    \newline \indented{$(k_j$, $k_{j+1}) := $ dequeue}
-    \newline \indented{if $\big($not is\_monotone$(k_j$, $k_{j+1}$, $f)\big)$}
-    \newline \indented{\indented{if $\big($changed$(j)\big) \qquad \ Df(k_j) := Df(k_j) -{}$step\_size$(j)$}}
-      \indented{\indented{if $\big($changed$(j+1)\big) \quad Df(k_{j+1}) := Df(k_{j+1}) -{}$step\_size$(j+1)$}}
-      \indented{\indented{\comment{If a breakpoint belongs to an interval that has been previously modified, then the enforced monotonicity conditions on the second derivatives of adjacent intervals must contradict one another.
-In turn, the involved first derivatives are decreased towards zero by a predetermined step size.}}}
-    \newline \indented{\indented{make\_monotone$(k_j,\ k_{j+1},\ f)$}}
-    \newline \indented{\indented{changed$(j) :=$ TRUE}}
-      \indented{\indented{changed$(j+1) :=$ TRUE}}
-    \newline \indented{\indented{if $\big(j > 1$ and $(k_{j-1}, k_j) \not\in$ queue$\big)$ enqueue $\bigl((k_{j-1}, k_j)\bigr)$}}
-      \indented{\indented{enqueue $\bigl((k_j, k_{j+1})\bigr)$}}
-      \indented{\indented{if $\big(j+1 < n$ and $(k_{j+1}, k_{j+2}) \not\in$ queue$\big)$ enqueue $\bigl((k_{j+1}, k_{j+2})\bigr)$}}
-      \indented{\indented{\comment{Step \ref{step:changed} records the endpoints of the current interval as having been changed, while \ref{step:requeue} adds adjacent intervals to {\tt queue} that may have inadvertently been made nonmonotone by the changes to the present interval.}}}
-      \indented{end if}
-      end while
-\vskip 2mm
-\hrule
-\vskip 3mm
+% lux2020thesis
+\reft{Lux, T. C. H.}{2020}{Interpolants and error bounds for modeling
+  and predicting variability in computer systems}{Thesis, Virginia
+  Polytechnic Institute and State University, Blacksburg, Virginia}
 
-%% ----------------------------------------------------------------------
+% piah2011improved
+\refj{Piah, A. R. M. and Unsworth, K.}{2011}{Improved sufficient
+  conditions for monotonic piecewise rational quartic
+  interpolation}{Sains Malaysiana}{40(10):1173–-1178}
 
-It is mentioned in \cite{ulrich1994positivity} that for sufficiently small $Df(k_i)$ and $Df(k_{i+1})$ the admissible solution interval of second derivative values becomes arbitrarily large.
-It can also be seen that decreasing the assigned derivative values towards zero will eventually cause Steps \ref{alg1-alpha} through \ref{alg1-gamma} of Algorithm \ref{alg:check-monotone} to all fail, resulting in Step \ref{alg1-simplified-true} returning {\tt TRUE}.
+% quint2003scalable
+\refj{Quint, A.}{2003}{Scalable vector graphics}{IEEE
+  MultiMedia}{10(3):99–102}
 
-If successive monotonicity fixes are required for all intervals, the worst case runtime of Algorithm \ref{alg:monotone-spline} is ${\cal O}(s n)$ for $n$ breakpoints and shrink factor $s$.
-In practice this worst case has been observed to be unlikely.
+% ramsay1988monotone
+\refj{Ramsay, J. O.}{1988}{Monotone regression splines in
+  action}{Statistical Science}{3(4):425-–441}
 
-\heading{3. EXPERIMENTAL RESULTS}
+% ulrich1994positivity
+\refj{Ulrich, G. and Watson, L. T.}{1994}{Positivity conditions for
+  quartic polynomials}{SIAM Journal on Scientific
+  Computing}{15(3):528–-544}
 
-Given the increased order of approximation, it is naturally expected that some accuracy benefit could be gained by using a quintic spline over a cubic spline.
-Experiments 1 and 2 below test some accuracy differences between monotone cubic and quintic splines.
-Experiment 3 analyzes the number of operations performed by Algorithm \ref{alg:monotone-spline} with increasingly large samples of monotone data.
+% wang2004rational
+\refj{Wang, Q. and Tan, J.}{2004}{Rational quartic spline involving
+  shape parameters}{Journal of Information and Computational
+  Science}{1(1):127–-130}
 
-\heading{3.1 \enspace Approximating a Trigonometric Function}
+% yao2018unconditionally
+\refj{Yao, J. and Nelson, K. E.}{2018}{An Unconditionally Monotone
+  $C^2$ Quartic Spline Method with Nonoscillation
+  Derivatives}{Advances in Pure Mathematics}{8(LLNL-JRNL-742107)}
 
-For this experiment, the function $\sin(x) + x$ is considered over the interval $[0,(5/2)\pi]$ as seen in Figure \ref{fig:cubic_quintic_sin}.
-Given an increasing number of points, the maximum error of cubic and quintic approximations is shown in Figure \ref{fig:experiment_1}.
-The quintic approximation consistently has a maximum error that is roughly half that of the cubic approximation, given the same number of points.
-There is an increase in the error gap between the two approximations as more points are added.
-The quintic spline requires only $700$ points to approximate the function to the same accuracy achieved by the cubic with $1000$ points.
-
-%% \begin{figure}[h]
-%%   \centering
-%%   \includegraphics[width=.7\textwidth]{cubic-quintic-sin}
-%%   \caption{Depicted above are the monotone cubic and quintic spline interpolants of the function $\sin(x) + x$ over the points $[0, (5/2) \pi]$.
-Notice that the maximum error of the cubic interpolant is larger, because it only captures first derivative information at the points.
-The quintic interpolant captures both first and second derivative information at the points.}
-%% \end{figure}
-
-%% \begin{figure}
-%%   \centering
-%%   \includegraphics[width=.7\textwidth]{experiment_1_errors}
-%%   \caption{The error in the quintic and cubic approximations of $\sin(x) + x$ with increasing number of points.
-Both axes are log scaled, where $\mu$ means $\times 10^{-6}$ and n means $\times 10^{-9}$ on the vertical axis.
-The quintic approximation generally provides an approximation with about half of the maximum error of the cubic approximation, while the relative gap between grows with increasing number of points.
-In order to achieve the same absolute error the quintic spline requires ten fewer points at max error $10^{-4}$, thirty fewer points at max error $10^{-6}$, and $300$ fewer points at max error $10^{-8}$.}
-%% \end{figure}
-
-\heading{3.2 \enspace Random Number Generation}
-
-Consider the cumulative distribution function (CDF) defined by a mixture of three Gaussian distributions and visualized in Figure \ref{fig:experiment_2_distribution}.
-The distributions have weights $(.3, .6, .1)$, means $(.2, .45, .85)$ and standard deviations $(.05, .08, .03)$.
-$200$ random samples are generated from this distribution with both cubic and quintic approximations from four points, and over $100$ trials the resulting spread of empirical distribution functions (EDFs) are visualized in Figure \ref{fig:experiment_2_results}.
-
-While an increased number of points could be used to decrease the error in either approximation, the purpose of this experiment is to demonstrate the effect approximation error has on random number generation.
-The errors in the CDFs are magnified by the variance involved in randomly sampling from a distribution.
-As a result, while the cubic and quintic spline approximations have similar maximum error, the cubic distribution is significantly distorted around the first and second modes.
-
-%% \begin{figure}
-%%   \centering
-%%   \includegraphics[width=.8\textwidth]{experiment_2_distribution}
-%%   \caption{The cumulative distribution function (top) and probability density function (bottom) for the Gaussian mixture distribution in Experiment 2.
-The distributions have weights $(.3, .6, .1)$, means $(.2, .45, .85)$ and standard deviations $(.05, .08, .03)$.
-Random samples are generated from this mixture distribution by evaluating the inverse of the CDF at random numbers uniformly distributed in the range $[0,1]$.}
-%% \end{figure}
-
-%% \begin{figure}
-%%   \centering
-%%   \includegraphics[width=.4\textwidth]{experiment_2_cubic_dist}
-%%   \includegraphics[width=.4\textwidth]{experiment_2_quintic_dist}
-%%   \caption{Both cubic and quintic splines are used to approximate the CDF of a Gaussian mixture distribution with three components.
-Four CDF points are used to build the approximation, $200$ random samples are generated with the approximate CDFs over $100$ trials.
-The resulting cloud of empirical distribution function (EDF) values is seen above.
-The maximum approximation error in the CDFs are similar, $.08$ for the cubic and $.05$ for the quintic.
-However, the maximum EDF differences observed are doubled by the variance of random samples, $.15$ for the cubic and $.1$ for the quintic.}
-%% \end{figure}
-
-%% - generate large sets of random monotone data, show Table of number
-%%   of metrics growing with increasing number of points
-
-%% - show visual of the \textit{error}
-
-\heading{3.3 \enspace Random Monotone Data}
-
-This experiment studies the number of times the algorithms {\tt is\_monotone} and {\tt make\_monotone} are executed for increasingly large sequences of random monotone data.
-The minimum, median, and maximum number of times that Algorithms \ref{alg:check-monotone} and \ref{alg:make-monotone} are executed is recorded, as well as the number of steps taken in all calls to {\tt binary\_search}.
-The number of calls and steps grows linearly with $n$ as expected, requiring roughly one call to {\tt make\_monotone} and $20$ binary search steps to create a monotone piece.
-Some (less common) problems require an average of three calls to {\tt make\_monotone} per interval.
-
-%% \begin{table}
-%%   \centering
-%%   \caption{The minimum, median, and maximum number of checks, fixes, and binary search steps required in the execution of {\tt make\_spline} for increasing size sequences, $n$, over $100$ randomly generated sets of monotone data.
-Notice the maximum for each counter is often significantly greater than the minimum and median, because the distribution of each counter is skewed right.}
-
-%%   \begin{tabular}{c|c c c|c c c|c c c}
-%%     \hline
-%%     \multirow{2}{*}{$n$}
-%%       & \multicolumn{3}{c|}{Checks} & \multicolumn{3}{c}{Fixes} & \multicolumn{3}{|c}{Search Steps} \\
-%%       & min & median & max & min & median & max & min & median & max \\
-%%     \hline
-%%     5 & 4 & 6 & 274 & 0 & 1 & 154 & 3 & 31 & 2006\\
-%%     25 & 28 & 47 & 378 & 2 & 14 & 225 & 80 & 348 & 3742\\
-%%     50 & 61 & 124 & 484 & 6 & 40 & 238 & 217 & 851 & 4445\\
-%%     75 & 102 & 216 & 842 & 14 & 74 & 411 & 463 & 1491 & 6333\\
-%%     100 & 142 & 306 & 776 & 21 & 107 & 380 & 629 & 2140 & 7414\\
-%%     \hline
-%%   \end{tabular}
-%% \end{table}
-
-\heading{4. DISCUSSION}
-
-The monotone quintic spline interpolant provides a distinct increase in accuracy over the monotone cubic variant.
-The computational complexity of this algorithm is identical to the existing cubic method and the cost of evaluating the spline after construction is unchanged.
-While the number of computations required per interval has increased, the number of points required to achieve the same level of accuracy has decreased.
-In the present state, the algorithm for enforcing monotonicity on a spline is not as trivially parallel as the cubic algorithm, however it can still be parallelized.
-The checks for monotonicity across all intervals are independent and the monotonicity adjustments could be computed independently and intelligently merged with minor changes to the serial algorithm.
-
-Acknowledging that the algorithm for constructing a monotone quintic spline interpolant is slightly more expensive than the cubic case, gains in accuracy or decreased necessary number of points are often worth the computational effort.
-In practice the costs of spline interpolation are often dominated by evaluation rather than construction, and the increased accuracy is afforded for a negligible increase in computational cost.
-
-\heading{5. CONCLUSION AND FUTURE WORK}
-
-This paper proposes and tests an algorithm for constructing monotone quintic spline interpolants.
-Experiments demonstrate an improvement in approximation accuracy over monotone cubic spline interpolants, as expected based on theory.
-There are still open avenues of research going forward, such as an alternative sufficient condition for enforcing monotonicity or increased order monotone approximations.
-If the monotonicity conditions can be generalized to any order or made linear, the search for a monotone interpolating spline could potentially be formulated as a convex optimization problem.
-Finally, this work could be used to improve cumulative distribution function (CDF) estimates as well as predictive models that use CDFs.
-
-\heading{Acknowledgments}
-This work was supported by the National Science Foundation Grants CNS-1565314 and CNS-1838271.
-
-\heading{\hlvVIII BIBLIOGRAPHY} %% {\let\caps=\capsVIII \let\sl=\slVIII
-%% \def\ys{\hskip 1em minus .5em}\let\rm=\rmeight \rmVIII
-%% \newcount\refnum \refnum=0 \parskip=2pt plus 2pt minus 2pt
-%% \frenchspacing
-
-\bibliography{mqsi20}
-\bibliographystyle{plain}
+%% -------------------------------------------------------------------
 
 \vfill\eject\end
-
-%% %\refj is for regular journal articles
-%% \def\refj#1#2#3#4#5{\noindent\hangindent=10pt\hangafter=1
-%% %\global\advance\refnum by 1[\number\refnum]
-%% {\caps #1}\ys #2.\ys {\rm #3}. {\sl #4}, #5.\par}
-
-%% %\refb is for books
-%% \def\refb#1#2#3#4{\noindent\hangindent=10pt\hangafter=1
-%% %\global\advance\refnum by 1[\number\refnum]
-%% {\caps #1}\ys #2.\ys{\sl #3}. {\rm #4}.\par}
-
-%% %\refp is for private communication
-%% \def\refp#1#2#3{\noindent\hangindent=10pt\hangafter=1
-%% %\global\advance\refnum by 1[\number\refnum]
-%% {\caps #1}\ys #2.\ys{\rm #3}.\par}
-
-%% %\reft is for tech. reports, theses, and conferences
-%% \def\reft#1#2#3#4{\noindent\hangindent=10pt\hangafter=1
-%% %\global\advance\refnum by 1[\number\refnum]
-%% {\caps #1}\ys #2.\ys{\rm #3}. {\rm #4}.\par}
-
-%% %%%%% top of refs
-%% \reft{Amos, B. D., Easterling, D. R., Watson, L. T., Thacker, W. I.,
-%% Castle, B. S., Trosset, M. W.}{2014}{Algorithm XXX: QNSTOP: Quasi-Newton
-%% algorithm for stochastic optimization}{Technical Report 2014-07,
-%% Virginia Polytechnic Institute and State University, Blacksburg, VA}
-
-%% \refb{Anderson, E., Bai, Z., Bischof, C., Blackford, S., Demmel, J.,
-%% Dongarra, J., Du Croz, J., Greenbaum, A., Hammarling, S., McKenney, A.,
-%% and Sorensen, D.}{1999}{LAPACK Users' Guide Third Edition}
-%% {SIAM, Philidelphia, PA}
-
-%% \refb{Aurenhammer, F., Klein, R., and Lee, D. T.}{2013}
-%% {Voronoi diagrams and Delaunay triangulations}
-%% {World Scientific Publishing Co., Hackensack, NJ}
-
-%% \refj{Barber, C. B., Dobkin, D. P., and Huhdanpaa, H.}{1996}
-%% {The Quickhull algorithm for convex hulls}{ACM Trans. Math.
-%% Softw. 22}{469--483}
-
-%% \reft{Boissonnat, J.-D., Devillers, O., and Hornus, S.}{2009}
-%% {Incremental construction of the Delaunay triangulation and the
-%% Delaunay graph in medium dimension}{In {\sl Proc. Twenty-fifth Annual Symp.
-%% on Computational Geometry}, ACM, New York, NY, 208--216}
-
-%% \refj{Bowyer, A.}{1981}{Computing Dirichlet tessellations}
-%% {The Computer Journal 24}{162--166}
-
-%% \refj{Cameron, K. W., Anwar, A., Cheng, Y., Xu, L., Li, B., Ananth, U.,
-%% Bernard, J., Jearls, C., Lux, T., Hong, Y., Watson, L. T.,
-%% and Butt, A. R.}{2019}{MOANA: modeling and analyzing I/O variability
-%% in parallel system experimental design}
-%% {IEEE Trans. Parallel Distrib. Systems}{to appear}
-
-%% \reft{Chang, T. H., Watson, L. T., Lux, T. C. H., Bernard, J., Li, B.,
-%% Xu, L., Back, G., Butt, A. R., Cameron, K. W., and Hong, Y.}{2018a}
-%% {Predicting system performance by interpolation using a high-dimensional
-%% Delaunay triangulation} {In {\sl Proc. 2018 Spring Simulation Multiconference,
-%% 26th High Performance Computing Symp.}, Soc. for Modelling and Simulation
-%% Internat., Vista, CA, Article No. 2}
-
-%% \reft{Chang, T. H., Watson, L. T., Lux, T. C. H., Li, B., Xu, L., 
-%% Butt, A. R., Cameron, K. W., and Hong, Y.}{2018b}{A polynomial time
-%% algorithm for multivariate interpolation in arbitrary dimension via
-%% the Delaunay triangulation} {In {\sl Proc. ACM 2018 Southeast Conference
-%% (ACMSE '18)}, ACM, New York, NY, Article No. 12}
-
-%% \refb{Cheney, E. W. and Light, W. A.}{2009}
-%% {A Course in Approximation Theory}{American Mathematical Soc., 
-%% Providence, RI}
-
-%% \refb{Cheng, S. W., Dey, T. K., and Shewchuk, J.}{2012}
-%% {Delaunay Mesh Generation}{Computer and Information Science
-%% Series, CRC Press, New York, NY}
-
-%% \refj{Cignoni, P., Montani, C., and Scopigno, R.}{1998}
-%% {DeWall: A fast divide \& conquer Delaunay triangulation algorithm in $E^d$}
-%% {Computer-Aided Design 30}{333--341}
-
-%% \refb{de Berg, M., Cheong, O., van Kreveld, M., and Overmars, M.}{2008}
-%% {Computational Geometry: Algorithms and Applications}{Springer-Verlag,
-%% Berlin, Germany}
-
-%% \refj{Devillers, O., Pion, S., and Teillaud, M.}{2002}
-%% {Walking in a triangulation}{Internat. Journal of Foundations of
-%% Computer Science 13}{181--199}
-
-%% \reft{Edelsbrunner, H.}{1989}{An acyclicity theorem for cell complexes in
-%% d dimensions} {In {\sl Proc. Fifth Annual Symp. on Computational Geometry}, 
-%% ACM, New York, 145--151}
-
-%% \refj{Hanson, R. J. and Haskell, K. H.}{1982}
-%% {Algorithm 587: Two algorithms for the linearly constrained least
-%% squares problem}{ACM Trans. Math. Softw. 8}{323--333}
-
-%% \refj{Klee, V.}{1980}{On the complexity of d-dimensional Voronoi diagrams}
-%% {Archiv der Mathematik 34}{75--80}
-
-%% \reft{Lux, T. C. H., Watson, L. T., Chang, T. H., Bernard, J., Li, B.,
-%% Xu, L., Back, G., Butt, A. R., Cameron, K. W., and Hong, Y.}{2018}
-%% {Predictive modeling of I/O characteristics in high performance computing
-%% systems}{In {\sl Proc. 2018 Spring Simulation Multiconference, 26th High 
-%% Performance Computing Symp.}, Soc. for Modelling and Simulation Internat.,
-%% Vista, CA, Article No. 8}
-
-%% \refj{M{\"u}cke, E. P., Saias, I., and Zhu, B.}{1999}
-%% {Fast randomized point location without preprocessing in two- and 
-%% three-dimensional Delaunay triangulations}
-%% {Computational Geometry 12}{63--83}
-
-%% \refj{Omohundro, S. M.}{1990}{Geometric learning algorithms}
-%% {Physica D: Nonlinear Phenomena 42}{307--321}
-
-%% \reft{OpenMP Architecture Review Board (ARB)}{2015}
-%% {OpenMP Application Programming Interface (version 4.5)}
-%% {OpenMP}
-
-%% \refj{Rajan, V. T.}{1994}
-%% {Optimality of the Delaunay triangulation in $R^d$}
-%% {Discrete \& Computational Geometry 12}{189--202}
-
-%% \refj{Schaap, W. E. and van de Weygaert, R.}{2000}
-%% {Continuous fields and discrete samples: reconstruction through
-%% Delaunay tessellations}{Astronomy and Astrophysics 363}{L29--L32}
-
-%% \refj{Su, P. and Drysdale, R. L. S.}{1997}
-%% {A comparison of sequential Delaunay triangulation algorithms}
-%% {Computational Geometry 7}{361--385}
-
-%% \refj{Watson, D. F.}{1981}{Computing the n-dimensional Delaunay tessellation
-%% with application to Voronoi polytopes}
-%% {The Computer Journal 24}{167--172}
-
-%% }\vfill\eject\end
-%% %%%%% bottom

### 2020-06-23 Research_[2020-05-18]_Piecewise_Monotone_Quintic_Fortran/mqsi/SPLINE.f90

@@ -1,3 +1,50 @@
+! ^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^
+!                           SPLINE.f90
+! 
+! DESCRIPTION:
+!   This file defines the subroutines FIT_SPLINE for computing the
+!   coefficients of a B-spline basis necessary to reproduce given
+!   function and derivative values, and EVAL_SPLINE for evaluating the
+!   value, integral, and derivatives of a spline defined in terms of
+!   its B-spline basis.
+! 
+! CONTAINS:
+!   SUBROUTINE FIT_SPLINE(XI, FX, T, BCOEF, INFO)
+!     USE REAL_PRECISION, ONLY: R8
+!     REAL(KIND=R8), INTENT(IN),  DIMENSION(:)   :: XI
+!     REAL(KIND=R8), INTENT(IN),  DIMENSION(:,:) :: FX
+!     REAL(KIND=R8), INTENT(OUT), DIMENSION(:)   :: T, BCOEF
+!     INTEGER, INTENT(OUT) :: INFO
+!   END SUBROUTINE FIT_SPLINE
+! 
+!   SUBROUTINE EVAL_SPLINE(T, BCOEF, XY, INFO, D)
+!     USE REAL_PRECISION, ONLY: R8
+!     REAL(KIND=R8), INTENT(IN),    DIMENSION(:) :: T, BCOEF
+!     REAL(KIND=R8), INTENT(INOUT), DIMENSION(:) :: XY
+!     INTEGER, INTENT(OUT) :: INFO
+!     INTEGER, INTENT(IN), OPTIONAL :: D
+!   END SUBROUTINE EVAL_SPLINE
+! 
+! EXTERNAL DEPENDENCIES:
+!   MODULE REAL_PRECISION
+!     INTEGER, PARAMETER :: R8
+!   END MODULE REAL_PRECISION
+! 
+!   SUBROUTINE EVAL_BSPLINE(T, XY, D)
+!     USE REAL_PRECISION, ONLY: R8
+!     REAL(KIND=R8), INTENT(IN),    DIMENSION(:) :: T
+!     REAL(KIND=R8), INTENT(INOUT), DIMENSION(:) :: XY
+!     INTEGER, INTENT(IN), OPTIONAL :: D
+!   END SUBROUTINE EVAL_BSPLINE
+! 
[…]

### 2020-06-23 Research_[2020-05-18]_Piecewise_Monotone_Quintic_Fortran/mqsi/sample_main.f90

@@ -1,6 +1,42 @@
-! This file (sample_main.f90) contains a sample main program that
-! calls MQSI to interpolate handcrafted nonmonotone data.
-
+! ^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^
+!                          sample_main.f90
+! 
+! DESCRIPTION:
+!   This file (sample_main.f90) contains a sample main program that
+!   calls MQSI to interpolate handcrafted nonmonotone data.
+!
+! CONTAINS:
+!   PROGRAM SAMPLE_MAIN
+! 
+! DEPENDENCIES:
+!   MODULE REAL_PRECISION
+!     INTEGER, PARAMETER :: R8
+!   END MODULE REAL_PRECISION
+! 
+!   SUBROUTINE MQSI(X, Y, T, BCOEF, INFO)
+!     USE REAL_PRECISION, ONLY: R8
+!     REAL(KIND=R8), INTENT(IN),    DIMENSION(:) :: X
+!     REAL(KIND=R8), INTENT(INOUT), DIMENSION(:) :: Y
+!     REAL(KIND=R8), INTENT(OUT),   DIMENSION(:) :: T, BCOEF
+!     INTEGER, INTENT(OUT) :: INFO
+!   END SUBROUTINE MQSI
+! 
+!   SUBROUTINE EVAL_SPLINE(T, BCOEF, XY, INFO, D)
+!     USE REAL_PRECISION, ONLY: R8
+!     REAL(KIND=R8), INTENT(IN),    DIMENSION(:) :: T, BCOEF
+!     REAL(KIND=R8), INTENT(INOUT), DIMENSION(:) :: XY
+!     INTEGER, INTENT(OUT) :: INFO
+!     INTEGER, INTENT(IN), OPTIONAL :: D
+!   END SUBROUTINE EVAL_SPLINE
+! 
+! CONTRIBUTORS:
+!   Thomas C.H. Lux (tchlux@vt.edu)
+!   Layne T. Watson (ltwatson@computer.org)
+!   William I. Thacker (thacker@winthrop.edu)
+! 
+! VERSION HISTORY:
+!   June 2020 -- (tchl) Created file, (ltw / wit) reviewed and revised.
+! 
 PROGRAM SAMPLE_MAIN
 USE REAL_PRECISION, ONLY: R8  
 IMPLICIT NONE
[…]

### 2020-06-23 Research_[2020-05-18]_Piecewise_Monotone_Quintic_Fortran/paper/1-introduction.tex
Imported newest Dr. Watson revisions.

@@
 (splines) provide the smooth approximations for animation in graphics
-[Herman et al. 2015; Quint 2003], aesthetic structural support in
+[Herman et al.\ 2015; Quint 2003], aesthetic structural support in
 architecture [Brennan 2020], efficient aerodynamic surfaces in
 automotive and aerospace engineering [Brennan 2020], prolonged
-effective operation of electric motors [Berglund et al 2009], and
+effective operation of electric motors [Berglund et al.\ 2009], and
 accurate nonparametric approximations in statistics [Knott
@@
 It is often the case that the true underlying function or phenomenon
-being modeled has known properties e.g., convexity, positivity,
+being modeled has known properties like convexity, positivity,
 various levels of continuity, or monotonicity. Given a reasonable
@@
 it can interpolate, which in turn determines the growth rate of error
-away from interpolated values. This work constructs quintic splines in
-order to match the function value and two derivatives at either end of
-an interval, which requires six degrees of freedom in an interpolating
-polynomial. This work provides a Fortran 2003 subroutine {\tt MQSI}
-based on the necessary and sufficient conditions in Ulrich and Watson
-[1994] for the construction of monotone quintic spline interpolants of
-monotone data.
+away from interpolated values. $C^2$ quintic (order six) splines
+match the function value and two given derivatives at each breakpoint.
+This work provides a Fortran 2003 subroutine {\tt MQSI} based on the
+necessary and sufficient conditions in Ulrich and Watson [1994] for the
+construction of monotone quintic spline interpolants of monotone data.
+Precisely, the problem is, given a strictly increasing sequence $X_1<X_2<
+\cdots <X_n$ of breakpoints with corresponding monotone increasing
+function values $Y_1\le Y_2\le \cdots \le Y_n$, find a $C^2$ monotone
+increasing quintic spline $Q(x)$ with the same breakpoints satisfying
+$Q(X_i)=Y_i$ for $1\le i\le n$. ({\tt MQSI} actually does something slightly
+more general, producing $Q(x)$ that is monotone increasing (decreasing)
+wherever the data is monotone increasing (decreasing).)
 

### 2020-06-23 Research_[2020-05-18]_Piecewise_Monotone_Quintic_Fortran/paper/2-monotone-quintic.tex
Imported newest Dr. Watson revisions.

@@
 
-Fritsch and Carlson [1980] originally proposed the use of central
-differences to estimate derivatives, however this often leads to extra
-and unnecessary {\it wiggles} in the spline when used to approximate
-second derivatives. In an attempt to capture the local shape of the
-data, this package uses a facet model [Haralick and Watson 1981] to
-estimate first and second derivatives at data. Rather than picking a
-local linear regressor with minimal residual, this work uses a
-quadratic facet model that selects the local quadratic interpolant
-with minimum magnitude curvature.
+Fritsch and Carlson [1980] originally proposed the use of central differences
+to estimate derivatives, however this often leads to extra and unnecessary
+{\it wiggles} in the spline when used to approximate second derivatives.
+In an attempt to capture the local shape of the data, this package uses a
+facet model from image processing [Haralick and Watson 1981] to estimate
+first and second derivatives at breakpoints. Rather than picking a local
+linear or quadratic fit with minimal residual, this work uses a quadratic
+facet model that selects the local quadratic interpolant with minimum
+magnitude curvature.
 
@@
 {\bf Algorithm 1:} {
-\tt QUADRATIC\_FACET}$\big(X(1{:}n)$, $Y(1{:}n)$, $i \big)$ \vskip 1mm
-where $X_j$,$Y_j$ $\in \bbb{R}$ for $j = 1$, $\ldots$, $n$, and $1 < i
+\tt QUADRATIC\_FACET}$\bigl(X(1{:}n)$, $Y(1{:}n)$, $i \bigr)$
+
+where $X_j$, $Y_j$ $\in \bbb{R}$ for $j = 1$, $\ldots$, $n$, and $1 < i
 < n$. Returns the slope and curvature at $X_i$ of the local quadratic
 interpolant with minimum magnitude curvature.
-\vskip 2mm}
-{\parindent=6mm
+
+}
+{\parindent=3mm
 %% -------------------------------------------------------------------
-\item{} {\tt if} $\big((Y_i \approx Y_{i-1})$ {\tt or} $(Y_i \approx
-  Y_{i+1})\big)$ {\tt then; return $(0,0)$}
-\item{} {\tt else if} $\big((Y_{i+1} - Y_i)(Y_i - Y_{i-1}) < 0\big)$ {\tt then}
+\item{} {\tt if} $\bigl((Y_i \approx Y_{i-1})$ {\tt or} $(Y_i \approx
+  Y_{i+1})\bigr)$ {\tt then; return $(0,0)$}
+\item{} {\tt else if} $\bigl((Y_{i+1} - Y_i)(Y_i - Y_{i-1}) < 0\bigr)$
+{\tt then}
 \itemitem{} The point $(X_i$, $Y_i)$ is an extreme point. The
-quadratic with minimum curvature that has slope $0$ at $X_i$ will be returned.
-\item{} \codent $f_1 := $ interpolate $(X_{i-1}$,$Y_{i-1})$, $(X_i$,$Y_i)$,
+quadratic with minimum magnitude
+curvature that has slope zero at $X_i$ will be the facet chosen.
+\itemitem{} $f_1\:\hbox{interpolant to }(X_{i-1},Y_{i-1})$, $(X_i,Y_i)$,
   and $Df_1(X_i) = 0$.
-\item{} \codent $f_2 := $ interpolate $(X_i$,$Y_i)$, $(X_{i+1}$,$Y_{i+1})$,
+\itemitem{} $f_2\:\hbox{interpolant to }(X_i,Y_i)$, $(X_{i+1},Y_{i+1})$,
   and $Df_2(X_i) = 0$.
-\item{} \codent {\tt if} $\big(|D^2f_1| \leq |D^2f_2|\big)$ {\tt then; return} $\big(Df_1$, $D^2f_1\big)$
-\item{} \codent {\tt else;} \hskip 34.25mm {\tt return} $\big(Df_2$, $D^2f_2\big)$
-\item{} \codent {\tt endif}
+\itemitem{} {\tt if} $\bigl(|D^2f_1| \leq |D^2f_2|\bigr)$ {\tt then;
+  return} $\bigl(Df_1$, $D^2f_1\bigr)$
+\itemitem{} {\tt else;} \hskip 34.25mm {\tt return} $\bigl(Df_2$,
+$D^2f_2\bigr)$
+\itemitem{} {\tt endif}
 \item{} {\tt else}
@@
 not exist because $i = 2$ or $i = n-1$. In those cases, the minimum
-curvature among existing quadratics is chosen.}
-\item{} \codent $f_1 := $ interpolate $(X_{i-2}$,$Y_{i-2})$,
-  $(X_{i-1}$,$Y_{i-1})$, and $(X_i$,$Y_i)$.
-\item{} \codent $f_2 := $ interpolate $(X_{i-1}$,$Y_{i-1})$,
-  $(X_i$,$Y_i)$, and $(X_{i+1}$,$Y_{i+1})$.
-\item{} \codent $f_3 := $ interpolate $(X_i$,$Y_i)$,
-  $(X_{i+1}$,$Y_{i+1})$, and $(X_{i+2}$,$Y_{i+2})$.
-\item{} \codent {\tt if} \hskip 9mm $\big(|D^2f_1| \leq |D^2f_2|, |D^2f_3|\big)$
-  {\tt then; return} $\big(Df_1$, $D^2f_1\big)$
-\item{} \codent {\tt else if} $\big(|D^2f_2| \leq |D^2f_1|, |D^2f_3|\big)$
-  {\tt then; return} $\big(Df_2$, $D^2f_2\big)$
-\item{} \codent {\tt else;} \hskip 55mm {\tt return} $\big(Df_3$, $D^2f_3\big)$
+magnitude curvature among existing quadratics is chosen.}
+\itemitem{} $f_1\:\hbox{interpolant to }(X_{i-2},Y_{i-2})$,
+  $(X_{i-1},Y_{i-1})$, and $(X_i,Y_i)$.
+\itemitem{} $f_2\:\hbox{interpolant to }(X_{i-1},Y_{i-1})$,
+  $(X_i,Y_i)$, and $(X_{i+1},Y_{i+1})$.
+\itemitem{} $f_3\:\hbox{interpolant to }(X_i,Y_i)$,
+  $(X_{i+1},Y_{i+1})$, and $(X_{i+2},Y_{i+2})$.
+\itemitem{} {\tt if} \hskip 9mm $\bigl(|D^2f_1| = \min\bigl\{ |D^2f_1|,
+ |D^2f_2|, |D^2f_3|\bigr\} \bigr)$ {\tt then; return}
+ $\bigl(Df_1$, $D^2f_1\bigr)$
+\itemitem{} {\tt else if} $\bigl(|D^2f_2| = \min\bigl\{ |D^2f_1|,
+ |D^2f_2|, |D^2f_3|\bigr\} \bigr)$ {\tt then; return}
+ $\bigl(Df_2$, $D^2f_2\bigr)$
+\itemitem{} {\tt else;} \codent {\tt return} $\bigl(Df_3$, $D^2f_3\bigr)$
 \item{} \codent {\tt endif}
@@
 {\parindent 0mm
-{\bf Algorithm 2:} {\tt IS\_MONOTONE}$\big(x_0, x_1, f \big)$ \vskip 1mm
+{\bf Algorithm 2:} {\tt IS\_MONOTONE}$\bigl(x_0, x_1, f \bigr)$
+
 where $x_0$, $x_1 \in {\bbb R}$, $x_0 < x_1$, and $f$ is an order six
@@
 increasing on $[x_0,x_1]$.
-\vskip 2mm}
+
+}
 {\parindent=6mm
 %% -------------------------------------------------------------------
-\item{1.} {\tt if} $\big(f(x_0) \approx f(x_1)\big)$ {\tt then}
-\item{2.} \codent {\tt return} $\big( 0 = Df(x_0) = Df(x_1)
-  = D^2f(x_0) = D^2f(x_1) \big)$
+\item{1.} {\tt if} $\bigl(f(x_0) \approx f(x_1)\bigr)$ {\tt then}
+\item{2.} \codent {\tt return} $\bigl( 0 = Df(x_0) = Df(x_1)
+  = D^2f(x_0) = D^2f(x_1) \bigr)$
 \item{3.} {\tt endif}
-\item{4.} {\tt if} $\big(Df(x_0) < 0$ {\tt or} $Df(x_1) < 0\big)$ {\tt
+\item{4.} {\tt if} $\bigl(Df(x_0) < 0$ {\tt or} $Df(x_1) < 0\bigr)$ {\tt
   then; return FALSE; endif}
 \item{5.} $w := x_1 - x_0$
-\item{6.} $v := f(x_1) - f(x_0)$ \vskip 1mm
-\item{} {The necessity of steps 2 and 4 follows directly from
-  the fact that $f$ is $C^2$. The following steps 7--13 coincide with
+\item{6.} $v := f(x_1) - f(x_0)$
+
+\item{} {The necessity of Steps 2 and 4 follows directly from
+  the fact that $f$ is $C^2$. The following Steps 7--13 coincide with
   a simplified condition for quintic monotonicity that reduces to one
@@
   \delta - 2 \sqrt{\alpha \delta}$.  Step 4 checked for $\delta < 0$,
-  step 8 checks $\alpha < 0$, step 10 checks $\beta < \alpha - 2
-  \sqrt{\alpha \delta}$, and step 11 checks $\gamma < \delta - 2
+  Step 8 checks $\alpha < 0$, Step 10 checks $\beta < \alpha - 2
+  \sqrt{\alpha \delta}$, and Step 11 checks $\gamma < \delta - 2
   \sqrt{\alpha \delta}$. If none of the monotonicity conditions were
-  violated, then the order six piece is monotone and step 12
-  concludes.} \vskip 1mm
-\item{7.} {\tt if} $\big(Df(x_0) \approx 0$ {\tt or} $Df(x_1) \approx 0\big)$ {\tt then}
-\item{8.} \codent {\tt if} $\big(D^2f(x_1)w > 4Df(x_1)$ {\tt then; return FALSE; endif}
-\item{9.} \codent $t := 2 \sqrt{Df(x_0) \left(4Df(x_1) - D^2f(x_1) w \right)}$
-\item{10.} \codent {\tt if} $\big(t + 3Df(x_0) + D^2f(x_0)w < 0 \big)$
+  violated, then the order six piece is monotone and Step 12
+  concludes.}
+
+\item{7.} {\tt if} $\bigl(Df(x_0) \approx 0$ {\tt or} $Df(x_1) \approx
+0\bigr)$ {\tt then}
+\item{8.} \codent {\tt if} $\bigl(D^2f(x_1)w > 4Df(x_1)$ {\tt then;
+return FALSE; endif}
+\item{9.} \codent $t := 2 \sqrt{Df(x_0) (4Df(x_1) - D^2f(x_1) w) }$
+\item{10.} \codent {\tt if} $\bigl(t + 3Df(x_0) + D^2f(x_0)w < 0 \bigr)$
   {\tt then; return FALSE; endif}
-\item{11.} \codent {\tt if} $\big(60v - w\big(24Df(x_0) + 32Df(x_1) - 2t
-  + w(3D^2f(x_0) - 5D^2f(x_1))\big) < 0\big)$
+\item{11.} \codent {\tt if} $\bigl(60v - w\bigl(24Df(x_0) + 32Df(x_1) - 2t
+  + w(3D^2f(x_0) - 5D^2f(x_1))\bigr) < 0\bigr)$
 \item{}   \codent \codent {\tt then; return FALSE; endif}
 \item{12.} \codent {\tt return TRUE}
-\item{13.} {\tt endif} \vskip 1mm
+\item{13.} {\tt endif}
+
 \item{} {The following code considers the full quintic monotonicity
@@
 $\beta$, and $\gamma$ as defined by Ulrich and Watson, a quintic
-  piece is proven to be monotone when $\tau_1 > 0$ and $\alpha, \gamma
-  \geq -(\beta+2)/2$ when $\beta \leq 6$, and $\alpha, \gamma \geq -2
-  \sqrt{\beta-2}$ when $\beta > 6$. Step 12 checks $\tau_1 <= 0$,
-  steps 18 and 19 determine monotonicity based on $\alpha$, $\beta$,
-  and $\gamma$.} \vskip 1mm
-\item{12.} {\tt if} $\big( w\big(2\sqrt{Df(x_0) Df(x_1)} - 3(Df(x_0) +
-  Df(x_1))\big) - 24v \leq 0 \big)$
+piece is proven to be monotone if and only if
+$\tau_1 > 0$, and $\alpha, \gamma > -(\beta+2)/2$ when $\beta \leq 6$,
+and $\alpha, \gamma > -2 \sqrt{\beta-2}$ when $\beta > 6$.
+Step 14 checks $\tau_1 \le 0$, Steps 19 and 20 determine monotonicity based
+on $\alpha$, $\beta$, and $\gamma$.}
+
+\item{14.} {\tt if} $\bigl( w\bigl(2\sqrt{Df(x_0)\,Df(x_1)} - 3(Df(x_0) +
+  Df(x_1))\bigr) - 24v \leq 0 \bigr)$
 \item{} \codent {\tt then; return FALSE; endif}
-\item{14.} $t := \root{3/4}\of{Df(x_0) Df(x_1)}$
-\item{15.} $\alpha := (4 Df(x_1) - D^2f(x_1)w) \sqrt{Df(x_0)} / t$
-\item{16.} $\gamma := (4 Df(x_0) - D^2f(x_0)w) \sqrt{Df(x_1)} / t$
-\item{17.} $\displaystyle \beta := {60v/w + 3\big(w(D^2f(x_1) -
-  D^2f(x_0)) - 8(Df(x_0) + Df(x_1))\big) \over 2 \sqrt{Df(x_0) Df(x_1)}}$
-\item{18.} {\tt if} $(\beta \leq 6)$ {\tt then; return} $\min(\alpha,\gamma) > - (\beta + 2) / 2$
-\item{19.} {\tt else;} \hskip 18mm {\tt return} $\min(\alpha,\gamma) > -2 \sqrt{\beta - 2}$
-\item{20.} {\tt endif}
+\item{15.} $t := \bigl(Df(x_0)\, Df(x_1)\bigr)^{3/4}$
+\item{16.} $\alpha := (4 Df(x_1) - D^2f(x_1)w) \sqrt{Df(x_0)} / t$
+\item{17.} $\gamma := (4 Df(x_0) - D^2f(x_0)w) \sqrt{Df(x_1)} / t$
+\item{18.} $\displaystyle \beta := {60v/w + 3\bigl(w(D^2f(x_1) -
+  D^2f(x_0)) - 8(Df(x_0) + Df(x_1))\bigr) \over 2 \sqrt{Df(x_0)\,Df(x_1)}}$
+\item{19.} {\tt if} $(\beta \leq 6)$ {\tt then; return}
+$\bigl( \min\{\alpha,\gamma\} > - (\beta + 2) / 2 \bigr)$
+\item{20.} {\tt else;} \hskip 18mm {\tt return}
+$\bigl( \min\{\alpha,\gamma\} > -2 \sqrt{\beta - 2}\,\bigr)$
+\item{21.} {\tt endif}
 }
@@
 
-It is shown by Ulrich and Watson [1994] that when $0 = DQ(X_i)$ $=
-DQ(X_{i+1})$ $= D^2Q(X_i)$ $= D^2Q(X_{i+1})$, the quintic polynomial
-over $[X_i$, $X_{i+1}]$ is guaranteed to be monotone. Using this fact,
-the following algorithm shrinks initial derivative estimates until a
-monotone spline is achieved and outlines the core routine in the
+It is shown by Ulrich and Watson [1994] that when $0 = DQ(X_i) =
+DQ(X_{i+1}) = D^2Q(X_i) = D^2Q(X_{i+1})$, the quintic polynomial
+over $[X_i$, $X_{i+1}]$ is guaranteed to be monotone. Using this fact, the
+following algorithm shrinks (in magnitude) initial derivative estimates
+until a monotone spline is achieved and outlines the core routine in the
 accompanying package.
@@
 {\parindent 0mm
-{\bf Algorithm 3:} {\tt MQSI}$\big(X(1{:}n), Y(1{:}n) \big)$ \vskip 1mm
-where $(X_i,Y_i) \in (\bbb{R},\bbb{R})$ $i = 1,\ldots,n$ are data
+{\bf Algorithm 3:} {\tt MQSI}$\bigl(X(1{:}n), Y(1{:}n) \bigr)$
+
+where $(X_i,Y_i) \in \bbb{R}\times\bbb{R}$, $i = 1$, $\ldots$, $n$ are data
 points. Returns monotone quintic spline interpolant $Q(x)$ such that
 $Q(X_i) = Y_i$ and is monotone increasing (decreasing) on all
-intervals that $Y_i$ is increasing (decreasing).
-\vskip 2mm}
+intervals that $Y_i$ is monotone increasing (decreasing).
+
+}
 {\parindent=6mm
-\item{} Approximate first and second derivatives at all $X_i$ with {\tt QUADRATIC\_FACET}.
+\item{} Approximate first and second derivatives at all $X_i$ with
+{\tt QUADRATIC\_FACET}.
 \item{} {\tt do} $i = 1, \ldots n-1$
-\item{} \codent $D Q(X_i)$, $D^2 Q(X_i) :=$ {\tt QUADRATIC\_FACET}$(X$, $Y$, $i)$
+\item{} \codent $\bigl(D Q(X_i)$, $D^2 Q(X_i)\bigr):=$
+{\tt QUADRATIC\_FACET}$(X$, $Y$, $i)$
 \item{} {\tt enddo}
@@
 \item{} \codent {\tt if not IS\_MONOTONE}$(X_i$, $X_{i+1}$, $Q)$ {\tt then}
-\item{} \codent \codent Add interval $\big(X_i$, $X_{i+1}\big)$ to {\tt queue}.
+\item{} \codent\codent Add interval $\bigl(X_i$, $X_{i+1}\bigr)$ to
+{\tt queue}.
 \item{} \codent {\tt endif}
 \item{} {\tt enddo}
-\item{} {\tt do while} $\big($ {\tt queue} of nonmonotone intervals is nonempty $\big)$
-\itemitem{} Shrink $DQ$ and $D^2Q$ that border nonmonotone intervals.
-\itemitem{} Identify and store all remaining nonmonotone intervals in {\tt queue}.
+\item{} {\tt do while} $\bigl($ {\tt queue} of nonmonotone intervals is
+nonempty $\bigr)$
+\itemitem{} Shrink (in magnitude) $DQ$ and $D^2Q$ that border nonmonotone
+intervals.
+\itemitem{} Identify and store all remaining nonmonotone intervals in
+{\tt queue}.
 \item{} {\tt enddo}
@@
 
-Given the minimum curvature nature of the initial derivative
+Given the minimum magnitude curvature nature of the initial derivative
 estimates, it is desirable to make the smallest necessary changes to
@@
 $DQ$ and $D^2Q$, adding an additional step that increases previously
-shrunken derivative values by a smaller step. As long as a step size
-schedule is set that allows the value $0$ to be obtained in a fixed
+shrunken derivative values by a smaller increment than the previous
+decrement. The goal is to converge on the boundary of the monotone region
+in the $(\tau_1$, $\alpha$, $\beta$, $\gamma)$ space. As long as a step size
+schedule is set that allows the value zero to be obtained in a fixed
 number of computations, this has no effect on computational
-complexity. Notably, Algorithm 3 is equally capable of ensuring a {\it
-  piecewise} monotone quintic spline interpolates piecewise monotone
-data.
+complexity. Notably, since {\tt IS\_MONOTONE} handles both nondecreasing and
+nonincreasing simultaneously by taking into account the sign of $v$,
+Algorithm 3 produces $Q(x)$ that is monotone increasing (decreasing) over
+exactly the same intervals that the data $(X_i$, $Y_i)$ is monotone
+increasing (decreasing).

### 2020-06-23 Research_[2020-05-18]_Piecewise_Monotone_Quintic_Fortran/paper/3-spline-rep.tex
Imported newest Dr. Watson revisions.

@@
 \heading{3. SPLINE REPRESENTATION}
-
-The monotone quintic spline interpolant is represented in terms of a
-B-spline basis. Two routines in this package are provided, {\tt
-  FIT\_SPLINE} and {\tt EVAL\_SPLINE} that compute the B-spline
-coefficients to match the piecewise quintic polynomial values and
-derivatives, and evaluate a spline represented in terms of a B-spline
-basis. A Fortran 95 implementation of the B-spline recurrence relation
-evaluation code by C. deBoor for the value, derivatives, and integral
-of a B-spline is also provided.
+The monotone quintic spline interpolant $Q(x)$ is represented in terms of
+a B-spline basis. The routine {\tt FIT\_SPLINE} in this package computes
+the B-spline coefficients $\alpha_i$ of $Q(x)=\sum_{i=1}^{3n} \alpha_i
+B_{i,6,t}(x)$ to match the piecewise quintic polynomial values and (first
+two) derivatives at the breakpoints $X_i$, where the spline order is six
+and the knot sequence $t$ has the breakpoint multiplicities (6, 3, $\ldots$,
+3, 6).  The routine {\tt EVAL\_SPLINE} evaluates a spline represented in
+terms of a B-spline basis. A Fortran 2003 implementation {\tt EVAL\_BSPLINE}
+of the B-spline recurrence relation evaluation code by C. deBoor [1978]
+for the value, derivatives, and integral of a B-spline is also provided.
 

### 2020-06-23 Research_[2020-05-18]_Piecewise_Monotone_Quintic_Fortran/paper/4-analysis.tex
Imported newest Dr. Watson revisions.

@@
 
-Both Algorithms 1 and 2 have ${\cal O}(1)$ runtime. Given a fixed
-schedule for shrinking derivative values, Algorithm 3 has a ${\cal
-  O}(n)$ runtime for $n$ data points. In execution, the majority of
-time is spent solving the banded linear system of equations for
-B-spline coefficients. The quadratic facet model produces a unique
+Both Algorithms 1 and 2 have ${\cal O}(1)$ runtime. Given a fixed schedule
+for shrinking derivative values, Algorithm 3 has a ${\cal O}(n)$ runtime
+for $n$ data points. In execution, the majority of the time, still ${\cal
+O}(n)$, is spent solving the banded linear system of equations for the
+B-spline coefficients.  Thus for $n$ data points, the overall execution
+time is ${\cal O}(n)$.  The quadratic facet model produces a unique
 sensitivity to input perturbation, as small changes in input may cause
-data points to be associated with different local quadratic
-interpolants. While this makes the technique susceptible to random
-perturbations, it generally produces high quality monotone quintic
-splines and is preferred for that reason.
+different quadratic facets to be associated with a breakpoint, and thus
+different initial derivative estimates.  However, the quasi-bisection
+search for a point near the monotone boundary in $(\tau_1$, $\alpha$,
+$\beta$, $\gamma)$ space usually results in a high quality visually appealing
+(meaning less wiggle) monotone quintic spline.  Despite this potential
+sensitivity, the quadratic facet model is still preferred because it
+generally provides excellent initial estimates of the first and second
+derivatives at the breakpoints, with few iterations required to find a
+monotone $Q(x)$.

### 2020-06-23 Research_[2020-05-18]_Piecewise_Monotone_Quintic_Fortran/paper/mqsi20.tex
Imported newest Dr. Watson revisions.

@@
 \def\thickrule{\hrule\hrule\hrule}
-\def\codent{\hskip 1.8em} % <- 'tab' for indentation.
+\def\codent{\hskip 1.0em} % <- 'tab' for indentation.
 
@@
 theoretical monotonicity constraints, first and second derivative estimates
-at data provided by a quadratic facet model are refined to produce a $C^2$
-monotone interpolant.  Algorithm and implementation details, complexity
-and sensitivity analyses, usage information, and a brief performance study
-are included.
+at data provided by a quadratic facet model are refined to produce a
+$\scriptstyle C^2$ monotone interpolant.  Algorithm and implementation
+details, complexity and sensitivity analyses, usage information, and a
+brief performance study are included.
 
@@
 
+}
 %% -------------------------------------------------------------------

### 2020-06-24 Research_[2020-05-18]_Piecewise_Monotone_Quintic_Fortran/python/testing/SPLINE.f90
Reorganized folders, started working on visuals.

@@ -1,0 +1,1 @@
+../../code/SPLINE.f90

### 2020-06-24 Research_[2020-05-18]_Piecewise_Monotone_Quintic_Fortran/python/testing/f_test_splines.py

@@ -1,8 +1,8 @@
 
 import fmodpy
-bspline = fmodpy.fimport("mqsi/EVAL_BSPLINE.f90")
-spline = fmodpy.fimport("mqsi/SPLINE.f90")
-mqsi = fmodpy.fimport("mqsi/MQSI.f90")
+bspline = fmodpy.fimport("EVAL_BSPLINE.f90")
+spline = fmodpy.fimport("SPLINE.f90")
+mqsi = fmodpy.fimport("MQSI.f90")
 # Define an object that has all of the expected functions.
 class splines:
     mqsi = mqsi.mqsi

### 2020-06-24 Research_[2020-05-18]_Piecewise_Monotone_Quintic_Fortran/visuals/mqsi.py

@@ -1,0 +1,1221 @@
+# Fraction class that provides infinite-precision rational numbers
+# (making use of Python default arbitrarily large integers).
+
+from math import gcd
+class UsageError(Exception): pass
+
+# This class-operator wrapper automatially returns a `float` type if
+# the operand is neither the same type as the operator nor an `int`.
+def float_fallback(operator):
+    def wrapped_operator(primary, other):
+        # Call the standard operator if the types match.
+        if (type(other) == type(primary)):
+            return operator(primary, other)
+        # Handle integers in a special way (so that users can put ints
+        # in code without breaking fractions).
+        elif (type(other) == int):
+            return operator(primary, type(primary)(other))
+        # Otherwise, cast "other" to the type of "primary", call the
+        # operator, then cast the result back to a 'float' type.
+        else:
+            return float(operator(primary, type(primary)(other)))
+    return wrapped_operator
+
+# This class-operator wrapper automatially casts to and returns an
+# expected type (in common usage here, `primary` will be `Fraction`).
+def auto_same_type(operator):
+    def wrapped_operator(primary, other):
+        # Call the standard operator if the types match.
+        if (type(other) == type(primary)):
+            return operator(primary, other)
+        # Otherwise, cast "other" into the same type.
+        else: return operator(primary, type(primary)(other))
+    return wrapped_operator
+
+# This can be 'float_fallback' or 'auto_same_type'.
+TYPE_HANDLER = auto_same_type
+ACCURACY = float('inf')
+
+# A fraction instance. Initialized as:
+# 
+#     Fraction(numerator=0, denominator=1)
+# 
+# The arguments to the Fraction must either be Python integers or
+# convertible to Python float via the 'float' built-in function.
+class Fraction: 
+    def __init__(self, numerator=0, denominator=1, reduce=True):
+        # If "None" was provided, then set this fraction to have "None" values.
+        if ((numerator is None) or (denominator is None)):
+            self._numerator = None
+            self._denominator = None
+            return
+        # If a Fraction was provided, shortcut and exit.
+        elif (type(numerator) == Fraction):
+            self._numerator   = numerator.numerator
+            self._denominator = numerator.denominator
+            return
+        # Convert input arguments to the appropriate type.
+        if (type(numerator) != int):
[…]

### 2020-06-24 Research_[2020-05-18]_Piecewise_Monotone_Quintic_Fortran/visuals/test.py

@@ -1,0 +1,33 @@
+from util.plot import Plot
+from mqsi import Polynomial, monotone_quintic_spline
+
+p = Plot()
+
+f1 = lambda x: x**2
+f2 = lambda x, m=1: m*(x-2)**2 + 6
+
+def linspace(l,r,n): return [l+(i/(n-1))*(r-l) for i in range(n)]
+
+eps = 1 - 2**(-51)
+x = linspace(1, 4, 5)
+x[1] = 2*x[1]/3 + x[2]/3
+x[3] = 2*x[3]/3 + x[2]/3
+y1 = [f1(x[0]), f1(x[1]), f1(x[2]), f2(x[3]), f2(x[4])]
+gap = f2(x[2]) - f2(x[2],eps)
+y2 = [f1(x[0]), f1(x[1]), f1(x[2]), f2(x[3],eps)+gap, f2(x[4],eps)+gap]
+
+q1 = monotone_quintic_spline(x, y1, monotone=True)
+q2 = monotone_quintic_spline(x, y2, monotone=True)
+# q1o = monotone_quintic_spline(x, y1, monotone=False)
+# q2o = monotone_quintic_spline(x, y2, monotone=False)
+
+p.add("data 1", x, y1, color=0)
+p.add("data 2", x, y2, color=0)
+bounds = [min(x), max(x)]
+p.add_function("f1", f1, bounds, color=3, opacity=.2, dash="dash")
+p.add_function("f2", f2, bounds, color=3, opacity=.2, dash="dash")
+p.add_function("Q1", q1, bounds, color=1)
+p.add_function("Q2", q2, bounds, color=2)
+# p.add_function("Q1 qfm", q1o, bounds, color=1, opacity=.5)
+# p.add_function("Q2 qfm", q2o, bounds, color=2, opacity=.5)
+p.show()

### 2020-06-26 Research_[2020-05-18]_Piecewise_Monotone_Quintic_Fortran/misc/example_tex/old_files/0-conclusions.tex
Changed some directory names and fixed some typos in code files. Still working on the paper.

--- 2019-12-16 Paper_[2019-11]_HPC/paper/paper.tex
@@
-%**************************************************************************
-%* SpringSim 2020 Author Kit
-%*
-%* Word Processing System: TeXnicCenter and MiKTeX
-%*
-%**************************************************************************
+\heading{5. CONCLUSION AND FUTURE WORK}
 
-\documentclass{scspaperproc}
-
-\usepackage{latexsym}
-\usepackage{graphicx}
-\usepackage{mathptmx}
-
-%
-%****************************************************************************
-% AUTHOR: You may want to use some of these packages. (Optional)
-\usepackage{amsmath}
-\usepackage{amsfonts}
-\usepackage{amssymb}
-\usepackage{amsbsy}
-\usepackage{amsthm}
-%****************************************************************************
-
-%
-%****************************************************************************
-% AUTHOR: If you do not wish to use hyperlinks, then just comment
-% out the hyperref usepackage commands below.
-
-%% This version of the command is used if you use pdflatex. In this case you
-%% cannot use ps or eps files for graphics, but pdf, jpeg, png etc are fine.
-
-\usepackage[pdftex,colorlinks=true,urlcolor=blue,citecolor=black,anchorcolor=black,linkcolor=black,bookmarks=false]{hyperref}
-
-%% The next versions of the hyperref command are used if you adopt the
-%% outdated latex-dvips-ps2pdf route in generating your pdf file. In
-%% this case you can use ps or eps files for graphics, but not pdf, jpeg, png etc.
-%% However, the final pdf file should embed all fonts required which means that you have to use file
-%% formats which can embed fonts. Please note that the final PDF file will not be generated on your computer!
-%% If you are using WinEdt or PCTeX, then use the following. If you are using
-%% Y&Y TeX then replace "dvips" with "dvipsone"
-
-%% \usepackage[dvips,colorlinks=true,urlcolor=blue,citecolor=black,%
-%% anchorcolor=black,linkcolor=black]{hyperref}
-
-%% The use of the long citation format (e.g. "Brown and Edwards (1993)" rather than "[5]") and at the same
-%% time using the hyperref package can lead to hard to trace bugs in case the citation is broken accross the
-%% line (usually this will mark the entire paragraph as a hyperlink (clickable) which is easily noticeable and fixed
-%% if using colorlinks, but not if the color is black -- as it is now). Worse yet, if a citation spans page boundary,
-%% LaTeX compilation can fail, with an obscure error message. Since this depends a lot on the flow of the text
-%% and wording, these bugs come and go and can be extremely hard for a beginner to trace. The error
-%% message can look like this:
-%%
-%%    ! pdfTeX error (ext4): \pdfendlink ended up in different nesting level than \pdfstartlink.
-%%    \AtBegShi@Output ...ipout \box \AtBeginShipoutBox 
-%%    \fi \fi 
-%%    l.174 
-%%    ! ==> Fatal error occurred, no output PDF file produced!
-%%
-%% and can be universally fixed by putting an \mbox{} around the citation in question (in this case, at line 174)
-%% and maybe adapting the wording a little bit to improve the paragraph typesetting, which is perhaps not
-%% immediately obvious.
-%****************************************************************************
-
-%****************************************************************************
-%*
-%* AUTHOR: YOUR CALL!  Document-specific macros can come here.
-%*
-%****************************************************************************
-
-%% Use a package to allow for multi-row cells in a table.
-\usepackage{multirow}
-
-%% Use a custom `algorithm` command for this paper.
-\input{lux-algorithm}
-
-% add custom hyphenation rules here
-\usepackage{hyphenat}
-\hyphenation{op-tical net-works semi-conduc-tor}
-
-% If you use theorems
-\newtheoremstyle{scsthe}% hnamei
-{8pt}% hSpace abovei
-{8pt}% hSpace belowi
-{\it}% hBody fonti
-{}% hIndent amounti1
-{\bf}% hTheorem head fontbf
-{.}% hPunctuation after theorem headi
-{.5em}% hSpace after theorem headi2
-{}% hTheorem head spec (can be left empty, meaning `normal')i
-
-\theoremstyle{scsthe}
-\newtheorem{theorem}{Theorem}
-\renewcommand{\thetheorem}{\arabic{theorem}}
-\newtheorem{corollary}[theorem]{Corollary}
-\renewcommand{\thecorollary}{\arabic{corollary}}
-\newtheorem{definition}{Definition}
-\renewcommand{\thedefinition}{\arabic{definition}}
-
-% avoid overrunning the right margin; you are welcome to remove this, provided that you take care not to overrun the right margin anywhere in your paper
-\sloppy
-
-%#########################################################
-%*
-%*  The Document.
-%*
-\begin{document}
-
-%***************************************************************************
-% AUTHOR: Thomas C.H. Lux et al.
-% FORMAT AUTHORS NAMES Like: Author1, Author2 and Author3 (last names)
-%
-%		You need to change the author listing below!
-%               Please list ALL authors using last name only, separate by a comma except
-%               for the last author, separate with "and"
-
-\SCSpagesetup{Lux, Watson, Chang, Xu, Wang, and Hong}
-
-% AUTHOR: Uncomment ONE of these correct conference names.
-\def\SCSconferenceacro{SpringSim'20}
-%\def\SCSconferenceacro{SummerSim}
-%\def\SCSconferenceacro{AutumnSim}
-%\def\SCSconferenceacro{PowerPlantSim}
-
-% AUTHOR: Set the correct year of the conference.
-\def\SCSpublicationyear{2020}
-
-% AUTHOR: Set the correct month and dates; the dates are separated by a single minus sign
-% with no spaces and no leading zeros, the month is a full name (e.g. April) with the first letter
-% capitalized. For example, "April 8-13".
-\def\SCSconferencedates{May 19-May 21}
-
-% AUTHOR: Set the correct venue in the form "City, State, Country", for example "Los Angeles, CA, USA".
-\def\SCSconferencevenue{Fairfax, VA, USA}
-
-% AUTHOR: Enter the title, all letters in upper case
-\title{An Algorithm for Constructing Monotone \\ Quintic Interpolating Splines}
-
-% AUTHOR: Enter the authors of the article, see end of the example document for further examples
-\author{
-Thomas C. H. Lux\\
-Layne T. Watson\\
-Tyler H. Chang\\ [12pt]
-Departments of Computer Science, Mathematics,\\
-and Aerospace and Ocean Engineering\\
-Virginia Polytechnic Institute \& State University\\
-2000 Torgersen Hall, Blacksburg, VA, USA\\
-\{tchlux,ltw,thchang\}@vt.edu\\
-\and
-Li Xu\\%
-Yueyao Wang\\%
-Yili Hong\\ [12pt]
-Department of Statistics\\
-Virginia Polytechnic Institute \& State University\\
-213 Hutcheson Hall, Blacksburg, VA, USA\\
-\{lix1992,yueyao94,yilihong\}@vt.edu\\
-}
-
-\maketitle
-
-%% Examples and notes.
-%% 
-%% \begin{table}[htb]
-%%   \centering
-%%   \caption{Caption above the table!}\label{tab:example}
-%%   \begin{tabular}{r|l}
-%%     First & row \\ \hline
-%%     second & row \\
-%%     third & row \\
-%%   \end{tabular}
-%% \end{table}
-%% 
-%% \begin{figure}[htb]
-%%   \centering
-%%   \includegraphics[width=0.50\textwidth]{path_without_extension}
-%%   \caption{Caption below the figure!}\label{fig:example}
-%% \end{figure}
-%% 
-%% \cite{ref}  - Make a citation that is in parenthesis
-%% \citeN{ref} - Make a citation out of parenthesis (for noun usage)
-%% \shortcite{ref}  - Make a citation for references with >3 authors.
-%% \shortciteN{ref} - Make a noun-form citation for references with >3 authors.
-%% 
-%% 
-%% 5 - 12 page paper mandated
-%% 150 word abstract
-%% 5 keywords maximum
-
-\section*{Abstract}
-
-An algorithm for computing monotone order six piecewise polynomial
-interpolants is proposed. Algebraic constraints for enforcing
-monotonicity are provided that align with quintic monotonicity
-theory. The algorithm is implemented, tested, and applied to several
-sample problems to demonstrate the improved accuracy of monotone
-quintic spline interpolants compared to that for monotone cubic spline
-interpolants.
-
-\textbf{Keywords:} monotone, quintic spline, Hermite interpolation, sixth order polynomial.
-
-\section{Introduction and Motivation}
-\label{sec:introduction}
-
-%% \begin{figure}[htb]
-%%   \centering
-%%   \includegraphics[width=0.50\textwidth]{demo_projection}
-%%   \caption{These are the projection mechanisms that make a cubic polynomial piece monotone.}\label{fig:projection}
-%% \end{figure}
-
-%% The ability to interpolate or regress a function with continuous value and derivatives has proven fundamentally useful.
-In many applications, there are known shapes or characteristics that are expected of approximations.
-For example, a cumulative distribution function in statistics cannot decrease, the entropy of a closed thermodynamic system can never increase, an electric motor should not suddenly change direction when slowing down or speeding up.
-In each of these cases it is important that any constructed spline approximation is monotone.
-
-Many domains of science rely on smooth approximations to real-valued functions over a closed interval.
-Piecewise polynomial functions (splines) provide the smooth approximations for animation in graphics \cite{herman2015techniques,quint2003scalable}, aesthetic structural support in architecture \cite{brennan2019measure}, efficient aerodynamic surfaces in automotive and aerospace engineering \cite{brennan2019measure}, prolonged effective operation of electric motors \shortcite{berglund2009planning}, and accurate nonparametric approximations in statistics \cite{knott2012interpolating}.
-While polynomial interpolants or regressors apply broadly, splines are often a good choice because they can approximate globally complex functions while minimizing the local complexity of an approximation.
-
-It is often the case that the true underlying function or phenomenon being modeled has known properties e.g., convexity, positivity, various levels of continuity, or monotonicity.
-Given a reasonable amount of data, it quickly becomes difficult to achieve desirable properties in a single polynomial function.
-In general, the maintenance of function properties through interpolation/regression is referred to as \textit{shape preserving} \cite{fritsch1980monotone,gregory1985shape}.
-The specific shapes this work will achieve in approximations are monotonicity and $C^2$ continuity.
-These properties are chiefly important to the approximation of cumulative distribution functions and subsequently the effective generation of random numbers from a specified distribution.
-
-In statistics especially, the construction of a monotone interpolating spline that is $C^2$ continuous is meaningfully useful \cite{ramsay1988monotone}.
-A function with these properties could approximate a cumulative distribution function to a high level of accuracy with relatively few intervals.
-A twice continuously differentiable approximation to a cumulative distribution function (CDF) would also produce a corresponding probability density function (PDF) that is continuously differentiable, which is a property many standard parametric distributions maintain.
-
-The currently available software for monotone piecewise polynomial interpolation includes quadratic \cite{he1998monotone}, cubic \cite{fritsch1980monotone}, and (with limited application) quartic \cite{wang2004rational,piah2011improved,yao2018unconditionally} cases.
-In addition, a statistical method for bootstrapping the construction of an arbitrarily smooth monotone fit exists \cite{leitenstorfer2006generalized}, but the method does not take advantage of known analytic properties related to quintic polynomials.
-Theory has been provided for the quintic case \cite{ulrich1994positivity,hess1994positive}, however this theory has not yet been used to construct a monotone quintic spline interpolation routine.
-Recent work suggests that the lack of quintic software may be due to a general unawareness of the theory \shortcite{xie2018semiparametric}.
-
-\begin{figure}
-  \centering
-  \includegraphics[width=.9\textwidth]{spline_demonstration}
-  \caption{Example polynomials that interpolate function values at the ends of the interval $[0,1]$.
-The first only interpolates the function values $g_2(0) = 1$ and $g_2(1) = 0$, making it the order two polynomial $g_2(x) = 1 - x$.
-For the second plot $g_4(x) = 8x^3 - 12x^2 + 4x + 1/2$, which is order four and interpolates the values $g_4(0) = 1/2$, $g_4'(0) = 4$, $g_4(1) = 1/2$, $g_4'(1) = 4$.
-Finally the third plot shows the order six polynomial $g_6(x) = - 64x^5 + 160x^4 - 140x^3 + 50x^2 - 7x + 1$ interpolating the function values $g_6(0) = 1$, $g_6'(0) = -7$, $g_6''(0) = 100$, $g_6(1) = 0$, $g_6'(1) = -7$, $g_6''(1) = -100$.
-Notice that interpolating the same fixed number of function values at each endpoint will always result in an even order interpolating polynomial.
-  }\label{fig:spline_demonstration}
-\end{figure}
-
-The importance of piecewise quintic interpolation over lower order approximations can be simply demonstrated.
-In general, the order of a polynomial determines the number of function values it can interpolate, and the growth rate of error away from the interpolated function values.
-As demonstrated in Figure \ref{fig:spline_demonstration}, it can be seen that matching a value at either end of the interval requires an order two (linear) approximation and each additional derivative at the ends of the interval raises the necessary polynomial order by two.
-The body of this work is composed of a novel algorithm for enforcing monotonicity on quintic polynomial pieces, then extending that solution to work on quintic splines.
-
-%% Software exists that does monotone piecewise quintic regression \cite{}, but the available routines do not computationally take advantage of the quintic monotonicity theory. ???
-
-The major contribution of this work is an algorithm for constructing monotone quintic interpolating splines that utilizes existing quintic monotonicity theory.
-The remainder of this paper is structured as follows: {\bf ?????}
-
-\subsection{Computing a Monotone Cubic Interpolant}
-
-The current state-of-the-art monotone interpolating spline with a mathematical software implementation is piecewise cubic, continuously differentiable, and was first proposed in \cite{fritsch1980monotone} then expanded upon in \cite{carlson1985monotone}.
-Let $\pi: x_0 = k_1 < k_2 < \cdots < k_n = x_1$ be a partition of the interval $[x_0,x_1]$.
-Let $f: \mathbb{R} \rightarrow \mathbb{R}$ be $C^2$, and $\bigl\{f(k_i)\bigr\}_{i=1}^n$ and $\bigl\{\Delta_i\bigr\}_{i=1}^n$ be given sets of function and derivative values at the partition points for a monotone function $f$.
-Either $f(k_i) \leq f(k_{i+1})$, $i=1$, $\ldots$, $n-1$, and $\Delta_i\ge0$, $i=1$, $\ldots$, $n$, or $f(k_i) \geq f(k_{i+1})$, $i=1$, $\ldots$, $n-1$, and $\Delta_i\le0$, $i=1$, $\ldots$, $n$.
-Let $\hat f$ be a piecewise cubic polynomial defined in each subinterval $I_i = [k_i, k_{i+1}]$ by
-
-\begin{align*}
-h_i &= k_{i+1} - k_{i}, \quad
-u(t) = 3t^2 - 2t^3, \quad
-p(t) = t^3 - t^2,\\
-\hat f(x) &= f(k_i) u\big((k_{i+1} - x) / h_i\big) +
-f(k_{i+1}) u\big((x - k_i) / h_i\big) \\
-&{} - h_i\Delta_i p\big((k_{i+1}-x)/h_i\big) + h_i\Delta_{i+1}
-p\big((x-k_i)/h_i\big).
-\end{align*}
-
-Notice that a trivially monotone spline results when $\Delta_i = 0$, for $i = 1$, $\ldots$, $n$.
-However, such a spline has too many \textit{wiggles} for most applications.
-\citeN{carlson1985monotone} show that simple conditions on the derivative values can guarantee monotonicity, and that these conditions can be enforced in a way that ensures modifications on one interval will not break the monotonicity of cubic polynomials over any neighboring intervals.
-Consider the terms $\alpha = \big(\Delta_i (k_{i+1}-k_i)\big) / \big(f(k_{i+1}) - f(k_i)\big)$ and $\beta = \big(\Delta_{i+1}(k_{i+1}-k_i)\big) / \big(f(k_{i+1}) - f(k_i)\big)$.
-If $f(k_{i+1}) - f(k_i)=0$, take $\Delta_i = \Delta_{i+1} =0$ and $\alpha =\beta =1$.
-In general, monotonicity of a cubic polynomial over a subinterval can be maintained by ensuring that $\alpha$ and $\beta$ reside in any of the regions depicted in Figure \ref{fig:projection}.
-
-\begin{figure}
-  \centering
-  \includegraphics[width=0.45\textwidth]{cubic_projection_demonstration}
-  \caption{These are the feasible regions of monotonicity for cubic splines and the projections that make a cubic polynomial piece monotone.
-The regions themselves are numbered 1--4 corresponding to their original description in Fritsch and Carlson (1980).
-One point is projected onto each region as a demonstration.}\label{fig:projection}
-\end{figure}
-
-The actual region of monotonicity for a cubic polynomial is larger, but projection of $(\alpha, \beta)$ into one of these regions ensures that monotonicity will be achieved and not violated for neighboring regions.
-The user must decide which region is most appropriate for the projections based on the application, Fritsch and Carlson recommend using Region 2.
-
-While the cubic monotonicity case affords such a concise solution, the region of monotonicity is not so simple in the quintic case.
-In the next section, an algorithm for performing a projection similar to those for cubic polynomials is proposed.
-
-\section{Monotone Quintic Interpolation}
-
-The following section is composed of three algorithms that together are used to construct a monotone quintic spline interpolant.
-Without loss of generality, the algorithms will only consider the monotone increasing (nondecreasing) case.
-The monotone decreasing (nonincreasing) case is handled similarly.
-Algorithm \ref{alg:check-monotone} checks monotonicity,
-Algorithm \ref{alg:make-monotone} enforces monotonicity on an order six
-polynomial piece of the quintic spline, and Algorithm
-\ref{alg:monotone-spline} uses the previous two algorithms to enforce
-monotonicity for the entire quintic spline.
-
-In pseudocode, a function \texttt{binary\_search($g,$ $a,$ $b$)} is used, where $a,b\in S\subset \mathbb{R}^p$ for convex $S$, $g: S \rightarrow \{$\texttt{FALSE}, \texttt{TRUE}$\}$ is a right continuous Boolean function, $g(b) = \hbox{\texttt{TRUE}}$, and for $\mu \in [0,1]$, $g\big((1-\mu)a + \mu b\big) =\hbox{\texttt{TRUE}} \Longrightarrow g\big((1-\nu)a + \nu b\big) =\hbox{\texttt{TRUE}}$ for $\mu \le \nu\le1$.
-The search returns $\big((1-c)a + c b\big)$ for the smallest $c \in [0,1]$ such that $g\big((1-c)a + c b\big) = \hbox{\texttt{TRUE}}$.
-
-These algorithms make use of first and second derivatives of the approximated function.
-In the case that the first and second derivative information is not provided along with function values, the derivatives are estimated with finite differences of the function values.
-The final quintic spline is represented as a piecewise polynomial using the Newton form for each polynomial piece, and ultimately converted to a $B$-spline representation for evaluation.
-
-\algorithm
-    {is\_monotone$\big(x_0, x_1, f \big)$}
-    {where $x_0$, $x_1 \in \mathbb{R}$, $x_0 < x_1$, and $f$ is an order six polynomial defined by $f(x_0)$, $Df(x_0)$, $D^2f(x_0)$, $f(x_1)$,
-  $Df(x_1)$, $D^2f(x_1)$. Returns \texttt{TRUE} if $f$ is monotone increasing on $[x_0,x_1]$.}
-    {\label{alg:check-monotone}
-  \item if $\big(f(x_0) = f(x_1)\big)$ and not $\big( 0 = Df(x_0) = Df(x_1) = D^2f(x_0) = D^2f(x_1) \big)$
-    \indented{return FALSE}
-    end if
-  \item if $\big(Df(x_0) < 0$ or $Df(x_1) < 0\big)$
-    \indented{return FALSE}
-    end if
-    \comment{The necessity of these first two steps follows directly from the fact that $f$ is $C^2$.
-The next case is in accordance with a simplified condition for quintic monotonicity that reduces to one of cubic positivity studied in \citeN{schmidt1988positivity}, where $\alpha$, $\beta$, $\gamma$, and $\delta$ are defined in terms of values and derivatives of $f$ at $x_0$ and $x_1$.
-Step \ref{alg1-alpha} checks for the necessary condition that $\alpha \ge 0$, Step \ref{alg1-beta} checks $\beta \geq \alpha$, and Step \ref{alg1-gamma} checks $\gamma \geq \delta$, all from \citeN{schmidt1988positivity}.
-If all necessary conditions are met, then the order six piece is monotone and Step \ref{alg1-simplified-true} concludes this check.}
-  \item if $\big(Df(x_0) = 0$ or $Df(x_1) = 0\big)$
-  \item \indented{$w := x_0 - x_1$\\$v := f(x_0) - f(x_1)$}
-  \item \indented{if $\big(D^2f(x_1) > {-4}Df(x_1) / w \big)$ return FALSE}
-    \label{alg1-alpha}
-  \item \indented{if $\big(D^2f(x_1) < (3w D^2f(x_0) - 24 Df(x_0) - 32
-    Df(x_1) + 60v/w) / (5w) \big)$ return FALSE} \label{alg1-beta}
-  \item \indented{if $\big(D^2f(x_0) < 3 Df(x_0) / w \big)$ return FALSE}
-    \label{alg1-gamma}
-  \item \indented{return TRUE} \label{alg1-simplified-true}
-    end if
-    \comment{The following code considers the remaining case where $Df(x_0) \neq 0$ and $Df(x_1) \neq 0$.}
-  \item $\displaystyle A := Df(x_0)\frac{x_1 - x_0}{f(x_1) - f(x_0)}$
-    \\ $\displaystyle B := Df(x_1) \frac{x_1 - x_0}{f(x_1) - f(x_0)}$
-    \comment{The variables $A$ and $B$ correspond directly to the theoretical foundation for positive quartic polynomials established in \citeN{ulrich1994positivity}, first defined after Equation (18).}
-  \item $\displaystyle \gamma_0 := 4 \frac{Df(x_0)}{Df(x_1)} (B/A)^{3/4}$
-    \\ $\displaystyle \gamma_1 := \frac{x_1 - x_0}{Df(x_1)} (B/A)^{3/4}$
-    \\ $\alpha_0 := 4 (B/A)^{1/4}$
-    \\ $\displaystyle \alpha_1 := -\frac{x_1 - x_0}{Df(x_1)} (B/A)^{1/4}$
-    \\ $\displaystyle \beta_0 := 30 - \frac{12 \big(Df(x_0) + Df(x_1)\big) (x_1 - x_0)}{\big(f(x_1) - f(x_0)\big) \sqrt{A}\sqrt{B}}$
-    \\ $\displaystyle \beta_1 := \frac{-3 (x_1 - x_0)^2}{2 \big(f(x_1) - f(x_0)\big) \sqrt{A} \sqrt{B}} $ \label{alg1-linearized-abg}
-    \comment{The $\gamma$, $\alpha$, and $\beta$ terms with subscripts $0$ and $1$ are algebraic reductions of the simplified conditions for satisfying Theorem 2 in Equation (16) of \citeN{ulrich1994positivity}.
-These terms with subscripts $0$ and $1$ make the computation of $\alpha$, $\beta$, and $\gamma$ functions of the second derivative terms, as seen in Step \ref{alg1-abg} below.}
-  \item $\gamma := \gamma_0 + \gamma_1 D^2f(x_0)$
-    \\ $\alpha := \alpha_0 + \alpha_1 D^2f(x_1)$
-    \\ $\beta := \beta_0 + \beta_1 \big(D^2f(x_0) - D^2f(x_1)\big)$
-    \label{alg1-abg}
-  \item if $(\beta \leq 6)$ return $\big(\alpha > - (\beta + 2) / 2\big)$
-  \item else $\qquad$ return $\big(\gamma > -2 \sqrt{\beta - 2}\big)$
-    \\ end if \label{alg1-last}
-}
-%% ----------------------------------------------------------------------
-
-The reason for structuring the $\alpha$, $\beta$, and $\gamma$ computations in terms of the second derivative of $f$ (seen in Step \ref{alg1-abg} of Algorithm \ref{alg:check-monotone}) will become more apparent later.
-The next problem to consider is that of making a nonmonotone order six polynomial piece into a monotone one by modifying its first and second derivative values at the ends of an interval.
-Note that the actual value of the function at the ends of the interval is not modified, as the resulting polynomial needs to interpolate.
-
-%% ----------------------------------------------------------------------
-\algorithm
-    {make\_monotone$\big (x_0$,$x_1$,$f \big)$}
-    {where $x_0,\ x_1 \in \mathbb{R}$, $x_0 < x_1$, and $f$ is an order six polynomial defined by $f(x_0)$, $Df(x_0)$, $D^2f(x_0)$, $f(x_1)$, $Df(x_1)$, $D^2f(x_1)$.
-Returns $f$ monotone on $[x_0,x_1]$.}
-    {\label{alg:make-monotone}
-    \item if $\big(f(x_1) = f(x_0)\big)$
-      \indented{$Df(x_0) := Df(x_1) := D^2f(x_0) := D^2f(x_1) := 0$}
-      \indented{return}
-      end if
-    \item $Df(x_0) :=\ $median$\displaystyle \bigg(0,\ Df(x_0),\ 14 \frac{f(x_1) - f(x_0)}{x_1 - x_0} \bigg)$
-      \\ $Df(x_1) :=\ $median$\displaystyle \bigg(0,\ Df(x_1),\ 14 \frac{f(x_1) - f(x_0)}{x_1 - x_0}\bigg)$ \label{alg2-init-df}
-      \comment{The selection of values in Step \ref{alg2-init-df} for $Df(x_0)$ and $Df(x_1)$ is suggested by \citeN{ulrich1994positivity} and also by \citeN{huynh1993accurate}.
-These rules quickly enforce upper and lower bounds on derivative values to ensure quintic monotonicity is obtainable.}
-      \comment{The next case that will be covered is in accordance with a simplified condition for quintic monotonicity that reduces to one of cubic positivity studied in \citeN{schmidt1988positivity}, where $\alpha$, $\beta$, $\gamma$, and $\delta$ are defined in terms of values and derivatives of $f$ at $x_0$ and $x_1$.
-Steps \ref{alg2-dx-conditions} and \ref{alg2-rescale-dx} ensure that the following bounds on $D^2f(x_0)$ are compatible while Step \ref{alg2-ddx0-upper} ensures that the conditions on $D^2f(x_1)$ are compatible.
-To conclude the simplified case, Step \ref{alg2-gamma} first enforces $\gamma \geq \delta$, then Steps \ref{alg2-alpha} and \ref{alg2-beta} enforce $\alpha \geq 0$ and $\beta \geq \alpha$ respectively.
-Combined, these three conditions derived from Equation (2.18) in \citeN{schmidt1988positivity} guarantee monotonicity of this polynomial piece.}
-    \item if $\big(Df(x_0) = 0$ or $Df(x_1) = 0\big)$
-    \item \indented{$w := x_0 - x_1$\\$v := f(x_0) - f(x_1)$}
-    \item \indented{if $\big(5 Df(x_0) + 4Df(x_1) > 20v / w)$} \label{alg2-dx-conditions}
-    \item \indented{\indented{$Df(x_0) := Df(x_0) \text{ max}\displaystyle \bigg(0, \frac{20v}{w \big(5 Df(x_0) + 4 Df(x_1)\big)}\bigg) $}}
-      \indented{\indented{$Df(x_1) := Df(x_1) \text{ max}\displaystyle \bigg(0, \frac{20v}{w \big(5 Df(x_0) + 4 Df(x_1)\big)}\bigg) $}} \label{alg2-rescale-dx}
-      \indented{end if}
-    \item \indented{$D^2f(x_0) :=\ $min$\displaystyle \bigg(D^2f(x_0), \frac{4 (2 Df(x_0)+Df(x_1))+20 v/w}{w} \bigg)$} \label{alg2-ddx0-upper}
-    \item  \indented{$D^2f(x_0) :=\ $max$\big(D^2f(x_0), 3 Df(x_0) / w \big)$} \label{alg2-gamma}
-    \item \indented{$D^2f(x_1) :=\ $min$\big(D^2f(x_1), {-4}Df(x_1) / w \big)$} \label{alg2-alpha}
-    \item  \indented{$D^2f(x_1) :=\ $max$\displaystyle \bigg(D^2f(x_1), \frac{3w D^2f(x_0) - 24 Df(x_0) - 32 Df(x_1) + 60v/w}{5w} \bigg)$} \label{alg2-beta}
-      \indented{return}
-      end if
-      \comment{The following code considers the case where $Df(x_0) \neq 0$ and $Df(x_1) \neq 0$.}
-    \item $\displaystyle A := Df(x_0)\frac{x_1 - x_0}{f(x_1) - f(x_0)}$
-      \\ $\displaystyle B := Df(x_1) \frac{x_1 - x_0}{f(x_1) - f(x_0)}$
-    \item if $\big($max$(A,B) > 6\big)$
-      \indented{$Df(x_0) := 6 Df(x_0) \big/ $max$(A,B)$}
-      \indented{$Df(x_1) := 6 Df(x_1) \big/ $max$(A,B)$}
-      end if
-      \comment{This ensures that $(A,B)$ remains within a viable region of monotonicity (satisfying Theorem 4, seen in Fig. 6 of \citeN{ulrich1994positivity}).}
-    \item $\eta := \big(D^2f(x_0), D^2f(x_1)\big)$
-      \\ $\displaystyle \eta_0 := \left({-\frac{\sqrt{A}}{4}}\big(7\sqrt{A} + 3\sqrt{B}\big),\ \frac{\sqrt{B}}{4}\big(3\sqrt{A} + 7\sqrt{B}\big) \right)$
-    \item $\bigl(D^2f(x_0)$, $D^2f(x_1)\bigr) :=$ binary\_search$\bigl(g,
-       \eta,\eta_0\bigr)$
-      \comment{Here $g(z) =$\texttt{ is\_monotone}$(x_0,x_1,f)$ where $x_0$, $x_1$, $f(x_0)$, $f(x_1)$, $Df(x_0)$, $Df(x_1)$ are fixed, and the variable $z=\bigl(D^2f(x_0),D^2f(x_1)\bigr)$.
-This binary search identifies feasible values of $D^2f(x_0)$ and $D^2f(x_1)$ that are nearest to the current values $\eta$.
-This search is valid because $\eta_0$ is guaranteed to produce a monotone function, which can be seen in Equation (23) of \citeN{ulrich1994positivity}.}
-    return
-}
-%% ----------------------------------------------------------------------
-
-Notice that both Algorithms \ref{alg:check-monotone} and \ref{alg:make-monotone} have $\mathcal{O}(1)$ runtime  assuming a fixed level of relative precision.
-A constant number of operations are needed for verifying monotonicity (Algorithm \ref{alg:check-monotone}), while a constant set of operations and a single binary search are performed for enforcing monotonicity (Algorithm \ref{alg:make-monotone}).
-The search requires a fixed number of steps to achieve any predetermined relative precision.
-Also note that an efficient implementation of Algorithm \ref{alg:make-monotone} only needs to recalculate Steps \ref{alg1-abg} through \ref{alg1-last} of Algorithm \ref{alg:check-monotone} during the binary search.
-Next, an algorithm for constructing a monotone quintic spline interpolant is presented.
-
-%% ----------------------------------------------------------------------
-
-\algorithm
-    {monotone\_spline$\big((k_1$, $\ldots$, $k_n)$, $f$, $s\big)$}
-    {where $(k_1$, $\ldots$, $k_n)$ is an increasing sequence of real numbers, $f$ is an order six piecewise polynomial with breakpoints $k_1$, $\ldots$, $k_n$ defined by the data $\bigl\{f(k_i)\bigr\}_{i=1}^n$, $\bigl\{Df(k_i)\bigr\}_{i=1}^n$, $\bigl\{D^2f(k_i)\bigr\}_{i=1}^n$, and $s > 1$ is an integer shrink factor.}
-    {\label{alg:monotone-spline}
-    \item create queue
-      \\ create changed$(1$, $\ldots$, $n) := $ FALSE
-      \\ create step\_size$(1$, $\ldots$, $n) := \bigl(Df(k_1)/s$, $\ldots$, $Df(k_n)/s\bigr)$ \label{step:create}
-      \comment{The three variables defined in Step \ref{step:create} are used to ensure the eventual achievement of monotonicity.
-The \texttt{queue} is a standard first in first out (FIFO) queue with \texttt{enqueue} and \texttt{dequeue} operations.
-The array \texttt{changed} contains Booleans describing whether or not a breakpoint belongs to an interval that has been modified to enforce monotonicity.
-The \texttt{step\_size} array contains the real-valued derivative decrement step sizes to use in the search for a valid monotone spline.}
-    \item for $i := 1$, $\ldots$, $n-1$
-      \indented{enqueue $\bigl((k_i, k_{i+1})\bigr)$}
-      \comment{Initially, all intervals must be checked for monotonicity. The following loop will continue until all intervals are verified as monotone without need for modification.}
-    \item while $($size$($queue$) > 0)$
-    \item \indented{$(x_j$, $x_{j+1}) := $ dequeue}
-    \item \indented{if $\big($not is\_monotone$(x_j$, $x_{j+1}$, $f)\big)$}
-    \item \indented{\indented{if $\big($changed$(j)\big) \qquad \ Df(x_j) := Df(x_j) -{}$step\_size$(j) \qquad \ \ $ end if}}
-      \indented{\indented{if $\big($changed$(j+1)\big) \quad Df(x_{j+1}) := Df(x_{j+1}) -{}$step\_size$(j+1)$ end if}}
-      \indented{\indented{\comment{If a breakpoint belongs to an interval that has been previously modified, then the enforced monotonicity conditions on the second derivatives of adjacent intervals must contradict one another.
-In turn, the involved first derivatives are decreased towards zero by a predetermined step size.}}}
-    \item \indented{\indented{make\_monotone$(x_j,\ x_{j+1},\ f)$}}
-    \item \indented{\indented{changed$(j) :=$ TRUE}}
-      \indented{\indented{changed$(j+1) :=$ TRUE}} \label{step:changed}
-    \item \indented{\indented{if $\big(j > 1$ and $(x_{j-1}, x_j) \not\in$ queue$\big)$ enqueue $\bigl((x_{j-1}, x_j)\bigr)$ end if}}
-      \indented{\indented{enqueue $\bigl((x_j, x_{j+1})\bigr)$}}
-      \indented{\indented{if $\big(j+1 < n$ and $(x_{j+1}, x_{j+2}) \not\in$ queue$\big)$ enqueue $\bigl((x_{j+1}, x_{j+2})\bigr)$ end if}} \label{step:requeue}
-      \indented{\indented{\comment{Step \ref{step:changed} records the endpoints of the current interval as having been changed, while \ref{step:requeue} adds adjacent intervals to \texttt{queue} that may have inadvertently been made nonmonotone by the changes to the present interval.}}}
-      \indented{end if}
-      end while
-}
-%% ----------------------------------------------------------------------
-
-It is mentioned in \citeN{ulrich1994positivity} that for sufficiently small $Df(k_i)$ and $Df(k_{i+1})$ the admissible solution interval of second derivative values becomes arbitrarily large.
-It can also be seen that decreasing the assigned derivative values towards zero will eventually cause Steps \ref{alg1-alpha} through \ref{alg1-gamma} of Algorithm \ref{alg:check-monotone} to all fail, resulting in Step \ref{alg1-simplified-true} returning \texttt{TRUE}.
-
-Given the potential for successive monotonicity fixes across all intervals, the worst case runtime of Algorithm \ref{alg:monotone-spline} is $\mathcal{O}(s n)$ for $n$ breakpoints and shrink factor $s$.
-In practice this worst case has been observed to be unlikely and difficult to produce.
-
-\section{Experimental Results}
-
-Given the increased order of approximation, it is naturally expected that some accuracy benefit could be gained by using a quintic spline over a cubic spline.
-Experiments 1 and 2 below test some accuracy differences between monotone cubic and quintic splines.
-Experiment 3 analyzes the number of operations performed by Algorithm \ref{alg:monotone-spline} with increasingly large samples of monotone data.
-
-\subsection{Approximating a Trigonometric Function}
-
-For this experiment, the function $\sin(x) + x$ is considered over the interval $[0,(5/2)\pi]$ as seen in Figure \ref{fig:cubic_quintic_sin}.
-Given an increasing number of points, the maximum error of cubic and quintic approximations is shown in Figure \ref{fig:experiment_1}.
-The quintic approximation consistently has a maximum error that is roughly half that of the cubic approximation, given the same number of points.
-There is an increase in the error gap between the two approximations as more points are added.
-The quintic spline requires only $700$ points to approximate the function to the same accuracy achieved by the cubic with $1000$ points.
-
-\begin{figure}
-  \centering
-  \includegraphics[width=.7\textwidth]{cubic-quintic-sin}
-  \caption{Depicted above are the monotone cubic and quintic spline interpolants of the function $\sin(x) + x$ over the points $[0, (5/2) \pi]$.
-Notice that the maximum error of the cubic interpolant is larger, because it only captures first derivative information at the points.
-The quintic interpolant captures both first and second derivative information at the points.}
-  \label{fig:cubic_quintic_sin}
-\end{figure}
-
-\begin{figure}
-  \centering
-  \includegraphics[width=.7\textwidth]{experiment_1_errors}
-  \caption{The error in the quintic and cubic approximations of $\sin(x) + x$ with increasing number of points.
-Both axes are log scaled.
-The quintic approximation generally provides an approximation with about half of the maximum error of the cubic approximation, while the relative gap between grows with increasing number of points.
-In order to achieve the same absolute error the quintic spline requires ten fewer points at max error $10^{-4}$, thirty fewer points at max error $10^{-6}$, and $300$ fewer points at error $10^{-8}$.}
-  \label{fig:experiment_1}
-\end{figure}
-
-\subsection{Random Number Generation}
-
-Consider the cumulative distribution function (CDF) defined by a mixture of three Gaussian distributions and visualized in Figure \ref{fig:experiment_2_distribution}.
-The distributions have weights $(.3, .6, .1)$, means $(.2, .45, .85)$ and standard deviations $(.05, .08, .03)$.
-$200$ random samples are generated from this distribution with both cubic and quintic approximations from four points, and over $100$ trials the resulting spread of empirical distribution functions (EDFs) are visualized in Figure \ref{fig:experiment_2_results}.
-
-While an increased number of points could be used to decrease the error in either approximation, the purpose of this experiment is to demonstrate the effect approximation error has on random number generation.
-The errors in the CDFs are magnified by the variance involved in randomly sampling from a distribution.
-As a result, while the cubic and quintic spline approximations have similar maximum error, the cubic distribution is significantly distorted around the first and second modes.
-
-\begin{figure}
-  \centering
-  \includegraphics[width=.8\textwidth]{experiment_2_distribution}
-  \caption{The cumulative distribution function (top) and probability density function (bottom) for the Gaussian mixture distribution in Experiment 2.
-The distributions have weights $(.3, .6, .1)$, means $(.2, .45, .85)$ and standard deviations $(.05, .08, .03)$.
-Random samples are generated from this mixture distribution by evaluating the inverse of the CDF at random numbers uniformly distributed in the range $[0,1]$.}
-  \label{fig:experiment_2_distribution}
-\end{figure}
-
-\begin{figure}
-  \centering
-  \includegraphics[width=.4\textwidth]{experiment_2_cubic_dist}
-  \includegraphics[width=.4\textwidth]{experiment_2_quintic_dist}
-  \caption{Both cubic and quintic splines are used to approximate the CDF of a Gaussian mixture distribution with three components.
-Four CDF points are used to build the approximation, $200$ random samples are generated with the approximate CDFs over $100$ trials.
-The resulting cloud of empirical distribution function (EDF) values is seen above.
-The maximum approximation error in the CDFs are similar, $.08$ for the cubic and $.05$ for the quintic.
-However, the maximum EDf differences observed are doubled by the variance of random samples, $.15$ for the cubic and $.1$ for the quintic.}
-  \label{fig:experiment_2_results}
-\end{figure}
-
-%% - generate large sets of random monotone data, show Table of number
-%%   of metrics growing with increasing number of points
-
-%% - show visual of the \textit{error}
-
-\subsection{Random Monotone Data}
-
-This experiment studies the number of times the algorithms \texttt{is\_monotone} and \texttt{make\_monotone} are executed for increasingly large sequences of random monotone data.
-The minimum, median, and maximum number of times that Algorithms \ref{alg:check-monotone} and \ref{alg:make-monotone} are executed is recorded, as well as the number of steps taken in all calls to \texttt{binary\_search}.
-The number of calls and steps grows linearly with $n$ as expected, requiring roughly one call to \texttt{make\_monotone} and $20$ binary search steps to create a monotone piece.
-Some (less common) problems require an average of three calls to \texttt{make\_monotone} per interval.
-
-\begin{table}
-  \centering
-  \begin{tabular}{c|c c c|c c c|c c c}
-    \hline
-    \multirow{2}{*}{$n$}
-      & \multicolumn{3}{c|}{Checks} & \multicolumn{3}{c}{Fixes} & \multicolumn{3}{|c}{Search Steps} \\
-      & min & median & max & min & median & max & min & median & max \\
-    \hline
-    5 & 4 & 6 & 274 & 0 & 1 & 154 & 3 & 31 & 2006\\
-    25 & 28 & 47 & 378 & 2 & 14 & 225 & 80 & 348 & 3742\\
-    50 & 61 & 124 & 484 & 6 & 40 & 238 & 217 & 851 & 4445\\
-    75 & 102 & 216 & 842 & 14 & 74 & 411 & 463 & 1491 & 6333\\
-    100 & 142 & 306 & 776 & 21 & 107 & 380 & 629 & 2140 & 7414\\
-    \hline
-  \end{tabular}
-  \caption{Here we see a table of the minimum, median, and maximum number of checks, fixes, and binary search steps required in the execution of \texttt{make\_spline} for increasing size sequences, $n$, over $100$ randomly generated sets of monotone data.
-Notice the maximum for each counter is often significantly greater than the minimum and median, because the distribution of each counter is skewed right.}
-  \label{table:e3_results}
-\end{table}
-
-\section{Discussion}
-
-The monotone quintic spline interpolant provides a distinct increase in accuracy of the monotone cubic variant.
-Although the complexity of the algorithm is increased, the number of points required to achieve the same level of accuracy has decreased.
-In the present state, the algorithm for enforcing monotonicity on a spline is not as trivially parallel as the cubic algorithm, however it can still be parallelized.
-
-Acknowledging that the algorithm for constructing a monotone quintic spline interpolant is slightly more complex than the cubic case, gains in accuracy or a decreased necessary number of points are often worth the computational effort.
-
-\section{Conclusion and Future Work}
-
-This paper proposes and tests an algorithm for constructing monotone quintic spline interpolants.
-Experiments demonstrate an improvement in approximation accuracy over monotone cubic spline interpolants, as expected based on theory.
-
-There are still open avenues of research going forward, such as an alternative sufficient condition for enforcing monotonicity or increased order monotone approximations.
-If the monotonicity conditions can be generalized to any order or made linear, the search for a monotone interpolating spline could potentially be formulated as a convex optimization problem.
-Finally, this work could be used to improve cumulative distribution function (CDF) estimates as well as predictive models that use CDFs.
-
-%% \section*{Acknowledgments}
-%% The NSF is acknowledged for its support of this project.
-
-% Please don't change the bibliographystyle style
-\bibliographystyle{scsproc}
-% AUTHOR: Include your bib file here
-\bibliography{paper}
-
-\section*{Author Biographies}
-
-\textbf{\uppercase{THOMAS C. H. LUX}} is a Ph.D. candidate in computer
-science at Virginia Polytechnic Institute and State University
-(VPI\&SU) studying under Dr. Layne Watson. His research interests
-include computational science, approximation theory, optimization, and
-artificial intelligence. His email address is \email{tchlux@vt.edu}.
-
-\textbf{\uppercase{Layne T. Watson}} (Ph.D., Michigan, 1974) has
-interests in numerical analysis, mathematical programming,
-bioinformatics, and data science.  He has been involved with the
-organization of HPCS since 2000.
-
-\textbf{\uppercase{Tyler H. Chang}} is a Ph.D. candidate at VPI\&SU
-studying computer science under Dr. Layne Watson.
-
-\textbf{\uppercase{Li Xu}} is a Ph.D. candidate at VPI\&SU studying
-statistics under Dr. Yili Hong.
-
-\textbf{\uppercase{Yueyao Wang}} is a Ph.D. candidate at VPI\&SU
-studying statistics under Dr. Yili Hong.
-
-\textbf{\uppercase{Yili Hong}} (Ph.D., Iowa State, 2009) has interests in
-engineering statistics, statistical modeling, and data analysis.
-
-%% \newpage
-
-%% \appendix
-
-%% \section{Author Checklist}
-%% We strive for a consistent appearance in all papers published in the proceedings.
-If you used the template and styles within this author’s kit, then almost all of the requirements in this checklist will be automatically satisfied, and there is very little to check.
-
-%% Please \textbf{print a hardcopy of your paper}, and go over your printed paper to make sure it adheres to the following requirements. \textit{Thank you!}
-
-%% \begin{enumerate}
-%% 	\item Abstract
-%%   \begin{enumerate}
-%% 	  \item 150 or fewer words.
-%% 	  \item Provide 3-5 keywords (mandatory). This set of keywords will identify your paper in indices and databases.
-%% 	\end{enumerate}
-%% 	\item Paper Length
-%%   \begin{enumerate}
-%% 	  \item At least 5, but no more than 12 pages.
-%% 	  \item Page size is letter size (8.5’’ x 11’’, or 216 mm x 279 mm).
-%% 	\end{enumerate}
-%% 	\item All text is in 11-Point Times New Roman except title, header and footer (default in this template).
-%% 	\item Paper title is in 12-Point Times New Roman \textbf{BOLDFACE ALL CAPS} (default in this template).
-%% 	\item The paper has been spellchecked using U.S. English. 
-%% 	\item Spacing and Margins
-%%   \begin{enumerate}
-%% 	  \item Single spaced.
-%% 	  \item Left and right margins are each 1 inch (default in this template).
-%% 	  \item Top and bottom margins are according to the template.
-%% 	  \item Title starts 1.25 inches from the top of the page (default in this template).
-%% 	\end{enumerate}
-%% 	\item Section Headings
-%%   \begin{enumerate}
-%% 	  \item Left justified and set in \textbf{BOLDFACE ALL CAPS} (default in this template).
-%% 	  \item Numbered, except for the abstract, acknowledgments, references and author biographies.
-%% 	  \item Subsection headings are set in \textbf{Boldface Headline Style} (default in this template).
-%% 	\end{enumerate}
-%% 	\item No footnotes or page numbers.
-%% 	\item The copyright notice on the first page contains the name of the symposium or a track where you want to submit your paper, and the running head on subsequent pages is the surnames of all authors.
-%% 	\item Multiple authors are formatted correctly.
-%% 	\item Equations are centered and any equation numbers are in parentheses and right-justified (default).
-%% 	\item Figures and Tables
-%%   \begin{enumerate}
-%% 	  \item All text in figures and tables is readable.
-%% 	  \item Table captions appear above the table.
-%% 	  \item Figure captions appear below the figure.
-%% 	\end{enumerate}
-%% 	\item Citations and References
-%%   \begin{enumerate}
-%% 	  \item Citations are by author and year, and are enclosed in parentheses, not brackets (default \BibTeX). 
-%% 	  \item References are in the hangref style, and are listed alphabetically by the last names(s) of the author(s) (also default \BibTeX\ setting in this template). 
-%% 	\end{enumerate}
-%% 	\item Author biographies are one paragraph per author.
-%% 	\item Hyperlinks
-%%   \begin{enumerate}
-%% 	  \item Be sure that hyperlinks will probably work in the future as well.
-%% 	  \item Live hyperlinks are blue. Nonlive hyperlinks are black (default in this template).
-%% 	\end{enumerate}
-%% 	\item All fonts must be embedded in the resulting \texttt{.pdf}.
-If you use \texttt{pdflatex}, this should already be true--unless you included figures in the \texttt{.eps} or \texttt{.pdf} format which may introduce additional font dependencies.
-You can use the \texttt{pdffonts} command to see if all the fonts are embedded (the column ``emb'' should say yes for all rows).
-You can use GhostScript to force embed the fonts, like so: {\texttt{gs -dNOPAUSE -dBATCH -sDEVICE=pdfwrite -dEmbedAllFonts=true -sOutputFile=out.pdf -f in.pdf}} (in Windows, use \texttt{gswin32} or \texttt{gswin32c} to invoke GhostScript).
-%% \end{enumerate}
-
-%% After verifying that your paper meets these requirements, please go to the final submission page at conference website and submit your paper.
-Be sure to complete the transfer of copyright form and upload the \texttt{.pdf} receipt.
-\textit{Thank you for contributing to the SCS conferences!}
-
-%% \section{Most Observed Mistakes}
-
-%% The following list comprises \textbf{the most common sources of error} that had to be corrected by previous editors.
-Please make sure to go through the following list and check that your paper is formatted correctly:
-%% \begin{enumerate}
-%% 	\item	The paper is less than 5 or more than 12 pages long.
-%% 	\item	Paper title and section titles are in \textbf{BOLD ALL CAPS}, subsections are \textbf{Bold and Capitalize the First Letter of Important Words}. Please use the templates.
-%% 	\item	Paper is in A4 format, not letter format. Please use the required margins.
-%% 	\item	The copyright notice is incorrect.
-%% 	\item	The running heads are incorrect. Do not forget that the LastNameLastAuthor is preceded by ``, and ''.
-%% 	\item	The citation format is incorrect. Use \BibTeX\ for citations and do not change the format used by the template.
-%% 	\item	The biographies are missing. Do not forget the ``author biographies'' section.
-%% 	\item	Figures or tables are not referenced in the text or have the incorrect caption format.
-%% 	\item	The author section after the title is not formatted correctly, the number of organizations does not define the number of blocks, or the number of blocks does not define the layout.
-%% 	\item In the heading on the title page, country names are in all capitals.
-%% 	\item	Paragraphs are not indented.
-%% 	\item Some fonts in the resulting \texttt{.pdf} are not embedded.
-%% \end{enumerate}
-
-\end{document}
+This paper proposes and tests an algorithm for constructing monotone
+quintic spline interpolants. Experiments demonstrate an improvement in
+approximation accuracy over monotone cubic spline interpolants, as
+expected based on theory. There are still open avenues of research
+going forward, such as an alternative sufficient condition for
+enforcing monotonicity or increased order monotone approximations. If
+the monotonicity conditions can be generalized to any order or made
+linear, the search for a monotone interpolating spline could
+potentially be formulated as a convex optimization problem. Finally,
+this work could be used to improve cumulative distribution function
+(CDF) estimates as well as predictive models that use CDFs.

### 2020-06-26 Research_[2020-05-18]_Piecewise_Monotone_Quintic_Fortran/misc/example_tex/old_files/0-discussion.tex
Changed some directory names and fixed some typos in code files. Still working on the paper.

\heading{4. DISCUSSION}

The monotone quintic spline interpolant provides a distinct increase
in accuracy over the monotone cubic variant. The computational
complexity of this algorithm is identical to the existing cubic method
and the cost of evaluating the spline after construction is
unchanged. While the number of computations required per interval has
increased, the number of points required to achieve the same level of
accuracy has decreased. In the present state, the algorithm for
enforcing monotonicity on a spline is not as trivially parallel as the
cubic algorithm, however it can still be parallelized. The checks for
monotonicity across all intervals are independent and the monotonicity
adjustments could be computed independently and intelligently merged
with minor changes to the serial algorithm.

Acknowledging that the algorithm for constructing a monotone quintic
spline interpolant is slightly more expensive than the cubic case,
gains in accuracy or decreased necessary number of points are often
worth the computational effort. In practice the costs of spline
interpolation are often dominated by evaluation rather than
construction, and the increased accuracy is afforded for a negligible
increase in computational cost.

### 2020-06-26 Research_[2020-05-18]_Piecewise_Monotone_Quintic_Fortran/paper/2-monotone-quintic.tex
Changed some directory names and fixed some typos in code files. Still working on the paper.

@@
   \sqrt{\alpha \delta}$, and Step 11 checks $\gamma < \delta - 2
-  \sqrt{\alpha \delta}$. If none of the monotonicity conditions were
+  \sqrt{\alpha \delta}$. If none of the monotonicity conditions are
   violated, then the order six piece is monotone and Step 12
@@
 practice a quasi-bisection search is used in place of solely shrinking
-$DQ$ and $D^2Q$, adding an additional step that increases previously
-shrunken derivative values by a smaller increment than the previous
-decrement. The goal is to converge on the boundary of the monotone region
-in the $(\tau_1$, $\alpha$, $\beta$, $\gamma)$ space. As long as a step size
-schedule is set that allows the value zero to be obtained in a fixed
-number of computations, this has no effect on computational
-complexity. Notably, since {\tt IS\_MONOTONE} handles both nondecreasing and
-nonincreasing simultaneously by taking into account the sign of $v$,
-Algorithm 3 produces $Q(x)$ that is monotone increasing (decreasing) over
-exactly the same intervals that the data $(X_i$, $Y_i)$ is monotone
-increasing (decreasing).
+$DQ$ and $D^2Q$. This algorithm is outlined below. The goal is to
+converge on the boundary of the monotone region in the $(\tau_1$,
+$\alpha$, $\beta$, $\gamma)$ space.
+
+%% ===================================================================
+%%              Algorithm 4: Quasi-bisection search
+\vskip 5mm
+{\parindent 0mm
+{\bf Algorithm 4:} {\tt QUASI-BISECTION}$\bigl(x_0, x_1, f \bigr)$
+
+where $x_0$, $x_1 \in {\bbb R}$, $x_0 < x_1$, and $f$ is an order six
+polynomial that is nonmonotone on the interval $[x_0$, $x_1]$. Defines
+a stepping procedure for function derivative values that ensures $f$
+can be made monotone in a fixed number of steps given a relative
+precision $e \in {\bbb R}$.
+
+}
+{\parindent=6mm
+\item{} Set a starting {\tt step\_size} and save the initial value of $f$.
+\item{} $\hat f := f$
+\item{} {\tt step\_size} $:= 1/2$
+\item{} {\tt shrinking} $:=$ {\tt TRUE}
+\item{} {\tt do while } $\big( ($ {\tt step\_size} $ > e)$ {\tt or not
+  IS\_MONOTONE}$(x_1$, $x_2$, $f)$ $\big)$
+\item{} \codent {\tt if shrinking then step\_size} $:=$ {\tt step\_size} $/ 2$
+
+}
+\vskip 5mm
+%% ----------------------------------------------------------------------
+
+As long as a step size schedule is
+set that allows the value zero to be obtained in a fixed number of
+computations, this has no effect on computational complexity. Notably,
+since {\tt IS\_MONOTONE} handles both nondecreasing and nonincreasing
+simultaneously by taking into account the sign of $v$, Algorithm 3
+produces $Q(x)$ that is monotone increasing (decreasing) over exactly
+the same intervals that the data $(X_i$, $Y_i)$ is monotone increasing
+(decreasing).

### 2020-06-26 Research_[2020-05-18]_Piecewise_Monotone_Quintic_Fortran/paper/4-analysis.tex
Changed some directory names and fixed some typos in code files. Still working on the paper.

@@
 monotone $Q(x)$.
+
+\topinsert
+\centerline{\epsfxsize=4truein \epsffile{vis/1-sensitivity.eps}}
+{\narrower\noindent\rmVIII Fig.\ 1.
+A demonstration of the quadratic facet model's sensitivity to small
+data perturbations. In this example two quadratic functions $f_1(x) =
+x^2$ over $\{1$,$2$,$5/2\}$ and $f_2(x) = (x-2)^2 + 6$ over
+$\{5/2$,$3$,$4\}$ have the same curvature and equal value at
+$x=5/2$. Given the five points above Algorithm 2 produces the slope of
+the solid blue line because $f_1$ and $f_2$ have equal curvature.
+However, perturbing the curvature of $f_2$ down by the machine
+precision at values $x=\{3$, $4\}$ causes Algorithm 2 to produce the
+slope of the red dashed line at $x = 5/2$. This is the nature of a
+facet model and a side effect of associating data with local facets.
+\par}
+\endinsert

### 2020-06-26 Research_[2020-05-18]_Piecewise_Monotone_Quintic_Fortran/paper/5-applications.tex
Changed some directory names and fixed some typos in code files. Still working on the paper.

\heading{5. PERFORMANCE AND APPLICATIONS}

This section contains graphs of sample {\tt MQSI} results given
various sources of data. Tables of computation times for various
problems sizes are also provided.

\midinsert
\centerline{\epsfxsize=4truein \epsffile{vis/2-piecewise-polynomial.eps}}
\centerline{\epsfxsize=4truein \epsffile{vis/3-large-tangent.eps}}
\centerline{\epsfxsize=4truein \epsffile{vis/4-signal-decay.eps}}
{\narrower\noindent\rmVIII Fig.\ 2.
{\ttVIII MQSI} results for three of the functions in the included test
suite. The {\itVIII piecewise polynomial} function (top) shows the
interpolant capturing local linear segments, local flats, and
alternating extreme points. The {\itVIII large tangent} (middle)
problem demonstrates outcomes on rapidly changing segments of data.
The {\itVIII signal decay} (bottom) alternates between extreme values
of steadily decreasing magnitude.
\par}
\endinsert

\midinsert
\centerline{\epsfxsize=4truein \epsffile{vis/5-random-monotone.eps}}
{\narrower\noindent\rmVIII Fig.\ 3.
The {\itVIII random monotone} test poses a particularly challenging
problem with large variations in slope. Notice that despite drastic
shifts in slope, the resulting {\ttVIII MQSI} provides smooth and
reasonable estimates to function values between data.
\par}
\endinsert

### 2020-06-26 Research_[2020-05-18]_Piecewise_Monotone_Quintic_Fortran/python/REAL_PRECISION.f90

@@ -1,3 +1,1 @@
-MODULE REAL_PRECISION  ! module for 64-bit real arithmetic
-  INTEGER, PARAMETER:: R8=SELECTED_REAL_KIND(13)
-END MODULE REAL_PRECISION
+../code/REAL_PRECISION.f90

### 2020-06-26 Research_[2020-05-18]_Piecewise_Monotone_Quintic_Fortran/python/lapack.f

@@ -1,2189 +1,1 @@
-      SUBROUTINE DGBSV( N, KL, KU, NRHS, AB, LDAB, IPIV, B, LDB, INFO )
-*
-*  -- LAPACK driver routine (version 3.2) --
-*  -- LAPACK is a software package provided by Univ. of Tennessee,    --
-*  -- Univ. of California Berkeley, Univ. of Colorado Denver and NAG Ltd..--
-*     November 2006
-*
-*     .. Scalar Arguments ..
-      INTEGER            INFO, KL, KU, LDAB, LDB, N, NRHS
-*     ..
-*     .. Array Arguments ..
-      INTEGER            IPIV( * )
-      DOUBLE PRECISION   AB( LDAB, * ), B( LDB, * )
-*     ..
-*
-*  Purpose
-*  =======
-*
-*  DGBSV computes the solution to a real system of linear equations
-*  A * X = B, where A is a band matrix of order N with KL subdiagonals
-*  and KU superdiagonals, and X and B are N-by-NRHS matrices.
-*
-*  The LU decomposition with partial pivoting and row interchanges is
-*  used to factor A as A = L * U, where L is a product of permutation
-*  and unit lower triangular matrices with KL subdiagonals, and U is
-*  upper triangular with KL+KU superdiagonals.  The factored form of A
-*  is then used to solve the system of equations A * X = B.
-*
-*  Arguments
-*  =========
-*
-*  N       (input) INTEGER
-*          The number of linear equations, i.e., the order of the
-*          matrix A.  N >= 0.
-*
-*  KL      (input) INTEGER
-*          The number of subdiagonals within the band of A.  KL >= 0.
-*
-*  KU      (input) INTEGER
-*          The number of superdiagonals within the band of A.  KU >= 0.
-*
-*  NRHS    (input) INTEGER
-*          The number of right hand sides, i.e., the number of columns
-*          of the matrix B.  NRHS >= 0.
-*
-*  AB      (input/output) DOUBLE PRECISION array, dimension (LDAB,N)
-*          On entry, the matrix A in band storage, in rows KL+1 to
-*          2*KL+KU+1; rows 1 to KL of the array need not be set.
-*          The j-th column of A is stored in the j-th column of the
-*          array AB as follows:
-*          AB(KL+KU+1+i-j,j) = A(i,j) for max(1,j-KU)<=i<=min(N,j+KL)
[…]

### 2020-06-26 Research_[2020-05-18]_Piecewise_Monotone_Quintic_Fortran/python/mqsi.py

@@ -139,7 +139,7 @@ class Fraction:
         elif (self.denominator == 1): return str(self.numerator)
         my_string = f"{self.numerator} / {self.denominator}"
         float_string = str(float(self))
-        if (len(float_string) < len(my_string)): return float_string
+        if (2*len(float_string) < len(my_string)): return float_string
         else:                                    return my_string
     # Format this Fraction by converting it to a float.
     def __format__(self, *args, **kwargs): return float(self).__format__(*args,**kwargs)
@@ -1036,7 +1036,7 @@ if __name__ == "__main__":
 # one with the lowest curvature.
 def quadratic_facet(x, y, i, extremes, flats):
     # If this is a local flat, estimate 0.
-    if (i in flats): return [0, 0]
+    if (i in flats): return (0, 0)
     # If this is a local maximum, force first derivative to zero, 
     # and assume that it is not an endpoint.
     elif (i in extremes):
@@ -1096,7 +1096,7 @@ def quadratic_facet(x, y, i, extremes, flats):
         ddxi = ddf(x[i])
         derivatives += [(dxi, ddxi)]
     # If there were no viable options, return 0's.
-    if (len(derivatives) == 0): return [0, 0]
+    if (len(derivatives) == 0): return (0, 0)
     # Sort the derivatives by magnitude of curvature, then return the 
     # values from the function with the least curvature.
     derivatives.sort(key=lambda d: abs(d[1]))
@@ -1254,6 +1254,40 @@ def is_monotone(U0, U1, F0, DF0, DDF0, F1, DF1, DDF1, EPS=2**(-26)):
     if (BETA <= 6): TEMP = -(BETA + 2) / 2
     else:           TEMP = -2*(BETA - 2)**(0.5)
     return (ALPHA > TEMP) and (GAMMA > TEMP)
+
+# Construct an interpolant that minimizes the L2 of the second derivative.
+def min_curve(x, y, t=10):
+    # Pick first and second derivative values that minimize the L2.
+    import numpy as np
+    from util.optimize import minimize, zero_on_line
+    # Define an objective function that minimizes L2 and makes the derivative positive.
+    def obj_func(v):
+        f = Spline(x, list(zip(y,v[0::2],v[1::2])))
+        df = f.derivative()
+        dfx = df(x)
+        ddf = df.derivative()
+        ddfx = ddf(x)
[…]

### 2020-06-26 Research_[2020-05-18]_Piecewise_Monotone_Quintic_Fortran/python/optimization_checkpoint.py

@@ -1,0 +1,2 @@
+solution = [0.000000e+00, 0.000000e+00, 2.010187e-01, 2.745939e+00, 4.210155e-01, 0.000000e+00, 4.146332e-01, 0.000000e+00, 6.163935e-01, 5.192494e+00, 9.494104e-01, 0.000000e+00, 1.225327e-01, 0.000000e+00, 1.522333e-01, 8.659047e-01, 2.100998e+01, 0.000000e+00, 1.367723e+00, -2.138334e+01, 8.096734e-01, -4.205948e+00, 8.354332e-01, -4.557465e+00, 6.665069e-01, 0.000000e+00, 3.551195e+00, 0.000000e+00, 1.338609e-01, -2.416559e+00, 0.000000e+00, 0.000000e+00]
+obj_value = 444009006.5371172

### 2020-07-20 Research_[2020-05-18]_Piecewise_Monotone_Quintic_Fortran/paper/2-monotone-quintic.tex
Updated again and sent to Dr. Watson, already received feedback and about to reincorporate it.

--- 2020-06-23 Research_[2020-05-18]_Piecewise_Monotone_Quintic_Fortran/paper/2-monotone-quintic.tex
@@
 pieces.
-
-% The remainder of this section outlines three algorithms for
-% constructing a quintic spline interpolant.  Algorithm 1 uses a
-% quadratic facet model to construct initial first and second
-% derivative estimates for the interpolating spline, Algorithm 2
-% checks monotonicity of an interval, and Algorithm 3 defines a
-% procedure for stepping initial first and second derivative estimates
-% towards values which result in a monotone quintic spline.
 
@@
 
-where $X_j$, $Y_j$ $\in \bbb{R}$ for $j = 1$, $\ldots$, $n$, and $1 < i
-< n$. Returns the slope and curvature at $X_i$ of the local quadratic
-interpolant with minimum magnitude curvature.
+where $X_j$, $Y_j$ $\in \bbb{R}$ for $j = 1$, $\ldots$, $n$, and $1
+\le i \le n$. Returns the slope and curvature at $X_i$ of the local
+quadratic interpolant with minimum magnitude curvature.
 
@@
 \itemitem{} {The point $(X_i$, $Y_i)$ is in a monotone segment of
-data. In the following, it is possible that either $f_1$ or $f_3$ does
-not exist because $i = 2$ or $i = n-1$. In those cases, the minimum
+data. In the following, it is possible that $f_1$, $f_2$, or $f_3$ does
+not exist because $i \in \{1, 2, n-1, n\}$. In those cases, the minimum
 magnitude curvature among existing quadratics is chosen.}
@@
   then; return FALSE; endif}
-\item{5.} $w := x_1 - x_0$
-\item{6.} $v := f(x_1) - f(x_0)$
+\item{5.} $w \: x_1 - x_0$
+\item{6.} $v \: f(x_1) - f(x_0)$
 
@@
 return FALSE; endif}
-\item{9.} \codent $t := 2 \sqrt{Df(x_0) (4Df(x_1) - D^2f(x_1) w) }$
+\item{9.} \codent $t \: 2 \sqrt{Df(x_0) (4Df(x_1) - D^2f(x_1) w) }$
 \item{10.} \codent {\tt if} $\bigl(t + 3Df(x_0) + D^2f(x_0)w < 0 \bigr)$
@@
 \item{} \codent {\tt then; return FALSE; endif}
-\item{15.} $t := \bigl(Df(x_0)\, Df(x_1)\bigr)^{3/4}$
-\item{16.} $\alpha := (4 Df(x_1) - D^2f(x_1)w) \sqrt{Df(x_0)} / t$
-\item{17.} $\gamma := (4 Df(x_0) - D^2f(x_0)w) \sqrt{Df(x_1)} / t$
-\item{18.} $\displaystyle \beta := {60v/w + 3\bigl(w(D^2f(x_1) -
+\item{15.} $t \: \bigl(Df(x_0)\, Df(x_1)\bigr)^{3/4}$
+\item{16.} $\alpha \: (4 Df(x_1) - D^2f(x_1)w) \sqrt{Df(x_0)} / t$
+\item{17.} $\gamma \: (4 Df(x_0) - D^2f(x_0)w) \sqrt{Df(x_1)} / t$
+\item{18.} $\displaystyle \beta \: {60v/w + 3\bigl(w(D^2f(x_1) -
   D^2f(x_0)) - 8(Df(x_0) + Df(x_1))\bigr) \over 2 \sqrt{Df(x_0)\,Df(x_1)}}$
@@
 %%       Algorithm 3: MQSI - Monotone quintic spline interpolant
-\vskip 5mm
+\vskip 3mm
 {\parindent 0mm
@@
 {\parindent=6mm
-\item{} Approximate first and second derivatives at all $X_i$ with
-{\tt QUADRATIC\_FACET}.
-\item{} {\tt do} $i = 1, \ldots n-1$
-\item{} \codent $\bigl(D Q(X_i)$, $D^2 Q(X_i)\bigr):=$
+\item{} Approximate first and second derivatives at $X_i$ with {\tt
+  QUADRATIC\_FACET}.
+\item{} {\tt do} $i = 1, \ldots n$
+\item{} \codent $\bigl(D Q(X_i)$, $D^2 Q(X_i)\bigr)\:$
 {\tt QUADRATIC\_FACET}$(X$, $Y$, $i)$
 \item{} {\tt enddo}
-\item{} Identify and store all nonmonotone intervals in a {\tt queue}.
+\item{} Identify and store all intervals where $Q$ is nonmonotone in a {\tt queue}.
 \item{} {\tt do} $i = 1, \ldots n-1$
@@
 \item{} {\tt enddo}
-\item{} {\tt do while} $\bigl($ {\tt queue} of nonmonotone intervals is
-nonempty $\bigr)$
-\itemitem{} Shrink (in magnitude) $DQ$ and $D^2Q$ that border nonmonotone
-intervals.
-\itemitem{} Identify and store all remaining nonmonotone intervals in
-{\tt queue}.
+\item{} {\tt do while} $\bigl($ {\tt queue} of intervals is nonempty $\bigr)$
+\itemitem{} Shrink (in magnitude) $DQ$ and $D^2Q$ that border intervals where $Q$ is nonmonotone.
+\itemitem{} Identify and store all remaining intervals where $Q$ is nonmonotone in {\tt queue}.
 \item{} {\tt enddo}
+\item{} Construct and return a spline representation of $Q(x)$.
+}
+\vskip 3mm
+%% ----------------------------------------------------------------------
+
+Since {\tt IS\_MONOTONE} handles both nondecreasing and
+nonincreasing simultaneously by taking into account the sign of $v$,
+Algorithm 3 produces $Q(x)$ that is monotone increasing (decreasing)
+over exactly the same intervals that the data $(X_i$, $Y_i)$ is
+monotone increasing (decreasing).
+
+Given the minimum magnitude curvature nature of the initial derivative
+estimates, it is desirable to make the smallest necessary changes to
+the initial interpolating spline $Q$ while enforcing monotonicity. In
+practice a binary search for the boundary of monotonicity is used in
+place of solely shrinking $DQ$ and $D^2Q$ at breakpoints adjoining
+{\it active} intervals, or intervals for which $Q$ is nonmonotone at
+least once during the search. The binary search considers a Boolean
+function $B(s)$, for $0 \le s \le 1$, that is true if the order six
+polynomial on $[x_i, x_{i+1}]$ matching derivatives $f(x_i)$, $s
+Df(x_i)$, $s D^2f(x_i)$ at $x_i$, and derivatives $f(x_{i+1})$, $s
+Df(x_{i+1})$, $s D^2f(x_{i+1})$ at $x_{i+1}$ is monotone, and false
+otherwise. As is outlined in Algorithm 3, the binary search is only
+applied at those breakpoints adjoining intervals for which $Q$ is
+nonmonotone and hence $B(1)$ is false. It is further assumed that
+there exists $0 \le s^* \le 1$ such that $B(s)$ is true for $0 \le s
+\le s^*$ and false for $s > s^*$. Since the derivative conditions at
+interior breakpoints are shared by intervals left and right of the
+breakpoint, the binary search is performed at all breakpoints
+simultaneously.  Specifically, the monotonicity of $Q$ is checked on
+all active intervals in each step of the binary search to determine
+the next derivative modification at each breakpoint. The goal of this
+search is to converge on the boundary of the monotone region in the
+$(\tau_1$, $\alpha$, $\beta$, $\gamma)$ space for all intervals. This
+multiple-interval binary search allows the value zero to be obtained
+for all derivative values in a fixed number of computations, hence has
+no effect on computational complexity. This binary search algorithm is
+outlined below.
+
+%% ===================================================================
+%%              Algorithm 4: Global binary search
+\vskip 5mm {\parindent 0mm {\bf Algorithm 4:} {\tt
+    MULTIPLE\_BINARY\_SEARCH}$\bigl(X(1{:}n), Y(1{:}n), Q \bigr)$
+
+where $(X_i,Y_i) \in \bbb{R}\times\bbb{R}$, $i = 1$, $\ldots$, $n$ are
+data points, and $Q(x)$ is a quintic spline interpolant such that
+$Q(X_i) = Y_i$. Modifies derivative values of $Q$ at data points to
+ensure {\tt IS\_MONOTONE} is true for all intervals defined by
+adjacent data points, given a desired precision $\mu \in {\bbb R}$.
+
+}
+{\parindent=6mm
+\item{} Initialize the step size $s$, make a copy of $Q$, and construct
+  three necessary queues for proceeding with the multiple-interval binary search.
+\item{} $s \: 1$
+\item{} $\hat Q \: Q$
+\item{} {\tt searching} $\:$ {\tt TRUE}
+\item{} {\tt checking} $\:$ empty queue for holding left indices of {\it intervals}
+\item{} {\tt growing} $\:$ empty queue for holding indices of {\it data points}
+\item{} {\tt shrinking} $\:$ empty queue for holding indices of {\it data points}
+\item{} {\tt do} $i = 1$, $n-1$
+\itemitem{} {\tt if not IS\_MONOTONE}$\bigl( X_i$, $X_{i+1}$, $Q \bigr)$ {\tt then}
+\itemitem{} \codent Add interval starting at $i$ to {\tt shrinking}.
+\itemitem{} {\tt endif}
+\item{} {\tt enddo}
+\item{} {\tt do while} $\bigl(${\tt searching or shrinking} is nonempty$\bigr)$
+\itemitem{} Compute the {\it step size} for this iteration of the search.
+\itemitem{} {\tt if  searching  then;} $s \: \max\{\mu, s / 2\}$
+\itemitem{} {\tt else;} \hskip 24mm $s \: 3s / 2$
+\itemitem{} {\tt endif}
+\itemitem{} {\tt if} $(s = \mu)$ {\tt then; searching} $\:$ {\tt FALSE; endif}
+\itemitem{} Increase $DQ$ and $D^2Q$ at all data points in {\tt growing}
+  that are strictly adjoining intervals for which $Q$ is monotone.
+\itemitem{} {\tt do} $(i \in$ {\tt growing}$)$ {\tt and} $(i \not \in$ {\tt shrinking}$)$
+\itemitem{} \codent $DQ(X_i) \: DQ(X_i) + s D\hat Q(X_i)$
+\itemitem{} \codent $D^2Q(X_i) \: D^2Q(X_i) + s D^2\hat Q(X_i)$
+\itemitem{} \codent Add intervals starting at $i-1$ and $i$ to {\tt checking}.
+\itemitem{} {\tt enddo}
+\itemitem{} Decrease $DQ$ and $D^2Q$ at all data points in {\tt shrinking} and
+  ensure those data point indices are placed into {\tt growing}.
+\itemitem{} {\tt do} $i \in$ {\tt shrinking}
+\itemitem{} \codent Add $i$ to {\tt growing} if it is not already present.
+\itemitem{} \codent $DQ(X_i) \: DQ(X_i) - s D\hat Q(X_i)$
+\itemitem{} \codent $D^2Q(X_i) \: D^2Q(X_i) - s D^2\hat Q(X_i)$
+\itemitem{} \codent Add intervals starting at $i-1$ and $i$ to {\tt checking}.
+\itemitem{} {\tt enddo}
+\itemitem{} {Empty {\tt shrinking} queue then check all intervals in
+  {\tt checking} for monotonicity with {\tt IS\_MONOTONE}, placing
+  data points adjacent to intervals for which $Q$ is nonmonotone into
+  {\tt shrinking}.}
+\itemitem{} {\tt do} $i \in$ {\tt checking}
+\itemitem{} \codent {\tt if not IS\_MONOTONE}$\bigl( X_i$, $X_{i+1}$, $Q \bigr)$ {\tt then}
+\itemitem{} \codent \codent Add $i$ and $i+1$ to {\tt shrinking}.
+\itemitem{} \codent {\tt endif}
+\itemitem{} {\tt enddo}
+\item{} {\tt enddo }
 }
@@
 
-Given the minimum magnitude curvature nature of the initial derivative
-estimates, it is desirable to make the smallest necessary changes to
-the initial interpolating spline $Q$ while enforcing monotonicity. In
-practice a quasi-bisection search is used in place of solely shrinking
-$DQ$ and $D^2Q$, adding an additional step that increases previously
-shrunken derivative values by a smaller increment than the previous
-decrement. The goal is to converge on the boundary of the monotone region
-in the $(\tau_1$, $\alpha$, $\beta$, $\gamma)$ space. As long as a step size
-schedule is set that allows the value zero to be obtained in a fixed
-number of computations, this has no effect on computational
-complexity. Notably, since {\tt IS\_MONOTONE} handles both nondecreasing and
-nonincreasing simultaneously by taking into account the sign of $v$,
-Algorithm 3 produces $Q(x)$ that is monotone increasing (decreasing) over
-exactly the same intervals that the data $(X_i$, $Y_i)$ is monotone
-increasing (decreasing).
+%% \itemitem{} Take search steps either modifying derivative values
+%%   towards monotonicity (zero) or towards the original values over all
+%%   intervals that are either nonmonotone or previously modified.
+
+In the provided package $\mu = 2^{-26}$, which results in a guaranteed
+$26$ search steps for all intervals that are initially nonmonotone. An
+additional $43$ steps could be required to reduce a derivative to zero
+with step size growth rate of $3/2$. This can only happen when $Q$
+becomes nonmonotone on an interval for the first time while the step
+size equals $\mu$, but for which the only viable solution is a
+derivative value of zero. The maximum number of steps is due to the
+fact that $\sum_{i=0}^{42} \mu (3/2)^i > 1$. In total the {\tt
+  MULTIPLE\_BINARY\_SEARCH} search could require $69$ steps.

### 2020-07-20 Research_[2020-05-18]_Piecewise_Monotone_Quintic_Fortran/paper/4-analysis.tex
Updated again and sent to Dr. Watson, already received feedback and about to reincorporate it.

--- 2020-06-23 Research_[2020-05-18]_Piecewise_Monotone_Quintic_Fortran/paper/4-analysis.tex
@@
 
-Both Algorithms 1 and 2 have ${\cal O}(1)$ runtime. Given a fixed
+\topinsert
+\centerline{\epsfxsize=4truein \epsffile{vis/1-sensitivity.eps}}
+{\everymath={\scriptstyle}
+\narrower\noindent\rmVIII Fig.\ 1. A demonstration of the quadratic
+  facet model's sensitivity to small data perturbations. This example is
+  composed of two quadratic functions $f_1(x) = x^2$ over points $\{1$,
+  $2$, $5/2\}$, and $f_2(x) = (x-2)^2 + 6$ over points $\{5/2$, $3$,
+  $4\}$. Notably, $f_1(5/2) = f_2(5/2)$ and $f_1$, $f_2$ have the same
+  curvature. Given the exact five data points seen above, the quadratic
+  facet model produces the slope seen in the solid blue line at $x = 5/2$.
+  However, by subtracting the value of $f_3$ $= \epsilon(x-2)^2$ from
+  points at $x$ $= \{3$, $4\}$, where $\epsilon$ is the machine precision
+  ($2^{-52}$ for an IEEE 64-bit real), the quadratic facet model produces
+  the slope seen in the dashed red line at $x = 5/2$. This is the nature
+  of a facet model and a side effect of associating data with local facets.
+\par} \endinsert
+
+Algorithms 1, 2, and 4 have ${\cal O}(1)$ runtime. Given a fixed
 schedule for shrinking derivative values, Algorithm 3 has a ${\cal
   O}(n)$ runtime for $n$ data points. In execution, the majority of
-time is spent solving the banded linear system of equations for
-B-spline coefficients. The quadratic facet model produces a unique
-sensitivity to input perturbation, as small changes in input may cause
-data points to be associated with different local quadratic
-interpolants. While this makes the technique susceptible to random
-perturbations, it generally produces high quality monotone quintic
-splines and is preferred for that reason.
+the time, still ${\cal O}(n)$, is spent solving the banded linear
+system of equations for the B-spline coefficients. Thus for $n$ data
+points, the overall execution time is ${\cal O}(n)$. The quadratic
+facet model produces a unique sensitivity to input perturbation, as
+small changes in input may cause different quadratic facets to be
+associated with a breakpoint, and thus different initial derivative
+estimates can be produced. This phenomenon is depicted in Figure
+1. Despite this sensitivity, the quadratic facet model is still
+preferred because it generally provides aesthetically pleasing (low
+{\it wiggle}) initial estimates of the first and second derivatives at
+the breakpoints while perfectly capturing local low-order phenomena
+(e.g., a run of points on a straight line) in the data. The binary
+search for a point near the monotone boundary in $(\tau_1$, $\alpha$,
+$\beta$, $\gamma)$ space is preferred because it results in monotone
+quintic spline interpolants that are nearer to initial estimates than
+a policy that strictly shrinks derivative values.

### 2020-07-20 Research_[2020-05-18]_Piecewise_Monotone_Quintic_Fortran/paper/5-applications.tex
Updated again and sent to Dr. Watson, already received feedback and about to reincorporate it.

@@
 
-This section contains graphs of sample {\tt MQSI} results given
-various sources of data. Tables of computation times for various
-problems sizes are also provided.
+This section contains graphs of sample {\tt MQSI} results given various
+data configurations. Computation times for various problem sizes are also
+provided. The files accompanying the subroutine {\tt MQSI} offer multiple
+usages, namely {\tt sample\_main.f90} that demonstrates Fortran 2003 usage
+and a command line interface {\tt cli.f90} that produces {\tt MQSI} estimates
+for points in batches from data files.  Compilation instructions and the
+full package contents are specified in the {\tt README} file.
 
-\midinsert
+\topinsert
 \centerline{\epsfxsize=4truein \epsffile{vis/2-piecewise-polynomial.eps}}
@@
 \par}
+\vskip 6mm
+\centerline{\epsfxsize=4truein \epsffile{vis/5-real-data.eps}}
+{\narrower\noindent\rmVIII Fig.\ 3.
+{\ttVIII MQSI} results when approximating the cumulative distribution
+function of system throughput (bytes per second) data for a computer
+with a 3.2 GHz CPU performing file read operations from Cameron et al.
+[2019]. The empirical distribution of 30 thousand throughput values is
+shown in the red dashed line, while the solid line with stylized
+markers denotes the approximation made with MQSI given equally spaced
+empirical distribution points from a sample of size 100.
+\par}
 \endinsert
 
-\midinsert
-\centerline{\epsfxsize=4truein \epsffile{vis/5-random-monotone.eps}}
-{\narrower\noindent\rmVIII Fig.\ 3.
+\topinsert
+\centerline{\epsfxsize=4truein \epsffile{vis/6-random-monotone.eps}}
+{\narrower\noindent\rmVIII  Fig.\ 4.
 The {\itVIII random monotone} test poses a particularly challenging
 problem with large variations in slope. Notice that despite drastic
-shifts in slope, the resulting {\ttVIII MQSI} provides smooth and
-reasonable estimates to function values between data.
+shifts in slope, the resulting monotone quintic spline interpolant
+provides smooth and reasonable estimates to function values between data.
+\par}
+\vskip 15mm
+\centerline{\epsfxsize=4truein \epsffile{vis/7-runtimes-10.eps}}
+\centerline{\epsfxsize=4truein \epsffile{vis/8-runtimes-100.eps}}
+{\everymath={\scriptstyle} \narrower\noindent\rmVIII Fig.\ 5.
+Median total runtime of the routine {\ttVIII MQSI} for number of data points
+$n=10$
+to 100 (top) and $n=100$ to 1000 (bottom), generated from 100
+repeated trials averaged over $14$ different testing functions. The
+timings are colored by major algorithmic component, but the vast
+majority of execution time is spent solving the banded linear system
+of equations in the {\ttVIII FIT\_SPLINE} routine. The runtimes for
+the quadratic facet (Algorithm 1) take roughly one microsecond
+($10^{-6}$ seconds) per breakpoint, while the binary search
+(Algorithm 4) takes roughly four microseconds per breakpoint.
 \par}
 \endinsert
+
+Throughout, all visuals have points that are stylized by local
+monotonicity conditions. Blue circles denote extreme points, purple
+squares are in {\it flat} regions with no change in function value,
+red down triangles are monotone decreasing, and green up triangles are
+monotone increasing.
+
+Figure 2 offers examples of the interpolating splines produced by the
+routine {\tt MQSI} on various hand-crafted sets of data. These same
+data sets are used for testing local installations in the provided
+program {\tt test\_all.f90}. Notice that the quadratic facet model
+perfectly captures the local linear segments of data in the piecewise
+polynomial test for Figure 2. Figure 3 depicts an approximation of a
+cumulative distribution function made by {\tt MQSI} on a computer
+systems application by Cameron et al. [2019] that studies the
+distribution of throughput (in bytes per second) when reading files
+from a storage medium. Figure 4 provides a particularly difficult
+monotone interpolation challenge using randomly generated monotone
+data.
+
+Finally, Figure 5 provides execution times on a computer running MacOS
+10.15.5 with a 2 GHz Intel Core I5 CPU. The total time is
+broken down by the major stages of {\tt MQSI}. Most of the computation
+time is spent solving the banded linear system for B-spline
+coefficients in {\tt FIT\_SPLINE}.
+
+% For large problems ($n > 100$) it would be faster
+% to construct splines over intervals independently (each interval
+% requiring a $6 \times 6$ linear system to be solved), however the
+% singular linear system is chosen here for the decreased redundancy
+% in the spline description.

### 2020-07-20 Research_[2020-05-18]_Piecewise_Monotone_Quintic_Fortran/paper/mqsi20.tex
Updated again and sent to Dr. Watson, already received feedback and about to reincorporate it.

@@
 
+% deboor1978practical
+\refb{de Boor, C.}{1978}{A Practical Guide to Splines}{Springer, Verlag,
+New York}
+
+% cameron2019moana
+\refj{Cameron, K.
+W., Anwar, A., Cheng, Y., Xu, L., Li, B., Ananth, U., Bernard, J., Jearls, C., Lux,  T.
+C.
+H., Hong, Y., Watson, L.
+T., and Butt, A.
+R.}{2019}{MOANA: Modeling and Analyzing I/O Variability in Parallel System Experimental Design}{IEEE Transactions on Parallel and Distributed Systems}{30(8):1843--1856}
+
 % fritsch1980monotone
 \refj{Fritsch., F and Carlson., R.}{1980}{Monotone piecewise cubic
-  interpolation}{SIAM Journal on Numerical Analysis}{17(2):238–-246}
+  interpolation}{SIAM Journal on Numerical Analysis}{17(2):238--246}
 

### 2020-07-20 Research_[2020-05-18]_Piecewise_Monotone_Quintic_Fortran/fortran/demo/run.sh

@@ -2,8 +2,8 @@
 
 SRC_DIR=".."
 CUR_DIR="$(pwd)"
-# Name the executable 'cli' for "command line interface".
-EXEC="cli"
+# Name the command line interface executable 'mqsi'.
+EXEC="mqsi"
 
 # Declare the variables for compilation.
 F03=gfortran

### 2020-07-23 Research_[2020-05-18]_Piecewise_Monotone_Quintic_Fortran/paper/2-monotone-quintic.tex
Made modifications including Dr. Watsons recent revisions. Working towards a submission version.

--- 2020-06-23 Research_[2020-05-18]_Piecewise_Monotone_Quintic_Fortran/paper/2-monotone-quintic.tex
@@
 
-where $X_j$, $Y_j$ $\in \bbb{R}$ for $j = 1$, $\ldots$, $n$, and $1 < i
-< n$. Returns the slope and curvature at $X_i$ of the local quadratic
-interpolant with minimum magnitude curvature.
+where $X_j$, $Y_j$ $\in \bbb{R}$ for $j = 1$, $\ldots$, $n$, $1 \le i
+\le n$, and $n \ge 3$. Returns the slope and curvature at $X_i$ of the
+local quadratic interpolant with minimum magnitude curvature.
 
@@
 %% -------------------------------------------------------------------
-\item{} {\tt if} $\bigl((Y_i \approx Y_{i-1})$ {\tt or} $(Y_i \approx
-  Y_{i+1})\bigr)$ {\tt then; return $(0,0)$}
-\item{} {\tt else if} $\bigl((Y_{i+1} - Y_i)(Y_i - Y_{i-1}) < 0\bigr)$
-{\tt then}
+\item{} {\tt if} $\bigl((i\ne1\wedge Y_i \approx Y_{i-1})$ {\tt or}
+$(i\ne n\wedge Y_i \approx Y_{i+1})\bigr)$ {\tt then return $(0,0)$}
+
+\item{} {\tt else if} $i=1$ {\tt then}
+\itemitem{} $f_1\:\hbox{interpolant to }(X_1,Y_1)$,
+  $(X_2,Y_2)$, and $(X_3,Y_3)$.
+\itemitem{} {\tt if} $\bigl(Df_1(X_1)(Y_2-Y_1)<0\bigr)$ {\tt then return}
+  $\bigl(0,0\bigr)$
+\itemitem{} {\tt else return} $\bigl(Df_1(X_1),D^2f_1\bigr)$
+\itemitem{} {\tt endif}
+
+\item{} {\tt else if} $i=n$ {\tt then}
+\itemitem{} $f_1\:\hbox{interpolant to }(X_{n-2},Y_{n-2})$,
+  $(X_{n-1},Y_{n-1})$, and $(X_n,Y_n)$.
+\itemitem{} {\tt if} $\bigl(Df_1(X_n)(Y_n-Y_{n-1})<0\bigr)$ {\tt then return}
+  $\bigl(0,0\bigr)$
+\itemitem{} {\tt else return} $\bigl(Df_1(X_n),D^2f_1\bigr)$
+\itemitem{} {\tt endif}
+
+\item{} {\tt else if} $\bigl(1<i<n\wedge (Y_{i+1} - Y_i)(Y_i - Y_{i-1}) < 0
+\bigr)$ {\tt then}
 \itemitem{} The point $(X_i$, $Y_i)$ is an extreme point. The
@@
   and $Df_2(X_i) = 0$.
-\itemitem{} {\tt if} $\bigl(|D^2f_1| \leq |D^2f_2|\bigr)$ {\tt then;
-  return} $\bigl(Df_1$, $D^2f_1\bigr)$
-\itemitem{} {\tt else;} \hskip 34.25mm {\tt return} $\bigl(Df_2$,
-$D^2f_2\bigr)$
+\itemitem{} {\tt if} $\bigl(|D^2f_1| \leq |D^2f_2|\bigr)$ {\tt then
+  return} $\bigl(0$, $D^2f_1\bigr)$
+\itemitem{} {\tt else return} $\bigl(0$, $D^2f_2\bigr)$
 \itemitem{} {\tt endif}
+
 \item{} {\tt else}
 \itemitem{} {The point $(X_i$, $Y_i)$ is in a monotone segment of
-data. In the following, it is possible that either $f_1$ or $f_3$ does
-not exist because $i = 2$ or $i = n-1$. In those cases, the minimum
+data. In the following, it is possible that $f_1$ or $f_3$ do
+not exist because $i \in \{2, n-1\}$. In those cases, the minimum
 magnitude curvature among existing quadratics is chosen.}
@@
   $(X_{i+1},Y_{i+1})$, and $(X_{i+2},Y_{i+2})$.
-\itemitem{} {\tt if} \hskip 9mm $\bigl(|D^2f_1| = \min\bigl\{ |D^2f_1|,
- |D^2f_2|, |D^2f_3|\bigr\} \bigr)$ {\tt then; return}
- $\bigl(Df_1$, $D^2f_1\bigr)$
-\itemitem{} {\tt else if} $\bigl(|D^2f_2| = \min\bigl\{ |D^2f_1|,
- |D^2f_2|, |D^2f_3|\bigr\} \bigr)$ {\tt then; return}
- $\bigl(Df_2$, $D^2f_2\bigr)$
-\itemitem{} {\tt else;} \codent {\tt return} $\bigl(Df_3$, $D^2f_3\bigr)$
-\item{} \codent {\tt endif}
+
+\itemitem{} {\tt if} $\bigl(Df_1(X_i)(Y_i-Y_{i-1}) \ge 0 \wedge
+  |D^2f_1| = \min\bigl\{ |D^2f_1|,
+  |D^2f_2|, |D^2f_3|\bigr\} \bigr)$ {\tt then}
+\itemitem{} \codent {\tt return} $\bigl(Df_1(X_i),D^2f_1\bigr)$
+
+\itemitem{} {\tt else if} $\bigl(Df_2(X_i)(Y_i-Y_{i-1}) \ge 0 \wedge
+  |D^2f_2| = \min\bigl\{ |D^2f_1|,
+  |D^2f_2|, |D^2f_3|\bigr\} \bigr)$ {\tt then}
+\itemitem{} \codent {\tt return} $\bigl(Df_2(X_i),D^2f_2\bigr)$
+
+\itemitem{} {\tt else if} $\bigl(Df_3(X_i)(Y_{i+1}-Y_i) \ge 0\bigr)$ {\tt then}
+\itemitem{} \codent {\tt return} $\bigl(Df_3(X_i),D^2f_3\bigr)$
+\itemitem{} {\tt else return} $(0, 0)$
+\itemitem{} {\tt endif}
 \item{} {\tt endif}
@@
 {\parindent 0mm
-{\bf Algorithm 2:} {\tt IS\_MONOTONE}$\bigl(x_0, x_1, f \bigr)$
+{\bf Algorithm 2:}
+
+{\tt IS\_MONOTONE}$\bigl(x_0$, $x_1$, $f(x_0)$, $Df(x_0)$,
+$D^2f(x_0)$, $f(x_1)$, $Df(x_1)$, $D^2f(x_1) \bigr)$
 
@@
 \item{4.} {\tt if} $\bigl(Df(x_0) < 0$ {\tt or} $Df(x_1) < 0\bigr)$ {\tt
-  then; return FALSE; endif}
-\item{5.} $w := x_1 - x_0$
-\item{6.} $v := f(x_1) - f(x_0)$
+  then return FALSE endif}
+\item{5.} $w \: x_1 - x_0$
+\item{6.} $z \: f(x_1) - f(x_0)$
 
-\item{} {The necessity of Steps 2 and 4 follows directly from
-  the fact that $f$ is $C^2$. The following Steps 7--13 coincide with
-  a simplified condition for quintic monotonicity that reduces to one
-  of cubic positivity studied by Schmidt and He{\ss} [1988]. Given
+\item{} {The necessity of Steps 1--4 follows directly from the
+  fact that $f$ is $C^2$. The following Steps 7--13 coincide with a
+  simplified condition for quintic monotonicity that reduces to one of
+  cubic positivity studied by Schmidt and He{\ss} [1988]. Given
   $\alpha$, $\beta$, $\gamma$, and $\delta$ as defined by Schmidt and
@@
 0\bigr)$ {\tt then}
-\item{8.} \codent {\tt if} $\bigl(D^2f(x_1)w > 4Df(x_1)$ {\tt then;
-return FALSE; endif}
-\item{9.} \codent $t := 2 \sqrt{Df(x_0) (4Df(x_1) - D^2f(x_1) w) }$
+\item{8.} \codent {\tt if} $\bigl(D^2f(x_1)w > 4Df(x_1)$ {\tt then
+return FALSE endif}
+\item{9.} \codent $t \: 2 \sqrt{Df(x_0) (4Df(x_1) - D^2f(x_1) w) }$
 \item{10.} \codent {\tt if} $\bigl(t + 3Df(x_0) + D^2f(x_0)w < 0 \bigr)$
-  {\tt then; return FALSE; endif}
-\item{11.} \codent {\tt if} $\bigl(60v - w\bigl(24Df(x_0) + 32Df(x_1) - 2t
+  {\tt then return FALSE endif}
+\item{11.} \codent {\tt if} $\bigl(60z - w\bigl(24Df(x_0) + 32Df(x_1) - 2t
   + w(3D^2f(x_0) - 5D^2f(x_1))\bigr) < 0\bigr)$
-\item{}   \codent \codent {\tt then; return FALSE; endif}
+\item{}   \codent \codent {\tt then return FALSE endif}
 \item{12.} \codent {\tt return TRUE}
@@
 \item{14.} {\tt if} $\bigl( w\bigl(2\sqrt{Df(x_0)\,Df(x_1)} - 3(Df(x_0) +
-  Df(x_1))\bigr) - 24v \leq 0 \bigr)$
-\item{} \codent {\tt then; return FALSE; endif}
-\item{15.} $t := \bigl(Df(x_0)\, Df(x_1)\bigr)^{3/4}$
-\item{16.} $\alpha := (4 Df(x_1) - D^2f(x_1)w) \sqrt{Df(x_0)} / t$
-\item{17.} $\gamma := (4 Df(x_0) - D^2f(x_0)w) \sqrt{Df(x_1)} / t$
-\item{18.} $\displaystyle \beta := {60v/w + 3\bigl(w(D^2f(x_1) -
+  Df(x_1))\bigr) - 24z \leq 0 \bigr)$
+\item{} \codent {\tt then return FALSE endif}
+\item{15.} $t \: \bigl(Df(x_0)\, Df(x_1)\bigr)^{3/4}$
+\item{16.} $\alpha \: (4 Df(x_1) - D^2f(x_1)w) \sqrt{Df(x_0)} / t$
+\item{17.} $\gamma \: (4 Df(x_0) - D^2f(x_0)w) \sqrt{Df(x_1)} / t$
+\item{18.} $\displaystyle \beta \: {60z/w + 3\bigl(w(D^2f(x_1) -
   D^2f(x_0)) - 8(Df(x_0) + Df(x_1))\bigr) \over 2 \sqrt{Df(x_0)\,Df(x_1)}}$
-\item{19.} {\tt if} $(\beta \leq 6)$ {\tt then; return}
+\item{19.} {\tt if} $(\beta \leq 6)$ {\tt then return}
 $\bigl( \min\{\alpha,\gamma\} > - (\beta + 2) / 2 \bigr)$
-\item{20.} {\tt else;} \hskip 18mm {\tt return}
+\item{20.} {\tt else} {\tt return}
 $\bigl( \min\{\alpha,\gamma\} > -2 \sqrt{\beta - 2}\,\bigr)$
@@
 {\parindent=6mm
-\item{} Approximate first and second derivatives at all $X_i$ with
-{\tt QUADRATIC\_FACET}.
-\item{} {\tt do} $i = 1, \ldots n-1$
-\item{} \codent $\bigl(D Q(X_i)$, $D^2 Q(X_i)\bigr):=$
-{\tt QUADRATIC\_FACET}$(X$, $Y$, $i)$
+\item{} Approximate first and second derivatives at $X_i$ with {\tt
+  QUADRATIC\_FACET}.
+\item{} {\tt for} $i\:1$ {\tt step} 1 {\tt until} $n$ {\tt do}
+\item{} \codent $(u_i$, $v_i)\:$ {\tt QUADRATIC\_FACET}$(X$, $Y$, $i)$
 \item{} {\tt enddo}
-\item{} Identify and store all nonmonotone intervals in a {\tt queue}.
-\item{} {\tt do} $i = 1, \ldots n-1$
-\item{} \codent {\tt if not IS\_MONOTONE}$(X_i$, $X_{i+1}$, $Q)$ {\tt then}
+\item{} Identify and store all intervals where $Q$ is nonmonotone in a
+queue $\cal Q$.
+\item{} {\tt for} $i\:1$ {\tt step} 1 {\tt until} $n-1$ {\tt do}
+\item{} \codent {\tt if not IS\_MONOTONE}$(X_i$, $X_{i+1}$, $Y_i$, $u_i$,
+$v_i$, $Y_{i+1}$, $u_{i+1}$, $v_{i+1})$ {\tt then}
 \item{} \codent\codent Add interval $\bigl(X_i$, $X_{i+1}\bigr)$ to
-{\tt queue}.
+queue $\cal Q$.
 \item{} \codent {\tt endif}
 \item{} {\tt enddo}
-\item{} {\tt do while} $\bigl($ {\tt queue} of nonmonotone intervals is
-nonempty $\bigr)$
-\itemitem{} Shrink (in magnitude) $DQ$ and $D^2Q$ that border nonmonotone
-intervals.
-\itemitem{} Identify and store all remaining nonmonotone intervals in
-{\tt queue}.
+\item{} {\tt do while} $\bigl($ queue $\cal Q$ of intervals is nonempty $\bigr)$
+\itemitem{} Shrink (in magnitude) $DQ$ (in $u$) and $D^2Q$ (in $v$)
+  that border intervals where $Q$ is nonmonotone.  
+\itemitem{} Identify and store remaining intervals where $Q$ is
+  nonmonotone in queue $\cal Q$.
 \item{} {\tt enddo}
+\item{} Construct and return a $B$-spline representation of $Q(x)$.
+}
+\vskip 3mm
+%% ----------------------------------------------------------------------
+
+Since {\tt IS\_MONOTONE} can handle both nondecreasing and
+nonincreasing simultaneously by taking into account the sign of $z$,
+Algorithm 3 produces $Q(x)$ that is monotone increasing (decreasing)
+over exactly the same intervals that the data $(X_i$, $Y_i)$ is
+monotone increasing (decreasing).
+
+Given the minimum magnitude curvature nature of the initial derivative
+estimates, it is desirable to make the smallest necessary changes to
+the initial interpolating spline $Q$ while enforcing monotonicity. In
+practice a binary search for the boundary of monotonicity is used in
+place of solely shrinking $DQ$ and $D^2Q$ at breakpoints adjoining
+{\it active\/} intervals, or intervals over which $Q$ is nonmonotone at
+least once during the search. The binary search considers a Boolean
+function $B_i(s)$, for $0 \le s \le 1$, that is true if the order six
+polynomial $f(x)$ on $[X_i, X_{i+1}]$ matching derivatives
+$f(X_i)=Y_i$, $Df(X_i)=s \,u_i$, $D^2f(X_i)=s \,v_i$ at $X_i$, and derivatives
+$f(X_{i+1})=Y_{i+1}$, $Df(X_{i+1})=s \,u_{i+1}$, $D^2f(X_{i+1})=s \,v_{i+1}$
+at $X_{i+1}$ is monotone, and false otherwise. As is outlined in Algorithm
+3, the binary search is only applied at those breakpoints adjoining intervals
+$[X_i, X_{i+1}]$ over which $Q$ is nonmonotone and hence $B_i(1)$ is false.
+It is further assumed that there exists $0 \le s^* \le 1$ such that $B_i(s)$
+is true for $0 \le s \le s^*$ and false for some $1 > s > s^*$. Since the
+derivative conditions at interior breakpoints are shared by intervals left
+and right of the breakpoint, the binary search is performed at all
+breakpoints simultaneously.  Specifically, the monotonicity of $Q$ is
+checked on all active intervals in each step of the binary search to
+determine the next derivative modification at each breakpoint. The goal
+of this search is to converge on the boundary of the monotone region in
+the $(\tau_1$, $\alpha$, $\beta$, $\gamma)$ space (described in Ulrich and
+Watson [1994]) for all intervals. This multiple-interval binary search
+allows the value zero to be obtained for all (first and second) derivative
+values in a fixed number of computations, hence has no effect on
+computational complexity order. This binary search algorithm is outlined
+below.
+
+%% ===================================================================
+%%              Algorithm 4: Global binary search
+\vskip 5mm {\parindent 0mm {\bf Algorithm 4:} {\tt
+    BINARY\_SEARCH}$\bigl(X(1{:}n), Y(1{:}n), u(1{:}n), v(1{:}n) \bigr)$
+
+where $(X_i,Y_i) \in \bbb{R}\times\bbb{R}$, $i = 1$, $\ldots$, $n$ are
+data points, and $Q(x)$ is a quintic spline interpolant such that
+$Q(X_i) = Y_i$, $DQ(X_i) = u_i$, $D^2Q(X_i) = v_i$. Modifies
+derivative values ($u$ and $v$) of $Q$ at data points to ensure
+{\tt IS\_MONOTONE} is true for all intervals defined by adjacent data
+points, given a desired precision $\mu \in {\bbb R}$. proj$\bigl(w$,
+int$(a,b)\bigr)$ denotes the projection of $w$ onto the closed
+interval with endpoints $a$ and $b$.
+
+}
+{\parindent=6mm
+\item{} Initialize the step size $s$, make a copy of data defining $Q$,
+and construct three queues necessary for the multiple-interval binary search.
+\item{} $s \: 1$
+\item{} $(\hat u,\hat v) \: (u,v)$
+\item{} {\tt searching} $\:$ {\tt TRUE}
+\item{} {\tt checking} $\:$ empty queue for holding left indices of {\it intervals}
+\item{} {\tt growing} $\:$ empty queue for holding indices of {\it data points}
+\item{} {\tt shrinking} $\:$ empty queue for holding indices of {\it data points}
+\item{} {\tt for} $i\:1$ {\tt step} 1 {\tt until} $n-1$ {\tt do}
+\itemitem{} {\tt if not IS\_MONOTONE}$\bigl( X_i$, $X_{i+1}$, $Y_i$, $u_i$,
+$v_i$, $Y_{i+1}$, $u_{i+1}$, $v_{i+1} \bigr)$ {\tt then}
+\itemitem{} \codent Add data indices $i$ and $i+1$ to queue {\tt shrinking}.
+\itemitem{} {\tt endif}
+\item{} {\tt enddo}
+\item{} {\tt do while} $\bigl(${\tt searching or} ({\tt shrinking} is
+nonempty)$\bigr)$
+\itemitem{} Compute the {\it step size} $s$ for this iteration of the search.
+\itemitem{} {\tt if  searching  then} $s \: \max\{\mu, s/2\}$
+{\tt else} $s \: 3s/2$ {\tt endif}
+\itemitem{} {\tt if} $(s=\mu)$ {\tt then searching} $\:$ {\tt FALSE};
+  clear queue {\tt growing endif}
+\itemitem{} Increase in magnitude $u_i$ and $v_i$ for all data indices
+  $i$ in {\tt growing} such that the points $X_i$ are strictly adjoining
+  intervals over which $Q$ is monotone.
+\itemitem{} {\tt for} ($i \in$ {\tt growing}) {\tt and} ($i \not\in$
+{\tt shrinking}) {\tt do}
+\itemitem{} \codent $u_i \: \hbox{proj}\bigl(u_i + s\,\hat u_i, \hbox{int}
+  (0,\hat u_i)\bigr)$
+\itemitem{} \codent $v_i \: \hbox{proj}\bigl(v_i + s\,\hat v_i, \hbox{int}
+  (0,\hat v_i)\bigr)$
+\itemitem{} \codent Add data indices $i-1$ (if not 0) and $i$ (if not $n$) to queue
+  {\tt checking}.
+\itemitem{} {\tt enddo}
+\itemitem{} Decrease in magnitude $u_i$ and $v_i$ for all data indices
+  $i$ in {\tt shrinking} and ensure those data point indices are placed
+  into {\tt growing} when {\tt searching}.
+\itemitem{} {\tt for} $i \in$ {\tt shrinking do}
+\itemitem{} \codent If {\tt searching}, then add index $i$ to queue
+  {\tt growing} if not already present.
+\itemitem{} \codent $u_i \: \hbox{proj}\bigl(u_i - s\,\hat u_i, \hbox{int}
+  (0,\hat u_i)\bigr)$
+\itemitem{} \codent $v_i \: \hbox{proj}\bigl(v_i - s\,\hat v_i, \hbox{int}
+  (0,\hat v_i)\bigr)$
+\itemitem{} \codent Add data indices $i-1$ (if not 0) and $i$ (if not $n$) to queue
+  {\tt checking}.
+\itemitem{} {\tt enddo}
+\itemitem{} Empty queue {\tt shrinking}, then check all intervals left-indexed
+  in queue {\tt checking} for monotonicity with {\tt IS\_MONOTONE},
+  placing data endpoint indices of intervals over which $Q$ is nonmonotone
+  into queue {\tt shrinking}.
+\itemitem{} Clear queue {\tt shrinking}.
+\itemitem{} {\tt for} $i \in$ {\tt checking do}
+\itemitem{} \codent
+  {\tt if not IS\_MONOTONE}$\bigl( X_i$, $X_{i+1}$, $Y_i$, $u_i$,
+  $v_i$, $Y_{i+1}$, $u_{i+1}$, $v_{i+1} \bigr)$ {\tt then}
+\itemitem{} \codent \codent Add data indices $i$ and $i+1$ to {\tt shrinking}.
+\itemitem{} \codent {\tt endif}
+\itemitem{} {\tt enddo}
+\itemitem{} Clear queue {\tt checking}.
+\item{} {\tt enddo }
 }
@@
 
-Given the minimum magnitude curvature nature of the initial derivative
-estimates, it is desirable to make the smallest necessary changes to
-the initial interpolating spline $Q$ while enforcing monotonicity. In
-practice a quasi-bisection search is used in place of solely shrinking
-$DQ$ and $D^2Q$, adding an additional step that increases previously
-shrunken derivative values by a smaller increment than the previous
-decrement. The goal is to converge on the boundary of the monotone region
-in the $(\tau_1$, $\alpha$, $\beta$, $\gamma)$ space. As long as a step size
-schedule is set that allows the value zero to be obtained in a fixed
-number of computations, this has no effect on computational
-complexity. Notably, since {\tt IS\_MONOTONE} handles both nondecreasing and
-nonincreasing simultaneously by taking into account the sign of $v$,
-Algorithm 3 produces $Q(x)$ that is monotone increasing (decreasing) over
-exactly the same intervals that the data $(X_i$, $Y_i)$ is monotone
-increasing (decreasing).
+%% \itemitem{} Take search steps either modifying derivative values
+%%   towards monotonicity (zero) or towards the original values over all
+%%   intervals that are either nonmonotone or previously modified.
+
+In the subroutine {\tt MQSI}, $\mu = 2^{-26}$, which results in 26
+guaranteed search steps for all intervals that are initially
+nonmonotone. An additional 43 steps could be required to reduce a
+derivative magnitude to zero with step size growth rate of $3/2$. This
+can only happen when $Q$ becomes nonmonotone on an interval for the
+first time while the step size equals $\mu$, but for which the only
+viable solution is a derivative value of zero. The maximum number of
+steps is due to the fact that $\sum_{i=0}^{42} \mu (3/2)^i > 1$. In
+total {\tt BINARY\_SEARCH} search could require 69 steps.

### 2020-07-23 Research_[2020-05-18]_Piecewise_Monotone_Quintic_Fortran/paper/5-applications.tex
Made modifications including Dr. Watsons recent revisions. Working towards a submission version.

@@
 \par}
+\iffalse
 \vskip 15mm
@@
 \par}
+\fi
 \endinsert
@@
 
-Finally, Figure 5 provides execution times on a computer running MacOS
-10.15.5 with a 2 GHz Intel Core I5 CPU. The total time is
-broken down by the major stages of {\tt MQSI}. Most of the computation
-time is spent solving the banded linear system for B-spline
-coefficients in {\tt FIT\_SPLINE}.
-
-% For large problems ($n > 100$) it would be faster
-% to construct splines over intervals independently (each interval
-% requiring a $6 \times 6$ linear system to be solved), however the
-% singular linear system is chosen here for the decreased redundancy
-% in the spline description.
+On a computer running MacOS 10.15.5 with a 2 GHz Intel Core i5 CPU, the
+quadratic facet (Algorithm 1) takes roughly one microsecond ($10^{-6}$
+seconds) per breakpoint, while the binary search (Algorithm 4) takes roughly
+four microseconds per breakpoint; these times were generated from 100
+repeated trials averaged over 14 different testing functions.  The vast
+majority of execution time is spent solving the banded linear system of
+equations in the routine {\tt FIT\_SPLINE} for the $B$-spline coefficients.
+For large problems ($n > 100$) it would be faster to construct splines
+over intervals independently (each interval requiring a $6 \times 6$ linear
+system to be solved), however the single linear system is chosen here for
+the decreased redundancy in the spline description.

### 2020-07-23 Research_[2020-05-18]_Piecewise_Monotone_Quintic_Fortran/paper/mqsi20.tex
Made modifications including Dr. Watsons recent revisions. Working towards a submission version.

@@
 \hrule\bigskip\smallskip \footnote{}{\hskip
--\parindent{\parindent=0pt\rmVIII Authors' addresses:
+-\parindent{\parindent=0pt\rmVIII
+This work was supported by National Science Foundation Grants CNS-1565314,
+CNS-1838271, and DGE-154362.
+Authors' addresses:
 T. C. H. Lux, L. T. Watson, Departments of Computer Science, Mathematics,
@@
 \medskip
-Additional Key Words and Phrases: Quintic Spline, interpolation
+Additional Key Words and Phrases: quintic spline, interpolation
 
@@
 %                            Bibliography
+% \vfil\goodbreak
 \heading{\hlvVIII BIBLIOGRAPHY} {
-\baselineskip=2pt \parskip=100mm
+\baselineskip=2pt%\parskip=100mm
 \frenchspacing
@@
 \def\refj#1#2#3#4#5{\noindent\hangindent=10pt\hangafter=1
-{\caps #1}\ys #2.\ys {\rm #3}. {\sl #4}, #5.\vskip 6pt}
+{\caps #1}\ys #2.\ys {\rm #3}. {\sl #4}, #5.\vskip 6pt minus 3pt}
 
@@
 \def\refb#1#2#3#4{\noindent\hangindent=10pt\hangafter=1
-{\caps #1}\ys #2.\ys{\sl #3}. {\rm #4}.\vskip 6pt}
+{\caps #1}\ys #2.\ys{\sl #3}. {\rm #4}.\vskip 6pt minus 3pt}
 
@@
 \def\reft#1#2#3#4{\noindent\hangindent=10pt\hangafter=1
-{\caps #1}\ys #2.\ys{\rm #3}. {\rm #4}.\vskip 6pt}
+{\caps #1}\ys #2.\ys{\rm #3}. {\rm #4}.\vskip 6pt minus 3pt}
 
@@
 % brennan2019measure
-\refj{Brennan, A.}{2020}{Measure, Modulation and Metadesign: NC
-  Fabrication in Industrial Design and Architecture}{Journal of Design
+\refj{Brennan, A.}{2020}{Measure, modulation and metadesign: NC
+  fabrication in industrial design and architecture}{Journal of Design
   History}{33(1):66--82}
@@
 % cameron2019moana
-\refj{Cameron, K.
-W., Anwar, A., Cheng, Y., Xu, L., Li, B., Ananth, U., Bernard, J., Jearls, C., Lux,  T.
-C.
-H., Hong, Y., Watson, L.
-T., and Butt, A.
-R.}{2019}{MOANA: Modeling and Analyzing I/O Variability in Parallel System Experimental Design}{IEEE Transactions on Parallel and Distributed Systems}{30(8):1843--1856}
+\refj{Cameron, K. W., Anwar, A., Cheng, Y., Xu, L., Li, B., Ananth, U.,
+Bernard, J., Jearls, C., Lux,  T. C. H., Hong, Y., Watson, L. T., and Butt,
+A. R.}{2019}{MOANA: modeling and analyzing I/O variability in parallel
+system experimental design}{IEEE Transactions on Parallel and Distributed
+Systems}{30(8):1843--1856}
 
@@
 \refj{He{\ss}, W. and Schmidt, J. W.}{1994}{Positive quartic, monotone
-  quintic $C^2$-spline interpolation in one and two dimensions}{Journal
-  of Computational and Applied Mathematics}{55(1):51–-67}
+  quintic $\scriptstyle C^2$-spline interpolation in one and two
+  dimensions}{Journal of Computational and Applied Mathematics}{55(1):51–-67}
 
@@
 % yao2018unconditionally
-\refj{Yao, J. and Nelson, K. E.}{2018}{An Unconditionally Monotone
-  $C^2$ Quartic Spline Method with Nonoscillation
-  Derivatives}{Advances in Pure Mathematics}{8(LLNL-JRNL-742107)}
+\refj{Yao, J. and Nelson, K. E.}{2018}{An unconditionally monotone
+  $\scriptstyle C^2$ quartic spline method with nonoscillation
+  derivatives}{Advances in Pure Mathematics}{8(LLNL-JRNL-742107)}
 

### 2020-07-28 3-Disseration/algs.tex
Updated thesis document, sent draft to Dr. Watson.

--- 2019-09-27 Paper_[2019-02]_NUMA/paper.tex
@@
 \end{align*}
-where $j$ is the iteration number, $1 \le p \le n$, and $B_l(x)$,
-$B_k(x)$ are basis functions from the previous iteration,
- $$w_+ = \begin{cases} w, & w \geq 0 \\ 0, & w < 0 \end{cases},$$
-and $w_- = (-w)_+$. After iteratively constructing a model, MARS then
+where $j \leq m$ is the iteration number, $m$ is the maximum
+  number of underlying basis functions, $1 \le p \le n$, and
+$B_l(x)$, $B_k(x)$ are basis functions from the previous iteration,
+ $$w_+ = \begin{cases} w, & w \geq 0 \\ 0, & w < 0 \end{cases},$$ and
+$w_- = (-w)_+$. After iteratively constructing a model, MARS then
 iteratively removes basis functions that do not contribute to goodness
@@
 product approximation of the data. The overall computational
-complexity of Fast MARS is $\mathcal{O}(n d m^3)$ where $m$ is the
-maximum number of underlying basis functions. This paper uses an
-implementation of Fast MARS \cite{rudy2017pyearth} with $m = 200$
-throughout all experiments.
+complexity of Fast MARS is $\mathcal{O}(n d m^3)$. This paper
+uses an implementation of Fast MARS \cite{rudy2017pyearth} with $m =
+200$ throughout all experiments.
 
@@
 with ten hidden layers each having thirty two nodes (a total of
-approximately ten thousand parameters), ReLU activation, and five
-thousand epochs of SGD for training.
+approximately ten thousand parameters), ReLU activation, and
+one hundred thousand steps of SGD
+for training. It should be noted that the choice of neural
+  network architecture affects approximation performance, but no
+  architecture tuning is performed here.
 
@@
 producing an interpolant \cite{lee1980two}. The Delaunay triangulation
-of a set of data points into simplices is such that there are no data
-points inside the sphere defined by the vertices of each simplex. For
-a $d$-simplex S with vertices $x^{(0)}$, $x^{(1)}$, $\ldots$,
-$x^{(d)}$, and function values $f\bigl(x^{(i)}\bigr)$, $i=0$,
-$\ldots$, $d$, $y \in S$ is a unique convex combination of the
-vertices,
+of a set of points into simplices is such that there
+are no points inside the sphere defined by the
+vertices of each simplex. For a $d$-simplex S with vertices $x^{(0)}$,
+$x^{(1)}$, $\ldots$, $x^{(d)}$, and function values
+$f\bigl(x^{(i)}\bigr)$, $i=0$, $\ldots$, $d$, $y \in S$ is a unique
+convex combination of the vertices,
  $$ y = \sum_{i=0}^{d} w_i x^{(i)}, \quad \sum_{i=0}^{d} w_i = 1, \quad w_i \geq 0, \quad i=0,\ldots,d, $$
@@
 
+\begin{figure}
+  \centering
+  \includegraphics[width=0.4\textwidth,]{Figures/NA/example_delaunay.pdf}
+  \includegraphics[width=0.4\textwidth,]{Figures/NA/example_not_delaunay.pdf}
+  \caption{On the left above is a depiction of a Delaunay
+    triangulation over four points, notice that the circumball (shaded
+    circle) for the left simplex does not contain the fourth point.
+    On the right above, a non-Delaunay mesh is depicted. Notice that
+    the circumball for the top simplex (shaded circle, clipped at
+    bottom edge of the visual) contains the fourth point which
+    violates the Delaunay condition for a simplex.
+  \vspace{-.1cm}}
+  \label{fig:delaunay}
+\end{figure}
+
 \subsection{Modified Shepard}
@@
  $$ p(x) = \frac{\sum\limits_{k=1}^{n}W_k(x)f\bigl(x^{(k)}\bigr)}
-     {\sum\limits_{k=1}^{n}W_k(x)} ,$$
-where $W_k(x) = \bigl(r_k - \bigl\|x - x^{(k)}\bigr\|\bigr)_+^2 \big/
-\bigl\|x - x^{(k)}\bigr\|_2^2$ and $r_k \in \mathbb{R}$ is the
-smallest radius such that at least $d+1$ other points $x^{(j)}$, $j
-\not = k$, are inside the closed Euclidean ball of radius $r_k$ about
-$x^{(k)}$. The computational complexity of ShepMod is
-$\mathcal{O}(n^2d)$. This paper uses a Fortran 95 implementation of
-ShepMod derived from SHEPPACK \cite{thacker2010algorithm}.
+     {\sum\limits_{k=1}^{n}W_k(x)} ,$$ where
+     $p\bigl(x^{(k)}\bigr) = f\bigl(x^{(k)}\bigr)$, $W_k(x) =
+       \bigl(\bigl(r_k - \bigl\|x - x^{(k)}\bigr\|_2\bigr)_+ \big/
+       \bigl( r_k \bigl\|x - x^{(k)}\bigr\|_2 \bigr)\bigr)^2$ for $x
+       \neq x^{(k)}$, and $r_k \in \mathbb{R}$ is the smallest
+     radius such that at least $d+1$ other points $x^{(j)}$, $j \not =
+     k$, are inside the closed Euclidean ball of radius $r_k$ about
+     $x^{(k)}$. The interpolant $p(x)$ is continuous because
+       the singularities at the $x^{(k)}$ are removable. The
+     computational complexity of ShepMod is $\mathcal{O}(n^2d)$. This
+     paper uses a Fortran 95 implementation of ShepMod derived from
+     SHEPPACK \cite{thacker2010algorithm}.
+
+\begin{figure}
+  \centering
+  \includegraphics[width=0.4\textwidth,]{Figures/NA/example_shepard_radius.pdf}
+  \hspace{.2cm}
+  \includegraphics[width=0.4\textwidth,]{Figures/NA/example_shepard_weight.pdf}
+  \caption{On the left above is a depiction of the radius
+    of influence for three chosen points of a collection in two
+    dimensions using the modified Shepard criteria. On the right a
+    third axis shows the relative weight for the center most
+    interpolation point $x^{(i)}$ with the solid line representing its
+    radius of influence, where $W_i(x)$ is $0$ for $\|x-x^{(i)}\|_2
+    \geq r_i$ and $W_i(x) / \sum_{k=1}^n W_k(x) \to 1$ as $x \to
+    x^{(i)}$.
+  \vspace{-.1cm}}
+  \label{fig:shepard}
+\end{figure}
 
@@
  {\sum\limits_{k=1}^{n}W_k(x)} ,$$
-where $W_k(x)$ is a locally supported weighting function and $P_k(x)$
-is a local linear approximation to the data satisfying
+where $W_k(x)$ is the same as for the modified Shepard method and
+$P_k(x)$ is a local linear approximation to the data satisfying
 $P_k\bigl(x^{(k)}\bigr) = f\bigl(x^{(x)}\bigr)$. The computational
@@
 \cite{thacker2010algorithm}.
-
-\subsection{Box Splines}
-\label{sec:box-splines}
-
-The box spline model used here is an interpolation technique built
-from overlapping box splines \cite{de2013box}. The box splines serve
-as basis functions that are shifted and scaled to have support over
-box shaped regions. The boxes are constructed in a way to guarantee a
-covering of the domain \cite{lux2018novel}. Given a set of box splines
-$\bigl\{b^{x^{(i)}}\bigr\}$ with the iterative box properties outlined
-in \cite{lux2018novel} and anchored at interior points
-$\bigl\{x^{(i)}\bigr\}$,
- $$ \hat f(y) = \frac{\sum\limits_{i=1}^n b^{x^{(i)}}(y) f\bigl(x^{(i)}\bigr)}{\sum\limits_{i=1}^n b^{x^{(i)}}(y)}. $$
-Note that the box splines always satisfy
-$b^{x^{(i)}}\bigl(x^{(j)}\bigr) = \delta_{ij}$ and $b^{x^{(i)}}(y)
-\geq 0$. The computational complexity of interpolation via the box
-spline model is $\mathcal{O}(n^2 d)$.
-
-\subsection{Voronoi}
-\label{sec:voronoi}
-
-The nearest neighbor algorithm \cite{cover1967nearest} is a
-well-studied technique for classification and approximation. Nearest
-neighbor inherently utilizes the convex region $C^{x^{(i)}}$ (Voronoi
-cell \cite{dirichlet1850reduction}) consisting of all points closer to
-$x^{(i)}$ than to any other point $x^{(j)}$. The Voronoi model smooths
-the nearest neighbor approximation by utilizing the Voronoi cells to
-define support via a generic basis function $v: \mathbb{R}^d
-\rightarrow \mathbb{R}_+$ given by
- $$ v^{x^{(i)}}(y) = \left(1 - \frac{\bigl\|y - x^{(i)}\bigr\|_2}
- {2 \ h \bigl(y - x^{(i)} \mid x^{(i)}\bigr)} \right)_+, $$
-where $h\bigl(w \mid x^{(i)}\bigr)$ is the distance between $x^{(i)}$
-and the boundary of the Voronoi cell $C^{x^{(i)}}$ in the direction
-$w$. $v^{x^{(i)}}\bigl(x^{(j)}\bigr) = \delta_{ij}$ and $v^{x^{(i)}}$
-has local support, giving the interpolated value
- $$ f(y) = \frac{\sum\limits_{i=1}^n v^{x^{(i)}}(y)
-  f\bigl(x^{(i)}\bigr)}{\sum\limits_{i=1}^n v^{x^{(i)}}(y)}, $$
-where $0 \leq v^{x^{(i)}}(y) \leq 1$. The computational complexity of
-interpolation via this Voronoi model is $\mathcal{O}(n^2 d)$. All of
-the approximations are an interpolant involving a convex combination
-of known function values $f\bigl(x^{(i)}\bigr)$.
-
-%% ===================================================================
-\section{Measuring Error}
-\label{sec:error}
-
-\begin{figure}[htb]
-  \centering
-  \begin{tikzpicture}
-    \node (img)  {\includegraphics[width=0.75\textwidth,]{Fig1.pdf}};
-    \node[below=of img, node distance=1cm, yshift=1cm] {I/O Throughput};
-    \node[left=of img, node distance=0cm, rotate=90, anchor=center,yshift=-0.7cm] {CDF Value};
-  \end{tikzpicture}
-  \vspace{-0.3cm}
-  \caption{In this HPC I/O example, the general methodology for
-    predicting a CDF and evaluating error can be seen. The Delaunay
-    method chose three source distributions (dotted lines) and
-    assigned weights \{.3, .4, .3\} (top to bottom at arrow). The
-    weighted sum of the three known CDFs produces the predicted CDF
-    (dashed line). The KS Statistic (arrow) computed between the true
-    CDF (solid line) and predicted CDF (dashed line) is 0.2 for this
-    example. The KS test null hypothesis is rejected at $p$-value
-    0.01, however it is not rejected at $p$-value 0.001.
-  \vspace{-.1cm}}
-  \label{fig:prediction-example}
-\end{figure}
-
-When the range of an approximation is the real numbers, error is
-reported with summary statistics including: min absolute error, max
-absolute error, and absolute error quartiles. When the range of an
-approximation is the space of cumulative distribution functions, the
-Kolmogorov-Smirnov statistic (max-norm difference between the
-functions) is used.
-
-A hurdle when modeling function-valued outputs such as cumulative
-distribution functions (CDFs) or probability density functions (PDFs)
-is that certain properties must be maintained. It is necessary that a
-PDF $f: \mathbb{R} \rightarrow \mathbb{R}$ have the properties $f(x)
-\geq 0$ and $\int_{-\infty}^{\infty}f(x)dx = 1$. However, for a CDF
-$F: \mathbb{R} \rightarrow \mathbb{R}$ the properties are
-$F(x) \in [0,1]$ and $F(x)$ is absolutely continuous and
-nondecreasing. This work utilizes the fact that a convex combination
-of CDFs (or PDFs) results in a valid CDF (or PDF). Given $G(x) =
-\sum_{i}w_i F_i(x)$, $\sum_{i} w_i = 1$, $w_i \geq 0$, and each $F_i$
-is a valid CDF, $G$ must also be a valid CDF. A demonstration of how
-this is applied can be seen in Figure \ref{fig:prediction-example}.
-
-The performance of approximation techniques that predict probability
-functions can be analyzed through a variety of summary
-statistics. This work uses the max absolute difference, also known as
-the Kolmogorov-Smirnov (KS) statistic \cite{lilliefors1967kolmogorov}
-for its compatibility with the KS test.
-
-The two-sample KS test is a useful nonparametric test for comparing
-two CDFs while only assuming stationarity, finite mean, and finite
-variance. The null hypothesis (that two CDFs come from the same
-underlying distribution) is rejected at level $p \in [0,1]$ when
- $$ KS > \sqrt{-\frac{1}{2}\ln\biggl(\frac{p}{2}\biggr)} \sqrt{\frac{1}{n_1} + \frac{1}{n_2}}, $$
-with distribution sample sizes $n_1,n_2 \in \mathcal{N}$. For all
-applications of the KS test presented in this work $n_1 = n_2$. An
-example of the round-trip prediction methodology from known and
-predicted distributions to the calculation of error can be seen in
-Figure \ref{fig:prediction-example}.
-
-\section{Theoretical Error Bound}
-\label{sec:theory}
-
-This section presents the theoretical results bounding the error of
-(piecewise) linear interpolation. The error analysis relies on linear
-interpolation for three reasons: (1) second order results can be
-obtained utilizing a Lipschitz constant on the gradient of a function,
-rather than standard Lipschitz bounds; (2) the results directly apply
-to Delaunay interpolation; and (3) multiple other interpolants in this
-paper compute predictions as convex combinations of observed function
-values, which may allow for straight forward extensions of this error
-bound.
-
-\begin{plainlemma}
-  \label{lemma:1}
-  Let $S \subset \mathbb{R}^d$ be open and convex, $f: \mathbb{R}^d
-  \rightarrow \mathbb{R}$, and $\nabla f \in
-  Lip_{(\gamma,\|\cdot\|_2)}(S)$, the set of $\gamma$-Lipschitz
-  continuous functions in the $2$-norm. Then for all $x,y \in S$
-  $$\big|f(y) - f(x) - \langle \nabla f(x), y - x \rangle \big| \leq \frac{\gamma \|y - x\|_2^2}{2}.$$
-\end{plainlemma}
-
-\begin{proofdot}
-  Consider the function $g(t) = f \big((1-t) x + t y \big)$, $0 \leq t
-  \leq 1$, whose derivative $g'(t) = \big\langle \nabla f \big((1-t) x
-  + t y \big), y - x \big\rangle$ is the directional derivative of $f$
-  in the direction $(y - x).$
-  \begin{align*}
-    \big|f(y) - f(x) - \langle \nabla f(x), y - x \rangle \big|
-       &= \big|g(1) - g(0) - g'(0) \big| & \\
-       &= \bigg| \int_0^1 g'(t) - g'(0)\ dt \bigg| \\
-       &\leq \int_0^1 \big|g'(t) - g'(0)\big|\ dt \\
-       &= \int_0^1 \bigg| \big \langle \nabla f\big((1-t)x + ty\big) - \nabla f(x), y - x \big \rangle \bigg|\ dt \\
-       &\leq \int_0^1 \big \| \nabla f\big((1-t)x + ty\big) - \nabla f(x) \big \|_2\ \| y - x \|_2\ dt \\
-       &\leq \int_0^1 \big ( \gamma\ \|y-x\|_2 \big) \ \big( \|y-x\|_2 \big) t\ dt \\
-       &= \frac{\gamma \|y - x\|_2^2}{2}.
-  \end{align*}
-  \qed
-\end{proofdot}
-
-\begin{plainlemma}
-  \label{lemma:2}
-  Let $x, y, v_i \in \mathbb{R}^d$, $c_i \in \mathbb{R}$, and
-  $|\langle y - x, v_i \rangle| \leq c_i$ for $i = 1$, $\ldots$, $d.$
-  If $M = (v_1$, $\ldots$, $v_d)$ is nonsingular, then
-  $$\|y - x\|_2^2 \leq \frac{1}{\sigma_d^2} \sum_{i=1}^d c_i^2,$$
-  where $\sigma_d$ is the smallest singular value of $M.$
-\end{plainlemma}
-
-\begin{proofdot}
-  Using the facts that $M$ and $M^t$ have the same singular values,
-  and $\|M^tw\|_2 \geq \sigma_d \|w\|_2$, gives
-  \begin{align*}
-    \|y - x\|_2^2 &\leq \frac{\|M^t (y - x)\|_2^2}{\sigma_d^2} \\
-                  &=    \frac{1}{\sigma_d^2} \sum_{i=1}^d \langle y - x, v_i \rangle^2 \\
-                  &\leq \frac{1}{\sigma_d^2} \sum_{i=1}^d c_i^2.
-  \end{align*}
-  \qed
-\end{proofdot}
-
-\begin{plainlemma}
-  \label{lemma:3}
-  Given $f$, $\gamma$, $S$ as in {\it Lemma \ref{lemma:1}}, let $X =
-  \{x_0$, $x_1$, $\ldots$, $x_d\}$ $\subset S$ be the vertices of a
-  $d$-simplex, and let $\hat f(x) = \langle c, x - x_0 \rangle +
-  f(x_0)$, $c \in \mathbb{R}^d$ be the linear function interpolating
-  $f$ on $X.$ Let $\sigma_d$ be the smallest singular value of the
-  matrix $M = (x_1 - x_0$, $\ldots$, $x_d - x_0)$, and $k =
-  \max\limits_{1\ \leq\ j\ \leq\ d} \|x_j - x_0\|_2.$ Then
-  $$\big\|\nabla f(x_0) - c\big\|_2 \leq \sqrt{d} \frac{\gamma k^2}{2 \sigma_d}.$$
-\end{plainlemma}
-
-\begin{proofdot}
-  Consider $f(x) - \hat f(x)$ along the line segment $z(t) = (1-t)x_0
-  + t x_j$, $0 \leq t \leq 1.$ By Rolle's Theorem, for some $0 < \hat
-  t < 1$, $\big\langle \nabla f\big(z(\ \hat t\ )\big) - c, x_j - x_0
-  \big\rangle = 0.$ Now
-  \begin{align*}
-    \big| \big< \nabla f(x_0) - c, x_j - x_0 \big \rangle \big|
-        &= \big| \big< \nabla f(x_0) - \nabla f\big(z(\ \hat t\ )\big) + \nabla f\big(z(\ \hat t\ )\big) - c, x_j - x_0 \big \rangle \big| \\
-	&= \big| \big\langle \nabla f(x_0) - \nabla f \big(z(\ \hat t\ )\big), x_j - x_0 \big \rangle \big| \\
-	&\leq \big \| \nabla f(x_0) - \nabla f\big(z(\ \hat t\ )\big) \big \|_2 \|x_j - x_0\|_2 \\
-	&\leq \gamma \|x_0 - z(\ \hat t\ )\|_2\ \|x_j - x_0\|_2 \\
-	&\leq \frac{\gamma}{2} \|x_j - x_0\|_2^2 \leq \frac{\gamma k^2}{2}, \\
-  \end{align*}
-  for all $1 \leq j \leq d.$ Using {\it Lemma \ref{lemma:2}},
-  $$\big \| \nabla f(x_i) - c \big \|_2^2 \leq \frac{d}{\sigma_d^2}
-  \big( \gamma k^2 / 2 \big)^2 \implies \big \| \nabla f(x_i) - c \big \|_2
-  \leq \sqrt{d} \frac{\gamma k^2}{2 \sigma_d}.$$
-  \qed
-\end{proofdot}
-
-\begin{onetheorem}
-  Under the assumptions of {\it Lemma \ref{lemma:1}} and {\it Lemma
-    \ref{lemma:3}}, for $z \in S$,
-  $$ \big|f(z) - \hat f(z)\big| \leq \frac{\gamma \|z - x_0\|_2^2}{2} + \sqrt{d} \frac{\gamma k^2}{2 \sigma_d} \|z - x_0\|_2.$$
-\end{onetheorem}
-
-\begin{proofdot}
-  Let $v = \nabla f(x_0) - c$, where $\|v\|_2 \leq \sqrt{d} \gamma k^2 / (2 \sigma_d)$ by {\it Lemma \ref{lemma:3}}. Now
-  \begin{align*}
-    \big|f(z) - \hat f(z)\big|
-       &= \big|f(z) - f(x_0) - \langle c, z - x_0 \rangle \big| \\ 
-       &= \big|f(z) - f(x_0) - \langle \nabla f(x_0) - v, z - x_0 \rangle \big| \\
-       &= \big|f(z) - f(x_0) - \langle \nabla f(x_0) , z - x_0 \rangle + \langle v , z - x_0 \rangle \big| \\
-       &\leq \big|f(z) - f(x_0) - \langle \nabla f(x_0) , z - x_0 \rangle \big| + \big| \langle v , z - x_0 \rangle \big| \\
-       &\leq \big|f(z) - f(x_0) - \langle \nabla f(x_0) , z - x_0 \rangle \big| + \|v\|_2 \|z - x_0\|_2 \\
-       &\leq \big|f(z) - f(x_0) - \langle \nabla f(x_0), z - x_0 \rangle \big| + \textstyle{\sqrt{d}\frac{\gamma k^2}{2 \sigma_d}} \|z - x_0\|_2 \\
-       &\leq \frac{\gamma \|z - x_0\|_2^2}{2} + \sqrt{d}\frac{\gamma k^2}{2 \sigma_d} \|z - x_0\|_2,
-  \end{align*}
-  where the last inequality follows from {\it Lemma \ref{lemma:1}}.
-  %
-  \qed
-\end{proofdot}
-
-In summary, the approximation error of a linear (simplicial)
-interpolant tends quadratically towards zero when approaching observed
-data only when the diameter of the simplex is also reduced at a
-proportional rate.  Only linear convergence to the true function can
-be achieved in practice, without the incorporation of additional
-observations. Notice that the approximation error is largely
-determined by the spacing of observed data. Predictions made by
-simplices whose vertices are not well-spaced (i.e., have large
-diameter, or are nearly contained in a hyperplane) have higher
-error.
-
-The error bound provided is tight and that is easily observed by
-considering the test function $q(x) = \|x\|_2^2,$ $x \in \mathbb{R}^d$
-for $d \in \mathcal{N},$ using the simplex $X = \{\mathbf{0}, I\},$
-where $I$ is the $d \times d$ identity matrix, and an approximation is
-made at $z = \mathbf{1/2}.$ Now this setup yields
-
-\vspace{-4mm}
-$$ \gamma = 1, \qquad \sigma_d = 1, \qquad k = 1, \qquad x_0 =
-\mathbf{0}, \qquad \hat q(x) = \langle\mathbf{1}, x - \mathbf{0}\rangle +
-q(\mathbf{0}). $$
-\vspace{-2mm}
-\noindent Plugging these values into the Theorem, we get
-\vspace{-2mm}
-
-\begin{align*}
-  \big|q(z) - \hat q(z)\big| \quad
-  =& \quad \big|\|z\|_2^2 - \langle\mathbf{1},z\rangle\big| \\
-  =& \quad \bigg| \frac{d}{4} - \frac{d}{2} \bigg| \quad = \quad \frac{d}{4} \\
-  \leq& \quad \frac{\gamma \|z - x_0\|_2^2}{2} + \sqrt{d} \frac{\gamma k^2}
-      {2 \sigma_d} \|z - x_0\|_2 \\
-  =& \quad \frac{\|z\|_2^2 + \sqrt{d} \|z\|_2}{2} \\
-  =& \quad \bigg(\frac{d}{4} + \frac{d}{4}\bigg)\ /\ 2 \quad = \quad \frac{d}{4}.
-\end{align*}
-
-In light of this error bound and analytic test, an empirical
-evaluation of presented algorithms follows.
-
-%% ===================================================================
-\vspace{-2mm}
-\section{Data and Empirical Analysis}
-\label{sec:data}
-
-The theoretical results constructed in Section \ref{sec:theory} for
-(piecewise) linear interpolation are promising and apply directly to
-Delaunay interpolation, however they are difficult to interpret in
-context with approximation algorithms that do not have similar known
-uniform error bounds. For that reason, this paper utilizes five
-different data sets of varying dimension and application to construct
-approximations and compare the accuracy of different techniques.
-
-In the following five subsections the sources and targets of each test
-data set are described, as well as challenges and limitations related
-to approximating these data. The distribution of response values being
-modeled is presented followed by the distribution of approximation
-errors for each algorithm. The plots for all five data sets have the
-same format.
-
-All five data sets are rescaled such that the domain of approximation
-is the unit hypercube. The range of the first four data sets is the
-real numbers, while the range of the fifth data set is the space of
-cumulative distribution functions. All approximation techniques are
-applied to the first four data sets, while only those interpolants
-whose approximations are convex combinations of observed data are
-applied to the final data set.
-
-All approximations are constructed using $k$-fold cross validation as
-described in \cite{kohavi1995study} with $k=10$. This approach
-randomly partitions data into $k$ (nearly) equal sized sets. Each
-algorithm is then evaluated by constructing an approximation over each
-unique union of $k-1$ elements of the partition, making predictions
-for points in the remaining element. As a result, each observed data
-point is used in the construction of $k-1$ different approximations
-and is approximated exactly once. The $k$-fold cross validation method
-is data-efficient and provides an unbiased estimate of the expected
-prediction error \cite{kohavi1995study}, however it should be noted
-that neither this method nor others can provide a universally unbiased
-estimator for the variance of prediction error \cite{bengio2004no}.
-
-In addition to the figures displaying approximation results for each
-data set, tables of accompanying numerical results are located in the
-Appendix. All of the test data sets capture underlying functions that
-are almost certainly stochastic. As described in Section
-\ref{sec:introduction}, regression techniques appear most appropriate
-for these problems. However, typically data grows exponentially more
-sparse with increasing dimension. Given sparse data, regressors tend
-towards interpolation.
-
-%% -------------------------------------------------------------------
-\subsection{Forest Fire ($n = 504$, $d = 12$)}
-
-\begin{figure}
-  \centering
-  \begin{tikzpicture}
-    \node (img)  {\includegraphics[width=0.75\textwidth,]{Fig2.pdf}};
-    \node[below=of img, node distance=1cm, yshift=1cm] {Forest Fire Area Burned};
-    \node[left=of img, node distance=0cm, rotate=90, anchor=center,yshift=-0.7cm] {Count};
-  \end{tikzpicture}
-  \caption{Histogram of forest fire area burned under recorded weather
-    conditions. The data is presented on a $\ln$ scale because
-    most values are small with exponentially fewer fires on record
-    that burn large areas.}
-  \label{fig:hist-forest-fire}
-\end{figure}
-
-\begin{figure}
-  \centering
-  \begin{tikzpicture}
-    \node (img)  {\includegraphics[width=0.8\textwidth,]{Fig3.pdf}};
-    \node[below=of img, node distance=1cm, yshift=1cm] {Area Burned Error};
-  \end{tikzpicture}
-  \caption{All models are applied to approximate the amount of area
-    that would be burned given environment conditions. $10$-fold cross
-    validation as described in the beginning of Section \ref{sec:data}
-    is used to evaluated each algorithm. This results in exactly one
-    prediction from each algorithm for each data point. These boxes
-    depict the median (middle bar), median $95\%$ confidence interval
-    (cones), quartiles (box edges), fences at $3/2$ interquartile
-    range (whiskers), and outliers (dots) of absolute prediction error
-    for each model. Similar to Figure \ref{fig:hist-forest-fire}, the
-    errors are presented on a $\ln$ scale. The numerical data
-    corresponding to this figure is provided in Table
-    \ref{table:error-forest-fire} in the Appendix.}
-  \label{fig:error-forest-fire}
-\end{figure}
-
-The forest fire data set \cite{cortez2007data} describes the area of
-Montesinho park burned over months of the year along with
-environmental conditions. The twelve dimensions being used to model
-burn area are the $x$ and $y$ spatial coordinates of burns in the
-park, month of year (mapped to $x$, $y$ coordinates on a unit circle),
-the FFMC, DMC, DC, and ISI indices (see source for details), the
-temperature, relative humidity, wind speed, and outdoor rain. The
-original analysis of this data set demonstrated it to be difficult to
-model, likely due to the skew in response values.
-
-As suggested by Figure \ref{fig:error-forest-fire}, the SVR has the
-lowest absolute prediction errors for $80\%$ of the data, with MLP and
-Delaunay being the nearest overall competitors. The effectiveness of
-SVR on this data suggests the underlying function can be defined by
-relatively few parameters, as well as the importance of capturing the
-low-burn-area data points.
-%% -------------------------------------------------------------------
-
-%% -------------------------------------------------------------------
-\subsection{Parkinson's Telemonitoring ($n = 5875$, $d = 19$)}
-
-\begin{figure}
-  \centering
-  \begin{tikzpicture}
-    \node (img)  {\includegraphics[width=0.75\textwidth,]{Fig4.pdf}};
-    \node[below=of img, node distance=1cm, yshift=1cm] {Parkinson's Total UPDRS Score};
-    \node[left=of img, node distance=0cm, rotate=90, anchor=center,yshift=-0.7cm] {Count};
-  \end{tikzpicture}
-  \caption{Histogram of the Parkinson's patient total UPDRS clinical
-    scores that will be approximated by each algorithm.}
-  \label{fig:hist-parkinsons}
-\end{figure}
-
-\begin{figure}
-  \centering
-  \begin{tikzpicture}
-    \node (img)  {\includegraphics[width=0.8\textwidth,]{Fig5.pdf}};
-    \node[below=of img, node distance=1cm, yshift=1cm] {Total UPDRS Score Error};
-  \end{tikzpicture}
-  \caption{All models are applied to approximate the total UPDRS score
-    given audio features from patients' home life, using $10$-fold
-    cross validation. These boxes depict the median (middle bar),
-    median $95\%$ confidence interval (cones), quartiles (box edges),
-    fences at $3/2$ interquartile range (whiskers), and outliers
-    (dots) of absolute prediction error for each model. The numerical
-    data corresponding to this figure is provided in Table
-    \ref{table:error-parkinsons} in the Appendix.}
-  \label{fig:error-parkinsons}
-\end{figure}
-
-The second data set for evaluation \cite{tsanas2010accurate} is
-derived from a speech monitoring study with the intent to
-automatically estimate Parkinson's disease symptom development in
-Parkinson's patients. The function to be approximated is a
-time-consuming clinical evaluation measure referred to as the UPDRS
-score. The total UPDRS score given by a clinical evaluation is
-estimated through 19 real numbers generated from biomedical voice
-measures of in-home sound recordings.
-
-Figure \ref{fig:error-parkinsons} shows the ShepMod algorithm has the
-lowest minimum, first quartile, and median of absolute errors for this
-problem, while providing the best prediction $66\%$ of the time. The
-MLP has the lowest third quartile and provides the best prediction for
-$32\%$ of approximations. The dominance of ShepMod may be due in part
-to regular-interval total UPDRS scores provided by clinicians,
-favoring a nearest-neighbor strategy of prediction.
-%% -------------------------------------------------------------------
-
-%% -------------------------------------------------------------------
-\subsection{Australian Daily Rainfall Volume ($n = 2609$, $d = 23$)}
-
-\begin{figure}
-  \centering
-  \begin{tikzpicture}
-    \node (img)  {\includegraphics[width=0.75\textwidth,]{Fig6.pdf}};
-    \node[below=of img, node distance=1cm, yshift=1cm] {Daily Rainfall in Sydney};
-    \node[left=of img, node distance=0cm, rotate=90, anchor=center,yshift=-0.7cm] {Count};
-  \end{tikzpicture}
-  \caption{Histogram of daily rainfall in Sydney, Australia, presented
-    on a $\ln$ scale because the frequency of larger amounts of
-    rainfall is significantly less. There is a peak in occurrence of
-    the value $0$, which has a notable effect on the resulting model
-    performance.}
-  \label{fig:hist-weather}
-\end{figure}
-
-\begin{figure}
-  \centering
-  \begin{tikzpicture}
-    \node (img)  {\includegraphics[width=0.8\textwidth,]{Fig7.pdf}};
-    \node[below=of img, node distance=1cm, yshift=1cm] {Sydney Rainfall Tomorrow Error};
-  \end{tikzpicture}
-  \caption{All models are applied to approximate the amount of
-    rainfall expected on the next calendar day given various sources
-    of local meteorological data, using $10$-fold cross validation.
-    These boxes depict the median (middle bar), median $95\%$
-    confidence interval (cones), quartiles (box edges), fences at
-    $3/2$ interquartile range (whiskers), and outliers (dots) of
-    absolute prediction error for each model. The errors are presented
-    on a $\ln$ scale, mimicking the presentation in Figure
-    \ref{fig:hist-weather}. The numerical data corresponding to this
-    figure is provided in Table \ref{table:error-weather} in the
-    Appendix.}
-  \label{fig:error-weather}
-\end{figure}
-
-The third data set for the total daily rainfall in Sydney, Australia
-\cite{williams2009rattle} provides a slightly higher dimensional
-challenge for the interpolants and regressors. This data is composed
-of many meteorological readings from the region in a day including:
-min and max temperatures, sunshine, wind speed directions (converted
-to coordinates on a circle), wind speeds, and humidities throughout
-the day, day of the year (converted to coordinates on a circle), and
-the model must predict the amount of rainfall tomorrow.
-
-While Figure \ref{fig:hist-weather} makes MARS look far better than
-other techniques, it only provides the best prediction for $11\%$ of
-points. The MLP has the lowest absolute error for $56\%$ of points and
-LSHEP is best for $28\%$. MARS likely achieves such a low first
-quartile due to the high occurrence of the value zero in the data.
-%% -------------------------------------------------------------------
-
-%% -------------------------------------------------------------------
-\subsection{Credit Card Transaction Amount ($n = 5562$, $d = 28$)}
-
-\begin{figure}
-  \centering
-  \begin{tikzpicture}
-    \node (img)  {\includegraphics[width=0.75\textwidth,]{Fig8.pdf}};
-    \node[below=of img, node distance=1cm, yshift=1cm] {Credit Card Transaction Amount};
-    \node[left=of img, node distance=0cm, rotate=90, anchor=center,yshift=-0.7cm] {Count};
-  \end{tikzpicture}
-  \caption{Histogram of credit card transaction amounts, presented on
-    a $\ln$ scale. The data contains a notable frequency peak around
-    $\$1$ transactions. Fewer large purchases are made, but some large
-    purchases are in excess of five orders of magnitude greater than
-    the smallest purchases.}
-  \label{fig:hist-credit-card}
-\end{figure}
-
-\begin{figure}
-  \centering
-  \begin{tikzpicture}
-    \node (img)  {\includegraphics[width=0.8\textwidth,]{Fig9.pdf}};
-    \node[below=of img, node distance=1cm, yshift=1cm] {Transaction Amount Error};
-  \end{tikzpicture}
-  \caption{All models are applied to approximate the expected
-    transaction amount given transformed (and obfuscated) vendor and
-    customer-descriptive features, using $10$-fold cross validation.
-    These boxes depict the median (middle bar), median $95\%$
-    confidence interval (cones), quartiles (box edges), fences at
-    $3/2$ interquartile range (whiskers), and outliers (dots) of
-    absolute prediction error for each model. The absolute errors in
-    transaction amount predictions are presented on a $\ln$ scale,
-    just as in Figure \ref{fig:hist-credit-card}. The numerical data
-    corresponding to this figure is provided in Table
-    \ref{table:error-credit-card} in the Appendix.}
-  \label{fig:error-credit-card}
-\end{figure}
-
-The fourth test data set, and the final with a real-valued range, is a
-collection of credit card transactions
-\cite{pozzolo2015calibrating}. The provided data carries no direct
-real-world meaning, being the output of a principle component analysis
-on the original hidden source data. This obfuscation is done to
-protect the information of the credit card users. This data has the
-largest dimension of all considered, at 28. A model for this data
-predicts the transaction amount given the vector of principle
-component information.
-
-As can be seen in Figure \ref{fig:error-credit-card}, the MLP
-outperforms all other algorithms at the first, second, third, and
-fourth quartiles. The MLP produces the lowest absolute error
-prediction for $80\%$ of transactions, Delaunay bests the remaining
-$20\%$. It is likely that with less data, Delaunay would be the best
-performer.
-%% -------------------------------------------------------------------
-
-%% -------------------------------------------------------------------
-\subsection{High Performance Computing I/O ($n = 3016$, $d = 4$)}
-
-\begin{figure}
-  \centering
-  \begin{tikzpicture}
-    \node (img)  {\includegraphics[width=0.75\textwidth,]{Fig10.pdf}};
-    \node[below=of img, node distance=1cm, yshift=1cm] {I/O Read Throughput};
-    \node[left=of img, node distance=0cm, rotate=90, anchor=center,yshift=-0.7cm] {Count};
-  \end{tikzpicture}
-  \caption{Histogram of the raw throughput values recorded during all
-    IOzone tests across all system configurations. The distribution is
-    skewed right, with few tests having significantly higher
-    throughput than most others. The data is presented on a $\ln$
-    scale.}
-  \label{fig:hist-throughput}
-\end{figure}
-
-\begin{figure}
-  \centering
-  \begin{tikzpicture}
-    \node (img)  {\includegraphics[width=0.8\textwidth,]{Fig11.pdf}};
-    \node[below=of img, node distance=1cm, yshift=1cm] {Read Throughput KS Statistic};
-  \end{tikzpicture}
-  \caption{The models directly capable of predicting distributions are
-    applied to predicting the expected CDF of I/O throughput at a
-    previously unseen system configuration, using $10$-fold cross
-    validation. The KS statistic between the observed distribution at
-    each system configuration and the predicted distribution is
-    recorded and presented above. Note that the above figure is
-    \textit{not} log-scaled like Figure \ref{fig:hist-throughput}. The
-    numerical data corresponding to this figure is provided in Table
-    \ref{table:error-throughput} in the Appendix.}
-  \label{fig:error-throughput}
-\end{figure}
-
-The final of five data sets is derived from \cite{cameron2019moana},
-which provides four-dimensional distribution data by executing the
-IOzone benchmark \cite{iozone} on a computer system and varying the
-system's file size, record size, thread count, and CPU frequency. At
-each configuration, IOzone samples the I/O file-read throughput (in
-bytes per second) 150 times. Empirical distribution function points
-are computed from each set of 150 executions, which are interpolated
-with a piecewise cubic Hermite interpolating polynomial
-\cite{fritsch1980monotone} to approximate the CDF. All interpolation
-algorithms with the exception of LSHEP are used to approximate these
-CDFs from system configurations.
-
-Delaunay achieves the lowest KS statistic for $62\%$ of
-approximations, while Voronoi is best for the remaining $38\%$. Figure
-\ref{fig:error-throughput} shows that while Delaunay may have more
-best predictions, the behavior of Voronoi may be preferable.
-%% -------------------------------------------------------------------
-
-%% -------------------------------------------------------------------
-\begin{figure}
-  \centering
-  \begin{tikzpicture}
-    \node (img)  {\includegraphics[width=0.8\textwidth]{Fig12.pdf}};
-    \node[below=of img, node distance=1cm, yshift=1cm] {KS Statistic for Predicted vs. Actual};
-    \node[left=of img, node distance=0cm, rotate=90, anchor=center,yshift=-0.7cm] {Count of KS Statistic};
-  \end{tikzpicture}
-  \caption{Histograms of the prediction error for each interpolant
-    that produces predictions as convex combinations of observed data,
-    using $10$-fold cross validation. The histograms show the KS
-    statistics for the predicted throughput distribution versus the
-    actual throughput distribution. The four vertical lines represent
-    cutoff KS statistics given $150$ samples for commonly used
-    $p$-values 0.05, 0.01, 0.001, $10^{-6}$, respectively. All
-    predictions to the right of a vertical line represent CDF
-    predictions that are significantly different (by respective
-    $p$-value) from the actual distribution according to the KS
-    test. The numerical counterpart to this figure is presented in
-    Table \ref{table:null-hypothesis-results}.}
-  \label{fig:throughput-prediction}
-\end{figure}
-
-\begin{table}
-  \renewcommand{\arraystretch}{1.3}
-  \centering
-  \begin{tabular}{c|c|c|c|c}
-     & $p = .05$ & $p = .01$ & $p = .001$ & $p = 10^{-6}$\\
-    \hline
-    Delaunay & $\mathbf{50.3}\%$ & $\mathit{43.5}\%$ & $\mathit{36.2}\%$ & $\mathit{24.7}\%$\\
-    ShepMod & $\mathit{51.4}\%$ & $44.8\%$ & $38.1\%$ & $27.7\%$\\
-    Voronoi & $52.6\%$ & $\mathbf{43.4}\%$ & $\mathbf{34.4}\%$ & $\mathbf{19.1}\%$\\
-    BoxSpline & $99.5\%$ & $98.5\%$ & $96.6\%$ & $89.5\%$\\
-  \end{tabular}
-  \caption{Numerical counterpart of the histogram data presented in
-    Figure \ref{fig:throughput-prediction}. The columns display the
-    percent of null hypothesis rejections by the KS-test when provided
-    different selections of $p$-values for each algorithm. The
-    algorithm with the lowest rejection rate at each $p$ is boldface,
-    while the second lowest is italicized.}
-  \label{table:null-hypothesis-results}
-\end{table}
-
-%% ===================================================================
-
-Figure \ref{fig:throughput-prediction} expands on the KS statistic
-results presented in Figure \ref{fig:error-throughput}. Agglomerate
-errors for each algorithm resemble a Gamma distribution. The
-percentages of significant prediction errors with varying $p$-values
-are on display in Table \ref{table:null-hypothesis-results}. When
-considering the $p=0.001$ results for each technique, a little over
-one third of the predicted CDFs are significantly different from the
-measured (and presumed) correct CDFs. However, it should be noted that
-with 150 samples, the error of an empirical distribution function
-(EDF) can reasonably be upwards of $.1$, which serves as a rough
-estimate for the lower limit of achievable error. Globally, only a
-third of Voronoi predictions fail to capture \textit{all} of the
-characteristics of the CDFs at new system configurations.
-
-\section{Discussion}
-\label{sec:discussion}
-
-\begin{table}
-  \centering
-  \begin{tabular}{c|r@{.}l|r@{.}l|r@{.}l}
-    Algorithm & \multicolumn{2}{c|}{Avg. \% Best} & \multicolumn{2}{c|}{\multilinecell{Avg. Fit or \\ Prep. Time (s)}} & \multicolumn{2}{c}{Avg. App. Time (s)}\\
-    \hline
-    MARS & \quad\quad$4$&$6$ & \,\quad$20$&$0$ & \quad\quad$0$&$001$\\
-    SVR & $\mathit{21}$&$\mathit{1}$ & $\mathbf{0}$&$\mathbf{5}$ & $\mathbf{0}$&$\mathbf{0001}$\\
-    MLP & $\mathbf{42}$&$\mathbf{1}$ & $200$&$0$ & $0$&$001$\\
-    Delaunay & $6$&$0$ & $1$&$0$ & $1$&$0$\\
-    ShepMod & $18$&$4$ & $\mathit{0}$&$\mathit{7}$ & $\mathbf{0}$&$\mathbf{0001}$\\
-    LSHEP & $8$&$4$ & $2$&$0$ & $\mathbf{0}$&$\mathbf{0001}$\\
-    Voronoi & $0$&$5$ & $1$&$0$ & $0$&$04$\\
-    BoxSpline & $0$&$2$ & $0$&$8$ & $\mathit{0}$&$\mathit{0005}$\\
-  \end{tabular}
-  \caption{This average of Appendix Tables
-    \ref{table:best-forest-fire}, \ref{table:best-parkinsons},
-    \ref{table:best-weather}, and \ref{table:best-credit-card}
-    provides a gross summary of overall results. The columns display
-    (weighted equally by data set, \textit{not} points) the average
-    frequency with which any algorithm provided the lowest absolute
-    error approximation, the average time to fit/prepare, and the
-    average time required to approximate one point. The times have
-    been rounded to one significant digit, as reasonably large
-    fluctuations may be observed due to implementation hardware.
-    Interpolants provide the lowest error approximation for nearly one
-    third of all data, while regressors occupy the other two
-    thirds. This result is obtained without any customized tuning or
-    preprocessing to maximize the performance of any given
-    algorithm. In practice, tuning and preprocessing may have large
-    effects on approximation performance.}
-  \label{table:avg-performance}
-\end{table}
-
-Table \ref{table:avg-performance} summarizes results across the four
-test data sets with real-valued ranges. The interpolants discussed in
-this paper produce the \textit{best} approximations roughly one third
-of the time, and produce competitive approximations for almost all
-data sets. These test problems are almost certainly
-\textit{stochastic} in nature, but the high dimension leads to greater
-data sparsity and model construction cost, making interpolation more
-competitive.
-
-The major advantages to interpolation lie in the near absence of
-\textit{fit} time. Delaunay, LSHEP, and ShepMod all require pairwise
-distance calculations, for numerical robustness (Delaunay) and to
-determine the radii of influence for data points (LSHEP and
-ShepMod). At least hundreds, and sometimes hundreds of thousands of
-predictions can be made by the interpolants before the most widely
-used regressor (MLP) finishes fitting these relatively small data
-sets.  However, the computational complexities of all interpolants
-presented are greater than linear in either dimension or number of
-points, whereas the regressors' nonlinear complexity in dimension
-generally comes from the model fitting optimization.
-
-The new theoretical results presented in Section \ref{sec:theory}
-directly apply to Delaunay interpolation, however the performance of
-Delaunay does not appear significantly better than other algorithms on
-these data sets.  This observation may be due to the stochastic nature
-of the data, but it also speaks to the power of the approximations
-generated by the different interpolation methods.  The strong
-performance of other interpolants suggests that theoretical results
-similar to those presented in this work can be achieved for the other
-interpolants under reasonable assumptions.
-
-Finally, most of the interpolants presented in this work benefit from
-the ability to model any function over real $d$-tuples with a range
-that is closed under convex combinations. The results of the
-distribution prediction case study indicate that interpolants can
-effectively predict CDFs. The error analysis for that work relies on
-the KS statistic, which captures the worst part of any prediction and
-hence provides a conservatively large estimate of approximation
-error. The average absolute errors in the predicted CDFs are always
-lower than the KS statistics. However, the KS statistic was chosen as
-a metric because of the important surrounding statistical theory. A
-nonnegligible volume of predictions provide impressively low levels of
-average absolute error in that final case study.
-
-\section{Conclusion}
-\label{sec:conclusion}
-
-The major contributions of this work are: 1) new uniform theoretical
-error bounds for piecewise linear interpolation in arbitrary dimension
-(Section \ref{sec:theory}); 2) an empirical evaluation across
-real-world problems that demonstrates interpolants produce
-competitively accurate models of multivariate phenomenon when compared
-with common regressors for sparse, moderately high dimensional
-problems (Section \ref{sec:data}); and 3) a demonstration that some
-interpolants generalize to interpolation in function spaces (Section
-\ref{sec:error}), preserving monotonicity (with CDFs, e.g.), neither
-of which common regressors can do.
-
-The various interpolants discussed in this paper have been
-demonstrated to effectively approximate multivariate phenomena up to
-$30$ dimensions. The underlying constructions are theoretically
-straightforward, interpretable, and yield reasonably accurate
-predictions. Most of the interpolants' computational complexities make
-them particularly suitable for applications in even higher
-dimension. The major benefits of interpolation are seen when only a
-small number of approximations ($\leq 1000$) are made from data and
-when there are relatively few data points for the dimension (for
-empirical results presented, $\log_d n \leq 5$). These findings
-encourage broader application of interpolants to multivariate
-approximation in science
-
-%\begin{acknowledgements}
-%If you'd like to thank anyone, place your comments here
-%and remove the percent signs.
-%\end{acknowledgements}
-
-%%====================================================================
-%%====================================================================
-
-\bibliographystyle{spmpsci}      % mathematics and physical sciences
-\bibliography{paper}
-
-\newpage
-
-\begin{appendix}
-\section*{Appendix}
-\label{sec:appendix}
-
-The tables that follow show the precise experimental results for all
-data sets presented in Section \ref{sec:data}. The tests were all run
-serially on an otherwise idle machine with a CentOS 6.10 operating
-system and an Intel i7-3770 CPU operating at 3.4 GHz. The detailed
-performance results in the tables that follow are very much dependent
-on the problem and the algorithm implementation (e.g., some codes are
-TOMS software, some industry distributions, and others are from
-conference paper venues). Different typeface is used to show best
-performers, however not much significance should be attached to
-ranking algorithms based on small time (millisecond) differences. The
-results serve as a demonstration of conceptual validity.
-
-\vspace*{\fill}
-
-\begin{table}[H]
-  \centering
-  \begin{tabular}{c|r@{.}l|r@{.}l|r@{.}l|r@{.}l|r@{.}l}
-    \hline
-    Algorithm & \multicolumn{2}{c|}{Min} & \multicolumn{2}{c|}{$25^{th}$} & \multicolumn{2}{c|}{$50^{th}$} & \multicolumn{2}{c|}{$75^{th}$} & \multicolumn{2}{c}{Max}\\
-    \hline
-    MARS & $0$&$00984$ & $3$&$11$ & $7$&$01$ & $\mathit{11}$&$\mathit{7}$ & $1090$&$0$\\
-    SVR & $\mathit{0}$&$\mathit{00402}$ & $\mathbf{0}$&$\mathbf{243}$ & $\mathbf{0}$&$\mathbf{416}$ & $\mathbf{6}$&$\mathbf{19}$ & $1090$&$0$\\
-    MLP & $0$&$0426$ & $2$&$63$ & $\mathit{5}$&$\mathit{27}$ & $14$&$0$ & $1090$&$0$\\
-    Delaunay & $\mathbf{0}$&$\mathbf{00}$ & $1$&$98$ & $5$&$37$ & $13$&$1$ & $\mathit{1080}$&$\mathit{0}$\\
-    ShepMod & $\mathbf{0}$&$\mathbf{00}$ & $\mathit{1}$&$\mathit{93}$ & $6$&$27$ & $16$&$0$ & $1090$&$0$\\
-    LSHEP & $0$&$0400$ & $2$&$17$ & $8$&$87$ & $19$&$1$ & $\mathbf{1070}$&$\mathbf{0}$\\
-    Voronoi & $0$&$00982$ & $3$&$65$ & $7$&$56$ & $15$&$6$ & $1090$&$0$\\
-    BoxSpline & $\mathbf{0}$&$\mathbf{00}$ & $2$&$27$ & $5$&$91$ & $16$&$4$ & $1090$&$0$\\
-    \hline
-  \end{tabular}
-  \caption{This numerical data accompanies the visual provided in
-    Figure \ref{fig:error-forest-fire}. The columns of absolute error
-    percentiles correspond to the minimum, first quartile, median,
-    third quartile, and maximum absolute errors respectively. The
-    minimum of each column is boldface, while the second lowest value
-    is italicized. All values are rounded to three significant
-    digits.}
-  \label{table:error-forest-fire}
-\end{table}
-
-\begin{table}[H]
-  \centering
-  \begin{tabular}{|c|r@{.}l| c |r@{.}l|r@{.}l|r@{.}l|}
-    \cline{1-3}\cline{5-10}
-    Algorithm & \multicolumn{2}{c|}{\% Best} &  & \multicolumn{2}{c|}{Fit/Prep. Time (s)} & \multicolumn{2}{c|}{App. Time (s)} & \multicolumn{2}{c|}{Total App. Time (s)}\\
-    \cline{1-3}\cline{5-10}
-    MARS & \quad$\mathit{7}$&$\mathit{7}$ &  & \quad\quad$29$&$1$ & \quad$0$&$00137$ & \quad\quad\quad$0$&$0686$\\
-    SVR & $\mathbf{80}$&$\mathbf{2}$ &  & $\mathbf{0}$&$\mathbf{00584}$ & $\mathit{0}$&$\mathit{0000620}$ & $\mathit{0}$&$\mathit{00310}$\\
-    MLP & $0$&$0$ &  & $32$&$8$ & $0$&$000871$ & $0$&$0436$\\
-    Delaunay & $3$&$5$ &  & $0$&$0151$ & $0$&$0234$ & $1$&$18$\\
-    ShepMod & $3$&$7$ &  & $\mathit{0}$&$\mathit{00634}$ & $0$&$0000644$ & $0$&$00322$\\
-    LSHEP & $5$&$1$ &  & $0$&$0275$ & $\mathbf{0}$&$\mathbf{0000463}$ & $\mathbf{0}$&$\mathbf{00231}$\\
-    Voronoi & $0$&$0$ &  & $0$&$0152$ & $0$&$000396$ & $0$&$0198$\\
-    BoxSpline & $0$&$4$ &  & $0$&$00724$ & $0$&$0000978$ & $0$&$00489$\\
-    \cline{1-3}\cline{5-10}
-  \end{tabular}
-  \caption{The left above shows how often each algorithm had the
-    lowest absolute error approximating forest fire data in Table
-    \ref{table:error-forest-fire}. On the right columns are median fit
-    time of 454 points, median time for one approximation, and median
-    time approximating 50 points.}
-  \label{table:best-forest-fire}
-\end{table}
-
-\vspace*{\fill}
-\newpage
-
-\begin{table}
-  \centering
-  \begin{tabular}{c|r@{.}l|r@{.}l|r@{.}l|r@{.}l|r@{.}l}
-    \hline
-    Algorithm & \multicolumn{2}{c|}{Min} & \multicolumn{2}{c|}{$25^{th}$} & \multicolumn{2}{c|}{$50^{th}$} & \multicolumn{2}{c|}{$75^{th}$} & \multicolumn{2}{c}{Max}\\
-    \hline
-    MARS & $0$&$00948$ & $9$&$98$ & $20$&$4$ & $32$&$9$ & $863$&$0$\\
-    SVR & $0$&$00138$ & $3$&$17$ & $7$&$31$ & $12$&$6$ & $\mathbf{27}$&$\mathbf{5}$\\
-    MLP & $0$&$0000239$ & $\mathit{0}$&$\mathit{533}$ & $\mathit{1}$&$\mathit{25}$ & $\mathbf{2}$&$\mathbf{84}$ & $39$&$3$\\
-    Delaunay & $\mathit{3}$&$\mathit{72 \times 10^{-12}}$ & $1$&$20$ & $3$&$50$ & $7$&$67$ & $30$&$7$\\
-    ShepMod & $\mathbf{0}$&$\mathbf{0}$ & $\mathbf{0}$&$\mathbf{255}$ & $\mathbf{0}$&$\mathbf{908}$ & $\mathit{3}$&$\mathit{43}$ & $34$&$5$\\
-    LSHEP & $0$&$00254$ & $2$&$93$ & $7$&$16$ & $13$&$1$ & $\mathit{29}$&$\mathit{0}$\\
-    Voronoi & $\mathbf{0}$&$\mathbf{0}$ & $1$&$29$ & $3$&$52$ & $6$&$87$ & $30$&$1$\\
-    BoxSpline & $0$&$00287$ & $4$&$39$ & $9$&$10$ & $14$&$8$ & $41$&$3$\\
-    \hline
-  \end{tabular}
-  \caption{This numerical data accompanies the visual provided in
-    Figure \ref{fig:error-parkinsons}. The columns of absolute error
-    percentiles correspond to the minimum, first quartile, median,
-    third quartile, and maximum absolute errors respectively. The
-    minimum of each column is boldface, while the second lowest value
-    is italicized. All values are rounded to three significant
-    digits.}
-  \label{table:error-parkinsons}
-\end{table}
-
-\begin{table}
-  \centering
-  \begin{tabular}{|c|r@{.}l| c |r@{.}l|r@{.}l|r@{.}l|}
-    \cline{1-3}\cline{5-10}
-    Algorithm & \multicolumn{2}{c|}{\% Best} &  & \multicolumn{2}{c|}{Fit/Prep. Time (s)} & \multicolumn{2}{c|}{App. Time (s)} & \multicolumn{2}{c|}{Total App. Time (s)}\\
-    \cline{1-3}\cline{5-10}
-    MARS & \quad\,$0$&$0$ &  & \quad\quad\quad$37$&$9$ & \quad\,\,$0$&$00253$ & \quad\quad\quad\,\,$1$&$48$\\
-    SVR & $0$&$1$ &  & $\mathbf{0}$&$\mathbf{859}$ & $\mathit{0}$&$\mathit{000181}$ & $\mathit{0}$&$\mathit{106}$\\
-    MLP & $\mathit{32}$&$\mathit{0}$ &  & $348$&$0$ & $0$&$00111$ & $0$&$653$\\
-    Delaunay & $0$&$0$ &  & $2$&$47$ & $1$&$22$ & $720$&$0$\\
-    ShepMod & $\mathbf{66}$&$\mathbf{4}$ &  & $\mathit{1}$&$\mathit{13}$ & $0$&$000182$ & $0$&$107$\\
-    LSHEP & $0$&$0$ &  & $2$&$39$ & $\mathbf{0}$&$\mathbf{000144}$ & $\mathbf{0}$&$\mathbf{0845}$\\
-    Voronoi & $1$&$6$ &  & $2$&$77$ & $0$&$0274$ & $16$&$1$\\
-    BoxSpline & $0$&$0$ &  & $1$&$26$ & $0$&$000643$ & $0$&$377$\\
-    \cline{1-3}\cline{5-10}
-  \end{tabular}
-  \caption{The left above shows how often each algorithm had the
-    lowest absolute error approximating Parkinson's data in Table
-    \ref{table:error-parkinsons}. On the right columns are median fit
-    time of 5288 points, median time for one approximation, and median
-    time approximating 587 points.}
-  \label{table:best-parkinsons}
-\end{table}
-
-\begin{table}
-  \centering
-  \begin{tabular}{c|r@{.}l|r@{.}l|r@{.}l|r@{.}l|r@{.}l}
-    \hline
-    Algorithm & \multicolumn{2}{c|}{Min} & \multicolumn{2}{c|}{$25^{th}$} & \multicolumn{2}{c|}{$50^{th}$} & \multicolumn{2}{c|}{$75^{th}$} & \multicolumn{2}{c}{Max}\\
-    \hline
-    MARS & $\mathit{6}$&$\mathit{45 \times 10^{-15}}$ & $\mathbf{2}$&$\mathbf{70 \times 10^{-14}}$ & $1$&$66$ & $1$&$96$ & $\mathit{53}$&$\mathit{3}$\\
-    SVR & $0$&$0420$ & $0$&$143$ & $0$&$148$ & $\mathit{0}$&$\mathit{860}$ & $119$&$0$\\
-    MLP & $0$&$0000689$ & $0$&$0337$ & $\mathbf{0}$&$\mathbf{0795}$ & $\mathbf{0}$&$\mathbf{264}$ & $\mathbf{5}$&$\mathbf{31}$\\
-    Delaunay & $\mathbf{0}$&$\mathbf{0}$ & $0$&$187$ & $0$&$919$ & $2$&$72$ & $56$&$3$\\
-    ShepMod & $\mathbf{0}$&$\mathbf{0}$ & $0$&$0957$ & $0$&$685$ & $2$&$90$ & $73$&$2$\\
-    LSHEP & $\mathbf{0}$&$\mathbf{0}$ & $\mathit{0}$&$\mathit{0153}$ & $\mathit{0}$&$\mathit{106}$ & $1$&$17$ & $113$&$0$\\
-    Voronoi & $\mathbf{0}$&$\mathbf{0}$ & $0$&$430$ & $1$&$28$ & $2$&$94$ & $83$&$8$\\
-    BoxSpline & $\mathbf{0}$&$\mathbf{0}$ & $0$&$789$ & $2$&$06$ & $5$&$25$ & $119$&$0$\\
-    \hline
-  \end{tabular}
-  \caption{This numerical data accompanies the visual provided in
-    Figure \ref{fig:error-weather}. The columns of absolute error
-    percentiles correspond to the minimum, first quartile, median,
-    third quartile, and maximum absolute errors respectively. The
-    minimum value of each column is boldface, while the second lowest
-    is italicized. All values are rounded to three significant
-    digits.}
-  \label{table:error-weather}
-\end{table}
-
-\begin{table}
-  \centering
-  \begin{tabular}{|c|r@{.}l| c |r@{.}l|r@{.}l|r@{.}l|}
-    \cline{1-3}\cline{5-10}
-    Algorithm & \multicolumn{2}{c|}{\% Best} &  & \multicolumn{2}{c|}{Fit/Prep. Time (s)} & \multicolumn{2}{c|}{App. Time (s)} & \multicolumn{2}{c|}{Total App. Time (s)}\\
-    \cline{1-3}\cline{5-10}
-    MARS & \,\,\,\,$10$&$7$ &  & \quad\quad\quad$\mathit{0}$&$\mathit{151}$ & \quad$0$&$000117$ & \quad\quad\quad\,\,$0$&$0304$\\
-    SVR & $4$&$1$ &  & $\mathbf{0}$&$\mathbf{133}$ & $\mathbf{0}$&$\mathbf{0000947}$ & $\mathbf{0}$&$\mathbf{0246}$\\
-    MLP & $\mathbf{56}$&$\mathbf{8}$ &  & $169$&$0$ & $0$&$00137$ & $0$&$356$\\
-    Delaunay & $0$&$1$ &  & $0$&$664$ & $0$&$886$ & $230$&$0$\\
-    ShepMod & $3$&$5$ &  & $0$&$265$ & $0$&$000128$ & $0$&$0333$\\
-    LSHEP & $\mathit{28}$&$\mathit{4}$ &  & $0$&$874$ & $\mathit{0}$&$\mathit{0000975}$ & $\mathit{0}$&$\mathit{0254}$\\
-    Voronoi & $0$&$2$ &  & $0$&$675$ & $0$&$0270$ & $7$&$01$\\
-    BoxSpline & $0$&$3$ &  & $0$&$330$ & $0$&$000406$ & $0$&$106$\\
-    \cline{1-3}\cline{5-10}
-  \end{tabular}
-  \caption{Left table shows how often each algorithm had the lowest
-    absolute error approximating Sydney rainfall data in Table
-    \ref{table:error-weather}. On the right columns are median fit
-    time of 2349 points, median time for one approximation, and median
-    time approximating 260 points.}
-  \label{table:best-weather}
-\end{table}
-
-\begin{table}
-  \centering
-  \begin{tabular}{c|r@{.}l|r@{.}l|r@{.}l|r@{.}l|r@{.}l}
-    \hline
-    Algorithm & \multicolumn{2}{c|}{Min} & \multicolumn{2}{c|}{$25^{th}$} & \multicolumn{2}{c|}{$50^{th}$} & \multicolumn{2}{c|}{$75^{th}$} & \multicolumn{2}{c}{Max}\\
-    \hline
-    MARS & $5$&$36$ & $3610$&$0$ & $4580$&$0$ & $5450$&$0$ & $13400$&$0$\\
-    SVR & $0$&$0138$ & $8$&$47$ & $15$&$4$ & $41$&$9$ & $7700$&$0$\\
-    MLP & $0$&$00151$ & $\mathbf{1}$&$\mathbf{60}$ & $\mathbf{3}$&$\mathbf{86}$ & $\mathbf{8}$&$\mathbf{34}$ & $\mathbf{604}$&$\mathbf{0}$\\
-    Delaunay & $\mathbf{0}$&$\mathbf{0}$ & $\mathit{2}$&$\mathit{69}$ & $\mathit{13}$&$\mathit{8}$ & $\mathit{35}$&$\mathit{0}$ & $\mathit{4840}$&$\mathit{0}$\\
-    ShepMod & $\mathbf{0}$&$\mathbf{0}$ & $4$&$21$ & $17$&$3$ & $51$&$6$ & $6510$&$0$\\
-    LSHEP & $1$&$27$ & $199$&$0$ & $260$&$0$ & $343$&$0$ & $6530$&$0$\\
-    Voronoi & $\mathit{2}$&$\mathit{89 \times 10^{-10}}$ & $14$&$7$ & $32$&$8$ & $52$&$9$ & $4860$&$0$\\
-    BoxSpline & $0$&$00740$ & $20$&$8$ & $40$&$9$ & $79$&$6$ & $7640$&$0$\\
-    \hline
-  \end{tabular}
-  \caption{This numerical data accompanies the visual provided in
-    Figure \ref{fig:error-credit-card}. The columns of absolute error
-    percentiles correspond to the minimum, first quartile, median,
-    third quartile, and maximum absolute errors respectively. The
-    minimum value of each column is boldface, while the second lowest
-    is italicized. All values are rounded to three significant
-    digits.}
-  \label{table:error-credit-card}
-\end{table}
-
-\begin{table}
-  \centering
-  \begin{tabular}{|c|r@{.}l| c |r@{.}l|r@{.}l|r@{.}l|}
-    \cline{1-3}\cline{5-10}
-    Algorithm & \multicolumn{2}{c|}{\% Best} &  & \multicolumn{2}{c|}{Fit/Prep. Time (s)} & \multicolumn{2}{c|}{App. Time (s)} & \multicolumn{2}{c|}{Total App. Time (s)}\\
-    \cline{1-3}\cline{5-10}
-    MARS & \quad\,$0$&$0$ &  & \quad\quad\quad$22$&$0$ & \quad$0$&$00148$ & \quad\quad\quad\,\,$0$&$820$\\
-    SVR & $0$&$0$ &  & $\mathbf{1}$&$\mathbf{01}$ & $\mathit{0}$&$\mathit{000210}$ & $\mathit{0}$&$\mathit{117}$\\
-    MLP & $\mathbf{79}$&$\mathbf{5}$ &  & $290$&$0$ & $0$&$000714$ & $0$&$397$\\
-    Delaunay & $\mathit{20}$&$\mathit{5}$ &  & $3$&$12$ & $3$&$71$ & $2070$&$0$\\
-    ShepMod & $0$&$1$ &  & $\mathit{1}$&$\mathit{4}5$ & $0$&$000220$ & $0$&$122$\\
-    LSHEP & $0$&$0$ &  & $3$&$47$ & $\mathbf{0}$&$\mathbf{000176}$ & $\mathbf{0}$&$\mathbf{0981}$\\
-    Voronoi & $0$&$0$ &  & $3$&$32$ & $0$&$0950$ & $52$&$8$\\
-    BoxSpline & $0$&$0$ &  & $1$&$66$ & $0$&$000956$ & $0$&$532$\\
-    \cline{1-3}\cline{5-10}
-  \end{tabular}
-  \caption{The left above shows how often each algorithm had the
-    lowest absolute error approximating credit card transaction data
-    in Table \ref{table:error-credit-card}. On the right columns are
-    median fit time of 5006 points, median time for one approximation,
-    and median time approximating 556 points.}
-  \label{table:best-credit-card}
-\end{table}
-
-\begin{table}
-  \centering
-  \begin{tabular}{c|r@{.}l|r@{.}l|r@{.}l|r@{.}l|r@{.}l}
-    \hline
-    Algorithm & \multicolumn{2}{c|}{Min} & \multicolumn{2}{c|}{$25^{th}$} & \multicolumn{2}{c|}{$50^{th}$} & \multicolumn{2}{c|}{$75^{th}$} & \multicolumn{2}{c}{Max}\\
-    \hline
-    Delaunay & $\mathbf{0}$&$\mathbf{0287}$ & $\mathbf{0}$&$\mathbf{0914}$ & $\mathbf{0}$&$\mathbf{158}$ & $\mathit{0}$&$\mathit{308}$ & $\mathit{0}$&$\mathit{897}$\\
-    ShepMod & $0$&$0307$ & $\mathit{0}$&$\mathit{0983}$ & $\mathit{0}$&$\mathit{164}$ & $0$&$335$ & $1$&$00$\\
-    Voronoi & $\mathit{0}$&$\mathit{0303}$ & $0$&$100$ & $0$&$165$ & $\mathbf{0}$&$\mathbf{274}$ & $\mathbf{0}$&$\mathbf{762}$\\
-    BoxSpline & $0$&$0703$ & $0$&$417$ & $0$&$561$ & $0$&$745$ & $1$&$00$\\
-    \hline
-  \end{tabular}
-  \caption{This numerical data accompanies the visual provided in
-    Figure \ref{fig:error-throughput}. The columns of absolute error
-    percentiles correspond to the minimum, first quartile, median,
-    third quartile, and maximum KS statistics respectively between
-    truth and guess for models predicting the distribution of I/O
-    throughput that will be observed at previously unseen system
-    configurations. The minimum value of each column is boldface,
-    while the second lowest is italicized. All values are rounded to
-    three significant digits.}
-  \label{table:error-throughput}
-\end{table}
-
-\begin{table}
-  \centering
-  \begin{tabular}{|c|r@{.}l| c |r@{.}l|r@{.}l|r@{.}l|}
-    \cline{1-3}\cline{5-10}
-    Algorithm & \multicolumn{2}{c|}{\% Best} &  & \multicolumn{2}{c|}{Fit/Prep. Time (s)} & \multicolumn{2}{c|}{App. Time (s)} & \multicolumn{2}{c|}{Total App. Time (s)}\\
-    \cline{1-3}\cline{5-10}
-    Delaunay & \,\,$\mathbf{62}$&$\mathbf{0}$ &  & \quad\quad\quad$0$&$344$ & \quad$0$&$00984$ & \quad\quad\,\,$2$&$71$\\
-    ShepMod & $0$&$0$ &  & $\mathbf{0}$&$\mathbf{0884}$ & $\mathbf{0}$&$\mathbf{000145}$ & $\mathbf{0}$&$\mathbf{0436}$\\
-    Voronoi & $\mathit{38}$&$\mathit{0}$ &  & $0$&$360$ & $0$&$00253$ & $0$&$762$\\
-    BoxSpline & $0$&$0$ &  & $\mathit{0}$&$\mathit{0972}$ & $\mathit{0}$&$\mathit{000210}$ & $\mathit{0}$&$\mathit{0633}$\\
-    \cline{1-3}\cline{5-10}
-  \end{tabular}
-  \caption{The left above shows how often each algorithm had the
-    lowest KS statistic on the I/O throughput distribution data in
-    Table \ref{table:error-throughput}. On the right columns are
-    median fit time of 2715 points, median time for one approximation,
-    and median time approximating 301 points.}
-  \label{table:best-throughput}
-\end{table}
-
-\end{appendix}
-
-\end{document}

### 2020-07-28 3-Disseration/apps.tex
Updated thesis document, sent draft to Dr. Watson.

@@
 \citet{grobelny2007fase} refer to statistical models as being oversimplified and not capable of capturing the true complexity of the underlying system.
-Historically, statistical models have been limited to modelling at most one or two system parameters and have therefore not been capable of modeling the complexity of the underlying system.
-\cite{snavely2002framework,bailey2005performance,barker2009using,ye2010analyzing}.
+Historically, statistical models have been limited to modelling at most one or two system parameters and have therefore not been capable of modeling the complexity of the underlying system \cite{snavely2002framework,bailey2005performance,barker2009using,ye2010analyzing}.
 These limited statistical models have provided satisfactory performance in narrow application settings.
@@
 splines \cite{de2013box}, of which one of the approximation techniques
-in this work is composed \cite{lux2018novel}.
+in Chapter \ref{ch:boxes} of this work is composed \cite{lux2018novel}.
 
@@
 
-Predicting the performance of an HPC system is a challenging problem
-that is primarily attempted in one of two ways: (1) build a
-statistical model of the performance by running experiments on the
-system at select settings, or (2) run artificial experiments using a
-simplified simulation of the target system to estimate architecture
-and application bottlenecks. In this paper the proposed multivariate
-modeling techniques rest in the first category, and they represent a
-notable increase in the ability to model complex interactions between
-system parameters.
-
-Many previous works attempting to model system performance have used
-simulated environments to estimate the performance of a system
-\cite{grobelny2007fase,wang2009simulation,wang2013towards}. Some
-of these works refer to statistical models as being oversimplified and
-not capable of capturing the true complexity of the underlying
-system. This claim is partially correct, noting that a large portion
-of predictive statistical models rely on simplifying the model to one
-or two parameters
-\cite{snavely2002framework,bailey2005performance,barker2009using,ye2010analyzing}.
-These limited statistical models have provided satisfactory
-performance in very narrow application settings. Many of the
-aforementioned statistical modeling techniques claim to generalize,
-while simultaneously requiring additional code annotations, hardware
-abstractions, or additional application level understandings in order
-to generate models. The approach presented here requires no
-modifications of the application, no architectural abstractions, nor
-any structural descriptions of the input data being modeled. The
-techniques used are purely mathematical and only need performance data
-as input.
-
-Among the statistical models presented in prior works,
+Among the statistical models presented in prior works for modeling computer systems,
 \cite{bailey2005performance} specifically mention that it is difficult
-for the simplified models to capture variability introduced by
+for simplified models to capture variability introduced by
 I/O. System variability in general has become a problem of increasing

### 2020-07-28 3-Disseration/boxes.tex
Updated thesis document, sent draft to Dr. Watson.

@@
 \section{Voronoi Mesh}
+
+\begin{figure}
+  \centering
+  \includegraphics[width=0.6\textwidth,]{Figures/NA/VM_Construction.pdf}
+  \caption{Above is a depiction of the Voronoi cell
+    boundaries (dashed lines) about a set of interpolation points
+    (dots) in two dimensions. In this example, the Voronoi mesh basis
+    function about the center most point has nonzero weight in the
+    shaded region and transitions from a value of one at the point to
+    zero at the boundary of the twice expanded Voronoi cell (solid
+    line).
+  \vspace{-.1cm}}
+  \label{fig:voronoi-basis}
+\end{figure}
 

### 2020-07-28 3-Disseration/error.tex
Updated thesis document, sent draft to Dr. Watson.

@@
-This chapter presents the theoretical results bounding the error of
+This chapter presents theoretical results bounding the error of
 (piecewise) linear interpolation. The error analysis relies on linear
@@
 to Delaunay interpolation; and (3) multiple other interpolants in this
-work compute predictions as convex combinations of observed function
-values, which may allow for straight forward extensions of this error
+paper compute predictions as convex combinations of observed function
+values, which may allow for straightforward extensions of this error
 bound.
 
-\begin{lemma}
+\begin{plainlemma}
   \label{lemma:1}
-  Let $S \subset \mathbb{R}^d$ be open and convex, $f: \mathbb{R}^d
-  \rightarrow \mathbb{R}$, and $\nabla f \in
-  Lip_{(\gamma,\|\cdot\|_2)}(S)$, the set of $\gamma$-Lipschitz
-  continuous functions in the $2$-norm. Then for all $x,y \in S$
+  Let $S \subset \mathbb{R}^d$ be open and convex, $f: S \rightarrow
+  \mathbb{R}$, and let $\nabla f(x)$ be $\gamma$-Lipschitz continuous
+  in the $2$-norm. Then for all $x,y \in S$
   $$\big|f(y) - f(x) - \langle \nabla f(x), y - x \rangle \big| \leq \frac{\gamma \|y - x\|_2^2}{2}.$$
-\end{lemma}
+\end{plainlemma}
 
-\begin{proof}
+\begin{proofdot}
   Consider the function $g(t) = f \big((1-t) x + t y \big)$, $0 \leq t
@@
   \begin{align*}
-    \big|f(y) - f(x) - \langle \nabla f(x), y - x \rangle \big|
-       &= \big|g(1) - g(0) - g'(0) \big| & \\
-       &= \bigg| \int_0^1 g'(t) - g'(0)\ dt \bigg| \\
-       &\leq \int_0^1 \big|g'(t) - g'(0)\big|\ dt \\
+    \big|f(y) - f(x) - \langle &\nabla f(x), y - x \rangle \big|
+        = \big|g(1) - g(0) - g'(0) \big| \\
+       &= \bigg| \int_0^1 g'(t) - g'(0)\ dt \bigg| \leq \int_0^1 \big|g'(t) - g'(0)\big|\ dt \\
        &= \int_0^1 \bigg| \big \langle \nabla f\big((1-t)x + ty\big) - \nabla f(x), y - x \big \rangle \bigg|\ dt \\
        &\leq \int_0^1 \big \| \nabla f\big((1-t)x + ty\big) - \nabla f(x) \big \|_2\ \| y - x \|_2\ dt \\
-       &\leq \int_0^1 \big ( \gamma\ \|y-x\|_2 \big) \ \big( \|y-x\|_2 \big) t\ dt \\
-       &= \frac{\gamma \|y - x\|_2^2}{2}.
+       &\leq \int_0^1 \big ( \gamma\ \|y-x\|_2 \big) \ \big( \|y-x\|_2 \big) t\ dt = \frac{\gamma \|y - x\|_2^2}{2}.
   \end{align*}
-\end{proof}
+  \qed
+\end{proofdot}
 
-\begin{lemma}
+\hfill
+
+\begin{plainlemma}
   \label{lemma:2}
@@
   where $\sigma_d$ is the smallest singular value of $M.$
-\end{lemma}
+\end{plainlemma}
 
-\begin{proof}
+\begin{proofdot}
   Using the facts that $M$ and $M^t$ have the same singular values,
@@
   \end{align*}
-\end{proof}
+  \qed
+\end{proofdot}
 
-\begin{lemma}
+%% \newpage
+
+\begin{plainlemma}
   \label{lemma:3}
@@
   \max\limits_{1\ \leq\ j\ \leq\ d} \|x_j - x_0\|_2.$ Then
-  $$\big\|\nabla f(x_0) - c\big\|_2 \leq \sqrt{d} \frac{\gamma k^2}{\sigma_d}.$$
-\end{lemma}
+  $$\big\|\nabla f(x_0) - c\big\|_2 \leq \frac{\sqrt{d} \, \gamma k^2}{2 \sigma_d}.$$
+\end{plainlemma}
 
-\begin{proof}
-  Consider $f(x) - \hat f(x)$ along the line segment $z(t) = (1-t)x_0
-  + t x_j$, $0 \leq t \leq 1.$ By Rolle's Theorem, for some $0 < \hat
-  t < 1$, $\big\langle \nabla f\big(z(\ \hat t\ )\big) - c, x_j - x_0
-  \big\rangle = 0.$ Now
+\begin{proofdot}
+  \def\g{\gamma_g}
+  Consider $g(t)=f(x(t)) - \hat f(x(t))$ along
+  the line segment $x(t)=(1-t)x_0 + tx_j$, $0\le t\le1$, from $x_0$ to
+  $x_j$.  Observe that $g(0)=f(x_0) - \hat f(x_0) = 0$, $g(1)=f(x_j) -
+  \hat f(x_j) = 0$, and $g'(t)$ is $\g$-Lipschitz continuous with $\g
+  = \gamma \|x_j-x_0\|_2^2$. The following is visualized in 
+  Figure \ref{fig:lemma3}.
+
+  Suppose $g'(0) > \g/2$. Then $|g'(0)-g'(t)| \le \g t \implies g'(0)
+  - \g t \le g'(t)$, and the line $w=g'(0)-\g t$ intersects the
+  $t$-axis at $\tilde t=g'(0)/\g > 1/2$. $\tilde t<1$ necessarily,
+  since by Rolle's Theorem there exists $0<z<1$ such that $g'(z)=0$
+  and so $g'(0) - \g z \le g'(z) = 0$. Now by integrating
+  $$\frac{g'(0)^2}{2\g} = \frac{g'(0)\tilde t}{2} = \int_0^{\tilde t}
+  (g'(0)-\g t)dt \le \int_0^{\tilde t} g'(t)dt = g(\tilde t),$$ and
+  using $g'(0) > \g / 2$
   \begin{align*}
-    \big| \big< \nabla f(x_0) - c, x_j - x_0 \big \rangle \big|
-        &= \big| \big< \nabla f(x_0) - \nabla f\big(z(\ \hat t\ )\big) + \nabla f\big(z(\ \hat t\ )\big) - c, x_j - x_0 \big \rangle \big| \\
-	&= \big| \big\langle \nabla f(x_0) - \nabla f \big(z(\ \hat t\ )\big), x_j - x_0 \big \rangle \big| \\
-	&\leq \big \| \nabla f(x_0) - \nabla f\big(z(\ \hat t\ )\big) \big \|_2 \|x_j - x_0\|_2 \\
-	&\leq \gamma \|x_0 - z(\ \hat t\ )\|_2\ \|x_j - x_0\|_2 \\
-	&\leq \gamma \|x_j - x_0\|_2^2 \leq \gamma k^2, \\
+    \frac{-g'(0)^2}{2\g} < g'(0) - \frac{g'(0)^2}{2\g} - \g/2 =
+    \frac{(1-\tilde t)(g'(0)-\g)}{2} &= \int_{\tilde t}^1 (g'(0)-\g
+    t)dt \\ &\le \int_{\tilde t}^1 g'(t)dt = -g(\tilde t),
   \end{align*}
-  for all $1 \leq j \leq d.$ Using {\it Lemma \ref{lemma:2}},
-  $$\big \| \nabla f(x_i) - c \big \|_2^2 \leq \frac{d}{\sigma_d^2}
-  \big( \gamma k^2\big)^2 \implies \big \| \nabla f(x_i) - c \big \|_2
-  \leq \sqrt{d} \frac{\gamma k^2}{\sigma_d}.$$
-\end{proof}
+  \noindent a contradiction for the value of $g(\tilde t)$. A similar
+  contradiction arises for $g'(0)<-\g/2$. Therefore $|g'(0)| \le
+  \g/2$. In terms of $f$,
+  $$\bigl|\langle \nabla f(x_0)-c,x_j-x_0\rangle\bigr| = |g'(0)| \le
+  \gamma \|x_j-x_0\|_2^2/2 \le \gamma k^2 / 2,$$ which holds for all
+  $1\le j\le d$.  Finally, using {\it Lemma \ref{lemma:2}},
+  $$\|\nabla f(x_0)-c\|_2^2 \le \frac{d}{\sigma_d^2}(\gamma k^2/2)^2
+  \implies \|\nabla f(x_0)-c\|_2 \le \frac{\sqrt{d} \, \gamma
+    k^2}{2\sigma_d}.$$ \qed
+\end{proofdot}
 
-\begin{theorem}
+\begin{figure}
+  \centering
+  \includegraphics[width=0.9\textwidth,]{Figures/NA/example_lemma_3.pdf}
+  \def\g{\gamma_g}
+  \caption{Three different scenarios visualizing {\it
+      Lemma \ref{lemma:3}}, where $g(t)$ is the difference between a
+    piecewise linear interpolant and the approximated function along a
+    normalized line segment between interpolation points, $g'(t)$ is
+    $\g$-Lipschitz continuous, and $w$ and $\tilde t$ are defined in
+    the proof. Leftmost is a randomly chosen permissible shape of $g$
+    and $g'$. The middle is the only possible shape of $g$ and $g'$
+    given $g'(0) = \g / 2$, establishing the case of equality in the
+    lemma. Rightmost is the resulting contradiction when $g'(0) > \g /
+    2$, notice it is impossible to ensure $g'(t)$ is $\g$-Lipschitz
+    continuous and satisfy $g(1) = 0$ (highlighted with red circle on
+    the right).
+  \vspace{-.1cm}}
+  \label{fig:lemma3}
+\end{figure}
+
+\begin{onetheorem}
   Under the assumptions of {\it Lemma \ref{lemma:1}} and {\it Lemma
     \ref{lemma:3}}, for $z \in S$,
-  $$ \big|f(z) - \hat f(z)\big| \leq \frac{\gamma \|z - x_0\|_2^2}{2} + \sqrt{d} \frac{\gamma k^2}{\sigma_d} \|z - x_0\|_2.$$
-\end{theorem}
+  $$ \big|f(z) - \hat f(z)\big| \leq \frac{\gamma \|z - x_0\|_2^2}{2} + \frac{\sqrt{d} \, \gamma k^2}{2 \sigma_d} \|z - x_0\|_2.$$
+\end{onetheorem}
 
-\begin{proof}
-  Let $v = \nabla f(x_0) - c$, where $\|v\|_2 \leq \sqrt{d} \gamma k^2 / \sigma_d$ by {\it Lemma \ref{lemma:3}}. Now
+\begin{proofdot}
+  Let $v = \nabla f(x_0) - c$, where $\|v\|_2 \leq \sqrt{d}\, \gamma k^2 / (2 \sigma_d)$ by {\it Lemma \ref{lemma:3}}. Now
   \begin{align*}
@@
        &\leq \big|f(z) - f(x_0) - \langle \nabla f(x_0) , z - x_0 \rangle \big| + \|v\|_2 \|z - x_0\|_2 \\
-       &\leq \big|f(z) - f(x_0) - \langle \nabla f(x_0), z - x_0 \rangle \big| + \textstyle{\frac{\gamma k^2 \sqrt{d}}{\sigma_d}} \|z - x_0\|_2 \\
-       &\leq \frac{\gamma \|z - x_0\|_2^2}{2} + \sqrt{d}\frac{\gamma k^2}{\sigma_d} \|z - x_0\|_2,
+       &\leq \big|f(z) - f(x_0) - \langle \nabla f(x_0), z - x_0 \rangle \big| + \big(\sqrt{d}\,\gamma k^2 / (2 \sigma_d)\big) \|z - x_0\|_2 \\
+       &\leq \frac{\gamma \|z - x_0\|_2^2}{2} + \frac{\sqrt{d}\,\gamma k^2}{2 \sigma_d} \|z - x_0\|_2,
   \end{align*}
   where the last inequality follows from {\it Lemma \ref{lemma:1}}.
-\end{proof}
+  %
+  \qed
+\end{proofdot}
 
@@
 data only when the diameter of the simplex is also reduced at a
-proportional rate.  Only linear convergence to the true function can
-be achieved in practice, without the incorporation of additional
-observations. Notice that the approximation error is largely
-determined by the spacing of observed data. Predictions made by
-simplices whose vertices are not well-spaced (i.e., have large
-diameter, or are nearly contained in a hyperplane) have higher
-error. In light of this error bound, an empirical evaluation of
-presented algorithms follows.
+proportional rate. Only linear convergence to the true function
+is achieved in practice without the
+incorporation of additional observations, by moving closer to
+  interpolation points. Notice that the approximation error is
+largely determined by the spacing of observed data. Predictions made
+by simplices whose vertices are not well-spaced (i.e., have large
+diameter, or are nearly contained in a hyperplane) have higher error.
+
+That the theoretical error bound is sharp can be observed with the
+test function $q(x) = \|x\|_2^2,$ $x \in \mathbb{R}^d,$ and the
+simplex defined by vertices $X = \{0, e_1, \ldots, e_d\},$ where $e_i$
+is the $i$-th standard basis vector in $\mathbb{R}^d,$ $e = (1$,
+$\ldots$, $1) \in \mathbb{R}^d,$ and $0$ denotes the zero vector in
+any dimension.  The constants relevant to the error bound are
+$$ \gamma = 2, \qquad \sigma_d = 1, \qquad k = 1, \qquad x_0 = 0,
+\qquad \hat q(x) = \langle e, x - 0\rangle + q(0). $$
+\noindent Noting that $q(0) = 0,$ the approximation error at $z =
+-(1/2)e$ is
+$$ \big|q(z) - \hat q(z)\big| = \big|\|z\|_2^2 - \langle e,z
+\rangle\big| = \big| d/4 + d/2 \big| = 3d/4, $$
+\noindent while the error bound from the theorem gives
+
+\vspace{-2mm}
+\begin{align*}
+  \big|q(z) - \hat q(z)\big|
+  &\leq \frac{\gamma \|z - x_0\|_2^2}{2} + \frac{\sqrt{d}\,\gamma k}
+      {2 \sigma_d} \|z - x_0\|_2 \\
+  &{}=  \|z\|_2^2 + \sqrt{d} \|z\|_2 \\
+  &{}=  d/4 + d/2 = 3d/4.
+\end{align*}
+\vspace{-2mm}
+
+\noindent Acknowledging that the error bound is sharp, it may be of
+interest to observe the error of piecewise linear approximation
+techniques on an analytic test function.
+
+\begin{figure}
+  \centering
+  \begin{tikzpicture}
+    \node (img)  {\includegraphics[width=0.9\textwidth,]{Figures/NA/Fig2.pdf}};
+    \node[below=of img, node distance=1cm, yshift=1cm] {Number of Data Points};
+    \node[left=of img, node distance=0cm, rotate=90, anchor=center,yshift=-0.7cm] {Absolute Error};
+  \end{tikzpicture}
+  \caption{Delaunay and MLP approximations are constructed from Fekete
+    points over the unit cube evaluating the test function $f(x) =
+    \cos(\|x\|_2)$ for $x \in \mathbb{R}^2$. The figure shows the
+    first/third quartiles at the box bottom/top, the second quartile
+    (median) at the white bar, median 95\% confidence interval (cones,
+    barely visible in figure), and whiskers at 3/2 of the adjacent
+    interquartile ranges, for the absolute prediction error for each
+    model at $1000$ random evaluation points. The left plot observes a
+    perfect interpolation problem with exact evaluations of $f.$ The
+    right plot observes a regression problem with uniform random noise
+    giving values in $[.9 f(x),\ 1.1f(x)]$ for each $x.$ Both axes are
+    log scaled.}
+  \label{fig:convergence-2d}
+\end{figure}
+
+%% ===================================================================
+\vspace{-2mm}
+\section{Demonstration on an Analytic Test Function}
+\label{sec:analytic}
+
+The theoretical results constructed in Section \ref{sec:theory} for
+(piecewise) linear interpolation are promising and apply directly to
+Delaunay interpolation, however they are difficult to interpret in
+context with approximation algorithms that do not have similar known
+uniform error bounds. For that reason, an analytic function is used to
+measure the error of a piecewise linear interpolant (Delaunay) and a
+piecewise linear regressor (the MLP) when provided an increasing
+amount of data with varying levels of random noise. Only
+Delaunay and the MLP are considered in this demonstration because
+all other mentioned techniques are not strictly piecewise linear
+approximations.
+
+The test function chosen here for analysis resembles the
+\textit{oscillatory} function used by \cite{barthelmann2000high}.
+However, a slight modification is made to remove the simple dot
+product structure (which is favorable for the MLP). Let $f(x)
+=\cos(\|x\|_2)$ for $x \in \mathbb{R}^d$ where $d$ is either $2$
+(Fig. \ref{fig:convergence-2d}) or $20$
+(Fig. \ref{fig:convergence-20d}). This function has a bounded change
+in gradient $2$-norm and hence meets the necessary Lipschitz condition
+for the error bound. Data points and approximation points for this
+test will be within the unit cube $[0,1]^d$.
+
+The error of a linear interpolant constructed from well spaced points
+is dominated by the distance to the nearest data point. In order to
+uniformly decrease the distance to the nearest data point across the
+unit cube, exponentially more data points must be available for
+approximation. For this experiment $N = 2(4^i),$ $i = 1,$ $\ldots,$
+$6,$ chosen points are kept well spaced by computing approximate
+Fekete points. The following experiment was also run with a
+  Latin hypercube design \cite{park1994optimal}, which produced almost
+  identical results that are not reported here. Fekete points have
+a history in potential theory \cite{kovari_pommerenke_1968} and are
+most generally defined as those points that maximize the absolute
+value of a Vandermonde determinant. Here, the QR method outlined in
+\cite{bos2010computing} is used to identify approximate Fekete points
+from a Vandermonde matrix. A multivariate polynomial basis of size
+$2N$ is constructed containing all polynomials of degree $\leq n$ for
+the largest $n$ such that ${d+n \choose n} \leq 2N$ and $2N - {d+n
+  \choose n}$ arbitrarily selected polynomials of degree $n + 1.$ The
+Vandermonde matrix for these $2N$ polynomials is used to select $N$
+approximate Fekete points.
+
+Finally an additional aspect is added to this test problem by
+incorporating random uniform noise into the evaluations of $f.$ For
+each test two experiments are executed, one with exact function
+evaluations (an interpolation problem) and one with a
+constant signal-to-noise ratio (SNR) of 10:1 (a regression
+  problem).
+
+\begin{figure}
+  \centering
+  \begin{tikzpicture}
+    \node (img)  {\includegraphics[width=0.9\textwidth,]{Figures/NA/Fig3.pdf}};
+    \node[below=of img, node distance=1cm, yshift=1cm] {Number of Data Points};
+    \node[left=of img, node distance=0cm, rotate=90, anchor=center,yshift=-0.7cm] {Absolute Error};
+  \end{tikzpicture}
+  \caption{Delaunay and MLP approximations are constructed from Fekete
+    points over the unit cube evaluating the test function $f(x) =
+    \cos(\|x\|_2)$ for $x \in \mathbb{R}^{20}$. The details are the
+    same as for Fig. \ref{fig:convergence-2d}.}
+  \label{fig:convergence-20d}
+\end{figure}
+
+The bound from the theorem suggests that by increasing the number of
+well-spaced data points, approximation error can be reliably
+decreased. The $d = 2$ test seen on the left half of Figure
+\ref{fig:convergence-2d} shows a consistent decrease in error for
+Delaunay and also shows the eventual accuracy plateau obtained by a
+parametric regression form (the MLP, at roughly 500 points). On the
+right hand side of Figure \ref{fig:convergence-2d} the random noise
+clearly prohibits Delaunay from converging to $f$, while the MLP is
+able to improve its approximation with more data points on average.
+The convergence result for a very low-dimensional problem like this is
+expected.  However, intuition fails for higher dimensional problems.
+
+Figure \ref{fig:convergence-20d} shows the test function with $d = 20$
+presents a significantly more challenging approximation problem than
+its counterpart in low dimension. The same increase in number of data
+points from 32 to 8192 causes no apparent improvement in approximation
+for either the noise-free or the noisy problems. Perhaps unexpectedly,
+the interpolation technique performs better than the regression
+technique on the noisy data (right) in Figure
+\ref{fig:convergence-20d}, and worse than the regression technique on
+the noise-free data (left). This result emphasizes the relevance of
+interpolation for problems in high dimension. It also reveals that the
+outcome of MLP regressions can get worse when adding more data points.
+
+\begin{figure}
+  \centering
+  \begin{tikzpicture}
+    \node (img)  {\includegraphics[width=0.9\textwidth,]{Figures/NA/theorem-convergence.pdf}};
+    \node[below=of img, node distance=1cm, yshift=1cm] {Number of Data Points};
+    \node[left=of img, node distance=0cm, rotate=90, anchor=center,yshift=-0.7cm] {Error, Distance, and Smallest SV };
+  \end{tikzpicture}
+  \caption{The distribution of absolute error, distance to
+    the nearest data point, smallest singular value (SV) and the
+    longest edge of the simplex containing each approximation point in
+    the tests from Fig. \ref{fig:convergence-2d} and
+    Fig. \ref{fig:convergence-20d} for Delaunay. In two dimensions it
+    can be seen that $\|z - x_0\|_2$, $\sigma_d$, and $k$ all shrink
+    at the same rate for well-spaced approximation points, resulting
+    in a faster rate of decrease for approximation error. Notice that
+    in higher dimension the data remains sparse even with thousands of
+    data points, and the decay in data spacing is more prominant. The
+    relatively small reduction in $k$ along with the decrease in
+    $\sigma_d$ explain the minimal reduction in error seen by Delaunay
+    in Fig. \ref{fig:convergence-20d}.}
+  \label{fig:data-spacing}
+\end{figure}
+
+The results achieved for both $d = 2$ and $d = 20$ align with the
+theoretical error bound as can be seen in Figure
+\ref{fig:data-spacing}. In low dimension, thousands of data points
+meaningfully reduce the sparsity of the approximation problem.
+However, in higher dimension it takes many more points to achieve a
+reduction in data sparsity while simultaneously being more difficult
+to produce well-conditioned simplices. This evidences the
+inherent challenge of data sparsity in high dimension approximation
+problems. The analytic results presented are not caused by the chosen
+test function, but rather the exponential increase in complexity that
+accompanies increased dimension.
+
+In summary, the regime of signal-to-noise ratios that result in
+competitively accurate interpolants is greater for moderate to high
+dimensional problems due to increased sparsity. Acknowledging the
+viability of interpolation for problems of moderate dimension, the
+next section will consider real-world problems of similar proportion
+(thousands of examples in tens of dimensions).
 
@@
 
-The theoretical results constructed at the start of this chapter for
-(piecewise) linear interpolation are promising and apply directly to
-Delaunay interpolation, however they are difficult to interpret in
-context with approximation algorithms that do not have similar known
-uniform error bounds. For that reason, this chapter utilizes five
-different data sets of varying dimension and application to construct
+This section extends the comparison of interpolation and regression
+algorithms to a sample of real-world problems. Five different data
+sets of varying dimension and application are utilized to construct
 approximations and compare the accuracy of different techniques.
@@
 In addition to the figures displaying approximation results for each
-data set, tables of accompanying numerical results are located in the
-Appendix. All of the test data sets capture underlying functions that
-are almost certainly stochastic. As described in Section
-\ref{sec:introduction}, regression techniques appear most appropriate
-for these problems. However, typically data grows exponentially more
-sparse with increasing dimension. Given sparse data, regressors tend
-towards interpolation.
+data set, scatter plots of predicted versus actual values
+  and tables of accompanying numerical results including
+  timings and quartiles of prediction error are located in the
+Appendix (Section \ref{sec:appendix}). All of the test data sets
+capture underlying functions that are almost certainly stochastic. As
+described in Section \ref{sec:introduction}, regression techniques
+appear most appropriate for these problems. However, typically data
+grows exponentially more sparse with increasing dimension. Given that
+sparse data regressors tend towards interpolation and, as demonstrated
+in Section \ref{sec:analytic}, interpolants produce similar (if not
+identical) results, it is presumed that interpolants are equally
+viable approximation techniques on these problems.
 
@@
   \begin{tikzpicture}
-    \node (img)  {\includegraphics[width=0.75\textwidth,]{Figures/NA/Fig2.pdf}};
+    \node (img)  {\includegraphics[width=0.75\textwidth,]{Figures/NA/Fig4.pdf}};
     \node[below=of img, node distance=1cm, yshift=1cm] {Forest Fire Area Burned};
@@
   \label{fig:hist-forest-fire}
-\end{figure}
 
-\begin{figure}
-  \centering
+  \vspace{.3cm}
+
   \begin{tikzpicture}
-    \node (img)  {\includegraphics[width=0.8\textwidth,]{Figures/NA/Fig3.pdf}};
+    \node (img)  {\includegraphics[width=0.8\textwidth,]{Figures/NA/Fig5.pdf}};
     \node[below=of img, node distance=1cm, yshift=1cm] {Area Burned Error};
@@
   \begin{tikzpicture}
-    \node (img)  {\includegraphics[width=0.75\textwidth,]{Figures/NA/Fig4.pdf}};
+    \node (img)  {\includegraphics[width=0.75\textwidth,]{Figures/NA/Fig6.pdf}};
     \node[below=of img, node distance=1cm, yshift=1cm] {Parkinson's Total UPDRS Score};
@@
   \label{fig:hist-parkinsons}
-\end{figure}
 
-\begin{figure}
-  \centering
+  \vspace{.3cm}
+
   \begin{tikzpicture}
-    \node (img)  {\includegraphics[width=0.8\textwidth,]{Figures/NA/Fig5.pdf}};
+    \node (img)  {\includegraphics[width=0.8\textwidth,]{Figures/NA/Fig7.pdf}};
     \node[below=of img, node distance=1cm, yshift=1cm] {Total UPDRS Score Error};
@@
   \begin{tikzpicture}
-    \node (img)  {\includegraphics[width=0.75\textwidth,]{Figures/NA/Fig6.pdf}};
+    \node (img)  {\includegraphics[width=0.75\textwidth,]{Figures/NA/Fig8.pdf}};
     \node[below=of img, node distance=1cm, yshift=1cm] {Daily Rainfall in Sydney};
@@
   \label{fig:hist-weather}
-\end{figure}
 
-\begin{figure}
-  \centering
+  \vspace{.3cm}
+
   \begin{tikzpicture}
-    \node (img)  {\includegraphics[width=0.8\textwidth,]{Figures/NA/Fig7.pdf}};
+    \node (img)  {\includegraphics[width=0.8\textwidth,]{Figures/NA/Fig9.pdf}};
     \node[below=of img, node distance=1cm, yshift=1cm] {Sydney Rainfall Tomorrow Error};
@@
   \begin{tikzpicture}
-    \node (img)  {\includegraphics[width=0.75\textwidth,]{Figures/NA/Fig8.pdf}};
+    \node (img)  {\includegraphics[width=0.75\textwidth,]{Figures/NA/Fig10.pdf}};
     \node[below=of img, node distance=1cm, yshift=1cm] {Credit Card Transaction Amount};
@@
   \label{fig:hist-credit-card}
-\end{figure}
 
-\begin{figure}
-  \centering
+  \vspace{.3cm}
+
   \begin{tikzpicture}
-    \node (img)  {\includegraphics[width=0.8\textwidth,]{Figures/NA/Fig9.pdf}};
+    \node (img)  {\includegraphics[width=0.8\textwidth,]{Figures/NA/Fig11.pdf}};
     \node[below=of img, node distance=1cm, yshift=1cm] {Transaction Amount Error};
@@
   \begin{tikzpicture}
-    \node (img)  {\includegraphics[width=0.75\textwidth,]{Figures/NA/Fig10.pdf}};
+    \node (img)  {\includegraphics[width=0.75\textwidth,]{Figures/NA/Fig12.pdf}};
     \node[below=of img, node distance=1cm, yshift=1cm] {I/O Read Throughput};
@@
   \begin{tikzpicture}
-    \node (img)  {\includegraphics[width=0.8\textwidth,]{Figures/NA/Fig11.pdf}};
+    \node (img)  {\includegraphics[width=0.8\textwidth,]{Figures/NA/Fig13.pdf}};
     \node[below=of img, node distance=1cm, yshift=1cm] {Read Throughput KS Statistic};
@@
   \begin{tikzpicture}
-    \node (img)  {\includegraphics[width=0.8\textwidth]{Figures/NA/Fig12.pdf}};
+    \node (img)  {\includegraphics[width=0.8\textwidth]{Figures/NA/Fig14.pdf}};
     \node[below=of img, node distance=1cm, yshift=1cm] {KS Statistic for Predicted vs. Actual};
@@
 \section{Discussion of Empirical Results}
+\label{sec:discussion}
 
@@
 test data sets with real-valued ranges. The interpolants discussed in
-this chapter produce the \textit{best} approximations roughly one third
+this paper produce the \textit{best} approximations roughly one third
 of the time, and produce competitive approximations for almost all
@@
 
-The theoretical results presented in this chapter directly apply to
-Delaunay interpolation, however the performance of Delaunay does not
-appear significantly better than other algorithms on these data sets.
-This observation may be due to the stochastic nature of the data, but
-it also speaks to the power of the approximations generated by the
-different interpolation methods.  The strong performance of other
-interpolants suggests that theoretical results similar to those
-presented in this work can be achieved for the other interpolants
-under reasonable assumptions.
+The new theoretical results presented at the beginning of this chapter
+directly apply to Delaunay interpolation, however the performance of
+Delaunay does not appear significantly better than other algorithms on
+these data sets.  This observation may be due to the stochastic nature
+of the data, but it also speaks to the power of the approximations
+generated by the different interpolation methods.  The strong
+performance of other interpolants suggests that theoretical results
+similar to those presented in this work can be achieved for the other
+interpolants under reasonable assumptions.
 
 Finally, most of the interpolants presented in this work benefit from
-the ability to model any function over real $d$-tuples with a range
-that is closed under convex combinations. The results of the
-distribution prediction case study indicate that interpolants can
-effectively predict CDFs. The error analysis for that work relies on
-the KS statistic, which captures the worst part of any prediction and
-hence provides a conservatively large estimate of approximation
-error. The average absolute errors in the predicted CDFs are always
-lower than the KS statistics. However, the KS statistic was chosen as
-a metric because of the important surrounding statistical theory. A
-nonnegligible volume of predictions provide impressively low levels of
-average absolute error in that final case study.
+the ability to model \textit{any} function over real $d$-tuples with a
+range that is closed under convex combinations. In general, error can
+be quantified by any measure (particularly of interest may be
+$L^2$, $L^\infty$, etc.). The results of the distribution prediction
+case study indicate that interpolants can effectively predict
+CDFs. The error analysis for that work relies on the KS statistic,
+which captures the worst part of any prediction and hence provides a
+conservatively large estimate of approximation error. The average
+absolute errors in the predicted CDFs are always lower than the KS
+statistics. However, the KS statistic was chosen as a metric because
+of the important surrounding statistical theory. A nonnegligible
+volume of predictions provide impressively low levels of average
+absolute error in that final case study.
 
-\section{Takeaway from Empirical Tests}
+\section{Takeaway from Empirical Results}
+\label{sec:error_conclusion}
 
 The major contributions of this work are: 1) new uniform theoretical
-error bounds for piecewise linear interpolation in arbitrary
-dimension; 2) an empirical evaluation across real-world problems that
-demonstrates interpolants produce competitively accurate models of
-multivariate phenomenon when compared with common regressors for
-sparse, moderately high dimensional problems (Section
-\ref{sec:error_data}); and 3) a demonstration that some interpolants
-generalize to interpolation in function spaces, preserving
-monotonicity (with CDFs, e.g.), neither of which common regressors can
-do.
+error bounds for piecewise linear interpolation in arbitrary dimension;
+2) an empirical evaluation across
+real-world problems that demonstrates interpolants produce
+competitively accurate models of multivariate phenomenon when compared
+with common regressors for sparse, moderately high dimensional
+problems (Section \ref{sec:error_data}); and 3) a demonstration that some
+interpolants generalize to interpolation in function spaces, 
+preserving monotonicity (with CDFs, e.g.), neither
+of which common regressors can do.
 
-The various interpolants discussed in this work have been
+The various interpolants discussed in this paper have been
 demonstrated to effectively approximate multivariate phenomena up to
@@
 encourage broader application of interpolants to multivariate
-approximation in science
+approximation in science.

### 2020-07-28 3-Disseration/strong.tex
Updated thesis document, sent draft to Dr. Watson.

@@
 
-\begin{figure}[htb]
+\begin{figure}
   \centering
-  \begin{tikzpicture}
-    \node (img)  {\includegraphics[width=0.75\textwidth,]{Figures/NA/Fig1.pdf}};
-    \node[below=of img, node distance=1cm, yshift=1cm] {I/O Throughput};
-    \node[left=of img, node distance=0cm, rotate=90, anchor=center,yshift=-0.7cm] {CDF Value};
-  \end{tikzpicture}
-  \vspace{-0.3cm}
+  \includegraphics[width=0.8\textwidth,]{Figures/NA/example_distribution_prediction.pdf}
   \caption{In this HPC I/O example, the general methodology for
-    predicting a CDF and evaluating error can be seen. The Delaunay
-    method chose three source distributions (dotted lines) and
-    assigned weights \{.3, .4, .3\} (top to bottom at arrow). The
-    weighted sum of the three known CDFs produces the predicted CDF
-    (dashed line). The KS Statistic (arrow) computed between the true
-    CDF (solid line) and predicted CDF (dashed line) is 0.2 for this
-    example. The KS test null hypothesis is rejected at $p$-value
-    0.01, however it is not rejected at $p$-value 0.001.}
+    predicting a CDF and evaluating error can be seen, where M means
+    $\times 10^6$. The Delaunay method chose three source
+    distributions (dotted lines) and assigned weights \{.1, .3, .6\}
+    (top to bottom at middle). The weighted sum of the three known
+    CDFs produces the predicted CDF (dashed line). The KS Statistic
+    (vertical line) computed between the true CDF (solid line) and
+    predicted CDF (dashed line) is 0.2 for this example. For this
+    example the KS test null hypothesis is rejected at $p$-value 0.01,
+    however it is not rejected at $p$-value 0.001.
+  \vspace{-.1cm}}
   \label{fig:prediction-example}
@@
 reported with summary statistics including: min absolute error, max
-absolute error, and absolute error quartiles. When the range of an
-approximation is the space of cumulative distribution functions, the
-Kolmogorov-Smirnov statistic (max-norm difference between the
+absolute error, and absolute error quartiles. When the range
+of an approximation is the space of cumulative distribution functions,
+the Kolmogorov-Smirnov statistic (max-norm difference between the
 functions) is used.
@@
 PDF $f: \mathbb{R} \rightarrow \mathbb{R}$ have the properties $f(x)
-\geq 0$ and $\int_{-\infty}^{\infty}f(x)dx = 1$. However, for a CDF
+\geq 0$ and $\int_{-\infty}^{\infty}f(x)dx = 1$. Instead, for a CDF
 $F: \mathbb{R} \rightarrow \mathbb{R}$ the properties are
@@
 this is applied can be seen in Figure \ref{fig:prediction-example}. 
+In this example the KS test null hypothesis is rejected at $p$-value
+0.01, however it is not rejected at $p$-value 0.001.
 
@@
 The two-sample KS test is a useful nonparametric test for comparing
-two CDFs while only assuming stationarity, finite mean, and finite
-variance. The null hypothesis (that two CDFs come from the same
-underlying distribution) is rejected at level $p \in [0,1]$ when
- $$ KS > \sqrt{-\frac{1}{2}\ln\biggl(\frac{p}{2}\biggr)} \sqrt{\frac{1}{n_1} + \frac{1}{n_2}}, $$
-with distribution sample sizes $n_1,n_2 \in \mathcal{N}$. For all
-applications of the KS test presented in this work $n_1 = n_2$. An
-example of the round-trip prediction methodology from known and
-predicted distributions to the calculation of error can be seen in
-Figure \ref{fig:prediction-example}.
+two empirical CDFs while only assuming stationarity,
+finite mean, and finite variance. The null hypothesis (that two
+empirical CDFs come from the same underlying distribution)
+is rejected at level $p \in [0,1]$ when
+ $$ KS > \sqrt{-\frac{1}{2}\ln\biggl(\frac{p}{2}\biggr)}
+\sqrt{\frac{1}{n_1} + \frac{1}{n_2}}, $$ with distribution sample
+sizes $n_1,n_2 \in \mathcal{N}$. For all applications of the KS test
+presented in this work $n_1 = n_2$. An example of the process of
+generating predicted distributions from known distributions and the
+subsequent calculation of error can be seen in Figure
+\ref{fig:prediction-example}. A brief listing of relevant statistical
+terms used throughout this work is provided in the Appendix (Section
+\ref{sec:appendix}).
 
 %% ===================================================================
-
 \subsection{Feature Weighting}

### 2020-07-28 3-Disseration/thesis.tex
Updated thesis document, sent draft to Dr. Watson.

--- 2019-05-03 Disseration/prelim.tex
@@
-\documentclass[doublespace,nopageskip]{VTthesis} % nopageskip = Removes arbitrary blank pages.
+\documentclass[singlespace,nopageskip]{VTthesis} % nopageskip = Removes arbitrary blank pages.
 
@@
 \usetikzlibrary{positioning}
+\usepackage{plain} % For plain TeX in the TOMS paper
 
@@
 
+%% Create a ``Lemma'' environment that does not have italics font.
+\spnewtheorem{plainlemma}{Lemma}{\bf}{}
+%% Create a ``Theorem'' environment that is not numbered.
+\spnewtheorem*{onetheorem}{Theorem}{\bf}{}
+%% Create a ``Proof'' environment that starts with 'Proof.'
+\spnewtheorem*{proofdot}{Proof.}{\it}{}
+
 % Using the following header instead will create a draft copy of your thesis
@@
 % Title of your thesis
-\title{Interpolation for Modeling and Predicting Variability}
+\title{Interpolants and Error Bounds for Modeling and Predicting Variability in Computer Systems}
 
@@
 % This should be your defense date:
-\submitdate{May 16, 2019}
+\submitdate{August 14, 2020}
 
@@
 %% Propose new method to improve predictions for distributions.
-\chapter{Improving Variability Estimates with Monotone $C^2$ Splines} \label{ch:splines}
+%% \chapter{Improving Variability Estimates with Monotone $C^2$ Splines} 
+\chapter{A Package for Monotone Quintic Spline Interpolation} \label{ch:splines}
 \input{splines}
@@
 %% Bibliography.
-\bibliography{prelim}
+\bibliography{thesis}
 \bibliographystyle{plainnat}
@@
 
-  The tables that follow show the precise experimental results for all
-  data sets presented in Section \ref{sec:data}. The tests were all run
-  serially on an otherwise idle machine with a CentOS 6.10 operating
-  system and an Intel i7-3770 CPU operating at 3.4 GHz. The detailed
-  performance results in the tables that follow are very much dependent
-  on the problem and the algorithm implementation (e.g., some codes are
-  TOMS software, some industry distributions, and others are from
-  conference paper venues). Different typeface is used to show best
-  performers, however not much significance should be attached to
-  ranking algorithms based on small time (millisecond) differences. The
-  results serve as a demonstration of conceptual validity.
+\begin{normalsize}
+\textbf{Statistical Terminology.} A random variable $X$ is precisely
+defined by its \textit{cumulative distribution function} (CDF) $F_X$
+and the derivative of the CDF, the \textit{probability density
+  function} (PDF) $f_X.$ For any possible value $x$ of $X$, the
+\textit{percentile} of $x$ is $100$ $F_X(x)$, the percentage of values
+drawn from $X$ that would be less than or equal to $x$ as the number
+of samples tends towards infinity. The \textit{quartiles} of $X$ are
+its $25$-th, $50$-th (median), and $75$-th percentiles. The absolute
+difference between the median and an adjacent quartile is an
+\textit{interquartile range}. Given an independent and identically
+distributed sample from $X$ and presuming that $X$ has finite mean and
+variance, a \textit{confidence interval} can be drawn about any
+percentile estimated from the sample. A \textit{confidence interval}
+describes the probability that a value lies within an interval. The
+\textit{null hypothesis} is a statement (derived from some test
+statistic) that the expected value of the observed statistic is equal
+to an assumed population statistic. The $p$ value is the probability
+of observing a given statistic if the \textit{null hypothesis} is
+true. For a more detailed introduction to statistics and 
+related terminology, see the work of Navidi \cite{navidi_2015}.
+
+\vspace{4mm}
+\noindent \textbf{Raw Numerical Results.} The tables that follow show the
+precise experimental results for all data sets presented in Section
+\ref{sec:data}. The tests were all run serially on an otherwise idle
+machine with a CentOS 6.10 operating system and an Intel i7-3770 CPU
+operating at 3.4 GHz. The detailed performance results in the tables
+that follow are very much dependent on the problem and the algorithm
+implementation (e.g., some codes are TOMS software, some industry
+distributions, and others are from conference paper venues). Different
+typeface is used to show best performers, however not much
+significance should be attached to ranking algorithms based on small
+time (millisecond) differences. The results serve as a demonstration
+of conceptual validity.
+
+\end{normalsize}
+
+\begin{figure}[h]
+  \centering
+  \includegraphics[width=0.45\textwidth,]{Figures/NA/scatter_forestfires.pdf}
+  \hspace{.6cm}
+  \includegraphics[width=0.45\textwidth,]{Figures/NA/scatter_parkinsons.pdf}
+  \vspace{.6cm}
+
+  \includegraphics[width=0.45\textwidth,]{Figures/NA/scatter_weather.pdf}
+  \hspace{.6cm}
+  \includegraphics[width=0.45\textwidth,]{Figures/NA/scatter_creditcard.pdf}
+  \vspace{.3cm}
+
+  \caption{Scatter plots for predicted versus actual values for the
+    top three models on each of the four real valued approximation
+    problems. Top left is forest fire data, top right is Parkinson's
+    data, bottom left is rainfall data, and bottom right is credit
+    card transaction data. The blue circles correspond to the best
+    model, green diamonds to the second best, and red crosses to the
+    third best model for each data set. There are a large of number of
+    0-valued entries in the forest fire and rainfall data sets that
+    are not included in the visuals making the true ranking of the
+    models appear to disagree with the observed outcomes.}
+  \label{fig:scatter_plots}
+\end{figure}
 
@@
     \hline
-      Algorithm & \multicolumn{2}{c|}{Min} & \multicolumn{2}{c|}{$25^{th}$} & \multicolumn{2}{c|}{$50^{th}$} & \multicolumn{2}{c|}{$75^{th}$} & \multicolumn{2}{c}{Max}\\
+    Algorithm & \multicolumn{2}{c}{Min}       & \multicolumn{2}{c}{$25^{th}$} & \multicolumn{2}{c}{$50^{th}$} & \multicolumn{2}{c}{$75^{th}$} & \multicolumn{2}{c}{Max}\\
     \hline
     MARS      & $0$&$00984$                   & $3$&$11$                      & $7$&$01$                      & $\mathit{11}$&$\mathit{7}$    & $1090$&$0$\\
-      SVR & $\mathit{0}$&$\mathit{00402}$ & $\mathbf{0}$&$\mathbf{243}$ & $\mathbf{0}$&$\mathbf{416}$ & $\mathbf{6}$&$\mathbf{19}$ & $1090$&$0$\\
-      MLP & $0$&$0426$ & $2$&$63$ & $\mathit{5}$&$\mathit{27}$ & $14$&$0$ & $1090$&$0$\\
-      Delaunay & $\mathbf{0}$&$\mathbf{00}$ & $1$&$98$ & $5$&$37$ & $13$&$1$ & $\mathit{1080}$&$\mathit{0}$\\
-      ShepMod & $\mathbf{0}$&$\mathbf{00}$ & $\mathit{1}$&$\mathit{93}$ & $6$&$27$ & $16$&$0$ & $1090$&$0$\\
-      LSHEP & $0$&$0400$ & $2$&$17$ & $8$&$87$ & $19$&$1$ & $\mathbf{1070}$&$\mathbf{0}$\\
-      Voronoi & $0$&$00982$ & $3$&$65$ & $7$&$56$ & $15$&$6$ & $1090$&$0$\\
-      BoxSpline & $\mathbf{0}$&$\mathbf{00}$ & $2$&$27$ & $5$&$91$ & $16$&$4$ & $1090$&$0$\\
+    SVR       & $0$&$0118$                    & $\mathbf{0}$&$\mathbf{615}$   & $\mathbf{0}$&$\mathbf{931}$   & $\mathbf{5}$&$\mathbf{89}$    & $1090$&$0$\\
+    MLP       & $0$&$0426$                    & $2$&$63$                      & $5$&$27$                      & $14$&$0$                      & $1090$&$0$\\
+    Delaunay  & $\mathbf{0}$&$\mathbf{0}$     & $1$&$98$                      & $5$&$37$                      & $13$&$1$                      & $\mathit{1080}$&$\mathit{0}$\\
+    ShepMod   & $\mathbf{0}$&$\mathbf{0}$     & $1$&$93$                      & $6$&$27$                      & $16$&$0$                      & $1090$&$0$\\
+    LSHEP     & $0$&$04$                      & $2$&$17$                      & $8$&$87$                      & $19$&$1$                      & $\mathbf{1070}$&$\mathbf{0}$\\
+    Voronoi   & $\mathit{0}$&$\mathit{00982}$ & $3$&$65$                      & $7$&$56$                      & $15$&$6$                      & $1090$&$0$\\
+    BoxSpline & $\mathbf{0}$&$\mathbf{0}$     & $\mathit{1}$&$\mathit{27}$    & $\mathit{4}$&$\mathit{61}$    & $12$&$6$                      & $1090$&$0$\\
     \hline
@@
     digits.}
+  \vspace{.2cm}
   \label{table:error-forest-fire}
-  \end{table}
 
-  \begin{table}[H]
-    \centering
   \begin{tabular}{|c|r@{.}l| c |r@{.}l|r@{.}l|r@{.}l|}
@@
     \cline{1-3}\cline{5-10}
-      MARS & \quad$\mathit{7}$&$\mathit{7}$ &  & \quad\quad$29$&$1$ & \quad$0$&$00137$ & \quad\quad\quad$0$&$0686$\\
-      SVR & $\mathbf{80}$&$\mathbf{2}$ &  & $\mathbf{0}$&$\mathbf{00584}$ & $\mathit{0}$&$\mathit{0000620}$ & $\mathit{0}$&$\mathit{00310}$\\
+    MARS      & \quad$7$&$3$                   &  & \quad\quad$29$&$1$                      & \quad$0$&$00137$                   & \quad\quad\quad$0$&$0686$\\
+    SVR       & $\mathbf{78}$&$\mathbf{0}$     &  & $\mathbf{0}$&$\mathbf{00584}$           & $\mathit{0}$&$\mathit{0000620}$    & $\mathit{0}$&$\mathit{00310}$\\
     MLP       & $0$&$0$                        &  & $32$&$8$                                & $0$&$000871$                       & $0$&$0436$\\
-      Delaunay & $3$&$5$ &  & $0$&$0151$ & $0$&$0234$ & $1$&$18$\\
-      ShepMod & $3$&$7$ &  & $\mathit{0}$&$\mathit{00634}$ & $0$&$0000644$ & $0$&$00322$\\
+    Delaunay  & $0$&$2$                        &  & $0$&$0151$                              & $0$&$0234$                         & $1$&$18$\\
+    ShepMod   & $2$&$0$                        &  & $\mathit{0}$&$\mathit{00634}$           & $0$&$0000644$                      & $0$&$00322$\\
     LSHEP     & $5$&$1$                        &  & $0$&$0275$                              & $\mathbf{0}$&$\mathbf{0000463}$    & $\mathbf{0}$&$\mathbf{00231}$\\
     Voronoi   & $0$&$0$                        &  & $0$&$0152$                              & $0$&$000396$                       & $0$&$0198$\\
-      BoxSpline & $0$&$4$ &  & $0$&$00724$ & $0$&$0000978$ & $0$&$00489$\\
+    BoxSpline & $\mathit{9}$&$\mathit{7}$      &  & $0$&$00724$                             & $0$&$0000978$                      & $0$&$00489$\\
     \cline{1-3}\cline{5-10}
@@
 \end{table}
+\vspace{-1cm}
 
-  \begin{table}
+\begin{table}[H]
   \centering
@@
     \hline
-      Algorithm & \multicolumn{2}{c|}{Min} & \multicolumn{2}{c|}{$25^{th}$} & \multicolumn{2}{c|}{$50^{th}$} & \multicolumn{2}{c|}{$75^{th}$} & \multicolumn{2}{c}{Max}\\
+    Algorithm & \multicolumn{2}{c}{Min}                   & \multicolumn{2}{c}{$25^{th}$} & \multicolumn{2}{c}{$50^{th}$} & \multicolumn{2}{c}{$75^{th}$} & \multicolumn{2}{c}{Max}\\
     \hline
     MARS      & $0$&$00948$                               & $9$&$98$                      & $20$&$4$                      & $32$&$9$                      & $863$&$0$\\
-      SVR & $0$&$00138$ & $3$&$17$ & $7$&$31$ & $12$&$6$ & $\mathbf{27}$&$\mathbf{5}$\\
+    SVR       & $0$&$00233$                               & $3$&$15$                      & $7$&$21$                      & $11$&$3$                      & $\mathbf{28}$&$\mathbf{6}$\\
     MLP       & $0$&$0000239$                             & $\mathit{0}$&$\mathit{533}$   & $\mathit{1}$&$\mathit{25}$    & $\mathbf{2}$&$\mathbf{84}$    & $39$&$3$\\
-      Delaunay & $\mathit{3}$&$\mathit{72 \times 10^{-12}}$ & $1$&$20$ & $3$&$50$ & $7$&$67$ & $30$&$7$\\
+    Delaunay  & $\mathit{3}$&$\mathit{72 \times 10^{-12}}$ & $1$&$2$                       & $3$&$5$                       & $7$&$67$                      & $30$&$7$\\
     ShepMod   & $\mathbf{0}$&$\mathbf{0}$                 & $\mathbf{0}$&$\mathbf{255}$   & $\mathbf{0}$&$\mathbf{908}$   & $\mathit{3}$&$\mathit{43}$    & $34$&$5$\\
@@
     Voronoi   & $\mathbf{0}$&$\mathbf{0}$                 & $1$&$29$                      & $3$&$52$                      & $6$&$87$                      & $30$&$1$\\
-      BoxSpline & $0$&$00287$ & $4$&$39$ & $9$&$10$ & $14$&$8$ & $41$&$3$\\
+    BoxSpline & $0$&$006$                                 & $4$&$3$                       & $8$&$91$                      & $15$&$1$                      & $45$&$3$\\
     \hline
@@
     digits.}
+  \vspace{.2cm}
   \label{table:error-parkinsons}
-  \end{table}
 
-  \begin{table}
-    \centering
   \begin{tabular}{|c|r@{.}l| c |r@{.}l|r@{.}l|r@{.}l|}
@@
     time approximating 587 points.}
+  \vspace{-1cm}
   \label{table:best-parkinsons}
@@
 
-  \begin{table}
+\begin{table}[H]
   \centering
@@
     \hline
-      Algorithm & \multicolumn{2}{c|}{Min} & \multicolumn{2}{c|}{$25^{th}$} & \multicolumn{2}{c|}{$50^{th}$} & \multicolumn{2}{c|}{$75^{th}$} & \multicolumn{2}{c}{Max}\\
+    Algorithm & \multicolumn{2}{c}{Min}                    & \multicolumn{2}{c}{$25^{th}$}              & \multicolumn{2}{c}{$50^{th}$}  & \multicolumn{2}{c}{$75^{th}$}  & \multicolumn{2}{c}{Max}\\
     \hline
     MARS      & $\mathit{6}$&$\mathit{45 \times 10^{-15}}$ & $\mathbf{2}$&$\mathbf{70 \times 10^{-14}}$ & $1$&$66$                       & $1$&$96$                       & $\mathit{53}$&$\mathit{3}$\\
-      SVR & $0$&$0420$ & $0$&$143$ & $0$&$148$ & $\mathit{0}$&$\mathit{860}$ & $119$&$0$\\
+    SVR       & $0$&$0000915$                              & $0$&$0833$                                 & $0$&$263$                      & $\mathit{1}$&$\mathit{13}$     & $109$&$0$\\
     MLP       & $0$&$0000689$                              & $0$&$0337$                                 & $\mathbf{0}$&$\mathbf{0795}$   & $\mathbf{0}$&$\mathbf{264}$    & $\mathbf{5}$&$\mathbf{31}$\\
     Delaunay  & $\mathbf{0}$&$\mathbf{0}$                  & $0$&$187$                                  & $0$&$919$                      & $2$&$72$                       & $56$&$3$\\
-      ShepMod & $\mathbf{0}$&$\mathbf{0}$ & $0$&$0957$ & $0$&$685$ & $2$&$90$ & $73$&$2$\\
+    ShepMod   & $\mathbf{0}$&$\mathbf{0}$                  & $0$&$0957$                                 & $0$&$685$                      & $2$&$9$                        & $73$&$2$\\
     LSHEP     & $\mathbf{0}$&$\mathbf{0}$                  & $\mathit{0}$&$\mathit{0153}$               & $\mathit{0}$&$\mathit{106}$    & $1$&$17$                       & $113$&$0$\\
-      Voronoi & $\mathbf{0}$&$\mathbf{0}$ & $0$&$430$ & $1$&$28$ & $2$&$94$ & $83$&$8$\\
-      BoxSpline & $\mathbf{0}$&$\mathbf{0}$ & $0$&$789$ & $2$&$06$ & $5$&$25$ & $119$&$0$\\
+    Voronoi   & $\mathbf{0}$&$\mathbf{0}$                  & $0$&$43$                                   & $1$&$28$                       & $2$&$94$                       & $83$&$8$\\
+    BoxSpline & $\mathbf{0}$&$\mathbf{0}$                  & $0$&$342$                                  & $1$&$59$                       & $5$&$53$                       & $119$&$0$\\
     \hline
@@
     digits.}
+  \vspace{.2cm}
   \label{table:error-weather}
-  \end{table}
 
-  \begin{table}
-    \centering
   \begin{tabular}{|c|r@{.}l| c |r@{.}l|r@{.}l|r@{.}l|}
@@
     MARS      & \,\,\,\,$10$&$7$             &  & \quad\quad\quad$\mathit{0}$&$\mathit{151}$ & \quad$0$&$000117$                  & \quad\quad\quad\,\,$0$&$0304$\\
-      SVR & $4$&$1$ &  & $\mathbf{0}$&$\mathbf{133}$ & $\mathbf{0}$&$\mathbf{0000947}$ & $\mathbf{0}$&$\mathbf{0246}$\\
-      MLP & $\mathbf{56}$&$\mathbf{8}$ &  & $169$&$0$ & $0$&$00137$ & $0$&$356$\\
+    SVR       & $0$&$0$                      &  & $\mathbf{0}$&$\mathbf{133}$                & $\mathbf{0}$&$\mathbf{0000947}$    & $\mathbf{0}$&$\mathbf{0246}$\\
+    MLP       & $\mathbf{60}$&$\mathbf{9}$   &  & $169$&$0$                                  & $0$&$00137$                        & $0$&$356$\\
     Delaunay  & $0$&$1$                      &  & $0$&$664$                                  & $0$&$886$                          & $230$&$0$\\
@@
     Voronoi   & $0$&$2$                      &  & $0$&$675$                                  & $0$&$0270$                         & $7$&$01$\\
-      BoxSpline & $0$&$3$ &  & $0$&$330$ & $0$&$000406$ & $0$&$106$\\
+    BoxSpline & $4$&$1$                      &  & $0$&$330$                                  & $0$&$000406$                       & $0$&$106$\\
     \cline{1-3}\cline{5-10}
@@
     time approximating 260 points.}
+  \vspace{-1cm}
   \label{table:best-weather}
@@
     \hline
-      Algorithm & \multicolumn{2}{c|}{Min} & \multicolumn{2}{c|}{$25^{th}$} & \multicolumn{2}{c|}{$50^{th}$} & \multicolumn{2}{c|}{$75^{th}$} & \multicolumn{2}{c}{Max}\\
+    Algorithm & \multicolumn{2}{c}{Min}                    & \multicolumn{2}{c}{$25^{th}$} & \multicolumn{2}{c}{$50^{th}$} & \multicolumn{2}{c}{$75^{th}$} & \multicolumn{2}{c}{Max}\\
     \hline
     MARS      & $5$&$36$                                   & $3610$&$0$                    & $4580$&$0$                    & $5450$&$0$                    & $13400$&$0$\\
-      SVR & $0$&$0138$ & $8$&$47$ & $15$&$4$ & $41$&$9$ & $7700$&$0$\\
-      MLP & $0$&$00151$ & $\mathbf{1}$&$\mathbf{60}$ & $\mathbf{3}$&$\mathbf{86}$ & $\mathbf{8}$&$\mathbf{34}$ & $\mathbf{604}$&$\mathbf{0}$\\
-      Delaunay & $\mathbf{0}$&$\mathbf{0}$ & $\mathit{2}$&$\mathit{69}$ & $\mathit{13}$&$\mathit{8}$ & $\mathit{35}$&$\mathit{0}$ & $\mathit{4840}$&$\mathit{0}$\\
+    SVR       & $0$&$00706$                                & $8$&$14$                      & $\mathit{13}$&$\mathit{0}$    & $41$&$6$                      & $7690$&$0$\\
+    MLP       & $0$&$00151$                                & $\mathbf{1}$&$\mathbf{6}$     & $\mathbf{3}$&$\mathbf{86}$    & $\mathbf{8}$&$\mathbf{34}$    & $\mathbf{604}$&$\mathbf{0}$\\
+    Delaunay  & $\mathbf{0}$&$\mathbf{0}$                  & $\mathit{2}$&$\mathit{69}$    & $13$&$8$                      & $\mathit{35}$&$\mathit{0}$    & $\mathit{4840}$&$\mathit{0}$\\
     ShepMod   & $\mathbf{0}$&$\mathbf{0}$                  & $4$&$21$                      & $17$&$3$                      & $51$&$6$                      & $6510$&$0$\\
@@
     Voronoi   & $\mathit{2}$&$\mathit{89 \times 10^{-10}}$ & $14$&$7$                      & $32$&$8$                      & $52$&$9$                      & $4860$&$0$\\
-      BoxSpline & $0$&$00740$ & $20$&$8$ & $40$&$9$ & $79$&$6$ & $7640$&$0$\\
+    BoxSpline & $\mathbf{0}$&$\mathbf{0}$                  & $12$&$4$                      & $35$&$0$                      & $90$&$1$                      & $7690$&$0$\\
     \hline
@@
     digits.}
+  \vspace{.2cm}
   \label{table:error-credit-card}
-  \end{table}
 
-  \begin{table}
-    \centering
   \begin{tabular}{|c|r@{.}l| c |r@{.}l|r@{.}l|r@{.}l|}
@@
     and median time approximating 556 points.}
+  \vspace{-1cm}
   \label{table:best-credit-card}
@@
     \hline
-      Algorithm & \multicolumn{2}{c|}{Min} & \multicolumn{2}{c|}{$25^{th}$} & \multicolumn{2}{c|}{$50^{th}$} & \multicolumn{2}{c|}{$75^{th}$} & \multicolumn{2}{c}{Max}\\
+    Algorithm & \multicolumn{2}{c}{Min}      & \multicolumn{2}{c}{$25^{th}$} & \multicolumn{2}{c}{$50^{th}$} & \multicolumn{2}{c}{$75^{th}$} & \multicolumn{2}{c}{Max}\\
     \hline
     Delaunay  & $\mathbf{0}$&$\mathbf{0287}$ & $\mathbf{0}$&$\mathbf{0914}$  & $\mathbf{0}$&$\mathbf{158}$   & $\mathit{0}$&$\mathit{308}$   & $\mathit{0}$&$\mathit{897}$\\
-      ShepMod & $0$&$0307$ & $\mathit{0}$&$\mathit{0983}$ & $\mathit{0}$&$\mathit{164}$ & $0$&$335$ & $1$&$00$\\
-      Voronoi & $\mathit{0}$&$\mathit{0303}$ & $0$&$100$ & $0$&$165$ & $\mathbf{0}$&$\mathbf{274}$ & $\mathbf{0}$&$\mathbf{762}$\\
-      BoxSpline & $0$&$0703$ & $0$&$417$ & $0$&$561$ & $0$&$745$ & $1$&$00$\\
+    ShepMod   & $0$&$0307$                   & $\mathit{0}$&$\mathit{0983}$  & $\mathit{0}$&$\mathit{164}$   & $0$&$335$                     & $1$&$0$\\
+    Voronoi   & $\mathit{0}$&$\mathit{0303}$ & $0$&$1$                       & $0$&$165$                     & $\mathbf{0}$&$\mathbf{274}$   & $\mathbf{0}$&$\mathbf{762}$\\
+    BoxSpline & $0$&$0679$                   & $0$&$429$                     & $0$&$571$                     & $0$&$752$                     & $1$&$0$\\
     \hline
@@
     three significant digits.}
+  \vspace{.2cm}
   \label{table:error-throughput}
-  \end{table}
 
-  \begin{table}
-    \centering
   \begin{tabular}{|c|r@{.}l| c |r@{.}l|r@{.}l|r@{.}l|}
@@
 \end{table}
+
+\vspace*{\fill}
 

### 2020-07-28 Research_[2020-05-18]_Piecewise_Monotone_Quintic_Fortran/paper/2-monotone-quintic.tex
Updated MQSI with new optional argument for returning derivative values. In the midst of discussing the CLI with coauthors.

@@
 function $B_i(s)$, for $0 \le s \le 1$, that is true if the order six
-polynomial $f(x)$ on $[X_i, X_{i+1}]$ matching derivatives
-$f(X_i)=Y_i$, $Df(X_i)=s \,u_i$, $D^2f(X_i)=s \,v_i$ at $X_i$, and derivatives
-$f(X_{i+1})=Y_{i+1}$, $Df(X_{i+1})=s \,u_{i+1}$, $D^2f(X_{i+1})=s \,v_{i+1}$
+polynomial $Q(x)$ on $[X_i, X_{i+1}]$ matching derivatives
+$Q(X_i)=Y_i$, $DQ(X_i)=s \,u_i$, $D^2Q(X_i)=s \,v_i$ at $X_i$, and derivatives
+$Q(X_{i+1})=Y_{i+1}$, $DQ(X_{i+1})=s \,u_{i+1}$, $D^2Q(X_{i+1})=s \,v_{i+1}$
 at $X_{i+1}$ is monotone, and false otherwise. As is outlined in Algorithm

### 2020-07-28 Research_[2020-05-18]_Piecewise_Monotone_Quintic_Fortran/fortran/MQSI.f90

@@ -6,12 +6,13 @@
 !   quintic spline interpolant of data in terms of its B-spline basis.
 ! 
 ! CONTAINS:
-!   SUBROUTINE MQSI(X, Y, T, BCOEF, INFO)
+!   SUBROUTINE MQSI(X, Y, T, BCOEF, INFO, UV)
 !     USE REAL_PRECISION, ONLY: R8
 !     REAL(KIND=R8), INTENT(IN),    DIMENSION(:) :: X
 !     REAL(KIND=R8), INTENT(INOUT), DIMENSION(:) :: Y
 !     REAL(KIND=R8), INTENT(OUT),   DIMENSION(:) :: T, BCOEF
 !     INTEGER, INTENT(OUT) :: INFO
+!     REAL(KIND=R8), INTENT(OUT), DIMENSION(:,:), OPTIONAL :: UV
 !   END SUBROUTINE MQSI
 ! 
 ! DEPENDENCIES:
@@ -35,7 +36,7 @@
 ! VERSION HISTORY:
 !   June 2020 -- (tchl) Created file, (ltw / wit) reviewed and revised.
 ! 
-SUBROUTINE MQSI(X, Y, T, BCOEF, INFO)
+SUBROUTINE MQSI(X, Y, T, BCOEF, INFO, UV)
 ! Computes a monotone quintic spline interpolant (MQSI), Q(X), to data
 ! in terms of spline coefficients BCOEF for a B-spline basis defined by
 ! knots T. Q(X) is theoretically guaranteed to be monotone increasing
@@ -65,8 +66,11 @@ SUBROUTINE MQSI(X, Y, T, BCOEF, INFO)
 !        function values and derivative values causes the linear system used
 !        to compute the final spline interpolant to have a prohibitively
 !        large condition number.
+!     8  The optional array UV must have size at least ND x 2.
 !   >10  20 plus the info flag as returned by DGBSV from LAPACK when
 !        computing the final spline interpolant.
+!   UV(1:ND,1:2) -- First and second derivatives of Q(X) at the breakpoints
+!     (optional argument).
 !   
 USE REAL_PRECISION, ONLY: R8
 IMPLICIT NONE
@@ -75,6 +79,7 @@ REAL(KIND=R8), INTENT(IN),    DIMENSION(:) :: X
 REAL(KIND=R8), INTENT(INOUT), DIMENSION(:) :: Y
[…]

### 2020-07-28 Research_[2020-05-18]_Piecewise_Monotone_Quintic_Fortran/fortran/sample_main.f90

@@ -23,12 +23,13 @@
 !     INTEGER, PARAMETER :: R8
 !   END MODULE REAL_PRECISION
 ! 
-!   SUBROUTINE MQSI(X, Y, T, BCOEF, INFO)
+!   SUBROUTINE MQSI(X, Y, T, BCOEF, INFO, UV)
 !     USE REAL_PRECISION, ONLY: R8
 !     REAL(KIND=R8), INTENT(IN),    DIMENSION(:) :: X
 !     REAL(KIND=R8), INTENT(INOUT), DIMENSION(:) :: Y
 !     REAL(KIND=R8), INTENT(OUT),   DIMENSION(:) :: T, BCOEF
 !     INTEGER, INTENT(OUT) :: INFO
+!     REAL(KIND=R8), INTENT(OUT), DIMENSION(:,:), OPTIONAL :: UV
 !   END SUBROUTINE MQSI
 ! 
 !   SUBROUTINE EVAL_SPLINE(T, BCOEF, XY, INFO, D)
@@ -53,12 +54,13 @@ IMPLICIT NONE
 
 ! Define the interfaces for relevant MQSI package subroutines.
 INTERFACE
- SUBROUTINE MQSI(X, Y, T, BCOEF, INFO)
+ SUBROUTINE MQSI(X, Y, T, BCOEF, INFO, UV)
    USE REAL_PRECISION, ONLY: R8
-   REAL(KIND=R8), INTENT(IN),  DIMENSION(:) :: X
-   REAL(KIND=R8), INTENT(INOUT),  DIMENSION(:) :: Y
-   REAL(KIND=R8), INTENT(OUT), DIMENSION(:) :: T, BCOEF
+   REAL(KIND=R8), INTENT(IN),    DIMENSION(:) :: X
+   REAL(KIND=R8), INTENT(INOUT), DIMENSION(:) :: Y
+   REAL(KIND=R8), INTENT(OUT),   DIMENSION(:) :: T, BCOEF
    INTEGER, INTENT(OUT) :: INFO
+   REAL(KIND=R8), INTENT(OUT), DIMENSION(:,:), OPTIONAL :: UV
  END SUBROUTINE MQSI
  SUBROUTINE EVAL_SPLINE(T, BCOEF, XY, INFO, D)
    USE REAL_PRECISION, ONLY: R8

### 2020-07-29 3-Disseration/1-apps.tex
Update.

--- 2020-07-28 3-Disseration/apps.tex
@@
 applied to any performance metric and any system.
+
+\section{Organization of this Document}
+
+This thesis is comprised of work from multiple conference and journal
+papers. Chapters \ref{ch:apps} and \ref{ch:algs} share the broader
+context for this work and derive from many papers
+\cite{lux2018predictive,lux2020effective,lux2020interpolation}. Chapter
+\ref{ch:naive} appears in \cite{lux2018predictive}. Chapter
+\ref{ch:boxes} appears in \cite{lux2018novel}. Chapter \ref{ch:strong}
+appears in \cite{lux2018nonparametric}. Chapter \ref{ch:error} is to
+appear in \cite{lux2020interpolation}. Finally Chapter
+\ref{ch:splines} was presented in part in \cite{lux2020algorithm} and
+is to be submitted in full to ACM Transactions on Mathematical
+Software (TOMS).

### 2020-07-29 3-Disseration/6-error.tex
Update.

--- 2020-07-28 3-Disseration/error.tex
@@
 to Delaunay interpolation; and (3) multiple other interpolants in this
-paper compute predictions as convex combinations of observed function
+work compute predictions as convex combinations of observed function
 values, which may allow for straightforward extensions of this error
@@
 test data sets with real-valued ranges. The interpolants discussed in
-this paper produce the \textit{best} approximations roughly one third
+this work produce the \textit{best} approximations roughly one third
 of the time, and produce competitive approximations for almost all
@@
 
-The various interpolants discussed in this paper have been
+The various interpolants discussed in this work have been
 demonstrated to effectively approximate multivariate phenomena up to

### 2020-07-29 3-Disseration/7-splines.tex
Update.

--- 2020-07-28 Research_[2020-05-18]_Piecewise_Monotone_Quintic_Fortran/paper/2-monotone-quintic.tex
@@
-\heading{2. MONOTONE QUINTIC INTERPOLATION}
+\def\codent{\hskip 1.0em} % <- 'tab' for indentation.
+\def\:{\mathrel{:\mathord=}}
+
+\section{Relevance of Quintic Splines}
+
+Many domains of science rely on smooth approximations to real-valued
+functions over a closed interval. Piecewise polynomial functions
+(splines) provide the smooth approximations for animation in graphics
+\cite{herman2006techniques,quint2003scalable}, aesthetic structural support in
+architecture \cite{brennan2020measure}, efficient aerodynamic surfaces in
+automotive and aerospace engineering \cite{brennan2020measure}, prolonged
+effective operation of electric motors \cite{berglund2009planning}, and
+accurate nonparametric approximations in statistics \cite{knott2000interpolating}.
+While polynomial interpolants and regressors apply broadly,
+splines are often a good choice because they can approximate globally
+complex functions while minimizing the local complexity of an
+approximation.
+
+It is often the case that the true underlying function or phenomenon
+being modeled has known properties like convexity, positivity,
+various levels of continuity, or monotonicity. Given a reasonable
+amount of data, it quickly becomes difficult to achieve desirable
+properties in a single polynomial function. In general, the
+maintenance of function properties through interpolation/regression is
+referred to as {\it shape preserving} \cite{fritsch1980monotone,gregory1985shape}. The specific properties the present algorithm will
+preserve in approximations are monotonicity and $C^2$ continuity. In
+addition to previously mentioned applications, these properties are
+crucially important in statistics to the approximation of a cumulative
+distribution function and subsequently the effective generation of
+random numbers from a specified distribution \cite{ramsay1988monotone}.  A spline
+function with these properties could approximate a cumulative
+distribution function to a high level of accuracy with relatively few
+intervals. A twice continuously differentiable approximation to a
+cumulative distribution function (CDF) would produce a corresponding
+probability density function (PDF) that is continuously
+differentiable, which is desirable.
+
+The currently available software for monotone piecewise polynomial
+interpolation includes quadratic \cite{he1998monotone},
+cubic \cite{fritsch1980monotone}, and (with limited application) quartic
+\cite{wang2004rational,abd2011improved,yao2018unconditionally} cases. In
+addition, a statistical method for bootstrapping the construction of
+an arbitrarily smooth monotone fit exists \cite{leitenstorfer2007generalized},
+but the method does not take advantage of known analytic
+properties of quintic polynomials. The code by Fritsch \cite{fritsch1982piecewise} for
+$C^1$ cubic spline interpolation is the predominantly utilized code
+for constructing monotone interpolants at present. Theory has been
+provided for the quintic case \cite{ulrich1994positivity,hess1994positive}
+and that theory was recently utilized in a proposed
+algorithm \cite{lux2020algorithm} for monotone quintic spline construction, however
+no published mathematical software exists.
+
+The importance of piecewise quintic interpolation over lower order
+approximations can be simply observed. In general, the order of a
+polynomial determines the number of function (and derivative) values
+it can interpolate, which in turn determines the growth rate of error
+away from interpolated values. $C^2$ quintic (order six) splines
+match the function value and two given derivatives at each breakpoint.
+This work provides a Fortran 2003 subroutine {\tt MQSI} based on the
+necessary and sufficient conditions in Ulrich and Watson \cite{ulrich1994positivity} for the
+construction of monotone quintic spline interpolants of monotone data.
+Precisely, the problem is, given a strictly increasing sequence $X_1<X_2<
+\cdots <X_n$ of breakpoints with corresponding monotone increasing
+function values $Y_1\le Y_2\le \cdots \le Y_n$, find a $C^2$ monotone
+increasing quintic spline $Q(x)$ with the same breakpoints satisfying
+$Q(X_i)=Y_i$ for $1\le i\le n$. ({\tt MQSI} actually does something slightly
+more general, producing $Q(x)$ that is monotone increasing (decreasing)
+wherever the data is monotone increasing (decreasing).)
+
+The remainder of this chapter is structured as follows: Section 2 provides
+the algorithms for constructing a $C^2$ monotone quintic spline interpolant
+to monotone data, Section 3 outlines the method of spline representation
+($B$-spline basis) and evaluation, Section 4 analyzes the complexity and
+sensitivity of the algorithms in {\tt MQSI}, and Section 5 presents an
+empirical performance study and some graphs of constructed interpolants.
+
+\section{Monotone Quintic Interpolation}
 
@@
 
-Fritsch and Carlson [1980] originally proposed the use of central differences
+Fritsch and Carlson \cite{fritsch1980monotone} originally proposed the use of central differences
 to estimate derivatives, however this often leads to extra and unnecessary
@@
 In an attempt to capture the local shape of the data, this package uses a
-facet model from image processing [Haralick and Watson 1981] to estimate
+facet model from image processing \cite{haralick1981facet} to estimate
 first and second derivatives at breakpoints. Rather than picking a local
@@
 
+\begin{plain}
+\parskip=0pt
 %% ===================================================================
@@
 
-where $X_j$, $Y_j$ $\in \bbb{R}$ for $j = 1$, $\ldots$, $n$, $1 \le i
+where $X_j$, $Y_j$ $\in \mathbb{R}$ for $j = 1$, $\ldots$, $n$, $1 \le i
 \le n$, and $n \ge 3$. Returns the slope and curvature at $X_i$ of the
@@
 %% ----------------------------------------------------------------------
+\end{plain}
 
@@
 guaranteed to produce monotone quintic polynomial segments. Ulrich and
-Watson [1994] established tight constraints on the monotonicity of a
+Watson \cite{ulrich1994positivity} established tight constraints on the monotonicity of a
 quintic polynomial piece, while deferring to He{\ss} and Schmidt
-[1994] for a relevant simplified case. The following algorithm
+\cite{hess1994positive} for a relevant simplified case. The following algorithm
 implements a sharp check for monotonicity by considering the
@@
 
-where $x_0$, $x_1 \in {\bbb R}$, $x_0 < x_1$, and $f$ is an order six
+where $x_0$, $x_1 \in {\mathbb R}$, $x_0 < x_1$, and $f$ is an order six
 polynomial defined by $f(x_0)$, $Df(x_0)$, $D^2f(x_0)$, $f(x_1)$,
@@
 \item{} {The following code considers the full quintic monotonicity
-case studied by Ulrich and Watson [1994]. Given $\tau_1$, $\alpha$,
+case studied by Ulrich and Watson \cite{ulrich1994positivity}. Given $\tau_1$, $\alpha$,
 $\beta$, and $\gamma$ as defined by Ulrich and Watson, a quintic
@@
 %% ----------------------------------------------------------------------
+\end{plain}
 
-It is shown by Ulrich and Watson [1994] that when $0 = DQ(X_i) =
+It is shown by Ulrich and Watson \cite{ulrich1994positivity} that when $0 = DQ(X_i) =
 DQ(X_{i+1}) = D^2Q(X_i) = D^2Q(X_{i+1})$, the quintic polynomial
@@
 
-where $(X_i,Y_i) \in \bbb{R}\times\bbb{R}$, $i = 1$, $\ldots$, $n$ are data
+where $(X_i,Y_i) \in \mathbb{R}\times\mathbb{R}$, $i = 1$, $\ldots$, $n$ are data
 points. Returns monotone quintic spline interpolant $Q(x)$ such that
@@
 the $(\tau_1$, $\alpha$, $\beta$, $\gamma)$ space (described in Ulrich and
-Watson [1994]) for all intervals. This multiple-interval binary search
+Watson \cite{ulrich1994positivity}) for all intervals. This multiple-interval binary search
 allows the value zero to be obtained for all (first and second) derivative
@@
 
-where $(X_i,Y_i) \in \bbb{R}\times\bbb{R}$, $i = 1$, $\ldots$, $n$ are
+where $(X_i,Y_i) \in \mathbb{R}\times\mathbb{R}$, $i = 1$, $\ldots$, $n$ are
 data points, and $Q(x)$ is a quintic spline interpolant such that
@@
 {\tt IS\_MONOTONE} is true for all intervals defined by adjacent data
-points, given a desired precision $\mu \in {\bbb R}$. proj$\bigl(w$,
+points, given a desired precision $\mu \in {\mathbb R}$. proj$\bigl(w$,
 int$(a,b)\bigr)$ denotes the projection of $w$ onto the closed
@@
 total {\tt BINARY\_SEARCH} search could require 69 steps.
+
+\section{Spline Representation}
+
+The monotone quintic spline interpolant $Q(x)$ is represented in terms of
+a B-spline basis. The routine {\tt FIT\_SPLINE} in this package computes
+the B-spline coefficients $\alpha_i$ of $Q(x)=\sum_{i=1}^{3n} \alpha_i
+B_{i,6,t}(x)$ to match the piecewise quintic polynomial values and (first
+two) derivatives at the breakpoints $X_i$, where the spline order is six
+and the knot sequence $t$ has the breakpoint multiplicities (6, 3, $\ldots$,
+3, 6).  The routine {\tt EVAL\_SPLINE} evaluates a spline represented in
+terms of a B-spline basis. A Fortran 2003 implementation {\tt EVAL\_BSPLINE}
+of the B-spline recurrence relation evaluation code by C. deBoor \cite{de1978practical}
+for the value, derivatives, and integral of a B-spline is also provided.
+
+\section{Complexity and Sensitivity}
+
+Algorithms 1, 2, and 4 have ${\cal O}(1)$ runtime. Given a fixed
+schedule for shrinking derivative values, Algorithm 3 has a ${\cal
+  O}(n)$ runtime for $n$ data points. In execution, the majority of
+the time, still ${\cal O}(n)$, is spent solving the banded linear
+system of equations for the B-spline coefficients. Thus for $n$ data
+points, the overall execution time is ${\cal O}(n)$. The quadratic
+facet model produces a unique sensitivity to input perturbation, as
+small changes in input may cause different quadratic facets to be
+associated with a breakpoint, and thus different initial derivative
+estimates can be produced. This phenomenon is depicted in Figure
+\ref{splines:fig-1}. Despite this sensitivity, the quadratic facet model is still
+preferred because it generally provides aesthetically pleasing (low
+{\it wiggle}) initial estimates of the first and second derivatives at
+the breakpoints while perfectly capturing local low-order phenomena
+(e.g., a run of points on a straight line) in the data. The binary
+search for a point near the monotone boundary in $(\tau_1$, $\alpha$,
+$\beta$, $\gamma)$ space is preferred because it results in monotone
+quintic spline interpolants that are nearer to initial estimates than
+a policy that strictly shrinks derivative values.
+
+\begin{figure}
+  \centering
+  \includegraphics[width=.8\textwidth]{Figures/TOMS/1-sensitivity.pdf}
+  \caption{A demonstration of the quadratic
+  facet model's sensitivity to small data perturbations. This example is
+  composed of two quadratic functions $f_1(x) = x^2$ over points $\{1$,
+  $2$, $5/2\}$, and $f_2(x) = (x-2)^2 + 6$ over points $\{5/2$, $3$,
+  $4\}$. Notably, $f_1(5/2) = f_2(5/2)$ and $f_1$, $f_2$ have the same
+  curvature. Given the exact five data points seen above, the quadratic
+  facet model produces the slope seen in the solid blue line at $x = 5/2$.
+  However, by subtracting the value of $f_3$ $= \epsilon(x-2)^2$ from
+  points at $x$ $= \{3$, $4\}$, where $\epsilon$ is the machine precision
+  ($2^{-52}$ for an IEEE 64-bit real), the quadratic facet model produces
+  the slope seen in the dashed red line at $x = 5/2$. This is the nature
+  of a facet model and a side effect of associating data with local facets.}
+  \label{splines:fig-1}
+\end{figure}
+
+\section{Performance and Applications}
+
+This section contains graphs of sample {\tt MQSI} results given various
+data configurations. Computation times for various problem sizes are also
+provided. The files accompanying the subroutine {\tt MQSI} offer multiple
+usages, namely {\tt sample\_main.f90} that demonstrates Fortran 2003 usage
+and a command line interface {\tt cli.f90} that produces {\tt MQSI} estimates
+for points in batches from data files.  Compilation instructions and the
+full package contents are specified in the {\tt README} file.
+
+\begin{figure}
+  \centering
+  \includegraphics[width=.7\textwidth]{Figures/TOMS/2-piecewise-polynomial.pdf}
+  \includegraphics[width=.7\textwidth]{Figures/TOMS/3-large-tangent.pdf}
+  \includegraphics[width=.7\textwidth]{Figures/TOMS/4-signal-decay.pdf}
+\caption{{\tt MQSI} results for three of the functions in the included test
+suite. The {\it piecewise polynomial} function (top) shows the
+interpolant capturing local linear segments, local flats, and
+alternating extreme points. The {\it large tangent} (middle)
+problem demonstrates outcomes on rapidly changing segments of data.
+The {\it signal decay} (bottom) alternates between extreme values
+of steadily decreasing magnitude.}
+\label{splines:test_funcs}
+\end{figure}
+
+\begin{figure}
+  \centering
+  \includegraphics[width=.7\textwidth]{Figures/TOMS/5-real-data.pdf}
+\caption{{\tt MQSI} results when approximating the cumulative distribution
+function of system throughput (bytes per second) data for a computer
+with a 3.2 GHz CPU performing file read operations from Cameron et al.
+\cite{cameron2019moana}. The empirical distribution of 30 thousand throughput values is
+shown in the red dashed line, while the solid line with stylized
+markers denotes the approximation made with MQSI given equally spaced
+empirical distribution points from a sample of size 100.}
+\label{splines:cdf}
+\end{figure}
+
+\begin{figure}
+  \centering
+  \includegraphics[width=.7\textwidth]{Figures/TOMS/6-random-monotone.pdf}
+\caption{The {\it random monotone} test poses a particularly challenging
+problem with large variations in slope. Notice that despite drastic
+shifts in slope, the resulting monotone quintic spline interpolant
+provides smooth and reasonable estimates to function values between data.}
+\label{splines:random}
+\end{figure}
+
+Throughout, all visuals have points that are stylized by local
+monotonicity conditions. Blue circles denote extreme points, purple
+squares are in {\it flat} regions with no change in function value,
+red down triangles are monotone decreasing, and green up triangles are
+monotone increasing.
+
+Figure \ref{splines:test_funcs} offers examples of the interpolating splines produced by the
+routine {\tt MQSI} on various hand-crafted sets of data. These same
+data sets are used for testing local installations in the provided
+program {\tt test\_all.f90}. Notice that the quadratic facet model
+perfectly captures the local linear segments of data in the piecewise
+polynomial test for Figure \ref{splines:test_funcs}. Figure
+\ref{splines:cdf} depicts an approximation of a
+cumulative distribution function made by {\tt MQSI} on a computer
+systems application by Cameron et al. \cite{cameron2019moana} that studies the
+distribution of throughput (in bytes per second) when reading files
+from a storage medium. Figure \ref{splines:random} provides a particularly difficult
+monotone interpolation challenge using randomly generated monotone
+data.
+
+On a computer running MacOS 10.15.5 with a 2 GHz Intel Core i5 CPU, the
+quadratic facet (Algorithm 1) takes roughly one microsecond ($10^{-6}$
+seconds) per breakpoint, while the binary search (Algorithm 4) takes roughly
+four microseconds per breakpoint; these times were generated from 100
+repeated trials averaged over 14 different testing functions.  The vast
+majority of execution time is spent solving the banded linear system of
+equations in the routine {\tt FIT\_SPLINE} for the $B$-spline coefficients.
+For large problems ($n > 100$) it would be faster to construct splines
+over intervals independently (each interval requiring a $6 \times 6$ linear
+system to be solved), however the single linear system is chosen here for
+the decreased redundancy in the spline description.

### 2020-07-29 3-Disseration/thesis.tex
Update.

@@
 % Title of your thesis
-\title{Interpolants and Error Bounds for Modeling and Predicting Variability in Computer Systems}
+\title{Interpolants, Error Bounds, and Mathematical Software for Modeling and Predicting Variability in Computer Systems}
 
@@
 Specifically, this work studies the prediction of distributions of random variables applied to computer system throughput variability.
-New methods for interpolating data with moderately high dimension are presented and novel theoretical error bounds are constructed for piecewise linear interpolation.
-Finally, a mathematical software for approximating distributions from data samples via monotone quintic splines is proposed.}
+Existing approximation methods including multivariate adaptive regression splines, support vector regressors, multilayer perceptrons, Shepard variants, and the Delaunay mesh are investigated in the context of computer variability modeling.
+New methods of approximation using Box splines, Voronoi cells, and Delaunay for interpolating distributions of data with moderately high dimension are presented and compared with existing approaches.
+Novel theoretical error bounds are constructed for piecewise linear interpolants over functions with a Lipschitz continuous gradient.
+Finally, a mathematical software that constructs monotone quintic spline interpolants for distribution approximation from data samples is proposed.}
 
@@
 \chapter{The Importance and Applications of Variability} \label{ch:apps}
-\input{apps}
+\input{1-apps}
 
@@
 \chapter{Algorithms for Constructing Approximations} \label{ch:algs}
-\input{algs}
+\input{2-algs}
 
@@
 \chapter{Na\"{\i}ve Approximations of Variability} \label{ch:naive}
-\input{naive}
+\input{3-naive}
 
@@
 \chapter{Box Splines: Uses, Constructions, and Applications} \label{ch:boxes}
-\input{boxes}
+\input{4-boxes}
 
@@
 \chapter{Stronger Approximations of Variability} \label{ch:strong}
-\input{strong}
+\input{5-strong}
 
@@
 %% Apply approximation algorithms to a final set of problems. Discuss results.
-\input{error}
+\input{6-error}
 
@@
 \chapter{A Package for Monotone Quintic Spline Interpolation} \label{ch:splines}
-\input{splines}
+\input{7-splines}
 

### 2020-08-17 3-Dissertation/1-apps.tex
Renamed dissertation folder to remove typo, finished final defense.

--- 2020-07-28 3-Disseration/apps.tex
@@
 applied to any performance metric and any system.
+
+\section{Organization of this Document}
+
+This thesis is comprised of work from multiple conference and journal
+papers. Chapters \ref{ch:apps} and \ref{ch:algs} share the broader
+context for this work and describe relevant existing algorithms for
+approximation, these chapters derive from multiple published papers
+\cite{lux2018predictive,lux2020effective,lux2020interpolation}. Chapter
+\ref{ch:naive} presents a first attempt at modeling variability and
+appears in \cite{lux2018predictive}. Chapter \ref{ch:boxes} appears in
+\cite{lux2018novel} and presents multiple novel approximation
+techniques. Chapter \ref{ch:strong} applies the novel approximation
+techniques to distribution prediction and appears in
+\cite{lux2018nonparametric}. Chapter \ref{ch:error} presents a novel
+error bound and is to appear in \cite{lux2020interpolation}. Finally
+Chapter \ref{ch:splines} proposes a mathematical software package for
+monotone quintic spline interpolation, was partially presented in
+\cite{lux2020algorithm}, and is to be submitted in full to ACM
+Transactions on Mathematical Software (TOMS).

### 2020-08-17 3-Dissertation/2-algs.tex
Renamed dissertation folder to remove typo, finished final defense.

--- 2020-07-28 3-Disseration/algs.tex
@@
 product approximation of the data. The overall computational
-complexity of Fast MARS is $\mathcal{O}(n d m^3)$. This paper
+complexity of Fast MARS is $\mathcal{O}(n d m^3)$. This work
 uses an implementation of Fast MARS \cite{rudy2017pyearth} with $m =
@@
 $m$ being determined by the minimization convergence criterion. This
-paper uses the scikit-learn SVR \cite{scikit-learn} with a radial
+work uses the scikit-learn SVR \cite{scikit-learn} with a radial
 basis function kernel.
@@
 and the stopping criterion of the minimization used for finding
-weights. This paper uses an MLP built from Keras and Tensorflow to
+weights. This work uses an MLP built from Keras and Tensorflow to
 perform regression \cite{tensorflow2015-whitepaper,chollet2015keras}
@@
      computational complexity of ShepMod is $\mathcal{O}(n^2d)$. This
-     paper uses a Fortran 95 implementation of ShepMod derived from
+     work uses a Fortran 95 implementation of ShepMod derived from
      SHEPPACK \cite{thacker2010algorithm}.
@@
 $P_k\bigl(x^{(k)}\bigr) = f\bigl(x^{(x)}\bigr)$. The computational
-complexity of LSHEP is $\mathcal{O}(n^2d^3)$. This paper uses the
+complexity of LSHEP is $\mathcal{O}(n^2d^3)$. This work uses the
 Fortran 95 implementation of LSHEP in SHEPPACK

### 2020-08-17 3-Dissertation/3-naive.tex
Renamed dissertation folder to remove typo, finished final defense.

--- 2019-05-03 Disseration/naive.tex
@@
 In order to evaluate the viability of multivariate models for
-predicting system performance, this paper presents a case study of a
+predicting system performance, this work presents a case study of a
 four-dimensional dataset produced by executing the IOzone benchmark

### 2020-08-17 3-Dissertation/7-splines.tex
Renamed dissertation folder to remove typo, finished final defense.

--- 2020-07-29 3-Disseration/7-splines.tex
@@
 It is often the case that the true underlying function or phenomenon
-being modeled has known properties like convexity, positivity,
-various levels of continuity, or monotonicity. Given a reasonable
-amount of data, it quickly becomes difficult to achieve desirable
-properties in a single polynomial function. In general, the
-maintenance of function properties through interpolation/regression is
-referred to as {\it shape preserving} \cite{fritsch1980monotone,gregory1985shape}. The specific properties the present algorithm will
-preserve in approximations are monotonicity and $C^2$ continuity. In
-addition to previously mentioned applications, these properties are
-crucially important in statistics to the approximation of a cumulative
+being modeled has known properties like convexity, positivity, various
+levels of continuity, or monotonicity. Given a reasonable amount of
+data, it quickly becomes difficult to achieve desirable properties in
+a single polynomial function. In general, the maintenance of function
+properties through interpolation/regression is referred to as
+{\it shape preserving} \cite{fritsch1980monotone,gregory1985shape}.
+The specific properties the present algorithm will preserve in
+approximations are monotonicity and $C^2$ continuity. In addition to
+previously mentioned applications, these properties are crucially
+important in statistics to the approximation of a cumulative
 distribution function and subsequently the effective generation of
-random numbers from a specified distribution \cite{ramsay1988monotone}.  A spline
-function with these properties could approximate a cumulative
-distribution function to a high level of accuracy with relatively few
-intervals. A twice continuously differentiable approximation to a
-cumulative distribution function (CDF) would produce a corresponding
-probability density function (PDF) that is continuously
-differentiable, which is desirable.
+random numbers from a specified distribution
+\cite{ramsay1988monotone}.  A spline function with these properties
+could approximate a cumulative distribution function to a high level
+of accuracy with relatively few intervals. A twice continuously
+differentiable approximation to a cumulative distribution function
+(CDF) would produce a corresponding probability density function (PDF)
+that is continuously differentiable, which is desirable.
 
@@
   However, by subtracting the value of $f_3$ $= \epsilon(x-2)^2$ from
-  points at $x$ $= \{3$, $4\}$, where $\epsilon$ is the machine precision
+  points at $x = 3$, $4$, where $\epsilon$ is the machine precision
   ($2^{-52}$ for an IEEE 64-bit real), the quadratic facet model produces

### 2020-08-17 3-Dissertation/Writeup/pqmip.tex
Renamed dissertation folder to remove typo, finished final defense.

--- 2019-09-09 Disseration/Writeup/pqmip.tex
@@
   \begin{itemize}
-    \item[] \textit{This can be seen clearly from the fact that $f$ is analytic; there will exist some nonempty ball about $x_0$ or $x_1$ on which $f$ is decreasing.}
+    \item[] \textit{This can be seen clearly from the fact that $f$ is analytic; there will exist some nonempty interval about $x_0$ or $x_1$ for which $f'$ is negative.}
   \end{itemize}
@@
 Given the same initial conditions there are special circumstances which allow for the usage of simpler monotonicity conditions.
-In this case, consider when the quintic function has either $f'(x_0) = 0$ or $f'(x_1) = 0$, which can be tested for monotonicity via the cubic positivity conditions established by \cite{schmidt1988positivity}.
+In this case, consider when the quintic function has either $f'(x_0) = 0$ or $f'(x_1) = 0.$ This reduces the problem of verifying monotonicity to one of cubics established by \cite{schmidt1988positivity}.
 
@@
   \begin{itemize}
-  \item[] \textit{This selection of value for $f'(x_0)$ and $f'(x_1)$ is suggested by \cite{ulrich1994positivity} (originally from \cite{huynh1993accurate}), and quickly enforces upper and lower bounds on derivative values to allow for quintic monotonicity.}
+  \item[] \textit{This selection of values for $f'(x_0)$ and $f'(x_1)$ is suggested by \cite{ulrich1994positivity} (originally from \cite{huynh1993accurate}), and quickly enforces upper and lower bounds on derivative values to ensure quintic monotonicity is obtainable.}
   \end{itemize}
@@
   \begin{itemize}
-  \item[] \textit{This simple box bound ensures that $(A,B)$ remains within a viable region of monotonicity (satisfying Theorem 4, seen in Fig. 6 of \cite{ulrich1994positivity}).}
+  \item[] \textit{This box bound ensures that $(A,B)$ remains within a viable region of monotonicity (satisfying Theorem 4, seen in Fig. 6 of \cite{ulrich1994positivity}).}
   \end{itemize}
@@
 
-\item[0:] $f''(x_0) = $\texttt{max}$\bigg(f''(x_0), \frac{-6 f'(x_0)}{x_1 - x_0}\bigg)$
+\item[0:] $f''(x_0) = $\texttt{ max}$\bigg(f''(x_0), \frac{-6 f'(x_0)}{x_1 - x_0}\bigg)$
 
@@
 
-\item[1:] $f''(x_1) = $max$\bigg( f''(x_1), f''(x_0) + \frac{14 f'(x_0) + 16 f'(x_1) + 60(f(x_1) - f(x_0)) / (x_1 - x_0)}{x_1 - x_0} \bigg)$
+\item[1:] $f''(x_1) = $\texttt{ max}$\bigg( f''(x_1), f''(x_0) + \frac{14 f'(x_0) + 16 f'(x_1) + 60(f(x_1) - f(x_0)) / (x_1 - x_0)}{x_1 - x_0} \bigg)$
 
@@
 
-\item[2:] $f''(x_1) = $max$\bigg( f''(x_1), 6 f'(x_0) - 4 f'(x_1) - f''(x_0) \bigg)$
+\item[2:] $f''(x_1) = $\texttt{ max}$\bigg( f''(x_1), 6 f'(x_0) - 4 f'(x_1) - f''(x_0) \bigg)$
 
@@
 %% ----------------------------------------------------------------------
+Notice that all above algorithms (assuming a fixed level of precision is desired) have $\mathcal{O}(1)$ runtime.
+Only a finite number of operations are needed for monotonicity verification.
+A line search is performed for enforcement, however that search requires a fixed number of steps to achieve any predetermined relative precision on the line.
 
@@
 Finally, the construction of a full piecewise quintic spline is outlined using the above algorithms.
-
 Let $f: \mathbb{R} \rightarrow \mathbb{R}$ be a function in $\mathcal{C}^2.$ Proceed given evaluation tuples $\big(x_i,$ $f(x_i),$ $f'(x_i),$ $f''(x_i)\big)$ for $i = 0,\ldots,N$ such that $x_i < x_{i+1}$ and (without loss of generality) $f(x_i) \leq f(x_{i+1})$ for $i = 1,$ $\ldots,$ $N-1$.
@@
 
-\item Compute
+\item[0:] \texttt{for }$i=0,\ldots,N-1$
+\item[1:] $\quad$\texttt{if $\big($not is\_monotone$(i,i+1)\big)$  make\_monotone$(i,i+1)$}
+
+  \begin{itemize}
+    \item[] \textit{In the shorthand notation above, $i$ and $i+1$ refer to the associated tuples of the form $(x_i,$ $f(x_i),$ $f'(x_i),$ $f''(x_i)).$ This notation will be used through the remainder of this algorithm.}
   \end{itemize}
+
+\item[2:] $\quad$\texttt{for $j=i-1,\ldots,0$}
+\item[3:] $\quad\quad$\texttt{if $\big($not is\_monotone$(j,j+1)\big)$  make\_monotone$(j,j+1)$}
+\item[4:] $\quad\quad$\texttt{else break}
+
+  \begin{itemize}
+    \item[] \textit{The above `for' loop will be referred to as the cascade fix.
+If the adjustment of second derivative values causes the previous interval to become nonmonotone, then the left-hand second derivative value must be updated.
+This may (abnormally) require adjustments across all previously visited intervals.}
+  \end{itemize}
+
+\item[5:] $\quad$\texttt{while $\big($not is\_monotone$(i,i+1)\big)$}
+\item[6:] $\quad\quad f'(x_i) = f'(x_i) / k$
+\item[7:] $\quad\quad f'(x_{i+1}) = f'(x_{i+1}) / k$
+\item[8:] $\quad\quad$\texttt{make\_monotone$(i,i+1)$}
+
+  \begin{itemize}
+    \item[] \textit{In the case that the two corrections to neighboring intervals contradict, the first derivative values of the active interval are decreased to enlarge the overlap of regions I and II (from \cite{ulrich1994positivity}) of the two intervals.}
+  \end{itemize}
+
+\item[9:] $\quad\quad$\texttt{for $j=i-1,\ldots,0$}
+\item[10:] $\quad\quad\quad$\texttt{if $\big($not is\_monotone$(j,j+1)\big)$  make\_monotone$(j,j+1)$}
+\item[11:] $\quad\quad\quad$\texttt{else break}
+
+  \begin{itemize}
+    \item[] \textit{Finally an additional `cascade fix' is performed to ensure that all previous intervals are still monotone after shrinking the derivative values of the current interval.}
+  \end{itemize}
+
+\end{itemize}
+\hrule
+\vspace{10pt}
 %% ----------------------------------------------------------------------
+
+It is mentioned in \cite{ulrich1994positivity} that for sufficiently small $f'(x_i)$ and $f'(x_{i+1})$ the admissible solution interval of second derivative values becomes arbitrarily large.
+It can also be seen that decreasing the assigned derivative on right-hand side of an interval always allows the achievement of monotonicity because shrinking $f'(x_1)$ results in $\gamma$ growing faster than $\sqrt{\beta}$ for \textit{algorithm 1a}, for \textit{algorithm 1b} $\beta$ and $\gamma$ will grow faster $\alpha$ and $\delta$ respectively.
+In application, one needs only to pick some $k > 1$ to ensure successful termination.
+
+Given the potential for successive cascade fixes, the runtime of \textit{algorithm 3} is $\mathcal{O}(N^2)$.
+In practice this has been observed to be incredibly unlikely and difficult to produce.
+While single cascade fixes can occur from hand crafted examples, the generously rounded values selected to ensure monotonicity frequently prevent successive cascade fixes.
 

### 2020-08-17 Research_[2020-05-18]_Piecewise_Monotone_Quintic_Fortran/paper/4-analysis.tex
Made modifications including Dr. Watsons recent revisions. Working towards a submission version.
Renamed dissertation folder to remove typo, finished final defense.

@@
 \heading{4. COMPLEXITY AND SENSITIVITY}
-
-\topinsert
-\centerline{\epsfxsize=4truein \epsffile{vis/1-sensitivity.eps}}
-{\everymath={\scriptstyle}
-\narrower\noindent\rmVIII Fig.\ 1. A demonstration of the quadratic
-  facet model's sensitivity to small data perturbations. This example is
-  composed of two quadratic functions $f_1(x) = x^2$ over points $\{1$,
-  $2$, $5/2\}$, and $f_2(x) = (x-2)^2 + 6$ over points $\{5/2$, $3$,
-  $4\}$. Notably, $f_1(5/2) = f_2(5/2)$ and $f_1$, $f_2$ have the same
-  curvature. Given the exact five data points seen above, the quadratic
-  facet model produces the slope seen in the solid blue line at $x = 5/2$.
-  However, by subtracting the value of $f_3$ $= \epsilon(x-2)^2$ from
-  points at $x$ $= \{3$, $4\}$, where $\epsilon$ is the machine precision
-  ($2^{-52}$ for an IEEE 64-bit real), the quadratic facet model produces
-  the slope seen in the dashed red line at $x = 5/2$. This is the nature
-  of a facet model and a side effect of associating data with local facets.
-\par} \endinsert
-
 Algorithms 1, 2, and 4 have ${\cal O}(1)$ runtime. Given a fixed
@@
 a policy that strictly shrinks derivative values.
+
+\topinsert
+\centerline{\epsfxsize=4truein \epsffile{vis/1-sensitivity.eps}}
+{\everymath={\scriptstyle}
+\narrower\noindent\rmVIII Fig.\ 1. A demonstration of the quadratic
+  facet model's sensitivity to small data perturbations. This example is
+  composed of two quadratic functions $f_1(x) = x^2$ over points $\{1$,
+  $2$, $5/2\}$, and $f_2(x) = (x-2)^2 + 6$ over points $\{5/2$, $3$,
+  $4\}$. Notably, $f_1(5/2) = f_2(5/2)$ and $f_1$, $f_2$ have the same
+  curvature. Given the exact five data points seen above, the quadratic
+  facet model produces the slope seen in the solid blue line at $x = 5/2$.
+  However, by subtracting the value of $f_3$ $= \epsilon(x-2)^2$ from
+  points at $x = 3$, $4$, where $\epsilon$ is the machine precision
+  ($2^{-52}$ for an IEEE 64-bit real), the quadratic facet model produces
+  the slope seen in the dashed red line at $x = 5/2$. This is the nature
+  of a facet model and a side effect of associating data with local facets.
+\par} \endinsert

### 2020-08-17 3-Disseration/Code/fmodpy_splines/OLD-fmodpy_splines(1)/simplified_fortran.f90

@@ -1,390 +1,0 @@
-
-MODULE SPLINES
-USE ISO_FORTRAN_ENV , ONLY : REAL64
-
-
-FUNCTION L2 ( KNOTS , VALUES1 , VALUES2 , DIVISIONS )
-! Given a set of knots and two sets of coefficients, where each
-! set defines a spline, compute the L2 difference between the two
-! splines using Simpson's rule.
-!
-!   KNOTS(K) -- The non-decreasing sequence of unique real-valued
-!               break points, with at least 2 unique values.
-!   VALUES1(K,C1) -- The function (and derivative) values of a polynomial at knots.
-!   VALUES2(K,C2) -- The function (and derivative) values of a polynomial at knots.
-!   DIVISIONS -- The number of divisions used to estimate the
-!                integral over EACH interval of the splines.
-REAL ( KIND = REAL64 ) , INTENT ( IN ) , DIMENSION ( : ) : : KNOTS
-REAL ( KIND = REAL64 ) , INTENT ( IN ) , DIMENSION ( : , : ) : : VALUES1 , VALUES2
-INTEGER , INTENT ( IN ) , OPTIONAL : : DIVISIONS
-! Local variables.
-! First define variables for the local spline fits of the values.
-REAL ( KIND = REAL64 ) : : KNOTS1 ( SIZE ( VALUES1 ) + 2 * SIZE ( VALUES1 , 2 ) ) , KNOTS2 ( SIZE ( VALUES2 ) + 2 * SIZE ( VALUES2 , 2 ) ) , COEFS1 ( SIZE ( VALUES1 ) ) , COEFS2 ( SIZE ( VALUES2 ) )
-! Now declare extra auxillary variables needed for computing L2.
-INTEGER : : D , STEP
-REAL ( KIND = REAL64 ) : : SQ_SUM , L2
-REAL ( KIND = REAL64 ) , ALLOCATABLE , DIMENSION ( : ) : : REL_X , X , EVALS1 , EVALS2
-! Declare the (even) number of divisions.
-! Allocate space for storing the spline evaluations on each interval.
-! Initialize a relative set of X to be evenly spaced on [0,1].
-! Fit the splines to the two sets of knots and values.
-! Cycle over all the intervals tracking the squared sum.
-! Evaluate the (squared) difference between the splines at points.
[…]

### 2020-08-17 3-Disseration/Code/test.py

@@ -1,486 +1,0 @@
-import numpy as np
-import fmodpy
-from util.plot import Plot
-
-splines = fmodpy.fimport("splines.f08", verbose=True,
-                         autocompile_extra_files=True,
-                         module_link_args=["-lblas", "-llapack"])
-
-
-TEST_BANDED_MATRIX = False
-TEST_B_SPLINE = True
-TEST_SPLINE = False
-TEST_SPLINE_ADD = False
-TEST_SPLINE_L2 = False
-TEST_MONOTONE = False
-
-if TEST_MONOTONE:
-    # Return a function that interpolates the given values at the given knots.
-    def fit_spline(x, y):
-        # Make sure "y" is two dimensional.
-        try: len(y[0])
-        except: y = [[v] for v in y]
-        # Fit the spline and get the knots / coefficients for the
-        # B-spline representation of the data.
-        sk, sc = splines.fit_spline(np.array(x, order='F', dtype=float), 
-                                    np.array(y, order='F', dtype=float))
-        # Define an interpolating function with some stored attributes.
-        def f(x, d=0):
-            try:    len(x)
-            except: x = np.array([x], order="f", dtype=float)
-            return splines.eval_spline(sk, sc, x.copy(), d=d)
-        f.x = x.copy()
-        f.y = y.copy()
-        f.knots = sk
-        f.coefs = sc
-        return f
-
-
-    SEED = 0
-    NODES = 30
-    SUBSET = slice(None)
-
-    # Generate random data to test the monotonic fit function.
-    import numpy as np
-    np.random.seed(SEED)
-    nodes = NODES + 2
-    x = np.linspace(0, 1, nodes)
-    y = sorted(np.random.normal(size=(nodes,)))
-    x -= min(x); x /= max(x)
-    y -= min(y); y /= max(y); y -= max(y)
-    # Convert these arrays to lists.
-    x, y = list(x)[SUBSET], list(y)[SUBSET]
-    # Convert these arrays to exact arithmetic.
-    from util.math import Fraction
-    x = list(map(Fraction, x))
-    y = list(map(Fraction, y))
-    interval = [float(min(x)), float(max(x))]
-    from polynomial import fit
-    f = fit(x, y, continuity=1)
-    x = np.array([float(v) for v in f.knots])
[…]

### 2020-08-17 3-Dissertation/Code/Old_Code/cubic.py

@@ -1,0 +1,503 @@
+
+# Custom exception for when there are data errors.
+class BadUsage(Exception): pass
+class BadData(Exception): pass
+class BadValues(Exception): pass
+class UnexpectedError(Exception): pass
+
+# Given the left and right slope ratios (alpha and beta in paper),
+# compute a new left and right slope ratio that rest in the feasible
+# region for a monotonic cubic interpolating polynomial.
+# 
+#   mode -- 1 (curvy)   Project onto the box inside of "3".
+#           2 (default) Project onto circle of radius "3".
+#           3 ( ~~ )    Project onto line "y = 3 - x".
+#           4 (linear)  Project onto max of "y = 3 - 2x" and "x = 3 - 2y".
+# 
+#   project -- 1 uses the intercept of a line with the region.
+#              2 uses the closest point on the region boundary.
+# 
+def compute_cubic_feasible(left_ratio, right_ratio, mode, project):
+    # First check to make sure we have valid ratios (no division by 0).
+    if (max(left_ratio, right_ratio) == 0):
+        pass
+    # Project the ratios onto the appropriate surface.
+    elif (mode == 1):
+        # Option 1: Project onto the square of side length 3.
+        if project == 1:
+            # Determine which line segment it will project onto.
+            mult = 3 / max(left_ratio, right_ratio)
+            # Perform the projection onto the line segment by
+            # shortening the vector to make the max coordinate 3.
+            left_ratio = min(left_ratio, mult*left_ratio)
+            right_ratio = min(right_ratio, mult*right_ratio)
+        elif (project == 2):
+            # The projection direction is naively determined by
+            # capping the value of the vector coordinates at 3.
+            left_ratio = min(left_ratio, 3)
+            right_ratio = min(right_ratio, 3)
+    elif (mode == 2):
+        # Option 2: Project onto the circle with radius 3.
+        #           Same behavior regardless of projection method.
+        length = (left_ratio**2 + right_ratio**2)**(1/2)
+        if (length > 3):
+            left_ratio /= (length / 3)
+            right_ratio /= (length / 3)
+    elif (mode == 3):
+        # Option 3: Project onto the line "y = 3 - x".
+        if (right_ratio != 0): slope = left_ratio / right_ratio
+        else:                  slope = float('inf')
+        if (project == 1):
+            # Compute the intersection of the line and the ratio vector.
[…]

### 2020-08-17 Research_[2020-05-18]_Piecewise_Monotone_Quintic_Fortran/fortran/SPLINE.f90

@@ -244,18 +244,21 @@ IF (INFO .NE. 0) THEN; INFO = INFO + 20; RETURN; END IF
 ! Check to see if the linear system was correctly solved by looking at
 ! the difference between produced B-spline values and provided values.
 DO DERIV = 0, NCC-1
-   ! Reuse the first row of AB as scratch space (the first column
-   ! might not be large enough, but the first row certainly is).
-   AB(1,1:NB) = XI(:)
-   ! Evaluate this spline at all breakpoints. Correct usage is
-   ! enforced here, so it is expected that INFO=0 always.
-   CALL EVAL_SPLINE(T, BCOEF, AB(1,1:NB), INFO, D=DERIV)
-   ! Check the precision of the reproduced values.
-   ! Return an error if the precision is too low.
-   IF (MAXVAL( ABS((AB(1,1:NB) - FX(:,DERIV+1)) &
-      / (1.0_R8 + ABS(FX(:,DERIV+1)))) ) .GT. MAX_ERROR) THEN
-      INFO = 7; RETURN
-   END IF
+   DO I = 1, NB
+      ! Compute the first and last B-spline that have a knot at this breakpoint.
+      I1 = MAX(1, (I-2)*NCC+1)
+      I2 = MIN(NB*NCC, (I+1)*NCC)
+      ! Evaluate this spline at this breakpoint. Correct usage is
+      ! enforced here, so it is expected that INFO=0 always.
+      AB(1,1) = XI(I)
+      CALL EVAL_SPLINE(T(I1:I2+2*NCC), BCOEF(I1:I2), AB(1,1:1), INFO, D=DERIV)
+      ! Check the precision of the reproduced value.
+      ! Return an error if the precision is too low.
+      IF (ABS( (AB(1,1) - FX(I,DERIV+1)) / &
+           (1.0_R8 + ABS(FX(I,DERIV+1))) ) .GT. MAX_ERROR) THEN
+         INFO = 7; RETURN
+      END IF
+   END DO
 END DO
 
 END SUBROUTINE FIT_SPLINE

### 2020-08-25 3-Dissertation/6-error.tex
Updated dissertation document according to Dr. Watson suggestions, included MQSI package code in an Appendix.

--- 2020-07-29 3-Disseration/6-error.tex
@@
 
-The theoretical results constructed in Section \ref{sec:theory} for
+The theoretical results constructed in Chapter \ref{ch:error} for
 (piecewise) linear interpolation are promising and apply directly to
@@
   timings and quartiles of prediction error are located in the
-Appendix (Section \ref{sec:appendix}). All of the test data sets
+Appendix (Section \ref{app:error}). All of the test data sets
 capture underlying functions that are almost certainly stochastic. As
-described in Section \ref{sec:introduction}, regression techniques
+described in Chapter \ref{ch:algs}, regression techniques
 appear most appropriate for these problems. However, typically data

### 2020-08-25 Research_[2020-03-20]_B-Spline_Code/test.py

@@ -1,5 +1,5 @@
 import numpy as np
-import fmodpy
+import og_fmodpy as fmodpy
 from util.plot import Plot
 
 splines = fmodpy.fimport("splines.f90", verbose=True,
@@ -8,8 +8,8 @@ splines = fmodpy.fimport("splines.f90", verbose=True,
 
 
 TEST_BANDED_MATRIX = False
-TEST_B_SPLINE = False
-TEST_SPLINE = True
+TEST_B_SPLINE = True
+TEST_SPLINE = False
 
 
 # =============================================
@@ -59,6 +59,9 @@ if TEST_B_SPLINE:
     r = 2
     knots = np.array([0.] + [1]*r + r*[2])
     knots *= len(knots)-2
+
+    knots = np.array([0.0000000000000000,        0.0000000000000000,        0.0000000000000000,        0.0000000000000000,        0.0000000000000000,        0.0000000000000000,       0.14285714285714285])
+
 
     print()
     print("len(knots) : ",len(knots))

### 2020-08-25 3-Dissertation/8-conclusions.tex
Added box spline code to the thesis document.

This body of work reveals many interesting findings related to modeling variability in computer systems and approximation in general. While generic ``off the shelf'' techniques from machine learning are immediately capable of being applied to the prediction of variability summary statistics, a little extra effort can lend significant improvements in accuracy and precision. The major takeaways from this work are threefold:

\begin{enumerate}
\item computer performance data is often \textit{not} normally distributed and accurate descriptions of performance should rely on nonparametric methods or mixtures of parametric distributions;
\item interpolants such as Delaunay and Shepard methods methods tend to perform better than regressors when used to predict variability-related performance measures, especially when the amount of available data is limited; and
\item in medium to high dimension regressors and interpolants produce equally accurate approximations on analytic test functions, but interpolants favor applications where data is either limited or rapidly changing (i.e., real time systems), because they do not have the \textit{fit} time required by regressors.
\end{enumerate}

Acknowledging these takeaways, there is some room to speculate on the most promising directions for future research related to variability approximation. Specifically, consider systems that need to operate in real time, making approximations in less than a tenth of a millisecond with dynamically updating data. For problems in less than tens of dimensions and with at most thousands of points, it appears best to use Delaunay to make predictions of either means, variances, or full distributions. When the number of points or dimension goes up to tens / hundreds of dimensions with tens of thousands of points, Shepard methods remain particularly well suited to making variability predictions. Finally if there are hundreds / thousands or more dimensions and tens to hundreds of thousands of data points, then a fast $k$ nearest neighbor lookup mixed with a Shepard method could give good results. Notably, all of these approximation methods are interpolants that (given nondegenerate data) require no concept of a \textit{fit}. In addition, all of these interpolants are readily capable of predicting distribution outcomes.

One of the most popular approximation methods in the research community at present is neural networks. While they show incredible promise in computer vision, reinforcement learning, and stream (text sequence) processing, they have failed to maintain superior performance when applied specifically to computer performance modeling. However, the most interesting line of research that ties together interpolants, real time systems, and neural networks is the application of a neural network as a preconditioner for interpolation methods. For instance, train a neural network on computer system data offline, then remove the final layer and instead use the transformed representation of data in the last internal layer as a new vector representation for standard interpolation. Fundamentally this takes advantage of the best components from each method, the structure processing / compositional nature of neural networks, and the ability to quickly make real time predictions from recently generated (or otherwise rapidly changing) runtime data.

Lastly, the improvements in distribution approximation provided by the \texttt{MQSI} software package are not immediately acquired for any data. There is work to be done investigating the best mixture of sampling and smoothing methods to generate a reasonable set of empirical distribution function (EDF) points that can be interpolated by a monotone quintic spline. While initial experiments offered the selection of uniformly distribution EDF points, this may not be optimal and the question of how many points is best remains open. Traditionally Gaussian smoothing or some other method may be used to translate an EDF into a good distribution approximation, however these methods require the (often arbitrary) selection of multiple parameters. Ideally, one would refine the EDF to a small set of points with high confidence that would afford the best possible estimates of first and second derivatives by the quadratic facet model in \texttt{MQSI}.

This work provides a deep dive into the best methodologies for approximating variability in computer systems. By comparing known approximation techniques, proposing new approximations, extending variability modeling from real-valued (parametric) prediction to distribution-valued (nonparametric) prediction, proving a novel error bound for piecewise linear interpolants, and writing a mathematical software package for constructing monotone quintic spline interpolants, this work has significantly contributed to the study of computer system variability and approximation by interpolants. The body of this work was possible largely due to funding and support from the National Science Foundation, and support and guidance from the faculty and staff of Virginia Polytechnic Institute and State University.

### 2020-08-25 3-Dissertation/9-error-app.tex
Added box spline code to the thesis document.

--- 2020-07-28 3-Disseration/thesis.tex
@@
-\documentclass[singlespace,nopageskip]{VTthesis} % nopageskip = Removes arbitrary blank pages.
-
-%% HPC
-\usepackage{latexsym}
-\usepackage{graphicx}
-\usepackage{mathptmx}
-\usepackage{amsmath}
-\usepackage{amsfonts}
-\usepackage{amssymb}
-\usepackage{amsbsy}
-\usepackage{amsthm}
-\usepackage{float}    % For controlling the placement of tables.
-%% 
-\usepackage{booktabs} % For formal tables
-\usepackage{tikz}     % For drawing example boxes
-\usetikzlibrary{positioning}
-\usepackage{plain} % For plain TeX in the TOMS paper
-
-\usepackage{hyphenat}
-\hyphenation{op-tical net-works semi-conduc-tor}
-
-%% Command for creating multiplie-line cells inside a table environment
-\newcommand{\multilinecell}[2][c]{%
-  \begin{tabular}[#1]{@{}c@{}}#2\end{tabular}}
-
-%% Create a ``Lemma'' environment that does not have italics font.
-\spnewtheorem{plainlemma}{Lemma}{\bf}{}
-%% Create a ``Theorem'' environment that is not numbered.
-\spnewtheorem*{onetheorem}{Theorem}{\bf}{}
-%% Create a ``Proof'' environment that starts with 'Proof.'
-\spnewtheorem*{proofdot}{Proof.}{\it}{}
-
-% Using the following header instead will create a draft copy of your thesis
-%\documentclass[doublespace,draft]{VTthesis}
-
-% Title of your thesis
-\title{Interpolants and Error Bounds for Modeling and Predicting Variability in Computer Systems}
-
-% You should include 3-5 keywords, separated by commas
-\keywords{Approximation Theory, Numerical Analysis, High Performance Computing, Computer Security, Nonparametric Statistics, Mathematical Software}
-
-% Your name, including middle initial(s)
-\author{Thomas C. H. Lux}
-
-% Change this to your program, e.g. Physics, Civil Engineering, etc.
-\program{Computer Science}
-
-% Change this to your degree, e.g. Master of Science, Master of Art, etc.
-\degree{Doctor of Philosophy}
-
-% This should be your defense date:
-\submitdate{August 14, 2020}
-
-% Committee members. Only have five readers and one chair available.
-% Only use the ones you need and don't include the ones you don't need.
-% You can also declare a Co-advisor. If you do, the principal and co-advisors
-% will be listed as co-advisors on the title page.  Per the VT ETD standards, 
-% you should not include titles or educational qualifications such as PhD or Dr.
-% You should, however, include middle initials if possible.
-\principaladvisor{Layne Watson}
-\coadvisor{Yili Hong}
-\firstreader{Kirk Cameron}
-\secondreader{Bert Huang}
-\thirdreader{Gang Wang}
-\fourthreader{Yang Cao}
-
-% The dedication and acknowledgement pages are optional. Comment them out to remove them.
-%% \dedication{...}
-%% \acknowledge{...}
-
-% The abstract is required and should be <=250 words for thesis, <=350 words for dissertation.
-\abstract{Function approximation is an important problem.
-This work presents applications of interpolants to modeling random variables.
-Specifically, this work studies the prediction of distributions of random variables applied to computer system throughput variability.
-New methods for interpolating data with moderately high dimension are presented and novel theoretical error bounds are constructed for piecewise linear interpolation.
-Finally, a mathematical software for approximating distributions from data samples via monotone quintic splines is proposed.}
-
-% The general audience abstract is required. There are currently no word limits.
-\abstractgenaud{It is common for scientists to collect data on something they are studying.
-Often scientists want to create a (predictive) model of that phenomenon based on the data, but the choice of \textit{how} to model the data is a difficult one to answer.
-This work proposes methods for modeling data that operate under very few assumptions that are broadly applicable across science.
-Finally, a software package is proposed that would allow scientists to better understand the \textit{true} distribution of their data given relatively few observations.}
-
-\begin{document}
-% The following lines set up the front matter of your thesis or dissertation and is required to ensure proper formatting per the VT ETD standards. 
-\frontmatter
-\maketitle
-\tableofcontents
-
-% The list of figures and tables are now optional per the official ETD standards.
-Unless you have a very good reason for removing them, you should leave these lists in the document.
-Comment them out to remove them.
-\listoffigures
-\listoftables
-
-%% \printnomenclature %creates a list of abbreviations. Comment out to remove it. 
-%% ABBR is described here.
-%% \nomenclature{ABBR}{abbreviation}
-
-% The following sets up the document for the main part of the thesis or dissertation. Do not comment out or remove this line.
-\mainmatter
-
-%% Introduction of variability modeling, the importance of modeling in science in general, then dive into the specific application of modeling techniques to systems (and security).
-\chapter{The Importance and Applications of Variability} \label{ch:apps}
-\input{apps}
-
-%% Describe all of the interpolants and regressors that will be used throughout this work. (except meshes)
-\chapter{Algorithms for Constructing Approximations} \label{ch:algs}
-\input{algs}
-
-%% Demonstrate the naive application of interpolants and regressors to data. Discuss results.
-\chapter{Na\"{\i}ve Approximations of Variability} \label{ch:naive}
-\input{naive}
-
-%% Introduce the theory on box splines. Introduce the algorithms that rely on box splines. Show the results of the code that produces box splines.
-%% Demonstrate the application of these meshes (and box splines particularly) to approximation problems. Discuss results.
-\chapter{Box Splines: Uses, Constructions, and Applications} \label{ch:boxes}
-\input{boxes}
-
-%% Introduce a measurement for error when predicting distribution outputs, then show what happens when system behavior is predicted as a distribution.
-\chapter{Stronger Approximations of Variability} \label{ch:strong}
-\input{strong}
-
-%% Construct error theory that works for all observed applications to this point.
-\chapter{An Error Bound on Piecewise Linear Interpolation} \label{ch:error}
-%% Apply approximation algorithms to a final set of problems. Discuss results.
-\input{error}
-
-%% Propose new method to improve predictions for distributions.
-%% \chapter{Improving Variability Estimates with Monotone $C^2$ Splines} 
-\chapter{A Package for Monotone Quintic Spline Interpolation} \label{ch:splines}
-\input{splines}
-
-%% Bibliography.
-\bibliography{thesis}
-\bibliographystyle{plainnat}
-
-% This formats the chapter name to appendix to properly define the headers:
-\appendix
-
-% Add your appendices here. You must leave the appendices enclosed in the appendices environment in order for the table of contents to be correct.
-\begin{appendices}
-  \chapter{Error Bound Appendices} \label{app:error}
-
 \begin{normalsize}
@@
 \vspace*{\fill}
-
-\end{appendices}
-
-\end{document}
-
-%****************************************************************************
-% Below are some general suggestions for writing your dissertation:
-%
-% 1. Label everything with a meaningful prefix so that you
-%    can refer back to sections, tables, figures, equations, etc.
-%    Usage \label{<prefix>:<label_name>} where some suggested
-%    prefixes are:
-%			ch: Chapter
-%     		se: Section
-%     		ss: Subsection
-%     		sss: Sub-subsection
-%			app: Appendix
-%     		ase: Appendix section
-%     		tab: Tables
-%     		fig: Figures
-%     		sfig: Sub-figures
-%     		eq: Equations
-%
-% 2. The VTthesis class provides for natbib citations. You should upload
-%	 one or more *.bib bibtex files. Suppose you have two bib files: some_refs.bib and 
-%    other_refs.bib.  Then your bibliography line to include them
-%    will be:
-%      \bibliography{some_refs, other_refs}
-%    where multiple files are separated by commas. In the body of 
-%    your work, you can cite your references using natbib citations.
-%    Examples:
-%      Citation                     Output
-%      -------------------------------------------------------
-%      \cite{doe_title_2016}        [18]
-%      \citet{doe_title_2016}       Doe et al. [18]
-%      \citet*{doe_title_2016}      Doe, Jones, and Smith [18]
-%
-%    For a complete list of options, see
-%      https://www.ctan.org/pkg/natbib?lang=en
-%
-% 3. Here is a sample table. Notice that the caption is centered at the top. Also
-%    notice that we use booktabs formatting. You should not use vertical lines
-%    in your tables.
-% 
-%				\begin{table}[htb]
-%					\centering
-%					\caption{Approximate computation times in hh:mm:ss for full order 						versus reduced order models.}
-%					\begin{tabular}{ccc}
-%						\toprule
-%						& \multicolumn{2}{c}{Computation Time}\\
-%						\cmidrule(r){2-3}
-%						$\overline{U}_{in}$ m/s & Full Model & ROM \\
-%						\midrule
-%						0.90 & 2:00:00 & 2:08:00\\
-%						0.88 & 2:00:00 & 0:00:03\\
-%						0.92 & 2:00:00 & 0:00:03\\
-%						\midrule
-%						Total & 6:00:00 & 2:08:06\\
-%						\bottomrule
-%					\end{tabular}
-%					\label{tab:time_rom}
-%				\end{table}
-% 
-% 4. Below are some sample figures. Notice the caption is centered below the
-%    figure.
-%    a. Single centered figure:
-%					\begin{figure}[htb]
-%						\centering
-%						\includegraphics[scale=0.5]{my_figure.eps}
-%						\caption{Average outlet velocity magnitude given an average  
-%				        input velocity magnitude of 0.88 m/s.} 
-%						\label{fig:output_rom}
-%					\end{figure}
-%    b. Two by two grid of figures with subcaptions
-%					\begin{figure}[htb]
-%						\centering
-%						\begin{subfigure}[h]{0.45\textwidth}
-%							\centering
-%							\includegraphics[scale=0.4]{figure_1_1.eps}
-%							\caption{Subcaption number one}
-%							\label{sfig:first_subfig}
-%						\end{subfigure}
-%						\begin{subfigure}[h]{0.45\textwidth}
-%							\centering
-%							\includegraphics[scale=0.4]{figure_1_2.png}
-%							\caption{Subcaption number two}
-%							\label{sfig:second_subfig}
-%						\end{subfigure}
-%
-%						\begin{subfigure}[h]{0.45\textwidth}
-%							\centering
-%							\includegraphics[scale=0.4]{figure_2_1.pdf}
-%							\caption{Subcaption number three}
-%							\label{sfig:third_subfig}
-%						\end{subfigure}
-%						\begin{subfigure}[h]{0.45\textwidth}
-%							\centering
-%							\includegraphics[scale=0.4]{figure_2_2.eps}
-%							\caption{Subcaption number four}
-%							\label{sfig:fourth_subfig}
-%						\end{subfigure}
-%						\caption{Here is my main caption describing the relationship 							between the 4 subimages}
-%						\label{fig:main_figure}
-%					\end{figure}
-%
-%----------------------------------------------------------------------------
-%
-% The following is a list of definitions and packages provided by VTthesis:
-%
-% A. The following packages are provided by the VTthesis class:
-%      amsmath, amsthm, amssymb, enumerate, natbib, hyperref, graphicx, 
-%      tikz (with shapes and arrows libraries), caption, subcaption,
-%      listings, verbatim
-%
-% B. The following theorem environments are defined by VTthesis:
-%      theorem, proposition, lemma, corollary, conjecture
-% 
-% C. The following definition environments are defined by VTthesis:
-%      definition, example, remark, algorithm
-%
-%----------------------------------------------------------------------------
-%
-%  I hope this template file and the VTthesis class will keep you from having 
-%  to worry about the formatting and allow you to focus on the actual writing.
-%  Good luck, and happy writing.
-%    Alan Lattimer, VT, 2016
-%
-%****************************************************************************

### 2020-08-25 3-Dissertation/MQSI/REAL_PRECISION.f90

@@ -1,0 +1,6 @@
+! ^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^
+!                         REAL_PRECISION.f90
+! 
+MODULE REAL_PRECISION  ! module for 64-bit real arithmetic
+  INTEGER, PARAMETER:: R8=SELECTED_REAL_KIND(13)
+END MODULE REAL_PRECISION

### 2020-08-25 3-Dissertation/MQSI/SPLINE.f90

@@ -1,0 +1,386 @@
+! ^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^
+!                           SPLINE.f90
+! 
+! DESCRIPTION:
+!   This file defines the subroutines FIT_SPLINE for computing the
+!   coefficients of a B-spline basis necessary to reproduce given
+!   function and derivative values, and EVAL_SPLINE for evaluating the
+!   value, integral, and derivatives of a spline defined in terms of
+!   its B-spline basis.
+! 
+! CONTAINS:
+!   SUBROUTINE FIT_SPLINE(XI, FX, T, BCOEF, INFO)
+!     USE REAL_PRECISION, ONLY: R8
+!     REAL(KIND=R8), INTENT(IN),  DIMENSION(:)   :: XI
+!     REAL(KIND=R8), INTENT(IN),  DIMENSION(:,:) :: FX
+!     REAL(KIND=R8), INTENT(OUT), DIMENSION(:)   :: T, BCOEF
+!     INTEGER, INTENT(OUT) :: INFO
+!   END SUBROUTINE FIT_SPLINE
+! 
+!   SUBROUTINE EVAL_SPLINE(T, BCOEF, XY, INFO, D)
+!     USE REAL_PRECISION, ONLY: R8
+!     REAL(KIND=R8), INTENT(IN),    DIMENSION(:) :: T, BCOEF
+!     REAL(KIND=R8), INTENT(INOUT), DIMENSION(:) :: XY
+!     INTEGER, INTENT(OUT) :: INFO
+!     INTEGER, INTENT(IN), OPTIONAL :: D
+!   END SUBROUTINE EVAL_SPLINE
+! 
+! EXTERNAL DEPENDENCIES:
+!   MODULE REAL_PRECISION
+!     INTEGER, PARAMETER :: R8
+!   END MODULE REAL_PRECISION
+! 
+!   SUBROUTINE EVAL_BSPLINE(T, XY, D)
+!     USE REAL_PRECISION, ONLY: R8
+!     REAL(KIND=R8), INTENT(IN),    DIMENSION(:) :: T
+!     REAL(KIND=R8), INTENT(INOUT), DIMENSION(:) :: XY
+!     INTEGER, INTENT(IN), OPTIONAL :: D
+!   END SUBROUTINE EVAL_BSPLINE
+! 
+! CONTRIBUTORS:
[…]

### 2020-08-29 3-Dissertation/Paper/1-apps.tex
Final thesis document completed.

--- 2020-08-17 3-Dissertation/1-apps.tex
@@
 \cite{lux2018nonparametric}. Chapter \ref{ch:error} presents a novel
-error bound and is to appear in \cite{lux2020interpolation}. Finally
-Chapter \ref{ch:splines} proposes a mathematical software package for
-monotone quintic spline interpolation, was partially presented in
+error bound and is to appear in \cite{lux2020interpolation}.  Chapter
+\ref{ch:splines} proposes a mathematical software package for monotone
+quintic spline interpolation, was partially presented in
 \cite{lux2020algorithm}, and is to be submitted in full to ACM
-Transactions on Mathematical Software (TOMS).
+Transactions on Mathematical Software (TOMS). Finally Chapter
+\ref{ch:concs} concludes, presenting the major takeaways of this work
+as well as promising future research directions.

### 2020-08-29 3-Dissertation/Paper/8-conclusions.tex
Final thesis document completed.

--- 2020-08-25 3-Dissertation/8-conclusions.tex
@@
 This body of work reveals many interesting findings related to modeling variability in computer systems and approximation in general.
-While generic ``off the shelf'' techniques from machine learning are immediately capable of being applied to the prediction of variability summary statistics, a little extra effort can lend significant improvements in accuracy and precision.
+While generic ``off the shelf'' techniques from machine learning are immediately capable of being applied to the prediction of variability summary statistics, a little extra effort can lend significant improvements in the accuracy and descriptiveness of the predictions being made.
+For example, interpolants like Delaunay yield lower error predictions with less data (improved accuracy), while also trivially extending to the prediction of full distributions (improved precision).
 The major takeaways from this work are threefold:
@@
 
-Lastly, the improvements in distribution approximation provided by the \texttt{MQSI} software package are not immediately acquired for any data.
-There is work to be done investigating the best mixture of sampling and smoothing methods to generate a reasonable set of empirical distribution function (EDF) points that can be interpolated by a monotone quintic spline.
-While initial experiments offered the selection of uniformly distribution EDF points, this may not be optimal and the question of how many points is best remains open.
-Traditionally Gaussian smoothing or some other method may be used to translate an EDF into a good distribution approximation, however these methods require the (often arbitrary) selection of multiple parameters.
-Ideally, one would refine the EDF to a small set of points with high confidence that would afford the best possible estimates of first and second derivatives by the quadratic facet model in \texttt{MQSI}.
+Lastly, the potential improvements in distribution approximation provided by the \texttt{MQSI} software package are significant.
+The ability to capture second derivative information in a piecewise monotone spline interpolant can greatly reduce the amount of data required to estimate a distribution from a sample.
+Extensions of this work should consider statistical importance sampling, smoothing, and other methods to provide robust estimates of points on the distribution functions that can be used in conjunction with \texttt{MQSI} to provide the best possible distribution estimates.
 

### 2020-08-29 3-Dissertation/Paper/thesis.tex
Final thesis document completed.

--- 2019-05-03 Disseration/prelim.tex
@@
 % Title of your thesis
-\title{Interpolation for Modeling and Predicting Variability}
+\title{Interpolants, Error Bounds, and Mathematical Software for Modeling and Predicting Variability in Computer Systems}
 
@@
 % The dedication and acknowledgement pages are optional. Comment them out to remove them.
-%% \dedication{...}
-%% \acknowledge{...}
+\dedication{
+For Alex Faustie Grant, my dream teammate.
+}
+\acknowledge{
+Finishing a Ph.D.
+is hard work.
+It's not always fun.
+And sometimes it's stressful.
+But the people I've met along the way, the things I've learned, and the people from all paths of life that have supported me throughout this journey made all the difference.
+My gratitudes are endless, but I'll do my best to enumerate some here.
+{\vspace{2mm} \\}
+First and foremost I want to thank my advisor Dr.
+Watson for his enduring push for deep scientific rigor and always reminding me of my most important job, to ``think!'' While reading that without context might fall short, Dr.
+Watson's incredible effort, earnest care, and impressive repertoire of knowledge afforded me many great lessons throughout my graduate tenure that I value dearly.
+His genuine feedback has helped me become a better researcher and he's someone I won't hesitate to consult in the future.
+{\vspace{2mm} \\}
+I want to thank my coadvisor Dr.
+Hong for deepening my knowledge of statistics and introducing me to countless relevant and interesting lines of research.
+His support and direction drove the statistical significance of my work.
+In addition I want to thank all of my committee members for their care and contributions, Dr.
+Cameron for asking important practical questions and encouraging a bigger picture cohesion throughout my work, Dr.
+Huang and Dr.
+Wang for ensuring theoretical groundedness and relevance to modern research as well as being great teachers, and Dr.
+Cao for motivating many connections between my work and other areas of applied computational science.
+{\vspace{-1mm} \\}
+I was uniquely lucky in that the entirety of my work at Virginia Tech was supported by the VarSys project.
+Those team members, both students and faculty, guided my growth as a researcher and my research direction as a whole.
+I thank them all for the many conversations about fascinating and diverse topics in computer science, as well as constant feedback and collaboration that broadened my interests and drove me to produce higher quality results.
+{\vspace{2mm} \\}
+Where would I be without my closest friends through graduate school.
+I sincerely thank all of Grad Council, those who helped found the group and see it to its great success as a leading computer science graduate council in the nation.
+I am especially grateful for my lab mates Tyler Chang and Ayaan Kazerouni for the lunches around Blacksburg, encouraging conversations, and general support that made graduate life fun and memorable in ways that it would never have been otherwise.
+{\vspace{2mm} \\}
+Another thanks goes to my family for their continued love and support, particularly my parents Karen and John Lux, brothers Dylan and Austen, sisters Amy, and Arial, and grandmothers Gooma and Mor Mor.
+They have made all the difference and I am unbelievably grateful for them all.
+{\vspace{2mm} \\}
+Lastly I would like to thank my fianc\'e Alex Grant for her enduring love and support over the last four years.
+I couldn't imagine getting this far without our fun trips to Kentucky, summers together, and Skype dinner and movie dates that made the distance between us from Blacksburg to St.
+Louis bearable.
+{\vspace{10mm} \\}
+This work was supported in part by the National Science Foundation under Grant Numbers CNS-1565314 and CNS-1838271.
+Any opinions, findings, and conclusions or recommendations expressed in this material are my own and do not necessarily reflect the views of the National Science Foundation.
+}
 
@@
 %% Apply approximation algorithms to a final set of problems. Discuss results.
-\input{error}
+\input{6-error}
 
 %% Propose new method to improve predictions for distributions.
-\chapter{Improving Variability Estimates with Monotone $C^2$ Splines} \label{ch:splines}
-\input{splines}
+%% \chapter{Improving Variability Estimates with Monotone $C^2$ Splines} 
+\chapter{A Package for Monotone Quintic Spline Interpolation} \label{ch:splines}
+\input{7-splines}
+
+%% Lessons learned, and future directions for research
+\chapter{Conclusions on Modeling and Approximating Variability in Computer Systems} \label{ch:concs}
+\input{8-conclusions}
 
 %% Bibliography.
-\bibliography{prelim}
+\bibliography{thesis}
 \bibliographystyle{plainnat}
@@
 \begin{appendices}
-  \chapter{Error Bound Appendices} \label{app:error}
+\chapter{Error Bound Empirical Test Results} \label{app:error}
 
-  The tables that follow show the precise experimental results for all
-  data sets presented in Section \ref{sec:data}. The tests were all run
-  serially on an otherwise idle machine with a CentOS 6.10 operating
-  system and an Intel i7-3770 CPU operating at 3.4 GHz. The detailed
-  performance results in the tables that follow are very much dependent
-  on the problem and the algorithm implementation (e.g., some codes are
-  TOMS software, some industry distributions, and others are from
-  conference paper venues). Different typeface is used to show best
-  performers, however not much significance should be attached to
-  ranking algorithms based on small time (millisecond) differences. The
-  results serve as a demonstration of conceptual validity.
+\input{9-error-app}
 
-  \begin{table}[H]
-    \centering
-    \begin{tabular}{c|r@{.}l|r@{.}l|r@{.}l|r@{.}l|r@{.}l}
-      \hline
-      Algorithm & \multicolumn{2}{c|}{Min} & \multicolumn{2}{c|}{$25^{th}$} & \multicolumn{2}{c|}{$50^{th}$} & \multicolumn{2}{c|}{$75^{th}$} & \multicolumn{2}{c}{Max}\\
-      \hline
-      MARS & $0$&$00984$ & $3$&$11$ & $7$&$01$ & $\mathit{11}$&$\mathit{7}$ & $1090$&$0$\\
-      SVR & $\mathit{0}$&$\mathit{00402}$ & $\mathbf{0}$&$\mathbf{243}$ & $\mathbf{0}$&$\mathbf{416}$ & $\mathbf{6}$&$\mathbf{19}$ & $1090$&$0$\\
-      MLP & $0$&$0426$ & $2$&$63$ & $\mathit{5}$&$\mathit{27}$ & $14$&$0$ & $1090$&$0$\\
-      Delaunay & $\mathbf{0}$&$\mathbf{00}$ & $1$&$98$ & $5$&$37$ & $13$&$1$ & $\mathit{1080}$&$\mathit{0}$\\
-      ShepMod & $\mathbf{0}$&$\mathbf{00}$ & $\mathit{1}$&$\mathit{93}$ & $6$&$27$ & $16$&$0$ & $1090$&$0$\\
-      LSHEP & $0$&$0400$ & $2$&$17$ & $8$&$87$ & $19$&$1$ & $\mathbf{1070}$&$\mathbf{0}$\\
-      Voronoi & $0$&$00982$ & $3$&$65$ & $7$&$56$ & $15$&$6$ & $1090$&$0$\\
-      BoxSpline & $\mathbf{0}$&$\mathbf{00}$ & $2$&$27$ & $5$&$91$ & $16$&$4$ & $1090$&$0$\\
-      \hline
-    \end{tabular}
-    \caption{This numerical data accompanies the visual provided in
-      Figure \ref{fig:error-forest-fire}. The columns of absolute error
-      percentiles correspond to the minimum, first quartile, median,
-      third quartile, and maximum absolute errors respectively. The
-      minimum of each column is boldface, while the second lowest value
-      is italicized. All values are rounded to three significant
-      digits.}
-    \label{table:error-forest-fire}
-  \end{table}
+\chapter{\texttt{MQSI} Package} \label{app:mqsi}
 
-  \begin{table}[H]
-    \centering
-    \begin{tabular}{|c|r@{.}l| c |r@{.}l|r@{.}l|r@{.}l|}
-      \cline{1-3}\cline{5-10}
-      Algorithm & \multicolumn{2}{c|}{\% Best} &  & \multicolumn{2}{c|}{Fit/Prep. Time (s)} & \multicolumn{2}{c|}{App. Time (s)} & \multicolumn{2}{c|}{Total App. Time (s)}\\
-      \cline{1-3}\cline{5-10}
-      MARS & \quad$\mathit{7}$&$\mathit{7}$ &  & \quad\quad$29$&$1$ & \quad$0$&$00137$ & \quad\quad\quad$0$&$0686$\\
-      SVR & $\mathbf{80}$&$\mathbf{2}$ &  & $\mathbf{0}$&$\mathbf{00584}$ & $\mathit{0}$&$\mathit{0000620}$ & $\mathit{0}$&$\mathit{00310}$\\
-      MLP & $0$&$0$ &  & $32$&$8$ & $0$&$000871$ & $0$&$0436$\\
-      Delaunay & $3$&$5$ &  & $0$&$0151$ & $0$&$0234$ & $1$&$18$\\
-      ShepMod & $3$&$7$ &  & $\mathit{0}$&$\mathit{00634}$ & $0$&$0000644$ & $0$&$00322$\\
-      LSHEP & $5$&$1$ &  & $0$&$0275$ & $\mathbf{0}$&$\mathbf{0000463}$ & $\mathbf{0}$&$\mathbf{00231}$\\
-      Voronoi & $0$&$0$ &  & $0$&$0152$ & $0$&$000396$ & $0$&$0198$\\
-      BoxSpline & $0$&$4$ &  & $0$&$00724$ & $0$&$0000978$ & $0$&$00489$\\
-      \cline{1-3}\cline{5-10}
-    \end{tabular}
-    \caption{The left above shows how often each algorithm had the
-      lowest absolute error approximating forest fire data in Table
-      \ref{table:error-forest-fire}. On the right columns are median fit
-      time of 454 points, median time for one approximation, and median
-      time approximating 50 points.}
-    \label{table:best-forest-fire}
-  \end{table}
+%% -------------------------------------------------------------------
+%%     Setup for including a code listing in a paper.
 
-  \begin{table}
-    \centering
-    \begin{tabular}{c|r@{.}l|r@{.}l|r@{.}l|r@{.}l|r@{.}l}
-      \hline
-      Algorithm & \multicolumn{2}{c|}{Min} & \multicolumn{2}{c|}{$25^{th}$} & \multicolumn{2}{c|}{$50^{th}$} & \multicolumn{2}{c|}{$75^{th}$} & \multicolumn{2}{c}{Max}\\
-      \hline
-      MARS & $0$&$00948$ & $9$&$98$ & $20$&$4$ & $32$&$9$ & $863$&$0$\\
-      SVR & $0$&$00138$ & $3$&$17$ & $7$&$31$ & $12$&$6$ & $\mathbf{27}$&$\mathbf{5}$\\
-      MLP & $0$&$0000239$ & $\mathit{0}$&$\mathit{533}$ & $\mathit{1}$&$\mathit{25}$ & $\mathbf{2}$&$\mathbf{84}$ & $39$&$3$\\
-      Delaunay & $\mathit{3}$&$\mathit{72 \times 10^{-12}}$ & $1$&$20$ & $3$&$50$ & $7$&$67$ & $30$&$7$\\
-      ShepMod & $\mathbf{0}$&$\mathbf{0}$ & $\mathbf{0}$&$\mathbf{255}$ & $\mathbf{0}$&$\mathbf{908}$ & $\mathit{3}$&$\mathit{43}$ & $34$&$5$\\
-      LSHEP & $0$&$00254$ & $2$&$93$ & $7$&$16$ & $13$&$1$ & $\mathit{29}$&$\mathit{0}$\\
-      Voronoi & $\mathbf{0}$&$\mathbf{0}$ & $1$&$29$ & $3$&$52$ & $6$&$87$ & $30$&$1$\\
-      BoxSpline & $0$&$00287$ & $4$&$39$ & $9$&$10$ & $14$&$8$ & $41$&$3$\\
-      \hline
-    \end{tabular}
-    \caption{This numerical data accompanies the visual provided in
-      Figure \ref{fig:error-parkinsons}. The columns of absolute error
-      percentiles correspond to the minimum, first quartile, median,
-      third quartile, and maximum absolute errors respectively. The
-      minimum of each column is boldface, while the second lowest value
-      is italicized. All values are rounded to three significant
-      digits.}
-    \label{table:error-parkinsons}
-  \end{table}
+%% Allow TT fonts to be scaled with LaTeX macros like '\tiny'.
+\newcommand{\ttscale}{s*[0.8]}
+\DeclareFontFamily{OT1}{cmtt}{\hyphenchar\font -1 }
+\DeclareFontShape{OT1}{cmtt}{m}{n}{
+        <-9>   \ttscale cmtt8
+        <9-10> \ttscale cmtt9
+        <10-12>\ttscale cmtt10
+        <12->  \ttscale cmtt12
+      }{}
+\DeclareFontShape{OT1}{cmtt}{m}{bf}{
+        <->    \ttscale cmbftt10
+      }{}
+\DeclareFontShape{OT1}{cmtt}{m}{it}{
+        <->    \ttscale cmitt10
+      }{}
+\DeclareFontShape{OT1}{cmtt}{m}{sl}{
+        <->    \ttscale cmsltt10
+      }{}
+\DeclareFontShape{OT1}{cmtt}{m}{sc}{
+        <->    \ttscale cmtcsc10
+      }{}
+\DeclareFontShape{OT1}{cmtt}{m}{ui}
+       {<->ssub*cmtt/m/it}{}
+\DeclareFontShape{OT1}{cmtt}{bx}{n}
+       {<->ssub*cmtt/m/n}{}
+\DeclareFontShape{OT1}{cmtt}{bx}{it}
+       {<->ssub*cmtt/m/it}{}
+\DeclareFontShape{OT1}{cmtt}{bx}{ui}
+       {<->ssub*cmtt/m/it}{}
 
-  \begin{table}
-    \centering
-    \begin{tabular}{|c|r@{.}l| c |r@{.}l|r@{.}l|r@{.}l|}
-      \cline{1-3}\cline{5-10}
-      Algorithm & \multicolumn{2}{c|}{\% Best} &  & \multicolumn{2}{c|}{Fit/Prep. Time (s)} & \multicolumn{2}{c|}{App. Time (s)} & \multicolumn{2}{c|}{Total App. Time (s)}\\
-      \cline{1-3}\cline{5-10}
-      MARS & \quad\,$0$&$0$ &  & \quad\quad\quad$37$&$9$ & \quad\,\,$0$&$00253$ & \quad\quad\quad\,\,$1$&$48$\\
-      SVR & $0$&$1$ &  & $\mathbf{0}$&$\mathbf{859}$ & $\mathit{0}$&$\mathit{000181}$ & $\mathit{0}$&$\mathit{106}$\\
-      MLP & $\mathit{32}$&$\mathit{0}$ &  & $348$&$0$ & $0$&$00111$ & $0$&$653$\\
-      Delaunay & $0$&$0$ &  & $2$&$47$ & $1$&$22$ & $720$&$0$\\
-      ShepMod & $\mathbf{66}$&$\mathbf{4}$ &  & $\mathit{1}$&$\mathit{13}$ & $0$&$000182$ & $0$&$107$\\
-      LSHEP & $0$&$0$ &  & $2$&$39$ & $\mathbf{0}$&$\mathbf{000144}$ & $\mathbf{0}$&$\mathbf{0845}$\\
-      Voronoi & $1$&$6$ &  & $2$&$77$ & $0$&$0274$ & $16$&$1$\\
-      BoxSpline & $0$&$0$ &  & $1$&$26$ & $0$&$000643$ & $0$&$377$\\
-      \cline{1-3}\cline{5-10}
-    \end{tabular}
-    \caption{The left above shows how often each algorithm had the
-      lowest absolute error approximating Parkinson's data in Table
-      \ref{table:error-parkinsons}. On the right columns are median fit
-      time of 5288 points, median time for one approximation, and median
-      time approximating 587 points.}
-    \label{table:best-parkinsons}
-  \end{table}
+%% Define a new color that is subtle but can be used for keywords.
+\definecolor{bluegray}{rgb}{0.2, 0.3, 0.5}
 
-  \begin{table}
-    \centering
-    \begin{tabular}{c|r@{.}l|r@{.}l|r@{.}l|r@{.}l|r@{.}l}
-      \hline
-      Algorithm & \multicolumn{2}{c|}{Min} & \multicolumn{2}{c|}{$25^{th}$} & \multicolumn{2}{c|}{$50^{th}$} & \multicolumn{2}{c|}{$75^{th}$} & \multicolumn{2}{c}{Max}\\
-      \hline
-      MARS & $\mathit{6}$&$\mathit{45 \times 10^{-15}}$ & $\mathbf{2}$&$\mathbf{70 \times 10^{-14}}$ & $1$&$66$ & $1$&$96$ & $\mathit{53}$&$\mathit{3}$\\
-      SVR & $0$&$0420$ & $0$&$143$ & $0$&$148$ & $\mathit{0}$&$\mathit{860}$ & $119$&$0$\\
-      MLP & $0$&$0000689$ & $0$&$0337$ & $\mathbf{0}$&$\mathbf{0795}$ & $\mathbf{0}$&$\mathbf{264}$ & $\mathbf{5}$&$\mathbf{31}$\\
-      Delaunay & $\mathbf{0}$&$\mathbf{0}$ & $0$&$187$ & $0$&$919$ & $2$&$72$ & $56$&$3$\\
-      ShepMod & $\mathbf{0}$&$\mathbf{0}$ & $0$&$0957$ & $0$&$685$ & $2$&$90$ & $73$&$2$\\
-      LSHEP & $\mathbf{0}$&$\mathbf{0}$ & $\mathit{0}$&$\mathit{0153}$ & $\mathit{0}$&$\mathit{106}$ & $1$&$17$ & $113$&$0$\\
-      Voronoi & $\mathbf{0}$&$\mathbf{0}$ & $0$&$430$ & $1$&$28$ & $2$&$94$ & $83$&$8$\\
-      BoxSpline & $\mathbf{0}$&$\mathbf{0}$ & $0$&$789$ & $2$&$06$ & $5$&$25$ & $119$&$0$\\
-      \hline
-    \end{tabular}
-    \caption{This numerical data accompanies the visual provided in
-      Figure \ref{fig:error-weather}. The columns of absolute error
-      percentiles correspond to the minimum, first quartile, median,
-      third quartile, and maximum absolute errors respectively. The
-      minimum value of each column is boldface, while the second lowest
-      is italicized. All values are rounded to three significant
-      digits.}
-    \label{table:error-weather}
-  \end{table}
+\lstset{% general command to set parameter(s)
+  frame=,                                  % disable the frame lines
+  basicstyle=\color{black}\small\ttfamily, % whole listing style
+  keywordstyle=\color{bluegray},           % language keyword color
+  identifierstyle=,                        % no identifier
+  commentstyle=\color{gray},               % comment color
+  showstringspaces=false,                  % no special string spaces
+  numbers=none                             % number lines
+}
+  %% numberstyle=\tiny,                       % make style of numbers tiny
+  %% stepnumber=3,                            % show numbers every * lines
+  %% numbersep=5pt%
 
-  \begin{table}
-    \centering
-    \begin{tabular}{|c|r@{.}l| c |r@{.}l|r@{.}l|r@{.}l|}
-      \cline{1-3}\cline{5-10}
-      Algorithm & \multicolumn{2}{c|}{\% Best} &  & \multicolumn{2}{c|}{Fit/Prep. Time (s)} & \multicolumn{2}{c|}{App. Time (s)} & \multicolumn{2}{c|}{Total App. Time (s)}\\
-      \cline{1-3}\cline{5-10}
-      MARS & \,\,\,\,$10$&$7$ &  & \quad\quad\quad$\mathit{0}$&$\mathit{151}$ & \quad$0$&$000117$ & \quad\quad\quad\,\,$0$&$0304$\\
-      SVR & $4$&$1$ &  & $\mathbf{0}$&$\mathbf{133}$ & $\mathbf{0}$&$\mathbf{0000947}$ & $\mathbf{0}$&$\mathbf{0246}$\\
-      MLP & $\mathbf{56}$&$\mathbf{8}$ &  & $169$&$0$ & $0$&$00137$ & $0$&$356$\\
-      Delaunay & $0$&$1$ &  & $0$&$664$ & $0$&$886$ & $230$&$0$\\
-      ShepMod & $3$&$5$ &  & $0$&$265$ & $0$&$000128$ & $0$&$0333$\\
-      LSHEP & $\mathit{28}$&$\mathit{4}$ &  & $0$&$874$ & $\mathit{0}$&$\mathit{0000975}$ & $\mathit{0}$&$\mathit{0254}$\\
-      Voronoi & $0$&$2$ &  & $0$&$675$ & $0$&$0270$ & $7$&$01$\\
-      BoxSpline & $0$&$3$ &  & $0$&$330$ & $0$&$000406$ & $0$&$106$\\
-      \cline{1-3}\cline{5-10}
-    \end{tabular}
-    \caption{Left table shows how often each algorithm had the lowest
-      absolute error approximating Sydney rainfall data in Table
-      \ref{table:error-weather}. On the right columns are median fit
-      time of 2349 points, median time for one approximation, and median
-      time approximating 260 points.}
-    \label{table:best-weather}
-  \end{table}
+%% -------------------------------------------------------------------
 
-  \begin{table}
-    \centering
-    \begin{tabular}{c|r@{.}l|r@{.}l|r@{.}l|r@{.}l|r@{.}l}
-      \hline
-      Algorithm & \multicolumn{2}{c|}{Min} & \multicolumn{2}{c|}{$25^{th}$} & \multicolumn{2}{c|}{$50^{th}$} & \multicolumn{2}{c|}{$75^{th}$} & \multicolumn{2}{c}{Max}\\
-      \hline
-      MARS & $5$&$36$ & $3610$&$0$ & $4580$&$0$ & $5450$&$0$ & $13400$&$0$\\
-      SVR & $0$&$0138$ & $8$&$47$ & $15$&$4$ & $41$&$9$ & $7700$&$0$\\
-      MLP & $0$&$00151$ & $\mathbf{1}$&$\mathbf{60}$ & $\mathbf{3}$&$\mathbf{86}$ & $\mathbf{8}$&$\mathbf{34}$ & $\mathbf{604}$&$\mathbf{0}$\\
-      Delaunay & $\mathbf{0}$&$\mathbf{0}$ & $\mathit{2}$&$\mathit{69}$ & $\mathit{13}$&$\mathit{8}$ & $\mathit{35}$&$\mathit{0}$ & $\mathit{4840}$&$\mathit{0}$\\
-      ShepMod & $\mathbf{0}$&$\mathbf{0}$ & $4$&$21$ & $17$&$3$ & $51$&$6$ & $6510$&$0$\\
-      LSHEP & $1$&$27$ & $199$&$0$ & $260$&$0$ & $343$&$0$ & $6530$&$0$\\
-      Voronoi & $\mathit{2}$&$\mathit{89 \times 10^{-10}}$ & $14$&$7$ & $32$&$8$ & $52$&$9$ & $4860$&$0$\\
-      BoxSpline & $0$&$00740$ & $20$&$8$ & $40$&$9$ & $79$&$6$ & $7640$&$0$\\
-      \hline
-    \end{tabular}
-    \caption{This numerical data accompanies the visual provided in
-      Figure \ref{fig:error-credit-card}. The columns of absolute error
-      percentiles correspond to the minimum, first quartile, median,
-      third quartile, and maximum absolute errors respectively. The
-      minimum value of each column is boldface, while the second lowest
-      is italicized. All values are rounded to three significant
-      digits.}
-    \label{table:error-credit-card}
-  \end{table}
+\lstinputlisting{MQSI/README}
+\lstinputlisting[language=Fortran]{MQSI/REAL_PRECISION.f90}
+\lstinputlisting[language=Fortran]{MQSI/MQSI.f90}
+\lstinputlisting[language=Fortran]{MQSI/SPLINE.f90}
+\lstinputlisting[language=Fortran]{MQSI/EVAL_BSPLINE.f90}
 
-  \begin{table}
-    \centering
-    \begin{tabular}{|c|r@{.}l| c |r@{.}l|r@{.}l|r@{.}l|}
-      \cline{1-3}\cline{5-10}
-      Algorithm & \multicolumn{2}{c|}{\% Best} &  & \multicolumn{2}{c|}{Fit/Prep. Time (s)} & \multicolumn{2}{c|}{App. Time (s)} & \multicolumn{2}{c|}{Total App. Time (s)}\\
-      \cline{1-3}\cline{5-10}
-      MARS & \quad\,$0$&$0$ &  & \quad\quad\quad$22$&$0$ & \quad$0$&$00148$ & \quad\quad\quad\,\,$0$&$820$\\
-      SVR & $0$&$0$ &  & $\mathbf{1}$&$\mathbf{01}$ & $\mathit{0}$&$\mathit{000210}$ & $\mathit{0}$&$\mathit{117}$\\
-      MLP & $\mathbf{79}$&$\mathbf{5}$ &  & $290$&$0$ & $0$&$000714$ & $0$&$397$\\
-      Delaunay & $\mathit{20}$&$\mathit{5}$ &  & $3$&$12$ & $3$&$71$ & $2070$&$0$\\
-      ShepMod & $0$&$1$ &  & $\mathit{1}$&$\mathit{4}5$ & $0$&$000220$ & $0$&$122$\\
-      LSHEP & $0$&$0$ &  & $3$&$47$ & $\mathbf{0}$&$\mathbf{000176}$ & $\mathbf{0}$&$\mathbf{0981}$\\
-      Voronoi & $0$&$0$ &  & $3$&$32$ & $0$&$0950$ & $52$&$8$\\
-      BoxSpline & $0$&$0$ &  & $1$&$66$ & $0$&$000956$ & $0$&$532$\\
-      \cline{1-3}\cline{5-10}
-    \end{tabular}
-    \caption{The left above shows how often each algorithm had the
-      lowest absolute error approximating credit card transaction data
-      in Table \ref{table:error-credit-card}. On the right columns are
-      median fit time of 5006 points, median time for one approximation,
-      and median time approximating 556 points.}
-    \label{table:best-credit-card}
-  \end{table}
+\chapter{Box Spline Subroutine} \label{app:boxspline}
 
-  \begin{table}
-    \centering
-    \begin{tabular}{c|r@{.}l|r@{.}l|r@{.}l|r@{.}l|r@{.}l}
-      \hline
-      Algorithm & \multicolumn{2}{c|}{Min} & \multicolumn{2}{c|}{$25^{th}$} & \multicolumn{2}{c|}{$50^{th}$} & \multicolumn{2}{c|}{$75^{th}$} & \multicolumn{2}{c}{Max}\\
-      \hline
-      Delaunay & $\mathbf{0}$&$\mathbf{0287}$ & $\mathbf{0}$&$\mathbf{0914}$ & $\mathbf{0}$&$\mathbf{158}$ & $\mathit{0}$&$\mathit{308}$ & $\mathit{0}$&$\mathit{897}$\\
-      ShepMod & $0$&$0307$ & $\mathit{0}$&$\mathit{0983}$ & $\mathit{0}$&$\mathit{164}$ & $0$&$335$ & $1$&$00$\\
-      Voronoi & $\mathit{0}$&$\mathit{0303}$ & $0$&$100$ & $0$&$165$ & $\mathbf{0}$&$\mathbf{274}$ & $\mathbf{0}$&$\mathbf{762}$\\
-      BoxSpline & $0$&$0703$ & $0$&$417$ & $0$&$561$ & $0$&$745$ & $1$&$00$\\
-      \hline
-    \end{tabular}
-    \caption{This numerical data accompanies the visual provided in
-      Figure \ref{fig:error-throughput}. The columns of absolute error
-      percentiles correspond to the minimum, first quartile, median,
-      third quartile, and maximum KS statistics respectively between
-      truth and guess for models predicting the distribution of I/O
-      throughput that will be observed at previously unseen system
-      configurations. The minimum value of each column is boldface,
-      while the second lowest is italicized. All values are rounded to
-      three significant digits.}
-    \label{table:error-throughput}
-  \end{table}
-
-  \begin{table}
-    \centering
-    \begin{tabular}{|c|r@{.}l| c |r@{.}l|r@{.}l|r@{.}l|}
-      \cline{1-3}\cline{5-10}
-      Algorithm & \multicolumn{2}{c|}{\% Best} &  & \multicolumn{2}{c|}{Fit/Prep. Time (s)} & \multicolumn{2}{c|}{App. Time (s)} & \multicolumn{2}{c|}{Total App. Time (s)}\\
-      \cline{1-3}\cline{5-10}
-      Delaunay & \,\,$\mathbf{62}$&$\mathbf{0}$ &  & \quad\quad\quad$0$&$344$ & \quad$0$&$00984$ & \quad\quad\,\,$2$&$71$\\
-      ShepMod & $0$&$0$ &  & $\mathbf{0}$&$\mathbf{0884}$ & $\mathbf{0}$&$\mathbf{000145}$ & $\mathbf{0}$&$\mathbf{0436}$\\
-      Voronoi & $\mathit{38}$&$\mathit{0}$ &  & $0$&$360$ & $0$&$00253$ & $0$&$762$\\
-      BoxSpline & $0$&$0$ &  & $\mathit{0}$&$\mathit{0972}$ & $\mathit{0}$&$\mathit{000210}$ & $\mathit{0}$&$\mathit{0633}$\\
-      \cline{1-3}\cline{5-10}
-    \end{tabular}
-    \caption{The left above shows how often each algorithm had the
-      lowest KS statistic on the I/O throughput distribution data in
-      Table \ref{table:error-throughput}. On the right columns are
-      median fit time of 2715 points, median time for one approximation,
-      and median time approximating 301 points.}
-    \label{table:best-throughput}
-  \end{table}
+\lstinputlisting[language=Fortran]{boxspline.f90}
 

### 2020-08-29 3-Dissertation/Code/Old_Code/cubic.py

@@ -1,503 +1,0 @@
-
-# Custom exception for when there are data errors.
-class BadUsage(Exception): pass
-class BadData(Exception): pass
-class BadValues(Exception): pass
-class UnexpectedError(Exception): pass
-
-# Given the left and right slope ratios (alpha and beta in paper),
-# compute a new left and right slope ratio that rest in the feasible
-# region for a monotonic cubic interpolating polynomial.
-# 
-#   mode -- 1 (curvy)   Project onto the box inside of "3".
-#           2 (default) Project onto circle of radius "3".
-#           3 ( ~~ )    Project onto line "y = 3 - x".
-#           4 (linear)  Project onto max of "y = 3 - 2x" and "x = 3 - 2y".
-# 
-#   project -- 1 uses the intercept of a line with the region.
-#              2 uses the closest point on the region boundary.
-# 
-def compute_cubic_feasible(left_ratio, right_ratio, mode, project):
-    # First check to make sure we have valid ratios (no division by 0).
-    if (max(left_ratio, right_ratio) == 0):
-        pass
-    # Project the ratios onto the appropriate surface.
-    elif (mode == 1):
-        # Option 1: Project onto the square of side length 3.
-        if project == 1:
-            # Determine which line segment it will project onto.
-            mult = 3 / max(left_ratio, right_ratio)
-            # Perform the projection onto the line segment by
-            # shortening the vector to make the max coordinate 3.
-            left_ratio = min(left_ratio, mult*left_ratio)
-            right_ratio = min(right_ratio, mult*right_ratio)
-        elif (project == 2):
-            # The projection direction is naively determined by
-            # capping the value of the vector coordinates at 3.
-            left_ratio = min(left_ratio, 3)
-            right_ratio = min(right_ratio, 3)
-    elif (mode == 2):
-        # Option 2: Project onto the circle with radius 3.
-        #           Same behavior regardless of projection method.
-        length = (left_ratio**2 + right_ratio**2)**(1/2)
-        if (length > 3):
-            left_ratio /= (length / 3)
-            right_ratio /= (length / 3)
-    elif (mode == 3):
-        # Option 3: Project onto the line "y = 3 - x".
-        if (right_ratio != 0): slope = left_ratio / right_ratio
-        else:                  slope = float('inf')
-        if (project == 1):
-            # Compute the intersection of the line and the ratio vector.
[…]

### 2020-08-29 3-Dissertation/MQSI/MQSI.f90

@@ -1,553 +1,0 @@
-! ^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^
-!                          MQSI.f90
-! 
-! DESCRIPTION:
-!   This file defines a subroutine MQSI for constructing a monotone
-!   quintic spline interpolant of data in terms of its B-spline basis.
-! 
-! CONTAINS:
-!   SUBROUTINE MQSI(X, Y, T, BCOEF, INFO, UV)
-!     USE REAL_PRECISION, ONLY: R8
-!     REAL(KIND=R8), INTENT(IN),    DIMENSION(:) :: X
-!     REAL(KIND=R8), INTENT(INOUT), DIMENSION(:) :: Y
-!     REAL(KIND=R8), INTENT(OUT),   DIMENSION(:) :: T, BCOEF
-!     INTEGER, INTENT(OUT) :: INFO
-!     REAL(KIND=R8), INTENT(OUT), DIMENSION(:,:), OPTIONAL :: UV
-!   END SUBROUTINE MQSI
-! 
-! DEPENDENCIES:
-!   MODULE REAL_PRECISION
-!     INTEGER, PARAMETER :: R8
-!   END MODULE REAL_PRECISION
-! 
-!   SUBROUTINE FIT_SPLINE(XI, FX, T, BCOEF, INFO)
-!     USE REAL_PRECISION, ONLY: R8
-!     REAL(KIND=R8), INTENT(IN),  DIMENSION(:)   :: XI
-!     REAL(KIND=R8), INTENT(IN),  DIMENSION(:,:) :: FX
-!     REAL(KIND=R8), INTENT(OUT), DIMENSION(:)   :: T, BCOEF
-!     INTEGER, INTENT(OUT) :: INFO
-!   END SUBROUTINE FIT_SPLINE
-! 
-! CONTRIBUTORS:
-!   Thomas C.H. Lux (tchlux@vt.edu)
-!   Layne T. Watson (ltwatson@computer.org)
-!   William I. Thacker (thackerw@winthrop.edu)
-! 
-! VERSION HISTORY:
-!   June 2020 -- (tchl) Created file, (ltw / wit) reviewed and revised.
-! 
-SUBROUTINE MQSI(X, Y, T, BCOEF, INFO, UV)
-! Computes a monotone quintic spline interpolant (MQSI), Q(X), to data
-! in terms of spline coefficients BCOEF for a B-spline basis defined by
[…]

### 2020-08-29 3-Dissertation/MQSI/SPLINE.f90

@@ -1,386 +1,0 @@
-! ^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^
-!                           SPLINE.f90
-! 
-! DESCRIPTION:
-!   This file defines the subroutines FIT_SPLINE for computing the
-!   coefficients of a B-spline basis necessary to reproduce given
-!   function and derivative values, and EVAL_SPLINE for evaluating the
-!   value, integral, and derivatives of a spline defined in terms of
-!   its B-spline basis.
-! 
-! CONTAINS:
-!   SUBROUTINE FIT_SPLINE(XI, FX, T, BCOEF, INFO)
-!     USE REAL_PRECISION, ONLY: R8
-!     REAL(KIND=R8), INTENT(IN),  DIMENSION(:)   :: XI
-!     REAL(KIND=R8), INTENT(IN),  DIMENSION(:,:) :: FX
-!     REAL(KIND=R8), INTENT(OUT), DIMENSION(:)   :: T, BCOEF
-!     INTEGER, INTENT(OUT) :: INFO
-!   END SUBROUTINE FIT_SPLINE
-! 
-!   SUBROUTINE EVAL_SPLINE(T, BCOEF, XY, INFO, D)
-!     USE REAL_PRECISION, ONLY: R8
-!     REAL(KIND=R8), INTENT(IN),    DIMENSION(:) :: T, BCOEF
-!     REAL(KIND=R8), INTENT(INOUT), DIMENSION(:) :: XY
-!     INTEGER, INTENT(OUT) :: INFO
-!     INTEGER, INTENT(IN), OPTIONAL :: D
-!   END SUBROUTINE EVAL_SPLINE
-! 
-! EXTERNAL DEPENDENCIES:
-!   MODULE REAL_PRECISION
-!     INTEGER, PARAMETER :: R8
-!   END MODULE REAL_PRECISION
-! 
-!   SUBROUTINE EVAL_BSPLINE(T, XY, D)
-!     USE REAL_PRECISION, ONLY: R8
-!     REAL(KIND=R8), INTENT(IN),    DIMENSION(:) :: T
-!     REAL(KIND=R8), INTENT(INOUT), DIMENSION(:) :: XY
-!     INTEGER, INTENT(IN), OPTIONAL :: D
-!   END SUBROUTINE EVAL_BSPLINE
-! 
-! CONTRIBUTORS:
[…]

### 2020-08-29 3-Dissertation/MQSI/lapack.f

@@ -1,2189 +1,0 @@
-      SUBROUTINE DGBSV( N, KL, KU, NRHS, AB, LDAB, IPIV, B, LDB, INFO )
-*
-*  -- LAPACK driver routine (version 3.2) --
-*  -- LAPACK is a software package provided by Univ. of Tennessee,    --
-*  -- Univ. of California Berkeley, Univ. of Colorado Denver and NAG Ltd..--
-*     November 2006
-*
-*     .. Scalar Arguments ..
-      INTEGER            INFO, KL, KU, LDAB, LDB, N, NRHS
-*     ..
-*     .. Array Arguments ..
-      INTEGER            IPIV( * )
-      DOUBLE PRECISION   AB( LDAB, * ), B( LDB, * )
-*     ..
-*
-*  Purpose
-*  =======
-*
-*  DGBSV computes the solution to a real system of linear equations
-*  A * X = B, where A is a band matrix of order N with KL subdiagonals
-*  and KU superdiagonals, and X and B are N-by-NRHS matrices.
-*
-*  The LU decomposition with partial pivoting and row interchanges is
-*  used to factor A as A = L * U, where L is a product of permutation
-*  and unit lower triangular matrices with KL subdiagonals, and U is
-*  upper triangular with KL+KU superdiagonals.  The factored form of A
-*  is then used to solve the system of equations A * X = B.
-*
-*  Arguments
-*  =========
-*
-*  N       (input) INTEGER
-*          The number of linear equations, i.e., the order of the
-*          matrix A.  N >= 0.
-*
-*  KL      (input) INTEGER
-*          The number of subdiagonals within the band of A.  KL >= 0.
-*
-*  KU      (input) INTEGER
-*          The number of superdiagonals within the band of A.  KU >= 0.
-*
-*  NRHS    (input) INTEGER
-*          The number of right hand sides, i.e., the number of columns
-*          of the matrix B.  NRHS >= 0.
-*
-*  AB      (input/output) DOUBLE PRECISION array, dimension (LDAB,N)
-*          On entry, the matrix A in band storage, in rows KL+1 to
-*          2*KL+KU+1; rows 1 to KL of the array need not be set.
-*          The j-th column of A is stored in the j-th column of the
-*          array AB as follows:
-*          AB(KL+KU+1+i-j,j) = A(i,j) for max(1,j-KU)<=i<=min(N,j+KL)
[…]

### 2020-08-29 3-Dissertation/Old/Code/Old_Code/cdf.py

@@ -1,0 +1,40 @@
+if __name__ == "__main__":
+    import numpy as np
+    from util.plot import Plot
+    from util.stats import cdf_points
+    from monotone import monotone_quintic_spline
+
+    np.random.seed(0)
+    N = 100
+    x = np.random.normal(size=(N,))
+
+    points = np.array(list(zip(*cdf_points(x))))
+    step = N // 3
+
+    p = Plot()
+    funcs = []
+    p.color_num += 1
+    for i in range(step):
+        indices = np.arange(i,len(points),step)
+        sub_points = points[indices]
+        print()
+        print(f"Constructing function {i+1}..", end=" ", flush=True)
+        funcs.append( monotone_quintic_spline(*(sub_points.T), mids=1, ends=0) )
+        print("adding..", end=" ", flush=True)
+        p.add_func(f"Fit {i+1}", funcs[-1], [min(sub_points[:,0]),
+                                             max(sub_points[:,1])], 
+                   dash='dash', opacity=.3)
+        print("done.", flush=True)
+
+    avg_dist = lambda x: sum(min(1,max(0,f(x))) for f in funcs) / len(funcs)
+    p.add_func(f"Avg", avg_dist, [min(x), max(x)], color=p.color(0))
+
+    f = monotone_quintic_spline(*cdf_points(x), mids=1, ends=0)
+    p.add_func(f"All", f, [min(x), max(x)], color=p.color(1))
+
+    from scipy.stats import norm
+    p.add_func(f"Truth", norm.cdf, [min(x), max(x)], color=p.color(3))
+
+    p.show(y_range=[-1,2])
+    exit()
+

### 2020-08-29 3-Dissertation/Paper/boxspline.f90

@@ -1,0 +1,634 @@
+! ^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^
+!                         boxspline.f90
+! 
+! DESCRIPTION:
+!   This file contains the subroutine BOXSPLEV that can be used
+!   to evaluate a box-spline, defined by its associated direction
+!   vector set, at given evaluation points.
+! 
+! CONTRIBUTORS:
+!   Thomas C.H. Lux (tchlux@vt.edu)
+!   Layne T. Watson (ltwatson@computer.org)
+! 
+! VERSION HISTORY:
+!   July 2018 -- (tchl) Created file, (ltw) reviewed and revised.
+! 
+! ====================================================================
+SUBROUTINE BOXSPLEV(UNIQUE_DVECS, DVEC_MULTS, EVAL_PTS, BOX_EVALS, ERROR)
+  ! BOXSPLEV evaluates a box-spline, defined by a direction vector set
+  ! in dimension S, at given evaluation points.
+  ! 
+  ! This implementation uses the numerically consistent algorithm for
+  ! evaluating box-splines originally presented in [1]. Most notably,
+  ! the evaluation of the box-spline near the boundaries of polynomial
+  ! pieces does not exhibit the random behavior that is seen in the
+  ! naive recursive implementation. As described in [1], the
+  ! computational complexity for direction vector sets with repeated
+  ! direction vectors is reduced from the naive recursive
+  ! implementation complexity
+  !     O( 2^{n - s} {n! / s!} ),
+  ! where "n" is the number of direction vectors and "s" is the
+  ! dimension, to
+  !     O( (k)^{n - s} ),
+  ! where "k" is the number of unique direction vectors, "n" is the
+  ! total number of direction vectors, and "s" is the dimension.
+  ! 
+  ! The memory complexity of the implementation provided here is
[…]
